### 2014-09-23 causality_tutorial.tex
adds causality_tuturial.tex with definition of SEM

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usetikzlibrary{arrows,positioning} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, very thick,
           text width=2.5em,
           minimum height=2em,
           text centered},
    % Define arrow style
    pil/.style={
           ->,
           thick,
           shorten <=2pt,
           shorten >=2pt,}
}

\author{Finnian Lattimore}
\title{A short introduction to causal inference}

\begin{document}

\section{Why is causality important}

\section{Models of causality}
To understand causality we first need a more formal definition of what we are talking about. In this post we will look at three slightly different definitions of causality and use them to describe the following (fictional) example.

We have developed a new drug for some illness and wish to determine how effective it is. We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group. The people in the treatment group get the drug, everyone else gets a placebo pill. The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo. We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome .
\subsection{Structural Equation Models}

Structural equation models (SEMs) describe a deterministic world, where underlying mechanisms determine the output of any process for a given input. The mechanism (but not the output) is assumed to be independent of what is fed into it. Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}. More recently, they have been formalized, generalized to the non-parametric setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.

Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables. The noise terms are required to be mutually independent. If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable. Structural equation models can be represented visually as a network. Each variable is a node and arrows are drawn from causes to their effects. For our example the SEM is:

\[
\begin {aligned}
X = & \epsilon_{x} \\
Y = & f(X,\epsilon_{y})
\end {aligned}
\qquad
\raisebox{-5mm}
{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$X$};
\node[main node, right=of 1](2){$Y$};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2);

\end{tikzpicture}
}
\]

This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$. This is justifiable on the grounds that $X$ is random; the outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).

We want to know the causal effect of giving the active drug versus the placebo. We can phrase this question in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$? We have adopted the do-notation, \cite{Pearl1995}, to distinguish the act of intervening to force X to a given value, from observing it to be that value. In our simple example $P(Y|do(X)) = P(Y|X)$. More generally, for a given SEM and interventional query, there is an algorithm \cite{Shpitser2012} that can:

a) determine if the query can be translated into an expression involving only distributions over observed variables. In other words, determine if the query is identifiable given the assumptions encoded by the SEM

b) if it is identiable, return the required expression

For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation. However a SEM is, by definition, more than a set of simultaneous equations. By declaring it to be structural we are saying that it represents assumptions about the relationships between variables. When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.

\subsection{Counterfactuals and the Neyman-Rubin framework}

The Neyman-Rubin model \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals. Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}. Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else.

In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{1,i}$ versus what would happen if they got the placebo, $y_{0,i}$. The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated. The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment. This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.

E[y1 – y0] = E[y1] – E[y0]

We require the assumption that …

which we can state in words as

Take our patients,

We can break the patients down into four groups. The first group will recover whether or not they receive treatment, the second group will recover if treated but not on the placebo, the third group will recover on the placebo and not if treated, and the last group will not recover on treatment or placebo. Unfortunately, the researches don't know which group each person belongs to.

\subsection{Causal Bayesian Networks}

\subsection{Conclusion}

\section{The Do Calculus}
\subsection{Independence in graphical models: D-separation}
\subsection{The three rules}
\subsection{Easy graphical tests for identifiability}
\subsubsection{The Back Door Criterion}
\subsubsection{The Front Door Criterion}

\section{It's not (non-parametrically) identifiable! What can I do?}
\subsection{Bounding causal effects}
\subsection{Additional assumptions}
\subsection{Instrumental variables}

\section{Structure Learning}

\bibliographystyle{plain}% Select the citation style e.g. ieeetr
\bibliography{library}% write the directory to the .bib file
\end{document}

### 2014-09-24 causality_tutorial.tex
completes definition of counter-factual model and initial draft of causal bayesian networks

@@
 \begin{document}
+\def\ci{\perp\!\!\!\perp} % from Wikipedia
 
@@
 To understand causality we first need a more formal definition of what we are talking about.
-In this post we will look at three slightly different definitions of causality and use them to describe the following (fictional) example.
+In this post we will look at three slightly different definitions of causality and use them to describe the following very simple example.
+This section aims to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
 
@@
 
-We want to know the causal effect of giving the active drug versus the placebo.
-We can phrase this question in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$?
-We have adopted the do-notation, \cite{Pearl1995}, to distinguish the act of intervening to force X to a given value, from observing it to be that value.
-In our simple example $P(Y|do(X)) = P(Y|X)$.
-More generally, for a given SEM and interventional query, there is an algorithm \cite{Shpitser2012} that can:
+We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
 
-a) determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the SEM
-
-b) if it is identiable, return the required expression
+CONSIDER SOLVING THIS JUST FROM THE FUNCTIONAL POINT OF VIEW. MOVE THE DO NOTATION DOWN TO CAUSAL BAYES NETS.
 
@@
 When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
+
+\subsection{Causal Bayesian Networks}
+Causal Bayesian Networks are, unsurprisingly, an extension of Bayesian Networks.
+Any joint probability distribution can be factorized into a product of conditional probabilities.
+There are multiple valid factorizations, corresponding to permutations of variable ordering.
+
+\begin{equation}
+P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
+\end{equation}
+
+We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}).
+If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.
+
+\begin{figure}[h]
+\centering
+\caption{General Bayesian Network over 3 variables}
+\label{fig:bayesnet}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, below left=of 1](2){$X_{2}$};
+\node[main node, below right=of 1](3){$X_{3}$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3);
+
+\end{tikzpicture}
+\end{figure}
+
+The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorized over the nodes and edges in the graph.
+There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges).
+If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
+
+\begin{equation}
+P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
+\end{equation}
+
+A CBN is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
+This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
+More generally, if $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network \cite{Pearl2000}.
+
+\[
+ P_{do(Z=z)}(X_{1}...X_{N}) =
+  \begin{cases}
+  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{$X_{1}...X_{N}$ consistant with $Z=z$}  \\
+   0       & \text{otherwise } 
+  \end{cases}
+\]
+
+Causal Bayesian networks encode information on all interventional distributions over a set of variables.
+A causal network represents (much) more information than a Bayesian network with identical structure due to the assumption that links can be interpreted causally.
+Causal Bayesian networks are Bayesian networks so results that apply to Bayesian networks carry directly across;the local markov property states that variables are independent of their non-effects given their direct causes.
+Similarly the global markov property and d-separation also hold in causal networks.
+There is an algorithm \cite{Shpitser2012} based on these properties that, for a given network and interventional (do-type) query, can:
+
+a) determine if the query can be translated into an expression involving only distributions over observed variables.
+In other words, determine if the query is identifiable given the assumptions encoded by the network
+
+b) if it is identiable, return the required expression
+
+Returning at last to our simple example, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$?
+
+assume the network has the same form as
+
+We have adopted the do-notation, \cite{Pearl1995}, to distinguish the act of intervening to force X to a given value, from observing it to be that value.
+In our simple example $P(Y|do(X)) = P(Y|X)$.
+More generally, for a given SEM and interventional query, there is an algorithm  that can:
 
@@
 
-In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{1,i}$ versus what would happen if they got the placebo, $y_{0,i}$.
+In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{i}^{1}$ versus what would happen if they got the placebo, $y_{i}^{0}$.
 The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
@@
 
-E[y1 – y0] = E[y1] – E[y0]
+Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.
+Let $Y^{1}$ be a random variable representing the potential outcome if treated.
+The distribution of $Y^{1}$ is the distribution we would see of $Y$ if everyone was treated.
+Similarly $Y^{0}$ represents the potential outcome for the placebo.
+We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $P(Y^{1})-P(Y^{0})$.
+We can estimate, from an experimental or observational study, the probability that people recover if treated $P(Y|X=1)$ and the probability that they recover if not treated $P(Y|X=0)$.
+Now if $X=0$ then $Y = Y^{0}$.
+Equivalently stated:
 
-We require the assumption that …
+\begin{equation}
+\begin{aligned}
+P(Y^{0}|X=0)&= P(Y|X=0)\\
+P(Y^{1}|X=1)&=P(Y|X=1)
+\end{aligned}
+\end{equation}
 
-which we can state in words as
+If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
 
-Take our patients,
+\begin{equation}
+\begin{aligned}
+P(Y^{1}) &= P(Y^{1}|X=1) = P(Y|X=1) \\
+P(Y^{0}) &= P(Y^{0}|X=0) = P(Y|X=0) 
+\end{aligned}
+\end{equation}
+
+\begin{equation}
+\implies P(Y^{1})-P(Y^{0}) = P(Y|X=1) - P(Y|X=0)
+\end{equation}
+
+The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{}.
+They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
+Again this is justified in our example due to the randomization of treatment assignment.
+In general these assumption do not hold.
+If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income, that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.
+
+\subsection{Tying it all together. How do these models relate?}
+acyclic SEMs are valid Causal Bayesian Networks so everything that applies to them also applies to SEMs.
+
+There is a direct relationship between these ignoreability assumptions and the functional or graphical assumptions encoded by SEM and Causal Bayesian Networks.
+
+Ignoreability holds when do seperation is valid.
+
+Lets link together the notation for SEMs and Counterfactuals here...
+
+counterfactuals are defined for SEMs and the Rubin-Neyman model but not for causal bayesian networks
 
@@
 Unfortunately, the researches don't know which group each person belongs to.
+Drawing this up as a table:
 
-\subsection{Causal Bayesian Networks}
+\begin{tabular}{c|c|c|c}
+group & treatment & placebo & probability of group\\
+\hline
+1 & recover & recover & $P(Y^{1}=1,Y^{0}=1)$\\
+2 & recover & die & $P(Y^{1}=1,Y^{0}=0)$\\
+3 & die & recover & $P(Y^{1}=0,Y^{0}=1)$\\
+4 & die & die & $P(Y^{1}=0,Y^{0}=0)$\\
+\end{tabular}
 
-\subsection{Conclusion}
+This difference does not result in different conclusions for interventional queries but there are subtle difference with regard to counterfactual queries, for example mediation.
 
@@
 \subsection{Independence in graphical models: D-separation}
+
 \subsection{The three rules}
+The cool thing is they are complete.
+For a given SEM, a do query is non-parametrically identifiable if and only if it can be transformed to do-less statement via repeated application of these three rules.
+\subsubsection{Rule 1}
+D-separation still applies
+
+Let $Z$ be the set of variables on which we are intervening and $G$ our graphical model.
+$G^{\dag}_{Z}$ represents the model where all links going into variables $Z elementof Z$ are deleted and replaced by a new parent node $Z$
+
+if $(W \ci Y | Z,X)$ in $G_{Z}$:
+\begin{equation}
+\label{eq:Do1}
+ P(Y|do(Z=z),X=x,W=w) = P(Y|do(Z=z),X=x) 
+\end{equation}
+\begin{equation}
+\label{eq:Do12}
+ P(Y|do(Z=z),W=w) = P(Y|do(Z=z)) 
+\end{equation}
+
+\subsubsection{Rule 2}
+If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.
+
+if $(W \ci Y | Z,X)$ in $G_{Z}$:
+\begin{equation}
+\label{eq:Do2}
+P(Y|do(Z=z),do(X=x),W=w) = P(Y|do(Z=z),X=x,W=w)
+\end{equation}
+
+This also implies that if $Y d-sep X|X$:
+\begin{equation}
+\label{eq:Do22}
+P(Y|do(X=x)) = P(Y|X=x)
+\end{equation}
+
+\subsubsection{Rule 3}
+if $(W \ci Y | Z,X)$ in $G_{Z}$:
+\begin{equation}
+\label{eq:Do3}
+P(Y|do(Z=z),do(X=x),W=w) = P(Y|do(Z=z),W=w)
+\end{equation}
+\begin{equation}
+\label{eq:Do32}
+P(Y|do(X=x)) = P(Y)
+\end{equation}
+
 \subsection{Easy graphical tests for identifiability}
-\subsubsection{The Back Door Criterion}
+The completeness of these rules does not make them easy to apply. There are some quick graphical tests that are sufficient (but not necessary) for non-parametric identifiability.
+
+\subsubsection{If all variables are observed}
+All interventional queries are identifiable. Just condition on the parents.
+
+\subsubsection{The Back Door criterion}
+The back door criterion nicely aligns to answer the problem, when can you calculate a causal effect by conditioning on or adjusting for confounding variables and answers which variables should be included.
+
 \subsubsection{The Front Door Criterion}
+Allows for identification of causal effects via sequentially identifying other causal effects.
 
@@
 \subsection{Additional assumptions}
+additive noise
+causal anti-causal \cite{Janzing2012}
+
 \subsection{Instrumental variables}
 
-\section{Structure Learning}
+\section{Structure learning}
+peters approach \cite{Peters2014}
+Pearl/Verma/Sprites approach \cite{Pearl2000}
 

### 2014-09-24 causality_review.tex
adds causality_review.tex

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{parskip}
\usepackage{tikz}

\usetikzlibrary{arrows,positioning} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           rectangle,
           rounded corners,
           draw=black, very thick,
           text width=6.5em,
           minimum height=2em,
           text centered},
    % Define arrow style
    pil/.style={
           ->,
           thick,
           shorten <=2pt,
           shorten >=2pt,}
}

\author{Finnian Lattimore}
\title{A short introduction to causal inference}
\begin{document}

\section{Abstract}
wide applicability makes it an inherently multidiciplinary field, economics, social science, philosophy, statistics and more recently, computer science.

cross-world independence cannot be determined experimentally - thus not subject to scientific scrutiny. The problem is if we are doing induction, we have already made assumptions not subject to scientific scrutiny. 
\section*{Introduction}

The notion of causality is deeply tied to questions about actions and interventions. Does high blood pressure cause heart attack – or are the two merely associated through a third causal factor? Would taking action to reduce blood pressure reduce the risk of heart attack? Sometimes these questions can be resolved by direct experiment, however this may be prohibitively expensive, ethically unconscionable or technically impossible. For these cases the question arises, can we infer the consequences of action from purely observational data? Without making any assumptions, this is impossible. The fundamental problem of causal inference can be illustrated by a simple diagram (Figure XXX).

Classical statistics deals with the question; Given samples from an unknown distribution, what can we say about that distribution. Causal inference asks; Given samples from an unknown distribution, what can be inferred about the distribution after I take some action? Without making any assumptions about how the action could effect the distribution, that question is unanswerable no matter how many samples are available.

There are two somewhat distinct problems commonly referred to as causal inference. The first deals with estimating causal effects where the structure of the relationships between variables is considered given by theory or other prior knowledge. This form of causal inference is prevalent, often implicitly, in almost every area of scientific research and techniques to address it have been developed in parallel in economics \cite{Heckman}, physcology \cite{Campbell}, statistics \cite{Rubin} and computer science \cite{Pearl}.

The second and more difficult problem is to infer the structure of the causal relationships between variables. Any causal inference requires either experiment or assumptions, thus determining causal structure from observational data is impossible without some assumptions. However, recent research, XXXX, XXXX,XXXX has demonstrated that some aspects of causal structure can be identified from very general assumptions.

\section*{Defining Causality}

In order to study causality, we need clearer definition of what we are talking about. There are a number of different approaches to defining causality.

\subsection*{The Neyman-Rubin model}

The Neyman-Rubin model defines causality in terms of potential outcomes, or counterfactuals. It was developed from the starting point of randomized experiment and shares much of this terminology. A causal effect is defined as the difference in outcome for the same subject under different possible treatments. Expressed mathematically, assume we wish to understand the causal effect of a variable we will call $T$ (for treatment) on an outcome variable $Y$. For simplicity and consistency with the majority of the literature, I will assume $T$ is a discrete, binary variable. The levels of $T$ are often referred to as active ($T=1$) and control ($T=0$). for an individual unit (person, location, etc) $i$  that could be assigned a value of $T$, the individual causal effect $e_{i}$ of $T=active$ relative to $T=control$ on outcome $Y$ is defined by:
\begin{equation}
\label{eq:ICE}
\tau_{i} = Y_{i}(T=active)-Y_{i}(T=control)
\end{equation}
These can not both be directly measured, since a single unit is can only be exposed to one of the two options. This is what Rubin calls the fundamental problem of causality. The problem can be resolved if you can assume based on prior knowledge that all units are identical or that units will revert exactly to their initial state some time after treatment. This type of assumption often holds to a good approximation in the natural sciences, and explains why researchers in these fields tend not to be concerned with causal theory.

Rather than trying to estimate individual causal effects, one can focus on the average causal effect, defined as:
\begin{equation}
\label{eq:ACE}
\tau = E[Y_{i}|T=active] - E[Y_{i}|T=control]
\end{equation}
It is essential to understand that the averages in equation (\ref{eq:ACE}) are over both potential outcomes for all the units in the population regardless of which treatment they actually received. What can directly be measured, in an experimental or observational study, is the difference in means of the observed outcomes for the treatment and control groups, sometimes referred to as the \textit{prima facie causal effect}. :
\begin{equation}
\label{eq:observed_outcome_mean_difference}
S^{*} = E[Y_{j: j \in treated}|T = active] - E[Y_{k: k \in control}|T = control]
\end{equation}
In general, the average causal effect (\ref{eq:ACE}) and the observed difference in means (\ref{eq:observed_outcome_mean_difference}) will not be equal. However, if the assignment of units to treatment is random, then in the limit of large numbers of units the two will converge (Rubin). This formalizes the case for using randomized experiments to answer causal questions. It is interesting to note that this analysis relies on properties of the mean operator. If one is interested in other measures of the distribution of individual causal effects, or the form of the distribution as a whole, then randomization is less powerful (Heckman).

The Neyman-Rubin model extends to observational studies by careful modelling of the mechanism by which units are assigned to treatments. In particular, propensity scores, the most widely used technique arising from the Neyman-Rubin framework, stem from theAt heart, the Rubin Causal Framework is primarily a definition of causality in terms of counterfactuals, along with the recomendation to conceptualize inference from observational studies in terms of trying to frame it as random experiment. Quote XXXX. idea that if one could stratify on all variables that effected assignment to treatment, then within each strata, the prima facie causal effect would equal the average causal effect. The rest of the technique is just a clever way to approximate this when covariates are continuous or high dimensional such that stratification is not possible.

Assumptions, ignorability, strong ignorability, SUTVA.

At heart, the Rubin Causal Framework is primarily a definition of causality in terms of counterfactuals, along with the recomendation to conceptualize inference from observational studies in terms of trying to frame it as random experiment. Quote XXXX.

\subsection*{Structural Equation Models}

A (causal) structural equation model is a set of $n$ equations of the form:
\begin{equation}
\label{eq:struct_eq_mod_definition}At heart, the Rubin Causal Framework is primarily a definition of causality in terms of counterfactuals, along with the recomendation to conceptualize inference from observational studies in terms of trying to frame it as random experiment. Quote XXXX.
x_{i} = f_{i}(pa_{i},\epsilon_{i})
\end{equation}

where $X = \{x_{1}...x_{n}\}$ are variables, $f_{i} \in \{f_{1}...f_{n}\}$ are deterministic functions, $pa_{i}$ stands for the set of variables that directly effect $x_{i}$ and $\epsilon_{i}$ represents noise introduced by not incorporating all sources of variation as separate variables. The noise terms are, by construction, mutually independent. Any variables that directly effect more than one variable in the model must be included in $X$.

Structural equation models can be represented graphically as a network. Each node represents a variable and its associated deterministic function and a link $x_{i}->x_{j}$ indicates $x_{i}$ causes $x_{j}$.

A causal inference perspective on the problem of granting bail.

Consider the following, greatly simplified, model of the bail process: People arrested on suspicion of committing a crime are brought before a judge, who decides if they should be granted bail (released into the community) or remanded in custody until they face court. The court must balance the rights of a person presumed innocent against the risk they fail to appear at a hearing, interfere with a witness or commit some other offence if released. For simplicity, we group all these issues into a single concept of 'bad behaviour'. The problem is now to decide, based on the available evidence, if a person is a high or low risk of bad behaviour.

Estimating risk is difficult for humans to do accurately, which motivates the application of machine learning and statistical techniques. Additionally, a computer algorithm can be designed to avoid particular forms of bias, for example based on race or physical appearance. The problem appears to be standard binary classification, suggesting that it should be straightforward to apply machine learning techniques and assess if they are better at predicting who will break bail than judges. However this ignores the way in which the data, on which a machine learning algorithm could be trained, was generated.

Image we have a large data set on what happened to people released on bail. For each person we have a set of variables, we will denote as Z, describing their personal attributes and the crime they are accused of. We also have a binary variable Y that specifies if they behaved badly or not. We train our machine learning algorithm on a subset of the data and then test on the remainder to see how it did and get out the following confusion matrix.

they are accused of and a binary variable

 age, sex, prior history, shoe size,

A few different possible causal models to consider

Estimation of the probability of bad behaviour is central to the decision, motivating the use of machine learning provide a predictive model.

he fact that the decition to grant bail is substantially based on the risk, that is probability, that a person will behave in a particular way motivates applying machine learning and statistical technqiues. Such techniques be additionally justified by the possibility that they can be designed to avoid particular bias - for example on race, or physical appearance. But is the problem fundamentally statistical?

What bounds?

Under what circumstances could we empircally show statistical techniques were fairer, reduced risk. etc.

 %nodes

\node[main node](1){$x_{1} = f_{1}(\epsilon_{1})$};
\node[main node, below left=of 1](2){$x_{2} = f_{2}(x_{1},x_{5},\epsilon_{2})$}; 
\node[main node, below right=of 1](3){$x_{3} = f_{3}(x_{1},x_{2},\epsilon_{3})$};
\node[main node, below right= of 2](4){$x_{4}=f_{4}(x_{2},x_{3},\epsilon_{4})$};
\node[main node, above left=of 2](5){$x_{5} = f_{5}(\epsilon_{5}$)};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (3)
        edge node {} (2)        
    (2) edge node {} (4)
    	edge node {} (3)        
    (3) edge node {} (4)
    (5) edge node {} (2);

\end{tikzpicture}
\end{figure}

Structural are ubiquitous across economics and social sciences. The causality comes from what we define the equations or links to mean. A causal structural equation model clearly specifies our assumptions on the causal structure of the problem. Missing links in the network (or equivalently, excluding a variable from the set $pa_{i}$ represents an assumption that x does not directly cause y). Asserting a structural equation model for a given problem cleanly encodes a set of assumptions.

Interventions in structural equation models correspond to external modification of functions and variable distributions. The do opperator.

Extensively applied and causaly interpreted in economics - exemplified by Wright 1921, Haavelmo 1943 and more recently Heckman. Generally with additional assumptions that the functions are linear and the noise additive.

Counterfactuals can be derived from structual models.

\subsection*{Causal Graphical Networks}

Graphical models (Pearl)
Causal Bayes nets are a subset of structural equation models for those who have philosophical objections to counterfactuals. 
Defined in terms of interventional distributions.

Potential outcomes and structural equation models allow statements about counterfactuals. Causal Bayes nets are

\subsection*{Differences, similarities and controversies}

Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed are definitionally (almost) equivalent. However, the practical guidance given by their proponents on applications differs. This has led to confusion and controversy.

The key difference between Rubin and Pearl is that Rubin does not make explicit how you decide which variables one should satisfy/condition on/include in a propensity score. In some cases Rubin has implied one should condition on as many variables as possible "quote". Using a graphical framework it is easy to demonstrate that it is possible to increase the bias in a causal estimate by conditioning on certain variables, for example, a collider.  XXXX gives some plausible examples of real cases where this could occur. The correct answer to which variables should be conditioned on relies on making some assumptions about the plausibility of the structural role played by that variable in the problem. It is possible that certain structures are more prevalent in nature or within particular research fields. Similarly for particular distributions and functional relations between them, conditioning on a variable that should have been excluded could do less harm than not conditioning on one that should have been included. (binary variable thing). However, such justifications for conditioning on a variable where you have 'no idea' about its true causal structure should be explicitly stated. Alternatively one could estimate the causal effect for each structural model possible with in the assumptions you were willing to make incorporate the resulting uncertainty into the reported effect.

Pearls work focuses on the large sample limit and non-parametric inference, and providing very clear theory. Rubin's papers contain a series of more practical examples, and advice and techniques, in particular propensity scores, that are useful for applied work.

Propensity scores are a purely statistical concept. They can be calculated directly from joint distributions over variables and do not contain any extra-statistical or causal information. However, they address a problem that is commonly encountered in practical causal inference with finite data: how do you condition on a continuous or high-dimensional set of variables?

Perhaps as a consequence of this difference in focus, Rubins work has been widely cited in applied research in XXXX, as opposed to Pearl (use some citiation data to check that). One provides clear, simple theoretical method for describing what assumptions one must make, which variables should be conditioned on. The other, a notation that can elegently capture the what if types of queries that humans often phrase as causal queries and propensity scores.

Also the focus on treatment assigment ignores other causal inference techniques such as instrumental variables that can be represented and described by structural equation models.

Rubin's focus on expectation values may contribute to its empirical success/use vs Pearls work is more general but may be more sensitive to sampling issues. More concrete, less general complete theory. Way in which the material is presented may be more accessible to practitioners.

warnings not to adjust for post-exposure variables (mentioned in Greenland  Robins 2009). Reference Cox DR: Planning of Experiments New York City: John Wiley and Sons
Inc; 1958.

Is it important to only consider interventions on variables that could 'potentially be modified'? : Epidemiologic measures and policy formulation:
lessons from potential outcomes,
Invited commentary: hypothetical interventions
to define causal effects--afterthought or prerequisite?

\section*{Estimating Causal Effects}

If you are using the do-calculus to estimate causal effects, the assumptions you are effectively making are the weak ones preferred by Robins, et al. Ie for any problem involving only interventional queries, the answer will be the same if you use weak or strong ignorablility.

Firstly, estimating the magnitude and direction of causal effects where the structure but not functional form of the relationships is known and secondly inferring the structure of cause and effect between variables.

Firstly the problem of estimating causal effects where the structure of the relationships between variables is known.

1) You assume you already know the causal structure of a set of variables and the goal is to infer the value of a given causal effect, or in other words the consequence of an intervention. Assuming the structure means you could draw a network of the variables with arrows from causes to effects. You can allow for variables you cannot measure but you still need to draw them into the network and specify how they are linked to observed variables.

In the large sample limit, this problem is solved. The do calculus (Pearl 2000) tells you when the problem is identifiable and if it is, gives a formula to calculate the causal effect. A causal effect that is not identifiable under the do calculus cannot be estimated purely from observational data, no matter how many samples you have, without making additional assumptions. Some problems that are not identifiable can still be solved if you are additionally willing to make parametric assumptions about the form of the distributions of variables and the functional relationships between them.

Where the causal effect is not non-parametrically identified, you may still be able to bound it.

An extremely short summary of Pearl on this matter. 1) Assume causality defined by a SEM (or a causal Bayes network). 2) This allows us to view interventions as modifications to the graph. 3) The 3 do-calculus rules can be derived from d-seperation on the modified graph. 4) These rules are complete. They can be used to determine if a interventional query is non-parametrically identifiable and if so, what the formula for calculating it is. 5) The back door and front door criterion are sufficient but, not necessary, for causal identifiability. They are handy shortcuts to applying the full do-calculus algorithm as they cover many types of causal structure. Also, back door lines up with when one can calculate causal effect by conditioning. Front door corresponds to identifying by sequentially cacluating other effects. A generlizing sufficient condition for identifying P(y|do(x)) is that there exists no path consisting of only bidirected arcs between X and any of its direct children. Tian  Pearl 2002a

2) Given only observational data on a set of variables and infer the causal structure between those variables. This is a much harder problem, perhaps surprisingly, some aspects of causal structure can non the less be inferred with fairly broad, Occum's Razor type assumptions.

Frameworks. Structural Equation Models SEM is where its at …

We will begin our consideration of causal frameworks with regard to the first question.

Holland “For causal inference, it is critical that each unit be potentially exposable to any one of the causes. As an example, the schooling a student receives can be a cause, in our sense, of the students performance on a test, whereas the student's race or gender cannot” This seems like a major draw back … Rubin would see this as still a meaningful causal effect.

First ingredient – a definition of causality.

Rubin defines causal effects in terms of counterfactuals on individual units. But the quantity actually estimated is the average causal effect across all the units. Rubin demonstrates that a classic randomized experiment gives an unbiased estimate of this quantity. Rubin – design experiments to approximate randomized experiments.

2 key parts – definition in of causal effects in terms of counterfactual statement about potential outcomes. 2) Explicit description/modelling of the assignment process that determines which units received which treatments. Rubin 2008

“The assignment mechanism gives the conditional probability of each vector of assignments given the covariates and the potential outcomes” ,”we attempt to assemble data with 
enough covariates that it becomes plausible (or initially arguable) that the 
unknown assignment mechanism is unconfounded given these covariates” 
Rubin 2007 Handbook of stats

The definition of causality in term of counterfactuals is consistent with the graphical model.

The transport safety causality study seems to be confused about the assumptions required for potential outcome modelling.

\section*{Random Thoughts}
Gelmnans objections to causal structure learning - no true 0's. Should not be nessessary to think in terms of conditional independencies to do causal inference. Models are styalized -> likely everything comes out linked to everything. I guess it is a very testing based approach... scoring or averaging might appeal more.

Accepting the limitations of science. (or statistical inference). Note parallel between requirement for assumption in statistics (due to the no-free lunch theorem) and the requirement for assumption in causal inference.

Another question, why does assuming things are linear (even when we know they are not) seem to work so well. Are many relationships in the world linear or roughly linear over the observed domain? Similarly the default assumption is independence of effect.

Practical implications - just how much data is required to do particular sets of tests. When do you have enough data to do things non-parametrically? When could you test conditional independencies. Is there a rule of thumb that would tell you (say you were willing to assume Markovian model) how many data points would be required to do reasonable inference - or to give confidence intervals on the results.

What other assumptions could you use to narrow down confidence bounds (what about if you were willing to rank variables/links by their magnitude of effects, or too specify if the effect would be positive or negative.

could you build something that calculated this for simple models (ie discrete ones ...)

What field of real data might find application (with some form of ability to varify) ... (why are longitudanal studies important ... Longitudanal study of Australian Children, Census data, Neuron firing data, genetic data,

How can we combine with reinforcement learning approaches ... if we get additional information back when we run an experiment, can we use causal inference theory to help tell us which experiment we should do next? What is the simplest (theoretical) model I can come up with that would allow exploration of this?

The whole idea of random variables in this way is not 'true'. If we can break things down enough you end up with quantum theory which, although contains randomness at a fundamental level, this randomness does not obey the standard rules of probability. The way we group events together and call things variables is clearly an approximation, but a powerful one. Relates to things like IID assumption...

When is dislike of black boxes justified. How does this relate to causality? What possible solutions are there to this (ie running multiple models in parralel and report when they diverge). Covariate shift - but you only care about shifts in covariates that effect the predicted outcome. Goal - find a bunch of predictive models that are diverse (in terms of the variables they rely on) but give 'high' predictive accuracy.

\section*{Inferring Causal Structure}
\end{document}

### 2014-09-26 causality_tutorial.tex
completes first draft of section 2 of causality_tutorial.tex

@@
 To understand causality we first need a more formal definition of what we are talking about.
-In this post we will look at three slightly different definitions of causality and use them to describe the following very simple example.
-This section aims to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
+In this section we will look at three slightly different definitions of causality and use them to describe the following very simple example.
+The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
 
@@
 
-\[
+\begin{equation}
+\label{eq:SEM}
 \begin {aligned}
@@
 }
-\]
+\end{equation}
 
@@
 
+The probability of observing $Y$ given $X$, $P(Y = y|X = x) = P(f(x,\epsilon_{y})=y)$ where $\epsilon_{y} ~ P(\epsilon_{y})$ as $\epsilon_{y} \ci \epsilon{x}$
+
 We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
-
-CONSIDER SOLVING THIS JUST FROM THE FUNCTIONAL POINT OF VIEW. MOVE THE DO NOTATION DOWN TO CAUSAL BAYES NETS.
+Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$.
+The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}.
+However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$.
+The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomized experiment.
 
@@
 \subsection{Causal Bayesian Networks}
-Causal Bayesian Networks are, unsurprisingly, an extension of Bayesian Networks.
+Causal Bayesian Networks are, unsurprisingly, an extension of Bayesian networks.
 Any joint probability distribution can be factorized into a product of conditional probabilities.
@@
 \centering
-\caption{General Bayesian Network over 3 variables}
+\caption{General Bayesian network over 3 variables}
 \label{fig:bayesnet}
@@
 
-A CBN is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
+A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
 This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-More generally, if $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network \cite{Pearl2000}.
+More generally, if $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network.
+This is referred to as the truncated product formula \cite{Pearl2000}.
 
-\[
+\begin{equation}
+\label{eq:truncatedproduct}
  P_{do(Z=z)}(X_{1}...X_{N}) =
   \begin{cases}
-  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{$X_{1}...X_{N}$ consistant with $Z=z$}  \\
+  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
    0       & \text{otherwise } 
   \end{cases}
-\]
+\end{equation}
 
@@
 Returning at last to our simple example, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$?
-
-assume the network has the same form as
-
-We have adopted the do-notation, \cite{Pearl1995}, to distinguish the act of intervening to force X to a given value, from observing it to be that value.
-In our simple example $P(Y|do(X)) = P(Y|X)$.
-More generally, for a given SEM and interventional query, there is an algorithm  that can:
+If we assume that $X$ causes $Y$ and there are no common causes then the causal Bayesian network is just $X \rightarrow Y$, the same structure as the network for the SEM specified by equation (\ref{eq:SEM}).
+This network represents the (causal) factorization  $P(X,Y) = P(X)P(Y|X)$, so from equation (\ref{eq:truncatedproduct}), $P(Y|do(X)) = P(Y|X)$.
+In this example, the interventional distribution is equivelent to the observational one as we found earlier using the structural equation model.
 
@@
 
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{}.
+The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{Rosenbaum1983}.
 They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
@@
 \subsection{Tying it all together. How do these models relate?}
-acyclic SEMs are valid Causal Bayesian Networks so everything that applies to them also applies to SEMs.
 
-There is a direct relationship between these ignoreability assumptions and the functional or graphical assumptions encoded by SEM and Causal Bayesian Networks.
+Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed are definitionally (almost) equivalent.
 
-Ignoreability holds when do seperation is valid.
+If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorization of the joint distribution over its variables.
+In other words, the network is a causal Bayesian network.
+All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
+Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
+This corresponds to dropping a term in the factorization and the truncated product formula (equation \ref{eq:truncatedproduct}).
+Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
+We can also connect this to counterfactuals via:
 
-Lets link together the notation for SEMs and Counterfactuals here...
+\begin{equation}
+\begin{aligned}
+&Y^{0} \equiv P(Y|do(X=0)) \\
+&Y^{1} \equiv P(Y|do(X=1))
+\end{aligned}
+\end{equation}
 
-counterfactuals are defined for SEMs and the Rubin-Neyman model but not for causal bayesian networks
+The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
+When discussing the counterfactual model, we actually made the slightly weaker assumption:
 
-We can break the patients down into four groups.
+\begin{equation}
+\label{eq:weakignore}
+X \ci Y^{0} \text{ and } X \ci Y^{1}
+\end{equation}
+
+It is possible to relax the independence of errors assumption we made for SEMs slightly to correspond exactly the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \cite{Richardson2013}.
+To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of functional models or bayesian networks or as conditional independence statements involving counterfactual variables (ignorability assumptions).
+By non-parametrically, I mean that we are not making any assumptions about the form of the relationships between variables.
+Models that are not non-parametrically identifiable can still be identified given assumptions about the distributions of variables and the functional relationship between them, for example, that the functions are linear or that the noise is additive \cite{Peters2014}.
+This form of assumption fits extremely naturally into the structural equation framework.
+
+We can also pose causal queries that are not interventional and cannot be phrased in terms of the do-notation.
+Our patients can be broken down into four groups.
 The first group will recover whether or not they receive treatment, the second group will recover if treated but not on the placebo, the third group will recover on the placebo and not if treated, and the last group will not recover on treatment or placebo.
-Unfortunately, the researches don't know which group each person belongs to.
+Unfortunately, we don't know which group each person belongs to.
 Drawing this up as a table:
@@
 \begin{tabular}{c|c|c|c}
-group & treatment & placebo & probability of group\\
+group & placebo & treatment & probability of group\\
 \hline
-1 & recover & recover & $P(Y^{1}=1,Y^{0}=1)$\\
-2 & recover & die & $P(Y^{1}=1,Y^{0}=0)$\\
-3 & die & recover & $P(Y^{1}=0,Y^{0}=1)$\\
-4 & die & die & $P(Y^{1}=0,Y^{0}=0)$\\
+1 & die & die & $\alpha=P(Y^{0}=0,Y^{1}=0)$\\
+2 & die & recover & $\beta=P(Y^{0}=0,Y^{1}=1)$\\
+3 & recover & die & $\gamma=P(Y^{0}=1,Y^{1}=0)$\\
+4 & recover & recover & $\delta=P(Y^{0}=1,Y^{1}=1)$\\
 \end{tabular}
 
-This difference does not result in different conclusions for interventional queries but there are subtle difference with regard to counterfactual queries, for example mediation.
+The queries we have been asking thus far are about $P(Y^{0}=1) = \gamma + \delta$ and $P(Y^{1}=1) = \beta + \delta$, but suppose we asked the question; what is the probability that this patient, who was not treated and died, would have recovered if they had been treated?
+We know they are in either group 1 or 2 since they died without treatment, so the answer is $\frac{\beta}{\alpha+\beta}$.
+Can we estimate the $\alpha, \beta, \gamma, \delta$ or in other words, identify the joint distribution over the counterfactuals $P(Y^{0},Y^{1})$ given the interventional distributions, $P(Y^{0})$ and $P(Y^{1})$?
+The answer is no, putting our constraints and unknowns in matrix form:
+
+\begin{equation}
+\left(
+\begin{array}{cccc}
+0&0&1&1\\
+1&1&0&0\\
+0&1&0&1\\
+1&0&1&0\\
+1&1&1&1\\
+\end{array}
+\right)
+\left(
+\begin{array}{c}
+\alpha\\
+\beta\\
+\gamma\\
+\delta\\
+\end{array}
+\right)= 
+\left(
+\begin{array}{c}
+P(Y^{0})\\
+1-P(Y^{0})\\
+P(Y^{1})\\
+1-P(Y^{1})\\
+1\\
+\end{array}
+\right)
+\implies
+\left(
+\begin{array}{c}
+\alpha\\
+\beta\\
+\gamma\\
+\delta\\
+\end{array}
+\right)=
+\left(
+\begin{array}{c}
+1-P(Y^{0})-P(Y^{1})+\delta\\
+P(Y^{1})-\delta\\
+P(Y^{0})-\delta\\
+\delta\\
+\end{array}
+\right)
+\end{equation}
+The value of $\delta$ is not determined so the query is not identifiable.
+However we do get bounds on the terms.
+Since probabilities cannot be negative, $P(Y{1})-P(Y^{0})-1 \leq \delta \leq min(P(Y^{1}),P(Y^{0})$.
+Note, if we made the additional assumption $gamma=0$; that the drug did not cause anyone to die who would otherwise have survived, then we can determine the joint distribution over counterfactuals.
+Alternatively, if we could assume that after treatment people returned to their initial state after some period of time, (say we were testing a drug for acne) then we could run a crossover study to determine the joint distribution.
+In a crossover study, the participants are randomly assigned to treatment and placebo, results are measured and then the groups are swapped.
+The scientific and philosophical validity of counterfactual queries remains under question \cite{Dawid2000,Dawid2014}, however they are nonetheless widely posed in the form of attribution of causal effects to different pathways and mediation \cite{Pearl2014,Imai2010a,VanderWeele2011}.
+
+There are differences between the models we have considered when it comes to counterfactual queries.
+Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.
+Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
+Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
+
+In practice, differences in focus and approach across different fields eclipse these actual differences in the models.
+The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
+The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment.
+This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}.
+In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field.
+Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
+
+The equivalence between models for interventional queries allows researchers to combine the best aspects of all of these models.
+You can define your assumptions and determine which variables must be controlled for using a graphical model and apply propensity scores to make the adjustment.
+If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
 
 \section{The Do Calculus}
+The do calculus provides the algorithm that determines if and how a given set of assumptions is sufficient to make an interventional query non-parametrically identifiable.
+
 \subsection{Independence in graphical models: D-separation}
@@
 The cool thing is they are complete.
-For a given SEM, a do query is non-parametrically identifiable if and only if it can be transformed to do-less statement via repeated application of these three rules.
+For a given causal network, a do-query is non-parametrically identifiable if and only if it can be transformed to do-less statement via repeated application of these three rules.
 \subsubsection{Rule 1}
@@
 \subsection{Easy graphical tests for identifiability}
-The completeness of these rules does not make them easy to apply. There are some quick graphical tests that are sufficient (but not necessary) for non-parametric identifiability.
+The completeness of these rules does not make them easy to apply. There are some quick graphical tests that are sufficient (and almost necessary) for non-parametric identifiability.
 
@@
 Allows for identification of causal effects via sequentially identifying other causal effects.
+
+\subsubsection{Examples}
+Show how we can represent all kinds specific biases and their solutions graphically and that the solution matches that given by do-calculus.
 

### 2014-10-07 causality_tutorial.tex
updates causality_tutorial with section 3. Includes conditioning on a confounder as a case that can increase bias

@@
 \usetikzlibrary{arrows,positioning} 
+\usepackage{subcaption}
 \tikzset{
@@
            rounded corners,
-           draw=black, very thick,
-           text width=2.5em,
-           minimum height=2em,
+           draw=black, thick,
+           text width=1em,
+           minimum height=1em,
+           text centered},
+    punktrect/.style={
+    rectangle, 
+    rounded corners, 
+    % fill=black!10,
+    draw=black, thick,
+    minimum height=1em, 
     text centered},
     % Define arrow style
-    pil/.style={
+    pil/.style={punktrect/.style={
+    rectangle, 
+    rounded corners, 
+    % fill=black!10,
+    draw=black, very thick,
+    text width=10em, 
+    minimum height=3em, 
+    text centered, 
+    on chain},
            ->,
@@
 
+TODO
+
+(this section should include explanation of why correlation is not causation)
+
+There are two ways in which correlation is not causation. Firstly correlation may not be statistically significant, - ie not even a genuine association. secondly it can be significant but spurious.
+
 \section{Models of causality}
+\label{sec:causalmodels}
 To understand causality we first need a more formal definition of what we are talking about.
@@
 The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomized experiment.
+In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.
 
@@
 In general these assumption do not hold.
-If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income, that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.
+If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model.
+that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.
 
@@
 Can we estimate the $\alpha, \beta, \gamma, \delta$ or in other words, identify the joint distribution over the counterfactuals $P(Y^{0},Y^{1})$ given the interventional distributions, $P(Y^{0})$ and $P(Y^{1})$?
-The answer is no, putting our constraints and unknowns in matrix form:
+The answer is no, putting our constraints and unknowns in matrix form:D-separation still applies in the augmented model.
 
@@
 Since probabilities cannot be negative, $P(Y{1})-P(Y^{0})-1 \leq \delta \leq min(P(Y^{1}),P(Y^{0})$.
-Note, if we made the additional assumption $gamma=0$; that the drug did not cause anyone to die who would otherwise have survived, then we can determine the joint distribution over counterfactuals.
+Note, if we made the additional assumption $\gamma=0$; that the drug did not cause anyone to die who would otherwise have survived, then we can determine the joint distribution over counterfactuals.
 Alternatively, if we could assume that after treatment people returned to their initial state after some period of time, (say we were testing a drug for acne) then we could run a crossover study to determine the joint distribution.
@@
 Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
+SAY SOMETHING ABOUT NON-RECURSIVE (IE CYCLIC MODELS) HERE.
 
@@
 You can define your assumptions and determine which variables must be controlled for using a graphical model and apply propensity scores to make the adjustment.
+\cite{Morgan2007} provides an excellent and pragmatic introduction to this combined approach.
 If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
@@
 \section{The Do Calculus}
-The do calculus provides the algorithm that determines if and how a given set of assumptions is sufficient to make an interventional query non-parametrically identifiable.
+If you state your assumptions in the form of an acyclic structural equation model or a causal bayesian network, the do calculus provides the methodology to determine if a given causal effect can be estimated from observational data and, if it can, provides to formula to do so.
+The calculus relies heavily on the understanding how conditional independence properties are encoded in graphical models.
+Therefore, in the next section we give a short review of independence in graphical models.
 
 \subsection{Independence in graphical models: D-separation}
+TODO
+Local Markov property implies global markov property (in other words d-separation).
 
 \subsection{The three rules}
-The cool thing is they are complete.
-For a given causal network, a do-query is non-parametrically identifiable if and only if it can be transformed to do-less statement via repeated application of these three rules.
+The do calculus consists of three rules.
+They derive from the causal information encoded in a causal network and the properties of d-separation and do not imply any addition assumptions other than that of specifying the causal network.
+The key property of these three rules is that they are complete.
+If an interventional, do-query, can be transformed to a do-less statement via repeated application of these rules we can get an unbiased estimate of the causal effect from observational data.
+Conversely, if the query is not identified via these rules we cannot get an unbiased estimate of the causal effect without additional assumptions, even with an infinite amount of observational data.
+
 \subsubsection{Rule 1}
-D-separation still applies
+This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention.
+The intervention $do(\boldsymbol{Z}=\boldsymbol{z})$ changes a causal network, $G$, in a simple way.
+Variables in $\boldsymbol{Z}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{z}$.
+This corresponds to deleting the edges with arrows into variables in $\boldsymbol{Z}$  see example Figure \ref{fig:mutilatednet}.
+The resulting 'mutilated' network $G_{\overline{\boldsymbol{Z}}}$ remains a causal network and d-separation still applies.
 
-Let $Z$ be the set of variables on which we are intervening and $G$ our graphical model.
-$G^{\dag}_{Z}$ represents the model where all links going into variables $Z elementof Z$ are deleted and replaced by a new parent node $Z$
+\begin{figure}[]
+\centering
+\caption{Visualizing intervention in a causal bayesian network. Left: original network, $G$. Right: network after the intervention $do(Z=z)$, $G_{\overline{Z}}$}
+\label{fig:mutilatednet}
+\begin{tabular}{cc}
 
-if $(W \ci Y | Z,X)$ in $G_{Z}$:
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$Z$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 2](4){$W$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3)
+    (4) edge node {} (2);
+
+\end{tikzpicture} &
+
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$z$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 2](4){$W$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (3)
+    (2) edge node {} (3);
+
+\end{tikzpicture}
+\end{tabular}
+
+\end{figure}
+
+If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{Z}}}$:
 \begin{equation}
 \label{eq:Do1}
- P(Y|do(Z=z),X=x,W=w) = P(Y|do(Z=z),X=x) 
+ P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x}) 
 \end{equation}
+
+$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z})$ in $G_{\overline{\boldsymbol{Z}}}$:
 \begin{equation}
 \label{eq:Do12}
- P(Y|do(Z=z),W=w) = P(Y|do(Z=z)) 
+ P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z})) 
 \end{equation}
 
+This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.
+
 \subsubsection{Rule 2}
+Rule 2 states when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening  $do(\boldsymbol{X}=\boldsymbol{x})$ have the same effect on the distribution of the outcome $\boldsymbol{Y}$.
+You can think of this as when correlation is causation.
+It is easiest to understand by explicitly including the intervention process in the graphical model.
+We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on.
+Let $\epsilon$ be some arbitrary value not in the set of possible values for $X$.
+If $X = \epsilon$ the distribution of $X$ is what it was without intervention.
+Otherwise, if $X=x$, $X$ deterministically takes the value $x$ and is independent of its previous parents, representing the intervention $do(X=x)$.
+We use the notation $G^{\dagger}$ to represent $G$ augmented with these decision nodes.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
+
+\begin{equation}
+\label{eq:Do2}
+P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w})
+\end{equation}
+
+$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X})$ in $G^{\dagger}$:
+\begin{equation}
+\label{eq:Do22}
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x})
+\end{equation}
+
 If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.
 
-if $(W \ci Y | Z,X)$ in $G_{Z}$:
+\subsubsection{Rule 3}
+This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
 \begin{equation}
-\label{eq:Do2}
-P(Y|do(Z=z),do(X=x),W=w) = P(Y|do(Z=z),X=x,W=w)
+\label{eq:Do3}
+P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
+\end{equation}
+$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}})$ in $G^{\dagger}$:
+\begin{equation}
+\label{eq:Do32}
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y})
+\end{equation}
+The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ without conditioning on $\boldsymbol{X}$ implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ which starts on an arrow leaving $\boldsymbol{X}$.
+This means there is no causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$ and the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ does not alter the distribution of $P(\boldsymbol{Y})$.
+
+\subsection{Easy graphical tests for identifiability}
+Determining if a network is identifiable via the two calculus still requires some algebra and it may not always be obvious what sequence of rules are required for a given network.
+From the do-calculus we can derive some quick graphical tests that are sufficient (but not necessary) for non-parametric identifiability, and align with intuitive ideas about estimating causal effect by adjusting for confounding variables.
+
+\subsubsection{If all variables are observed}
+If all the variables in a causal network can be measured any interventional query on that network are identifiable.
+You just have to condition on the parents of the variables on which you intervene.
+For discrete variables, the causal effect is given by:
+
+\begin{equation}
+P(\boldsymbol{Y}|do(\boldsymbol{X})) = \sum_{\boldsymbol{Pa_{X}}}{P(\boldsymbol{Y}|\boldsymbol{X},\boldsymbol{Pa_{X}})P(\boldsymbol{Pa_{X}})}
 \end{equation}
 
-This also implies that if $Y d-sep X|X$:
+where $\boldsymbol{Pa_{X}}$ are the parents of $\boldsymbol{X}$.
+
+Conditioning on the parents blocks all information that does not flow causally (along the arrows) from the variables on which we intervened which eliminates spurious correlations and gives the causal effect.
+
+\subsubsection{The Back Door criterion}
+The back door criterion generalizes from the case where all variables can be measured by observing we don't need to condition on all the parents of the intervened on variables: it is sufficient to condition on a set of variables that block all spurious paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$.
+A spurious, or back door, path is one that leaves $\boldsymbol{X}$ against the direction of the causal arrows.
+If we have a set of variables $\boldsymbol{Z}$ that block all back door paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$, where no variable in $\boldsymbol{Z}$ is a descendent of $\boldsymbol{X}$ then:
+
 \begin{equation}
-\label{eq:Do22}
-P(Y|do(X=x)) = P(Y|X=x)
+P(\boldsymbol{Y}|do(\boldsymbol{X})) = \sum_{Z}{P(\boldsymbol{Y}|\boldsymbol{X},\boldsymbol{Z})P(\boldsymbol{Z})}
 \end{equation}
 
-\subsubsection{Rule 3}
-if $(W \ci Y | Z,X)$ in $G_{Z}$:
+The requirement that $\boldsymbol{Z}$ should not include descendants of $\boldsymbol{X}$ stems from the fact that we don't want to open any new spurious paths.
+If $Z$ descends from $X$ then it may lie on a path of the form $X \rightarrow...\rightarrow Z \leftarrow...
+Y$, which conditioning on $Z$ would open.
+The back door criterion corresponds exactly to the strong ignoreability assumption \cite{Rosenbaum1983}, used to justify the validity of a causal estimate in the counterfactual approach .
+If $\boldsymbol{Z}$ satisfies the backdoor criterion then, for a single binary treatment variable $X$ and outcome variable $Y$, $\{Y^{0},Y^{1}\} \ci X | \boldsymbol{Z}$.
+
+\subsubsection{The Front Door Criterion}
+It we have an unobservable variable creating a spurious path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ then we cannot apply the back door criterion to directly estimate $P(\boldsymbol{Y}|do(\boldsymbol{X}))$.
+However, if there is a set of mediating variables $\boldsymbol{M}$ that intercepts all all the direct (non-spurious) paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and we can estimate the causal effect of $\boldsymbol{X}$ on $\boldsymbol{M}$ and the causal effect of $\boldsymbol{M}$ on $\boldsymbol{Y}$ then we can combine these terms to get the causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$.
+
 \begin{equation}
-\label{eq:Do3}
-P(Y|do(Z=z),do(X=x),W=w) = P(Y|do(Z=z),W=w)
-\end{equation}
-\begin{equation}
-\label{eq:Do32}
-P(Y|do(X=x)) = P(Y)
+\begin {aligned}
+P(\boldsymbol{Y}|do(\boldsymbol{X})) & = \sum_{\boldsymbol{M}}{P(\boldsymbol{M}|do(\boldsymbol{X}))P(\boldsymbol{Y}|do(\boldsymbol{M}))} \\
+&= \sum_{\boldsymbol{M}}{P(\boldsymbol{M}|\boldsymbol{X})} \sum_{\boldsymbol{X}}P(\boldsymbol{Y}|\boldsymbol{X},\boldsymbol{M})P(\boldsymbol{X}) 
+\end {aligned}
 \end{equation}
 
-\subsection{Easy graphical tests for identifiability}
-The completeness of these rules does not make them easy to apply. There are some quick graphical tests that are sufficient (and almost necessary) for non-parametric identifiability.
+\subsubsection{Examples}
 
-\subsubsection{If all variables are observed}
-All interventional queries are identifiable. Just condition on the parents.
+We are now in a position to consider some more interesting examples than the one in section \ref{sec:causalmodels}.
+Consider the question; does eating chocolate increase or decrease the risk of obesity?
+Assume we have data from a cross-sectional study where a large group of people were weighed and asked how much chocolate they ate.
+It is easy to come up with a whole lot of variables that may causally effect chocolate consumption and obesity, ie.
+age, income and gender, so we cannot conclude that the causal effect is equivalent to the correlation.
+If we assume there may be unknown variables (denoted by $U$ in all the following examples) that effect both obesity and chocolate consumption, (figure \ref{fig:chockNI}), the causal effect cannot be identified as there is no way to block the spurious path through the unobserved variables.
+If we assume the only confounding variables are income, age and gender, (figure \ref{fig:chockI}), we can calculate the causal effect by conditioning on these variables.
 
-\subsubsection{The Back Door criterion}
-The back door criterion nicely aligns to answer the problem, when can you calculate a causal effect by conditioning on or adjusting for confounding variables and answers which variables should be included.
+\begin{figure}[h]
+\centering
+\caption{Chocolate and Obesity}
+\label{fig:chocolate}
+\begin{subfigure}[t]{0.4\textwidth}
+\caption{Not identifiable}
+\label{fig:chockNI}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.5cm,  thick,main node/.style={punktrect}]
 
-\subsubsection{The Front Door Criterion}
-Allows for identification of causal effects via sequentially identifying other causal effects.
+ %nodes
+\node[main node](1){$\boldsymbol{U}$};
+\node[main node, below left=of 1](2){Chocolate};
+\node[main node, below right=of 1](3){Obesity};
 
-\subsubsection{Examples}
-Show how we can represent all kinds specific biases and their solutions graphically and that the solution matches that given by do-calculus.
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.4\textwidth}
+\caption{Identifiable}
+\label{fig:chockI}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punktrect}]
+
+ %nodes
+\node[main node](1){Age};
+\node[main node, left= of 1](4){Income};
+\node[main node, right=of 1](5){Gender};
+\node[main node, below left=of 1](2){Chocolate};
+\node[main node, below right=of 1](3){Obesity};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3)
+    (4) edge node {} (2)
+        edge node {} (3)
+    (5) edge node {} (2)
+    	edge node {} (3);
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+In reality of course there are many other variables involved  \cite{Golomb2012} and it may not be obvious from prior knowledge or theory what role the variables play in the network.
+This is important as conditioning on the wrong variables can increase the bias in a causal estimate: Conditioning on a collider variable, $\rightarrow Z \leftarrow$, may open a previously blocked, spurious path \cite{Sjolander2009a} (figure \ref{fig:colliderCondition}) .
+Conditioning on a variable that mediates between the cause and effect biases the causal estimate by blocking a true casual path (figure \ref{fig:mediationBias}).
+Conditioning on a variable that is a true confounder, that is a common cause of both $X$ and $Y$, may increase the bias in our causal estimate if it moves the causal estimate in the opposite direction than would correcting for a hidden (and stronger) confounder $U$.
+Finally, (although it is not obvious graphically) conditioning on a variable that is a cause of $X$ but not (or only weakly) a cause of $Y$ can amplify or introduce bias \cite{Pearl2012b,Wooldridge2009}.
+
+\begin{figure}[h]
+		\caption{Conditioning on a collider variable can introduce bias}\label{fig:colliderCondition}
+       \centering
+        \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+                \caption{post-treatment bias: conditioning on $Z$ opens the spurious path $X,Z,Y$}
+                \label{fig:postTreatmentBias}
+                \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[](1){}; 
+\node[main node, left = of 1](2){X};
+\node[main node, right= of 1](3){Y};
+\node[main node, below =of 1](4){Z};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (3)
+    	edge node {} (4)
+    (3) edge node {} (4);
+\end{tikzpicture}
+        \end{subfigure}%
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+                \caption{M-bias: conditioning on $Z$ opens the path $X,U_{1},Z,U_{2},Y$}
+                \label{fig:Mbias}
+                \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){Z}; 
+\node[main node, below left = of 1](2){$X$};
+\node[main node, below right= of 1](3){$Y$};
+\node[main node, above left = of 1](5){$U_{1}$};
+\node[main node, above right = of 1](6){$U_{2}$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (3)
+    (5) edge node {} (1)
+    	edge node {} (2)
+    (6) edge node {} (1)
+    	edge node {} (3)
+    (1) edge node {} (2)
+    	edge node {} (3);
+\end{tikzpicture}
+        \end{subfigure}    
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\caption{Conditioning on a mediating variable ($Z$) can introduce bias}
+\label{fig:mediationBias}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){$X$}; 
+\node[main node, above right = of 1](2){$Z$};
+\node[main node, below right= of 1](3){$U$};
+\node[main node, above right = of 3](4){$Y$};
+\path[every node/.style={font=\sffamily\small}]
+	(1) edge node {} (2)
+		edge node {} (3)
+	(2) edge node {} (4)
+	(3) edge node {} (4);
+
+\end{tikzpicture}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\caption{Conditioning on a confounder ($Z$) can increase bias}
+\label{fig:biasAmplification}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){Z}; 
+\node[main node, below left = of 1](2){X};
+\node[main node, right = of 1](4){U};
+\node[main node, below right= of 4](3){Y};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (3)
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    	edge node {} (3);
+
+\end{tikzpicture}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\caption{Bias amplification}
+\label{fig:biasAmplification}
+\begin{subfigure}[t]{0.4\textwidth}
+\centering
+\caption{Conditioning on an instrumental variable $Z$ can increase bias in estimate of effect of $X$ on $Y$}
+\label{fig:instrumentalBias}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){U}; 
+\node[main node, below left = of 1](2){X};
+\node[main node, below right = of 1](3){Y};
+\node[main node, above left= of 2](4){Z};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (3)
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2);
+
+\end{tikzpicture}
+\end{subfigure}
+ \begin{subfigure}[t]{0.4\textwidth}
+\centering
+\caption{Conditioning on a confounding variable that is strongly linked to $X$ and weakly linked to $Y$ can increase bias.}
+\label{fig:semiinstrumentalbias}
+
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){U}; 
+\node[main node, below left = of 1](2){X};
+\node[main node, below right = of 1](3){Y};
+\node[main node, above left= of 2](4){Z};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (3)
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    	edge [dashed] node {} (3);
+
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+For some problems, the temporal ordering of events imposes restrictions on the causal structure.
+Consider the example of a non-randomized trial.
+A new treatment is developed and made available to a large group of patients some of whom decide to take.
+Researchers have variables on the health, income, gender, age, etc of the patients before the treatment was offered (pre-treatment variables) and collect further data on the health of all patients after treatment has begun (post-treatment variables).
+It has been understood for a long time that conditioning on post-treatment variables can increase bias \cite{Rosenbaum1984a}.
+The realization that that conditioning on pre-treatement can also introduce bias is more recent.
+Not conditioning on post-treatment variables avoids the biases in figure \ref{fig:postTreatmentBias} and figure \ref{fig:mediationBias}.
+However, M-bias (figure \ref{fig:Mbias}) and bias amplification (figure \ref{fig:biasAmplification}) are consequences of conditioning on pre-treatment variables.
+
+It is clear we need to be wary of the condition on everything approach and the condition on all pre-treatment variables approach.
+But what is the alternative if the structure of the causal network is not clear from prior knowledge?
+One approach would be to construct a number of different plausible structures and evaluate the desired causal effect for each of them.
+This would give more accurate uncertainty bounds for the causal effect.
+Alternatively, one could apply a causal discovery algorithm (see section \ref{sec:causaldiscovery}).
+
+It may be that there are some general conditions under which certain types of biases are less likely or weaker than other.
+For example, in the situation in figure \ref{fig:Mbias}, where $Z$ is both a confounding variable and part of an M-structure.
+Conditioning on $Z$ blocks the spurious path $X \leftarrow Z \rightarrow Y$ and opens the spurious path $X \leftarrow U_{1} \rightarrow Z \leftarrow U_{2} \rightarrow Y$.
+Should we condition on $Z$ or not?
+Under what general assumptions is one of these paths weaker than the other.
+There has been relatively little work in this area.
+\cite{Vansteelandt2012} gives a description of the problem and links it to selecting covariates when the goal is causal estimation.
+\cite{Myers2011} use simulation studies to consider if one should adjust for a variable that could be either an instrumental variable or a confounder for the effect of $X$ on $Y$.
 
@@
 \subsection{Bounding causal effects}
+Causal effects that cannot be identified can often still be bounded. In other words, bounds require weaker assumptions than point identification.
+
 \subsection{Additional assumptions}
@@
 
-\section{Structure learning}
+\section{Causal structure learning (causal discovery)}
+\label{sec:causaldiscovery}
 peters approach \cite{Peters2014}
 Pearl/Verma/Sprites approach \cite{Pearl2000}
+
+\section{Causality and prediction}
+Refresh the difference between a causal query and predictive query.
+Point out the importance of accurate prediction for causal inference. For example, if we can predict the control response, we can run experimental studies without controls (\cite{Brodersen2013})
 

### 2014-10-08 causality_tutorial.tex
first completed draft of section 3

@@
 \def\ci{\perp\!\!\!\perp} % from Wikipedia
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{definition}{Definition}[section]
 
@@
 \subsection{Causal Bayesian Networks}
+\label{sec:causalbayesnets}
 Causal Bayesian Networks are, unsurprisingly, an extension of Bayesian networks.
@@
 Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
-SAY SOMETHING ABOUT NON-RECURSIVE (IE CYCLIC MODELS) HERE.
+In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
 
@@
 If you state your assumptions in the form of an acyclic structural equation model or a causal bayesian network, the do calculus provides the methodology to determine if a given causal effect can be estimated from observational data and, if it can, provides to formula to do so.
-The calculus relies heavily on the understanding how conditional independence properties are encoded in graphical models.
-Therefore, in the next section we give a short review of independence in graphical models.
+The calculus relies heavily on the understanding how conditional independence properties are encoded in bayesian networks.
+Therefore, in the next section we give a short review of independence in bayesian networks.
 
-\subsection{Independence in graphical models: D-separation}
-TODO
-Local Markov property implies global markov property (in other words d-separation).
+\subsection{Independence in bayesian networks: D-separation}
+
+Recall from section \ref{sec:causalbayesnets} the local markov condition:
+
+\begin{theorem}
+\textbf{(Local markov condition)} Given a bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decendents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
+\end{theorem}
+
+The set of conditional independence relations given by the local markov condition can enforce additional independences that also hold in all distributions that are compatible with $G$.
+D-separation extends the local markov property to find these additional independences and allows us to read from a network if a given conditional independence statement is true in all distributions compatible with that network.
+
+\begin{theorem}
+\textbf{(D-separation)} If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
+Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected given $\boldsymbol{Z}$ then there will be at least one distribution $P'$ that is compatible with $G$ where they are dependent.
+\end{theorem}
+
+$X$ conditionally independent of $Y$ given $Z$ means, if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
+You can also think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
+Let's examine the four possible bayesian networks over the variables ${X,Y,Z}$ that do not have a direct link  ${X \rightarrow Y}$ (figure \ref{fig:dseptriples}).
+
+In figure (\ref{fig:dseptripple1}) conditioning on $Z$ renders $X$ and $Y$ conditionally independent.
+For example, $X$ could be number of years a person smoked, $Z$ the amount of tar in their lungs and $Y$ lung cancer.
+If you don't know $Z$, knowing how much they smoked changes your estimate of the likelyhood of their having lung cancer, but, if smoking causes cancer purely through the build up of tar in the lungs as we have assumed in this model, then once you measure the amount of tar in the lungs knowing how many years a person smoked for gives you no additional information on the probability that they will develop lung cancer.
+Figure (\ref{fig:dseptripple2}) is the same as (\ref{fig:dseptripple1}) with $X$ and $Y$ exchanged, which does not change anything since conditional independence is symmetric $(X \ci Y|Z) \Leftrightarrow (Y \ci X|Z)$.
+
+In figure (\ref{fig:dseptripple3}), $Z$ could represent having an illness and $X$ and $Y$ could be two independent tests that can detect (with some error) that illness.
+Without conditioning on $Z$ you would expect results from the two different tests to be highly correlated, but if you already know someone has the illness then learning they tested positive to the test $X$ doesn't change the likelyhood they test positive in test $Y$.
+Again, conditioning on $Z$ blocks the path from $X$ to $Y$.
+
+The arrangement of variables in figure (\ref{fig:dseptripple4}), often referred to as a collider or v-structure, is the odd one out.
+In this case $X$ is independent of $Y$ without conditioning on $Z$ but if you do condition on $Z$ then $X$ and $Y$ actually become dependent.
+Suppose $X$ is socio-economic status, $Y$ is gender and $Z$ is receipt of a scholarship available to female or disadvantaged students.
+Roughly equal numbers of women and men grow up in poorer families so $X$ is independent of $Y$, but if you know that a given student received a scholarship, then learning that they are male tells you they must have a low socio-economic background.
+Conditioning on $Z$ unblocks the path from $X$ to $Y$.
+
+\begin{figure}[h]
+\centering
+\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
+\label{fig:dseptriples}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y| Z$}
+\label{fig:dseptripple1}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={punkt}]
+\node[main node](1){$X$};
+\node[main node, below =of 1](2){$Z$};
+\node[main node, below =of 2](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    (2) edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|Z$}
+\label{fig:dseptripple2}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={punkt}]
+\node[main node](1){$Y$};
+\node[main node, below =of 1](2){$Z$};
+\node[main node, below =of 2](3){$X$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    (2) edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|Z$}
+\label{fig:dseptripple3}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={punkt}]
+\node[main node](1){$Z$};
+\node[main node, below left =of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+        edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|\{\}$}
+\label{fig:dseptripple4}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={punkt}]
+\node[main node](1){$Z$};
+\node[main node, above left =of 1](2){$X$};
+\node[main node, above right =of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (1)
+    (3) edge node {} (1);	
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+For a general bayesian network a path between two variables may involve more than three variables.
+
+\begin{definition}
+\textbf{(unblocked path)} A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
+It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
+\end{definition}
+
+If all paths between two nodes are blocked by conditioning on $\boldsymbol{Z}$ then no information can flow between them and they are d-separated given $\boldsymbol{Z}$.
+
+\begin{definition}
+The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, after conditioning on $\boldsymbol{Z}$ there are no unblocked paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$.
+\end{definition}
+
+For a more complete introduction to d-separation (with proofs) see \cite{Koller2009}.
 
@@
 The key property of these three rules is that they are complete.
-If an interventional, do-query, can be transformed to a do-less statement via repeated application of these rules we can get an unbiased estimate of the causal effect from observational data.
+If an interventional (do-type) query can be transformed to a do-less statement via repeated application of these rules we can get an unbiased point estimate of the causal effect from observational data.
 Conversely, if the query is not identified via these rules we cannot get an unbiased estimate of the causal effect without additional assumptions, even with an infinite amount of observational data.
@@
 This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention.
-The intervention $do(\boldsymbol{Z}=\boldsymbol{z})$ changes a causal network, $G$, in a simple way.
-Variables in $\boldsymbol{Z}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{z}$.
-This corresponds to deleting the edges with arrows into variables in $\boldsymbol{Z}$  see example Figure \ref{fig:mutilatednet}.
-The resulting 'mutilated' network $G_{\overline{\boldsymbol{Z}}}$ remains a causal network and d-separation still applies.
+The intervention $do(\boldsymbol{X}=\boldsymbol{x})$ changes a causal network, $G$, in a simple way.
+Variables in $\boldsymbol{X}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{x}$.
+This corresponds to deleting the edges with arrows into variables in $\boldsymbol{X}$  (see figure \ref{fig:mutilatednet}).
+The resulting 'mutilated' network $G_{\overline{\boldsymbol{X}}}$ remains a causal network and d-separation still applies.
 
-\begin{figure}[]
+If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
+\begin{equation}
+\label{eq:Do1}
+ P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
+\end{equation}
+
+$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
+\begin{equation}
+\label{eq:Do12}
+ P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) 
+\end{equation}
+
+This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.
+
+\begin{figure}[h]
 \centering
-\caption{Visualizing intervention in a causal bayesian network. Left: original network, $G$. Right: network after the intervention $do(Z=z)$, $G_{\overline{Z}}$}
+\caption{Intervention in a causal bayesian network}
 \label{fig:mutilatednet}
-\begin{tabular}{cc}
-
+\begin{subfigure}[t]{0.3\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:mutilatedOriginal}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \node[main node](1){$U$};
-\node[main node, below left=of 1](2){$Z$};
+\node[main node, below left=of 1](2){$X$};
 \node[main node, below right=of 1](3){$Y$};
@@
 
-\end{tikzpicture} &
-
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.5\textwidth}
+\centering
+\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
+\label{fig:mutilatedAfter}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \node[main node](1){$U$};
-\node[main node, below left=of 1](2){$z$};
+\node[main node, below left=of 1](2){$x$};
 \node[main node, below right=of 1](3){$Y$};
@@
 \end{tikzpicture}
-\end{tabular}
-
+\end{subfigure}
 \end{figure}
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{Z}}}$:
-\begin{equation}
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x}) 
-\end{equation}
-
-$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z})$ in $G_{\overline{\boldsymbol{Z}}}$:
-\begin{equation}
-\label{eq:Do12}
- P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z})) 
-\end{equation}
-
-This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.
 
@@
 It is easiest to understand by explicitly including the intervention process in the graphical model.
-We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on.
+We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on (figure \ref{fig:augmentedNetwork}).
 Let $\epsilon$ be some arbitrary value not in the set of possible values for $X$.
@@
 If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.
+
+\begin{figure}[h]
+\centering
+\caption{Augmentated causal network with intervention}
+\label{fig:augmentedNetwork}
+\begin{subfigure}[t]{0.4\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:augmentedOriginal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left = of 1](4){$U_{1}$};
+\node[main node, above right = of 1](5){$U_{2}$};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(2) edge node {} (3)
+ 	(4) edge node {} (2)
+ 		edge node {} (1)
+ 	(5) edge node {} (3)
+ 		edge node {} (1)
+ 	(1) edge node {} (2)
+ 		edge node {} (3);
+
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.4\textwidth}
+\centering
+\caption{augmented network after the intervention $do(Z=z)$, $G^{\dagger}_{\overline{Z}}$}
+\label{fig:augmentedAfter}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.3cm,
+  thick,main node/.style={punkt}]
+
+\node[main node](1){$z$};
+\node[below left=of 1](2){};
+\node[below right=of 1](3){};
+\node[main node, below left=of 2](4){$X$};
+\node[main node, below right=of 3](5){$Y$};
+\node[above right=of 1](6){};
+\node[above left=of 1](7){};
+\node[main node, above left=of 7](8){$U_{1}$};
+\node[main node, above right=of 6](9){$U_{2}$};
+\node[main node, above left=of 4](10){$\hat{X}$};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(4) edge node {} (5)
+ 	(10) edge node {} (4)
+ 	(8) edge node {} (4)
+ 	(9) edge node {} (5)
+ 	(1) edge node {} (4)
+ 		edge node {} (5);
+
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[]
 		\caption{Conditioning on a collider variable can introduce bias}\label{fig:colliderCondition}
@@
 
-\begin{figure}[h]
+\begin{figure}[]
 \centering
@@
 \cite{Vansteelandt2012} gives a description of the problem and links it to selecting covariates when the goal is causal estimation.
-\cite{Myers2011} use simulation studies to consider if one should adjust for a variable that could be either an instrumental variable or a confounder for the effect of $X$ on $Y$.
+\cite{Myers2011, Pearl2012b} consider if one should adjust for a variable that could be either an instrumental variable or a confounder for the effect of $X$ on $Y$.
 
@@
 causal anti-causal \cite{Janzing2012}
+learning what indicates causality.
 
@@
 \label{sec:causaldiscovery}
+If you have a method for doing things pairwise, then you can extend it to a full network.
+
 peters approach \cite{Peters2014}

### 2014-10-14 causality_tutorial.tex
updates causality_tutorial and bib file

@@
 \usepackage{amsmath}
+\usepackage{listings}
 \usepackage{amsfonts}
 \usepackage{amssymb}
+\usepackage{tabularx}
+\usepackage{enumitem}
+\usepackage{algorithm}% http://ctan.org/pkg/algorithm
+\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
 \usepackage{parskip}
@@
 \usepackage{subcaption}
+
+\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
+
+\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
+
 \tikzset{
@@
     % Define arrow style
-    pil/.style={punktrect/.style={
-    rectangle, 
-    rounded corners, 
-    % fill=black!10,
-    draw=black, very thick,
-    text width=10em, 
-    minimum height=3em, 
-    text centered, 
-    on chain},
-           ->,
+    pil/.style={
+           o->,
            thick,
@@
 TODO
+1BVAEU
 
@@
 \subsection{Structural Equation Models}
-
+\label{sec:SEM}
 Structural equation models (SEMs) describe a deterministic world, where underlying mechanisms determine the output of any process for a given input.
@@
 This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random; the outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-
-The probability of observing $Y$ given $X$, $P(Y = y|X = x) = P(f(x,\epsilon_{y})=y)$ where $\epsilon_{y} ~ P(\epsilon_{y})$ as $\epsilon_{y} \ci \epsilon{x}$
+This is justifiable on the grounds that $X$ is random.
+The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
 
@@
 \subsection{Independence in bayesian networks: D-separation}
+\label{sec:dsep}
 
@@
 \end{definition}
+
+Let us conclude with an example. Assuming the model in figure \ref{fig:chocindepstress}: Is obesity independent of stress given chocolate consumption?
+\begin{figure}[h]
+\centering
+\caption{D-separation example}
+\label{fig:chocindepstress}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.5cm,  thick,main node/.style={punktrect}]
+ %nodes
+\node[main node](1){Cholocate};
+\node[main node, above left=of 1](2){Stress};
+\node[main node, above right=of 1](3){Genotype};
+\node[main node, below = of 1](4){Obesity};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(2) edge node {} (1)
+ 	(3) edge node {} (1)
+ 		edge node {} (4)
+ 	(1) edge node {} (4);
+
+\end{tikzpicture}
+\end{figure}
+
+There are two paths from stress to obesity in figure \ref{fig:chocindepstress}, $Stress \rightarrow Chocolate \rightarrow Obesity$ and $Stress \rightarrow Chocolate \leftarrow Genotype \rightarrow Obesity$.
+In the first, the variable chocolate is an intermediary as in figure \ref{fig:dseptripple1} so this path is blocked when we condition on chocolate.
+However, in the second path chocolate is a collider as in figure \ref{fig:dseptripple4} so this path is unblocked when we condition on chocolate.
+Since there is an unblocked path from stress to obesity after conditioning on chocolate, obesity is not independent of stress given chocolate consumption.
+This example shows the role of a variable (if it is a collider or not) is relative to the path you are considering.
+Conditioning on a variable can block some paths and unblock others.
 
@@
 \begin{equation}
+\label{eq:allvarsobserved}
 P(\boldsymbol{Y}|do(\boldsymbol{X})) = \sum_{\boldsymbol{Pa_{X}}}{P(\boldsymbol{Y}|\boldsymbol{X},\boldsymbol{Pa_{X}})P(\boldsymbol{Pa_{X}})}
@@
 \subsubsection{The Back Door criterion}
-The back door criterion generalizes from the case where all variables can be measured by observing we don't need to condition on all the parents of the intervened on variables: it is sufficient to condition on a set of variables that block all spurious paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$.
+The back door criterion is simplified version of rule 2.
+It generalizes from the case where all variables can be measured by observing we don't need to condition on all the parents of the intervened on variables: it is sufficient to condition on a set of variables that block all spurious paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$.
 A spurious, or back door, path is one that leaves $\boldsymbol{X}$ against the direction of the causal arrows.
@@
 \label{sec:causaldiscovery}
-If you have a method for doing things pairwise, then you can extend it to a full network.
-
-peters approach \cite{Peters2014}
-Pearl/Verma/Sprites approach \cite{Pearl2000}
+
+In previous sections we discussed when assumptions about the structure of the variables in a specific problem is sufficient to identify a causal effect.
+This approach relies on having enough prior knowledge or theory about the problem to allow you to, at least partially, specify the causal network.
+In this section, we consider the much harder problem of causal inference where you need to learn the network.
+Causal inference might seem impossible without specific assumptions about the structure of the variables involved but, amazingly, some aspects of causal structure can be determined from much more general assumptions.
+
+In this section we consider variants of the following problem: Assume there is some acyclic causal network $G$ that generated the distribution $P(\boldsymbol{V})$ from which our data has been sampled.
+Our goal is to recover the network from this data.
+
+What can we look for in the distribution that could give us clues as to the structure of the network?
+
+\subsection{Causal discovery with conditional independence tests}
+One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
+We know, (section \ref{sec:dsep}), if $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in $P$ .
+However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
+This requires that we assume the reverse condition:$(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
+This assumption, commonly referred to as  \textbf{\textit{faithfulness}}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
+Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} for $P$.
+
+Faithfulness is an assumption.
+It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
+However, most distributions generated by a causal bayesian network will be faithful to that network.
+For faithfulness to be violated, different causal effects must exactly balance one-another out.
+For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
+If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
+However, this independence is not stable.
+It would disappear under a small perturbation to any of the parameters.
+
+Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$.
+This set can also be referred to as the markov equivalence class compatible with $P$.
+
+\subsubsection{Without hidden common causes}
+The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of bayesian networks we must consider.
+
+We will begin with a brute force algorithm (described as the SGS algorithm in \cite{Sprites} and IC algorithm in \cite{Pearl2000}).
+While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.
+
+\begin{table}[H]
+ \begin{tabularx}{\textwidth}{X}
+ \hline
+\rule{0pt}{2.5ex} 
+ \textbf{The SGS (or IC) Algorithm}\\
+ \hline
+ \rule{0pt}{2.5ex}
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
+\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
+ \begin{enumerate}[itemsep=8pt]
+  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an undirected link to form a complete graph.
+  \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent.
+If such a set (including the empty set) exists then $a$ and $b$ cannot be directly connected in $G$ so delete the link.
+  \label{alg:SGSexponential}
+
+  \item For all pairs of non-linked variables $(\alpha,\beta)$ with a common neighbour, $c$, if $c \notin \boldsymbol{S}_{\alpha,\beta}$, then $c$ must be a collider in the path $\alpha,c,\beta$ so  add arrows to direct the links $\alpha-c$ and $\beta-c$ towards $c$.
+  \label{alg:SGScolliders}  
+  \item Recursively try to orient any edges that remain undirected to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
+  \label{alg:SGSfinal}
+\end{enumerate}\\
+ \hline
+\end{tabularx}
+\end{table}
+
+The SGS algorithm utilizes the fact that a collider structure (f
+igure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
+Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
+Unfortunately the number of conditional independence tests required for step \ref{alg:SGSexponential} grows exponentially (in the worst case) with the number of variables.
+Not only that, but for each edge that is in the true network, the algorithm will always tests all other possible subsets of variables.
+If the assumption that there are no hidden common causes or that the distribution is faithful are violated, step \ref{alg:SGScolliders} of the SGS algorithm can produce double headed arrows.
+
+The PC algorithm \cite{Sprites} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilize the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
+It also checks for low order conditional independence relations before higher order ones.
+This allows it to exploit any sparsity in the true network, leading to much better average case performance  \cite{Sprites} (although the worst case, where the true network is complete, is still exponential).
+With finite data, the order in which the links are considered can change the output (unlike for SGS).
+The effect of wrongly removing a link early on flows through to later conditional independence tests by changing which nodes are considered adjacent.
+
+\begin{table}[H]
+ \begin{tabularx}{\textwidth}{X}
+ \hline
+\rule{0pt}{2.5ex} 
+ \textbf{The PC Algorithm}\\
+ \hline
+ \rule{0pt}{2.5ex}
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
+\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
+ \begin{enumerate}[itemsep=8pt]
+  \item As for SGS
+  \item \textbf{for} each link $a-b$:
+  \begin{itemize}[label={}]
+   \item $n = 0$
+   \item $\boldsymbol{A}_{a,b} = \{A_{1}...A_{j}\}$ be the set of nodes adjacent to $a$ and/or $b$
+   \item \textbf{while} $a$ and $b$ are connected and $n < j$:
+   \item 
+    	\begin{itemize}[label={}]
+    	\item \textbf{if} any subset of size $n$ of $\boldsymbol{A}$ makes $a$ and $b$ conditionally independent:
+    	\item \begin{itemize}[label={}]
+    			delete the link
+    		  \end{itemize}
+
+    	\item $n = n+1$
+    	\end{itemize}
+  \end{itemize}   
+  \item as for SGS
+  \item as for SGS
+\end{enumerate}\\
+ \hline
+\end{tabularx}
+\end{table}
+
+The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution.
+Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects with equation \ref{eq:allvarsobserved}.
+We can then bound the true causal effect by combining the results for the all the networks.
+This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
+An implementation is available in the R package \cite{Kalisch2012}
+
+\subsubsection{With hidden variables}
+There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables.
+With an unknown number of hidden variables there are an infinity many possible structures to search over.
+In addition, the space of causal networks is not closed under marginalization.
+If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalizing over $\boldsymbol{U}$, may not be faithful to any bayesian network (see figure \ref{fig:DAGSnotclosed}).
+
+\begin{figure}
+\centering
+\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
+\label{fig:DAGSnotclosed}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){U}; 
+\node[main node, below left = of 1](2){B};
+\node[main node, below right = of 1](3){C};
+\node[main node, above left= of 2](4){A};
+\node[main node, above right = of 3](5){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    (5) edge node {} (3);
+
+\end{tikzpicture}
+
+\end{figure}
+
+The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables). See example figure XXX.
+
+\begin{theorem}
+\cite{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
+\end{theorem}
+
+Since we only care about the causal relationships between observed variables, it is sufficient to search over networks where any hidden variables have no parents and directly cause two of the observed variables.
+Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable.
+These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalization and conditioning.
+
+For any DAG with latent (and selection) variables there is a unique MAG \cite{Richardson2002}.
+This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \cite{Sprites}.
+The logic is very similar.
+Certain structures are ruled out by being inconstant with the observed conditional independence relations.
+The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG.
+Assuming there are no selection variables (see \cite xxx), the PAG can contain four types of link:
+
+\begin{enumerate}
+\item $X \rightarrow Y$, meaning $X$ causes $Y$
+\item $X \leftrightarrow Y$, meaning there is a latent variable that causes $X$ and $Y$.
+\item $X\ {\circ} {\rightarrow}\ Y$, either $X$ causes $Y$ or a latent variable causes both.
+\item $X\ {\circ} {-}  {\circ}\ Y$, either $X$ causes $Y$ or $Y$ causes $X$ or a latent variable causes both.
+\end{enumerate}
+
+The circles indicate where it is ambiguous if there should be an arrowhead (ie where there is one in some MAGs and not in others in the equivalence class).
+Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).
+
+\begin{figure}
+\centering
+\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
+\label{fig:FCIExamples}
+ \begin{subfigure}[t]{0.8\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){U}; 
+\node[main node, below left = of 1](2){B};
+\node[main node, below right = of 1](3){C};
+\node[main node, above left= of 2](4){A};
+\node[main node, above right = of 3](5){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    (5) edge node {} (3);
+
+\end{tikzpicture}
+ \end{subfigure}
+ \begin{subfigure}[t]{0.8\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+ %nodes
+
+\node[main node](1){A};
+\node[main node, below right=of 1](2){B};
+\node[main node, right=of 2](3){C};
+\node[main node, above right=of 3](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge[pil] node {} (2)
+    (2) edge node {} (3)
+    (3) edge node {} (2)
+    (4) edge [pil] node {} (3);    	
+\end{tikzpicture}
+\end{subfigure}
+ \begin{subfigure}[t]{0.4\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+ %nodes
+
+\node[main node](1){C};
+\node[main node, above left=of 1](2){A};
+\node[main node, above right=of 1](3){B};
+\node[main node, below=of 1](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (4)
+    (2) edge node {} (1)
+    (3) edge node {} (1);    	
+\end{tikzpicture}
+\end{subfigure}
+ \begin{subfigure}[t]{0.4\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+ %nodes
+
+\node[main node](1){C};
+\node[main node, above left=of 1](2){A};
+\node[main node, above right=of 1](3){B};
+\node[main node, below=of 1](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(1) edge node {} (4)
+    (2) edge [pil] node {} (1)
+    (3) edge [pil] node {} (1);    	
+\end{tikzpicture}
+\end{subfigure}
+
+\end{figure}
+
+The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \cite{Zhang2008}.
+More recently \cite{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
+\cite{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
+
+With latent variables we are not using all the information - so we could go further (to nested markov models and inequalities.) \cite{Richardson2012} \cite{Shipster2014}
+
+These are all constraint based methods ... efficient because they stop early, but also may not be robust to errors early on.
+
+\subsubsection{Doing conditional independence tests}
+The off-diagonal elements of the standardized inverse of the correlation
+matrix are the negatives of the partial correlation coefficients between the corresponding
+variables given the remaining variables (see e.g. Whittaker, 1990). Hence in the linear case, the
+independence graph can be efficiently constructed by placing an edge between A and B if and
+only if the entry in the standardized inverse correlation matrix is non-zero. In the discrete case,
+Fung and Crawford (1990) have recently proposed a fast algorithm for constructing an
+independence graph from discrete data. We have not tested their procedure as a preprocessor for
+the PC algorithm. (COPIED FROM SPRITES)
+
+\cite{Zhang2012} Kernel independence tests
+
+HSIC \cite{Gretton2008}
+
+\subsection{Discovery with functional models}
+All of the algorithms we have considered so far return a Markov equivalence class.
+They cannot distinguish between two models that result in the same set of conditional independence relations.
+Consider the very simple case where we have only two variables and the only possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
+These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
+No algorithm relying purely on conditional independence relations can separate these two cases.
+
+Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$.
+What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from.
+Recall the functional definition of causality (section \ref{sec:SEM}).
+There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \ref{}, post-non-linear additive noise
+
+Linear models with non-gaussian noise \cite{Hoyer2012}
+
+Allow us to make assumptions about the form of the relationships between variables.
+With just two variables:
+
+Additive noise, \cite{Hoyer2009}
+
+post-non-linear additive noise \cite{Zhang2008a},
+
+causal anti-causal,
+
+deterministic functions,
+
+IGCI \cite{Janzing2012a}
+
+learning what causality looks like.
+
+If we have a method to solve this case we can solve the more general case (with no hidden variables) Putting things together to learn the full network.
+
+\subsection{Scoring}
 
@@
 
+\section{Concluding remarks}
+
+Section 3 tells empirical researchers how to formalize what they already do when forced to work with observational data.
+The results should not be taken to mean that experimentation should be abandoned.
+Similarly the fact that causal discovery from general assumptions is possible (section 5) does not mean we should throw away careful analysis of a specific problem in favour of simply throwing all the variables into a causal discovery algorithm.
+A human may likely still be the state of the art for constructing a qualitative network (or better set of plausible networks) in fields like economics, ecology and social science where the variables are subject to human intuition and there exists some some theory to constrain models.
+However, we should not assume this will always be the case.
+As causal discovery methods are improved and the ability of AI to automatically incorporate large bodies of theory or prior knowledge into models develops computers should become better at science than us.
+Even if you don't believe in such dreams, there are already problems with large number of variables where we don't have a lot of theoretical understanding of the relationships between them.
+For these problems, even very naive causal discovery algorithms can improve on the alternative of standard feature selection.
+
 \bibliographystyle{plain}% Select the citation style e.g. ieeetr
@@
 \end{document}
+
+\begin{algorithm}
+  \caption{The SGS (or IC) Algorithm}\label{alg:IC}
+  \label{alg:PC}
+  \begin{algorithmic}[1]
+    \For{all pairs of variables $(a,b)$}
+    	\State  {all other sets of variables excluding a,b}
+
+    \EndFor
+    \If{$n < 0$}
+      \State $X \leftarrow 1 / x$
+    \EndIf
+    \If{$n > 0$}
+      \State $X \leftarrow 1$
+    \EndIf
+    \For{$N \neq 0$}
+      \State $X \leftarrow X \times X$
+    \EndFor
+    \While{$N \neq 0$}
+      \State $X \leftarrow X \times X$
+    \EndWhile
+  \end{algorithmic}
+\end{algorithm}

### 2014-10-24 reinforcement.tex
gets basic thompson sampling working in libcause

--- 2014-10-07 causality_tutorial.tex
@@
 \author{Finnian Lattimore}
-\title{A short introduction to causal inference}
+\title{Notes on Re-inforcment learning}
 
@@
 \def\ci{\perp\!\!\!\perp} % from Wikipedia
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{definition}{Definition}[section]
 
-\section{Why is causality important}
-
-TODO
-
-(this section should include explanation of why correlation is not causation)
-
-There are two ways in which correlation is not causation. Firstly correlation may not be statistically significant, - ie not even a genuine association. secondly it can be significant but spurious.
-
-\section{Models of causality}
-\label{sec:causalmodels}
-To understand causality we first need a more formal definition of what we are talking about.
-In this section we will look at three slightly different definitions of causality and use them to describe the following very simple example.
-The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
-
-We have developed a new drug for some illness and wish to determine how effective it is.
-We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group.
-The people in the treatment group get the drug, everyone else gets a placebo pill.
-The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo.
-We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome .
-\subsection{Structural Equation Models}
-
-Structural equation models (SEMs) describe a deterministic world, where underlying mechanisms determine the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalized, generalized to the non-parametric setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-For our example the SEM is:
-
-\begin{equation}
-\label{eq:SEM}
-\begin {aligned}
-X = & \epsilon_{x} \\
-Y = & f(X,\epsilon_{y})
-\end {aligned}
-\qquad
-\raisebox{-5mm}
-{
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
-
- %nodes
-\node[main node](1){$X$};
-\node[main node, right=of 1](2){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2);
-
-\end{tikzpicture}
-}
-\end{equation}
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random; the outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-
-The probability of observing $Y$ given $X$, $P(Y = y|X = x) = P(f(x,\epsilon_{y})=y)$ where $\epsilon_{y} ~ P(\epsilon_{y})$ as $\epsilon_{y} \ci \epsilon{x}$
-
-We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
-Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$.
-The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}.
-However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$.
-The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomized experiment.
-In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.
-
-For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-
-\subsection{Causal Bayesian Networks}
-Causal Bayesian Networks are, unsurprisingly, an extension of Bayesian networks.
-Any joint probability distribution can be factorized into a product of conditional probabilities.
-There are multiple valid factorizations, corresponding to permutations of variable ordering.
-
-\begin{equation}
-P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
-\end{equation}
-
-We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}).
-If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.
-
-\begin{figure}[h]
-\centering
-\caption{General Bayesian network over 3 variables}
-\label{fig:bayesnet}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below left=of 1](2){$X_{2}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{figure}
-
-The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorized over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges).
-If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
-
-\begin{equation}
-P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
-\end{equation}
-
-A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
-This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-More generally, if $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network.
-This is referred to as the truncated product formula \cite{Pearl2000}.
-
-\begin{equation}
-\label{eq:truncatedproduct}
- P_{do(Z=z)}(X_{1}...X_{N}) =
-  \begin{cases}
-  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
-   0       & \text{otherwise } 
-  \end{cases}
-\end{equation}
-
-Causal Bayesian networks encode information on all interventional distributions over a set of variables.
-A causal network represents (much) more information than a Bayesian network with identical structure due to the assumption that links can be interpreted causally.
-Causal Bayesian networks are Bayesian networks so results that apply to Bayesian networks carry directly across;the local markov property states that variables are independent of their non-effects given their direct causes.
-Similarly the global markov property and d-separation also hold in causal networks.
-There is an algorithm \cite{Shpitser2012} based on these properties that, for a given network and interventional (do-type) query, can:
-
-a) determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network
-
-b) if it is identiable, return the required expression
-
-Returning at last to our simple example, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$?
-If we assume that $X$ causes $Y$ and there are no common causes then the causal Bayesian network is just $X \rightarrow Y$, the same structure as the network for the SEM specified by equation (\ref{eq:SEM}).
-This network represents the (causal) factorization  $P(X,Y) = P(X)P(Y|X)$, so from equation (\ref{eq:truncatedproduct}), $P(Y|do(X)) = P(Y|X)$.
-In this example, the interventional distribution is equivelent to the observational one as we found earlier using the structural equation model.
-
-\subsection{Counterfactuals and the Neyman-Rubin framework}
-
-The Neyman-Rubin model \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}.
-Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else.
-
-In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{i}^{1}$ versus what would happen if they got the placebo, $y_{i}^{0}$.
-The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment.
-This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
-
-Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.
-Let $Y^{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $Y^{1}$ is the distribution we would see of $Y$ if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $P(Y^{1})-P(Y^{0})$.
-We can estimate, from an experimental or observational study, the probability that people recover if treated $P(Y|X=1)$ and the probability that they recover if not treated $P(Y|X=0)$.
-Now if $X=0$ then $Y = Y^{0}$.
-Equivalently stated:
-
-\begin{equation}
-\begin{aligned}
-P(Y^{0}|X=0)&= P(Y|X=0)\\
-P(Y^{1}|X=1)&=P(Y|X=1)
-\end{aligned}
-\end{equation}
-
-If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
-
-\begin{equation}
-\begin{aligned}
-P(Y^{1}) &= P(Y^{1}|X=1) = P(Y|X=1) \\
-P(Y^{0}) &= P(Y^{0}|X=0) = P(Y|X=0) 
-\end{aligned}
-\end{equation}
-
-\begin{equation}
-\implies P(Y^{1})-P(Y^{0}) = P(Y|X=1) - P(Y|X=0)
-\end{equation}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{Rosenbaum1983}.
-They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-Again this is justified in our example due to the randomization of treatment assignment.
-In general these assumption do not hold.
-If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model.
-that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.
-
-\subsection{Tying it all together. How do these models relate?}
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed are definitionally (almost) equivalent.
-
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorization of the joint distribution over its variables.
-In other words, the network is a causal Bayesian network.
-All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
-Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorization and the truncated product formula (equation \ref{eq:truncatedproduct}).
-Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
-We can also connect this to counterfactuals via:
-
-\begin{equation}
-\begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
-\end{aligned}
-\end{equation}
-
-The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
-
-\begin{equation}
-\label{eq:weakignore}
-X \ci Y^{0} \text{ and } X \ci Y^{1}
-\end{equation}
-
-It is possible to relax the independence of errors assumption we made for SEMs slightly to correspond exactly the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \cite{Richardson2013}.
-To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of functional models or bayesian networks or as conditional independence statements involving counterfactual variables (ignorability assumptions).
-By non-parametrically, I mean that we are not making any assumptions about the form of the relationships between variables.
-Models that are not non-parametrically identifiable can still be identified given assumptions about the distributions of variables and the functional relationship between them, for example, that the functions are linear or that the noise is additive \cite{Peters2014}.
-This form of assumption fits extremely naturally into the structural equation framework.
-
-We can also pose causal queries that are not interventional and cannot be phrased in terms of the do-notation.
-Our patients can be broken down into four groups.
-The first group will recover whether or not they receive treatment, the second group will recover if treated but not on the placebo, the third group will recover on the placebo and not if treated, and the last group will not recover on treatment or placebo.
-Unfortunately, we don't know which group each person belongs to.
-Drawing this up as a table:
-
-\begin{tabular}{c|c|c|c}
-group & placebo & treatment & probability of group\\
-\hline
-1 & die & die & $\alpha=P(Y^{0}=0,Y^{1}=0)$\\
-2 & die & recover & $\beta=P(Y^{0}=0,Y^{1}=1)$\\
-3 & recover & die & $\gamma=P(Y^{0}=1,Y^{1}=0)$\\
-4 & recover & recover & $\delta=P(Y^{0}=1,Y^{1}=1)$\\
-\end{tabular}
-
-The queries we have been asking thus far are about $P(Y^{0}=1) = \gamma + \delta$ and $P(Y^{1}=1) = \beta + \delta$, but suppose we asked the question; what is the probability that this patient, who was not treated and died, would have recovered if they had been treated?
-We know they are in either group 1 or 2 since they died without treatment, so the answer is $\frac{\beta}{\alpha+\beta}$.
-Can we estimate the $\alpha, \beta, \gamma, \delta$ or in other words, identify the joint distribution over the counterfactuals $P(Y^{0},Y^{1})$ given the interventional distributions, $P(Y^{0})$ and $P(Y^{1})$?
-The answer is no, putting our constraints and unknowns in matrix form:D-separation still applies in the augmented model.
-
-\begin{equation}
-\left(
-\begin{array}{cccc}
-0&0&1&1\\
-1&1&0&0\\
-0&1&0&1\\
-1&0&1&0\\
-1&1&1&1\\
-\end{array}
-\right)
-\left(
-\begin{array}{c}
-\alpha\\
-\beta\\
-\gamma\\
-\delta\\
-\end{array}
-\right)= 
-\left(
-\begin{array}{c}
-P(Y^{0})\\
-1-P(Y^{0})\\
-P(Y^{1})\\
-1-P(Y^{1})\\
-1\\
-\end{array}
-\right)
-\implies
-\left(
-\begin{array}{c}
-\alpha\\
-\beta\\
-\gamma\\
-\delta\\
-\end{array}
-\right)=
-\left(
-\begin{array}{c}
-1-P(Y^{0})-P(Y^{1})+\delta\\
-P(Y^{1})-\delta\\
-P(Y^{0})-\delta\\
-\delta\\
-\end{array}
-\right)
-\end{equation}
-The value of $\delta$ is not determined so the query is not identifiable.
-However we do get bounds on the terms.
-Since probabilities cannot be negative, $P(Y{1})-P(Y^{0})-1 \leq \delta \leq min(P(Y^{1}),P(Y^{0})$.
-Note, if we made the additional assumption $\gamma=0$; that the drug did not cause anyone to die who would otherwise have survived, then we can determine the joint distribution over counterfactuals.
-Alternatively, if we could assume that after treatment people returned to their initial state after some period of time, (say we were testing a drug for acne) then we could run a crossover study to determine the joint distribution.
-In a crossover study, the participants are randomly assigned to treatment and placebo, results are measured and then the groups are swapped.
-The scientific and philosophical validity of counterfactual queries remains under question \cite{Dawid2000,Dawid2014}, however they are nonetheless widely posed in the form of attribution of causal effects to different pathways and mediation \cite{Pearl2014,Imai2010a,VanderWeele2011}.
-
-There are differences between the models we have considered when it comes to counterfactual queries.
-Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.
-Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
-Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
-
-In practice, differences in focus and approach across different fields eclipse these actual differences in the models.
-The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field.
-Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
-SAY SOMETHING ABOUT NON-RECURSIVE (IE CYCLIC MODELS) HERE.
-
-The equivalence between models for interventional queries allows researchers to combine the best aspects of all of these models.
-You can define your assumptions and determine which variables must be controlled for using a graphical model and apply propensity scores to make the adjustment.
-\cite{Morgan2007} provides an excellent and pragmatic introduction to this combined approach.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-
-\section{The Do Calculus}
-If you state your assumptions in the form of an acyclic structural equation model or a causal bayesian network, the do calculus provides the methodology to determine if a given causal effect can be estimated from observational data and, if it can, provides to formula to do so.
-The calculus relies heavily on the understanding how conditional independence properties are encoded in graphical models.
-Therefore, in the next section we give a short review of independence in graphical models.
-
-\subsection{Independence in graphical models: D-separation}
-TODO
-Local Markov property implies global markov property (in other words d-separation).
-
-\subsection{The three rules}
-The do calculus consists of three rules.
-They derive from the causal information encoded in a causal network and the properties of d-separation and do not imply any addition assumptions other than that of specifying the causal network.
-The key property of these three rules is that they are complete.
-If an interventional, do-query, can be transformed to a do-less statement via repeated application of these rules we can get an unbiased estimate of the causal effect from observational data.
-Conversely, if the query is not identified via these rules we cannot get an unbiased estimate of the causal effect without additional assumptions, even with an infinite amount of observational data.
-
-\subsubsection{Rule 1}
-This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention.
-The intervention $do(\boldsymbol{Z}=\boldsymbol{z})$ changes a causal network, $G$, in a simple way.
-Variables in $\boldsymbol{Z}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{z}$.
-This corresponds to deleting the edges with arrows into variables in $\boldsymbol{Z}$  see example Figure \ref{fig:mutilatednet}.
-The resulting 'mutilated' network $G_{\overline{\boldsymbol{Z}}}$ remains a causal network and d-separation still applies.
-
-\begin{figure}[]
-\centering
-\caption{Visualizing intervention in a causal bayesian network. Left: original network, $G$. Right: network after the intervention $do(Z=z)$, $G_{\overline{Z}}$}
-\label{fig:mutilatednet}
-\begin{tabular}{cc}
-
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$Z$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2);
-
-\end{tikzpicture} &
-
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$z$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{tabular}
-
-\end{figure}
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{Z}}}$:
-\begin{equation}
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x}) 
-\end{equation}
-
-$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z})$ in $G_{\overline{\boldsymbol{Z}}}$:
-\begin{equation}
-\label{eq:Do12}
- P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z})) 
-\end{equation}
-
-This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.
-
-\subsubsection{Rule 2}
-Rule 2 states when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening  $do(\boldsymbol{X}=\boldsymbol{x})$ have the same effect on the distribution of the outcome $\boldsymbol{Y}$.
-You can think of this as when correlation is causation.
-It is easiest to understand by explicitly including the intervention process in the graphical model.
-We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on.
-Let $\epsilon$ be some arbitrary value not in the set of possible values for $X$.
-If $X = \epsilon$ the distribution of $X$ is what it was without intervention.
-Otherwise, if $X=x$, $X$ deterministically takes the value $x$ and is independent of its previous parents, representing the intervention $do(X=x)$.
-We use the notation $G^{\dagger}$ to represent $G$ augmented with these decision nodes.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
-
-\begin{equation}
-\label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w})
-\end{equation}
-
-$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X})$ in $G^{\dagger}$:
-\begin{equation}
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x})
-\end{equation}
-
-If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.
-
-\subsubsection{Rule 3}
-This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
-\begin{equation}
-\label{eq:Do3}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-\end{equation}
-$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}})$ in $G^{\dagger}$:
-\begin{equation}
-\label{eq:Do32}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y})
-\end{equation}
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ without conditioning on $\boldsymbol{X}$ implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ which starts on an arrow leaving $\boldsymbol{X}$.
-This means there is no causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$ and the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ does not alter the distribution of $P(\boldsymbol{Y})$.
-
-\subsection{Easy graphical tests for identifiability}
-Determining if a network is identifiable via the two calculus still requires some algebra and it may not always be obvious what sequence of rules are required for a given network.
-From the do-calculus we can derive some quick graphical tests that are sufficient (but not necessary) for non-parametric identifiability, and align with intuitive ideas about estimating causal effect by adjusting for confounding variables.
-
-\subsubsection{If all variables are observed}
-If all the variables in a causal network can be measured any interventional query on that network are identifiable.
-You just have to condition on the parents of the variables on which you intervene.
-For discrete variables, the causal effect is given by:
-
-\begin{equation}
-P(\boldsymbol{Y}|do(\boldsymbol{X})) = \sum_{\boldsymbol{Pa_{X}}}{P(\boldsymbol{Y}|\boldsymbol{X},\boldsymbol{Pa_{X}})P(\boldsymbol{Pa_{X}})}
-\end{equation}
-
-where $\boldsymbol{Pa_{X}}$ are the parents of $\boldsymbol{X}$.
-
-Conditioning on the parents blocks all information that does not flow causally (along the arrows) from the variables on which we intervened which eliminates spurious correlations and gives the causal effect.
-
-\subsubsection{The Back Door criterion}
-The back door criterion generalizes from the case where all variables can be measured by observing we don't need to condition on all the parents of the intervened on variables: it is sufficient to condition on a set of variables that block all spurious paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$.
-A spurious, or back door, path is one that leaves $\boldsymbol{X}$ against the direction of the causal arrows.
-If we have a set of variables $\boldsymbol{Z}$ that block all back door paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$, where no variable in $\boldsymbol{Z}$ is a descendent of $\boldsymbol{X}$ then:
-
-\begin{equation}
-P(\boldsymbol{Y}|do(\boldsymbol{X})) = \sum_{Z}{P(\boldsymbol{Y}|\boldsymbol{X},\boldsymbol{Z})P(\boldsymbol{Z})}
-\end{equation}
-
-The requirement that $\boldsymbol{Z}$ should not include descendants of $\boldsymbol{X}$ stems from the fact that we don't want to open any new spurious paths.
-If $Z$ descends from $X$ then it may lie on a path of the form $X \rightarrow...\rightarrow Z \leftarrow...
-Y$, which conditioning on $Z$ would open.
-The back door criterion corresponds exactly to the strong ignoreability assumption \cite{Rosenbaum1983}, used to justify the validity of a causal estimate in the counterfactual approach .
-If $\boldsymbol{Z}$ satisfies the backdoor criterion then, for a single binary treatment variable $X$ and outcome variable $Y$, $\{Y^{0},Y^{1}\} \ci X | \boldsymbol{Z}$.
-
-\subsubsection{The Front Door Criterion}
-It we have an unobservable variable creating a spurious path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ then we cannot apply the back door criterion to directly estimate $P(\boldsymbol{Y}|do(\boldsymbol{X}))$.
-However, if there is a set of mediating variables $\boldsymbol{M}$ that intercepts all all the direct (non-spurious) paths from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and we can estimate the causal effect of $\boldsymbol{X}$ on $\boldsymbol{M}$ and the causal effect of $\boldsymbol{M}$ on $\boldsymbol{Y}$ then we can combine these terms to get the causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$.
-
-\begin{equation}
-\begin {aligned}
-P(\boldsymbol{Y}|do(\boldsymbol{X})) & = \sum_{\boldsymbol{M}}{P(\boldsymbol{M}|do(\boldsymbol{X}))P(\boldsymbol{Y}|do(\boldsymbol{M}))} \\
-&= \sum_{\boldsymbol{M}}{P(\boldsymbol{M}|\boldsymbol{X})} \sum_{\boldsymbol{X}}P(\boldsymbol{Y}|\boldsymbol{X},\boldsymbol{M})P(\boldsymbol{X}) 
-\end {aligned}
-\end{equation}
-
-\subsubsection{Examples}
-
-We are now in a position to consider some more interesting examples than the one in section \ref{sec:causalmodels}.
-Consider the question; does eating chocolate increase or decrease the risk of obesity?
-Assume we have data from a cross-sectional study where a large group of people were weighed and asked how much chocolate they ate.
-It is easy to come up with a whole lot of variables that may causally effect chocolate consumption and obesity, ie.
-age, income and gender, so we cannot conclude that the causal effect is equivalent to the correlation.
-If we assume there may be unknown variables (denoted by $U$ in all the following examples) that effect both obesity and chocolate consumption, (figure \ref{fig:chockNI}), the causal effect cannot be identified as there is no way to block the spurious path through the unobserved variables.
-If we assume the only confounding variables are income, age and gender, (figure \ref{fig:chockI}), we can calculate the causal effect by conditioning on these variables.
-
-\begin{figure}[h]
-\centering
-\caption{Chocolate and Obesity}
-\label{fig:chocolate}
-\begin{subfigure}[t]{0.4\textwidth}
-\caption{Not identifiable}
-\label{fig:chockNI}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.5cm,  thick,main node/.style={punktrect}]
-
- %nodes
-\node[main node](1){$\boldsymbol{U}$};
-\node[main node, below left=of 1](2){Chocolate};
-\node[main node, below right=of 1](3){Obesity};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-\caption{Identifiable}
-\label{fig:chockI}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punktrect}]
-
- %nodes
-\node[main node](1){Age};
-\node[main node, left= of 1](4){Income};
-\node[main node, right=of 1](5){Gender};
-\node[main node, below left=of 1](2){Chocolate};
-\node[main node, below right=of 1](3){Obesity};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2)
-        edge node {} (3)
-    (5) edge node {} (2)
-    	edge node {} (3);
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-In reality of course there are many other variables involved  \cite{Golomb2012} and it may not be obvious from prior knowledge or theory what role the variables play in the network.
-This is important as conditioning on the wrong variables can increase the bias in a causal estimate: Conditioning on a collider variable, $\rightarrow Z \leftarrow$, may open a previously blocked, spurious path \cite{Sjolander2009a} (figure \ref{fig:colliderCondition}) .
-Conditioning on a variable that mediates between the cause and effect biases the causal estimate by blocking a true casual path (figure \ref{fig:mediationBias}).
-Conditioning on a variable that is a true confounder, that is a common cause of both $X$ and $Y$, may increase the bias in our causal estimate if it moves the causal estimate in the opposite direction than would correcting for a hidden (and stronger) confounder $U$.
-Finally, (although it is not obvious graphically) conditioning on a variable that is a cause of $X$ but not (or only weakly) a cause of $Y$ can amplify or introduce bias \cite{Pearl2012b,Wooldridge2009}.
-
-\begin{figure}[h]
-		\caption{Conditioning on a collider variable can introduce bias}\label{fig:colliderCondition}
-       \centering
-        \begin{subfigure}[t]{0.4\textwidth}
-        \centering
-                \caption{post-treatment bias: conditioning on $Z$ opens the spurious path $X,Z,Y$}
-                \label{fig:postTreatmentBias}
-                \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
- %nodes
-\node[](1){}; 
-\node[main node, left = of 1](2){X};
-\node[main node, right= of 1](3){Y};
-\node[main node, below =of 1](4){Z};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (3)
-    	edge node {} (4)
-    (3) edge node {} (4);
-\end{tikzpicture}
-        \end{subfigure}%
-        ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
-          %(or a blank line to force the subfigure onto a new line)
-        \begin{subfigure}[t]{0.4\textwidth}
-        \centering
-                \caption{M-bias: conditioning on $Z$ opens the path $X,U_{1},Z,U_{2},Y$}
-                \label{fig:Mbias}
-                \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){Z}; 
-\node[main node, below left = of 1](2){$X$};
-\node[main node, below right= of 1](3){$Y$};
-\node[main node, above left = of 1](5){$U_{1}$};
-\node[main node, above right = of 1](6){$U_{2}$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (3)
-    (5) edge node {} (1)
-    	edge node {} (2)
-    (6) edge node {} (1)
-    	edge node {} (3)
-    (1) edge node {} (2)
-    	edge node {} (3);
-\end{tikzpicture}
-        \end{subfigure}    
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\caption{Conditioning on a mediating variable ($Z$) can introduce bias}
-\label{fig:mediationBias}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){$X$}; 
-\node[main node, above right = of 1](2){$Z$};
-\node[main node, below right= of 1](3){$U$};
-\node[main node, above right = of 3](4){$Y$};
-\path[every node/.style={font=\sffamily\small}]
-	(1) edge node {} (2)
-		edge node {} (3)
-	(2) edge node {} (4)
-	(3) edge node {} (4);
-
-\end{tikzpicture}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\caption{Conditioning on a confounder ($Z$) can increase bias}
-\label{fig:biasAmplification}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){Z}; 
-\node[main node, below left = of 1](2){X};
-\node[main node, right = of 1](4){U};
-\node[main node, below right= of 4](3){Y};
-
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (3)
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    	edge node {} (3);
-
-\end{tikzpicture}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\caption{Bias amplification}
-\label{fig:biasAmplification}
-\begin{subfigure}[t]{0.4\textwidth}
-\centering
-\caption{Conditioning on an instrumental variable $Z$ can increase bias in estimate of effect of $X$ on $Y$}
-\label{fig:instrumentalBias}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){U}; 
-\node[main node, below left = of 1](2){X};
-\node[main node, below right = of 1](3){Y};
-\node[main node, above left= of 2](4){Z};
-
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (3)
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2);
-
-\end{tikzpicture}
-\end{subfigure}
- \begin{subfigure}[t]{0.4\textwidth}
-\centering
-\caption{Conditioning on a confounding variable that is strongly linked to $X$ and weakly linked to $Y$ can increase bias.}
-\label{fig:semiinstrumentalbias}
-
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){U}; 
-\node[main node, below left = of 1](2){X};
-\node[main node, below right = of 1](3){Y};
-\node[main node, above left= of 2](4){Z};
-
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (3)
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    	edge [dashed] node {} (3);
-
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-For some problems, the temporal ordering of events imposes restrictions on the causal structure.
-Consider the example of a non-randomized trial.
-A new treatment is developed and made available to a large group of patients some of whom decide to take.
-Researchers have variables on the health, income, gender, age, etc of the patients before the treatment was offered (pre-treatment variables) and collect further data on the health of all patients after treatment has begun (post-treatment variables).
-It has been understood for a long time that conditioning on post-treatment variables can increase bias \cite{Rosenbaum1984a}.
-The realization that that conditioning on pre-treatement can also introduce bias is more recent.
-Not conditioning on post-treatment variables avoids the biases in figure \ref{fig:postTreatmentBias} and figure \ref{fig:mediationBias}.
-However, M-bias (figure \ref{fig:Mbias}) and bias amplification (figure \ref{fig:biasAmplification}) are consequences of conditioning on pre-treatment variables.
-
-It is clear we need to be wary of the condition on everything approach and the condition on all pre-treatment variables approach.
-But what is the alternative if the structure of the causal network is not clear from prior knowledge?
-One approach would be to construct a number of different plausible structures and evaluate the desired causal effect for each of them.
-This would give more accurate uncertainty bounds for the causal effect.
-Alternatively, one could apply a causal discovery algorithm (see section \ref{sec:causaldiscovery}).
-
-It may be that there are some general conditions under which certain types of biases are less likely or weaker than other.
-For example, in the situation in figure \ref{fig:Mbias}, where $Z$ is both a confounding variable and part of an M-structure.
-Conditioning on $Z$ blocks the spurious path $X \leftarrow Z \rightarrow Y$ and opens the spurious path $X \leftarrow U_{1} \rightarrow Z \leftarrow U_{2} \rightarrow Y$.
-Should we condition on $Z$ or not?
-Under what general assumptions is one of these paths weaker than the other.
-There has been relatively little work in this area.
-\cite{Vansteelandt2012} gives a description of the problem and links it to selecting covariates when the goal is causal estimation.
-\cite{Myers2011} use simulation studies to consider if one should adjust for a variable that could be either an instrumental variable or a confounder for the effect of $X$ on $Y$.
-
-\section{It's not (non-parametrically) identifiable! What can I do?}
-\subsection{Bounding causal effects}
-Causal effects that cannot be identified can often still be bounded. In other words, bounds require weaker assumptions than point identification.
-
-\subsection{Additional assumptions}
-additive noise
-causal anti-causal \cite{Janzing2012}
-
-\subsection{Instrumental variables}
-
-\section{Causal structure learning (causal discovery)}
-\label{sec:causaldiscovery}
-peters approach \cite{Peters2014}
-Pearl/Verma/Sprites approach \cite{Pearl2000}
-
-\section{Causality and prediction}
-Refresh the difference between a causal query and predictive query.
-Point out the importance of accurate prediction for causal inference. For example, if we can predict the control response, we can run experimental studies without controls (\cite{Brodersen2013})
+In both classical experimental design and in reinforcement learning the number of possible configurations explodes as we add variables.
+Key strategy of bandits is 'Randomized probability matching'; play the arm according to the probability its optimal
 

### 2014-10-24 causalreinforcement.py

@@ -84,7 +84,7 @@ for i in range(simulations):
 # 
 
 #hmmm this doesn't get to do final optimization as ultimately can't narrow down enough - It should be setting each arm ... eventually. Ie it should learn ...
-interventions = [{"X1":'0'},{"X1":'1'},{"X2":'0'},{"X2":'1'},{"X3":'0'},{"X3":'1'}] # number of arms the same in this case as 2*2 = 2^2
+interventions = [{"X1":'0'},{"X1":'1'},{"X2":'0'},{"X2":'1'},{"X3":'0'},{"X3":'1'}] 
 X1arms = {0:0,1:1}
 X2arms = {0:2,1:3}
 X3arms = {0:4,1:5}

### 2014-10-24 causalreinforcement2.py

@@ -1,0 +1,32 @@
+from libcause import BinaryNetworkBandit
+from numpy import *
+from scipy.stats import *
+
+
+# A second attempt at causal reinforcement learning
+# restricted to the simpler case where we can only intervene on one variable at a time
+
+interventions = [{"X1":'0'},{"X1":'1'},{"X2":'0'},{"X2":'1'},{"X3":'0'},{"X3":'1'}] 
+bandit = BinaryNetworkBandit("bayesnet.json",interventions,'Y')
+
+simulations = 1
+samples = 12
+best_arm = 0.4
+o = open("causalreinforcementresults2.txt","w")
+
+for i in xrange(simulations):
+    bandit.sample(samples)
+    regret = bandit.regret(best_arm)
+    o.write("Thompson,"+str(regret)+"\n")
+    bandit.reset()
+
+    bandit.sample_with_network_info(10,["X1","X2","X3"])
+    regret = bandit.regret(best_arm)
+    o.write("CThompson,"+str(regret)+"\n") 
+    bandit.reset()
+    
+o.close()
+
+
+
+print "DONE"

### 2014-10-24 libcause.py

@@ -1,0 +1,174 @@
+from numpy import *
+from scipy.stats import *
+from libpgm.nodedata import NodeData
+from libpgm.graphskeleton import GraphSkeleton
+from libpgm.discretebayesiannetwork import DiscreteBayesianNetwork
+from libpgm.tablecpdfactorization import TableCPDFactorization
+import itertools
+import matplotlib.pyplot as plt
+import time
+        
+
+class BinaryNetworkBandit(object):
+    """Class represents a Discrete Bayesian Network that supports Thompson sampling.
+        Currently only supports interventions on a single variable at a time"""
+    def __init__(self, bayesnetfile, targetVar = 'Y', prior=(1.0,1.0)):
+        self.file = bayesnetfile
+        self.theta_range = linspace(0.0001,0.9999,100)
+        self.set_bayesnet()
+        self.y = targetVar
+        self.reset(prior = prior)
+
+    def getCPD(self):
+        """ required as querying a TableCPDFactorization leads modifies the underlying Bayesian network (refresh() appears broken) """
+        self.set_bayesnet()
+        return TableCPDFactorization(self.bn)
+
+    def set_bayesnet(self):
+        nd = NodeData()
+        skel = GraphSkeleton()
+        nd.load(self.file)
+        skel.load(self.file)
+        skel.toporder()
+        self.bn = DiscreteBayesianNetwork(skel, nd)
+        
+    def reset(self,prior=(1.0,1.0)):
+        """ Clears all the data on samples that have been taken so far but keeps graph structure.
+            You can optionally specify a new prior"""  
+        # possible single interventions on non-target variable
+        self.prior = prior
+        self.interventions = [] # a list of possible assignments
+        self.variables = [] # store the variables - defines an ordering
+        values = []
+        for variable, data in self.bn.Vdata.iteritems():
+            if variable != self.y:
+                self.variables.append(variable)
+                vals = data.get("vals")
+                values.append(vals)
+                for value in vals:
+                    self.interventions.append({variable:value})
+
+        # lets calculate and print the actual value of theta for each arm (since we know it)
+        
+        truth = []
+        for i in self.interventions:
+             cpd = self.getCPD()
+             answer = cpd.specificquery({self.y:"1"},i)
+             truth.append(answer) 
+        print "THETA",truth
+        cpd = self.getCPD() # reset the network to its original state
+                    
+
[…]

### 2014-10-24 libcause.py
fixes plot labels

@@ -6,7 +6,9 @@ from libpgm.discretebayesiannetwork import DiscreteBayesianNetwork
 from libpgm.tablecpdfactorization import TableCPDFactorization
 import itertools
 import matplotlib.pyplot as plt
+import matplotlib
 import time
+import json
         
 
 class BinaryNetworkBandit(object):
@@ -120,14 +122,18 @@ class BinaryNetworkBandit(object):
         # put labels under each plot
         f,sp = plt.subplots(1,len(self.variables),sharey=False,figsize=(15,5))
         for i in range(len(self.variables)):
-            dist = beta(self.observed_true[i], self.observed_trials[i]-self.observed_true[i])
+            dist = beta(self.prior[0]+self.observed_true[i], self.prior[1]+self.observed_trials[i]-self.observed_true[i])
             sp[i].plot(self.theta_range,dist.pdf(self.theta_range))
         plt.show()
 
     def plot_assignments(self):
+        print self.atrials
+        print self.asuccesses
         f,sp = plt.subplots(1,len(self.assingment_map),sharey=False,figsize=(15,5))
-        for i in range(len(self.assingment_map)):
-            dist = beta(self.asuccesses[i], self.atrials[i]-self.asuccesses[i])
+        titles = [json.dumps(x) for x in self.assignements]
+        for i in range(len(self.assingment_map)): # need to get rid of the unicode tags so things fit - dirty way is s.encode('ascii')
+            dist = beta(self.prior[0]+self.asuccesses[i]+1, self.prior[1]+self.atrials[i]-self.asuccesses[i])
+            sp[i].set_title(titles[i])
             sp[i].plot(self.theta_range,dist.pdf(self.theta_range))
         plt.show()
         
@@ -165,10 +171,10 @@ class BinaryNetworkBandit(object):
         optimal = bestprob
         return 1 - reward/bestprob
 
-
+matplotlib.rcParams.update({'font.size': 8})
 bandit = BinaryNetworkBandit("bayesnet.json")
 bandit.sample(500,plot=500)
 print bandit.trials
 print bandit.successes
 bandit.plot_observed()
-bandit.plot_assignments()
+bandit.plot_assignments() 

### 2014-11-05 causal-bandit/c-bandit.tex
added tex files

\documentclass{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\title{Causal Inference and Bandits}
\author{Blah blah}

\begin{document}

\maketitle

\begin{abstract}
An abstract.
\end{abstract}

\section{Introduction}

Useful references are: \cite{BC12}.

\section{Notation}

We are considering a variant of the multi-armed bandit problem.
Let $p \in [0,1]^K$ be a fixed and known vector.
In each time-step $t$ 
\begin{enumerate}
\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
\eq{
\tilde X_{t,i} = \begin{cases}
X_{t,i} &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}
\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
\end{enumerate}
The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|X_i = j]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
}

\section{Estimating $\mu_{i,j}$}
The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
outperform this approach by exploiting the extra structure in the problem.

Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
We define estimator $\hat \mu_t$ of $\mu_{i,j}$ by
\eq{
n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
\hat \mu_t &= \eta_i \frac{m_{i,j}}{n_{i,j}}  + \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right] \\
\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \\
n_a        &= \begin{cases} 
  n_{i,j}                                           & \text{if } a = i \\
  \min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
\end{cases}
}
\begin{theorem} With probability at least $1 - \delta$ we have that:
$\displaystyle \left|\hat \mu_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
where $\beta > 0$ is some constant.
\end{theorem}

\section{Algorithm}

\begin{algorithm}
\caption{UCB}
\begin{algorithmic}[1]
\State {\bf Input:} Number of variables $K$, vector $p \in [0,1]^K$, horizon $n$
\For{$t \in 1,\ldots,n$}
\For{$i \in 1,\ldots,K$}
\For{$j \in \set{0,1}$}
\State Compute $\tilde \mu_{i,j} = \hat \mu_{i,j} + \sqrt{\frac{\alpha}{\sum_{a} n_a} \log n}$ 
\EndFor
\EndFor
\State Choose $I_t, J_t = \argmax_{i,j} \tilde \mu_{i,j}$
\EndFor
\end{algorithmic}
\end{algorithm}

\section{Theorems}
\section{Experiments}
\section{Conclusion}

\bibliographystyle{plainnat}
\bibliography{c-bandit}

\end{document}

### 2014-11-06 causal-bandit/c-bandit.tex
started to add proof
fixed some typos in proof
updates c-bandit to include experimental figures

@@
 \usepackage[capitalize]{cleveref}
+\usepackage{graphicx}
 
@@
 \newcommand{\Ber}{\operatorname{Bernoulli}}
+\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
 
@@
 
-\title{Causal Inference and Bandits}
+\title{Intervention Bandits}
 \author{Blah blah}
@@
 
-We are considering a variant of the multi-armed bandit problem.
-Let $p \in [0,1]^K$ be a fixed and known vector.
-In each time-step $t$ 
+Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$.
+We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
+
+Let $p \in [0,1]^K$ be a fixed and known vector. In each time-step $t$:
+
 \begin{enumerate}
@@
 \eq{
-n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
-m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
+m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
+n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
 \hat \mu_t &= \eta_i \frac{m_{i,j}}{n_{i,j}}  + \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right] \\
@@
   n_{i,j}                                           & \text{if } a = i \\
-  \min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
+  \frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
 \end{cases}
@@
 \end{theorem}
+
+\begin{proof}
+First note that $n_{a,b}$ is a random variable that is bounded by $t$ for all $a,b$.
+We use the short-hand $\mu^{a,b}_{i,j} = \E[q(X) | X_i = j, X_a = b]$. Then
+\eq{
+\mu_{i,j} = p_a \mu_{i,j}^{a,1} + (1 - p_a) \mu_{i,j}^{a,0}\,.
+}
+Now we can apply Hoeffding's bound and the union bound to show that
+\eq{
+\P{\left|\frac{m_{a,b}}{n_{a,b}} - \mu_{i,j}^{a,b}\right| \geq \sqrt{\frac{1}{2n_{a,b}} \log \frac{4t}{\delta}}} \leq \frac{\delta}{2}\,.
+}
+Therefore by the union bound
+\eq{
+\P{\left|p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}} - \mu_{i,j}\right| \geq p_a \sqrt{\frac{1}{2n_{a,1}} \log\frac{4t}{\delta}}
++ (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}} \leq \delta\,
+}
+Now by Jensen's inequality 
+\eq{
+p_a \sqrt{\frac{1}{2n_{a,1}}\log\frac{4t}{\delta}} + (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}
+&\leq \sqrt{\left(\frac{p_a}{2n_{a,1}} + \frac{1-p_a}{2n_{a,0}}\right)\log \frac{4t}{\delta}} \\
+&\leq \sqrt{\max\set{\frac{p_a}{n_{a,1}}, \frac{1-p_a}{n_{a,0}}} \log\frac{4t}{\delta}} \\
+&= \sqrt{{1 \over 2n_a} \log\frac{4t}{\delta}}\,.
+}
+Similarly,
+\eq{
+\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{4t}{\delta}}} \leq
+\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{2t}{\delta}}} \leq \delta\,.
+}
+\end{proof}
 
@@
 \section{Experiments}
+
+\begin{figure}[h]
+\caption{Testing values of beta - 100 experiments for each value of beta, with a 4 armed bandit.}
+\includegraphics[width=0.7\textwidth]{regretvsbeta.pdf}
+\end{figure}
+
+Simululations to compare the performance of standard UCB with our modified algorithm.
+For each number of arms, $100$ bandits of each type were created and run upto to a horizon of $1000$ timesteps.
+The mean regret and its standard error from these simulations is plotted in figure \ref{fig:regretvsarms} The true data was generated from a model where:
+
+\eq{
+&p = [0.5]^{K}\\
+&q(\boldsymbol{X}) = \begin{cases}0.5 & \text{if $X_{1}=0$}\\ 0.6 & \text{otherwise} \end{cases}
+}
+
+For small number of arms there is not much difference.
+As the number of arms increases, cusb shows significantly lower regret.
+As the number of arms increases above 100 the regret of the two algorithms begins to converge again.
+I suspect this is due to the fixed time horizon of $1000$ steps.
+With a time horizon of $100$ steps the two algorithms begin to converge again at a smaller number of arms.
+
+\begin{figure}[h]
+\caption{Shows relationship between the regret and the number of bandit arms $2K$ for the standard UCB algorithm and our modified algorithm.  }
+\label{fig:regretvsarms}
+\includegraphics[width=0.9\textwidth]{regretvsarms.pdf}
+\end{figure}
+
 \section{Conclusion}

### 2014-11-14 causal-bandit/c-bandit.tex
updates notation and adds latest plots to c-bandit.aux

@@
 \begin{document}
-
+\def\ci{\perp\!\!\!\perp}
 \maketitle
@@
 \end{enumerate}
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|X_i = j]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
+The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
 where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
@@
 lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
-outperform this approach by exploiting the extra structure in the problem.
+outperform this approach by exploiting the known causal structure of the problem.
+
+\eq {
+P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
+ &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
+ &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
+ &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
+}
 
 Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
-We define estimator $\hat \mu_t$ of $\mu_{i,j}$ by
-\eq{
-m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
-n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
-\hat \mu_t &= \eta_i \frac{m_{i,j}}{n_{i,j}}  + \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right] \\
-\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \\
-n_a        &= \begin{cases} 
+
+Let $\hat{\mu_a}$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $a$.
+
+\eq {
+\hat{\mu_a} &=
+\begin{cases}
+\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
+\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
+\end{cases}\\
+\text{where:}\\
+m_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
+n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s \\
+}
+
+This gives $K$ estimators $\{\hat{\mu_1}...\hat{\mu_K}\}$ to be pooled into a single estimator $\hat{\mu}$.
+
+\eq {
+\hat{\mu} &= \sum^{K}_{a=1}\eta_a\hat{\mu}_a=\eta_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
+\text{where:}\\
+\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
 n_{i,j} & \text{if } a = i \\
@@
 }
+
+If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p_a}$
+
+\eq{
+\hat{p_a} = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
+}
+
 \begin{theorem} With probability at least $1 - \delta$ we have that:
-$\displaystyle \left|\hat \mu_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
+$\displaystyle \left|\hat{\mu}_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
 where $\beta > 0$ is some constant.
@@
 \begin{figure}[h]
-\caption{Testing values of beta - 100 experiments for each value of beta, with a 4 armed bandit.}
-\includegraphics[width=0.7\textwidth]{regretvsbeta.pdf}
+\caption{Comparison of the performance of standard UCB versus causal UCB with $\beta=3$ and $\beta = 5$.
+100 simulations were run for each algorithm up to a horizon of $10^5$ per value of $K$.
+Error bars span the 1st to 3rd quantile of the regret, round points mark the median and triangular points show the mean.
+For standard UCB the regret increases linearly with the number of arms $K$.
+For causal UCB the increase is sub-linear.
+Increasing $\beta$ leads to slower convergence but lower variance.}
+\includegraphics[width=1\textwidth]{regret_vs_K_series_beta.pdf}
+\end{figure}
+
+\begin{figure}[h]
+\caption{The distribution of regret varies with the $\beta$ parameter in the bound in the estimator.
+As beta increases, the mean regret increases but the variance decreases.
+The plot shows the results of running $100$ independent bandits, with $K=256$ and $\epsilon=0.1$, up to a horizon $h=10^5$ for each value of $\beta$.
+}
+\includegraphics[width=0.7\textwidth]{regret_vs_beta_boxplot.pdf}
 \end{figure}
@@
 
-For small number of arms there is not much difference.
-As the number of arms increases, cusb shows significantly lower regret.
-As the number of arms increases above 100 the regret of the two algorithms begins to converge again.
-I suspect this is due to the fixed time horizon of $1000$ steps.
-With a time horizon of $100$ steps the two algorithms begin to converge again at a smaller number of arms.
+\section{Conclusion}
+
+\bibliographystyle{plainnat}
+\bibliography{c-bandit}
+
+\end{document}
+
+\begin{figure}[h]
+\caption{Testing values of beta - 100 experiments for each value of beta, with a 4 armed bandit.}
+\includegraphics[width=0.7\textwidth]{regretvsbeta.pdf}
+\end{figure}
 
@@
 
-\section{Conclusion}
-
-\bibliographystyle{plainnat}
-\bibliography{c-bandit}
-
-\end{document}
+\begin{figure}[h]
+\caption{Regret vs number of arms with horizon $100$}
+\label{fig:regretvsarmsshorthorizon}
+\includegraphics[width=0.8\textwidth]{regregvsamrs100.pdf}
+\end{figure}

### 2014-11-21 wiml_poster/FCI_figure.tex
adds first draft of wiml_poster

\documentclass{article}
\usepackage{tikz} 
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx

\usepackage{tikz}

\usepackage{graphicx}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}
\thispagestyle{empty}
\usepackage{multicol,caption}
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }

}

\begin{document}
\def\ci{\perp\!\!\!\perp} % from Wikipedia

\begin{figure}
\centering
\begin{subfigure}[t]{0.44\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]
\node[main node](1){C};
\node[main node, above left=of 1](2){A};
\node[main node, above right=of 1](3){B};
\node[main node, below=of 1](4){D};

 \path[]
 	(1) edge node {} (4)
    (2) edge [->]  (1)
    (3) edge [->]  (1);

\node[main node,right of=3,xshift=1.2em](5){A};
\node[main node, below right=of 5](6){C};
\node[main node, above right=of 6](7){B};
\node[main node, below=of 6](8){D};
\path[every node/.style={font=\sffamily\footnotesize}]
	(1) edge [-, >=stealth', blue!30!white, line width=.9em,shorten <=1em,shorten >= 1.7em,
	preaction={line width = .3em, blue!30!white,-triangle 90,draw,shorten >=.6em,shorten <=1em}
	] node[black]{FCI} (6)
	(5) edge [o->] (6)
	(7) edge [o->] (6)
	(6) edge [->] (8);

\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.55\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]

\node[main node](1){A};
\node[main node, below right=of 1](2){B};
\node[shaded, above right=of 2](3){U};
\node[main node, below right=of 3](4){C};
\node[main node, above right=of 4](5){D};
\node[ below=of 3,yshift=.2cm](11){};
\path[] (1) edge (2)
		(3) edge (2) edge (4)
		(5) edge (4);

\node[main node,right=of 5,xshift=-.5em](6){A};
\node[main node, below right=of 6](7){B};
\node[above right=of 7](8){};
\node[main node, below right=of 8](9){C};
\node[main node, above right=of 9](10){D};
\node[below = of 8,yshift=0.2cm](12){};	
\path[every node/.style={font=\sffamily\footnotesize}]
(11) edge [-, >=stealth', blue!30!white, line width=.9em,shorten <=3.3em,shorten >= 4.5em,
	preaction={line width = .3em, blue!30!white,-triangle 90,draw,shorten >=3.7em,shorten <=4em}
	] node[black,xshift=-.5em]{FCI} (12)
(6) edge [o->] (7)
(7) edge [<->] (9)
(10) edge [o->](9);

\end{tikzpicture}
\end{subfigure}
\end{figure}

\end{document}

### 2014-11-21 wiml_poster/conference_poster_5.tex
adds first draft of wiml_poster

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% a0poster Landscape Poster
% LaTeX Template
% Version 1.0 (22/06/13)
%
% The a0poster class was created by:
% Gerlinde Kettl and Matthias Weiser (tex@kettl.de)
% 
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[b0,landscape,25pt]{sciposter}

\usepackage{multicol,caption} % This is so we can have multiple columns of text side-by-side
\columnsep=100pt % This is the amount of white space between the columns in the poster
\columnseprule=3pt % This is the thickness of the black line between the columns in the poster

\usepackage[svgnames]{xcolor} % Specify colors by their 'svgnames', for a full list of all colors available see here: http://www.latextemplates.com/svgnames-colors

%\usepackage{times} % Use the times font
%\usepackage{palatino} % Uncomment to use the Palatino font

\usepackage{graphicx} % Required for including images
\graphicspath{{figures/}} % Location of the graphics files
\usepackage{booktabs} % Top and bottom rules for table
\usepackage[font=small,labelfont=bf]{caption} % Required for specifying captions to tables and figures
\usepackage{amsfonts, amsmath, amsthm, amssymb} % For math fonts, symbols and environments
%\usepackage{wrapfig} % Allows wrapping text around tables and figures

\usepackage{listings}
\usepackage{amsfonts}
\usepackage{tabularx}
\usepackage{enumitem}

\usepackage{tikz}
\usepackage{graphicx}
\usepackage{subfig}

\usepackage{lipsum}

%\usepackage[boxruled]{algorithm2e}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\newtheorem{Def}{Definition}
\newtheorem{Theorem}{Theorem}

\usetikzlibrary{arrows,positioning}

\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=.5em,
           minimum height=.5em,
           font=\small,
           text centered},
    punktrect/.style={
    rectangle, 
    rounded corners, 
    % fill=black!10,
    draw=black, thick,
    minimum height=1em, 
    text centered},
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,}
}

\begin{document}
\def\ci{\perp\!\!\!\perp}
%----------------------------------------------------------------------------------------
%	POSTER HEADER 
%----------------------------------------------------------------------------------------

% The header is divided into three boxes:
% The first is 55% wide and houses the title, subtitle, names and university/organization
% The second is 25% wide and houses contact information
% The third is 19% wide and houses a logo for your university/organization or a photo of you
% The widths of these boxes can be easily edited to accommodate your content as you see fit
\begin{minipage}[t]{0.15\linewidth}%
\includegraphics[width=10cm]{nicta_logo2.jpg} % Logo or a photo of you, adjust its dimensions here
\end{minipage}
\begin{minipage}[b]{0.70\linewidth}
\centering
\veryHuge \color{NavyBlue} \textbf{Causal Inference in Machine Learning} \color{Black}\\ % Title
\Huge\textit{From Prediction to Decision Making: A review}\\[1cm] % Subtitle
\huge \textbf{Finnian Lattimore}\\ % Author(s)
\huge Australian National University\\ % University/organization
\end{minipage}
\begin{minipage}[t]{0.15\linewidth}\end{minipage}

%----------------------------------------------------------------------------------------

\begin{multicols}{5} % This is how many columns your poster will be broken into, a poster with many figures may benefit from less columns whereas a text-heavy poster benefits from more

%----------------------------------------------------------------------------------------
%	INTRODUCTION
%----------------------------------------------------------------------------------------

\color{DarkSlateGray} % DarkSlateGray color for the rest of the content

\section*{Introduction}

Would reducing salt intake reduce heart attack risk? Would upping the minimum wage increase unemployment? Causal problems differ from the traditional machine learning setting in that they require us to predict the consequences of an intervention, which may change the properties of the distribution from which our data is sampled. Without experimentally testing the intervention or making assumptions to constrain its effect, such inference is impossible. There are many important questions where direct experimentation is expensive, unethical or impossible. Recent research on causality has clarified what assumptions allow causality to be determined and has demonstrated causal inference and discovery is possible with some very general assumptions.

\section*{Causal Frameworks}
\subsection*{Causal Directed Acyclic Graphs}
\begin{itemize}
\item A causal DAG is a Bayesian network where $A \rightarrow B$ is defined to mean $A$ causes $B$.
\item Variables are independent of their non-effects given their direct causes (Causal Markov Property)
\item An intervention that sets a subset of variables $\boldsymbol{X}$ to $\boldsymbol{x}$, denoted $do(\boldsymbol{X} = \boldsymbol{x})$, has a simple graphical representation in a causal DAG, $G$. All links entering intervened on variables, $\boldsymbol{X}$, are deleted, resulting in the mutilated network $G_{\overline{\boldsymbol{X}}}$ (figure \ref{fig:intervention}). Thus, a causal DAG represents the set of all possible interventional distributions over its variables.
\end{itemize}

\begin{figure}
\caption{Intervention in a causal DAG}
\label{fig:intervention}
\centering
{\resizebox{0.6\columnwidth}{!}{\includegraphics{intervention_figure-crop}}}
\end{figure}

\subsection*{(Causal) Structural Equation Models (SEMs)}
\begin{itemize}
\item Represent each variable as a deterministic function of its direct causes and a noise term, where the noise terms are mutually independent.

\item If the set of equations does not create a cycle then the Causal Markov Property holds and the SEM is a causal DAG, (but not visa-versa - SEMs can encode more information).
\end{itemize}

\subsection*{Counterfactuals}
\begin{itemize}
\item Counterfactuals are statements about what would happen under alternate realities where some specified thing differs. For example, consider people taking a medical drug:

For an individual, $i$, let: $\begin{cases}y^{0}_{i} =& \text{outcome if $x_{i}=0$ (not treated)} \\ y^{1}_{i} =& \text{outcome if $x_{i}=1$ (treated)}
\end{cases}$

\item We can define a random variable $Y^{1}$, where $P(Y^{1})$ is the distribution of outcome, $Y$, that would occur if everyone was treated. Similarly $P(Y^{0})$ is the distribution of outcome if no-one was treated. 
\item If   $(X \ci Y^{0}|\boldsymbol{Z})$ \&  $(X\ci Y^{1}|\boldsymbol{Z})$:  $\longleftarrow$ Ignoreability Assumption \\*
we can calculate counterfactual distributions from observed ones:
\begin{equation*}
P(Y^{1}|Z) = P(Y|X=1,Z) \text{ and } P(Y^{0}|Z) = P(Y|X=0,Z)
\end{equation*}

\item Distributions over counterfactual variables that correspond to interventions can be translated directly to the do notation $P(Y^{1}) = P(Y|do(X=1))$. However we can phrase queries with counterfactual variables that are not interventional. For example: \textit{what is the proability that Joe, who was not treated and died, would have recovered had he been treated?}. This query asks about the joint distribution of $P(Y^{0},Y^{1})$.

\begin{center}
\begin{tabular}{c|c|c|c}
group & placebo & treatment & probability of group\\
\hline
1 & die & die & $\alpha=P(Y^{0}=0,Y^{1}=0)$\\
2 & die & recover & $\beta=P(Y^{0}=0,Y^{1}=1)$\\
3 & recover & die & $\gamma=P(Y^{0}=1,Y^{1}=0)$\\
4 & recover & recover & $\delta=P(Y^{0}=1,Y^{1}=1)$\\
\end{tabular}
\end{center}

\item Counterfactuals can be defined in terms of SEMs with a slight relaxation of the independence of errors assumption \cite{Richardson2013}. 
\end{itemize}

\section*{Causal Inference}

Consider the problem where the causal DAG is known, due to theory or prior knowledge and we wish to infer the outcome of an intervention of the form $P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}))$ using observational data.

\begin{itemize}
\item If there are no latent variables, we can compute outcome of any intervention by simply multiply the factors in the mutliated network (figure \ref{fig:intervention}).
\end{itemize}

\subsection*{The Do Calculus}

The do-calculus consists of three rules, which result from d-separation in a causal DAG \cite{Pearl2000}. It is complete: a causal effect is non-parametrically identifiable if and only if the interventional query can be reduced to an observational one via these rules \cite{Shpitser2008}.

\begin{figure}
\caption{d-separation allows us to read conditional independences off a DAG. If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$. }
\label{fig:dsep}
\centering
{\resizebox{0.6\columnwidth}{!}{\includegraphics{dsepfigure-crop}}}
\end{figure}

\fbox{\begin{minipage}{0.99\columnwidth}%
\textbf{The Three Rules}
\begin{enumerate}
\item A causal DAG remains a causal DAG after an intervention so d-separation still applies.
\begin{equation*}
\label{eq:Do1}
\begin{aligned}
\text{if }  (\boldsymbol{Y} \ci & \boldsymbol{W}|\boldsymbol{X})  \text{ in } G_{\overline{\boldsymbol{X}}}  \\
& P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) 
\end{aligned}
\end{equation*}
\item If $\boldsymbol{Y}$ is independent of \textit{how} variables $\boldsymbol{X}$ take their values then the effect on $\boldsymbol{Y}$ of setting $\boldsymbol{X}$ to some value is equivalent to observing it take that value. If this rule is satisfied the corresponding ignoreability assumption in the counterfactual framework is satisfied.
\begin{equation*}
\label{eq:Do2}
\begin{aligned}
\text{if } (\boldsymbol{Y} \ci  \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z}) & \text{ in } G^{\dagger}\\
& P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{Z})
\end{aligned}
\end{equation*}
\item If there is no direct causal path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ then intervention on $\boldsymbol{X}$ does not change the distribution of $\boldsymbol{Y}$.
\begin{equation*}
\label{eq:Do2}
\begin{aligned}
\text{if }  (\boldsymbol{Y} \ci  \boldsymbol{\hat{X}}|\boldsymbol{Z}) & \text{ in } G^{\dagger}\\
& P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}) = P(\boldsymbol{Y}|\boldsymbol{Z})
\end{aligned}
\end{equation*}
\end{enumerate}
\footnotesize (For readability, this is a simplified version of the do-calculus that covers interventions on a single variable or cases where it is sufficient for identifiability to consider intervention on all variables together. The fully general version is only slightly more complex see \cite{Pearl2000})
\normalsize
\end{minipage}}

\section*{Causal Discovery}

Causal discovery attempts to infer causal structure from data based on more general assumptions.

\subsection*{Indepdendence Based Methods}
\subsubsection*{Without Latent Variables}
\begin{enumerate}
\item We assume our distribution $P$ was generated by some (unknown) causal DAG over our observed variables (causal sufficiency) 
\item We assume that all the conditional independences in $P$ are implied by d-separation in the true causal network (\textbf{\textit{faithfulness}})
\item Finding the causal structure equates to finding perfect maps for $P$
\end{enumerate}

\begin{algorithm}
\caption{SGS or IC Algorithm \cite{Sprites,Pearl2000}. \label{alg:spect}}
\textbf{Input:} A Distribution $P$ over variables $\boldsymbol{V}$

\textbf{Output:} A partially directed network representing the Markov equivalence class for the generating causal model.
\begin{enumerate}
\item Create a complete undirected graph over $V$. For all pairs of variables $(a,b) \in V$ search for a set $S_{ab}$ s.t $a \ci b|S_{ab}$. If such a set exists, delete the link $a-b$
\item For all pairs of unlinked-nodes $(\alpha, \beta)$ with a common neighbour $c$, if $c \notin S_{\alpha \beta}$ direct links towards $c$.
\item Recursively direct any remaining links for which there is only one orientation that does not create a cycle or any additional v-structures ( $\bullet \rightarrow \bullet \leftarrow \bullet$). 
\end{enumerate}
\end{algorithm}

\begin{figure}
\caption{The SGS Algorithm}
\label{fig:SGS}
\centering
{\resizebox{0.4\columnwidth}{!}{\includegraphics{SGS_figure-crop}}}
\end{figure}

\begin{itemize}
\item The SGS algorithm is infeasible in practise due to the exponential number of (high order) conditional independence tests it requires. 
\item 
The PC algorithm \cite{Sprites} modifies the SGS algorithm to exploit any sparsity in the true network, leading to much better average case performance.
\end{itemize}

\subsubsection*{With Latent Variables}
\begin{itemize}
\item For every latent structure there is a dependency equivalent structure in which every latent variable is a root node with exactly two children \cite{Verma1993}. This key to the FCI algorithm \cite{Sprites}, which generalizes the PC algorithm to handle latent and selection variables.

\item The FCI algorithm returns an equivalence class of Maximal Ancestral Graphs (a generalization of DAGs) since DAGs are not closed under marginalization (figure \ref{fig:FCI}b).
\end{itemize}

\newcommand{\imsize}{.5\columnwidth}

\begin{figure}
\caption{The FCI Algorithm}
\label{fig:FCI}
\centering
{\resizebox{0.9\columnwidth}{!}{\includegraphics{FCI_figure-crop}}}
\end{figure}

\begin{itemize}
\item The FCI algorithm discovers all aspects of causal structure identifiable from conditional independence relations \cite{Zhang2008}.
\item It can be made to require a worst case polynomial (rather than exponential) number of conditional independence tests for sparse graphs \cite{Claassen2013}. 
\item Implementations of both the PC and FCI algorithm are available in the R package pcalg \cite{Kalisch2012}
\end{itemize}

\subsection*{Beyond independence}
Independence based methods have the advantage that they require only very general, non-parametric, assumptions. However they cannot distinguish between causal graphs with equivalent dependency structure; for example, between $X \rightarrow Y$ and $X \leftarrow Y$.

\begin{figure}
\centering
\caption{Figure from {\cite{Hoyer2009}}. Additive noise models, $Y = f(X)+\epsilon$, are identifiable for most combinations of $f$ and $P(\epsilon)$ but not in linear-gaussian case}
{\resizebox{\imsize}{!}{\includegraphics{Hoyer1}}}
\end{figure}

\begin{figure}
\centering
\caption{Figure from {\cite{Daniusis2010}}. The causal direction can be identifiable even where the relationship between $X$ and $Y$ is deterministic and invertible. Let $Y = f(X) \Leftrightarrow X = g(Y)$. For most input distributions, $p_{X}(x)$, the distribution $p_{Y}(y)$ will be higher where $f'$ is small and a larger region of $X$ maps to similar values of $Y$. If $X$ causes $Y$ (but not if $Y$ causes $X$) we would expect $f$ and $p_{X}(x)$ to be independent and $p_{Y}(y)$ should be correlated with $f'$.}

{\resizebox{\imsize}{!}{\includegraphics{Daniusis1}}}
\end{figure}

\begin{figure}
\centering
\caption{Figure from {\cite{Scholkopf2012}}. The idea of independence of mechanism and input can be generalized to the non-deterministic setting. If $X \rightarrow Y$ then $P(X)$ and $P(Y|X)$ should be independent, but $P(Y)$ and $P(X|Y)$ are not. Therefore, semi-supervised learning should yield no benefit when trying to learn in the causal direction, (estimating $P(Y|X)$) but could help when learning in the anti-causal direction.}
{\resizebox{\imsize}{!}{\includegraphics{Janzig6}}}
\end{figure}

\subsubsection*{Learning what causality looks like {\cite{LopezPaz2014}}}

Suppose we had $M$ different causal pairs data sets.

\begin{equation*}
D= \{\{x_{j},y_{j}\}_{j=1}^{N_{i}},l_{i}\}_{i=1}^{M}
\end{equation*}

where $l_{i}$ is a binary label that indicates if $X \rightarrow Y$ or $Y \rightarrow X$ in dataset $i$.

\begin{itemize}
\item Kernel mean embedding allows us to take a distribution $P$ and transform it to a point in some Hilbert space.
\item We expect there to be differences in the relationships between $P(X)$, $P(Y)$ and $P(Y|X)$ for $X \rightarrow Y$ and $Y \rightarrow X$
\end{itemize}

\begin{algorithm}
\caption{\label{alg:causallearn}}
\begin{enumerate}
\item Let $\mu$ be a kernel mean embedding that maps a distribution $P$ into some Hilbert space.
\item For each data set $i={1...M}$, construct a feature vector that approximates ${\mu(P(X)), \mu(P(Y)),\mu(P(X,Y)) }$
\item Apply a standard classification algorithm to learn if $X\rightarrow Y$ or $Y \rightarrow X$
\end{enumerate}
\end{algorithm}
\subsubsection*{With more than two variables}
\begin{itemize}
\item If you can come up with a condition, on the triple $(P(Y), P(\epsilon), f)$, that guarantees identifiability for the bi-variate SEM $Y = f(X) + \epsilon$, you can extend that result to get the conditions under which the multivariate case is identifiable \cite{Peters2014}.

\end{itemize}
%----------------------------------------------------------------------------------------
%	CONCLUSIONS
%----------------------------------------------------------------------------------------

\color{SaddleBrown} % SaddleBrown color for the conclusions to make them stand out

\color{DarkSlateGray} % Set the color back to DarkSlateGray for the rest of the content

 %----------------------------------------------------------------------------------------
%	REFERENCES
%----------------------------------------------------------------------------------------

\bibliographystyle{plain} % Plain referencing style
\bibliography{library} % Use the example bibliography file sample.bib

\end{multicols}
\end{document}

We present a survey of recent results, focusing on the problems of inferring causal effects from observational data and causal discovery.

  \begin{minipage}{\columnwidth}
    \makeatletter
    \newcommand{\@captype}{figure}
    \makeatother
    \centering
    \subfloat[Market revenue]{%
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, above left=of 2](4){$W$};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    	edge node {} (3)
    (2) edge node {} (3)
    (4) edge node {} (2);

\end{tikzpicture}

      \label{fig:evaluation:revenue}%
    }\qquad%
    \subfloat[Average price]{%
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, above left=of 2](4){$W$};

 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    	edge node {} (3)
    (2) edge node {} (3)
    (4) edge node {} (2);

\end{tikzpicture}

      \label{fig:evaluation:avgPrice}%
    }
    \caption{Simulation results}
  \end{minipage}

### 2014-11-21 wiml_poster/sciposter-example.tex
adds first draft of wiml_poster

\documentclass{sciposter}

\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{multicol}
%\usepackage{fancybullets}

\newtheorem{Def}{Definition}

%\definecolor{BoxCol}{rgb}{0.9,0.9,0.9}
% uncomment for grey background to \section boxes 
% for use with default option boxedsections

%\definecolor{BoxCol}{rgb}{0.9,0.9,1}
% uncomment for light blue background to \section boxes 
% for use with default option boxedsections

%\definecolor{SectionCol}{rgb}{0,0,0.5}
% uncomment for dark blue \section text

\title{Generalized Pattern Spectra Sensitive to Spatial Information}

% Note: only give author names, not institute
\author{Michael H. F. Wilkinson}

% insert correct institute name
\institute{Institute for Mathematics and Computing Science,\\
           University of Groningen\\}

\email{michael@cs.rug.nl}  % shows author email address below institute

%\date is unused by the current \maketitle

% The following commands can be used to alter the default logo settings
%\leftlogo[0.9]{logoWenI}{  % defines logo to left of title (with scale factor)
%\rightlogo[0.52]{RuGlogo}  % same but on right

% NOTE: This will require presence of files logoWenI.eps and RuGlogo.eps, 
% or other supported format in the current directory  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Begin of Document

\begin{document}
%define conference poster is presented at (appears as footer)

\conference{{\bf ICPR 2002}, 16th International Conference on Pattern
  Recognition, 11-15 August 2002, Qu\'ebec City, Canada}

%\LEFTSIDEfootlogo  
% Uncomment to put footer logo on left side, and 
% conference name on right side of footer

% Some examples of caption control (remove % to check result)

%\renewcommand{\algorithmname}{Algoritme} % for Dutch

%\renewcommand{\mastercapstartstyle}[1]{\textit{\textbf{#1}}}
%\renewcommand{\algcapstartstyle}[1]{\textsc{\textbf{#1}}}
%\renewcommand{\algcapbodystyle}{\bfseries}
%\renewcommand{\thealgorithm}{\Roman{algorithm}}

\maketitle

%%% Begin of Multicols-Enviroment
\begin{multicols}{3}

%%% Abstract
\begin{abstract}
Morphological pattern spectra computed from granulometries are frequently used
to classify the size classes of details in textures and images. An extension 
of this technique, which retains information on the spatial 
distribution of the details in each size class is developed. Algorithms for
computation of these spatial pattern spectra for a large number of 
granulometries on binary images are presented. 
\end{abstract}

%%% Introduction
\section{Introduction}

\PARstart{G}{ranulometries} are ordered sets of morphological openings or closings, each of
which removes image details below a certain size. These can be used for texture
analysis
through the use of \emph{pattern spectra}, which show how the number of 
foreground pixels in the image changes as a function of the size parameter 
\cite{maragos89:_patter}.
A drawback of the classical definition of pattern spectra is that spatial 
information is not included in a pattern spectrum as shown below.
 In this paper, \emph{spatial pattern spectra} are developed which retain information on the distribution of these details at different scales.

\newcommand{\imsize}{0.45\columnwidth}
\begin{figure}
\begin{center}
\begin{tabular}{c c}
3 &
3\\
(a) & (b) \\
4 &
5\\
(c) & (d) \\
\end{tabular}
\end{center}
\caption{ Parts (a) through (c) show three images consisting of squares of
different sizes;
(d) shows the pattern spectra, denoting the number of foreground pixels 
 removed by openings by reconstruction by $\lambda \times \lambda$ squares. No 
granulometry is capable of separating the patterns, because the only 
differences between the images lie in the distributions of the 
connected components. }\label{fig:blocks}
\end{figure}

\section{Theory}

Let binary images $X$ and $Y$ be defined as a subset of the image domain 
${\mathbf M}\subset {\mathbb Z}^n$ or ${\mathbb R}^n$ (usually $n=2$). 
\begin{Def}
A binary 
granulometry is a set of operators $\{\alpha_r\}$ with $r$ from some ordered 
set $\Lambda$ (usually $\Lambda \subset {\mathbb R}$ or ${\mathbb Z}$), with 
the following three properties
\begin{align}
   \alpha_r(X) & \subset  X \label{eq:antiext} \\
   X \subset Y & \Rightarrow \alpha_r(X) \subset \alpha_r(Y) 
   \label{eq:increasing} \\
   \alpha_r(\alpha_s(X)) & =  \alpha_{\max(r,s)}(X) \label{eq:idempot},
\end{align}   
for all $r,s \in \Lambda$.
\end{Def}

\begin{Def}
The pattern spectrum $s_{\alpha}(X)$ obtained by applying 
granulometry $\{\alpha_r\}$ to a binary image $X$ is defined as
\begin{equation}
    (s_{\alpha}(X))(u) = 
    - \frac{\partial A(\alpha_r(X))}{\partial r}\bigg{\vert}_{r=u}
\end{equation}
in which $A(X)$ is a function denoting the Lebesgue measure in 
${\mathbb R}^n$.
\end{Def} 
In the case of discrete images, and with $r \in \Lambda \subset {\mathbb Z}$, 
this differentiation reduces to
\begin{align}
    (s_{\alpha}(X))(r)  & = \#(\alpha_{r}(X) \setminus \alpha_{r^+}(X)) \\ 
                        & = \#(\alpha_{r}(X)) - \#(\alpha_{r^+}(X)), 
\end{align}
with $r^+ = \min\{ r' \in \Lambda \vert r' > r \}$, and $\#(X)$ the 
numnber of elements of $X$.

The opening transform \cite{Nacken:thesis} $\Omega_X$ of a binary image $X$ 
for a granulometry ${\alpha_r}$ is
\begin{equation}
  \Omega_X(x) = \max\{ r \in \Lambda \vert x \in \alpha_r(X) \}
\end{equation}

The pattern spectrum of a binary image $X$ using granulometry 
$\{\alpha_r\}$ is the histogram of $\Omega_X$ obtained with the same 
size distribution \cite{Nacken:thesis}, disregarding the bin for grey level 0.

\begin{figure}
\begin{center}
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}
\end{center}
\caption{ \label{fig:opentransf} Opening transform with $\{\alpha_r\}$ as in 
 Fig. \ref{fig:blocks}: (left) original image; (right) opening transform
(contrast stretched for clarity). 
}
\end{figure}

\section{Spatial pattern spectra}
Pattern spectra only retain the amount of detail present at  scale $r$.
This can be amended by computing some parameterization of the spatial 
distribution in an image $\alpha_r(X) \setminus \alpha_{r+}(X)$ as a function of $r$.

\begin{Def}
Let ${M}(X)$ be some parameterization of the spatial distribution of detail
in the image $X$. The spatial pattern spectrum ${S}_{{M},\alpha}$ is
then defined as
\begin{equation}
  ({S}_{{M},\alpha}(X))(r) = {M}(\alpha_r(X) \setminus \alpha_{r+}(X)).
\end{equation}  
\end{Def}

An obvious parameterization of the spatial distribution is through 
the use of moments. Focusing on the case of 2-D binary images, the 
moment $m_{ij}$ of order $ij$ of an image $X$ is given by
\begin{equation}
  m_{ij}(X) = \sum_{(x,y) \in \mathbf X} x^i y^j.
\end{equation}  
The spatial moment spectrum $S_{m_{ij},\alpha}$ of order $ij$ is
\begin{equation}
  (S_{m_{ij},\alpha}(X))(r) = m_{i,j}(\alpha_r(X) \setminus \alpha_{r^+}(X)).
\end{equation}  
For $i=0$ and $j=0$ we obtain the standard pattern spectrum. 
For each $r$, $(S_{m_{ij},\alpha}(X))(r)$ is just the moment of an image, 
therefore, derived parameters such as coordinates of the centre of mass, 
(co-)variances, skewness and kurtosis of the distribution of details at each
scale can be computed easily. We can then define pattern mean
spectra, pattern (co-)variance spectra, pattern kurtosis spectra, etc. The 
pattern mean-$x$ and variance-$x$ spectra 
($S_{\bar x,\alpha}$ and $S_{\sigma(x),\alpha}$) are defined as: 
\begin{align}
  S_{\bar x,\alpha} & = \frac{S_{m_{10},\alpha}} {S_{m_{00},\alpha}} \\
\intertext{and}
   S_{\sigma(x),\alpha} & = \sqrt{\frac{S_{m_{20},\alpha}}
                                 {S_{m_{00},\alpha}} 
                                 - S_{\bar x, \alpha}}.  
 \end{align}
These two are shown in Figures \ref{fig:tauspect} and \ref{fig:binspect}. Note that 
these definitions hold only where $(S_{m_{00},\alpha}(f))(r) \neq 0$. For all 
other values of $r$ they will be defined as zero. Further post-processing can
be done to compute central moments and moment invariant from pattern moment 
spectra \cite{Flusser:Suk:93,Hu:62}.

\section{An Algorithm}

Nacken \cite{Nacken:thesis} derived an algorithm for computation
of pattern spectra for granulometries based on openings by discs of increasing
radius for various metrics, using the opening transform. After the
opening transform has been computed, it is straightforward to compute the 
pattern spectrum:
\begin{itemize}
\item Set all elements of array {\tt S} to zero
\item For all $x \in X$ increment {\tt S}[$\Omega_X(x)$] by one. 
\end{itemize}

To compute the pattern \emph{moment} spectrum, the only thing that needs to be
changed is the way {\tt S}[$\Omega_X(x)$] is incremented. As shown in Algorithm
\ref{alg:spect}.

\begin{algorithm}
\begin{itemize}
\item Set all elements of array {\tt S} to zero
\item For all $(x,y) \in X$ increment {\tt S}[$\Omega_X(x,y)$] by 
$x^iy^j$. 
\end{itemize}
\caption{ Algorithm for computation of pattern moment
spectrum of order $ij$. \label{alg:spect}}
\end{algorithm}

This algorithm can 
readily be adapted to other granulometries, simply by computing the 
appropriate opening transform.

\begin{figure}
\begin{center}
\begin{tabular}{c c}
{\resizebox{\imsize}{!}{\includegraphics{f.png}}} &
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}\\
(a) & (b)\\
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}&
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}\\
(c) & (d) \\
\end{tabular}
\end{center}
\caption{ \label{fig:tauspect} 
The opening transform using city-block metric: (a) opening transform of
Fig. 1(c); (b) pattern spectrum; (c) pattern variance-$x$; 
(d) variance-$y$ spectra.}
\end{figure}

\renewcommand{\imsize}{0.3\columnwidth}
\begin{figure}
\begin{center}
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}\\
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}
{\resizebox{\imsize}{!}{\includegraphics{f.png}}}
\end{center}
\caption{ \label{fig:binspect} Pattern mean-$x$ (top) and variance-$x$ 
(bottom) spectra: the three collumns show spectra for Fig. 1(a), (b) and (c) 
from left to right respectively.  Unlike the standard pattern spectra, 
these spatial pattern spectra can distinguish the three images.}
\end{figure}

\section{Discussion}
Spatial pattern spectra form a useful supplement to ordinary pattern 
spectra, because of their ability to retain spatial information.  
Pattern moment spectra, in particular, are easily computed concurrently with 
computation of the standard pattern spectrum. Post-processing of these pattern
moment spectra can be done to yield a number of easily interpreted spectra, 
such as pattern mean, variance, skew, and kurtosis spectra, which have reduced 
covariance compared to the ``raw'' pattern moment spectra. Invariance to 
rotation, translation or scale change can also be achieved by post-processing
\cite{Flusser:Suk:93,Hu:62}.

In the future grey scale versions of these spatial pattern spectra will be 
developed. I expect that the efficient grey level algorithms for area and
attribute pattern spectra  
\cite{Meijster:Wilkinson:PAMI}
can be adapted to spatial pattern spectra as well.

%%% References

%% Note: use of BibTeX als works!!

\bibliographystyle{plain}
\begin{thebibliography}{1}

\bibitem{Flusser:Suk:93}
J.~Flusser and T.~Suk.
\newblock Pattern recognition by affine moment invariants.
\newblock {\em Pattern Recognition}, 26:167--174, 1993.

\bibitem{Hu:62}
M.~K. Hu.
\newblock Visual pattern recognition by moment invariants.
\newblock {\em IRE Transactions on Information Theory}, IT-8:179--187, 1962.

\bibitem{maragos89:_patter}
P.~Maragos.
\newblock Pattern spectrum and multiscale shape representation.
\newblock {\em IEEE Trans. Patt. Anal. Mach. Intell.}, 11:701--715, 1989.

\bibitem{Meijster:Wilkinson:PAMI}
A.~Meijster and M.~H.~F. Wilkinson.
\newblock A comparison of algorithms for connected set openings and closings.
\newblock {\em IEEE Trans. Patt. Anal. Mach. Intell.}, 24(4):484--494, 2002.

\bibitem{Nacken:thesis}
P.~F.~M. Nacken.
\newblock {\em Image Analysis Methods Based on Hierarchies of Graphs and
  Multi-Scale Mathematical Morphology}.
\newblock PhD thesis, University of Amsterdam, Amsterdam, The Netherlands,
  1994.

\end{thebibliography}

\end{multicols}

\end{document}

### 2014-11-21 causal-bandit/c-bandit.tex
updates causal bandit stuff

@@
 
-Let $\hat{\mu_a}$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $a$.
+Let $\hat{\mu}_a$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $X_a$.
 
 \eq {
-\hat{\mu_a} &=
+\hat{\mu}_a &=
 \begin{cases}
@@
 
-If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p_a}$
+If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p}_a$
 
 \eq{
-\hat{p_a} = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
+\hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
 }

### 2014-11-21 causal-bandit/ucb_regret.tex
updates causal bandit stuff

--- 2014-11-06 causal-bandit/c-bandit.tex
@@
 
-\title{Intervention Bandits}
-\author{Blah blah}
+\title{Regret Bounds for UCB}
+\author{Finnian Lattimore}
 
 \begin{document}
-
+\def\ci{\perp\!\!\!\perp}
 \maketitle
 
-\begin{abstract}
-An abstract.
-\end{abstract}
+Assume $\exists$ a convex function, $\psi$, such that:
 
-\section{Introduction}
+\begin{equation}
+\begin{aligned}
+log(E[e^{\lambda(X-E[X]}]) & \leq \psi(\lambda)\\
+log(E[e^{\lambda(E[X]-X}]) & \leq \psi(\lambda)
+\end{aligned}
+\end{equation}
+This ensures that the moments of the distribution of X are defined. Assume that at time $t$ we select arm $I_t$ with the highest upper confidence bound:
 
-Useful references are: \cite{BC12}.
+\begin{equation}
+\label{eq:armSelection}
+I_{t} = argmax_{i=1...K}\left[\hat{\mu}_{it} + (\psi^{*})^{-1}\frac{\alpha log(t)}{T_{it}}\right]
+\end{equation}
 
-\section{Notation}
+Then if $\alpha > 2$,
 
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
+\begin{equation}
+R_n \leq \sum_{i:\Delta i > 0}\left(\frac{\alpha \Delta i}{\psi^{*}(\Delta i/2)}log(n)+\frac{\alpha}{\alpha-2}\right)
+\end{equation}
 
-Let $p \in [0,1]^K$ be a fixed and known vector. In each time-step $t$:
-
+\begin{theorem} If $I_t = i \neq i^*$ at least one of the following statements is true:
 \begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
+\item The estimated UCB on the best arm, $i^*$, is less than or equal to the actual reward for that arm: $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \hat \mu$
+\item The estimated reward for arm $i$ is greater than or equal to the estimated CI higher than the true reward for that arm: $\hat \mu_{it} \geq \mu_i + \hat \epsilon_{it}$
+\item The number of times we have selected arm $i$ in previous timesteps, $T_{it}$, is less than some bound (that grows logarithmically with $n$). $T_{it} < \frac{\alpha log(n)}{\psi^*(\Delta i/2)}$
 \end{enumerate}
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|X_i = j]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
+\end{theorem}
+\begin{proof}
+Assume statements 1-3 are all false.
+\eq {
+3. & \implies T_{it} > \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \\
+  & \implies \Delta i > 2 (\psi^*)^{-1} \frac{\alpha log(n)}{T_{it}} \geq 2 (\psi^*)^{-1} \frac{\alpha log(t)}{T_{it}} = 2 \hat \epsilon_{it}\\
+  & \implies \Delta i > 2 \hat \epsilon_{it} \\
+1. & \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \hat \mu  = \mu_i+\Delta i > \mu_i +  2 \hat \epsilon_{it}\\
+& \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \mu_i +  2 \hat \epsilon_{it}\\
+2. & \implies \hat \mu_{it} < \mu_i + \hat \epsilon_{it} \\
+   & \implies \hat \mu_{it} + \hat \epsilon_{it} < \mu_i +2 \hat \epsilon_{it} \\
+   & \implies \hat \mu_{it} + \hat \epsilon_{it} <  \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \longleftarrow \text { UCB for arm $i < $ UCB for arm $i^*$, which contradicts equation (\ref{eq:armSelection}})
 }
 
-\section{Estimating $\mu_{i,j}$}
-The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
-lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
-outperform this approach by exploiting the extra structure in the problem.
-
-Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
-We define estimator $\hat \mu_t$ of $\mu_{i,j}$ by
-\eq{
-m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
-n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
-\hat \mu_t &= \eta_i \frac{m_{i,j}}{n_{i,j}}  + \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right] \\
-\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \\
-n_a        &= \begin{cases} 
-  n_{i,j}                                           & \text{if } a = i \\
-  \frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
-\end{cases}
-}
-\begin{theorem} With probability at least $1 - \delta$ we have that:
-$\displaystyle \left|\hat \mu_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
-where $\beta > 0$ is some constant.
-\end{theorem}
-
-\begin{proof}
-First note that $n_{a,b}$ is a random variable that is bounded by $t$ for all $a,b$.
-We use the short-hand $\mu^{a,b}_{i,j} = \E[q(X) | X_i = j, X_a = b]$. Then
-\eq{
-\mu_{i,j} = p_a \mu_{i,j}^{a,1} + (1 - p_a) \mu_{i,j}^{a,0}\,.
-}
-Now we can apply Hoeffding's bound and the union bound to show that
-\eq{
-\P{\left|\frac{m_{a,b}}{n_{a,b}} - \mu_{i,j}^{a,b}\right| \geq \sqrt{\frac{1}{2n_{a,b}} \log \frac{4t}{\delta}}} \leq \frac{\delta}{2}\,.
-}
-Therefore by the union bound
-\eq{
-\P{\left|p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}} - \mu_{i,j}\right| \geq p_a \sqrt{\frac{1}{2n_{a,1}} \log\frac{4t}{\delta}}
-+ (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}} \leq \delta\,
-}
-Now by Jensen's inequality 
-\eq{
-p_a \sqrt{\frac{1}{2n_{a,1}}\log\frac{4t}{\delta}} + (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}
-&\leq \sqrt{\left(\frac{p_a}{2n_{a,1}} + \frac{1-p_a}{2n_{a,0}}\right)\log \frac{4t}{\delta}} \\
-&\leq \sqrt{\max\set{\frac{p_a}{n_{a,1}}, \frac{1-p_a}{n_{a,0}}} \log\frac{4t}{\delta}} \\
-&= \sqrt{{1 \over 2n_a} \log\frac{4t}{\delta}}\,.
-}
-Similarly,
-\eq{
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{4t}{\delta}}} \leq
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{2t}{\delta}}} \leq \delta\,.
-}
 \end{proof}
 
-\section{Algorithm}
-
-\begin{algorithm}
-\caption{UCB}
-\begin{algorithmic}[1]
-\State {\bf Input:} Number of variables $K$, vector $p \in [0,1]^K$, horizon $n$
-\For{$t \in 1,\ldots,n$}
-\For{$i \in 1,\ldots,K$}
-\For{$j \in \set{0,1}$}
-\State Compute $\tilde \mu_{i,j} = \hat \mu_{i,j} + \sqrt{\frac{\alpha}{\sum_{a} n_a} \log n}$ 
-\EndFor
-\EndFor
-\State Choose $I_t, J_t = \argmax_{i,j} \tilde \mu_{i,j}$
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\section{Theorems}
-\section{Experiments}
-
-\begin{figure}[h]
-\caption{Testing values of beta - 100 experiments for each value of beta, with a 4 armed bandit.}
-\includegraphics[width=0.7\textwidth]{regretvsbeta.pdf}
-\end{figure}
-
-Simululations to compare the performance of standard UCB with our modified algorithm.
-For each number of arms, $100$ bandits of each type were created and run upto to a horizon of $1000$ timesteps.
-The mean regret and its standard error from these simulations is plotted in figure \ref{fig:regretvsarms} The true data was generated from a model where:
-
-\eq{
-&p = [0.5]^{K}\\
-&q(\boldsymbol{X}) = \begin{cases}0.5 & \text{if $X_{1}=0$}\\ 0.6 & \text{otherwise} \end{cases}
-}
-
-For small number of arms there is not much difference.
-As the number of arms increases, cusb shows significantly lower regret.
-As the number of arms increases above 100 the regret of the two algorithms begins to converge again.
-I suspect this is due to the fixed time horizon of $1000$ steps.
-With a time horizon of $100$ steps the two algorithms begin to converge again at a smaller number of arms.
-
-\begin{figure}[h]
-\caption{Shows relationship between the regret and the number of bandit arms $2K$ for the standard UCB algorithm and our modified algorithm.  }
-\label{fig:regretvsarms}
-\includegraphics[width=0.9\textwidth]{regretvsarms.pdf}
-\end{figure}
-
-\section{Conclusion}
-
-\bibliographystyle{plainnat}
-\bibliography{c-bandit}
+If statements 1. and 2. are both false, then statement 3. places a bound on the number of times we can previously have selected the incorrect arm $i$ in order to select it in this timestep.
 

### 2014-11-21 UCBbandit.py
commits python scripts etc for causal reinforcment

@@ -1,0 +1,5 @@
+
+def epsilon(N,alpha):
+    return sqrt((1.0/float(2.0*N))*math.log(2.0/alpha))
+
+

### 2014-11-21 libcause.py

@@ -41,14 +41,22 @@ class BinaryNetworkBandit(object):
         self.prior = prior
         self.interventions = [] # a list of possible assignments
         self.variables = [] # store the variables - defines an ordering
+        self.intervention_to_arm = {}
+        self.variable_name_to_index = {}
         values = []
+        index = 0
+        variable_index = 0
         for variable, data in self.bn.Vdata.iteritems():
             if variable != self.y:
                 self.variables.append(variable)
+                self.variable_name_to_index[variable] = variable_index
+                variable_index +=1
                 vals = data.get("vals")
                 values.append(vals)
                 for value in vals:
                     self.interventions.append({variable:value})
+                    self.intervention_to_arm[(variable,value)]=index
+                    index +=1
 
         # lets calculate and print the actual value of theta for each arm (since we know it)
         
@@ -80,6 +88,10 @@ class BinaryNetworkBandit(object):
         # record distributions for P(X1), P(X2) ... - they update only when we observe Xn not when we do it
         self.observed_trials = zeros(shape=(self.num_arms,), dtype=int)
         self.observed_true = zeros(shape=(self.num_arms,), dtype=int)
+
+        # records how many times each variable was set by intervention
+        self.intervened_trials = zeros(shape=(self.num_arms,), dtype=int)
+        self.intervened_true = zeros(shape=(self.num_arms,), dtype=int)
                 
     
     def sample(self,n,plot=-1):
@@ -91,7 +103,7 @@ class BinaryNetworkBandit(object):
                 do_plot = False
             arm = self.get_recommendation(do_plot)
             intervention = self.interventions[arm]
-            # note evidence is equivelent to do in libpgm
+            # note: evidence is equivelent to do in libpgm
             result = self.bn.randomsample(1,evidence=intervention)[0] # returns a dictionary containing values for each variable
             reward = int(result.get(self.y))
            
@@ -100,6 +112,15 @@ class BinaryNetworkBandit(object):
             if (reward == 1):
                 self.successes[arm] = self.successes[arm] + 1
 
+            # for variable we intervened on record that the we intervened
+            assert(len(interventions.keys())==1)
+            do_variable = intervention.keys()[0] # ASSUMING SINGLE INTERVENTIONS AT THIS POINT
+            do_value = intervention.values()[0]
+            do_variable_indx = self.variable_name_to_index[do_variable]
+            self.intervened_trials[do_variable_indx] +=1
+            self.intervened_true[do_variable_indx] +=1
+            
+            
             
[…]

### 2014-11-21 multicausesim.py

@@ -130,7 +130,7 @@ class Arm(object):
                 totalW += n
      
             varID +=1
-
+        print "arm",self.id,"u:",pr,"nSum:",totalW
         pr = pr/totalW
         er = math.sqrt((beta/totalW)*math.log(1/self.alpha))    
         return (pr,er)
@@ -179,6 +179,7 @@ class CausalBandit2(object):
         upperBounds = [arm.upperCI(self.model.pxlist,self.beta) for arm in self.arms]
         self.interventions +=1
         selectedArmIndx = upperBounds.index(max(upperBounds))
+        print "Arm:",selectedArmIndx
         self.sumExpectedReward += self.model.py[selectedArmIndx]
         return selectedArmIndx
                
@@ -467,59 +468,64 @@ def tryBeta(betalst):
 
 
 beta = 0.6
-n = 1000
-experiments = 100
-cmeans = []
-cstds = []
-means = []
-stds = []
-o = open("manyarmedbandit.txt","w")
-armslst = [pow(2,x) for x in range(2,11)]
-for numArms in armslst:
-    print numArms,"arms",
-    model = TrivialProbabilityModel(numArms,0.3)
-
-    data = np.zeros((experiments,2))
-    for i in xrange(experiments):
-        bandit = CausalBinaryBandit(model)
-        bandit.UCBSample(n,0.05)
-        bandit2 = CausalBandit2(model,beta)
-        bandit2.sample(n,0.05)
-        data[i,0] = bandit2.regret()
-        data[i,1] = bandit.regret()
-        if i % 5 == 0:
-            print i,
-    print ""
-        
-    
-    m = np.mean(data,axis=0)
-    sd = np.std(data,axis=0)/math.sqrt(experiments)
-    l = list(m)
-    s = list(sd)
-    l.extend(s)
-    l.append(numArms)
-    l = [str(x) for x in l]
-    o.write(",".join(l)+"\n")
-    o.flush()
-    cmeans.append(m[0])
-    cstds.append(sd[0])
-    means.append(m[1])
-    stds.append(sd[1])
-
-o.close()
-
-f,axis = plt.subplots()
-axis.errorbar(armslst,means,yerr=stds,fmt="o",label="ucb")
-axis.errorbar(armslst,cmeans,yerr=cstds,fmt="D",label="cucb")
-axis.set_xlabel("number of arms")
-axis.set_ylabel("regret")
[…]

### 2014-11-21 profile.py

@@ -1,0 +1,9 @@
+import cProfile
+from multicausesim import compareBandits
+o = open("tmp.txt","w")
+numArms = 8
+px = [0.5]*numArms
+pyGivenX = [0.5]*pow(2,numArms)
+pyGivenX[0:pow(2,numArms-1)] = [0.6]*pow(2,numArms-1) # think further about this step ...
+cProfile.run('compareBandits(1,1,o,px,pyGivenX)')
+        

### 2014-11-21 tmpcause.py

@@ -1,0 +1,4 @@
+from scipy.stats import bernoulli
+from scipy.stats import beta
+import numpy as np
+

### 2014-11-26 causal-bandit/c-bandit.tex
adds sketch for ucb regret bound proof

@@
 \usepackage{graphicx}
-
+\usepackage{parskip}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\subsection{Estimators}
 Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
@@
 }
+\subsection{Random sampling variant}
 
-\begin{theorem} With probability at least $1 - \delta$ we have that:
+Basic idea - explore randomly until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB).
+Bounds on regret: when exploring regret is at most 1, when exploiting regret it at most $\epsilon$.
+If the horizon is $n$, the cost of exploiting $< \epsilon n$ and the cost of exploring is $O(1/\epsilon^2)$ (this comes from how quickly we converge to within $\epsilon$ for all arms).
+
+\eqn{
+R_n = O(n\epsilon+\frac{1}{\epsilon^2})
+}
+
+Differentiating and selecting $\epsilon = (\frac{2}{n})^{1/3}$ to minimize the regret yields:
+
+\eqn{
+R_n = O(n^{2/3})
+}
+
+At each time step we randomly select a variable $X_i$ on which to intervene and set it to $1$ with probability $p_i$.
+
+The estimators from section XX are unbiased. So we can get an estimate of the error in the estimate by
+
+we are pooling estimators - so what is the error in a pooled estimate - is it the same as what Tor wrote before (because he was assuming unbiased estimators?
+Or is it tighter because we don't need to worry about the the non-fixed nature of ...) I could try a couple of different bounds and see what happens.
+
+\subsection{UCB variant}
+In this case the estimators from section XX are biased - so its a lot harder to prove a regret bound.
+
+\begin{theorem} (Probably False) With probability at least $1 - \delta$ we have that:
 $\displaystyle \left|\hat{\mu}_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
 \caption{The distribution of regret varies with the $\beta$ parameter in the bound in the estimator.
@@
 \section{Conclusion}
+
+\newpage
+\section{Notes}
+
+\subsection{Why Hoeffdings bound doesn't hold if $n$ is a random variable}
+
+Let $\{Z_1,Z_2...Z_n\} \sim Bernoulli(\frac{1}{2}) $ and $X_i = 2Z_i-1 \implies X_i \in \{-1,1\} \text{ and } E[X_i] = 0$.
+
+For a fixed $n$, Hoeffdings inequality says: 
+\begin{equation}
+\label{eq:HoeffdingsRandom}
+\begin{aligned}
+& P\left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \epsilon\right) \leq 2e^{-n\epsilon^2} \\
+\implies & P \left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \sqrt{\frac{\log{4}}{n}}\right) \leq \frac{1}{2} \\
+\implies & P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \leq \frac{1}{2}\\
+\end{aligned}
+\end{equation}
+
+If $n$ is allowed to be dependent on the sequence of values sampled, this inequality no longer holds.
+
+\begin{proof}
+Choose $n$ based on the sequence of samples seen so far such that:
+\eq {
+n = min \{ n: n > 4 \text{ and } \sum_{i=1}^{n}X_i > \sqrt{n \log \log n}\}
+}
+By the law of iterated logarithms this quantity is finite with probability $\sim 1$.
+
+\eq {
+& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{\log{n}}}\right) \sim 1 \\
+& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \sim 1 \text{, Since } \log{\log{n}} > \log{4} \;\; \forall n > 4 \\
+}
+
+Thus Hoeffdings inequality (equation \ref{eq:HoeffdingsRandom}) does not hold in general if $n$ is not independent of the samples $\{X_i\}$ (The bound does not work if you decide to stop sampling as soon as you reach a point where random walk fluctuations take you outside it)
+
+\end{proof}
 

### 2014-11-26 causal-bandit/ucb_regret.tex
adds sketch for ucb regret bound proof

@@
 
-Assume $\exists$ a convex function, $\psi$, such that:
+Assume for each arm $i \in \{1...K\} $ there is an unknown distribution of rewards $P(X)$ and a convex function, $\psi$, such that:
 
@@
 \end{equation}
-This ensures that the moments of the distribution of X are defined. Assume that at time $t$ we select arm $I_t$ with the highest upper confidence bound:
+This ensures that the moments of the distribution of $X$ are defined. If we select arm $i$ a fixed number of times $s$:
+\eqn {
+\label{eq:hoeff1}
+& P(|\hat \mu_{is} - \mu_{i}| > \epsilon) \leq 2e^{-s\psi^*(\epsilon)} \\
+\label{eq:hoeff2}
+ \implies & P(|\hat \mu_{is} - \mu_{i}| > (\psi^{*})^{-1}\frac{\log(\frac{2}{\delta})}{s}) \leq \delta
+}
+
+Assume that at time $t$ we select arm $I_t$ with the highest upper confidence bound:
 
@@
 \label{eq:armSelection}
-I_{t} = argmax_{i=1...K}\left[\hat{\mu}_{it} + (\psi^{*})^{-1}\frac{\alpha log(t)}{T_{it}}\right]
+I_{t} = argmax_{i=1...K}\left[\hat{\mu}_{it} + (\psi^{*})^{-1}\frac{\alpha \log(t)}{T_{it}}\right]
 \end{equation}
@@
 \begin{enumerate}
-\item The estimated UCB on the best arm, $i^*$, is less than or equal to the actual reward for that arm: $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \hat \mu$
+\item The estimated UCB on the best arm, $i^*$, is less than or equal to the actual reward for that arm: $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu^*$
 \item The estimated reward for arm $i$ is greater than or equal to the estimated CI higher than the true reward for that arm: $\hat \mu_{it} \geq \mu_i + \hat \epsilon_{it}$
-\item The number of times we have selected arm $i$ in previous timesteps, $T_{it}$, is less than some bound (that grows logarithmically with $n$). $T_{it} < \frac{\alpha log(n)}{\psi^*(\Delta i/2)}$
+\item The number of times we have selected arm $i$ in previous timesteps, $T_{it}$, is less than some bound (that grows logarithmically with $n$).
+$T_{it} < \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \;\;\; \color{red} \longleftarrow \text{feels odd that this grows with $n$, not $t$}$
 \end{enumerate}
@@
   & \implies \Delta i > 2 (\psi^*)^{-1} \frac{\alpha log(n)}{T_{it}} \geq 2 (\psi^*)^{-1} \frac{\alpha log(t)}{T_{it}} = 2 \hat \epsilon_{it}\\
-  & \implies \Delta i > 2 \hat \epsilon_{it} \\
-1. & \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \hat \mu  = \mu_i+\Delta i > \mu_i +  2 \hat \epsilon_{it}\\
-& \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \mu_i +  2 \hat \epsilon_{it}\\
+  & \implies \Delta i > 2 \hat \epsilon_{it} \\[10pt]
+1. & \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  >  \mu^*  = \mu_i+\Delta i > \mu_i +  2 \hat \epsilon_{it}\\
+& \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \mu_i +  2 \hat \epsilon_{it}\\[10pt]
 2. & \implies \hat \mu_{it} < \mu_i + \hat \epsilon_{it} \\
    & \implies \hat \mu_{it} + \hat \epsilon_{it} < \mu_i +2 \hat \epsilon_{it} \\
-   & \implies \hat \mu_{it} + \hat \epsilon_{it} <  \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \longleftarrow \text { UCB for arm $i < $ UCB for arm $i^*$, which contradicts equation (\ref{eq:armSelection}})
+   & \implies \hat \mu_{it} + \hat \epsilon_{it} <  \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \longleftarrow \text { UCB for arm $i < $ UCB for arm $i^*$, which contradicts $i \neq i^*$}
 }
@@
 
-If statements 1. and 2. are both false, then statement 3. places a bound on the number of times we can previously have selected the incorrect arm $i$ in order to select it in this timestep.
+If statements (1) and (2) are both false, then statement (3) places a bound on the number of times we can previously have selected the incorrect arm $i$ in order to select it in this timestep.
+We can write the regret in terms of the number of times we select each arm and its sub-optimality:
+
+\eq {
+\bar{R}_n = & n\mu^* - \sum_{t=1}^{n}E[\mu_{I_t}] \\
+= & \sum_{i=1}^{K}\Delta_i E[T_{in}]\\
+= & \sum_{i=1}^{K}\Delta_i E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i\}\right] \;\; \leftarrow \text{Expected number of times selected arm $I_t$ is $i$ }\\
+}
+
+Let $\gamma = \ceil{\frac{\alpha \log(n)}{\psi^*(\Delta i/2)}}$ and suppose we had selected arm $i$ in all timesteps until $\gamma$.
+In the remaining timesteps, we can only select $i$ if statement 3) is false
+
+\eq{
+\implies E\left[T_{in}\right] & \leq \gamma + E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i \text{ and (3) is false} \}\right] \\[3pt]
+& \leq \gamma + E\left[\sum_{t=\gamma+1}^{n} \mathds{1} \{\text{(1) or (2) is true} \}\right] \leftarrow \text{since if (3) is false, (1) or (2) must be true}\\[3pt]
+& \leq \gamma + \sum_{t=\gamma+1}^{n} \left[ \mathds{P}(\text{(1) is true}) + \mathds{P}(\text{(2) is true})  \}\right] \color{red} \leftarrow \text{Bubeck has $=$ here but are (1) and (2) disjoint?} 
+}
+
+\eq {
+P(\text{(1) is true})  = & P(\hat{\mu}_{i^*t}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{t} \right) \leq \mu^*)\\
+\leq & P(\exists s \in \{1...t\} : \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right) \leq \mu^*) \color{red} \leftarrow \text {to get around the problem that t is random}\\
+\leq & \sum_{s=1}^t P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right) \leq \mu^* \right) \color{red} \leftarrow \text{union bound}\\
+\text{ From equation (\ref{eq:hoeff2}) we have:}\\
+& P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{ \log{\frac{1}{\delta}}}{s} \right) \leq \mu^* \right) < \delta\\
+\text{Let } \delta = t^{-\alpha} \implies & P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right)\leq \mu^* \right) < t^{-\alpha}\\
+\implies & P(\text{(1) is true}) \leq \sum_{s=1}^t t^{-\alpha} = t*t^{-\alpha} =  t^{1-\alpha}
+}
+
+Similarly, $P(\text{(2) is true}) \leq t^{1-\alpha} \implies E[T_{in}] \leq \gamma + \sum_{t=\gamma+1}^n 2t^{1-\alpha}$
 

### 2014-11-26 causal-bandit/bad_algorithm.py

@@ -1,0 +1,75 @@
+from math import *
+import random
+
+def randb(p):
+    """ returns 1 with probability p and 0 with probability (1-p)"""
+    r = random.random()
+    return int(r < p)
+
+class Estimate(object):
+    def __init__(self,delta):
+        self.m = 1
+        self.n = 1
+        self.delta = delta
+        self.w = w
+    def update(self,value):
+        self.n+=1
+        self.m+=value
+    def get(self):
+        return self.m/float(self.n)
+
+    def count(self):
+        return self.n
+
+    def e(self):
+        return sqrt((1/(2.0*self.n))*log(2.0/self.delta))
+
+
+def n_a(v1,v0,p):
+    return .5*min(v1.count()/float(p),v0.count()/float(1-p))
+
+def weights(v21,v20,v31,v30,p2,p3):
+    na2 = n_a(v21,v20,p2)
+    na3 = n_a(v31,v30,p3)
+    t = float(na2+na3)
+    return (na2/t,na3/t)
+
+def algorithm(v21,v20,v31,v30,p2,p3):
+    """ runs the algorithm for a given step, chosing some action to take and doing updates. """
+    #### insert action selection to make it as biased as possible
+    
+
+    ####
+
+    v2 = v21.get()*p+v20.get()*(1-p)
+    v3 = v31.get()*p+v30.get()*(1-p)
+    w2,w3 = weights(v21,v20,v31,v30,p2,p3)
+    v = w2*v2+w3*v3
+    return v
+
+
+# We can calculate the expectation of v2, and v3 after t timesteps (for a given algorithm)
+# We can calculate how far the true estimate is from the actual one. 
+
+
+delta = .05
+p2 = 0.5
+p3 = 0.5
+v21 = Estimate(delta)
+v20 = Estimate(delta)
+v31 = Estimate(delta)
+v30 = Estimate(delta)
+
+
+
+ 
+
+n = 1000
+for i in range(n):
+    v = randb(.7)
+    v11.update(v)
+
+print v11.get()
+print v11.e()
+        
+

### 2014-12-04 causal-bandit/c-bandit.tex
adds McDiarmid inequality bounds

@@
 The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
-lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
-outperform this approach by exploiting the known causal structure of the problem.
+lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$.
+
+In this instance we can significantly
+outperform the standard approach by exploiting the known causal structure of the problem.
 
@@
 \subsection{Estimators}
+\label{sec:estimators}
 Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
@@
 \eq {
-\hat{\mu} &= \sum^{K}_{a=1}\eta_a\hat{\mu}_a=\eta_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
-\text{where:}\\
-\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
-n_{i,j} & \text{if } a = i \\
-\frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
-\end{cases}
+\hat{\mu} &= \sum^{K}_{a=1}w_a\hat{\mu}_a=w_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} w_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
 }
@@
 }
-\subsection{Random sampling variant}
+\subsection{Observe then exploit}
 
-Basic idea - explore randomly until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB).
-Bounds on regret: when exploring regret is at most 1, when exploiting regret it at most $\epsilon$.
-If the horizon is $n$, the cost of exploiting $< \epsilon n$ and the cost of exploring is $O(1/\epsilon^2)$ (this comes from how quickly we converge to within $\epsilon$ for all arms).
+Basic idea - observe until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB).
+Bounds on regret: when observing the regret for each timestep is at most 1 (since the reward is $\in [0,1]$).
+When exploiting regret it at most $\epsilon$.
+If the horizon is $n$, the cost of exploiting $< \epsilon n$ and the cost of observing is $O(1/\epsilon^2)$ (this comes from how quickly we converge to within $\epsilon$ for all arms).
 
@@
 
-At each time step we randomly select a variable $X_i$ on which to intervene and set it to $1$ with probability $p_i$.
+For each arm $X_i = j$, we have:
 
-The estimators from section XX are unbiased. So we can get an estimate of the error in the estimate by
+\eqn{
+\hat{\mu}_{ij} = \frac{m_{ij}}{n_{ij}}
+}
 
-we are pooling estimators - so what is the error in a pooled estimate - is it the same as what Tor wrote before (because he was assuming unbiased estimators?
-Or is it tighter because we don't need to worry about the the non-fixed nature of ...) I could try a couple of different bounds and see what happens.
+where
+
+\eqn{
+m_{i,j}    &= \sum_{s=1}^t \ind{X_i = j, Y = 1} \\
+n_{i,j}    &= \sum_{s=1}^t \ind{X_i = j} \\
+}
+
+and
+
+\subsection{Bounding a weighted sum of unbiased estimators with McDiarmid's Ineqaulity}
+
+McDiarmid's Inequality states:
+If $X_i \ci X_j$ and 
+\eqn{
+|\phi(X_1...X_i...X_N) - \phi(X_1...X_i^{'}...X_N)| < c_i \;\; \forall i
+}
+\eqn{
+& P\left(|\phi(\boldsymbol{X}) - E[\phi(\boldsymbol{X})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
+\label{eqn:McDiarmid}
+& P\left(|\phi(\boldsymbol{X}) - E[\phi(\boldsymbol{X})]| \geq 
+\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
+\right) \leq \delta
+}
+
+In our problem we have:
+
+\eqn{
+\hat{\mu}_a = \begin{cases}
+\frac{p_a}{n_{a1}}\sum_{j=1}^{n_{a1}}X_{a1,j} + \frac{1-p_a}{n_{a0}}\sum_{j=1}^{n_{a0}}X_{a0,j} & \text{ if $a \neq i$}\\
+\sum_{j=1}^{n_i}X_j & \text{if $a = i$}
+\end{cases}
+}
+
+\eqn {
+\mu = \sum_a w_a \mu_a \text{ where } \sum_a w_a = 1
+}
+
+Let
+
+\eqn {
+\hat{\mu} = \phi(\boldsymbol{X}) = \sum_{a \neq i} w_a \left[ 
+\frac{p_a}{n_{a1}}\sum_{j=1}^{n_{a1}}X_{a1,j} + \frac{1-p_a}{n_{a0}}\sum_{j=1}^{n_{a0}}X_{a0,j}
+\right] + \frac{w_i}{n_i} \sum_{j=1}^{n_i}X_{i,j}
+}
+
+Note: this treats the number of samples $n_i$,$n_{a1}$, and $n_{a0}$ as fixed.
+If the choice of action is random they are in fact random variables (but do not depend on previous action choices).
+Will this bounds still hold?
+
+\eqn {
+&\phi(...X_{\alpha 1,j}...) - \phi(...X_{\alpha 1,j}^{'}...) \leq w_\alpha \frac{p_\alpha}{n_{\alpha 1}} \\
+&\phi(...X_{\alpha 0,j}...) - \phi(...X_{\alpha 0,j}^{'}...) \leq w_\alpha \frac{1-p_\alpha}{n_{\alpha 0}}\\
+& \phi(...X_i...) - \phi(...X_i^{'}...) \leq \frac{w_i}{n_i}
+}
+
+\eqn{
+\sum_i c_i^2 = & n_i \left(\frac{w_i}{n_i}\right)^2+ \sum_a \left[ n_{a1}\left(w_a \frac{p_a}{n_{a1}}\right)^2 + n_{a0}\left(w_a \frac{1-p_a}{n_{a0}}\right)^2\right]\\
+= & \frac{w_i^2}{n_i}+ \sum_a w_a^2 \left(\frac{p_a^2}{n_{a1}}+  \frac{(1-p_a)^2}{n_{a0}} \right) \\
+\label{eq:tominimize}
+= & \sum_a w_a^2 f(a) \text{, where }\\ 
+f(a) =& \begin{cases}
+\frac{p_a^2}{n_{a1}}+  \frac{(1-p_a)^2}{n_{a0}}  & a \neq i\\
+\frac{1}{n_i} & a = i
+\end{cases}
+}
+
+We want to choose weights $\boldsymbol{w}$ so as to minimize \ref{eq:tominimize} subject to the constraint $\sum_a w_a = 1$.
+
+The minimum (assuming it exists) should occur at a critical point of:
+
+\eqn {
+L(w_1...w_k,\lambda) = \sum_a w_a^2 f(a) + \lambda((\sum_a w_a)-1)
+}
+
+\eqn{
+& \frac{\partial L}{\partial w_a}  = 2 w_a f(a) + \lambda = 0 \\
+& \implies w_a =  \frac{-\lambda}{2 f(a)}\\
+& \frac{\partial L}{\partial \lambda}  = (\sum_a w_a)-1= 0 \\
+& \implies -\frac{\lambda}{2}\sum_a \frac{1}{f(a)} = 1
+ \implies \lambda = -\frac{2}{\sum_a \frac{1}{f(a)}}\\
+& \implies w_a = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
+\label{eqn:sumofsquares}
+& \implies \sum_i c_i^2 = \frac{1}{\sum_a \frac{1}{f(a)}} \\
+}
+
+Substituting \ref{eqn:sumofsquares} into the McDiarmid inequality \ref{eqn:McDiarmid}:
+
+\eqn{
+P\left(|\hat{\mu} - \mu| > \sqrt{\frac{1}{2\sum_a \eta_a} \log{\frac{2}{\delta}}}\right) \leq \delta
+}
+
+where:
+
+\eqn {
+\eta_a = \begin{cases}
+\frac{n_{a1}n_{a0}}{n_{a1}(1-p_a)^2+n_{a0}p_a^2} & a \neq i\\
+n_i & a = i
+\end{cases}
+}
+
+\begin{figure}[H]
+\centering
+\caption{The effective number of samples $\eta_a$ versus $n_{a1}$, where $p_a = .8$ and the total number of samples,  $ n_{a1}+n_{a0}=100$.
+The effective number of samples is maximized (and equals the total) if we sample each side according to its probability.}
+\includegraphics[width=.5\textwidth]{samplehalves}
+\end{figure}
+
+\subsection{Targeted Sampling}
+Sample a equal number of points from each half of each arm - until the error on all arms is smaller than some threshold.
+Then exploit.
+This should do better than observe-exploit, particularly where the probability of some events occurring without intervention is low.
 
 \subsection{UCB variant}
-In this case the estimators from section XX are biased - so its a lot harder to prove a regret bound.
+Combine estimators according to:
+
+\eq {
+w_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
+n_{i,j} & \text{if } a = i \\
+\frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
+\end{cases}
+}
+
+In this case the estimators from section \ref{sec:estimators} are biased - so its a lot harder to prove a regret bound.
 
@@
 
-\begin{figure}[h]
-\caption{Testing values of beta - 100 experiments for each value of beta, with a 4 armed bandit.}
-\includegraphics[width=0.7\textwidth]{regretvsbeta.pdf}
-\end{figure}
+\eqn {
+P(|\hat{\mu}_{a0} -\mu_{a0}| > \epsilon_{a0}) \leq \delta/2 \;\; \text{ where } 
+\epsilon_{a0}=\sqrt{\frac{1}{2n_{a0}}\log{\frac{4}{\delta}}}\\ 
+P(|\hat{\mu}_{a1} -\mu_{a1}| > \epsilon_{a1}) \leq \delta/2 \;\; \text{ where } 
+\epsilon_{a1}=\sqrt{\frac{1}{2n_{a1}}\log{\frac{4}{\delta}}}\\ 
+}
 
-\begin{figure}[h]
-\caption{Shows relationship between the regret and the number of bandit arms $2K$ for the standard UCB algorithm and our modified algorithm.  }
-\label{fig:regretvsarms}
-\includegraphics[width=0.9\textwidth]{regretvsarms.pdf}
-\end{figure}
+We should not need $t$ in here since, although $n_{aj}$ is still random, its not dependent on previous results.
 
-\begin{figure}[h]
-\caption{Regret vs number of arms with horizon $100$}
-\label{fig:regretvsarmsshorthorizon}
-\includegraphics[width=0.8\textwidth]{regregvsamrs100.pdf}
-\end{figure}
+Combining these two halves and applying the union bound gives:
+
+\eqn{
+P\left(|\hat{\mu}_a - \mu | > \sqrt{\frac{1}{2} \log{\frac{4}{\delta}}}
+\left(p_a \sqrt{\frac{1}{2n_{a1}}}+(1-p_a) \sqrt{\frac{1}{2n_{a0}}} \right)\right) \leq \delta
+}
+
+I now need to pool these estimators for each $a \in \{1...K\}/I$ but I can't just use a union bound - I need the bound to shrink by combining the multiple estimates.
+
+For some distributions,

### 2015-01-15 causal-bandit/c-bandit.tex
adds final version of WIML poster and first draft of observe approach bandit in simplest setting

@@
 
-Basic idea - observe until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB).
+Basic idea - observe until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB). This is only really going to work if $p_a \sim 0.5$.
+
 Bounds on regret: when observing the regret for each timestep is at most 1 (since the reward is $\in [0,1]$).
@@
 
-and
+and Hoeffding's Inequality gives:
+
+\eqn {
+P\left(|\hat{\mu}_{ij} - \mu_{ij}| > \sqrt{\frac{1}{2n_{ij}}\log{\frac{2}{\delta}}}\right) \leq \delta
+}
+
+\section{Observe until number of plausibly optional arms $< \alpha$}
+
+Define the set of plausibly optimal arms to be those with an upper confidence bound that is higher than the largest lower confidence bound.
+
+\eqn {
+A_t = \{k : UB(k) \geq max_{k'}(LB(k'))  \}
+}
+
+We then observe until the size of this set is less than some value $\alpha$ before switching to UCB.
+This takes into account that we don't need narrow confidence bounds on the arms with very low expected reward.
+Once we start doing UCB, we keep track of the ucb bound separately and for each arm use whichever bounds are narrower.
+
+\subsection{Regret during observe phase}
+
+\eqn {
+R_n = & \max_{i=1...K} E\left[\sum_{t=1}^n Y_{it} - \sum_{t=1}^n Y_{I_t,t} \right] \\
+= & E[n]\left(P(Y|X_{i^*}={j^*})  - P(Y)\right) \text{ where } X_{i^*}={j^*} \text{ is the best arm}
+}
+
+\subsubsection{A very simple model}
+
+Lets try to analyse a very simple model in which the reward depends only on the value of $X_1$.
+
+\eqn{
+&P(Y|X_1=0,X_2,...X_N) = \frac{1}{2}-a\\
+&P(Y|X_1=1,X_2,...X_N) = \frac{1}{2}+a\\
+&P(X_i = j) = \frac{1}{2} \;\;\; \forall (i,j)\\
+&\implies P(Y) = \frac{1}{2} \\
+&\implies R_n = aE[n]
+}
+
+The goal is now to find $E[n]$ (which will be a function of $a$)
+
+Lets start by considering how long on average we need to observe until the lower bound on the optimal arm is higher than the upper bound on a single other arm, $i$, (with expected reward $\frac{1}{2}$)
+
+If the following three statements hold, then there is no overlap $LB^* = \hat{\mu}^* - \epsilon > UB_i = \hat{\mu}_i + \epsilon$.
+
+\eqn {
+\label{eq:overlap1}
+&\mu^* - \hat{\mu}^* \leq \epsilon \\
+\label{eq:overlap2}
+&\hat{\mu}_i - \mu_i \leq \epsilon \\
+\label{eq:overlap3}
+&\epsilon \leq a/4
+}
+
+If \ref{eq:overlap3} is true, we can bound the probability of overlap at some timestep $t$ with the union bound of the probability that either \ref{eq:overlap1} or \ref{eq:overlap2} are false.
+
+\eqn {
+P(overlap) = P(n \geq t) \leq 
+P(\mu^* - \hat{\mu}^* \geq \epsilon) + P(\hat{\mu}_i - \mu_i \geq \epsilon)  \;\;\;\text{ if } \epsilon \leq a/4
+}
+
+Let $\epsilon = a/4$
+
+\eqn {
+P(n > t) \leq & e^{-nD(p_{ij}+p (a/4) || p_{ij})} +e^{-nD(p_{ij}-p (a/4) || p_{ij})} \\
+\leq & \;\;\;?
+}
+
+Where $p_ij = P(Y = 1, X_i = j) \leq P(X_i=j) = p \;\;\; \forall(i,j)$
+
+Now
+
+\eqn {
+E[n] = \sum_{t=1}^\infty P(n \geq t)
+}
+
+But this is just the expected number of steps until one arm does not overlap. With more arms its more complicated - as we need the probability that $\alpha$ of them overlap (another union bound?).
 
@@
 
-In our problem we have:
+In our problem, for each variable $V_i$ and value $j$ we have:
 
 \eqn{
-\hat{\mu}_a = \begin{cases}
-\frac{p_a}{n_{a1}}\sum_{j=1}^{n_{a1}}X_{a1,j} + \frac{1-p_a}{n_{a0}}\sum_{j=1}^{n_{a0}}X_{a0,j} & \text{ if $a \neq i$}\\
-\sum_{j=1}^{n_i}X_j & \text{if $a = i$}
+\hat{\mu}_a = \tilde{P}(Y|V_i = j)= \begin{cases}
+\frac{p_a}{n_{a1}}\sum_{s=1}^{n_{a1}}X_{a1,s} + \frac{1-p_a}{n_{a0}}\sum_{s=1}^{n_{a0}}X_{a0,s} & \text{ if $a \neq i$}\\
+\sum_{s=1}^{n_i}X_s & \text{if $a = i$}
 \end{cases}
@@
 
+where $X_{a1,s}  \sim P(Y|V_i = j, do(V_a = 1))$ and $X_{a0,s}  \sim P(Y|V_i = j, do(V_a = 0))$, both $\in [0,1]$. We then pool $\{\hat{\mu}_1 ... \hat{\mu}_K \}$ to get a single estimate.
+
 \eqn {
-\mu = \sum_a w_a \mu_a \text{ where } \sum_a w_a = 1
+\hat{\mu} = \tilde{P}(Y|V_i = j)= \sum_a w_a \hat{\mu}_a \text{, where } \sum_a w_a = 1
 }
@@
 \hat{\mu} = \phi(\boldsymbol{X}) = \sum_{a \neq i} w_a \left[ 
-\frac{p_a}{n_{a1}}\sum_{j=1}^{n_{a1}}X_{a1,j} + \frac{1-p_a}{n_{a0}}\sum_{j=1}^{n_{a0}}X_{a0,j}
-\right] + \frac{w_i}{n_i} \sum_{j=1}^{n_i}X_{i,j}
+\frac{p_a}{n_{a1}}\sum_{s=1}^{n_{a1}}X_{a1,s} + \frac{1-p_a}{n_{a0}}\sum_{s=1}^{n_{a0}}X_{a0,s}
+\right] + \frac{w_i}{n_i} \sum_{s=1}^{n_i}X_{i,s}
 }
@@
 Note: this treats the number of samples $n_i$,$n_{a1}$, and $n_{a0}$ as fixed.
-If the choice of action is random they are in fact random variables (but do not depend on previous action choices).
-Will this bounds still hold?
+Even for fixed actions, the latter two are still random variables that depend on $\boldsymbol{p}$  Will this bound still hold?
 
 \eqn {
-&\phi(...X_{\alpha 1,j}...) - \phi(...X_{\alpha 1,j}^{'}...) \leq w_\alpha \frac{p_\alpha}{n_{\alpha 1}} \\
-&\phi(...X_{\alpha 0,j}...) - \phi(...X_{\alpha 0,j}^{'}...) \leq w_\alpha \frac{1-p_\alpha}{n_{\alpha 0}}\\
+&\phi(...X_{\alpha 1,s}...) - \phi(...X_{\alpha 1,s}^{'}...) \leq w_\alpha \frac{p_\alpha}{n_{\alpha 1}} \\
+&\phi(...X_{\alpha 0,s}...) - \phi(...X_{\alpha 0,s}^{'}...) \leq w_\alpha \frac{1-p_\alpha}{n_{\alpha 0}}\\
 & \phi(...X_i...) - \phi(...X_i^{'}...) \leq \frac{w_i}{n_i}
@@
 \label{eqn:sumofsquares}
-& \implies \sum_i c_i^2 = \frac{1}{\sum_a \frac{1}{f(a)}} \\
+& \implies \sum_i c_i^2 = \frac{1}{\sum_a \frac{1}{f(a)}} 
 }
@@
 \eqn{
-P\left(|\hat{\mu} - \mu| > \sqrt{\frac{1}{2\sum_a \eta_a} \log{\frac{2}{\delta}}}\right) \leq \delta
+P\left(|\hat{\mu}^{ij} - \mu^{ij}| > \sqrt{\frac{1}{2\sum_a \eta_a^{ij}} \log{\frac{2}{\delta}}}\right) \leq \delta
 }
@@
 \eqn {
-\eta_a = \begin{cases}
-\frac{n_{a1}n_{a0}}{n_{a1}(1-p_a)^2+n_{a0}p_a^2} & a \neq i\\
-n_i & a = i
-\end{cases}
+& \mu^{ij} = P(Y | V_i = j) \\
+& \eta_a^{ij} = \begin{cases}
+\frac{n_{a1}^{ij} n_{a0}^{ij}}{n_{a1}^{ij}(1-p_a)^2+n_{a0}^{ij}p_a^2} & a \neq i\\
+n_{ij} & a = i
+\end{cases} \\
+&n_{al}^{ij} = \sum_{s=1}^t \ind{do(V_a = l), V_i = j} \\
+& n_{ij} = \sum_{s=1}^t \ind{(do(V_i = j)}
+}
+
+All of the above is to get estimates for $P(Y|V_i = j)$ for some fixed $i,j$.
+Suppose we explore for some fixed total number of rounds $h$, the goal is to minimize the worst confidence bound - that doesn't quite make sense as a goal - we just need to be sure that all actions are worse than our estimate of the best with probability $\delta$.
+...
+anyway
+
+We have control of are the number of times we do each action, $n_{al}$, which obviously also influences the random variables $n_{al}^{ij}$, subject to the constraint $\sum n_{al} = h$.
+
+Objective: minimize the expectation of the largest confidence bound after a fixed number of rounds $n$.
+
+Let $\boldsymbol{n} = [n_{11},n_{10},n_{21},n_{20}...n_{k1},n_{k0}]$
+
+\eqn{
+\tilde{\boldsymbol{n}} = \argmin_{\{\boldsymbol{n}:\|\boldsymbol{n}\|=h\}} \max_{i,j} \left( E \left[ \sqrt{\frac{1}{2\sum_a \eta_a^{ij}} \log{\frac{2}{\delta}}} \right] \right)
+}
+
+We can bound the first part of $\eta_a^{ij}$
+
+\eqn {
+\eta_a^{ij} = \frac{n_{a1}^{ij}n_{a0}^{ij}}{n_{a1}^{ij}(1-p_a)^2+n_{a0}^{ij}p_a^2} & \leq \frac{n_{a1}^{ij}n_{a0}^{ij}}{\max(n_{a1}^{ij}(1-p_a)^2+n_{a0}^{ij}p_a^2)} \\
+& = \min \left( \frac{n_{a0}^{ij}}{(1-p_a)^2}, \frac{n_{a1}^{ij}}{p_a^2} \right)
+}
+
+\eqn {
+E \left[ \min \left( \frac{n_{a0}^{ij}}{(1-p_a)^2}, \frac{n_{a1}^{ij}}{p_a^2} \right)  \right] & \leq 
+\min \left( E\left[\frac{n_{a0}^{ij}}{(1-p_a)^2} \right], E\left[\frac{n_{a1}^{ij}}{p_a^2} \right] \right) \\
+& = P(V_i = j) *\min \left( \frac{n_{a0}}{(1-p_a)^2} , \frac{n_{a1}}{p_a^2}  \right) \\
 }
@@
 \subsection{Targeted Sampling}
-Sample a equal number of points from each half of each arm - until the error on all arms is smaller than some threshold.
+Sample a equal number of points from each arm - until the error on all arms is smaller than some threshold.
 Then exploit.
 This should do better than observe-exploit, particularly where the probability of some events occurring without intervention is low.
+
+A more sophisticated version:
 
@@
 For a fixed $n$, Hoeffdings inequality says: 
-\begin{equation}
+\eqn{
 \label{eq:HoeffdingsRandom}
-\begin{aligned}
 & P\left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \epsilon\right) \leq 2e^{-n\epsilon^2} \\
 \implies & P \left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \sqrt{\frac{\log{4}}{n}}\right) \leq \frac{1}{2} \\
-\implies & P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \leq \frac{1}{2}\\
-\end{aligned}
-\end{equation}
+\implies & P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \leq \frac{1}{2}
+}
 

### 2015-01-15 causal-bandit/importance_sampling.tex
adds final version of WIML poster and first draft of observe approach bandit in simplest setting

--- 2014-11-26 causal-bandit/c-bandit.tex
@@
 
-\title{Intervention Bandits}
-\author{Blah blah}
+\title{Estimators}
 
@@
 
-\begin{abstract}
-An abstract.
-\end{abstract}
+Suppose we have samples $(X_t,Y_t)$ for $t = 1... n$ drawn from the joint distribution $P(X,Y)$ and we want to estimate $\mu = P(Y=1|X=1)$
 
-\section{Introduction}
+\section{The standard estimator}
 
-Useful references are: \cite{BC12}.
-
-\section{Notation}
-
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
-
-Let $p \in [0,1]^K$ be a fixed and known vector. In each time-step $t$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
-\end{enumerate}
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
+\eqn{
+\hat{\mu} = \frac{\sum_{t=1}^n Y_t \ind{X_t=1}}{\sum_{t=1}^n\ind{X_t=1}}
 }
 
-\section{Estimating $\mu_{i,j}$}
-The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
-lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
-outperform this approach by exploiting the known causal structure of the problem.
+We can't actually say that this estimator is unbiased - as it can be undefined if $n = 0$
 
-\eq {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
+\color{red}{Is there an unbiased estimator for this simple problem if the $P(X = 1)$ is not known?}\color{black}
+
+Let's consider variance and bounds.
+
+\section{The importance sampling estimator}
+If we know the probability that $X=1$ we can instead use importance sampling.
+
+\eqn{
+\hat{\mu} = \frac{1}{n}\sum_{t=1}^n\frac{Y_t \ind{X_t=1}}{P(X_t=1)}
 }
 
-\subsection{Estimators}
-Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
+The estimator is unbiased:
 
-Let $\hat{\mu}_a$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $X_a$.
-
-\eq {
-\hat{\mu}_a &=
-\begin{cases}
-\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
-\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
-\end{cases}\\
-\text{where:}\\
-m_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
-n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s \\
+\eqn{
+E[\hat{\mu}] = &\frac{1}{n}\sum_{t=1}^n E\left[\frac{Y_t \ind{X_t=1}}{P(X_t=1)}\right] = \frac{1}{np}\sum_{t=1}^n E\left[Y_t \ind{X_t=1}\right] \\
+= & \frac{1}{n} \sum_{t=1}^n \frac{\left( 0*P(Y_t=0 \text{ or } X_t=0) + 1*P(Y_t = 1 \text{ and } X_t=1) \right)}{P(X_t=1)} \\
+= & \frac{1}{n}\sum_{i=1}^n\frac{P(Y=1,X=1)}{P(X=1)} = \frac{1}{n}\sum_{i=1}^{n}P(Y=1|X=1) = \frac{1}{n} \sum_{i=1}^n \mu = \mu
 }
 
-This gives $K$ estimators $\{\hat{\mu_1}...\hat{\mu_K}\}$ to be pooled into a single estimator $\hat{\mu}$.
+We can use Hoeffding's inequality to get a bound on how far the estimator is likely to be from the true value. Let:
 
-\eq {
-\hat{\mu} &= \sum^{K}_{a=1}\eta_a\hat{\mu}_a=\eta_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
-\text{where:}\\
-\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
-n_{i,j} & \text{if } a = i \\
-\frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
-\end{cases}
+\eqn{
+Z_t = \frac{Y_t \ind{X_t=1}}{p} \in \{0,\frac{1}{p}\}
 }
 
-If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p}_a$
+\textbf{Hoeffding's inequality:} If $X_1...X_n$ are independent observations such that $a_i < X_i < b_i$ and $\bar{X} = \frac{1}{n}\sum_{i=1}^n X_i$ then:
 
-\eq{
-\hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
+\eqn {
+P(\bar{X} - E[\bar{X}] \geq \epsilon) \leq \exp{\frac{-2n^2\epsilon^2}{\sum_{i=1}^n(b_i - a_i)^2}}
 }
-\subsection{Random sampling variant}
 
-Basic idea - explore randomly until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB).
-Bounds on regret: when exploring regret is at most 1, when exploiting regret it at most $\epsilon$.
-If the horizon is $n$, the cost of exploiting $< \epsilon n$ and the cost of exploring is $O(1/\epsilon^2)$ (this comes from how quickly we converge to within $\epsilon$ for all arms).
+In our case this gives:
+\eqn{
+P(\frac{1}{n}\sum_{i=1}^n Z_t - \mu \geq \epsilon) \leq \exp{(-2n\epsilon^2p^2)} 
+\implies P(\frac{1}{n}\sum_{i=1}^n Z_t - \mu \geq \frac{1}{p}\sqrt{\frac{1}{2n}\log{\frac{1}{\delta}}}) \leq \delta
+}
+
+This is not so good because $\frac{1}{p}$ can be very large if $p$ is small and its outside the log - so the bounds will grow quickly as $p$ gets small.
+
+We can get a tighter bound by using an Chernoff's inequality that takes account of the variance of $X$. Let $W_t = \ind{Y_t = 1,X_t=1} = pZ_t$, then:
 
 \eqn{
-R_n = O(n\epsilon+\frac{1}{\epsilon^2})
+P(\frac{1}{n}\sum_{i=1}^n Z_t - \mu \geq \epsilon) = P(\frac{1}{n}\sum_{i=1}^n pZ_t - p\mu \geq p\epsilon) = P(\frac{1}{n}\sum_{i=1}^n W_t - p\mu \geq p\epsilon)
 }
 
-Differentiating and selecting $\epsilon = (\frac{2}{n})^{1/3}$ to minimize the regret yields:
-
+Now $W_t$ is a bernoulli random variable so:
 \eqn{
-R_n = O(n^{2/3})
+V[W_t] = P(W_t=1)P(W_t=0) \leq P(W_t=1) = P(Y_t=1,X_t=1) \leq P(X_t=1) = p
 }
 
-At each time step we randomly select a variable $X_i$ on which to intervene and set it to $1$ with probability $p_i$.
-
-The estimators from section XX are unbiased. So we can get an estimate of the error in the estimate by
-
-we are pooling estimators - so what is the error in a pooled estimate - is it the same as what Tor wrote before (because he was assuming unbiased estimators?
-Or is it tighter because we don't need to worry about the the non-fixed nature of ...) I could try a couple of different bounds and see what happens.
-
-\subsection{UCB variant}
-In this case the estimators from section XX are biased - so its a lot harder to prove a regret bound.
-
-\begin{theorem} (Probably False) With probability at least $1 - \delta$ we have that:
-$\displaystyle \left|\hat{\mu}_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
-where $\beta > 0$ is some constant.
-\end{theorem}
-
-\begin{proof}
-First note that $n_{a,b}$ is a random variable that is bounded by $t$ for all $a,b$.
-We use the short-hand $\mu^{a,b}_{i,j} = \E[q(X) | X_i = j, X_a = b]$. Then
-\eq{
-\mu_{i,j} = p_a \mu_{i,j}^{a,1} + (1 - p_a) \mu_{i,j}^{a,0}\,.
-}
-Now we can apply Hoeffding's bound and the union bound to show that
-\eq{
-\P{\left|\frac{m_{a,b}}{n_{a,b}} - \mu_{i,j}^{a,b}\right| \geq \sqrt{\frac{1}{2n_{a,b}} \log \frac{4t}{\delta}}} \leq \frac{\delta}{2}\,.
-}
-Therefore by the union bound
-\eq{
-\P{\left|p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}} - \mu_{i,j}\right| \geq p_a \sqrt{\frac{1}{2n_{a,1}} \log\frac{4t}{\delta}}
-+ (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}} \leq \delta\,
-}
-Now by Jensen's inequality 
-\eq{
-p_a \sqrt{\frac{1}{2n_{a,1}}\log\frac{4t}{\delta}} + (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}
-&\leq \sqrt{\left(\frac{p_a}{2n_{a,1}} + \frac{1-p_a}{2n_{a,0}}\right)\log \frac{4t}{\delta}} \\
-&\leq \sqrt{\max\set{\frac{p_a}{n_{a,1}}, \frac{1-p_a}{n_{a,0}}} \log\frac{4t}{\delta}} \\
-&= \sqrt{{1 \over 2n_a} \log\frac{4t}{\delta}}\,.
-}
-Similarly,
-\eq{
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{4t}{\delta}}} \leq
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{2t}{\delta}}} \leq \delta\,.
-}
-\end{proof}
-
-\section{Algorithm}
-
-\begin{algorithm}
-\caption{UCB}
-\begin{algorithmic}[1]
-\State {\bf Input:} Number of variables $K$, vector $p \in [0,1]^K$, horizon $n$
-\For{$t \in 1,\ldots,n$}
-\For{$i \in 1,\ldots,K$}
-\For{$j \in \set{0,1}$}
-\State Compute $\tilde \mu_{i,j} = \hat \mu_{i,j} + \sqrt{\frac{\alpha}{\sum_{a} n_a} \log n}$ 
-\EndFor
-\EndFor
-\State Choose $I_t, J_t = \argmax_{i,j} \tilde \mu_{i,j}$
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\section{Theorems}
-\section{Experiments}
-
-\begin{figure}[h]
-\caption{Comparison of the performance of standard UCB versus causal UCB with $\beta=3$ and $\beta = 5$.
-100 simulations were run for each algorithm up to a horizon of $10^5$ per value of $K$.
-Error bars span the 1st to 3rd quantile of the regret, round points mark the median and triangular points show the mean.
-For standard UCB the regret increases linearly with the number of arms $K$.
-For causal UCB the increase is sub-linear.
-Increasing $\beta$ leads to slower convergence but lower variance.}
-\includegraphics[width=1\textwidth]{regret_vs_K_series_beta.pdf}
-\end{figure}
-
-\begin{figure}[H]
-\caption{The distribution of regret varies with the $\beta$ parameter in the bound in the estimator.
-As beta increases, the mean regret increases but the variance decreases.
-The plot shows the results of running $100$ independent bandits, with $K=256$ and $\epsilon=0.1$, up to a horizon $h=10^5$ for each value of $\beta$.
-}
-\includegraphics[width=0.7\textwidth]{regret_vs_beta_boxplot.pdf}
-\end{figure}
-
-Simululations to compare the performance of standard UCB with our modified algorithm.
-For each number of arms, $100$ bandits of each type were created and run upto to a horizon of $1000$ timesteps.
-The mean regret and its standard error from these simulations is plotted in figure \ref{fig:regretvsarms} The true data was generated from a model where:
-
-\eq{
-&p = [0.5]^{K}\\
-&q(\boldsymbol{X}) = \begin{cases}0.5 & \text{if $X_{1}=0$}\\ 0.6 & \text{otherwise} \end{cases}
-}
-
-\section{Conclusion}
-
-\newpage
-\section{Notes}
-
-\subsection{Why Hoeffdings bound doesn't hold if $n$ is a random variable}
-
-Let $\{Z_1,Z_2...Z_n\} \sim Bernoulli(\frac{1}{2}) $ and $X_i = 2Z_i-1 \implies X_i \in \{-1,1\} \text{ and } E[X_i] = 0$.
-
-For a fixed $n$, Hoeffdings inequality says: 
-\begin{equation}
-\label{eq:HoeffdingsRandom}
-\begin{aligned}
-& P\left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \epsilon\right) \leq 2e^{-n\epsilon^2} \\
-\implies & P \left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \sqrt{\frac{\log{4}}{n}}\right) \leq \frac{1}{2} \\
-\implies & P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \leq \frac{1}{2}\\
-\end{aligned}
-\end{equation}
-
-If $n$ is allowed to be dependent on the sequence of values sampled, this inequality no longer holds.
-
-\begin{proof}
-Choose $n$ based on the sequence of samples seen so far such that:
-\eq {
-n = min \{ n: n > 4 \text{ and } \sum_{i=1}^{n}X_i > \sqrt{n \log \log n}\}
-}
-By the law of iterated logarithms this quantity is finite with probability $\sim 1$.
-
-\eq {
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{\log{n}}}\right) \sim 1 \\
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \sim 1 \text{, Since } \log{\log{n}} > \log{4} \;\; \forall n > 4 \\
-}
-
-Thus Hoeffdings inequality (equation \ref{eq:HoeffdingsRandom}) does not hold in general if $n$ is not independent of the samples $\{X_i\}$ (The bound does not work if you decide to stop sampling as soon as you reach a point where random walk fluctuations take you outside it)
-
-\end{proof}
+\color{red} Now do the same for the standard estimator \color{black}
 
@@
 \end{document}
-
-\begin{figure}[h]
-\caption{Testing values of beta - 100 experiments for each value of beta, with a 4 armed bandit.}
-\includegraphics[width=0.7\textwidth]{regretvsbeta.pdf}
-\end{figure}
-
-\begin{figure}[h]
-\caption{Shows relationship between the regret and the number of bandit arms $2K$ for the standard UCB algorithm and our modified algorithm.  }
-\label{fig:regretvsarms}
-\includegraphics[width=0.9\textwidth]{regretvsarms.pdf}
-\end{figure}
-
-\begin{figure}[h]
-\caption{Regret vs number of arms with horizon $100$}
-\label{fig:regretvsarmsshorthorizon}
-\includegraphics[width=0.8\textwidth]{regregvsamrs100.pdf}
-\end{figure}

### 2015-01-15 causal-bandit/ucb_regret.tex
adds final version of WIML poster and first draft of observe approach bandit in simplest setting

@@
 \usepackage{graphicx}
+\usepackage{parskip}
 
@@
 
-\title{Regret Bounds for UCB}
+\title{Regret Bounds for $(\alpha,\gamma)$-UCB}
 \author{Finnian Lattimore}
@@
 
-Assume for each arm $i \in \{1...K\} $ there is an unknown distribution of rewards $P(X)$ and a convex function, $\psi$, such that:
+The basic setting is the stochastic bandit problem:
+
+\begin{itemize}
+\item We have a set of arms $i \in \{1...K\}$
+\item For each arm $i$, there is an unknown distribution of rewards $P_{i}(X)$
+\item Each time we select a given arm, $i$, the reward is sampled i.i.d from $P_{i}(X)$.
+This is a big assumption - it states that the reward at a given timestep depends only on the action selected at that timestep, not on the sequence of previous actions.
+\item \color{red}Can the reward distributions be (fixed) functions of the timestep? (No I think - otherwise the idea of an optimal arm independent of t doesn't make sense)
+\end{itemize}
+
+Some notation:
+
+\begin{itemize}
+\item $K$ the number of arms
+\item $i$ identifies an arm, $i \in \{1...K\}$
+\item $i^*$ the arm with the highest true expected reward
+\item $I_t$ the arm selected by the algorithm at timestep $t$
+\item $\hat{\mu}_i$ an estimator for the expected reward of arm $i$ based on the sample mean
+\item $\mu_i$ the true expected reward for an arm $i$
+\item $\mu^*$ the true expected reward of  $i^*$ (the best arm)
+\item $\Delta_i = \mu^* - \mu_i$ how much worse arm $i$ is than the best arm
+\item $T_i(s) = \sum_{t=1}^s \ind{I_t = i}$ the number of times arm $i$ was selected upto timestep $s$
+\item $\hat \epsilon_{it} =  (\psi^*)^{-1} \left(\frac{\alpha log(t)}{T_{i}(t-1)}\right)$ an estimate of the uncertainty in the empirical estimator for the expected reward
+
+\end{itemize}
+
+The goal is to get a bound on the pseudo-regret, defined as:
+
+\eqn{
+R_n  = & n\mu^* - E\left[\sum_{t = 1}^n \mu_{It}\right] \\
+\label{eq:regretK}
+= & \sum_{i=1}^K \Delta_i E[T_i(n)] 
+}
+
+For the UCB algorithm we make the additional assumption that for each $P(X)$ there exists a convex function $\psi$ such that:
 
@@
 \end{equation}
-This ensures that the moments of the distribution of $X$ are defined. If we select arm $i$ a fixed number of times $s$:
+This ensures that the moments of the distribution of $X$ are defined and gives us a high probability bound on the how much our sample based estimate of the expected reward can be below the true expected reward.
+Let  $\hat \mu_{is} = \frac{1}{s}\sum_{t=1}^s X_t$ be the sample average and $\mu_{i} = E[P_i(X)]$ if we select arm $i$ a fixed number of times $s$:
+
 \eqn {
 \label{eq:hoeff1}
-& P(|\hat \mu_{is} - \mu_{i}| > \epsilon) \leq 2e^{-s\psi^*(\epsilon)} \\
+& P( \mu_{i} - \hat \mu_{is}  > \epsilon) \leq e^{-s\psi^*(\epsilon)} \\
 \label{eq:hoeff2}
- \implies & P(|\hat \mu_{is} - \mu_{i}| > (\psi^{*})^{-1}\frac{\log(\frac{2}{\delta})}{s}) \leq \delta
+ \implies & P\left(\mu_{i} - \hat \mu_{is} > (\psi^{*})^{-1}\left( \frac{1}{s}\log\frac{1}{\delta}\right)\right) \leq \delta
 }
 
-Assume that at time $t$ we select arm $I_t$ with the highest upper confidence bound:
+\fbox{\begin{minipage}{\textwidth}%
+\textbf{The UCB Algorithm}
+
+Define the upper confidence bound for each arm $i$ at timestep $t$ as:
+
+\eqn {
+ucb_{it} & = \hat \mu_{it} + (\psi^{*})^{-1}\left( \frac{\alpha}{T_{i}(t-1)}\log t\right) \text{ , where $\alpha$ is an input parameter} \\
+& =  \hat \mu_{it} + \hat \epsilon_{it}
+}
+
+At time $t$ select arm $I_t$ with the highest upper confidence bound:
 
@@
 \label{eq:armSelection}
-I_{t} = argmax_{i=1...K}\left[\hat{\mu}_{it} + (\psi^{*})^{-1}\frac{\alpha \log(t)}{T_{it}}\right]
+I_{t} = argmax_{i=1...K}\left(ucb_{it} \right)
 \end{equation}
@@
 \begin{equation}
-R_n \leq \sum_{i:\Delta i > 0}\left(\frac{\alpha \Delta i}{\psi^{*}(\Delta i/2)}log(n)+\frac{\alpha}{\alpha-2}\right)
+\label{eq:regretBound}
+R_n \leq \sum_{i:\Delta i > 0}\left(\frac{\alpha \Delta i \log(n)}{\psi^{*}(\Delta i/2)}+\frac{\alpha}{\alpha-2}\right)
 \end{equation}
+
+For $[0,1]$ random variables, with $\psi(\lambda) = \frac{\lambda^2}{8}$ gives $\psi^{*}(\epsilon)=2\epsilon^2$,  $(\psi^{*})^{-1}(x)=\sqrt{\frac{x}{2}}$ and:
+
+\eqn{
+ucb_{it} & = \hat \mu_{it} +  \sqrt{\frac{\alpha}{2 T_{i}(t-1)}\log t}
+}
+
+\begin{equation}
+\label{eq:Bernoulli}
+R_n \leq \sum_{i:\Delta i > 0}\left(\frac{2 \alpha \log(n)}{\Delta_i}+\frac{\alpha}{\alpha-2}\right)
+\end{equation}
+
+\end{minipage}}
+
+The confidence bound we use in the UCB algorithm is clearly related but not identical to the bound in equation \ref{eq:hoeff2}.
+The difference is due to the fact that when we use the UCB algorithm, the number of times each arm is selected is not fixed in advance but depends on the results of previous actions.
+
+We now want to prove the bound in equation \ref{eq:regretBound} holds.
 
@@
 \item The estimated reward for arm $i$ is greater than or equal to the estimated CI higher than the true reward for that arm: $\hat \mu_{it} \geq \mu_i + \hat \epsilon_{it}$
-\item The number of times we have selected arm $i$ in previous timesteps, $T_{it}$, is less than some bound (that grows logarithmically with $n$).
-$T_{it} < \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \;\;\; \color{red} \longleftarrow \text{feels odd that this grows with $n$, not $t$}$
+\item The number of times we have selected arm $i$ in previous timesteps, $T_{i}(t-1)$, is less than some bound (that grows logarithmically with $n$).
+$T_{i}(t-1) < \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \;\;\; \color{red} \longleftarrow \text{feels odd that this grows with $n$, not $t$.
+Consequence of union bound?}$
 \end{enumerate}
@@
 \eq {
-3. & \implies T_{it} > \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \\
-  & \implies \Delta i > 2 (\psi^*)^{-1} \frac{\alpha log(n)}{T_{it}} \geq 2 (\psi^*)^{-1} \frac{\alpha log(t)}{T_{it}} = 2 \hat \epsilon_{it}\\
+3. & \implies T_{i}(t-1) > \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \\
+  & \implies \Delta i > 2 (\psi^*)^{-1} \frac{\alpha log(n)}{T_{i}(t-1)} \geq 2 (\psi^*)^{-1} \frac{\alpha log(t)}{T_{i}(t-1)} = 2 \hat \epsilon_{it}\\
   & \implies \Delta i > 2 \hat \epsilon_{it} \\[10pt]
@@
 If statements (1) and (2) are both false, then statement (3) places a bound on the number of times we can previously have selected the incorrect arm $i$ in order to select it in this timestep.
-We can write the regret in terms of the number of times we select each arm and its sub-optimality:
-
-\eq {
-\bar{R}_n = & n\mu^* - \sum_{t=1}^{n}E[\mu_{I_t}] \\
-= & \sum_{i=1}^{K}\Delta_i E[T_{in}]\\
-= & \sum_{i=1}^{K}\Delta_i E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i\}\right] \;\; \leftarrow \text{Expected number of times selected arm $I_t$ is $i$ }\\
-}
 
@@
 \eq{
-\implies E\left[T_{in}\right] & \leq \gamma + E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i \text{ and (3) is false} \}\right] \\[3pt]
+\implies E\left[T_{i}(n)\right] & \leq \gamma + E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i \text{ and (3) is false} \}\right] \\[3pt]
 & \leq \gamma + E\left[\sum_{t=\gamma+1}^{n} \mathds{1} \{\text{(1) or (2) is true} \}\right] \leftarrow \text{since if (3) is false, (1) or (2) must be true}\\[3pt]
@@
 
-Similarly, $P(\text{(2) is true}) \leq t^{1-\alpha} \implies E[T_{in}] \leq \gamma + \sum_{t=\gamma+1}^n 2t^{1-\alpha}$
+Similarly, $P(\text{(2) is true}) \leq t^{1-\alpha}$ so we have proved:
+
+\eqn {
+ E\left[T_{i}(n)\right] & \leq \gamma + \sum_{t=\gamma+1}^n 2t^{1-\alpha}
+}
+
+Plugging this into the definition of pseudo-regret in equation \ref{eq:regretK} gives:
+
+\eqn {
+R_n \leq \sum_{i=1}^K \Delta_i  \left( \ceil{\frac{\alpha \log(n)}{\psi^*(\Delta i/2)}} + \sum_{t=\gamma+1}^n 2t^{1-\alpha}\right) 
+}
 

### 2015-01-15 simpsons_christmas.tex
adds final version of WIML poster and first draft of observe approach bandit in simplest setting

--- 2014-11-26 causal-bandit/c-bandit.tex
@@
 
-\title{Intervention Bandits}
-\author{Blah blah}
+\title{A Christmas Problem}
+\author{Finnian Lattimore (CECS)}
 
@@
 
-\begin{abstract}
-An abstract.
-\end{abstract}
+The results of the trial presented in table \ref{table:results} provide estimators for the entries in the conditional table $P(Outcome|Drug)$.
+In addition, the totals give an estimates of $P(Drug = A)$ and $P(Drug = B)$.
 
-\section{Introduction}
+\begin{table}[H]
+\caption{Clinical Trial Results}
+\label{table:results}
+\centering
+\begin{tabular}{c|c|c|c}
+ & Drug A & Drug B & Total\\
+\hline
+Success & 65 & 50 & 115\\
+Fail & 35 & 50 & 85\\
+Total & 100 & 100 & 200\\
+\end{tabular}
+\end{table}
 
-Useful references are: \cite{BC12}.
-
-\section{Notation}
-
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
-
-Let $p \in [0,1]^K$ be a fixed and known vector. In each time-step $t$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
-\end{enumerate}
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
-}
-
-\section{Estimating $\mu_{i,j}$}
-The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
-lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
-outperform this approach by exploiting the known causal structure of the problem.
-
-\eq {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
-}
-
-\subsection{Estimators}
-Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
-
-Let $\hat{\mu}_a$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $X_a$.
-
-\eq {
-\hat{\mu}_a &=
-\begin{cases}
-\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
-\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
-\end{cases}\\
-\text{where:}\\
-m_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
-n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s \\
-}
-
-This gives $K$ estimators $\{\hat{\mu_1}...\hat{\mu_K}\}$ to be pooled into a single estimator $\hat{\mu}$.
-
-\eq {
-\hat{\mu} &= \sum^{K}_{a=1}\eta_a\hat{\mu}_a=\eta_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
-\text{where:}\\
-\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
-n_{i,j} & \text{if } a = i \\
-\frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
-\end{cases}
-}
-
-If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p}_a$
-
-\eq{
-\hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
-}
-\subsection{Random sampling variant}
-
-Basic idea - explore randomly until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB).
-Bounds on regret: when exploring regret is at most 1, when exploiting regret it at most $\epsilon$.
-If the horizon is $n$, the cost of exploiting $< \epsilon n$ and the cost of exploring is $O(1/\epsilon^2)$ (this comes from how quickly we converge to within $\epsilon$ for all arms).
-
-\eqn{
-R_n = O(n\epsilon+\frac{1}{\epsilon^2})
-}
-
-Differentiating and selecting $\epsilon = (\frac{2}{n})^{1/3}$ to minimize the regret yields:
-
-\eqn{
-R_n = O(n^{2/3})
-}
-
-At each time step we randomly select a variable $X_i$ on which to intervene and set it to $1$ with probability $p_i$.
-
-The estimators from section XX are unbiased. So we can get an estimate of the error in the estimate by
-
-we are pooling estimators - so what is the error in a pooled estimate - is it the same as what Tor wrote before (because he was assuming unbiased estimators?
-Or is it tighter because we don't need to worry about the the non-fixed nature of ...) I could try a couple of different bounds and see what happens.
-
-\subsection{UCB variant}
-In this case the estimators from section XX are biased - so its a lot harder to prove a regret bound.
-
-\begin{theorem} (Probably False) With probability at least $1 - \delta$ we have that:
-$\displaystyle \left|\hat{\mu}_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
-where $\beta > 0$ is some constant.
-\end{theorem}
-
-\begin{proof}
-First note that $n_{a,b}$ is a random variable that is bounded by $t$ for all $a,b$.
-We use the short-hand $\mu^{a,b}_{i,j} = \E[q(X) | X_i = j, X_a = b]$. Then
-\eq{
-\mu_{i,j} = p_a \mu_{i,j}^{a,1} + (1 - p_a) \mu_{i,j}^{a,0}\,.
-}
-Now we can apply Hoeffding's bound and the union bound to show that
-\eq{
-\P{\left|\frac{m_{a,b}}{n_{a,b}} - \mu_{i,j}^{a,b}\right| \geq \sqrt{\frac{1}{2n_{a,b}} \log \frac{4t}{\delta}}} \leq \frac{\delta}{2}\,.
-}
-Therefore by the union bound
-\eq{
-\P{\left|p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}} - \mu_{i,j}\right| \geq p_a \sqrt{\frac{1}{2n_{a,1}} \log\frac{4t}{\delta}}
-+ (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}} \leq \delta\,
-}
-Now by Jensen's inequality 
-\eq{
-p_a \sqrt{\frac{1}{2n_{a,1}}\log\frac{4t}{\delta}} + (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}
-&\leq \sqrt{\left(\frac{p_a}{2n_{a,1}} + \frac{1-p_a}{2n_{a,0}}\right)\log \frac{4t}{\delta}} \\
-&\leq \sqrt{\max\set{\frac{p_a}{n_{a,1}}, \frac{1-p_a}{n_{a,0}}} \log\frac{4t}{\delta}} \\
-&= \sqrt{{1 \over 2n_a} \log\frac{4t}{\delta}}\,.
-}
-Similarly,
-\eq{
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{4t}{\delta}}} \leq
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{2t}{\delta}}} \leq \delta\,.
-}
-\end{proof}
-
-\section{Algorithm}
-
-\begin{algorithm}
-\caption{UCB}
-\begin{algorithmic}[1]
-\State {\bf Input:} Number of variables $K$, vector $p \in [0,1]^K$, horizon $n$
-\For{$t \in 1,\ldots,n$}
-\For{$i \in 1,\ldots,K$}
-\For{$j \in \set{0,1}$}
-\State Compute $\tilde \mu_{i,j} = \hat \mu_{i,j} + \sqrt{\frac{\alpha}{\sum_{a} n_a} \log n}$ 
-\EndFor
-\EndFor
-\State Choose $I_t, J_t = \argmax_{i,j} \tilde \mu_{i,j}$
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\section{Theorems}
-\section{Experiments}
-
-\begin{figure}[h]
-\caption{Comparison of the performance of standard UCB versus causal UCB with $\beta=3$ and $\beta = 5$.
-100 simulations were run for each algorithm up to a horizon of $10^5$ per value of $K$.
-Error bars span the 1st to 3rd quantile of the regret, round points mark the median and triangular points show the mean.
-For standard UCB the regret increases linearly with the number of arms $K$.
-For causal UCB the increase is sub-linear.
-Increasing $\beta$ leads to slower convergence but lower variance.}
-\includegraphics[width=1\textwidth]{regret_vs_K_series_beta.pdf}
-\end{figure}
-
-\begin{figure}[H]
-\caption{The distribution of regret varies with the $\beta$ parameter in the bound in the estimator.
-As beta increases, the mean regret increases but the variance decreases.
-The plot shows the results of running $100$ independent bandits, with $K=256$ and $\epsilon=0.1$, up to a horizon $h=10^5$ for each value of $\beta$.
-}
-\includegraphics[width=0.7\textwidth]{regret_vs_beta_boxplot.pdf}
-\end{figure}
-
-Simululations to compare the performance of standard UCB with our modified algorithm.
-For each number of arms, $100$ bandits of each type were created and run upto to a horizon of $1000$ timesteps.
-The mean regret and its standard error from these simulations is plotted in figure \ref{fig:regretvsarms} The true data was generated from a model where:
-
-\eq{
-&p = [0.5]^{K}\\
-&q(\boldsymbol{X}) = \begin{cases}0.5 & \text{if $X_{1}=0$}\\ 0.6 & \text{otherwise} \end{cases}
-}
-
-\section{Conclusion}
-
-\newpage
-\section{Notes}
-
-\subsection{Why Hoeffdings bound doesn't hold if $n$ is a random variable}
-
-Let $\{Z_1,Z_2...Z_n\} \sim Bernoulli(\frac{1}{2}) $ and $X_i = 2Z_i-1 \implies X_i \in \{-1,1\} \text{ and } E[X_i] = 0$.
-
-For a fixed $n$, Hoeffdings inequality says: 
-\begin{equation}
-\label{eq:HoeffdingsRandom}
-\begin{aligned}
-& P\left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \epsilon\right) \leq 2e^{-n\epsilon^2} \\
-\implies & P \left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \sqrt{\frac{\log{4}}{n}}\right) \leq \frac{1}{2} \\
-\implies & P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \leq \frac{1}{2}\\
-\end{aligned}
-\end{equation}
-
-If $n$ is allowed to be dependent on the sequence of values sampled, this inequality no longer holds.
-
-\begin{proof}
-Choose $n$ based on the sequence of samples seen so far such that:
-\eq {
-n = min \{ n: n > 4 \text{ and } \sum_{i=1}^{n}X_i > \sqrt{n \log \log n}\}
-}
-By the law of iterated logarithms this quantity is finite with probability $\sim 1$.
-
-\eq {
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{\log{n}}}\right) \sim 1 \\
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \sim 1 \text{, Since } \log{\log{n}} > \log{4} \;\; \forall n > 4 \\
-}
-
-Thus Hoeffdings inequality (equation \ref{eq:HoeffdingsRandom}) does not hold in general if $n$ is not independent of the samples $\{X_i\}$ (The bound does not work if you decide to stop sampling as soon as you reach a point where random walk fluctuations take you outside it)
-
-\end{proof}
+\begin{table}[H]
+\caption{Clinical Trial Results}
+\label{table:results}
+\centering
+\begin{tabular}{|c|c|c}
+Male & Female & Total\\
+\hline
+100 & 100 & 200\\
+\end{tabular}
+\end{table}
 

### 2015-01-15 wiml_poster/Causal_Inference_Lattimore.tex
adds final version of WIML poster and first draft of observe approach bandit in simplest setting

--- 2014-11-21 wiml_poster/conference_poster_5.tex
@@
 \begin{minipage}[t]{0.15\linewidth}%
-\includegraphics[width=10cm]{nicta_logo2.jpg} % Logo or a photo of you, adjust its dimensions here
+\centering
+\includegraphics[height=12cm]{nicta_logo2.jpg} % Logo or a photo of you, adjust its dimensions here
 \end{minipage}
-\begin{minipage}[b]{0.70\linewidth}
+\begin{minipage}[b]{0.65\linewidth}
 \centering
-\veryHuge \color{NavyBlue} \textbf{Causal Inference in Machine Learning} \color{Black}\\ % Title
-\Huge\textit{From Prediction to Decision Making: A review}\\[1cm] % Subtitle
+\resizebox{\linewidth}{!}{\veryHuge \color{NavyBlue} \textbf{Causal Inference in Machine Learning: A review}}
+ \color{Black}\\
+
 \huge \textbf{Finnian Lattimore}\\ % Author(s)
 \huge Australian National University\\ % University/organization
+finnlattimore@gmail.com
+
 \end{minipage}
-\begin{minipage}[t]{0.15\linewidth}\end{minipage}
+\begin{minipage}[t]{0.15\linewidth}
+\centering
+\includegraphics[height=12cm]{anu-logo-notext}
+\end{minipage}
 
@@
 
-\color{DarkSlateGray} % DarkSlateGray color for the rest of the content
+%\color{DarkSlateGray} % DarkSlateGray color for the rest of the content
 
@@
 
-Would reducing salt intake reduce heart attack risk?
+Would cutting salt intake reduce heart attack risk?
 Would upping the minimum wage increase unemployment?
@@
 \centering
-{\resizebox{0.6\columnwidth}{!}{\includegraphics{intervention_figure-crop}}}
+{\resizebox{\columnwidth}{!}{\includegraphics{intervention_figure-crop}}}
 \end{figure}
@@
 
-\item Counterfactuals can be defined in terms of SEMs with a slight relaxation of the independence of errors assumption \cite{Richardson2013}. 
+\item Counterfactuals can be defined in terms of SEMs $Y^0 \sim f(X=0,\epsilon_Y)$ 
+\item The ignorability assumption $(X \ci Y^{0}|\boldsymbol{Z})$ \&  $(X\ci Y^{1}|\boldsymbol{Z})$ is weaker than that implied by mutually independent errors in a SEM, $(X \ci Y^0,Y^1 | \boldsymbol{Z})$.
+The assumptions can be made equivalent by utilizing SEMs with a modified independence of errors assumption \cite{Richardson2013}.
+
 \end{itemize}
@@
 
-Consider the problem where the causal DAG is known, due to theory or prior knowledge and we wish to infer the outcome of an intervention of the form $P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}))$ using observational data.
+Consider the problem where the causal DAG is known, due to theory or prior knowledge, and we wish to infer the outcome of an intervention of the form $P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}))$ using observational data.
 
@@
 \centering
-{\resizebox{0.6\columnwidth}{!}{\includegraphics{dsepfigure-crop}}}
+{\resizebox{\columnwidth}{!}{\includegraphics{dsepfigure-crop}}}
 \end{figure}
@@
 
+\columnbreak
 \section*{Causal Discovery}
@@
 \begin{enumerate}
-\item We assume our distribution $P$ was generated by some (unknown) causal DAG over our observed variables (causal sufficiency) 
-\item We assume that all the conditional independences in $P$ are implied by d-separation in the true causal network (\textbf{\textit{faithfulness}})
+\item We assume our distribution $P$ was generated by some (unknown) causal DAG over our observed variables (\emph{causal sufficiency}) 
+\item We assume that all the conditional independences in $P$ are implied by d-separation in the true causal network (\emph{faithfulness}) 
 \item Finding the causal structure equates to finding perfect maps for $P$
@@
 \centering
-{\resizebox{0.4\columnwidth}{!}{\includegraphics{SGS_figure-crop}}}
+{\resizebox{0.45\columnwidth}{!}{\includegraphics{SGS_figure-crop}}}
 \end{figure}
@@
 
-\newcommand{\imsize}{.5\columnwidth}
+\newcommand{\imsize}{.70\columnwidth}
 
@@
 \centering
-{\resizebox{0.9\columnwidth}{!}{\includegraphics{FCI_figure-crop}}}
+{\resizebox{\columnwidth}{!}{\includegraphics{FCI_figure-crop}}}
 \end{figure}
@@
 Therefore, semi-supervised learning should yield no benefit when trying to learn in the causal direction, (estimating $P(Y|X)$) but could help when learning in the anti-causal direction.}
-{\resizebox{\imsize}{!}{\includegraphics{Janzig6}}}
+{\resizebox{\imsize}{!}{\includegraphics{Janzig61}}}
 \end{figure}

### 2015-01-15 wiml_poster/a4version.tex
adds final version of WIML poster and first draft of observe approach bandit in simplest setting

--- 2015-01-15 wiml_poster/Causal_Inference_Lattimore.tex
@@
 
-\documentclass[b0,landscape,25pt]{sciposter}
+%\documentclass[a4,portrait,25pt]{sciposter}
+%\documentclass{article}
+\documentclass[20pt]{extreport}
 
-\usepackage{multicol,caption} % This is so we can have multiple columns of text side-by-side
-\columnsep=100pt % This is the amount of white space between the columns in the poster
-\columnseprule=3pt % This is the thickness of the black line between the columns in the poster
-
+%\usepackage{multicol,caption} % This is so we can have multiple columns of text side-by-side
+%\columnsep=100pt % This is the amount of white space between the columns in the poster
+%\columnseprule=3pt % This is the thickness of the black line between the columns in the poster
+\usepackage[margin=0.3in]{geometry}
 \usepackage[svgnames]{xcolor} % Specify colors by their 'svgnames', for a full list of all colors available see here: http://www.latextemplates.com/svgnames-colors
@@
 \usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{algorithm}% http://ctan.org/pkg/algorithms
+
 \newtheorem{Def}{Definition}
@@
 
+\title{Causal Inference in Machine Learning: A Review}
+\author{Finnian Lattimore\\ Australian National University\\finnlattimore@gmail.com\\\textbf{Apologies,}\\ \textbf{the poster is in Air Canada's lost baggage area} }
+
 \begin{document}
 \def\ci{\perp\!\!\!\perp}
+\maketitle
 %----------------------------------------------------------------------------------------
@@
 % The widths of these boxes can be easily edited to accommodate your content as you see fit
-\begin{minipage}[t]{0.15\linewidth}%
-\centering
-\includegraphics[height=12cm]{nicta_logo2.jpg} % Logo or a photo of you, adjust its dimensions here
-\end{minipage}
-\begin{minipage}[b]{0.65\linewidth}
-\centering
-\resizebox{\linewidth}{!}{\veryHuge \color{NavyBlue} \textbf{Causal Inference in Machine Learning: A review}}
- \color{Black}\\
+%\begin{minipage}[t]{0.15\linewidth}%
+%\centering
+%\includegraphics[height=12cm]{nicta_logo2.jpg} % Logo or a photo of you, adjust its dimensions here
+%\end{minipage}
+%\begin{minipage}[b]{0.65\linewidth}
+%\centering
+%\resizebox{\linewidth}{!}{\veryHuge \color{NavyBlue} \textbf{Causal Inference in Machine Learning: %A review}}
+% \color{Black}\\
 
-\huge \textbf{Finnian Lattimore}\\ % Author(s)
-\huge Australian National University\\ % University/organization
-finnlattimore@gmail.com
+%\huge \textbf{Finnian Lattimore}\\ % Author(s)
+%\huge Australian National University\\ % University/organization
+%finnlattimore@gmail.com
 
-\end{minipage}
-\begin{minipage}[t]{0.15\linewidth}
-\centering
-\includegraphics[height=12cm]{anu-logo-notext}
-\end{minipage}
+%\end{minipage}
+%\begin{minipage}[t]{0.15\linewidth}
+%\centering
+%\includegraphics[height=12cm]{anu-logo-notext}
+%\end{minipage}
 
@@
 
-\begin{multicols}{5} % This is how many columns your poster will be broken into, a poster with many figures may benefit from less columns whereas a text-heavy poster benefits from more
+%\begin{multicols}{2} % This is how many columns your poster will be broken into, a poster with many figures may benefit from less columns whereas a text-heavy poster benefits from more
 
@@
 
-\fbox{\begin{minipage}{0.99\columnwidth}%
+\begin{center} 
+\fbox{\begin{minipage}{0.9\textwidth}%
 \textbf{The Three Rules}
@@
 \begin{aligned}
-\text{if }  (\boldsymbol{Y} \ci & \boldsymbol{W}|\boldsymbol{X})  \text{ in } G_{\overline{\boldsymbol{X}}}  \\
+& \text{if }  (\boldsymbol{Y} \ci  \boldsymbol{W}|\boldsymbol{X})  \text{ in } G_{\overline{\boldsymbol{X}}}  \\
 & P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) 
@@
 \begin{aligned}
-\text{if } (\boldsymbol{Y} \ci  \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z}) & \text{ in } G^{\dagger}\\
+& \text{if } (\boldsymbol{Y} \ci  \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z})  \text{ in } G^{\dagger}\\
 & P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{Z})
@@
 \begin{aligned}
-\text{if }  (\boldsymbol{Y} \ci  \boldsymbol{\hat{X}}|\boldsymbol{Z}) & \text{ in } G^{\dagger}\\
+&\text{if }  (\boldsymbol{Y} \ci  \boldsymbol{\hat{X}}|\boldsymbol{Z})  \text{ in } G^{\dagger}\\
 & P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}) = P(\boldsymbol{Y}|\boldsymbol{Z})
@@
 \end{minipage}}
+\end{center}
 
-\columnbreak
+%\columnbreak
 \section*{Causal Discovery}
@@
 \centering
-{\resizebox{0.45\columnwidth}{!}{\includegraphics{SGS_figure-crop}}}
+{\resizebox{0.55\columnwidth}{!}{\includegraphics{SGS_figure-crop}}}
 \end{figure}
@@
 
-\end{multicols}
+%\end{multicols}
 \end{document}

### 2015-01-15 causal-bandit/two_vars.py

@@ -1,0 +1,83 @@
+# lets run a simulation to see how we should allocate our samples
+# to get within epsilon bounds for all arms as quick as possible
+from scipy.stats import binom
+import numpy as np
+import matplotlib.pyplot as plt
+import math
+import time
+
+def split_samples(n,p1,p2):
+    """ return the number of times to take each action, subject to a total of n"""
+    equal = .25*n
+    return (equal,equal,equal,equal) # do(x1=1), do(x1=0), do(x2=1), do(x2=0)
+
+def epsilon(a):
+    return np.sqrt(np.log(2/.5)/(2.0*a))
+
+def a(n,na1,na0,pa):
+    nom = na1*na0
+    den = na1*pow(1-pa,2)+na0*pow(pa,2)
+    try:
+        assert((den > 0.0).all())
+    except:
+        print min(den)
+    return n+nom/den
+
+def simulate(split,p1,p2,runs):
+    n11,n10,n21,n20 = split
+    n11_21 = binom.rvs(n11, p2, size=runs)
+    n11_20 = (n11 - n11_21)
+    n10_21 = binom.rvs(n10,p2,size=runs)
+    n10_20 = (n10-n10_21)
+    n21_11 = binom.rvs(n21,p1,size=runs)
+    n21_10 = (n21- n21_11)
+    n20_11 = binom.rvs(n20,p1,size=runs)
+    n20_10 = (n20- n20_11)
+
+    # deal with the fact some of these n's can be 0.
+
+    a11 = a(n11,n21_11,n20_11,p2)
+    a10 = a(n10,n21_10,n20_10,p2)
+    a21 = a(n21,n11_21,n10_21,p1)
+    a20 = a(n20,n11_20,n10_20,p1)
+   
+
+    e11 = epsilon(a11)
+    e10 = epsilon(a10)
+    e21 = epsilon(a21)
+    e20 = epsilon(a20)
+
+    e = [np.mean(e11),np.mean(e10),np.mean(e21),np.mean(e20)]
+    worst = max(e)
+    return worst
+
+    #plt.hist(a11)
+    #plt.show()
+    
+    
+n = 100.0 # total number of samples
+p1 = .5 # P(X1=1)
+p2 = .5 # P(X2=1)
+
+
+results = []
+for f1 in np.linspace(.1,.9,9):
+    for f11 in np.linspace(.1,.9,9):
+        n11 = n*f1*f11
+        n10 = n*f1*(1-f11)
+        for f21 in np.linspace(.1,.9,9):
[…]

### 2015-01-16 causal-bandit/inequalities.tex
gets regret bound for observe stage for simplest possible model

--- 2014-11-26 causal-bandit/c-bandit.tex
@@
 
-\title{Intervention Bandits}
+\title{Inequalities}
 \author{Blah blah}
@@
 
-\begin{abstract}
-An abstract.
-\end{abstract}
+Assume we have i.i.d samples $X_i \in \{0,1\}$ from a Bernoulli distribution with mean $p$
 
-\section{Introduction}
-
-Useful references are: \cite{BC12}.
-
-\section{Notation}
-
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
-
-Let $p \in [0,1]^K$ be a fixed and known vector. In each time-step $t$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
-\end{enumerate}
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
+Hoeffding's inequality
+\eqn {
+P(\frac{1}{n}\sum_{i=1}^n X_i -p \geq \epsilon) \leq e^{-2n\epsilon^2} 
 }
 
-\section{Estimating $\mu_{i,j}$}
-The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
-lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$. In this instance we can significantly
-outperform this approach by exploiting the known causal structure of the problem.
-
-\eq {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
+Chebyshev
+\eqn {
+P(\frac{1}{n}\sum_{i=1}^n X_i -p \geq \epsilon) \leq \frac{1}{\epsilon}\sqrt{\frac{p(1-p)}{n}}
 }
 
-\subsection{Estimators}
-Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
-
-Let $\hat{\mu}_a$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $X_a$.
-
-\eq {
-\hat{\mu}_a &=
-\begin{cases}
-\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
-\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
-\end{cases}\\
-\text{where:}\\
-m_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
-n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s \\
+Chernoff 
+\eqn {
+P(\frac{1}{n}\sum_{i=1}^n X_i -p \geq \epsilon) \leq e^{-nD(p+\epsilon||p)} 
 }
-
-This gives $K$ estimators $\{\hat{\mu_1}...\hat{\mu_K}\}$ to be pooled into a single estimator $\hat{\mu}$.
-
-\eq {
-\hat{\mu} &= \sum^{K}_{a=1}\eta_a\hat{\mu}_a=\eta_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
-\text{where:}\\
-\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
-n_{i,j} & \text{if } a = i \\
-\frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
-\end{cases}
-}
-
-If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p}_a$
-
-\eq{
-\hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
-}
-\subsection{Random sampling variant}
-
-Basic idea - explore randomly until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB).
-Bounds on regret: when exploring regret is at most 1, when exploiting regret it at most $\epsilon$.
-If the horizon is $n$, the cost of exploiting $< \epsilon n$ and the cost of exploring is $O(1/\epsilon^2)$ (this comes from how quickly we converge to within $\epsilon$ for all arms).
-
-\eqn{
-R_n = O(n\epsilon+\frac{1}{\epsilon^2})
-}
-
-Differentiating and selecting $\epsilon = (\frac{2}{n})^{1/3}$ to minimize the regret yields:
-
-\eqn{
-R_n = O(n^{2/3})
-}
-
-At each time step we randomly select a variable $X_i$ on which to intervene and set it to $1$ with probability $p_i$.
-
-The estimators from section XX are unbiased. So we can get an estimate of the error in the estimate by
-
-we are pooling estimators - so what is the error in a pooled estimate - is it the same as what Tor wrote before (because he was assuming unbiased estimators?
-Or is it tighter because we don't need to worry about the the non-fixed nature of ...) I could try a couple of different bounds and see what happens.
-
-\subsection{UCB variant}
-In this case the estimators from section XX are biased - so its a lot harder to prove a regret bound.
-
-\begin{theorem} (Probably False) With probability at least $1 - \delta$ we have that:
-$\displaystyle \left|\hat{\mu}_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
-where $\beta > 0$ is some constant.
-\end{theorem}
-
-\begin{proof}
-First note that $n_{a,b}$ is a random variable that is bounded by $t$ for all $a,b$.
-We use the short-hand $\mu^{a,b}_{i,j} = \E[q(X) | X_i = j, X_a = b]$. Then
-\eq{
-\mu_{i,j} = p_a \mu_{i,j}^{a,1} + (1 - p_a) \mu_{i,j}^{a,0}\,.
-}
-Now we can apply Hoeffding's bound and the union bound to show that
-\eq{
-\P{\left|\frac{m_{a,b}}{n_{a,b}} - \mu_{i,j}^{a,b}\right| \geq \sqrt{\frac{1}{2n_{a,b}} \log \frac{4t}{\delta}}} \leq \frac{\delta}{2}\,.
-}
-Therefore by the union bound
-\eq{
-\P{\left|p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}} - \mu_{i,j}\right| \geq p_a \sqrt{\frac{1}{2n_{a,1}} \log\frac{4t}{\delta}}
-+ (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}} \leq \delta\,
-}
-Now by Jensen's inequality 
-\eq{
-p_a \sqrt{\frac{1}{2n_{a,1}}\log\frac{4t}{\delta}} + (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}
-&\leq \sqrt{\left(\frac{p_a}{2n_{a,1}} + \frac{1-p_a}{2n_{a,0}}\right)\log \frac{4t}{\delta}} \\
-&\leq \sqrt{\max\set{\frac{p_a}{n_{a,1}}, \frac{1-p_a}{n_{a,0}}} \log\frac{4t}{\delta}} \\
-&= \sqrt{{1 \over 2n_a} \log\frac{4t}{\delta}}\,.
-}
-Similarly,
-\eq{
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{4t}{\delta}}} \leq
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{2t}{\delta}}} \leq \delta\,.
-}
-\end{proof}
-
-\section{Algorithm}
-
-\begin{algorithm}
-\caption{UCB}
-\begin{algorithmic}[1]
-\State {\bf Input:} Number of variables $K$, vector $p \in [0,1]^K$, horizon $n$
-\For{$t \in 1,\ldots,n$}
-\For{$i \in 1,\ldots,K$}
-\For{$j \in \set{0,1}$}
-\State Compute $\tilde \mu_{i,j} = \hat \mu_{i,j} + \sqrt{\frac{\alpha}{\sum_{a} n_a} \log n}$ 
-\EndFor
-\EndFor
-\State Choose $I_t, J_t = \argmax_{i,j} \tilde \mu_{i,j}$
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\section{Theorems}
-\section{Experiments}
-
-\begin{figure}[h]
-\caption{Comparison of the performance of standard UCB versus causal UCB with $\beta=3$ and $\beta = 5$.
-100 simulations were run for each algorithm up to a horizon of $10^5$ per value of $K$.
-Error bars span the 1st to 3rd quantile of the regret, round points mark the median and triangular points show the mean.
-For standard UCB the regret increases linearly with the number of arms $K$.
-For causal UCB the increase is sub-linear.
-Increasing $\beta$ leads to slower convergence but lower variance.}
-\includegraphics[width=1\textwidth]{regret_vs_K_series_beta.pdf}
-\end{figure}
-
-\begin{figure}[H]
-\caption{The distribution of regret varies with the $\beta$ parameter in the bound in the estimator.
-As beta increases, the mean regret increases but the variance decreases.
-The plot shows the results of running $100$ independent bandits, with $K=256$ and $\epsilon=0.1$, up to a horizon $h=10^5$ for each value of $\beta$.
-}
-\includegraphics[width=0.7\textwidth]{regret_vs_beta_boxplot.pdf}
-\end{figure}
-
-Simululations to compare the performance of standard UCB with our modified algorithm.
-For each number of arms, $100$ bandits of each type were created and run upto to a horizon of $1000$ timesteps.
-The mean regret and its standard error from these simulations is plotted in figure \ref{fig:regretvsarms} The true data was generated from a model where:
-
-\eq{
-&p = [0.5]^{K}\\
-&q(\boldsymbol{X}) = \begin{cases}0.5 & \text{if $X_{1}=0$}\\ 0.6 & \text{otherwise} \end{cases}
-}
-
-\section{Conclusion}
-
-\newpage
-\section{Notes}
-
-\subsection{Why Hoeffdings bound doesn't hold if $n$ is a random variable}
-
-Let $\{Z_1,Z_2...Z_n\} \sim Bernoulli(\frac{1}{2}) $ and $X_i = 2Z_i-1 \implies X_i \in \{-1,1\} \text{ and } E[X_i] = 0$.
-
-For a fixed $n$, Hoeffdings inequality says: 
-\begin{equation}
-\label{eq:HoeffdingsRandom}
-\begin{aligned}
-& P\left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \epsilon\right) \leq 2e^{-n\epsilon^2} \\
-\implies & P \left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \sqrt{\frac{\log{4}}{n}}\right) \leq \frac{1}{2} \\
-\implies & P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \leq \frac{1}{2}\\
-\end{aligned}
-\end{equation}
-
-If $n$ is allowed to be dependent on the sequence of values sampled, this inequality no longer holds.
-
-\begin{proof}
-Choose $n$ based on the sequence of samples seen so far such that:
-\eq {
-n = min \{ n: n > 4 \text{ and } \sum_{i=1}^{n}X_i > \sqrt{n \log \log n}\}
-}
-By the law of iterated logarithms this quantity is finite with probability $\sim 1$.
-
-\eq {
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{\log{n}}}\right) \sim 1 \\
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \sim 1 \text{, Since } \log{\log{n}} > \log{4} \;\; \forall n > 4 \\
-}
-
-Thus Hoeffdings inequality (equation \ref{eq:HoeffdingsRandom}) does not hold in general if $n$ is not independent of the samples $\{X_i\}$ (The bound does not work if you decide to stop sampling as soon as you reach a point where random walk fluctuations take you outside it)
-
-\end{proof}
 

### 2015-01-16 causal-bandit/kl_inequality.py

@@ -1,0 +1,9 @@
+def kl(x,y):
+    return x*log(x/y)+(1-x)*log((1-x)/(1-y))
+
+p = .5
+p2 = .4
+e = 1
+
+kl(p2+e*p,p2)
+kl

### 2015-01-16 causal-bandit/c-bandit.tex
gets regret bound for observe stage for simplest possible model
working on regret consdiering overlap of more than one arm

@@
 
-If \ref{eq:overlap3} is true, we can bound the probability of overlap at some timestep $t$ with the union bound of the probability that either \ref{eq:overlap1} or \ref{eq:overlap2} are false.
+If \ref{eq:overlap3} is true, we can bound the probability of overlap  been the confidence intervals of the best arm and one of the other arms, $k$, at some timestep $t$ with the union bound of the probability that either \ref{eq:overlap1} or \ref{eq:overlap2} are false.
 
 \eqn {
-P(overlap) = P(n \geq t) \leq 
-P(\mu^* - \hat{\mu}^* \geq \epsilon) + P(\hat{\mu}_i - \mu_i \geq \epsilon)  \;\;\;\text{ if } \epsilon \leq a/4
+\rho(t) \equiv P(overlap_k) \leq 
+P(\mu^* - \hat{\mu}^* \geq \epsilon) + P(\hat{\mu}_k - \mu_k \geq \epsilon)  \;\;\;\text{ if } \epsilon \leq a/4
 }
 
-Let $\epsilon = a/4$
+Now for random variables $W_{sij} = P(Y|X_i=j)$ drawn from a Bernoulli distribution with mean $p_{ij}$ the Chernoff inequality gives:
 
 \eqn {
-P(n > t) \leq & e^{-nD(p_{ij}+p (a/4) || p_{ij})} +e^{-nD(p_{ij}-p (a/4) || p_{ij})} \\
-\leq & \;\;\;?
+& P(\frac{1}{t}\sum_s W_{sij} - p_{ij} \geq \gamma) \leq \exp(-\frac{t\gamma^2}{2p_{ij}(1-p_{ij})+2\gamma/3}) \\
+\implies & P(\frac{1}{t}\sum_s \frac{W_{sij}}{p} - \frac{p_{ij}}{p} \geq \epsilon) = P(\hat{\mu}_i - \mu_i \geq \epsilon) \leq \exp(-\frac{tp\epsilon^2}{2+2\epsilon/3}) = \exp(-\frac{t\epsilon^2}{4+4\epsilon/3})
+}
+Where $p_{ij} = P(Y = 1, X_i = j) \leq P(X_i=j) = p = 1/2 \;\;\; \forall(i,j)$
+
+Similarly
+
+\eqn {
+P(\mu^* - \hat{\mu}^* \geq \epsilon) \leq exp(-tp^2\epsilon^2) = exp(-\frac{t\epsilon^2}{4})
 }
 
-Where $p_ij = P(Y = 1, X_i = j) \leq P(X_i=j) = p \;\;\; \forall(i,j)$
-
-Now
+Now the estimators and their bounds are independent, as they are composed from sums of independent variables $X_i$ so the probability that $\alpha$ of them have overlapping bounds is given by a binomial distribution.
 
 \eqn {
-E[n] = \sum_{t=1}^\infty P(n \geq t)
+P(O > \alpha)  = & P(n \geq t) = 1 - \sum_{i=1}^\alpha \binom{K-1}{i}\rho(t)^{i}(1-\rho(t))^{K-1-i} \\
+\leq & \exp(-(K-1)D(\frac{\alpha}{K-1} || \rho(t))
+}
+
+\eqn {
+E[n] = \sum_{t=1}^\infty P(n \geq t) \leq & \sum_{t=1}^\infty \left( \exp(-\frac{t\epsilon^2}{4+4\epsilon/3}) +exp(-\frac{t\epsilon^2}{4}) \right) \\
+= & \frac{1}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} + \frac{1}{\exp(\epsilon^2/4) - 1} \\
+\leq & \frac{2}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} \\
+= & \frac{2}{\exp(a^2/(64+4a/3)) - 1}
+}
+
+So the regret for the observe stage is bounded by:
+
+\eqn {
+R \leq \frac{2a}{\exp(a^2/(64+4a/3)) - 1}
 }

### 2015-01-19 causal-bandit/c-bandit.tex
regret bound for observe until plausible set smaller than alpha

@@
 
+Let $n$ be the number of timesteps until the size of the set $A_t$ falls below $\alpha$ or we reach the overall horizon $T$
+
+\eqn {
+n = \min(\{t:size(A_t)< \alpha, T\})
+}
+
+Let $Q(k)$ be the expected reward for arm $k$ and $\mu^*$ be the expected reward for the best arm.
+
 \eqn {
 R_n = & \max_{i=1...K} E\left[\sum_{t=1}^n Y_{it} - \sum_{t=1}^n Y_{I_t,t} \right] \\
-= & E[n]\left(P(Y|X_{i^*}={j^*})  - P(Y)\right) \text{ where } X_{i^*}={j^*} \text{ is the best arm}
+= & E[n]\left(\mu^*  - E[Q]\right) 
 }
 
-\subsubsection{A very simple model}
-
-Lets try to analyse a very simple model in which the reward depends only on the value of $X_1$.
+Without loss of generality, assume the arms are ordered from best to worst.
 
 \eqn{
-&P(Y|X_1=0,X_2,...X_N) = \frac{1}{2}-a\\
-&P(Y|X_1=1,X_2,...X_N) = \frac{1}{2}+a\\
-&P(X_i = j) = \frac{1}{2} \;\;\; \forall (i,j)\\
-&\implies P(Y) = \frac{1}{2} \\
-&\implies R_n = aE[n]
+& Q(k = 1...K) = [\mu^*, \mu^*-d_2 ... \mu^*-d_k] \\
+&\implies E[Q] = \mu^* - \frac{1}{K}\sum_{k=2}^K d_k \\
+&\implies  R_n = E[n]\left(\frac{1}{K}\sum_{k=2}^K d_k \right)
 }
 
-The goal is now to find $E[n]$ (which will be a function of $a$)
-
-Lets start by considering how long on average we need to observe until the lower bound on the optimal arm is higher than the upper bound on a single other arm, $i$, (with expected reward $\frac{1}{2}$)
-
-If the following three statements hold, then there is no overlap $LB^* = \hat{\mu}^* - \epsilon > UB_i = \hat{\mu}_i + \epsilon$.
+Whenever the following two statements hold only arms with indexes less than $\alpha$ can be in the set $A_t$:
 
 \eqn {
-\label{eq:overlap1}
-&\mu^* - \hat{\mu}^* \leq \epsilon \\
-\label{eq:overlap2}
-&\hat{\mu}_i - \mu_i \leq \epsilon \\
-\label{eq:overlap3}
-&\epsilon \leq a/4
+& \mu^* - \hat{\mu}^* \leq \frac{d_\alpha}{2} \\
+& \hat{\mu_k} - \mu_k \leq \frac{d_k}{2} \; \forall k \geq \alpha 
 }
 
-If \ref{eq:overlap3} is true, we can bound the probability of overlap  been the confidence intervals of the best arm and one of the other arms, $k$, at some timestep $t$ with the union bound of the probability that either \ref{eq:overlap1} or \ref{eq:overlap2} are false.
+\eqn{
+\label{eq:proboverlap}
+\implies \rho(t) \equiv P(\exists k \in A_t: k \geq \alpha)  \leq P(\mu^* - \hat{\mu}^* > \frac{d_\alpha}{2}) + \sum_{k=\alpha}^K P( \hat{\mu_k} - \mu_k > \frac{d_k}{2})
+}
 
 \eqn {
-\rho(t) \equiv P(overlap_k) \leq 
-P(\mu^* - \hat{\mu}^* \geq \epsilon) + P(\hat{\mu}_k - \mu_k \geq \epsilon)  \;\;\;\text{ if } \epsilon \leq a/4
+P(n > t) = \begin{cases} 0 & \text{if } t \geq T \\
+P(\exists k \in A_t: k \geq \alpha) & \text{otherwise}
+\end{cases}
 }
 
-Now for random variables $W_{sij} = P(Y|X_i=j)$ drawn from a Bernoulli distribution with mean $p_{ij}$ the Chernoff inequality gives:
+We now concentrate bounding \ref{eq:proboverlap}. For random variables $W_{sij} = P(Y|X_i=j)$ drawn from a Bernoulli distribution with mean $p_{ij}$ the Chernoff inequality gives:
 
@@
 & P(\frac{1}{t}\sum_s W_{sij} - p_{ij} \geq \gamma) \leq \exp(-\frac{t\gamma^2}{2p_{ij}(1-p_{ij})+2\gamma/3}) \\
-\implies & P(\frac{1}{t}\sum_s \frac{W_{sij}}{p} - \frac{p_{ij}}{p} \geq \epsilon) = P(\hat{\mu}_i - \mu_i \geq \epsilon) \leq \exp(-\frac{tp\epsilon^2}{2+2\epsilon/3}) = \exp(-\frac{t\epsilon^2}{4+4\epsilon/3})
+\implies & P(\frac{1}{t}\sum_s \frac{W_{sij}}{p} - \frac{p_{ij}}{p} \geq \epsilon) = P(\hat{\mu}_k - \mu_k \geq \epsilon) \leq \exp(-\frac{tp\epsilon^2}{2+2\epsilon/3}) = \exp(-\frac{t\epsilon^2}{4+4\epsilon/3})
 }
@@
 
-Similarly
-
+Similarly, 
 \eqn {
@@
 
-Now the estimators and their bounds are independent, as they are composed from sums of independent variables $X_i$ so the probability that $\alpha$ of them have overlapping bounds is given by a binomial distribution.
+\eqn {
+\implies \rho(t) \leq \exp(-\frac{ d_\alpha^2 t}{16}) + \sum_{k=\alpha}^K exp(-\frac{d_k^2 t}{16(1+d_k/6)})
+}
+
+\eqn{ 
+E[n] = & \sum_{t=1}^T P(n \geq t) \leq \sum_{t=1}^T \rho(t) \\
+= & \min \left(T, \frac{1 - e^{-d_\alpha^2 T/16}}{e^{d_\alpha^2 /16} -1 } +  \sum_{k=\alpha}^K \frac{1 - e^{-d_k^2 T/16(1+d_k/6)}}{e^{d_k^2 /16(1+d_k/6)} -1 } \right)
+}
+
+Since the regret is $R_n = E[n]\left(\frac{1}{K}\sum_{k=2}^K d_k \right)$ the worst case is that just enough of the differences $d_k$ are $0$ to make $E[n] = T$ and the remainder are $1$.
+This occurs if $d_\alpha = 0$, In this case the regret is:
 
 \eqn {
-P(O > \alpha)  = & P(n \geq t) = 1 - \sum_{i=1}^\alpha \binom{K-1}{i}\rho(t)^{i}(1-\rho(t))^{K-1-i} \\
-\leq & \exp(-(K-1)D(\frac{\alpha}{K-1} || \rho(t))
-}
-
-\eqn {
-E[n] = \sum_{t=1}^\infty P(n \geq t) \leq & \sum_{t=1}^\infty \left( \exp(-\frac{t\epsilon^2}{4+4\epsilon/3}) +exp(-\frac{t\epsilon^2}{4}) \right) \\
-= & \frac{1}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} + \frac{1}{\exp(\epsilon^2/4) - 1} \\
-\leq & \frac{2}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} \\
-= & \frac{2}{\exp(a^2/(64+4a/3)) - 1}
+R_n = T \left(\frac{K-(\alpha+1)}{K}\right)
 }
@@
 \eqn {
+E[n] = \sum_{t=1}^T P(n \geq t) \leq & \sum_{t=1}^\infty \left( \exp(-\frac{t\epsilon^2}{4+4\epsilon/3}) +exp(-\frac{t\epsilon^2}{4}) \right) \\
+= & \frac{1}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} + \frac{1}{\exp(\epsilon^2/4) - 1} \\
+\leq & \frac{2}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} \\
+= & \frac{2}{\exp(a^2/(64+4a/3)) - 1}
+}
+
+\subsubsection{A very simple model}
+
+Lets try to analyse a very simple model in which the reward depends only on the value of $X_1$.
+
+\eqn{
+&P(Y|X_1=0,X_2,...X_N) = \frac{1}{2}-a\\
+&P(Y|X_1=1,X_2,...X_N) = \frac{1}{2}+a\\
+&P(X_i = j) = \frac{1}{2} \;\;\; \forall (i,j)\\
+&\implies P(Y) = \frac{1}{2} \\
+&\implies R_n = aE[n]
+}
+
+The goal is now to find $E[n]$ (which will be a function of $a$)
+
+Lets start by considering how long on average we need to observe until the lower bound on the optimal arm is higher than the upper bound on a single other arm, $i$, (with expected reward $\frac{1}{2}$)
+
+If the following three statements hold, then there is no overlap $LB^* = \hat{\mu}^* - \epsilon > UB_i = \hat{\mu}_i + \epsilon$.
+
+\eqn {
+\label{eq:overlap1}
+&\mu^* - \hat{\mu}^* \leq \epsilon \\
+\label{eq:overlap2}
+&\hat{\mu}_i - \mu_i \leq \epsilon \\
+\label{eq:overlap3}
+&\epsilon \leq a/4
+}
+
+If \ref{eq:overlap3} is true, we can bound the probability of overlap  been the confidence intervals of the best arm and one of the other arms, $k$, at some timestep $t$ with the union bound of the probability that either \ref{eq:overlap1} or \ref{eq:overlap2} are false FOR ALL ARMS.
+
+\eqn {
+\rho(t) \equiv P(overlap_k) \leq 
+P(\mu^* - \hat{\mu}^* \geq \epsilon) + P(\hat{\mu}_k - \mu_k \geq \epsilon)  \;\;\;\text{ if } \epsilon \leq a/4
+}
+
+Now the estimators and their bounds are independent, as they are composed from sums of independent variables $X_i$ so the probability that $\alpha$ of them have overlapping bounds is given by a binomial distribution.
+
+\eqn {
+P(O > \alpha)  = & P(n \geq t) = 1 - \sum_{i=1}^\alpha \binom{K-1}{i}\rho(t)^{i}(1-\rho(t))^{K-1-i} \\
+\leq & \exp(-(K-1)D(\frac{\alpha}{K-1} || \rho(t))
+}
+
+\eqn {
 P(|\hat{\mu}_{a0} -\mu_{a0}| > \epsilon_{a0}) \leq \delta/2 \;\; \text{ where } 

### 2015-01-22 causal-bandit/c-bandit.tex
regret bound for observe until plausible set smaller than alpha
attempted implementation of information direced sampling

@@
 
-So the regret for the observe stage is bounded by:
+\subsection{Regret during UCB stage}
 
-\eqn {
-R \leq \frac{2a}{\exp(a^2/(64+4a/3)) - 1}
-}
+On completion of the observe stage we have tight bounds on all the arms except the first $\alpha$
 
-But this is just the expected number of steps until one arm does not overlap. With more arms its more complicated - as we need the probability that $\alpha$ of them overlap (another union bound?).
+Essentially, we want to know how large the regret is if we initialize with a certain number of counts already in place...
+
+\subsection{General rough thoughts}
+
+When observing the confidence interval on all the arms drops approximately with $\sqrt{\frac{1}{n}}$ (provided that the probability of each occurring is the same) - otherwise for each arm it will be weighted by this drop will be weighted by a function of the probability that it occurs.
+When acting, the interval on a single arm we intervene on drops.
+What is the informational value of reducing the bounds on various arms (it is higher for arms that are more likely to be good).
 
@@
 
+So the regret for the observe stage is bounded by:
+
+\eqn {
+R \leq \frac{2a}{\exp(a^2/(64+4a/3)) - 1}
+}
+
+But this is just the expected number of steps until one arm does not overlap. With more arms its more complicated - as we need the probability that $\alpha$ of them overlap (another union bound?).
+
 \eqn {
@@
 
+So the regret for the observe stage is bounded by:
+
+\eqn {
+R \leq \frac{2a}{\exp(a^2/(64+4a/3)) - 1}
+}
+
+But this is just the expected number of steps until one arm does not overlap. With more arms its more complicated - as we need the probability that $\alpha$ of them overlap (another union bound?).
 Lets try to analyse a very simple model in which the reward depends only on the value of $X_1$.

### 2015-01-22 causal-bandit/beta.py

@@ -1,0 +1,15 @@
+from scipy.stats import beta
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.stats import bernoulli
+from math import log
+from scipy.stats import entropy
+
+x = np.zeros(10)
+y = np.zeros(10)
+y[1] = 3
+y[2] = -3
+z = np.maximum(x,y)
+print x
+print y
+print z

### 2015-03-13 causal-bandit/c-bandit.tex
adds work trying to figure out what happens to UCB regret if you initialize with bounds from observational data

@@
 \usepackage{parskip}
+\usepackage{tikz} 
+\usetikzlibrary{arrows,positioning} 
+\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
+%\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+
+\tikzset{
+    %Define standard arrow tip
+    >=stealth',
+    %Define style for boxes
+    observed/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=1em,
+           minimum height=1em,
+           font=\footnotesize,
+           text centered,
+           fill=blue!20!white},
+     latent/.style={
+           circle,
+           rounded corners,
+           draw=black, thick, dashed,
+           minimum width=.5em,
+           minimum height=.5em,
+           font=\footnotesize,
+           text centered,
+           fill=black!10!white
+           },
+    % Define arrow style
+    pil/.style={
+           o->,
+           thick,
+           shorten <=2pt,
+           shorten >=2pt,},
+    sh/.style={ shade, shading=axis, left color=red, right color=green,
+    shading angle=45 }  
+}
 
@@
 
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
+Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$, figure .
+We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
 
@@
 
-\eq {
+\eqn {
 P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
@@
  &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
+ \label{eqn:marginalizedEstimator}
  &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
@@
 
-Let $\hat{\mu}_a$ be an empirical estimator for $P(Y|do(X_i=j))$ obtained via marginalization over $X_a$.
+Considering equation \ref{eqn:marginalizedEstimator}, a natural way to construct an empirical estimator,  $\hat{\mu}_a$, for $P(Y|do(X_i=j))$ utilizing the information obtained by intervening on variable $a$ is:
 
-\eq {
+\eqn {
 \hat{\mu}_a &=
 \begin{cases}
+\label{eqn:marginalizedEstimator2}
 \frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
@@
 \end{cases}\\
-\text{where:}\\
+}
+where $\frac{m_{a,b}}{n_{a,b}}$ is an empirical estimator for $P(Y|X_i = j,do(X_a = b))$,
+\eqn{
 m_{a,b}   &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
-n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s \\
+n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s 
 }
@@
 
-\eq {
+\eqn {
+\label{eqn:pooledEstimators}
 \hat{\mu} &= \sum^{K}_{a=1}w_a\hat{\mu}_a=w_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} w_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
@@
 
-If $p$ is not known, these expression are unchanged except that $p_a$ is replaced with $\hat{p}_a$
+If $p$ is not known $p_a$ could be replaced with its empirical estimate $\hat{p}_a$
 
-\eq{
+\eqn{
 \hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
 }
-\subsection{Observe then exploit}
+
+Now I want to select weights $w$ to minimize the variance in $\hat{\mu}$ and then get some kind of high probability bound for it.
+However, if I'm not very careful about how I pool things in equation \ref{eqn:pooledEstimators}, the estimator could get very biased, ie estimators $\hat{\mu_a}$ where I have data only for one state of the variable $a$ are not very informative, so should tend to get low weight, but if we choose to sample the second state based on previous payoffs - then we could end up assigning high weights to those estimators that had fluctuated above the true value...
+
+This complicates getting correct uncertainty bounds for a UCB type algorithm/proof.
+
+Consider, what exactly you mean by not unbiased ...
+
+\section{Importance Sampling Approach}
+
+Consider down weighting estimators proportional to the number of times we played them
+
+\section{Observe then exploit}
+
+\subsection{Basic idea}
 
@@
 
-\section{Observe until number of plausibly optional arms $< \alpha$}
+\subsection{Alternate Observe and Exploit With Some Decay}
+Keep track of two separate bounds and always choose the narrower of the two.
+
+\subsection{Observe until number of plausibly optional arms $< \alpha$}
 
@@
 
-\subsection{Regret during observe phase}
+\subsubsection{Regret during observe phase}
 
@@
 
-\subsection{Regret during UCB stage}
+Clearly, this is not quite the right criteria to decide when to switch ...
+\subsubsection{Regret during UCB stage}
 
@@
 
-\subsection{General rough thoughts}
+Suppose we have $O_i$ observations for arm $i$.
+We will now begin doing standard UCB - storing a separate set of bounds, only each time we choose an action we will pick the narrower of the observational/interventional confidence bounds.
+
+\eqn{
+\hat{\epsilon}_{it} &= \min \left( \sqrt{\frac{\alpha \log(t)}{2 T_{it}}},\sqrt{\frac{log(1/\delta)}{2 O_i}} \right)\\
+\label{eqn:combinedBoundsEpsilon}
+& = \sqrt{\frac{\alpha \log(t)}{2 \max(T_{it},O_i)}} \;\; \text{if we let $\delta = t^{-\alpha}$}
+}
+
+When running UCB, in order to select a non-optimal arm $i$ at timestep $t$, at least one of the following statements must be true:
+
+\begin{enumerate}
+\item $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu_i + \Delta_i$
+\item $\hat{\mu}_{it} + \hat{\epsilon}_{it} \geq \mu_i + 2\hat{\epsilon}_{it}$
+\item $\Delta_i < 2 \hat{\epsilon}_{it}$
+\end{enumerate}
+
+Given our expression for the confidence bound in equation \ref{eqn:combinedBoundsEpsilon} we can write statement (3) as:
+\eqn{
+\max(T_{it},O_i) < \frac{2\alpha \log(t)}{\Delta_i^2} := \beta
+}
+
+If $O_i \geq \beta$, statement (3) is guaranteed to be false for all $t$ and we can bound the expected number of times we select $i$ with how many times we expect the bounds given by (1) and (2) to be true.
+
+For arms $i$ such that $O_i \geq \beta$
+
+\eqn{
+&P(\text{1 is true}) < t^{1-\alpha}\\
+&P(\text{2 is true}) < t^{1-\alpha}\\
+& \implies P(I_t = i) < 2t^{1-\alpha}\\
+& \implies R_n(i) \leq \frac{\Delta_i \alpha}{\alpha -2}
+}
+
+For all other arms with $\Delta_i > 0$
+\eqn{
+R_n(i) \leq  \frac{2\alpha \log(n)}{\Delta_i} + \frac{\Delta_i \alpha}{\alpha -2} 
+}
+
+Assume we observed for $h$ steps. The expected regret from this period is just $h E[\Delta_i]$. Let $T = h+n$ be the total time.
+
+\eqn{
+R_T < \sum_{i=1}^K \Delta_i \left(\frac{h}{K} + \frac{\alpha}{\alpha-2}\right) + \sum_{i:\Delta_i > 0}P(O_i \leq \beta)\frac{2\alpha \log(n)}{\Delta_i}
+}
+
+Using Hoeffding's Inequality:
+
+\eqn {
+P(O_i \leq \beta) \leq \begin{cases}
+2 e^{-2h(\frac{1}{2}-\frac{\beta}{h})^2} & \text{if $h > 2\beta$} \\
+1 & \text{otherwise}
+\end{cases}
+}
+
+Its tricky - selecting the $h$ to switch at based on the inequality can certainly yield a non-optimal answer.
+Maybe I should figure this out via simulation??
+But some theory to tell me what to simulate would certainly help.
+
+\pagebreak
+
+\subsubsection{General rough thoughts}
 

### 2015-03-13 causal-bandit/c-bandit.tex
fixes a couple of unclear bits in ucb regret section

@@
 Clearly, this is not quite the right criteria to decide when to switch ...
-\subsubsection{Regret during UCB stage}
+\subsubsection{Regret on observing for h rounds before switching to UCB}
 
-On completion of the observe stage we have tight bounds on all the arms except the first $\alpha$
-
-Essentially, we want to know how large the regret is if we initialize with a certain number of counts already in place...
+How large the regret is if we initialize with a certain number of counts already in place...
 
@@
 \eqn{
-\max(T_{it},O_i) < \frac{2\alpha \log(t)}{\Delta_i^2} := \beta
+\max(T_{it},O_i) < \frac{2\alpha \log(t)}{\Delta_i^2}
+}
+
+Let:
+
+\eqn{
+\beta = \frac{2\alpha \log(n)}{\Delta_i^2} 
 }
@@
 
-For all other arms with $\Delta_i > 0$
+For all other arms with $\Delta_i > 0$, the bound is the same as standard UCB
 \eqn{

### 2015-04-10 causal-bandit/c-bandit.tex
lots of changes - laptop has battery issues

@@
 }
+\pagebreak
 
@@
 Clearly, this is not quite the right criteria to decide when to switch ...
+
+What about if I
+
 \subsubsection{Regret on observing for h rounds before switching to UCB}
@@
 
+\subsection{Importance weighted estimators}
+
+Now Hoeffdings bound gives us $ P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}|n_i)$
+
+\eqn{
+P(\Delta_{\hat{i^*}} > D) & \leq K \sum_{\boldsymbol{n}}  P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}|n_i)P(\boldsymbol{n}) \\
+& = K \sum_{\boldsymbol{n}}  e^{-n_iD^2/2}P(\boldsymbol{n})
+}
+
+Where $\boldsymbol{n} = [n_1...n_K]$ is a vector of the number of observations we have for each arm and $P(\boldsymbol{n}) \sim multinomial(\boldsymbol{p},h)$ To approximate this, we break the space of $\boldsymbol{n}$ up into those vectors where $n_i > h/4 \forall i$ and those where $\exists i$ such that $n_i < h/4$.
+
 \bibliographystyle{plainnat}

### 2015-04-10 causal-bandit/observe_then_best.tex
lots of changes - laptop has battery issues

\documentclass{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{tikz} 
\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
%\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=1em,
           minimum height=1em,
           font=\footnotesize,
           text centered,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\begin{document}
\def\ci{\perp\!\!\!\perp}

\section*{Observe then pick best}
\begin{itemize}

\item Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.

\item The goal is to bound the psuedo-regret upto a total number of timesteps $T$

\item Our algorithm will explore by playing uniformly at random for $h$ timesteps and then select the arm with the highest estimated reward for the remaining timesteps $T-h$.

\item We assume each arm corresponds to setting one binary input variable $X_j$ to a given value. All the input variables are assumed to be independent causes of the binary reward variable $Y$. With this structure the probability of a reward is the same under the observation that a variable takes a given configuration as under the action that assigns it. Therefore, for each exploration timestep, we get data on the performance of half of the arms.

\item We assume $P(X_j = 1) = \frac{1}{2} \forall j$. With this assumption we will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$. Relaxing this assumption will require us to have a more targeted exploration phase - as otherwise we do not gain any information about the value of arms that do not occur naturally with reasonable probability. 
\end{itemize}

\subsection*{Regret during explore phase}
Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
\eqn{
R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta]
}

\subsection*{Regret during exploit phase}
The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.

\eqn{
\hat{i^*} = argmax_i [\hat{\mu}_i]
}
\eqn{
R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
}

The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$. Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
\eqn{
R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
}
The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.

Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$. If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$

\eqn{
\label{eqn:probDeltaToLarge}
P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
}

If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution. Instead we will use an importance weighted estimator.

\eqn{
\label{eqn:importance_weighted_estimator}
\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
}

where $q_i = P(\text{arm $i$ active})$

Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:

\eqn{
\label{eqn:balancedHoeffdings}
P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
}

\eqn{
R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
}

Let $D = \sqrt{\frac{8}{h}\log(hk)}$

\eqn{
R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hk)} + \frac{1}{h}\right)
}

\subsection*{Total Regret}

Putting together the regret from the exploration and exploitation phases,

\eqn{
R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hk)} + \frac{1}{h}\right)\\
& \leq h + T\left(\sqrt{\frac{8}{h}\log(Tk)} + \frac{1}{h}\right)
}

Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,

\eqn {
R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
}

If $T \geq 2$ and $K \geq 2$, the first term dominates and,

\eqn {
R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
}

The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$

\subsection*{Empirical results}
\begin{figure}[H]
\centering
\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
\includegraphics[width=.5\textwidth]{explore_exploit}
\end{figure}

\pagebreak

\section{Generalizing to unbalanced $\boldsymbol{q}$}

\subsection{Option 1: Targeted sampling during exploration phase}

The key fact we were utilizing to draw conclusions about multiple arms during each timestep of the explore phase is that, given our assumed causal structure, $P(Y|do(X_i=j)) = P(Y|X_i=j)$

If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ some specified number of times $\tau_{ab}$, then we can no longer estimate $P(Y|X_i=j)$ simply by from the proportion of successes given $X_i=j$.

\eqn {
P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
 &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
 \label{eqn:marginalizedEstimator}
 &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
}

Define $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$

Let $\hat{\mu}_a^{ij}$ be an estimator for $\mu^{ij}$ based on samples where the intervention was on variable $a$.
\eqn {
\hat{\mu}_a^{ij} &=
\begin{cases}
\label{eqn:marginalizedEstimator2}
\frac{1}{q_i(j)}\left(\frac{m_{a,1}^{ij}}{\tau_{a1}}q_a+\frac{m_{a,0}^{ij}}{\tau_{a0}}(1-q_a)\right) & \text{if } a \neq{i} \\
\frac{m_{i,j}^{ij}}{\tau_{ij}} & \text{if } a=i \\
\end{cases}\\
}

where

\eqn{
m_{a,b}^{ij} &= \sum_{s \in \{t:I_t=(a,b)\}} Z_{ab,s}^{ij} \text{ and,}\\
Z_{ab,s}^{ij} &= \ind{X_{i,s} = j,Y_s=1} \in \{0,1\}
}

For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$. We will pool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.

\eqn{
\hat{\mu}^{ij} = \sum_{a=1}^K w_{a} \hat{\mu}_{a}^{ij} \;\;\text{, where } \sum_{a=1}^K w_a = 1
}

Putting everything together,

\eqn{
\hat{\mu}^{ij} = \frac{w_i}{\tau_{ij}}\sum_{s \in \{t: I_t = (i,j)\}} Z_{ij,s}^{ij} + \sum_{a \neq i} \left(\frac{w_a}{q_i(j)}\left[ \frac{q_a}{\tau_{a1}}\sum_{s \in \{t:I_t=(a,1)\}} Z_{a1,s}^{ij}+  \frac{1-q_a}{\tau_{a0}}\sum_{s \in \{t:I_t = (a,0)\}} Z_{a0,s}^{ij} \right] \right)
}

We now need to show $E[\hat{\mu}^{ij}] = \mu^{ij}$ and get a high probability bound for their difference. For the latter, lets try and use McDiarmids Inequality.

McDiarmid's Inequality states:
If $Z_i \ci Z_j$ and 
\eqn{
|\phi(Z_1...Z_i...Z_N) - \phi(Z_1...Z_i^{'}...Z_N)| < c_i \;\; \forall i
}
\eqn{
\label{eqn:McDiarmid}
& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq 
\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
\right) \leq \delta
}

For our problem, $\hat{\mu}^{ij} = \phi(\boldsymbol{Z}^{ij})$. All the $Z's$ are independent as they correspond indicator functions over the results at different timesteps.

\eqn{
|\phi(...Z_{ij,s}^{ij}...) - \phi(...Z_{ij,s}^{'ij}...)| & \leq \frac{w_i}{\tau{ij}}\;\;\;\; \leftarrow \tau_{ij} \text{ such $Z$'s}\\
|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a q_a}{q_i(j)\tau_{a1}} \;\;\;\; \leftarrow \tau_{a1} \text{ such $Z$'s for each $a$}\\
|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a (1-q_a)}{q_i(j)\tau_{a0}} \;\;\;\; \leftarrow \tau_{a0} \text{ such $Z$'s for each $a$}
}

\eqn{
\label{eqn:sum_ci}
\sum_l c_l^2 = & \frac{w_i^2}{\tau_{ij}} + \sum_{a \neq i}\frac{w_a^2}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}} +\frac{(1-q_a)^2}{\tau_{a0}}  \right)\\
= & \sum_{a=1}^K w_a^2 f(a) \text{, where} \\
f(a) =& \begin{cases}
\frac{1}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}}+  \frac{(1-q_a)^2}{\tau_{a0}}\right)  & a \neq i\\
\frac{1}{\tau_{ij}} & a = i
\end{cases} 
}

We want to select weights to minimize equation \ref{eqn:sum_ci} so as to achieve as tight a bound as possible in equation \ref{eqn:McDiarmid}.

Applying Legrange Multipliers

\eqn {
w_a & = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
\label{eqn:min_sum_ci}
\sum_i c_i^2 & = \frac{1}{\sum_a \frac{1}{f(a)}}
}

Substitution \ref{eqn:min_sum_ci} back into \ref{eqn:McDiarmid}

\eqn {
& P\left(|\hat{\mu}^{ij} - \mu^{ij}| \geq \epsilon \right) \leq 2 \exp \left(-2 \epsilon^2 \sum_{a=1}^K \eta_a^{ij} \right)\\
& \eta_a^{ij} = \begin{cases}
\frac{\tau_{a1} \tau_{a0} q_i(j)^2}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2} & a \neq i\\
\tau_{ij} & a = i
\end{cases} \\
}

Substituting this into \ref{eqn:probDeltaToLarge}

\eqn{
P(\Delta_{\hat{i^*}} > D) \leq 2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right)
}

\eqn{
R_2 \leq (T-h)\left(D+2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right) \right)
}

Looking at the second term

\eqn{
 \sum_{(i,j)} exp\left(-\frac{D^2}{2} \left( \tau_{ij} + q_{ij}^2 \sum_{a \neq i} \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}  \right) \right)
}

Now let $\tau_a = \tau_{a1}+\tau_{a0}$ then $\frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}$ is maximized if $\tau_{a1} = q_a \tau_a$

Changing the ratio will only otherwise effect the terms that have where $a = i$. So compare those two terms ...

\eqn{
exp(-\frac{D^2}{2}(\tau_{\alpha 1}+q_\alpha g) + exp(-\frac{D^2}{2}(\tau_{\alpha 0}+(1-q_\alpha)g)
}

minimize this subject to the constraint, hmm seems to give and on the face of it different answer.

Consider splitting it up so that we sum first over the terms where $j = 1$ then over the terms where $j = 0$ (may be clearer).

\pagebreak

\subsection{Option 2: Three stage algorithm}
\begin{enumerate}
\item Observe randomly
\item Play low probability arms
\item Pick best and exploit
\end{enumerate}

\pagebreak

\end{document}

### 2015-04-10 causal-bandit/regret_bounds.tex
lots of changes - laptop has battery issues

--- 2014-11-21 causal-bandit/ucb_regret.tex
@@
 
-Assume $\exists$ a convex function, $\psi$, such that:
+The basic setting is the stochastic bandit problem:
 
-\begin{equation}
-\begin{aligned}
-log(E[e^{\lambda(X-E[X]}]) & \leq \psi(\lambda)\\
-log(E[e^{\lambda(E[X]-X}]) & \leq \psi(\lambda)
-\end{aligned}
-\end{equation}
-This ensures that the moments of the distribution of X are defined. Assume that at time $t$ we select arm $I_t$ with the highest upper confidence bound:
+\begin{itemize}
+\item We have a set of arms $i \in \{1...K\}$
+\item For each arm $i$, there is an unknown distribution of rewards $P_{i}(X)$
+\item Each time we select a given arm, $i$, the reward is sampled i.i.d from $P_{i}(X)$.
+This is a big assumption - it states that the reward at a given timestep depends only on the action selected at that timestep, not on the sequence of previous actions.
+\item \color{red}Can the reward distributions be (fixed) functions of the timestep? (No I think - otherwise the idea of an optimal arm independent of t doesn't make sense)
+\end{itemize}
+
+Some notation:
+
+\begin{itemize}
+\item $K$ the number of arms
+\item $i$ identifies an arm, $i \in \{1...K\}$
+\item $i^*$ the arm with the highest true expected reward
+\item $I_t$ the arm selected by the algorithm at timestep $t$
+\item $\hat{\mu}_i$ an estimator for the expected reward of arm $i$ based on the sample mean
+\item $\mu_i$ the true expected reward for an arm $i$
+\item $\mu^*$ the true expected reward of  $i^*$ (the best arm)
+\item $\Delta_i = \mu^* - \mu_i$ how much worse arm $i$ is than the best arm
+\item $T_i(s) = \sum_{t=1}^s \ind{I_t = i}$ the number of times arm $i$ was selected upto timestep $s$
+\item $\hat \epsilon_{it}$ an estimate of the uncertainty in the empirical estimator for the expected reward
+
+\end{itemize}
+
+The goal is to get a bound on the pseudo-regret, defined as:
+
+\eqn{
+R_n  = & n\mu^* - E\left[\sum_{t = 1}^n \mu_{It}\right] \\
+\label{eq:regretK}
+= & \sum_{i=1}^K \Delta_i E[T_i(n)] 
+}
+
+\subsection*{UCB-1}
+
+For simplicity just consider Bernoulli bandits: $P(X_i) \sim Bernoulli(p_i)$.
+Hoeffding's inequality gives us a high probability bound on the how much our sample based estimate of the expected reward can be below the true expected reward.
+Let  $\hat \mu_{is} = \frac{1}{s}\sum_{t=1}^s X_t$ be the sample average and $\mu_{i} = E[P_i(X)]$ if we select arm $i$ a fixed number of times $s$:
+
+\eqn {
+\label{eq:hoeff1}
+& P( \mu_{i} - \hat \mu_{is}  > \epsilon) \leq e^{-2s\epsilon^2} \\
+\label{eq:hoeff2}
+ \implies & P\left(\mu_{i} - \hat \mu_{is} > \sqrt{\frac{\log(1/\delta)}{2s}}\right) \leq \delta
+}
+
+\fbox{\begin{minipage}{\textwidth}%
+\textbf{The UCB Algorithm}
+
+Define the upper confidence bound for each arm $i$ at timestep $t$ as:
+
+\eqn {
+ucb_{it} & = \hat \mu_{it} + \sqrt{\frac{2 \log(t)}{T_i(t)}}\\
+& =  \hat \mu_{it} + \hat \epsilon_{it}
+}
+
+At time $t$ select arm $I_t$ with the highest upper confidence bound:
 
@@
 \label{eq:armSelection}
-I_{t} = argmax_{i=1...K}\left[\hat{\mu}_{it} + (\psi^{*})^{-1}\frac{\alpha log(t)}{T_{it}}\right]
+I_{t} = argmax_{i=1...K}\left(ucb_{it} \right)
 \end{equation}
 
-Then if $\alpha > 2$,
+\eqn{
+\label{eq:regretBound}
+R_n & \leq \sum_{i:\Delta i > 0}\left(\frac{8\log(n)}{\Delta_i}+(1+\frac{\pi}{3})\Delta_i\right)\\
+R_n & < O(\sqrt{K n \log(n)}), \; \text{ provided $n >> K$} 
+}
 
-\begin{equation}
-R_n \leq \sum_{i:\Delta i > 0}\left(\frac{\alpha \Delta i}{\psi^{*}(\Delta i/2)}log(n)+\frac{\alpha}{\alpha-2}\right)
-\end{equation}
+\end{minipage}}
+
+The confidence bound we use in the UCB algorithm is clearly related but not identical to the bound in equation \ref{eq:hoeff2}.
+The difference is due to the fact that when we use the UCB algorithm, the number of times each arm is selected is not fixed in advance but depends on the results of previous actions.
+
+We now want to prove the bound in equation \ref{eq:regretBound} holds.
 
@@
 \begin{enumerate}
-\item The estimated UCB on the best arm, $i^*$, is less than or equal to the actual reward for that arm: $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \hat \mu$
-\item The estimated reward for arm $i$ is greater than or equal to the estimated CI higher than the true reward for that arm: $\hat \mu_{it} \geq \mu_i + \hat \epsilon_{it}$
-\item The number of times we have selected arm $i$ in previous timesteps, $T_{it}$, is less than some bound (that grows logarithmically with $n$). $T_{it} < \frac{\alpha log(n)}{\psi^*(\Delta i/2)}$
+\item The estimated UB on the best arm, $i^*$, is less than or equal to the actual reward for that arm:
+\eq{
+\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu_i+\Delta_i
+}
+\item The UB on the non-optimal arm is too high: 
+\eqn{
+\hat \mu_{it}+ \hat \epsilon_{it} \geq \mu_i + 2 \hat \epsilon_{it}
+}
+\item The uncertainty bound is too wide compared the difference between the payoff of this arm and the optimal one.
+Since $\hat{\epsilon}_{it}$ is a function of the number of times we have selected arm $i$, this can also be thought of as 'we have not selected arm $i$ enough yet'.
+\eqn{
+\Delta_i < 2\hat{\epsilon}_{it}  
+}
 \end{enumerate}
@@
 Assume statements 1-3 are all false.
-\eq {
-3. & \implies T_{it} > \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \\
-  & \implies \Delta i > 2 (\psi^*)^{-1} \frac{\alpha log(n)}{T_{it}} \geq 2 (\psi^*)^{-1} \frac{\alpha log(t)}{T_{it}} = 2 \hat \epsilon_{it}\\
-  & \implies \Delta i > 2 \hat \epsilon_{it} \\
-1. & \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \hat \mu  = \mu_i+\Delta i > \mu_i +  2 \hat \epsilon_{it}\\
-& \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \mu_i +  2 \hat \epsilon_{it}\\
-2. & \implies \hat \mu_{it} < \mu_i + \hat \epsilon_{it} \\
-   & \implies \hat \mu_{it} + \hat \epsilon_{it} < \mu_i +2 \hat \epsilon_{it} \\
-   & \implies \hat \mu_{it} + \hat \epsilon_{it} <  \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \longleftarrow \text { UCB for arm $i < $ UCB for arm $i^*$, which contradicts equation (\ref{eq:armSelection}})
+
+\eq{
+& UB(i^*) \equiv \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} >  \mu_i+\Delta_i > \mu_i+2\hat{\epsilon}_{it} \\
+& UB(i)\equiv \hat{\mu}_{it} + \hat{\epsilon}_{it} < \mu_i+2\hat{\epsilon}_{it}\\
+& \implies UB(i) < UB(i^*) \; \text{ which contradicts the statement that we will play arm $i$, $I_t = i$}
 }
@@
 
-If statements 1. and 2. are both false, then statement 3. places a bound on the number of times we can previously have selected the incorrect arm $i$ in order to select it in this timestep.
+Suppose we had selected a non-optimal arm $i$ in all timesteps until $\gamma$, where we choose a value $\gamma$ that ensures statement 3 is false.
+In the remaining time-steps, we can then only select arm $i$ if either statement 1 or statement 2 is true.
+
+\eq{
+E\left[T_{i}(n)\right] & \leq \gamma + E\left[\sum_{t=\gamma+1}^{n} \mathds{1} \{\text{(1) or (2) is true} \}\right] \leftarrow \text{since if (3) is false, (1) or (2) must be true}\\[3pt]
+& \leq \gamma + \sum_{t=\gamma+1}^{n} \left[ \mathds{P}(\text{(1) is true}) + \mathds{P}(\text{(2) is true})  \}\right]
+}
+
+\eq {
+P(\text{(1) is true})  = & P(\hat{\mu}_{i^*t} + \sqrt{\frac{2 \log(t)}{T_i(t)}} \leq \mu^*)\\
+\leq & P(\exists s \in \{1...t\} : \hat{\mu}_{i^*s}+ \sqrt{\frac{2 \log(t)}{s}} \leq \mu^*) \color{red} \leftarrow \text {to get around the problem that $T_i(t)$ is random}\\
+\leq & \sum_{s=1}^t P\left( \hat{\mu}_{i^*s}+ \sqrt{\frac{2 \log(t)}{s}} \leq \mu^* \right) \color{red} \leftarrow \text{union bound}\\
+\text{ From equation (\ref{eq:hoeff2}) we have:}\\
+& P\left(\mu_{i} - \hat \mu_{is} > \sqrt{\frac{\log(1/\delta)}{2s}}\right) \leq \delta \text{, letting $\delta = t^{-4}$,}\\
+\implies & P\left( \hat{\mu}_{i^*s}+\sqrt{\frac{2 \log(t)}{s}}\leq \mu^* \right) < t^{-4}\\
+\implies & P(\text{(1) is true}) \leq \sum_{s=1}^t t^{-4} = t*t^{-4} =  t^{-3}
+}
+
+Similarly, $P(\text{(2) is true}) \leq t^{-3}$ so we have proved:
+
+\eqn {
+ E\left[T_{i}(n)\right] & \leq \gamma + \sum_{t=\gamma+1}^n 2t^{-3}
+}
+
+Plugging this into the definition of pseudo-regret in equation \ref{eq:regretK} gives:
+
+\eqn {
+R_n \leq \sum_{i=1}^K \Delta_i  \left(\gamma + \sum_{t=\gamma+1}^n 2t^{-3}\right) 
+}
 

### 2015-04-10 causal_inference_without_do.tex
lots of changes - laptop has battery issues

\documentclass{article}

\usepackage[hscale=0.8,vscale=0.8]{geometry}
\usepackage{graphicx}
\usepackage{amsfonts, amsmath, amsthm, amssymb} % For math fonts, symbols and environments

\pagenumbering{gobble}

\author{Finnian Lattimore}
\title{Causal Inference in one page}

\begin{document}
\def\ci{\perp\!\!\!\perp}

\subsection*{Causal Inference without the do-opperator}

One way we can encode our assumptions about the state of the world now and how a given intervention modifies that state is with graphical models. Specifically, assume that the data we have is generated from a directed acyclic graph (DAG) and that we know the structure of the graph (although we may not be able to observe all the variables). An arrow $A \rightarrow B$ means $A$ causes $B$.

If we intervene in the system and force a particular variable, $X$ to some value, $x$, the result is a new graph that is identical to the original one except that the conditional probability distribution associated with $X$ is replaced by $X=x$ and any links coming into $X$ are removed (since under the assumption that some external intervention is forcing $X=x$ it will no longer depend on its original causes). The classic example for the question of the causal effect of smoking on lung cancer is shown in figure XXX.

We want to estimate the probability of cancer given smoking after the intervention (ie in the graph $G_X$) using observational data (data generated by the original model $G$). The problem of causal inference in general is to determine if it is possible to estimate the desired conditional probabilities in the graph representing the system after a given intervention from conditional probabilities in the original graph.

This can be determined by repeated applications of Pearl's 3 rules, which we will restate (without the do notation). For simplicity we will only consider the effect of intervention on a single variable, $X$, on a single target $Y$.

\subsubsection*{Rule 1}
 D-separation still applies in the graph $G_X$:

 \begin{equation}
 Y \text{ d-sep } W \text{ given } Z \text{ in } G_X \implies P_{G_X}(Y|Z,W) = P_{G_X}(Y|Z)
 \end{equation}

\subsubsection*{Rule 2}
If, after conditioning on $Z$, $G$ and $G_X$ have exactly the same active trails between $X$ and $Y$ then $P(Y|X,Z)$ will be the same in both graphs $P_G(Y|X,Z) = P_{G_X}(Y|X,Z)$.

Equivalently, we can state this in terms of D-separation by adding an additional variable $\hat{X}$ as a parent of $X$ in the original graph. If $\hat{X}$ d-sep $Y$ given $X$ and $Z$ in $G$ then there are no active trails from $X$ to $Y$ that begin with an arrow into $X$. This means no active trails will be removed in the process of creating graph $G_X$. You can think of the variable $\hat{X}$ as a switch that determines if $X$ takes its value as determined by its other parents or from some external source.

\begin{equation}
Y\text{ d-sep }\hat{X}\text{ given }X,Z \implies P_G(Y|X,Z) = P_{G_X}(Y|X,Z)
\end{equation}

\subsubsection*{Rule 3}
If, after conditioning on $Z$, the only active trails from $X$ to $Y$ in the original graph $G$ begin with an arrow pointing into $X$ then $Y$ will be independent of $X$ given $Z$ after the intervention (since we will have deleted all the arrows going into $X$).

Stated in terms of D-separation:
\begin{equation}
Y \text{ d-sep } \hat{X} \text{ given } Z \implies P_{G_X}(Y|X,Z) = P_{G_X}(Y|Z)
\end{equation}

\subsubsection*{These rules are complete}
We can estimate the conditional distribution of interest in $G_X$ if and only if we can re-write it in terms of observable conditional independences in $G$ via repeated application of these rules.

You might have been hoping for a single, simple graphical criterion for identification - but sadly, to my knowledge, there isn't one. There are some that are sufficient and some that are necessary - but none that are both.

\subsubsection*{The causal effect of smoking}

\subsection*{Causal Inference}
Causal inference covers problems where we want to predict the outcome of an intervention given some known causal structure
\begin{itemize}
\item Assume that the true causal structure is a directed graphical model where $X \rightarrow Y$ means $X$ causes $Y$. We know the structure of the network but may not be able to observe the values of all the variables.
\item Intervening to set $X = x$, denoted $do(X=x)$ in a causal network, has the effect of removing all links coming into the variable $X$ (since its value is no longer determined by its parents) and setting its value to $x$.
\item We can use d-separation to derive the 3 rules of the do-calculus that allow us write queries about interventions in terms of the pre-interventional distribution. 
\item These rules are complete.  If an expression containing do's can be converted into one that does not by repeated application of the rules then you can determine the post-interventional distribution from observational data. If you can't, its not possible without additional assumptions.
\end{itemize}
The rules (simplified to cover only the case where we intervene on a single variable) can all be illustrated by the graph in figure \ref{fig:dorules}. Assume our query is $P(Y|do(X))$. $G^{\dagger}$ is the original graph, $G$, augmented to include a variable $\hat{X}$  that acts as a switch determining if $X$ takes its value as determined by its parents or from an external intervention. $G_{\overline{X}}$ is $G$ after the intervention.

\begin{figure}[h]
\caption{The do calculus}
\label{fig:dorules}
\centering
\includegraphics[scale=0.9]{do_rules_figure-crop}
\end{figure}

\begin{enumerate}
\item D-separation still applies after an intervention. $Y \ci V|X$ in $G_{\overline{X}}$ $\implies P(Y|do(X),V) = P(Y|do(X))$
\item If the outcome variable is independent of the way in which the intervened on variable takes its value, intervention = observation. $Y \ci \hat{X}|X,Z$ in $G^{\dagger} \implies P(Y|do(X),Z) = P(Y|X,Z)$
\item If there is no direct path from the intervened on variable to the outcome variable, the intervention does not effect the distribution of the outcome variable. $Y \ci \hat{X}|U$ in  $G^{\dagger} \implies P(Y|do(X),U) = P(Y|U)$
\end{enumerate}
In figure \ref{fig:dorules} we can identify the causal effect of $X$ on $Y$ provided we can observe either $Z$ or $U$.

Key general references: \cite{Pearl2000,Koller2009}

\subsection*{Causal Structure Learning - learn causal structure from observational data}
\subsubsection*{Conditional independence approach}

\begin{itemize}
\item Assume that the data was generated by an unknown causal directed graphical model
\item Assume that all dependencies in the graph are reflected by conditional dependencies in the distribution of the data (ie no dependencies exactly cancel one-another out) $\leftarrow$ faithfullness assumption
\item Finding the true causal structure is equivalent to finding the set of graphs that are perfect maps for the data distribution. In general, result is not unique but some links may have the same direction in all graphs in the set.
\item Approach can be extended to allow for latent variables
\item Key general reference: \cite{Sprites}
\end{itemize}
\subsubsection*{Functional form approach}
Conditional independence based methods cannot distinguish between graphs with the same dependency structure, for example,  $X \rightarrow Y$ vs $Y \rightarrow X$. We can infer causality in such cases if we make braod assumptions about the relationship between the function mapping cause to effect and any noise. Suppose $X \rightarrow Y$:
\begin{itemize}
\item If we assume noise is additive in the causal direction, $Y = f(X)+\epsilon$ cannot generally be inverted to give an additive model in the other direction (linear $f$ and gaussian $\epsilon$ are an unfortunate exception)\cite{Hoyer2009}.
\item If $f$ is deterministic and invertible, for most input distributions $p_X(x)$ the distribution of $p_Y(y)$ will be higher where $f'$ is small and a large region of $X$ maps to similar values of $Y$. We expect that $f$ and $p_X$ are independent but $f'$ and $p_Y$ are correlated \cite{Daniusis2010}.
\item In general, $P(X)$ and $P(Y|X)$ should be independent but $P(Y)$ and $P(X|Y)$ will not be \cite{Scholkopf2012}
\end{itemize}

\bibliographystyle{plain} % Plain referencing style
\bibliography{library} % Use the example bibliography file sample.bib
\end{document}

### 2015-04-10 one_page_summary.tex
lots of changes - laptop has battery issues

--- 2015-04-10 causal_inference_without_do.tex
@@
 \def\ci{\perp\!\!\!\perp}
-
-\subsection*{Causal Inference without the do-opperator}
-
-One way we can encode our assumptions about the state of the world now and how a given intervention modifies that state is with graphical models.
-Specifically, assume that the data we have is generated from a directed acyclic graph (DAG) and that we know the structure of the graph (although we may not be able to observe all the variables).
-An arrow $A \rightarrow B$ means $A$ causes $B$.
-
-If we intervene in the system and force a particular variable, $X$ to some value, $x$, the result is a new graph that is identical to the original one except that the conditional probability distribution associated with $X$ is replaced by $X=x$ and any links coming into $X$ are removed (since under the assumption that some external intervention is forcing $X=x$ it will no longer depend on its original causes).
-The classic example for the question of the causal effect of smoking on lung cancer is shown in figure XXX.
-
-We want to estimate the probability of cancer given smoking after the intervention (ie in the graph $G_X$) using observational data (data generated by the original model $G$).
-The problem of causal inference in general is to determine if it is possible to estimate the desired conditional probabilities in the graph representing the system after a given intervention from conditional probabilities in the original graph.
-
-This can be determined by repeated applications of Pearl's 3 rules, which we will restate (without the do notation).
-For simplicity we will only consider the effect of intervention on a single variable, $X$, on a single target $Y$.
-
-\subsubsection*{Rule 1}
- D-separation still applies in the graph $G_X$:
-
- \begin{equation}
- Y \text{ d-sep } W \text{ given } Z \text{ in } G_X \implies P_{G_X}(Y|Z,W) = P_{G_X}(Y|Z)
- \end{equation}
-
-\subsubsection*{Rule 2}
-If, after conditioning on $Z$, $G$ and $G_X$ have exactly the same active trails between $X$ and $Y$ then $P(Y|X,Z)$ will be the same in both graphs $P_G(Y|X,Z) = P_{G_X}(Y|X,Z)$.
-
-Equivalently, we can state this in terms of D-separation by adding an additional variable $\hat{X}$ as a parent of $X$ in the original graph.
-If $\hat{X}$ d-sep $Y$ given $X$ and $Z$ in $G$ then there are no active trails from $X$ to $Y$ that begin with an arrow into $X$.
-This means no active trails will be removed in the process of creating graph $G_X$.
-You can think of the variable $\hat{X}$ as a switch that determines if $X$ takes its value as determined by its other parents or from some external source.
-
-\begin{equation}
-Y\text{ d-sep }\hat{X}\text{ given }X,Z \implies P_G(Y|X,Z) = P_{G_X}(Y|X,Z)
-\end{equation}
-
-\subsubsection*{Rule 3}
-If, after conditioning on $Z$, the only active trails from $X$ to $Y$ in the original graph $G$ begin with an arrow pointing into $X$ then $Y$ will be independent of $X$ given $Z$ after the intervention (since we will have deleted all the arrows going into $X$).
-
-Stated in terms of D-separation:
-\begin{equation}
-Y \text{ d-sep } \hat{X} \text{ given } Z \implies P_{G_X}(Y|X,Z) = P_{G_X}(Y|Z)
-\end{equation}
-
-\subsubsection*{These rules are complete}
-We can estimate the conditional distribution of interest in $G_X$ if and only if we can re-write it in terms of observable conditional independences in $G$ via repeated application of these rules.
-
-You might have been hoping for a single, simple graphical criterion for identification - but sadly, to my knowledge, there isn't one.
-There are some that are sufficient and some that are necessary - but none that are both.
-
-\subsubsection*{The causal effect of smoking}
-
+\setlength\parindent{0pt}
 \subsection*{Causal Inference}
@@
 \item If there is no direct path from the intervened on variable to the outcome variable, the intervention does not effect the distribution of the outcome variable.
-$Y \ci \hat{X}|U$ in  $G^{\dagger} \implies P(Y|do(X),U) = P(Y|U)$
+$Y \ci \hat{X}|U,Z$ in  $G^{\dagger} \implies P(Y|do(X),U,Z) = P(Y|U,Z)$ (note: we condition on $Z$ because conditioning on $U$ opens the path $\hat{X} \leftarrow Z \rightarrow Y$)
 \end{enumerate}
-In figure \ref{fig:dorules} we can identify the causal effect of $X$ on $Y$ provided we can observe either $Z$ or $U$.
+
+A single application of the rules show we can identify the causal effect of $X$ on $Y$ if we can observe $Z$.
+By repeated application of the rules the query is also identifiable if we can observe only $U$.
 

### 2015-04-10 bandit_utils.py

@@ -22,6 +22,9 @@ class BernoulliEstimator(object):
         self.fail+= (1-reward)
         self.n +=1
 
+
+    
+
    
 def ucbBound(model,alpha,horizon):
     t = np.arange(horizon)
@@ -54,7 +57,7 @@ class TrivialProbabilityModel(object):
         
 
     def sampleX(self):
-        """ sample a vector X from P(X)"""
+        """ sample a vector X from P(X). Returns a vector of length numVars specifying the value (0 or 1) of each variable"""
         return self.px.rvs(size=self.numVars)
         
     def sampleYGivenX(self,X):

### 2015-04-10 information_directed_sampling.py

@@ -1,0 +1,246 @@
+import scipy.integrate as integrate
+from scipy.stats import beta,bernoulli
+from scipy.optimize import minimize_scalar
+from numpy import *
+from matplotlib.pyplot import *
+
+def kl(p,q):
+    return p*log(p/q) + (1-p)*log((1-p)/(1-q))
+
+def M_ij(i,j):
+    # calculate inner integral for each value of t
+    f_inner = x*pdfs[j]
+    F_inner_t = integrate.cumtrapz(f_inner,x,initial=0)
+    
+    # now do the outer integral and return the result ...
+    #f = pdfs[i]*prod(cdfs,axis=0)/cdfs_dezerod[i]*F_inner_t #TODO Check this line - should be dividing by cdfs_dezerod[j] too
+    f = pdfs[i]*prod(cdfs,axis=0)*F_inner_t/(cdfs_dezerod[i]*cdfs_dezerod[j])
+    return 1.0/a[i]*integrate.simps(f,dx = 1/float(n_samples)) 
+
+class BetaBernoulliIDS:
+    def __init__(self, num_arms, prior, model = None, num_samples = pow(2,8)+1):
+        """ prior is a 2-value tuple indicating the shape of the beta prior over arm rewards
+            you can optionally pass in a model, which should be an array of bernoulli random variables of length num_arms"""
+        if model:
+            assert len(model) == num_arms
+            self.model = model
+        else:
+            self.model = [bernoulli(p) for  p in beta.rvs(prior[0],prior[1],size=num_arms)] # the true reward probability for each arm
+        self.K = num_arms
+        self.n = num_samples
+        self.best_reward = max([p.mean() for p in self.model])
+        self.beta1 = zeros(self.K)+prior[0] # number of successes + 1 
+        self.beta2 = zeros(self.K)+prior[1] # number of fails + 1
+        self.x = linspace(0,1,num_samples)
+        self.posteriors = [beta(self.beta1[i],self.beta2[i]) for i in range(self.K)]
+        self.pdfs = asarray([p.pdf(self.x) for p in self.posteriors]) # creats a num_arms * n_samples array, where each row is the pdf for an arm
+        self.cdfs = asarray([p.cdf(self.x) for p in self.posteriors])
+        self.cdfs_dezerod = copy(self.cdfs)
+        self.cdfs_dezerod[self.cdfs==0] = 1
[…]

### 2015-04-10 libpgm.py

@@ -1,0 +1,31 @@
+# we will start just covering binary models. Not optimized for performance, assumes we will work with reasonably small networks.
+
+
+class CPD():
+    def __init__(self,inlinks):
+        
+
+class Graph():
+    self.initialized = False
+
+    def add_link(self,start_node,end_node):
+        
+
+    def initialize(self):
+        """Must be called after all links have been created """
+        self.initialized = True
+
+
+    def setCPD(self,node):
+        assert(self.initialized,True,"Graph not initialized")
+
+    
+
+    
+
+
+    
+        
+    
+
+    

### 2015-04-10 slearn.py

@@ -1,0 +1,42 @@
+from pgmpy.models import BayesianModel
+from pgmpy.factors import TabularCPD
+import numpy as np
+
+
+
+def set_uniform_cpds(G, cards = {}):
+    """ cards is an optional mapping from variable name to cardinality.
+        assumed to be 2 for any variables for which its not specified"""
+    for n in G.nodes():
+        parents  = G.predecessors(n)
+        parents_card = [cards.get(p,2) for p in parents]
+        card = cards.get(n,2)
+        values = np.full((card,np.prod(parents_card)),1.0/card)
+        cpd = TabularCPD(n,card,values,parents,parents_card)
+        G.add_cpds(cpd)
+                    
+
+    
+# now I want to be able to do sampling and updating ...
+
+
+
+G = BayesianModel()
+G.add_edges_from([("X1","Y"),("X2","Y"),("X3","Y")])
+set_uniform_cpds(G)
+
+# idea
+
+#1) sample a model
+
+#2) sample a beleif about the state of that model from current data we have
+    # requires posterior not just maximum likelyhood estimates.
+
+    
+
+
+#3) take best action according to that belief
+
+#4) add result to data set
+
+

### 2015-04-10 tmp.py

@@ -1,0 +1,9 @@
+class Test:
+    def todo(self,thing):
+        print thing
+    def todo(self):
+        print "nothng"
+
+t = Test()
+t.todo(1)
+t.todo()

### 2015-04-22 causal-bandit/grandad_problem.tex
updates observe exploit theory to include mixture of perfectly balanced and unbalanced arms. Adds greedy search code for optimal solution

--- 2015-04-10 causal-bandit/observe_then_best.tex
@@
 
-\section*{Observe then pick best}
-\begin{itemize}
-
-\item Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
-
-\item The goal is to bound the psuedo-regret upto a total number of timesteps $T$
-
-\item Our algorithm will explore by playing uniformly at random for $h$ timesteps and then select the arm with the highest estimated reward for the remaining timesteps $T-h$.
-
-\item We assume each arm corresponds to setting one binary input variable $X_j$ to a given value.
-All the input variables are assumed to be independent causes of the binary reward variable $Y$.
-With this structure the probability of a reward is the same under the observation that a variable takes a given configuration as under the action that assigns it.
-Therefore, for each exploration timestep, we get data on the performance of half of the arms.
-
-\item We assume $P(X_j = 1) = \frac{1}{2} \forall j$.
-With this assumption we will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage.
-Note that this is independent of the number of arms $K$.
-Relaxing this assumption will require us to have a more targeted exploration phase - as otherwise we do not gain any information about the value of arms that do not occur naturally with reasonable probability.
-\end{itemize}
-
-\subsection*{Regret during explore phase}
-Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
-\eqn{
-R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta]
-}
-
-\subsection*{Regret during exploit phase}
-The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
+The goal is, given a vector $[q_1...q_N]$ and a constant $A$, find an assignment to $\boldsymbol{\tau}$ to minimize (or approximately minimize) the following function subject to the constraint that the $\tau$'s sum to $1$, $\sum_{i=1}^N(\tau_{i0}+\tau_{i1}) = h$
 
 \eqn{
-\hat{i^*} = argmax_i [\hat{\mu}_i]
-}
-\eqn{
-R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
-}
-
-The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$.
-Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
-\eqn{
-R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
-}
-The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.
-
-Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$.
-If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$
-
-\eqn{
-\label{eqn:probDeltaToLarge}
-P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
-}
-
-If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution.
-Instead we will use an importance weighted estimator.
-
-\eqn{
-\label{eqn:importance_weighted_estimator}
-\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
-}
-
-where $q_i = P(\text{arm $i$ active})$
-
-Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:
-
-\eqn{
-\label{eqn:balancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
+f(\boldsymbol{\tau}) =  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \sum_{a \neq i} c_a  )}+\sum_{i=1}^N e^{-A ( \tau_{i0} + (1-q_{i})^2 \sum_{a \neq i} c_a)}
 }
@@
 \eqn{
-R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
+\text{,where }c_a :=  \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}
 }
 
-Let $D = \sqrt{\frac{8}{h}\log(hk)}$
+$\boldsymbol{\tau}$ is a vector of positive integers of length $2N$, $[\tau_{10},\tau_{11},\tau_{20},\tau_{21}...\tau_{N0},\tau_{N1}]$
 
-\eqn{
-R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hk)} + \frac{1}{h}\right)
-}
+$[q_1...q_N]$ is a vector of length $N$, where $0 \leq q_i \leq 1$ for all $i$
 
-\subsection*{Total Regret}
+$A$ is constant, $0 < A < 1$
 
-Putting together the regret from the exploration and exploitation phases,
-
-\eqn{
-R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hk)} + \frac{1}{h}\right)\\
-& \leq h + T\left(\sqrt{\frac{8}{h}\log(Tk)} + \frac{1}{h}\right)
-}
-
-Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,
-
-\eqn {
-R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
-}
-
-If $T \geq 2$ and $K \geq 2$, the first term dominates and,
-
-\eqn {
-R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
-}
-
-The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$
-
-\subsection*{Empirical results}
-\begin{figure}[H]
-\centering
-\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
-\includegraphics[width=.5\textwidth]{explore_exploit}
-\end{figure}
-
-\pagebreak
-
-\section{Generalizing to unbalanced $\boldsymbol{q}$}
-
-\subsection{Option 1: Targeted sampling during exploration phase}
-
-The key fact we were utilizing to draw conclusions about multiple arms during each timestep of the explore phase is that, given our assumed causal structure, $P(Y|do(X_i=j)) = P(Y|X_i=j)$
-
-If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ some specified number of times $\tau_{ab}$, then we can no longer estimate $P(Y|X_i=j)$ simply by from the proportion of successes given $X_i=j$.
-
-\eqn {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- \label{eqn:marginalizedEstimator}
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
-}
-
-Define $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$
-
-Let $\hat{\mu}_a^{ij}$ be an estimator for $\mu^{ij}$ based on samples where the intervention was on variable $a$.
-\eqn {
-\hat{\mu}_a^{ij} &=
-\begin{cases}
-\label{eqn:marginalizedEstimator2}
-\frac{1}{q_i(j)}\left(\frac{m_{a,1}^{ij}}{\tau_{a1}}q_a+\frac{m_{a,0}^{ij}}{\tau_{a0}}(1-q_a)\right) & \text{if } a \neq{i} \\
-\frac{m_{i,j}^{ij}}{\tau_{ij}} & \text{if } a=i \\
-\end{cases}\\
-}
-
-where
-
-\eqn{
-m_{a,b}^{ij} &= \sum_{s \in \{t:I_t=(a,b)\}} Z_{ab,s}^{ij} \text{ and,}\\
-Z_{ab,s}^{ij} &= \ind{X_{i,s} = j,Y_s=1} \in \{0,1\}
-}
-
-For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$.
-We will pool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{a=1}^K w_{a} \hat{\mu}_{a}^{ij} \;\;\text{, where } \sum_{a=1}^K w_a = 1
-}
-
-Putting everything together,
-
-\eqn{
-\hat{\mu}^{ij} = \frac{w_i}{\tau_{ij}}\sum_{s \in \{t: I_t = (i,j)\}} Z_{ij,s}^{ij} + \sum_{a \neq i} \left(\frac{w_a}{q_i(j)}\left[ \frac{q_a}{\tau_{a1}}\sum_{s \in \{t:I_t=(a,1)\}} Z_{a1,s}^{ij}+  \frac{1-q_a}{\tau_{a0}}\sum_{s \in \{t:I_t = (a,0)\}} Z_{a0,s}^{ij} \right] \right)
-}
-
-We now need to show $E[\hat{\mu}^{ij}] = \mu^{ij}$ and get a high probability bound for their difference. For the latter, lets try and use McDiarmids Inequality.
-
-McDiarmid's Inequality states:
-If $Z_i \ci Z_j$ and 
-\eqn{
-|\phi(Z_1...Z_i...Z_N) - \phi(Z_1...Z_i^{'}...Z_N)| < c_i \;\; \forall i
-}
-\eqn{
-\label{eqn:McDiarmid}
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq 
-\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
-\right) \leq \delta
-}
-
-For our problem, $\hat{\mu}^{ij} = \phi(\boldsymbol{Z}^{ij})$. All the $Z's$ are independent as they correspond indicator functions over the results at different timesteps.
-
-\eqn{
-|\phi(...Z_{ij,s}^{ij}...) - \phi(...Z_{ij,s}^{'ij}...)| & \leq \frac{w_i}{\tau{ij}}\;\;\;\; \leftarrow \tau_{ij} \text{ such $Z$'s}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a q_a}{q_i(j)\tau_{a1}} \;\;\;\; \leftarrow \tau_{a1} \text{ such $Z$'s for each $a$}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a (1-q_a)}{q_i(j)\tau_{a0}} \;\;\;\; \leftarrow \tau_{a0} \text{ such $Z$'s for each $a$}
-}
-
-\eqn{
-\label{eqn:sum_ci}
-\sum_l c_l^2 = & \frac{w_i^2}{\tau_{ij}} + \sum_{a \neq i}\frac{w_a^2}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}} +\frac{(1-q_a)^2}{\tau_{a0}}  \right)\\
-= & \sum_{a=1}^K w_a^2 f(a) \text{, where} \\
-f(a) =& \begin{cases}
-\frac{1}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}}+  \frac{(1-q_a)^2}{\tau_{a0}}\right)  & a \neq i\\
-\frac{1}{\tau_{ij}} & a = i
-\end{cases} 
-}
-
-We want to select weights to minimize equation \ref{eqn:sum_ci} so as to achieve as tight a bound as possible in equation \ref{eqn:McDiarmid}.
-
-Applying Legrange Multipliers
-
-\eqn {
-w_a & = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
-\label{eqn:min_sum_ci}
-\sum_i c_i^2 & = \frac{1}{\sum_a \frac{1}{f(a)}}
-}
-
-Substitution \ref{eqn:min_sum_ci} back into \ref{eqn:McDiarmid}
-
-\eqn {
-& P\left(|\hat{\mu}^{ij} - \mu^{ij}| \geq \epsilon \right) \leq 2 \exp \left(-2 \epsilon^2 \sum_{a=1}^K \eta_a^{ij} \right)\\
-& \eta_a^{ij} = \begin{cases}
-\frac{\tau_{a1} \tau_{a0} q_i(j)^2}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2} & a \neq i\\
-\tau_{ij} & a = i
-\end{cases} \\
-}
-
-Substituting this into \ref{eqn:probDeltaToLarge}
-
-\eqn{
-P(\Delta_{\hat{i^*}} > D) \leq 2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right)
-}
-
-\eqn{
-R_2 \leq (T-h)\left(D+2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right) \right)
-}
-
-Looking at the second term
-
-\eqn{
- \sum_{(i,j)} exp\left(-\frac{D^2}{2} \left( \tau_{ij} + q_{ij}^2 \sum_{a \neq i} \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}  \right) \right)
-}
-
-Now let $\tau_a = \tau_{a1}+\tau_{a0}$ then $\frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}$ is maximized if $\tau_{a1} = q_a \tau_a$
-
-Changing the ratio will only otherwise effect the terms that have where $a = i$. So compare those two terms ...
-
-\eqn{
-exp(-\frac{D^2}{2}(\tau_{\alpha 1}+q_\alpha g) + exp(-\frac{D^2}{2}(\tau_{\alpha 0}+(1-q_\alpha)g)
-}
-
-minimize this subject to the constraint, hmm seems to give and on the face of it different answer.
-
-Consider splitting it up so that we sum first over the terms where $j = 1$ then over the terms where $j = 0$ (may be clearer).
-
-\pagebreak
-
-\subsection{Option 2: Three stage algorithm}
-\begin{enumerate}
-\item Observe randomly
-\item Play low probability arms
-\item Pick best and exploit
-\end{enumerate}
-
-\pagebreak
+$h$ is a positive constant. You can assume $h > 2N$.
 

### 2015-04-22 causal-bandit/observe_then_best.tex
updates observe exploit theory to include mixture of perfectly balanced and unbalanced arms. Adds greedy search code for optimal solution

@@
 
+TODO check/show this estimator is unbiased
+
 For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$.
@@
 \eqn{
+\label{eqn:targeted_prob_delta_too_big}
 P(\Delta_{\hat{i^*}} > D) \leq 2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right)
@@
 
-Looking at the second term
+Looking at equation \ref{eqn:targeted_prob_delta_too_big}, let $A = \frac{D^2}{2}$ and
 
 \eqn{
- \sum_{(i,j)} exp\left(-\frac{D^2}{2} \left( \tau_{ij} + q_{ij}^2 \sum_{a \neq i} \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}  \right) \right)
+c_a =  \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}
 }
 
-Now let $\tau_a = \tau_{a1}+\tau_{a0}$ then $\frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}$ is maximized if $\tau_{a1} = q_a \tau_a$
+\eqn{
+f(\boldsymbol{\tau}) := \frac{1}{2}P(\Delta_{\hat{i^*}} > D)  =  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \sum_{a \neq i} c_a  )}+\sum_{i=1}^N e^{-A ( \tau_{i0} + (1-q_{i})^2 \sum_{a \neq i} c_a)}
+}
 
-Changing the ratio will only otherwise effect the terms that have where $a = i$. So compare those two terms ...
+The goal is to find some assignment $\boldsymbol{\tau}(\boldsymbol{q})$ to minimize this subject to the constraint that $\sum_i \tau_{i1} + \sum_i \tau_{i0} = h$
+
+No luck attempting to do this directly (via Lagrange multipliers ...)
+
+Note, letting $\tau_a = \tau_{a1}+\tau_{a0}$ then $c_a$ is maximized to equal $\tau_a$ if $\tau_{a1} = q_a \tau_a$
+
+Consider the case where we have only 2 variables.
 
 \eqn{
-exp(-\frac{D^2}{2}(\tau_{\alpha 1}+q_\alpha g) + exp(-\frac{D^2}{2}(\tau_{\alpha 0}+(1-q_\alpha)g)
+f(\boldsymbol{\tau}) = e^{-A(\tau_{11} + q_1^2 c_2)}+
+e^{-A(\tau_{21} + q_2^2 c_1)}+
+e^{-A(\tau_{10} + (1-q_1)^2 c_2)}+
+e^{-A(\tau_{20} + (1-q_2)^2 c_1)}
 }
 
-minimize this subject to the constraint, hmm seems to give and on the face of it different answer.
+If we consider the extreme case where $q_1=q_2=1$, ie the arms $X_1=0$ and $X_2=0$ never occur naturally.
 
-Consider splitting it up so that we sum first over the terms where $j = 1$ then over the terms where $j = 0$ (may be clearer).
+\eqn{
+f(\boldsymbol{\tau}) = 2 e^{-A(\tau_{11} + \tau_{21})}+
+e^{-A(\tau_{10})}+
+e^{-A(\tau_{20})}
+}
+
+Which is minimized subject to the constraint $\tau_{11}+\tau_{10}+\tau_{21}+\tau_{20}=h$ if we let $\tau_{10}=\tau_{20} = \frac{Ah-log(2)}{3A}$.
+Note that as $h$ gets large, we play the arms that do not occur naturally roughly $\frac{1}{3}$ of the time as opposed to $\frac{1}{4}$ if we were to play uniformly.
+We only learn anything about the arms that do not naturally occur when we play them directly.
+For the other arms we learn something all the time (provided we have some level of balance in all the other arms...) so it makes sense to play the arms that don't occur naturally a little bit more to keep the overall bounds tight.
+
+Ok, what happens if we just let $\tau_{ij} = \frac{h}{K}$
+
+\eqn {
+c_a = \frac{h}{K}\left(\frac{1}{(1-q_a)^2+q_a^2} \right) \geq \frac{h}{K}
+}
+
+\eqn {
+f \leq \sum_{i=1}^N \left(e^{-A(\frac{h}{K}+q_i^2\frac{(K-1)}{K}h)}+ e^{-A(\frac{h}{K}+(1-q_i)^2\frac{(K-1)}{K}h)}\right) \\
+\leq \sum_{i=1}^N \left(e^{-Ah}+ e^{-A(\frac{h}{K})}\right)
+}
+
+Basically worst case, we learn about half the arms fast.
+
+More generally, it would seem by targeted sampling we should be able to learn all the arms with reasonable natural probability fast - and the remaining arms at the rate $h/K$ (or better if we play them more often).
+
+Approaches to finding some better sampling function.
+
+\begin{enumerate}
+\item Find something motivated by the two variable example and show that it performs better than the uniform sampling approach.
+\item  Consider some other edge cases 
+\end{enumerate}
+
+Lets consider the case $q_1 = 1, q_2...q_N = \frac{1}{2}$ We will have $\tau_{11},\tau_{10},\tau_{ij} = \tau \; \forall i > 1,j$
+
+\eqn {
+c_a = \begin{cases}
+\tau_{11} & \text{ if } a=1\\
+2 \tau & \text{ otherwise}
+\end{cases}
+}
+
+\eqn{
+f(\boldsymbol{\tau}) & = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\tau + \frac{1}{4}(\tau_{11}+2(N-2)\tau))} \\
+& = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})} \\
+& <  e^{-A\tau_{10}} + (2 (N-1)+1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}\\
+& =  e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}
+}
+
+and the constraint is $\tau_{11}+\tau_{10}+2(N-1)\tau = h$
+
+If we maximize $(\frac{\tau_{11}}{4} + \frac{N\tau}{2})$ subject to $\tau_{11}+2(N-1)\tau = (h-\tau_{10})$ we find $\tau = \frac{h-\tau_{10}}{2(N-1)}$ and $\tau_{11} = 0$.
+Note, as $N$ gets large the sensitivity to the trade off between $\tau$ and $\tau_{11}$ becomes negligible.
+
+\eqn {
+\frac{\tau_{11}}{4} + \frac{N\tau}{2} = \begin{cases}
+\frac{(h-\tau_{10})N}{4(N-1)} & \text{if } \tau_{11}=0 \\
+\frac{(h-\tau_{10})N}{4(N-1)} - \frac{1}{4(N-1)} &\text{if } \tau_{11}=(h-\tau_{10})
+\end{cases}
+}
+
+\eqn {
+f(\boldsymbol{\tau}) & <e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{1}{4}(h-\tau_{10}))}
+}
+
+Which is minimized if 
+\eqn {
+\tau_{10} = \frac{h}{5} - \frac{4}{5A}log(\frac{2N-1}{4})
+}
+
+Ok - lets look at the reverse case, where $q_1=\frac{1}{2}, q_2...q_N = 1$
+
+\eqn{
+\sum_a c_a = \begin{cases}
+(N-1)\tau_{11} & \text{ if } a = 1 \\
+2\tau+(N-2)\tau_{11} &\text{ otherwise }
+\end{cases}
+}
+
+The constraint is 
+\eqn {
+2\tau+(N-1)\tau_{11} + (N-1)\tau_{10} = h
+}
+
+\eqn{
+f(\boldsymbol{\tau}) = 2e^{-A(\tau+\frac{1}{4}(N-1)\tau_{11})} + (N-1)e^{-A(2\tau+(N-1)\tau_{11})}+(N-1)e^{-A\tau_{10}}
+}
+
+Holding $\tau_{10}$ constant, the 2nd and 3rd terms are constant. The first is maximized by letting $\tau_{11} = 0, \tau = \frac{h - (N-1)\tau_{10}}{2}$
+
+\eqn{
+f(\boldsymbol{\tau}) & = 2e^{-A\tau} + (N-1)e^{-2A\tau}+(N-1)e^{-A\frac{h - 2\tau}{N-1}} \leftarrow \text{ can't minimize analytically} \\
+\label{eqn:case1balancedapprox}
+& \leq (N+1)e^{-A\tau} +(N-1)e^{-A\frac{h - 2\tau}{N-1}}
+}
+
+which is minimized if
+
+\eqn {
+\tau & = \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N(N+1)}{2(N-1)})\right) \\
+\label{eqn:case1balancedapproxsolution}
+& \sim \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N}{2})\right)
+}
+
+Equation \ref{eqn:case1balancedapproxsolution} is also the solution that leads to the exponentiated terms in equation \ref{eqn:case1balancedapprox} being equal - which is should intuitively be optimal since the coefficients in front of them are very similar.
+
+Yet another case study - if $q_1...q_N = 1$ So none of the arms $X_i = 0$ occur naturally.
+
+\eqn{
+&c_a  = \tau_1 \; \forall a\\
+&\sum_{i \neq a}c_a  = (N-1)\tau_1 \; \forall a\\
+& N(\tau_0+\tau_1) = h\\
+& f(\boldsymbol{\tau}) = N e^{-A N\tau_1} + N e^{-A\tau_0}
+}
+
+minimized if 
+\eqn{
+\tau_0 = \frac{h}{N+1}-\frac{log(N)}{A(N+1)}
+}
+
+Note as $N$ get large - this equates to $\tau_0 \sim \frac{h}{N}$ ie almost all the actions are allocated to the arms that do not occur naturally.
+
+Another case study $q_1...q_{N_1} = 1,q_{N_1+1}...q_N = \frac{1}{2}$ (should encompass all previous examples ...)
+
+\eqn{
+&c_a  = \begin{cases}
+\tau_{11} &\text{ if } a \leq N_1
+ \\
+2\tau & \text{otherwise}
+\end{cases}\\
+&\sum_{i \neq a}c_a  = \begin{cases}
+(N_1-1)\tau_{11} + 2 N_2 \tau & \text{ if } a \leq N \\
+N_1\tau_{11} + 2(N_2-1)\tau & \text{ otherwise }
+\end{cases}\\
+&N_1\tau_{10}+N_1\tau_{11}+2N_2\tau = h\\
+&f(\boldsymbol{\tau}) =  N_1 e^{-A\tau_{10}}+ N_1 e^{-A(N_1\tau_{11}+2N_2\tau)} + 2N_2 e^{-A(\frac{N_1}{4}\tau_{11}+\frac{N_2+1}{2}\tau)}
+}
+
+The 2nd and 3rd terms are minimized by letting $\tau_{11} = 0$ (although the result is only weakly dependnent on the trade off between $\tau_{11}$ and $\tau$ provided $N_2 >> 2$).
+
+\eqn{
+f(\boldsymbol{\tau}) &=  N_1 e^{-A\tau_{10}}+ N_1 e^{-A2N_2\tau} + 2N_2 e^{-A\frac{N_2+1}{2}\tau}\\
+& \leq N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A\frac{N_2+1}{2}\tau}\\
+& = N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A(\frac{(N_2+1)(h-N_1\tau_{10})}{4N_2})} 
+}
+
+\eqn{
+\tau_{10} = \left(\frac{N_2+1}{(N_2+1)N_1+4N_2}\right)h-\frac{4N_2\log(\frac{(N_2+1)(2N_2+N_2)}{4N_2})}{A((N_2+1)N_1+4N_2))}
+}
+
+It seems like this function can be optimized in a greedy way.
+That is, start with $\boldsymbol{\tau} = \boldsymbol{0}$, and in each timestep select the $\tau_{ij}$ that leads to the greatest reduction in the error bound.
+This appears to yield broadly consistent results to the analytic solution.
+
+\includegraphics[width=.5\textwidth]{greedy_minimization_close}
+\includegraphics[width=.5\textwidth]{greedy_minimization}
+
+Ok - so when we have many unbalanced variables, we assign almost all actions to the arms corresponding to the rare side.
+However, this will still be $< h/N$ which is only a factor of $2$ better than uniform assignment.
+Doing something smarter than uniform assignment will make more difference where we have a only a few unbalanced variables.
+In this case - smart assignment will lead to bounds almost as tight as the perfectly balanced case.
+Uniform assignment may yield solutions that are almost as bad as the worst case.
+(still better than without using the structure at all)
+
+I still need to cover more fractional $q$'s...
+
+What about if I repeated one of the above analysis but let the unbalanced q = .75?
+
+What can I say overall about worst/best cases/any symmetries in the problem??? Do these solutions lead to the terms in the exponentials being similar?
+
+Goal is to choose something that will give us a simple bound and will be ok for the worst case.
+Its still not entirely clear to me what the worst case will be - after all it depends on the way you assign to $\tau$.
+
+What happens if we assume $\sum_{a\neq i} c_a$ is constant ... this is reasonable if $N$ is large
+
+When you look at each term in $c_a$, its maximized if we play the arms according to their natural probabilities.
+If the arm is balanced, this will also minimize the sum of the two terms for doing $x_i = 0$ and $x_i = 1$.
+If the arm is unbalanced, then we have to trade of balancing this sum versus contributing to the sum of $c_a$.
+Consider just how bad the sum of $c_a$ can be even if we focus entirely on balancing the two terms ...
+maybe its already enough.
+
+Fundamentally the relationship between arms is complex. Its not that we get any direct link from one arm to another - rather pairs of actions tell us something about other arms.
 
@@
 
-\subsection{Option 2: Three stage algorithm}
+\subsection{Option 2: Some measure of information gain}
+
+\subsection{Option 3: Three stage algorithm}
 \begin{enumerate}

### 2015-04-22 binary_independence.py

@@ -1,0 +1,7 @@
+import numpy as np
+for p in np.arange(0,1.2,.2):
+    for q in np.arange(0,1.2,.2):
+        print "----------"
+        print (1-p)*(1-q),(1-p)*q
+        print p*(1-q), p*q
+        print "----------"

### 2015-05-01 causal-bandit/observe_then_best.tex
explore exploit bounds for unbalanced but known q

@@
 \usepackage{pgfplots}
-\usepackage[ruled]{algorithm}
-\usepackage{algpseudocode}
 \usepackage{dsfont}
@@
 \begin{itemize}
+\item Avi's point - of all the space of possible functions $f(X_1...X_N)$, most of them will have very similar marginals- so limited?
+opportunity for regret?
+Does this mean we effectively only care about a particular subset of the functions where the marginals are significantly different.
+We could assume we are in that subspace (if we are not then all actions are equally good anyway)
+
+\item How do assumptions/priors about the function $f(X_1...X_N)$ flow through to information about the marginals?%
 
@@
 
+\section{Generalizing to unbalanced $\boldsymbol{q}$}
+
+When some arms have low natural probability we cannot rely on exploring them adequately by pure observation. We need to explicitly play them during the exploration phase.
+
+We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
+
+Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
+\leq q_N$.
+Let $m \in [2,N] = \{m:q_m > \frac{1}{m}\}$ Ie if the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
+If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$
+
+Suppose we observe for the first $h/2$ timesteps. This is at worst half the optimal.
+
+We then have estimates
+
+\eqn {
+\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y=1,X_i=1}}{\frac{h}{2}q_{ij}}
+}
+
+We take this as our estimate for those arms for which $q_{ij} > \frac{1}{m}$
+
+For these arms the Hoeffdings gives
+
+\eqn{
+P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/2m^2}
+}
+
+We play each of the remaining $m$ arms $\frac{h}{2m}$ times so for them we get
+
+\eqn{
+P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m}
+}
+
+So for all the arms
+
+\eqn{
+P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m^2}
+}
+
+and
+
+\eqn{
+\label{eqn:unbalancedHoeffdings}
+P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/4m^2}
+}
+
+If we let $D = \sqrt{\frac{4m^2\log(hK)}{h}}$
+
+\eqn{
+R_T  \leq h + T\left(\sqrt{\frac{4m^2\log(hK)}{h}} + \frac{1}{h}\right)
+}
+
+Let $h = T^{2/3}m^{2/3}\log(hk)$
+
+\eqn {
+R_T  \leq 4T^{\frac{2}{3}}m^{2/3}(log(KT))^{\frac{1}{3}}
+}
+
+Note, I think the $m^{2/3}$ should be improvable to something close to $m^{1/3}$ is I use Berstein's instead of Hoeffdings to bound the estimator for the arms where $q > 1/m$ but I haven't got the equations to quite work out for that yet.
+
+It should also be possible to generalize to handle the case were the $q$'s are unknown, since we should be able to get a reasonable estimate for them while we are observing.
+They key will be how the resulting uncertainty in $m$ effects the bounds.
+
+To get results that degrade to similar order bounds as UCB when the arms are very unbalanced, I will need to drop the explore/exploit strategy.
+
 \pagebreak
 
-\section{Generalizing to unbalanced $\boldsymbol{q}$}
+\end{document}
 
@@
 
-If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ some specified number of times $\tau_{ab}$, then we can no longer estimate $P(Y|X_i=j)$ simply by from the proportion of successes given $X_i=j$.
+If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ with some probability not equal to $P(X_a=b)$, then we can no longer estimate $P(Y|X_i=j)$ directly from the proportion of successes given $X_i=j$.
 
@@
  &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- \label{eqn:marginalizedEstimator}
+ \label{eqn:marginalizedProbability}
  &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b)
@@
 
-Define $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$
+Define:
+\begin{itemize}
+\item  $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$
+\item $q_{ab} = P(X_a = b)$, the underlying probability that variable $a$ takes the value $b$
+\item $p_{ab} = P(do(X_a=b)|I=a)$, the probability that we set variable $a$ to $b$ given we are acting on variable $a$
+\end{itemize}
 
-Let $\hat{\mu}_a^{ij}$ be an estimator for $\mu^{ij}$ based on samples where the intervention was on variable $a$.
-\eqn {
-\hat{\mu}_a^{ij} &=
-\begin{cases}
-\label{eqn:marginalizedEstimator2}
-\frac{1}{q_i(j)}\left(\frac{m_{a,1}^{ij}}{\tau_{a1}}q_a+\frac{m_{a,0}^{ij}}{\tau_{a0}}(1-q_a)\right) & \text{if } a \neq{i} \\
-\frac{m_{i,j}^{ij}}{\tau_{ij}} & \text{if } a=i \\
-\end{cases}\\
+\eqn{
+\hat{\mu}_a^{ij} = \begin{cases}
+\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
+\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
+\end{cases}
 }
 
-where
+This estimator is unbiased and has variance bounded by (proofs in section \ref{seq:proof_expect_var}) :
+
+\eqn{
+\label{eqn:variance}
+V[\hat{\mu}_a^{ij}] \leq \begin{cases}
+\frac{1}{p_{ij}} & \text{ if } a = i \\
+\frac{1}{q_{ij}}\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right) & \text{ if } a \neq i
+\end{cases}
+}
+
+At each time step, we select and arm $do(X_a=b)$ and then receive estimates for all the arms $i,j$ with variances given by \ref{eqn:variance}.
+Note that $\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right)$ is minimized to equal $1$ if we set the intervened on variable according to its natural probability, that is we effectively take no action.
+
+Define  $\bar{q}_i = min_j q_{ij}$, the probability of the most unlikely value for each variable.
+We can assume without loss of generality that $\bar{q}_1 \leq ...
+\leq \bar{q}_N$.
+Define $m = max(N,min\{m: \frac{1}{m}\leq \bar{q}_m\})$ We allocate half our exploration timesteps $\frac{h}{2}$ to observing and allocate the remainder equally to play the unbalanced arms with $i \leq m$.
+
+At the end of the exploration phase, we combine the estimators from each timestep as a weighted sum with weights selected to minimize the variance of the combined estimator.
+
+\eqn{
+\hat{\mu}^{ij} = \sum_{t=1}^h w_t \hat{\mu}_{a(t)}^{ij}, \text{ where } w_t = \frac{V_t^{-1}}{\sum_s V_s^{-1}}
+}
+
+We can then use Bernstein's inequality to bound the probability of large deviations
+
+\eqn{
+P\left(\sum_{t=1}^h w_t  \hat{\mu}_{a(t)}^{ij} - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
+}
+
+\eqn {
+\sum_{t=1}^h\frac{1}{V_t} & =  \begin{cases}
+\frac{h}{2}q_{kb} & \text{ if } q_{kb} \geq \frac{1}{m} \\
+\frac{h}{2}q_{kb} + \frac{h}{2m} & \text{ otherwise }
+\end{cases}\\
+& \geq \frac{h}{2m} \; \forall q_{kb}
+}
+
+\eqn{
+P\left(\sum_{t=1}^n w_t \hat{\mu}_{a(t)}^{ij} - \mu^{ij} \geq \frac{4}{3}\sqrt{\frac{4m\log(1/\delta)}{h}}\right) \leq \delta \; \forall (i,j)
+}
+
+Lower bound?
+
+\pagebreak
+
+\subsection{Proofs}
+\label{seq:proof_expect_var}
+\subsubsection{Expectation and variance of the estimators}
+\eq{
+\hat{\mu}_a^{ij} = \begin{cases}
+\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
+\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
+\end{cases}
+}
+
+If $a = i$:
+\eq{
+E[\hat{\mu}_a^{ij}] = \frac{1}{p_{ij}}E[Y\ind{X_i=j}] = \frac{1}{p_{ij}}P(Y,do(X_i=j)) = \frac{1}{p_{ij}}P(Y|X_i=j)P(do(X_i=j)) = P(Y|X_i=j)
+}
+
+\eq{
+V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2  = \frac{1}{p_{ij}^2}E[Y\ind{X_i=j}] - (\mu^{ij})^2  = \frac{1}{p_{ij}}\mu^{ij} - (\mu^{ij})^2 = \frac{\mu^{ij}(1-p_{ij}\mu^{ij})}{p_{ij}} \leq \frac{\mu^{ij}}{p_{ij}} \\
+\leq & \frac{1}{p_{ij}}
+}
+
+if $a \neq i$
+
+\eq{
+E[\hat{\mu}_a^{ij}] & = \frac{q_{a0}}{q_{ij}p_{a0}}E[Y\ind{X_i=j,X_a=0}]+\frac{q_{a1}}{q_{ij}p_{a1}}E[Y\ind{X_i=j,X_a=1}] \\
+& = \frac{q_{a0}}{q_{ij}p_{a0}}P(Y|X_i=j,do(X_i=0))q_{ij}p_{i0}+\frac{q_{a1}}{q_{ij}p_{a1}}P(Y|X_i=j,do(X_i=1))q_{ij}p_{i1} \\
+& =P(Y|X_i=j,do(X_i=0))P(X_i=0)+P(Y|X_i=j,do(X_i=1))P(X_i=1)\\
+& = P(Y|X_i=j)
+}
+
+\eq{
+V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2 \\
+ = & E[\ind{Y=1,X_i = j,X_a=0}](\frac{q_{a0}}{q_{ij}p_{a0}})^2 + E[\ind{Y=1,X_i = j,X_a=1}](\frac{q_{a1}}{q_{ij}p_{a1}})^2 \\
+& +  2 E[\ind{Y=1,X_i = j,X_a=0}\ind{Y=1,X_i = j,X_a=1}\frac{q_{a0}}{q_{ij}p_{a0}}\frac{q_{a1}}{q_{ij}p_{a1}}] -  (\mu^{ij})^2\\
+=& P(Y|X_i=j,X_a=0)q_{ij}p_{a0}(\frac{q_{a0}}{q_{ij}p_{a0}})^2 + P(Y|X_i=j,X_a=1)q_{ij}p_{a1}(\frac{q_{a1}}{q_{ij}p_{a1}})^2] - (\mu^{ij})^2 \\
+=& \frac{1}{q_{ij}}\left(P(Y|X_i=j,X_a=0)\frac{q_{a0}^2}{p_{a0}} + P(Y|X_i=j,X_a=1)\frac{q_{a1}^2}{p_{a1}}\right)- (\mu^{ij})^2  \\
+\leq & \frac{1}{q_{ij}}\left(\frac{q_{a0}^2}{p_{a0}} + \frac{q_{a1}^2}{p_{a1}}\right) \\
+}
+
+\subsubsection{Bernstein Bound}
+\begin{theorem}{Berstein's Inequality}
+Let $X_t \in \{X_1 ... X_n\}$  be independent, bounded random variables with mean $\mu$ and variance $V_t$ satisfying $|X_t - \mu| \leq b \; \forall t$. Let $V= \frac{1}{n}\sum_{t=1}^n V_t$ 
+\eq{
+& P(\sum_{t=1}^n X_t - \mu \geq \epsilon) \leq \exp(-\frac{\epsilon^2}{2nV+2b\epsilon/3})\\
+\implies & P\left(\sum_{t=1}^n X_t - \mu \geq \frac{2b\log(1/\delta)}{3} + \sqrt{2nV\log(1/\delta)}\right) \leq \delta
+}
+\end{theorem}
+Let $Z_t = w_t V_t$ where $w_t = V_t^{-1}/\sum V_t^{-1}$.
+\eq{
+V[Z_t] = \frac{V_t^{-1}}{(\sum V_t^{-1})^2}\\
+V = \frac{1}{n}\sum V[Z_t] = \frac{1}{n\sum V_t^{-1}}
+}
+
+\eq{
+& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2b_z\log(1/\delta)}{3} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta, \text{ where } b_z = max_t \frac{V_t^{-1}b}{\sum V^{-1}}
+}
+
+While we are observing, $|\hat{\mu}_{kb}| = |\frac{Y\ind{X_k=b}}{q_{kb}}|\leq \frac{1}{q_{kb}}$ and $V_t^{-1} = q_{kb}$ so $V_t^{-1}b = 1$.
+
+When we are explicitly setting $X_k=b$, $|\hat{\mu}_{kb}| = |Y|\leq 1$ and $V_t^{-1} =1$ so $V_t^{-1}b = 1$
+
+\eq{
+b_z = \frac{1}{\sum V_t^{-1}}
+}
+
+\eq{
+& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2\log(1/\delta)}{3\sum V_t^{-1}} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta \\
+\implies &  P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
+}
+
+\subsection{Notes}
+Based on equation \ref{eqn:marginalizedProbability}, a natural estimator for $\mu^{ij}$, given we acted to set the variable $a$, would be
+
+\eqn{
+\hat{\mu}_a^{ij} = \begin{cases}
+\frac{Y\ind{X_i=j}}{\ind{X_i=j}} & \text{ if } a = i \\
+\frac{Y\ind{X_i=j,X_a=0}}{\ind{X_i=j,X_a=0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{\ind{X_i=j,X_a=1}}q_{a1} & \text{ if } a\neq i
+\end{cases}
+}
+
+However, this forces us to deal with random variables (which may be 0) in the denominator.
+We can avoid this issue by using the importance weighted estimator, which replaces these indicators with their expectation.
+
+\eqn{
+\hat{\mu}_a^{ij} & = \begin{cases}
+\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
+\frac{Y\ind{X_i=j,X_a=0}}{q_{ij}p_{a0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{q_{ij}p_{a1}}q_{a1} & \text{ if } a\neq i
+\end{cases} \\
+&= \begin{cases}
+\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
+\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
+\end{cases}
+}
+
+\subsection{Previous attempt}
+
+For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$. We will pwhere
 
@@
 
-TODO check/show this estimator is unbiased
-
-For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$.
-We will pool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.
+TODO check/show this estimator is unbiasedool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.
 
@@
 
+Ok, suppose we spend half our actions, $\frac{h}{2}$, observing - that is selecting variables on which to act uniformly and setting them according to their natural probabilities.
+This contributes $\frac{h}{2N}$ to each $c_a$.
+So $\sum_{i\neq a} c_a \sim \frac{h}{2}$ provided $N-1 \sim N$.
+We don't allocate any more pulls to the more likely arm associated with each variable (this is at most a factor of 2 two few pulls).
+We now decide how to spend the remaining budget.
+
+We can assume without loss of generality that $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq ... \leq q_N$
+
+\eqn{
+f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+\sum_{i=1}^N e^{-A (\frac{h}{8})}
+}
+
+Evenly balanced allocation is $\frac{h}{2N}$.
+For those arms for which $q_i^2 \geq \frac{1}{N}$ they will already get this much information during the observe phase so we will not allocate them any more (again - at worst a factor or 2 two few).
+Lets just allocate the remainder evenly.
+So let $m = min\{m:m\leq N,\frac{1}{m} \leq q_m \}$
+
+\eqn{
+f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+N e^{-A (\frac{h}{8})}
+}
+
 \pagebreak
@@
 \end{enumerate}
-
-\pagebreak
-
-\end{document}

### 2015-05-20 causal-bandit/importance_sampling.tex
lots of changes - laptop has battery issues
prior to re-write of observeThenBest approach

@@
 
-\title{Estimators}
+\title{Concentration Inequalities}
 
@@
 \def\ci{\perp\!\!\!\perp}
-\maketitle
 
-Suppose we have samples $(X_t,Y_t)$ for $t = 1... n$ drawn from the joint distribution $P(X,Y)$ and we want to estimate $\mu = P(Y=1|X=1)$
+\section{Hoeffding's Inequality}
+If $X_1...X_n$ are independent, bounded random variables with $a_i < X_i < b_i$ and means $\mu_1...\mu_n$, and let $\mu = \frac{1}{n} \sum_{i=1}^n \mu_i$
 
-\section{The standard estimator}
+\eqn {
+P(\frac{1}{n}\sum_{i=1}^n X_i - \mu] \geq \epsilon) \leq \exp{\frac{-2n^2\epsilon^2}{\sum_{i=1}^n(b_i - a_i)^2}}
+}
 
+\section{Bernstein's Inequality}
+$X_1...X_N$ are independent, bounded random variables with $0 \leq X_i \leq E[X_i]+b$ Let $V = \sum[V[X_i]$
+
+\eqn{
+&P(\sum X_i - \sum E[X_i] \geq \epsilon) \leq e^{-\frac{\epsilon^2}{2\sum V[X_i]+2b\epsilon/3}} \implies P\left(\boldsymbol{.} \geq \frac{2b\log(1/\delta)}{3})+\sqrt{2\log(1/\delta)\sum V[X_i]}\right)\leq \delta\\
+&P(\frac{\sum X_i}{n} - \frac{\sum E[X_i]}{n} \geq \epsilon) \leq e^{-\frac{n\epsilon^2}{2\frac{1}{n}\sum V[X_i]+2b\epsilon/3}} \implies P\left(\boldsymbol{.} \geq \frac{2b\log(1/\delta)}{3n})+\frac{\sqrt{2\log(1/\delta)\sum V[X_i]}}{n}\right)\leq \delta
+}
+
+\subsection*{Estimating bounds for sums of bernoulli RV's with random denominator}
+Suppose we have samples $(X_t,Y_t)$ for $t = 1... n$ drawn from the joint distribution $P(X,Y)$, where $X$ and $Y$ are binary random variables and $P(X) = p$ and we want to estimate $\mu = P(Y=1|X=1)$
+
+\subsubsection*{Importance sampling estimate}
+ Let: 
+\eqn{
+&\hat{\mu} = \frac{1}{n}\sum_{t=1}^nZ_t, \text{ where }
+Z_t = \frac{Y_t \ind{X_t=1}}{p} \in \{0,\frac{1}{p}\}\\
+&W_t = pZ_t, \; W_t \in \{0,1\},\; V[W_t] \leq p
+}
+
+\eqn{
+&\text{Hoeff. } \implies P(\hat{\mu} - \mu \geq \epsilon) \leq e^{-2np^2\epsilon^2} 
+\implies P(\hat{\mu} - \mu \geq \frac{1}{p}\sqrt{\frac{1}{2n}\log{\frac{1}{\delta}}}) \leq \delta \\
+&\text{Berns. } \implies P(\mu -\hat{\mu}\geq \epsilon) \leq e^{-\frac{np\epsilon^2}{2(1+\epsilon/3)}} 
+\implies P(\mu - \hat{\mu}\geq \frac{2\log(1/\delta)}{3np}+\sqrt{\frac{2}{np}\log{\frac{1}{\delta}}}) \leq \delta
+}
+
+\subsubsection*{Standard Estimator}
+Let 
 \eqn{
@@
 
-We can't actually say that this estimator is unbiased - as it can be undefined if $n = 0$
+We have a problem if we don't have a lower bound on $\sum_{t=1}^n\ind{X_t=1}$ as the estimator may be undefined. Assume $s = \sum_{t=1}^n\ind{X_t=1} > 0$, then:
 
-\color{red}{Is there an unbiased estimator for this simple problem if the $P(X = 1)$ is not known?}\color{black}
-
-Let's consider variance and bounds.
+\eqn{
+\text{Hoeff.} \implies P(\hat{\mu} - \mu \geq \epsilon) \leq e^{-2s\epsilon^2}
+ \implies P\left(\hat{\mu} - \mu \geq \sqrt{\frac{1}{2s}\log(\frac{1}{\delta})}\right)
+}
 

### 2015-05-20 causal-bandit/observe_then_best.tex
prior to re-write of observeThenBest approach

@@
 
+\subsection{Generalizing to unknown q}
+\begin{itemize}
+\item Assume without loss of generality that $q_i \in [0,\frac{1}{2}]$
+
+\item Assume we observe for the first $h/2$ timesteps.
+
+\item Let $\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}$
+
+\item Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ We order the $\bar{q}$'s from smallest to largest and estimate $m$ with $\hat{m} = min_j : \bar{q}_j \geq \frac{1}{j}$
+
+\item The worst case $\boldsymbol{q}$ for a given fixed $m$ is 
+\eqn{
+q_i = \begin{cases}
+0 & \text{ if } i < m\\
+\frac{1}{m} & \text{ if } i \geq m
+\end{cases}
+}
+
+In this case $\bar{q}_0 ... \bar{q}_{m-1} = 0$
+
+For $i \geq m$
+\eq{
+P(q_i - \bar{q}_i \geq C) = P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})
+}
+
+The first term will dominate for all but very small $m$. Lets just work with that one for now.
+
+Via Bernstein's Inequality
+\eqn{
+P(q_i - \hat{q}_i \geq C) \leq \exp(-\frac{hC^2}{4q_i+4C/3}) := \gamma \\
+\implies P(q_i - \hat{q}_i \geq \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4q_i\log(1/\gamma)}{h}})\leq \gamma
+}
+
+Actually I think I can use the Lower tail version  of Bernsteins
+
+\eqn{
+P(q_i - \hat{q}_i \geq C) \leq \exp(-\frac{hC^2}{4q_i}) := \gamma \\
+\implies P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(1/\gamma)}{mh}})\leq \gamma
+}
+
+Define $\hat{m}_i = \frac{1}{\hat{q}_i}$.
+
+\eq{
+\hat{q}_i \geq q_i - C \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
+\implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}
+}
+
+If we let $\gamma =  \frac{1}{N}\exp(-{\frac{h}{16m}})$
+
+\eqn {
+C = 2\sqrt{\frac{m\log N}{hm^2}+\frac{1}{16m^2}} \leq \frac{1}{2m} \text{, if } h > 16m\log N \\
+}
+
+Then
+
+\eqn{
+ q_i - \hat{q}_i \leq \frac{1}{2m} \implies \hat{m}_i - m \leq m \\
+ \implies P(\hat{m}_i - m \geq m) \leq \frac{1}{N}\exp(-{\frac{h}{16m}})
+}
+
+If $\hat{m}_i - m \geq \varphi$ for at most $\varphi$ of the variables $i \geq m$, then $\hat{m} - m \leq \varphi$
+
+Let $W_i = \ind{\hat{m_i}-m \geq \varphi}$, $E[W_i] \leq \gamma$,$V[W_i] \leq \gamma$ 
+\eqn{
+P(\sum_{i=m}^N W_i \geq (N-m)\gamma + \varepsilon) \leq \exp(-\frac{\varepsilon^2}{2(N-m)\gamma+2\varepsilon/3})
+}
+Letting $(N-m)\gamma + \varepsilon = \varphi \implies \varepsilon = \varphi - (N-m)\gamma$
+\eqn{
+P(\hat{m}-m \geq \varphi) =  P(\sum_{i=m}^N W_i \geq \varphi) &\leq \exp({-\frac{(\varphi-(N-m)\gamma)^2}{2(N-m)\gamma + 2(\varphi-(N-m)\gamma)/3}})\\
+& \leq \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \text{, where } N\gamma < \varphi
+}
+
+If $P(\hat{m}-m \geq \varphi) \leq \zeta$.
+
+\eqn{
+P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq & P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \leq \varphi )P(\hat{m}-m \leq \varphi) \\
+& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \beta)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
+}
+
+Putting it together
+
+\eqn {
+P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}}) := \delta
+}
+The first term will dominate if
+
+\eqn{
+\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3} \geq \frac{h\epsilon^2}{m+\varphi}
+}
+
+Setting them equal and solving for $\varphi$. Roughly,
+
+\eqn{
+\varphi = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = \frac{1}{2}(\sqrt{4\epsilon^2 h+ m^2} - m)\\
+\implies \epsilon = \sqrt{\frac{\log(1/\delta)(\log(1/\delta)+m)}{h}} \leq \frac{\log(1/\delta)}{\sqrt{h}} + \sqrt{\frac{m}{h}\log(1/\delta)}
+}
+
+Now repeating the above but more exactly ...
+
+\eqn{
+\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}  \geq & \frac{\varphi^2-2\varphi N\gamma}{4\varphi/3} \; \text{, if } \varphi > 3N\gamma \\
+= & \frac{3}{4}(\varphi - 2N\gamma)
+}
+
+\eqn{
+\frac{3}{4}(\varphi - 2N\gamma) = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = &\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}\\
+\leq & \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}
+}
+
+If we let $\varphi =\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2} $
+
+\eqn{
+P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp(-\frac{3}{4}(\varphi - 2N\gamma))\\
+\implies \epsilon = \sqrt{\frac{4\log^2(2/\delta)+(3m+6N\gamma)\log(2/\delta)}{3h}}\\
+\leq \frac{2\log(2/\delta)}{\sqrt{3h}}+\sqrt{\frac{m+2N\gamma}{h}\log(2/\delta)}
+}
+Ok - but I still need to ensure there exists a $\gamma$ such that  $N\gamma < \frac{\varphi}{3}$ and $P(\hat{m}_i - m > \varphi) \leq \gamma$
+
+\eqn{
+P(\hat{m}_i - m > a) \leq \exp(-\frac{h a^2}{4m(m+a)^2})
+}
+
+It feels like I have just gone round and round in circles and pushed the problem to somewhere else ... my expression for $\varphi$ now contains both $\epsilon$ and $\gamma$ ...
+
+\eqn{
+\varphi \geq \frac{1}{2}\sqrt{m^2+4N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}
+}
+
+\end{itemize}
+
 \pagebreak
+
+\eqn{
+ \implies & C = \frac{\varphi}{m(\varphi+m)}  \geq \frac{\varphi}{\alpha m^2} \text{ if } \varphi < (\alpha-1)m\\
+\implies & P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4\log(1/\gamma)}{mh}} \right)\right) \leq \gamma
+}
+
+If we let $\gamma = \frac{e^{-h/m}}{N}$
+
+\eqn{
+P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{\log N}{h}+\frac{1}{m})+2\sqrt{\frac{\log N}{mh}+\frac{1}{m^2}} \right)\right) \leq \gamma \\
+\implies P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{2}{m})+2\sqrt{\frac{2}{m^2}} \right)\right) \leq \gamma \; \text{, if } h > m\log N \\
+\implies P\left(\hat{m}_i - m \geq 6\alpha m \right) \leq \gamma
+}
+
+This doesn't work. We need $\hat{m}_i - m < (\alpha-1)m$ in order to get this expression.
+
+Or we could let $\varphi = \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}$
+
+\eqn{
+P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp({-\frac{h\epsilon^2}{m+\frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}}})
+}
+\eqn{
+\implies \epsilon = & \frac{1}{\sqrt{2}}\sqrt{\frac{3 \log^2(2/\delta)+2m\log(2/\delta)}{h}+\frac{\sqrt{3}}{h^2}\sqrt{h^2\log^2(2/\delta)(3\log^2(2/\delta)+4m\log(2/\delta)+3mN\gamma)}} \\
+\leq & 
+}
 

### 2015-09-01 causal-bandit/observe_then_best.tex

@@
 }
+\usepackage{comment}
 \usepackage{amsthm}
@@
 \pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
-%\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
+\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
            draw=black, thick,
-           minimum width=1em,
-           minimum height=1em,
+           minimum width=2.5em,
+           minimum height=2.5em,
            font=\footnotesize,
@@
            },
+     empty/.style={
+           circle,
+           rounded corners,
+           minimum width=.5em,
+           minimum height=.5em,
+           font=\footnotesize,
+           text centered,
+           },
     % Define arrow style
@@
 
-\section*{Observe then pick best}
-\begin{itemize}
-\item Avi's point - of all the space of possible functions $f(X_1...X_N)$, most of them will have very similar marginals- so limited?
-opportunity for regret?
-Does this mean we effectively only care about a particular subset of the functions where the marginals are significantly different.
-We could assume we are in that subspace (if we are not then all actions are equally good anyway)
-
-\item How do assumptions/priors about the function $f(X_1...X_N)$ flow through to information about the marginals?%
-
-\item Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
-
-\item The goal is to bound the psuedo-regret upto a total number of timesteps $T$
-
-\item Our algorithm will explore by playing uniformly at random for $h$ timesteps and then select the arm with the highest estimated reward for the remaining timesteps $T-h$.
-
-\item We assume each arm corresponds to setting one binary input variable $X_j$ to a given value.
-All the input variables are assumed to be independent causes of the binary reward variable $Y$.
-With this structure the probability of a reward is the same under the observation that a variable takes a given configuration as under the action that assigns it.
-Therefore, for each exploration timestep, we get data on the performance of half of the arms.
-
-\item We assume $P(X_j = 1) = \frac{1}{2} \forall j$.
-With this assumption we will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage.
-Note that this is independent of the number of arms $K$.
-Relaxing this assumption will require us to have a more targeted exploration phase - as otherwise we do not gain any information about the value of arms that do not occur naturally with reasonable probability.
-\end{itemize}
-
-\subsection*{Regret during explore phase}
+\begin{comment}
+\section{Random notes}
+Information directed sampling have developed a general framework to construct algorithms that leverege generic structure between arms to minimize Bayesian regret, by explicitly trading off expected regret and information gain at each timestep.
+
+What can be expressed in the partial monitoring framework that cannot be defined by the feedback graph?
+
+We consider a variant of the stochastic multi-armed bandit problem when we have prior knowledge of the causal structure, but not the functional form, of the relationship between variables.
+
+sequential decision process,exploration-exploitation trade off
+
+Points to present. Why we use regret against single best arm - it makes sense in the stochastic setting. 
+The difference between observational and interventional distributions.
+
+terminology for intervention vs observation.
+Recast bandit algorithm in terms of this.
+Show limitations to existing approaches by pointing to places were existing algorithms eg UCB/explore-exploit/exp3 require unbiased estimates.
+Are unbiased estimates actually impossible in our setting?
+
+This paper is about sequentially choosing actions. 
+Modelling sequntial decition processes.
+
+In the classic multi-armed bandit setting, an ag
+
+General bandit part ...
+define multi-armed bandit, and importance.
+Decision making is fundamental.
+Goal is to select an action to maximize a reward (or minimize a loss).
+In many settings, such as drug trials, online ad placement, etc we can choose actions sequentially.
+This introduces an exploration-exploitation trade-off.
+
+The problem of sequentially choosing actions to minimize some loss wide range of applications in XXXX, YYYY. The multi-armed bandit setting, in which .... .
+
+Now the required causal stuff should go here.
+
+Causal structure can be represented as a directed acyclic graph with nodes representing the variables and an edge from $A \rightarrow B$ meaning $A$ causes $B$.
+Deterministic interventions, consisting of setting variables to some value, are modelled by setting the nodes to the desired values and removing any links coming into them, as their value is now no longer set by other variables in the graph according to the original mechanism.
+
+If the causal graph is known, such that there are no hidden variables causing 2 or more of the variables and the probabilities are all > 0 then the distribution of any variable after any intervention can be inferred (asymptotically) from data collected in absence of any interventions.
+\end{comment}
+
+\section*{Introduction}
+
+Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
+For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
+A widely used framework for sequential decision making is the multi-armed bandit.
+In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary and independent of the reward distribution of other actions.
+At each timestep the agent selects an action and receives a reward sampled iid from the corresponding reward distribution.
+
+An an alternate approach to selecting actions is causal inference.
+Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
+This allows an agent to predict the outcome of an action based on non-experimental data.
+This approach is common in social science, demography, and economics where explicit experimentation may be difficult.
+For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.
+
+We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
+This structure creates dependencies between the rewards of different arms such that selecting one action can provide information on the reward for other actions.
+
+There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
+Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
+There is also an active line of research into bandits with feedback defined by a graph.
+Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
+The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
+
+\section*{Problem Formulation}
+Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
+We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
+As an example, consider a farmer wishing to optimize the yield of her crop.
+She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
+We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
+After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
+This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
+
+\begin{figure}[h]
+\centering
+\caption{Assumed Causal Structure}
+\label{fig:causalStructure}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){Y};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5);
+
+\end{tikzpicture}
+\end{figure}
+
+Let $q \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known end point $T$:
+
+\begin{enumerate}
+\item The learner chooses an $I_t \in \set{1,\ldots, N}$ and $J_t \in \set{0,1}$.
+\item Then $X_t \in \set{0,1}^N$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(q_i)$ 
+\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
+\eq{
+\tilde X_{t,i} = \begin{cases}
+X_{t,i} &\text{if } i \neq I_t \\
+J_t & \text{otherwise}\,.
+\end{cases}
+}
+\item The learner receives reward $Y_t \sim \Ber(r(\tilde X))$ where $r:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
+\end{enumerate}
+
+The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[r(X)|do(X_i = j)]$. The optimal reward and action are denoted $\mu^*$ and $(i^*,j^*)$ respectively,
+where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
+\eq{
+R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
+}
+
+The problem can be treated as a classical multi-armed bandit with $K = 2N$ arms. However, this does not utilize the information provided by the causal assumption.
+
+Now need to expand upon how the assumption gives us extra information.
+
+The difficulties faced due to bias.
+
+To deal with this issue we use a simple explore-exploit algorithm.
+Our algorithm will explore for $h$ time-steps, sampling actions in a way that depends on our prior knowledge of $q$ but is independent of the observed rewards.
+We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
+
+\section*{Results}
+
+Summarize results here and note differences to classic bandit results
+
+\subsection*{Known and Balanced q}
+
+We begin with the simplest case were we assume that $q_i=\frac{1}{2} \forall i$.
+During the exploration phase we sample actions uniformly at random.
+In this case, this is equivalent to purely observing, that is taking no action and allowing all input variables to take their value randomly as  $X_{t,i} \sim \Ber(\frac{1}{2})$.
+
+We will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$.
+
+Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
+
+\subsubsection*{Regret during explore phase}
 Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
@@
 
-\subsection*{Regret during exploit phase}
+\subsubsection*{Regret during exploit phase}
 The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
@@
 \subsection{Generalizing to unknown q}
-\begin{itemize}
-\item Assume without loss of generality that $q_i \in [0,\frac{1}{2}]$
-
-\item Assume we observe for the first $h/2$ timesteps.
-
-\item Let $\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}$
-
-\item Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ We order the $\bar{q}$'s from smallest to largest and estimate $m$ with $\hat{m} = min_j : \bar{q}_j \geq \frac{1}{j}$
-
-\item The worst case $\boldsymbol{q}$ for a given fixed $m$ is 
+We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.
+
+\eqn{
+\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}
+}
+
+Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ and construct $\bar{\boldsymbol{q}}$ such that $\bar{q}_1 \leq \bar{q}_2 \leq ...\leq \bar{q}_N$.
+We then estimate $m$ with $\hat{m} = min_i : \bar{q}_i \geq \frac{1}{i}$
+
+We estimate the reward for the apparently common arms, $i \geq \hat{m}$, as:
+
+\eqn{
+\hat{\mu}_i = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\frac{h}{2}\hat{q}_i} = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\sum_{t=1}^{h/2}\ind{X_i=1}}
+}
+
+We play the remaining arms $\frac{h}{2\hat{m}}$ times and estimate their reward as:
+
+\eqn{
+\hat{\mu}_i = \frac{2\hat{m}}{h}\sum_{t=1}^{h/2\hat{m}}\ind{Y = 1|X_i=1}
+}
+
+We now consider how errors in the estimation of $m$ effect our estimates of the arm rewards.
+For the arms with $i \geq \hat{m}$ we know $\hat{q}_i \geq \frac{1}{\hat{m}}$ and thus our estimate is based on at least $\frac{h}{2\hat{m}}$ observations.
+Similarly we explicitly play the infrequently observed arms, $i < m$,  $\frac{h}{2\hat{m}}$ times.
+Thus our estimates will only be worse than the known $\boldsymbol{q}$ case if $\hat{m} > m$.
+
+For a fixed $m$ the $\boldsymbol{q}$ most likely to lead us to overestimate $m$ is:
+
 \eqn{
@@
 
-In this case $\bar{q}_0 ... \bar{q}_{m-1} = 0$
-
-For $i \geq m$
-\eq{
-P(q_i - \bar{q}_i \geq C) = P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})
-}
-
-The first term will dominate for all but very small $m$. Lets just work with that one for now.
-
-Via Bernstein's Inequality
-\eqn{
-P(q_i - \hat{q}_i \geq C) \leq \exp(-\frac{hC^2}{4q_i+4C/3}) := \gamma \\
-\implies P(q_i - \hat{q}_i \geq \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4q_i\log(1/\gamma)}{h}})\leq \gamma
-}
-
-Actually I think I can use the Lower tail version  of Bernsteins
-
-\eqn{
-P(q_i - \hat{q}_i \geq C) \leq \exp(-\frac{hC^2}{4q_i}) := \gamma \\
-\implies P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(1/\gamma)}{mh}})\leq \gamma
+We now bound $P(\hat{m} - m > \varphi)$ for this worst case.
+
+For $i < m$,we have $\bar{q}_i = q_i = 0$. For $i \geq m$
+\eq{
+P(q_i - \bar{q}_i \geq C) & =  P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})\\
+& \leq  2P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})
+}
+
+Via Bernstein's Inequality:
+
+\eqn{
+&2P(q_i - \hat{q}_i \geq C | \hat{q}_i \leq \frac{1}{2}) \leq 2\exp(-\frac{hC^2}{4q_i}) := \gamma \\
+\implies &P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(2/\gamma)}{mh}})\leq \gamma
 }
@@
 
-\eq{
-\hat{q}_i \geq q_i - C \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
-\implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}
-}
-
-If we let $\gamma =  \frac{1}{N}\exp(-{\frac{h}{16m}})$
-
-\eqn {
-C = 2\sqrt{\frac{m\log N}{hm^2}+\frac{1}{16m^2}} \leq \frac{1}{2m} \text{, if } h > 16m\log N \\
-}
-
-Then
-
-\eqn{
- q_i - \hat{q}_i \leq \frac{1}{2m} \implies \hat{m}_i - m \leq m \\
- \implies P(\hat{m}_i - m \geq m) \leq \frac{1}{N}\exp(-{\frac{h}{16m}})
+\eqn{
+\hat{q}_i \geq q_i - C & \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
+& \implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}\\
+& \implies P(\hat{m}_i - m \geq \frac{m^2C}{1-mC}) \leq \gamma
+}
+
+Let $\varphi = \frac{m^2C}{1-mC} \implies C = \frac{\varphi}{m(\varphi+m)} \implies \gamma = 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2}) \implies \varphi = \frac{2m\sqrt{m\log(2/\gamma)}}{\sqrt{h}-2\sqrt{m \log(2/\gamma)}} $
+
+Note this implies that if $h \geq 16m\log(2/\gamma)$ then $\varphi \leq m$
+
+\eqn {
+P(\hat{m}_i - m \geq \varphi) \leq 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2})
 }
@@
 P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq & P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \leq \varphi )P(\hat{m}-m \leq \varphi) \\
-& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \beta)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
+& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \varphi)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
 }
@@
 \eqn {
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}}) := \delta
-}
+P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}})\\
+& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \\
+& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}))^2}{2N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}) + 2\varphi/3}})
+}
+
+Assume $h > 16m\log(2N)$ and . Let $\varphi = \epsilon\sqrt{h}$. Assume $\epsilon \geq \frac{m}{\sqrt{h}}$ so as to ensure $\varphi > m$
+\eqn{
+P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h}{16m}))^2}{2N\exp(-\frac{h}{16m}) + 2\varphi/3}})\\
+& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-\varphi/2)^2}{\varphi + 2\varphi/3}})\\
+& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\varphi}{20}})\\
+& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\epsilon \sqrt{h}}{20}}) := \delta\\
+& \leq \begin{cases}
+2 \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) & \text{ if } h < \frac{9m^2}{289\epsilon^2} \\
+2\exp({-\frac{3\epsilon\sqrt{h}}{20})} & \text{ otherwise } 
+\end{cases}
+}
+
+\eqn{
+\implies \epsilon \leq \begin{cases}
+\frac{1}{\sqrt{h}}\log(2/\delta) + \sqrt{\frac{m}{h}\log(2/\delta)} & \text{ if } h < \frac{9m^2}{289\epsilon^2}\\
+\frac{20}{3\sqrt{h}}\log(2/\delta) & \text{ otherwise }
+\end{cases}\\
+}
+
+\eqn{
+\implies \epsilon \leq \frac{20}{3\sqrt{h}}\log(2/\delta)+ \sqrt{\frac{m}{h}\log(2/\delta)} 
+}
+
+Does this fit with the assumption we made about $\epsilon$???
+
+\pagebreak
+
 The first term will dominate if
@@
 }
-
-\end{itemize}
 

### 2015-09-25 unknown_q.py
adds lots of figures for TPR

@@ -1,0 +1,47 @@
+from numpy import *
+from scipy.stats import binom
+import matplotlib.pyplot as plt
+from scipy import stats
+
+def calc_m(array): # assumes array already sorted smallest to largest
+    for i,v in enumerate(array):
+        if v >= 1/(i+1):
+            return i+1
+    print("returning len")
+    return len(array)
+
+N = 100 # Number of variables
+m = 10 # Number of unbalanced variables
+q = zeros(N)
+q[m-1:] = 1/m
+
+# simulate observering for h rounds and estimating bar(q_i)
+delta = .05
+h = floor(N*m*log(1/delta))
+print(h)
+simulations = 1000
+
+est_m = zeros(simulations)
+for i in range(simulations):
+    counts = binom.rvs(h,q) #For each q_i the the count will be ~ binomial(h,q_i)
+    est_q = counts/h
+    est_bar_q = sort(minimum(est_q,1-est_q)) #the algorithm doesn't know q_i < .5
+    est_m[i] = calc_m(est_bar_q)
+
+freq = stats.itemfreq(est_m)
+print(freq)
+fig,ax = plt.subplots(1,1)
+plt.bar(freq[:,0],freq[:,1],align='center')
+plt.show()
+
+plt.subplots(1,1)
+p = linspace(0,1,50)
+v = p*(1-p)
+plt.plot(p,v)
+plt.show()
+
+
+
+
+
+

### 2015-09-29 causal-bandit/observe_then_best.tex
adds lots of figures for TPR
updated causal bandits

@@
 \documentclass{article}
+
+\usepackage{todonotes}
+
+%\usepackage[disable]{todonotes}
+\newcommand{\todot}[2][]{\todo[color=red!20!white,#1]{#2}}
+\newcommand{\todof}[2][]{\todo[color=orange!20!white,#1]{#2}}
 
@@
 \usepackage{amssymb}
+\usepackage[boxed]{algorithm}
+\usepackage{algpseudocode}
 \usepackage{mathtools}
@@
 
+\section{Bounding the Simple Regret}
+
+\begin{itemize}
+\item $N$ is number of variables.
+\item $q_i$ is probability that $X_{i,t} = 1$ for any $t$
+\item $s_i = \min\set{q_i, 1 - q_i}$
+\item $h = T/2$
+\item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
+\item $I_t \in \set{1,\ldots,N} \times \set{0,1}$ is the intervention
+\end{itemize}
+
+Let $r_T$ be the simple regret after $T$ time steps, which is
+\eq{
+r_q(T) = \E\left[\mu^* - \mu_{I_T}\right]\,,
+}
+where $\mu^*$ is the mean payoff of the optimal intervention and $I_T$ is the intervention chosen in the $T$th round.
+We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.
+
+\begin{algorithm}[H]
+\caption{Simple Regret Algorithm}\label{alg:simple}
+\begin{algorithmic}[1]
+\State {\bf Input:} $T, N$
+\State $h = T/2$
+\For{$t \in 1,\ldots,h$}
+\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
+\EndFor
+\State Compute $\hat q_i = \frac{1}{h} \sum_{t=1}^n X_{i,t}$
+\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
+\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$
+\State Compute $A$ is subset of insufficiently explored arms with $|A| = \hat m$
+\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
+\eq{
+\hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_i = j} r_t}{\sum_{t=1}^h \ind{X_i = j}}\,.
+}
+\For{$i \in A$}
+\State $j = \ind{\hat q_{I_t} \geq 1/2}$
+\For{$t \in 1,\ldots,h/\hat m$}
+\State Choose $I = i$ and $J = j$
+\State Observe reward $r_t$
+\EndFor
+\State Recompute $\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t$ 
+\EndFor
+\end{algorithmic}
+\end{algorithm}
+
+\begin{theorem}\label{thm:simple-regret}
+Define $m = \min\set{i : q_i \geq 1/i}$.
+Then the algorithm given in \cref{alg:simple} satisfies
+\eq{
+r_q(T) \in O\left(\sqrt{\frac{m}{T} \log \left(\frac{NT}{m}\right)}\right)\,.
+}
+\end{theorem}
+
+\begin{lemma}\label{lem:conc1}
+$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{3q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
+\end{lemma}
+
+\begin{proof}
+Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
+Then
+\eq{
+\hat q_i = \frac{1}{h} \sum_{t=1}^{h} Z_t\,.
+}
+Now $Z_1,\ldots,Z_h$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
+\end{proof}
+
+\begin{lemma}\label{lem:conc2}
+Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
+define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
+If
+\eq{
+h \geq 12m \log\frac{2N}{\delta}\,,
+}
+then
+\eq{
+\P{\hat m \leq 2m} \geq 1 - \delta\,.
+}
+\end{lemma}
+
+\begin{proof}
+It is easy to see that the worst case occurs when 
+\eq{
+q_i = \begin{cases}
+0 & \text{if } i < m \\
+\frac{1}{m} & \text{otherwise}\,.
+\end{cases}
+}
+Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that 
+\eq{
+(\forall i) \qquad \left| \hat q_i - q_i\right| 
+&\leq \sqrt{\frac{3q_i}{h} \log \frac{2N}{\delta}} \\
+&= \sqrt{\frac{3}{mh} \log\frac{2N}{\delta}}\,. 
+}
+By assumption 
+\eq{
+h \geq 12m \log \frac{2N}{\delta}\,.
+}
+Therefore with probability at least $1 - \delta$ we have
+\eq{
+(\forall i) \qquad \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.
+}
+Note that for $i$ with $q_i = 0$ it is easy to see that $\hat q_i = 0$ (the variance is zero). The above guarantee that
+for $i$ with $q_i = 1/m$ we have $\hat q_i \in [1/(2m), 3/(2m)]$ for all $i$ with probability at least $1 - \delta$,
+which implies that $\hat s_i \geq 1/(2m)$ for all $i$ where $q_i = 1/m$.
+Therefore $\hat m \leq 2m$ with probability at least $1 - \delta$. 
+\end{proof}
+
+\begin{proof}[Proof of \cref{thm:simple-regret}]
+Given the assumption that
+\eq{
+h \geq 12m \log \frac{N}{\delta}
+}
+we have shown with probability at least $1 - \delta$ that
+\eq{
+(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{m}{h} \log\frac{N}{\delta}}\,.
+}
+Suppose $h < 12m \log \frac{N}{\delta}$. Then
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{12m}{h} \log \frac{N}{\delta}}\,.
+}
+Therefore
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{12m}{h} \log \frac{N}{\delta}}\,.
+}
+Therefore 
+\eq{
+r_q(T) 
+&\leq \delta + \sqrt{\frac{12m}{h} \log \frac{N}{\delta}} \\
+&= \delta + \sqrt{\frac{24m}{T} \log \frac{N}{\delta}} \\
+&\leq \frac{m}{T} + \sqrt{\frac{24m}{T} \log \left(\frac{NT}{m}\right)}
+}
+as required.
+\end{proof}
+
 \section*{Results}
@@
 
-For a fixed $m$ the $\boldsymbol{q}$ most likely to lead us to overestimate $m$ is:
+For a fixed $m$ the $\boldsymbol{q}$ most likely to lead us to overestimate $m$ is: \todot{Justify this claim.}
 
@@
 
-For $i < m$,we have $\bar{q}_i = q_i = 0$. For $i \geq m$
+For $i < m$,we have $\bar{q}_i = q_i = 0$. For $i \geq m$ \todot{Is this really true, justify}
 \eq{
@@
 Via Bernstein's Inequality:
-
 \eqn{

### 2015-09-29 randomized_trial.tex

--- 2014-11-21 wiml_poster/FCI_figure.tex
@@
 \usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{tikz}
+\usepackage{graphicx}
 
-\usepackage{tikz}
+\usepackage{subcaption}
+\usepackage{multicol,caption}
+\usepackage{geometry}
+ \geometry{
+ a4paper,
+ total={210mm,297mm},
+ left=20mm,
+ right=20mm,
+ top=20mm,
+ bottom=20mm,
+ }
 
-\usepackage{graphicx}
-\usetikzlibrary{arrows,positioning} 
-\usepackage{subcaption}
+\usetikzlibrary{arrows,positioning,shapes} 
+\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
 \thispagestyle{empty}
-\usepackage{multicol,caption}
-\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
-
 \DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
@@
     observed/.style={
-           circle,
+           rectangle,
            rounded corners,
            draw=black, thick,
-           minimum width=.5em,
-           minimum height=.5em,
+           minimum width=3em,
+           minimum height=1.5em,
            font=\footnotesize,
@@
 
-\begin{figure}
+%\multicolumn{'num_cols'}{'alignment'}{'contents'}
+
+\begin{figure}[h]
+\begin{subfigure}[t]{0.45\textwidth}
 \centering
-\begin{subfigure}[t]{0.44\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
-
+\caption{Unrandomized}
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.7cm,  thick,main node/.style={observed}, shaded/.style={latent}]
+\node[cloud,draw,aspect=3](C){Confounders};
+\node[main node, below left=of C](T){Treatment};
+\node[main node, below right=of C](O){Outcome};
 \path[]
- 	(1) edge node {} (4)
-    (2) edge [->]  (1)
-    (3) edge [->]  (1);
-
-\node[main node,right of=3,xshift=1.2em](5){A};
-\node[main node, below right=of 5](6){C};
-\node[main node, above right=of 6](7){B};
-\node[main node, below=of 6](8){D};
-\path[every node/.style={font=\sffamily\footnotesize}]
-	(1) edge [-, >=stealth', blue!30!white, line width=.9em,shorten <=1em,shorten >= 1.7em,
-	preaction={line width = .3em, blue!30!white,-triangle 90,draw,shorten >=.6em,shorten <=1em}
-	] node[black]{FCI} (6)
-	(5) edge [o->] (6)
-	(7) edge [o->] (6)
-	(6) edge [->] (8);
-
+	(C) edge (T) edge (O)
+	(T) edge (O);
 \end{tikzpicture}
 \end{subfigure}
-\begin{subfigure}[t]{0.55\textwidth}
+\begin{subfigure}[t]{0.45\textwidth}
 \centering
-\caption{}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]
-
-\node[main node](1){A};
-\node[main node, below right=of 1](2){B};
-\node[shaded, above right=of 2](3){U};
-\node[main node, below right=of 3](4){C};
-\node[main node, above right=of 4](5){D};
-\node[ below=of 3,yshift=.2cm](11){};
-\path[] (1) edge (2)
-		(3) edge (2) edge (4)
-		(5) edge (4);
-
-\node[main node,right=of 5,xshift=-.5em](6){A};
-\node[main node, below right=of 6](7){B};
-\node[above right=of 7](8){};
-\node[main node, below right=of 8](9){C};
-\node[main node, above right=of 9](10){D};
-\node[below = of 8,yshift=0.2cm](12){};	
-\path[every node/.style={font=\sffamily\footnotesize}]
-(11) edge [-, >=stealth', blue!30!white, line width=.9em,shorten <=3.3em,shorten >= 4.5em,
-	preaction={line width = .3em, blue!30!white,-triangle 90,draw,shorten >=3.7em,shorten <=4em}
-	] node[black,xshift=-.5em]{FCI} (12)
-(6) edge [o->] (7)
-(7) edge [<->] (9)
-(10) edge [o->](9);
-
+\caption{Randomized}
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.7cm,  thick,main node/.style={observed}, shaded/.style={latent}]
+\node[cloud,draw,aspect=3](C){Confounders};
+\node[main node, below left=of C](T){Treatment};
+\node[main node, below right=of C](O){Outcome};
+\node[above left=of T](R){\includegraphics[width=1cm]{dice.jpg}};
+\path[]
+	(C) edge (O)
+	(T) edge (O)
+	(R) edge (T);
 \end{tikzpicture}

### 2015-09-29 smoking.tex

\documentclass{article}
\usepackage{tikz} 
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{tikz}
\usepackage{graphicx}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}
\usepackage{multicol,caption}
\usepackage{geometry}
 \geometry{
 a4paper,
 total={210mm,297mm},
 left=20mm,
 right=20mm,
 top=20mm,
 bottom=20mm,
 }

\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
\thispagestyle{empty}
\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=3em,
           minimum height=1.5em,
           font=\footnotesize,
           text centered,
           fill=blue!20!white},
     latent/.style={	(Switch) edge (X)
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }

}

\begin{document}
\def\ci{\perp\!\!\!\perp} % from Wikipedia

\begin{enumerate}
\item D-separation still applies after intervention.

${(Cancer \ci Asthma|Smoke)_{G_{\overline{X}}} \implies P(Cancer|do(Smoke),Asthma) = P(Cancer|do(Smoke)}$
\item If there are no backdoor paths from $X$ to $Y$ then intervention$\equiv$observation.

${(\hat{X} \ci Cancer|X,Poverty)_{G^{\dagger}} \implies P(Cancer|do(Smoke),Poverty) = P(Cancer|Smoke,Poverty)}$
\item If there are only backdoor paths from $X$ to $Y$ then intervention doesn't change $P(Y)$.

$(\hat{X} \ci Diet)_{G^{\dagger}} \implies P(Diet|do(Smoke)) = P(Diet)$
\end{enumerate}
\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\caption{$G^{\dagger}$}
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,thick,main node/.style={observed}]
\node[main node](X){X=Smoke};
\node[main node, above right=of X](P){Poverty};
\node[main node, below right=of P](D){Diet};
\node[main node, below right=of X](C){Cancer};
\node[main node, fill=red!20!yellow, above left=of X](Switch){$\hat{X}$};
\node[main node, below left=of X](A){Asthma};
\path[]
	(Switch) edge (X)
	(X) edge (C)
	(D) edge (C)
	(X) edge (A)
	(P) edge (D) edge (X);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\centering
\caption{$G_{\overline{X}}$}
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]
\node[main node](X){X=Smoke};
\node[main node, above right=of X](P){Poverty};
\node[main node, below right=of P](D){Diet};
\node[main node, below right=of X](C){Cancer};
\node[main node, below left=of X](A){Asthma};
\path[]
	(X) edge (C) edge (A)
	(D) edge (C)
	(P) edge (D);
\end{tikzpicture}
\end{subfigure}
\end{figure}

\end{document}

### 2015-09-30 causal-bandit/observe_then_best.tex
updates algorithm description for simple regret

@@
 \EndFor
-\State Compute $\hat q_i = \frac{1}{h} \sum_{t=1}^n X_{i,t}$
+\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
+\eq{
+\hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_i = j} r_t}{\sum_{t=1}^h \ind{X_i = j}}\,.
+}
+\State Compute $\hat q_i = \frac{1}{h} \sum_{t=1}^h X_{i,t}$
 \State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$
-\State Compute $A$ is subset of insufficiently explored arms with $|A| = \hat m$
-\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
-\eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_i = j} r_t}{\sum_{t=1}^h \ind{X_i = j}}\,.
-}
-\For{$i \in A$}
-\State $j = \ind{\hat q_{I_t} \geq 1/2}$
+\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
+\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
+\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
+\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
+
+\For{$(i,j) \in A$}
 \For{$t \in 1,\ldots,h/\hat m$}
@@
 \begin{theorem}\label{thm:simple-regret}
-Define $m = \min\set{i : q_i \geq 1/i}$.
+Define $m = \min\set{2 \leq i \leq N : q_i \geq 1/i}$.
 Then the algorithm given in \cref{alg:simple} satisfies
@@
 }
+
+\eq {
+\text{ for } (i,j)\in A \qquad \hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t
+}
+Prove bound for this case
+
+\eq {
+\text{ for } (i,j)\notin A \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_i = j} r_t}{\sum_{t=1}^h \ind{X_i = j}}
+}
+
+Prove bound for this case
+
 we have shown with probability at least $1 - \delta$ that
@@
 
-Let $D = \sqrt{\frac{8}{h}\log(hk)}$
-
-\eqn{
-R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hk)} + \frac{1}{h}\right)
+Let $D = \sqrt{\frac{8}{h}\log(hK)}$
+
+\eqn{
+R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)
 }
@@
 \eqn{
-R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hk)} + \frac{1}{h}\right)\\
-& \leq h + T\left(\sqrt{\frac{8}{h}\log(Tk)} + \frac{1}{h}\right)
+R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)\\
+& \leq h + T\left(\sqrt{\frac{8}{h}\log(TK)} + \frac{1}{h}\right)
 }

### 2015-10-01 causal-bandit/observe_then_best.tex
completes proof of simple regret bound for unknown q

@@
 
-Let $r_T$ be the simple regret after $T$ time steps, which is
-\eq{
-r_q(T) = \E\left[\mu^* - \mu_{I_T}\right]\,,
+Let $R_T$ be the simple regret after $T$ time steps, which is
+\eq{
+R_q(T) = \E\left[\mu^* - \mu_{I_T}\right]\,,
 }
@@
 \eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_i = j} r_t}{\sum_{t=1}^h \ind{X_i = j}}\,.
+\hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_{i,t} = j} r_t}{\sum_{t=1}^h \ind{X_{i,t} = j}}\,.
 }
@@
 \For{$t \in 1,\ldots,h/\hat m$}
-\State Choose $I = i$ and $J = j$
-\State Observe reward $r_t$
+\State Choose action $X_{i,t} = j$
+\State Observe reward $r_t(X_{i,t}=j)$
 \EndFor
-\State Recompute $\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t$ 
+\State Recompute $\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t(X_{i,t}=j)$ 
 \EndFor
@@
 \eq{
-h \geq 12m \log\frac{2N}{\delta}\,,
+h \geq 12m \log\frac{4N}{\delta}\,,
 }
@@
 \eq{
-h \geq 12m \log \frac{2N}{\delta}\,.
+h \geq 12m \log \frac{4N}{\delta}\,.
 }
@@
 \begin{proof}[Proof of \cref{thm:simple-regret}]
-Given the assumption that
-\eq{
-h \geq 12m \log \frac{N}{\delta}
-}
-
+
+for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{\hat m}$ times.
 \eq {
-\text{ for } (i,j)\in A \qquad \hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t
-}
-Prove bound for this case
+\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t(X_i = j)
+}
+
+Via Hoeffding's Inequality
 
 \eq {
-\text{ for } (i,j)\notin A \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_i = j} r_t}{\sum_{t=1}^h \ind{X_i = j}}
-}
-
-Prove bound for this case
-
-we have shown with probability at least $1 - \delta$ that
-\eq{
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{m}{h} \log\frac{N}{\delta}}\,.
-}
-Suppose $h < 12m \log \frac{N}{\delta}$. Then
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{12m}{h} \log \frac{N}{\delta}}\,.
+\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{2h\epsilon^2}{\hat{m}}}
+}
+
+for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{\hat m}$ times.
+
+\eq {
+(i,j)\notin A & \implies \hat{s}_i \geq \frac{1}{\hat m} \\
+& \implies \sum_{t=1}^h \ind{X_i = j} \geq \frac{h}{\hat{m}}
+}
+
+Let  $Z_{ij} = \sum_{t=1}^h \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
+
+\eq {
+\hat \mu_{i,j} = \frac{1}{Z_{ij}}\sum_{t'=1}^{Z_{ij}}r_{t'}
+}
+
+\eq {
+\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
+= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
+\leq & \P{\left| \frac{\hat m}{h} \sum_{t'=1}^{h/\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
+\leq & 2\exp{-\frac{2h\epsilon^2}{\hat{m}}}\\
+}
+
+Applying the union bound over all $2N$ actions.
+\eq{
+& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{2h\epsilon^2}{\hat{m}}} \\
+\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{2h}\log{\frac{4N}{\delta}}}} \leq \delta \\
+}
+
+Now by \cref{lem:conc2}, 
+\eq{
+h \geq 12m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
+}
+
+Therefore if $h \geq 12m \log \frac{4N}{\delta}$, we have shown with probability at least $1 - 2\delta$ that
+
+\eq{
+(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{m}{h} \log\frac{4N}{\delta}}\,.
+}
+Suppose $h < 12m \log \frac{4N}{\delta}$. Then
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}}\,.
 }
@@
 \eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{12m}{h} \log \frac{N}{\delta}}\,.
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
 }
@@
 r_q(T) 
-&\leq \delta + \sqrt{\frac{12m}{h} \log \frac{N}{\delta}} \\
-&= \delta + \sqrt{\frac{24m}{T} \log \frac{N}{\delta}} \\
-&\leq \frac{m}{T} + \sqrt{\frac{24m}{T} \log \left(\frac{NT}{m}\right)}
+&\leq 2\delta + \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}} \\
+&= 2\delta + \sqrt{\frac{24m}{T} \log \frac{4N}{\delta}} \\
+&\leq \frac{8m}{T} + \sqrt{\frac{24m}{T} \log \left(\frac{NT}{m}\right)}
 }

### 2015-10-02 causal-bandit/observe_then_best.tex
updates known but unbalanced q case to get O(T^2/3 m^1/3) regret

@@
 
-\section{Bounding the Simple Regret}
+\section*{Results}
+
+Summarize results here and note differences to classic bandit results
+
+\subsection*{Known and Balanced q}
+
+We begin with the simplest case were we assume that $q_i=\frac{1}{2} \forall i$.
+During the exploration phase we sample actions uniformly at random.
+In this case, this is equivalent to purely observing, that is taking no action and allowing all input variables to take their value randomly as  $X_{t,i} \sim \Ber(\frac{1}{2})$.
+
+We will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$.
+
+Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
+
+\subsubsection*{Regret during explore phase}
+Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
+\eqn{
+R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta] < h
+}
+
+\subsubsection*{Regret during exploit phase}
+The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
+
+\eqn{
+\hat{i^*} = argmax_i [\hat{\mu}_i]
+}
+\eqn{
+R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
+}
+
+The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$.
+Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
+\eqn{
+R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
+}
+The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.
+
+Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$.
+If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$
+
+\eqn{
+\label{eqn:probDeltaToLarge}
+P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
+}
+
+If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution.
+Instead we will use an importance weighted estimator.
+
+\eqn{
+\label{eqn:importance_weighted_estimator}
+\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
+}
+
+where $q_i = P(\text{arm $i$ active})$
+
+Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:
+
+\eqn{
+\label{eqn:balancedHoeffdings}
+P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
+}
+
+\eqn{
+R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
+}
+
+Let $D = \sqrt{\frac{8}{h}\log(hK)}$
+
+\eqn{
+R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)
+}
+
+\subsection*{Total Regret}
+
+Putting together the regret from the exploration and exploitation phases,
+
+\eqn{
+R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)\\
+& \leq h + T\left(\sqrt{\frac{8}{h}\log(TK)} + \frac{1}{h}\right)
+}
+
+Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,
+
+\eqn {
+R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
+}
+
+If $T \geq 2$ and $K \geq 2$, the first term dominates and,
+
+\eqn {
+R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
+}
+
+The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$
+
+\subsection*{Empirical results}
+\begin{figure}[H]
+\centering
+\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
+\includegraphics[width=.5\textwidth]{explore_exploit}
+\end{figure}
+
+\section{Generalizing to unbalanced $\boldsymbol{q}$}
+
+When some arms have low natural probability we cannot rely on exploring them adequately by pure observation. We need to explicitly play them during the exploration phase.
+
+We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
+
+Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
+\leq q_N$.
+Let $m \in [2,N] = \{m:q_m > \frac{1}{m}\}$ Ie if the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
+If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$.
+Let $q_{ij} = \P{X_i = j}$
+
+Suppose we observe for the first $h/2$ timesteps. This is at worst half the optimal.
+
+We then have estimates
+
+\eqn {
+\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
+}
+
+We take this as our estimate for those arms for which $q_{ij} > \frac{1}{m}$
+
+Let $Z_{t,ij} = \ind{Y_t=1,X_{t,i}=j} \sim Bernoulli(q_{ij}\mu_{ij})$,
+
+Chernoff's inequality gives
+
+\eqn{
+P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
+}
+
+We play each of the remaining $m$ arms $\frac{h}{2m}$ times so for them we get
+
+\eqn{
+P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m}
+}
+
+So for all the arms
+
+\eqn{
+P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
+}
+
+and
+
+\eqn{
+\label{eqn:unbalancedHoeffdings}
+P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/24m}
+}
+
+If we let $D = \sqrt{\frac{24m\log(hK)}{h}}$
+
+\eqn{
+R_T &  \leq h + T\left(\sqrt{\frac{24m\log(hK)}{h}} + \frac{1}{h}\right) \\
+& \leq h + T\left(\sqrt{\frac{24m\log(TK)}{h}} + \frac{1}{h}\right) 
+}
+
+Let $h = T^{2/3}m^{1/3}\log(TK)^{1/3}$
+
+\eqn {
+R_T & \leq 6T^{2/3}m^{1/3}log(KT)^{1/3} + T^{1/3}m^{-1/3}log(KT)^{-1/3}\\ 
+& \leq 7T^{2/3}m^{1/3}log(KT)^{1/3}
+}
+
+It should also be possible to generalize to handle the case were the $q$'s are unknown, since we should be able to get a reasonable estimate for them while we are observing.
+They key will be how the resulting uncertainty in $m$ effects the bounds.
+
+To get results that degrade to similar order bounds as UCB when the arms are very unbalanced, I will need to drop the explore/exploit strategy.
+
+\subsection{Generalizing to unknown q}
+
+\subsubsection{Bounding the Simple Regret}
 
@@
 
-\section*{Results}
-
-Summarize results here and note differences to classic bandit results
-
-\subsection*{Known and Balanced q}
-
-We begin with the simplest case were we assume that $q_i=\frac{1}{2} \forall i$.
-During the exploration phase we sample actions uniformly at random.
-In this case, this is equivalent to purely observing, that is taking no action and allowing all input variables to take their value randomly as  $X_{t,i} \sim \Ber(\frac{1}{2})$.
-
-We will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$.
-
-Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
-
-\subsubsection*{Regret during explore phase}
-Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
-\eqn{
-R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta]
-}
-
-\subsubsection*{Regret during exploit phase}
-The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
-
-\eqn{
-\hat{i^*} = argmax_i [\hat{\mu}_i]
-}
-\eqn{
-R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
-}
-
-The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$.
-Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
-\eqn{
-R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
-}
-The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.
-
-Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$.
-If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$
-
-\eqn{
-\label{eqn:probDeltaToLarge}
-P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
-}
-
-If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution.
-Instead we will use an importance weighted estimator.
-
-\eqn{
-\label{eqn:importance_weighted_estimator}
-\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
-}
-
-where $q_i = P(\text{arm $i$ active})$
-
-Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:
-
-\eqn{
-\label{eqn:balancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
-}
-
-\eqn{
-R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
-}
-
-Let $D = \sqrt{\frac{8}{h}\log(hK)}$
-
-\eqn{
-R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)
-}
-
-\subsection*{Total Regret}
-
-Putting together the regret from the exploration and exploitation phases,
-
-\eqn{
-R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)\\
-& \leq h + T\left(\sqrt{\frac{8}{h}\log(TK)} + \frac{1}{h}\right)
-}
-
-Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,
-
-\eqn {
-R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
-}
-
-If $T \geq 2$ and $K \geq 2$, the first term dominates and,
-
-\eqn {
-R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
-}
-
-The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$
-
-\subsection*{Empirical results}
-\begin{figure}[H]
-\centering
-\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
-\includegraphics[width=.5\textwidth]{explore_exploit}
-\end{figure}
-
-\section{Generalizing to unbalanced $\boldsymbol{q}$}
-
-When some arms have low natural probability we cannot rely on exploring them adequately by pure observation. We need to explicitly play them during the exploration phase.
-
-We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
-
-Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
-\leq q_N$.
-Let $m \in [2,N] = \{m:q_m > \frac{1}{m}\}$ Ie if the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
-If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$
-
-Suppose we observe for the first $h/2$ timesteps. This is at worst half the optimal.
-
-We then have estimates
-
-\eqn {
-\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y=1,X_i=1}}{\frac{h}{2}q_{ij}}
-}
-
-We take this as our estimate for those arms for which $q_{ij} > \frac{1}{m}$
-
-For these arms the Hoeffdings gives
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/2m^2}
-}
-
-We play each of the remaining $m$ arms $\frac{h}{2m}$ times so for them we get
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m}
-}
-
-So for all the arms
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m^2}
-}
-
-and
-
-\eqn{
-\label{eqn:unbalancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/4m^2}
-}
-
-If we let $D = \sqrt{\frac{4m^2\log(hK)}{h}}$
-
-\eqn{
-R_T  \leq h + T\left(\sqrt{\frac{4m^2\log(hK)}{h}} + \frac{1}{h}\right)
-}
-
-Let $h = T^{2/3}m^{2/3}\log(hk)$
-
-\eqn {
-R_T  \leq 4T^{\frac{2}{3}}m^{2/3}(log(KT))^{\frac{1}{3}}
-}
-
-Note, I think the $m^{2/3}$ should be improvable to something close to $m^{1/3}$ is I use Berstein's instead of Hoeffdings to bound the estimator for the arms where $q > 1/m$ but I haven't got the equations to quite work out for that yet.
-
-It should also be possible to generalize to handle the case were the $q$'s are unknown, since we should be able to get a reasonable estimate for them while we are observing.
-They key will be how the resulting uncertainty in $m$ effects the bounds.
-
-To get results that degrade to similar order bounds as UCB when the arms are very unbalanced, I will need to drop the explore/exploit strategy.
-
-\subsection{Generalizing to unknown q}
 We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.

### 2015-10-07 causal-bandit/observe_then_best.tex
works on bandit case for unknown q and standardizes some notation accross cases

@@
 
-It should also be possible to generalize to handle the case were the $q$'s are unknown, since we should be able to get a reasonable estimate for them while we are observing.
-They key will be how the resulting uncertainty in $m$ effects the bounds.
+Again comparing to the UCB bound of $\sqrt{KT}$, we expect the causal algorithm to do better if $K > m^{2/3}T^{1/3}$, or alternatively if $m \leq \frac{K^{3/2}}{\sqrt{T}}$
 
@@
 \subsection{Generalizing to unknown q}
+We now consider the case where $\boldsymbol{q}$, and thus $m$ is not known in advance. We begin by considering the simple regret
 
@@
 \item $s_i = \min\set{q_i, 1 - q_i}$
-\item $h = T/2$
 \item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
@@
 
-Let $R_T$ be the simple regret after $T$ time steps, which is
-\eq{
-R_q(T) = \E\left[\mu^* - \mu_{I_T}\right]\,,
-}
-where $\mu^*$ is the mean payoff of the optimal intervention and $I_T$ is the intervention chosen in the $T$th round.
+Let $R_s(h)$ be the simple regret after $h$ time steps, which is
+\eq{
+R_s(h) = \E\left[\mu^* - \mu_{I_h}\right]\,,
+}
+where $\mu^*$ is the mean payoff of the optimal intervention and $I_h$ is the intervention chosen in the $h$th round.
 We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.
@@
 \begin{algorithmic}[1]
-\State {\bf Input:} $T, N$
-\State $h = T/2$
-\For{$t \in 1,\ldots,h$}
+\State {\bf Input:} $h, N$
+\For{$t \in 1,\ldots,h/2$}
 \State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
@@
 \eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^h \ind{X_{i,t} = j} r_t}{\sum_{t=1}^h \ind{X_{i,t} = j}}\,.
-}
-\State Compute $\hat q_i = \frac{1}{h} \sum_{t=1}^h X_{i,t}$
+\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
+}
+\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
 \State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
@@
 \For{$(i,j) \in A$}
-\For{$t \in 1,\ldots,h/\hat m$}
+\For{$t \in 1,\ldots,h/2\hat m$}
 \State Choose action $X_{i,t} = j$
@@
 \EndFor
-\State Recompute $\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t(X_{i,t}=j)$ 
+\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
 \EndFor
@@
 \eq{
-r_q(T) \in O\left(\sqrt{\frac{m}{T} \log \left(\frac{NT}{m}\right)}\right)\,.
+R_s(h) \in O\left(\sqrt{\frac{m}{h} \log \left(\frac{NT}{m}\right)}\right)\,.
 }
@@
 \begin{lemma}\label{lem:conc1}
-$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{3q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
+$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
 \end{lemma}
@@
 \eq{
-\hat q_i = \frac{1}{h} \sum_{t=1}^{h} Z_t\,.
-}
-Now $Z_1,\ldots,Z_h$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
+\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} Z_t\,.
+}
+Now $Z_1,\ldots,Z_{h/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
 \end{proof}
@@
 \eq{
-h \geq 12m \log\frac{4N}{\delta}\,,
+h \geq 24m \log\frac{4N}{\delta}\,,
 }
@@
 (\forall i) \qquad \left| \hat q_i - q_i\right| 
-&\leq \sqrt{\frac{3q_i}{h} \log \frac{2N}{\delta}} \\
-&= \sqrt{\frac{3}{mh} \log\frac{2N}{\delta}}\,. 
+&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
+&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
 }
@@
 \eq{
-h \geq 12m \log \frac{4N}{\delta}\,.
+h \geq 24m \log \frac{4N}{\delta}\,.
 }
@@
 
-for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{\hat m}$ times.
+for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
 \eq {
-\hat \mu_{i,j} = \frac{\hat m}{h} \sum_{t=1}^{\hat m/h} r_t(X_i = j)
+\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
 }
@@
 \eq {
-\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{2h\epsilon^2}{\hat{m}}}
-}
-
-for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{\hat m}$ times.
+\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
+}
+
+for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.
 
@@
 
-Let  $Z_{ij} = \sum_{t=1}^h \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
+Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
 
@@
 = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
-\leq & \P{\left| \frac{\hat m}{h} \sum_{t'=1}^{h/\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
-\leq & 2\exp{-\frac{2h\epsilon^2}{\hat{m}}}\\
+\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
+\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
 }
@@
 \eq{
-& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{2h\epsilon^2}{\hat{m}}} \\
-\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{2h}\log{\frac{4N}{\delta}}}} \leq \delta \\
+& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
+\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
 }
@@
 \eq{
-h \geq 12m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
-}
-
-Therefore if $h \geq 12m \log \frac{4N}{\delta}$, we have shown with probability at least $1 - 2\delta$ that
-
-\eq{
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{m}{h} \log\frac{4N}{\delta}}\,.
-}
-Suppose $h < 12m \log \frac{4N}{\delta}$. Then
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}}\,.
+h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
+}
+
+Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that
+
+\eqn{
+\label{eqn:joint_bound_estimates_m}
+(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
+}
+Suppose $h < 24m \log \frac{4N}{\delta}$. Then
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
 }
@@
 \eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
 }
@@
 \eq{
-r_q(T) 
-&\leq 2\delta + \sqrt{\frac{12m}{h} \log \frac{4N}{\delta}} \\
-&= 2\delta + \sqrt{\frac{24m}{T} \log \frac{4N}{\delta}} \\
-&\leq \frac{8m}{T} + \sqrt{\frac{24m}{T} \log \left(\frac{NT}{m}\right)}
+R_s(h) 
+&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
+&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
 }
@@
 
+\subsubsection{Bounding the Bandit Regret}
+
+\begin{itemize}
+\item Use $24Nlog(4N/\delta)$ samples to estimate $m$
+\item If $\hat m > \frac{N^{3/2}}{\sqrt{T}}$ then stop and do UCB with remaining rounds.
+\item Else use causal explore-exploit and let the exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
+\end{itemize}
+
+Since we don't know $\boldsymbol{q}$, and thus $m$, we can't set the exploration time $h$ in advance based on $m$ as we did for the known $\boldsymbol{q}$ case.
+Instead we first use $24Nlog(4N/\delta)$ samples to estimate $m$.
+If $\hat m < \frac{N^{3/2}}{\sqrt{T}}$ we will continue with the causal algorithm and let the total exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$.
+
+Since $m \leq N$, by \cref{eqn:joint_bound_estimates_m}, we have that with probability at least $1-2\delta$,
+
+\eqn{
+\label{eqn:bounds_hold}
+(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
+}
+
+So the regret in this case is bounded by,
+
+\eq {
+R(T) & =  2\delta E\left[R(T)|\text{ not } \cref{eqn:bounds_hold}\right] + (1-2\delta)E[R(T)| \cref{eqn:bounds_hold}]  \\
+& \leq 2\delta T + \left(h + T\sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}} \right)\\
+ &  \leq \frac{4 \hat m^{1/3}T}{h}  + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(\frac{2Nh}{\hat m^{1/3}}\right)}\right), \text { letting } \delta = \frac{2 \hat m^{1/3}}{h} \\
+&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(TK\right)}\right) \\
+&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + 2T^{2/3}(2m)^{1/3}log(TK)^{1/3}  \\
+&  \leq 8T^{2/3}m^{1/3}log(TK)^{1/3}  \\
+}
+
+Now make sure I have all the cases right - then have a go with an implementation.
+
+\subsubsection{ Older Attempts}
 We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.

### 2015-10-09 causal-bandit/observe_then_best.tex
completes sketch of retret bound proof for bandit case

@@
 
+\begin{lemma}\label{lem:m_est_small_enough}
+Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
+define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
+If
+\eq{
+h \geq 24m \log\frac{4N}{\delta}\,,
+}
+then
+\eq{
+\P{\hat m < \frac{2}{3}m} \leq \delta\,.
+}
+\end{lemma}
+
+\begin{proof}
+
+To over-estimate $m$, we must identify as balanced some unbalanced arms (ie those with $i < m$. For all arms with $i < m$, by definition, $q_i < \frac{1}{m}$
+
+Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that: \color{red}(this step not quite right - because I haven't said $q_i = 1/m$ as the worst case is that $q_i$ is infintesimally smaller than $1/m$.
+With smaller $q_i$ the bounds would be slightly tighter but not made up for by the decrease in the mean - ie we are more likely to over-estimate if $q_i = 1/m$ than if $q_i < 1/m$.
+Need a better way to say this ...
+).
+Should probably just combine this with the other side into one lemma (even though we don't need this side for simple regret.
+\color{black}
+
+\eq{
+(\forall i) \qquad \left| \hat q_i - q_i\right| 
+&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
+&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
+}
+
+By assumption 
+\eq{
+h \geq 24m \log \frac{4N}{\delta}\,.
+}
+Therefore with probability at least $1 - \delta$ we have
+\eq{
+(\forall i < m) \qquad & \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.\\
+ & \hat{q}_i < \frac{3}{2}m \\
+ \implies & \hat{m} > \frac{2}{3}m
+}
+
+\end{proof}
+
 \begin{proof}[Proof of \cref{thm:simple-regret}]
@@
 \item Use $24Nlog(4N/\delta)$ samples to estimate $m$
-\item If $\hat m > \frac{N^{3/2}}{\sqrt{T}}$ then stop and do UCB with remaining rounds.
+\item If $\hat m > \lambda = \frac{N^{3/2}}{\sqrt{T}}$ then stop and do UCB with remaining rounds.
 \item Else use causal explore-exploit and let the exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
@@
 
+There will be 3 cases.
+\begin{enumerate}
+\item $m < \lambda/2 \implies \hat{m} < \lambda$ with prob $1-\delta$
+\eq {
+R_T \sim O(T^{2/3}m^{1/3}) + O(\sqrt{NT})\delta \implies \text { want } \delta \leq \frac{T^{1/6}}{\sqrt{N}}
+}
+\item $m > 3\lambda/2 \implies \hat{m} > \lambda$ with prob $1-\delta$
+\eq {
+R_T \sim O(T^{2/3}N^{1/3})\delta + O(\sqrt{NT}) \implies \text { want } \delta \leq \frac{N^{1/6}}{T^{1/6}}
+}
+\item $\lambda/2 < m < 3\lambda/2$, algorithm could end up doing UCB or Explore-Exploit.
+\eq{
+R_T \sim  & O(T^{2/3}m^{1/3}) + O(\sqrt{NT}) \\
+= &  O(\sqrt{NT}) \text { as } m = O(\frac{N^{3/2}}{\sqrt{T}})
+}
+\end{enumerate}
+
+So if $\delta = \frac{1}{T^{1/6}\sqrt{N}}$ (or $\delta = \frac{1}{T^{1/3}}$) that would be enough concentration around $m$ to choose the correct algorithm ...
+Is it enough to choose a reasonable total exploration time $h$?
+It seems that $\delta = \frac{1}{T^{1/3}}$ works.
+
+\begin{algorithm}[H]
+\caption{Bandit Regret Algorithm}\label{alg:bandit}
+\begin{algorithmic}[1]
+\State {\bf Input:} $T, N$
+\State $T_1 = 24Nlog\left(4N/\delta\right)$ TODO figure out what $\delta$ should be 
+\For{$t \in 1,T_1$}
+\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
+\EndFor
+\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
+\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
+\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
+\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
+\If {$\hat{m} > \frac{N^{3/2}}{\sqrt{T}}$}
+\State Switch to the standard UCB algorithm with $K = 2N$ arms.
+\Else
+\State $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
+\For{$t \in T_1,h/2$}
+\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
+\EndFor
+\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
+\eq{
+\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
+}
+\State Recompute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
+\State Recompute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
+\State Recompute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
+\State Recompute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
+\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
+\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
+
+\For{$(i,j) \in A$}
+\For{$t \in 1,\ldots,h/2\hat m$}
+\State Choose action $X_{i,t} = j$
+\State Observe reward $r_t(X_{i,t}=j)$
+\EndFor
+\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
+\EndFor
+\EndIf
+\end{algorithmic}
+\end{algorithm}
+
 Since we don't know $\boldsymbol{q}$, and thus $m$, we can't set the exploration time $h$ in advance based on $m$ as we did for the known $\boldsymbol{q}$ case.
 Instead we first use $24Nlog(4N/\delta)$ samples to estimate $m$.
-If $\hat m < \frac{N^{3/2}}{\sqrt{T}}$ we will continue with the causal algorithm and let the total exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$.
+If $\hat m < \frac{N^{3/2}}{\sqrt{T}}$ we will continue with the causal algorithm and let the total exploration time $h = T^{2/3}\hat{m}^{1/3}log(2TK)^{1/3}$.
 
@@
 & \leq 2\delta T + \left(h + T\sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}} \right)\\
- &  \leq \frac{4 \hat m^{1/3}T}{h}  + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(\frac{2Nh}{\hat m^{1/3}}\right)}\right), \text { letting } \delta = \frac{2 \hat m^{1/3}}{h} \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(TK\right)}\right) \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + 2T^{2/3}(2m)^{1/3}log(TK)^{1/3}  \\
-&  \leq 8T^{2/3}m^{1/3}log(TK)^{1/3}  \\
-}
-
-Now make sure I have all the cases right - then have a go with an implementation.
+ &  \leq 2T^{2/3}  + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(4NT^{1/3}\right)}\right), \text { letting } \delta = \frac{1}{T^{1/3}} \\
+&  \leq 2T^{2/3} + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(2TK\right)}\right) \\
+&  \leq 2T^{2/3}  + 2T^{2/3}(2m)^{1/3}log(2TK)^{1/3}  \\
+&  \leq 6T^{2/3}m^{1/3}log(TK)^{1/3}  \\
+}
+
+Now lets try an implementation ...
 

### 2015-10-22 causal-bandit/observe_then_best.tex
prep for tpr talk

@@
 }
-\item The learner receives reward $Y_t \sim \Ber(r(\tilde X))$ where $r:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
+\item The learner receives reward $Y_t \sim \Ber(r(\tilde X))$ where $r:\set{0,1}^K \to [0,1]$ is unknown and arbitrary. THIS LOOKS WRONG SHOULD BE N not K.
 \end{enumerate}
@@
 
-We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
+We now have an additional trade off to make, which is how much should be observe (learning something about half the arms each timestep) versus playing the low probability arms.
 

### 2015-10-22 causal-bandit/problem_formulation.tex
prep for tpr talk

--- 2015-10-07 causal-bandit/observe_then_best.tex
@@
 \renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+\pagestyle{empty}
 
@@
 
-\begin{comment}
-\section{Random notes}
-Information directed sampling have developed a general framework to construct algorithms that leverege generic structure between arms to minimize Bayesian regret, by explicitly trading off expected regret and information gain at each timestep.
-
-What can be expressed in the partial monitoring framework that cannot be defined by the feedback graph?
-
-We consider a variant of the stochastic multi-armed bandit problem when we have prior knowledge of the causal structure, but not the functional form, of the relationship between variables.
-
-sequential decision process,exploration-exploitation trade off
-
-Points to present. Why we use regret against single best arm - it makes sense in the stochastic setting. 
-The difference between observational and interventional distributions.
-
-terminology for intervention vs observation.
-Recast bandit algorithm in terms of this.
-Show limitations to existing approaches by pointing to places were existing algorithms eg UCB/explore-exploit/exp3 require unbiased estimates.
-Are unbiased estimates actually impossible in our setting?
-
-This paper is about sequentially choosing actions. 
-Modelling sequntial decition processes.
-
-In the classic multi-armed bandit setting, an ag
-
-General bandit part ...
-define multi-armed bandit, and importance.
-Decision making is fundamental.
-Goal is to select an action to maximize a reward (or minimize a loss).
-In many settings, such as drug trials, online ad placement, etc we can choose actions sequentially.
-This introduces an exploration-exploitation trade-off.
-
-The problem of sequentially choosing actions to minimize some loss wide range of applications in XXXX, YYYY. The multi-armed bandit setting, in which .... .
-
-Now the required causal stuff should go here.
-
-Causal structure can be represented as a directed acyclic graph with nodes representing the variables and an edge from $A \rightarrow B$ meaning $A$ causes $B$.
-Deterministic interventions, consisting of setting variables to some value, are modelled by setting the nodes to the desired values and removing any links coming into them, as their value is now no longer set by other variables in the graph according to the original mechanism.
-
-If the causal graph is known, such that there are no hidden variables causing 2 or more of the variables and the probabilities are all > 0 then the distribution of any variable after any intervention can be inferred (asymptotically) from data collected in absence of any interventions.
-\end{comment}
-
-\section*{Introduction}
-
-Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
-For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
-A widely used framework for sequential decision making is the multi-armed bandit.
-In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary and independent of the reward distribution of other actions.
-At each timestep the agent selects an action and receives a reward sampled iid from the corresponding reward distribution.
-
-An an alternate approach to selecting actions is causal inference.
-Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
-This allows an agent to predict the outcome of an action based on non-experimental data.
-This approach is common in social science, demography, and economics where explicit experimentation may be difficult.
-For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.
-
-We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
-This structure creates dependencies between the rewards of different arms such that selecting one action can provide information on the reward for other actions.
-
-There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
-Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
-There is also an active line of research into bandits with feedback defined by a graph.
-Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
-The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
-
-\section*{Problem Formulation}
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
-As an example, consider a farmer wishing to optimize the yield of her crop.
-She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
-We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
-After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
-This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
-
-\begin{figure}[h]
-\centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\end{figure}
-
 Let $q \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known end point $T$:
@@
 \begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, N}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^N$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(q_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(r(\tilde X))$ where $r:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
+\item The learner chooses an $I_t \in \set{0,\ldots, N}$ and $J_t \in \set{0,1}$, setting $X_{I_t,t} = J_t$. Selecting $I_t = 0$ corresponds to taking the nothing action $do()$ and just observing. 
+\item For $i \neq I_t$, $X_{i,t} \sim Bernoulli(q_i)$
+
+\item The learner observes $ X_t = [X_{1,t} ... X_{N,t}] $
+\item The learner receives reward $Y_t \sim \Ber(r(X_t))$, where $r:\set{0,1}^N \to [0,1]$ is unknown and arbitrary.
 \end{enumerate}
 
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[r(X)|do(X_i = j)]$. The optimal reward and action are denoted $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
+\vspace{5mm}
+
+The causal structure gives us:
+\eq {
+P(Y|do(X_i = j)) = & P(Y|X_i = j) \\
+= & P(Y|do(X_a = 1),X_i = j)q_a + P(Y|do(X_a = 0),X_i=j)(1-q_a)
 }
 
-The problem can be treated as a classical multi-armed bandit with $K = 2N$ arms. However, this does not utilize the information provided by the causal assumption.
-
-Now need to expand upon how the assumption gives us extra information.
-
-The difficulties faced due to bias.
-
-To deal with this issue we use a simple explore-exploit algorithm.
-Our algorithm will explore for $h$ time-steps, sampling actions in a way that depends on our prior knowledge of $q$ but is independent of the observed rewards.
-We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
-
-\section*{Results}
-
-Summarize results here and note differences to classic bandit results
-
-\subsection*{Known and Balanced q}
-
-We begin with the simplest case were we assume that $q_i=\frac{1}{2} \forall i$.
-During the exploration phase we sample actions uniformly at random.
-In this case, this is equivalent to purely observing, that is taking no action and allowing all input variables to take their value randomly as  $X_{t,i} \sim \Ber(\frac{1}{2})$.
-
-We will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$.
-
-Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
-
-\subsubsection*{Regret during explore phase}
-Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
-\eqn{
-R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta] < h
-}
-
-\subsubsection*{Regret during exploit phase}
-The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
-
-\eqn{
-\hat{i^*} = argmax_i [\hat{\mu}_i]
-}
-\eqn{
-R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
-}
-
-The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$.
-Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
-\eqn{
-R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
-}
-The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.
-
-Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$.
-If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$
-
-\eqn{
-\label{eqn:probDeltaToLarge}
-P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
-}
-
-If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution.
-Instead we will use an importance weighted estimator.
-
-\eqn{
-\label{eqn:importance_weighted_estimator}
-\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
-}
-
-where $q_i = P(\text{arm $i$ active})$
-
-Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:
-
-\eqn{
-\label{eqn:balancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
-}
-
-\eqn{
-R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
-}
-
-Let $D = \sqrt{\frac{8}{h}\log(hK)}$
-
-\eqn{
-R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)
-}
-
-\subsection*{Total Regret}
-
-Putting together the regret from the exploration and exploitation phases,
-
-\eqn{
-R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)\\
-& \leq h + T\left(\sqrt{\frac{8}{h}\log(TK)} + \frac{1}{h}\right)
-}
-
-Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,
-
-\eqn {
-R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
-}
-
-If $T \geq 2$ and $K \geq 2$, the first term dominates and,
-
-\eqn {
-R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
-}
-
-The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$
-
-\subsection*{Empirical results}
-\begin{figure}[H]
-\centering
-\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
-\includegraphics[width=.5\textwidth]{explore_exploit}
-\end{figure}
-
-\section{Generalizing to unbalanced $\boldsymbol{q}$}
-
-When some arms have low natural probability we cannot rely on exploring them adequately by pure observation. We need to explicitly play them during the exploration phase.
-
-We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
-
-Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
-\leq q_N$.
-Let $m \in [2,N] = \{m:q_m > \frac{1}{m}\}$ Ie if the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
-If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$.
-Let $q_{ij} = \P{X_i = j}$
-
-Suppose we observe for the first $h/2$ timesteps. This is at worst half the optimal.
-
-We then have estimates
-
-\eqn {
-\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
-}
-
-We take this as our estimate for those arms for which $q_{ij} > \frac{1}{m}$
-
-Let $Z_{t,ij} = \ind{Y_t=1,X_{t,i}=j} \sim Bernoulli(q_{ij}\mu_{ij})$,
-
-Chernoff's inequality gives
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
-}
-
-We play each of the remaining $m$ arms $\frac{h}{2m}$ times so for them we get
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m}
-}
-
-So for all the arms
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
-}
-
-and
-
-\eqn{
-\label{eqn:unbalancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/24m}
-}
-
-If we let $D = \sqrt{\frac{24m\log(hK)}{h}}$
-
-\eqn{
-R_T &  \leq h + T\left(\sqrt{\frac{24m\log(hK)}{h}} + \frac{1}{h}\right) \\
-& \leq h + T\left(\sqrt{\frac{24m\log(TK)}{h}} + \frac{1}{h}\right) 
-}
-
-Let $h = T^{2/3}m^{1/3}\log(TK)^{1/3}$
-
-\eqn {
-R_T & \leq 6T^{2/3}m^{1/3}log(KT)^{1/3} + T^{1/3}m^{-1/3}log(KT)^{-1/3}\\ 
-& \leq 7T^{2/3}m^{1/3}log(KT)^{1/3}
-}
-
-Again comparing to the UCB bound of $\sqrt{KT}$, we expect the causal algorithm to do better if $K > m^{2/3}T^{1/3}$, or alternatively if $m \leq \frac{K^{3/2}}{\sqrt{T}}$
-
-To get results that degrade to similar order bounds as UCB when the arms are very unbalanced, I will need to drop the explore/exploit strategy.
-
-\subsection{Generalizing to unknown q}
-We now consider the case where $\boldsymbol{q}$, and thus $m$ is not known in advance. We begin by considering the simple regret
-
-\subsubsection{Bounding the Simple Regret}
-
-\begin{itemize}
-\item $N$ is number of variables.
-\item $q_i$ is probability that $X_{i,t} = 1$ for any $t$
-\item $s_i = \min\set{q_i, 1 - q_i}$
-\item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
-\item $I_t \in \set{1,\ldots,N} \times \set{0,1}$ is the intervention
-\end{itemize}
-
-Let $R_s(h)$ be the simple regret after $h$ time steps, which is
-\eq{
-R_s(h) = \E\left[\mu^* - \mu_{I_h}\right]\,,
-}
-where $\mu^*$ is the mean payoff of the optimal intervention and $I_h$ is the intervention chosen in the $h$th round.
-We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.
-
-\begin{algorithm}[H]
-\caption{Simple Regret Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\State {\bf Input:} $h, N$
-\For{$t \in 1,\ldots,h/2$}
-\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
-\EndFor
-\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
-\eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
-}
-\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
-\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
-\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
-
-\For{$(i,j) \in A$}
-\For{$t \in 1,\ldots,h/2\hat m$}
-\State Choose action $X_{i,t} = j$
-\State Observe reward $r_t(X_{i,t}=j)$
-\EndFor
-\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:simple-regret}
-Define $m = \min\set{2 \leq i \leq N : q_i \geq 1/i}$.
-Then the algorithm given in \cref{alg:simple} satisfies
-\eq{
-R_s(h) \in O\left(\sqrt{\frac{m}{h} \log \left(\frac{NT}{m}\right)}\right)\,.
-}
-\end{theorem}
-
-\begin{lemma}\label{lem:conc1}
-$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
-\end{lemma}
-
-\begin{proof}
-Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
-Then
-\eq{
-\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} Z_t\,.
-}
-Now $Z_1,\ldots,Z_{h/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
-\end{proof}
-
-\begin{lemma}\label{lem:conc2}
-Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
-define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
-If
-\eq{
-h \geq 24m \log\frac{4N}{\delta}\,,
-}
-then
-\eq{
-\P{\hat m \leq 2m} \geq 1 - \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-It is easy to see that the worst case occurs when 
-\eq{
-q_i = \begin{cases}
-0 & \text{if } i < m \\
-\frac{1}{m} & \text{otherwise}\,.
-\end{cases}
-}
-Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that 
-\eq{
-(\forall i) \qquad \left| \hat q_i - q_i\right| 
-&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
-&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
-}
-By assumption 
-\eq{
-h \geq 24m \log \frac{4N}{\delta}\,.
-}
-Therefore with probability at least $1 - \delta$ we have
-\eq{
-(\forall i) \qquad \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.
-}
-Note that for $i$ with $q_i = 0$ it is easy to see that $\hat q_i = 0$ (the variance is zero). The above guarantee that
-for $i$ with $q_i = 1/m$ we have $\hat q_i \in [1/(2m), 3/(2m)]$ for all $i$ with probability at least $1 - \delta$,
-which implies that $\hat s_i \geq 1/(2m)$ for all $i$ where $q_i = 1/m$.
-Therefore $\hat m \leq 2m$ with probability at least $1 - \delta$. 
-\end{proof}
-
-\begin{proof}[Proof of \cref{thm:simple-regret}]
-
-for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
-\eq {
-\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
-}
-
-Via Hoeffding's Inequality
-
-\eq {
-\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
-}
-
-for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.
-
-\eq {
-(i,j)\notin A & \implies \hat{s}_i \geq \frac{1}{\hat m} \\
-& \implies \sum_{t=1}^h \ind{X_i = j} \geq \frac{h}{\hat{m}}
-}
-
-Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
-
-\eq {
-\hat \mu_{i,j} = \frac{1}{Z_{ij}}\sum_{t'=1}^{Z_{ij}}r_{t'}
-}
-
-\eq {
-\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
-= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
-\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
-\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
-}
-
-Applying the union bound over all $2N$ actions.
-\eq{
-& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
-\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
-}
-
-Now by \cref{lem:conc2}, 
-\eq{
-h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
-}
-
-Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that
-
-\eqn{
-\label{eqn:joint_bound_estimates_m}
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
-}
-Suppose $h < 24m \log \frac{4N}{\delta}$. Then
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
-}
-Therefore
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
-}
-Therefore 
-\eq{
-R_s(h) 
-&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
-&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
-}
-as required.
-\end{proof}
-
-\subsubsection{Bounding the Bandit Regret}
-
-\begin{itemize}
-\item Use $24Nlog(4N/\delta)$ samples to estimate $m$
-\item If $\hat m > \frac{N^{3/2}}{\sqrt{T}}$ then stop and do UCB with remaining rounds.
-\item Else use causal explore-exploit and let the exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
-\end{itemize}
-
-Since we don't know $\boldsymbol{q}$, and thus $m$, we can't set the exploration time $h$ in advance based on $m$ as we did for the known $\boldsymbol{q}$ case.
-Instead we first use $24Nlog(4N/\delta)$ samples to estimate $m$.
-If $\hat m < \frac{N^{3/2}}{\sqrt{T}}$ we will continue with the causal algorithm and let the total exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$.
-
-Since $m \leq N$, by \cref{eqn:joint_bound_estimates_m}, we have that with probability at least $1-2\delta$,
-
-\eqn{
-\label{eqn:bounds_hold}
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
-}
-
-So the regret in this case is bounded by,
-
-\eq {
-R(T) & =  2\delta E\left[R(T)|\text{ not } \cref{eqn:bounds_hold}\right] + (1-2\delta)E[R(T)| \cref{eqn:bounds_hold}]  \\
-& \leq 2\delta T + \left(h + T\sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}} \right)\\
- &  \leq \frac{4 \hat m^{1/3}T}{h}  + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(\frac{2Nh}{\hat m^{1/3}}\right)}\right), \text { letting } \delta = \frac{2 \hat m^{1/3}}{h} \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(TK\right)}\right) \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + 2T^{2/3}(2m)^{1/3}log(TK)^{1/3}  \\
-&  \leq 8T^{2/3}m^{1/3}log(TK)^{1/3}  \\
-}
-
-Now make sure I have all the cases right - then have a go with an implementation.
-
-\subsubsection{ Older Attempts}
-We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.
-
-\eqn{
-\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}
-}
-
-Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ and construct $\bar{\boldsymbol{q}}$ such that $\bar{q}_1 \leq \bar{q}_2 \leq ...\leq \bar{q}_N$.
-We then estimate $m$ with $\hat{m} = min_i : \bar{q}_i \geq \frac{1}{i}$
-
-We estimate the reward for the apparently common arms, $i \geq \hat{m}$, as:
-
-\eqn{
-\hat{\mu}_i = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\frac{h}{2}\hat{q}_i} = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\sum_{t=1}^{h/2}\ind{X_i=1}}
-}
-
-We play the remaining arms $\frac{h}{2\hat{m}}$ times and estimate their reward as:
-
-\eqn{
-\hat{\mu}_i = \frac{2\hat{m}}{h}\sum_{t=1}^{h/2\hat{m}}\ind{Y = 1|X_i=1}
-}
-
-We now consider how errors in the estimation of $m$ effect our estimates of the arm rewards.
-For the arms with $i \geq \hat{m}$ we know $\hat{q}_i \geq \frac{1}{\hat{m}}$ and thus our estimate is based on at least $\frac{h}{2\hat{m}}$ observations.
-Similarly we explicitly play the infrequently observed arms, $i < m$,  $\frac{h}{2\hat{m}}$ times.
-Thus our estimates will only be worse than the known $\boldsymbol{q}$ case if $\hat{m} > m$.
-
-For a fixed $m$ the $\boldsymbol{q}$ most likely to lead us to overestimate $m$ is: \todot{Justify this claim.}
-
-\eqn{
-q_i = \begin{cases}
-0 & \text{ if } i < m\\
-\frac{1}{m} & \text{ if } i \geq m
-\end{cases}
-}
-
-We now bound $P(\hat{m} - m > \varphi)$ for this worst case.
-
-For $i < m$,we have $\bar{q}_i = q_i = 0$. For $i \geq m$ \todot{Is this really true, justify}
-\eq{
-P(q_i - \bar{q}_i \geq C) & =  P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})\\
-& \leq  2P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})
-}
-
-Via Bernstein's Inequality:
-\eqn{
-&2P(q_i - \hat{q}_i \geq C | \hat{q}_i \leq \frac{1}{2}) \leq 2\exp(-\frac{hC^2}{4q_i}) := \gamma \\
-\implies &P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(2/\gamma)}{mh}})\leq \gamma
-}
-
-Define $\hat{m}_i = \frac{1}{\hat{q}_i}$.
-
-\eqn{
-\hat{q}_i \geq q_i - C & \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
-& \implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}\\
-& \implies P(\hat{m}_i - m \geq \frac{m^2C}{1-mC}) \leq \gamma
-}
-
-Let $\varphi = \frac{m^2C}{1-mC} \implies C = \frac{\varphi}{m(\varphi+m)} \implies \gamma = 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2}) \implies \varphi = \frac{2m\sqrt{m\log(2/\gamma)}}{\sqrt{h}-2\sqrt{m \log(2/\gamma)}} $
-
-Note this implies that if $h \geq 16m\log(2/\gamma)$ then $\varphi \leq m$
-
-\eqn {
-P(\hat{m}_i - m \geq \varphi) \leq 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2})
-}
-
-If $\hat{m}_i - m \geq \varphi$ for at most $\varphi$ of the variables $i \geq m$, then $\hat{m} - m \leq \varphi$
-
-Let $W_i = \ind{\hat{m_i}-m \geq \varphi}$, $E[W_i] \leq \gamma$,$V[W_i] \leq \gamma$ 
-\eqn{
-P(\sum_{i=m}^N W_i \geq (N-m)\gamma + \varepsilon) \leq \exp(-\frac{\varepsilon^2}{2(N-m)\gamma+2\varepsilon/3})
-}
-Letting $(N-m)\gamma + \varepsilon = \varphi \implies \varepsilon = \varphi - (N-m)\gamma$
-\eqn{
-P(\hat{m}-m \geq \varphi) =  P(\sum_{i=m}^N W_i \geq \varphi) &\leq \exp({-\frac{(\varphi-(N-m)\gamma)^2}{2(N-m)\gamma + 2(\varphi-(N-m)\gamma)/3}})\\
-& \leq \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \text{, where } N\gamma < \varphi
-}
-
-If $P(\hat{m}-m \geq \varphi) \leq \zeta$.
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq & P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \leq \varphi )P(\hat{m}-m \leq \varphi) \\
-& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \varphi)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
-}
-
-Putting it together
-
-\eqn {
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}))^2}{2N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}) + 2\varphi/3}})
-}
-
-Assume $h > 16m\log(2N)$ and . Let $\varphi = \epsilon\sqrt{h}$. Assume $\epsilon \geq \frac{m}{\sqrt{h}}$ so as to ensure $\varphi > m$
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h}{16m}))^2}{2N\exp(-\frac{h}{16m}) + 2\varphi/3}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-\varphi/2)^2}{\varphi + 2\varphi/3}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\varphi}{20}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\epsilon \sqrt{h}}{20}}) := \delta\\
-& \leq \begin{cases}
-2 \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) & \text{ if } h < \frac{9m^2}{289\epsilon^2} \\
-2\exp({-\frac{3\epsilon\sqrt{h}}{20})} & \text{ otherwise } 
-\end{cases}
-}
-
-\eqn{
-\implies \epsilon \leq \begin{cases}
-\frac{1}{\sqrt{h}}\log(2/\delta) + \sqrt{\frac{m}{h}\log(2/\delta)} & \text{ if } h < \frac{9m^2}{289\epsilon^2}\\
-\frac{20}{3\sqrt{h}}\log(2/\delta) & \text{ otherwise }
-\end{cases}\\
-}
-
-\eqn{
-\implies \epsilon \leq \frac{20}{3\sqrt{h}}\log(2/\delta)+ \sqrt{\frac{m}{h}\log(2/\delta)} 
-}
-
-Does this fit with the assumption we made about $\epsilon$???
-
-\pagebreak
-
-The first term will dominate if
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3} \geq \frac{h\epsilon^2}{m+\varphi}
-}
-
-Setting them equal and solving for $\varphi$. Roughly,
-
-\eqn{
-\varphi = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = \frac{1}{2}(\sqrt{4\epsilon^2 h+ m^2} - m)\\
-\implies \epsilon = \sqrt{\frac{\log(1/\delta)(\log(1/\delta)+m)}{h}} \leq \frac{\log(1/\delta)}{\sqrt{h}} + \sqrt{\frac{m}{h}\log(1/\delta)}
-}
-
-Now repeating the above but more exactly ...
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}  \geq & \frac{\varphi^2-2\varphi N\gamma}{4\varphi/3} \; \text{, if } \varphi > 3N\gamma \\
-= & \frac{3}{4}(\varphi - 2N\gamma)
-}
-
-\eqn{
-\frac{3}{4}(\varphi - 2N\gamma) = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = &\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}\\
-\leq & \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}
-}
-
-If we let $\varphi =\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2} $
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp(-\frac{3}{4}(\varphi - 2N\gamma))\\
-\implies \epsilon = \sqrt{\frac{4\log^2(2/\delta)+(3m+6N\gamma)\log(2/\delta)}{3h}}\\
-\leq \frac{2\log(2/\delta)}{\sqrt{3h}}+\sqrt{\frac{m+2N\gamma}{h}\log(2/\delta)}
-}
-Ok - but I still need to ensure there exists a $\gamma$ such that  $N\gamma < \frac{\varphi}{3}$ and $P(\hat{m}_i - m > \varphi) \leq \gamma$
-
-\eqn{
-P(\hat{m}_i - m > a) \leq \exp(-\frac{h a^2}{4m(m+a)^2})
-}
-
-It feels like I have just gone round and round in circles and pushed the problem to somewhere else ... my expression for $\varphi$ now contains both $\epsilon$ and $\gamma$ ...
-
-\eqn{
-\varphi \geq \frac{1}{2}\sqrt{m^2+4N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}
-}
-
-\pagebreak
-
-\eqn{
- \implies & C = \frac{\varphi}{m(\varphi+m)}  \geq \frac{\varphi}{\alpha m^2} \text{ if } \varphi < (\alpha-1)m\\
-\implies & P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4\log(1/\gamma)}{mh}} \right)\right) \leq \gamma
-}
-
-If we let $\gamma = \frac{e^{-h/m}}{N}$
-
-\eqn{
-P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{\log N}{h}+\frac{1}{m})+2\sqrt{\frac{\log N}{mh}+\frac{1}{m^2}} \right)\right) \leq \gamma \\
-\implies P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{2}{m})+2\sqrt{\frac{2}{m^2}} \right)\right) \leq \gamma \; \text{, if } h > m\log N \\
-\implies P\left(\hat{m}_i - m \geq 6\alpha m \right) \leq \gamma
-}
-
-This doesn't work. We need $\hat{m}_i - m < (\alpha-1)m$ in order to get this expression.
-
-Or we could let $\varphi = \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}$
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp({-\frac{h\epsilon^2}{m+\frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}}})
-}
-\eqn{
-\implies \epsilon = & \frac{1}{\sqrt{2}}\sqrt{\frac{3 \log^2(2/\delta)+2m\log(2/\delta)}{h}+\frac{\sqrt{3}}{h^2}\sqrt{h^2\log^2(2/\delta)(3\log^2(2/\delta)+4m\log(2/\delta)+3mN\gamma)}} \\
-\leq & 
-}
+At each timestep, observing will reveal the reward for half the arms.
 
 \end{document}
-
-\subsection{Option 1: Targeted sampling during exploration phase}
-
-The key fact we were utilizing to draw conclusions about multiple arms during each timestep of the explore phase is that, given our assumed causal structure, $P(Y|do(X_i=j)) = P(Y|X_i=j)$
-
-If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ with some probability not equal to $P(X_a=b)$, then we can no longer estimate $P(Y|X_i=j)$ directly from the proportion of successes given $X_i=j$.
-
-\eqn {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- \label{eqn:marginalizedProbability}
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b)
-}
-
-Define:
-\begin{itemize}
-\item  $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$
-\item $q_{ab} = P(X_a = b)$, the underlying probability that variable $a$ takes the value $b$
-\item $p_{ab} = P(do(X_a=b)|I=a)$, the probability that we set variable $a$ to $b$ given we are acting on variable $a$
-\end{itemize}
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-This estimator is unbiased and has variance bounded by (proofs in section \ref{seq:proof_expect_var}) :
-
-\eqn{
-\label{eqn:variance}
-V[\hat{\mu}_a^{ij}] \leq \begin{cases}
-\frac{1}{p_{ij}} & \text{ if } a = i \\
-\frac{1}{q_{ij}}\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right) & \text{ if } a \neq i
-\end{cases}
-}
-
-At each time step, we select and arm $do(X_a=b)$ and then receive estimates for all the arms $i,j$ with variances given by \ref{eqn:variance}.
-Note that $\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right)$ is minimized to equal $1$ if we set the intervened on variable according to its natural probability, that is we effectively take no action.
-
-Define  $\bar{q}_i = min_j q_{ij}$, the probability of the most unlikely value for each variable.
-We can assume without loss of generality that $\bar{q}_1 \leq ...
-\leq \bar{q}_N$.
-Define $m = max(N,min\{m: \frac{1}{m}\leq \bar{q}_m\})$ We allocate half our exploration timesteps $\frac{h}{2}$ to observing and allocate the remainder equally to play the unbalanced arms with $i \leq m$.
-
-At the end of the exploration phase, we combine the estimators from each timestep as a weighted sum with weights selected to minimize the variance of the combined estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{t=1}^h w_t \hat{\mu}_{a(t)}^{ij}, \text{ where } w_t = \frac{V_t^{-1}}{\sum_s V_s^{-1}}
-}
-
-We can then use Bernstein's inequality to bound the probability of large deviations
-
-\eqn{
-P\left(\sum_{t=1}^h w_t  \hat{\mu}_{a(t)}^{ij} - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\eqn {
-\sum_{t=1}^h\frac{1}{V_t} & =  \begin{cases}
-\frac{h}{2}q_{kb} & \text{ if } q_{kb} \geq \frac{1}{m} \\
-\frac{h}{2}q_{kb} + \frac{h}{2m} & \text{ otherwise }
-\end{cases}\\
-& \geq \frac{h}{2m} \; \forall q_{kb}
-}
-
-\eqn{
-P\left(\sum_{t=1}^n w_t \hat{\mu}_{a(t)}^{ij} - \mu^{ij} \geq \frac{4}{3}\sqrt{\frac{4m\log(1/\delta)}{h}}\right) \leq \delta \; \forall (i,j)
-}
-
-Lower bound?
-
-\pagebreak
-
-\subsection{Proofs}
-\label{seq:proof_expect_var}
-\subsubsection{Expectation and variance of the estimators}
-\eq{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-If $a = i$:
-\eq{
-E[\hat{\mu}_a^{ij}] = \frac{1}{p_{ij}}E[Y\ind{X_i=j}] = \frac{1}{p_{ij}}P(Y,do(X_i=j)) = \frac{1}{p_{ij}}P(Y|X_i=j)P(do(X_i=j)) = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2  = \frac{1}{p_{ij}^2}E[Y\ind{X_i=j}] - (\mu^{ij})^2  = \frac{1}{p_{ij}}\mu^{ij} - (\mu^{ij})^2 = \frac{\mu^{ij}(1-p_{ij}\mu^{ij})}{p_{ij}} \leq \frac{\mu^{ij}}{p_{ij}} \\
-\leq & \frac{1}{p_{ij}}
-}
-
-if $a \neq i$
-
-\eq{
-E[\hat{\mu}_a^{ij}] & = \frac{q_{a0}}{q_{ij}p_{a0}}E[Y\ind{X_i=j,X_a=0}]+\frac{q_{a1}}{q_{ij}p_{a1}}E[Y\ind{X_i=j,X_a=1}] \\
-& = \frac{q_{a0}}{q_{ij}p_{a0}}P(Y|X_i=j,do(X_i=0))q_{ij}p_{i0}+\frac{q_{a1}}{q_{ij}p_{a1}}P(Y|X_i=j,do(X_i=1))q_{ij}p_{i1} \\
-& =P(Y|X_i=j,do(X_i=0))P(X_i=0)+P(Y|X_i=j,do(X_i=1))P(X_i=1)\\
-& = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2 \\
- = & E[\ind{Y=1,X_i = j,X_a=0}](\frac{q_{a0}}{q_{ij}p_{a0}})^2 + E[\ind{Y=1,X_i = j,X_a=1}](\frac{q_{a1}}{q_{ij}p_{a1}})^2 \\
-& +  2 E[\ind{Y=1,X_i = j,X_a=0}\ind{Y=1,X_i = j,X_a=1}\frac{q_{a0}}{q_{ij}p_{a0}}\frac{q_{a1}}{q_{ij}p_{a1}}] -  (\mu^{ij})^2\\
-=& P(Y|X_i=j,X_a=0)q_{ij}p_{a0}(\frac{q_{a0}}{q_{ij}p_{a0}})^2 + P(Y|X_i=j,X_a=1)q_{ij}p_{a1}(\frac{q_{a1}}{q_{ij}p_{a1}})^2] - (\mu^{ij})^2 \\
-=& \frac{1}{q_{ij}}\left(P(Y|X_i=j,X_a=0)\frac{q_{a0}^2}{p_{a0}} + P(Y|X_i=j,X_a=1)\frac{q_{a1}^2}{p_{a1}}\right)- (\mu^{ij})^2  \\
-\leq & \frac{1}{q_{ij}}\left(\frac{q_{a0}^2}{p_{a0}} + \frac{q_{a1}^2}{p_{a1}}\right) \\
-}
-
-\subsubsection{Bernstein Bound}
-\begin{theorem}{Berstein's Inequality}
-Let $X_t \in \{X_1 ... X_n\}$  be independent, bounded random variables with mean $\mu$ and variance $V_t$ satisfying $|X_t - \mu| \leq b \; \forall t$. Let $V= \frac{1}{n}\sum_{t=1}^n V_t$ 
-\eq{
-& P(\sum_{t=1}^n X_t - \mu \geq \epsilon) \leq \exp(-\frac{\epsilon^2}{2nV+2b\epsilon/3})\\
-\implies & P\left(\sum_{t=1}^n X_t - \mu \geq \frac{2b\log(1/\delta)}{3} + \sqrt{2nV\log(1/\delta)}\right) \leq \delta
-}
-\end{theorem}
-Let $Z_t = w_t V_t$ where $w_t = V_t^{-1}/\sum V_t^{-1}$.
-\eq{
-V[Z_t] = \frac{V_t^{-1}}{(\sum V_t^{-1})^2}\\
-V = \frac{1}{n}\sum V[Z_t] = \frac{1}{n\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2b_z\log(1/\delta)}{3} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta, \text{ where } b_z = max_t \frac{V_t^{-1}b}{\sum V^{-1}}
-}
-
-While we are observing, $|\hat{\mu}_{kb}| = |\frac{Y\ind{X_k=b}}{q_{kb}}|\leq \frac{1}{q_{kb}}$ and $V_t^{-1} = q_{kb}$ so $V_t^{-1}b = 1$.
-
-When we are explicitly setting $X_k=b$, $|\hat{\mu}_{kb}| = |Y|\leq 1$ and $V_t^{-1} =1$ so $V_t^{-1}b = 1$
-
-\eq{
-b_z = \frac{1}{\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2\log(1/\delta)}{3\sum V_t^{-1}} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta \\
-\implies &  P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\subsection{Notes}
-Based on equation \ref{eqn:marginalizedProbability}, a natural estimator for $\mu^{ij}$, given we acted to set the variable $a$, would be
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{\ind{X_i=j}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{\ind{X_i=j,X_a=0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{\ind{X_i=j,X_a=1}}q_{a1} & \text{ if } a\neq i
-\end{cases}
-}
-
-However, this forces us to deal with random variables (which may be 0) in the denominator.
-We can avoid this issue by using the importance weighted estimator, which replaces these indicators with their expectation.
-
-\eqn{
-\hat{\mu}_a^{ij} & = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{q_{ij}p_{a0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{q_{ij}p_{a1}}q_{a1} & \text{ if } a\neq i
-\end{cases} \\
-&= \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-\subsection{Previous attempt}
-
-For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$. We will pwhere
-
-\eqn{
-m_{a,b}^{ij} &= \sum_{s \in \{t:I_t=(a,b)\}} Z_{ab,s}^{ij} \text{ and,}\\
-Z_{ab,s}^{ij} &= \ind{X_{i,s} = j,Y_s=1} \in \{0,1\}
-}
-
-TODO check/show this estimator is unbiasedool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{a=1}^K w_{a} \hat{\mu}_{a}^{ij} \;\;\text{, where } \sum_{a=1}^K w_a = 1
-}
-
-Putting everything together,
-
-\eqn{
-\hat{\mu}^{ij} = \frac{w_i}{\tau_{ij}}\sum_{s \in \{t: I_t = (i,j)\}} Z_{ij,s}^{ij} + \sum_{a \neq i} \left(\frac{w_a}{q_i(j)}\left[ \frac{q_a}{\tau_{a1}}\sum_{s \in \{t:I_t=(a,1)\}} Z_{a1,s}^{ij}+  \frac{1-q_a}{\tau_{a0}}\sum_{s \in \{t:I_t = (a,0)\}} Z_{a0,s}^{ij} \right] \right)
-}
-
-We now need to show $E[\hat{\mu}^{ij}] = \mu^{ij}$ and get a high probability bound for their difference. For the latter, lets try and use McDiarmids Inequality.
-
-McDiarmid's Inequality states:
-If $Z_i \ci Z_j$ and 
-\eqn{
-|\phi(Z_1...Z_i...Z_N) - \phi(Z_1...Z_i^{'}...Z_N)| < c_i \;\; \forall i
-}
-\eqn{
-\label{eqn:McDiarmid}
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq 
-\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
-\right) \leq \delta
-}
-
-For our problem, $\hat{\mu}^{ij} = \phi(\boldsymbol{Z}^{ij})$. All the $Z's$ are independent as they correspond indicator functions over the results at different timesteps.
-
-\eqn{
-|\phi(...Z_{ij,s}^{ij}...) - \phi(...Z_{ij,s}^{'ij}...)| & \leq \frac{w_i}{\tau{ij}}\;\;\;\; \leftarrow \tau_{ij} \text{ such $Z$'s}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a q_a}{q_i(j)\tau_{a1}} \;\;\;\; \leftarrow \tau_{a1} \text{ such $Z$'s for each $a$}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a (1-q_a)}{q_i(j)\tau_{a0}} \;\;\;\; \leftarrow \tau_{a0} \text{ such $Z$'s for each $a$}
-}
-
-\eqn{
-\label{eqn:sum_ci}
-\sum_l c_l^2 = & \frac{w_i^2}{\tau_{ij}} + \sum_{a \neq i}\frac{w_a^2}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}} +\frac{(1-q_a)^2}{\tau_{a0}}  \right)\\
-= & \sum_{a=1}^K w_a^2 f(a) \text{, where} \\
-f(a) =& \begin{cases}
-\frac{1}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}}+  \frac{(1-q_a)^2}{\tau_{a0}}\right)  & a \neq i\\
-\frac{1}{\tau_{ij}} & a = i
-\end{cases} 
-}
-
-We want to select weights to minimize equation \ref{eqn:sum_ci} so as to achieve as tight a bound as possible in equation \ref{eqn:McDiarmid}.
-
-Applying Legrange Multipliers
-
-\eqn {
-w_a & = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
-\label{eqn:min_sum_ci}
-\sum_i c_i^2 & = \frac{1}{\sum_a \frac{1}{f(a)}}
-}
-
-Substitution \ref{eqn:min_sum_ci} back into \ref{eqn:McDiarmid}
-
-\eqn {
-& P\left(|\hat{\mu}^{ij} - \mu^{ij}| \geq \epsilon \right) \leq 2 \exp \left(-2 \epsilon^2 \sum_{a=1}^K \eta_a^{ij} \right)\\
-& \eta_a^{ij} = \begin{cases}
-\frac{\tau_{a1} \tau_{a0} q_i(j)^2}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2} & a \neq i\\
-\tau_{ij} & a = i
-\end{cases} \\
-}
-
-Substituting this into \ref{eqn:probDeltaToLarge}
-
-\eqn{
-\label{eqn:targeted_prob_delta_too_big}
-P(\Delta_{\hat{i^*}} > D) \leq 2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right)
-}
-
-\eqn{
-R_2 \leq (T-h)\left(D+2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right) \right)
-}
-
-Looking at equation \ref{eqn:targeted_prob_delta_too_big}, let $A = \frac{D^2}{2}$ and
-
-\eqn{
-c_a =  \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) := \frac{1}{2}P(\Delta_{\hat{i^*}} > D)  =  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \sum_{a \neq i} c_a  )}+\sum_{i=1}^N e^{-A ( \tau_{i0} + (1-q_{i})^2 \sum_{a \neq i} c_a)}
-}
-
-The goal is to find some assignment $\boldsymbol{\tau}(\boldsymbol{q})$ to minimize this subject to the constraint that $\sum_i \tau_{i1} + \sum_i \tau_{i0} = h$
-
-No luck attempting to do this directly (via Lagrange multipliers ...)
-
-Note, letting $\tau_a = \tau_{a1}+\tau_{a0}$ then $c_a$ is maximized to equal $\tau_a$ if $\tau_{a1} = q_a \tau_a$
-
-Consider the case where we have only 2 variables.
-
-\eqn{
-f(\boldsymbol{\tau}) = e^{-A(\tau_{11} + q_1^2 c_2)}+
-e^{-A(\tau_{21} + q_2^2 c_1)}+
-e^{-A(\tau_{10} + (1-q_1)^2 c_2)}+
-e^{-A(\tau_{20} + (1-q_2)^2 c_1)}
-}
-
-If we consider the extreme case where $q_1=q_2=1$, ie the arms $X_1=0$ and $X_2=0$ never occur naturally.
-
-\eqn{
-f(\boldsymbol{\tau}) = 2 e^{-A(\tau_{11} + \tau_{21})}+
-e^{-A(\tau_{10})}+
-e^{-A(\tau_{20})}
-}
-
-Which is minimized subject to the constraint $\tau_{11}+\tau_{10}+\tau_{21}+\tau_{20}=h$ if we let $\tau_{10}=\tau_{20} = \frac{Ah-log(2)}{3A}$.
-Note that as $h$ gets large, we play the arms that do not occur naturally roughly $\frac{1}{3}$ of the time as opposed to $\frac{1}{4}$ if we were to play uniformly.
-We only learn anything about the arms that do not naturally occur when we play them directly.
-For the other arms we learn something all the time (provided we have some level of balance in all the other arms...) so it makes sense to play the arms that don't occur naturally a little bit more to keep the overall bounds tight.
-
-Ok, what happens if we just let $\tau_{ij} = \frac{h}{K}$
-
-\eqn {
-c_a = \frac{h}{K}\left(\frac{1}{(1-q_a)^2+q_a^2} \right) \geq \frac{h}{K}
-}
-
-\eqn {
-f \leq \sum_{i=1}^N \left(e^{-A(\frac{h}{K}+q_i^2\frac{(K-1)}{K}h)}+ e^{-A(\frac{h}{K}+(1-q_i)^2\frac{(K-1)}{K}h)}\right) \\
-\leq \sum_{i=1}^N \left(e^{-Ah}+ e^{-A(\frac{h}{K})}\right)
-}
-
-Basically worst case, we learn about half the arms fast.
-
-More generally, it would seem by targeted sampling we should be able to learn all the arms with reasonable natural probability fast - and the remaining arms at the rate $h/K$ (or better if we play them more often).
-
-Approaches to finding some better sampling function.
-
-\begin{enumerate}
-\item Find something motivated by the two variable example and show that it performs better than the uniform sampling approach.
-\item  Consider some other edge cases 
-\end{enumerate}
-
-Lets consider the case $q_1 = 1, q_2...q_N = \frac{1}{2}$ We will have $\tau_{11},\tau_{10},\tau_{ij} = \tau \; \forall i > 1,j$
-
-\eqn {
-c_a = \begin{cases}
-\tau_{11} & \text{ if } a=1\\
-2 \tau & \text{ otherwise}
-\end{cases}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) & = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\tau + \frac{1}{4}(\tau_{11}+2(N-2)\tau))} \\
-& = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})} \\
-& <  e^{-A\tau_{10}} + (2 (N-1)+1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}\\
-& =  e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}
-}
-
-and the constraint is $\tau_{11}+\tau_{10}+2(N-1)\tau = h$
-
-If we maximize $(\frac{\tau_{11}}{4} + \frac{N\tau}{2})$ subject to $\tau_{11}+2(N-1)\tau = (h-\tau_{10})$ we find $\tau = \frac{h-\tau_{10}}{2(N-1)}$ and $\tau_{11} = 0$.
-Note, as $N$ gets large the sensitivity to the trade off between $\tau$ and $\tau_{11}$ becomes negligible.
-
-\eqn {
-\frac{\tau_{11}}{4} + \frac{N\tau}{2} = \begin{cases}
-\frac{(h-\tau_{10})N}{4(N-1)} & \text{if } \tau_{11}=0 \\
-\frac{(h-\tau_{10})N}{4(N-1)} - \frac{1}{4(N-1)} &\text{if } \tau_{11}=(h-\tau_{10})
-\end{cases}
-}
-
-\eqn {
-f(\boldsymbol{\tau}) & <e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{1}{4}(h-\tau_{10}))}
-}
-
-Which is minimized if 
-\eqn {
-\tau_{10} = \frac{h}{5} - \frac{4}{5A}log(\frac{2N-1}{4})
-}
-
-Ok - lets look at the reverse case, where $q_1=\frac{1}{2}, q_2...q_N = 1$
-
-\eqn{
-\sum_a c_a = \begin{cases}
-(N-1)\tau_{11} & \text{ if } a = 1 \\
-2\tau+(N-2)\tau_{11} &\text{ otherwise }
-\end{cases}
-}
-
-The constraint is 
-\eqn {
-2\tau+(N-1)\tau_{11} + (N-1)\tau_{10} = h
-}
-
-\eqn{
-f(\boldsymbol{\tau}) = 2e^{-A(\tau+\frac{1}{4}(N-1)\tau_{11})} + (N-1)e^{-A(2\tau+(N-1)\tau_{11})}+(N-1)e^{-A\tau_{10}}
-}
-
-Holding $\tau_{10}$ constant, the 2nd and 3rd terms are constant. The first is maximized by letting $\tau_{11} = 0, \tau = \frac{h - (N-1)\tau_{10}}{2}$
-
-\eqn{
-f(\boldsymbol{\tau}) & = 2e^{-A\tau} + (N-1)e^{-2A\tau}+(N-1)e^{-A\frac{h - 2\tau}{N-1}} \leftarrow \text{ can't minimize analytically} \\
-\label{eqn:case1balancedapprox}
-& \leq (N+1)e^{-A\tau} +(N-1)e^{-A\frac{h - 2\tau}{N-1}}
-}
-
-which is minimized if
-
-\eqn {
-\tau & = \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N(N+1)}{2(N-1)})\right) \\
-\label{eqn:case1balancedapproxsolution}
-& \sim \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N}{2})\right)
-}
-
-Equation \ref{eqn:case1balancedapproxsolution} is also the solution that leads to the exponentiated terms in equation \ref{eqn:case1balancedapprox} being equal - which is should intuitively be optimal since the coefficients in front of them are very similar.
-
-Yet another case study - if $q_1...q_N = 1$ So none of the arms $X_i = 0$ occur naturally.
-
-\eqn{
-&c_a  = \tau_1 \; \forall a\\
-&\sum_{i \neq a}c_a  = (N-1)\tau_1 \; \forall a\\
-& N(\tau_0+\tau_1) = h\\
-& f(\boldsymbol{\tau}) = N e^{-A N\tau_1} + N e^{-A\tau_0}
-}
-
-minimized if 
-\eqn{
-\tau_0 = \frac{h}{N+1}-\frac{log(N)}{A(N+1)}
-}
-
-Note as $N$ get large - this equates to $\tau_0 \sim \frac{h}{N}$ ie almost all the actions are allocated to the arms that do not occur naturally.
-
-Another case study $q_1...q_{N_1} = 1,q_{N_1+1}...q_N = \frac{1}{2}$ (should encompass all previous examples ...)
-
-\eqn{
-&c_a  = \begin{cases}
-\tau_{11} &\text{ if } a \leq N_1
- \\
-2\tau & \text{otherwise}
-\end{cases}\\
-&\sum_{i \neq a}c_a  = \begin{cases}
-(N_1-1)\tau_{11} + 2 N_2 \tau & \text{ if } a \leq N \\
-N_1\tau_{11} + 2(N_2-1)\tau & \text{ otherwise }
-\end{cases}\\
-&N_1\tau_{10}+N_1\tau_{11}+2N_2\tau = h\\
-&f(\boldsymbol{\tau}) =  N_1 e^{-A\tau_{10}}+ N_1 e^{-A(N_1\tau_{11}+2N_2\tau)} + 2N_2 e^{-A(\frac{N_1}{4}\tau_{11}+\frac{N_2+1}{2}\tau)}
-}
-
-The 2nd and 3rd terms are minimized by letting $\tau_{11} = 0$ (although the result is only weakly dependnent on the trade off between $\tau_{11}$ and $\tau$ provided $N_2 >> 2$).
-
-\eqn{
-f(\boldsymbol{\tau}) &=  N_1 e^{-A\tau_{10}}+ N_1 e^{-A2N_2\tau} + 2N_2 e^{-A\frac{N_2+1}{2}\tau}\\
-& \leq N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A\frac{N_2+1}{2}\tau}\\
-& = N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A(\frac{(N_2+1)(h-N_1\tau_{10})}{4N_2})} 
-}
-
-\eqn{
-\tau_{10} = \left(\frac{N_2+1}{(N_2+1)N_1+4N_2}\right)h-\frac{4N_2\log(\frac{(N_2+1)(2N_2+N_2)}{4N_2})}{A((N_2+1)N_1+4N_2))}
-}
-
-It seems like this function can be optimized in a greedy way.
-That is, start with $\boldsymbol{\tau} = \boldsymbol{0}$, and in each timestep select the $\tau_{ij}$ that leads to the greatest reduction in the error bound.
-This appears to yield broadly consistent results to the analytic solution.
-
-\includegraphics[width=.5\textwidth]{greedy_minimization_close}
-\includegraphics[width=.5\textwidth]{greedy_minimization}
-
-Ok - so when we have many unbalanced variables, we assign almost all actions to the arms corresponding to the rare side.
-However, this will still be $< h/N$ which is only a factor of $2$ better than uniform assignment.
-Doing something smarter than uniform assignment will make more difference where we have a only a few unbalanced variables.
-In this case - smart assignment will lead to bounds almost as tight as the perfectly balanced case.
-Uniform assignment may yield solutions that are almost as bad as the worst case.
-(still better than without using the structure at all)
-
-I still need to cover more fractional $q$'s...
-
-What about if I repeated one of the above analysis but let the unbalanced q = .75?
-
-What can I say overall about worst/best cases/any symmetries in the problem??? Do these solutions lead to the terms in the exponentials being similar?
-
-Goal is to choose something that will give us a simple bound and will be ok for the worst case.
-Its still not entirely clear to me what the worst case will be - after all it depends on the way you assign to $\tau$.
-
-What happens if we assume $\sum_{a\neq i} c_a$ is constant ... this is reasonable if $N$ is large
-
-When you look at each term in $c_a$, its maximized if we play the arms according to their natural probabilities.
-If the arm is balanced, this will also minimize the sum of the two terms for doing $x_i = 0$ and $x_i = 1$.
-If the arm is unbalanced, then we have to trade of balancing this sum versus contributing to the sum of $c_a$.
-Consider just how bad the sum of $c_a$ can be even if we focus entirely on balancing the two terms ...
-maybe its already enough.
-
-Fundamentally the relationship between arms is complex. Its not that we get any direct link from one arm to another - rather pairs of actions tell us something about other arms.
-
-Ok, suppose we spend half our actions, $\frac{h}{2}$, observing - that is selecting variables on which to act uniformly and setting them according to their natural probabilities.
-This contributes $\frac{h}{2N}$ to each $c_a$.
-So $\sum_{i\neq a} c_a \sim \frac{h}{2}$ provided $N-1 \sim N$.
-We don't allocate any more pulls to the more likely arm associated with each variable (this is at most a factor of 2 two few pulls).
-We now decide how to spend the remaining budget.
-
-We can assume without loss of generality that $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq ... \leq q_N$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+\sum_{i=1}^N e^{-A (\frac{h}{8})}
-}
-
-Evenly balanced allocation is $\frac{h}{2N}$.
-For those arms for which $q_i^2 \geq \frac{1}{N}$ they will already get this much information during the observe phase so we will not allocate them any more (again - at worst a factor or 2 two few).
-Lets just allocate the remainder evenly.
-So let $m = min\{m:m\leq N,\frac{1}{m} \leq q_m \}$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+N e^{-A (\frac{h}{8})}
-}
-
-\pagebreak
-
-\subsection{Option 2: Some measure of information gain}
-
-\subsection{Option 3: Three stage algorithm}
-\begin{enumerate}
-\item Observe randomly
-\item Play low probability arms
-\item Pick best and exploit
-\end{enumerate}

### 2015-10-22 causal-bandit/unbalanced_q_slide.tex
prep for tpr talk

--- 2015-10-07 causal-bandit/observe_then_best.tex
@@
 
-\begin{comment}
-\section{Random notes}
-Information directed sampling have developed a general framework to construct algorithms that leverege generic structure between arms to minimize Bayesian regret, by explicitly trading off expected regret and information gain at each timestep.
-
-What can be expressed in the partial monitoring framework that cannot be defined by the feedback graph?
-
-We consider a variant of the stochastic multi-armed bandit problem when we have prior knowledge of the causal structure, but not the functional form, of the relationship between variables.
-
-sequential decision process,exploration-exploitation trade off
-
-Points to present. Why we use regret against single best arm - it makes sense in the stochastic setting. 
-The difference between observational and interventional distributions.
-
-terminology for intervention vs observation.
-Recast bandit algorithm in terms of this.
-Show limitations to existing approaches by pointing to places were existing algorithms eg UCB/explore-exploit/exp3 require unbiased estimates.
-Are unbiased estimates actually impossible in our setting?
-
-This paper is about sequentially choosing actions. 
-Modelling sequntial decition processes.
-
-In the classic multi-armed bandit setting, an ag
-
-General bandit part ...
-define multi-armed bandit, and importance.
-Decision making is fundamental.
-Goal is to select an action to maximize a reward (or minimize a loss).
-In many settings, such as drug trials, online ad placement, etc we can choose actions sequentially.
-This introduces an exploration-exploitation trade-off.
-
-The problem of sequentially choosing actions to minimize some loss wide range of applications in XXXX, YYYY. The multi-armed bandit setting, in which .... .
-
-Now the required causal stuff should go here.
-
-Causal structure can be represented as a directed acyclic graph with nodes representing the variables and an edge from $A \rightarrow B$ meaning $A$ causes $B$.
-Deterministic interventions, consisting of setting variables to some value, are modelled by setting the nodes to the desired values and removing any links coming into them, as their value is now no longer set by other variables in the graph according to the original mechanism.
-
-If the causal graph is known, such that there are no hidden variables causing 2 or more of the variables and the probabilities are all > 0 then the distribution of any variable after any intervention can be inferred (asymptotically) from data collected in absence of any interventions.
-\end{comment}
-
-\section*{Introduction}
-
-Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
-For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
-A widely used framework for sequential decision making is the multi-armed bandit.
-In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary and independent of the reward distribution of other actions.
-At each timestep the agent selects an action and receives a reward sampled iid from the corresponding reward distribution.
-
-An an alternate approach to selecting actions is causal inference.
-Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
-This allows an agent to predict the outcome of an action based on non-experimental data.
-This approach is common in social science, demography, and economics where explicit experimentation may be difficult.
-For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.
-
-We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
-This structure creates dependencies between the rewards of different arms such that selecting one action can provide information on the reward for other actions.
-
-There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
-Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
-There is also an active line of research into bandits with feedback defined by a graph.
-Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
-The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
-
-\section*{Problem Formulation}
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
-As an example, consider a farmer wishing to optimize the yield of her crop.
-She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
-We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
-After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
-This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
-
-\begin{figure}[h]
-\centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\end{figure}
-
-Let $q \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known end point $T$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, N}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^N$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(q_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(r(\tilde X))$ where $r:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
-\end{enumerate}
-
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[r(X)|do(X_i = j)]$. The optimal reward and action are denoted $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
-}
-
-The problem can be treated as a classical multi-armed bandit with $K = 2N$ arms. However, this does not utilize the information provided by the causal assumption.
-
-Now need to expand upon how the assumption gives us extra information.
-
-The difficulties faced due to bias.
-
-To deal with this issue we use a simple explore-exploit algorithm.
-Our algorithm will explore for $h$ time-steps, sampling actions in a way that depends on our prior knowledge of $q$ but is independent of the observed rewards.
-We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
-
-\section*{Results}
-
-Summarize results here and note differences to classic bandit results
-
-\subsection*{Known and Balanced q}
-
-We begin with the simplest case were we assume that $q_i=\frac{1}{2} \forall i$.
-During the exploration phase we sample actions uniformly at random.
-In this case, this is equivalent to purely observing, that is taking no action and allowing all input variables to take their value randomly as  $X_{t,i} \sim \Ber(\frac{1}{2})$.
-
-We will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$.
-
-Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
-
-\subsubsection*{Regret during explore phase}
-Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
-\eqn{
-R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta] < h
-}
-
-\subsubsection*{Regret during exploit phase}
-The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
-
-\eqn{
-\hat{i^*} = argmax_i [\hat{\mu}_i]
-}
-\eqn{
-R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
-}
-
-The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$.
-Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
-\eqn{
-R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
-}
-The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.
-
-Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$.
-If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$
-
-\eqn{
-\label{eqn:probDeltaToLarge}
-P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
-}
-
-If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution.
-Instead we will use an importance weighted estimator.
-
-\eqn{
-\label{eqn:importance_weighted_estimator}
-\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
-}
-
-where $q_i = P(\text{arm $i$ active})$
-
-Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:
-
-\eqn{
-\label{eqn:balancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
-}
-
-\eqn{
-R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
-}
-
-Let $D = \sqrt{\frac{8}{h}\log(hK)}$
-
-\eqn{
-R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)
-}
-
-\subsection*{Total Regret}
-
-Putting together the regret from the exploration and exploitation phases,
-
-\eqn{
-R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)\\
-& \leq h + T\left(\sqrt{\frac{8}{h}\log(TK)} + \frac{1}{h}\right)
-}
-
-Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,
-
-\eqn {
-R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
-}
-
-If $T \geq 2$ and $K \geq 2$, the first term dominates and,
-
-\eqn {
-R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
-}
-
-The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$
-
-\subsection*{Empirical results}
-\begin{figure}[H]
-\centering
-\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
-\includegraphics[width=.5\textwidth]{explore_exploit}
-\end{figure}
-
-\section{Generalizing to unbalanced $\boldsymbol{q}$}
-
-When some arms have low natural probability we cannot rely on exploring them adequately by pure observation. We need to explicitly play them during the exploration phase.
-
-We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
-
-Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
-\leq q_N$.
-Let $m \in [2,N] = \{m:q_m > \frac{1}{m}\}$ Ie if the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
-If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$.
-Let $q_{ij} = \P{X_i = j}$
-
-Suppose we observe for the first $h/2$ timesteps. This is at worst half the optimal.
-
-We then have estimates
-
-\eqn {
-\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
-}
-
-We take this as our estimate for those arms for which $q_{ij} > \frac{1}{m}$
-
-Let $Z_{t,ij} = \ind{Y_t=1,X_{t,i}=j} \sim Bernoulli(q_{ij}\mu_{ij})$,
-
-Chernoff's inequality gives
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
-}
-
-We play each of the remaining $m$ arms $\frac{h}{2m}$ times so for them we get
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m}
-}
-
-So for all the arms
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
-}
-
-and
-
-\eqn{
-\label{eqn:unbalancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/24m}
-}
-
-If we let $D = \sqrt{\frac{24m\log(hK)}{h}}$
-
-\eqn{
-R_T &  \leq h + T\left(\sqrt{\frac{24m\log(hK)}{h}} + \frac{1}{h}\right) \\
-& \leq h + T\left(\sqrt{\frac{24m\log(TK)}{h}} + \frac{1}{h}\right) 
-}
-
-Let $h = T^{2/3}m^{1/3}\log(TK)^{1/3}$
-
-\eqn {
-R_T & \leq 6T^{2/3}m^{1/3}log(KT)^{1/3} + T^{1/3}m^{-1/3}log(KT)^{-1/3}\\ 
-& \leq 7T^{2/3}m^{1/3}log(KT)^{1/3}
-}
-
-Again comparing to the UCB bound of $\sqrt{KT}$, we expect the causal algorithm to do better if $K > m^{2/3}T^{1/3}$, or alternatively if $m \leq \frac{K^{3/2}}{\sqrt{T}}$
-
-To get results that degrade to similar order bounds as UCB when the arms are very unbalanced, I will need to drop the explore/exploit strategy.
-
-\subsection{Generalizing to unknown q}
-We now consider the case where $\boldsymbol{q}$, and thus $m$ is not known in advance. We begin by considering the simple regret
-
-\subsubsection{Bounding the Simple Regret}
-
 \begin{itemize}
-\item $N$ is number of variables.
-\item $q_i$ is probability that $X_{i,t} = 1$ for any $t$
-\item $s_i = \min\set{q_i, 1 - q_i}$
-\item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
-\item $I_t \in \set{1,\ldots,N} \times \set{0,1}$ is the intervention
+\item Need to trade of observing vs explicitly playing low probability arms.
+\item Spend half our exploration time $h$ doing each.
+\item Assume $q_i \in [0,\frac{1}{2}]$ and order variables such that $q_1 \leq q_2 \leq ... \leq q_N$ 
+\item Let $m \in [2,N] = {\min i: q_i \geq \frac{1}{i}}$
+\item Divide the arms into low probability $\{(i,1) : q_i < m\}$ and frequent $\{(i,0) \forall i  \cup (i,1) : q_i > m\}$ 
+\item Divide the $h/2$ explicit play budget between the $m$ low probability arms, giving $\frac{h}{2m}$ samples each.
+\item For the frequent arms, we expect $\sim q_i\frac{h}{2} \geq \frac{h}{2m}$ samples from the observe phase.
 \end{itemize}
 
-Let $R_s(h)$ be the simple regret after $h$ time steps, which is
-\eq{
-R_s(h) = \E\left[\mu^* - \mu_{I_h}\right]\,,
-}
-where $\mu^*$ is the mean payoff of the optimal intervention and $I_h$ is the intervention chosen in the $h$th round.
-We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.
-
-\begin{algorithm}[H]
-\caption{Simple Regret Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\State {\bf Input:} $h, N$
-\For{$t \in 1,\ldots,h/2$}
-\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
-\EndFor
-\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
-\eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
-}
-\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
-\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
-\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
-
-\For{$(i,j) \in A$}
-\For{$t \in 1,\ldots,h/2\hat m$}
-\State Choose action $X_{i,t} = j$
-\State Observe reward $r_t(X_{i,t}=j)$
-\EndFor
-\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:simple-regret}
-Define $m = \min\set{2 \leq i \leq N : q_i \geq 1/i}$.
-Then the algorithm given in \cref{alg:simple} satisfies
-\eq{
-R_s(h) \in O\left(\sqrt{\frac{m}{h} \log \left(\frac{NT}{m}\right)}\right)\,.
-}
-\end{theorem}
-
-\begin{lemma}\label{lem:conc1}
-$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
-\end{lemma}
-
-\begin{proof}
-Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
-Then
-\eq{
-\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} Z_t\,.
-}
-Now $Z_1,\ldots,Z_{h/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
-\end{proof}
-
-\begin{lemma}\label{lem:conc2}
-Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
-define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
-If
-\eq{
-h \geq 24m \log\frac{4N}{\delta}\,,
-}
-then
-\eq{
-\P{\hat m \leq 2m} \geq 1 - \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-It is easy to see that the worst case occurs when 
-\eq{
-q_i = \begin{cases}
-0 & \text{if } i < m \\
-\frac{1}{m} & \text{otherwise}\,.
-\end{cases}
-}
-Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that 
-\eq{
-(\forall i) \qquad \left| \hat q_i - q_i\right| 
-&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
-&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
-}
-By assumption 
-\eq{
-h \geq 24m \log \frac{4N}{\delta}\,.
-}
-Therefore with probability at least $1 - \delta$ we have
-\eq{
-(\forall i) \qquad \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.
-}
-Note that for $i$ with $q_i = 0$ it is easy to see that $\hat q_i = 0$ (the variance is zero). The above guarantee that
-for $i$ with $q_i = 1/m$ we have $\hat q_i \in [1/(2m), 3/(2m)]$ for all $i$ with probability at least $1 - \delta$,
-which implies that $\hat s_i \geq 1/(2m)$ for all $i$ where $q_i = 1/m$.
-Therefore $\hat m \leq 2m$ with probability at least $1 - \delta$. 
-\end{proof}
-
-\begin{proof}[Proof of \cref{thm:simple-regret}]
-
-for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
-\eq {
-\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
-}
-
-Via Hoeffding's Inequality
-
-\eq {
-\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
-}
-
-for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.
-
-\eq {
-(i,j)\notin A & \implies \hat{s}_i \geq \frac{1}{\hat m} \\
-& \implies \sum_{t=1}^h \ind{X_i = j} \geq \frac{h}{\hat{m}}
-}
-
-Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
-
-\eq {
-\hat \mu_{i,j} = \frac{1}{Z_{ij}}\sum_{t'=1}^{Z_{ij}}r_{t'}
-}
-
-\eq {
-\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
-= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
-\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
-\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
-}
-
-Applying the union bound over all $2N$ actions.
-\eq{
-& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
-\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
-}
-
-Now by \cref{lem:conc2}, 
-\eq{
-h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
-}
-
-Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that
-
-\eqn{
-\label{eqn:joint_bound_estimates_m}
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
-}
-Suppose $h < 24m \log \frac{4N}{\delta}$. Then
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
-}
-Therefore
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
-}
-Therefore 
-\eq{
-R_s(h) 
-&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
-&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
-}
-as required.
-\end{proof}
-
-\subsubsection{Bounding the Bandit Regret}
-
-\begin{itemize}
-\item Use $24Nlog(4N/\delta)$ samples to estimate $m$
-\item If $\hat m > \frac{N^{3/2}}{\sqrt{T}}$ then stop and do UCB with remaining rounds.
-\item Else use causal explore-exploit and let the exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
-\end{itemize}
-
-Since we don't know $\boldsymbol{q}$, and thus $m$, we can't set the exploration time $h$ in advance based on $m$ as we did for the known $\boldsymbol{q}$ case.
-Instead we first use $24Nlog(4N/\delta)$ samples to estimate $m$.
-If $\hat m < \frac{N^{3/2}}{\sqrt{T}}$ we will continue with the causal algorithm and let the total exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$.
-
-Since $m \leq N$, by \cref{eqn:joint_bound_estimates_m}, we have that with probability at least $1-2\delta$,
-
-\eqn{
-\label{eqn:bounds_hold}
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
-}
-
-So the regret in this case is bounded by,
-
-\eq {
-R(T) & =  2\delta E\left[R(T)|\text{ not } \cref{eqn:bounds_hold}\right] + (1-2\delta)E[R(T)| \cref{eqn:bounds_hold}]  \\
-& \leq 2\delta T + \left(h + T\sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}} \right)\\
- &  \leq \frac{4 \hat m^{1/3}T}{h}  + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(\frac{2Nh}{\hat m^{1/3}}\right)}\right), \text { letting } \delta = \frac{2 \hat m^{1/3}}{h} \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(TK\right)}\right) \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + 2T^{2/3}(2m)^{1/3}log(TK)^{1/3}  \\
-&  \leq 8T^{2/3}m^{1/3}log(TK)^{1/3}  \\
-}
-
-Now make sure I have all the cases right - then have a go with an implementation.
-
-\subsubsection{ Older Attempts}
-We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.
-
-\eqn{
-\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}
-}
-
-Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ and construct $\bar{\boldsymbol{q}}$ such that $\bar{q}_1 \leq \bar{q}_2 \leq ...\leq \bar{q}_N$.
-We then estimate $m$ with $\hat{m} = min_i : \bar{q}_i \geq \frac{1}{i}$
-
-We estimate the reward for the apparently common arms, $i \geq \hat{m}$, as:
-
-\eqn{
-\hat{\mu}_i = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\frac{h}{2}\hat{q}_i} = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\sum_{t=1}^{h/2}\ind{X_i=1}}
-}
-
-We play the remaining arms $\frac{h}{2\hat{m}}$ times and estimate their reward as:
-
-\eqn{
-\hat{\mu}_i = \frac{2\hat{m}}{h}\sum_{t=1}^{h/2\hat{m}}\ind{Y = 1|X_i=1}
-}
-
-We now consider how errors in the estimation of $m$ effect our estimates of the arm rewards.
-For the arms with $i \geq \hat{m}$ we know $\hat{q}_i \geq \frac{1}{\hat{m}}$ and thus our estimate is based on at least $\frac{h}{2\hat{m}}$ observations.
-Similarly we explicitly play the infrequently observed arms, $i < m$,  $\frac{h}{2\hat{m}}$ times.
-Thus our estimates will only be worse than the known $\boldsymbol{q}$ case if $\hat{m} > m$.
-
-For a fixed $m$ the $\boldsymbol{q}$ most likely to lead us to overestimate $m$ is: \todot{Justify this claim.}
-
-\eqn{
-q_i = \begin{cases}
-0 & \text{ if } i < m\\
-\frac{1}{m} & \text{ if } i \geq m
-\end{cases}
-}
-
-We now bound $P(\hat{m} - m > \varphi)$ for this worst case.
-
-For $i < m$,we have $\bar{q}_i = q_i = 0$. For $i \geq m$ \todot{Is this really true, justify}
-\eq{
-P(q_i - \bar{q}_i \geq C) & =  P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})\\
-& \leq  2P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})
-}
-
-Via Bernstein's Inequality:
-\eqn{
-&2P(q_i - \hat{q}_i \geq C | \hat{q}_i \leq \frac{1}{2}) \leq 2\exp(-\frac{hC^2}{4q_i}) := \gamma \\
-\implies &P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(2/\gamma)}{mh}})\leq \gamma
-}
-
-Define $\hat{m}_i = \frac{1}{\hat{q}_i}$.
-
-\eqn{
-\hat{q}_i \geq q_i - C & \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
-& \implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}\\
-& \implies P(\hat{m}_i - m \geq \frac{m^2C}{1-mC}) \leq \gamma
-}
-
-Let $\varphi = \frac{m^2C}{1-mC} \implies C = \frac{\varphi}{m(\varphi+m)} \implies \gamma = 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2}) \implies \varphi = \frac{2m\sqrt{m\log(2/\gamma)}}{\sqrt{h}-2\sqrt{m \log(2/\gamma)}} $
-
-Note this implies that if $h \geq 16m\log(2/\gamma)$ then $\varphi \leq m$
-
-\eqn {
-P(\hat{m}_i - m \geq \varphi) \leq 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2})
-}
-
-If $\hat{m}_i - m \geq \varphi$ for at most $\varphi$ of the variables $i \geq m$, then $\hat{m} - m \leq \varphi$
-
-Let $W_i = \ind{\hat{m_i}-m \geq \varphi}$, $E[W_i] \leq \gamma$,$V[W_i] \leq \gamma$ 
-\eqn{
-P(\sum_{i=m}^N W_i \geq (N-m)\gamma + \varepsilon) \leq \exp(-\frac{\varepsilon^2}{2(N-m)\gamma+2\varepsilon/3})
-}
-Letting $(N-m)\gamma + \varepsilon = \varphi \implies \varepsilon = \varphi - (N-m)\gamma$
-\eqn{
-P(\hat{m}-m \geq \varphi) =  P(\sum_{i=m}^N W_i \geq \varphi) &\leq \exp({-\frac{(\varphi-(N-m)\gamma)^2}{2(N-m)\gamma + 2(\varphi-(N-m)\gamma)/3}})\\
-& \leq \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \text{, where } N\gamma < \varphi
-}
-
-If $P(\hat{m}-m \geq \varphi) \leq \zeta$.
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq & P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \leq \varphi )P(\hat{m}-m \leq \varphi) \\
-& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \varphi)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
-}
-
-Putting it together
-
-\eqn {
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}))^2}{2N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}) + 2\varphi/3}})
-}
-
-Assume $h > 16m\log(2N)$ and . Let $\varphi = \epsilon\sqrt{h}$. Assume $\epsilon \geq \frac{m}{\sqrt{h}}$ so as to ensure $\varphi > m$
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h}{16m}))^2}{2N\exp(-\frac{h}{16m}) + 2\varphi/3}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-\varphi/2)^2}{\varphi + 2\varphi/3}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\varphi}{20}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\epsilon \sqrt{h}}{20}}) := \delta\\
-& \leq \begin{cases}
-2 \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) & \text{ if } h < \frac{9m^2}{289\epsilon^2} \\
-2\exp({-\frac{3\epsilon\sqrt{h}}{20})} & \text{ otherwise } 
-\end{cases}
-}
-
-\eqn{
-\implies \epsilon \leq \begin{cases}
-\frac{1}{\sqrt{h}}\log(2/\delta) + \sqrt{\frac{m}{h}\log(2/\delta)} & \text{ if } h < \frac{9m^2}{289\epsilon^2}\\
-\frac{20}{3\sqrt{h}}\log(2/\delta) & \text{ otherwise }
-\end{cases}\\
-}
-
-\eqn{
-\implies \epsilon \leq \frac{20}{3\sqrt{h}}\log(2/\delta)+ \sqrt{\frac{m}{h}\log(2/\delta)} 
-}
-
-Does this fit with the assumption we made about $\epsilon$???
-
-\pagebreak
-
-The first term will dominate if
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3} \geq \frac{h\epsilon^2}{m+\varphi}
-}
-
-Setting them equal and solving for $\varphi$. Roughly,
-
-\eqn{
-\varphi = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = \frac{1}{2}(\sqrt{4\epsilon^2 h+ m^2} - m)\\
-\implies \epsilon = \sqrt{\frac{\log(1/\delta)(\log(1/\delta)+m)}{h}} \leq \frac{\log(1/\delta)}{\sqrt{h}} + \sqrt{\frac{m}{h}\log(1/\delta)}
-}
-
-Now repeating the above but more exactly ...
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}  \geq & \frac{\varphi^2-2\varphi N\gamma}{4\varphi/3} \; \text{, if } \varphi > 3N\gamma \\
-= & \frac{3}{4}(\varphi - 2N\gamma)
-}
-
-\eqn{
-\frac{3}{4}(\varphi - 2N\gamma) = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = &\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}\\
-\leq & \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}
-}
-
-If we let $\varphi =\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2} $
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp(-\frac{3}{4}(\varphi - 2N\gamma))\\
-\implies \epsilon = \sqrt{\frac{4\log^2(2/\delta)+(3m+6N\gamma)\log(2/\delta)}{3h}}\\
-\leq \frac{2\log(2/\delta)}{\sqrt{3h}}+\sqrt{\frac{m+2N\gamma}{h}\log(2/\delta)}
-}
-Ok - but I still need to ensure there exists a $\gamma$ such that  $N\gamma < \frac{\varphi}{3}$ and $P(\hat{m}_i - m > \varphi) \leq \gamma$
-
-\eqn{
-P(\hat{m}_i - m > a) \leq \exp(-\frac{h a^2}{4m(m+a)^2})
-}
-
-It feels like I have just gone round and round in circles and pushed the problem to somewhere else ... my expression for $\varphi$ now contains both $\epsilon$ and $\gamma$ ...
-
-\eqn{
-\varphi \geq \frac{1}{2}\sqrt{m^2+4N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}
-}
-
-\pagebreak
-
-\eqn{
- \implies & C = \frac{\varphi}{m(\varphi+m)}  \geq \frac{\varphi}{\alpha m^2} \text{ if } \varphi < (\alpha-1)m\\
-\implies & P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4\log(1/\gamma)}{mh}} \right)\right) \leq \gamma
-}
-
-If we let $\gamma = \frac{e^{-h/m}}{N}$
-
-\eqn{
-P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{\log N}{h}+\frac{1}{m})+2\sqrt{\frac{\log N}{mh}+\frac{1}{m^2}} \right)\right) \leq \gamma \\
-\implies P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{2}{m})+2\sqrt{\frac{2}{m^2}} \right)\right) \leq \gamma \; \text{, if } h > m\log N \\
-\implies P\left(\hat{m}_i - m \geq 6\alpha m \right) \leq \gamma
-}
-
-This doesn't work. We need $\hat{m}_i - m < (\alpha-1)m$ in order to get this expression.
-
-Or we could let $\varphi = \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}$
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp({-\frac{h\epsilon^2}{m+\frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}}})
-}
-\eqn{
-\implies \epsilon = & \frac{1}{\sqrt{2}}\sqrt{\frac{3 \log^2(2/\delta)+2m\log(2/\delta)}{h}+\frac{\sqrt{3}}{h^2}\sqrt{h^2\log^2(2/\delta)(3\log^2(2/\delta)+4m\log(2/\delta)+3mN\gamma)}} \\
-\leq & 
-}
-
 \end{document}
-
-\subsection{Option 1: Targeted sampling during exploration phase}
-
-The key fact we were utilizing to draw conclusions about multiple arms during each timestep of the explore phase is that, given our assumed causal structure, $P(Y|do(X_i=j)) = P(Y|X_i=j)$
-
-If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ with some probability not equal to $P(X_a=b)$, then we can no longer estimate $P(Y|X_i=j)$ directly from the proportion of successes given $X_i=j$.
-
-\eqn {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- \label{eqn:marginalizedProbability}
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b)
-}
-
-Define:
-\begin{itemize}
-\item  $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$
-\item $q_{ab} = P(X_a = b)$, the underlying probability that variable $a$ takes the value $b$
-\item $p_{ab} = P(do(X_a=b)|I=a)$, the probability that we set variable $a$ to $b$ given we are acting on variable $a$
-\end{itemize}
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-This estimator is unbiased and has variance bounded by (proofs in section \ref{seq:proof_expect_var}) :
-
-\eqn{
-\label{eqn:variance}
-V[\hat{\mu}_a^{ij}] \leq \begin{cases}
-\frac{1}{p_{ij}} & \text{ if } a = i \\
-\frac{1}{q_{ij}}\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right) & \text{ if } a \neq i
-\end{cases}
-}
-
-At each time step, we select and arm $do(X_a=b)$ and then receive estimates for all the arms $i,j$ with variances given by \ref{eqn:variance}.
-Note that $\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right)$ is minimized to equal $1$ if we set the intervened on variable according to its natural probability, that is we effectively take no action.
-
-Define  $\bar{q}_i = min_j q_{ij}$, the probability of the most unlikely value for each variable.
-We can assume without loss of generality that $\bar{q}_1 \leq ...
-\leq \bar{q}_N$.
-Define $m = max(N,min\{m: \frac{1}{m}\leq \bar{q}_m\})$ We allocate half our exploration timesteps $\frac{h}{2}$ to observing and allocate the remainder equally to play the unbalanced arms with $i \leq m$.
-
-At the end of the exploration phase, we combine the estimators from each timestep as a weighted sum with weights selected to minimize the variance of the combined estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{t=1}^h w_t \hat{\mu}_{a(t)}^{ij}, \text{ where } w_t = \frac{V_t^{-1}}{\sum_s V_s^{-1}}
-}
-
-We can then use Bernstein's inequality to bound the probability of large deviations
-
-\eqn{
-P\left(\sum_{t=1}^h w_t  \hat{\mu}_{a(t)}^{ij} - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\eqn {
-\sum_{t=1}^h\frac{1}{V_t} & =  \begin{cases}
-\frac{h}{2}q_{kb} & \text{ if } q_{kb} \geq \frac{1}{m} \\
-\frac{h}{2}q_{kb} + \frac{h}{2m} & \text{ otherwise }
-\end{cases}\\
-& \geq \frac{h}{2m} \; \forall q_{kb}
-}
-
-\eqn{
-P\left(\sum_{t=1}^n w_t \hat{\mu}_{a(t)}^{ij} - \mu^{ij} \geq \frac{4}{3}\sqrt{\frac{4m\log(1/\delta)}{h}}\right) \leq \delta \; \forall (i,j)
-}
-
-Lower bound?
-
-\pagebreak
-
-\subsection{Proofs}
-\label{seq:proof_expect_var}
-\subsubsection{Expectation and variance of the estimators}
-\eq{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-If $a = i$:
-\eq{
-E[\hat{\mu}_a^{ij}] = \frac{1}{p_{ij}}E[Y\ind{X_i=j}] = \frac{1}{p_{ij}}P(Y,do(X_i=j)) = \frac{1}{p_{ij}}P(Y|X_i=j)P(do(X_i=j)) = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2  = \frac{1}{p_{ij}^2}E[Y\ind{X_i=j}] - (\mu^{ij})^2  = \frac{1}{p_{ij}}\mu^{ij} - (\mu^{ij})^2 = \frac{\mu^{ij}(1-p_{ij}\mu^{ij})}{p_{ij}} \leq \frac{\mu^{ij}}{p_{ij}} \\
-\leq & \frac{1}{p_{ij}}
-}
-
-if $a \neq i$
-
-\eq{
-E[\hat{\mu}_a^{ij}] & = \frac{q_{a0}}{q_{ij}p_{a0}}E[Y\ind{X_i=j,X_a=0}]+\frac{q_{a1}}{q_{ij}p_{a1}}E[Y\ind{X_i=j,X_a=1}] \\
-& = \frac{q_{a0}}{q_{ij}p_{a0}}P(Y|X_i=j,do(X_i=0))q_{ij}p_{i0}+\frac{q_{a1}}{q_{ij}p_{a1}}P(Y|X_i=j,do(X_i=1))q_{ij}p_{i1} \\
-& =P(Y|X_i=j,do(X_i=0))P(X_i=0)+P(Y|X_i=j,do(X_i=1))P(X_i=1)\\
-& = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2 \\
- = & E[\ind{Y=1,X_i = j,X_a=0}](\frac{q_{a0}}{q_{ij}p_{a0}})^2 + E[\ind{Y=1,X_i = j,X_a=1}](\frac{q_{a1}}{q_{ij}p_{a1}})^2 \\
-& +  2 E[\ind{Y=1,X_i = j,X_a=0}\ind{Y=1,X_i = j,X_a=1}\frac{q_{a0}}{q_{ij}p_{a0}}\frac{q_{a1}}{q_{ij}p_{a1}}] -  (\mu^{ij})^2\\
-=& P(Y|X_i=j,X_a=0)q_{ij}p_{a0}(\frac{q_{a0}}{q_{ij}p_{a0}})^2 + P(Y|X_i=j,X_a=1)q_{ij}p_{a1}(\frac{q_{a1}}{q_{ij}p_{a1}})^2] - (\mu^{ij})^2 \\
-=& \frac{1}{q_{ij}}\left(P(Y|X_i=j,X_a=0)\frac{q_{a0}^2}{p_{a0}} + P(Y|X_i=j,X_a=1)\frac{q_{a1}^2}{p_{a1}}\right)- (\mu^{ij})^2  \\
-\leq & \frac{1}{q_{ij}}\left(\frac{q_{a0}^2}{p_{a0}} + \frac{q_{a1}^2}{p_{a1}}\right) \\
-}
-
-\subsubsection{Bernstein Bound}
-\begin{theorem}{Berstein's Inequality}
-Let $X_t \in \{X_1 ... X_n\}$  be independent, bounded random variables with mean $\mu$ and variance $V_t$ satisfying $|X_t - \mu| \leq b \; \forall t$. Let $V= \frac{1}{n}\sum_{t=1}^n V_t$ 
-\eq{
-& P(\sum_{t=1}^n X_t - \mu \geq \epsilon) \leq \exp(-\frac{\epsilon^2}{2nV+2b\epsilon/3})\\
-\implies & P\left(\sum_{t=1}^n X_t - \mu \geq \frac{2b\log(1/\delta)}{3} + \sqrt{2nV\log(1/\delta)}\right) \leq \delta
-}
-\end{theorem}
-Let $Z_t = w_t V_t$ where $w_t = V_t^{-1}/\sum V_t^{-1}$.
-\eq{
-V[Z_t] = \frac{V_t^{-1}}{(\sum V_t^{-1})^2}\\
-V = \frac{1}{n}\sum V[Z_t] = \frac{1}{n\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2b_z\log(1/\delta)}{3} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta, \text{ where } b_z = max_t \frac{V_t^{-1}b}{\sum V^{-1}}
-}
-
-While we are observing, $|\hat{\mu}_{kb}| = |\frac{Y\ind{X_k=b}}{q_{kb}}|\leq \frac{1}{q_{kb}}$ and $V_t^{-1} = q_{kb}$ so $V_t^{-1}b = 1$.
-
-When we are explicitly setting $X_k=b$, $|\hat{\mu}_{kb}| = |Y|\leq 1$ and $V_t^{-1} =1$ so $V_t^{-1}b = 1$
-
-\eq{
-b_z = \frac{1}{\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2\log(1/\delta)}{3\sum V_t^{-1}} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta \\
-\implies &  P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\subsection{Notes}
-Based on equation \ref{eqn:marginalizedProbability}, a natural estimator for $\mu^{ij}$, given we acted to set the variable $a$, would be
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{\ind{X_i=j}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{\ind{X_i=j,X_a=0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{\ind{X_i=j,X_a=1}}q_{a1} & \text{ if } a\neq i
-\end{cases}
-}
-
-However, this forces us to deal with random variables (which may be 0) in the denominator.
-We can avoid this issue by using the importance weighted estimator, which replaces these indicators with their expectation.
-
-\eqn{
-\hat{\mu}_a^{ij} & = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{q_{ij}p_{a0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{q_{ij}p_{a1}}q_{a1} & \text{ if } a\neq i
-\end{cases} \\
-&= \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-\subsection{Previous attempt}
-
-For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$. We will pwhere
-
-\eqn{
-m_{a,b}^{ij} &= \sum_{s \in \{t:I_t=(a,b)\}} Z_{ab,s}^{ij} \text{ and,}\\
-Z_{ab,s}^{ij} &= \ind{X_{i,s} = j,Y_s=1} \in \{0,1\}
-}
-
-TODO check/show this estimator is unbiasedool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{a=1}^K w_{a} \hat{\mu}_{a}^{ij} \;\;\text{, where } \sum_{a=1}^K w_a = 1
-}
-
-Putting everything together,
-
-\eqn{
-\hat{\mu}^{ij} = \frac{w_i}{\tau_{ij}}\sum_{s \in \{t: I_t = (i,j)\}} Z_{ij,s}^{ij} + \sum_{a \neq i} \left(\frac{w_a}{q_i(j)}\left[ \frac{q_a}{\tau_{a1}}\sum_{s \in \{t:I_t=(a,1)\}} Z_{a1,s}^{ij}+  \frac{1-q_a}{\tau_{a0}}\sum_{s \in \{t:I_t = (a,0)\}} Z_{a0,s}^{ij} \right] \right)
-}
-
-We now need to show $E[\hat{\mu}^{ij}] = \mu^{ij}$ and get a high probability bound for their difference. For the latter, lets try and use McDiarmids Inequality.
-
-McDiarmid's Inequality states:
-If $Z_i \ci Z_j$ and 
-\eqn{
-|\phi(Z_1...Z_i...Z_N) - \phi(Z_1...Z_i^{'}...Z_N)| < c_i \;\; \forall i
-}
-\eqn{
-\label{eqn:McDiarmid}
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq 
-\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
-\right) \leq \delta
-}
-
-For our problem, $\hat{\mu}^{ij} = \phi(\boldsymbol{Z}^{ij})$. All the $Z's$ are independent as they correspond indicator functions over the results at different timesteps.
-
-\eqn{
-|\phi(...Z_{ij,s}^{ij}...) - \phi(...Z_{ij,s}^{'ij}...)| & \leq \frac{w_i}{\tau{ij}}\;\;\;\; \leftarrow \tau_{ij} \text{ such $Z$'s}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a q_a}{q_i(j)\tau_{a1}} \;\;\;\; \leftarrow \tau_{a1} \text{ such $Z$'s for each $a$}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a (1-q_a)}{q_i(j)\tau_{a0}} \;\;\;\; \leftarrow \tau_{a0} \text{ such $Z$'s for each $a$}
-}
-
-\eqn{
-\label{eqn:sum_ci}
-\sum_l c_l^2 = & \frac{w_i^2}{\tau_{ij}} + \sum_{a \neq i}\frac{w_a^2}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}} +\frac{(1-q_a)^2}{\tau_{a0}}  \right)\\
-= & \sum_{a=1}^K w_a^2 f(a) \text{, where} \\
-f(a) =& \begin{cases}
-\frac{1}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}}+  \frac{(1-q_a)^2}{\tau_{a0}}\right)  & a \neq i\\
-\frac{1}{\tau_{ij}} & a = i
-\end{cases} 
-}
-
-We want to select weights to minimize equation \ref{eqn:sum_ci} so as to achieve as tight a bound as possible in equation \ref{eqn:McDiarmid}.
-
-Applying Legrange Multipliers
-
-\eqn {
-w_a & = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
-\label{eqn:min_sum_ci}
-\sum_i c_i^2 & = \frac{1}{\sum_a \frac{1}{f(a)}}
-}
-
-Substitution \ref{eqn:min_sum_ci} back into \ref{eqn:McDiarmid}
-
-\eqn {
-& P\left(|\hat{\mu}^{ij} - \mu^{ij}| \geq \epsilon \right) \leq 2 \exp \left(-2 \epsilon^2 \sum_{a=1}^K \eta_a^{ij} \right)\\
-& \eta_a^{ij} = \begin{cases}
-\frac{\tau_{a1} \tau_{a0} q_i(j)^2}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2} & a \neq i\\
-\tau_{ij} & a = i
-\end{cases} \\
-}
-
-Substituting this into \ref{eqn:probDeltaToLarge}
-
-\eqn{
-\label{eqn:targeted_prob_delta_too_big}
-P(\Delta_{\hat{i^*}} > D) \leq 2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right)
-}
-
-\eqn{
-R_2 \leq (T-h)\left(D+2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right) \right)
-}
-
-Looking at equation \ref{eqn:targeted_prob_delta_too_big}, let $A = \frac{D^2}{2}$ and
-
-\eqn{
-c_a =  \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) := \frac{1}{2}P(\Delta_{\hat{i^*}} > D)  =  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \sum_{a \neq i} c_a  )}+\sum_{i=1}^N e^{-A ( \tau_{i0} + (1-q_{i})^2 \sum_{a \neq i} c_a)}
-}
-
-The goal is to find some assignment $\boldsymbol{\tau}(\boldsymbol{q})$ to minimize this subject to the constraint that $\sum_i \tau_{i1} + \sum_i \tau_{i0} = h$
-
-No luck attempting to do this directly (via Lagrange multipliers ...)
-
-Note, letting $\tau_a = \tau_{a1}+\tau_{a0}$ then $c_a$ is maximized to equal $\tau_a$ if $\tau_{a1} = q_a \tau_a$
-
-Consider the case where we have only 2 variables.
-
-\eqn{
-f(\boldsymbol{\tau}) = e^{-A(\tau_{11} + q_1^2 c_2)}+
-e^{-A(\tau_{21} + q_2^2 c_1)}+
-e^{-A(\tau_{10} + (1-q_1)^2 c_2)}+
-e^{-A(\tau_{20} + (1-q_2)^2 c_1)}
-}
-
-If we consider the extreme case where $q_1=q_2=1$, ie the arms $X_1=0$ and $X_2=0$ never occur naturally.
-
-\eqn{
-f(\boldsymbol{\tau}) = 2 e^{-A(\tau_{11} + \tau_{21})}+
-e^{-A(\tau_{10})}+
-e^{-A(\tau_{20})}
-}
-
-Which is minimized subject to the constraint $\tau_{11}+\tau_{10}+\tau_{21}+\tau_{20}=h$ if we let $\tau_{10}=\tau_{20} = \frac{Ah-log(2)}{3A}$.
-Note that as $h$ gets large, we play the arms that do not occur naturally roughly $\frac{1}{3}$ of the time as opposed to $\frac{1}{4}$ if we were to play uniformly.
-We only learn anything about the arms that do not naturally occur when we play them directly.
-For the other arms we learn something all the time (provided we have some level of balance in all the other arms...) so it makes sense to play the arms that don't occur naturally a little bit more to keep the overall bounds tight.
-
-Ok, what happens if we just let $\tau_{ij} = \frac{h}{K}$
-
-\eqn {
-c_a = \frac{h}{K}\left(\frac{1}{(1-q_a)^2+q_a^2} \right) \geq \frac{h}{K}
-}
-
-\eqn {
-f \leq \sum_{i=1}^N \left(e^{-A(\frac{h}{K}+q_i^2\frac{(K-1)}{K}h)}+ e^{-A(\frac{h}{K}+(1-q_i)^2\frac{(K-1)}{K}h)}\right) \\
-\leq \sum_{i=1}^N \left(e^{-Ah}+ e^{-A(\frac{h}{K})}\right)
-}
-
-Basically worst case, we learn about half the arms fast.
-
-More generally, it would seem by targeted sampling we should be able to learn all the arms with reasonable natural probability fast - and the remaining arms at the rate $h/K$ (or better if we play them more often).
-
-Approaches to finding some better sampling function.
-
-\begin{enumerate}
-\item Find something motivated by the two variable example and show that it performs better than the uniform sampling approach.
-\item  Consider some other edge cases 
-\end{enumerate}
-
-Lets consider the case $q_1 = 1, q_2...q_N = \frac{1}{2}$ We will have $\tau_{11},\tau_{10},\tau_{ij} = \tau \; \forall i > 1,j$
-
-\eqn {
-c_a = \begin{cases}
-\tau_{11} & \text{ if } a=1\\
-2 \tau & \text{ otherwise}
-\end{cases}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) & = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\tau + \frac{1}{4}(\tau_{11}+2(N-2)\tau))} \\
-& = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})} \\
-& <  e^{-A\tau_{10}} + (2 (N-1)+1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}\\
-& =  e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}
-}
-
-and the constraint is $\tau_{11}+\tau_{10}+2(N-1)\tau = h$
-
-If we maximize $(\frac{\tau_{11}}{4} + \frac{N\tau}{2})$ subject to $\tau_{11}+2(N-1)\tau = (h-\tau_{10})$ we find $\tau = \frac{h-\tau_{10}}{2(N-1)}$ and $\tau_{11} = 0$.
-Note, as $N$ gets large the sensitivity to the trade off between $\tau$ and $\tau_{11}$ becomes negligible.
-
-\eqn {
-\frac{\tau_{11}}{4} + \frac{N\tau}{2} = \begin{cases}
-\frac{(h-\tau_{10})N}{4(N-1)} & \text{if } \tau_{11}=0 \\
-\frac{(h-\tau_{10})N}{4(N-1)} - \frac{1}{4(N-1)} &\text{if } \tau_{11}=(h-\tau_{10})
-\end{cases}
-}
-
-\eqn {
-f(\boldsymbol{\tau}) & <e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{1}{4}(h-\tau_{10}))}
-}
-
-Which is minimized if 
-\eqn {
-\tau_{10} = \frac{h}{5} - \frac{4}{5A}log(\frac{2N-1}{4})
-}
-
-Ok - lets look at the reverse case, where $q_1=\frac{1}{2}, q_2...q_N = 1$
-
-\eqn{
-\sum_a c_a = \begin{cases}
-(N-1)\tau_{11} & \text{ if } a = 1 \\
-2\tau+(N-2)\tau_{11} &\text{ otherwise }
-\end{cases}
-}
-
-The constraint is 
-\eqn {
-2\tau+(N-1)\tau_{11} + (N-1)\tau_{10} = h
-}
-
-\eqn{
-f(\boldsymbol{\tau}) = 2e^{-A(\tau+\frac{1}{4}(N-1)\tau_{11})} + (N-1)e^{-A(2\tau+(N-1)\tau_{11})}+(N-1)e^{-A\tau_{10}}
-}
-
-Holding $\tau_{10}$ constant, the 2nd and 3rd terms are constant. The first is maximized by letting $\tau_{11} = 0, \tau = \frac{h - (N-1)\tau_{10}}{2}$
-
-\eqn{
-f(\boldsymbol{\tau}) & = 2e^{-A\tau} + (N-1)e^{-2A\tau}+(N-1)e^{-A\frac{h - 2\tau}{N-1}} \leftarrow \text{ can't minimize analytically} \\
-\label{eqn:case1balancedapprox}
-& \leq (N+1)e^{-A\tau} +(N-1)e^{-A\frac{h - 2\tau}{N-1}}
-}
-
-which is minimized if
-
-\eqn {
-\tau & = \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N(N+1)}{2(N-1)})\right) \\
-\label{eqn:case1balancedapproxsolution}
-& \sim \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N}{2})\right)
-}
-
-Equation \ref{eqn:case1balancedapproxsolution} is also the solution that leads to the exponentiated terms in equation \ref{eqn:case1balancedapprox} being equal - which is should intuitively be optimal since the coefficients in front of them are very similar.
-
-Yet another case study - if $q_1...q_N = 1$ So none of the arms $X_i = 0$ occur naturally.
-
-\eqn{
-&c_a  = \tau_1 \; \forall a\\
-&\sum_{i \neq a}c_a  = (N-1)\tau_1 \; \forall a\\
-& N(\tau_0+\tau_1) = h\\
-& f(\boldsymbol{\tau}) = N e^{-A N\tau_1} + N e^{-A\tau_0}
-}
-
-minimized if 
-\eqn{
-\tau_0 = \frac{h}{N+1}-\frac{log(N)}{A(N+1)}
-}
-
-Note as $N$ get large - this equates to $\tau_0 \sim \frac{h}{N}$ ie almost all the actions are allocated to the arms that do not occur naturally.
-
-Another case study $q_1...q_{N_1} = 1,q_{N_1+1}...q_N = \frac{1}{2}$ (should encompass all previous examples ...)
-
-\eqn{
-&c_a  = \begin{cases}
-\tau_{11} &\text{ if } a \leq N_1
- \\
-2\tau & \text{otherwise}
-\end{cases}\\
-&\sum_{i \neq a}c_a  = \begin{cases}
-(N_1-1)\tau_{11} + 2 N_2 \tau & \text{ if } a \leq N \\
-N_1\tau_{11} + 2(N_2-1)\tau & \text{ otherwise }
-\end{cases}\\
-&N_1\tau_{10}+N_1\tau_{11}+2N_2\tau = h\\
-&f(\boldsymbol{\tau}) =  N_1 e^{-A\tau_{10}}+ N_1 e^{-A(N_1\tau_{11}+2N_2\tau)} + 2N_2 e^{-A(\frac{N_1}{4}\tau_{11}+\frac{N_2+1}{2}\tau)}
-}
-
-The 2nd and 3rd terms are minimized by letting $\tau_{11} = 0$ (although the result is only weakly dependnent on the trade off between $\tau_{11}$ and $\tau$ provided $N_2 >> 2$).
-
-\eqn{
-f(\boldsymbol{\tau}) &=  N_1 e^{-A\tau_{10}}+ N_1 e^{-A2N_2\tau} + 2N_2 e^{-A\frac{N_2+1}{2}\tau}\\
-& \leq N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A\frac{N_2+1}{2}\tau}\\
-& = N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A(\frac{(N_2+1)(h-N_1\tau_{10})}{4N_2})} 
-}
-
-\eqn{
-\tau_{10} = \left(\frac{N_2+1}{(N_2+1)N_1+4N_2}\right)h-\frac{4N_2\log(\frac{(N_2+1)(2N_2+N_2)}{4N_2})}{A((N_2+1)N_1+4N_2))}
-}
-
-It seems like this function can be optimized in a greedy way.
-That is, start with $\boldsymbol{\tau} = \boldsymbol{0}$, and in each timestep select the $\tau_{ij}$ that leads to the greatest reduction in the error bound.
-This appears to yield broadly consistent results to the analytic solution.
-
-\includegraphics[width=.5\textwidth]{greedy_minimization_close}
-\includegraphics[width=.5\textwidth]{greedy_minimization}
-
-Ok - so when we have many unbalanced variables, we assign almost all actions to the arms corresponding to the rare side.
-However, this will still be $< h/N$ which is only a factor of $2$ better than uniform assignment.
-Doing something smarter than uniform assignment will make more difference where we have a only a few unbalanced variables.
-In this case - smart assignment will lead to bounds almost as tight as the perfectly balanced case.
-Uniform assignment may yield solutions that are almost as bad as the worst case.
-(still better than without using the structure at all)
-
-I still need to cover more fractional $q$'s...
-
-What about if I repeated one of the above analysis but let the unbalanced q = .75?
-
-What can I say overall about worst/best cases/any symmetries in the problem??? Do these solutions lead to the terms in the exponentials being similar?
-
-Goal is to choose something that will give us a simple bound and will be ok for the worst case.
-Its still not entirely clear to me what the worst case will be - after all it depends on the way you assign to $\tau$.
-
-What happens if we assume $\sum_{a\neq i} c_a$ is constant ... this is reasonable if $N$ is large
-
-When you look at each term in $c_a$, its maximized if we play the arms according to their natural probabilities.
-If the arm is balanced, this will also minimize the sum of the two terms for doing $x_i = 0$ and $x_i = 1$.
-If the arm is unbalanced, then we have to trade of balancing this sum versus contributing to the sum of $c_a$.
-Consider just how bad the sum of $c_a$ can be even if we focus entirely on balancing the two terms ...
-maybe its already enough.
-
-Fundamentally the relationship between arms is complex. Its not that we get any direct link from one arm to another - rather pairs of actions tell us something about other arms.
-
-Ok, suppose we spend half our actions, $\frac{h}{2}$, observing - that is selecting variables on which to act uniformly and setting them according to their natural probabilities.
-This contributes $\frac{h}{2N}$ to each $c_a$.
-So $\sum_{i\neq a} c_a \sim \frac{h}{2}$ provided $N-1 \sim N$.
-We don't allocate any more pulls to the more likely arm associated with each variable (this is at most a factor of 2 two few pulls).
-We now decide how to spend the remaining budget.
-
-We can assume without loss of generality that $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq ... \leq q_N$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+\sum_{i=1}^N e^{-A (\frac{h}{8})}
-}
-
-Evenly balanced allocation is $\frac{h}{2N}$.
-For those arms for which $q_i^2 \geq \frac{1}{N}$ they will already get this much information during the observe phase so we will not allocate them any more (again - at worst a factor or 2 two few).
-Lets just allocate the remainder evenly.
-So let $m = min\{m:m\leq N,\frac{1}{m} \leq q_m \}$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+N e^{-A (\frac{h}{8})}
-}
-
-\pagebreak
-
-\subsection{Option 2: Some measure of information gain}
-
-\subsection{Option 3: Three stage algorithm}
-\begin{enumerate}
-\item Observe randomly
-\item Play low probability arms
-\item Pick best and exploit
-\end{enumerate}

### 2015-10-22 causality_tutorial.tex
lots of changes - laptop has battery issues
prep for tpr talk

@@
 
-TODO
-1BVAEU
-
 (this section should include explanation of why correlation is not causation)
@@
 Can we estimate the $\alpha, \beta, \gamma, \delta$ or in other words, identify the joint distribution over the counterfactuals $P(Y^{0},Y^{1})$ given the interventional distributions, $P(Y^{0})$ and $P(Y^{1})$?
-The answer is no, putting our constraints and unknowns in matrix form:D-separation still applies in the augmented model.
+The answer is no, putting our constraints and unknowns in matrix form:
 
@@
 
+A comparison of algorithms
+
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{| c | c | p{4cm} | c | c | c |}
+\hline
+  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
+  \hline
+  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
+  \hline
+  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
+  \hline
+  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
+  \hline
+  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
+  \hline
+  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
+  \hline
+  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
+ \hline
+\end{tabular}
+
 \subsubsection{Doing conditional independence tests}

### 2015-10-22 graph_diagram_cholesterol.tex
prep for tpr talk

--- 2014-11-21 wiml_poster/FCI_figure.tex
@@
 \usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
-
+\usepackage[margin=0.7in]{geometry}
 \usepackage{tikz}
@@
     observed/.style={
-           circle,
+           rectangle,
            rounded corners,
            draw=black, thick,
-           minimum width=.5em,
-           minimum height=.5em,
+           minimum width=6em,
+           minimum height=2.2em,
            font=\footnotesize,
            text centered,
-           fill=blue!20!white},
+           },
      latent/.style={
@@
            draw=black, thick, dashed,
-           minimum width=.5em,
-           minimum height=.5em,
+           minimum width=2.2em,
+           minimum height=2.2em,
            font=\footnotesize,
@@
 \begin{document}
+%
 \def\ci{\perp\!\!\!\perp} % from Wikipedia
@@
 \centering
-\begin{subfigure}[t]{0.44\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
-
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, lt/.style={latent}]
+\node[main node](1){Health concious};
+\node[main node, below left=of 1](2){Diet};
+\node[main node, below right=of 1](3){Exercise};
+\node[main node, below right=of 2](4){Weight};
+\node[main node, fill=black!20!white,below left=of 4](5){Y=Cholesterol};
+\node[main node, below =of 5](6){Test result};
+\node[main node, above right=of 3](7){Free time};
 \path[]
- 	(1) edge node {} (4)
-    (2) edge [->]  (1)
-    (3) edge [->]  (1);
-
-\node[main node,right of=3,xshift=1.2em](5){A};
-\node[main node, below right=of 5](6){C};
-\node[main node, above right=of 6](7){B};
-\node[main node, below=of 6](8){D};
-\path[every node/.style={font=\sffamily\footnotesize}]
-	(1) edge [-, >=stealth', blue!30!white, line width=.9em,shorten <=1em,shorten >= 1.7em,
-	preaction={line width = .3em, blue!30!white,-triangle 90,draw,shorten >=.6em,shorten <=1em}
-	] node[black]{FCI} (6)
-	(5) edge [o->] (6)
-	(7) edge [o->] (6)
-	(6) edge [->] (8);
-
+	(1) edge (2) edge (3)
+	(2) edge (4) edge (5)
+	(3) edge (4)
+	(5) edge (6)
+	(7) edge (3);
 \end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.55\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]
-
-\node[main node](1){A};
-\node[main node, below right=of 1](2){B};
-\node[shaded, above right=of 2](3){U};
-\node[main node, below right=of 3](4){C};
-\node[main node, above right=of 4](5){D};
-\node[ below=of 3,yshift=.2cm](11){};
-\path[] (1) edge (2)
-		(3) edge (2) edge (4)
-		(5) edge (4);
-
-\node[main node,right=of 5,xshift=-.5em](6){A};
-\node[main node, below right=of 6](7){B};
-\node[above right=of 7](8){};
-\node[main node, below right=of 8](9){C};
-\node[main node, above right=of 9](10){D};
-\node[below = of 8,yshift=0.2cm](12){};	
-\path[every node/.style={font=\sffamily\footnotesize}]
-(11) edge [-, >=stealth', blue!30!white, line width=.9em,shorten <=3.3em,shorten >= 4.5em,
-	preaction={line width = .3em, blue!30!white,-triangle 90,draw,shorten >=3.7em,shorten <=4em}
-	] node[black,xshift=-.5em]{FCI} (12)
-(6) edge [o->] (7)
-(7) edge [<->] (9)
-(10) edge [o->](9);
-
-\end{tikzpicture}
-\end{subfigure}
 \end{figure}

### 2015-10-22 identify.py

@@ -1,46 +1,125 @@
-
-
-
-# district - start at a node, explore outwards until all found.
-        
-# a recursive algorithm is the obvious approach, is there an alternative?
-
-bi_edges = [('A','B'),('B','C'),('D','C')]
-
-def add_value(dictionary,key,value):
+# code based on "On the Identification of Causal Effects", Tian & Pearl
+import matplotlib.pyplot as plt
+def add_value(dictionary,key,value = None):
     """ adds the value to list of values for relevent key """
     if key in dictionary:
-        dictionary[key].append(value)
+        if value is not None:
+            dictionary[key].append(value)
     else:
-        dictionary[key]=[value]
+        if value is None:
+            dictionary[key] = []
+        else:
+            dictionary[key]=[value]
+
+def intersection(lst1,lst2):
+    return [x for x in lst1 if x in lst2]
     
 
 class Graph():
-    def __init__(self,directed_edges,undirected_edges):
-        self.directed={}
-        self.undirected={}
-        nodes = set([])
+    def __init__(self, graph = None):
+        if graph:
+            self.children = graph.children.copy()
+            self.parents = graph.parents.copy()
+            self.siblings = graph.siblings.copy()
+        else:
+            self.children = {} # each node in graph points to a list of children
+            self.parents = {} # each node in the graph points to a list of parents
+            self.siblings = {} # each node in the graph points to a list of siblings (ie nodes connected with a bidirectional link
+
+    @classmethod
+    def from_edge_lists(cls,directed_edges,undirected_edges):
+        g = cls()
+        g.load_edge_lists(directed_edges,undirected_edges)
+        return g
+
+    def load_edge_lists(self,directed_edges,undirected_edges):
         for from_node,to_node in directed_edges:
-            add_value(self.directed,from_node,to_node)
-            nodes.add(from_node)
-            nodes.add(to_node)
+            self.add_edge(from_node,to_node)
+            
         for a,b in undirected_edges:
-            add_value(self.undirected,a,b)
-            add_value(self.undirected,b,a)
-            nodes.add(a)
-            nodes.add(b)
-        self.nodelist = sorted(list(nodes))
+            self.add_bidirected_edge(a,b)
+        
+
+    def add_edge(self,from_node, to_node):
+        add_value(self.children,from_node,to_node)
+        add_value(self.children,to_node)
[…]

### 2015-10-22 unknown_q.py

@@ -10,35 +10,60 @@ def calc_m(array): # assumes array already sorted smallest to largest
     print("returning len")
     return len(array)
 
-N = 100 # Number of variables
-m = 10 # Number of unbalanced variables
-q = zeros(N)
-q[m-1:] = 1/m
+def simulate_estimation_of_q_and_m(N,m,delta,simulations):
+    # simulate observering for h rounds and estimating bar(q_i)
+    q = zeros(N)
+    q[m-1:] = 1/m
+    h = 12*m*log(4*N/delta)
+    print("h",h)
+    est_m = zeros(simulations)
+    for i in range(simulations):
+        counts = binom.rvs(h,q) #For each q_i the the count will be ~ binomial(h,q_i)
+        est_q = counts/h
+        est_bar_q = sort(minimum(est_q,1-est_q)) #the algorithm doesn't know q_i < .5
+        est_m[i] = calc_m(est_bar_q)
 
-# simulate observering for h rounds and estimating bar(q_i)
-delta = .05
-h = floor(N*m*log(1/delta))
-print(h)
-simulations = 1000
+    freq = stats.itemfreq(est_m)
+    print(freq)
+    fig,ax = plt.subplots(1,1)
+    plt.bar(freq[:,0],freq[:,1],align='center')
+    plt.show()
 
-est_m = zeros(simulations)
-for i in range(simulations):
-    counts = binom.rvs(h,q) #For each q_i the the count will be ~ binomial(h,q_i)
-    est_q = counts/h
-    est_bar_q = sort(minimum(est_q,1-est_q)) #the algorithm doesn't know q_i < .5
-    est_m[i] = calc_m(est_bar_q)
+def m_versus_q(Nvals,simulations):
+    results = zeros((simulations, len(Nvals)))
+    for i,s in enumerate(range(simulations)):
+        for j,N in enumerate(Nvals):
+            q = random.uniform(high=0.5,size = N)
+            q = sort(q)
+            m = calc_m(q)
+            results[i,j] = m
+    f, ax = plt.subplots(1, 1)
+    error = results.std(axis=0)
+    ax.errorbar(Nvals, results.mean(axis=0), yerr=error, fmt='o')
+    #ax.plot(Nvals,results.mean(axis=0))
+    ax.set_xscale('log')
+    ax.set_yscale('log')
+    plt.show()
+    return results
 
[…]

### 2015-10-30 TPR_report.tex
adds substantially to TPR

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}

\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=5em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=5em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }

}

\author{Finnian Lattimore}
\title{Learning how to act: making good decisions with machine learning}

\begin{document}
\def\ci{\perp\!\!\!\perp} % from Wikipedia
\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]
\maketitle

\section{Introduction}

Learning the outcome of an action is central to many real world problems. Does de-worming children in poor countries improve health and educational outcomes [worm-wars]? Would increasing the minimum wage benefit lead to higher unemployment? Will offering this customer a discount improve my revenue? These questions are difficult as they require more than identifying a pattern in data.  There are two, very different, major approaches to this problem with the machine learning community: reinforcement learning and causal inference.

Reinforcement learning addresses the problem of learning from explicitly taking actions. There is typically some state or environment, an agent chooses an action from those available in the current state, the state evolves stochastically as a function of the selected action and the agent then receives some feedback or reward that is a function of the new state. This setting differs from the standard classification problem in that the agent must learn from feedback on the selected action, rather than being presented with the correct action for a given state. A common modelling assumption is to assume that the state evolves only as a function of the previous state and the action chosen, and given these, is independent of the previous history of states and actions. This is known as a Markov decision process or MDP.  A particularly well studied and understood model are MDPs with only a single state. In this case, there are a set of actions, each associated with a fixed but unknown reward distribution and at each time step our agent selects and action and receives corresponding feedback. This is known as the multi-armed bandit problem.

Causal inference makes use of assumptions to allow the outcome of actions to be predicted from observational data. The key to causal inference is a framework that can model how actions change the state of the world. This framework then allows us to map information collected in one setting to another.

Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process. Causal inference with the problem that full randomisation is not always feasible, affordable or ethical.The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined. The goal of my thesis is to explore these questions. In the next sections I briefly review the key literature in causal inference and bandits. I then present a general approach to  how causal models might be incorporated into bandit settings and conclude by demonstrating an algorithm that leverages causal assumptions to improve performance in a specific bandit setting.

\section{Causal Inference}

\subsection{Models of causality and intervention}

Predicting the outcome of actions without explicitly taking them requires assumptions about how the actions will change the system. A very powerful and general model that underlies much of the recent work in causality is the causal bayesian network. A causal Bayesian network, or directed acyclic graph (DAG), is a Bayesian network in which a link $V_{i} \rightarrow V_{j}$, is defined to mean $V_{i}$ directly causes $V_{j}$. This means that if we intervene and change the value of $V_{i}$, we expect $V_{j}$ to change, but if we intervene to change $V_{j}$, $V_{i}$ will not change. More generally, if $G$ is a causal network for a distribution $P$ defined over variables $V_{1}...V_{N}$, then the distribution after an intervention where we set $X \subset V$ to $x$, denoted $do(X=x)$ is obtained by simply dropping the terms for each of the variables in $X$ from the factorization given by the network. This is referred to as the truncated product formula \cite{Pearl2000}.

If the network contains latent (unobservable) variables we will not be able to calculate all the terms in the truncated product formula. However, it may still be possible to determine the post-interventional distribution of specific variables of interest. A general causal query $P(Y|do(X=x))$ is identifiable if it can be shown to be equivalent to an expression containing only pre-interventional quantities. This means that asymptotically we obtain an unbiased estimate for the distribution after an intervention based on purely observational data.

The Do Calculus is a set of three rules that allow transformations of interventional terms to non-interventional ones, given a causal graph \cite{}. They are derived directly from d-seperation properties of graphical models and the definition of intervention in causal DAGs. These rules are complete. A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus and there is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {}. I have implemented this algorithm as a simple javascript application to demonstrate XXXX. If a query is not identifiable, it may still be possible to get bounds for causal effects, for example using instrumental variables \cite{} or by making additional assumptions.

There are two other key frameworks that arise in causal inference. Counterfactuals and structural equation models.  Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}. Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008}.

For example, if we wanted to estimate the causal effect of a medical treatment, then we might let $Y^{1}$ be a counterfactual random variable representing the (binary) potential outcome if treated. The distribution of $Y^{1}$ is the distribution we would see of the outcome, $Y$, if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. The causal effect of the drug is the difference between the probability of recovery, across the population, if everyone was treated, and the probability of recovery given placebo $P(Y^{1})-P(Y^{0})$. This quantity is estimable from observational data if we assume $X \ci Y^{0}$ and $X \ci Y^{1}$. These assumptions are referred to as ignoreability assumptions \cite{Rosenbaum1983}. They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated. Graphically, this is equivalent to the assumption that there is no variable that is a parent of both the treatment $X$ and the outcome $Y$.

Structural equation models (SEMs) describe a deterministic world, where underlying mechanisms determine the output of any process for a given input. The mechanism (but not the output) is assumed to be independent of what is fed into it. Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}. Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables. The noise terms are required to be mutually independent. If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable. Structural equation models can be represented visually as a network. Each variable is a node and arrows are drawn from causes to their effects. If the network for a structural equation model is acyclic then it implies a recursive factorization of the joint distribution over its variables. In other words, it is a causal bayesian network.

Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed are almost equivalent. To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of bayesian networks or as structural equation models or as conditional independence statements involving counterfactual variables (ignorability assumptions).

It is possible to pose causal queries in terms of counterfactuals that are not interventional and cannot be phrased in terms of the do-notation. The scientific and philosophical validity of such counterfactual queries remains under question \cite{Dawid2000,Dawid2014}, however they are nonetheless widely posed in the form of attribution of causal effects to different pathways and mediation \cite{Pearl2014,Imai2010a,VanderWeele2011}.

There are differences between the models we have considered when it comes to these non-interventional queries. Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.  Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made. Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.

In practice, differences in focus and approach across different fields eclipse these actual differences in the models. The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations. The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment. This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}. In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field. Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}. In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.

\subsection{Discovering causal structure}

In the previous section we discussed when assumptions about the structure of the variables in a specific problem is sufficient to identify a causal effect. This approach relies on having enough prior knowledge or theory about the problem to allow you to, at least partially, specify the causal network. In this section, we consider the much harder problem of causal inference where you need to learn the network. Causal inference might seem impossible without specific assumptions about the structure of the variables involved but, amazingly, some aspects of causal structure can be determined from much more general assumptions.

\subsubsection{Discovery based on conditional independence}
One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution. Assume there is some acyclic causal network $G$ that generated the distribution $P(\boldsymbol{V})$ from which our data has been sampled. Our goal is to recover the network from this data.

Since $G$ is a bayesian network, if $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in $P$ . However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network. This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$. This assumption, commonly referred to as  \textbf{\textit{faithfulness}}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$. Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} for $P$.

Faithfulness is an assumption. It does not always hold and we cannot verify it from the observational data we wish to use for causal inference. However, most distributions generated by a causal bayesian network will be faithful to that network. For faithfulness to be violated, different causal effects must exactly balance one-another out. For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure). If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption. However, this independence is not stable. It would disappear under a small perturbation to any of the parameters.

Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$. A wide range of algorithms have been developed based on this key observation, see table \ref{table:discovery_algorithms}. Constraint based methods such as the PC algorithm \cite{}, FCI algorithm \cite{} and RFCI algorithm \cite{}, perform sequential conditional independence tests and eliminate inconsistent graphs. Search and score based methods, such as GES \cite{}, search over the space of graphs and score them according to how well they fit the independences given a complexity penalising prior. Constraint based methods are faster, particularly for sparse graphs, but can lack robustness as errors in early conditional independence tests can propagate. Search and score based methods are more robust for small samples sizes but difficult to scale to larger graphs. This has led to the development of hybrid approaches, such as the MMHC algorithm \cite{}. A key component of causal discovery is the ability to do high dimensional non-parametric conditional independence tests. Developments in kernalized conditional independence tests \cite{},\cite{Zhang2012},\cite{Gretton2008} have made this possible.

\begin{table}[h]
\centering
\caption{A comparison of key causal discovery algorithms}
\label{table:discovery_algorithms}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{| c | c | p{4cm} | c | c |}
\hline
  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent }  \\
  \hline
  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No  \\ 
  \hline
  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes\\
  \hline
  RFCI & Constraint based & ? & 500 & Yes  \\
  \hline
  GES & Search \& Score  & Worst case exponential & 50 & No  \\
  \hline
  MMHC & Hybrid & ? & 5000 & No  \\
 \hline
\end{tabular}
\end{table}

If the end goal of causal discovery is to estimate causal effects, then it may not be necessary to learn the entire graph, only the subset of the graph surrounding target variables of interest. Such local causal discovery techniques can be scaled to problems with many more variables \cite{}. Once a set of causal graphs has been identified, causal effects of interest can be bounded by combining the results for the all the networks. This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}. An implementation is available in the R package \cite{Kalisch2012}

\subsubsection{Discovery with functional models}
All of the algorithms we have considered so far return a Markov equivalence class. They cannot distinguish between two models that result in the same set of conditional independence relations. Consider the very simple case where we have only two variables and the only possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$. These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$. No algorithm relying purely on conditional independence relations can separate these two cases.

One solution is to utilize structural equation models to specify additional assumptions. For example, if we assume that noise is additive, such that $X \rightarrow Y \implies Y = f(X)+\epsilon$, then this will only be invertible such that $X = g(Y) + \epsilon'$ for specific pairs of functions $f$ and noise distributions $\epsilon$. Thus in general we will be able to identify the causal direction \cite{Hoyer2009}. This can be extended to post-non-linear additive noise, $Y = h(f(X)+ \epsilon)$, \cite{Zhang2008a}. These techniques can also be applied over more than two variables \cite{}.

A more general approach is to leverage the assumption that the functions are independent of inputs \cite{}. This leads to the idea that $P(X)$ and $P(Y|X)$ are independent if $X \rightarrow Y$ but not if $Y \rightarrow X$. \cite{} propose testing for this by applying both semi-supervised and standard techniques to estimate $P(Y|X)$. Semi-supervised methods, which utilize additional points from $P(X)$ to learn $P(Y|X)$ should only be able to outperform standard methods if $Y \rightarrow X$.

Finally, rather than explicitly developing an algorithm based on a specific asymmetry between cause and effect, \cite{} propose learning what causality looks like from data. They take as input a dataset where each row of data is itself a dataset in which either $X \rightarrow Y$ or $Y \rightarrow X$ and a corresponding label. Estimates of the distributions $P(X)$, $P(Y)$ and $P()$ for each row are then mapped to features in some kernel space via mean kernel embeddings and finally a standard classification algorithm can be trained to learn label. New datasets where the direction of causality is unknown are then simply mapped to the kernel space and the causal direction is classified according to the trained classifier. In practice, the classifier is trained mostly on simulated data, as it is difficult to find a sufficient set of causal problems with only two variables where the direction of causality is known.

\section{Multi-armed Bandits}

Classical framework. Forms of feedback and structure. Key results graph feedback, etc.

stochastic vs adversarial.

\section{Unifying the frameworks}

stochastic most natural fit because graphical models are stochasitc.

what kinds of stuctural assumptions lead to what?

\subsection{Results for a specific problem}

\subsection{Open questions}

\bibliographystyle{plain}% Select the citation style e.g. ieeetr
\bibliography{library}% write the directory to the .bib file
\end{document}

### 2015-10-30 identify.py

@@ -13,6 +13,12 @@ def add_value(dictionary,key,value = None):
 
 def intersection(lst1,lst2):
     return [x for x in lst1 if x in lst2]
+
+def math_sum(variables,exclude):
+    sum_over  = variables - exclude
+    if len(sum_over) > 0:
+        return "\sum_{%s}" % ",".join(sum_over)
+    return ""
     
 
 class Graph():
@@ -132,6 +138,15 @@ class Graph():
             if dest not in district:
                 district.add(dest)
                 self._find_district(dest,district)
+                
+
+    def indentify2(self,y,x,P,G):
+        V = G.nodes()
+        if len(x) == 0:
+            return math_sum(V,y)+P
+
+
+        
 
 
     def identify(self,C,T,Q):

### 2015-10-30 identify2.py

@@ -1,0 +1,317 @@
+# code based on "On the Identification of Causal Effects", Tian & Pearl
+import matplotlib.pyplot as plt
+
+def add_value(dictionary,key,value = None):
+    """ adds the value to list of values for relevent key """
+    if key in dictionary:
+        if value is not None:
+            dictionary[key].append(value)
+    else:
+        if value is None:
+            dictionary[key] = []
+        else:
+            dictionary[key]=[value]
+
+def intersection(lst1,lst2):
+    return [x for x in lst1 if x in lst2]
+
+def math_sum(variables,exclude):
+    sum_over  = variables - exclude
+    if len(sum_over) > 0:
+        return "\sum_{%s}" % ",".join(sum_over)
+    return ""
+    
+
+class Graph():
+    def __init__(self, graph = None):
+        if graph:
+            self.children = graph.children.copy()
+            self.parents = graph.parents.copy()
+            self.siblings = graph.siblings.copy()
+        else:
+            self.children = {} # each node in graph points to a list of children
+            self.parents = {} # each node in the graph points to a list of parents
+            self.siblings = {} # each node in the graph points to a list of siblings (ie nodes connected with a bidirectional link
+
+    @classmethod
+    def from_edge_lists(cls,directed_edges,undirected_edges):
+        g = cls()
+        g.load_edge_lists(directed_edges,undirected_edges)
+        return g
+
+    def load_edge_lists(self,directed_edges,undirected_edges):
+        for from_node,to_node in directed_edges:
+            self.add_edge(from_node,to_node)
+            
+        for a,b in undirected_edges:
+            self.add_bidirected_edge(a,b)
+        
+
+    def add_edge(self,from_node, to_node):
+        add_value(self.children,from_node,to_node)
+        add_value(self.children,to_node)
+        add_value(self.parents, to_node, from_node)
+        add_value(self.parents,from_node)
+        add_value(self.siblings,to_node)
+        add_value(self.siblings,from_node)
+        self.verify_structure()  
+
+    def add_bidirected_edge(self, a,b):
+        add_value(self.siblings,a,b)
+        add_value(self.siblings,b,a)
+        add_value(self.children,a)
+        add_value(self.children,b)
[…]

### 2015-10-30 test.py

@@ -1,0 +1,30 @@
+
+class Test():
+    def __init__(self):
+        self.parents = 1
+
+    def parents(self,values):
+        return values
+
+t = Test()
+
+def math_condition(left,*right):
+    result = "P("+left
+    first = True
+    for cond in right:
+        print(cond)
+        if len(cond) > 0:
+            if first:
+                result+="|"
+                first = False
+            else:
+                result+=","
+            result+=",".join(cond)
+    result +=")"
+    return result
+
+
+
+
+print(math_condition('x',[],['y'],['z','w']))
+            

### 2015-11-04 TPR_report.tex
adds notes for bandits
adds unification of frameworks section to TPR report

@@
 \usepackage{amsmath}
+\usepackage{dsfont}
 \usepackage{listings}
@@
     observed/.style={
-           rectangle,
+           circle,
            rounded corners,
            draw=black, thick,
-           minimum width=5em,
+           minimum width=2.2em,
            minimum height=2.2em,
@@
      latent/.style={
-           rectangle,
+           circle,
            rounded corners,
            draw=black, thick, dashed,
-           minimum width=5em,
+           minimum width=2.2em,
            minimum height=2.2em,
@@
            fill=black!10!white
+           },
+    target/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!20!white,
+           },
+    observedrect/.style={
+           rectangle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=6em,
+           minimum height=2em,
+           font=\footnotesize,
+           text centered,
+           },
+     targetrect/.style={
+           rectangle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=6em,
+           minimum height=2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!20!white,
+           },
+     empty/.style={
+           circle,
+           rounded corners,
+           minimum width=.5em,
+           minimum height=.5em,
+           font=\footnotesize,
+           text centered,
            },
@@
     shading angle=45 }   
+}
 
-}
+\newcommand{\defined}{\vcentcolon =}
+\newcommand{\rdefined}{=\vcentcolon}
+\newcommand{\E}{\mathbb E}
+\newcommand{\Var}{\operatorname{Var}}
+\newcommand{\calF}{\mathcal F}
+\newcommand{\sr}[1]{\stackrel{#1}}
+\newcommand{\set}[1]{\left\{#1\right\}}
+\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
+\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
+\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
+\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
+\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
+\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
+\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
+\newcommand{\Ber}{\operatorname{Bernoulli}}
+\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
+\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
+\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
+\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
+\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
+
+\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
 
@@
 The goal of my thesis is to explore these questions.
-In the next sections I briefly review the key literature in causal inference and bandits.
+In the next sections I review the key literature in causal inference and bandits.
 I then present a general approach to  how causal models might be incorporated into bandit settings and conclude by demonstrating an algorithm that leverages causal assumptions to improve performance in a specific bandit setting.
@@
 
-Classical framework. Forms of feedback and structure. Key results graph feedback, etc.
+In its classic formulation \cite{Robbins1952} the K-armed bandit ...
 
-stochastic vs adversarial.
+Definition of regret
+
+An algorithm is learning if it can obtain sublinear regret.
+
+Mention of adversarial and markovian alternatives.
+
+\subsection{Regret bounds for classic K-armed bandit}
+
+\begin{itemize}
+\item the lower bound for any algorithm for the K-armed bandit problem (stochastic or adversarial) is $\bigo{\sqrt{nK}}$ \cite{Auer1995}
+
+\item UCB problem dependent upper bound \cite{Auer2002}
+
+\eqn {
+R^{ucb}_\mu(n) \in \bigo{\sum_{i=2}^K \frac{1}{\Delta_i}\log(n)}
+}
+\item UCB problem independent upper bound $\bigo{\sqrt{nK\log(n)}}$ \cite{Bubeck2012}
+
+\item Modifications of UCB can close this gap.
+\cite{Lattimore2015, Audibert2009}.
+Subtle modification of the UCB algorithm closes the gap with the lower bound (ie problem independent upper bound of $\bigo{\sqrt{nK}}$, whilst yielding a similar problem dependent upper bound to UCB.
+\cite{Lattimore2015}
+
+\item Exp3 achieves $\bigo{\sqrt{nK\log(n)}}$ regret \cite{Auer1995}
+
+\item $\bigo{\sqrt{nK}}$ has also been achieved for the adversarial setting. \cite{Audibert2009}
+
+\end{itemize}
+
+\subsection{Extensions to K-armed bandits}
+
+The classic multi-armed bandit is a powerful tool for sequential decision making.
+However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example linear bandits, dependent bandits, X-armed bandits and gaussian process bandits, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
+
+Feedback with expert advice (full feedback) on all actions get $R_n = O(\sqrt{n\log(K)})$
+
+Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
+It has demonstrated that problems fall into one of four categories, trivial with no regret, easy with $R_n = \bigthetatilde{\sqrt{n}}$ , hard with $R_n = \bigtheta{n^{2/3}}$ and hopeless with $R_n = \bigomega{n}$ \cite{Bartok2014}.
+Focus on finite sets of actions and feedback signals.
+Analysis yields algorithms that are optimal with respect to the horizon $n$ but not necessarily other parameters (ie $K$).
 
@@
 
-stochastic most natural fit because graphical models are stochasitc.
+A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph.
+Each possible assignment of variables to values is an action (bandit arm).
+See figure \ref{fig:unify_frameworks} for a simple example.
+The reward could be a general function of the action selected and the final state of the graph.
+However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action.
+The number of actions or arms grows exponentially with the number of variables in the graph, making it important to use algorithms that take account of the graph structure to reduce the search space.
 
-what kinds of stuctural assumptions lead to what?
+Modelling a problem as a causal graph only makes sense when rewards are generated stochastically - since causal graphs fundamentally model probability distributions over variables.
+Thus the connection is to stochastic bandit problems (although adversarial bandits algorithms may be applied to stochastic problems).
 
-\subsection{Results for a specific problem}
+\begin{figure}[h]
+\caption{A simple causal graphical model and corresponding complete action space. A and B represent binary variables that can be intervened on and Y represents the reward.}
+\label{fig:unify_frameworks}
+\centering
+\begin{subfigure}[c]{0.3\textwidth}
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
+\node[main node](1){A};
+\node[tg, below right=of 1](2){Y};
+\node[main node, above right=of 2](3){B};
+\path[]
+	(1) edge (2)
+	(3) edge (2);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.4\textwidth}
+Actions = \begin{tabular}{|c|}
+	\hline
+  do(A=0,B=0) \\
+  do(A=0,B=1) \\
+  do(A=1,B=0) \\
+  do(A=1,B=1) \\
+  \hline
+  do(A=0) \\
+  do(A=1) \\
+  do(B=0) \\
+  do(B=1) \\
+  do() \\
+  \hline
+\end{tabular}
+\end{subfigure}
+\end{figure}
+
+If we begin by considering the case where the causal graph is known, problems then take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected.
+
+If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
+See figure \ref{fig:cholesterol_graph} as an example.
+
+\begin{figure}[h]
+\caption{Example causal graph (based on \cite{Koller2009}) where the outcome of interest (reward) is cholesterol level .
+The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
+For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
+\label{fig:cholesterol_graph}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
+\node[node](1){Health concious};
+\node[node, below left=of 1](2){Diet};
+\node[node, below right=of 1](3){Exercise};
+\node[node, below right=of 2](4){Weight};
+\node[tg,below left=of 4](5){Y=Cholesterol};
+\node[node, below =of 5](6){Test result};
+\node[node, above right=of 3](7){Free time};
+\path[]
+	(1) edge (2) edge (3)
+	(2) edge (4) edge (5)
+	(3) edge (4)
+	(5) edge (6)
+	(7) edge (3);
+\end{tikzpicture}
+\end{figure}
+
+If we receive feedback on additional nodes, the problem can be more interesting.
+In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
+Consider again the model in figure \ref{fig:unify_frameworks}.
+The causal structure implies:
+
+\eqn {
+P(Y|do(A = 0)) &= P(Y|do(),A = 0) \\
+&= P(Y|do(B=0),A=0)P(B=0)+P(Y|do(B=1),A=0)P(B=1) 
+}
+
+Thus we gain information about the reward for the action $do(A=0)$ from selecting the action $do()$ or $do(B = b)$ and then observing $A = 0$.
+
+We only get this form of side information for actions that don't specify the value of every variable, ie those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
+Since the reward distribution for actions that set a subset of the variables is the result of marginalizing out other variables, they can only be optimal if they have lower cost.
+So if the cost of all actions is constant (no matter how many variables must be set), then the problem has the same characteristics as if only the reward node were observable.
+
+If the information on the value of additional nodes is available prior to selecting an action the problem resembles a contextual bandit.
+For example if we observe $A = 0$ then, in deciding between the actions $do(B=0)$ and $do(B=1)$, we would want information on $P(Y|A=0,B=0)$ and $P(Y|A=0,B=1)$.
+Note, side information can still arise if we learn the value of some variables prior to selecting an action and some afterwards.
+
+\subsection{Incorporating side information induced by casual graph}
+
+As a first step towards more general problems, we develop and evaluate a bandit algorithm that incorporates the side information induced by a specific causal graph.
+
+Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
+This is an expansion of the model shown in figure \ref{fig:unify_frameworks}.
+We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
+
+As an example, consider a farmer wishing to optimize the yield of her crop.
+She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
+We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
+After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
+This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
+
+\begin{figure}[h]
+\centering
+\caption{Assumed Causal Structure}
+\label{fig:causalStructure}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty},tg/.style={target}]
+
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[tg, below right=of 2](5){Y};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5);
+
+\end{tikzpicture}
+\end{figure}
+
+Let $\boldsymbol{q} \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known horizon $T$:
+
+\begin{enumerate}
+\item The learner chooses an $I_t \in \set{0,\ldots, N}$ and $J_t \in \set{0,1}$, where $I_t = 0$ represents $do()$.
+
+\item Variables not set by the learner take their values probabilistically according to $\boldsymbol{q}$ 
+\eq{
+X_{t,i} = \begin{cases}
+\sim \Ber(q_i) &\text{if } i \neq I_t \\
+J_t & \text{otherwise}\,.
+\end{cases}
+}
+\item The learner observes $(\boldsymbol{X}_{t},Y_{t})$ and receives reward $Y_t \sim \Ber(r(\boldsymbol{X}_{t}))$ where $r:\set{0,1}^N \to [0,1]$ is unknown and arbitrary. 
+\end{enumerate}
+
+The problem could be treated as a standard bandit problem with $K = 2N + 1$ arms,
+
+\eq {
+\{do(), do(X_1=0), do(X_1 = 1)\;\; ... \;\;do(X_N = 1)\}
+}
+
+yielding a regret $\bigo{\sqrt{TN}}$. However, this doesn't leverage the side information induced by the causal structure:
+
+\eqn {
+\label{eq:observe}
+P(Y_t|do(X_{i,t} = j)) &= P(Y_t|do(),X_{i,t} = j) \\
+\label{eq:estimation_transfer}
+&= P(Y_t|do(X_{l,t}=0),X_{i,t}=j)(1-q_l)+P(Y_t|do(X_{l,t}=1),X_{i,t}=j)q_l 
+}
+
+The hardness of this problem depends strongly on the vector $\boldsymbol{q}$.
+This is intuitive from a consideration of the extreme cases.
+If $\boldsymbol{q} = \boldsymbol{0}$ (or $\boldsymbol{1}$), then half of the actions we wish to explore will only occur if we explicitly select them.
+For these actions we get no side information by playing other actions.
+So at best we could expect a factor of two improvement over simply running a standard bandit algorithm.
+If $\boldsymbol{q} = \boldsymbol{\frac{1}{2}}$, the observe action, $do()$, effectively becomes a revealing action and we would anticipate obtaining a regret with minimal dependence on the number of variables $N$.
+
+We consider a simple explore-exploit based algorithm, that will explore for $h$ timesteps, sampling actions in a way that depends on $\boldsymbol{q}$.
+We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
+During the exploration phase, we will balance purely observing, $I_t = 0$, which takes advantage of equation \ref{eq:observe} to reveal the reward for frequently occurring arms, with explicit sampling of infrequently occurring arms.
+
+Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ... \leq q_N$. Let:
+
+\eqn {
+m \in [2,N] = \set{m:q_m > \frac{1}{m}}
+}
+
+$m$ is a measure of the number of infrequently occurring arms.
+If the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
+If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$.
+Let $q_{ij} = \P{X_i = j}$ and $\mu_{ij} = P(Y|do(X_i = j))$.
+
+Suppose we select $I_t = 0$ for the first $h/2$ timesteps.
+This is at worst half the optimal.
+We use these observations to estimate the rewards for all the frequently occurring arms, $\set{do(X_i = 0) \; \forall i}$, and $\set{X_i = 1:i \geq m}$
+
+\eqn {
+\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
+}
+
+We then explicitly play each of the $m$ infrequently occurring arms $\frac{h}{2m}$ times and estimate their rewards as:
+
+\eqn {
+\hat{\mu}_{ij} = \frac{2m}{h}\sum_{t = 1}^{h/2m}\ind{Y_t = 1|X_i=j}
+}
+
+We then simply play the arm with the highest estimated reward for the remaining $T-h$ timesteps.
+We can show that we can select an $h$ (based on $T$, $N$ and $m$) that leads to worst case regret $R_T = \bigotilde{T^{2/3}m^{1/3}}$.
+This result can still be obtained in the case where $\boldsymbol{q}$ is unknown and must be learned during the exploration time.
+Comparing this to the standard bandit regret, $R_T = \bigo{\sqrt{TK}}$, we expect the causal algorithm to do better if $m < \frac{K^{3/2}}{\sqrt{T}}$.
+
+In this algorithm, we have only used the side information provided by the $do()$ action about other actions.
+Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
+The feedback graph is distinct from the causal graph.
+A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
+For this specific problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
+The Exp3-IX algorithm \cite{Kocak2014} was developed for the adversarial version of this problem and has regret $\bigo{\sqrt{\bar{\alpha}T}}$, where $\bar{\alpha}$ is the average independence number of $G_t$.
+In our case $\bar{\alpha} = \frac{N}{2}$ so we again obtain the regret of the standard bandit algorithm.
+The issue here is that a malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
+This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$
+
+\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
+In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
+\cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
+
+More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
+In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
+The quality of the estimate notably depends not only on the number of times that action was selected.
+For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$.
+This more complex side information does not fit within the graph feedback framework.
 
@@
 
-\bibliographystyle{plain}% Select the citation style e.g. ieeetr
-\bibliography{library}% write the directory to the .bib file
+\begin{itemize}
+\item derive lower bounds
+\item generalize to arbitrary causal graphs
+\item leverage the more complex form of side information as in equation XXX, that provides only partial information about additional actions.
+\item consider cases where the causal graph is not known.
+\end{itemize}
+
+\section{Appendices}
+
+\subsection{Summary of Notation}
+\begin{itemize}
+\item $\bigotilde{.}$ hides polylogarithmic factors in $\bigo{.}$
+\item $\bigthetatilde{.}$ hides polylogarmic factors in $\bigtheta{.}$
+\end{itemize}
+
+\bibliographystyle{unsrt}% Select the citation style e.g. ieeetr
+\bibliography{../library}% write the directory to the .bib file
 \end{document}

### 2015-11-05 TPR_report.tex
completes first draft of TPR

@@
 \newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}{\mathbb E}
+\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
 \newcommand{\Var}{\operatorname{Var}}
@@
 Learning the outcome of an action is central to many real world problems.
-Does de-worming children in poor countries improve health and educational outcomes [worm-wars]?
-Would increasing the minimum wage benefit lead to higher unemployment?
+Does de-worming children in poor countries improve health and educational outcomes \cite{miguel2004,davey2015}?
+Would increasing the minimum wage lead to higher unemployment?
 Will offering this customer a discount improve my revenue?
 These questions are difficult as they require more than identifying a pattern in data.
-There are two, very different, major approaches to this problem with the machine learning community: reinforcement learning and causal inference.
+There are two, very different, approaches to this problem within the machine learning community: reinforcement learning and causal inference.
 
 Reinforcement learning addresses the problem of learning from explicitly taking actions.
-There is typically some state or environment, an agent chooses an action from those available in the current state, the state evolves stochastically as a function of the selected action and the agent then receives some feedback or reward that is a function of the new state.
+There is typically some state or environment.
+An agent chooses an action from those available in the current state.
+The state then evolves stochastically as a function of the selected action and the agent receives some feedback or reward that is a function of the new state.
 This setting differs from the standard classification problem in that the agent must learn from feedback on the selected action, rather than being presented with the correct action for a given state.
-A common modelling assumption is to assume that the state evolves only as a function of the previous state and the action chosen, and given these, is independent of the previous history of states and actions.
+A common modelling assumption is that the state evolves only as a function of the previous state and the action chosen, and given these, is independent of the previous history of states and actions.
 This is known as a Markov decision process or MDP.
-A particularly well studied and understood model are MDPs with only a single state.
+A particularly well studied and understood model is MDPs with only a single state.
 In this case, there are a set of actions, each associated with a fixed but unknown reward distribution and at each time step our agent selects and action and receives corresponding feedback.
@@
 A very powerful and general model that underlies much of the recent work in causality is the causal bayesian network.
-A causal Bayesian network, or directed acyclic graph (DAG), is a Bayesian network in which a link $V_{i} \rightarrow V_{j}$, is defined to mean $V_{i}$ directly causes $V_{j}$.
+A causal Bayesian network, or directed acyclic graph (DAG), is a Bayesian network in which a link $V_{i} \rightarrow V_{j}$ is defined to mean $V_{i}$ directly causes $V_{j}$.
 This means that if we intervene and change the value of $V_{i}$, we expect $V_{j}$ to change, but if we intervene to change $V_{j}$, $V_{i}$ will not change.
@@
 A general causal query $P(Y|do(X=x))$ is identifiable if it can be shown to be equivalent to an expression containing only pre-interventional quantities.
-This means that asymptotically we obtain an unbiased estimate for the distribution after an intervention based on purely observational data.
+This means that, asymptotically, we obtain an unbiased estimate for the distribution after an intervention based on purely observational data.
 
-The Do Calculus is a set of three rules that allow transformations of interventional terms to non-interventional ones, given a causal graph \cite{}.
+The do calculus is a set of three rules that allow transformations of interventional terms to non-interventional ones, given a causal graph \cite{Pearl2000}.
 They are derived directly from d-seperation properties of graphical models and the definition of intervention in causal DAGs.
 These rules are complete.
-A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus and there is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {}.
-I have implemented this algorithm as a simple javascript application to demonstrate XXXX.
-If a query is not identifiable, it may still be possible to get bounds for causal effects, for example using instrumental variables \cite{} or by making additional assumptions.
+A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus \cite{Shpitser2006,Huang2006} and there is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {Shpitser2008} (see XXXX for a javascript demonstration of this algorithm).
+If a query is not identifiable, it may still be possible to get bounds for causal effects, for example using instrumental variables \cite{Angrist2009} or by making additional assumptions.
 
@@
 For example, if we wanted to estimate the causal effect of a medical treatment, then we might let $Y^{1}$ be a counterfactual random variable representing the (binary) potential outcome if treated.
-The distribution of $Y^{1}$ is the distribution we would see of the outcome, $Y$, if everyone was treated.
+The distribution of $Y^{1}$ is the distribution we would see in the outcome $Y$ if everyone was treated.
 Similarly $Y^{0}$ represents the potential outcome for the placebo.
 The causal effect of the drug is the difference between the probability of recovery, across the population, if everyone was treated, and the probability of recovery given placebo $P(Y^{1})-P(Y^{0})$.
-This quantity is estimable from observational data if we assume $X \ci Y^{0}$ and $X \ci Y^{1}$.
+This quantity can be estimated from observational data if we assume $X \ci Y^{0}$ and $X \ci Y^{1}$.
 These assumptions are referred to as ignoreability assumptions \cite{Rosenbaum1983}.
-They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
+They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
 Graphically, this is equivalent to the assumption that there is no variable that is a parent of both the treatment $X$ and the outcome $Y$.
@@
 
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed are almost equivalent.
+Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed are functionally very similar.
 To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of bayesian networks or as structural equation models or as conditional independence statements involving counterfactual variables (ignorability assumptions).
@@
 Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
-Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
+Structural equation models with independent errors allow the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
 
-In practice, differences in focus and approach across different fields eclipse these actual differences in the models.
+In practice, differences in focus and approach across different fields eclipse the actual differences in the models.
 The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
@@
 For faithfulness to be violated, different causal effects must exactly balance one-another out.
-For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
+For example, consider a simple binary variable model of chocolate consumption, income and obesity, figure \ref{fig:unfaithful}.
 If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
@@
 
+\begin{figure}[h]
+\caption{Example of a data generated from a causal graph violating the faithfulness assumption}
+\label{fig:unfaithful}
+\centering
+\begin{subfigure}[t]{0.99\textwidth}
+\centering
+\caption{True causal graph generating $P$}
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,thick,main node/.style={observedrect}]
+\node[main node, align=center,minimum height=4.3em](C){\textbf{Chocolate} \\ \begin{tabular}{| c | c | c |}
+  \hline			
+  Income & H & L \\
+  \hline
+  P(C=H) & .8639 & .2\\
+  \hline
+\end{tabular}};
+\node[main node, above right=of C, align=center](I){\textbf{Income} \\ $P(I=High) = 0.5$};
+\node[main node, below right=of I, align=center,minimum height=4.3em](O){\textbf{Obesity} \\ \begin{tabular}{| c | c | c | c | c |}
+  \hline			
+  Income & \multicolumn{2}{| c |}{H} & \multicolumn{2}{| c |}{L} \\
+  \hline
+  Chocolate & H & L & H & L \\
+  \hline
+  $P(O=Y)$ & .4 & .2 & .7 & .5 \\
+  \hline
+\end{tabular}};
+\path[]
+	(I) edge (C) edge (O)
+	(C) edge (O);
+\end{tikzpicture}
+\vspace{1em}
+\end{subfigure}
+
+\begin{subfigure}[b]{0.99\textwidth}
+\centering
+\caption{Perfect map for $P, \;(C \ci O)$}
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observedrect}, shaded/.style={latent}]
+\node[main node](C){Chocolate};
+\node[main node, below right=of C](I){Income};
+\node[main node, above right=of I](O){Obesity};
+\path[]
+	(C) edge (I)
+	(O) edge (I);
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
 Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$.
 A wide range of algorithms have been developed based on this key observation, see table \ref{table:discovery_algorithms}.
-Constraint based methods such as the PC algorithm \cite{}, FCI algorithm \cite{} and RFCI algorithm \cite{}, perform sequential conditional independence tests and eliminate inconsistent graphs.
-Search and score based methods, such as GES \cite{}, search over the space of graphs and score them according to how well they fit the independences given a complexity penalising prior.
+Constraint based methods such as the PC algorithm \cite{Sprites2000}, FCI algorithm \cite{Sprites2000} and RFCI algorithm \cite{Colombo2012}, perform sequential conditional independence tests and eliminate inconsistent graphs.
+Search and score based methods, such as GES \cite{Chickering2002}, search over the space of graphs and score them according to how well they fit the independences given a complexity penalising prior.
 Constraint based methods are faster, particularly for sparse graphs, but can lack robustness as errors in early conditional independence tests can propagate.
 Search and score based methods are more robust for small samples sizes but difficult to scale to larger graphs.
-This has led to the development of hybrid approaches, such as the MMHC algorithm \cite{}.
+This has led to the development of hybrid approaches, such as the MMHC algorithm \cite{Tsamardinos2006}.
 A key component of causal discovery is the ability to do high dimensional non-parametric conditional independence tests.
-Developments in kernalized conditional independence tests \cite{},\cite{Zhang2012},\cite{Gretton2008} have made this possible.
+Developments in kernalized conditional independence tests,\cite{Gretton2008,Zhang2012} have made this possible.
 
@@
 If the end goal of causal discovery is to estimate causal effects, then it may not be necessary to learn the entire graph, only the subset of the graph surrounding target variables of interest.
-Such local causal discovery techniques can be scaled to problems with many more variables \cite{}.
+Such local causal discovery techniques can be scaled to problems with many more variables \cite{Aliferis2002}.
 Once a set of causal graphs has been identified, causal effects of interest can be bounded by combining the results for the all the networks.
@@
 This can be extended to post-non-linear additive noise, $Y = h(f(X)+ \epsilon)$, \cite{Zhang2008a}.
-These techniques can also be applied over more than two variables \cite{}.
+These techniques can also be applied over more than two variables \cite{Peters2014}.
 
-A more general approach is to leverage the assumption that the functions are independent of inputs \cite{}.
+A more general approach is to leverage the assumption that the functions are independent of inputs \cite{Janzing2012a}.
 This leads to the idea that $P(X)$ and $P(Y|X)$ are independent if $X \rightarrow Y$ but not if $Y \rightarrow X$.
-\cite{} propose testing for this by applying both semi-supervised and standard techniques to estimate $P(Y|X)$.
+\cite{Janzing2012} propose testing for this by applying both semi-supervised and standard techniques to estimate $P(Y|X)$.
 Semi-supervised methods, which utilize additional points from $P(X)$ to learn $P(Y|X)$ should only be able to outperform standard methods if $Y \rightarrow X$.
 
-Finally, rather than explicitly developing an algorithm based on a specific asymmetry between cause and effect, \cite{} propose learning what causality looks like from data.
+Finally, rather than explicitly developing an algorithm based on a specific asymmetry between cause and effect, \cite{LopezPaz2014} propose learning what causality looks like from data.
 They take as input a dataset where each row of data is itself a dataset in which either $X \rightarrow Y$ or $Y \rightarrow X$ and a corresponding label.
@@
 
-In its classic formulation \cite{Robbins1952} the K-armed bandit ...
+In its classic formulation \cite{Robbins1952} the (stochastic) K-armed bandit describes a sequential decision making problem, with $K$ possible actions or arms.
+Each arm $i$ is associated with a fixed but unknown reward distribution $\nu_i$.
+For each timestep $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$.
+The goal of the learner is to maximize the total reward they receive.
+This problem introduces the fundamental exploration-exploitation trade-off.
+The learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
 
-Definition of regret
-
-An algorithm is learning if it can obtain sublinear regret.
-
-Mention of adversarial and markovian alternatives.
-
-\subsection{Regret bounds for classic K-armed bandit}
-
-\begin{itemize}
-\item the lower bound for any algorithm for the K-armed bandit problem (stochastic or adversarial) is $\bigo{\sqrt{nK}}$ \cite{Auer1995}
-
-\item UCB problem dependent upper bound \cite{Auer2002}
+The performance of bandit algorithms is generally described by the (pseudo) regret, $R(T)$.
+This is the difference between the expected reward obtained by the algorithm and the expected reward of selecting the best action in every timestep.
 
 \eqn {
-R^{ucb}_\mu(n) \in \bigo{\sum_{i=2}^K \frac{1}{\Delta_i}\log(n)}
+R(T) = max_{\set{i=1...K}} \E{\sum_{t=1}^T g_{i,t}} - \E{\sum_{t=1}^T g_{I_t,t}}
 }
-\item UCB problem independent upper bound $\bigo{\sqrt{nK\log(n)}}$ \cite{Bubeck2012}
 
-\item Modifications of UCB can close this gap.
-\cite{Lattimore2015, Audibert2009}.
-Subtle modification of the UCB algorithm closes the gap with the lower bound (ie problem independent upper bound of $\bigo{\sqrt{nK}}$, whilst yielding a similar problem dependent upper bound to UCB.
-\cite{Lattimore2015}
+If we let $\mu_i = \E{\nu_i}$ denote the expected reward for each arm $i$ and $\mu^* = max_{\set{i=1...K}}(\mu_i)$ denote the reward for the best arm:
 
-\item Exp3 achieves $\bigo{\sqrt{nK\log(n)}}$ regret \cite{Auer1995}
+\eqn {
+R(T) = T\mu^* - \sum_{t=1}^T E[\mu_{I_t}] 
+}
 
-\item $\bigo{\sqrt{nK}}$ has also been achieved for the adversarial setting. \cite{Audibert2009}
+An algorithm is learning if it obtains regret that is sublinear in $T$.
+The lower bound on the worst case regret for any algorithm (stochastic or adversarial) for the K-armed bandit problem is  $\bigomega{\sqrt{TK}}$ \cite{Auer1995}.
 
-\end{itemize}
+A key algorithm for stochastic bandits, with tractable analysis and strong performance guarantees, is the UCB algorithm \cite{Auer2002}.
+They key to this algorithm is that it keeps track of an upper confidence bound (hence UCB) on the expected reward for each arm and selects the arm with the highest one.
+This balances exploration and exploitation as an arm with a high upper confidence bound must have either a high expected reward or large uncertainty on the expected reward.
+Assume for notational simplicity that $\mu_1 > \mu_2 > ...
+> \mu_K$, such that $\mu^* = \mu_1$, and let $\Delta_i = \mu_i - \mu^*$ be the sub-optimality for each arm.
+The (problem dependent) regret for UCB is bounded by:
 
-\subsection{Extensions to K-armed bandits}
+\eqn {
+R^{ucb}(T) \in \bigo{\sum_{i=2}^K \frac{1}{\Delta_i}\log(T)}
+}
+
+This bound blows up as differences $\Delta_i \rightarrow 0$, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
+In the worst case, $R^{ucb}(T) = \bigo{\sqrt{TK\log(T)}}$ \cite{Bubeck2012}.
+Subtle modifications to the UCB algorithm can eliminate the logarithmic term in this worst case regret bound.
+This yields $R^{ucb}(T) = \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.
+
+Adversarial bandits are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
+Instead, simultaneously with the learner selecting an action $I_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{g}_t$.
+As in the stochastic setting, the learner then receives reward $g_{I_t,t}$.
+The seminal algorithm for adversarial bandits is Exp-3, which, like UCB, obtains regret $\bigo{\sqrt{TK\log(T)}}$ regret \cite{Auer1995}.
+Optimal algorithms, with $R(T) = \bigo{\sqrt{TK}}$, have also been demonstrated for the adversarial setting \cite{Audibert2009}.
 
@@
 This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example linear bandits, dependent bandits, X-armed bandits and gaussian process bandits, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
-
-Feedback with expert advice (full feedback) on all actions get $R_n = O(\sqrt{n\log(K)})$
-
-Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It has demonstrated that problems fall into one of four categories, trivial with no regret, easy with $R_n = \bigthetatilde{\sqrt{n}}$ , hard with $R_n = \bigtheta{n^{2/3}}$ and hopeless with $R_n = \bigomega{n}$ \cite{Bartok2014}.
-Focus on finite sets of actions and feedback signals.
-Analysis yields algorithms that are optimal with respect to the horizon $n$ but not necessarily other parameters (ie $K$).
 
@@
 Comparing this to the standard bandit regret, $R_T = \bigo{\sqrt{TK}}$, we expect the causal algorithm to do better if $m < \frac{K^{3/2}}{\sqrt{T}}$.
+\footnote{ $\bigotilde{.}$ and $\bigthetatilde{.}$ suppress polylogarithmic factors}
 
@@
 
+Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
+It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
+Partial monitoring algorithms yield results that are optimal with respect the the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
+
 \subsection{Open questions}
 
-\begin{itemize}
-\item derive lower bounds
-\item generalize to arbitrary causal graphs
-\item leverage the more complex form of side information as in equation XXX, that provides only partial information about additional actions.
-\item consider cases where the causal graph is not known.
-\end{itemize}
+There are many open questions still to be addressed.
+For the specific setting described above I need to derive lower bounds for the worst case regret so as to determine how close the current approach is to optimal.
 
-\section{Appendices}
+It is clearly desirable to obtain an algorithm that could be generalized to an arbitrary causal graph.
+This will likely require the development of a bandit algorithm that can leverage more complex forms of side information like that in equation \ref{eq:estimation_transfer}, in addition to an efficient algorithm for eliminating actions that need not be considered at all.
 
-\subsection{Summary of Notation}
-\begin{itemize}
-\item $\bigotilde{.}$ hides polylogarithmic factors in $\bigo{.}$
-\item $\bigthetatilde{.}$ hides polylogarmic factors in $\bigtheta{.}$
-\end{itemize}
+Another open class of problem to consider are cases where the causal structure is not (fully) known and also needs to be learnt within the online setting.
 

### 2015-11-13 causal-bandit/lowerbound_auer.tex
typed up auer lowerbound proof

\documentclass{article}

\usepackage{todonotes}

%\usepackage[disable]{todonotes}
\newcommand{\todot}[2][]{\todo[color=red!20!white,#1]{#2}}
\newcommand{\todof}[2][]{\todo[color=orange!20!white,#1]{#2}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[boxed]{algorithm}
\usepackage{algpseudocode}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{comment}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{tikz} 
\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}

\renewcommand{\Pi}[1]{P_i\left( #1 \right)}
\newcommand{\Pu}[1]{P_{unif}\left( #1 \right)}
\newcommand{\Ps}[1]{P_{*}\left( #1 \right)}
\newcommand{\Ei}[1]{\E_i\left[ #1 \right]}
\newcommand{\Eu}[1]{\E_{unif}\left[ #1 \right]}
\newcommand{\Es}[1]{\E_{*}\left[ #1 \right]}
\renewcommand{\r}{\boldsymbol{r}}
\renewcommand{\log}[1]{lg\left( #1 \right)}
\newcommand{\kl}[2]{KL\left(#1 || #2 \right)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\begin{document}
\def\ci{\perp\!\!\!\perp}

\section{Lower bound on regret for K-armed bernoulli bandits}

For any horizon $T$ and number of arms $K$, there exists a strategy for choosing rewards such that the expected regret for any algorithm is $\bigomega{\sqrt{TK}}$. This strategy is oblivious to the algorithm and assigns rewards at random according to some distribution. Thus the regret bound applies to stochastic and adversarial bandits.

\begin{theorem}
There exits a distribution over rewards such that: 
\eqn {
R(T) \geq \frac{1}{20}\min{\{\sqrt{KT},T\}}
}
\end{theorem}

\begin{proof}

Consider a set of environments indexed by $i$. In environment $i$ action $i$ is the 'good' action. All other arms are slightly worse. 
\begin{itemize}
\item  In environment $i$, $r_{i,t} \sim Bernoulli(\frac{1}{2}+\epsilon)$ and $r_{j,t} \sim Bernoulli(\frac{1}{2}) \;\; \forall j \neq i$ 
\item $\Pi{.}$ is the probability with respect to environment $i$
\item $\Pu{.}$ is the probability with respect to an environment in which the expected reward for \textbf{all} arms is $\frac{1}{2}$
\item $\Ps{.}$ is the probability with respect to an environment sampled uniformly at random from $\set{1...K}$
\item $r_t$ is the reward received at time $t$
\item $\boldsymbol{r}^t = <r_1,...,r_t>$ is the history of rewards received upto time $t$
\item A is the algorithm, maps $\boldsymbol{r}^{t-1} \rightarrow i_t$
\item $N_i$ is a random variable representing the number of times action $i$ is selected by the algorithm.
\item $G_A = \sum_{t=1}^T r_t$ is the total reward for the algorithm
\item $G_{max}$ is the total reward for playing the optimal arm in every timestep.

\end{itemize}

\begin{lemma}
\label{lem:N_not_too_big}
For any arm $i$,
\eqn{
\Ei{N_i} -\Eu{N_i} \leq  \frac{T}{2}\sqrt{-\ln({1-4\epsilon^2})\Eu{N_i}}
}

This says that the number of times we expect to play arm $i$ in the environment in which it is optimal is not too much greater than the number of times we expect to play it if all arms are equal.

\begin{proof}
\eqn{
\Ei{N_i} -\Eu{N_i} = & \sum_{\r \in \set{0,1}^T}N_i(\r)\left(\Pi{\r} - \Pu{\r} \right) \qquad \leftarrow \text{definition of expectation}\\
\leq & \sum_{\r:\Pi{\r} \geq \Pu{\r}}N_i(\r)\left(\Pi{\r} - \Pu{\r} \right) \qquad \leftarrow \text{dropped only -ive terms}\\
\leq & T\sum_{\r:\Pi{\r} \geq \Pu{\r}}\left(\Pi{\r} - \Pu{\r} \right) \qquad \leftarrow N_i \leq T \;\; \forall \r\\
 = & \frac{T}{2}|| \Pi{\r} - \Pu{\r}||_1 \qquad \leftarrow \text{see Thomas\&Cover 11.137}\\
 \leq & \frac{T}{2}\sqrt{2\ln({2})\kl{\Pu{\r}}{\Pi{\r}}} \qquad \leftarrow \text{see Thomas\&Cover 11.138}\\
  = & \frac{T}{2}\sqrt{-ln(2)\log{1-4\epsilon^2} \Eu{N_i}} \qquad \leftarrow \text{see section \ref{sec:calculation_of_divergence}}\\
    = & \frac{T}{2}\sqrt{-\ln({1-4\epsilon^2}) \Eu{N_i}} \qquad \leftarrow \text{change of base}
}

\end{proof}
\end{lemma}

\begin{theorem}
For any algorithm $A$, if the distribution over rewards is selected unifirmly at random from environments $\set{1...K}$:

\eqn {
E_*[G_{max} - G_A] \geq \epsilon\left(T - \frac{T}{K} - \frac{T}{2}\sqrt{-\frac{T}{K}ln(1-4\epsilon^2)}  \right) 
}

\begin{proof}

\eqn{
\Ei{r_t} &= \left(\frac{1}{2}+\epsilon \right)\Pi{i_t = i} + \frac{1}{2}\left(1-\Pi{i_t = i} \right) \\
& = \frac{1}{2} + \epsilon \Pi{i_t = i}\\
\implies \Ei{G_A} &= \sum_{t=1}^T \Ei{r_t} = \frac{T}{2}+\epsilon \Ei{N_i}
}

This gives us the expected gain given action $i$ is the good action in terms of the number of times the algorithm $A$ selects action $i$. The expected gain over all the environments $i$ is:

\eqn{
\Es{G_A} &= \frac{1}{K}\sum_{i=1}^K \Ei{G_A} \\
&= \frac{T}{2} + \frac{\epsilon}{K}\sum_{i=1}^K \Ei{N_i}\\
\Es{G_{max}} &= \left(\frac{1}{2}+\epsilon \right)T
}

From lemma \ref{lem:N_not_too_big}

\eqn{
\sum_{i=1}^K\Ei{N_i} & \leq \sum_{i=1}^K \left(\Eu{N_i}+ \frac{T}{2}\sqrt{-\ln({1-4\epsilon^2})\Eu{N_i}}\right)
}

Now $\sum_{i=1}^K\Eu{N_i} = T$ (because $\sum_{i=1}^K N_i = T$ ? but doesn't that imply $\sum_{i=1}^K\Ei{N_i} = T$?)

\eqn{
&\implies \sum_{i=1}^K\Ei{N_i} \leq T+ \frac{T}{2} \sqrt{-\ln(1-4\epsilon^2)KT} \qquad \leftarrow \text{via Jenson's Inequality} \\
&\implies \Es{G_A} \leq \frac{T}{2} + \epsilon\left(\frac{T}{K}+ \frac{T}{2} \sqrt{-\frac{T}{K}\ln(1-4\epsilon^2)} \right) \\ 
&\implies \Es{G_{max} - G_A} \geq \epsilon\left(T - \frac{T}{K} - \frac{T}{2}\sqrt{-\frac{T}{K}ln(1-4\epsilon^2)}  \right) 
}

\end{proof}

\end{theorem}

\end{proof}

\subsection{Calculation of KL divergence}
\label{sec:calculation_of_divergence}
\eqn{
& \kl{p(x_1,...,x_n)}{q(x_1,...,x_n)} =  \sum_{i=1}^N \kl{p(x_i|\boldsymbol{x}^{i-1})}{q(x_i|\boldsymbol{x}^{i-1})} \qquad \leftarrow \text{chain rule for KL divergence} \\
& \text{where } \kl{p(x_i|\boldsymbol{x}^{i-1})}{q(x_i|\boldsymbol{x}^{i-1})} = 
\sum_{\boldsymbol{x}^{i-1}}p(\boldsymbol{x}^{i-1})\sum_{x_i}p(x_i|\boldsymbol{x}^{i-1})\log{\frac{p(x_i|\boldsymbol{x}^{i-1})}{q(x_i|\boldsymbol{x}^{i-1})}}
}

So 
\eqn {
\kl{P_{unif}}{P_i} = \sum_{t=1}^T\left(
\underbrace{\sum_{\r^{t-1}}\Pu{\r^{t-1}}}_{\text{expectation over history}}
\underbrace{\sum_{r_t \in \set{0,1}}\Pu{r_t|\r^{t-1}}\log{\frac{\Pu{r_t|\r^{t-1}}}{\Pi{r_t|\r^{t-1}}}}}_{\text{KL divergence at time t}}
\right)
}

Now 
\eqn{
\Pu{r_t|\r^{t-1}}& = \frac{1}{2} \;\; \forall \;{r_t,\r^{t-1}} \\
\Pi{r_t|\r^{t-1}}&=
\begin{cases}
(\frac{1}{2}+\epsilon)^{r_t}+(\frac{1}{2}-\epsilon)^{1-r_t} & \text{ if } A(\r^{t-1}) = i\\
\frac{1}{2} & \text{ otherwise}\\
\end{cases}
}

Let $B$ be the set of histories that lead the algorithm to select the good arm, $B =\set{ \r^{t-1}:A(\r^{t-1})=i}$

\eqn {
\kl{P_{unif}}{P_i} &= \sum_{t=1}^T\left(
\sum_{B}\Pu{\r^{t-1}}
\kl{\frac{1}{2}}{\frac{1}{2}+\epsilon}
+ \sum_{B^c}\Pu{\r^{t-1}}
\kl{\frac{1}{2}}{\frac{1}{2}}
\right)\\
&= \kl{\frac{1}{2}}{\frac{1}{2}+\epsilon}\sum_{t=1}^T\left(
\sum_{B}\Pu{\r^{t-1}}
\right)\\
&=\kl{\frac{1}{2}}{\frac{1}{2}+\epsilon} \sum_{t=1}^T\left(
\Pu{i_t=i}
\right)\\
&=\kl{\frac{1}{2}}{\frac{1}{2}+\epsilon}\Eu{N_i}\\
&=-\frac{1}{2}\log{1-4\epsilon^2}\Eu{N_i}
}
\pagebreak
\subsection{Jenson's inequality}
Jenson's inequality states that for a concave function $\phi$:

\eqn{
\frac{\sum_{i=1}^K \phi(x_i)}{K} \leq \phi \left(\frac{\sum_{i=1}^K x_i}{K} \right) \\
\implies  \sum_{i=1}^K \sqrt{\Eu{N_i}} \leq \sqrt{KT} \\
\implies \sum_{i=1}^K\Ei{N_i} \leq T+ \frac{T}{2} \sqrt{-\ln(1-4\epsilon^2)KT}
}

\end{document}

### 2015-11-19 causal-bandit/lowerbound_auer.tex
adds KL divergence for specific feedback graph

@@
 \newtheorem{example}[theorem]{Example}
+ \tikzset{
+    %Define standard arrow tip
+    >=stealth',
+    %Define style for boxes
+    observed/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered,
+           },
+     latent/.style={
+           circle,
+           rounded corners,
+           draw=black, thick, dashed,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!10!white
+           },
+    target/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!20!white,
+           },
+    observedrect/.style={
+           rectangle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=6em,
+           minimum height=2em,
+           font=\footnotesize,
+           text centered,
+           },
+     targetrect/.style={
+           rectangle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=6em,
+           minimum height=2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!20!white,
+           },
+     empty/.style={
+           circle,
+           rounded corners,
+           minimum width=.5em,
+           minimum height=.5em,
+           font=\footnotesize,
+           text centered,
+           },
+    % Define arrow style
+    pil/.style={
+           o->,
+           thick,
+           shorten <=2pt,
+           shorten >=2pt,},
+    sh/.style={ shade, shading=axis, left color=red, right color=green,
+    shading angle=45 }   
+}
 
@@
 
-Now $\sum_{i=1}^K\Eu{N_i} = T$ (because $\sum_{i=1}^K N_i = T$ ? but doesn't that imply $\sum_{i=1}^K\Ei{N_i} = T$?)
+Now $\sum_{i=1}^K\Eu{N_i} = \Eu{\sum_{i=1}^K N_i} = \Eu{T} = T$
+
+Note: we cannot do the same with $\sum_{i=1}^K\Ei{N_i}$ as $\Ei{.}$ is with respect to a different distribution for each $i$.
 
@@
 
-Let $B$ be the set of histories that lead the algorithm to select the good arm, $B =\set{ \r^{t-1}:A(\r^{t-1})=i}$
+Let $B$ be the set of histories that lead the algorithm to select the good arm, $B =\set{ \r^{t-1}:A(\r^{t-1})=i}$.
+Note: $\r^{t-1}$ is sufficient to determine $i_t$ for a deterministic algorithm $A$ in the bandit setting.
+At the first timestep $A$ will select $i_1(A)$, it then receives reward $r_1$, selects $i_2(A,r_1)$ and so on.
 
@@
 \sum_{B}\Pu{\r^{t-1}}
-\right)\\
+\right) \qquad \leftarrow \text{ as $kl(\frac{1}{2}||\frac{1}{2}) = 0$}\\
 &=\kl{\frac{1}{2}}{\frac{1}{2}+\epsilon} \sum_{t=1}^T\left(
@@
 \frac{\sum_{i=1}^K \phi(x_i)}{K} \leq \phi \left(\frac{\sum_{i=1}^K x_i}{K} \right) \\
+\implies \sum_{i=1}^K \sqrt{x_i} \leq \sqrt{K\sum_{i=1}^K x_i}\\
 \implies  \sum_{i=1}^K \sqrt{\Eu{N_i}} \leq \sqrt{KT} \\
@@
 
+\section{Lower bound for specific feedback graph}
+
+Now we move to considering a Bernoulli bandit problem with the feedback graph shown in figure \ref{fig:feedbackgraph}.
+The environments specifying rewards over arms $\set{1...K}$ are defined identically to the standard bandit case.
+The expected reward for $a_0$ (in all environments) is $0$ (this is the worst case).
+
+\begin{figure}[h]
+\centering
+\caption{Feedback graph}
+\label{fig:feedbackgraph}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+
+ %nodes
+\node[main node](1){$a_{1}$};
+\node[main node, right=of 1](2){$a_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$a_{K}$};
+\node[main node, above right=of 2](5){$a_0$};
+ \path[every node/.style={font=\sffamily\small}]
+    (5) edge (1)
+    	(5) edge (2)
+    (5) edge (4);
+
+\end{tikzpicture}
+\end{figure}
+
+As for the standard bandit case,
+
+\eqn{
+\Ei{N_i} -\Eu{N_i} \leq & \frac{T}{2}\sqrt{2\ln({2})\kl{\Pu{\boldsymbol{h}}}{\Pi{\boldsymbol{h}}}} 
+}
+
+The only difference is that the sequence of rewards $\r$ is no longer sufficient to determine $N_i$ because when we select $a_0$ we get feedback on all the other arms that does not contribute to the reward but can be leveraged by an algorithm.
+Therefore the KL divergence is now over distributions over the history $\boldsymbol{h}$ that includes this additional information.
+
+\subsection{Calculation of KL divergence}
+
+Define
+
+\eqn{
+h_t = 
+\begin{cases}
+(r_t,o_{1t},...,o_{Kt}) & \text{ if } i = 0 \\
+r_t & \text{ otherwise}\\
+\end{cases}
+}
+
+Where $o_{jt}$ is the observed reward for arm $j$ at timestep $t$.
+
+\eqn {
+\kl{\Pu{\boldsymbol{h}} }{\Pi{\boldsymbol{h}}} = 
+\sum_{t=1}^T \kl{\Pu{h_t|\boldsymbol h^{t-1}}}{\Pi{h_t|\boldsymbol h^{t-1}}}
+}
+
+At each time step $t$, $i_t = A(\boldsymbol h^{t-1})$ There are three cases for the value of the KL divergence.
+
+\begin{enumerate}
+\item The algorithm selects the optimal action. $i_t = i$ , (as for the standard bandit case)
+\eqn{
+\Pu{h_t|\boldsymbol h^{t-1}} &= \Pu{r_t|\boldsymbol h^{t-1}} = \frac{1}{2} \\
+\Pi{h_t|\boldsymbol h^{t-1}} &= \Pi{r_t|\boldsymbol h^{t-1}} = 
+(\frac{1}{2}+\epsilon)^{r_t}+(\frac{1}{2}-\epsilon)^{1-r_t} \\
+\implies kl(\frac{1}{2} || \frac{1}{2} + \epsilon)
+}
+
+\item The algorithm selects the revealing action. $i_t = 0$
+
+\eqn {
+\Pu{h_t|\boldsymbol h^{t-1}} &= B_0(r_t)\prod_{j \in \set{1...K}} B_{\frac{1}{2}}(o_j) \qquad \text{where} B_\alpha(x) \implies x \sim Bernoulli(\alpha) \\
+\Pi{h_t|\boldsymbol h^{t-1}} &= B_0(r_t) \prod_{j \in \set{1...K}/i} B_{\frac{1}{2}}(o_j)B_{\frac{1}{2}+\epsilon}(o_i) \\
+\implies kl(\frac{1}{2} || \frac{1}{2} + \epsilon)
+}
+
+Since the only term that differs $P_{unif}$ and $P_i$ is that for $o_i$ and,
+
+\eqn{
+\kl{\prod_{i=1}^nP_i(x_i)}{\prod_{i=1}^n Q_i(x_i)} = \sum_{i=1}^n \kl{P_i(x_i)}{Q_i(x_i)}
+}
+
+\item Otherwise. $i_t \in \set{1,..,K}/i$ , (as for the standard bandit case)
+\eqn {
+\Pu{h_t|\boldsymbol h^{t-1}} = \Pi{h_t|\boldsymbol h^{t-1}} = \frac{1}{2}\\
+\implies kl(\frac{1}{2} ||\frac{1}{2}) = 0
+}
+\end{enumerate}
+
+We divide the histories into corresponding sets.
+$S_1 = \set{\boldsymbol h^{t-1}:A(\boldsymbol h^{t-1})=i}$, $S_2 = \set{\boldsymbol h^{t-1}:A(\boldsymbol h^{t-1})=0}$,$S_3 = \set{\boldsymbol h^{t-1}:A(\boldsymbol h^{t-1})\in \set{1,..,K}/i}$
+
+\eqn {
+\kl{P_{unif}}{P_i} =& \sum_{t=1}^T\left(\sum_{S_1}\Pu{\boldsymbol h^{t-1}}kl(\frac{1}{2} || \frac{1}{2} + \epsilon) + 
+\sum_{S_2}\Pu{\boldsymbol h^{t-1}}kl(\frac{1}{2} || \frac{1}{2} + \epsilon) +
+\sum_{S_3}\Pu{\boldsymbol h^{t-1}}kl(\frac{1}{2} ||\frac{1}{2})\right) \\
+=& kl(\frac{1}{2} || \frac{1}{2} + \epsilon)\sum_{t=1}^T\left(\sum_{S_1}\Pu{\boldsymbol h^{t-1}}+\sum_{S_2}\Pu{\boldsymbol h^{t-1}}  \right)\\
+=& kl(\frac{1}{2} || \frac{1}{2} + \epsilon)\sum_{t=1}^T\left(\Pu{i_t = i}+\Pu{i_t=0}  \right)\\
+& kl(\frac{1}{2} || \frac{1}{2} + \epsilon)\left(\Eu{N_i}+\Eu{N_0}  \right)
+}
+
+\subsection{Bounding regret}
+
+\eqn {
+\Ei{r_t} =& (\frac{1}{2}+\epsilon)\Pi{i_t=i)} + 0\Pi{i_t = 0} + \frac{1}{2}\Pi{i_t \in \set{1...K}/i}\\
+\Ei{G_A} =& \sum_{t=1}^T\Ei{r_t} = \frac{T}{2} + \epsilon \Ei{N_i} - \frac{T}{2}\Ei{N_0}\\
+\Es{G_A}  =& \frac{1}{K}\sum_{i = 1}^K\Ei{G_A} = \frac{T}{2} + \frac{\epsilon}{K}\sum_{i=1}^K\Ei{N_i} - \frac{T}{2K}\sum_{i=1}^K\Ei{N_0}\\
+\Es{G_{max}} =& (\frac{1}{2}+\epsilon)T \\
+\implies R(T) =& \epsilon \left(T -\frac{1}{K}\sum_{i=1}^K\Ei{N_i}\right) + \frac{T}{2K}\sum_{i=1}^K\Ei{N_0}\\
+\geq & \epsilon \left(T -\frac{T}{K} - \frac{T}{2}\sqrt{-\frac{1}{K}ln(1-4\epsilon^2) (T + \sum_{i=1}^K \Eu{N_0})} \right) + \frac{T}{2K}\sum_{i=1}^K\Ei{N_0}  
+}
+
 \end{document}

### 2015-12-04 causal-bandit/icml-2015/causal-bandits.tex
Start ICML draft

 %\documentclass[wcp,gray]{jmlr} % test grayscale version
\documentclass[wcp]{jmlr}

 % The following packages will be automatically loaded:
 % amsmath, amssymb, natbib, graphicx, url, algorithm2e

 %\usepackage{rotating}% for sideways figures and tables
\usepackage{longtable}% for long tables
\usepackage{booktabs}
% \usepackage[load-configurations=version-1]{siunitx} % newer version
 %\usepackage{siunitx}

% change the arguments, as appropriate, in the following:
\jmlrvolume{NNN}
\jmlryear{2015}
\jmlrworkshop{ICML}

\title[Causal Bandits]{Causal Bandits}

 % Use \Name{Author Name} to specify the name.
 % If the surname contains spaces, enclose the surname
 % in braces, e.g. \Name{John {Smith Jones}} similarly
 % if the name has a "von" part, e.g \Name{Jane {de Winter}}.
 % If the first letter in the forenames is a diacritic
 % enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

 % Authors with different addresses:
 % \author{\Name{Author Name1} \Email{abc@sample.com}\\
 % \addr Address 1
 % \AND
 % \Name{Author Name2} \Email{xyz@sample.com}\\
 % \addr Address 2
 %}
\author{\Name{Author's Name}}

\editor{Editor's name}
 % \editors{List of editors' names}

\begin{document}

\maketitle

\begin{abstract}
	The abstract
\end{abstract}
\begin{keywords}
	Causality, Bandits, Regret Bounds
\end{keywords}

\section{Introduction}
\label{sec:intro}

The aim of this paper is to demonstrate theoretically the utility of having 
causal knowledge when solving multi-armed bandit (MAB) problems.

We do this by:
\begin{enumerate}
	\item Providing some motivating examples as to why it is reasonable to expect
		to have some knowledge of causal structures when solving a sequential
		decision making problem such as MAB problems.
	\item Formalising what we mean by ``causal information'' for a simple
		class of stochastic multi-armed bandit (MAB) problems. 
	\item Proposing a simple explore-then-exploit algorithm for this class of
		problems that makes use of the causal information and proving it
		has a sublinear regret bound.
	\item Deriving a matching lower bound for the same class of problems for 
		algorithms that make use of the causal structure.
	\item Showing that the upper bound for the ``causally informed'' algorithm
		is better than the corresponding lower bound for algorithms on this 
		class of problems that do not use the extra causal information.
\end{enumerate}

We identify a problem-dependent constant that appears in the upper and lower 
bounds that can be roughly interpreted as a measure of how much actions reveal
about other actions via the causal structure.

\subsection{Related Work}
\label{sub:related_work}

\begin{itemize}
	\item Bareinboim, Forney \& Pearl, \emph{Bandits with unobserved 
			confounders}, NIPS 2015
	\item Salomon, Audibert \& Alaoui, \emph{Lower Bounds and Selectivity of 
			Weak-Consistent Policies in Stochastic Multi-Armed Bandit Problems},
			JMLR 2013.
	\item Alon, Cesa-Bianchi, Dekel, Koren, \emph{Online Learning with Feedback
			Graphs: Beyond Bandits}, COLT 2015.
\end{itemize}

% \acks{Acknowledgements go here.}

\bibliography{../c-bandit.bib}

\end{document}

### 2015-12-04 causal-bandit/lowerbound_auer.tex

@@
 \begin{cases}
-(\frac{1}{2}+\epsilon)^{r_t}+(\frac{1}{2}-\epsilon)^{1-r_t} & \text{ if } A(\r^{t-1}) = i\\
+(\frac{1}{2}+\epsilon)^{r_t}(\frac{1}{2}-\epsilon)^{1-r_t} & \text{ if } A(\r^{t-1}) = i\\
 \frac{1}{2} & \text{ otherwise}\\
@@
 \Pi{h_t|\boldsymbol h^{t-1}} &= \Pi{r_t|\boldsymbol h^{t-1}} = 
-(\frac{1}{2}+\epsilon)^{r_t}+(\frac{1}{2}-\epsilon)^{1-r_t} \\
+(\frac{1}{2}+\epsilon)^{r_t}(\frac{1}{2}-\epsilon)^{1-r_t} \\
 \implies kl(\frac{1}{2} || \frac{1}{2} + \epsilon)
@@
 =& kl(\frac{1}{2} || \frac{1}{2} + \epsilon)\sum_{t=1}^T\left(\Pu{i_t = i}+\Pu{i_t=0}  \right)\\
-& kl(\frac{1}{2} || \frac{1}{2} + \epsilon)\left(\Eu{N_i}+\Eu{N_0}  \right)
+=& kl(\frac{1}{2} || \frac{1}{2} + \epsilon)\left(\Eu{N_i}+\Eu{N_0}  \right)\\
+=& -\frac{1}{2}lg(1-4\epsilon^2)\left(\Eu{N_i}+\Eu{N_0}  \right)
+}
+
+Similarly (I think)
+
+\eqn {
+\kl{P_{i}}{P_{unif}} 
+=& kl(\frac{1}{2}+\epsilon || \frac{1}{2})\left(\Ei{N_i}+\Ei{N_0}  \right)\\
+=& (\frac{1}{2}lg((1+2\epsilon)(1-2\epsilon))+ \epsilon lg\left(\frac{1+2\epsilon}{1-2\epsilon}\right))\left(\Ei{N_i}+\Ei{N_0}  \right)\\
+\leq & -\frac{1}{2}lg(1-4\epsilon^2)\left(\Ei{N_i}+\Ei{N_0}  \right)
 }
@@
 \subsection{Bounding regret}
+We will consider only algorithms that select $N_0$ at most $\epsilon T$ times.
+Once an algorithm has selected $N_0$ $\epsilon T$ times, we will truncate it and play any other action at random.
+The regret for the non-truncated algorithm is at least $\epsilon T(\frac{1}{2}+\epsilon) \geq \frac{\epsilon T}{2}$.
+Truncating adds at most $(T - \epsilon T)\epsilon \leq \epsilon T$.
+Thus, the assumption that $N_0 \leq \epsilon T$ increases the regret by at most a constant factor of $3$.
 
 \eqn {
-\Ei{r_t} =& (\frac{1}{2}+\epsilon)\Pi{i_t=i)} + 0\Pi{i_t = 0} + \frac{1}{2}\Pi{i_t \in \set{1...K}/i}\\
-\Ei{G_A} =& \sum_{t=1}^T\Ei{r_t} = \frac{T}{2} + \epsilon \Ei{N_i} - \frac{T}{2}\Ei{N_0}\\
-\Es{G_A}  =& \frac{1}{K}\sum_{i = 1}^K\Ei{G_A} = \frac{T}{2} + \frac{\epsilon}{K}\sum_{i=1}^K\Ei{N_i} - \frac{T}{2K}\sum_{i=1}^K\Ei{N_0}\\
-\Es{G_{max}} =& (\frac{1}{2}+\epsilon)T \\
-\implies R(T) =& \epsilon \left(T -\frac{1}{K}\sum_{i=1}^K\Ei{N_i}\right) + \frac{T}{2K}\sum_{i=1}^K\Ei{N_0}\\
-\geq & \epsilon \left(T -\frac{T}{K} - \frac{T}{2}\sqrt{-\frac{1}{K}ln(1-4\epsilon^2) (T + \sum_{i=1}^K \Eu{N_0})} \right) + \frac{T}{2K}\sum_{i=1}^K\Ei{N_0}  
+\Ei{r_t} =& (\frac{1}{2}+\epsilon)\Pi{i_t=i} + 0\Pi{i_t = 0} + \frac{1}{2}\Pi{i_t \in \set{1...K}/i}\\
+=& (\frac{1}{2}+\epsilon)\Pi{i_t=i} + \frac{1}{2}(1-\Pi{i_t=i} - \Pi{i_t=0})\\
+=& \frac{1}{2} + \epsilon \Pi{i_t=i} - \frac{1}{2}\Pi{i_t=0}\\
+\Ei{G_A} =& \sum_{t=1}^T\Ei{r_t} = \frac{T}{2} + \epsilon \Ei{N_i} - \frac{1}{2}\Ei{N_0}\\
+\Es{G_A}  =& \frac{1}{K}\sum_{i = 1}^K\Ei{G_A} = \frac{T}{2} + \frac{\epsilon}{K}\sum_{i=1}^K\Ei{N_i} - \frac{1}{2K}\sum_{i=1}^K\Ei{N_0}\\
+\Es{G_{max}} =& (\frac{1}{2}+\epsilon)T}
+
+\eqn {
+\implies R(T) =& \epsilon \left(T -\frac{1}{K}\sum_{i=1}^K\Ei{N_i}\right) + \frac{1}{2K}\sum_{i=1}^K\Ei{N_0}\\
+\geq & \epsilon \left(T -\frac{T}{K} - \frac{T}{2}\sqrt{-\frac{1}{K}ln(1-4\epsilon^2) (T + \sum_{i=1}^K \Eu{N_0})} \right) + \frac{1}{2K}\sum_{i=1}^K\Ei{N_0}  \\
+\label{eqn:graph_regret1}
+=& \epsilon \left(T -\frac{T}{K} - \frac{T}{2}\sqrt{-ln(1-4\epsilon^2) (\frac{T}{K} +  \Eu{N_0})} \right) + \frac{1}{2}\Es{N_0}\\
+\geq & \epsilon \left(T -\frac{T}{K} - T\epsilon \sqrt{2(\frac{T}{K} +  \Eu{N_0})} \right) + \frac{1}{2}\Es{N_0} \qquad \text{ if } \epsilon < \frac{2}{5} \\
+\geq & \frac{\epsilon}{3} \left(T -\frac{T}{K} - T\epsilon \sqrt{2T(\frac{1}{K} +  \epsilon)} \right) \qquad \text{ assuming } \Eu{N_0} \leq \epsilon T 
+}
+
+Let
+
+\eqn {
+\epsilon = \begin{cases}
+\frac{1}{4}\sqrt{\frac{K}{T}} & \text{ if } K < (16T)^{1/3}\\
+(16T)^{-1/3} & \text { otherwise }
+\end{cases}
+}
+
+If $K < (16T)^{1/3}$
+
+\eqn {
+R(T) \geq & \frac{1}{3}\left(\frac{1}{4}\sqrt{KT} - \frac{1}{4}\sqrt{\frac{T}{K}} - \frac{1}{16\sqrt{2}}\sqrt{4KT + K^{5/2}T^{1/2}}\right)\\
+\geq & \frac{1}{3}\left(\frac{1}{4}\sqrt{KT} - \frac{1}{4}\sqrt{\frac{T}{K}} - \frac{1}{16\sqrt{2}}\sqrt{8KT}\right)\\
+=& \frac{1}{3}\left(\frac{1}{8}\sqrt{KT} - \frac{1}{4}\sqrt{\frac{T}{K}}\right) \\
+\geq & \frac{1}{72}\sqrt{KT} \qquad \text{ if } K > 2
+}
+
+If  $K \geq (16T)^{1/3}$
+\eqn {
+R(T) \geq & \frac{T^{2/3}}{3(16^{1/3})}\left(1  - \frac{1}{K} - \frac{1}{2(2^{5/6})}\sqrt{\frac{4T^{1/3}}{K}+2^{2/3}}\right)\\
+ \geq & \frac{T^{2/3}}{3(16^{1/3})}\left(1  - \frac{1}{K} - \frac{1}{2(2^{5/6})}\sqrt{2(2^{2/3})}\right)\\
+ = & \frac{T^{2/3}}{3(16^{1/3})}\left(1  - \frac{1}{K} - \frac{1}{2}\right)\\
+ \geq & \frac{1}{46}T^{2/3} \qquad \text {if } K > 2\\
+}
+
+\pagebreak
+
+\section{Graph changes with time}
+
+Same as before we have a single revealing action. However, now at each timestep it reveals the rewards for a half the arms at random according to the prior probability vector $\boldsymbol{q}$
+
+\begin{itemize}
+\item Assume that $q_j \in [0,\frac{1}{2}]$ and $q_1 \leq q_2, \leq ...\leq,q_N$ 
+\item Let $m \in \set{2...N} = min_i : q_j \geq \frac{1}{j}$
+\item Let $M = \set{j:j \leq m}$, (the set of unbalanced arms).
+\item let $D_t$ be the set of arms revealed by the revealing action at timestep $t$
+\end{itemize}
+
+\subsection{KL divergence}
+As for previous section except now if the revealing action is selected, $\Pu{h_t|\boldsymbol h^{t-1}} = \Pi{h_t|\boldsymbol h^{t-1}}$ unless $i \in D_t$
+
+The probability $P(i \in D_t)$ depends only on $\boldsymbol{q}$ (and is independent of the history at anytime step $\boldsymbol{h}$).
+
+So:
+
+\eqn {
+\kl{P_{unif}}{P_i} & = kl(\frac{1}{2}||\frac{1}{2}+\epsilon)\left(\Eu{N_i}+\sum_{t=1}^T\Pu{i_t=0}P(i \in D_t)\right)\\
+&\geq \begin{cases}
+kl(\frac{1}{2}||\frac{1}{2}+\epsilon)\left(\Eu{N_i}+\frac{1}{m}\Eu{N_0}\right) & \text{ if } i \notin M \\
+kl(\frac{1}{2}||\frac{1}{2}+\epsilon)\Eu{N_i} & \text{ otherwise }
+\end{cases}
+}
+
+\subsection{Bounding regret}
+Assuming $ \epsilon < \frac{2}{5}$,
+\eqn {
+R(T) &= \epsilon  \left(T -\frac{1}{K}\sum_{i=1}^K\Ei{N_i}\right) + \frac{1}{2K}\sum_{i=1}^K\Ei{N_0} \qquad \leftarrow \text{as before}\\
+&\leq \epsilon \left(T - \frac{T}{K} - T\epsilon\sqrt{\frac{2}{K}\left(\sum_{i\in M}\Eu{N_i} + \sum_{i \notin M}(\Eu{N_i}+\frac{1}{m}\Eu{N_0})  \right)}   \right)+\frac{1}{2}\Es{N_0}\\
+&= \epsilon \left(T - \frac{T}{K} - T\epsilon\sqrt{2\left(\frac{T}{K} + \frac{1}{m}\Eu{N_0}\right)}   \right)+\frac{1}{2}\Es{N_0}\\
+&\geq \frac{\epsilon}{3} \left(T - \frac{T}{K} - T\epsilon\sqrt{2T\left(\frac{1}{K} + \frac{\epsilon}{m}\right)}   \right) \qquad \text{assuming } \Eu{N_0} \leq \epsilon T
+}
+
+Let:
+
+\eqn{
+\epsilon = \begin{cases}
+\frac{1}{4}\sqrt{\frac{K}{T}} & \text{if } K > m^{2/3}(16T)^{1/3} \\
+m^{1/3}(16T)^{-1/3} & \text{otherwise}
+\end{cases}
+}
+If $K < m^{2/3}(16T)^{1/3}$
+\eqn {
+R(T) \geq \frac{\sqrt{KT}}{12}\left(\frac{1}{2} - \frac{1}{K}\right)\\
+\geq \frac{1}{72}\sqrt{KT} \qquad \text{if } K > 2
+}
+
+If  $K \geq m^{2/3}(16T)^{1/3}$
+\eqn{
+R(T)& \geq \frac{m^{1/3}T^{2/3}}{3(16^{1/3})}\left(\frac{1}{2} - \frac{1}{K}\right) \\
+& \geq \frac{1}{46}m^{1/3}T^{2/3} \qquad \text{ if } K > 2
+}
+
+\pagebreak
+\subsection{Previous Attempt at bounding regret}
+
+If I just let $\Eu{N_0} = T$ and $\Es{N_0} = 0$, I get $R \sim \bigomega{\sqrt{T}}$ - so not tight enough.
+
+By similar appeal to KL divergence, we get
+
+\eqn {
+\Ei{N_0} & \geq \Eu{N_0} - \frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\Eu{N_i}+\Eu{N_0})} \\
+\implies \frac{1}{K}\sum_{i=1}^K\Ei{N_0} & \geq \frac{1}{K}\sum_{i=1}^K\Eu{N_0} - \frac{T}{2K}\sum_{i=1}^K\sqrt{-ln(1-4\epsilon^2)(\Eu{N_i}+\Eu{N_0})}\\
+\implies \Es{N_0} & \geq \Eu{N_0} - \frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\frac{1}{K}\sum_{i=1}^K\Eu{N_i}+\frac{1}{K}\sum_{i=1}^K\Eu{N_0})}\\
+ & = \Eu{N_0} - \frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\frac{T}{K}+\Eu{N_0})}
+}
+
+\eqn{
+\label{eqn:graph_regret2}
+\implies R(T) \geq & 
+\epsilon(T - \frac{T}{K})-(\epsilon + \frac{1}{2})\frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\frac{T}{K}+\Eu{N_0})} +\frac{1}{2}\Eu{N_0} 
+}
+
+For comparison, if we assumed $\Eu{N_0}=\Es{N_0}$ in equation \ref{eqn:graph_regret1} we would have:
+
+\eqn{
+R(T) \geq & 
+\epsilon(T - \frac{T}{K})-\epsilon\frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\frac{T}{K}+\Eu{N_0})} +\frac{1}{2}\Eu{N_0} 
+}
+
+So the regret bound is pulled down by an extra  $\frac{T}{4}\sqrt{-ln(1-4\epsilon^2)(\frac{T}{K}+\Eu{N_0})}$
+
+Argument that we cannot get a regret bound starting from equation \ref{eqn:graph_regret2}
+
+Let: 
+\eqn {
+R_1 &= \epsilon(T - \frac{T}{K})-(\epsilon+\frac{1}{2})\frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\frac{T}{K}+\Eu{N_0})} +\frac{1}{2}\Eu{N_0} 
+}
+
+such that  $R(T) \geq R_1$
+
+\eqn {
+R_1 & \leq \epsilon T-(\epsilon+\frac{1}{2})\frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\frac{T}{K}+\Eu{N_0})} +\frac{1}{2}\Eu{N_0} \\
+& \leq \epsilon T-\frac{1}{2}\frac{T}{2}\sqrt{-ln(1-4\epsilon^2)(\frac{T}{K}+\Eu{N_0})} +\frac{1}{2}\Eu{N_0} \\
+& \leq \epsilon T-\frac{T}{4}\sqrt{4\epsilon^2(\frac{T}{K}+\Eu{N_0})} +\frac{1}{2}\Eu{N_0} \\
+& \leq \epsilon T-\epsilon\frac{T}{2}\sqrt{\Eu{N_0}} +\frac{1}{2}\Eu{N_0} \\
+& \leq \frac{1}{2}\left(T-\frac{T}{2}\sqrt{\Eu{N_0}}\right) +\frac{1}{2}\Eu{N_0} \\
+&= \frac{1}{2}\left(2T-\frac{T^{3/2}}{2}\right) \qquad \text{if $\Eu{N_0} = T$}\\ 
+& \leq 0 \qquad \text{if $T > 16$}
 }

### 2015-12-04 causal-bandit/icml-2015/causal-bandits.tex
adds jmlr.css, adds very quick notes on where current results fit with key points

@@
 \usepackage{booktabs}
+\usepackage{tikz}
+\usetikzlibrary{arrows,positioning} 
 % \usepackage[load-configurations=version-1]{siunitx} % newer version
  %\usepackage{siunitx}
+
+ \tikzset{
+    %Define standard arrow tip
+    >=stealth',
+    %Define style for boxes
+    observed/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered,
+           },
+     latent/.style={
+           circle,
+           rounded corners,
+           draw=black, thick, dashed,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!10!white
+           },
+    target/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!20!white,
+           },
+    observedrect/.style={
+           rectangle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=6em,
+           minimum height=2em,
+           font=\footnotesize,
+           text centered,
+           },
+     targetrect/.style={
+           rectangle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=6em,
+           minimum height=2em,
+           font=\footnotesize,
+           text centered,
+           fill=black!20!white,
+           },
+     empty/.style={
+           circle,
+           rounded corners,
+           minimum width=.5em,
+           minimum height=.5em,
+           font=\footnotesize,
+           text centered,
+           },
+    % Define arrow style
+    pil/.style={
+           o->,
+           thick,
+           shorten <=2pt,
+           shorten >=2pt,},
+    sh/.style={ shade, shading=axis, left color=red, right color=green,
+    shading angle=45 }   
+}
 
@@
 \begin{enumerate}
-	\item Providing some motivating examples as to why it is reasonable to expect
-		to have some knowledge of causal structures when solving a sequential
-		decision making problem such as MAB problems.
-	\item Formalising what we mean by ``causal information'' for a simple
-		class of stochastic multi-armed bandit (MAB) problems. 
-	\item Proposing a simple explore-then-exploit algorithm for this class of
-		problems that makes use of the causal information and proving it
-		has a sublinear regret bound.
-	\item Deriving a matching lower bound for the same class of problems for 
-		algorithms that make use of the causal structure.
-	\item Showing that the upper bound for the ``causally informed'' algorithm
-		is better than the corresponding lower bound for algorithms on this 
-		class of problems that do not use the extra causal information.
+	\item Providing some motivating examples as to why it is reasonable to expect to have some knowledge of causal structures when solving a sequential decision making problem such as MAB problems.
+	\item Formalising what we mean by ``causal information'' for a simple 	class of stochastic multi-armed bandit (MAB) problems.
+
+	We can formalize a connection between causal graphs and fairly general stochastic MAB problems (section 4 of my TPR)
+
+	\item Demonstrating that knowledge of causal structure can induce interesting dependencies between the rewards of different MAB arms.
+
+	Described in section 4 of my TPR
+
+	\item Introducing a specific causal MAB problem demonstrating this structure and proposing a simple explore-then-exploit algorithm that makes use of the causal information and yields significant improvements over the corresponding lower bounds for algorithms that do not use the causal information.
+
+Given the causal structure in figure \ref{fig:causalStructure} we get an upper bound $R(T) = m^{1/3}T^{2/3}log(KT)^{1/3}$ (section 1 of observe-then-best, section 1.1 if $\boldsymbol{q}$ is not known in advance).
+Where $m$ can be interpreted as the number of arms corresponding to variable settings that occur only rarely if not explicitly set.
+The corresponding lower bound for algorithms that do not leverage causal structure is  $\sqrt{KT}$ (Auer1995).
+This demonstrates the causal algorithm will do significantly better where there are large number of arms, provided $m$ is not too large.
+\begin{figure}[H]
+\centering
+\caption{Assumed Causal Structure}
+\label{fig:causalStructure}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty},tg/.style={target}]
+
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[tg, below right=of 2](5){Y};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5);
+
+\end{tikzpicture}
+\end{figure}
+
+	\item Deriving a matching lower bound for this problem for algorithms that make use of the causal structure.
+
+	We currently have (lowerbounds-auer section 3) a lower bound of $m^{1/3}T^{2/3}$ for algorithms that use information provided by purely observing.
+Ie the structure introduced by equation 6 in my TPR.
+This bound does not apply to algorithms using information in the form provided by equation 7 in my TPR.
+(ie we have not yet shown that we are fully utilizing the causal information available).
+
+	\item Demonstrating we can also use the causal structure to gain significant improvement for the closely related best-arm identification problem.
+
+	We obtain simple regret $R_s(T) = \sqrt{m/T}$ (observe-then-best section 1.1) versus $R_s(T) = \sqrt{K/T}$ for the non-causal version of the problem (Bubeck et al 2009)
+
 \end{enumerate}

### 2015-12-24 causal-bandit/icml-2015/causal-bandits-icml.tex
begins paper outline

\documentclass{article}

% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

% For citations
\usepackage{natbib}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2015} with
% \usepackage[nohyperref]{icml2015} above.
\usepackage{hyperref}

% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{icml2015stylefiles/icml2015}

% Employ this version of the ``usepackage'' statement after the paper has
% been accepted, when creating the final version.  This will set the
% note in the first column to ``Proceedings of the...''
%\usepackage[accepted]{icml2015}

\usepackage{tikz}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{cleveref}
\usetikzlibrary{arrows,positioning} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.5em,
           minimum height=2.5em,
           font=\footnotesize,
           text centered,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[#1\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Causal Bandits}

\begin{document}

\twocolumn[
\icmltitle{Casual Bandits}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2015
% package.
\icmlauthor{Your Name}{email@yourdomain.edu}
\icmladdress{Your Fantastic Institute,
            314159 Pi St., Palo Alto, CA 94306 USA}
\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
\icmladdress{Their Fantastic Institute,
            27182 Exp St., Toronto, ON M6H 2T1 CANADA}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{causal,bandit}

\vskip 0.3in
]

\begin{abstract}

\end{abstract}

\section{Introduction}
Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour. For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.

A widely used framework for sequential decision making is the multi-armed bandit. In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary. At each timestep the agent selects an action and receives a reward sampled i.i.d from the corresponding reward distribution. The performance of bandit algorithms is described by the regret: the difference in the expected reward obtained by the algorithm and the reward that could be obtained if the optimal action was selected at every timestep.

An an alternate approach to selecting actions is causal inference. Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones. This allows an agent to predict the outcome of an action based on non-experimental data. This approach is common in social science, demography, and economics where explicit experimentation may be difficult. For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.

We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.

A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph. Each possible assignment of variables to values is an action (bandit arm). The reward could be a general function of the action selected and the final state of the graph. However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action. The number of actions grows exponentially with the number of variables in the graph, making it important to use algorithms that take account of the graph structure to reduce the search space.

Problems framed in this way take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected. If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.

If we receive feedback on additional nodes the problem can be more interesting. In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.

We consider a bandit problem where the actions and reward are represented by a specific causal graph that demonstrates this interesting structure. We develop an algorithm to leverage the information provided by this structure and demonstrate it substantially outperforms standard bandit algorithms applied to the same problem where the number of actions is large.

\section{Problem setup}

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}. We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.

As an example, consider a farmer wishing to optimize the yield of her crop. She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc. We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example). After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield. This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.

\begin{figure}[h]
\centering
\caption{Assumed Causal Structure}
\label{fig:causalStructure}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5);

\end{tikzpicture}
\end{figure}

Let $\boldsymbol{q} \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known end point $T$:

\begin{enumerate}
\item The learner chooses $I_t \in \set{0...N}$ and $J_t \in \set{0,1}$, corresponding to setting the variable $X_{I_t}=J_t$, also denoted $do(X_{I_t} = J_t)$. Selecting $I_t = 0$ represents not intervening on any variable and simply observing, also denoted $do()$.

\item The remaining variables take their values stochastically according to their natural probabilities $\boldsymbol{q}$

\eq{
X_{i,t} = \begin{cases}
\sim Bernoulli(q_i) &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}

\item The learner receives reward $Y_t \sim \Ber(r(\boldsymbol{X}_t))$ where $r:\set{0,1}^N \to [0,1]$ is unknown and arbitrary.
\end{enumerate}

The expected reward of taking action $i,j$ is $\mu_{i,j} = \E{r(X)|do(X_i = j)}$. The optimal reward and action are denoted $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E {\sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right)}.
}

The problem could be treated as a classical multi-armed bandit with $K = 2N+1$ arms, yielding a regret $\bigo{\sqrt{TN}}$. However, this doesn't leverage the side information induced by the causal structure.

\section{Causal Bandits}

Given the assumed causal structure, the probability of $Y_t$ given we intervene to set $X_{i,t} = j$ is the same as if we don't set any variables and observe $X_{i,t} = j$.

\eqn {
\label{eq:observe}
P(Y_t|do(X_{i,t} = j)) = P(Y_t|do(),X_{i,t} = j)
}

This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph. It is not the case in general. For example if there was a variable $X'$ that caused both $X_i$ and $Y$ (or $X_i$ and any other variable $X_l$), that would introduce a backdoor path from $X_i \rightarrow Y$ and we would have to condition on $X'$ to derive the interventional distribution of $Y$ from the observational one.

\color{red} Probably should define causal model and back-door rule to properly show this.\color{black}

We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.

\eqn {
\label{eq:estimation_transfer}
P(Y_t|do(X_{i,t} = j))=\\
 \sum_{j'} P(Y_t|do(X_{l,t} = j'),X_{i,t} = j)P(X_{l_t} = j')
}

We propose a simple explore-exploit based algorithm that leverages  \eqref{eq:observe}. Without loss of generality, we assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ... \leq q_N$.

\begin{algorithm}[h]
   \caption{Causal Explore-Exploit}
   \label{alg:known_q}
\begin{algorithmic}
   \STATE {\bfseries Input:} $T,\boldsymbol{q}$
   \STATE Let $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
   \STATE Let $h = T^{2/3}m^{1/3}log(TK)^{1/3}$
   \STATE Let $A = \set{(i,j) : i \leq m, j = 1}$ be the set of infrequently observed arms
   \FOR{$t=1$ {\bfseries to} $h/2$}
   \STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
   \ENDFOR
   \STATE Compute for all arms $(i,j) \notin A$:
   \eq{
   \hat{\mu}_{i,j} = \frac{2}{h}\frac{\sum_{t=1}^{h/2}\ind{X_{i,t}=j}r_t}{q_i^j(1-q_i)^{1-j}}
   }
   \FOR{$(i,j) \in A$}
    \FOR{$t'=1$ {\bfseries to} $h/2m$}
    \STATE Choose the action $do(X_{i,t'} = j)$ and observe $r_t$
    \ENDFOR
    \STATE Compute $\hat{\mu}_{i,j} = \frac{2m}{h}\sum_{t'=1}^{h/2m}\ind{X_{i,t'} = j}r_{t'}$
   \ENDFOR
   \STATE Compute $(\hat{i}^*,\hat{j}^*) = \argmax_{(i,j)} \hat{\mu}_{i,j}$
   \FOR{$t=h$ {\bfseries to} $T$}
   \STATE Choose the action $do(X_{\hat{i}^*,t} = \hat{j}^*)$
   \ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:known_q_regret}
Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
Then \cref{alg:known_q} satisfies
\eq{
R(T) \in \bigo{T^{2/3}m^{1/3}log(KT)^{1/3}}\,.
}
\end{theorem}

The lower bound for the standard bandit problem is $R_t \in \bigomega{\sqrt{TK}}$ \cite{Auer1995}. Comparing these results shows exploiting the extra information provided by the causal structure should outperform standard bandit algorithms when the number of arms is large, $K >> m^{2/3}T^{1/3}$. The parameter $m$ summarizes the vector $\boldsymbol{q}$, and represents the number of actions that occur rarely naturally and thus must be explicitly explored. If $q_1,...,q_N = 0$, the problem is completely unbalanced and $m = N$. If $q_1,...,q_N = \frac{1}{2}$, the problem is completely balanced and $m = 1$.

Algorithm \ref{alg:known_q} tries to learn the rewards for all the arms during an exploration phase $h$ and then picks the arm with the highest empirical mean for all remaining timesteps. During its exploration phase, it learns all the frequently occurring actions by observation and the remaining, infrequently occurring actions, by explicitly playing them. This leads to Chernoff type high probability bounds on the difference between the empirical and true rewards for all arms of the form $P(\hat{\mu}_{i,j} - \mu_{i,j} > D) \leq e^{-hD^2/m}$. By choosing optimal values for $D$ and $h$ we obtain the regret bound given in \cref{thm:known_q_regret}. A full proof is given in the supplementary materials. (\color{red} or should we put it here? \color{black})

Algorithm \ref{alg:known_q} relies on $\boldsymbol{q}$ and thus $m$ being known. We now consider the case where $\boldsymbol{q}$ is unknown. We begin by considering the simple regret, defined as the expected difference between the mean payoff of the optimal action and that of the action estimated to be optimal in $T$th round.

\eqn{
R^{simple}(h) = \E{\mu^* - max_{(i,j)}\hat{\mu}_{i,j}(h)}
}

\begin{algorithm}[h]
\caption{Causal Best Arm Identification}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} $T, N$
\FOR{$t \in 1,\ldots,T/2$}
\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
\ENDFOR
\STATE Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
\eq{
\hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{i,t} = j}}\,.
}
\STATE Compute $\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} X_{i,t}$
\STATE Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
\STATE Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
\STATE Compute $\hat m = \min\set{1 \leq i \leq N : \hat s'_{i+1} \geq \frac{1}{i}}$
\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
\STATE Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$

\FOR{$(i,j) \in A$}
\FOR{$t \in 1,\ldots,T/2\hat m$}
\STATE Choose action $do(X_{i,t} = j)$ and observe $r_t$
\ENDFOR
\STATE Recompute $\hat \mu_{i,j} = \frac{2\hat m}{T} \sum_{t=1}^{ T/2\hat m} r_t(X_{i,t}=j)$ 
\ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:unknown_q_simpleregret}
Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$.
Then \cref{alg:simple} satisfies
\eq{
R^{simple}(T) \in \bigo{\sqrt{\frac{m}{T}log\left(\frac{NT}{m}\right)}}\,.
}
\end{theorem}

The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}. By utilizing the causal structure the dependence on $K$ is reduced to a dependence on $m$, which may be much smaller. The intuition behind \cref{alg:simple} is that we can use data collected during rounds where we select the $do()$ action to estimate $m$ and determine which arms we must play explicitly.

For the bandit problem problem we need to collect some data to estimate $m$ before determining if we should use the causal explore-exploit algorithm or a standard bandit algorithm.

\begin{algorithm}[h]
\caption{Bandit Regret Algorithm}\label{alg:bandit}
\begin{algorithmic}[1]
\STATE {\bf Input:} $T, N$
\STATE $\delta = \frac{1}{T^{1/3}}$ 
\STATE $T_1 = 48Nlog\left(4N/\delta\right)$ 
\STATE Run \cref{alg:simple} to line 11 with input $T_1,N$.
\IF {$\hat{m} > \frac{N^{3/2}}{\sqrt{T}}$}
\STATE Switch to the standard UCB algorithm.
\ELSE
\STATE $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
\STATE Run \cref{alg:simple} with input $h,N$.
\STATE Compute $(\hat{i}^*,\hat{j}^*) = \argmax_{(i,j)} \hat{\mu}_{i,j}$
\FOR{$t=h$ {\bfseries to} $T$}
\STATE Choose the action $do(X_{\hat{i}^*,t} = \hat{j}^*)$
\ENDFOR
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:unknown_q_regret}
Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
Then \cref{alg:bandit} satisfies
\eq{
R(T) \in \bigo{T^{2/3}m^{1/3}log(KT)^{1/3}}\,.
}
\end{theorem}

\subsection{Lower Bounds}

\section{Experiments}

\section{Related Work}
There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures. Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc. There is also an active line of research into bandits with feedback defined by a graph. Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action []. The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.

In our algorithm, we have only used the side information provided by the $do()$ action about other actions. Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen. The feedback graph is distinct from the causal graph. A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$. For this specific problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions. The Exp3-IX algorithm \cite{Kocak2014} was developed for the adversarial version of this problem and has regret $\bigo{\sqrt{\bar{\alpha}T}}$, where $\bar{\alpha}$ is the average independence number of $G_t$. In our case $\bar{\alpha} = \frac{N}{2}$ so we again obtain the regret of the standard bandit algorithm. The issue here is that a malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action. This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$

\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action. In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case. \cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.

More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}. In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate. The quality of the estimate notably depends not only on the number of times that action was selected. For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$. This more complex side information does not fit within the graph feedback framework.

Partial monitoring is a very general framework for for decoupling the feedback from the action and reward. It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}. Partial monitoring algorithms yield results that are optimal with respect the the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.

ALSO NEED TO MENTION ANY OTHER COMBINATIONS OF BANDITS+CAUSAL (eg the Elias NIPS paper and Generalized Thompson Sampling paper)

Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action. Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context. (Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results. It is easy to come up with a counter example in which it does not).

\section{Conclusion}

\end{document}

### 2015-12-24 causal-bandit/icml-2015/icml2015stylefiles/example_paper.tex
begins paper outline

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ICML 2015 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use the following line _only_ if you're still using LaTeX 2.09.
%\documentstyle[icml2015,epsf,natbib]{article}
% If you rely on Latex2e packages, like most moden people use this:
\documentclass{article}

% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

% For citations
\usepackage{natbib}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2015} with
% \usepackage[nohyperref]{icml2015} above.
\usepackage{hyperref}

% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{icml2015}

% Employ this version of the ``usepackage'' statement after the paper has
% been accepted, when creating the final version.  This will set the
% note in the first column to ``Proceedings of the...''
%\usepackage[accepted]{icml2015}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Submission and Formatting Instructions for ICML 2015}

\begin{document}

\twocolumn[
\icmltitle{Submission and Formatting Instructions for \\ 
           International Conference on Machine Learning (ICML 2015)}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2015
% package.
\icmlauthor{Your Name}{email@yourdomain.edu}
\icmladdress{Your Fantastic Institute,
            314159 Pi St., Palo Alto, CA 94306 USA}
\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
\icmladdress{Their Fantastic Institute,
            27182 Exp St., Toronto, ON M6H 2T1 CANADA}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{boring formatting information, machine learning, ICML}

\vskip 0.3in
]

\begin{abstract} 
The purpose of this document is to provide both the basic paper template and
submission guidelines.
\end{abstract}

\section{Electronic Submission}
\label{submission}

Submission to ICML 2015 will be entirely electronic, via a web site
(not email).  Information about the submission process and \LaTeX\ templates
are available on the conference web site at:
\begin{center}
\textbf{\texttt{http://icml.cc/2015/}}
\end{center}
Send questions about submission and electronic templates to
\texttt{francis.bach@inria.fr, david.blei@columbia.edu}.

The guidelines below will be enforced for initial submissions and
camera-ready copies.  Here is a brief summary:
\begin{itemize}
\item Submissions must be in PDF.
\item The maximum paper length is \textbf{8 pages excluding references, and 10 pages
  including references} (pages 9 and 10 must contain only references).
\item Do \textbf{not include author information or acknowledgments} in your initial
submission. 
\item Your paper should be in \textbf{10 point Times font}.
\item Make sure your PDF file only uses Type-1 fonts.
\item Place figure captions {\em under} the figure (and omit titles from inside
the graphic file itself).  Place table captions {\em over} the table.
\item References must include page numbers whenever possible and be as complete
as possible.  Place multiple citations in chronological order.  
\item Do not alter the style template; in particular, do not compress the paper
format by reducing the vertical spaces.
\end{itemize}

\subsection{Submitting Papers}

{\bf Paper Deadline:} The deadline for paper submission to ICML 2015
is at \textbf{23:59 Universal Time (3:59 Pacific Daylight Time) on February 6, 2015}.
If your full submission does not reach us by this time, it will 
not be considered for publication. There is no separate abstract submission.

{\bf Anonymous Submission:} To facilitate blind review, no identifying
author information should appear on the title page or in the paper
itself.  Section~\ref{author info} will explain the details of how to
format this.

{\bf Simultaneous Submission:} ICML will not accept any paper which,
at the time of submission, is under review for another conference or
has already been published. This policy also applies to papers that
overlap substantially in technical content with conference papers
under review or previously published. ICML submissions must not be
submitted to other conferences during ICML's review period. Authors
may submit to ICML substantially different versions of journal papers
that are currently under review by the journal, but not yet accepted
at the time of submission. Informal publications, such as technical
reports or papers in workshop proceedings which do not appear in
print, do not fall under these restrictions.

\medskip

To ensure our ability to print submissions, authors must provide their
manuscripts in \textbf{PDF} format.  Furthermore, please make sure
that files contain only Type-1 fonts (e.g.,~using the program {\tt
  pdffonts} in linux or using File/DocumentProperties/Fonts in
Acrobat).  Other fonts (like Type-3) might come from graphics files
imported into the document.

Authors using \textbf{Word} must convert their document to PDF.  Most
of the latest versions of Word have the facility to do this
automatically.  Submissions will not be accepted in Word format or any
format other than PDF. Really. We're not joking. Don't send Word.

Those who use \textbf{\LaTeX} to format their accepted papers need to pay close
attention to the typefaces used.  Specifically, when producing the PDF by first
converting the dvi output of \LaTeX\ to Postscript the default behavior is to
use non-scalable Type-3 PostScript bitmap fonts to represent the standard
\LaTeX\ fonts. The resulting document is difficult to read in electronic form;
the type appears fuzzy. To avoid this problem, dvips must be instructed to use
an alternative font map.  This can be achieved with the following two commands:

{\footnotesize
\begin{verbatim}
dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
ps2pdf paper.ps
\end{verbatim}}
Note that it is a zero following the ``-G''.  This tells dvips to use
the config.pdf file (and this file refers to a better font mapping).

A better alternative is to use the \textbf{pdflatex} program instead of
straight \LaTeX. This program avoids the Type-3 font problem, however you must
ensure that all of the fonts are embedded (use {\tt pdffonts}). If they are
not, you need to configure pdflatex to use a font map file that specifies that
the fonts be embedded. Also you should ensure that images are not downsampled
or otherwise compressed in a lossy way.

Note that the 2015 style files use the {\tt hyperref} package to
make clickable links in documents.  If this causes problems for you,
add {\tt nohyperref} as one of the options to the {\tt icml2015}
usepackage statement.

\subsection{Reacting to Reviews}

We will continue the ICML tradition in which the authors are given the
option of providing a short reaction to the initial reviews. These
reactions will be taken into account in the discussion among the
reviewers and area chairs.

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except of
course that the normal author information (names and affiliations)
should be given.  See Section~\ref{final author} for details of how to
format this.

The footnote, ``Preliminary work.  Under review by the International
Conference on Machine Learning (ICML).  Do not distribute.'' must be
modified to ``\textit{Proceedings of the
$\mathit{31}^{st}$ International Conference on Machine Learning},
Lille, France, 2015.  JMLR: W\&CP volume 37. 
Copyright 2015 by the author(s).''

For those using the \textbf{\LaTeX} style file, simply change
$\mathtt{\backslash usepackage\{icml2015\}}$ to

$$\mathtt{\backslash usepackage[accepted]\{icml2015\}}$$

\noindent
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one.  The running title consists of a
single line centered above a horizontal rule which is $1$ point thick.
The running head should be centered, bold and in $9$ point type.  The
rule should be $10$ points above the main text.  For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the {\tt fancyhdr} package which is included in the ICML
2015 style file package.  In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\icmltitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the same format to ensure the printer can
reproduce them without problems and to let readers more easily find
the information that they desire.

\subsection{Length and Dimensions}

Papers must not exceed eight (8) pages, including all figures, tables,
and appendices, but excluding references. When references are included,
the paper must not exceed ten (10) pages. Any submission that exceeds 
this page limit or that diverges significantly from the format specified 
herein will be rejected without review.

The text of the paper should be formatted in two columns, with an
overall width of 6.75 inches, height of 9.0 inches, and 0.25 inches
between the columns. The left margin should be 0.75 inches and the top
margin 1.0 inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

To facilitate blind review, author information must not appear.  If
you are using \LaTeX\/ and the \texttt{icml2015.sty} file, you may use
\verb+\icmlauthor{...}+ to specify authors.  The author information
will simply not be printed until {\tt accepted} is an argument to the
style file. Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If your are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we 
have shown \ldots'').

Do not anonymize citations in the reference section by removing or
blacking out author names. The only exception are manuscripts that are
not yet published (e.g. under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have 
to be submitted
as Supplementary Material via CMT. However, keep in mind that an ICML
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look a the Supplementary Material when writing their
review.

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches
below the bottom rule surrounding the title. The authors' names should
appear in 10~point bold type, electronic mail addresses in 10~point
small capitals, and physical addresses in ordinary 10~point type.
Each author's name should be flush left, whereas the email address
should be flush right on the same line. The author's physical address
should appear flush left on the ensuing line, on a single line if
possible. If successive authors have the same affiliation, then give
their physical address only once.

A sample file (in PDF) with author names is included in the ICML2015 
style file package.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below
the final address. The heading `Abstract' should be centered, bold,
and in 11~point type. The abstract body should use 10~point type, with
a vertical spacing of 11~points, and should be indented 0.25~inches
more than normal on left-hand and right-hand margins. Insert
0.4~inches of blank space after the body. Keep your abstract brief and 
self-contained,
limiting it to one paragraph and no more than six or seven sentences.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}

Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{For the sake of readability, footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper. 
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
\caption{Historical locations and number of accepted papers for International
  Machine Learning Conferences (ICML 1993 -- ICML 2008) and
  International Workshops on Machine Learning (ML 1988 -- ML
  1992). At the time this figure was produced, the number of
  accepted papers for ICML 2008 was unknown and instead estimated.}
\label{icml-historical}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Figures}

You may want to include figures in the paper to help readers visualize
your approach and your results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
{\it after\/} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it, as in
Figure~\ref{icml-historical}.  The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left.  You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment {\tt figure*} in \LaTeX), but always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic'' 
environments to format pseudocode. These require 
the corresponding stylefiles, algorithm.sty and 
algorithmic.sty, which are supplied with this package. 
Algorithm~\ref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$} 
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like 
figures, these should be centered, legible, and numbered consecutively. 
However, place the title {\it above\/} the table with at least 
0.1~inches of space before the title and the same after it, as in 
Table~\ref{sample-table}. The table title should be set in 9~point 
type and centered unless it runs two or more lines, in which case it
should be flush left.

% Note use of \abovespace and \belowspace to get reasonable spacing 
% above and below tabular lines.

\begin{table}[t]
\caption{Classification accuracies for naive Bayes and flexible 
Bayes on various data sets.}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\hline
\abovespace\belowspace
Data set & Naive & Flexible & Better? \\
\hline
\abovespace
Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
\belowspace
Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
\hline
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

Tables contain textual material that can be typeset, as contrasted 
with figures, which contain graphical material that must be drawn. 
Specify the contents of each row and column in the table's topmost
row. Again, you may float tables to a column's top or bottom, and set
wide tables across both columns, but place two-column tables at the
top or bottom of the page.

\subsection{Citations and References}

Please use APA reference format regardless of your formatter
or word processor. If you rely on the \LaTeX\/ bibliographic 
facility, use {\tt natbib.sty} and {\tt icml2015.bst} 
included in the style-file package to obtain this format.

Citations within the text should include the authors' last names and
year. If the authors' names are included in the sentence, place only
the year in parentheses, for example when referencing Arthur Samuel's
pioneering work \yrcite{Samuel59}. Otherwise place the entire
reference in parentheses with the authors and year separated by a
comma \cite{Samuel59}. List multiple references separated by
semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
construct only for citations with three or more authors or after
listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

Authors should cite their own work in the third person
in the initial version of their paper submitted for blind review.
Please refer to Section~\ref{author info} for detailed instructions on how to
cite your own papers.

Use an unnumbered first-level section heading for the references, and 
use a hanging indent style, with the first line of the reference flush
against the left margin and subsequent lines indented by 10 points. 
The references at the end of this document give examples for journal
articles \cite{Samuel59}, conference publications \cite{langley00}, book chapters \cite{Newell81}, books \cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, 
technical reports \cite{mitchell80}, and dissertations \cite{kearns89}.

Alphabetize references by the surnames of the first authors, with
single author entries preceding multiple author entries. Order
references for the same authors by year of publication, with the
earliest first. Make sure that each reference includes all relevant
information (e.g., page numbers).

\subsection{Software and Data}

We strongly encourage the publication of software and data with the
camera-ready version of the paper whenever appropriate.  This can be
done by including a URL in the camera-ready copy.  However, do not
include URLs that reveal your institution or identity in your
submission for review.  Instead, provide an anonymous URL or upload
the material as ``Supplementary Material'' into the CMT reviewing
system.  Note that reviewers are not required to look a this material
when writing their review.

% Acknowledgements should only appear in the accepted version. 
\section*{Acknowledgments}

\textbf{Do not} include acknowledgements in the initial version of
the paper submitted for blind review.

If a paper is accepted, the final camera-ready version can (and
probably should) include acknowledgements. In this case, please
place such acknowledgements in an unnumbered section at the
end of the paper. Typically, this will include thanks to reviewers
who gave useful comments, to colleagues who contributed to the ideas, 
and to funding agencies and corporate sponsors that provided financial 
support.

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{example_paper}
\bibliographystyle{icml2015}

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was
% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
% slightly modified from the 2009 version by Kiri Wagstaff and 
% Sam Roweis's 2008 version, which is slightly modified from 
% Prasad Tadepalli's 2007 version which is a lightly 
% changed version of the previous year's version by Andrew Moore, 
% which was in turn edited from those of Kristian Kersting and 
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2015-12-24 causal-bandit/icml-2015/suplementary_proofs.tex
begins paper outline

--- 2015-10-09 causal-bandit/observe_then_best.tex
@@
 \newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}{\mathbb E}
+\newcommand{\E}[1]{\mathbb E\left[#1\right]}
 \newcommand{\Var}{\operatorname{Var}}
@@
 \newcommand{\Ber}{\operatorname{Bernoulli}}
-\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+\renewcommand{\P}[1]{\operatorname{P}\left(#1\right)}
+\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
+\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
+\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
+\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
+\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
 
@@
 
-\begin{comment}
-\section{Random notes}
-Information directed sampling have developed a general framework to construct algorithms that leverege generic structure between arms to minimize Bayesian regret, by explicitly trading off expected regret and information gain at each timestep.
+\section{Proof of Theorem 1}
+\begin{theorem}\label{thm:known_q_regret}
+Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
+Then Algorithm 1 satisfies
+\eq{
+R(T) \in \bigo{T^{2/3}m^{1/3}log(KT)^{1/3}}\,.
+}
+\end{theorem}
 
-What can be expressed in the partial monitoring framework that cannot be defined by the feedback graph?
+Let $A = \set{(i,j) : i \leq m, j = 1}$ be the set of infrequently observed arms
 
-We consider a variant of the stochastic multi-armed bandit problem when we have prior knowledge of the causal structure, but not the functional form, of the relationship between variables.
-
-sequential decision process,exploration-exploitation trade off
-
-Points to present. Why we use regret against single best arm - it makes sense in the stochastic setting. 
-The difference between observational and interventional distributions.
-
-terminology for intervention vs observation.
-Recast bandit algorithm in terms of this.
-Show limitations to existing approaches by pointing to places were existing algorithms eg UCB/explore-exploit/exp3 require unbiased estimates.
-Are unbiased estimates actually impossible in our setting?
-
-This paper is about sequentially choosing actions. 
-Modelling sequntial decition processes.
-
-In the classic multi-armed bandit setting, an ag
-
-General bandit part ...
-define multi-armed bandit, and importance.
-Decision making is fundamental.
-Goal is to select an action to maximize a reward (or minimize a loss).
-In many settings, such as drug trials, online ad placement, etc we can choose actions sequentially.
-This introduces an exploration-exploitation trade-off.
-
-The problem of sequentially choosing actions to minimize some loss wide range of applications in XXXX, YYYY. The multi-armed bandit setting, in which .... .
-
-Now the required causal stuff should go here.
-
-Causal structure can be represented as a directed acyclic graph with nodes representing the variables and an edge from $A \rightarrow B$ meaning $A$ causes $B$.
-Deterministic interventions, consisting of setting variables to some value, are modelled by setting the nodes to the desired values and removing any links coming into them, as their value is now no longer set by other variables in the graph according to the original mechanism.
-
-If the causal graph is known, such that there are no hidden variables causing 2 or more of the variables and the probabilities are all > 0 then the distribution of any variable after any intervention can be inferred (asymptotically) from data collected in absence of any interventions.
-\end{comment}
-
-\section*{Introduction}
-
-Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
-For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
-A widely used framework for sequential decision making is the multi-armed bandit.
-In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary and independent of the reward distribution of other actions.
-At each timestep the agent selects an action and receives a reward sampled iid from the corresponding reward distribution.
-
-An an alternate approach to selecting actions is causal inference.
-Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
-This allows an agent to predict the outcome of an action based on non-experimental data.
-This approach is common in social science, demography, and economics where explicit experimentation may be difficult.
-For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.
-
-We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
-This structure creates dependencies between the rewards of different arms such that selecting one action can provide information on the reward for other actions.
-
-There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
-Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
-There is also an active line of research into bandits with feedback defined by a graph.
-Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
-The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
-
-\section*{Problem Formulation}
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
-As an example, consider a farmer wishing to optimize the yield of her crop.
-She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
-We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
-After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
-This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
-
-\begin{figure}[h]
-\centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\end{figure}
-
-Let $q \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known end point $T$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, N}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^N$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(q_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(r(\tilde X))$ where $r:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
-\end{enumerate}
-
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[r(X)|do(X_i = j)]$. The optimal reward and action are denoted $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
-}
-
-The problem can be treated as a classical multi-armed bandit with $K = 2N$ arms. However, this does not utilize the information provided by the causal assumption.
-
-Now need to expand upon how the assumption gives us extra information.
-
-The difficulties faced due to bias.
-
-To deal with this issue we use a simple explore-exploit algorithm.
-Our algorithm will explore for $h$ time-steps, sampling actions in a way that depends on our prior knowledge of $q$ but is independent of the observed rewards.
-We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
-
-\section*{Results}
-
-Summarize results here and note differences to classic bandit results
-
-\subsection*{Known and Balanced q}
-
-We begin with the simplest case were we assume that $q_i=\frac{1}{2} \forall i$.
-During the exploration phase we sample actions uniformly at random.
-In this case, this is equivalent to purely observing, that is taking no action and allowing all input variables to take their value randomly as  $X_{t,i} \sim \Ber(\frac{1}{2})$.
-
-We will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$.
-
-Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
-
-\subsubsection*{Regret during explore phase}
-Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
-\eqn{
-R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta] < h
-}
-
-\subsubsection*{Regret during exploit phase}
-The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
-
-\eqn{
-\hat{i^*} = argmax_i [\hat{\mu}_i]
-}
-\eqn{
-R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
-}
-
-The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$.
-Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
-\eqn{
-R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
-}
-The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.
-
-Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$.
-If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$
-
-\eqn{
-\label{eqn:probDeltaToLarge}
-P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
-}
-
-If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution.
-Instead we will use an importance weighted estimator.
-
-\eqn{
-\label{eqn:importance_weighted_estimator}
-\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
-}
-
-where $q_i = P(\text{arm $i$ active})$
-
-Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:
-
-\eqn{
-\label{eqn:balancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
-}
-
-\eqn{
-R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
-}
-
-Let $D = \sqrt{\frac{8}{h}\log(hK)}$
-
-\eqn{
-R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)
-}
-
-\subsection*{Total Regret}
-
-Putting together the regret from the exploration and exploitation phases,
-
-\eqn{
-R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)\\
-& \leq h + T\left(\sqrt{\frac{8}{h}\log(TK)} + \frac{1}{h}\right)
-}
-
-Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,
+For the frequently observed arms, $(i,j) \notin A$ we have:
 
 \eqn {
-R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
+\hat{\mu}_{i,j} = \frac{2}{h}\frac{\sum_{t=1}^{h/2}\ind{X_{i,t}=j}r_t}{q_i^j(1-q_i)^{1-j}}
 }
-
-If $T \geq 2$ and $K \geq 2$, the first term dominates and,
-
-\eqn {
-R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
-}
-
-The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$
-
-\subsection*{Empirical results}
-\begin{figure}[H]
-\centering
-\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
-\includegraphics[width=.5\textwidth]{explore_exploit}
-\end{figure}
-
-\section{Generalizing to unbalanced $\boldsymbol{q}$}
-
-When some arms have low natural probability we cannot rely on exploring them adequately by pure observation. We need to explicitly play them during the exploration phase.
-
-We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
-
-Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
-\leq q_N$.
-Let $m \in [2,N] = \{m:q_m > \frac{1}{m}\}$ Ie if the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
-If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$.
-Let $q_{ij} = \P{X_i = j}$
-
-Suppose we observe for the first $h/2$ timesteps. This is at worst half the optimal.
-
-We then have estimates
-
-\eqn {
-\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
-}
-
-We take this as our estimate for those arms for which $q_{ij} > \frac{1}{m}$
 
@@
 
-We play each of the remaining $m$ arms $\frac{h}{2m}$ times so for them we get
+The algorithm explicitly plays each of the infrequently observed arms, $(i,j) \in A$, $\frac{h}{2m}$ times. So:
 
@@
 
-Again comparing to the UCB bound of $\sqrt{KT}$, we expect the causal algorithm to do better if $K > m^{2/3}T^{1/3}$, or alternatively if $m \leq \frac{K^{3/2}}{\sqrt{T}}$
-
-To get results that degrade to similar order bounds as UCB when the arms are very unbalanced, I will need to drop the explore/exploit strategy.
-
-\subsection{Generalizing to unknown q}
-We now consider the case where $\boldsymbol{q}$, and thus $m$ is not known in advance. We begin by considering the simple regret
-
-\subsubsection{Bounding the Simple Regret}
-
-\begin{itemize}
-\item $N$ is number of variables.
-\item $q_i$ is probability that $X_{i,t} = 1$ for any $t$
-\item $s_i = \min\set{q_i, 1 - q_i}$
-\item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
-\item $I_t \in \set{1,\ldots,N} \times \set{0,1}$ is the intervention
-\end{itemize}
-
-Let $R_s(h)$ be the simple regret after $h$ time steps, which is
-\eq{
-R_s(h) = \E\left[\mu^* - \mu_{I_h}\right]\,,
-}
-where $\mu^*$ is the mean payoff of the optimal intervention and $I_h$ is the intervention chosen in the $h$th round.
-We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.
-
-\begin{algorithm}[H]
-\caption{Simple Regret Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\State {\bf Input:} $h, N$
-\For{$t \in 1,\ldots,h/2$}
-\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
-\EndFor
-\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
-\eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
-}
-\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
-\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
-\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
-
-\For{$(i,j) \in A$}
-\For{$t \in 1,\ldots,h/2\hat m$}
-\State Choose action $X_{i,t} = j$
-\State Observe reward $r_t(X_{i,t}=j)$
-\EndFor
-\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
-\EndFor
-\end{algorithmic}
-\end{algorithm}
+\section{Proof of Theorem 2}
 
@@
 Define $m = \min\set{2 \leq i \leq N : q_i \geq 1/i}$.
-Then the algorithm given in \cref{alg:simple} satisfies
+Then Algorithm 2 satisfies
 \eq{
-R_s(h) \in O\left(\sqrt{\frac{m}{h} \log \left(\frac{NT}{m}\right)}\right)\,.
+R^{simple}(T) \in O\left(\sqrt{\frac{m}{T} \log \left(\frac{NT}{m}\right)}\right)\,.
 }
@@
 \begin{lemma}\label{lem:conc1}
-$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
+$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta$.
 \end{lemma}
@@
 \eq{
-\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} Z_t\,.
+\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} Z_t\,.
 }
-Now $Z_1,\ldots,Z_{h/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
+Now $Z_1,\ldots,Z_{T/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
 \end{proof}
+\begin{lemma}\label{lem:m_est}
+Let $\delta >0.$ 
+If $h \geq 24m \log\frac{4N}{\delta}$
 
-\begin{lemma}\label{lem:conc2}
-Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
-define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
-If
-\eq{
-h \geq 24m \log\frac{4N}{\delta}\,,
-}
 then
 \eq{
-\P{\hat m \leq 2m} \geq 1 - \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-It is easy to see that the worst case occurs when 
-\eq{
-q_i = \begin{cases}
-0 & \text{if } i < m \\
-\frac{1}{m} & \text{otherwise}\,.
-\end{cases}
-}
-Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that 
-\eq{
-(\forall i) \qquad \left| \hat q_i - q_i\right| 
-&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
-&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
-}
-By assumption 
-\eq{
-h \geq 24m \log \frac{4N}{\delta}\,.
-}
-Therefore with probability at least $1 - \delta$ we have
-\eq{
-(\forall i) \qquad \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.
-}
-Note that for $i$ with $q_i = 0$ it is easy to see that $\hat q_i = 0$ (the variance is zero). The above guarantee that
-for $i$ with $q_i = 1/m$ we have $\hat q_i \in [1/(2m), 3/(2m)]$ for all $i$ with probability at least $1 - \delta$,
-which implies that $\hat s_i \geq 1/(2m)$ for all $i$ where $q_i = 1/m$.
-Therefore $\hat m \leq 2m$ with probability at least $1 - \delta$. 
-\end{proof}
-
-\begin{lemma}\label{lem:m_est_small_enough}
-Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
-define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
-If
-\eq{
-h \geq 24m \log\frac{4N}{\delta}\,,
-}
-then
-\eq{
-\P{\hat m < \frac{2}{3}m} \leq \delta\,.
+\P{\hat m < \frac{2}{3}m} \leq \delta\ \text{ and } \P{\hat m > 2m} \leq \delta\,.
 }
@@
 
-To over-estimate $m$, we must identify as balanced some unbalanced arms (ie those with $i < m$. For all arms with $i < m$, by definition, $q_i < \frac{1}{m}$
-
-Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that: \color{red}(this step not quite right - because I haven't said $q_i = 1/m$ as the worst case is that $q_i$ is infintesimally smaller than $1/m$.
-With smaller $q_i$ the bounds would be slightly tighter but not made up for by the decrease in the mean - ie we are more likely to over-estimate if $q_i = 1/m$ than if $q_i < 1/m$.
-Need a better way to say this ...
-).
-Should probably just combine this with the other side into one lemma (even though we don't need this side for simple regret.
-\color{black}
+Let $\boldsymbol{q}^b$ and $\boldsymbol{q}^{ub}$ and be the maximally balanced and unbalanced $\boldsymbol{q}$ for a given $m$ respectively.
 
 \eq{
-(\forall i) \qquad \left| \hat q_i - q_i\right| 
-&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
-&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
+q^{ub}_i = \begin{cases}
+0 & \text{if } i \leq m \\
+\frac{1}{m} & \text{otherwise}\,.
+\end{cases}\\
+q^b_i < \begin{cases}
+\frac{1}{m} & \text{if } i \leq m \\
+1 & \text{otherwise}\,.
+\end{cases}  
+}
+For $\hat{m}$ to over-estimate $m$, we must identify some balanced arms as unbalanced. For $\hat{m}$ to under-estimate $m$, we must identify some unbalanced arms as balanced.
+\eq{
+\P{\hat m > 2m} \leq \P{\hat m > 2m|\boldsymbol{q} = \boldsymbol{q}^{ub}}\\
+\P{\hat m < \frac{2}{3}m} \leq \P{\hat m < \frac{2}{3}m|\boldsymbol{q} = \boldsymbol{q}^{b}}
 }
 
-By assumption 
+Given $\boldsymbol{q} = \boldsymbol{q}^{ub}$, we have by \cref{lem:conc1}, with probability at least $1 - \delta$ that:
+
 \eq{
-h \geq 24m \log \frac{4N}{\delta}\,.
+& \left| \hat q_i - q_i\right| 
+ \leq \begin{cases}
+0 & \text{if } i \leq m\\
+\sqrt{\frac{6}{mh} \log\frac{2}{\delta}} & \text{otherwise} \\
+\end{cases}\\
+\implies &
+\begin{cases}
+(\forall i \leq m) \qquad \left| \hat q_i - 0\right| = 0 \\
+(\forall i > m) \qquad \left| \hat q_i - \frac{1}{m}\right| \leq \frac{1}{2m} \; \text{, taking the union bound and assuming } h \geq 24m \log\frac{4N}{\delta}\\
+\end{cases}\\
+\implies &
+\begin{cases}
+(\forall i \leq m) \qquad  \hat q_i  = 0 \\
+(\forall i > m) \qquad \hat q_i \in [\frac{1}{2m}, \frac{3}{2m}]\\
+\end{cases}\\
+\implies & \hat{m} \leq 2m
 }
-Therefore with probability at least $1 - \delta$ we have
+
+Given $\boldsymbol{q} = \boldsymbol{q}^{b}$, we have by \cref{lem:conc1}, with probability at least $1 - \delta$ that:
+
 \eq{
-(\forall i < m) \qquad & \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.\\
- & \hat{q}_i < \frac{3}{2}m \\
- \implies & \hat{m} > \frac{2}{3}m
+& \left| \hat q_i - q_i\right| 
+ \leq \sqrt{\frac{6}{mh} \log\frac{2}{\delta}} \qquad \text{if } i \leq m \\
+\implies &
+(\forall i \leq m) \qquad  \hat q_i \leq \frac{3}{2m} \\
+\implies & \hat{m} \geq \frac{2m}{3}
 }
@@
 
-Now by \cref{lem:conc2}, 
+Now by \cref{lem:m_est}, 
 \eq{
@@
 
-\subsubsection{Bounding the Bandit Regret}
+\section{Proof of Theorem 3}
+
+\begin{theorem}\label{thm:unknown_q_regret}
+Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
+Then Algorithm 3 satisfies
+\eq{
+R(T) \in \bigo{T^{2/3}m^{1/3}log(KT)^{1/3}}\,.
+}
+\end{theorem}
 
@@
 
-\begin{algorithm}[H]
-\caption{Bandit Regret Algorithm}\label{alg:bandit}
-\begin{algorithmic}[1]
-\State {\bf Input:} $T, N$
-\State $T_1 = 24Nlog\left(4N/\delta\right)$ TODO figure out what $\delta$ should be 
-\For{$t \in 1,T_1$}
-\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
-\EndFor
-\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
-\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
-\If {$\hat{m} > \frac{N^{3/2}}{\sqrt{T}}$}
-\State Switch to the standard UCB algorithm with $K = 2N$ arms.
-\Else
-\State $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
-\For{$t \in T_1,h/2$}
-\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
-\EndFor
-\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
-\eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
-}
-\State Recompute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
-\State Recompute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\State Recompute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\State Recompute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
-\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
-
-\For{$(i,j) \in A$}
-\For{$t \in 1,\ldots,h/2\hat m$}
-\State Choose action $X_{i,t} = j$
-\State Observe reward $r_t(X_{i,t}=j)$
-\EndFor
-\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
-\EndFor
-\EndIf
-\end{algorithmic}
-\end{algorithm}
-
 Since we don't know $\boldsymbol{q}$, and thus $m$, we can't set the exploration time $h$ in advance based on $m$ as we did for the known $\boldsymbol{q}$ case.
@@
 
-Now lets try an implementation ...
-
-\subsubsection{ Older Attempts}
-We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.
-
-\eqn{
-\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}
-}
-
-Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ and construct $\bar{\boldsymbol{q}}$ such that $\bar{q}_1 \leq \bar{q}_2 \leq ...\leq \bar{q}_N$.
-We then estimate $m$ with $\hat{m} = min_i : \bar{q}_i \geq \frac{1}{i}$
-
-We estimate the reward for the apparently common arms, $i \geq \hat{m}$, as:
-
-\eqn{
-\hat{\mu}_i = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\frac{h}{2}\hat{q}_i} = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\sum_{t=1}^{h/2}\ind{X_i=1}}
-}
-
-We play the remaining arms $\frac{h}{2\hat{m}}$ times and estimate their reward as:
-
-\eqn{
-\hat{\mu}_i = \frac{2\hat{m}}{h}\sum_{t=1}^{h/2\hat{m}}\ind{Y = 1|X_i=1}
-}
-
-We now consider how errors in the estimation of $m$ effect our estimates of the arm rewards.
-For the arms with $i \geq \hat{m}$ we know $\hat{q}_i \geq \frac{1}{\hat{m}}$ and thus our estimate is based on at least $\frac{h}{2\hat{m}}$ observations.
-Similarly we explicitly play the infrequently observed arms, $i < m$,  $\frac{h}{2\hat{m}}$ times.
-Thus our estimates will only be worse than the known $\boldsymbol{q}$ case if $\hat{m} > m$.
-
-For a fixed $m$ the $\boldsymbol{q}$ most likely to lead us to overestimate $m$ is: \todot{Justify this claim.}
-
-\eqn{
-q_i = \begin{cases}
-0 & \text{ if } i < m\\
-\frac{1}{m} & \text{ if } i \geq m
-\end{cases}
-}
-
-We now bound $P(\hat{m} - m > \varphi)$ for this worst case.
-
-For $i < m$,we have $\bar{q}_i = q_i = 0$. For $i \geq m$ \todot{Is this really true, justify}
-\eq{
-P(q_i - \bar{q}_i \geq C) & =  P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})\\
-& \leq  2P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})
-}
-
-Via Bernstein's Inequality:
-\eqn{
-&2P(q_i - \hat{q}_i \geq C | \hat{q}_i \leq \frac{1}{2}) \leq 2\exp(-\frac{hC^2}{4q_i}) := \gamma \\
-\implies &P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(2/\gamma)}{mh}})\leq \gamma
-}
-
-Define $\hat{m}_i = \frac{1}{\hat{q}_i}$.
-
-\eqn{
-\hat{q}_i \geq q_i - C & \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
-& \implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}\\
-& \implies P(\hat{m}_i - m \geq \frac{m^2C}{1-mC}) \leq \gamma
-}
-
-Let $\varphi = \frac{m^2C}{1-mC} \implies C = \frac{\varphi}{m(\varphi+m)} \implies \gamma = 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2}) \implies \varphi = \frac{2m\sqrt{m\log(2/\gamma)}}{\sqrt{h}-2\sqrt{m \log(2/\gamma)}} $
-
-Note this implies that if $h \geq 16m\log(2/\gamma)$ then $\varphi \leq m$
-
-\eqn {
-P(\hat{m}_i - m \geq \varphi) \leq 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2})
-}
-
-If $\hat{m}_i - m \geq \varphi$ for at most $\varphi$ of the variables $i \geq m$, then $\hat{m} - m \leq \varphi$
-
-Let $W_i = \ind{\hat{m_i}-m \geq \varphi}$, $E[W_i] \leq \gamma$,$V[W_i] \leq \gamma$ 
-\eqn{
-P(\sum_{i=m}^N W_i \geq (N-m)\gamma + \varepsilon) \leq \exp(-\frac{\varepsilon^2}{2(N-m)\gamma+2\varepsilon/3})
-}
-Letting $(N-m)\gamma + \varepsilon = \varphi \implies \varepsilon = \varphi - (N-m)\gamma$
-\eqn{
-P(\hat{m}-m \geq \varphi) =  P(\sum_{i=m}^N W_i \geq \varphi) &\leq \exp({-\frac{(\varphi-(N-m)\gamma)^2}{2(N-m)\gamma + 2(\varphi-(N-m)\gamma)/3}})\\
-& \leq \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \text{, where } N\gamma < \varphi
-}
-
-If $P(\hat{m}-m \geq \varphi) \leq \zeta$.
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq & P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \leq \varphi )P(\hat{m}-m \leq \varphi) \\
-& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \varphi)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
-}
-
-Putting it together
-
-\eqn {
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}))^2}{2N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}) + 2\varphi/3}})
-}
-
-Assume $h > 16m\log(2N)$ and . Let $\varphi = \epsilon\sqrt{h}$. Assume $\epsilon \geq \frac{m}{\sqrt{h}}$ so as to ensure $\varphi > m$
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h}{16m}))^2}{2N\exp(-\frac{h}{16m}) + 2\varphi/3}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-\varphi/2)^2}{\varphi + 2\varphi/3}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\varphi}{20}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\epsilon \sqrt{h}}{20}}) := \delta\\
-& \leq \begin{cases}
-2 \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) & \text{ if } h < \frac{9m^2}{289\epsilon^2} \\
-2\exp({-\frac{3\epsilon\sqrt{h}}{20})} & \text{ otherwise } 
-\end{cases}
-}
-
-\eqn{
-\implies \epsilon \leq \begin{cases}
-\frac{1}{\sqrt{h}}\log(2/\delta) + \sqrt{\frac{m}{h}\log(2/\delta)} & \text{ if } h < \frac{9m^2}{289\epsilon^2}\\
-\frac{20}{3\sqrt{h}}\log(2/\delta) & \text{ otherwise }
-\end{cases}\\
-}
-
-\eqn{
-\implies \epsilon \leq \frac{20}{3\sqrt{h}}\log(2/\delta)+ \sqrt{\frac{m}{h}\log(2/\delta)} 
-}
-
-Does this fit with the assumption we made about $\epsilon$???
-
-\pagebreak
-
-The first term will dominate if
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3} \geq \frac{h\epsilon^2}{m+\varphi}
-}
-
-Setting them equal and solving for $\varphi$. Roughly,
-
-\eqn{
-\varphi = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = \frac{1}{2}(\sqrt{4\epsilon^2 h+ m^2} - m)\\
-\implies \epsilon = \sqrt{\frac{\log(1/\delta)(\log(1/\delta)+m)}{h}} \leq \frac{\log(1/\delta)}{\sqrt{h}} + \sqrt{\frac{m}{h}\log(1/\delta)}
-}
-
-Now repeating the above but more exactly ...
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}  \geq & \frac{\varphi^2-2\varphi N\gamma}{4\varphi/3} \; \text{, if } \varphi > 3N\gamma \\
-= & \frac{3}{4}(\varphi - 2N\gamma)
-}
-
-\eqn{
-\frac{3}{4}(\varphi - 2N\gamma) = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = &\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}\\
-\leq & \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}
-}
-
-If we let $\varphi =\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2} $
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp(-\frac{3}{4}(\varphi - 2N\gamma))\\
-\implies \epsilon = \sqrt{\frac{4\log^2(2/\delta)+(3m+6N\gamma)\log(2/\delta)}{3h}}\\
-\leq \frac{2\log(2/\delta)}{\sqrt{3h}}+\sqrt{\frac{m+2N\gamma}{h}\log(2/\delta)}
-}
-Ok - but I still need to ensure there exists a $\gamma$ such that  $N\gamma < \frac{\varphi}{3}$ and $P(\hat{m}_i - m > \varphi) \leq \gamma$
-
-\eqn{
-P(\hat{m}_i - m > a) \leq \exp(-\frac{h a^2}{4m(m+a)^2})
-}
-
-It feels like I have just gone round and round in circles and pushed the problem to somewhere else ... my expression for $\varphi$ now contains both $\epsilon$ and $\gamma$ ...
-
-\eqn{
-\varphi \geq \frac{1}{2}\sqrt{m^2+4N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}
-}
-
-\pagebreak
-
-\eqn{
- \implies & C = \frac{\varphi}{m(\varphi+m)}  \geq \frac{\varphi}{\alpha m^2} \text{ if } \varphi < (\alpha-1)m\\
-\implies & P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4\log(1/\gamma)}{mh}} \right)\right) \leq \gamma
-}
-
-If we let $\gamma = \frac{e^{-h/m}}{N}$
-
-\eqn{
-P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{\log N}{h}+\frac{1}{m})+2\sqrt{\frac{\log N}{mh}+\frac{1}{m^2}} \right)\right) \leq \gamma \\
-\implies P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{2}{m})+2\sqrt{\frac{2}{m^2}} \right)\right) \leq \gamma \; \text{, if } h > m\log N \\
-\implies P\left(\hat{m}_i - m \geq 6\alpha m \right) \leq \gamma
-}
-
-This doesn't work. We need $\hat{m}_i - m < (\alpha-1)m$ in order to get this expression.
-
-Or we could let $\varphi = \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}$
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp({-\frac{h\epsilon^2}{m+\frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}}})
-}
-\eqn{
-\implies \epsilon = & \frac{1}{\sqrt{2}}\sqrt{\frac{3 \log^2(2/\delta)+2m\log(2/\delta)}{h}+\frac{\sqrt{3}}{h^2}\sqrt{h^2\log^2(2/\delta)(3\log^2(2/\delta)+4m\log(2/\delta)+3mN\gamma)}} \\
-\leq & 
-}
-
 \end{document}
-
-\subsection{Option 1: Targeted sampling during exploration phase}
-
-The key fact we were utilizing to draw conclusions about multiple arms during each timestep of the explore phase is that, given our assumed causal structure, $P(Y|do(X_i=j)) = P(Y|X_i=j)$
-
-If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ with some probability not equal to $P(X_a=b)$, then we can no longer estimate $P(Y|X_i=j)$ directly from the proportion of successes given $X_i=j$.
-
-\eqn {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- \label{eqn:marginalizedProbability}
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b)
-}
-
-Define:
-\begin{itemize}
-\item  $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$
-\item $q_{ab} = P(X_a = b)$, the underlying probability that variable $a$ takes the value $b$
-\item $p_{ab} = P(do(X_a=b)|I=a)$, the probability that we set variable $a$ to $b$ given we are acting on variable $a$
-\end{itemize}
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-This estimator is unbiased and has variance bounded by (proofs in section \ref{seq:proof_expect_var}) :
-
-\eqn{
-\label{eqn:variance}
-V[\hat{\mu}_a^{ij}] \leq \begin{cases}
-\frac{1}{p_{ij}} & \text{ if } a = i \\
-\frac{1}{q_{ij}}\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right) & \text{ if } a \neq i
-\end{cases}
-}
-
-At each time step, we select and arm $do(X_a=b)$ and then receive estimates for all the arms $i,j$ with variances given by \ref{eqn:variance}.
-Note that $\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right)$ is minimized to equal $1$ if we set the intervened on variable according to its natural probability, that is we effectively take no action.
-
-Define  $\bar{q}_i = min_j q_{ij}$, the probability of the most unlikely value for each variable.
-We can assume without loss of generality that $\bar{q}_1 \leq ...
-\leq \bar{q}_N$.
-Define $m = max(N,min\{m: \frac{1}{m}\leq \bar{q}_m\})$ We allocate half our exploration timesteps $\frac{h}{2}$ to observing and allocate the remainder equally to play the unbalanced arms with $i \leq m$.
-
-At the end of the exploration phase, we combine the estimators from each timestep as a weighted sum with weights selected to minimize the variance of the combined estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{t=1}^h w_t \hat{\mu}_{a(t)}^{ij}, \text{ where } w_t = \frac{V_t^{-1}}{\sum_s V_s^{-1}}
-}
-
-We can then use Bernstein's inequality to bound the probability of large deviations
-
-\eqn{
-P\left(\sum_{t=1}^h w_t  \hat{\mu}_{a(t)}^{ij} - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\eqn {
-\sum_{t=1}^h\frac{1}{V_t} & =  \begin{cases}
-\frac{h}{2}q_{kb} & \text{ if } q_{kb} \geq \frac{1}{m} \\
-\frac{h}{2}q_{kb} + \frac{h}{2m} & \text{ otherwise }
-\end{cases}\\
-& \geq \frac{h}{2m} \; \forall q_{kb}
-}
-
-\eqn{
-P\left(\sum_{t=1}^n w_t \hat{\mu}_{a(t)}^{ij} - \mu^{ij} \geq \frac{4}{3}\sqrt{\frac{4m\log(1/\delta)}{h}}\right) \leq \delta \; \forall (i,j)
-}
-
-Lower bound?
-
-\pagebreak
-
-\subsection{Proofs}
-\label{seq:proof_expect_var}
-\subsubsection{Expectation and variance of the estimators}
-\eq{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-If $a = i$:
-\eq{
-E[\hat{\mu}_a^{ij}] = \frac{1}{p_{ij}}E[Y\ind{X_i=j}] = \frac{1}{p_{ij}}P(Y,do(X_i=j)) = \frac{1}{p_{ij}}P(Y|X_i=j)P(do(X_i=j)) = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2  = \frac{1}{p_{ij}^2}E[Y\ind{X_i=j}] - (\mu^{ij})^2  = \frac{1}{p_{ij}}\mu^{ij} - (\mu^{ij})^2 = \frac{\mu^{ij}(1-p_{ij}\mu^{ij})}{p_{ij}} \leq \frac{\mu^{ij}}{p_{ij}} \\
-\leq & \frac{1}{p_{ij}}
-}
-
-if $a \neq i$
-
-\eq{
-E[\hat{\mu}_a^{ij}] & = \frac{q_{a0}}{q_{ij}p_{a0}}E[Y\ind{X_i=j,X_a=0}]+\frac{q_{a1}}{q_{ij}p_{a1}}E[Y\ind{X_i=j,X_a=1}] \\
-& = \frac{q_{a0}}{q_{ij}p_{a0}}P(Y|X_i=j,do(X_i=0))q_{ij}p_{i0}+\frac{q_{a1}}{q_{ij}p_{a1}}P(Y|X_i=j,do(X_i=1))q_{ij}p_{i1} \\
-& =P(Y|X_i=j,do(X_i=0))P(X_i=0)+P(Y|X_i=j,do(X_i=1))P(X_i=1)\\
-& = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2 \\
- = & E[\ind{Y=1,X_i = j,X_a=0}](\frac{q_{a0}}{q_{ij}p_{a0}})^2 + E[\ind{Y=1,X_i = j,X_a=1}](\frac{q_{a1}}{q_{ij}p_{a1}})^2 \\
-& +  2 E[\ind{Y=1,X_i = j,X_a=0}\ind{Y=1,X_i = j,X_a=1}\frac{q_{a0}}{q_{ij}p_{a0}}\frac{q_{a1}}{q_{ij}p_{a1}}] -  (\mu^{ij})^2\\
-=& P(Y|X_i=j,X_a=0)q_{ij}p_{a0}(\frac{q_{a0}}{q_{ij}p_{a0}})^2 + P(Y|X_i=j,X_a=1)q_{ij}p_{a1}(\frac{q_{a1}}{q_{ij}p_{a1}})^2] - (\mu^{ij})^2 \\
-=& \frac{1}{q_{ij}}\left(P(Y|X_i=j,X_a=0)\frac{q_{a0}^2}{p_{a0}} + P(Y|X_i=j,X_a=1)\frac{q_{a1}^2}{p_{a1}}\right)- (\mu^{ij})^2  \\
-\leq & \frac{1}{q_{ij}}\left(\frac{q_{a0}^2}{p_{a0}} + \frac{q_{a1}^2}{p_{a1}}\right) \\
-}
-
-\subsubsection{Bernstein Bound}
-\begin{theorem}{Berstein's Inequality}
-Let $X_t \in \{X_1 ... X_n\}$  be independent, bounded random variables with mean $\mu$ and variance $V_t$ satisfying $|X_t - \mu| \leq b \; \forall t$. Let $V= \frac{1}{n}\sum_{t=1}^n V_t$ 
-\eq{
-& P(\sum_{t=1}^n X_t - \mu \geq \epsilon) \leq \exp(-\frac{\epsilon^2}{2nV+2b\epsilon/3})\\
-\implies & P\left(\sum_{t=1}^n X_t - \mu \geq \frac{2b\log(1/\delta)}{3} + \sqrt{2nV\log(1/\delta)}\right) \leq \delta
-}
-\end{theorem}
-Let $Z_t = w_t V_t$ where $w_t = V_t^{-1}/\sum V_t^{-1}$.
-\eq{
-V[Z_t] = \frac{V_t^{-1}}{(\sum V_t^{-1})^2}\\
-V = \frac{1}{n}\sum V[Z_t] = \frac{1}{n\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2b_z\log(1/\delta)}{3} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta, \text{ where } b_z = max_t \frac{V_t^{-1}b}{\sum V^{-1}}
-}
-
-While we are observing, $|\hat{\mu}_{kb}| = |\frac{Y\ind{X_k=b}}{q_{kb}}|\leq \frac{1}{q_{kb}}$ and $V_t^{-1} = q_{kb}$ so $V_t^{-1}b = 1$.
-
-When we are explicitly setting $X_k=b$, $|\hat{\mu}_{kb}| = |Y|\leq 1$ and $V_t^{-1} =1$ so $V_t^{-1}b = 1$
-
-\eq{
-b_z = \frac{1}{\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2\log(1/\delta)}{3\sum V_t^{-1}} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta \\
-\implies &  P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\subsection{Notes}
-Based on equation \ref{eqn:marginalizedProbability}, a natural estimator for $\mu^{ij}$, given we acted to set the variable $a$, would be
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{\ind{X_i=j}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{\ind{X_i=j,X_a=0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{\ind{X_i=j,X_a=1}}q_{a1} & \text{ if } a\neq i
-\end{cases}
-}
-
-However, this forces us to deal with random variables (which may be 0) in the denominator.
-We can avoid this issue by using the importance weighted estimator, which replaces these indicators with their expectation.
-
-\eqn{
-\hat{\mu}_a^{ij} & = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{q_{ij}p_{a0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{q_{ij}p_{a1}}q_{a1} & \text{ if } a\neq i
-\end{cases} \\
-&= \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-\subsection{Previous attempt}
-
-For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$. We will pwhere
-
-\eqn{
-m_{a,b}^{ij} &= \sum_{s \in \{t:I_t=(a,b)\}} Z_{ab,s}^{ij} \text{ and,}\\
-Z_{ab,s}^{ij} &= \ind{X_{i,s} = j,Y_s=1} \in \{0,1\}
-}
-
-TODO check/show this estimator is unbiasedool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{a=1}^K w_{a} \hat{\mu}_{a}^{ij} \;\;\text{, where } \sum_{a=1}^K w_a = 1
-}
-
-Putting everything together,
-
-\eqn{
-\hat{\mu}^{ij} = \frac{w_i}{\tau_{ij}}\sum_{s \in \{t: I_t = (i,j)\}} Z_{ij,s}^{ij} + \sum_{a \neq i} \left(\frac{w_a}{q_i(j)}\left[ \frac{q_a}{\tau_{a1}}\sum_{s \in \{t:I_t=(a,1)\}} Z_{a1,s}^{ij}+  \frac{1-q_a}{\tau_{a0}}\sum_{s \in \{t:I_t = (a,0)\}} Z_{a0,s}^{ij} \right] \right)
-}
-
-We now need to show $E[\hat{\mu}^{ij}] = \mu^{ij}$ and get a high probability bound for their difference. For the latter, lets try and use McDiarmids Inequality.
-
-McDiarmid's Inequality states:
-If $Z_i \ci Z_j$ and 
-\eqn{
-|\phi(Z_1...Z_i...Z_N) - \phi(Z_1...Z_i^{'}...Z_N)| < c_i \;\; \forall i
-}
-\eqn{
-\label{eqn:McDiarmid}
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq 
-\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
-\right) \leq \delta
-}
-
-For our problem, $\hat{\mu}^{ij} = \phi(\boldsymbol{Z}^{ij})$. All the $Z's$ are independent as they correspond indicator functions over the results at different timesteps.
-
-\eqn{
-|\phi(...Z_{ij,s}^{ij}...) - \phi(...Z_{ij,s}^{'ij}...)| & \leq \frac{w_i}{\tau{ij}}\;\;\;\; \leftarrow \tau_{ij} \text{ such $Z$'s}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a q_a}{q_i(j)\tau_{a1}} \;\;\;\; \leftarrow \tau_{a1} \text{ such $Z$'s for each $a$}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a (1-q_a)}{q_i(j)\tau_{a0}} \;\;\;\; \leftarrow \tau_{a0} \text{ such $Z$'s for each $a$}
-}
-
-\eqn{
-\label{eqn:sum_ci}
-\sum_l c_l^2 = & \frac{w_i^2}{\tau_{ij}} + \sum_{a \neq i}\frac{w_a^2}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}} +\frac{(1-q_a)^2}{\tau_{a0}}  \right)\\
-= & \sum_{a=1}^K w_a^2 f(a) \text{, where} \\
-f(a) =& \begin{cases}
-\frac{1}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}}+  \frac{(1-q_a)^2}{\tau_{a0}}\right)  & a \neq i\\
-\frac{1}{\tau_{ij}} & a = i
-\end{cases} 
-}
-
-We want to select weights to minimize equation \ref{eqn:sum_ci} so as to achieve as tight a bound as possible in equation \ref{eqn:McDiarmid}.
-
-Applying Legrange Multipliers
-
-\eqn {
-w_a & = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
-\label{eqn:min_sum_ci}
-\sum_i c_i^2 & = \frac{1}{\sum_a \frac{1}{f(a)}}
-}
-
-Substitution \ref{eqn:min_sum_ci} back into \ref{eqn:McDiarmid}
-
-\eqn {
-& P\left(|\hat{\mu}^{ij} - \mu^{ij}| \geq \epsilon \right) \leq 2 \exp \left(-2 \epsilon^2 \sum_{a=1}^K \eta_a^{ij} \right)\\
-& \eta_a^{ij} = \begin{cases}
-\frac{\tau_{a1} \tau_{a0} q_i(j)^2}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2} & a \neq i\\
-\tau_{ij} & a = i
-\end{cases} \\
-}
-
-Substituting this into \ref{eqn:probDeltaToLarge}
-
-\eqn{
-\label{eqn:targeted_prob_delta_too_big}
-P(\Delta_{\hat{i^*}} > D) \leq 2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right)
-}
-
-\eqn{
-R_2 \leq (T-h)\left(D+2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right) \right)
-}
-
-Looking at equation \ref{eqn:targeted_prob_delta_too_big}, let $A = \frac{D^2}{2}$ and
-
-\eqn{
-c_a =  \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) := \frac{1}{2}P(\Delta_{\hat{i^*}} > D)  =  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \sum_{a \neq i} c_a  )}+\sum_{i=1}^N e^{-A ( \tau_{i0} + (1-q_{i})^2 \sum_{a \neq i} c_a)}
-}
-
-The goal is to find some assignment $\boldsymbol{\tau}(\boldsymbol{q})$ to minimize this subject to the constraint that $\sum_i \tau_{i1} + \sum_i \tau_{i0} = h$
-
-No luck attempting to do this directly (via Lagrange multipliers ...)
-
-Note, letting $\tau_a = \tau_{a1}+\tau_{a0}$ then $c_a$ is maximized to equal $\tau_a$ if $\tau_{a1} = q_a \tau_a$
-
-Consider the case where we have only 2 variables.
-
-\eqn{
-f(\boldsymbol{\tau}) = e^{-A(\tau_{11} + q_1^2 c_2)}+
-e^{-A(\tau_{21} + q_2^2 c_1)}+
-e^{-A(\tau_{10} + (1-q_1)^2 c_2)}+
-e^{-A(\tau_{20} + (1-q_2)^2 c_1)}
-}
-
-If we consider the extreme case where $q_1=q_2=1$, ie the arms $X_1=0$ and $X_2=0$ never occur naturally.
-
-\eqn{
-f(\boldsymbol{\tau}) = 2 e^{-A(\tau_{11} + \tau_{21})}+
-e^{-A(\tau_{10})}+
-e^{-A(\tau_{20})}
-}
-
-Which is minimized subject to the constraint $\tau_{11}+\tau_{10}+\tau_{21}+\tau_{20}=h$ if we let $\tau_{10}=\tau_{20} = \frac{Ah-log(2)}{3A}$.
-Note that as $h$ gets large, we play the arms that do not occur naturally roughly $\frac{1}{3}$ of the time as opposed to $\frac{1}{4}$ if we were to play uniformly.
-We only learn anything about the arms that do not naturally occur when we play them directly.
-For the other arms we learn something all the time (provided we have some level of balance in all the other arms...) so it makes sense to play the arms that don't occur naturally a little bit more to keep the overall bounds tight.
-
-Ok, what happens if we just let $\tau_{ij} = \frac{h}{K}$
-
-\eqn {
-c_a = \frac{h}{K}\left(\frac{1}{(1-q_a)^2+q_a^2} \right) \geq \frac{h}{K}
-}
-
-\eqn {
-f \leq \sum_{i=1}^N \left(e^{-A(\frac{h}{K}+q_i^2\frac{(K-1)}{K}h)}+ e^{-A(\frac{h}{K}+(1-q_i)^2\frac{(K-1)}{K}h)}\right) \\
-\leq \sum_{i=1}^N \left(e^{-Ah}+ e^{-A(\frac{h}{K})}\right)
-}
-
-Basically worst case, we learn about half the arms fast.
-
-More generally, it would seem by targeted sampling we should be able to learn all the arms with reasonable natural probability fast - and the remaining arms at the rate $h/K$ (or better if we play them more often).
-
-Approaches to finding some better sampling function.
-
-\begin{enumerate}
-\item Find something motivated by the two variable example and show that it performs better than the uniform sampling approach.
-\item  Consider some other edge cases 
-\end{enumerate}
-
-Lets consider the case $q_1 = 1, q_2...q_N = \frac{1}{2}$ We will have $\tau_{11},\tau_{10},\tau_{ij} = \tau \; \forall i > 1,j$
-
-\eqn {
-c_a = \begin{cases}
-\tau_{11} & \text{ if } a=1\\
-2 \tau & \text{ otherwise}
-\end{cases}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) & = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\tau + \frac{1}{4}(\tau_{11}+2(N-2)\tau))} \\
-& = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})} \\
-& <  e^{-A\tau_{10}} + (2 (N-1)+1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}\\
-& =  e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}
-}
-
-and the constraint is $\tau_{11}+\tau_{10}+2(N-1)\tau = h$
-
-If we maximize $(\frac{\tau_{11}}{4} + \frac{N\tau}{2})$ subject to $\tau_{11}+2(N-1)\tau = (h-\tau_{10})$ we find $\tau = \frac{h-\tau_{10}}{2(N-1)}$ and $\tau_{11} = 0$.
-Note, as $N$ gets large the sensitivity to the trade off between $\tau$ and $\tau_{11}$ becomes negligible.
-
-\eqn {
-\frac{\tau_{11}}{4} + \frac{N\tau}{2} = \begin{cases}
-\frac{(h-\tau_{10})N}{4(N-1)} & \text{if } \tau_{11}=0 \\
-\frac{(h-\tau_{10})N}{4(N-1)} - \frac{1}{4(N-1)} &\text{if } \tau_{11}=(h-\tau_{10})
-\end{cases}
-}
-
-\eqn {
-f(\boldsymbol{\tau}) & <e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{1}{4}(h-\tau_{10}))}
-}
-
-Which is minimized if 
-\eqn {
-\tau_{10} = \frac{h}{5} - \frac{4}{5A}log(\frac{2N-1}{4})
-}
-
-Ok - lets look at the reverse case, where $q_1=\frac{1}{2}, q_2...q_N = 1$
-
-\eqn{
-\sum_a c_a = \begin{cases}
-(N-1)\tau_{11} & \text{ if } a = 1 \\
-2\tau+(N-2)\tau_{11} &\text{ otherwise }
-\end{cases}
-}
-
-The constraint is 
-\eqn {
-2\tau+(N-1)\tau_{11} + (N-1)\tau_{10} = h
-}
-
-\eqn{
-f(\boldsymbol{\tau}) = 2e^{-A(\tau+\frac{1}{4}(N-1)\tau_{11})} + (N-1)e^{-A(2\tau+(N-1)\tau_{11})}+(N-1)e^{-A\tau_{10}}
-}
-
-Holding $\tau_{10}$ constant, the 2nd and 3rd terms are constant. The first is maximized by letting $\tau_{11} = 0, \tau = \frac{h - (N-1)\tau_{10}}{2}$
-
-\eqn{
-f(\boldsymbol{\tau}) & = 2e^{-A\tau} + (N-1)e^{-2A\tau}+(N-1)e^{-A\frac{h - 2\tau}{N-1}} \leftarrow \text{ can't minimize analytically} \\
-\label{eqn:case1balancedapprox}
-& \leq (N+1)e^{-A\tau} +(N-1)e^{-A\frac{h - 2\tau}{N-1}}
-}
-
-which is minimized if
-
-\eqn {
-\tau & = \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N(N+1)}{2(N-1)})\right) \\
-\label{eqn:case1balancedapproxsolution}
-& \sim \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N}{2})\right)
-}
-
-Equation \ref{eqn:case1balancedapproxsolution} is also the solution that leads to the exponentiated terms in equation \ref{eqn:case1balancedapprox} being equal - which is should intuitively be optimal since the coefficients in front of them are very similar.
-
-Yet another case study - if $q_1...q_N = 1$ So none of the arms $X_i = 0$ occur naturally.
-
-\eqn{
-&c_a  = \tau_1 \; \forall a\\
-&\sum_{i \neq a}c_a  = (N-1)\tau_1 \; \forall a\\
-& N(\tau_0+\tau_1) = h\\
-& f(\boldsymbol{\tau}) = N e^{-A N\tau_1} + N e^{-A\tau_0}
-}
-
-minimized if 
-\eqn{
-\tau_0 = \frac{h}{N+1}-\frac{log(N)}{A(N+1)}
-}
-
-Note as $N$ get large - this equates to $\tau_0 \sim \frac{h}{N}$ ie almost all the actions are allocated to the arms that do not occur naturally.
-
-Another case study $q_1...q_{N_1} = 1,q_{N_1+1}...q_N = \frac{1}{2}$ (should encompass all previous examples ...)
-
-\eqn{
-&c_a  = \begin{cases}
-\tau_{11} &\text{ if } a \leq N_1
- \\
-2\tau & \text{otherwise}
-\end{cases}\\
-&\sum_{i \neq a}c_a  = \begin{cases}
-(N_1-1)\tau_{11} + 2 N_2 \tau & \text{ if } a \leq N \\
-N_1\tau_{11} + 2(N_2-1)\tau & \text{ otherwise }
-\end{cases}\\
-&N_1\tau_{10}+N_1\tau_{11}+2N_2\tau = h\\
-&f(\boldsymbol{\tau}) =  N_1 e^{-A\tau_{10}}+ N_1 e^{-A(N_1\tau_{11}+2N_2\tau)} + 2N_2 e^{-A(\frac{N_1}{4}\tau_{11}+\frac{N_2+1}{2}\tau)}
-}
-
-The 2nd and 3rd terms are minimized by letting $\tau_{11} = 0$ (although the result is only weakly dependnent on the trade off between $\tau_{11}$ and $\tau$ provided $N_2 >> 2$).
-
-\eqn{
-f(\boldsymbol{\tau}) &=  N_1 e^{-A\tau_{10}}+ N_1 e^{-A2N_2\tau} + 2N_2 e^{-A\frac{N_2+1}{2}\tau}\\
-& \leq N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A\frac{N_2+1}{2}\tau}\\
-& = N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A(\frac{(N_2+1)(h-N_1\tau_{10})}{4N_2})} 
-}
-
-\eqn{
-\tau_{10} = \left(\frac{N_2+1}{(N_2+1)N_1+4N_2}\right)h-\frac{4N_2\log(\frac{(N_2+1)(2N_2+N_2)}{4N_2})}{A((N_2+1)N_1+4N_2))}
-}
-
-It seems like this function can be optimized in a greedy way.
-That is, start with $\boldsymbol{\tau} = \boldsymbol{0}$, and in each timestep select the $\tau_{ij}$ that leads to the greatest reduction in the error bound.
-This appears to yield broadly consistent results to the analytic solution.
-
-\includegraphics[width=.5\textwidth]{greedy_minimization_close}
-\includegraphics[width=.5\textwidth]{greedy_minimization}
-
-Ok - so when we have many unbalanced variables, we assign almost all actions to the arms corresponding to the rare side.
-However, this will still be $< h/N$ which is only a factor of $2$ better than uniform assignment.
-Doing something smarter than uniform assignment will make more difference where we have a only a few unbalanced variables.
-In this case - smart assignment will lead to bounds almost as tight as the perfectly balanced case.
-Uniform assignment may yield solutions that are almost as bad as the worst case.
-(still better than without using the structure at all)
-
-I still need to cover more fractional $q$'s...
-
-What about if I repeated one of the above analysis but let the unbalanced q = .75?
-
-What can I say overall about worst/best cases/any symmetries in the problem??? Do these solutions lead to the terms in the exponentials being similar?
-
-Goal is to choose something that will give us a simple bound and will be ok for the worst case.
-Its still not entirely clear to me what the worst case will be - after all it depends on the way you assign to $\tau$.
-
-What happens if we assume $\sum_{a\neq i} c_a$ is constant ... this is reasonable if $N$ is large
-
-When you look at each term in $c_a$, its maximized if we play the arms according to their natural probabilities.
-If the arm is balanced, this will also minimize the sum of the two terms for doing $x_i = 0$ and $x_i = 1$.
-If the arm is unbalanced, then we have to trade of balancing this sum versus contributing to the sum of $c_a$.
-Consider just how bad the sum of $c_a$ can be even if we focus entirely on balancing the two terms ...
-maybe its already enough.
-
-Fundamentally the relationship between arms is complex. Its not that we get any direct link from one arm to another - rather pairs of actions tell us something about other arms.
-
-Ok, suppose we spend half our actions, $\frac{h}{2}$, observing - that is selecting variables on which to act uniformly and setting them according to their natural probabilities.
-This contributes $\frac{h}{2N}$ to each $c_a$.
-So $\sum_{i\neq a} c_a \sim \frac{h}{2}$ provided $N-1 \sim N$.
-We don't allocate any more pulls to the more likely arm associated with each variable (this is at most a factor of 2 two few pulls).
-We now decide how to spend the remaining budget.
-
-We can assume without loss of generality that $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq ... \leq q_N$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+\sum_{i=1}^N e^{-A (\frac{h}{8})}
-}
-
-Evenly balanced allocation is $\frac{h}{2N}$.
-For those arms for which $q_i^2 \geq \frac{1}{N}$ they will already get this much information during the observe phase so we will not allocate them any more (again - at worst a factor or 2 two few).
-Lets just allocate the remainder evenly.
-So let $m = min\{m:m\leq N,\frac{1}{m} \leq q_m \}$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+N e^{-A (\frac{h}{8})}
-}
-
-\pagebreak
-
-\subsection{Option 2: Some measure of information gain}
-
-\subsection{Option 3: Three stage algorithm}
-\begin{enumerate}
-\item Observe randomly
-\item Play low probability arms
-\item Pick best and exploit
-\end{enumerate}

### 2016-01-04 TPR_report.tex
Mutiple commits rolled together to fix large file that was accidentally committed.

@@
 \usepackage{graphicx}
+\usepackage{hyperref}
 \usetikzlibrary{arrows,positioning} 
@@
 \DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
-
+\DeclareUnicodeCharacter{00A0}{ }
 \tikzset{
@@
 This is known as a Markov decision process or MDP.
-A particularly well studied and understood model is MDPs with only a single state.
-In this case, there are a set of actions, each associated with a fixed but unknown reward distribution and at each time step our agent selects and action and receives corresponding feedback.
+A particularly well studied and understood model is the singe state MDP.
+In this case, there are a set of actions, each associated with a fixed but unknown reward distribution and at each time step our agent selects an action and receives corresponding feedback.
 This is known as the multi-armed bandit problem.
@@
 Both approaches can be seen as extensions to the concept of randomised controlled trials.
-Bandit algorithms deal with the sequential nature of the decision making process.
-Causal inference with the problem that full randomisation is not always feasible, affordable or ethical.The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
+Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem that full randomisation is not always feasible, affordable or ethical.
+The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
 The goal of my thesis is to explore these questions.
@@
 A very powerful and general model that underlies much of the recent work in causality is the causal bayesian network.
-A causal Bayesian network, or directed acyclic graph (DAG), is a Bayesian network in which a link $V_{i} \rightarrow V_{j}$ is defined to mean $V_{i}$ directly causes $V_{j}$.
+A causal bayesian network, or directed acyclic graph (DAG), is a bayesian network in which a link $V_{i} \rightarrow V_{j}$ is defined to mean $V_{i}$ directly causes $V_{j}$.
 This means that if we intervene and change the value of $V_{i}$, we expect $V_{j}$ to change, but if we intervene to change $V_{j}$, $V_{i}$ will not change.
@@
 A general causal query $P(Y|do(X=x))$ is identifiable if it can be shown to be equivalent to an expression containing only pre-interventional quantities.
-This means that, asymptotically, we obtain an unbiased estimate for the distribution after an intervention based on purely observational data.
+This means that, asymptotically, we can obtain an unbiased estimate for the distribution after an intervention based on purely observational data.
 
@@
 These rules are complete.
-A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus \cite{Shpitser2006,Huang2006} and there is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {Shpitser2008} (see XXXX for a javascript demonstration of this algorithm).
+A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus \cite{Shpitser2006,Huang2006}.
+There is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {Shpitser2008} (see \url{http://finnhacks42.github.io/causal_identify} for a javascript demonstration of this algorithm).
 If a query is not identifiable, it may still be possible to get bounds for causal effects, for example using instrumental variables \cite{Angrist2009} or by making additional assumptions.
@@
 There are differences between the models we have considered when it comes to these non-interventional queries.
-Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.
+Counterfactuals are not defined in causal bayesian networks, as they only encode information on the interventional distribution over variables.
 Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
@@
 In practice, differences in focus and approach across different fields eclipse the actual differences in the models.
-The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
+The work on causal graphical models \cite{Pearl2000,Sprites2000} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
 The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment.
@@
 This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
-An implementation is available in the R package \cite{Kalisch2012}
+An implementation is available in the R package pcalg \cite{Kalisch2012}
 
@@
 They take as input a dataset where each row of data is itself a dataset in which either $X \rightarrow Y$ or $Y \rightarrow X$ and a corresponding label.
-Estimates of the distributions $P(X)$, $P(Y)$ and $P()$ for each row are then mapped to features in some kernel space via mean kernel embeddings and finally a standard classification algorithm can be trained to learn label.
-New datasets where the direction of causality is unknown are then simply mapped to the kernel space and the causal direction is classified according to the trained classifier.
-In practice, the classifier is trained mostly on simulated data, as it is difficult to find a sufficient set of causal problems with only two variables where the direction of causality is known.
+Estimates of the distributions $P(X)$, $P(Y)$ and $P(X,Y)$ for each row are then mapped to features in some kernel space via mean kernel embeddings and finally a standard classification algorithm can be trained to learn the labels.
+New datasets, where the direction of causality is unknown, are then simply mapped to the kernel space and the causal direction is classified according to the trained classifier.
+In practice, the classifier is trained mostly on simulated data as it is difficult to find a sufficient set of causal problems with only two variables, where the direction of causality is known.
 
@@
 In its classic formulation \cite{Robbins1952} the (stochastic) K-armed bandit describes a sequential decision making problem, with $K$ possible actions or arms.
-Each arm $i$ is associated with a fixed but unknown reward distribution $\nu_i$.
+Each arm $i$ is associated with a fixed but unknown reward distribution $\nu_i$ \footnote{In order to obtain regret bounds, some assumptions are required on the distributions $\nu_i$.
+It sufficient to assume they are sub-gaussian}.
 For each timestep $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$.
@@
 
+Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
+In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
+The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm, $R_s(T) = \mu^* - \E{\mu_i}$ .
+This problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialization or exploitation phase.
+For example, many strategies might be assessed via simulation prior to one being selected and deployed.
+The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}.
+
 The classic multi-armed bandit is a powerful tool for sequential decision making.
 However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example linear bandits, dependent bandits, X-armed bandits and gaussian process bandits, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalized linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
 
@@
 This is an expansion of the model shown in figure \ref{fig:unify_frameworks}.
-We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
+We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
 
@@
 Comparing this to the standard bandit regret, $R_T = \bigo{\sqrt{TK}}$, we expect the causal algorithm to do better if $m < \frac{K^{3/2}}{\sqrt{T}}$.
+We can also show that the simple regret $R_s \in \bigotilde{\sqrt{m/T}}$.
+If $m << K$, this represents a significant improvement over standard bandit setting.
 \footnote{ $\bigotilde{.}$ and $\bigthetatilde{.}$ suppress polylogarithmic factors}
@@
 In this algorithm, we have only used the side information provided by the $do()$ action about other actions.
-Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
+Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
 The feedback graph is distinct from the causal graph.
@@
 
+\section{Paid work statement}
+I am not currently undertaking or planning to undertake any paid work.
+
 \bibliographystyle{unsrt}% Select the citation style e.g. ieeetr

### 2016-01-04 causal-bandit/icml-2015/causal-bandits-icml.tex
Mutiple commits rolled together to fix large file that was accidentally committed.

@@
 \usepackage{cleveref}
+
+\graphicspath{ {figures/} }
 \usetikzlibrary{arrows,positioning} 
@@
 
+There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
+Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
+There is also an active line of research into bandits with feedback defined by a graph.
+Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
+The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
+
+Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
+It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
+Partial monitoring algorithms yield results that are optimal with respect the the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
+
+ALSO NEED TO MENTION ANY OTHER COMBINATIONS OF BANDITS+CAUSAL (eg the Elias NIPS paper and Generalized Thompson Sampling paper)
+
+Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
+Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
+(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
+It is easy to come up with a counter example in which it does not).
+
 \section{Problem setup}
@@
 Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
-We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
+We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$ \footnote{A note on notation: In the bandit community it is implicit that algorithms selecting actions are intervening in the system.
+So it is sufficient to index actions according to the variable and value.
+However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from intervening to set that variable.
+Although in the specific causal graph we consider, observation and intervention are the same, we deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the bandit and causal inference communities.} , or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
 
@@
 \label{eq:estimation_transfer}
-P(Y_t|do(X_{i,t} = j))=\\
- \sum_{j'} P(Y_t|do(X_{l,t} = j'),X_{i,t} = j)P(X_{l_t} = j')
+P(Y_t|do(X_{i,t} = j))= & \nonumber \\
+ \sum_{j'}  P(Y_t|do(X_{l,t} & = j'),X_{i,t} = j)P(X_{l_t} = j')
 }
@@
 A full proof is given in the supplementary materials.
-(\color{red} or should we put it here?
-\color{black})
 
@@
 
-\section{Related Work}
-There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
-Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
-There is also an active line of research into bandits with feedback defined by a graph.
-Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
-The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
+\begin{figure}
+\caption{Final regret versus number of variables $N$ for UCB with $\alpha = 2$, Causal-Explore-Exploit with $m=2$ and with $m=N$ and horizon $T = 10,000$ .
+Error bars show standard deviation over 100 simulations.
+The regret for UCB grows linearly with the number of variables, whist for Causal-Explore-Exploit with fixed $m$, the growth is sub-logarthmic.
+}
+\label{fig:known_q_r_vs_N}
+\centering
+\includegraphics[width=.5\textwidth]{exp_regret_vs_N_T10000_sims100_20151229_113550.pdf}
+\end{figure}
+
+\begin{figure}
+\caption{Cumulative regret over time for $N = 17$ for UCB with $\alpha=2$, Causal-Explore-Exploit with $m=2$ and Causal-Explore-Exploit with $m=N$.
+Shaded region shows standard deviation over 100 simulations.
+The Causal-Explore-Exploit algorithm incurs linear regret during the exploration phase, after which it selects the optimal arm with high probability.
+For $m=2$, we have $K \sim m^{2/3}T^{1/3}$ and see that we are in the regime in which Causal-Explore-Exploit outperforms UCB.}
+\label{fig:known_q_r_vs_t}
+\centering
+\includegraphics[width=.5\textwidth]{exp_regret_vs_t_T10000_N17_sims100_20151229_120647.pdf}
+\end{figure}
+
+\section{Discussion}
 
@@
 
-Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
-Partial monitoring algorithms yield results that are optimal with respect the the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
-
-ALSO NEED TO MENTION ANY OTHER COMBINATIONS OF BANDITS+CAUSAL (eg the Elias NIPS paper and Generalized Thompson Sampling paper)
-
-Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
-Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
-(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
-It is easy to come up with a counter example in which it does not).
-
+\section{Future Open Questions}
+\begin{itemize}
+\item Known but arbitrary structure
+\item Learning structure then exploiting
+\end{itemize}
 \section{Conclusion}

### 2016-01-04 causal-bandit/lowerbound_auer.tex
Mutiple commits rolled together to fix large file that was accidentally committed.

@@
 \kl{P_{unif}}{P_i} & = kl(\frac{1}{2}||\frac{1}{2}+\epsilon)\left(\Eu{N_i}+\sum_{t=1}^T\Pu{i_t=0}P(i \in D_t)\right)\\
-&\geq \begin{cases}
-kl(\frac{1}{2}||\frac{1}{2}+\epsilon)\left(\Eu{N_i}+\frac{1}{m}\Eu{N_0}\right) & \text{ if } i \notin M \\
-kl(\frac{1}{2}||\frac{1}{2}+\epsilon)\Eu{N_i} & \text{ otherwise }
+& = kl(\frac{1}{2}||\frac{1}{2}+\epsilon)\left(\Eu{N_i}+q_i\Eu{N_0}\right)\\
+\implies \Ei{N_i} & \leq \Eu{N_i} + \sqrt{2}T\epsilon \sqrt{\Eu{N_i}+q_i\Eu{N_0}} \qquad \text{assuming }  \epsilon < \frac{2}{5}
+}
+Let $\Es{N_i} = \sum_{i=1}^K w_i\Ei{N_i}$
+
+\eqn {
+\Es{N_i} & \leq \sum_i w_i\Eu{N_i} + \sqrt{2}T\epsilon \sqrt{\sum_i w_i\Eu{N_i}+\Eu{N_0}\sum_i w_i q_i} 
+}
+
+Choose the weights such that $P(i \in M) = \frac{1}{2}$ ie, there is a 50\% chance of choosing an environment in which the optimal arm is in the set $M$.
+\eqn {
+w_i = \begin{cases}
+\frac{1}{2(m-1)} & \text{ if } i \leq m \\
+\frac{1}{2(K-m+1)} & \text{ otherwise}
 \end{cases}
@@
 
+Now $\sum_i w_i\Eu{N_i} = \Eu{\frac{1}{2(m-1)}\sum_{i=1}^{m-1}N_i+\frac{1}{2(K-m+1)}\sum_{i=m}^{K}N_i} \leq \frac{T}{2(m-1)}$, as  $m-1 < K-m+1$, since for every unbalanced arm in $m$ its partner is balanced.
+
+\eqn {
+\Es{N_i} & \leq \frac{T}{2(m-1)} + \sqrt{2}T\epsilon \sqrt{\frac{T}{2(m-1)}+\Eu{N_0}\sum_i w_i q_i}
+}
+
+\eqn{
+\sum_i w_i q_i & \leq \frac{1}{2(m-1)}\sum_{i < m} \frac{1}{m} + \frac{1}{2(k-m+1)}\sum_{i \geq m} 1 = \frac{1}{2}\left(\frac{1}{m} +1\right)
+}
+
 \subsection{Bounding regret}
-Assuming $ \epsilon < \frac{2}{5}$,
+
+We can now assume that the algorithm doesn't select $N_0$ more than $\epsilon T(1-\frac{1}{2(m-1)})$ times, since we can truncate all algorithms at that point and then play arms from $i \in M$ at random.
+
+\eqn {
+R(T) &= \epsilon  \left(T -\Es{N_i}\right) + \frac{1}{2}\Es{N_0}\\
+& \geq \epsilon \left (T - \frac{T}{2(m-1)} - \sqrt{2}T\epsilon \sqrt{\frac{T}{2(m-1)}+\frac{1}{2}\left(\frac{1}{m} + 1\right)\Eu{N_0}} \right)\\
+& \geq \frac{\epsilon}{3} \left (T - \frac{T}{2(m-1)} - \sqrt{2}T\epsilon \sqrt{\frac{T}{2(m-1)}+\frac{1}{2}\left(\frac{1}{m} + 1\right)\epsilon T (1 - \frac{1}{2(m-1)})} \right)\\
+& = \frac{\epsilon}{3} \left (T - \frac{T}{2(m-1)} - \sqrt{2}T\epsilon \sqrt{\frac{T}{2(m-1)}+\epsilon T \frac{m(2m-1)-3}{4m(m-1)}} \right)\\
+& \sim \frac{\epsilon}{3} \left (T - \frac{T}{2(m-1)} - \sqrt{2}T\epsilon \sqrt{\frac{T}{2(m-1)}+\frac{\epsilon T}{2} } \right) \qquad \frac{m(2m-1)-3}{4m(m-1)} \leq \frac{9}{17} \sim \frac{1}{2}
+}
+
+No dependence on $K$ this just cannot work.
+
+\pagebreak
+
+\eqn {
+R(T) &= \epsilon  \left(T -\frac{1}{K}\sum_{i=1}^K\Ei{N_i}\right) + \frac{1}{2K}\sum_{i=1}^K\Ei{N_0} \qquad \leftarrow \text{as before}\\
+&\geq \epsilon \left(T - \frac{T}{K} - T\epsilon\sqrt{\frac{2}{K}\left(\sum_i \Eu{N_i} + \sum_{i\in M}\frac{1}{m}\Eu{N_0} + \sum_{i \notin M}\Eu{N_0} \right)}   \right)+\frac{1}{2}\Es{N_0}\\
+&= \epsilon \left(T - \frac{T}{K} - T\epsilon\sqrt{2\left(\frac{T}{K} + \frac{1}{m}\Eu{N_0}\right)}   \right)+\frac{1}{2}\Es{N_0}\\
+&\geq \frac{\epsilon}{3} \left(T - \frac{T}{K} - T\epsilon\sqrt{2T\left(\frac{1}{K} + \frac{\epsilon}{m}\right)}   \right) \qquad \text{assuming } \Eu{N_0} \leq \epsilon T
+}
+
 \eqn {

### 2016-01-10 causal-bandit/icml-2015/causal-bandits-icml.tex
many updates and restructering

@@
 % \usepackage[nohyperref]{icml2015} above.
+\usepackage{cleveref}
 \usepackage{hyperref}
@@
 \usepackage{amsthm}
-\usepackage{cleveref}
+
+\setlength{\marginparwidth}{10ex}%added by Yifan so that the todos can fit in the margin
+%\usepackage[obeyFinal]{todonotes}
+\usepackage{todonotes}
+
+\newcommand{\tinytodo}[2][]{\todo[size=\tiny]{#2}}
+%\newcommand{\tinytodo}[2][]{\todo[size=\tiny, #1]{\begin{spacing}{1.0}#2\end{spacing}}}
+\newcommand{\todot}[2][]{\tinytodo[color=blue!20, #1]{T: #2}} % Csaba
+\newcommand{\todof}[2][]{\tinytodo[color=red!20, #1]{F:\@#2}} % Roshan
 
@@
 \newcommand{\E}[1]{\mathbb E\left[#1\right]}
+\newcommand{\R}{\mathbb R}
 \newcommand{\Var}{\operatorname{Var}}
@@
 \begin{abstract} 
-
+\todot{This can use some work}
+We study the problem of learning an optimal intervention. Our approach combines the theory of multi-armed bandits
+and causal inference. The main contribution is the new setting as well as algorithms that minimise both the simple and cumulative regret.
 \end{abstract}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% INTRODUCTION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
-Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
+
+A classical problem in economics and operations research is how to choose interventions that maximise utility. 
+We study an idealised model of such interventions using a combination of the multi-armed bandit framework \citep{Robbins1952} and the theory/language of causal inference 
+\citep{Pearl2000}.
+
+The idea is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. In each season she can invest in a green house (to control
+the temperature), cover her plants with nets (to prevent loss to birds), use expensive fertilizers, or install a high-tech watering system. We will assume for
+simplicity that she does not have the resources to make more than one intervention. Of course, the watering system will only be useful if the rainfall is low and
+the nets will be useful if the bird population is large. These things are random events that change from season to season, which makes the impact of each intervention
+difficulty to measure. The goal of this paper is to design an optimal strategy for the farmer to learn which intervention will be most profitable.
+
+%Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
 For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
@@
 
-An an alternate approach to selecting actions is causal inference.
+An alternate approach to selecting actions is causal inference.
 Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
@@
 It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
-Partial monitoring algorithms yield results that are optimal with respect the the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
+Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
 
@@
 
-\section{Problem setup}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PROBLEM SETUP
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Problem Setup}
 
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
-We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$ \footnote{A note on notation: In the bandit community it is implicit that algorithms selecting actions are intervening in the system.
-So it is sufficient to index actions according to the variable and value.
-However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from intervening to set that variable.
-Although in the specific causal graph we consider, observation and intervention are the same, we deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the bandit and causal inference communities.} , or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
+\newcommand{\bernoulli}{\operatorname{Bernoulli}}
+\newcommand{\dirac}{\operatorname{Dirac}}
 
-As an example, consider a farmer wishing to optimize the yield of her crop.
-She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
-We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
-After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
-This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
-
+Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1},\ldots,X_{N}\}$ that independently cause a 
+target variable of interest $Y \in \R$ (see Figure \ref{fig:causalStructure}).
 \begin{figure}[h]
@@
 
-Let $\boldsymbol{q} \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known end point $T$:
-
-\begin{enumerate}
-\item The learner chooses $I_t \in \set{0...N}$ and $J_t \in \set{0,1}$, corresponding to setting the variable $X_{I_t}=J_t$, also denoted $do(X_{I_t} = J_t)$.
-Selecting $I_t = 0$ represents not intervening on any variable and simply observing, also denoted $do()$.
-
-\item The remaining variables take their values stochastically according to their natural probabilities $\boldsymbol{q}$
-
+The game proceeds over $T$ identical rounds (or time-steps).
+In each round $t$ the learner can choose either to do nothing or they can choose a variable $I_t \in \set{1,\ldots,N}$ and
+an intervention $J_t \in \set{0,1}$. After the learner has chosen an intervention they observe $X_{t,i} \in \set{0,1}$ for all $i$ where
 \eq{
-X_{i,t} = \begin{cases}
-\sim Bernoulli(q_i) &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
+X_{t,i} \sim \begin{cases}
+\dirac(J_t) & \text{if } I_t = i \\
+\bernoulli(q_i) & \text{otherwise}\,.
 \end{cases}
 }
+where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. Note that if the learner did not choose an intervention then $I_t$ is undefined
+and $X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
+Finally the learner observes the reward $Y_t = r(X_t) + \eta_t$ where 
+\eq{
+r : \set{0,1}^N \to \R
+}
+is arbitrary (and unknown) 
+and $\eta_t$ is a $1$-subgaussian noise
+term (with the distribution possibly dependent on $X_t$).
 
-\item The learner receives reward $Y_t \sim \Ber(r(\boldsymbol{X}_t))$ where $r:\set{0,1}^N \to [0,1]$ is unknown and arbitrary.
-\end{enumerate}
+The expected reward for intervening on variable $i$ by setting it to $j$ is defined by
+\eq{
+\mu_{i,j} 
+&= \E{r(X)|do(X_i = j)} \\
+&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k} \,. 
+}
+The optimal intervention is $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
+Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
 
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E{r(X)|do(X_i = j)}$. The optimal reward and action are denoted $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E {\sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right)}.
+It is worth mentioning that the problem may be treated as a multi-armed bandit with $2N$ arms, one corresponding to each intervention.
+As we shall shortly see, this approach is usually not practical because the resulting algorithms do not exploit the structure in
+the covariates.
+
+We consider two standard performance measures. The first is the cumulative regret, which measures the difference between the reward expected under
+the omnipotent strategy that knows the optimal action in advance and the expected reward of the learner. 
+\eqn{
+\label{eq:regret}
+R_T = T \mu^* - \E{\sum_{t=1}^T Y_t}\,.
 }
+The second performance measure is the simple regret, which measures the performance of the learner on the final round only.
+\eqn{
+\label{eq:regret-simple}
+r_T = \mu^* - \E{\mu_{I_T,J_T}}\,.
+}
+Both versions of the regret are useful in different circumstances. The first is most useful when the learner is truly online and can change their policy over time.
+The second is useful when the exploration budget is limited and a fixed policy should eventually be chosen. \todot{make this nicer}
 
-The problem could be treated as a classical multi-armed bandit with $K = 2N+1$ arms, yielding a regret $\bigo{\sqrt{TN}}$.
-However, this doesn't leverage the side information induced by the causal structure.
+\todot{this remark feels out of place, but where to put it?}
+\begin{remark}
+In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
+the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
+In the bandit community it is implicit that 
+algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
+However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
+intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
+deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
+bandit and causal inference communities.
+\end{remark}
 
-\section{Causal Bandits}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% UPPER BOUND ON SIMPLE REGRET
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Upper Bounds on the Simple Regret}
 
-Given the assumed causal structure, the probability of $Y_t$ given we intervene to set $X_{i,t} = j$ is the same as if we don't set any variables and observe $X_{i,t} = j$.
+In this section we develop and analyse an algorithm for minimising the simple regret in the
+causal bandit problem.
 
+From a causal perspective the key insight is that the assumed causal structure implies that
+the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
+on $X_{t,i} = j$. Formally,
 \eqn {
 \label{eq:observe}
-P(Y_t|do(X_{i,t} = j)) = P(Y_t|do(),X_{i,t} = j)
+\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
 }
+This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph. It is not the case in general. 
+For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 
+that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the 
+interventional distribution of $Y_t$ from the observational one.
 
-This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph.
-It is not the case in general.
-For example if there was a variable $X'$ that caused both $X_i$ and $Y$ (or $X_i$ and any other variable $X_l$), that would introduce a backdoor path from $X_i \rightarrow Y$ and we would have to condition on $X'$ to derive the interventional distribution of $Y$ from the observational one.
-
+\todot{Or just give a reference or hide in the appendix depending on space in the end}
 \color{red} Probably should define causal model and back-door rule to properly show this.\color{black}
@@
 \label{eq:estimation_transfer}
-P(Y_t|do(X_{i,t} = j))= & \nonumber \\
- \sum_{j'}  P(Y_t|do(X_{l,t} & = j'),X_{i,t} = j)P(X_{l_t} = j')
+P(Y_t|do(X_{t,i} = j))= & \nonumber \\
+ \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
 }
+
+The main idea of the algorithm is that by making no intervention (doing nothing) it is possible to estimate simultaneously the returns of all interventions.
+Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and should be
+explored separately by actually making the intervention. The algorithm works by optimally balancing the collection of observational data (when
+no action is taken) and making interventions to learn about low probability events. The problem is more challenging because $\boldsymbol{q}$ is
+unknown and must also be estimated.
+
+\begin{algorithm}[h]
+\caption{Causal Best Arm Identification}\label{alg:simple}
+\begin{algorithmic}[1]
+\STATE {\bf Input:} $T, N$
+\FOR{$t \in 1,\ldots,(T - 1) / 2$}
+\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
+\ENDFOR
+\STATE Estimate $\mu$ using observation data:
+\eq{
+(\forall i,j) \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{t,i} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{t,i} = j}}\,.
+}
+\STATE Compute $\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} X_{t,i}$
+\STATE Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
+\STATE Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
+\STATE Compute $\hat m = \min\set{1 \leq i \leq N : \hat s'_{i+1} \geq \frac{1}{i}}$
+\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
+\STATE Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
+
+\FOR{$(i,j) \in A$}
+\FOR{$t \in 1,\ldots,\frac{T - 1)}{2\hat m}$}
+\STATE Choose action $do(X_{t,i} = j)$ and observe $r_t$
+\ENDFOR
+\STATE Recompute $\hat \mu_{i,j} = \frac{2\hat m}{T} \sum_{t=1}^{ T/2\hat m} r_t(X_{t,i}=j)$ 
+\ENDFOR
+\STATE Estimated optimal action is $\hat i^*, \hat j^* = \argmax_{i,j} \hat \mu_{i,j}$
+\STATE Choose action $do(X_{t,\hat i^*} = \hat j^*)$
+\end{algorithmic}
+\end{algorithm}
+
+\begin{theorem}\label{thm:unknown_q_simpleregret}
+Define $m = \min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$.
+Then Algorithm \ref{alg:simple} satisfies
+\eq{
+r_T \in \bigo{\sqrt{\frac{m}{T}\log\left(\frac{NT}{m}\right)}}\,.
+}
+\end{theorem}
+
+Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega\left(\sqrt{N/T}\right)$.
+This is a significant improvement, since $m \leq N$ is usually substantially smaller than $N$.
+
+\todot{Do we have a proof for this in another doc?}
+\begin{proof}
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% UPPER BOUND ON CUMULATIVE REGRET
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Upper Bound on the Cumulative Regret}
+
+We now tackle the case where the objective is to minimise the cumulative regret.
+This is significantly more challenging since the algorithm can no longer explore naively for $O(T)$ rounds without suffering linear regret.
 
@@
    \STATE Let $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
-   \STATE Let $h = T^{2/3}m^{1/3}log(TK)^{1/3}$
+   \STATE Let $h = T^{2/3}m^{1/3} \log(TK)^{1/3}$
    \STATE Let $A = \set{(i,j) : i \leq m, j = 1}$ be the set of infrequently observed arms
@@
 Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
-Then \cref{alg:known_q} satisfies
+Then Algorithm \ref{alg:known_q} satisfies
 \eq{
-R(T) \in \bigo{T^{2/3}m^{1/3}log(KT)^{1/3}}\,.
+R(T) \in \bigo{T^{2/3}m^{1/3} \log(KT)^{1/3}}\,.
 }
@@
 
-\eqn{
-R^{simple}(h) = \E{\mu^* - max_{(i,j)}\hat{\mu}_{i,j}(h)}
-}
-
-\begin{algorithm}[h]
-\caption{Causal Best Arm Identification}\label{alg:simple}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} $T, N$
-\FOR{$t \in 1,\ldots,T/2$}
-\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
-\ENDFOR
-\STATE Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
-\eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{i,t} = j}}\,.
-}
-\STATE Compute $\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} X_{i,t}$
-\STATE Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\STATE Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\STATE Compute $\hat m = \min\set{1 \leq i \leq N : \hat s'_{i+1} \geq \frac{1}{i}}$
-\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\STATE Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
-
-\FOR{$(i,j) \in A$}
-\FOR{$t \in 1,\ldots,T/2\hat m$}
-\STATE Choose action $do(X_{i,t} = j)$ and observe $r_t$
-\ENDFOR
-\STATE Recompute $\hat \mu_{i,j} = \frac{2\hat m}{T} \sum_{t=1}^{ T/2\hat m} r_t(X_{i,t}=j)$ 
-\ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:unknown_q_simpleregret}
-Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$.
-Then \cref{alg:simple} satisfies
-\eq{
-R^{simple}(T) \in \bigo{\sqrt{\frac{m}{T}log\left(\frac{NT}{m}\right)}}\,.
-}
-\end{theorem}
-
-The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}.
-By utilizing the causal structure the dependence on $K$ is reduced to a dependence on $m$, which may be much smaller.
-The intuition behind \cref{alg:simple} is that we can use data collected during rounds where we select the $do()$ action to estimate $m$ and determine which arms we must play explicitly.
-
-For the bandit problem problem we need to collect some data to estimate $m$ before determining if we should use the causal explore-exploit algorithm or a standard bandit algorithm.
-
 \begin{algorithm}[h]
@@
 \STATE $T_1 = 48Nlog\left(4N/\delta\right)$ 
-\STATE Run \cref{alg:simple} to line 11 with input $T_1,N$.
+\STATE Run Algorithm \ref{alg:simple} to line 11 with input $T_1,N$.
 \IF {$\hat{m} > \frac{N^{3/2}}{\sqrt{T}}$}
@@
 \STATE $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
-\STATE Run \cref{alg:simple} with input $h,N$.
+\STATE Run Algorithm \ref{alg:simple} with input $h,N$.
 \STATE Compute $(\hat{i}^*,\hat{j}^*) = \argmax_{(i,j)} \hat{\mu}_{i,j}$
@@
 Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
-Then \cref{alg:bandit} satisfies
+Then Algorithm \ref{alg:bandit} satisfies
 \eq{
@@
 
-\subsection{Lower Bounds}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LOWER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Lower Bounds}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% EXPERIMENTS
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Experiments}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% DISCUSSION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
@@
 
+\bibliography{library}
+\bibliographystyle{icml2015}
 \end{document}

### 2016-01-10 causal-bandit/icml-2015/causal-bandits-icml.tex
minor fixes

@@
 
-Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega\left(\sqrt{N/T}\right)$.
-This is a significant improvement, since $m \leq N$ is usually substantially smaller than $N$.
+Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega(\sqrt{N/T})$.
+Since $m$ is typically much smaller than $N$ we anticipate that the new algorithm will significantly outperform classical bandit algorithms in
+this setting.
 

### 2016-01-11 causal-bandit/icml-2015/causal-bandits-icml.tex
Tweaked abstract and opening example.

@@
 %\newcommand{\tinytodo}[2][]{\todo[size=\tiny, #1]{\begin{spacing}{1.0}#2\end{spacing}}}
-\newcommand{\todot}[2][]{\tinytodo[color=blue!20, #1]{T: #2}} % Csaba
-\newcommand{\todof}[2][]{\tinytodo[color=red!20, #1]{F:\@#2}} % Roshan
+\newcommand{\todot}[2][]{\tinytodo[color=blue!20, #1]{T: #2}} % Tor
+\newcommand{\todof}[2][]{\tinytodo[color=red!20, #1]{F:\@#2}} % Finn
+\newcommand{\todom}[2][]{\tinytodo[color=green!20, #1]{M:\@#2}} % Mark
 
@@
 
+\newcommand{\etc}{\textit{etc}}
+\newcommand{\ie}{\textit{i.e.}}
+\newcommand{\eg}{\textit{e.g.}}
+
 \theoremstyle{plain}
@@
 \begin{abstract} 
-\todot{This can use some work}
-We study the problem of learning an optimal intervention. Our approach combines the theory of multi-armed bandits
-and causal inference. The main contribution is the new setting as well as algorithms that minimise both the simple and cumulative regret.
+We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
+Our formalism combines multi-arm bandits and causal inference to models a novel type of bandit feedback that is not exploited by existing approaches.
+We propose a new algorithm that can make use of the causal feedback and prove that it achieves bounds on both simple and cumulative regret that are strictly better (in dependence on the number of possible interventions and number of rounds) than algorithms which do not use the additional causal information.
 \end{abstract}
@@
 
-The idea is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. In each season she can invest in a green house (to control
-the temperature), cover her plants with nets (to prevent loss to birds), use expensive fertilizers, or install a high-tech watering system. We will assume for
-simplicity that she does not have the resources to make more than one intervention. Of course, the watering system will only be useful if the rainfall is low and
-the nets will be useful if the bird population is large. These things are random events that change from season to season, which makes the impact of each intervention
-difficulty to measure. The goal of this paper is to design an optimal strategy for the farmer to learn which intervention will be most profitable.
+The challenge is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. 
+She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
+In each season the farmer has enough time and money to intervene and control exactly one of these variables:
+deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed a through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
+When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
+How might the farmer best experiment to identify the single, highest yielding intervention without sacrificing too much crop yield in the process?
 
+% protect her plants from birds with nets, use expensive fertilizers, or deploy a high-tech watering system. We will assume for
+% simplicity that she does not have the resources to make more than one intervention per season and that each intervention only lasts a single season.
+Of course, the watering system will only be useful if the rainfall is low, the nets will be useful if the bird population is large, and fertilizing only helps when the soil is poor.
+% Rainfall, bird populations, and soil quality are extrinsic random events that change from season to season. This makes the impact of each intervention difficult to measure. 
+% Crucially however, the farmer is able to observe the rainfall, birds, and soil and their effect on yield regardless of whether an intervention was made to control for those factors.
+% Using these observations and the knowledge that water, birds, and soil are the only factors which cause the crop yield to change, how might the farmer identify the single most cost-effective intervention to make every season?
+% The goal of this paper is to design an optimal strategy for the farmer to solve this learning problem.
+
+%\todom{Other examples of why we would want to model this (e.g., medical).}
 %Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.

### 2016-01-11 causal-bandit/icml-2015/causal-bandits-icml.tex
minor fixes plus experiment details

--- 2016-01-10 causal-bandit/icml-2015/causal-bandits-icml.tex
@@
 
-The idea is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. In each season she can invest in a green house (to control
+The challenge is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. In each season she can invest in a green house (to control
 the temperature), cover her plants with nets (to prevent loss to birds), use expensive fertilizers, or install a high-tech watering system. We will assume for
@@
 
+\begin{remark}
+In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
+the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
+In the bandit community it is implicit that 
+algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
+However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
+intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
+deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
+bandit and causal inference communities.
+\end{remark}
+
 We consider two standard performance measures. The first is the cumulative regret, which measures the difference between the reward expected under
@@
 The second is useful when the exploration budget is limited and a fixed policy should eventually be chosen. \todot{make this nicer}
-
-\todot{this remark feels out of place, but where to put it?}
-\begin{remark}
-In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
-the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
-In the bandit community it is implicit that 
-algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
-However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
-intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
-deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
-bandit and causal inference communities.
-\end{remark}
 
@@
 
-We propose a simple explore-exploit based algorithm that leverages  \eqref{eq:observe}. Without loss of generality, we assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ... \leq q_N$.
+We begin by considering the case where $\boldsymbol{q}$ is known and propose a simple explore-exploit based algorithm that leverages  \eqref{eq:observe}.
+Without loss of generality, we assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
+\leq q_N$.
 
@@
 
-Algorithm \ref{alg:known_q} relies on $\boldsymbol{q}$ and thus $m$ being known.
-We now consider the case where $\boldsymbol{q}$ is unknown.
-We begin by considering the simple regret, defined as the expected difference between the mean payoff of the optimal action and that of the action estimated to be optimal in $T$th round.
+We now demonstrate we can obtain the same order of regret for the case where $\boldsymbol{q}$ is unknown.
 
@@
 \section{Experiments}
+In all experiments, the reward is generated according to:
+\eq{
+Y_t \sim \begin{cases}
+\bernoulli(.5+\epsilon) & \text{if } X_1 = 1 \\
+\bernoulli(\frac{.5-(.5+\epsilon)q_1}{1-q_1}) & \text{otherwise}\,.
+\end{cases}
+}
+
+With $\epsilon \in (0,.5)$. This ensures that 
+\eq {
+\P{Y|do(X_i = j)} = \begin{cases}
+.5+\epsilon & \text{if } (i,j)=(1,1) \\
+\frac{.5-(.5+\epsilon)q_1}{1-q_1} & \text{if } (i,j) = (1,0) \\
+.5 & \text{otherwise}
+\end{cases}
+}
+
+This allows us to rapidly simulate rewards and conditional rewards for large numbers of variables and to change the value of $\boldsymbol{q}$ with minimal impact on the reward function \footnote{$\boldsymbol{q}$ only effects the reward of a single, non-optimal arm, which becomes insignificant with large numbers of variables.}.
 

### 2016-01-12 causal-bandit/icml-2015/causal-bandits-icml.tex
Reworking intro

@@
 We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
-Our formalism combines multi-arm bandits and causal inference to models a novel type of bandit feedback that is not exploited by existing approaches.
+Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
 We propose a new algorithm that can make use of the causal feedback and prove that it achieves bounds on both simple and cumulative regret that are strictly better (in dependence on the number of possible interventions and number of rounds) than algorithms which do not use the additional causal information.
@@
 
+\todom{Describe feature of class of problem; highlight with examples.}
+
 The challenge is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. 
 She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
-In each season the farmer has enough time and money to intervene and control exactly one of these variables:
+In each season the farmer has enough time and money to intervene and control at most one of these variables:
 deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed a through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
 When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
-How might the farmer best experiment to identify the single, highest yielding intervention without sacrificing too much crop yield in the process?
+How might the farmer best experiment to identify the single, highest yielding intervention without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
 
-% protect her plants from birds with nets, use expensive fertilizers, or deploy a high-tech watering system. We will assume for
-% simplicity that she does not have the resources to make more than one intervention per season and that each intervention only lasts a single season.
-Of course, the watering system will only be useful if the rainfall is low, the nets will be useful if the bird population is large, and fertilizing only helps when the soil is poor.
-% Rainfall, bird populations, and soil quality are extrinsic random events that change from season to season. This makes the impact of each intervention difficult to measure. 
-% Crucially however, the farmer is able to observe the rainfall, birds, and soil and their effect on yield regardless of whether an intervention was made to control for those factors.
-% Using these observations and the knowledge that water, birds, and soil are the only factors which cause the crop yield to change, how might the farmer identify the single most cost-effective intervention to make every season?
-% The goal of this paper is to design an optimal strategy for the farmer to solve this learning problem.
+One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem~\cite{Robbins1952} -- each of the six possible interventions is an arm and the crop yield is the reward -- and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
+However, as we show in \S\ref{sec:results}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret with a sub-optimal dependence on the number of possible interventions.
 
-%\todom{Other examples of why we would want to model this (e.g., medical).}
+There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \cite{TODO} as well as results for partial monitoring which, as first glance, may appear applicable.
+As we show in Section~\ref{}, applying these results to our problem also exhibit a sub-optimal dependence on the number of available interventions.
+
+\todom{Rework rest of intro}
+
 %Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
@@
 
-A widely used framework for sequential decision making is the multi-armed bandit.
+% A widely used framework for sequential decision making is the multi-armed bandit.
 In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary.
@@
 We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
+
+\todom{Summarise results \& give intuition for algorithm}
 
@@
 term (with the distribution possibly dependent on $X_t$). 
+\todom{Connect to opening ex.}
 

### 2016-01-12 causal-bandit/icml-2015/causal-bandits-icml.tex
adds figures from simple regret experiment
adds figures

@@
 
+\begin{figure}
+\caption{Simple regret vs horizon, $T$, for $N = 30$ for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$.
+Error bars show estimated 95\% confidence interval from 1000 simulations.}
+\label{fig:simple_vs_T}
+\centering
+\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_T_N30_sims1000_20160113_010512.pdf}
+\end{figure}
+
+\begin{figure}
+\caption{Simple regret vs number of variables, $N$, for $T=500$, for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$.
+Error bars show estimated 95\% confidence interval from 1000 simulations.}
+\label{fig:simple_vs_N}
+\centering
+\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_N_T500_sims1000_20160113_010511.pdf}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-01-13 causal-bandit/icml-2015/lower-bound.tex
added lower bound proof

\newcommand{\simpleregret}{R_T^{\text{simple}}}

We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in \cref{thm:simple}

\begin{theorem}\label{thm:lower}
Suppose $\eta_t \sim \mathcal N(0, 1)$ and $\boldsymbol{q}$ satisfies $m(\boldsymbol{q}) = m$.
Then for all strategies there exists a reward function such that
\eq{
\simpleregret \geq \frac{\frac{m}{2e} - 1}{m} \sqrt{\frac{2m}{3T}} = \Omega\left(\sqrt{\frac{m}{T}}\right)\,.
}
\end{theorem}

\begin{proof}
Assume without loss of generality that $q_i \leq 1/2$ for all $i$ and that $q_1 \leq q_2 \leq \ldots \leq q_N$.
For each $i$ define reward function $r_i$ by
\eq{
r_0(\boldsymbol{X}) &= 0 &
r_i(\boldsymbol{X}) &= \begin{cases}
\epsilon & \text{if } X_i = 1 \\
0 & \text{otherwise}\,,
\end{cases}
}
where $\epsilon > 0$ is some constant to be chosen later.
We abbreviate $R_{T,i}^{\text{simple}}$ to be the expected simple regret incurred when interacting with the
environment determined by $\boldsymbol{q}$ and $r_i$. Let $P_i$ be the corresponding measure
on all observations over all $T$ rounds and $\mathbb E_i$ the expectation with respect to $P_i$. By Lemma 2.6 by \cite{Tsy08} we have
\eq{
\Prz{I_t = i} + \Pri{I_t \neq i} \geq \exp\left(-\KL(P_0, P_i)\right)\,,
}
where $\KL(P_0, P_i)$ is the KL divergence between measures $P_0$ and $P_i$.
Let $T_i(T) = \sum_{t=1}^T \ind{I_t = i}$ be the total number of times the learner intervenes on variable $i$ (with any intervention). 
Then for $i \leq m$ a computation shows that 
\eq{
\KL(P_0, P_i) 
&= \frac{\epsilon^2}{2} \mathbb E_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
&\leq \frac{\epsilon^2}{2} \left(\mathbb E_0 T_i(T) + q_i T\right) \\
&\leq \frac{\epsilon^2}{2} \left(\mathbb E_0 T_i(T) + \frac{T}{m}\right)\,.
}
Define set $A$ by
\eq{
A = \set{i \leq m : \mathbb E_0 T_i(T) \leq 2T / m}\,.
}
Then for $i \in A$ and choosing $\epsilon = \sqrt{2m/3T}$ we have
\eq{
\KL(P_0, P_i) \leq \frac{3T\epsilon^2}{2m} = 1\,. 
}
Now $\sum_{i=1}^m \mathbb E_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
Therefore
\eq{
\sum_{i \in A} \Pri{I_t \neq i} 
&\geq \sum_{i \in A} \exp\left(-\KL(P_0, P_i)\right) - 1\\
&\geq \frac{|A|}{e} - 1 
\geq \frac{m}{2e} - 1\,.
}
Therefore there exists an $i \in A$ such that
\eq{
\Pri{I_t \neq i} \geq \frac{\frac{m}{2e} - 1}{m}\,.
}
Therefore
\eq{
R^{\text{simple}}_{T,i} \geq \frac{1}{2} P_i(I_t \neq i) \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{2m}{3T}}
}
as required.
\end{proof}

### 2016-01-13 causal-bandit/icml-2015/causal-bandits-icml.tex
added lower bound proof
broke up main paper into separate tex

--- 2016-01-12 causal-bandit/icml-2015/causal-bandits-icml.tex
@@
 
-% use Times
 \usepackage{times}
-% For figures
 \usepackage{graphicx} % more modern
-%\usepackage{epsfig} % less modern
 \usepackage{subfigure} 
-
-% For citations
 \usepackage{natbib}
-
-% For algorithms
 \usepackage{algorithm}
@@
 
-% As of 2011, we use the hyperref package to produce hyperlinks in the
-% resulting PDF.  If this breaks your system, please commend out the
-% following usepackage line and replace \usepackage{icml2015} with
-% \usepackage[nohyperref]{icml2015} above.
 \usepackage{cleveref}
 \usepackage{hyperref}
-
-% Packages hyperref and algorithmic misbehave sometimes.  We can fix
-% this with the following command.
 \newcommand{\theHalgorithm}{\arabic{algorithm}}
-
-% Employ the following version of the ``usepackage'' statement for
-% submitting the draft version of the paper for review.  This will set
-% the note in the first column to ``Under review.  Do not distribute.''
 \usepackage{icml2015stylefiles/icml2015} 
-
-% Employ this version of the ``usepackage'' statement after the paper has
-% been accepted, when creating the final version.  This will set the
-% note in the first column to ``Proceedings of the...''
-%\usepackage[accepted]{icml2015}
-
 \usepackage{tikz}
@@
 \renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
+\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
 \newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
@@
 \newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
+\newcommand{\KL}{\operatorname{KL}}
 
@@
 
-A classical problem in economics and operations research is how to choose interventions that maximise utility. 
-We study an idealised model of such interventions using a combination of the multi-armed bandit framework \citep{Robbins1952} and the theory/language of causal inference 
-\citep{Pearl2000}.
-
-\todom{Describe feature of class of problem; highlight with examples.}
-
-The challenge is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. 
-She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
-In each season the farmer has enough time and money to intervene and control at most one of these variables:
-deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed a through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
-When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
-How might the farmer best experiment to identify the single, highest yielding intervention without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
-
-One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem~\cite{Robbins1952} -- each of the six possible interventions is an arm and the crop yield is the reward -- and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
-However, as we show in \S\ref{sec:results}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret with a sub-optimal dependence on the number of possible interventions.
-
-There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \cite{TODO} as well as results for partial monitoring which, as first glance, may appear applicable.
-As we show in Section~\ref{}, applying these results to our problem also exhibit a sub-optimal dependence on the number of available interventions.
-
-\todom{Rework rest of intro}
-
-%Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
-For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
-
-% A widely used framework for sequential decision making is the multi-armed bandit.
-In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary.
-At each timestep the agent selects an action and receives a reward sampled i.i.d from the corresponding reward distribution.
-The performance of bandit algorithms is described by the regret: the difference in the expected reward obtained by the algorithm and the reward that could be obtained if the optimal action was selected at every timestep.
-
-An alternate approach to selecting actions is causal inference.
-Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
-This allows an agent to predict the outcome of an action based on non-experimental data.
-This approach is common in social science, demography, and economics where explicit experimentation may be difficult.
-For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.
-
-We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
-
-\todom{Summarise results \& give intuition for algorithm}
-
-A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph.
-Each possible assignment of variables to values is an action (bandit arm).
-The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action.
-The number of actions grows exponentially with the number of variables in the graph, making it important to use algorithms that take account of the graph structure to reduce the search space.
-
-Problems framed in this way take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected.
-If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
-
-If we receive feedback on additional nodes the problem can be more interesting.
-In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
-
-We consider a bandit problem where the actions and reward are represented by a specific causal graph that demonstrates this interesting structure.
-We develop an algorithm to leverage the information provided by this structure and demonstrate it substantially outperforms standard bandit algorithms applied to the same problem where the number of actions is large.
-
-There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
-Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
-There is also an active line of research into bandits with feedback defined by a graph.
-Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
-The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
-
-Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
-Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
-
-ALSO NEED TO MENTION ANY OTHER COMBINATIONS OF BANDITS+CAUSAL (eg the Elias NIPS paper and Generalized Thompson Sampling paper)
-
-Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
-Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
-(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
-It is easy to come up with a counter example in which it does not).
+\input{introduction}
 
@@
 
-\newcommand{\bernoulli}{\operatorname{Bernoulli}}
-\newcommand{\dirac}{\operatorname{Dirac}}
-
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1},\ldots,X_{N}\}$ that independently cause a 
-target variable of interest $Y \in \R$ (see Figure \ref{fig:causalStructure}).
-\begin{figure}[h]
-\centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\end{figure}
-
-The game proceeds over $T$ identical rounds (or time-steps).
-In each round $t$ the learner can choose either to do nothing or they can choose a variable $I_t \in \set{1,\ldots,N}$ and
-an intervention $J_t \in \set{0,1}$. After the learner has chosen an intervention they observe $X_{t,i} \in \set{0,1}$ for all $i$ where
-\eq{
-X_{t,i} \sim \begin{cases}
-\dirac(J_t) & \text{if } I_t = i \\
-\bernoulli(q_i) & \text{otherwise}\,.
-\end{cases}
-}
-where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. Note that if the learner did not choose an intervention then $I_t$ is undefined
-and $X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
-Finally the learner observes the reward $Y_t = r(X_t) + \eta_t$ where 
-\eq{
-r : \set{0,1}^N \to \R
-}
-is arbitrary (and unknown) 
-and $\eta_t$ is a $1$-subgaussian noise
-term (with the distribution possibly dependent on $X_t$). 
-\todom{Connect to opening ex.}
-
-The expected reward for intervening on variable $i$ by setting it to $j$ is defined by
-\eq{
-\mu_{i,j} 
-&= \E{r(X)|do(X_i = j)} \\
-&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k} \,. 
-}
-The optimal intervention is $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
-Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
-
-It is worth mentioning that the problem may be treated as a multi-armed bandit with $2N$ arms, one corresponding to each intervention.
-As we shall shortly see, this approach is usually not practical because the resulting algorithms do not exploit the structure in
-the covariates.
-
-\begin{remark}
-In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
-the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
-In the bandit community it is implicit that 
-algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
-However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
-intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
-deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
-bandit and causal inference communities.
-\end{remark}
-
-We consider two standard performance measures. The first is the cumulative regret, which measures the difference between the reward expected under
-the omnipotent strategy that knows the optimal action in advance and the expected reward of the learner. 
-\eqn{
-\label{eq:regret}
-R_T = T \mu^* - \E{\sum_{t=1}^T Y_t}\,.
-}
-The second performance measure is the simple regret, which measures the performance of the learner on the final round only.
-\eqn{
-\label{eq:regret-simple}
-r_T = \mu^* - \E{\mu_{I_T,J_T}}\,.
-}
-Both versions of the regret are useful in different circumstances. The first is most useful when the learner is truly online and can change their policy over time.
-The second is useful when the exploration budget is limited and a fixed policy should eventually be chosen. \todot{make this nicer}
+\input{setup}
 
@@
 
-In this section we develop and analyse an algorithm for minimising the simple regret in the
-causal bandit problem.
-
-From a causal perspective the key insight is that the assumed causal structure implies that
-the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
-on $X_{t,i} = j$. Formally,
-\eqn {
-\label{eq:observe}
-\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
-}
-This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph. It is not the case in general. 
-For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 
-that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the 
-interventional distribution of $Y_t$ from the observational one.
-
-\todot{Or just give a reference or hide in the appendix depending on space in the end}
-\color{red} Probably should define causal model and back-door rule to properly show this.\color{black}
-
-We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
-
-\eqn {
-\label{eq:estimation_transfer}
-P(Y_t|do(X_{t,i} = j))= & \nonumber \\
- \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
-}
-
-The main idea of the algorithm is that by making no intervention (doing nothing) it is possible to estimate simultaneously the returns of all interventions.
-Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and should be
-explored separately by actually making the intervention. The algorithm works by optimally balancing the collection of observational data (when
-no action is taken) and making interventions to learn about low probability events. The problem is more challenging because $\boldsymbol{q}$ is
-unknown and must also be estimated.
-
-\begin{algorithm}[h]
-\caption{Causal Best Arm Identification}\label{alg:simple}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} $T, N$
-\FOR{$t \in 1,\ldots,(T - 1) / 2$}
-\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
-\ENDFOR
-\STATE Estimate $\mu$ using observation data:
-\eq{
-(\forall i,j) \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{t,i} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{t,i} = j}}\,.
-}
-\STATE Compute $\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} X_{t,i}$
-\STATE Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\STATE Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\STATE Compute $\hat m = \min\set{1 \leq i \leq N : \hat s'_{i+1} \geq \frac{1}{i}}$
-\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\STATE Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
-
-\FOR{$(i,j) \in A$}
-\FOR{$t \in 1,\ldots,\frac{T - 1)}{2\hat m}$}
-\STATE Choose action $do(X_{t,i} = j)$ and observe $r_t$
-\ENDFOR
-\STATE Recompute $\hat \mu_{i,j} = \frac{2\hat m}{T} \sum_{t=1}^{ T/2\hat m} r_t(X_{t,i}=j)$ 
-\ENDFOR
-\STATE Estimated optimal action is $\hat i^*, \hat j^* = \argmax_{i,j} \hat \mu_{i,j}$
-\STATE Choose action $do(X_{t,\hat i^*} = \hat j^*)$
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:unknown_q_simpleregret}
-Define $m = \min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$.
-Then Algorithm \ref{alg:simple} satisfies
-\eq{
-r_T \in \bigo{\sqrt{\frac{m}{T}\log\left(\frac{NT}{m}\right)}}\,.
-}
-\end{theorem}
-
-Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega(\sqrt{N/T})$.
-Since $m$ is typically much smaller than $N$ we anticipate that the new algorithm will significantly outperform classical bandit algorithms in
-this setting.
-
-\todot{Do we have a proof for this in another doc?}
-\begin{proof}
-\end{proof}
+\input{simple-regret}
 
@@
 
-We now tackle the case where the objective is to minimise the cumulative regret.
-This is significantly more challenging since the algorithm can no longer explore naively for $O(T)$ rounds without suffering linear regret.
-
-We begin by considering the case where $\boldsymbol{q}$ is known and propose a simple explore-exploit based algorithm that leverages  \eqref{eq:observe}.
-Without loss of generality, we assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
-\leq q_N$.
-
-\begin{algorithm}[h]
-   \caption{Causal Explore-Exploit}
-   \label{alg:known_q}
-\begin{algorithmic}
-   \STATE {\bfseries Input:} $T,\boldsymbol{q}$
-   \STATE Let $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
-   \STATE Let $h = T^{2/3}m^{1/3} \log(TK)^{1/3}$
-   \STATE Let $A = \set{(i,j) : i \leq m, j = 1}$ be the set of infrequently observed arms
-   \FOR{$t=1$ {\bfseries to} $h/2$}
-   \STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
-   \ENDFOR
-   \STATE Compute for all arms $(i,j) \notin A$:
-   \eq{
-   \hat{\mu}_{i,j} = \frac{2}{h}\frac{\sum_{t=1}^{h/2}\ind{X_{i,t}=j}r_t}{q_i^j(1-q_i)^{1-j}}
-   }
-   \FOR{$(i,j) \in A$}
-    \FOR{$t'=1$ {\bfseries to} $h/2m$}
-    \STATE Choose the action $do(X_{i,t'} = j)$ and observe $r_t$
-    \ENDFOR
-    \STATE Compute $\hat{\mu}_{i,j} = \frac{2m}{h}\sum_{t'=1}^{h/2m}\ind{X_{i,t'} = j}r_{t'}$
-   \ENDFOR
-   \STATE Compute $(\hat{i}^*,\hat{j}^*) = \argmax_{(i,j)} \hat{\mu}_{i,j}$
-   \FOR{$t=h$ {\bfseries to} $T$}
-   \STATE Choose the action $do(X_{\hat{i}^*,t} = \hat{j}^*)$
-   \ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:known_q_regret}
-Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
-Then Algorithm \ref{alg:known_q} satisfies
-\eq{
-R(T) \in \bigo{T^{2/3}m^{1/3} \log(KT)^{1/3}}\,.
-}
-\end{theorem}
-
-The lower bound for the standard bandit problem is $R_t \in \bigomega{\sqrt{TK}}$ \cite{Auer1995}.
-Comparing these results shows exploiting the extra information provided by the causal structure should outperform standard bandit algorithms when the number of arms is large, $K >> m^{2/3}T^{1/3}$.
-The parameter $m$ summarizes the vector $\boldsymbol{q}$, and represents the number of actions that occur rarely naturally and thus must be explicitly explored.
-If $q_1,...,q_N = 0$, the problem is completely unbalanced and $m = N$.
-If $q_1,...,q_N = \frac{1}{2}$, the problem is completely balanced and $m = 1$.
-
-Algorithm \ref{alg:known_q} tries to learn the rewards for all the arms during an exploration phase $h$ and then picks the arm with the highest empirical mean for all remaining timesteps.
-During its exploration phase, it learns all the frequently occurring actions by observation and the remaining, infrequently occurring actions, by explicitly playing them.
-This leads to Chernoff type high probability bounds on the difference between the empirical and true rewards for all arms of the form $P(\hat{\mu}_{i,j} - \mu_{i,j} > D) \leq e^{-hD^2/m}$.
-By choosing optimal values for $D$ and $h$ we obtain the regret bound given in \cref{thm:known_q_regret}.
-A full proof is given in the supplementary materials.
-
-We now demonstrate we can obtain the same order of regret for the case where $\boldsymbol{q}$ is unknown.
-
-\begin{algorithm}[h]
-\caption{Bandit Regret Algorithm}\label{alg:bandit}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} $T, N$
-\STATE $\delta = \frac{1}{T^{1/3}}$ 
-\STATE $T_1 = 48Nlog\left(4N/\delta\right)$ 
-\STATE Run Algorithm \ref{alg:simple} to line 11 with input $T_1,N$.
-\IF {$\hat{m} > \frac{N^{3/2}}{\sqrt{T}}$}
-\STATE Switch to the standard UCB algorithm.
-\ELSE
-\STATE $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
-\STATE Run Algorithm \ref{alg:simple} with input $h,N$.
-\STATE Compute $(\hat{i}^*,\hat{j}^*) = \argmax_{(i,j)} \hat{\mu}_{i,j}$
-\FOR{$t=h$ {\bfseries to} $T$}
-\STATE Choose the action $do(X_{\hat{i}^*,t} = \hat{j}^*)$
-\ENDFOR
-\ENDIF
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:unknown_q_regret}
-Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
-Then Algorithm \ref{alg:bandit} satisfies
-\eq{
-R(T) \in \bigo{T^{2/3}m^{1/3}log(KT)^{1/3}}\,.
-}
-\end{theorem}
+\input{cumulative-regret}
 
@@
 
+\input{lower-bound}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Experiments}
-In all experiments, the reward is generated according to:
-\eq{
-Y_t \sim \begin{cases}
-\bernoulli(.5+\epsilon) & \text{if } X_1 = 1 \\
-\bernoulli(\frac{.5-(.5+\epsilon)q_1}{1-q_1}) & \text{otherwise}\,.
-\end{cases}
-}
 
-With $\epsilon \in (0,.5)$. This ensures that 
-\eq {
-\P{Y|do(X_i = j)} = \begin{cases}
-.5+\epsilon & \text{if } (i,j)=(1,1) \\
-\frac{.5-(.5+\epsilon)q_1}{1-q_1} & \text{if } (i,j) = (1,0) \\
-.5 & \text{otherwise}
-\end{cases}
-}
-
-This allows us to rapidly simulate rewards and conditional rewards for large numbers of variables and to change the value of $\boldsymbol{q}$ with minimal impact on the reward function \footnote{$\boldsymbol{q}$ only effects the reward of a single, non-optimal arm, which becomes insignificant with large numbers of variables.}.
-
-\begin{figure}
-\caption{Final regret versus number of variables $N$ for UCB with $\alpha = 2$, Causal-Explore-Exploit with $m=2$ and with $m=N$ and horizon $T = 10,000$ .
-Error bars show standard deviation over 100 simulations.
-The regret for UCB grows linearly with the number of variables, whist for Causal-Explore-Exploit with fixed $m$, the growth is sub-logarthmic.
-}
-\label{fig:known_q_r_vs_N}
-\centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_N_T10000_sims100_20151229_113550.pdf}
-\end{figure}
-
-\begin{figure}
-\caption{Cumulative regret over time for $N = 17$ for UCB with $\alpha=2$, Causal-Explore-Exploit with $m=2$ and Causal-Explore-Exploit with $m=N$.
-Shaded region shows standard deviation over 100 simulations.
-The Causal-Explore-Exploit algorithm incurs linear regret during the exploration phase, after which it selects the optimal arm with high probability.
-For $m=2$, we have $K \sim m^{2/3}T^{1/3}$ and see that we are in the regime in which Causal-Explore-Exploit outperforms UCB.}
-\label{fig:known_q_r_vs_t}
-\centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_t_T10000_N17_sims100_20151229_120647.pdf}
-\end{figure}
+\input{experiments}
 
@@
 
-In our algorithm, we have only used the side information provided by the $do()$ action about other actions.
-Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
-The feedback graph is distinct from the causal graph.
-A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
-For this specific problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
-The Exp3-IX algorithm \cite{Kocak2014} was developed for the adversarial version of this problem and has regret $\bigo{\sqrt{\bar{\alpha}T}}$, where $\bar{\alpha}$ is the average independence number of $G_t$.
-In our case $\bar{\alpha} = \frac{N}{2}$ so we again obtain the regret of the standard bandit algorithm.
-The issue here is that a malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
-This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$
-
-\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
-In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
-\cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
-In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
-The quality of the estimate notably depends not only on the number of times that action was selected.
-For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$.
-This more complex side information does not fit within the graph feedback framework.
-
-\section{Future Open Questions}
-\begin{itemize}
-\item Known but arbitrary structure
-\item Learning structure then exploiting
-\end{itemize}
-\section{Conclusion}
+\input{discussion}
 

### 2016-01-13 causal-bandit/icml-2015/cumulative-regret.tex
broke up main paper into separate tex

We now tackle the case where the objective is to minimise the cumulative regret.
This is significantly more challenging since the algorithm can no longer explore naively for $O(T)$ rounds without suffering linear regret.

We begin by considering the case where $\boldsymbol{q}$ is known and propose a simple explore-exploit based algorithm that leverages  \eqref{eq:observe}. Without loss of generality, we assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ... \leq q_N$.

\begin{algorithm}[h]
   \caption{Causal Explore-Exploit}
   \label{alg:known_q}
\begin{algorithmic}
   \STATE {\bfseries Input:} $T,\boldsymbol{q}$
   \STATE Let $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
   \STATE Let $h = T^{2/3}m^{1/3} \log(TK)^{1/3}$
   \STATE Let $A = \set{(i,j) : i \leq m, j = 1}$ be the set of infrequently observed arms
   \FOR{$t=1$ {\bfseries to} $h/2$}
   \STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
   \ENDFOR
   \STATE Compute for all arms $(i,j) \notin A$:
   \eq{
   \hat{\mu}_{i,j} = \frac{2}{h}\frac{\sum_{t=1}^{h/2}\ind{X_{i,t}=j}r_t}{q_i^j(1-q_i)^{1-j}}
   }
   \FOR{$(i,j) \in A$}
    \FOR{$t'=1$ {\bfseries to} $h/2m$}
    \STATE Choose the action $do(X_{i,t'} = j)$ and observe $r_t$
    \ENDFOR
    \STATE Compute $\hat{\mu}_{i,j} = \frac{2m}{h}\sum_{t'=1}^{h/2m}\ind{X_{i,t'} = j}r_{t'}$
   \ENDFOR
   \STATE Compute $(\hat{i}^*,\hat{j}^*) = \argmax_{(i,j)} \hat{\mu}_{i,j}$
   \FOR{$t=h$ {\bfseries to} $T$}
   \STATE Choose the action $do(X_{\hat{i}^*,t} = \hat{j}^*)$
   \ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:known_q_regret}
Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
Then Algorithm \ref{alg:known_q} satisfies
\eq{
R(T) \in \bigo{T^{2/3}m^{1/3} \log(KT)^{1/3}}\,.
}
\end{theorem}

The lower bound for the standard bandit problem is $R_t \in \bigomega{\sqrt{TK}}$ \cite{Auer1995}. Comparing these results shows exploiting the extra information provided by the causal structure should outperform standard bandit algorithms when the number of arms is large, $K >> m^{2/3}T^{1/3}$. The parameter $m$ summarizes the vector $\boldsymbol{q}$, and represents the number of actions that occur rarely naturally and thus must be explicitly explored. If $q_1,...,q_N = 0$, the problem is completely unbalanced and $m = N$. If $q_1,...,q_N = \frac{1}{2}$, the problem is completely balanced and $m = 1$.

Algorithm \ref{alg:known_q} tries to learn the rewards for all the arms during an exploration phase $h$ and then picks the arm with the highest empirical mean for all remaining timesteps. During its exploration phase, it learns all the frequently occurring actions by observation and the remaining, infrequently occurring actions, by explicitly playing them. This leads to Chernoff type high probability bounds on the difference between the empirical and true rewards for all arms of the form $P(\hat{\mu}_{i,j} - \mu_{i,j} > D) \leq e^{-hD^2/m}$. By choosing optimal values for $D$ and $h$ we obtain the regret bound given in \cref{thm:known_q_regret}. A full proof is given in the supplementary materials.

We now demonstrate we can obtain the same order of regret for the case where $\boldsymbol{q}$ is unknown.

\begin{algorithm}[h]
\caption{Bandit Regret Algorithm}\label{alg:bandit}
\begin{algorithmic}[1]
\STATE {\bf Input:} $T, N$
\STATE $\delta = \frac{1}{T^{1/3}}$ 
\STATE $T_1 = 48Nlog\left(4N/\delta\right)$ 
\STATE Run Algorithm \ref{alg:simple} to line 11 with input $T_1,N$.
\IF {$\hat{m} > \frac{N^{3/2}}{\sqrt{T}}$}
\STATE Switch to the standard UCB algorithm.
\ELSE
\STATE $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
\STATE Run Algorithm \ref{alg:simple} with input $h,N$.
\STATE Compute $(\hat{i}^*,\hat{j}^*) = \argmax_{(i,j)} \hat{\mu}_{i,j}$
\FOR{$t=h$ {\bfseries to} $T$}
\STATE Choose the action $do(X_{\hat{i}^*,t} = \hat{j}^*)$
\ENDFOR
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:unknown_q_regret}
Define $m =   min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$
Then Algorithm \ref{alg:bandit} satisfies
\eq{
R(T) \in \bigo{T^{2/3}m^{1/3}log(KT)^{1/3}}\,.
}
\end{theorem}

### 2016-01-13 causal-bandit/icml-2015/discussion.tex
broke up main paper into separate tex

--- 2015-11-04 TPR_report.tex
@@
-\documentclass[11pt,a4paper]{article}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{dsfont}
-\usepackage{listings}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{tabularx}
-\usepackage{enumitem}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithm
-\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
-\usepackage{parskip}
-\usepackage{tikz}
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\usetikzlibrary{arrows,positioning} 
-\usepackage{subcaption}
-
-\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
-
-\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
-
-\tikzset{
-    %Define standard arrow tip
-    >=stealth',
-    %Define style for boxes
-    observed/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           },
-     latent/.style={
-           circle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-    target/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-    observedrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           },
-     targetrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-     empty/.style={
-           circle,
-           rounded corners,
-           minimum width=.5em,
-           minimum height=.5em,
-           font=\footnotesize,
-           text centered,
-           },
-    % Define arrow style
-    pil/.style={
-           o->,
-           thick,
-           shorten <=2pt,
-           shorten >=2pt,},
-    sh/.style={ shade, shading=axis, left color=red, right color=green,
-    shading angle=45 }   
-}
-
-\newcommand{\defined}{\vcentcolon =}
-\newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}{\mathbb E}
-\newcommand{\Var}{\operatorname{Var}}
-\newcommand{\calF}{\mathcal F}
-\newcommand{\sr}[1]{\stackrel{#1}}
-\newcommand{\set}[1]{\left\{#1\right\}}
-\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
-\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
-\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
-\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
-\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
-\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
-\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
-\newcommand{\Ber}{\operatorname{Bernoulli}}
-\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
-\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
-\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
-\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
-\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
-
-\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
-
-\author{Finnian Lattimore}
-\title{Learning how to act: making good decisions with machine learning}
-
-\begin{document}
-\def\ci{\perp\!\!\!\perp} % from Wikipedia
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{definition}{Definition}[section]
-\maketitle
-
-\section{Introduction}
-
-Learning the outcome of an action is central to many real world problems.
-Does de-worming children in poor countries improve health and educational outcomes [worm-wars]?
-Would increasing the minimum wage benefit lead to higher unemployment?
-Will offering this customer a discount improve my revenue?
-These questions are difficult as they require more than identifying a pattern in data.
-There are two, very different, major approaches to this problem with the machine learning community: reinforcement learning and causal inference.
-
-Reinforcement learning addresses the problem of learning from explicitly taking actions.
-There is typically some state or environment, an agent chooses an action from those available in the current state, the state evolves stochastically as a function of the selected action and the agent then receives some feedback or reward that is a function of the new state.
-This setting differs from the standard classification problem in that the agent must learn from feedback on the selected action, rather than being presented with the correct action for a given state.
-A common modelling assumption is to assume that the state evolves only as a function of the previous state and the action chosen, and given these, is independent of the previous history of states and actions.
-This is known as a Markov decision process or MDP.
-A particularly well studied and understood model are MDPs with only a single state.
-In this case, there are a set of actions, each associated with a fixed but unknown reward distribution and at each time step our agent selects and action and receives corresponding feedback.
-This is known as the multi-armed bandit problem.
-
-Causal inference makes use of assumptions to allow the outcome of actions to be predicted from observational data.
-The key to causal inference is a framework that can model how actions change the state of the world.
-This framework then allows us to map information collected in one setting to another.
-
-Both approaches can be seen as extensions to the concept of randomised controlled trials.
-Bandit algorithms deal with the sequential nature of the decision making process.
-Causal inference with the problem that full randomisation is not always feasible, affordable or ethical.The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
-The goal of my thesis is to explore these questions.
-In the next sections I review the key literature in causal inference and bandits.
-I then present a general approach to  how causal models might be incorporated into bandit settings and conclude by demonstrating an algorithm that leverages causal assumptions to improve performance in a specific bandit setting.
-
-\section{Causal Inference}
-
-\subsection{Models of causality and intervention}
-
-Predicting the outcome of actions without explicitly taking them requires assumptions about how the actions will change the system.
-A very powerful and general model that underlies much of the recent work in causality is the causal bayesian network.
-A causal Bayesian network, or directed acyclic graph (DAG), is a Bayesian network in which a link $V_{i} \rightarrow V_{j}$, is defined to mean $V_{i}$ directly causes $V_{j}$.
-This means that if we intervene and change the value of $V_{i}$, we expect $V_{j}$ to change, but if we intervene to change $V_{j}$, $V_{i}$ will not change.
-More generally, if $G$ is a causal network for a distribution $P$ defined over variables $V_{1}...V_{N}$, then the distribution after an intervention where we set $X \subset V$ to $x$, denoted $do(X=x)$ is obtained by simply dropping the terms for each of the variables in $X$ from the factorization given by the network.
-This is referred to as the truncated product formula \cite{Pearl2000}.
-
-If the network contains latent (unobservable) variables we will not be able to calculate all the terms in the truncated product formula.
-However, it may still be possible to determine the post-interventional distribution of specific variables of interest.
-A general causal query $P(Y|do(X=x))$ is identifiable if it can be shown to be equivalent to an expression containing only pre-interventional quantities.
-This means that asymptotically we obtain an unbiased estimate for the distribution after an intervention based on purely observational data.
-
-The Do Calculus is a set of three rules that allow transformations of interventional terms to non-interventional ones, given a causal graph \cite{}.
-They are derived directly from d-seperation properties of graphical models and the definition of intervention in causal DAGs.
-These rules are complete.
-A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus and there is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {}.
-I have implemented this algorithm as a simple javascript application to demonstrate XXXX.
-If a query is not identifiable, it may still be possible to get bounds for causal effects, for example using instrumental variables \cite{} or by making additional assumptions.
-
-There are two other key frameworks that arise in causal inference.
-Counterfactuals and structural equation models.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}.
-Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008}.
-
-For example, if we wanted to estimate the causal effect of a medical treatment, then we might let $Y^{1}$ be a counterfactual random variable representing the (binary) potential outcome if treated.
-The distribution of $Y^{1}$ is the distribution we would see of the outcome, $Y$, if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-The causal effect of the drug is the difference between the probability of recovery, across the population, if everyone was treated, and the probability of recovery given placebo $P(Y^{1})-P(Y^{0})$.
-This quantity is estimable from observational data if we assume $X \ci Y^{0}$ and $X \ci Y^{1}$.
-These assumptions are referred to as ignoreability assumptions \cite{Rosenbaum1983}.
-They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-Graphically, this is equivalent to the assumption that there is no variable that is a parent of both the treatment $X$ and the outcome $Y$.
-
-Structural equation models (SEMs) describe a deterministic world, where underlying mechanisms determine the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-If the network for a structural equation model is acyclic then it implies a recursive factorization of the joint distribution over its variables.
-In other words, it is a causal bayesian network.
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed are almost equivalent.
-To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of bayesian networks or as structural equation models or as conditional independence statements involving counterfactual variables (ignorability assumptions).
-
-It is possible to pose causal queries in terms of counterfactuals that are not interventional and cannot be phrased in terms of the do-notation.
-The scientific and philosophical validity of such counterfactual queries remains under question \cite{Dawid2000,Dawid2014}, however they are nonetheless widely posed in the form of attribution of causal effects to different pathways and mediation \cite{Pearl2014,Imai2010a,VanderWeele2011}.
-
-There are differences between the models we have considered when it comes to these non-interventional queries.
-Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.
-Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
-Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
-
-In practice, differences in focus and approach across different fields eclipse these actual differences in the models.
-The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field.
-Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
-In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
-
-\subsection{Discovering causal structure}
-
-In the previous section we discussed when assumptions about the structure of the variables in a specific problem is sufficient to identify a causal effect.
-This approach relies on having enough prior knowledge or theory about the problem to allow you to, at least partially, specify the causal network.
-In this section, we consider the much harder problem of causal inference where you need to learn the network.
-Causal inference might seem impossible without specific assumptions about the structure of the variables involved but, amazingly, some aspects of causal structure can be determined from much more general assumptions.
-
-\subsubsection{Discovery based on conditional independence}
-One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
-Assume there is some acyclic causal network $G$ that generated the distribution $P(\boldsymbol{V})$ from which our data has been sampled.
-Our goal is to recover the network from this data.
-
-Since $G$ is a bayesian network, if $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in $P$ .
-However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
-This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
-This assumption, commonly referred to as  \textbf{\textit{faithfulness}}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
-Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} for $P$.
-
-Faithfulness is an assumption.
-It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
-However, most distributions generated by a causal bayesian network will be faithful to that network.
-For faithfulness to be violated, different causal effects must exactly balance one-another out.
-For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
-If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
-However, this independence is not stable.
-It would disappear under a small perturbation to any of the parameters.
-
-Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$.
-A wide range of algorithms have been developed based on this key observation, see table \ref{table:discovery_algorithms}.
-Constraint based methods such as the PC algorithm \cite{}, FCI algorithm \cite{} and RFCI algorithm \cite{}, perform sequential conditional independence tests and eliminate inconsistent graphs.
-Search and score based methods, such as GES \cite{}, search over the space of graphs and score them according to how well they fit the independences given a complexity penalising prior.
-Constraint based methods are faster, particularly for sparse graphs, but can lack robustness as errors in early conditional independence tests can propagate.
-Search and score based methods are more robust for small samples sizes but difficult to scale to larger graphs.
-This has led to the development of hybrid approaches, such as the MMHC algorithm \cite{}.
-A key component of causal discovery is the ability to do high dimensional non-parametric conditional independence tests.
-Developments in kernalized conditional independence tests \cite{},\cite{Zhang2012},\cite{Gretton2008} have made this possible.
-
-\begin{table}[h]
-\centering
-\caption{A comparison of key causal discovery algorithms}
-\label{table:discovery_algorithms}
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{| c | c | p{4cm} | c | c |}
-\hline
-  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent }  \\
-  \hline
-  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No  \\ 
-  \hline
-  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes\\
-  \hline
-  RFCI & Constraint based & ? & 500 & Yes  \\
-  \hline
-  GES & Search \& Score  & Worst case exponential & 50 & No  \\
-  \hline
-  MMHC & Hybrid & ? & 5000 & No  \\
- \hline
-\end{tabular}
-\end{table}
-
-If the end goal of causal discovery is to estimate causal effects, then it may not be necessary to learn the entire graph, only the subset of the graph surrounding target variables of interest.
-Such local causal discovery techniques can be scaled to problems with many more variables \cite{}.
-Once a set of causal graphs has been identified, causal effects of interest can be bounded by combining the results for the all the networks.
-This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
-An implementation is available in the R package \cite{Kalisch2012}
-
-\subsubsection{Discovery with functional models}
-All of the algorithms we have considered so far return a Markov equivalence class.
-They cannot distinguish between two models that result in the same set of conditional independence relations.
-Consider the very simple case where we have only two variables and the only possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
-These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
-No algorithm relying purely on conditional independence relations can separate these two cases.
-
-One solution is to utilize structural equation models to specify additional assumptions.
-For example, if we assume that noise is additive, such that $X \rightarrow Y \implies Y = f(X)+\epsilon$, then this will only be invertible such that $X = g(Y) + \epsilon'$ for specific pairs of functions $f$ and noise distributions $\epsilon$.
-Thus in general we will be able to identify the causal direction \cite{Hoyer2009}.
-This can be extended to post-non-linear additive noise, $Y = h(f(X)+ \epsilon)$, \cite{Zhang2008a}.
-These techniques can also be applied over more than two variables \cite{}.
-
-A more general approach is to leverage the assumption that the functions are independent of inputs \cite{}.
-This leads to the idea that $P(X)$ and $P(Y|X)$ are independent if $X \rightarrow Y$ but not if $Y \rightarrow X$.
-\cite{} propose testing for this by applying both semi-supervised and standard techniques to estimate $P(Y|X)$.
-Semi-supervised methods, which utilize additional points from $P(X)$ to learn $P(Y|X)$ should only be able to outperform standard methods if $Y \rightarrow X$.
-
-Finally, rather than explicitly developing an algorithm based on a specific asymmetry between cause and effect, \cite{} propose learning what causality looks like from data.
-They take as input a dataset where each row of data is itself a dataset in which either $X \rightarrow Y$ or $Y \rightarrow X$ and a corresponding label.
-Estimates of the distributions $P(X)$, $P(Y)$ and $P()$ for each row are then mapped to features in some kernel space via mean kernel embeddings and finally a standard classification algorithm can be trained to learn label.
-New datasets where the direction of causality is unknown are then simply mapped to the kernel space and the causal direction is classified according to the trained classifier.
-In practice, the classifier is trained mostly on simulated data, as it is difficult to find a sufficient set of causal problems with only two variables where the direction of causality is known.
-
-\section{Multi-armed Bandits}
-
-In its classic formulation \cite{Robbins1952} the K-armed bandit ...
-
-Definition of regret
-
-An algorithm is learning if it can obtain sublinear regret.
-
-Mention of adversarial and markovian alternatives.
-
-\subsection{Regret bounds for classic K-armed bandit}
-
-\begin{itemize}
-\item the lower bound for any algorithm for the K-armed bandit problem (stochastic or adversarial) is $\bigo{\sqrt{nK}}$ \cite{Auer1995}
-
-\item UCB problem dependent upper bound \cite{Auer2002}
-
-\eqn {
-R^{ucb}_\mu(n) \in \bigo{\sum_{i=2}^K \frac{1}{\Delta_i}\log(n)}
-}
-\item UCB problem independent upper bound $\bigo{\sqrt{nK\log(n)}}$ \cite{Bubeck2012}
-
-\item Modifications of UCB can close this gap.
-\cite{Lattimore2015, Audibert2009}.
-Subtle modification of the UCB algorithm closes the gap with the lower bound (ie problem independent upper bound of $\bigo{\sqrt{nK}}$, whilst yielding a similar problem dependent upper bound to UCB.
-\cite{Lattimore2015}
-
-\item Exp3 achieves $\bigo{\sqrt{nK\log(n)}}$ regret \cite{Auer1995}
-
-\item $\bigo{\sqrt{nK}}$ has also been achieved for the adversarial setting. \cite{Audibert2009}
-
-\end{itemize}
-
-\subsection{Extensions to K-armed bandits}
-
-The classic multi-armed bandit is a powerful tool for sequential decision making.
-However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example linear bandits, dependent bandits, X-armed bandits and gaussian process bandits, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
-
-Feedback with expert advice (full feedback) on all actions get $R_n = O(\sqrt{n\log(K)})$
-
-Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It has demonstrated that problems fall into one of four categories, trivial with no regret, easy with $R_n = \bigthetatilde{\sqrt{n}}$ , hard with $R_n = \bigtheta{n^{2/3}}$ and hopeless with $R_n = \bigomega{n}$ \cite{Bartok2014}.
-Focus on finite sets of actions and feedback signals.
-Analysis yields algorithms that are optimal with respect to the horizon $n$ but not necessarily other parameters (ie $K$).
-
-\section{Unifying the frameworks}
-
-A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph.
-Each possible assignment of variables to values is an action (bandit arm).
-See figure \ref{fig:unify_frameworks} for a simple example.
-The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action.
-The number of actions or arms grows exponentially with the number of variables in the graph, making it important to use algorithms that take account of the graph structure to reduce the search space.
-
-Modelling a problem as a causal graph only makes sense when rewards are generated stochastically - since causal graphs fundamentally model probability distributions over variables.
-Thus the connection is to stochastic bandit problems (although adversarial bandits algorithms may be applied to stochastic problems).
-
-\begin{figure}[h]
-\caption{A simple causal graphical model and corresponding complete action space. A and B represent binary variables that can be intervened on and Y represents the reward.}
-\label{fig:unify_frameworks}
-\centering
-\begin{subfigure}[c]{0.3\textwidth}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
-\node[main node](1){A};
-\node[tg, below right=of 1](2){Y};
-\node[main node, above right=of 2](3){B};
-\path[]
-	(1) edge (2)
-	(3) edge (2);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-Actions = \begin{tabular}{|c|}
-	\hline
-  do(A=0,B=0) \\
-  do(A=0,B=1) \\
-  do(A=1,B=0) \\
-  do(A=1,B=1) \\
-  \hline
-  do(A=0) \\
-  do(A=1) \\
-  do(B=0) \\
-  do(B=1) \\
-  do() \\
-  \hline
-\end{tabular}
-\end{subfigure}
-\end{figure}
-
-If we begin by considering the case where the causal graph is known, problems then take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected.
-
-If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
-See figure \ref{fig:cholesterol_graph} as an example.
-
-\begin{figure}[h]
-\caption{Example causal graph (based on \cite{Koller2009}) where the outcome of interest (reward) is cholesterol level .
-The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
-For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
-\label{fig:cholesterol_graph}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
-\node[node](1){Health concious};
-\node[node, below left=of 1](2){Diet};
-\node[node, below right=of 1](3){Exercise};
-\node[node, below right=of 2](4){Weight};
-\node[tg,below left=of 4](5){Y=Cholesterol};
-\node[node, below =of 5](6){Test result};
-\node[node, above right=of 3](7){Free time};
-\path[]
-	(1) edge (2) edge (3)
-	(2) edge (4) edge (5)
-	(3) edge (4)
-	(5) edge (6)
-	(7) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-If we receive feedback on additional nodes, the problem can be more interesting.
-In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
-Consider again the model in figure \ref{fig:unify_frameworks}.
-The causal structure implies:
-
-\eqn {
-P(Y|do(A = 0)) &= P(Y|do(),A = 0) \\
-&= P(Y|do(B=0),A=0)P(B=0)+P(Y|do(B=1),A=0)P(B=1) 
-}
-
-Thus we gain information about the reward for the action $do(A=0)$ from selecting the action $do()$ or $do(B = b)$ and then observing $A = 0$.
-
-We only get this form of side information for actions that don't specify the value of every variable, ie those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-Since the reward distribution for actions that set a subset of the variables is the result of marginalizing out other variables, they can only be optimal if they have lower cost.
-So if the cost of all actions is constant (no matter how many variables must be set), then the problem has the same characteristics as if only the reward node were observable.
-
-If the information on the value of additional nodes is available prior to selecting an action the problem resembles a contextual bandit.
-For example if we observe $A = 0$ then, in deciding between the actions $do(B=0)$ and $do(B=1)$, we would want information on $P(Y|A=0,B=0)$ and $P(Y|A=0,B=1)$.
-Note, side information can still arise if we learn the value of some variables prior to selecting an action and some afterwards.
-
-\subsection{Incorporating side information induced by casual graph}
-
-As a first step towards more general problems, we develop and evaluate a bandit algorithm that incorporates the side information induced by a specific causal graph.
-
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
-This is an expansion of the model shown in figure \ref{fig:unify_frameworks}.
-We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
-
-As an example, consider a farmer wishing to optimize the yield of her crop.
-She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
-We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
-After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
-This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
-
-\begin{figure}[h]
-\centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty},tg/.style={target}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[tg, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\end{figure}
-
-Let $\boldsymbol{q} \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known horizon $T$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{0,\ldots, N}$ and $J_t \in \set{0,1}$, where $I_t = 0$ represents $do()$.
-
-\item Variables not set by the learner take their values probabilistically according to $\boldsymbol{q}$ 
-\eq{
-X_{t,i} = \begin{cases}
-\sim \Ber(q_i) &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner observes $(\boldsymbol{X}_{t},Y_{t})$ and receives reward $Y_t \sim \Ber(r(\boldsymbol{X}_{t}))$ where $r:\set{0,1}^N \to [0,1]$ is unknown and arbitrary. 
-\end{enumerate}
-
-The problem could be treated as a standard bandit problem with $K = 2N + 1$ arms,
-
-\eq {
-\{do(), do(X_1=0), do(X_1 = 1)\;\; ... \;\;do(X_N = 1)\}
-}
-
-yielding a regret $\bigo{\sqrt{TN}}$. However, this doesn't leverage the side information induced by the causal structure:
-
-\eqn {
-\label{eq:observe}
-P(Y_t|do(X_{i,t} = j)) &= P(Y_t|do(),X_{i,t} = j) \\
-\label{eq:estimation_transfer}
-&= P(Y_t|do(X_{l,t}=0),X_{i,t}=j)(1-q_l)+P(Y_t|do(X_{l,t}=1),X_{i,t}=j)q_l 
-}
-
-The hardness of this problem depends strongly on the vector $\boldsymbol{q}$.
-This is intuitive from a consideration of the extreme cases.
-If $\boldsymbol{q} = \boldsymbol{0}$ (or $\boldsymbol{1}$), then half of the actions we wish to explore will only occur if we explicitly select them.
-For these actions we get no side information by playing other actions.
-So at best we could expect a factor of two improvement over simply running a standard bandit algorithm.
-If $\boldsymbol{q} = \boldsymbol{\frac{1}{2}}$, the observe action, $do()$, effectively becomes a revealing action and we would anticipate obtaining a regret with minimal dependence on the number of variables $N$.
-
-We consider a simple explore-exploit based algorithm, that will explore for $h$ timesteps, sampling actions in a way that depends on $\boldsymbol{q}$.
-We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
-During the exploration phase, we will balance purely observing, $I_t = 0$, which takes advantage of equation \ref{eq:observe} to reveal the reward for frequently occurring arms, with explicit sampling of infrequently occurring arms.
-
-Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ... \leq q_N$. Let:
-
-\eqn {
-m \in [2,N] = \set{m:q_m > \frac{1}{m}}
-}
-
-$m$ is a measure of the number of infrequently occurring arms.
-If the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
-If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$.
-Let $q_{ij} = \P{X_i = j}$ and $\mu_{ij} = P(Y|do(X_i = j))$.
-
-Suppose we select $I_t = 0$ for the first $h/2$ timesteps.
-This is at worst half the optimal.
-We use these observations to estimate the rewards for all the frequently occurring arms, $\set{do(X_i = 0) \; \forall i}$, and $\set{X_i = 1:i \geq m}$
-
-\eqn {
-\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
-}
-
-We then explicitly play each of the $m$ infrequently occurring arms $\frac{h}{2m}$ times and estimate their rewards as:
-
-\eqn {
-\hat{\mu}_{ij} = \frac{2m}{h}\sum_{t = 1}^{h/2m}\ind{Y_t = 1|X_i=j}
-}
-
-We then simply play the arm with the highest estimated reward for the remaining $T-h$ timesteps.
-We can show that we can select an $h$ (based on $T$, $N$ and $m$) that leads to worst case regret $R_T = \bigotilde{T^{2/3}m^{1/3}}$.
-This result can still be obtained in the case where $\boldsymbol{q}$ is unknown and must be learned during the exploration time.
-Comparing this to the standard bandit regret, $R_T = \bigo{\sqrt{TK}}$, we expect the causal algorithm to do better if $m < \frac{K^{3/2}}{\sqrt{T}}$.
-
-In this algorithm, we have only used the side information provided by the $do()$ action about other actions.
-Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
+In our algorithm, we have only used the side information provided by the $do()$ action about other actions.
+Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
 The feedback graph is distinct from the causal graph.
@@
 
-\subsection{Open questions}
-
+\section{Future Open Questions}
 \begin{itemize}
-\item derive lower bounds
-\item generalize to arbitrary causal graphs
-\item leverage the more complex form of side information as in equation XXX, that provides only partial information about additional actions.
-\item consider cases where the causal graph is not known.
+\item Known but arbitrary structure
+\item Learning structure then exploiting
 \end{itemize}
-
-\section{Appendices}
-
-\subsection{Summary of Notation}
-\begin{itemize}
-\item $\bigotilde{.}$ hides polylogarithmic factors in $\bigo{.}$
-\item $\bigthetatilde{.}$ hides polylogarmic factors in $\bigtheta{.}$
-\end{itemize}
-
-\bibliographystyle{unsrt}% Select the citation style e.g. ieeetr
-\bibliography{../library}% write the directory to the .bib file
-\end{document}
+\section{Conclusion}

### 2016-01-13 causal-bandit/icml-2015/experiments.tex
broke up main paper into separate tex

In all experiments, the reward is generated according to:
\eq{
Y_t \sim \begin{cases}
\bernoulli(.5+\epsilon) & \text{if } X_1 = 1 \\
\bernoulli(\frac{.5-(.5+\epsilon)q_1}{1-q_1}) & \text{otherwise}\,.
\end{cases}
}

With $\epsilon \in (0,.5)$. This ensures that 
\eq {
\P{Y|do(X_i = j)} = \begin{cases}
.5+\epsilon & \text{if } (i,j)=(1,1) \\
\frac{.5-(.5+\epsilon)q_1}{1-q_1} & \text{if } (i,j) = (1,0) \\
.5 & \text{otherwise}
\end{cases}
}

This allows us to rapidly simulate rewards and conditional rewards for large numbers of variables and to change the value of $\boldsymbol{q}$ with minimal impact on the reward function \footnote{$\boldsymbol{q}$ only effects the reward of a single, non-optimal arm, which becomes insignificant with large numbers of variables.}.

\begin{figure}
\caption{Final regret versus number of variables $N$ for UCB with $\alpha = 2$, Causal-Explore-Exploit with $m=2$ and with $m=N$ and horizon $T = 10,000$ . Error bars show standard deviation over 100 simulations. The regret for UCB grows linearly with the number of variables, whist for Causal-Explore-Exploit with fixed $m$, the growth is sub-logarthmic.  }
\label{fig:known_q_r_vs_N}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_N_T10000_sims100_20151229_113550.pdf}
\end{figure}

\begin{figure}
\caption{Cumulative regret over time for $N = 17$ for UCB with $\alpha=2$, Causal-Explore-Exploit with $m=2$ and Causal-Explore-Exploit with $m=N$. Shaded region shows standard deviation over 100 simulations. The Causal-Explore-Exploit algorithm incurs linear regret during the exploration phase, after which it selects the optimal arm with high probability. For $m=2$, we have $K \sim m^{2/3}T^{1/3}$ and see that we are in the regime in which Causal-Explore-Exploit outperforms UCB.}
\label{fig:known_q_r_vs_t}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_t_T10000_N17_sims100_20151229_120647.pdf}
\end{figure}

\begin{figure}
\caption{Simple regret vs horizon, $T$, for $N = 30$ for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$. Error bars show estimated 95\% confidence interval from 1000 simulations.}
\label{fig:simple_vs_T}
\centering
\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_T_N30_sims1000_20160113_010512.pdf}
\end{figure}

\begin{figure}
\caption{Simple regret vs number of variables, $N$, for $T=500$, for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$. Error bars show estimated 95\% confidence interval from 1000 simulations.}
\label{fig:simple_vs_N}
\centering
\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_N_T500_sims1000_20160113_010511.pdf}
\end{figure}

### 2016-01-13 causal-bandit/icml-2015/introduction.tex
broke up main paper into separate tex

A classical problem in economics and operations research is how to choose interventions that maximise utility. 
We study an idealised model of such interventions using a combination of the multi-armed bandit framework \citep{Robbins1952} and the theory/language of causal inference 
\citep{Pearl2000}.

\todom{Describe feature of class of problem; highlight with examples.}

The challenge is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. 
She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
In each season the farmer has enough time and money to intervene and control at most one of these variables:
deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed a through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
How might the farmer best experiment to identify the single, highest yielding intervention without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?

One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem~\cite{Robbins1952} -- each of the six possible interventions is an arm and the crop yield is the reward -- and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
However, as we show in \S\ref{sec:results}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret with a sub-optimal dependence on the number of possible interventions.

There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \cite{TODO} as well as results for partial monitoring which, as first glance, may appear applicable. As we show in Section~\ref{}, applying these results to our problem also exhibit a sub-optimal dependence on the number of available interventions.

\todom{Rework rest of intro}

%Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour. For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.

% A widely used framework for sequential decision making is the multi-armed bandit. In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary. At each timestep the agent selects an action and receives a reward sampled i.i.d from the corresponding reward distribution. The performance of bandit algorithms is described by the regret: the difference in the expected reward obtained by the algorithm and the reward that could be obtained if the optimal action was selected at every timestep.

An alternate approach to selecting actions is causal inference. Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones. This allows an agent to predict the outcome of an action based on non-experimental data. This approach is common in social science, demography, and economics where explicit experimentation may be difficult. For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.

We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.

\todom{Summarise results \& give intuition for algorithm}

A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph. Each possible assignment of variables to values is an action (bandit arm). The reward could be a general function of the action selected and the final state of the graph. However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action. The number of actions grows exponentially with the number of variables in the graph, making it important to use algorithms that take account of the graph structure to reduce the search space.

Problems framed in this way take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected. If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.

If we receive feedback on additional nodes the problem can be more interesting. In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.

We consider a bandit problem where the actions and reward are represented by a specific causal graph that demonstrates this interesting structure. We develop an algorithm to leverage the information provided by this structure and demonstrate it substantially outperforms standard bandit algorithms applied to the same problem where the number of actions is large.

There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures. Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc. There is also an active line of research into bandits with feedback defined by a graph. Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action []. The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.

Partial monitoring is a very general framework for for decoupling the feedback from the action and reward. It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}. Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.

ALSO NEED TO MENTION ANY OTHER COMBINATIONS OF BANDITS+CAUSAL (eg the Elias NIPS paper and Generalized Thompson Sampling paper)

Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action. Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context. (Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results. It is easy to come up with a counter example in which it does not).

### 2016-01-13 causal-bandit/icml-2015/setup.tex
broke up main paper into separate tex

\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1},\ldots,X_{N}\}$ that independently cause a 
target variable of interest $Y \in \R$ (see Figure \ref{fig:causalStructure}).
\begin{figure}[h]
\centering
\caption{Assumed Causal Structure}
\label{fig:causalStructure}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5);

\end{tikzpicture}
\end{figure}

The game proceeds over $T$ identical rounds (or time-steps).
In each round $t$ the learner can choose either to do nothing or they can choose a variable $I_t \in \set{1,\ldots,N}$ and
an intervention $J_t \in \set{0,1}$. After the learner has chosen an intervention they observe $X_{t,i} \in \set{0,1}$ for all $i$ where
\eq{
X_{t,i} \sim \begin{cases}
\dirac(J_t) & \text{if } I_t = i \\
\bernoulli(q_i) & \text{otherwise}\,.
\end{cases}
}
where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. Note that if the learner did not choose an intervention then $I_t$ is undefined
and $X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
Finally the learner observes the reward $Y_t = r(X_t) + \eta_t$ where 
\eq{
r : \set{0,1}^N \to \R
}
is arbitrary (and unknown) 
and $\eta_t$ is a $1$-subgaussian noise
term (with the distribution possibly dependent on $X_t$). 
\todom{Connect to opening ex.}

The expected reward for intervening on variable $i$ by setting it to $j$ is defined by
\eq{
\mu_{i,j} 
&= \E{r(X)|do(X_i = j)} \\
&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k} \,. 
}
The optimal intervention is $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.

It is worth mentioning that the problem may be treated as a multi-armed bandit with $2N$ arms, one corresponding to each intervention.
As we shall shortly see, this approach is usually not practical because the resulting algorithms do not exploit the structure in
the covariates.

\begin{remark}
In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
In the bandit community it is implicit that 
algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
bandit and causal inference communities.
\end{remark}

We consider two standard performance measures. The first is the cumulative regret, which measures the difference between the reward expected under
the omnipotent strategy that knows the optimal action in advance and the expected reward of the learner. 
\eqn{
\label{eq:regret}
R_T = T \mu^* - \E{\sum_{t=1}^T Y_t}\,.
}
The second performance measure is the simple regret, which measures the performance of the learner on the final round only.
\eqn{
\label{eq:regret-simple}
r_T = \mu^* - \E{\mu_{I_T,J_T}}\,.
}
Both versions of the regret are useful in different circumstances. The first is most useful when the learner is truly online and can change their policy over time.
The second is useful when the exploration budget is limited and a fixed policy should eventually be chosen. \todot{make this nicer}

### 2016-01-13 causal-bandit/icml-2015/simple-regret.tex
broke up main paper into separate tex

In this section we develop and analyse an algorithm for minimising the simple regret in the
causal bandit problem.

From a causal perspective the key insight is that the assumed causal structure implies that
the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
on $X_{t,i} = j$. Formally,
\eqn {
\label{eq:observe}
\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
}
This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph. It is not the case in general. 
For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 
that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the 
interventional distribution of $Y_t$ from the observational one.

\todot{Or just give a reference or hide in the appendix depending on space in the end}
\color{red} Probably should define causal model and back-door rule to properly show this.\color{black}

We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.

\eqn {
\label{eq:estimation_transfer}
P(Y_t|do(X_{t,i} = j))= & \nonumber \\
 \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
}

The main idea of the algorithm is that by making no intervention (doing nothing) it is possible to estimate simultaneously the returns of all interventions.
Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and should be
explored separately by actually making the intervention. The algorithm works by optimally balancing the collection of observational data (when
no action is taken) and making interventions to learn about low probability events. The problem is more challenging because $\boldsymbol{q}$ is
unknown and must also be estimated.

\begin{algorithm}[h]
\caption{Causal Best Arm Identification}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} $T, N$
\FOR{$t \in 1,\ldots,(T - 1) / 2$}
\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
\ENDFOR
\STATE Estimate $\mu$ using observation data:
\eq{
(\forall i,j) \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{t,i} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{t,i} = j}}\,.
}
\STATE Compute $\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} X_{t,i}$
\STATE Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
\STATE Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
\STATE Compute $\hat m = \min\set{1 \leq i \leq N : \hat s'_{i+1} \geq \frac{1}{i}}$
\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
\STATE Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$

\FOR{$(i,j) \in A$}
\FOR{$t \in 1,\ldots,\frac{T - 1)}{2\hat m}$}
\STATE Choose action $do(X_{t,i} = j)$ and observe $r_t$
\ENDFOR
\STATE Recompute $\hat \mu_{i,j} = \frac{2\hat m}{T} \sum_{t=1}^{ T/2\hat m} r_t(X_{t,i}=j)$ 
\ENDFOR
\STATE Estimated optimal action is $\hat i^*, \hat j^* = \argmax_{i,j} \hat \mu_{i,j}$
\STATE Choose action $do(X_{t,\hat i^*} = \hat j^*)$
\end{algorithmic}
\end{algorithm}

\begin{theorem}\label{thm:unknown_q_simpleregret}
Define $m = \min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$.
Then Algorithm \ref{alg:simple} satisfies
\eq{
r_T \in \bigo{\sqrt{\frac{m}{T}\log\left(\frac{NT}{m}\right)}}\,.
}
\end{theorem}

Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega(\sqrt{N/T})$.
Since $m$ is typically much smaller than $N$ we anticipate that the new algorithm will significantly outperform classical bandit algorithms in
this setting.

\todot{Do we have a proof for this in another doc?}
\begin{proof}
\end{proof}

### 2016-01-13 causal-bandit/icml-2015/simple-regret.tex
incorporated the simple-regret upper bound and continued to cleanup the proof

@@
 
+\begin{definition}
+Let $\boldsymbol{q} \in [0,1]^N$. Then define $q^{(i)}$ to be the $i$th largest value of the set $\set{\min \set{q_j, 1 - q_j} : 1 \leq j \leq N}$.
+For example, if $\boldsymbol{q} = (0,\, 2/3,\, 1/2)$, then $q^{(1)} = 0$, $q^{(2)} = 1/3$, $q^{(3)} = 1/2$.
+Then define 
+\eqn{
+\label{def:m}
+m(\boldsymbol{q}) = \min\set{1 \leq i \leq N: q^{(i+1)} \geq \frac{1}{i}}\,.
+}
+\end{definition}
+
 \begin{theorem}\label{thm:unknown_q_simpleregret}
-Define $m = \min\set{1 \leq i \leq N:q_{i+1} \geq \frac{1}{i}}$.
-Then Algorithm \ref{alg:simple} satisfies
+Algorithm \ref{alg:simple} satisfies
 \eq{
-r_T \in \bigo{\sqrt{\frac{m}{T}\log\left(\frac{NT}{m}\right)}}\,.
+\simpleregret \in \bigo{\sqrt{\frac{m}{T}\log\left(\frac{NT}{m}\right)}}\,.
 }
@@
 
-\todot{Do we have a proof for this in another doc?}
+The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound.
+
+\begin{lemma}\label{lem:conc1}
+Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
+\eq{
+\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
+}
+\end{lemma}
+
+\iffalse
 \begin{proof}
+Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
+Then
+\eq{
+\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} Z_t\,.
+}
+Now $Z_1,\ldots,Z_{T/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
 \end{proof}
+\fi
+
+\begin{lemma}\label{lem:m_est}
+Let $\delta >0$ and assume $T \geq 48m \log\frac{4N}{\delta}$. Then
+\eq{
+\P{\hat m < \frac{2}{3}m} \leq \delta\ \text{ and } \P{\hat m > 2m} \leq \delta\,.
+}
+\end{lemma}
+
+\begin{proof}
+Let $\boldsymbol{q}^b$ and $\boldsymbol{q}^{ub}$ and be the maximally balanced and unbalanced $\boldsymbol{q}$ for a given $m$ respectively.
+\eq{
+q^{ub}_i = \begin{cases}
+0 & \text{if } i \leq m \\
+\frac{1}{m} & \text{otherwise}\,.
+\end{cases} \\
+q^b_i < \begin{cases}
+\frac{1}{m} & \text{if } i \leq m \\
+1 & \text{otherwise}\,.
+\end{cases}  
+}
+For $\hat{m}$ to over-estimate $m$, we must identify some balanced arms as unbalanced. For $\hat{m}$ to under-estimate $m$, we must identify some unbalanced arms as balanced.
+\eq{
+\P{\hat m > 2m} \leq \P{\hat m > 2m|\boldsymbol{q} = \boldsymbol{q}^{ub}}\\
+\P{\hat m < \frac{2}{3}m} \leq \P{\hat m < \frac{2}{3}m|\boldsymbol{q} = \boldsymbol{q}^{b}}
+}
+
+Given $\boldsymbol{q} = \boldsymbol{q}^{ub}$, we have by Lemma \ref{lem:conc1}, with probability at least $1 - \delta$ that:
+
+\eq{
+& \left| \hat q_i - q_i\right| 
+ \leq \begin{cases}
+0 & \text{if } i \leq m\\
+\sqrt{\frac{6}{mh} \log\frac{2}{\delta}} & \text{otherwise} \\
+\end{cases}\\
+\implies &
+\begin{cases}
+(\forall i \leq m) \qquad \left| \hat q_i - 0\right| = 0 \\
+(\forall i > m) \qquad \left| \hat q_i - \frac{1}{m}\right| \leq \frac{1}{2m} \\ %\; \text{, taking the union bound and assuming } h \geq 24m \log\frac{4N}{\delta}\\
+\end{cases}\\
+\implies &
+\begin{cases}
+(\forall i \leq m) \qquad  \hat q_i  = 0 \\
+(\forall i > m) \qquad \hat q_i \in [\frac{1}{2m}, \frac{3}{2m}]\\
+\end{cases}\\
+\implies & \hat{m} \leq 2m
+}
+
+Given $\boldsymbol{q} = \boldsymbol{q}^{b}$, we have by Lemma \ref{lem:conc1}, with probability at least $1 - \delta$ that:
+
+\eq{
+& \left| \hat q_i - q_i\right| 
+ \leq \sqrt{\frac{6}{mh} \log\frac{2}{\delta}} \qquad \text{if } i \leq m \\
+\implies &
+(\forall i \leq m) \qquad  \hat q_i \leq \frac{3}{2m} \\
+\implies & \hat{m} \geq \frac{2m}{3}
+}
+
+\end{proof}
+
+\begin{proof}[Proof of \cref{thm:simple-regret}]
+
+for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
+\eq {
+\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
+}
+
+Via Hoeffding's Inequality
+
+\eq {
+\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
+}
+
+for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.
+
+\eq {
+(i,j)\notin A & \implies \hat{s}_i \geq \frac{1}{\hat m} \\
+& \implies \sum_{t=1}^h \ind{X_i = j} \geq \frac{h}{\hat{m}}
+}
+
+Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
+
+\eq {
+\hat \mu_{i,j} = \frac{1}{Z_{ij}}\sum_{t'=1}^{Z_{ij}}r_{t'}
+}
+
+\eq {
+\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
+= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
+\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
+\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
+}
+
+Applying the union bound over all $2N$ actions.
+\eq{
+& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
+\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
+}
+
+Now by \cref{lem:m_est}, 
+\eq{
+h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
+}
+
+Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that
+
+\eqn{
+\label{eqn:joint_bound_estimates_m}
+(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
+}
+Suppose $h < 24m \log \frac{4N}{\delta}$. Then
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
+}
+Therefore
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
+}
+Therefore 
+\eq{
+R_s(h) 
+&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
+&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
+}
+as required.
+\end{proof}

### 2016-01-13 causal-bandit/icml-2015/lower-bound.tex
added some todos
incorporated the simple-regret upper bound and continued to cleanup the proof
updates

@@
-\newcommand{\simpleregret}{R_T^{\text{simple}}}
+We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}
 
-We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in \cref{thm:simple}
-
+\todot{Remark that the Gaussian assumption is convenient but that other noise models will also work. Eg., Bernoulli.}
 \begin{theorem}\label{thm:lower}
@@
 Then for $i \leq m$ a computation shows that 
+\todot{Need to add at minimum some citations here}
 \eq{

### 2016-01-13 causal-bandit/icml-2015/setup.tex
updates

@@
 The game proceeds over $T$ identical rounds (or time-steps).
-In each round $t$ the learner can choose either to do nothing or they can choose a variable $I_t \in \set{1,\ldots,N}$ and
+In each round $t$ the learner can choose either to do nothing (denoted by $I_t = \bot$) or they can choose $I_t \in \set{1,\ldots,N}$ and
 an intervention $J_t \in \set{0,1}$. After the learner has chosen an intervention they observe $X_{t,i} \in \set{0,1}$ for all $i$ where
@@
 }
-where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. Note that if the learner did not choose an intervention then $I_t$ is undefined
-and $X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
+where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. 
+Note that if the learner did not choose an intervention ($I_t =  \bot$), then
+$X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
 Finally the learner observes the reward $Y_t = r(X_t) + \eta_t$ where 
@@
 term (with the distribution possibly dependent on $X_t$). 
-\todom{Connect to opening ex.}
+In the farming example $X_{t,1}$ might represent whether the plants had sufficient water in the $t$th season (controllable by installing irrigation) 
+while $X_{t,2}$ might be whether or not there was a spring frost (controllable with heat-lamps or covers).
 
@@
 \label{eq:regret-simple}
-r_T = \mu^* - \E{\mu_{I_T,J_T}}\,.
+\simpleregret = \mu^* - \E{\mu_{I_T,J_T}}\,.
 }
-Both versions of the regret are useful in different circumstances. The first is most useful when the learner is truly online and can change their policy over time.
-The second is useful when the exploration budget is limited and a fixed policy should eventually be chosen. \todot{make this nicer}
+The simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. The cumulative
+regret is more appropriate for online agents that may continue to adapt their policy.

### 2016-01-14 causal-bandit/icml-2015/lower-bound.tex
added unused

@@
 \eq{
-\Prz{I_t = i} + \Pri{I_t \neq i} \geq \exp\left(-\KL(P_0, P_i)\right)\,,
+\Prz{I_T = i} + \Pri{I_T \neq i} \geq \exp\left(-\KL(P_0, P_i)\right)\,,
 }

### 2016-01-15 causal-bandit/icml-2015/introduction.tex
Introduce general causal bandit problem.

@@
 
+This paper can be seen as a first step towards formalising and solving problems such as these. In \S\ref{sec:defs} we introduce \emph{causal bandit problems} and focus on a
+
+\subsection{Related Work}
+
 One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem~\cite{Robbins1952} -- each of the six possible interventions is an arm and the crop yield is the reward -- and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
-However, as we show in \S\ref{sec:results}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret with a sub-optimal dependence on the number of possible interventions.
+However, as we show in \S\ref{sec:results}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret after $T$ round of $\bigomega{\sqrt{NT}}$ with a sub-optimal dependence on $N$, the number of possible interventions.
 
-There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \cite{TODO} as well as results for partial monitoring which, as first glance, may appear applicable.
+\todom{Note that feedback work as uses graphs}
+
+There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \cite{TODO} as well as results for partial monitoring which, at first glance, may appear applicable.
 As we show in Section~\ref{}, applying these results to our problem also exhibit a sub-optimal dependence on the number of available interventions.

### 2016-01-15 causal-bandit/icml-2015/setup.tex
Introduce general causal bandit problem.

@@
 \newcommand{\dirac}{\operatorname{Dirac}}
+\renewcommand{\vec}[1]{\boldsymbol{#1}}
+
+We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}.
+In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}.
+The \emph{causal model} is specified using a directed acyclic graph over a set of random variables $\mathcal{X}$ that defines a Bayesian network over $\mathcal{X}$ where an edge from variable $X$ to $X'$ is interpreted to mean that a change in the value of $X$ directly causes a change to the value of $X'$.
+An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\}$ with the empty intervention (where no variable is set) denoted $do()$.
+An intervention ``mutilates'' the original graph by removing edges into the variables in $\vec{X}$ and the resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$.
+Further details can be found in Chapter 21 of~\cite{Koller2009}.
+
+A learner for a casual bandit problem is given the casual model, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{settable} variables $\mathcal{S} \subset \mathcal{X}$, a \emph{reward variable} $Y \in \mathcal{O} \cap \mathcal{S}^c$, and the following game is played over $T$ rounds.
+In round $t$, the learner selects an intervention $do(\vec{X}_t = \vec{x}_t)$ for $\vec{X}_t \subseteq \mathcal{S}$.
+Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$ and the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$, and recevies the reward $Y_t$.
+We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a causal model with an edge connecting a single settable variable $X$ that can take on $K$ values to a reward variable $Y = r(X) + \eta$ for some arbitrary but unknown, real-valued function $r$ and a noise term $\eta$ that may depend on $X$.
+
+The expected reward for playing $do(\vec{X}_t = \vec{x}_t)$ is $\mu_{\vec{x}_t} := \E{Y_t | do(\vec{X}_t = \vec{x}_t)}$.
+
+The \emph{cumulative regret} measures the difference between the reward expected under
+the omnipotent strategy that knows the optimal action in advance and the expected reward of the learner and
+is denoted
+\eqn{
+\label{eq:regret}
+R_T = T \mu^* - \E{\sum_{t=1}^T Y_t}\,.
+}
+The second performance measure is the \emph{simple regret}, which measures the performance of the learner on the final round only.
+\eqn{
+\label{eq:regret-simple}
+\simpleregret = \mu^* - \E{\mu_{I_T,J_T}}\,.
+}
+The simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. The cumulative
+regret is more appropriate for online agents that may continue to adapt their policy.
+
+\subsection{Simple Causal Bandit Problem}
+
+In the remainder of this paper we focus on a causal model that is sufficient to demonstrate the separation of regret bounds between algorithms that make use of the additional causal information and those which do not.
 
@@
 \end{remark}
-
-We consider two standard performance measures. The first is the cumulative regret, which measures the difference between the reward expected under
-the omnipotent strategy that knows the optimal action in advance and the expected reward of the learner. 
-\eqn{
-\label{eq:regret}
-R_T = T \mu^* - \E{\sum_{t=1}^T Y_t}\,.
-}
-The second performance measure is the simple regret, which measures the performance of the learner on the final round only.
-\eqn{
-\label{eq:regret-simple}
-\simpleregret = \mu^* - \E{\mu_{I_T,J_T}}\,.
-}
-The simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. The cumulative
-regret is more appropriate for online agents that may continue to adapt their policy.

### 2016-01-17 causal-bandit/icml-2015/setup.tex
More on setup

@@
 In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}.
-The \emph{causal model} is specified using a directed acyclic graph over a set of random variables $\mathcal{X}$ that defines a Bayesian network over $\mathcal{X}$ where an edge from variable $X$ to $X'$ is interpreted to mean that a change in the value of $X$ directly causes a change to the value of $X'$.
+The \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
+An edge from variable $X$ to $X'$ is interpreted to mean that a change in the value of $X$ directly causes a change to the value of $X'$.
 An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\}$ with the empty intervention (where no variable is set) denoted $do()$.
@@
 
-A learner for a casual bandit problem is given the casual model, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{settable} variables $\mathcal{S} \subset \mathcal{X}$, a \emph{reward variable} $Y \in \mathcal{O} \cap \mathcal{S}^c$, and the following game is played over $T$ rounds.
+A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{settable} variables $\mathcal{S} \subset \mathcal{X}$, a \emph{reward variable} $Y \in \mathcal{O} \cap \mathcal{S}^c$, and the following game is played over $T$ rounds.
 In round $t$, the learner selects an intervention $do(\vec{X}_t = \vec{x}_t)$ for $\vec{X}_t \subseteq \mathcal{S}$. 
 Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$ and the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$, and recevies the reward $Y_t$.
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a causal model with an edge connecting a single settable variable $X$ that can take on $K$ values to a reward variable $Y = r(X) + \eta$ for some arbitrary but unknown, real-valued function $r$ and a noise term $\eta$ that may depend on $X$.
 
-The expected reward for playing $do(\vec{X}_t = \vec{x}_t)$ is $\mu_{\vec{x}_t} := \E{Y_t | do(\vec{X}_t = \vec{x}_t)}$.
-
-The \emph{cumulative regret} measures the difference between the reward expected under
-the omnipotent strategy that knows the optimal action in advance and the expected reward of the learner and
-is denoted
+We will consider and analyse two different goals for the learner: 1) identifying the intervention with highest expected reward by round $T$, and 2) maximising the total expected reward over all $T$ rounds.
+We denote the expected reward for the intervention $do(\vec{X} = \vec{x})$ by $\mu_{\vec{x}} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{\vec{x}} \mu_{\vec{x}}$.
+Performance for the first goal can be measured by \emph{simple regret}, \ie, the gap between the optimal expected reward and the intervention chosen by the learner on round $T$:
+\eqn{
+\label{eq:regret-simple}
+\simpleregret = \mu^* - \E{\mu_{\vec{x}_T}}\,.
+}
+Performance for the second goal is measured by \emph{cumulative regret}, the difference between the total reward obtained if the optimal intervention were known in advance and the expected reward of the learner:
 \eqn{
@@
 }
-The second performance measure is the \emph{simple regret}, which measures the performance of the learner on the final round only.
-\eqn{
-\label{eq:regret-simple}
-\simpleregret = \mu^* - \E{\mu_{I_T,J_T}}\,.
-}
-The simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. The cumulative
+Simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. Cumulative
 regret is more appropriate for online agents that may continue to adapt their policy.
+
+We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single settable variable $X$ that can take on $K$ values to a reward variable $Y = r(X) + \eta$ for some arbitrary but unknown, real-valued function $r$ and a noise term $\eta$ that may depend on $X$.
+Furthermore, any causal bandit problem can reduced to a classical stochastic bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
+Intuitively though, one would expect
 

### 2016-01-18 causal-bandit/icml-2015/setup.tex
Defined parallel banit problem in setup.

@@
 An edge from variable $X$ to $X'$ is interpreted to mean that a change in the value of $X$ directly causes a change to the value of $X'$.
-An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\}$ with the empty intervention (where no variable is set) denoted $do()$.
+An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
 An intervention ``mutilates'' the original graph by removing edges into the variables in $\vec{X}$ and the resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$.
@@
 
-A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{settable} variables $\mathcal{S} \subset \mathcal{X}$, a \emph{reward variable} $Y \in \mathcal{O} \cap \mathcal{S}^c$, and the following game is played over $T$ rounds.
-In round $t$, the learner selects an intervention $do(\vec{X}_t = \vec{x}_t)$ for $\vec{X}_t \subseteq \mathcal{S}$. 
-Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$ and the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$, and recevies the reward $Y_t$.
+A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{allowed interventions} $\mathcal{A}$, a \emph{reward variable} $Y$.
+The following game is played over $T$ rounds.
+In round $t$, the learner selects an allowed intervention $do(\vec{X}_t = \vec{x}_t)$ from $\mathcal{A}$.
+Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$.
+Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t$.
 
@@
 Simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. Cumulative
-regret is more appropriate for online agents that may continue to adapt their policy.
+regret is more appropriate for online agents that may continue to adapt their policy. We analyse both cases in
+\S\ref{sec:simple-regret} and \S\ref{sec:cum-regret}, respectively.
 
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single settable variable $X$ that can take on $K$ values to a reward variable $Y = r(X) + \eta$ for some arbitrary but unknown, real-valued function $r$ and a noise term $\eta$ that may depend on $X$.
+\todom{Mention pruning of actions?}
+We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y = r(X) + \eta$ for some arbitrary but unknown, real-valued function $r$ and a noise term $\eta$ that may depend on $X$.
+The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
 Furthermore, any causal bandit problem can reduced to a classical stochastic bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
-Intuitively though, one would expect
-
-\subsection{Simple Causal Bandit Problem}
+Intuitively though, one would expect to perform better by making use of the extra observations.
 
@@
 
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1},\ldots,X_{N}\}$ that independently cause a 
-target variable of interest $Y \in \R$ (see Figure \ref{fig:causalStructure}).
+\subsection{The Parallel Bandit Problem}
+
+We now consider a class of a causal bandit problems which we call \emph{parallel bandits}.
+It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example.
+
+The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
+All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable interventions are all size 1 interventions on a given $X_i$ variable (\ie, $\mathcal{A} = \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}$).
+In the farming example from the introduction, $X_1$ represent temperature with $X_1=0$ denoting low and $X_1=1$ high temperature.
+In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.
+
 \begin{figure}[h]
 \centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \end{tikzpicture}
+\caption{Causal model for the parallel bandits problem.\label{fig:causalStructure}}
 \end{figure}
 
-The game proceeds over $T$ identical rounds (or time-steps).
-In each round $t$ the learner can choose either to do nothing (denoted by $I_t = \bot$) or they can choose $I_t \in \set{1,\ldots,N}$ and
-an intervention $J_t \in \set{0,1}$. After the learner has chosen an intervention they observe $X_{t,i} \in \set{0,1}$ for all $i$ where
-\eq{
-X_{t,i} \sim \begin{cases}
-\dirac(J_t) & \text{if } I_t = i \\
-\bernoulli(q_i) & \text{otherwise}\,.
-\end{cases}
-}
-where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. 
-Note that if the learner did not choose an intervention ($I_t =  \bot$), then
-$X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
-Finally the learner observes the reward $Y_t = r(X_t) + \eta_t$ where 
-\eq{
-r : \set{0,1}^N \to \R
-}
-is arbitrary (and unknown) 
-and $\eta_t$ is a $1$-subgaussian noise
-term (with the distribution possibly dependent on $X_t$). 
-In the farming example $X_{t,1}$ might represent whether the plants had sufficient water in the $t$th season (controllable by installing irrigation) 
-while $X_{t,2}$ might be whether or not there was a spring frost (controllable with heat-lamps or covers).
+When not intervened upon, we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
+The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
+In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.
 
-The expected reward for intervening on variable $i$ by setting it to $j$ is defined by
+% \eq{
+% X_{t,i} \sim \begin{cases}
+% \dirac(J_t) & \text{if } I_t = i \\
+% \bernoulli(q_i) & \text{otherwise}\,.
+% \end{cases}
+% }
+% where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. 
+% Note that if the learner did not choose an intervention ($I_t =  \bot$), then
+% $X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
+% Finally the learner observes the reward $Y_t = r(X_t) + \eta_t$ where 
+% \eq{
+% r : \set{0,1}^N \to \R
+% }
+% is arbitrary (and unknown) 
+% and $\eta_t$ is a $1$-subgaussian noise
+% term (with the distribution possibly dependent on $X_t$). 
+% In the farming example $X_{t,1}$ might represent whether the plants had sufficient water in the $t$th season (controllable by installing irrigation) 
+% while $X_{t,2}$ might be whether or not there was a spring frost (controllable with heat-lamps or covers).
+
+In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
+We will also make use of the expansion of the expected reward for $do(X_i = j)$:
 \eq{
@@
 &= \E{r(X)|do(X_i = j)} \\
-&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k} \,. 
+&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
 }
-The optimal intervention is $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
+The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
+The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
 Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
 
-It is worth mentioning that the problem may be treated as a multi-armed bandit with $2N$ arms, one corresponding to each intervention.
-As we shall shortly see, this approach is usually not practical because the resulting algorithms do not exploit the structure in
-the covariates.
+% It is worth mentioning that the problem may be treated as a multi-armed bandit with $2N$ arms, one corresponding to each intervention.
+% As we shall shortly see, this approach is usually not practical because the resulting algorithms do not exploit the structure in
+% the covariates.
 
-\begin{remark}
-In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
-the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
-In the bandit community it is implicit that 
-algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
-However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
-intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
-deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
-bandit and causal inference communities.
-\end{remark}
+% \begin{remark}
+% In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
+% the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
+% In the bandit community it is implicit that 
+% algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
+% However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
+% intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
+% deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
+% bandit and causal inference communities.
+% \end{remark}
+
+\subsection{Exploiting the Causal Model}

### 2016-01-18 causal-bandit/icml-2015/setup.tex
Moved key insight bit into setup.

@@
 \subsection{Exploiting the Causal Model}
+
+From a causal perspective the key insight is that the assumed causal structure implies that
+the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
+on $X_{t,i} = j$. Formally,
+\eqn {
+\label{eq:observe}
+\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
+}
+This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph. It is not the case in general. 
+For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 
+that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the 
+interventional distribution of $Y_t$ from the observational one.
+
+\todot{Or just give a reference or hide in the appendix depending on space in the end}
+\color{red} Probably should define causal model and back-door rule to properly show this.\color{black}
+
+We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
+
+\eqn {
+\label{eq:estimation_transfer}
+P(Y_t|do(X_{t,i} = j))= & \nonumber \\
+ \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
+}

### 2016-01-18 causal-bandit/icml-2015/simple-regret.tex
updates
Moved key insight bit into setup.

@@
 causal bandit problem.
-
-From a causal perspective the key insight is that the assumed causal structure implies that
-the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
-on $X_{t,i} = j$. Formally,
-\eqn {
-\label{eq:observe}
-\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
-}
-This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph. It is not the case in general. 
-For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 
-that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the 
-interventional distribution of $Y_t$ from the observational one.
-
-\todot{Or just give a reference or hide in the appendix depending on space in the end}
-\color{red} Probably should define causal model and back-door rule to properly show this.\color{black}
-
-We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
-
-\eqn {
-\label{eq:estimation_transfer}
-P(Y_t|do(X_{t,i} = j))= & \nonumber \\
- \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
-}
 
@@
 Let $\boldsymbol{q} \in [0,1]^N$. Then define $q^{(i)}$ to be the $i$th largest value of the set $\set{\min \set{q_j, 1 - q_j} : 1 \leq j \leq N}$.
-For example, if $\boldsymbol{q} = (0,\, 2/3,\, 1/2)$, then $q^{(1)} = 0$, $q^{(2)} = 1/3$, $q^{(3)} = 1/2$.
-Then define 
+For example
+\eq{
+\boldsymbol{q} &= (0,\, 2/3,\, 1/2,\, 1) \\
+q^{(1)} &= 0,\,\, q^{(2)} = 0,\,\, q^{(3)} = 1/3,\,\, q^{(4)} = 1/2\,.
+}
+Define 
 \eqn{
@@
 
-\begin{theorem}\label{thm:unknown_q_simpleregret}
+\begin{theorem}\label{thm:uq-simple}
 Algorithm \ref{alg:simple} satisfies

### 2016-01-18 causal-bandit/icml-2015/setup.tex
More on parallel bandits.

@@
 }
-Simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. Cumulative
-regret is more appropriate for online agents that may continue to adapt their policy. We analyse both cases in
-\S\ref{sec:simple-regret} and \S\ref{sec:cum-regret}, respectively.
+Simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
+Cumulative regret is more appropriate for online agents that may continue to adapt their policy. 
+We analyse both cases in \S\ref{sec:simple-regret} and \S\ref{sec:cum-regret}, respectively.
 
@@
 The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Furthermore, any causal bandit problem can reduced to a classical stochastic bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
+Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
 Intuitively though, one would expect to perform better by making use of the extra observations.
@@
 
-When not intervened upon, we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
+In each round of the game, the learner selects a single variable to intervene on and the value to set it to while the remainder of the variables are simultaneously set by independently biased coin flips.
+The value of all variables are then used to determine the distribution of rewards for that round.
+Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
 The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
 In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.
-
-% \eq{
-% X_{t,i} \sim \begin{cases}
-% \dirac(J_t) & \text{if } I_t = i \\
-% \bernoulli(q_i) & \text{otherwise}\,.
-% \end{cases}
-% }
-% where $\boldsymbol{q} \in [0,1]^N$ is a (possibly unknown) vector of probabilities with $q_i = \P{X_i = 1}$. 
-% Note that if the learner did not choose an intervention ($I_t =  \bot$), then
-% $X_{t,i} \sim \bernoulli(q_i)$ for all variables $i \in \set{1,\ldots,N}$.
-% Finally the learner observes the reward $Y_t = r(X_t) + \eta_t$ where 
-% \eq{
-% r : \set{0,1}^N \to \R
-% }
-% is arbitrary (and unknown) 
-% and $\eta_t$ is a $1$-subgaussian noise
-% term (with the distribution possibly dependent on $X_t$). 
-% In the farming example $X_{t,1}$ might represent whether the plants had sufficient water in the $t$th season (controllable by installing irrigation) 
-% while $X_{t,2}$ might be whether or not there was a spring frost (controllable with heat-lamps or covers).
 
@@
 Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
-
-% It is worth mentioning that the problem may be treated as a multi-armed bandit with $2N$ arms, one corresponding to each intervention.
-% As we shall shortly see, this approach is usually not practical because the resulting algorithms do not exploit the structure in
-% the covariates.
-
-% \begin{remark}
-% In order to be consistent with the literature on causal inference we use the notation $do()$ to denote the action of doing nothing and $do(X_{t,I_t} = J_t)$
-% the action of intervening on the $I_t$th variable and setting it to equal $J_t$. 
-% In the bandit community it is implicit that 
-% algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. 
-% However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from 
-% intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we 
-% deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the 
-% bandit and causal inference communities.
-% \end{remark}
 
@@
 }
+
+\begin{definition}
+Let $\boldsymbol{q} \in [0,1]^N$. Then define $q^{(i)}$ to be the $i$th largest value of the set $\set{\min \set{q_j, 1 - q_j} : 1 \leq j \leq N}$.
+For example
+\eq{
+\boldsymbol{q} &= (0,\, 2/3,\, 1/2,\, 1) \\
+q^{(1)} &= 0,\,\, q^{(2)} = 0,\,\, q^{(3)} = 1/3,\,\, q^{(4)} = 1/2\,.
+}
+Define 
+\eqn{
+\label{def:m}
+m(\boldsymbol{q}) = \min\set{1 \leq i \leq N: q^{(i+1)} \geq \frac{1}{i}}\,.
+}
+\end{definition}

### 2016-01-19 causal-bandit/icml-2015/introduction.tex
Minor tweaks to setup

@@
 \todom{Note that feedback work as uses graphs}
+
+Our framework also bears a superficial similarity to contextual bandit problems \cite{Agarwal2014} since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
+However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intevention and cannot be used as context.
 

### 2016-01-19 causal-bandit/icml-2015/setup.tex
Minor tweaks to setup

@@
 A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{allowed interventions} $\mathcal{A}$, a \emph{reward variable} $Y$.
-The following game is played over $T$ rounds. 
+The following game is then played over $T$ rounds. 
 In round $t$, the learner selects an allowed intervention $do(\vec{X}_t = \vec{x}_t)$ from $\mathcal{A}$.
@@
 The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
-All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable interventions are all size 1 interventions on a given $X_i$ variable (\ie, $\mathcal{A} = \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}$).
-In the farming example from the introduction, $X_1$ represent temperature with $X_1=0$ denoting low and $X_1=1$ high temperature. 
+All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable actions are all size 0 and size 1 interventions: 
+\[
+	\mathcal{A} = \{do()\} \cup \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}.
+\]
+In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
 In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.
@@
 
-In each round of the game, the learner selects a single variable to intervene on and the value to set it to while the remainder of the variables are simultaneously set by independently biased coin flips.
+In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
+The remainder of the variables are simultaneously set by independently biased coin flips. 
 The value of all variables are then used to determine the distribution of rewards for that round.
@@
 }
-This follows from application of the do-calculus \cite{Pearl2000} to our specific causal graph. It is not the case in general. 
+This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (Fig.~\ref{fig:causalStructure}). It is not the case in general. 
 For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 

### 2016-01-21 causal-bandit/icml-2015/experiments.tex
code to run simple_regret experiments and results

@@
 \begin{figure}
-\caption{Simple regret vs horizon, $T$, for $N = 30$ for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$.
-Error bars show estimated 95\% confidence interval from 1000 simulations.}
+\caption{Simple regret vs horizon, $T$, with $N = 20$ and fixed $\epsilon = .4$ for Successive Rejects and Causal-Best-Arm-Identification. Error bars show standard error over 100000 simulations.}
 \label{fig:simple_vs_T}
 \centering
-\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_T_N30_sims1000_20160113_010512.pdf}
+\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_T_N20_sims100000_20160121_140227}
 \end{figure}
@@
 \begin{figure}
-\caption{Simple regret vs number of variables, $N$, for $T=500$, for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$.
-Error bars show estimated 95\% confidence interval from 1000 simulations.}
+\caption{Simple regret vs horizon, $T$, with $N = 20$ and $\epsilon \propto \sqrt{1/T}$ for Successive Rejects and Causal Best Arm Identification.
+Error bars show standard error over 10000 simulations.}
+\label{fig:simple_vs_T_vary_epsilon}
+\centering
+\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_T_N20_sims10000_20160122_010802.pdf}
+\end{figure}
+
+\begin{figure}
+\caption{Simple regret vs number of variables, $N$, for $T=250$, for Successive Rejects and, Causal-Best-Arm-Identification. Error bars show standard error from 10000 simulations.}
 \label{fig:simple_vs_N}
 \centering
-\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_N_T500_sims1000_20160113_010511.pdf}
+\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_N_T250_sims10000_20160122_024459.pdf}
 \end{figure}
+
+\begin{figure}
+\caption{Simple regret vs number of variables, $m$, for $T=300$ and $N = 50$, for Successive Rejects and Causal-Best-Arm-Identification. Error bars show standard error over 10000 simulations.}
+\label{fig:simple_vs_m}
+\centering
+\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_m_T300_N50_sims10000_10000.pdf}
+\end{figure}

### 2016-01-23 causal-bandit/icml-2015/discussion.tex
sketches roadmap to solve general graph problem

@@
 
+\cite{wu2015online} consider a relevant feedback model.
+
 \section{Future Open Questions}
+
+\subsection{General graphs without latent variables}
+
+The query $P(Y|do(V)$ is identifiable for any graph with known structure and fully observable variables.
+This means we can create an estimator for the reward of any action in terms only of results observed when the $do()$ action was selected (although the variance of these estimators can be infinite if any node conditional probabilities are zero).
+For example, consider the graph in \cref{fig:causalStructure_confounded}.
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+
+ %nodes
+
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){Y};
+\node[main node, above=of 3](6){$Z$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5)
+    (6) edge (1) edge (2) edge(4);
+
+\end{tikzpicture}
+\caption{Causal model for the confounded parallel bandits problem.\label{fig:causalStructure_confounded}}
+\end{figure}
+
+\eq {
+P(Y|do(X_i = j) = \sum_z P(Y|X_i = j,Z)P(Z)
+}
+
+In this case, the variance of the estimator is no longer simply a function of the number of times we observe $X_i = j$ but instead depends on the joint probability $P(X_i,Z)$.
+For example, if $P(z=1) = P(z=0) = .5$, $P(X_i = 1|z = 1) = 1$, and $P(X_i = 0|z = 0) = 1$ then $P(X_i = 1) = .5$ but we will will be unable to estimate $P(Y|do(X_i = 1))$ as we will never get any data for the term $P(Y|X_i = 1, Z = 0)$.
+
+We can generalize the (simple regret) algorithm developed for the parallel bandit problem to general graphs as follows.
+
 \begin{itemize}
-\item Known but arbitrary structure
-\item Learning structure then exploiting
+\item Apply the do-calculus to write an expression for the reward of each action in term of observational data: $P(Y|do(X_i=j)) = \sum_{pa_i} P(Y|X_i = j,pa_i)P(pa_i)$ \cite{Pearl2000} Theorem 3.2.2
+\item Create an estimator from this expression
+\item Observe for $T/2$ rounds.
+\item Calculate the variance in the estimators from $P(pa_i)$ and $P(X_i|pa_i)$ if they are known (corresponds to known $q$ case) or estimate the variance from the observations (as for unknown $q$ case).
+\item Compute the set of poorly estimated actions, $A$ based on the inverses of the variances.
+\item Explicitly play actions in $A$ for the remaining $T/2$ rounds.
+\end{itemize}
+
+The key challenge here is probably in computing the variance of the estimators. This would give us an upper bound for the regret for general graphs.
+
+It is unclear if this approach remains optimal.
+The $do()$ action is not the only one that can reveal information about the rewards of multiple other actions.
+In the example in figure \ref{fig:causalStructure_confounded}, $do(Z)$ is another obvious candidate.
+We would need to establish a matching lower bound - or prove that $do()$ is always the most revealing action.
+We are also not using the fact that the rewards are not entirely independent (They are 'almost' independent in the parallel bandit example).
+To what extent does this apply in general (provided there are many actions?).
+
+\subsection{General graphs with known structure but latent variables}
+In this case, some queries will be non-identifiable.
+The corresponding actions can be immediately added to the set $A$ prior to collecting any data.
+We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
+
+\subsection{Graphs with partially or unknown structure}
+
+\begin{itemize}
+
+\item Partially known structure
+There has been substantial work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class \cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014}.
+This problem arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques.
+Key results $bigo(n)$ singleton or $bigo(log log n)$ multi-variate experiments are required.
+- Focus on minimizing the number of experiments that must be performed. Examples of active versus online learning. The cost of experiments constant or at least known.
+
+\item Unknown Structure
+If we need to learn the structure, in an online environment. 
+Experiment is much, much more revealing than inference from observational data. We would expect it to dominate.
+
 \end{itemize}

### 2016-01-27 causal-bandit/icml-2015/general.tex
added new algorithm

\newcommand{\calP}{\mathcal P}

Assume $Y \in \set{0,1}$. Let $\calP_i : \set{0,1}^N \to \set{0,1}^{N_i}$ be the
projection that selections the $N_i$ parents of variable $i$. Then we can write
\eq{
\P{X = x} = \prod_{k=1}^N \P{X_k = x_k|\calP_k(X) = \calP_k(x)}\,.
}
If we intervene by setting $X_i = j$ where $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$, then
\eq{
\P{X = x|do(X_i = j)} = \prod_{k \neq i} \P{X_k = x_k|\calP_k(X) = \calP_k(x)}\,.
}
Let $r(x) = \P{Y = 1|X = x}$. Then
\eq{
\P{Y = 1} = \sum_{x \in \set{0,1}^N} \P{X = x} r(x)
}
and
\eq{
\mu_{ij} 
& = \P{Y = 1|do(X_i = j)} \\
& = \sum_{x \in \set{0,1}^N : x_i = j} \P{X = x|do(X_i = j)} r(x)\,.
}
We wish to devise an estimator for $\mu_{ij}$ using $X$ and $Y$ without
intervening to set $X_i = j$. A natural choice is the importance weighted estimator
\eq{
Z_{ij} 
&= \frac{Y \ind{X_i = j}}{\P{X_i = j|\calP_i(X)}}\,. 
}
Checking:
\eq{
&\E{Z_{ij}} 
= \sum_x \frac{\P{X = x} r(x) \P{X = x | do(X_i = j)}}{\P{X = x}} \\
&= \P{Y = 1| do(X_i = j)} \,. 
}
The variance of $Z_{ij}$ may also be bounded
\eq{
&\Var[Z_{ij}]
\leq \E{Z_{ij}^2} \\
&\leq \sum_x \P{X = x} \left(\frac{\ind{X_i = j}}{\P{X_i = j|\calP_i(X) = \calP_i(x)}}\right)^2 \\
&= \E{\frac{1}{\P{X_i = j|\calP_i(X)}}} \\
&= V_{ij}\,.
}
In the general graph setting the dimension $m$ has a slightly different definition.
\eq{
m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}\,.
}

\begin{algorithm}[H]
\caption{General Algorithm}
\begin{algorithmic}
\STATE {\bf Input:} $N$, $\mathcal P$
\STATE Compute $m$:
\eq{
m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}
}
\STATE Collect $T/2$ observational samples and estimate $\hat \mu_{ij}$:
\eq{
\hat \mu_{ij} = \sum_{t=1}^{T/2} \frac{Y_t \ind{X_{t,i} = j}}{\P{X_{t,i} = j | \mathcal P_i(X_t)}}
}
\FOR{$i \in \set{1,\ldots,m}$}
\STATE For each $i,j $ with $V_{ij} < 1/m$ take action $do(X_i = j)$ for $T/(2m)$ samples $Y^{(ij)}_1,\ldots,Y^{(ij)}_{T/2m}$
\STATE Update estimate:
\eq{
\eta &= \frac{V_{ij}}{V_{ij} + m} \\
\hat \mu_{ij} &= \eta \left(\frac{2m}{T} \sum_{t=1}^{T/(2m)} Y^{(ij)}_t\right) + (1 - \eta) \hat \mu_{ij}
}
\ENDFOR
\STATE Choose $do(X_i = j)$ for $ij = \argmax_{ij} \hat \mu_{ij}$
\end{algorithmic}
\end{algorithm}

### 2016-01-28 causal-bandit/icml-2015/introduction.tex
Intro paragraph.

@@
-A classical problem in economics and operations research is how to choose interventions that maximise utility. 
-We study an idealised model of such interventions using a combination of the multi-armed bandit framework \citep{Robbins1952} and the theory/language of causal inference 
-\citep{Pearl2000}.
+Medical drug testing, economic policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1958}.
+In this framework, single actions or experiments from a pre-determined set are repeatedly performed in order to determine the most effective one via feedback from a single, real-valued reward signal. 
+We consider the more general setting for sequential experiments where the feedback after each action consists of the reward signal as well as the values of other variables drawn from a probabilistic causal model~\citep{Pearl2000}.
+By explicitly modeling this extra information (\eg, drug side-effects, secondary economic indicators) we show, theoretically and empircally, how it can be used to improve the rate at which effective actions can be identified.
 
-\todom{Describe feature of class of problem; highlight with examples.}
+% A classical problem in economics and operations research is how to choose interventions that maximise utility. 
+% We study an idealised model of such interventions using a combination of the multi-armed bandit framework \citep{Robbins1952} and the theory/language of causal inference 
+% \citep{Pearl2000}.
 
@@
 
-This paper can be seen as a first step towards formalising and solving problems such as these. In \S\ref{sec:defs} we introduce \emph{causal bandit problems} and focus on a
+This paper can be seen as a first step towards formalising and solving problems such as these.
+In \S\ref{sec:defs} we introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their inlfuence on the reward --- along with any other non-interventional observations --- is assumed to conform to a known causal graph.
+Although the general problem subsumes classical stochastic bandit problems as well as stochastic contextual bandits ... focus on parallel bandit problem.
+
+Our algorithm simply observes without intervening for some initial period and then plays the undersampled actions in order to get good estimates of their reward. 
+These estimates rely on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
+
+We provide matching upper and lower bounds on the simple regret of this algorithm that have a better dependence on the number of actions than any algorithm that ignores the extra observations. 
+Central to our new bounds is a quantity that measures how ``balanced'' the distributions over the observed variables are.
 
@@
 One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem~\cite{Robbins1952} -- each of the six possible interventions is an arm and the crop yield is the reward -- and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
-However, as we show in \S\ref{sec:results}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret after $T$ round of $\bigomega{\sqrt{NT}}$ with a sub-optimal dependence on $N$, the number of possible interventions.
+However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret after $T$ round of $\bigomega{\sqrt{NT}}$ with a sub-optimal dependence on $N$, the number of possible interventions.
 

### 2016-01-28 causal-bandit/icml-2015/general.tex

@@
 \newcommand{\calP}{\mathcal P}
+\newcommand{\x}{\boldsymbol{x}}
 
-Assume $Y \in \set{0,1}$. Let $\calP_i : \set{0,1}^N \to \set{0,1}^{N_i}$ be the
-projection that selections the $N_i$ parents of variable $i$. Then we can write
-\eq{
-\P{X = x} = \prod_{k=1}^N \P{X_k = x_k|\calP_k(X) = \calP_k(x)}\,.
+We now consider single variable interventions on general causal graphs over variables $\boldsymbol{V}$.
+Let $Y \in \set{0,1}$ be the target variable and $\boldsymbol{X} = V/\set{Y \cup Desendants(Y)}$.
+
+If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ from data collected without intervening via the truncated product formula \cite{Pearl2000} 1.37.
+
+\eq {
+P(Y,X) = \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\, \text{  $\forall \,\x$ consistent with $X_i = j$}  
 }
-If we intervene by setting $X_i = j$ where $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$, then
-\eq{
-\P{X = x|do(X_i = j)} = \prod_{k \neq i} \P{X_k = x_k|\calP_k(X) = \calP_k(x)}\,.
+
+Where $\calP_i : \set{0,1}^N \to \set{0,1}^{N_i}$ is the
+projection that selections the $N_i$ parents of variable $i$.
+
+We can then marginalize to obtain $P(Y|do(X_i = j))$
+
+\eq {
+P(Y|do(X_i = j)) =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\\
+ =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\x}\\
+ =& \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}\P{Y|\x}
 }
+
 Let $r(x) = \P{Y = 1|X = x}$. Then
-\eq{
-\P{Y = 1} = \sum_{x \in \set{0,1}^N} \P{X = x} r(x)
+
+\eq {
+\mu_{ij} = P(Y = 1|do(X_i = j)) = \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}r(x)
 }
-and
-\eq{
-\mu_{ij} 
-& = \P{Y = 1|do(X_i = j)} \\
-& = \sum_{x \in \set{0,1}^N : x_i = j} \P{X = x|do(X_i = j)} r(x)\,.
-}
+
 We wish to devise an estimator for $\mu_{ij}$ using $X$ and $Y$ without
 intervening to set $X_i = j$. A natural choice is the importance weighted estimator
+
 \eq{
@@
 }
+
 Checking:

### 2016-01-29 causal-bandit/icml-2015/introduction.tex
Summarise NIPS 2015 paper

@@
-Medical drug testing, economic policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1958}.
-In this framework, single actions or experiments from a pre-determined set are repeatedly performed in order to determine the most effective one via feedback from a single, real-valued reward signal. 
-We consider the more general setting for sequential experiments where the feedback after each action consists of the reward signal as well as the values of other variables drawn from a probabilistic causal model~\citep{Pearl2000}.
-By explicitly modeling this extra information (\eg, drug side-effects, secondary economic indicators) we show, theoretically and empircally, how it can be used to improve the rate at which effective actions can be identified.
+Medical drug testing, policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1959}.
+In this framework, single actions (\eg, experiments or interventions) from a pre-determined set are repeatedly performed in order to determine the most effective one via feedback from a single, real-valued reward signal.
+We propose generalizing such sequential experiments by assuming that, in additional to the reward signal, values of other variables are drawn from a probabilistic causal model~\citep{Pearl2000} and observed.
+These causal models are common in social science, demography, and economics where explicit experimentation may be difficult. 
+For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades. 
+Results from causal inference relate observational distributions to interventional ones, allowing us to predict the outcome of intervention without explicitly performing it. 
+By combining causal inference using a causal graph we show, theoretically and empirically, how non-interventional observations can be used to improve the rate at which high-reward actions can be identified.
 
@@
 
-The challenge is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. 
+The type of problem we are concerned with is best illustrated with an example. 
+Consider a farmer wishing to optimise the yield of her crop. 
 She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
@@
 
+\subsection{Contributions}
+
 This paper can be seen as a first step towards formalising and solving problems such as these. 
-In \S\ref{sec:defs} we introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their inlfuence on the reward --- along with any other non-interventional observations --- is assumed to conform to a known causal graph.
-Although the general problem subsumes classical stochastic bandit problems as well as stochastic contextual bandits ... focus on parallel bandit problem.
+In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph.
+We show that our general model subsumes classical stochastic bandit problems as well as stochastic contextual bandits before focusing on a special case where, like the above example, the
 
@@
 
-One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem~\cite{Robbins1952} -- each of the six possible interventions is an arm and the crop yield is the reward -- and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
+One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem (\eg, the six possible farming interventions is an arm and the crop yield is the reward) and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
 However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret after $T$ round of $\bigomega{\sqrt{NT}}$ with a sub-optimal dependence on $N$, the number of possible interventions.
 
-\todom{Note that feedback work as uses graphs}
-
-Our framework also bears a superficial similarity to contextual bandit problems \cite{Agarwal2014} since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
+Our framework also bears a superficial similarity to contextual bandit problems \cite{Langford2008,Agarwal2014} since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
 However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intevention and cannot be used as context.
@@
 There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \cite{TODO} as well as results for partial monitoring which, at first glance, may appear applicable.
-As we show in Section~\ref{}, applying these results to our problem also exhibit a sub-optimal dependence on the number of available interventions.
+Indeed, we show in Section~\ref{}, that it is possible to applying these existing results to our problem, however the resulting regret is $\bigo{\sqrt{NT}}$, again with a sub-optimal dependence on $N$.
 
@@
 
-An alternate approach to selecting actions is causal inference.
-Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
-This allows an agent to predict the outcome of an action based on non-experimental data.
-This approach is common in social science, demography, and economics where explicit experimentation may be difficult.
-For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.
-
 We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
@@
 
-A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph.
+% A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph.
 Each possible assignment of variables to values is an action (bandit arm).
@@
 
-Problems framed in this way take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected.
+% Problems framed in this way take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected.
 If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
 
-If we receive feedback on additional nodes the problem can be more interesting.
+% If we receive feedback on additional nodes the problem can be more interesting.
 In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
 
-We consider a bandit problem where the actions and reward are represented by a specific causal graph that demonstrates this interesting structure.
+% We consider a bandit problem where the actions and reward are represented by a specific causal graph that demonstrates this interesting structure.
 We develop an algorithm to leverage the information provided by this structure and demonstrate it substantially outperforms standard bandit algorithms applied to the same problem where the number of actions is large.
@@
 
-Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
+% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
 It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
@@
 
-ALSO NEED TO MENTION ANY OTHER COMBINATIONS OF BANDITS+CAUSAL (eg the Elias NIPS paper and Generalized Thompson Sampling paper)
+\citet{Bareinboim2015} very recently presented work that also combines bandits and causal inference.
+Their work differs from our in two key ways.
+Firstly, their focus is on correcting the effects of unobserved confounding variables whereas we assume all variables are observable.
+Secondly, they do not assume their learning algorithm has access to a causal graph and instead use a form of contextual randomization to improve reward estimates.
+Combining their handling of unobserved variables with our analysis is left as future work.
 
-Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
+\citet{Ortega2014}
+
+% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
 Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.

### 2016-01-29 causal-bandit/icml-2015/introduction.tex
Add Ortega summary

@@
 
-\citet{Bareinboim2015} very recently presented work that also combines bandits and causal inference.
-Their work differs from our in two key ways.
+Two pieces of recent work also consider using causal models in bandit problems.
+Although \citet{Bareinboim2015} also use causal inference to improve the chance of choosing an optimal arm, their work differs from our in two key ways.
 Firstly, their focus is on correcting the effects of unobserved confounding variables whereas we assume all variables are observable.
-Secondly, they do not assume their learning algorithm has access to a causal graph and instead use a form of contextual randomization to improve reward estimates.
-Combining their handling of unobserved variables with our analysis is left as future work.
-
-\citet{Ortega2014}
-
+Secondly, they do not assume their learning algorithm has access to a causal graph and instead use a form of contextual randomization to improve reward estimates in a Thompson sampling-style algorithm.
+\citet{Ortega2014} also present an analysis and extension of Thompson sampling assuming actions are causal interventions.
+Their focus, however, is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
+Neither work provides a regret analysis.
+Combining their handling of unobserved variables and causal induction with our analysis is left as future work.
 % Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.

### 2016-01-29 causal-bandit/icml-2015/setup.tex
Note on ordering of act-observe-reward

@@
 Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t$.
+
+Although we will focus on the intervene-observe-reward ordering of events within each round, others are also possible. 
+If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already well understood~\citep{Agarwal2014}. 
+Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of allowable interventions thereby reducing the complexity.
 
@@
 It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example.
+It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
 

### 2016-01-29 causal-bandit/icml-2015/unused/pgm.py
adds a model build with libpgm

@@ -1,0 +1,129 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Fri Jan 29 06:53:48 2016
+
+@author: finn
+"""
+from libpgm.nodedata import NodeData
+from libpgm.graphskeleton import GraphSkeleton
+from libpgm.discretebayesiannetwork import DiscreteBayesianNetwork
+from libpgm.tablecpdfactorization import TableCPDFactorization
+from libpgm.pgmlearner import PGMLearner
+import numpy as np
+import itertools
+import pandas as pd
+
+class DiscretePGM(object):
+    def __init__(self):
+        self.nodes = {}
+    
+    def addNode(self,name,values,parents,cprob):
+        """     "numoutcomes": <number of possible outcome values>,
+            "vals": ["<name of value 1>", ... , "<name of value n>"],
+            "parents": ["<name of parent 1>", ... , "<name of parent n>"],
+            "children": ["<name of child 1>", ... , "<name of child n>"],   
+            "cprob": an m * numoutcomes array, where m is number of joint assingments for the parents.  """
+        if parents == None:
+            assert sum(cprob) == 1, "probabilities do not sum to 1"
+            assert len(cprob) == len(values),"number of columns of cprob do not match numoutcomes"
+        else:
+            parents = [str(p) for p in parents]
+            parent_values = [self.nodes[p]["vals"] for p in parents]
+            combinations = list(itertools.product(*parent_values))
+            assert len(combinations) == cprob.shape[0], "number of rows in cprob inconsistent with number of parent outcomes"
+            assert len(values) == cprob.shape[1],"number of columns of cprob does not match numoutcomes"
+            cprob_dict = {}
+            for row,c in enumerate(combinations):
+                cprob_dict[str(list(c))] = cprob[row]
+                assert sum(cprob[row]) == 1, "probabilities do not sum to 1"
+            cprob = cprob_dict
+        
+        self.nodes[name] = {"numoutcomes":len(values),"vals":[str(v) for v in values],"parents":parents,"cprob":cprob,"children":[]}
+        
+    def construct(self):
+        skel = GraphSkeleton()
+        skel.V = self.nodes.keys()
+        skel.E = []
[…]

### 2016-01-29 causal-bandit/icml-2015/introduction.tex

@@
 
-This paper can be seen as a first step towards formalising and solving problems such as these. 
+This paper can be seen as a first step towards formalising and solving problems such as the one above. 
 In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph.
-We show that our general model subsumes classical stochastic bandit problems as well as stochastic contextual bandits before focusing on a special case where, like the above example, the
+We show that our causal bandit framework subsumes the classical bandits (no additional observations) and contextual stochastic bandit problems (observations are revealed before an intervention is chosen) before focusing on the case where, like the above example, observations occur \emph{after} each intervention is made.
 
-Our algorithm simply observes without intervening for some initial period and then plays the undersampled actions in order to get good estimates of their reward. 
-These estimates rely on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
+Our main results for this case are: an upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?); an upper bound on the cumulative regret (Theorem~\ref{thm:}) for a specific causal model that has a better dependence on the number of available actions then the best possible algorithm which ignores the causal model; and a matching lower bound (Theorem~\ref{thm:}) for any algorithm which does use the additional causal information.
+Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
+Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.
 
-We provide matching upper and lower bounds on the simple regret of this algorithm that have a better dependence on the number of actions than any algorithm that ignores the extra observations. 
-Central to our new bounds is a quantity that measures how ``balanced'' the distributions over the observed variables are.
+Although the witness (Algorithm~\ref{alg:}) for our upper bound is very simple (\ie, observe without intervening for some initial period and then play undersampled actions in order to get good estimates) our key point is that algorithms which use causal information when available have a clear, theoretical improvement over those which do not.
+These results are experimentally validated in \S\ref{sec}.
 
@@
 Secondly, they do not assume their learning algorithm has access to a causal graph and instead use a form of contextual randomization to improve reward estimates in a Thompson sampling-style algorithm.
-\citet{Ortega2014} also present an analysis and extension of Thompson sampling assuming actions are causal interventions.
+\citet{Ortega2014thompson} also present an analysis and extension of Thompson sampling assuming actions are causal interventions.
 Their focus, however, is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.

### 2016-01-31 causal-bandit/icml-2015/tech.tex
updates

\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
\newcommand{\EE}{\mathbb E}
\newcommand{\Pn}[2]{\operatorname{P}_{#1}\left\{#2\right\}}

\subsection*{Truncated Importance Sampling}

Let $P$ and $Q$ be distributions on $\Omega = \set{0,1}^N$ and let $Y \in \set{0,1}$ be defined in terms of its conditional expectation given $X \in \Omega$.
\eq{
\forall x \in \Omega \qquad \EE[Y|X = x] = r(x)\,,
}
where $r:\Omega \to [0,1]$ is some unknown function.
We tackle the problem of estimating $\mu = \EE_Q r = \sum_x \Q{x} r(x)$ using
samples of $X$ and $Y$ with $X$ sampled from $P$.
Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
\eqn{
\label{eq:truncated}
Z = Y R \ind{R < B}\,,
}
which trivially satisfies $|Z| \leq B$ surely.
Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
Computing the expectation we see
\eq{
\EE_P[Z] = \mu - \beta\,.
}
where $\beta \geq 0$ is the negative bias and satisfies
\eq{
\beta \leq \Q{R \geq B}\,.
}
Bounding the variance is also straightforward.
\eq{
\Var[Z] 
\leq \EE[Z^2] 
=\sum_x \P{X = x} Z(x)^2 
\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 
\leq \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
= V\,.
}
The bias can be controlled in terms of $V$ via Markov's inequality.
\eq{
\beta 
\leq \Q{R \geq B} 
\leq \frac{\EE_Q[R]}{B}
= \frac{V}{B}\,.
}
Finally let $\delta > 0$ be the desired confidence level and choose 
\eq{
B = \sqrt{\frac{V n}{\log\frac{1}{\delta}}}\,,
}
which implies that the negative bias satisfies
\eq{
\beta \leq \sqrt{\frac{V}{n} \log \frac{1}{\delta}}\,. 
}
Let $Z_1,\ldots,X_n$ be $n$ i.i.d.\ copies of $Z$. Then by the Bernstein inequality we have
\eq{
\P{\left|\frac{1}{n} \sum_{t=1}^n Z - \mu\right| \geq \sqrt{\frac{V}{n} \log\frac{1}{\delta}}} \leq \delta\,.
}

\subsection*{Mixing Truncate Importance Sampling}

Now we study the problem of estimating $\mu = \EE_Q \P{Y = 1|X}$ from samples drawn
from a set of sampling distributions $\operatorname{P}_1,\ldots,\operatorname{P}_k$ with $N_i$ samples
drawn of $\operatorname{P}_i$.
Let $R_i(x) = \Q{x} / \Pn{i}{x}$ and
\eq{
V_i = \sum_{x \in \Omega} \Q{X = x} \frac{\Q{X = x}}{\Pn{i}{X = x}}
}
and
\eq{
Z_i = Y R_i \ind{R_i < B_i}\,,
}
where
\eq{
B_i = V_i \sqrt{\frac{\sum_{i=1}^k \frac{N_i}{V_i}}{\log \frac{1}{\delta}}}\,,
}
which for $k = 1$ reduces to the choice of $B$ in the previous section.
Then $\EE_{\operatorname{P_i}} Z_i = \mu - \beta$ where
\eq{
0 \leq \beta_i 
\leq \frac{V_i}{B_i}
= \sqrt{\frac{1}{\sum_{i=1}^k \frac{N_i}{V_i}} \log\frac{1}{\delta}}\,.
}
Let $\eta_i \in [0,1]$ be a mixture coefficient that minimises the variance of the mixture 
estimator, which weights samples from $\operatorname{P_i}$ by a bound on its inverse variance, which is $V_i / N_i$.
\eq{
\eta_i = \frac{\frac{N_i}{V_i}}{\sum_k \frac{N_k}{V_k}}\,.
}
Let $Z_{i,1},\ldots,Z_{i,N_i}$ be $N_i$ copies of $Z_i$ for each $i$.
Then we define the mixture estimator by
\eq{
\hat \mu = \frac{\sum_i \eta_i \sum_{t=1}^{N_i} \frac{Z_{it}}{N_i}}{\sum_i \eta_i} 
}
Now we prepare to use Bernstein's inequality to control the concentration of $\hat \mu$ about its mean.
\eq{
b 
= \max_i \frac{\frac{\eta_i B_i}{N_i}}{\sum_i \eta_i} 
= \max_i \frac{\alpha}{\sum_i \frac{N_i}{V_i}} 
= \frac{1}{\alpha \sqrt{\log\frac{1}{\delta}}}\,.
}
Therefore
\eq{
\P{\left|\hat \mu - \mu\right| \geq \sqrt{\frac{1}{\sum_{i=1}^k \frac{N_k}{V_k}} \log \frac{1}{\delta}}} \leq \delta\,.
}

\subsection*{General Case}

Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.

\begin{algorithm}[H]
\caption{CheckValid}\label{alg:alloc}
\begin{algorithmic}
\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_1,\ldots,\operatorname P_N$
\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calP_i \subseteq \set{o_k : k < i}$ for all $i$
\FOR{$i \in \set{o_1,\ldots,o_N}$}
\STATE Compute variances:
\eq{
\forall k \in \calP_i \text{ and } l \in \set{0,1} \qquad
V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
}
\STATE Compute variance for learning from do nothing action:
\eq{
V_{ij} = \frac{2}{T} \EE_{\operatorname P_{ij}} \!\!\left[\frac{\Pij{x}}{\P{x}}\right]
}
\STATE Compute required budget: 
\eq{
N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calP_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
} 
\ENDFOR
\IF{$\sum_{i=1}^N \sum_{j \in \set{0,1}} N_{ij} \leq T / 2$}
\STATE {\bf return true}
\ELSE
\STATE {\bf return false}
\ENDIF
\end{algorithmic}
\end{algorithm}

\begin{theorem}
Let $\delta = 1/(2NT)$ and $\epsilon$ be the smallest value for which \cref{alg:alloc} returns true. 
Let $N_{ij}$ as computed by \cref{alg:alloc} be the number of times the algorithm selects the action $do(X_i = j)$ 
and computes estimates using the mixture estimator described in the previous section. Choosing $I_T = \argmax_{ij} \hat \mu_{ij}$ leads
to
\eq{
\EE[r_T] \leq \epsilon + \frac{1}{T}\,.
}
\end{theorem}

### 2016-01-31 causal-bandit/icml-2015/general.tex
modifies definition of m for general case
some explaination of general case, changes to earlier sections to improve flow to general case

@@
 
-If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ from data collected without intervening via the truncated product formula \cite{Pearl2000} 1.37.
+If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ via the truncated product formula \cite{Pearl2000} 3.11
 
 \eq {
-P(Y,X) = \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\, \text{  $\forall \,\x$ consistent with $X_i = j$}  
+\P{\boldsymbol{x},y|do(X_i = j)} = 
+\begin{cases}
+\frac{\P{\boldsymbol{x},y}}{\P{x_i=j|\calP_i(\boldsymbol{x})}} & \text{if} x_i = j  \\
+0 & \text{otherwise}
+\end{cases}
 }
@@
 
-We can then marginalize to obtain $P(Y|do(X_i = j))$
+Since estimating $\P{Y|do(X_i=j)}$ from observational data requires conditioning on the parents of $X_i$, the variance of our estimators depends on $\P{X_i=j|\calP_i}$, rather than simply being a function of the number of times we observe $X_i=j$ as in the parallel bandit case.
+For example consider the graph in \cref{fig:causalStructure_confounded}.
+In this model
+$P(Y|do(X_i = j) = \sum_z P(Y|X_i = j,Z)P(Z)$.
+Let $P(z=1)=.5$, $P(X_i = 1|z = 1) = 1$, and $P(X_i = 0|z = 0) = 1$ then $P(X_i = 1) = .5$ but we will be unable to estimate $P(Y|do(X_i = 1))$ as we will never get any data for the term $P(Y|X_i = 1, Z = 0)$.
 
-\eq {
-P(Y|do(X_i = j)) =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\\
- =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\x}\\
- =& \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}\P{Y|\x}
+\begin{figure}[h]
+\centering
+\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+
+ %nodes
+
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){Y};
+\node[main node, above=of 3](6){$Z$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5)
+    (6) edge (1) edge (2) edge(4);
+
+\end{tikzpicture}
+\end{figure}
+
+For general models, it is also insufficient to only trade off learning about multiple actions by observing against learning individual actions by explicitly playing them.
+We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
+
+\eqn {
+\label{eq:estimation_transfer}
+P(Y_t|do(X_{t,i} = j))= & \nonumber \\
+ \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
 }
+
+Consider the graph in \cref{fig:causalchain}, where each variable deterministically takes the value of its parent,$X_k = X_{k-1}$ for $k\in {N...2}$ and $\P{X_1} = 0$.
+We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
+If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to explicitly play each of these arms individually to estimate their rewards.
+
+\begin{figure}[h]
+\centering
+\caption{A causal chain graph}.
+\label{fig:causalchain}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, right=of 4](5){Y};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2)
+    	(2) edge (3)
+    (3) edge (4)
+    (4) edge (5);
+\end{tikzpicture}
+\end{figure}
+
+\todof{From here onwards needs to be replaced by truncated importance sampling version}
 
@@
 
+\todof{I think the above should be}
+
+\eq {
+m = \min \set{m: \sum_{ij} \ind{V_{ij} \geq m} = m}\,.
+}
+
+That is, $m$ should be chosen such that the size of the set variances greater than $m$ is $m$.
+This ensures all the variances of the remaining arms are smaller than $m$.
+This definition would also seem equivalent to sorting the variances such that $V^{(0)} \geq V^{(1)} ...
+\geq V^{(2N)}$ and then defining $m = \min m: V^{(m)} \leq m$ which is in turn equivalent to the definition in terms of $q$ for parallel bandits.
+
 \begin{algorithm}[H]

### 2016-01-31 causal-bandit/icml-2015/setup.tex
some explaination of general case, changes to earlier sections to improve flow to general case

@@
 }
-This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (Fig.~\ref{fig:causalStructure}). It is not the case in general. 
-For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 
-that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the 
-interventional distribution of $Y_t$ from the observational one.
+This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (\cref{fig:causalStructure}). It is not the case in general. 
+For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$),that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the interventional distribution of $Y_t$ from the observational one.
 
-\todot{Or just give a reference or hide in the appendix depending on space in the end}
-\color{red} Probably should define causal model and back-door rule to properly show this.\color{black}
+This enables us to estimate the the rewards for multiple interventions simultaneously by observing the values taken by the variables when we select the $do()$ action.
+Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and need to be explored separately, by explicitly making the intervention.
 
-We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
-
-\eqn {
-\label{eq:estimation_transfer}
-P(Y_t|do(X_{t,i} = j))= & \nonumber \\
- \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
-}
+To exploit the causal structure we must optimally balance the collection of observational data against making interventions to learn about low probability events.
+The problem is more challenging because the probabilities $P(X_i=j)$ are unknown and must also be estimated.
 
 \begin{definition}
-Let $\boldsymbol{q} \in [0,1]^N$. Then define $q^{(i)}$ to be the $i$th largest value of the set $\set{\min \set{q_j, 1 - q_j} : 1 \leq j \leq N}$.
-For example
-\eq{
-\boldsymbol{q} &= (0,\, 2/3,\, 1/2,\, 1) \\
-q^{(1)} &= 0,\,\, q^{(2)} = 0,\,\, q^{(3)} = 1/3,\,\, q^{(4)} = 1/2\,.
-}
+Let $V_{ij} = \frac{1}{P(X_i = j)}$, for $i \in \set{1...N}$ and $j \in {0,1}$. 
 Define 
@@
 \label{def:m}
-m(\boldsymbol{q}) = \min\set{1 \leq i \leq N: q^{(i+1)} \geq \frac{1}{i}}\,.
+m(\boldsymbol{V}) = \min \set{m: \sum_{ij} \ind{V_{ij} \geq m} = m}\,.
 }

### 2016-01-31 causal-bandit/icml-2015/simple-regret.tex
More on parallel bandits.
some explaination of general case, changes to earlier sections to improve flow to general case

@@
 causal bandit problem.
-
-The main idea of the algorithm is that by making no intervention (doing nothing) it is possible to estimate simultaneously the returns of all interventions.
-Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and should be
-explored separately by actually making the intervention. The algorithm works by optimally balancing the collection of observational data (when
-no action is taken) and making interventions to learn about low probability events. The problem is more challenging because $\boldsymbol{q}$ is
-unknown and must also be estimated.
 
@@
 }
-\STATE Compute $\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} X_{t,i}$
-\STATE Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\STATE Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\STATE Compute $\hat m = \min\set{1 \leq i \leq N : \hat s'_{i+1} \geq \frac{1}{i}}$
+\STATE Compute $\hat V_{ij} = \frac{T}{2 \sum_{t=1}^{T/2}\ind{X_{t,i}=j}} $
+\STATE Compute $\hat m = m(\hat{\boldsymbol{V}})$
 \STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\STATE Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
+\STATE Compute $A$ as the subset of infrequently observed arms $\set{(i,j):V_{ij} \geq m}$ with $|A| = \hat m$
 
@@
 \end{algorithm}
-
-\begin{definition}
-Let $\boldsymbol{q} \in [0,1]^N$. Then define $q^{(i)}$ to be the $i$th largest value of the set $\set{\min \set{q_j, 1 - q_j} : 1 \leq j \leq N}$.
-For example
-\eq{
-\boldsymbol{q} &= (0,\, 2/3,\, 1/2,\, 1) \\
-q^{(1)} &= 0,\,\, q^{(2)} = 0,\,\, q^{(3)} = 1/3,\,\, q^{(4)} = 1/2\,.
-}
-Define 
-\eqn{
-\label{def:m}
-m(\boldsymbol{q}) = \min\set{1 \leq i \leq N: q^{(i+1)} \geq \frac{1}{i}}\,.
-}
-\end{definition}
 

### 2016-01-31 causal-bandit/icml-2015/discussion.tex
some explaination of general case, changes to earlier sections to improve flow to general case
moved stuff into more relevent places

--- 2016-01-13 causal-bandit/icml-2015/discussion.tex
@@
 
+\cite{wu2015online} consider a relevant feedback model.
+
 \section{Future Open Questions}
+
+\subsection{General graphs with known structure but latent variables}
+In this case, some queries will be non-identifiable.
+The corresponding actions can be immediately added to the set $A$ prior to collecting any data.
+We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
+
+\subsection{Graphs with partially or unknown structure}
+
 \begin{itemize}
-\item Known but arbitrary structure
-\item Learning structure then exploiting
+
+\item Partially known structure
+There has been substantial work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class \cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014}.
+This problem arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques.
+Key results $bigo(n)$ singleton or $bigo(log log n)$ multi-variate experiments are required.
+- Focus on minimizing the number of experiments that must be performed. Examples of active versus online learning. The cost of experiments constant or at least known.
+
+\item Unknown Structure
+If we need to learn the structure, in an online environment. 
+Experiment is much, much more revealing than inference from observational data. We would expect it to dominate.
+
 \end{itemize}

### 2016-01-31 causal-bandit/icml-2015/general.tex
much shuffling

@@
 Since estimating $\P{Y|do(X_i=j)}$ from observational data requires conditioning on the parents of $X_i$, the variance of our estimators depends on $\P{X_i=j|\calP_i}$, rather than simply being a function of the number of times we observe $X_i=j$ as in the parallel bandit case.
-For example consider the graph in \cref{fig:causalStructure_confounded}.
+For example consider the graph in Figure \ref{fig:causalStructure_confounded}.
 In this model
@@
 
-Consider the graph in \cref{fig:causalchain}, where each variable deterministically takes the value of its parent,$X_k = X_{k-1}$ for $k\in {N...2}$ and $\P{X_1} = 0$.
+Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent,$X_k = X_{k-1}$ for $k\in {N...2}$ and $\P{X_1} = 0$.
 We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.

### 2016-01-31 causal-bandit/icml-2015/setup.tex
much shuffling

@@
 If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already well understood~\citep{Agarwal2014}. 
-Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of allowable interventions thereby reducing the complexity.
+Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of 
+allowable interventions thereby reducing the complexity.
 
@@
 }
-This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (\cref{fig:causalStructure}). It is not the case in general. 
+This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (Figure \ref{fig:causalStructure}). It is not the case in general. 
 For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$),that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the interventional distribution of $Y_t$ from the observational one.

### 2016-01-31 causal-bandit/icml-2015/simple-regret.tex
much shuffling

@@
 this setting.
+We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}
 
-The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound.
+\begin{theorem}\label{thm:lower}
+Suppose $\boldsymbol{q}$ satisfies $m(\boldsymbol{q}) = m$.
+Then for all strategies there exists a reward function such that
+\eq{
+\simpleregret \geq \frac{\frac{m}{2e} - 1}{m} \sqrt{\frac{2m}{3T}} = \Omega\left(\sqrt{\frac{m}{T}}\right)\,.
+}
+\end{theorem}
 
-\begin{lemma}\label{lem:conc1}
-Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
-\eq{
-\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
-}
-\end{lemma}
-
-\iffalse
-\begin{proof}
-Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
-Then
-\eq{
-\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} Z_t\,.
-}
-Now $Z_1,\ldots,Z_{T/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
-\end{proof}
+\ifsup
+The proof of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower}.
+\else
+The proof of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in the supplementary material.
 \fi
-
-\begin{lemma}\label{lem:m_est}
-Let $\delta >0$ and assume $T \geq 48m \log\frac{4N}{\delta}$. Then
-\eq{
-\P{\hat m < \frac{2}{3}m} \leq \delta\ \text{ and } \P{\hat m > 2m} \leq \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-Let $\boldsymbol{q}^b$ and $\boldsymbol{q}^{ub}$ and be the maximally balanced and unbalanced $\boldsymbol{q}$ for a given $m$ respectively.
-\eq{
-q^{ub}_i = \begin{cases}
-0 & \text{if } i \leq m \\
-\frac{1}{m} & \text{otherwise}\,.
-\end{cases} \\
-q^b_i < \begin{cases}
-\frac{1}{m} & \text{if } i \leq m \\
-1 & \text{otherwise}\,.
-\end{cases}  
-}
-For $\hat{m}$ to over-estimate $m$, we must identify some balanced arms as unbalanced. For $\hat{m}$ to under-estimate $m$, we must identify some unbalanced arms as balanced.
-\eq{
-\P{\hat m > 2m} \leq \P{\hat m > 2m|\boldsymbol{q} = \boldsymbol{q}^{ub}}\\
-\P{\hat m < \frac{2}{3}m} \leq \P{\hat m < \frac{2}{3}m|\boldsymbol{q} = \boldsymbol{q}^{b}}
-}
-
-Given $\boldsymbol{q} = \boldsymbol{q}^{ub}$, we have by Lemma \ref{lem:conc1}, with probability at least $1 - \delta$ that:
-
-\eq{
-& \left| \hat q_i - q_i\right| 
- \leq \begin{cases}
-0 & \text{if } i \leq m\\
-\sqrt{\frac{6}{mh} \log\frac{2}{\delta}} & \text{otherwise} \\
-\end{cases}\\
-\implies &
-\begin{cases}
-(\forall i \leq m) \qquad \left| \hat q_i - 0\right| = 0 \\
-(\forall i > m) \qquad \left| \hat q_i - \frac{1}{m}\right| \leq \frac{1}{2m} \\ %\; \text{, taking the union bound and assuming } h \geq 24m \log\frac{4N}{\delta}\\
-\end{cases}\\
-\implies &
-\begin{cases}
-(\forall i \leq m) \qquad  \hat q_i  = 0 \\
-(\forall i > m) \qquad \hat q_i \in [\frac{1}{2m}, \frac{3}{2m}]\\
-\end{cases}\\
-\implies & \hat{m} \leq 2m
-}
-
-Given $\boldsymbol{q} = \boldsymbol{q}^{b}$, we have by Lemma \ref{lem:conc1}, with probability at least $1 - \delta$ that:
-
-\eq{
-& \left| \hat q_i - q_i\right| 
- \leq \sqrt{\frac{6}{mh} \log\frac{2}{\delta}} \qquad \text{if } i \leq m \\
-\implies &
-(\forall i \leq m) \qquad  \hat q_i \leq \frac{3}{2m} \\
-\implies & \hat{m} \geq \frac{2m}{3}
-}
-
-\end{proof}
-
-\begin{proof}[Proof of \cref{thm:simple-regret}]
-
-for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
-\eq {
-\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
-}
-
-Via Hoeffding's Inequality
-
-\eq {
-\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
-}
-
-for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.
-
-\eq {
-(i,j)\notin A & \implies \hat{s}_i \geq \frac{1}{\hat m} \\
-& \implies \sum_{t=1}^h \ind{X_i = j} \geq \frac{h}{\hat{m}}
-}
-
-Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
-
-\eq {
-\hat \mu_{i,j} = \frac{1}{Z_{ij}}\sum_{t'=1}^{Z_{ij}}r_{t'}
-}
-
-\eq {
-\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
-= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
-\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
-\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
-}
-
-Applying the union bound over all $2N$ actions.
-\eq{
-& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
-\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
-}
-
-Now by \cref{lem:m_est}, 
-\eq{
-h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
-}
-
-Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that
-
-\eqn{
-\label{eqn:joint_bound_estimates_m}
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
-}
-Suppose $h < 24m \log \frac{4N}{\delta}$. Then
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
-}
-Therefore
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
-}
-Therefore 
-\eq{
-R_s(h) 
-&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
-&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
-}
-as required.
-\end{proof}

### 2016-01-31 causal-bandit/icml-2015/tech.tex
much shuffling

@@
 
-\subsection*{Truncated Importance Sampling}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PROOF OF SIMPLE REGRET UPPER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}
+
+The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound.
+
+\begin{lemma}\label{lem:conc1}
+Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
+\eq{
+\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
+}
+\end{lemma}
+
+\begin{lemma}\label{lem:m_est}
+Let $\delta >0$ and assume $T \geq 48m \log\frac{4N}{\delta}$. Then
+\eq{
+\P{\hat m < \frac{2}{3}m} \leq \delta\ \text{ and } \P{\hat m > 2m} \leq \delta\,.
+}
+\end{lemma}
+
+\begin{proof}
+Let $\boldsymbol{q}^b$ and $\boldsymbol{q}^{ub}$ and be the maximally balanced and unbalanced $\boldsymbol{q}$ for a given $m$ respectively.
+\eq{
+q^{ub}_i = \begin{cases}
+0 & \text{if } i \leq m \\
+\frac{1}{m} & \text{otherwise}\,.
+\end{cases} \\
+q^b_i < \begin{cases}
+\frac{1}{m} & \text{if } i \leq m \\
+1 & \text{otherwise}\,.
+\end{cases}  
+}
+For $\hat{m}$ to over-estimate $m$, we must identify some balanced arms as unbalanced. For $\hat{m}$ to under-estimate $m$, we must identify some unbalanced arms as balanced.
+\eq{
+\P{\hat m > 2m} \leq \P{\hat m > 2m|\boldsymbol{q} = \boldsymbol{q}^{ub}}\\
+\P{\hat m < \frac{2}{3}m} \leq \P{\hat m < \frac{2}{3}m|\boldsymbol{q} = \boldsymbol{q}^{b}}
+}
+
+Given $\boldsymbol{q} = \boldsymbol{q}^{ub}$, we have by Lemma \ref{lem:conc1}, with probability at least $1 - \delta$ that:
+
+\eq{
+& \left| \hat q_i - q_i\right| 
+ \leq \begin{cases}
+0 & \text{if } i \leq m\\
+\sqrt{\frac{6}{mh} \log\frac{2}{\delta}} & \text{otherwise} \\
+\end{cases}\\
+\implies &
+\begin{cases}
+(\forall i \leq m) \qquad \left| \hat q_i - 0\right| = 0 \\
+(\forall i > m) \qquad \left| \hat q_i - \frac{1}{m}\right| \leq \frac{1}{2m} \\ %\; \text{, taking the union bound and assuming } h \geq 24m \log\frac{4N}{\delta}\\
+\end{cases}\\
+\implies &
+\begin{cases}
+(\forall i \leq m) \qquad  \hat q_i  = 0 \\
+(\forall i > m) \qquad \hat q_i \in [\frac{1}{2m}, \frac{3}{2m}]\\
+\end{cases}\\
+\implies & \hat{m} \leq 2m
+}
+
+Given $\boldsymbol{q} = \boldsymbol{q}^{b}$, we have by Lemma \ref{lem:conc1}, with probability at least $1 - \delta$ that:
+
+\eq{
+& \left| \hat q_i - q_i\right| 
+ \leq \sqrt{\frac{6}{mh} \log\frac{2}{\delta}} \qquad \text{if } i \leq m \\
+\implies &
+(\forall i \leq m) \qquad  \hat q_i \leq \frac{3}{2m} \\
+\implies & \hat{m} \geq \frac{2m}{3}
+}
+
+\end{proof}
+
+\begin{proof}[Proof of Theorem \ref{thm:simple-regret}]
+
+for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
+\eq {
+\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
+}
+
+Via Hoeffding's Inequality
+
+\eq {
+\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
+}
+
+for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.
+
+\eq {
+(i,j)\notin A & \implies \hat{s}_i \geq \frac{1}{\hat m} \\
+& \implies \sum_{t=1}^h \ind{X_i = j} \geq \frac{h}{\hat{m}}
+}
+
+Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
+
+\eq {
+\hat \mu_{i,j} = \frac{1}{Z_{ij}}\sum_{t'=1}^{Z_{ij}}r_{t'}
+}
+
+\eq {
+\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
+= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
+\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
+\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
+}
+
+Applying the union bound over all $2N$ actions.
+\eq{
+& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
+\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
+}
+
+Now by Lemma \ref{lem:m_est}, 
+\eq{
+h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
+}
+
+Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that
+
+\eqn{
+\label{eqn:joint_bound_estimates_m}
+(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
+}
+Suppose $h < 24m \log \frac{4N}{\delta}$. Then
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
+}
+Therefore
+\eq{
+\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
+}
+Therefore 
+\eq{
+R_s(h) 
+&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
+&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
+}
+as required.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LOWER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}
+
+\todot{Change this to Bernoulli}
+Assume without loss of generality that $q_i \leq 1/2$ for all $i$ and that $q_1 \leq q_2 \leq \ldots \leq q_N$.
+For each $i$ define reward function $r_i$ by
+\eq{
+r_0(\boldsymbol{X}) &= 0 &
+r_i(\boldsymbol{X}) &= \begin{cases}
+\epsilon & \text{if } X_i = 1 \\
+0 & \text{otherwise}\,,
+\end{cases}
+}
+where $\epsilon > 0$ is some constant to be chosen later.
+We abbreviate $R_{T,i}^{\text{simple}}$ to be the expected simple regret incurred when interacting with the
+environment determined by $\boldsymbol{q}$ and $r_i$. Let $P_i$ be the corresponding measure
+on all observations over all $T$ rounds and $\mathbb E_i$ the expectation with respect to $P_i$. By Lemma 2.6 by \cite{Tsy08} we have
+\eq{
+\Prz{I_T = i} + \Pri{I_T \neq i} \geq \exp\left(-\KL(P_0, P_i)\right)\,,
+}
+where $\KL(P_0, P_i)$ is the KL divergence between measures $P_0$ and $P_i$.
+Let $T_i(T) = \sum_{t=1}^T \ind{I_t = i}$ be the total number of times the learner intervenes on variable $i$ (with any intervention). 
+Then for $i \leq m$ a computation shows that 
+\todot{Need to add at minimum some citations here}
+\eq{
+\KL(P_0, P_i) 
+&= \frac{\epsilon^2}{2} \mathbb E_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
+&\leq \frac{\epsilon^2}{2} \left(\mathbb E_0 T_i(T) + q_i T\right) \\
+&\leq \frac{\epsilon^2}{2} \left(\mathbb E_0 T_i(T) + \frac{T}{m}\right)\,.
+}
+Define set $A$ by
+\eq{
+A = \set{i \leq m : \mathbb E_0 T_i(T) \leq 2T / m}\,.
+}
+Then for $i \in A$ and choosing $\epsilon = \sqrt{2m/3T}$ we have
+\eq{
+\KL(P_0, P_i) \leq \frac{3T\epsilon^2}{2m} = 1\,. 
+}
+Now $\sum_{i=1}^m \mathbb E_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
+Therefore
+\eq{
+\sum_{i \in A} \Pri{I_T \neq i} 
+&\geq \sum_{i \in A} \exp\left(-\KL(P_0, P_i)\right) - 1\\
+&\geq \frac{|A|}{e} - 1 
+\geq \frac{m}{2e} - 1\,.
+}
+Therefore there exists an $i \in A$ such that
+\eq{
+\Pri{I_T \neq i} \geq \frac{\frac{m}{2e} - 1}{m}\,.
+}
+Therefore
+\eq{
+R^{\text{simple}}_{T,i} \geq \frac{1}{2} P_i(I_T \neq i) \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{2m}{3T}}
+}
+as required.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TRUNCATED IMPORTANCE SAMPLING
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Truncated Importance Sampling}
 
@@
 \begin{theorem}
-Let $\delta = 1/(2NT)$ and $\epsilon$ be the smallest value for which \cref{alg:alloc} returns true. 
-Let $N_{ij}$ as computed by \cref{alg:alloc} be the number of times the algorithm selects the action $do(X_i = j)$ 
+Let $\delta = 1/(2NT)$ and $\epsilon$ be the smallest value for which Algorithm \ref{alg:alloc} returns true. 
+Let $N_{ij}$ as computed by Algorithm \ref{alg:alloc} be the number of times the algorithm selects the action $do(X_i = j)$ 
 and computes estimates using the mixture estimator described in the previous section. Choosing $I_T = \argmax_{ij} \hat \mu_{ij}$ leads

### 2016-01-31 causal-bandit/icml-2015/general.tex
starting to clean up general section

@@
 
-We now consider single variable interventions on general causal graphs over variables $\boldsymbol{V}$.
-Let $Y \in \set{0,1}$ be the target variable and $\boldsymbol{X} = V/\set{Y \cup Desendants(Y)}$.
-
-If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ via the truncated product formula \cite{Pearl2000} 3.11
-
-\eq {
-\P{\boldsymbol{x},y|do(X_i = j)} = 
-\begin{cases}
-\frac{\P{\boldsymbol{x},y}}{\P{x_i=j|\calP_i(\boldsymbol{x})}} & \text{if} x_i = j  \\
-0 & \text{otherwise}
-\end{cases}
+We now consider single variable interventions on general causal graphs over binary variables $Y, X_1,\ldots,X_N$. 
+As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
+We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of $i$ so the joint distribution over $X$ may be written as
+\eq{
+\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
 }
-
-Where $\calP_i : \set{0,1}^N \to \set{0,1}^{N_i}$ is the
-projection that selections the $N_i$ parents of variable $i$.
+Throughout this section we are going to assume that the causal structure and conditional distributions of $X$ are known,
+which allows us to compute the above quantity for arbitrary $x$. As in the previous sections we allow interventions on
+a single variable by setting $X_i = j$ for any $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$. Then for $x \in \set{0,1}^N$ 
+with $x_i = j$ we have
+\eq{
+&\P{X = x|do(X_i = j)} \\
+&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
+}
+As before the expected reward from intervening by setting variable $X_i = j$ is
+\eq{
+\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,,
+}
+where $\Omega = \set{0,1}^N$. The problem of learning the optimal intervention is now significantly complicated
+by the dependencies between the random variables. Learning from observational data is no longer straightforward
+because variables can be correlated through their parents. Consider, for example, the causual model below where $X_1$
+deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
+then intervening on either $X_2$ or $X_3$ is optimal, but this positive is never observed from observational samples only. This is despite
+the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
+\begin{figure}[h]
+\centering
+\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+\node[main node](1){$X_1$};
+\node[main node, below left=of 1](2){$X_2$};
+\node[main node, below right=of 1](3){$X_3$};
+\node[main node, below right=of 2](4){Y};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2)
+    (1) edge (3)
+    (3) edge (4)
+    (2) edge (4);
+\end{tikzpicture}
+\eq{
+\P{X_1 = 1} &= \frac{1}{2} \\
+\P{X_2 = j|X_1 = j} &= 
+\P{X_3 = j|X_1 = j} = 1\,.
+}
+\end{figure}
 
@@
 Let $P(z=1)=.5$, $P(X_i = 1|z = 1) = 1$, and $P(X_i = 0|z = 0) = 1$ then $P(X_i = 1) = .5$ but we will be unable to estimate $P(Y|do(X_i = 1))$ as we will never get any data for the term $P(Y|X_i = 1, Z = 0)$.
-
-\begin{figure}[h]
-\centering
-\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
-
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
-\node[main node, above=of 3](6){$Z$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5)
-    (6) edge (1) edge (2) edge(4);
-
-\end{tikzpicture}
-\end{figure}
 

### 2016-01-31 causal-bandit/icml-2015/setup.tex
starting to clean up general section

@@
 Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$.
-Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t$.
+Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t \in \set{0,1}$.
 
@@
 \todom{Mention pruning of actions?}
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y = r(X) + \eta$ for some arbitrary but unknown, real-valued function $r$ and a noise term $\eta$ that may depend on $X$.
+\todot{I don't understand this at all. Later we are assuming $X_1,\ldots,X_N$ are binary and there are $N$ of them}
+We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
 The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.

### 2016-01-31 causal-bandit/icml-2015/general.tex
updates

@@
 deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
-then intervening on either $X_2$ or $X_3$ is optimal, but this positive is never observed from observational samples only. This is despite
+then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples only, despite
 the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
@@
 
-Since estimating $\P{Y|do(X_i=j)}$ from observational data requires conditioning on the parents of $X_i$, the variance of our estimators depends on $\P{X_i=j|\calP_i}$, rather than simply being a function of the number of times we observe $X_i=j$ as in the parallel bandit case.
-For example consider the graph in Figure \ref{fig:causalStructure_confounded}.
-In this model
-$P(Y|do(X_i = j) = \sum_z P(Y|X_i = j,Z)P(Z)$.
-Let $P(z=1)=.5$, $P(X_i = 1|z = 1) = 1$, and $P(X_i = 0|z = 0) = 1$ then $P(X_i = 1) = .5$ but we will be unable to estimate $P(Y|do(X_i = 1))$ as we will never get any data for the term $P(Y|X_i = 1, Z = 0)$.
-
-For general models, it is also insufficient to only trade off learning about multiple actions by observing against learning individual actions by explicitly playing them.
+Another difficulty is that it is no longer sufficient to learn about $\mu_{ij}$ exclusivly from observational data and when $do(X_i = j)$ is chosen.
 We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
@@
 
-Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent,$X_k = X_{k-1}$ for $k\in {N...2}$ and $\P{X_1} = 0$.
+Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
+for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
 We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. 
-If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to explicitly play each of these arms individually to estimate their rewards.
+If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to 
+explicitly play each of these arms individually to estimate their rewards.
 
@@
   thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
 \node[main node](1){$X_{1}$};
@@
 
-\todof{From here onwards needs to be replaced by truncated importance sampling version}
-
-Let $r(x) = \P{Y = 1|X = x}$. Then
-
-\eq {
-\mu_{ij} = P(Y = 1|do(X_i = j)) = \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}r(x)
+Before the algorithm we introduce the truncated importance weighted estimator, which we will use to learn about the returns of 
+specific interventions from observational data, or by intervening on other variables. Let $P$ and $Q$ be distributions on $\Omega$.
+We would like to estimate
+\eq{
+\mu_Q = \EE_Q r(x) = \sum_{x \in \Omega} Q(x) r(x) 
 }
-
-We wish to devise an estimator for $\mu_{ij}$ using $X$ and $Y$ without
-intervening to set $X_i = j$. A natural choice is the importance weighted estimator
-
-\eq{
-Z_{ij} 
-&= \frac{Y \ind{X_i = j}}{\P{X_i = j|\calP_i(X)}}\,. 
-}
-
-Checking:
-\eq{
-&\E{Z_{ij}} 
-= \sum_x \frac{\P{X = x} r(x) \P{X = x | do(X_i = j)}}{\P{X = x}} \\
-&= \P{Y = 1| do(X_i = j)} \,. 
-}
-The variance of $Z_{ij}$ may also be bounded
-\eq{
-&\Var[Z_{ij}]
-\leq \E{Z_{ij}^2} \\
-&\leq \sum_x \P{X = x} \left(\frac{\ind{X_i = j}}{\P{X_i = j|\calP_i(X) = \calP_i(x)}}\right)^2 \\
-&= \E{\frac{1}{\P{X_i = j|\calP_i(X)}}} \\
-&= V_{ij}\,.
-}
-In the general graph setting the dimension $m$ has a slightly different definition.
-\eq{
-m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}\,.
-}
-
-\todof{I think the above should be}
-
-\eq {
-m = \min \set{m: \sum_{ij} \ind{V_{ij} \geq m} = m}\,.
-}
-
-That is, $m$ should be chosen such that the size of the set variances greater than $m$ is $m$.
-This ensures all the variances of the remaining arms are smaller than $m$.
-This definition would also seem equivalent to sorting the variances such that $V^{(0)} \geq V^{(1)} ...
-\geq V^{(2N)}$ and then defining $m = \min m: V^{(m)} \leq m$ which is in turn equivalent to the definition in terms of $q$ for parallel bandits.
-
-\begin{algorithm}[H]
-\caption{General Algorithm}
-\begin{algorithmic}
-\STATE {\bf Input:} $N$, $\mathcal P$
-\STATE Compute $m$:
-\eq{
-m = \min\set{m : \sum_{ij} \ind{V_{ij} \leq 1/m} = m}
-}
-\STATE Collect $T/2$ observational samples and estimate $\hat \mu_{ij}$:
-\eq{
-\hat \mu_{ij} = \sum_{t=1}^{T/2} \frac{Y_t \ind{X_{t,i} = j}}{\P{X_{t,i} = j | \mathcal P_i(X_t)}}
-}
-\FOR{$i \in \set{1,\ldots,m}$}
-\STATE For each $i,j $ with $V_{ij} < 1/m$ take action $do(X_i = j)$ for $T/(2m)$ samples $Y^{(ij)}_1,\ldots,Y^{(ij)}_{T/2m}$
-\STATE Update estimate:
-\eq{
-\eta &= \frac{V_{ij}}{V_{ij} + m} \\
-\hat \mu_{ij} &= \eta \left(\frac{2m}{T} \sum_{t=1}^{T/(2m)} Y^{(ij)}_t\right) + (1 - \eta) \hat \mu_{ij}
-}
-\ENDFOR
-\STATE Choose $do(X_i = j)$ for $ij = \argmax_{ij} \hat \mu_{ij}$
-\end{algorithmic}
-\end{algorithm}
+using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$.

### 2016-01-31 causal-bandit/icml-2015/general.tex
updates

@@
 We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
-
-\eqn {
-\label{eq:estimation_transfer}
-P(Y_t|do(X_{t,i} = j))= & \nonumber \\
- \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
-}
-
 Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
@@
 }
-using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$.
+using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$. Ultimately we will be interested
+in the special case that $Q(x) = P(x|do(X_i = j))$ and where $P(x) = P(x|do())$ or $P(x|do(X_k = l))$, but for now we keep things general.
+Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
+\eqn{
+\label{eq:truncated}
+Z = Y R \ind{R < B}\,,
+}
+which trivially satisfies $|Z| \leq B$ surely.
+Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
+this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
+Computing the expectation we see
+\eq{
+\EE_P[Z] = \mu - \beta\,.
+}
+where $\beta \geq 0$ is the negative bias and satisfies
+\eq{
+\beta \leq \Q{R \geq B}\,.
+}
+Bounding the variance is also straightforward.
+\eq{
+\Var[Z] 
+&\leq \EE[Z^2] 
+=\sum_x \P{X = x} Z(x)^2 \\
+&\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 \\
+&\leq \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
+= V\,.
+}
+The bias can be controlled in terms of $V$ via Markov's inequality.
+\eq{
+\beta 
+\leq \Q{R \geq B} 
+\leq \frac{\EE_Q[R]}{B}
+= \frac{V}{B}\,.
+}
+The point is that $B$ can be tuned to trade the bias against the concentration of the estimator,
+the latter of which can be controlled via Bernstein's inequality.
+
+Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.
+
+\begin{algorithm}[H]
+\caption{CheckValid}\label{alg:alloc}
+\begin{algorithmic}
+\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_1,\ldots,\operatorname P_N$
+\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calP_i \subseteq \set{o_k : k < i}$ for all $i$
+\FOR{$i \in \set{o_1,\ldots,o_N}$ and $j \in \set{0,1}$}
+\STATE Compute variances:
+\eq{
+\forall k \in \calP_i,\, l \in \set{0,1} \quad
+V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
+}
+\STATE Compute variance for learning from observational data:
+\eq{
+V_{ij} = \frac{2}{T} \EE_{\operatorname P_{ij}} \!\!\left[\frac{\Pij{x}}{\P{x}}\right]
+}
+\STATE Compute required budget: 
+\eq{
+N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calP_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
+} 
+\ENDFOR
+\IF{$\sum_{i=1}^N \sum_{j \in \set{0,1}} N_{ij} \leq T / 2$}
+\STATE {\bf return true}
+\ELSE
+\STATE {\bf return false}
+\ENDIF
+\end{algorithmic}
+\end{algorithm}

### 2016-01-31 causal-bandit/icml-2015/general.tex
many updates

@@
 \newcommand{\calP}{\mathcal P}
+\newcommand{\calA}{\mathcal A}
 \newcommand{\x}{\boldsymbol{x}}
@@
 \centering
-\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
+\caption{}\label{fig:causalStructure_confounded}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 The point is that $B$ can be tuned to trade the bias against the concentration of the estimator,
-the latter of which can be controlled via Bernstein's inequality.
+the latter of which will be controlled via Bernstein's inequality.
 
 Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.
+As in the previous section, the algorithm for learning an optimal intervention on a general graph will 
+collect $T/2$ observational samples by choosing $do()$. It then chooses $do(X_i = j)$ 
+exactly $N_{ij}$ times, where $N_{ij}$ will be chosen shortly. For each $i$ it computes an estimate $\hat \mu_{ij}$ of $\mu_{ij}$ via
+a mixture of truncated importance weighted estimators.
+\eq{
+\hat \mu_{ij} 
+= 
+\frac{
+  \eta_{ij} \sum_{t=1}^{T/2} \frac{Z_{ij}(t)}{T/2} + \sum_{kl \in \calA_i} \eta_{ijkl} \sum_{t=1}^{N_{kl}} \frac{Z_{ijkl}(t)}{N_{kl}}}
+    {\eta_{ij} + \sum_{kl \in \calA_i} \eta_{ijkl}}\,,
+}
+where $\calA_i = \set{k, l : k \text{ is a ancestor of } i \text{ and } l \in \set{0,1}}$ and
+\eq{
+\eta_{ij} &= \frac{T/2}{V_{ij}} &
+V_{ij} &= \EE_{\operatorname{P}_{ij}} \left[\frac{\Pn{ij}{X = x}}{\P{X = x}}\right] \\
+\eta_{ijkl} &= \frac{N_{kl}}{V_{ijkl}} &
+V_{ijkl} &= \EE_{\operatorname{P}_{ij}} \left[\frac{\Pn{ij}{X = x}}{\Pn{kl}{X = x}}\right]
+}
+
+The following theorem controls the concentration of $\hat \mu_{ij}$ about $\mu_{ij}$. 
+\ifsup
+The proof may be found in Appendix \ref{sec:thm:estimate}.
+\else
+The proof follows from an application of Bernstein's inequality and may be found in the supplementary material.
+\fi
+
+\begin{theorem}\label{thm:estimate}
+The estimator $\hat \mu_{ij}$ satisfies:
+\eq{
+\P{\left|\hat \mu_{ij} - \mu\right| \geq \sqrt{\frac{1}{\sum_{k \in \calA_i, l \in \set{0,1}} \frac{N_{kl}}{V_{ijkl}}} \log\frac{1}{\delta}}} \leq \delta\,.
+}
+\end{theorem}
+
+All that remains is to choose $N_{ij}$ to minimise the worst-case approximation error in the bound above over all $i$ and $j$.
+This is acheived by finding the smallest $\epsilon > 0$ such that {\sc CheckValid} returns true (see Algorithm \ref{alg:alloc}).
 
@@
 \begin{algorithmic}
-\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_1,\ldots,\operatorname P_N$
-\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calP_i \subseteq \set{o_k : k < i}$ for all $i$
+\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_{ij}$ for all $ij$
+\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calA_i \subseteq \set{o_k : k < i}$ for all $i$
 \FOR{$i \in \set{o_1,\ldots,o_N}$ and $j \in \set{0,1}$}
@@
 \eq{
-\forall k \in \calP_i,\, l \in \set{0,1} \quad
+\forall k \in \calA_i,\, l \in \set{0,1} \quad
 V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
@@
 \eq{
-N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calP_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
+N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calA_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
 } 
 \ENDFOR
-\IF{$\sum_{i=1}^N \sum_{j \in \set{0,1}} N_{ij} \leq T / 2$}
-\STATE {\bf return true}
-\ELSE
-\STATE {\bf return false}
-\ENDIF
+\STATE {\bf if $\sum_{ij} N_{ij} \leq T/2$} {\bf return} true {\bf else return} false
 \end{algorithmic}
 \end{algorithm}
+
+\todot{discuss role of $delta = 1/(NT)$}
+\todot{define $Z_{ijkl}(t)$ etc.}
+\todot{Define $B_{ijkl}$}
+\begin{theorem}
+Let $\epsilon > 0$ be the smallest value of $\epsilon$ such that Algorithm \ref{alg:alloc} returns true and let $IJ = \argmax_{ij} \hat \mu_{ij}$.
+Then $\mu^* - \E \mu_{IJ} \leq \epsilon + 1/T$.
+\end{theorem}
+
+\subsection*{Comparison to Simple Setting}
+
+In Section \ref{sec:simple-regret} we assumed a simple graph structure and unknown distributions on $X_i$, which were then learned from observational
+data. In contrast, in this section we have allow arbitrary (known) graphs, but assumed the distribution on $X$ was known. 
+Using the approach in this section on the simple graph with known distribution leads to 
+\eq{
+\EE \mu^* - \EE \mu_{IJ} \in \bigo{\sqrt{\frac{m}{T} \log \frac{N}{\delta}}}\,, 
+}
+which is the same bound as in the previous section. 
+\todot{finish this}

### 2016-01-31 causal-bandit/icml-2015/tech.tex
updates
many updates

@@
-\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
-\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
-\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
-\newcommand{\EE}{\mathbb E}
-\newcommand{\Pn}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Truncated Importance Sampling}
+\section*{Truncated Importance Sampling}\label{sec:truncated}
 
@@
 Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
-\eqn{
-\label{eq:truncated}
+\eq{
 Z = Y R \ind{R < B}\,,
@@
 
-\subsection*{General Case}
+\section*{Proof of Theorem \ref{thm:estimate}}\label{sec:thm:estimate}
 
-Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.
+Apply Bernstein's inequality.
 
-\begin{algorithm}[H]
-\caption{CheckValid}\label{alg:alloc}
-\begin{algorithmic}
-\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_1,\ldots,\operatorname P_N$
-\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calP_i \subseteq \set{o_k : k < i}$ for all $i$
-\FOR{$i \in \set{o_1,\ldots,o_N}$}
-\STATE Compute variances:
-\eq{
-\forall k \in \calP_i \text{ and } l \in \set{0,1} \qquad
-V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
-}
-\STATE Compute variance for learning from do nothing action:
-\eq{
-V_{ij} = \frac{2}{T} \EE_{\operatorname P_{ij}} \!\!\left[\frac{\Pij{x}}{\P{x}}\right]
-}
-\STATE Compute required budget: 
-\eq{
-N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calP_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
-} 
-\ENDFOR
-\IF{$\sum_{i=1}^N \sum_{j \in \set{0,1}} N_{ij} \leq T / 2$}
-\STATE {\bf return true}
-\ELSE
-\STATE {\bf return false}
-\ENDIF
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}
-Let $\delta = 1/(2NT)$ and $\epsilon$ be the smallest value for which Algorithm \ref{alg:alloc} returns true. 
-Let $N_{ij}$ as computed by Algorithm \ref{alg:alloc} be the number of times the algorithm selects the action $do(X_i = j)$ 
-and computes estimates using the mixture estimator described in the previous section. Choosing $I_T = \argmax_{ij} \hat \mu_{ij}$ leads
-to
-\eq{
-\EE[r_T] \leq \epsilon + \frac{1}{T}\,.
-}
-\end{theorem}
+\todot{Write this up}

### 2016-01-31 causal-bandit/icml-2015/general.tex
updates

@@
 deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
-then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples only, despite
+then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
 the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
@@
 \frac{
-  \eta_{ij} \sum_{t=1}^{T/2} \frac{Z_{ij}(t)}{T/2} + \sum_{kl \in \calA_i} \eta_{ijkl} \sum_{t=1}^{N_{kl}} \frac{Z_{ijkl}(t)}{N_{kl}}}
-    {\eta_{ij} + \sum_{kl \in \calA_i} \eta_{ijkl}}\,,
+  \eta_{ij} \sum_{t=1}^{T/2} \frac{Z_{ij}(t)}{T/2} + \sum_{kl \in \calA_{ij}} \eta_{ijkl} \sum_{t=1}^{N_{kl}} \frac{Z_{ijkl}(t)}{N_{kl}}}
+    {\eta_{ij} + \sum_{kl \in \calA_{ij}} \eta_{ijkl}}\,,
 }
-where $\calA_i = \set{k, l : k \text{ is a ancestor of } i \text{ and } l \in \set{0,1}}$ and
+where 
+\eq{
+\calA_{ij} = \set{ij} \cup \set{kl : k \text{ is a ancestor of } i \text{ and } l \in \set{0,1}}
+}
+and
 \eq{
@@
 \eq{
-\P{\left|\hat \mu_{ij} - \mu\right| \geq \sqrt{\frac{1}{\sum_{k \in \calA_i, l \in \set{0,1}} \frac{N_{kl}}{V_{ijkl}}} \log\frac{1}{\delta}}} \leq \delta\,.
+\P{\left|\hat \mu_{ij} - \mu\right| \geq \sqrt{\frac{1}{\sum_{kl \in \calA_{ij}} \frac{N_{kl}}{V_{ijkl}}} \log\frac{1}{\delta}}} \leq \delta\,.
 }
@@
 \STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_{ij}$ for all $ij$
-\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calA_i \subseteq \set{o_k : k < i}$ for all $i$
+\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$
 \FOR{$i \in \set{o_1,\ldots,o_N}$ and $j \in \set{0,1}$}

### 2016-02-01 causal-bandit/icml-2015/general.tex
updates

@@
 }
-Throughout this section we are going to assume that the causal structure and conditional distributions of $X$ are known,
-which allows us to compute the above quantity for arbitrary $x$. As in the previous sections we allow interventions on
-a single variable by setting $X_i = j$ for any $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$. Then for $x \in \set{0,1}^N$ 
-with $x_i = j$ we have
+Throughout this section we assume that the causal structure and conditional distributions of $X$ are known,
+which allows us to compute the above quantity for arbitrary $x \in \Omega = \set{0,1}^N$. Given the learner intervens by $do(X_i = j)$
+leads to
 \eq{
-&\P{X = x|do(X_i = j)} \\
+&\Pn{ij}{X = x} = \P{X = x|do(X_i = j)} \\
 &\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
 }
-As before the expected reward from intervening by setting variable $X_i = j$ is
+The expected reward by making the intervention $do(X_i = j)$ is
 \eq{
-\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,,
+\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,.
 }
-where $\Omega = \set{0,1}^N$. The problem of learning the optimal intervention is now significantly complicated
-by the dependencies between the random variables. Learning from observational data is no longer straightforward
-because variables can be correlated through their parents. Consider, for example, the causual model below where $X_1$
+The problem of learning the optimal intervention is made more complicated than the simple case because of 
+the dependencies between the random variables. Learning from observational data is no longer straightforward
+because variables can be correlated through their parents. Consider, for example, the causual graph below where $X_1$
 deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
@@
 
+\todot{Topological sort}
 \begin{algorithm}[H]

### 2016-02-01 causal-bandit/icml-2015/general.tex
minor fixes

--- 2016-01-31 causal-bandit/icml-2015/general.tex
@@
 
-We now consider single variable interventions on general causal graphs over binary variables $Y, X_1,\ldots,X_N$. 
+We now consider the more general problem where the graph structure is known, but arbitrary.
 As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
-We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of $i$ so the joint distribution over $X$ may be written as
+We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of variable $X_i$ so the joint distribution over $X$ may be written as
 \eq{
@@
 }
-As before the expected reward from intervening by setting variable $X_i = j$ is
+As before the expected reward from intervening by $do(X_i = j)$.
 \eq{
@@
 \centering
-\caption{}\label{fig:causalStructure_confounded}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 }
+\caption{}\label{fig:causalStructure_confounded}
 \end{figure}
@@
 for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
-We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. 
-If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to 
-explicitly play each of these arms individually to estimate their rewards.
+We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
 
@@
 using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$. Ultimately we will be interested
-in the special case that $Q(x) = P(x|do(X_i = j))$ and where $P(x) = P(x|do())$ or $P(x|do(X_k = l))$, but for now we keep things general.
+in the case that 
+$Q(x) = P(x|do(X_i = j))$
+and where $P(x) = P(x|do())$ or $P(x|do(X_k = l))$, but for now we keep things general.
 Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
@@
 \eq{
-\calA_{ij} = \set{ij} \cup \set{kl : k \text{ is a ancestor of } i \text{ and } l \in \set{0,1}}
+\calA_{ij} = \set{ij} \cup \set{kl : k \text{ is an ancestor of } i \text{ and } l \in \set{0,1}}
 }

### 2016-02-01 causal-bandit/icml-2015/setup.tex
Start restructure

@@
 In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
-The \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
-An edge from variable $X$ to $X'$ is interpreted to mean that a change in the value of $X$ directly causes a change to the value of $X'$. 
+Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
+An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
+The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$ or just $\parents{i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
 An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-An intervention ``mutilates'' the original graph by removing edges into the variables in $\vec{X}$ and the resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$.
-Further details can be found in Chapter 21 of~\cite{Koller2009}.
+The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from the parents of $X_i$ to $X_i$ for each $X_i \in \vec{X}$. 
+The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
+Details can be found in Chapter 21 of~\cite{Koller2009}.
 
@@
 
-We will consider and analyse two different goals for the learner: 1) identifying the intervention with highest expected reward by round $T$, and 2) maximising the total expected reward over all $T$ rounds.
+The learning goal we will consider and analyse is identifying the intervention with highest expected reward by round $T$. 
+This objective is most appropriate when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
+\todom{Citation}
 We denote the expected reward for the intervention $do(\vec{X} = \vec{x})$ by $\mu_{\vec{x}} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{\vec{x}} \mu_{\vec{x}}$.
@@
 }
-Performance for the second goal is measured by \emph{cumulative regret}, the difference between the total reward obtained if the optimal intervention were known in advance and the expected reward of the learner:
-\eqn{
-\label{eq:regret}
-R_T = T \mu^* - \E{\sum_{t=1}^T Y_t}\,.
-}
-Simple regret typically makes more sense when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
-Cumulative regret is more appropriate for online agents that may continue to adapt their policy. 
-We analyse both cases in \S\ref{sec:simple-regret} and \S\ref{sec:cum-regret}, respectively.
 
-\todom{Mention pruning of actions?}
-\todot{I don't understand this at all. Later we are assuming $X_1,\ldots,X_N$ are binary and there are $N$ of them}
+Another common objective in bandit problems is maximising the \emph{cumulative regret}, the total expected reward over all $T$ rounds. 
+We have established upper bounds for cumulative regret for parallel bandits but not the general causal bandit case.
+Due to space limitations we focus on simple regret and defer these to Appendix~\ref{app:cumulatve}.
+
 We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.

### 2016-02-01 causal-bandit/icml-2015/introduction.tex

@@
 
-Our main results for this case are: an upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?); an upper bound on the cumulative regret (Theorem~\ref{thm:}) for a specific causal model that has a better dependence on the number of available actions then the best possible algorithm which ignores the causal model; and a matching lower bound (Theorem~\ref{thm:}) for any algorithm which does use the additional causal information.
+In \S\ref{sec:parallel} we analyse a specific family of causal bandit problems which we call \emph{parallel bandit} problems.
+We  upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?)
+\todom{Finish describing new structure}
+
+; an upper bound on the cumulative regret (Theorem~\ref{thm:}) for a specific causal model that has a better dependence on the number of available actions then the best possible algorithm which ignores the causal model; and a matching lower bound (Theorem~\ref{thm:}) for any algorithm which does use the additional causal information.
 Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.

### 2016-02-01 causal-bandit/icml-2015/setup.tex
Moved parallel to S3. Re-did Algorithm 1 using a instead of (i,j).

@@
 Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
+We will assume each variable only takes on a finite number of distinct values.
 An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
@@
 
-A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$, a set of \emph{observable} variables $\mathcal{O} \subseteq \mathcal{X}$, a set of \emph{allowed interventions} $\mathcal{A}$, a \emph{reward variable} $Y$.
+A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed interventions} $\mathcal{A}$.
+One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
 The following game is then played over $T$ rounds.
-In round $t$, the learner selects an allowed intervention $do(\vec{X}_t = \vec{x}_t)$ from $\mathcal{A}$.
-Values for all non-intervened variables $\vec{X}^c_t$ are sampled from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$.
-Finally, the learner is shown the sampled values of all the observable variables in $\vec{X}^c_t \cap \mathcal{O}$ and receives the reward $Y_t \in \set{0,1}$.
+In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, including the \emph{reward} $Y_t \in \{0,1\}$.
 
-Although we will focus on the intervene-observe-reward ordering of events within each round, others are also possible. 
+Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are also possible. 
 If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already well understood~\citep{Agarwal2014}. 
@@
 
-\subsection{The Parallel Bandit Problem}
-
-We now consider a class of a causal bandit problems which we call \emph{parallel bandits}.
-It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example.
-It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
-
-The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
-All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable actions are all size 0 and size 1 interventions: 
-\[
-	\mathcal{A} = \{do()\} \cup \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}.
-\]
-In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
-In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\caption{Causal model for the parallel bandits problem.\label{fig:causalStructure}}
-\end{figure}
-
-In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
-The remainder of the variables are simultaneously set by independently biased coin flips. 
-The value of all variables are then used to determine the distribution of rewards for that round.
-Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
-The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
-In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.
-
-In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
-We will also make use of the expansion of the expected reward for $do(X_i = j)$:
-\eq{
-\mu_{i,j} 
-&= \E{r(X)|do(X_i = j)} \\
-&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
-}
-The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
-The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
-Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
-
 \subsection{Exploiting the Causal Model}
@@
 
-This enables us to estimate the the rewards for multiple interventions simultaneously by observing the values taken by the variables when we select the $do()$ action.
+This enables us to estimate the rewards for multiple interventions simultaneously by observing the values taken by the variables when we select the $do()$ action.
 Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and need to be explored separately, by explicitly making the intervention.

### 2016-02-01 causal-bandit/icml-2015/simple-regret.tex
Removed line 8
Moved parallel to S3. Re-did Algorithm 1 using a instead of (i,j).

@@
-In this section we develop and analyse an algorithm for minimising the simple regret in the
-causal bandit problem.
+In this section we develop an algorithm for achieving the optimal simple regret in a special case of a causal bandit problem.
+
+\subsection{The Parallel Bandit Problem}
+
+We now consider a class of a causal bandit problems which we call \emph{parallel bandits}.
+It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example.
+It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
+
+The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
+All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable actions are all size 0 and size 1 interventions: 
+\[
+	\mathcal{A} = \{do()\} \cup \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}.
+\]
+In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
+In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){Y};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5);
+
+\end{tikzpicture}
+\caption{Causal model for the parallel bandits problem.\label{fig:causalStructure}}
+\end{figure}
+
+In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
+The remainder of the variables are simultaneously set by independently biased coin flips. 
+The value of all variables are then used to determine the distribution of rewards for that round.
+Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
+\todom{Should $Y$ be 0-1 valued or real everywhere?}
+The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
+In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.
+
+\todom{$i,j$ notation to $a \in \actions$}
+In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
+We will also make use of the expansion of the expected reward for $do(X_i = j)$:
+\eq{
+\mu_{i,j} 
+&= \E{r(X)|do(X_i = j)} \\
+&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
+}
+The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
+The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
+Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
+
+\subsection{The Parallel Bandit Algorithm}
+\label{sub:par-bandit-alg}
 
@@
 \begin{algorithmic}[1]
-\STATE {\bf Input:} $T, N$
+\STATE {\bf Input:} Total rounds $T$, causal graph $\mathcal{G}$, allowable interventions $\actions$. 
 \FOR{$t \in 1,\ldots,(T - 1) / 2$}
-\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
+\STATE Perform empty intervention $do()$; observe $\vec{x}_t$ and $r_t$
 \ENDFOR
-\STATE Estimate $\mu$ using observation data:
-\eq{
-(\forall i,j) \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{t,i} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{t,i} = j}}\,.
-}
-\STATE Compute $\hat V_{ij} = \frac{T}{2 \sum_{t=1}^{T/2}\ind{X_{t,i}=j}} $
-\STATE Compute $\hat m = m(\hat{\boldsymbol{V}})$
-\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\STATE Compute $A$ as the subset of infrequently observed arms $\set{(i,j):V_{ij} \geq m}$ with $|A| = \hat m$
-
-\FOR{$(i,j) \in A$}
-\FOR{$t \in 1,\ldots,\frac{T - 1)}{2\hat m}$}
-\STATE Choose action $do(X_{t,i} = j)$ and observe $r_t$
+\FOR{$a = do(X_i = x) \in \actions$}
+\STATE Count times $a$ was seen $T_a = \sum_{t=1}^{T/2} \ind{x_{t,i} = x}$
+\STATE Estimate reward $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{x_{t,i} = x} r_t$
+\STATE Estimate probabilities $\hat{p}_a = \frac{2 T_a}{T-1}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
 \ENDFOR
-\STATE Recompute $\hat \mu_{i,j} = \frac{2\hat m}{T} \sum_{t=1}^{ T/2\hat m} r_t(X_{t,i}=j)$ 
+\STATE Compute $\hat{m} = m(\hat{\vec{V}})$ and $A = \set{a \in \actions \colon \hat{V}_a \geq \hat{m}}$.
+\STATE Let $T_A := \frac{T-1}{2 |A|}$ be times to sample each $a\in\actions$.
+\FOR{$a = do(X_i = x) \in A$}
+\FOR{$t \in 1,\ldots,T_A$}
+\STATE Intervene with $a$ and observe $r_t$
 \ENDFOR
-\STATE Estimated optimal action is $\hat i^*, \hat j^* = \argmax_{i,j} \hat \mu_{i,j}$
-\STATE Choose action $do(X_{t,\hat i^*} = \hat j^*)$
+\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} r_t$
+\ENDFOR
+\RETURN estimated optimal $\hat{a}^* \in \argmax_{a\in\actions} \hat{\mu}_a$
 \end{algorithmic}

### 2016-02-02 causal-bandit/icml-2015/introduction.tex
Tweaks

@@
 
-In \S\ref{sec:parallel} we analyse a specific family of causal bandit problems which we call \emph{parallel bandit} problems.
-We  upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?)
+In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems which we call \emph{parallel bandit} problems.
+We propose a causal best arm identification algorithm (Algorithm~\ref{alg:simple}) for this problem and show it is optimal (Theorems~\ref{thm:uq-simple} \& \ref{thm:lower}).
+
+for a general, fully observable causal model (TODO?)
 \todom{Finish describing new structure}

### 2016-02-02 causal-bandit/icml-2015/setup.tex
Tweaks

@@
 The learning goal we will consider and analyse is identifying the intervention with highest expected reward by round $T$.
+This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009} or ``best-arm identification'' problem~\citep{Gabillon2012}.
 This objective is most appropriate when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
-\todom{Citation}
-We denote the expected reward for the intervention $do(\vec{X} = \vec{x})$ by $\mu_{\vec{x}} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{\vec{x}} \mu_{\vec{x}}$.
-Performance for the first goal can be measured by \emph{simple regret}, \ie, the gap between the optimal expected reward and the intervention chosen by the learner on round $T$:
-\eqn{
+We denote the expected reward for the intervention $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
+Performance for the first goal can be measured by \emph{simple regret}, \ie, the gap between the optimal expected reward and the intervention $a_T$ chosen by the learner on round $T$:
+\begin{equation}
 \label{eq:regret-simple}
-\simpleregret = \mu^* - \E{\mu_{\vec{x}_T}}\,.
-}
+	\simpleregret = \mu^* - \E{\mu_{a_T}}.
+\end{equation}
 

### 2016-02-02 causal-bandit/icml-2015/simple-regret.tex
Tweaks

@@
-In this section we develop an algorithm for achieving the optimal simple regret in a special case of a causal bandit problem.
+In this section we propose and analyse an algorithm for achieving the optimal simple regret in a special case of a causal bandit problem.
 
@@
 The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
-All variables are observable (\ie, $\mathcal{O} = \{Y, X_1, \ldots, X_N\}$) and the set of allowable actions are all size 0 and size 1 interventions: 
+All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: 
 \[
@@
 \FOR{$a = do(X_i = x) \in \actions$}
-\STATE Count times $a$ was seen $T_a = \sum_{t=1}^{T/2} \ind{x_{t,i} = x}$
-\STATE Estimate reward $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{x_{t,i} = x} r_t$
-\STATE Estimate probabilities $\hat{p}_a = \frac{2 T_a}{T-1}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
+\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{x_{t,i} = x}$
+\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{x_{t,i} = x} r_t$
+\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T-1}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
 \ENDFOR
 \STATE Compute $\hat{m} = m(\hat{\vec{V}})$ and $A = \set{a \in \actions \colon \hat{V}_a \geq \hat{m}}$.
-\STATE Let $T_A := \frac{T-1}{2 |A|}$ be times to sample each $a\in\actions$.
+\STATE Let $T_A := \frac{T-1}{2 |A|}$ be times to sample each $a\in A$.
 \FOR{$a = do(X_i = x) \in A$}

### 2016-02-02 causal-bandit/icml-2015/general.tex
Replace \leq with = in variance bound.
Updated contributions

@@
 &\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 \\
-&\leq \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
+&= \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
 = V\,.
@@
 \todot{Define $B_{ijkl}$}
-\begin{theorem}
+\begin{theorem}\label{thm:general-upper}
 Let $\epsilon > 0$ be the smallest value of $\epsilon$ such that Algorithm \ref{alg:alloc} returns true and let $IJ = \argmax_{ij} \hat \mu_{ij}$.

### 2016-02-02 causal-bandit/icml-2015/introduction.tex
Updated contributions

--- 2016-01-29 causal-bandit/icml-2015/introduction.tex
@@
 
-Our main results for this case are: an upper bound on the (simple) regret (Theorem~\ref{thm:}) for a general, fully observable causal model (TODO?); an upper bound on the cumulative regret (Theorem~\ref{thm:}) for a specific causal model that has a better dependence on the number of available actions then the best possible algorithm which ignores the causal model; and a matching lower bound (Theorem~\ref{thm:}) for any algorithm which does use the additional causal information.
-Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
-Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.
+In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently.
+We propose a simple causal best arm identification algorithm (Algorithm~\ref{alg:simple}) for this problem and show it is optimal up to log factors (Theorems~\ref{thm:uq-simple} \& \ref{thm:lower}) with $\tilde{\Theta}(\sqrt{m/T})$ simple regret after $T$ rounds where $m$ depends on the casual model and may be much smaller than $N$.
+In contrast, the best a bandit algorithm can do if the causal model is ignored is $\Omega(\sqrt{N/T})$.
+This shows theoretically the value of our framework over the traditional bandit problem. 
+Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in bandit problems.
 
-Although the witness (Algorithm~\ref{alg:}) for our upper bound is very simple (\ie, observe without intervening for some initial period and then play undersampled actions in order to get good estimates) our key point is that algorithms which use causal information when available have a clear, theoretical improvement over those which do not.
-These results are experimentally validated in \S\ref{sec}.
+Interventions and observations may have a complex relationship in the general causal model. 
+However, in \S\ref{sec:simple-regret-general}, Theorem~\ref{thm:general-upper} shows how an importance-sampling based algorithm (Algorithm~\ref{alg:alloc}) can obtain non-trivial guarantees on simple regret.
+
+% Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
+% Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.
+
+% Although the witness (Algorithm~\ref{alg:}) for our upper bound is very simple (\ie, observe without intervening for some initial period and then play undersampled actions in order to get good estimates) our key point is that algorithms which use causal information when available have a clear, theoretical improvement over those which do not.
+% These results are experimentally validated in \S\ref{sec}.
 

### 2016-02-02 causal-bandit/icml-2015/introduction.tex
Tweaks

@@
 Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in bandit problems.
+\todom{Mention cumulative regret?}
 
-Interventions and observations may have a complex relationship in the general causal model. 
-However, in \S\ref{sec:simple-regret-general}, Theorem~\ref{thm:general-upper} shows how an importance-sampling based algorithm (Algorithm~\ref{alg:alloc}) can obtain non-trivial guarantees on simple regret.
+In the general casual model interventions and observations may have a complex relationship. 
+However, in \S\ref{sec:simple-regret-general}, Theorem~\ref{thm:general-upper} shows how an importance-sampling based algorithm that exploits the causal graph (Algorithm~\ref{alg:alloc}) can obtain non-trivial guarantees on simple regret.
+While this upper bound reduces to the one for parallel bandits, a matching lower bound for the general case is left as an open problem.
 

### 2016-02-02 causal-bandit/icml-2015/introduction.tex
More tweaks

@@
 
-One obvious approach would be to cast the above example as a stochastic multi-armed bandit problem (\eg, the six possible farming interventions is an arm and the crop yield is the reward) and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
-However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields a worst-case regret after $T$ round of $\bigomega{\sqrt{NT}}$ with a sub-optimal dependence on $N$, the number of possible interventions.
+As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
+In the farming example, the farmer could treat the six possible interventions are arms, ignore the observations of the non-intervened variables, and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
+However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.
 
-Our framework also bears a superficial similarity to contextual bandit problems \cite{Langford2008,Agarwal2014} since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
-However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intevention and cannot be used as context.
+Bandit problems with side information are commonly referred to as ``contextual bandits'' and have been well studied in recent years~\cite{Langford2008,Agarwal2014}.
+Our framework bears a superficial similarity to contextual bandit problems since the extra observations on non-intervened variables might be viewed as context for selecting an intervention. 
+However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and cannot be used as context.
+We discuss the relationship between causal and contextual bandits further in \S\ref{sec:defs} and leave their combination as future work.
 
-There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \cite{TODO} as well as results for partial monitoring which, at first glance, may appear applicable.
-Indeed, we show in Section~\ref{}, that it is possible to applying these existing results to our problem, however the resulting regret is $\bigo{\sqrt{NT}}$, again with a sub-optimal dependence on $N$.
+There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \citep{Alon2015} which, at first glance, may appear applicable.
+Indeed, we show in \S~\ref{sec:discussion}, that it is possible to applying these existing results to our problem, however the resulting regret is $\bigo{\sqrt{NT}}$, again with a sub-optimal dependence on $N$.
 
@@
 
-We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
-
-\todom{Summarise results \& give intuition for algorithm}
+% We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
 

### 2016-02-02 causal-bandit/icml-2015/general.tex
updates

@@
 \newcommand{\x}{\boldsymbol{x}}
+\newcommand{\Ps}{\operatorname{P}}
 
 We now consider the more general problem where the graph structure is known, but arbitrary.
+We use the notation $\P{\parents{Y}(X)}$ to denote a random variable that is the probability that the parents of $Y$ are $X$.
+Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
+Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ be the mixture distribution.
+Let $a \in \calA$ be an intervention and define random variable
+\eq{
+Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
+}
+where $B_a \geq 0$ is some constant to be chosen subsequently and
+\eq{
+R_a(X) = \frac{\P{\parents{Y}(X)|a}}{\Q{\parents{Y}(X)}}\,.
+}
+\begin{algorithm}[H]
+\begin{algorithmic}
+\STATE {\bf Input:} $\eta$, $B$, $T$
+\FOR{$t \in \set{1,\ldots,T}$}
+\STATE Sample action $I_t = a$ with probability $\eta$
+\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
+\ENDFOR
+\STATE For each $a \in \calA$ compute an estimate of its return:
+\eq{
+\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
+}
+\STATE Choose $I = \argmax_a \hat \mu_a$
+\end{algorithmic}
+\end{algorithm}
+
+\begin{theorem}
+Blah
+\end{theorem}
+
 As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.

### 2016-02-02 causal-bandit/icml-2015/general.tex
updaets

@@
 We now consider the more general problem where the graph structure is known, but arbitrary.
-We use the notation $\P{\parents{Y}(X)}$ to denote a random variable that is the probability that the parents of $Y$ are $X$.
-Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
-Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ be the mixture distribution.
-Let $a \in \calA$ be an intervention and define random variable
-\eq{
-Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
-}
-where $B_a \geq 0$ is some constant to be chosen subsequently and
-\eq{
-R_a(X) = \frac{\P{\parents{Y}(X)|a}}{\Q{\parents{Y}(X)}}\,.
-}
-\begin{algorithm}[H]
-\begin{algorithmic}
-\STATE {\bf Input:} $\eta$, $B$, $T$
-\FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $I_t = a$ with probability $\eta$
-\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
-\ENDFOR
-\STATE For each $a \in \calA$ compute an estimate of its return:
-\eq{
-\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
-}
-\STATE Choose $I = \argmax_a \hat \mu_a$
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}
-Blah
-\end{theorem}
-
-As before $Y$ is the target variable and $\P{Y=1|X} = r(X)$ is the expected reward given $X$.
-We let $\calP_i \subseteq \set{1,\ldots,N}$ denote the parents of variable $X_i$ so the joint distribution over $X$ may be written as
-\eq{
-\P{X = x} = \prod_{i=1}^N \P{X_i = x_i|X_j = x_j \text{ for all } j \in \calP_i}\,.
-}
-Throughout this section we assume that the causal structure and conditional distributions of $X$ are known,
-which allows us to compute the above quantity for arbitrary $x \in \Omega = \set{0,1}^N$. Given the learner intervens by $do(X_i = j)$
-leads to
-\eq{
-&\Pn{ij}{X = x} = \P{X = x|do(X_i = j)} \\
-&\qquad =\prod_{k \neq i} \P{X_k = x_k|X_j = x_j \text{ for all } j \in \calP_k}\,.
-}
-The expected reward by making the intervention $do(X_i = j)$ is
-\eq{
-\mu_{ij} = \sum_{x \in \Omega} \P{X = x|do(X_i = j)} r(x)\,.
-}
 The problem of learning the optimal intervention is made more complicated than the simple case because of 
@@
 
-Before the algorithm we introduce the truncated importance weighted estimator, which we will use to learn about the returns of 
-specific interventions from observational data, or by intervening on other variables. Let $P$ and $Q$ be distributions on $\Omega$.
-We would like to estimate
+We use the notation $\P{\parents{Y}(X)}$ to denote a random variable that is the probability that the parents of $Y$ are $X$.
+Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
+Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ be the mixture distribution.
+Let $a \in \calA$ be an intervention and define random variable
 \eq{
-\mu_Q = \EE_Q r(x) = \sum_{x \in \Omega} Q(x) r(x) 
+Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
 }
-using samples $X \in \Omega$ and $Y \in \set{0,1}$ where $X \sim P$ and $Y = 1$ with probability $r(X)$. Ultimately we will be interested
-in the case that 
-$Q(x) = P(x|do(X_i = j))$
-and where $P(x) = P(x|do())$ or $P(x|do(X_k = l))$, but for now we keep things general.
-Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
-\eqn{
-\label{eq:truncated}
-Z = Y R \ind{R < B}\,,
+where $B_a \geq 0$ is some constant to be chosen subsequently and
+\eq{
+R_a(X) = \frac{\P{\parents{Y}(X)|a}}{\Q{\parents{Y}(X)}}\,.
 }
-which trivially satisfies $|Z| \leq B$ surely.
-Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
-this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
-Computing the expectation we see
+Define $m(\eta)$ by
 \eq{
-\EE_P[Z] = \mu - \beta\,.
+m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
 }
-where $\beta \geq 0$ is the negative bias and satisfies
+
+\begin{algorithm}[H]
+\caption{General Algorithm}\label{alg:general}
+\begin{algorithmic}
+\STATE {\bf Input:} $\eta$, $T$
+\FOR{$t \in \set{1,\ldots,T}$}
+\STATE Sample action $I_t = a$ with probability $\eta$
+\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
+\ENDFOR
+\STATE For each $a \in \calA$ compute an estimate of its return:
 \eq{
-\beta \leq \Q{R \geq B}\,.
+\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
 }
-Bounding the variance is also straightforward.
+\STATE Choose $I = \argmax_a \hat \mu_a$
+\end{algorithmic}
+\end{algorithm}
+
+\begin{theorem}\label{thm:general}
+If Algorithm \ref{alg:general} is run with $B \in \R^{|\calA|}$ given by
 \eq{
-\Var[Z] 
-&\leq \EE[Z^2] 
-=\sum_x \P{X = x} Z(x)^2 \\
-&\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 \\
-&= \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
-= V\,.
+B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.
 }
-The bias can be controlled in terms of $V$ via Markov's inequality.
+Then for $C = \sqrt{2} + 4$ we have
 \eq{
-\beta 
-\leq \Q{R \geq B} 
-\leq \frac{\EE_Q[R]}{B}
-= \frac{V}{B}\,.
+\mu^* - \EE[\mu_I] \leq C\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}\,.
 }
-The point is that $B$ can be tuned to trade the bias against the concentration of the estimator,
-the latter of which will be controlled via Bernstein's inequality.
+\end{theorem}
 
-Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.
-As in the previous section, the algorithm for learning an optimal intervention on a general graph will 
-collect $T/2$ observational samples by choosing $do()$. It then chooses $do(X_i = j)$ 
-exactly $N_{ij}$ times, where $N_{ij}$ will be chosen shortly. For each $i$ it computes an estimate $\hat \mu_{ij}$ of $\mu_{ij}$ via
-a mixture of truncated importance weighted estimators.
+\begin{proof}
+First note that $|Z_a(X_t)| \leq B_a$ and that
 \eq{
-\hat \mu_{ij} 
-= 
-\frac{
-  \eta_{ij} \sum_{t=1}^{T/2} \frac{Z_{ij}(t)}{T/2} + \sum_{kl \in \calA_{ij}} \eta_{ijkl} \sum_{t=1}^{N_{kl}} \frac{Z_{ijkl}(t)}{N_{kl}}}
-    {\eta_{ij} + \sum_{kl \in \calA_{ij}} \eta_{ijkl}}\,,
+\Var_Q[Z_a(X_t)] 
+&\leq \EE_Q[Z_a(X_t)^2] \\
+&\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] \\
+&\leq m(\eta)\,.
 }
-where 
+Checking the expectation we have
 \eq{
-\calA_{ij} = \set{ij} \cup \set{kl : k \text{ is an ancestor of } i \text{ and } l \in \set{0,1}}
+\EE_Q[Z] 
+&= \EE_{\Ps_a}\left[Y \ind{R_a \leq B_a}\right] \\
+&= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] \\
+&= \mu_a - \beta_a\,,
+}
+where $0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a > B_a}$ is the negative bias. 
+The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
+\eq{
+\beta_a \leq \Pn{a}{R_a > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
+}
+Let $\epsilon_a > 0$ be given by
+\eq{
+\epsilon = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
+}
+Then by the union bound and Bernstein's inequality we have
+\eq{
+&\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \\ 
+&\qquad\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
+}
+Assuming this event does not occur and letting $a^* = \argmax_{a \in \calA} \mu_a$ we have
+\eq{
+\mu_I \geq \hat \mu_I - \epsilon_I  
+\geq \hat \mu_{a^*} - \epsilon_I  
+\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
+}
+By the definition of the truncation
+we have
+\eq{
+\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
 }
@@
 \eq{
-\eta_{ij} &= \frac{T/2}{V_{ij}} &
-V_{ij} &= \EE_{\operatorname{P}_{ij}} \left[\frac{\Pn{ij}{X = x}}{\P{X = x}}\right] \\
-\eta_{ijkl} &= \frac{N_{kl}}{V_{ijkl}} &
-V_{ijkl} &= \EE_{\operatorname{P}_{ij}} \left[\frac{\Pn{ij}{X = x}}{\Pn{kl}{X = x}}\right]
+\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
 }
+Therefore for $C = \sqrt{2} + 4$ we have
+\eq{
+\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
+}
+Therefore
+\eq{
+\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
+}
+as required.
+\end{proof}
 
-The following theorem controls the concentration of $\hat \mu_{ij}$ about $\mu_{ij}$. 
-\ifsup
-The proof may be found in Appendix \ref{sec:thm:estimate}.
-\else
-The proof follows from an application of Bernstein's inequality and may be found in the supplementary material.
-\fi
+\subsection*{Choosing the Sampling Distribution}
 
-\begin{theorem}\label{thm:estimate}
-The estimator $\hat \mu_{ij}$ satisfies:
+Algorithm \ref{alg:general} depends on a choice of sampling distribution $\eta$. In light of Theorem \ref{thm:general}
+a natural choice of $\eta$ is the minimiser of $m(\eta)$.
 \eq{
-\P{\left|\hat \mu_{ij} - \mu\right| \geq \sqrt{\frac{1}{\sum_{kl \in \calA_{ij}} \frac{N_{kl}}{V_{ijkl}}} \log\frac{1}{\delta}}} \leq \delta\,.
+\eta^* 
+&= \argmin_\eta m(\eta) \\
+&= \argmin_\eta \max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
+&= \argmin_\eta \underbrace{\max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
 }
-\end{theorem}
+Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
+Therefore the minimisation problem may be tackled using standard techniques from convex optimisation. In the experimental section we present some special cases,
+but for now we give two simple results. The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
 
-All that remains is to choose $N_{ij}$ to minimise the worst-case approximation error in the bound above over all $i$ and $j$.
-This is acheived by finding the smallest $\epsilon > 0$ such that {\sc CheckValid} returns true (see Algorithm \ref{alg:alloc}).
+\begin{lemma}
+$m(\eta^*) \leq |\calA|$.
+\end{lemma}
 
-\todot{Topological sort}
-\begin{algorithm}[H]
-\caption{CheckValid}\label{alg:alloc}
-\begin{algorithmic}
-\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_{ij}$ for all $ij$
-\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$
-\FOR{$i \in \set{o_1,\ldots,o_N}$ and $j \in \set{0,1}$}
-\STATE Compute variances:
+\begin{proof}
+By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|$. Then
 \eq{
-\forall k \in \calA_i,\, l \in \set{0,1} \quad
-V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
+m(\eta) 
+&= \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
+&\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] \\
+&\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA|
 }
-\STATE Compute variance for learning from observational data:
-\eq{
-V_{ij} = \frac{2}{T} \EE_{\operatorname P_{ij}} \!\!\left[\frac{\Pij{x}}{\P{x}}\right]
-}
-\STATE Compute required budget: 
-\eq{
-N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calA_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
-} 
-\ENDFOR
-\STATE {\bf if $\sum_{ij} N_{ij} \leq T/2$} {\bf return} true {\bf else return} false
-\end{algorithmic}
-\end{algorithm}
+as required.
+\end{proof}
 
-\todot{discuss role of $delta = 1/(NT)$}
-\todot{define $Z_{ijkl}(t)$ etc.}
-\todot{Define $B_{ijkl}$}
-\begin{theorem}\label{thm:general-upper}
-Let $\epsilon > 0$ be the smallest value of $\epsilon$ such that Algorithm \ref{alg:alloc} returns true and let $IJ = \argmax_{ij} \hat \mu_{ij}$.
-Then $\mu^* - \E \mu_{IJ} \leq \epsilon + 1/T$.
-\end{theorem}
-
-\subsection*{Comparison to Simple Setting}
-
-In Section \ref{sec:simple-regret} we assumed a simple graph structure and unknown distributions on $X_i$, which were then learned from observational
-data. In contrast, in this section we have allow arbitrary (known) graphs, but assumed the distribution on $X$ was known. 
-Using the approach in this section on the simple graph with known distribution leads to 
-\eq{
-\EE \mu^* - \EE \mu_{IJ} \in \bigo{\sqrt{\frac{m}{T} \log \frac{N}{\delta}}}\,, 
-}
-which is the same bound as in the previous section. 
-\todot{finish this}
+The second observation is that in the parallel bandit setting the $m$ given in this section coincides with the $m$ given there.
+\todot{Finish this}

### 2016-02-02 causal-bandit/icml-2015/general.tex
updates

@@
 the dependencies between the random variables. Learning from observational data is no longer straightforward
-because variables can be correlated through their parents. Consider, for example, the causual graph below where $X_1$
+because variables can be correlated through their parents. Consider, for example, the causual graph in Figure \ref{fig:causalStructure_confounded} where $X_1$
 deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
@@
 Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
-Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ be the mixture distribution.
+Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.
+Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
+a truncated importance weighted estimator.
+
 Let $a \in \calA$ be an intervention and define random variable
@@
 \begin{algorithmic}
-\STATE {\bf Input:} $\eta$, $T$
+\STATE {\bf Input:} $T$, $\eta \in [0,1]^{|\calA|}$, $B \in [0,\infty)^{|\calA|}$
 \FOR{$t \in \set{1,\ldots,T}$}
@@
 \begin{proof}
-First note that $|Z_a(X_t)| \leq B_a$ and that
+We abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
+First note that $|Z_{at}| \leq B_a$ and that
 \eq{
-\Var_Q[Z_a(X_t)] 
-&\leq \EE_Q[Z_a(X_t)^2] \\
+\Var_Q[Z_{at}] 
+&\leq \EE_Q[Z_{at}^2] \\
 &\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] \\
@@
 \eq{
-\EE_Q[Z] 
-&= \EE_{\Ps_a}\left[Y \ind{R_a \leq B_a}\right] \\
+\EE_Q[Z_{at}] 
+&= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] \\
 &= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] \\
@@
 }
-where $0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a > B_a}$ is the negative bias. 
+where $0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}$ is the negative bias. 
 The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
 \eq{
-\beta_a \leq \Pn{a}{R_a > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
+\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
 }
@@
 }
-Then by the union bound and Bernstein's inequality we have
+Then by the union bound and Bernstein's inequality 
 \eq{

### 2016-02-02 causal-bandit/icml-2015/general.tex
typos

@@
 \begin{proof}
+First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
 We abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
-First note that $|Z_{at}| \leq B_a$ and that
+By definition we have $|Z_{at}| \leq B_a$ and 
 \eq{
 \Var_Q[Z_{at}] 
-&\leq \EE_Q[Z_{at}^2] \\
-&\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] \\
-&\leq m(\eta)\,.
+\leq \EE_Q[Z_{at}^2] 
+\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] 
+\leq m(\eta)\,.
 }
@@
 }
-where $0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}$ is the negative bias. 
+where 
+\eq{
+0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}
+}
+is the negative bias. 
 The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
@@
 
-Algorithm \ref{alg:general} depends on a choice of sampling distribution $\eta$. In light of Theorem \ref{thm:general}
+Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$. In light of Theorem \ref{thm:general}
 a natural choice of $\eta$ is the minimiser of $m(\eta)$.
@@
 The second observation is that in the parallel bandit setting the $m$ given in this section coincides with the $m$ given there.
+Note that in the previous section we did not assume knowledge of the conditional distributions on $X$ and instead only the graph structure.
+Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
 \todot{Finish this}
+
+Finally, we believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
+for future work.
+\todot{Mention obvious modification to deal with large gaps}

### 2016-02-02 causal-bandit/icml-2015/causal-bandits-icml.tex
incorporated the simple-regret upper bound and continued to cleanup the proof
Defined parallel banit problem in setup.
added new algorithm
updates calculations for general regret upper bound to be consistent with causal liturature
Intro paragraph.
updates
some explaination of general case, changes to earlier sections to improve flow to general case
much shuffling
starting to clean up general section
Moved parallel to S3. Re-did Algorithm 1 using a instead of (i,j).
Updated contributions
More tweaks
updaets
started to clean up introduction

@@
 
-\usepackage{cleveref}
+\newif\ifsup\suptrue
+
 \usepackage{hyperref}
@@
 \newcommand{\KL}{\operatorname{KL}}
+\newcommand{\simpleregret}{R_T^{\text{simple}}}
+\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
+\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
+\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
+\newcommand{\EE}{\mathbb E}
+\newcommand{\Pn}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
+\newcommand{\parents}[1]{\operatorname{Pa}_{#1}}
+\newcommand{\actions}{\mathcal{A}}
+\newcommand{\calA}{\mathcal A}
 
@@
 \newtheorem{example}[theorem]{Example}
+\let\temp\epsilon
+\let\epsilon\varepsilon
 
@@
 \twocolumn[
-\icmltitle{Casual Bandits}
+\icmltitle{Causal Bandits: Learning Good Interventions via Causal Inference}
+% ALTERNATIVES:
+% - Causal Bandits: Sequential Decision Making with Causal Models
 
@@
 Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
-We propose a new algorithm that can make use of the causal feedback and prove that it achieves bounds on both simple and cumulative regret that are strictly better (in dependence on the number of possible interventions and number of rounds) than algorithms which do not use the additional causal information.
+We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
+than algorithms that do not use the additional causal information.
 \end{abstract}
@@
 \section{Introduction}
-
+\label{sec:intro}
 \input{introduction}
@@
 \section{Problem Setup}
-
+\label{sec:defs}
 \input{setup}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Upper Bounds on the Simple Regret}
+\section{Regret Bounds for Parallel Bandit}
+\label{sec:simple-regret}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% UPPER BOUND ON CUMULATIVE REGRET
+% UPPER BOUND ON SIMPLE REGRET GENERAL CASE
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Upper Bound on the Cumulative Regret}
-
-\input{cumulative-regret}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LOWER BOUND
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Lower Bounds}
-
-\input{lower-bound}
+\section{Regret Bounds for General Graphs}
+\label{sec:simple-regret-general}
+\input{general}
 
@@
 \section{Experiments}
-
+\label{sec:experiments}
 \input{experiments}
@@
 
-\bibliography{library}
+\bibliography{library,library2}
 \bibliographystyle{icml2015}
+
+\onecolumn
+
+\input{tech}
+
 \end{document}

### 2016-02-02 causal-bandit/icml-2015/introduction.tex
started to clean up introduction

--- 2016-02-02 causal-bandit/icml-2015/introduction.tex
@@
 Medical drug testing, policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1959}.
-In this framework, single actions (\eg, experiments or interventions) from a pre-determined set are repeatedly performed in order to determine the most effective one via feedback from a single, real-valued reward signal.
-We propose generalizing such sequential experiments by assuming that, in additional to the reward signal, values of other variables are drawn from a probabilistic causal model~\citep{Pearl2000} and observed.
-These causal models are common in social science, demography, and economics where explicit experimentation may be difficult. 
-For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades. 
-Results from causal inference relate observational distributions to interventional ones, allowing us to predict the outcome of intervention without explicitly performing it. 
-By combining causal inference using a causal graph we show, theoretically and empirically, how non-interventional observations can be used to improve the rate at which high-reward actions can be identified.
-
-% A classical problem in economics and operations research is how to choose interventions that maximise utility. 
-% We study an idealised model of such interventions using a combination of the multi-armed bandit framework \citep{Robbins1952} and the theory/language of causal inference 
-% \citep{Pearl2000}.
+In this framework, single actions (\eg, experiments or interventions) from a pre-determined set are repeatedly performed in 
+order to evaluate their effectiveness via feedback from a single, real-valued reward signal.
+We propose a generalisation of the standard model by assuming that, in additional to the reward signal, the learner observes the values of a number of covariates 
+drawn from a probabilistic causal model~\citep{Pearl2000}.
+Causal models are commonly used in disciplines where explicit experimentation may be difficult such as social science, demography and economics.
+For example, when predicting the effect of changes to childcare subsidies on workforce participation, or school choice on grades. 
+Results from causal inference relate observational distributions to interventional ones, allowing the outcome of an intervention to be predicted without
+explicitly performing it.
+By exploiting the causal information we show, theoretically and empirically, how non-interventional observations can be used to improve the rate at 
+which high-reward actions can be identified.
 
@@
 When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
-How might the farmer best experiment to identify the single, highest yielding intervention without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
+How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
+%without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
 
@@
 
-This paper can be seen as a first step towards formalising and solving problems such as the one above. 
+This paper takes the first step towards formalising and solving problems such as the one above. 
 In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph.
@@
 
+We focus on the simple regret, which measures the difference between the return of the optimal action and that of the action chosen by the algorithm after $T$ rounds.
 In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently.
-We propose a simple causal best arm identification algorithm (Algorithm~\ref{alg:simple}) for this problem and show it is optimal up to log factors (Theorems~\ref{thm:uq-simple} \& \ref{thm:lower}) with $\tilde{\Theta}(\sqrt{m/T})$ simple regret after $T$ rounds where $m$ depends on the casual model and may be much smaller than $N$.
-In contrast, the best a bandit algorithm can do if the causal model is ignored is $\Omega(\sqrt{N/T})$.
+We propose a simple causal best arm identification algorithm (Algorithm~\ref{alg:simple}) for this problem and show that up to logarithmic factors it enjoys minimax optimal
+simple regret guarantees of $\tilde\Theta(\sqrt{m/T})$ where $m$ depends on the causal model and may be much smaller than $N$.
+In contrast, existing best arm identification algorithms suffer $\Omega(\sqrt{N/T})$ simple regret.
 This shows theoretically the value of our framework over the traditional bandit problem. 
@@
 Interventions and observations may have a complex relationship in the general causal model. 
-However, in \S\ref{sec:simple-regret-general}, Theorem~\ref{thm:general-upper} shows how an importance-sampling based algorithm (Algorithm~\ref{alg:alloc}) can obtain non-trivial guarantees on simple regret.
-
-% Our analysis relies on results from causal inference that relate the distributions obtained through interventions to those based only on observations.
-% Central to our new bounds is a quantity that replaces the number of possible actions with a measure of how ``informative'' the observations are for estimating the rewards for interventions.
-
-% Although the witness (Algorithm~\ref{alg:}) for our upper bound is very simple (\ie, observe without intervening for some initial period and then play undersampled actions in order to get good estimates) our key point is that algorithms which use causal information when available have a clear, theoretical improvement over those which do not.
-% These results are experimentally validated in \S\ref{sec}.
+In \S\ref{sec:simple-regret-general} we propose an algorithm inspired by importance-sampling that (a) enjoys sub-linear regret equivalent to the optimal rate in the parallel 
+bandit setting and (b) captures many (all?) of the intricacies of sharing information in a causal graph in the general case. As in the parallel bandit case, the regret guarantee
+is often significantly better than $O(\sqrt{|\calA|/T})$ and never worse, where $|\calA|$ is the number of available interventions.
 

### 2016-02-02 causal-bandit/icml-2015/setup.tex
Tweaks
started to clean up introduction

@@
 We have established upper bounds for cumulative regret for parallel bandits but not the general causal bandit case.
-Due to space limitations we focus on simple regret and defer these to Appendix~\ref{app:cumulatve}.
+%Due to space limitations we focus on simple regret and defer these to Appendix~\ref{app:cumulatve}.
 
@@
 
-In the remainder of this paper we focus on a causal model that is sufficient to demonstrate the separation of regret bounds between algorithms that make use of the additional causal information and those which do not.
+% In the remainder of this paper we focus on a causal model that is sufficient to demonstrate the separation of regret bounds between algorithms that make use of the additional causal information and those which do not.
 

### 2016-02-02 causal-bandit/icml-2015/general.tex
started to clean up introduction
started to clean up Sec 3

@@
 \newcommand{\calP}{\mathcal P}
-\newcommand{\calA}{\mathcal A}
 \newcommand{\x}{\boldsymbol{x}}
@@
 \FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $I_t = a$ with probability $\eta$
-\STATE Do action $I_t$ and observe $X_t$ and $Y_t$
+\STATE Sample action $a_t = a$ with probability $\eta$
+\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
 \ENDFOR
@@
 }
-\STATE Choose $I = \argmax_a \hat \mu_a$
+\STATE {\bf return} $\hat a^* = \argmax_a \hat \mu_a$
 \end{algorithmic}
@@
 &\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] \\
-&\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA|
+&\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA| \qedhere
 }
-as required.
 \end{proof}
 
-The second observation is that in the parallel bandit setting the $m$ given in this section coincides with the $m$ given there.
-Note that in the previous section we did not assume knowledge of the conditional distributions on $X$ and instead only the graph structure.
-Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
-\todot{Finish this}
+\todo{give reference}
+The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section coincides with the $m$ in Equation blah.
+Recall in that setting that 
+\eq{
+\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
+}
+For $a = do()$ we let $\eta_a = 1/2$. Assume without loss of generality that $\P{X_i = 1} \geq \P{X_i = 0}$ for all $i$ and
+$\P{X_i = 0}$ is monotone increasing in $i$. Then let $m = \max\set{i : \P{X_i = 0} \leq 1 / i}$, which is the same as in \S3. 
+For $i \leq m$ and $a = do(X_i = 0)$ let $\eta_a = 1/(2m)$. Then it is easy to check that $m(\eta) \leq 2m$.
 
+\begin{remark}
+The algorithm in the previous section for the parallel bandit problem did not assume knowledge of the conditional distributions on $X$ and 
+instead only the graph structure. Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
 Finally, we believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
 for future work.
-\todot{Mention obvious modification to deal with large gaps}
+\end{remark}

### 2016-02-02 causal-bandit/icml-2015/setup.tex
started to clean up Sec 3

--- 2016-02-02 causal-bandit/icml-2015/setup.tex
@@
 One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
-The following game is then played over $T$ rounds.
-In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, including the \emph{reward} $Y_t \in \{0,1\}$.
+We denote the expected reward for the intervention $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
+the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
 
-Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are also possible. 
-If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already well understood~\citep{Agarwal2014}. 
+The game proceeds over $T$ rounds.
+In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations. 
+It then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, 
+including the \emph{reward} $Y_t \in \{0,1\}$. 
+After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T$.
+
+The objective of the learner is to minimise the simple regret
+\begin{equation}
+\label{eq:regret-simple}
+	\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.
+\end{equation}
+This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009} or ``best-arm identification'' problem~\citep{Gabillon2012} and
+is most appropriate when the learner has a fixed learning budget after which its policy will be fixed indefintely.
+
+Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are possible. 
+If the non-intervened variables are observed before an intervention is selected our framework 
+reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}. 
 Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of 
@@
 
-The learning goal we will consider and analyse is identifying the intervention with highest expected reward by round $T$.
-This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009} or ``best-arm identification'' problem~\citep{Gabillon2012}.
-This objective is most appropriate when the learning agent has a fixed learning budget after which its policy will be fixed indefintely. 
-We denote the expected reward for the intervention $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
-Performance for the first goal can be measured by \emph{simple regret}, \ie, the gap between the optimal expected reward and the intervention $a_T$ chosen by the learner on round $T$:
-\begin{equation}
-\label{eq:regret-simple}
-	\simpleregret = \mu^* - \E{\mu_{a_T}}.
-\end{equation}
-
-Another common objective in bandit problems is maximising the \emph{cumulative regret}, the total expected reward over all $T$ rounds. 
-We have established upper bounds for cumulative regret for parallel bandits but not the general causal bandit case.
-Due to space limitations we focus on simple regret and defer these to Appendix~\ref{app:cumulatve}.
-
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
-The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
+We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one 
+edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some 
+arbitrary but unknown, real-valued function $r$. The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
+Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by 
+treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
 Intuitively though, one would expect to perform better by making use of the extra observations.
 
-In the remainder of this paper we focus on a causal model that is sufficient to demonstrate the separation of regret bounds between algorithms that make use of the additional causal information and those which do not.
-
+\iffalse
 \subsection{Exploiting the Causal Model}
@@
 \end{definition}
+
+\fi

### 2016-02-02 causal-bandit/icml-2015/simple-regret.tex
started to clean up Sec 3

@@
-In this section we propose and analyse an algorithm for achieving the optimal simple regret in a special case of a causal bandit problem.
-
-\subsection{The Parallel Bandit Problem}
-
-We now consider a class of a causal bandit problems which we call \emph{parallel bandits}.
+In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special 
+case of the causal bandit problem which we call the {\it parallel bandit}.
 It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example. 
@@
 
-The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \R$, as shown in Figure~\ref{fig:causalStructure}.
+The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a 
+reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:causalStructure}.
 All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: 
-\[
-	\mathcal{A} = \{do()\} \cup \{ do(X_i = j) \colon i \in \{1, \ldots, N\}, j \in \{0,1\}\}.
-\]
+\eq{
+	\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
+}
 In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
@@
 Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
-\todom{Should $Y$ be 0-1 valued or real everywhere?}
-The value of the reward variable is distributed as $Y = r(\vec{X}) + \eta$ where $r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function and $\eta$ is $1$-subgaussian noise with a distribution that may depend on $\vec{X}$.
-In the farming example, this choice of $Y$ models the fixed, noisy but unknown dependence of crop yield on the temperature, soil, and moisture.
+%\todom{Should $Y$ be 0-1 valued or real everywhere?} YES!
+The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
+$r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function. 
+In the farming example, this choice of $Y$ models the success or failure of a seasons crop, which depends stochastically on the various environment variables.
+Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
 
-\todom{$i,j$ notation to $a \in \actions$}
-In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
-We will also make use of the expansion of the expected reward for $do(X_i = j)$:
-\eq{
-\mu_{i,j} 
-&= \E{r(X)|do(X_i = j)} \\
-&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
-}
-The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
-The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$. 
-Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
+%\todom{$i,j$ notation to $a \in \actions$}
+%In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
+%We will also make use of the expansion of the expected reward for $do(X_i = j)$:
+%\eq{
+%\mu_{i,j} 
+%&= \E{r(X)|do(X_i = j)} \\
+%&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
+%}
+%The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
+%The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$.
 
@@
 
+The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
+The fact that $X_1,\ldots,X_N$ are independent can be used to create good estimators of the returns of those
+actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. These are the interventions that are essentially observed
+from observational data. Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
+the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
+the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
+carefully chosen to minimise the worst-case simple regret.
+
 \begin{algorithm}[h]
-\caption{Causal Best Arm Identification}\label{alg:simple}
+\caption{Parallel Bandit Algorithm}\label{alg:simple}
 \begin{algorithmic}[1]
-\STATE {\bf Input:} Total rounds $T$, causal graph $\mathcal{G}$, allowable interventions $\actions$. 
-\FOR{$t \in 1,\ldots,(T - 1) / 2$}
-\STATE Perform empty intervention $do()$; observe $\vec{x}_t$ and $r_t$
+\STATE {\bf Input:} Total rounds $T$ and $N$.
+\FOR{$t \in 1,\ldots,T / 2$}
+\STATE Perform empty intervention $do()$; observe $\vec{X}_t$ and $Y_t$
 \ENDFOR
 \FOR{$a = do(X_i = x) \in \actions$}
-\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{x_{t,i} = x}$
-\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{x_{t,i} = x} r_t$
-\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T-1}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
+\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
+\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
+\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
 \ENDFOR

### 2016-02-02 causal-bandit/icml-2015/introduction.tex

@@
 We focus on the simple regret, which measures the difference between the return of the optimal action and that of the action chosen by the algorithm after $T$ rounds.
-In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently.
+In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently and there are $2N$ possible interventions.
 We propose a simple causal best arm identification algorithm (Algorithm~\ref{alg:simple}) for this problem and show that up to logarithmic factors it enjoys minimax optimal
@@
 This shows theoretically the value of our framework over the traditional bandit problem. 
-Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in bandit problems.
+Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in this framework.
+\todom{Mention cumulative regret?}
 
-Interventions and observations may have a complex relationship in the general causal model. 
-In \S\ref{sec:simple-regret-general} we propose an algorithm inspired by importance-sampling that (a) enjoys sub-linear regret equivalent to the optimal rate in the parallel 
-bandit setting and (b) captures many (all?) of the intricacies of sharing information in a causal graph in the general case. As in the parallel bandit case, the regret guarantee
-is often significantly better than $O(\sqrt{|\calA|/T})$ and never worse, where $|\calA|$ is the number of available interventions.
+In the general casual model interventions and observations may have a complex relationship. 
+In \S\ref{sec:simple-regret-general} we propose an algorithm (Algorithm~\ref{alg:alloc}) inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
+As in the parallel bandit case, the regret guarantee is often significantly better than $O(\sqrt{N/T})$ and never worse, where $N$ is the number of available interventions.
 

### 2016-02-02 causal-bandit/icml-2015/simple-regret.tex
starting to clean up Section 3

@@
 In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
-In this case, the interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ might represent the use of shades or heat lamps to keep the temperature low or high, respectively.
+The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.
 
@@
 
-In each round of the game, the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
+In each round the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
 The remainder of the variables are simultaneously set by independently biased coin flips. 
@@
 The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
-$r : \{0,1\}^N \to \R$ is an arbitrary, fixed, and unknown function. 
+$r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
 In the farming example, this choice of $Y$ models the success or failure of a seasons crop, which depends stochastically on the various environment variables.
@@
 the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
-carefully chosen to minimise the worst-case simple regret.
+carefully chosen to minimise the worst-case simple regret. The difficulty of the problem depends on $\vec{q}$. 
+Define $m \in \set{1, \ldots, N}$ by
+\eq{
+m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, (1 - q_k)}} \leq 1/i}\,.
+}
 
@@
 \FOR{$t \in 1,\ldots,T / 2$}
-\STATE Perform empty intervention $do()$; observe $\vec{X}_t$ and $Y_t$
+\STATE Perform empty intervention $do()$
+\STATE Observe $\vec{X}_t$ and $Y_t$
 \ENDFOR
@@
 \STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
-\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$ and $\hat{V}_a = \frac{1}{\hat{p}_a}$
+\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$ 
+\STATE $\hat q_i = \hat{q}_{do(X_i = 1)}$
 \ENDFOR
-\STATE Compute $\hat{m} = m(\hat{\vec{V}})$ and $A = \set{a \in \actions \colon \hat{V}_a \geq \hat{m}}$.
-\STATE Let $T_A := \frac{T-1}{2 |A|}$ be times to sample each $a\in A$.
+\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq 1/\hat m}$.
+\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
 \FOR{$a = do(X_i = x) \in A$}
 \FOR{$t \in 1,\ldots,T_A$}
-\STATE Intervene with $a$ and observe $r_t$
+\STATE Intervene with $a$ and observe $Y_t$
 \ENDFOR
-\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} r_t$
+\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
 \ENDFOR
@@
 \eq{
-\simpleregret \in \bigo{\sqrt{\frac{m}{T}\log\left(\frac{NT}{m}\right)}}\,.
+\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m}\right)}}\,.
 }

### 2016-02-02 causal-bandit/icml-2015/introduction.tex
minor updates

@@
 In the general casual model interventions and observations may have a complex relationship. 
-In \S\ref{sec:simple-regret-general} we propose an algorithm (Algorithm~\ref{alg:alloc}) inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
+In \S\ref{sec:simple-regret-general} we propose an novel algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
 As in the parallel bandit case, the regret guarantee is often significantly better than $O(\sqrt{N/T})$ and never worse, where $N$ is the number of available interventions.

### 2016-02-03 causal-bandit/icml-2015/code/causal_bandits.py
basic implementation to find eta. Convergence not great

@@ -1,0 +1,120 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Feb  3 09:51:47 2016
+
+@author: finn
+"""
+import numpy as np
+from itertools import product
+from scipy.optimize import minimize
+from scipy.optimize import fmin_l_bfgs_b
+import sys
+
+class BalancedParallel(object):
+    def __init__(self,N,q):
+        self.N = N
+        self.pX = np.vstack((1-q,q))
+       
+    def P(self,x):
+        """ calculate vector of P_a for each action a """
+        indx = np.arange(len(x))
+        ps = self.pX[x,indx] #probability of xi for each i
+        joint = ps.prod()
+        pi = np.true_divide(joint,ps) # will be nan for elements for which ps is 0 -(should never happen if sampling)
+        for j in np.where(np.isnan(pi))[0]:
+            pi[j] = np.prod(ps[indx != j])
+        
+        pij = np.vstack((pi,pi))
+        pij[1-x,indx] = 0
+        pij = pij.reshape((len(x)*2,)) #flatten first N-1 will be px=0,2nd px=1
+        pobserve = joint# the probability of x given do()
+        return np.hstack((pij,pobserve))
+        
+    def V(self,eta):
+        xvals = map(np.asarray,product([0,1],repeat = self.N)) # all possible assignments to our N variables that are parents of Y
+        va = np.zeros(2*self.N + 1)  
+        for x in xvals:
+            pa = self.P(x)
+            Q = (eta*pa).sum()
+            ratio = np.true_divide(pa**2,Q)
+            ratio[np.isnan(ratio)] = 0 # we get nan when 0/0 but should just be 0 in this case
+            va += ratio
+            
+        return va 
+
+
+BIG_FLOAT = sys.float_info.max   
+        
+N = 6 # number of parents of Y
+n = 2*N+1 # number of actions
+q = np.full(N,.5)
+q[0:2] = 1
+print q
+model = BalancedParallel(N,q)
+
+eta0 = np.full(n,1.0/n)
+eta1 = np.zeros(n)
+eta1[-1]=1 
+eta2 = np.zeros(n)
+eta2[0] = 1
+
+
+print model.V(eta0)
+
+def random_eta(n):
+    eta = np.random.random(n)
+    eta = eta/eta.sum()
+    return eta
+
+def maxV(eta):
+    penalty = 5000*(eta.sum()-1)**2
+    maxV = model.V(eta).max()
[…]

### 2016-02-03 causal-bandit/icml-2015/general.tex

@@
 }
-Define $m(\eta)$ by
+Abusing notation we define $m(\eta)$ by
 \eq{
@@
 }
+We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately cooincides 
+with the version given in the previous section, which justifies the name overloading.
 
@@
 Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
-Therefore the minimisation problem may be tackled using standard techniques from convex optimisation. In the experimental section we present some special cases,
+Therefore the minimisation problem may be tackled using standard techniques from convex optimisation. 
+An interpretation of $m(\eta^*)$ is the minimum achievable worst-case variance of the importance weighted estimator.
+In the experimental section we present some special cases,
 but for now we give two simple results. The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
 
-\begin{lemma}
+\begin{proposition}
 $m(\eta^*) \leq |\calA|$.
-\end{lemma}
+\end{proposition}
 
@@
 
-\todo{give reference}
-The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section coincides with the $m$ in Equation blah.
+The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
 Recall in that setting that 
@@
 }
-For $a = do()$ we let $\eta_a = 1/2$. Assume without loss of generality that $\P{X_i = 1} \geq \P{X_i = 0}$ for all $i$ and
-$\P{X_i = 0}$ is monotone increasing in $i$. Then let $m = \max\set{i : \P{X_i = 0} \leq 1 / i}$, which is the same as in \S3. 
-For $i \leq m$ and $a = do(X_i = 0)$ let $\eta_a = 1/(2m)$. Then it is easy to check that $m(\eta) \leq 2m$.
+For $a = do()$ choose $\eta_a = 1/2$. 
+Let $m = m(\vec{q})$ and for $a = do(X_i = j)$ let $\eta_a = \ind{\P{X_i = j} \leq 1/m} / (2m)$. 
+Then it is easy to check that $m(\eta) \leq 2m$.
 

### 2016-02-03 causal-bandit/icml-2015/code/causal_bandits.py
run 1

@@ -7,14 +7,62 @@ Created on Wed Feb  3 09:51:47 2016
 import numpy as np
 from itertools import product
 from scipy.optimize import minimize
-from scipy.optimize import fmin_l_bfgs_b
-import sys
+from math import sqrt
+from math import log
+from numpy.random import binomial
+import matplotlib.pyplot as plt
+from time import time
 
-class BalancedParallel(object):
-    def __init__(self,N,q):
-        self.N = N
-        self.pX = np.vstack((1-q,q))
-       
+
+def random_eta(n):
+    eta = np.random.random(n)
+    eta = eta/eta.sum()
+    return eta
+    
+def most_unbalanced_q(N,m):
+    q = np.full(N,1.0/m,dtype=float)
+    q[0:m] = 0
+    return q
+    
+
+def maxV(eta):
+    maxV = model.V(eta).max()
+    if not np.isfinite(maxV):
+        return np.inf
+    return maxV#+penalty # consider penalizing eta going beyond bounds 
+
+class Parallel(object):
+    def __init__(self,q,epsilon):
+        """ actions are do(x_1 = 0)...do(x_N = 0),do(x_1=1)...do(N_1=1), do() """
+        assert q[0] <= .5, "P(x_1 = 1) should be <= .5 to ensure worst case reward distribution can be created"
+        self.epsilon = epsilon
+        self.epsilon_minus = self.epsilon*q[0]/(1.0-q[0]) 
+        self.N = len(q) # number of variables
+        self.K = 2*self.N+1 #number of actions
+        self.pX = np.vstack((1.0-q,q))
+        
+        self.expected_rewards = np.full(self.K,.5)
+        self.expected_rewards[0] = .5 - self.epsilon_minus
+        self.expected_rewards[self.N] = .5+self.epsilon
+        self.optimal = .5+self.epsilon
+         
+    def sample(self,action):
+        x = binomial(1,self.pX[1,:])
+        if action != self.K - 1: # everything except the do() action
+            i,j = action % self.N, action/self.N
+            x[i] = j
+        y = binomial(1,self.pYgivenX(x))
+        return x,y
+    
+    def sample_multiple(self,actions,n):
+        """ sample the specified actions, n times each """
+        return binomial(n,self.expected_rewards[actions])
+        
+    def pYgivenX(self,x):
+        if x[0] == 1:
+            return .5+self.epsilon
[…]

### 2016-02-03 causal-bandit/icml-2015/introduction.tex

@@
 Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in this framework.
-\todom{Mention cumulative regret?}
 
-In the general casual model interventions and observations may have a complex relationship. 
-In \S\ref{sec:simple-regret-general} we propose an novel algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
+In the general casual bandit problem interventions and observations may have a complex relationship. 
+In \S\ref{sec:simple-regret-general} we propose a novel algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
 As in the parallel bandit case, the regret guarantee is often significantly better than $O(\sqrt{N/T})$ and never worse, where $N$ is the number of available interventions.
@@
 As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
-In the farming example, the farmer could treat the six possible interventions are arms, ignore the observations of the non-intervened variables, and apply an algorithm such as UCB~\cite{Auer1995} with well understood regret guarantees.
+In the farming example, the farmer could treat the six possible interventions are arms, ignore the observations of the non-intervened variables, and apply an existing best-arm identificaion algorithm~\citep{Jamieson2013} with well understood simple regret guarantees.
 However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.
 
-Bandit problems with side information are commonly referred to as ``contextual bandits'' and have been well studied in recent years~\cite{Langford2008,Agarwal2014}.
+A well-studied class of bandit problems with side information are ``contextual bandits''~\cite{Langford2008,Agarwal2014}.
 Our framework bears a superficial similarity to contextual bandit problems since the extra observations on non-intervened variables might be viewed as context for selecting an intervention. 
-However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and cannot be used as context.
-We discuss the relationship between causal and contextual bandits further in \S\ref{sec:defs} and leave their combination as future work.
+However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
+% We discuss the relationship between causal and contextual bandits further in \S\ref{sec:defs} and leave their combination as future work.
 
-There has been a significant amount of recent work which proposes alternative modes of feedback within the bandit setting \citep{Alon2015} which, at first glance, may appear applicable.
-Indeed, we show in \S~\ref{sec:discussion}, that it is possible to applying these existing results to our problem, however the resulting regret is $\bigo{\sqrt{NT}}$, again with a sub-optimal dependence on $N$.
-
-\todom{Rework rest of intro}
-
-%Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
-For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
-
-% A widely used framework for sequential decision making is the multi-armed bandit.
-In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary.
-At each timestep the agent selects an action and receives a reward sampled i.i.d from the corresponding reward distribution.
-The performance of bandit algorithms is described by the regret: the difference in the expected reward obtained by the algorithm and the reward that could be obtained if the optimal action was selected at every timestep.
-
-% We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
-
-% A natural way to connect the causal framework with the bandit setting is to model the problem as a causal directed acyclic graph.
-Each possible assignment of variables to values is an action (bandit arm).
-The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action.
-The number of actions grows exponentially with the number of variables in the graph, making it important to use algorithms that take account of the graph structure to reduce the search space.
-
-% Problems framed in this way take on characteristics of different bandit settings depending on the assumptions we make about what subset of actions can be taken, what variables are observable and whether they are observed before or after an action is selected.
-If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
-
-% If we receive feedback on additional nodes the problem can be more interesting.
-In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
-
-% We consider a bandit problem where the actions and reward are represented by a specific causal graph that demonstrates this interesting structure.
-We develop an algorithm to leverage the information provided by this structure and demonstrate it substantially outperforms standard bandit algorithms applied to the same problem where the number of actions is large.
-
-There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
-Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
-There is also an active line of research into bandits with feedback defined by a graph.
-Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
-The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
+There have been several proposals for bandit problems where extra feedback is received after an action is taken.
+\citet{Alon2015} consider a very general model where rewards on unplayed actions are revealed according to a feedback graph.
+As we discuss in \S\ref{sec:discussion}, the problem they consider shares some similarity with ours however their results are for cumulative regret and thus cannot be used to guarantee low simple regret~\citep{Bubeck2009}.
+Even in the case of cumulative regret we show that an application of their results to the parallel bandit problem with $N$ variables yields a $\bigo{\sqrt{NT}}$ bound.
+\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
+They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
+\citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled. 
+The main difference from our present work is our focus on simple regret and exploitation of the more complex information related rewarding via the causal graph.
+To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.
 
@@
 Two pieces of recent work also consider using causal models in bandit problems.
-Although \citet{Bareinboim2015} also use causal inference to improve the chance of choosing an optimal arm, their work differs from our in two key ways.
+\citet{Bareinboim2015} also use causal inference to improve the chance of choosing an optimal arm.
+Their work differs from ours in two key ways.
 Firstly, their focus is on correcting the effects of unobserved confounding variables whereas we assume all variables are observable.
 Secondly, they do not assume their learning algorithm has access to a causal graph and instead use a form of contextual randomization to improve reward estimates in a Thompson sampling-style algorithm.
-\citet{Ortega2014thompson} also present an analysis and extension of Thompson sampling assuming actions are causal interventions.
-Their focus, however, is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
+\citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions.
+Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
 Neither work provides a regret analysis.

### 2016-02-03 causal-bandit/icml-2015/code/causal_bandits.py
as completes run 1

@@ -14,6 +14,7 @@ import matplotlib.pyplot as plt
 from time import time
 
 
+
 def random_eta(n):
     eta = np.random.random(n)
     eta = eta/eta.sum()
@@ -210,10 +211,10 @@ baseline  = SuccessiveRejects()
 
 ts = time()   
 regret = np.zeros((len(T_vals),3,simulations))
-for T_indx,T in enumerate(T_vals):
-    for s in xrange(simulations):
-        if s % 100 == 0:
+for s in xrange(simulations):
+    if s % 100 == 0:
             print s
+    for T_indx,T in enumerate(T_vals):     
         regret[T_indx,0,s] = causal.run(T,model,eta,4)
         regret[T_indx,1,s] = causal.run(T,model2,eta2,2)
         regret[T_indx,2,s] = baseline.run(T,model)
@@ -223,7 +224,12 @@ print 'took: %2.4f sec' % (te-ts)
 mean = regret.mean(axis=2)       
 plt.plot(T_vals,mean)
 
-
+error = 3*regret.std(axis=2)/sqrt(simulations)
+fig,ax = plt.subplots()
+ax.errorbar(T_vals,mean[:,0],yerr=error[:,0], label="Causal m(eta) = 4",linestyle="",marker="o")    
+ax.errorbar(T_vals,mean[:,2],yerr=error[:,2], label="Successive Rejects",linestyle="",marker="D") 
+ax.legend(loc="lower left",numpoints=1)
+   
 
 #
 ##constraints=({'type':'eq','fun':lambda eta: eta.sum()-1.0})

### 2016-02-03 causal-bandit/icml-2015/random_notes.tex
replaces 2015 style with 2016

Strictly better performance is not true for the case where q is unknown (due to the hard transition between exploration and exploitation). We have performance of the same order or better than algorithms that don't use the causal structure. Maybe an algorithm that did offer strictly better performance is possible - although there could be a trade off against performance on 'easy' problems?

We can then marginalize to obtain $P(Y|do(X_i = j))$

\eq {
P(Y|do(X_i = j)) =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\\
 =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\x}\\
 =& \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}\P{Y|\x}
}

The query $P(Y|do(V)$ is identifiable for any graph with known structure and fully observable variables. This means we can create an estimator for the reward of any action in terms only of results observed when the $do()$ action was selected (although the variance of these estimators can be infinite if any node conditional probabilities are zero). For example, consider the graph in \cref{fig:causalStructure_confounded}.

We can generalize the (simple regret) algorithm developed for the parallel bandit problem to general graphs as follows.

\begin{itemize}
\item Apply the do-calculus to write an expression for the reward of each action in term of observational data: $P(Y|do(X_i=j)) = \sum_{pa_i} P(Y|X_i = j,pa_i)P(pa_i)$ \cite{Pearl2000} Theorem 3.2.2
\item Create an estimator from this expression
\item Observe for $T/2$ rounds.
\item Calculate the variance in the estimators from $P(pa_i)$ and $P(X_i|pa_i)$ if they are known (corresponds to known $q$ case) or estimate the variance from the observations (as for unknown $q$ case). 
\item Compute the set of poorly estimated actions, $A$ based on the inverses of the variances.
\item Explicitly play actions in $A$ for the remaining $T/2$ rounds.
\end{itemize}

It is unclear if this approach remains optimal. The $do()$ action is not the only one that can reveal information about the rewards of multiple other actions. In the example in figure \ref{fig:causalStructure_confounded}, $do(Z)$ is another obvious candidate. We would need to establish a matching lower bound - or prove that $do()$ is always the most revealing action.

We are also not using the fact that the rewards are not entirely independent (They are 'almost' independent in the parallel bandit example). This does not apply in general. Consider a chain graph $X_1 \rightarrow X_2 \rightarrow, ..., X_N \rightarrow Y$. In this case $k > i \implies P(Y|do(X_i)) \leq P(Y|do(X_k))$.

### 2016-02-03 causal-bandit/icml-2015/unused/test_code.py

@@ -1,0 +1,37 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Fri Jan 29 16:58:32 2016
+
+@author: finn
+"""
+
+
+    def test_sample(self, simulations):
+        xdata = np.zeros((simulations,self.N),dtype=float)
+        ydata = np.zeros(simulations)
+        for s in xrange(simulations):
+            x,y,p = self.sample()
+            xdata[s] = x
+            ydata[s] = y
+        xm = np.mean(xdata,axis=0)
+        xse = 3*np.std(xdata,axis=0)/sqrt(simulations)
+        xtrue = np.concatenate(([self.pZ],self.pZ*self.pXgivenZ[0] + (1-self.pZ)*self.pXgivenZ[1]))
+        print xtrue
+        diffs = np.abs(xtrue-xm)
+        over = np.sum(diffs > xse)
+        p = (1-.997)
+        assert over <= (self.N*p + self.N*p*(1-p)),over
+        
+            def sample_given_action(self,arm):
+        k,j = arm%self.N,1-arm/self.N
+        assert 0 <= k <= self.N, "k not a variable indx"
+        assert j in [0,1], "j must be 0 or 1"
+        
+        V = np.empty(self.N,dtype=int)
+        V[0] = binomial(1,self.pZ)
+        V[k] = j
+        V[1:] = binomial(1,self.pXgivenZ[V[0]])
+        V[k] = j
+        Y = binomial(1,self.pYgivenXi[V[self.i]])
+        return V,Y
+        

### 2016-02-03 causal-bandit/icml-2015/tech.tex
updates
started to clean up Sec 3
updates to Theorem 1

--- 2016-01-31 causal-bandit/icml-2015/tech.tex
@@
-\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
-\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
-\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
-\newcommand{\EE}{\mathbb E}
-\newcommand{\Pn}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PROOF OF SIMPLE REGRET UPPER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}
 
-\subsection*{Truncated Importance Sampling}
+Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$. The assumption is non-restrictive since all variables
+are independent and permutations of the variables can be pushed to the reward function.
+The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound.
 
-Let $P$ and $Q$ be distributions on $\Omega = \set{0,1}^N$ and let $Y \in \set{0,1}$ be defined in terms of its conditional expectation given $X \in \Omega$.
+\begin{lemma}\label{lem:conc1}
+Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
 \eq{
-\forall x \in \Omega \qquad \EE[Y|X = x] = r(x)\,,
+\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
 }
-where $r:\Omega \to [0,1]$ is some unknown function.
-We tackle the problem of estimating $\mu = \EE_Q r = \sum_x \Q{x} r(x)$ using
-samples of $X$ and $Y$ with $X$ sampled from $P$.
-Define random variable $R(x) = Q(x)/P(x)$ and estimator $Z$ by
-\eqn{
-\label{eq:truncated}
-Z = Y R \ind{R < B}\,,
+\end{lemma}
+
+\begin{lemma}\label{lem:conc2}
+Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE X_i = p$ and $\delta \in [0,1]$.
+Then 
+\eq{
+\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \leq 4\delta\,.
 }
-which trivially satisfies $|Z| \leq B$ surely.
-Note that the usual importance weighted estimator would be $YR$, which if $Q$ is absolutely continuous with respect to $P$ satisfies $\EE_P YR = \mu$. Unfortunately
-this estimator may suffer from poor concentration, a problem that can be mitigated via the truncation in (\ref{eq:truncated}) at the cost of some bias.
-Computing the expectation we see
+\end{lemma}
+
+\begin{proof}
+For $\delta \geq 1/4$ the result is trivial. Otherwise 
+by Hoeffding's bound and the union bound:
 \eq{
-\EE_P[Z] = \mu - \beta\,.
+\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} 
+&\leq \sum_{t = n_0}^\infty \P{\left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \\
+&\leq 2\sum_{t=n_0}^\infty \exp\left(-\frac{t}{n_0} \log \frac{2}{\delta}\right) 
+\leq 4\delta\,. \qedhere
 }
-where $\beta \geq 0$ is the negative bias and satisfies
+\end{proof}
+
+\begin{lemma}\label{lem:m_est}
+Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
 \eq{
-\beta \leq \Q{R \geq B}\,.
+\P{2m(\vec{q}) / 3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})} \geq 1 - \delta\,.
 }
-Bounding the variance is also straightforward.
+\end{lemma}
+
+\begin{proof}
+Let $F$ be the event that there exists and $1 \leq i \leq N$ for which
 \eq{
-\Var[Z] 
-\leq \EE[Z^2] 
-=\sum_x \P{X = x} Z(x)^2 
-\leq \sum_x \P{X = x} \left(\frac{\Q{X = x}}{\P{X = x}}\right)^2 
-\leq \sum_x \Q{X = x} \frac{\Q{X = x}}{\P{X = x}} 
-= V\,.
+\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}\,.
 }
-The bias can be controlled in terms of $V$ via Markov's inequality.
+Then by the union bound and Lemma \ref{lem:conc1} we have $\P{F} \leq \delta$. The result will be completed by showing that
+when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
+Towards this end recall that $m(\vec{q}) = \min\set{i : q_i \geq 1 / i}$.
+Therefore for $i > m$ we have $q_i \geq q_m \geq 1/m$ and so by Lemma \ref{lem:conc1} we have
 \eq{
-\beta 
-\leq \Q{R \geq B} 
-\leq \frac{\EE_Q[R]}{B}
-= \frac{V}{B}\,.
+\frac{3}{4} 
+&\geq \frac{1}{2} + \sqrt{\frac{3}{T} \log \frac{2N}{\delta}} 
+\geq q_i + \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}} 
+\geq \hat q_i \\
+&\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
+\geq q_i - \sqrt{\frac{q_i}{8m}}
+\geq \frac{1}{2m}\,.
 }
-Finally let $\delta > 0$ be the desired confidence level and choose 
+Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m$.
+For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m$ that
 \eq{
-B = \sqrt{\frac{V n}{\log\frac{1}{\delta}}}\,,
+\hat q_i 
+\leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
+\leq \frac{1}{m} \left(1 + \sqrt{\frac{1}{8}}\right)
+\leq \frac{3}{2m}\,.
 }
-which implies that the negative bias satisfies
+Therefore $m(\vec{\hat q}) \geq 2m(\vec{q}) / 3$ as required. 
+\end{proof}
+
+\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
+Let $\delta = m = m(\vec{q}) / N$. Then by Lemma \ref{lem:m_est} we have 
 \eq{
-\beta \leq \sqrt{\frac{V}{n} \log \frac{1}{\delta}}\,. 
+\P{2m/3 \leq m(\vec{\hat q}) \leq 2m} \geq 1 - \delta\,.
 }
-Let $Z_1,\ldots,X_n$ be $n$ i.i.d.\ copies of $Z$. Then by the Bernstein inequality we have
+Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then
+for $a \in A$ the algorithm estimates $\mu_a$ from $T/(2m(\vec{\hat q})) \geq T/(4m)$ samples.
+Therefore by Hoeffding's inequality and the union bound we have
 \eq{
-\P{\left|\frac{1}{n} \sum_{t=1}^n Z - \mu\right| \geq \sqrt{\frac{V}{n} \log\frac{1}{\delta}}} \leq \delta\,.
+\P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{8m}{T} \log\frac{2N}{\delta}}} \leq \delta\,.
 }
-
-\subsection*{Mixing Truncate Importance Sampling}
-
-Now we study the problem of estimating $\mu = \EE_Q \P{Y = 1|X}$ from samples drawn
-from a set of sampling distributions $\operatorname{P}_1,\ldots,\operatorname{P}_k$ with $N_i$ samples
-drawn of $\operatorname{P}_i$.
-Let $R_i(x) = \Q{x} / \Pn{i}{x}$ and
+For arms not in $a$ we have $\hat p_a \geq 1/m(\vec{\hat q}) \geq 1/(2m)$.
+Therefore if $a = do(X_i = j)$, then 
 \eq{
-V_i = \sum_{x \in \Omega} \Q{X = x} \frac{\Q{X = x}}{\Pn{i}{X = x}}
+\hat p_a = \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{1}{2m}\,. 
 }
-and
+Therefore $\sum_{t=1}^{T/2} \ind{X_{t,i} = j} \geq T/4m$
+and by Lemma \ref{lem:conc2} we have
 \eq{
-Z_i = Y R_i \ind{R_i < B_i}\,,
+\P{\sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{T}{4m} \text{ and } \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}}} \leq 4\delta / N\,.
 }
-where
+Therefore with probability at least $1 - 6\delta$ we have
 \eq{
-B_i = V_i \sqrt{\frac{\sum_{i=1}^k \frac{N_i}{V_i}}{\log \frac{1}{\delta}}}\,,
+(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{8m}{T} \log \frac{N}{\delta}} = \epsilon\,.
 }
-which for $k = 1$ reduces to the choice of $B$ in the previous section.
-Then $\EE_{\operatorname{P_i}} Z_i = \mu - \beta$ where
+If this occurs, then 
 \eq{
-0 \leq \beta_i 
-\leq \frac{V_i}{B_i}
-= \sqrt{\frac{1}{\sum_{i=1}^k \frac{N_i}{V_i}} \log\frac{1}{\delta}}\,.
-}
-Let $\eta_i \in [0,1]$ be a mixture coefficient that minimises the variance of the mixture 
-estimator, which weights samples from $\operatorname{P_i}$ by a bound on its inverse variance, which is $V_i / N_i$.
-\eq{
-\eta_i = \frac{\frac{N_i}{V_i}}{\sum_k \frac{N_k}{V_k}}\,.
-}
-Let $Z_{i,1},\ldots,Z_{i,N_i}$ be $N_i$ copies of $Z_i$ for each $i$.
-Then we define the mixture estimator by
-\eq{
-\hat \mu = \frac{\sum_i \eta_i \sum_{t=1}^{N_i} \frac{Z_{it}}{N_i}}{\sum_i \eta_i} 
-}
-Now we prepare to use Bernstein's inequality to control the concentration of $\hat \mu$ about its mean.
-\eq{
-b 
-= \max_i \frac{\frac{\eta_i B_i}{N_i}}{\sum_i \eta_i} 
-= \max_i \frac{\alpha}{\sum_i \frac{N_i}{V_i}} 
-= \frac{1}{\alpha \sqrt{\log\frac{1}{\delta}}}\,.
+\mu_{\hat a^*} \geq \hat \mu_{\hat a^*} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
 }
@@
 \eq{
-\P{\left|\hat \mu - \mu\right| \geq \sqrt{\frac{1}{\sum_{i=1}^k \frac{N_k}{V_k}} \log \frac{1}{\delta}}} \leq \delta\,.
+\mu^* - \EE[\mu_{\hat a^*}] 
+\leq 6\delta + \epsilon 
+\leq \frac{6m}{T} + \sqrt{\frac{32m}{T} \log \frac{NT}{m}}\,, 
 }
+which completes the result.
+\end{proof}
 
-\subsection*{General Case}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LOWER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}
 
-Let $\operatorname P_{ij}$ be the measure on $\Omega$ with $\Pij{X = x} = \P{X = x|do(X_i = j)}$.
-
-\begin{algorithm}[H]
-\caption{CheckValid}\label{alg:alloc}
-\begin{algorithmic}
-\STATE {\bf Input:} $\epsilon > 0$, $\delta > 0$, $\operatorname P_1,\ldots,\operatorname P_N$
-\STATE Compute $o_1,\ldots,o_N \in \set{1,\ldots,N}$ such that $\calP_i \subseteq \set{o_k : k < i}$ for all $i$
-\FOR{$i \in \set{o_1,\ldots,o_N}$}
-\STATE Compute variances:
+\todot{Change this to Bernoulli}
+Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
+For each $i$ define reward function $r_i$ by
 \eq{
-\forall k \in \calP_i \text{ and } l \in \set{0,1} \qquad
-V_{ijkl} = \frac{1}{N_{kl}} \EE_{\operatorname P_{ij}}\!\! \left[\frac{\Pij{x}}{\Pkl{x}}\right] 
+r_0(\boldsymbol{X}) &= \frac{1}{2} &
+r_i(\boldsymbol{X}) &= \begin{cases}
+\frac{1}{2} + \epsilon & \text{if } X_i = 1 \\
+\frac{1}{2} & \text{otherwise}\,,
+\end{cases}
 }
-\STATE Compute variance for learning from do nothing action:
+where $\epsilon > 0$ is some constant to be chosen later.
+We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
+environment determined by $\boldsymbol{q}$ and $r_i$. Let $P_i$ be the corresponding measure
+on all observations over all $T$ rounds and $\mathbb E_i$ the expectation with respect to $P_i$. By Lemma 2.6 by \cite{Tsy08} we have
 \eq{
-V_{ij} = \frac{2}{T} \EE_{\operatorname P_{ij}} \!\!\left[\frac{\Pij{x}}{\P{x}}\right]
+\Prz{I_T = i} + \Pri{I_T \neq i} \geq \exp\left(-\KL(P_0, P_i)\right)\,,
 }
-\STATE Compute required budget: 
+where $\KL(P_0, P_i)$ is the KL divergence between measures $P_0$ and $P_i$.
+Let $T_i(T) = \sum_{t=1}^T \ind{I_t = i}$ be the total number of times the learner intervenes on variable $i$ (with any intervention). 
+Then for $i \leq m$ a computation shows that 
+\todot{Need to add at minimum some citations here}
 \eq{
-N_{ij} = \max\set{0, \ceil{\frac{1}{\epsilon^2} - \sum_{k \in \calP_i} \sum_{l \in \set{0,1}} \frac{1}{V_{ijkl}} - \frac{1}{V_{ij}}}}\,.
+\KL(P_0, P_i) 
+&\leq 4\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
+%&= \frac{\epsilon^2}{2} \mathbb E_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
+&\leq 4\epsilon^2 \left(\mathbb E_0 T_i(T) + q_i T\right) \\
+&\leq 4\epsilon^2 \left(\mathbb E_0 T_i(T) + \frac{T}{m}\right)\,.
 }
-\ENDFOR
-\IF{$\sum_{i=1}^N \sum_{j \in \set{0,1}} N_{ij} \leq T / 2$}
-\STATE {\bf return true}
-\ELSE
-\STATE {\bf return false}
-\ENDIF
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}
-Let $\delta = 1/(2NT)$ and $\epsilon$ be the smallest value for which \cref{alg:alloc} returns true. 
-Let $N_{ij}$ as computed by \cref{alg:alloc} be the number of times the algorithm selects the action $do(X_i = j)$ 
-and computes estimates using the mixture estimator described in the previous section. Choosing $I_T = \argmax_{ij} \hat \mu_{ij}$ leads
-to
+Define set $A$ by
 \eq{
-\EE[r_T] \leq \epsilon + \frac{1}{T}\,.
+A = \set{i \leq m : \mathbb E_0 T_i(T) \leq 2T / m}\,.
 }
-\end{theorem}
+Then for $i \in A$ and choosing $\epsilon = \sqrt{2m/3T}$ we have
+\eq{
+\KL(P_0, P_i) \leq \frac{3T\epsilon^2}{2m} = 1\,. 
+}
+Now $\sum_{i=1}^m \mathbb E_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
+Therefore
+\eq{
+\sum_{i \in A} \Pri{I_T \neq i} 
+&\geq \sum_{i \in A} \exp\left(-\KL(P_0, P_i)\right) - 1\\
+&\geq \frac{|A|}{e} - 1 
+\geq \frac{m}{2e} - 1\,.
+}
+Therefore there exists an $i \in A$ such that
+\eq{
+\Pri{I_T \neq i} \geq \frac{\frac{m}{2e} - 1}{m}\,.
+}
+Therefore
+\eq{
+R_{T,i} \geq \frac{1}{2} P_i(I_T \neq i) \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{2m}{3T}}
+}
+as required.

### 2016-02-03 causal-bandit/icml-2015/experiments.tex
puts latest figures into experiments

@@
 \end{figure}
+
+\begin{figure}
+\caption{Simple regret vs number of variables, $N$, for $T=250$, for Successive Rejects and, Causal-Best-Arm-Identification on the confounded bandit problem.
+Error bars show standard error from 10000 simulations.}
+\label{fig:simple_vs_N_confounded}
+\centering
+\includegraphics[width=.5\textwidth]{exp_general_simpleregret_vs_N_T250_sims10000_20160131_080208.pdf}
+\end{figure}
+
+\subsection{Experiments on more general graphs}
+We now demonstrate the general simple regret algorithm on some graphs that display some interesting behaviour.
+
+\begin{figure}[h]
+\centering
+\caption{Causal model for the confounded bandits problem.}
+\label{fig:causalStructure_confounded_N}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+\node[main node](1){$X_{2}$};
+\node[main node, right=of 1](2){$X_{3}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){Y};
+\node[main node,above right=of 2](6){$X_1$}; 
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5)
+    (6) edge (1) edge (2) edge (4);
+\end{tikzpicture}
+\end{figure}
+
+\subsubsection{X1-dominated confounded model}
+A model under which the value of the confounding variable $X_1$ plays the central role in determining $P(Y)$.  Let:
+\eq {
+\P{Z = 1} &= q \\
+\P{X_k = 1|Z=j} &=  q_j \;\; \forall k \in \set{1...n}\\
+\P{Y|x_1 ... x_n} &= \bar{\boldsymbol{x}}
+} 
+This leads to rewards:
+\eq {
+\P{Y|do(Z = j)} = q^j(1-q)^{1-j}\\
+\P{Y|do(X_i = j)} = \begin{cases}
+\frac{1}{n} + \frac{n-1}{n}\left(q q_1 + (1-q)q_0 \right) & \text{ if } j = 1 \\
+\frac{n-1}{n}\left(q q_1 + (1-q)q_0 \right) & \text { if } j = 0
+\end{cases}
+}
+
+We can select the probabilities to create a scenario close to the worst case problem such that there is a single arm $do(Z=1)$ that is optimal with expected reward ...

### 2016-02-04 causal-bandit/icml-2015/setup.tex
Tweaks to related work and set up

@@
 This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009} or ``best-arm identification'' problem~\citep{Gabillon2012} and
-is most appropriate when the learner has a fixed learning budget after which its policy will be fixed indefintely.
+is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefintely.
 

### 2016-02-04 causal-bandit/icml-2015/discussion.tex
Redoing discussion

@@
+Both our general causal model algorithm and algorithm for parallel bandits are adaptive reward estimators.
+
 In our algorithm, we have only used the side information provided by the $do()$ action about other actions.
@@
 This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$
+
+% Non-observable variables? 
+% Currently assume all P_a are available (implied by observablity)
+%   - do calculus' identifability
+%   - non-ident variables into infinite variance case.
+
+%
 
@@
 
-\section{Future Open Questions}
+\subsection{Open Questions}
 
-\subsection{General graphs with known structure but latent variables}
+\paragraph{General graphs with known structure but latent variables}
 In this case, some queries will be non-identifiable.
@@
 
-\subsection{Graphs with partially or unknown structure}
+\paragraph{Graphs with partially or unknown structure}
 
 \begin{itemize}
+
+\item Unknown in general case - e.g., like algorithm 1
 
@@
 \end{itemize}
-\section{Conclusion}

### 2016-02-04 causal-bandit/icml-2015/code/causal_bandits.py
experiment 1

@@ -7,13 +7,14 @@ Created on Wed Feb  3 09:51:47 2016
 import numpy as np
 from itertools import product
 from scipy.optimize import minimize
-from math import sqrt
-from math import log
+from math import sqrt,log,ceil
 from numpy.random import binomial
 import matplotlib.pyplot as plt
 from time import time
+from datetime import datetime as dt
 
-
+def now_string():
+    return dt.now().strftime('%Y%m%d_%H%M')
 
 def random_eta(n):
     eta = np.random.random(n)
@@ -25,28 +26,46 @@ def most_unbalanced_q(N,m):
     q[0:m] = 0
     return q
     
+def part_balanced_q(N,m):
+    """ all but m of the variables have probability .5"""
+    q = np.full(N,.5,dtype=float)
+    q[0:m] = 0
+    return q
+    
+    
+def argmax_rand(x):
+    """ return the index of the maximum element in the array, ignoring nans. 
+    If there are multiple max valued elements return 1 at random"""
+    max_val = np.nanmax(x)
+    indicies = np.where(x == max_val)[0]
+    return np.random.choice(indicies) 
 
-def maxV(eta):
-    maxV = model.V(eta).max()
-    if not np.isfinite(maxV):
-        return np.inf
-    return maxV#+penalty # consider penalizing eta going beyond bounds 
+def calculate_m(qij_sorted):
+    for indx,value in enumerate(qij_sorted):
+        if value >= 1.0/(indx+1):
+            return indx
+    return len(qij_sorted)/2      
+    
 
 class Parallel(object):
     def __init__(self,q,epsilon):
         """ actions are do(x_1 = 0)...do(x_N = 0),do(x_1=1)...do(N_1=1), do() """
         assert q[0] <= .5, "P(x_1 = 1) should be <= .5 to ensure worst case reward distribution can be created"
-        self.epsilon = epsilon
-        self.epsilon_minus = self.epsilon*q[0]/(1.0-q[0]) 
+        self.q = q
         self.N = len(q) # number of variables
         self.K = 2*self.N+1 #number of actions
         self.pX = np.vstack((1.0-q,q))
+        self.set_epsilon(epsilon)
         
+    def set_epsilon(self,epsilon):
+        assert epsilon <= .5 ,"epsilon cannot exceed .5"
+        self.epsilon = epsilon
[…]

### 2016-02-04 causal-bandit/icml-2015/discussion.tex
Re-org discussion

@@
-Both our general causal model algorithm and algorithm for parallel bandits are adaptive reward estimators.
+Both our general causal model algorithm and algorithm for parallel bandits adaptively estimate the reward for allowable actions over $T$ rounds.
+For general causal bandit problem with $N$ allowable interventions we have shown that Algorithm~\ref{alg:general} has simple regret that is never worse than $\bigo{\sqrt{N/T}}$.
+
+\paragraph{Cumulative Regret}
+In the special case of parallel bandits, we were able to remove the assumption that the conditional model
+
+Cumulative regret... use \citep{wu2015online} to get optimal cumulative regret bound (should be $\Omega(T^{2/3})$?)
+
+General discussion of how to map parallel bandit problem to graph feedback.
 
@@
 
+\paragraph{Causal Models with Non-Observable Variables}
+In this case, some queries will be non-identifiable.
+The corresponding actions can be immediately added to the set $A$ prior to collecting any data.
+We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
 %
@@
 
-\cite{wu2015online} consider a relevant feedback model.
+\paragraph{Partially or Completely Unknown Causal Graph}
 
-\subsection{Open Questions}
+Unknown in general case - e.g., like algorithm 1
 
-\paragraph{General graphs with known structure but latent variables}
-In this case, some queries will be non-identifiable.
-The corresponding actions can be immediately added to the set $A$ prior to collecting any data.
-We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
-
-\paragraph{Graphs with partially or unknown structure}
-
-\begin{itemize}
-
-\item Unknown in general case - e.g., like algorithm 1
-
-\item Partially known structure
+(Partially known structure)
 There has been substantial work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class \cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014}.
@@
 
-\item Unknown Structure
+(Unknown Structure)
 If we need to learn the structure, in an online environment. 
 Experiment is much, much more revealing than inference from observational data. We would expect it to dominate.
-
-\end{itemize}

### 2016-02-04 causal-bandit/icml-2015/discussion.tex
Tweaks

@@
-Both our general causal model algorithm and algorithm for parallel bandits adaptively estimate the reward for allowable actions over $T$ rounds.
-For general causal bandit problem with $N$ allowable interventions we have shown that Algorithm~\ref{alg:general} has simple regret that is never worse than $\bigo{\sqrt{N/T}}$.
+Our general causal bandit algorithm adaptively estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
+For general causal bandit problem we have shown that Algorithm~\ref{alg:general} has simple regret that is $\bigo{\sqrt{m(\eta)/T}}$ up to log factors and that the parameter $m(\eta)$ is always less than $N$.
+The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$.
+Intuitively, problems where variables' values in the causal model all ``occur naturally'' will have
+
+The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator terms $Z_a$ and the optimal sampling distribution $\eta^*$ require knowledge of the distributions $P_a$ for all $a \in \calA$.
+In contrast, in the special case of parallel bandits, we were able to remove the assumption that the conditional model ...  matching lower bound (up to log factors).
+Finding a version that estimates $\eta$ in general case ... and lower bounds is left as future work.
 
 \paragraph{Cumulative Regret}
-In the special case of parallel bandits, we were able to remove the assumption that the conditional model
-
 Cumulative regret... use \citep{wu2015online} to get optimal cumulative regret bound (should be $\Omega(T^{2/3})$?)

### 2016-02-04 causal-bandit/icml-2015/discussion.tex
Work on discussion

--- 2016-02-04 causal-bandit/icml-2015/discussion.tex
@@
-Both our general causal model algorithm and algorithm for parallel bandits adaptively estimate the reward for allowable actions over $T$ rounds.
-For general causal bandit problem with $N$ allowable interventions we have shown that Algorithm~\ref{alg:general} has simple regret that is never worse than $\bigo{\sqrt{N/T}}$.
+Algorithm~\ref{alg:general} for general causal bandit problems adaptively estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
+Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\bigo{\sqrt{m(\eta)/T}}$ and that the parameter $m(\eta)$ is always less than $N$.
+The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.
 
+The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator terms $Z_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $P_a$ for all $a \in \calA$.
+In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.
+Despite these extra estimates, Theorem~\ref{thm:lower} shows that this approach is optimal (up to log factors).
+Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
+
+\todom{Still cleaning up below}
 \paragraph{Cumulative Regret}
-In the special case of parallel bandits, we were able to remove the assumption that the conditional model
 
@@
 
+% \cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
+In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
+\cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
+
 % Non-observable variables? 
@@
 
-\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
-In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
-\cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
+% More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
 In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.

### 2016-02-04 causal-bandit/icml-2015/tech.tex
updates to lower bound

@@
 
-\todot{Change this to Bernoulli}
 Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
@@
 }
-where $\epsilon > 0$ is some constant to be chosen later.
+where $1/4 \geq \epsilon > 0$ is some constant to be chosen later.
 We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
 environment determined by $\boldsymbol{q}$ and $r_i$. Let $P_i$ be the corresponding measure
-on all observations over all $T$ rounds and $\mathbb E_i$ the expectation with respect to $P_i$. By Lemma 2.6 by \cite{Tsy08} we have
+on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $P_i$. By Lemma 2.6 by \cite{Tsy08} we have
 \eq{
@@
 where $\KL(P_0, P_i)$ is the KL divergence between measures $P_0$ and $P_i$.
-Let $T_i(T) = \sum_{t=1}^T \ind{I_t = i}$ be the total number of times the learner intervenes on variable $i$ (with any intervention). 
-Then for $i \leq m$ a computation shows that 
-\todot{Need to add at minimum some citations here}
+Let $T_i(T) = \sum_{t=1}^T \ind{a_t = do(X_i = 1)}$ be the total number of times the learner intervenes on variable $i$ by setting it to $1$.
+Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $P_0$ and $P_i$ may be bounded using the telescoping property (chain rule) and
+by bounding the local KL divergence by the $\chi$-squared distance as by \cite{Auer1995}. This leads to 
 \eq{
 \KL(P_0, P_i) 
-&\leq 4\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
-%&= \frac{\epsilon^2}{2} \mathbb E_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] \\
-&\leq 4\epsilon^2 \left(\mathbb E_0 T_i(T) + q_i T\right) \\
-&\leq 4\epsilon^2 \left(\mathbb E_0 T_i(T) + \frac{T}{m}\right)\,.
+&\leq 6\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] 
+\leq 6\epsilon^2 \left(\EE_0 T_i(T) + q_i T\right) 
+\leq 6\epsilon^2 \left(\EE_0 T_i(T) + \frac{T}{m}\right)\,.
 }
-Define set $A$ by
+Define set $A$ by $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
+Then for $i \in A$ and choosing $\epsilon = \sqrt{m/(18T)}$ we have
 \eq{
-A = \set{i \leq m : \mathbb E_0 T_i(T) \leq 2T / m}\,.
+\KL(P_0, P_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
 }
-Then for $i \in A$ and choosing $\epsilon = \sqrt{2m/3T}$ we have
-\eq{
-\KL(P_0, P_i) \leq \frac{3T\epsilon^2}{2m} = 1\,. 
-}
-Now $\sum_{i=1}^m \mathbb E_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
+Now $\sum_{i=1}^m \EE_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
 Therefore

### 2016-02-04 causal-bandit/icml-2015/tech.tex
updated lower bound

@@
 \eq{
-\mu_{\hat a^*} \geq \hat \mu_{\hat a^*} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
+\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
 }
@@
 \eq{
-\mu^* - \EE[\mu_{\hat a^*}] 
+\mu^* - \EE[\mu_{\hat a^*_T}] 
 \leq 6\delta + \epsilon 
@@
 We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
-environment determined by $\boldsymbol{q}$ and $r_i$. Let $P_i$ be the corresponding measure
-on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $P_i$. By Lemma 2.6 by \cite{Tsy08} we have
+environment determined by $\boldsymbol{q}$ and $r_i$. Let $\operatorname{P}_i$ be the corresponding measure
+on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \cite{Tsy08} we have
 \eq{
-\Prz{I_T = i} + \Pri{I_T \neq i} \geq \exp\left(-\KL(P_0, P_i)\right)\,,
+\Prz{\hat a^*_T = a^*} + \Pri{\hat a^*_T \neq a^*} \geq \exp\left(-\KL(\operatorname{P}_0, \operatorname{P}_i)\right)\,,
 }
-where $\KL(P_0, P_i)$ is the KL divergence between measures $P_0$ and $P_i$.
+where $\KL(\Ps_0, \Ps_i)$ is the KL divergence between measures $\operatorname{P}_0$ and $\operatorname{P}_i$.
 Let $T_i(T) = \sum_{t=1}^T \ind{a_t = do(X_i = 1)}$ be the total number of times the learner intervenes on variable $i$ by setting it to $1$.
-Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $P_0$ and $P_i$ may be bounded using the telescoping property (chain rule) and
+Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $\Ps_0$ and $\Ps_i$ may be bounded using the telescoping property (chain rule) and
 by bounding the local KL divergence by the $\chi$-squared distance as by \cite{Auer1995}. This leads to 
 \eq{
-\KL(P_0, P_i) 
+\KL(\Ps_0, \Ps_i) 
 &\leq 6\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] 
@@
 }
-Define set $A$ by $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
-Then for $i \in A$ and choosing $\epsilon = \sqrt{m/(18T)}$ we have
+Define set $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
+Then for $i \in A$ and choosing $\epsilon = \min\set{1/4, \sqrt{m/(18T)}}$ we have
 \eq{
-\KL(P_0, P_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
+\KL(\Ps_0, \Ps_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
 }
@@
 \sum_{i \in A} \Pri{I_T \neq i} 
-&\geq \sum_{i \in A} \exp\left(-\KL(P_0, P_i)\right) - 1\\
-&\geq \frac{|A|}{e} - 1 
+\geq \sum_{i \in A} \exp\left(-\KL(\Ps_0, \Ps_i)\right) - 1
+\geq \frac{|A|}{e} - 1 
 \geq \frac{m}{2e} - 1\,.
@@
 Therefore there exists an $i \in A$ such that
+$\Pri{I_T \neq i} \geq \frac{\frac{m}{2e} - 1}{m}$. 
+Therefore if $\epsilon < 1/4$ we have
 \eq{
-\Pri{I_T \neq i} \geq \frac{\frac{m}{2e} - 1}{m}\,.
+R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{m}{18T}}\,.
 }
-Therefore
+Otherwise $m \geq 18T$ so $\sqrt{m/T} = \Omega(1)$ and
 \eq{
-R_{T,i} \geq \frac{1}{2} P_i(I_T \neq i) \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{2m}{3T}}
+R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{1}{4} \frac{\frac{m}{2e} - 1}{2m} \in \Omega(1) 
 }

### 2016-02-04 causal-bandit/icml-2015/discussion.tex
More on discussion

@@
 
-\todom{Still cleaning up below}
 \paragraph{Cumulative Regret}
+Another open question is whether cumulative regret bounds can be obtained for the general causal bandit problem we have proposed.
+In the case of the parallel bandit problem however, we can make use of existing work on bandits with graph feedback to achieve this.
+Each time an action $do(X_i = j) \in \calA$ is chosen
 
@@
 \paragraph{Causal Models with Non-Observable Variables}
-In this case, some queries will be non-identifiable.
+We have assumed through this paper that all variables in the causal model were observable after an intervention is made.
+A natural generalisation of our model would be for some variables to be unobservable.
+In this case, some conditional distributions may be non-identifiable. 
 The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
@@
 \paragraph{Partially or Completely Unknown Causal Graph}
+A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
+The latter case arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques. 
+Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm as part of a pre-processing step.
+Simultaneously learning a completely unknown causal model while estimating the rewards of interventions would be much more challenging.
+\todom{Ideas? Relevant refs?}
 
-Unknown in general case - e.g., like algorithm 1
+% (Partially known structure)
+% Key results $bigo(n)$ singleton or $bigo(log log n)$ multi-variate experiments are required.
+% - Focus on minimizing the number of experiments that must be performed. Examples of active versus online learning. The cost of experiments constant or at least known.
 
-(Partially known structure)
-There has been substantial work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class \cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014}.
-This problem arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques.
-Key results $bigo(n)$ singleton or $bigo(log log n)$ multi-variate experiments are required.
-- Focus on minimizing the number of experiments that must be performed. Examples of active versus online learning. The cost of experiments constant or at least known.
-
-(Unknown Structure)
-If we need to learn the structure, in an online environment. 
-Experiment is much, much more revealing than inference from observational data. We would expect it to dominate.
+% (Unknown Structure)
+% If we need to learn the structure, in an online environment. 
+% Experiment is much, much more revealing than inference from observational data. We would expect it to dominate.

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
Tweak

@@
 \paragraph{Cumulative Regret}
-Another open question is whether cumulative regret bounds can be obtained for the general causal bandit problem we have proposed.
-In the case of the parallel bandit problem however, we can make use of existing work on bandits with graph feedback to achieve this.
-Each time an action $do(X_i = j) \in \calA$ is chosen
+Although we have focused on simple regret in our analysis, a natural question is whether cumulative regret bounds can be obtained for the general causal bandit problem we have proposed.
+% In the case of the parallel bandit problem however, we can make use of existing work on bandits with graph feedback to achieve this.
 
+\todom{Not sure about reduction here}
 Cumulative regret... use \citep{wu2015online} to get optimal cumulative regret bound (should be $\Omega(T^{2/3})$?)
 
-General discussion of how to map parallel bandit problem to graph feedback.
+General discussion of how to map parallel bandit problem to graph feedback?
 
-In our algorithm, we have only used the side information provided by the $do()$ action about other actions.
+% In our algorithm, we have only used the side information provided by the $do()$ action about other actions.
 Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
@@
 A natural generalisation of our model would be for some variables to be unobservable.
-In this case, some conditional distributions may be non-identifiable. 
-The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
-We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
-%
+\todom{More to say here?}
+If the conditional models are known then the $Z_a$ terms in Algorithm~\ref{alg:general} could still be obtained via marginalising.
+However, if this is not the case the estimation problem becomes more difficult.
+% In this case, some conditional distributions may be non-identifiable. 
+% The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
+% We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
+Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
 

### 2016-02-05 causal-bandit/icml-2015/code/causal_bandits.py
runs r vs T fixed epsilon

@@ -287,21 +287,27 @@ class SuccessiveRejects(object):
     label = "Successive Reject"
     
     def run(self,T,model):
-        if T <= model.K:
-            return np.nan
-        allocations = self.allocate(T,model.K)
         self.trials = np.zeros(model.K)
         self.success = np.zeros(model.K)
         self.actions = range(0,model.K)
-        self.rejected = np.zeros((model.K),dtype=bool)
-        for k in range(0,model.K-1):
-            nk = allocations[k]
-            self.success[self.actions] += model.sample_multiple(self.actions,nk)
-            self.trials[self.actions] += nk
-            self.reject()
-        assert len(self.actions == 1), "number of arms remaining is: {0}, not 1.".format(len(self.actions))
-        assert sum(self.trials <= T),"number of pulls = {0}, exceeds T = {1}".format(sum(self.trials),T)
-        best_action = self.actions[0]
+        if T <= model.K:
+            to_sample = np.random.choice(self.actions,size=T,replace=False)
+            self.trials[to_sample] +=1
+            self.success[to_sample] += model.sample_multiple(to_sample,1)
+            u = np.true_divide(self.success,self.trials)
+            best_action = argmax_rand(u)
+            #return np.nan
+        else:
+            allocations = self.allocate(T,model.K)
+            self.rejected = np.zeros((model.K),dtype=bool)
+            for k in range(0,model.K-1):
+                nk = allocations[k]
+                self.success[self.actions] += model.sample_multiple(self.actions,nk)
+                self.trials[self.actions] += nk
+                self.reject()
+            assert len(self.actions == 1), "number of arms remaining is: {0}, not 1.".format(len(self.actions))
+            assert sum(self.trials <= T),"number of pulls = {0}, exceeds T = {1}".format(sum(self.trials),T)
+            best_action = self.actions[0]
         return model.optimal - model.expected_rewards[best_action]
     
     def allocate(self,T,K):
@@ -324,36 +330,48 @@ class SuccessiveRejects(object):
[…]

### 2016-02-05 causal-bandit/icml-2015/introduction.tex

@@
 There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-\citet{Alon2015} consider a very general model where rewards on unplayed actions are revealed according to a feedback graph.
+\citet{Alon2015}, \citet{Kocak2014} and \citet{wu2015online} consider very general models where rewards on unplayed actions are revealed according to a feedback graph.
 As we discuss in \S\ref{sec:discussion}, the problem they consider shares some similarity with ours however their results are for cumulative regret and thus cannot be used to guarantee low simple regret~\citep{Bubeck2009}.
-Even in the case of cumulative regret we show that an application of their results to the parallel bandit problem with $N$ variables yields a $\bigo{\sqrt{NT}}$ bound.
+% Even in the case of cumulative regret we show that an application of their results to the parallel bandit problem with $N$ variables yields a $\bigo{\sqrt{NT}}$ bound.
 \citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.

### 2016-02-05 causal-bandit/icml-2015/introduction.tex
Fix to related work

@@
 There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-\citet{Alon2015}, \citet{Kocak2014} and \citet{wu2015online} consider very general models where rewards on unplayed actions are revealed according to a feedback graph.
+Most recently, \citet{Alon2015}, \citet{Kocak2014} and \citet{wu2015online} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
 As we discuss in \S\ref{sec:discussion}, the problem they consider shares some similarity with ours however their results are for cumulative regret and thus cannot be used to guarantee low simple regret~\citep{Bubeck2009}.
@@
 \citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled. 
-The main difference from our present work is our focus on simple regret and exploitation of the more complex information related rewarding via the causal graph.
+The main difference from our present work is our focus on simple regret and exploitation of the more complex information related rewards via the causal graph.
 To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
commented out some experiments
udates
updates to experiments

--- 2016-01-13 causal-bandit/icml-2015/experiments.tex
@@
-In all experiments, the reward is generated according to:
+\subsection{Parallel bandits problem}
+
+For the parallel bandit problem, we use a model in which, unknown to the algorithms, $Y$ depends only on a single variable.
 \eq{
 Y_t \sim \begin{cases}
-\bernoulli(.5+\epsilon) & \text{if } X_1 = 1 \\
-\bernoulli(\frac{.5-(.5+\epsilon)q_1}{1-q_1}) & \text{otherwise}\,.
+\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
+\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
 \end{cases}
@@
 
-With $\epsilon \in (0,.5)$. This ensures that 
-\eq {
-\P{Y|do(X_i = j)} = \begin{cases}
-.5+\epsilon & \text{if } (i,j)=(1,1) \\
-\frac{.5-(.5+\epsilon)q_1}{1-q_1} & \text{if } (i,j) = (1,0) \\
-.5 & \text{otherwise}
-\end{cases}
-}
+This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\left(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon\right)$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
+We set $\boldsymbol{q} = 0$ for $q_i \leq m$ and $\frac{1}{2}$ otherwise.
+Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
 
-This allows us to rapidly simulate rewards and conditional rewards for large numbers of variables and to change the value of $\boldsymbol{q}$ with minimal impact on the reward function \footnote{$\boldsymbol{q}$ only effects the reward of a single, non-optimal arm, which becomes insignificant with large numbers of variables.}.
+We compare the performance of the Algorithm 1, which is specific to the parallel problem, but does not require knowledge of $\boldsymbol{q}$, with that of Algorithm 2 and the Successive Reject algorithm of \cite{}.
+For each experiment, the results are over 10,000 simulation and error bars show three standard errors.
+
+In figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
+The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure.
+For the causal algorithms it increases approximately with $\sqrt{m}$.
+As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not take into account the observed rewards to focus sampling effort on actions with likely high pay-offs.
+
+\begin{figure}[h]
+\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=300$ and number of variables $N = 50$}
+\label{fig:simple_vs_m}
+\centering
+\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_m_T300_N50_sims10000_10000.pdf}
+\end{figure}
+
+If figures \ref{fig:simple_vs_T_vary_epsilon} and \ref{fig:simple_vs_T} we compare the performance as a function of $T$. \todof{finish this}
+
+For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
 
 \begin{figure}
-\caption{Final regret versus number of variables $N$ for UCB with $\alpha = 2$, Causal-Explore-Exploit with $m=2$ and with $m=N$ and horizon $T = 10,000$ .
-Error bars show standard deviation over 100 simulations.
-The regret for UCB grows linearly with the number of variables, whist for Causal-Explore-Exploit with fixed $m$, the growth is sub-logarthmic.
-}
-\label{fig:known_q_r_vs_N}
+\caption{Simple regret vs horizon, $T$, with $N = 50$ and $\epsilon = \sqrt{K/4T}$}
+\label{fig:simple_vs_T_vary_epsilon}
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_N_T10000_sims100_20151229_113550.pdf}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
 \end{figure}
@@
 \begin{figure}
-\caption{Cumulative regret over time for $N = 17$ for UCB with $\alpha=2$, Causal-Explore-Exploit with $m=2$ and Causal-Explore-Exploit with $m=N$.
-Shaded region shows standard deviation over 100 simulations.
-The Causal-Explore-Exploit algorithm incurs linear regret during the exploration phase, after which it selects the optimal arm with high probability.
-For $m=2$, we have $K \sim m^{2/3}T^{1/3}$ and see that we are in the regime in which Causal-Explore-Exploit outperforms UCB.}
-\label{fig:known_q_r_vs_t}
+\caption{Simple regret vs horizon, $T$, with $N = 50$ and fixed $\epsilon = .3$}
+\label{fig:simple_vs_T}
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_t_T10000_N17_sims100_20151229_120647.pdf}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s1000_20160205_1155}
 \end{figure}
 
-\begin{figure}
-\caption{Simple regret vs horizon, $T$, for $N = 30$ for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$.
-Error bars show estimated 95\% confidence interval from 1000 simulations.}
-\label{fig:simple_vs_T}
+\subsection{Highly confounded problem}
+
+\begin{figure}[h]
 \centering
-\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_T_N30_sims1000_20160113_010512.pdf}
+\caption{Causal model for the highly confounded problem.}
+\label{fig:causalStructure_highly_confounded}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
+\node[main node](1){$V_{2}$};
+\node[main node, right=of 1](2){$X_{3}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){Y};
+\node[main node,above right=of 2](6){$X_1$}; 
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5)
+    (6) edge (1) edge (2) edge (4);
+\end{tikzpicture}
 \end{figure}
 
-\begin{figure}
-\caption{Simple regret vs number of variables, $N$, for $T=500$, for UCB with $\alpha=2$, Causal-Best-Arm-Identification with $m=2$ and Causal-Best-Arm-Identification with $m=N$.
-Error bars show estimated 95\% confidence interval from 1000 simulations.}
-\label{fig:simple_vs_N}
-\centering
-\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_N_T500_sims1000_20160113_010511.pdf}
-\end{figure}
+A model under which the value of the confounding variable $X_1$ plays the central role in determining $P(Y)$.  Let:
+\eq {
+\P{Z = 1} &= q \\
+\P{X_k = 1|Z=j} &=  q_j \;\; \forall k \in \set{1...n}\\
+\P{Y|x_1 ... x_n} &= \bar{\boldsymbol{x}}
+} 
+This leads to rewards:
+\eq {
+\P{Y|do(Z = j)} = q^j(1-q)^{1-j}\\
+\P{Y|do(X_i = j)} = \begin{cases}
+\frac{1}{n} + \frac{n-1}{n}\left(q q_1 + (1-q)q_0 \right) & \text{ if } j = 1 \\
+\frac{n-1}{n}\left(q q_1 + (1-q)q_0 \right) & \text { if } j = 0
+\end{cases}
+}

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
Tweak

@@
 
+\paragraph{Using the Reward Signal}
+
 \paragraph{Cumulative Regret}
-Although we have focused on simple regret in our analysis, a natural question is whether cumulative regret bounds can be obtained for the general causal bandit problem we have proposed.
-% In the case of the parallel bandit problem however, we can make use of existing work on bandits with graph feedback to achieve this.
+Although we have focused on simple regret in our analysis, a natural but open question is whether cumulative regret bounds can be obtained for the general causal bandit problem we have proposed.
+In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with graph feedback to get optimal cumulative regret guarantees.
+Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
+We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
+Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
 
-\todom{Not sure about reduction here}
-Cumulative regret... use \citep{wu2015online} to get optimal cumulative regret bound (should be $\Omega(T^{2/3})$?)
-
-General discussion of how to map parallel bandit problem to graph feedback?
+Exploiting the variance structure in this way seems necessary even in the parallel bandit case.
+If instead we choose to use other graph feedback models for cumulative regret we would require those with time varying graphs~\citep{Alon2015,Kocak2014}.
+Results for these models give a cumulative regret of $\bigo{\sqrt{\bar{\alpha} T}}$ where $\bar{\alpha}$ is the average independence number of the time varying graphs.
+On each round, the feedback graph over the $2N$ possible interventions over $N$ binary variables would connect the chosen intervention to the corresponding intervention for each of the other $N-1$ observed variables.
+Unfortunately, in this case the resulting star graphs have an $\bar{\alpha} = N$ and so the cumulative regret bounds would improve over those obtained by ignoring the causal graph structure.
 
@@
 Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm as part of a pre-processing step.
+\todom{Ideas? Relevant refs?}
 Simultaneously learning a completely unknown causal model while estimating the rewards of interventions would be much more challenging.
-\todom{Ideas? Relevant refs?}
 

### 2016-02-05 causal-bandit/icml-2015/introduction.tex

@@
 
-This paper takes the first step towards formalising and solving problems such as the one above. 
+We take the first step towards formalising and solving problems such as the one above. 
 In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph.
@@
 
-We focus on the simple regret, which measures the difference between the return of the optimal action and that of the action chosen by the algorithm after $T$ rounds.
+Our focus is on the simple regret, which measures the difference between the return of the optimal action and that of the action chosen by the algorithm after $T$ rounds.
 In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently and there are $2N$ possible interventions.
@@
 In the general casual bandit problem interventions and observations may have a complex relationship. 
-In \S\ref{sec:simple-regret-general} we propose a novel algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
+In \S\ref{sec:simple-regret-general} we propose a new algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
 As in the parallel bandit case, the regret guarantee is often significantly better than $O(\sqrt{N/T})$ and never worse, where $N$ is the number of available interventions.
@@
 As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
-In the farming example, the farmer could treat the six possible interventions are arms, ignore the observations of the non-intervened variables, and apply an existing best-arm identificaion algorithm~\citep{Jamieson2013} with well understood simple regret guarantees.
+In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables, and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
 However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.
@@
 Two pieces of recent work also consider using causal models in bandit problems.
-\citet{Bareinboim2015} also use causal inference to improve the chance of choosing an optimal arm.
+\citet{Bareinboim2015} use causal inference to improve the chance of choosing an optimal arm.
 Their work differs from ours in two key ways.

### 2016-02-05 causal-bandit/icml-2015/setup.tex
Tweaks

@@
 An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from the parents of $X_i$ to $X_i$ for each $X_i \in \vec{X}$. 
+The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from $\parents{i}$ to $X_i$ for each $X_i \in \vec{X}$. 
 The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
@@
 the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$. 
-
-The game proceeds over $T$ rounds.
+The causal bandit game proceeds over $T$ rounds.
 In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations. 
@@
 including the \emph{reward} $Y_t \in \{0,1\}$. 
-After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T$.
+After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T \in \actions$ based on its prior observations.
 

### 2016-02-05 causal-bandit/icml-2015/code/causal_bandits.py
runs vs m experiment

@@ -16,6 +16,7 @@ from datetime import datetime as dt
 
 REGRET_LABEL = "Regret"
 HORIZON_LABEL = "T"
+M_LABEL = "m(q)"
 
 def now_string():
     return dt.now().strftime('%Y%m%d_%H%M')
@@ -75,7 +76,7 @@ class VeryConfounded(object):
         
         self.pW0givenA = np.vstack((1-self.pW0givenA,self.pW0givenA))
         self.pW1givenA = np.vstack((1-self.pW1givenA,self.pW1givenA))
-        self.wCombinations = np.asarray([(0, 0), (0, 1), (1, 0), (1, 1)])
+        self.parent_vals = np.asarray([(0, 0), (0, 1), (1, 0), (1, 1)])
         self.expected_rewards = self.estimate_rewards(100000)
         self.optimal = np.max(self.expected_rewards)
         
@@ -99,7 +100,7 @@ class VeryConfounded(object):
         
     def V(self,eta):
         va = np.zeros(self.K)  
-        for x in self.wCombinations:
+        for x in self.parent_vals:
             pa = self.P(x)
             Q = (eta*pa).sum()
             ratio = np.true_divide(pa**2,Q)
@@ -148,6 +149,11 @@ class VeryConfounded(object):
         
 
 
+        
+        
+        
+    
+
 class Parallel(object):
     def __init__(self,q,epsilon):
         """ actions are do(x_1 = 0)...do(x_N = 0),do(x_1=1)...do(N_1=1), do() """
@@ -157,6 +163,8 @@ class Parallel(object):
         self.K = 2*self.N+1 #number of actions
         self.pX = np.vstack((1.0-q,q))
         self.set_epsilon(epsilon)
+        self.parent_vals = None
+
         
     def set_epsilon(self,epsilon):
         assert epsilon <= .5 ,"epsilon cannot exceed .5"
@@ -196,7 +204,8 @@ class Parallel(object):
         pij[1-x,indx] = 0
         pij = pij.reshape((len(x)*2,)) #flatten first N-1 will be px=0,2nd px=1
         pobserve = joint# the probability of x given do()
-        return np.hstack((pij,pobserve))
+        result = np.hstack((pij,pobserve))
+        return result
         
     def R(self,x,eta):
         pa = self.P(x)
@@ -206,9 +215,10 @@ class Parallel(object):
         return ratio
         
[…]

### 2016-02-05 causal-bandit/icml-2015/discussion.tex

@@
 
-\paragraph{Using the Reward Signal}
+\paragraph{Making Better Use of the Reward Signal}
+Several existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
+In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
+Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
+This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
 
@@
 We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
+\todom{Check this!}
 Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
@@
 Exploiting the variance structure in this way seems necessary even in the parallel bandit case.
-If instead we choose to use other graph feedback models for cumulative regret we would require those with time varying graphs~\citep{Alon2015,Kocak2014}.
+If instead we choose to use other graph feedback models we would require those with time varying graphs~\citep{Alon2015,Kocak2014}.
 Results for these models give a cumulative regret of $\bigo{\sqrt{\bar{\alpha} T}}$ where $\bar{\alpha}$ is the average independence number of the time varying graphs.
@@
 
-% In our algorithm, we have only used the side information provided by the $do()$ action about other actions.
-Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
-The feedback graph is distinct from the causal graph.
-A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
-For this specific problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
-The Exp3-IX algorithm \cite{Kocak2014} was developed for the adversarial version of this problem and has regret $\bigo{\sqrt{\bar{\alpha}T}}$, where $\bar{\alpha}$ is the average independence number of $G_t$.
-In our case $\bar{\alpha} = \frac{N}{2}$ so we again obtain the regret of the standard bandit algorithm.
-The issue here is that a malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
-This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$
-
 % \cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
@@
 \cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-% Non-observable variables? 
-% Currently assume all P_a are available (implied by observablity)
-%   - do calculus' identifability
-%   - non-ident variables into infinite variance case.
 
@@
 Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm as part of a pre-processing step.
-\todom{Ideas? Relevant refs?}
 Simultaneously learning a completely unknown causal model while estimating the rewards of interventions would be much more challenging.
+\citet{Hu2014b} show that ...
 

### 2016-02-05 causal-bandit/icml-2015/general.tex
Tweaks

@@
 where $B_a \geq 0$ is some constant to be chosen subsequently and
+\todom{Symbol clash with regret}
 \eq{
-R_a(X) = \frac{\P{\parents{Y}(X)|a}}{\Q{\parents{Y}(X)}}\,.
+R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\,.
 }
@@
 }
-We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately cooincides 
-with the version given in the previous section, which justifies the name overloading.
+We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
 
+\todom{Change to $[0,1]^{\calA}$}
 \begin{algorithm}[H]

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
adds regret vs m general

@@
 \centering
-\includegraphics[width=.5\textwidth]{exp_simpleregret_vs_m_T300_N50_sims10000_10000.pdf}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s100_20160205_1637.pdf}
 \end{figure}
@@
 
-For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
+For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more 
+general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data. 
+This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator
+is known to be minimax optimal asymptotically \citep{LMS14}.
 
 \begin{figure}
-\caption{Simple regret vs horizon, $T$, with $N = 50$ and $\epsilon = \sqrt{K/4T}$}
+\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{K/4T}$}
 \label{fig:simple_vs_T_vary_epsilon}
@@
 \begin{figure}
-\caption{Simple regret vs horizon, $T$, with $N = 50$ and fixed $\epsilon = .3$}
+\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
 \label{fig:simple_vs_T}
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s1000_20160205_1155}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1532}
 \end{figure}

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
updates

@@
 \begin{figure}[h]
-\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=300$ and number of variables $N = 50$}
+\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
 \label{fig:simple_vs_m}
@@
 
-If figures \ref{fig:simple_vs_T_vary_epsilon} and \ref{fig:simple_vs_T} we compare the performance as a function of $T$. \todof{finish this}
+\begin{figure}[h]
+\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$, without truncation}
+\label{fig:simple_vs_m2}
+\centering
+\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s100_20160205_1701.pdf}
+\end{figure}
+
+If figures \ref{fig:simple_vs_T_vary_epsilon} and \ref{fig:simple_vs_T} we compare the performance as a function of $T$.
 

### 2016-02-05 causal-bandit/icml-2015/code/causal_bandits.py
update

@@ -284,7 +284,7 @@ class GeneralCausal(object):
             a = np.random.choice(actions,p=eta)
             x,y = model.sample(a) #x is an array containing values for each variable
             r = model.R(x,eta)
-            z = (r <= self.B)*r*y
+            z = r*y #(r <= self.B)*r*y #Truncation turned off to run parallel bandit experiments
             u += z
         self.u = u/float(T)
         best_action = np.argmax(u)
@@ -508,7 +508,7 @@ def experiment2():
 
 N = 50
 epsilon = .3
-simulations = 100
+simulations = 10000
 T = 400
 regret,mean,error = regret_vs_m(N,epsilon,simulations,T)
 

### 2016-02-05 causal-bandit/icml-2015/general.tex
Notation tweaks. Expand explanation after Prop. 4.

@@
 
-\todom{Change to $[0,1]^{\calA}$}
 \begin{algorithm}[H]
@@
 \begin{algorithmic}
-\STATE {\bf Input:} $T$, $\eta \in [0,1]^{|\calA|}$, $B \in [0,\infty)^{|\calA|}$
+\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
 \FOR{$t \in \set{1,\ldots,T}$}
@@
 \begin{theorem}\label{thm:general}
-If Algorithm \ref{alg:general} is run with $B \in \R^{|\calA|}$ given by
+If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by
 \eq{
@@
 
-\begin{proposition}
+\begin{proposition}\label{pro:m-bound}
 $m(\eta^*) \leq |\calA|$.
@@
 For $a = do()$ choose $\eta_a = 1/2$. 
-Let $m = m(\vec{q})$ and for $a = do(X_i = j)$ let $\eta_a = \ind{\P{X_i = j} \leq 1/m} / (2m)$. 
-Then it is easy to check that $m(\eta) \leq 2m$.
+\todom{Check please!}
+Let $m = m(\vec{q})$ and note that there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
+Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
+It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.
 

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
Note on unknown graph

@@
 % We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
-Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
+% Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
 
@@
 The latter case arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm as part of a pre-processing step.
-Simultaneously learning a completely unknown causal model while estimating the rewards of interventions would be much more challenging.
-\citet{Hu2014b} show that ...
+Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014b} could potentially be incorporated into a causal bandit algorithm.
+In particular, \citet{Hu2014b} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
+Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational dataset would be much more challenging.
 

### 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
minor updates
updates
minor fixes
updates to Theorem 1
updated lower bound
Re-expressed m(q) and added some explanation.

@@
 The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-The fact that $X_1,\ldots,X_N$ are independent can be used to create good estimators of the returns of those
-actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. These are the interventions that are essentially observed
-from observational data. Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
+The fact that $X_1,\ldots,X_N$ are independent can be used to create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. 
+These are the interventions that are easily estimated ``passively'' (\ie, without playing them explicitly). 
+Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
 the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
 the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
-carefully chosen to minimise the worst-case simple regret. The difficulty of the problem depends on $\vec{q}$. 
-Define $m \in \set{1, \ldots, N}$ by
-\eq{
-m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, (1 - q_k)}} \leq 1/i}\,.
+carefully chosen to minimise the worst-case simple regret. 
+The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occuring value (\ie, small $q_i$ or $1-q_i$). 
+Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
+Define
+\eqn{
+\label{eq:m-simple}
+m(\vec{q}) = \max \set{ d : |I_{\frac{1}{d}}| \leq d}\,.
+% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
 }
+This captures a trade off between identifying the low probability variables and not having too many of them.
+When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.
 
@@
 \STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
-\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$ 
-\STATE $\hat q_i = \hat{q}_{do(X_i = 1)}$
+\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
 \ENDFOR
-\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq 1/\hat m}$.
+\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A$ = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
 \STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
@@
 \begin{theorem}\label{thm:lower}
-Suppose $\boldsymbol{q}$ satisfies $m(\boldsymbol{q}) = m$.
-Then for all strategies there exists a reward function such that
+Let $\vec{q}$ be fixed and satisfies $m = m(\boldsymbol{q})$. Then for all strategies there exists a reward function such that
 \eq{
-\simpleregret \geq \frac{\frac{m}{2e} - 1}{m} \sqrt{\frac{2m}{3T}} = \Omega\left(\sqrt{\frac{m}{T}}\right)\,.
+\simpleregret 
+%\geq \frac{\frac{m}{2e} - 1}{m} \sqrt{\frac{2m}{3T}} 
+\in \Omega\left(\sqrt{\frac{m}{T}}\right)\,.
 }

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
Add FIXME for non-observable discussion

@@
 A natural generalisation of our model would be for some variables to be unobservable.
-\todom{More to say here?}
+\todom{FIX ME!}
 If the conditional models are known then the $Z_a$ terms in Algorithm~\ref{alg:general} could still be obtained via marginalising.

### 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
Added explanation of Algorithm 1

@@
 
+To identify a high-reward intervention, Algorithm~\ref{alg:simple} break $T$ rounds into halves. 
+In the first half, the non-interventional action $do()$ is played and the rewards observed.
+These are used to estimate the occurrance probabilities $\vec{q}$ and the value of $m(\vec{q})$.
+The remainder of the rounds are then evenly split to estimate the infrequently observed interventions, as determined by the estimate for $m(\vec{q})$.
+Although very simple, the following two theorems show that this algorithm is effectively optimal.
+
 \begin{theorem}\label{thm:uq-simple}

### 2016-02-05 causal-bandit/icml-2015/introduction.tex
Tweak wording
Fixes to bibliography in library2

@@
 simple regret guarantees of $\tilde\Theta(\sqrt{m/T})$ where $m$ depends on the causal model and may be much smaller than $N$.
+\todom{Need ref or proof?}
 In contrast, existing best arm identification algorithms suffer $\Omega(\sqrt{N/T})$ simple regret.
@@
 Most recently, \citet{Alon2015}, \citet{Kocak2014} and \citet{wu2015online} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
-As we discuss in \S\ref{sec:discussion}, the problem they consider shares some similarity with ours however their results are for cumulative regret and thus cannot be used to guarantee low simple regret~\citep{Bubeck2009}.
+As we discuss in \S\ref{sec:discussion}, the problem they consider shares some similarity with ours however their results are for cumulative regret and thus cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
 % Even in the case of cumulative regret we show that an application of their results to the parallel bandit problem with $N$ variables yields a $\bigo{\sqrt{NT}}$ bound.

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
updates experiments section

--- 2016-02-05 causal-bandit/icml-2015/experiments.tex
@@
 We compare the performance of the Algorithm 1, which is specific to the parallel problem, but does not require knowledge of $\boldsymbol{q}$, with that of Algorithm 2 and the Successive Reject algorithm of \cite{}.
-For each experiment, the results are over 10,000 simulation and error bars show three standard errors.
+For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
 
@@
 For the causal algorithms it increases approximately with $\sqrt{m}$.
-As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not take into account the observed rewards to focus sampling effort on actions with likely high pay-offs.
+As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not take into account the observed rewards to focus sampling effort on actions with high pay-offs.
 
 \begin{figure}[h]
-\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=300$ and number of variables $N = 50$}
+\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
 \label{fig:simple_vs_m}
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s100_20160205_1637.pdf}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s1000_20160205_0910.pdf}
 \end{figure}
 
-If figures \ref{fig:simple_vs_T_vary_epsilon} and \ref{fig:simple_vs_T} we compare the performance as a function of $T$. \todof{finish this}
+Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms,$\epsilon \sim \sqrt{\frac{N}{8T}}$ , is just too small to be learned.
+This gap is learn-able by the causal algorithms, for which the worst case depends on the $m \\l N$.
+In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
+The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.
 
-For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more 
-general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data. 
-This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator
-is known to be minimax optimal asymptotically \citep{LMS14}.
+For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
+This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be minimax optimal asymptotically \citep{LMS14}.
 
 \begin{figure}
-\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{K/4T}$}
+\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
 \label{fig:simple_vs_T_vary_epsilon}
@@
 \end{figure}
-
-\subsection{Highly confounded problem}
-
-\begin{figure}[h]
-\centering
-\caption{Causal model for the highly confounded problem.}
-\label{fig:causalStructure_highly_confounded}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-\node[main node](1){$V_{2}$};
-\node[main node, right=of 1](2){$X_{3}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
-\node[main node,above right=of 2](6){$X_1$}; 
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5)
-    (6) edge (1) edge (2) edge (4);
-\end{tikzpicture}
-\end{figure}
-
-A model under which the value of the confounding variable $X_1$ plays the central role in determining $P(Y)$.  Let:
-\eq {
-\P{Z = 1} &= q \\
-\P{X_k = 1|Z=j} &=  q_j \;\; \forall k \in \set{1...n}\\
-\P{Y|x_1 ... x_n} &= \bar{\boldsymbol{x}}
-} 
-This leads to rewards:
-\eq {
-\P{Y|do(Z = j)} = q^j(1-q)^{1-j}\\
-\P{Y|do(X_i = j)} = \begin{cases}
-\frac{1}{n} + \frac{n-1}{n}\left(q q_1 + (1-q)q_0 \right) & \text{ if } j = 1 \\
-\frac{n-1}{n}\left(q q_1 + (1-q)q_0 \right) & \text { if } j = 0
-\end{cases}
-}

### 2016-02-05 causal-bandit/icml-2015/general.tex
Tweak wording
Fixed typo

@@
 
-Another difficulty is that it is no longer sufficient to learn about $\mu_{ij}$ exclusivly from observational data and when $do(X_i = j)$ is chosen.
+Another difficulty is that it is no longer sufficient to learn about $\mu_{a}$ exclusively from observational data and when $a = do(X_i = j)$ is chosen.
 We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
@@
 \todom{Check please!}
-Let $m = m(\vec{q})$ and note that there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
+Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
 Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.

### 2016-02-05 causal-bandit/icml-2015/experiments.tex

@@
 For the causal algorithms it increases approximately with $\sqrt{m}$.
-As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not take into account the observed rewards to focus sampling effort on actions with high pay-offs.
+As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.
 
@@
 
-Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms,$\epsilon \sim \sqrt{\frac{N}{8T}}$ , is just too small to be learned.
-This gap is learn-able by the causal algorithms, for which the worst case depends on the $m \\l N$.
+Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\epsilon = \sqrt{\frac{N}{8T}}$ , is just too small to be learned.
+This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m << N$.
 In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
puts captions below figures
combines references to single bib file

@@
 
-We compare the performance of the Algorithm 1, which is specific to the parallel problem, but does not require knowledge of $\boldsymbol{q}$, with that of Algorithm 2 and the Successive Reject algorithm of \cite{}.
+We compare the performance of the Algorithm 1, which is specific to the parallel problem, but does not require knowledge of $\boldsymbol{q}$, with that of Algorithm 2 and the Successive Reject algorithm of \cite{audibert2010best}.
 For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
@@
 \begin{figure}[h]
+\centering
+\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s1000_20160205_0910.pdf}
 \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
 \label{fig:simple_vs_m}
-\centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s1000_20160205_0910.pdf}
 \end{figure}
@@
 \begin{figure}
+\centering
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
 \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
 \label{fig:simple_vs_T_vary_epsilon}
-\centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
 \end{figure}
@@
 \begin{figure}
+\centering
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1532}
 \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
 \label{fig:simple_vs_T}
-\centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1532}
 \end{figure}

### 2016-02-05 causal-bandit/icml-2015/code/causal_bandits.py

@@ -376,18 +376,19 @@ def sample_pa(model,p_indx,sims):
     
 def regret_vs_T(N,simulations,epsilon=None,Tstep=None,TperK=10):
     
-    a = 4.0
+    a = 9.0
     q = part_balanced_q(N,2) 
     if epsilon:
         model = Parallel(q,epsilon)
         Tmin = 10
+        vary_epsilon = False
     else:
         model =  Parallel(q,.5)
         Tmin = int(ceil(4*model.K/a))
+        vary_epsilon = True
     
     Tstep = model.K if not Tstep else Tstep
     eta,mq = model.analytic_eta()
-    print mq
      
     T_vals = range(Tmin,TperK*model.K,Tstep)
     
@@ -396,17 +397,18 @@ def regret_vs_T(N,simulations,epsilon=None,Tstep=None,TperK=10):
     baseline  = SuccessiveRejects()
     
     ts = time()   
-    regret = np.zeros((len(T_vals),3,simulations))
+    regret = np.zeros((len(T_vals),4,simulations))
     for s in xrange(simulations):
         if s % 100 == 0:
                 print s
         for T_indx,T in enumerate(T_vals): 
-            if not epsilon: #variable epsilon
+            if vary_epsilon: #variable epsilon
                 epsilon = sqrt(model.K/(a*T))
                 model.set_epsilon(epsilon)
             regret[T_indx,0,s] = causal.run(T,model,eta,mq)
             regret[T_indx,1,s] = causal_parallel.run(T,model)
             regret[T_indx,2,s] = baseline.run(T,model)
+            regret[T_indx,3,s] = epsilon
             
     te = time()
     print 'took: %2.4f sec' % (te-ts)
@@ -418,6 +420,7 @@ def regret_vs_T(N,simulations,epsilon=None,Tstep=None,TperK=10):
     ax.errorbar(T_vals,mean[:,0],yerr=error[:,0], label="Algorithm 2",linestyle="",marker="s",markersize=4) 
     ax.errorbar(T_vals,mean[:,1],yerr=error[:,1], label="Algorithm 1",linestyle="",marker="o",markersize=5)    
     ax.errorbar(T_vals,mean[:,2],yerr=error[:,2], label="Successive Rejects",linestyle="",marker="D",markersize=4) 
[…]

### 2016-02-05 causal-bandit/icml-2015/general.tex
puts captions below figures
adds truncated product formula

@@
 We now consider the more general problem where the graph structure is known, but arbitrary.
-The problem of learning the optimal intervention is made more complicated than the simple case because of 
-the dependencies between the random variables. Learning from observational data is no longer straightforward
-because variables can be correlated through their parents. Consider, for example, the causual graph in Figure \ref{fig:causalStructure_confounded} where $X_1$
-deterministically causes $X_2$ and $X_3$, which ensures that $X_2 = X_3$ for all observational data. Now suppose $r(X) = \ind{X_2 \neq X_3}$,
-then intervening on either $X_2$ or $X_3$ is optimal. But the positive reward is never observed from observational samples, despite
-the fact that $\P{X_2 = 1} = \P{X_3 = 1} = \frac{1}{2}$.
+For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
+However if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}.
+\eq{
+\P{X_1...X_N|do(X_i=j)} = 
+\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)  
+} 
+Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
+For example, given the causual graph in Figure \ref{fig:causalStructure_confounded}, by applying the truncated factorization and marginalizing, we obtain $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1}$.
+Note that the quality of any estimator we build for $P{Y|do(X_2= j)}$ is not just a function of the number of times we observe $X_2 = j$.
+
 \begin{figure}[h]
@@
 \end{tikzpicture}
-\eq{
-\P{X_1 = 1} &= \frac{1}{2} \\
-\P{X_2 = j|X_1 = j} &= 
-\P{X_3 = j|X_1 = j} = 1\,.
-}
 \caption{}\label{fig:causalStructure_confounded}
@@
 
-Another difficulty is that it is no longer sufficient to learn about $\mu_{a}$ exclusively from observational data and when $a = do(X_i = j)$ is chosen.
-We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
+It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones.
+We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
 Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
@@
 \centering
-\caption{A causal chain graph}.
-\label{fig:causalchain}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \end{tikzpicture}
+\caption{A causal chain graph}.
+\label{fig:causalchain}
 \end{figure}
 
-We use the notation $\P{\parents{Y}(X)}$ to denote a random variable that is the probability that the parents of $Y$ are $X$.
-Let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
+Let $P_a$ be a shorthand for $\P{.|a}$ and let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
 Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.

### 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
fixes missing dollar sign
adds truncated product formula

@@
 The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-The fact that $X_1,\ldots,X_N$ are independent can be used to create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. 
+The fact that only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
+Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
 These are the interventions that are easily estimated ``passively'' (\ie, without playing them explicitly). 
@@
 carefully chosen to minimise the worst-case simple regret. 
-The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occuring value (\ie, small $q_i$ or $1-q_i$). 
+The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$). 
 Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
@@
 \ENDFOR
-\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A$ = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
+\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
 \STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.

### 2016-02-05 causal-bandit/icml-2015/code/causal_bandits.py
makes successive rejects run nan for T less than K

@@ -297,17 +297,13 @@ class SuccessiveRejects(object):
     label = "Successive Reject"
     
     def run(self,T,model):
-        self.trials = np.zeros(model.K)
-        self.success = np.zeros(model.K)
-        self.actions = range(0,model.K)
+        
         if T <= model.K:
-            to_sample = np.random.choice(self.actions,size=T,replace=False)
-            self.trials[to_sample] +=1
-            self.success[to_sample] += model.sample_multiple(to_sample,1)
-            u = np.true_divide(self.success,self.trials)
-            best_action = argmax_rand(u)
-            #return np.nan
+            return np.nan
         else:
+            self.trials = np.zeros(model.K)
+            self.success = np.zeros(model.K)
+            self.actions = range(0,model.K)
             allocations = self.allocate(T,model.K)
             self.rejected = np.zeros((model.K),dtype=bool)
             for k in range(0,model.K-1):
@@ -543,10 +539,10 @@ def experiment2():
 #regret2,mean2,error2 = regret_vs_T(N,simulations,epsilon=None,Tstep=None,TperK=10)
 
 ## Experiment 3
-#simulations = 1000
-#N = 50
-#epsilon = .3
-#regret3,mean3,error3 = regret_vs_T(N,simulations,epsilon=epsilon,Tstep=25,TperK=6)
+simulations = 10000
+N = 50
+epsilon = .3
+regret3,mean3,error3 = regret_vs_T(N,simulations,epsilon=epsilon,Tstep=25,TperK=6)
 
 
 

### 2016-02-05 causal-bandit/icml-2015/code/causal_bandits.py
updates to save data

@@ -12,6 +12,7 @@ from numpy.random import binomial
 import matplotlib.pyplot as plt
 from time import time
 from datetime import datetime as dt
+import pickle
 
 
 REGRET_LABEL = "Regret"
@@ -442,7 +443,7 @@ def worst_case_constant():
     mean = regret.mean(axis=1)
     plt.plot(a_vals,mean)
 
-worst_case_constant()
+
         
 
 def regret_vs_m(N,epsilon,simulations,T):
@@ -532,18 +533,20 @@ def experiment2():
 #simulations = 100
 #T = 400
 #regret,mean,error = regret_vs_m(N,epsilon,simulations,T)
+#pickle.dump(regret, open("experiment1_{0}.pickle".format(now_string()),"wb"))
 
 # Experiment 2
 #N= 50
 #simulations = 100
 #regret2,mean2,error2 = regret_vs_T(N,simulations,epsilon=None,Tstep=None,TperK=10)
+#pickle.dump(regret2, open("experiment2_{0}.pickle".format(now_string()),"wb"))
 
 ## Experiment 3
 simulations = 10000
 N = 50
 epsilon = .3
 regret3,mean3,error3 = regret_vs_T(N,simulations,epsilon=epsilon,Tstep=25,TperK=6)
-
+pickle.dump(regret3, open("experiment3_{0}.pickle".format(now_string()),"wb"))
 
 
 

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
brief comment on causal models with non-interventional variables

@@
 
+\todof {This paragraph could be be clearer. These results are sub-optimal as they apply to adversarial case such that opponent can effectively choose m = N}
 Exploiting the variance structure in this way seems necessary even in the parallel bandit case.
@@
 \paragraph{Causal Models with Non-Observable Variables}
-We have assumed through this paper that all variables in the causal model were observable after an intervention is made.
-A natural generalisation of our model would be for some variables to be unobservable.
-\todom{FIX ME!}
-If the conditional models are known then the $Z_a$ terms in Algorithm~\ref{alg:general} could still be obtained via marginalising.
-However, if this is not the case the estimation problem becomes more difficult.
+If we assume knowledge of the conditional \textit{interventional} distributions $P_a$ our analysis applies unchanged to the case of causal models with non-observable variables.
+For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
+Some of the interventional distributions $P_a$ may be non-identifiable meaning we can not obtain prior estimates for $P_a$ from even an infinite amount of observational data.
+\todof{tie in to brief discussion of approach if had to estimate $P_a$ terms within algorithm.
+
 % In this case, some conditional distributions may be non-identifiable. 

### 2016-02-05 causal-bandit/icml-2015/tech.tex
Tweak wording
many minor improvements and checking Mark's comments

@@
 when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
-Towards this end recall that $m(\vec{q}) = \min\set{i : q_i \geq 1 / i}$.
-Therefore for $i > m$ we have $q_i \geq q_m \geq 1/m$ and so by Lemma \ref{lem:conc1} we have
+From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m$ that $q_i \geq q_m \geq 1/m$ and so by Lemma \ref{lem:conc1} we have
 \eq{
@@
 
+We follow a relatively standard path by choosing multiple environments that have different optimal arms, but which cannot all be statistically
+separated in $T$ rounds.
 Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
@@
 environment determined by $\boldsymbol{q}$ and $r_i$. Let $\operatorname{P}_i$ be the corresponding measure
-on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \cite{Tsy08} we have
+on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \citet{Tsy08} we have
 \eq{
@@
 Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $\Ps_0$ and $\Ps_i$ may be bounded using the telescoping property (chain rule) and
-by bounding the local KL divergence by the $\chi$-squared distance as by \cite{Auer1995}. This leads to 
+by bounding the local KL divergence by the $\chi$-squared distance as by \citet{Auer1995}. This leads to 
 \eq{
@@
 \eq{
-\sum_{i \in A} \Pri{I_T \neq i} 
+\sum_{i \in A} \Pri{\hat a^*_T \neq a} 
 \geq \sum_{i \in A} \exp\left(-\KL(\Ps_0, \Ps_i)\right) - 1
@@
 Therefore there exists an $i \in A$ such that
-$\Pri{I_T \neq i} \geq \frac{\frac{m}{2e} - 1}{m}$. 
+$\Pri{\hat a^*_T \neq a^*} \geq \frac{\frac{m}{2e} - 1}{m}$. 
 Therefore if $\epsilon < 1/4$ we have

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
add a couple of todos

@@
 \paragraph{Making Better Use of the Reward Signal}
-Several existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
+Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
 In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
@@
 This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
+For the parallel bandit the required modifications should be quite straightforward. The idea would be to adapt the algorithm to essentially use successive elimination in
+the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability. In the general case a similar modification is also possible
+by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping. Note
+that these modifications will not improve the minimax regret, which at least for the parallel bandit is already optimal. For this reason we prefer to emphasize 
+the main point that causal structure should be exploited when available. Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which
+in some cases may be preferred to a sequential design for logistical reasons. This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.
 
 \paragraph{Cumulative Regret}
-Although we have focused on simple regret in our analysis, a natural but open question is whether cumulative regret bounds can be obtained for the general causal bandit problem we have proposed.
-In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with graph feedback to get optimal cumulative regret guarantees.
-Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
-We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
-\todom{Check this!}
-Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
+Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
+In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information 
+to get near-optimal cumulative regret guarantees. They consider a finite-armed bandit model with side information where in reach round
+the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
+In this way the learner can gain information about actions it does not take with varying levels of accuracy. The reduction follows by substituting
+the importance weighted estimators in place of the Gaussian reward. In the case that $\vec{q}$ is known this would lead to a known variance
+and the only (insignificant) difference is the Bernoulli noise model. In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
+at least asymptotically.
+
+%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
+%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
+%\todom{Check this!}
+%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
 
@@
 Some of the interventional distributions $P_a$ may be non-identifiable meaning we can not obtain prior estimates for $P_a$ from even an infinite amount of observational data.
-\todof{tie in to brief discussion of approach if had to estimate $P_a$ terms within algorithm.
+\todof{tie in to brief discussion of approach if had to estimate $P_a$ terms within algorithm.}
 

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
adds truncated product formula
experiments thanks to AWS

@@
-\subsection{Parallel bandits problem}
-
-For the parallel bandit problem, we use a model in which, unknown to the algorithms, $Y$ depends only on a single variable.
+We compare Algorithms \ref{alg:simple} and \ref{alg:general} with Successive Elimination on the parallel bandit problem
+under a variety of conditions, including where the importance weighted estimator used by Algorithm \ref{alg:general} is not truncated,
+which is justified in this setting by Remark \ref{rem:truncate}. Throughout we use a model in which $Y$ depends only on a single variable (this is unknown to the algorithms). 
 \eq{
@@
 }
-
 This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\left(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon\right)$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
@@
 
-In figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
+In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
 The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure.
@@
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s1000_20160205_0910.pdf}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160205_2230}
 \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
@@
 This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m << N$.
-In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
+In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
 The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.
@@
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1316}
 \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
@@
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1532}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1303}
 \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}

### 2016-02-05 causal-bandit/icml-2015/general.tex
add a couple of todos

@@
 \end{figure} 
+\todof{connect truncated product formula to here by noting that the truncated formula allows us to estimate $P_a$ from observational quantities - which could plausible be available prior to running the experiment, allowing us to assume the $P_a$'s are known.}
 
@@
 Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
-a truncated importance weighted estimator.
+a truncated importance weighted estimator. The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail
+that is so detrimental to its concentration guarantees.
 
@@
 where $B_a \geq 0$ is some constant to be chosen subsequently and
-\todom{Symbol clash with regret}
 \eq{
@@
 }
-We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
+We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately 
+coincides with the version for parallel bandits, justifying the name overloading.
 
@@
 }
-\STATE {\bf return} $\hat a^* = \argmax_a \hat \mu_a$
+\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
 \end{algorithmic}
@@
 
+\begin{remark}\label{rem:truncate}
+The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
+often possible to choose $B_a$ significantly
+larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
+and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
+to learn from observational data.
+\end{remark}
+
 \subsection*{Choosing the Sampling Distribution}
@@
 For $a = do()$ choose $\eta_a = 1/2$. 
-\todom{Check please!}
 Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.

### 2016-02-05 causal-bandit/icml-2015/general.tex
reworks intro to general graphs

--- 2016-02-05 causal-bandit/icml-2015/general.tex
@@
 Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
-For example, given the causual graph in Figure \ref{fig:causalStructure_confounded}, by applying the truncated factorization and marginalizing, we obtain $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1}$.
-Note that the quality of any estimator we build for $P{Y|do(X_2= j)}$ is not just a function of the number of times we observe $X_2 = j$.
+For example, given the causual graph in Figure \ref{fig:causalStructure_confounded}, by applying the truncated factorization and marginalizing, we obtain $\P{Y|do(X_2= j)} = \sum_{X_1}\P{Y|X_1,X_2=j}\P{X_1}$.
 
@@
 
-It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones.
-We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor.
+However, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
+In our example, suppose $\P{X_1=\frac{1}{2}}$ and $X_2 = X_1$ deterministically.
+$\P{X_2 = 1} = \frac{1}{2}$, however we will never observe $(X_2=1,X_1 = 0)$ so we cannot get a good estimate for $\P{Y|do(X_2=1)}$.
+
+Additionally, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
 Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$
 for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
-We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
+We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
 
@@
 
+To overcome these difficulties we simplify the problem by assuming the conditional interventional distributions $\P{\boldsymbol{X}|a}$ are known for all variables except $Y$.
 Let $P_a$ be a shorthand for $\P{.|a}$ and let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.

### 2016-02-05 causal-bandit/icml-2015/introduction.tex
reworks intro to general graphs

@@
 
-Two pieces of recent work also consider using causal models in bandit problems.
-\citet{Bareinboim2015} use causal inference to improve the chance of choosing an optimal arm.
-Their work differs from ours in two key ways.
-Firstly, their focus is on correcting the effects of unobserved confounding variables whereas we assume all variables are observable.
-Secondly, they do not assume their learning algorithm has access to a causal graph and instead use a form of contextual randomization to improve reward estimates in a Thompson sampling-style algorithm.
+Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
+\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been intervened on can provide important contextual information.
+\todof{write one sentence to explain how their work is different to ours}
+
  \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions.
 Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
-Neither work provides a regret analysis.
-Combining their handling of unobserved variables and causal induction with our analysis is left as future work.
+Combining their handling of  causal induction with our analysis is left as future work.
 % Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.

### 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
reworks intro to general graphs

@@
 The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-The fact that only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
-Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
-These are the interventions that are easily estimated ``passively'' (\ie, without playing them explicitly). 
-Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
-the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
-the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
-carefully chosen to minimise the worst-case simple regret. 
+As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
+We use this to get good estimates for the actions $do(X_i=j)$ for which $\P{X_i = j}$ is large, without playing them explicitly.
+
+Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so the estimates of the returns of these actions from the first $T/2$ samples could be quite poor.
+For the remaining $T/2$ rounds the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small.
+The precise allocation is carefully chosen to minimise the worst-case simple regret.
 The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$).

### 2016-02-05 causal-bandit/icml-2015/experiments.tex
final version of figs for experiments
fixes

--- 2016-02-05 causal-bandit/icml-2015/experiments.tex
@@
 \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160205_1532}
+\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_s10000_20160206_0533}
 \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}

### 2016-02-05 causal-bandit/icml-2015/general.tex
fixes

--- 2016-02-05 causal-bandit/icml-2015/general.tex
@@
 
-It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones.
-We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor.
+However, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
+In our example, suppose $\P{X_1=\frac{1}{2}}$ and $X_2 = X_1$ deterministically.
+$\P{X_2 = 1} = \frac{1}{2}$, however we will never observe $(X_2=1,X_1 = 0)$ so we cannot get a good estimate for $\P{Y|do(X_2=1)}$.
+
+Additionally, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
 Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$
 for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
-We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
+We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
+
+%It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones.
+We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
+%Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
+%for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
+%We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
 

### 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
fixes

--- 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
@@
 \ENDFOR
-\RETURN estimated optimal $\hat{a}^* \in \argmax_{a\in\actions} \hat{\mu}_a$
+\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
 \end{algorithmic}
@@
 this setting.
-We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}
+We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}.
 
 \begin{theorem}\label{thm:lower}
-Let $\vec{q}$ be fixed and satisfies $m = m(\boldsymbol{q})$. Then for all strategies there exists a reward function such that
+For all $T$, $\vec{q}$ and all strategies there exists a reward function such that
 \eq{
 \simpleregret 
-%\geq \frac{\frac{m}{2e} - 1}{m} \sqrt{\frac{2m}{3T}} 
-\in \Omega\left(\sqrt{\frac{m}{T}}\right)\,.
+\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.
 }
@@
 \ifsup
-The proof of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower}.
+The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
 \else
-The proof of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in the supplementary material.
+The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
+of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
 \fi

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
updated comment in discussion

@@
 \paragraph{Causal Models with Non-Observable Variables}
-If we assume knowledge of the conditional \textit{interventional} distributions $P_a$ our analysis applies unchanged to the case of causal models with non-observable variables.
-For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
-Some of the interventional distributions $P_a$ may be non-identifiable meaning we can not obtain prior estimates for $P_a$ from even an infinite amount of observational data.
-\todof{tie in to brief discussion of approach if had to estimate $P_a$ terms within algorithm.}
+If we assume knowledge of the conditional \textit{interventional} distributions $P_a$ our analysis applies unchanged to the case of causal models with 
+non-observable variables. For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not 
+available from which to build estimates of $P_a$. Some of the interventional distributions $P_a$ may be non-identifiable meaning we can not obtain prior estimates for $P_a$ from 
+even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
+\ref{alg:general} cannot be used. Estimating these quantities will simultaneously minimising the simple regret is an interesting and challenging open problem.
 

### 2016-02-05 causal-bandit/icml-2015/discussion.tex
fixed typo
updates graph feedback para

@@
 
-\todof {This paragraph could be be clearer. These results are sub-optimal as they apply to adversarial case such that opponent can effectively choose m = N}
-Exploiting the variance structure in this way seems necessary even in the parallel bandit case.
-If instead we choose to use other graph feedback models we would require those with time varying graphs~\citep{Alon2015,Kocak2014}.
-Results for these models give a cumulative regret of $\bigo{\sqrt{\bar{\alpha} T}}$ where $\bar{\alpha}$ is the average independence number of the time varying graphs.
-On each round, the feedback graph over the $2N$ possible interventions over $N$ binary variables would connect the chosen intervention to the corresponding intervention for each of the other $N-1$ observed variables.
-Unfortunately, in this case the resulting star graphs have an $\bar{\alpha} = N$ and so the cumulative regret bounds would improve over those obtained by ignoring the causal graph structure.
+The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
+The feedback graph is distinct from the causal graph.
+A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
+For this parallel bandit problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
+However, \citet{Alon2015,Kocak2014} give adversarial algorithms, which when applied to the parallel bandit problem obtain the standard bandit regret.
+A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
+This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
 
@@
 even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
-\ref{alg:general} cannot be used. Estimating these quantities will simultaneously minimising the simple regret is an interesting and challenging open problem.
+\ref{alg:general} cannot be used. Estimating these quantities while simultaneously minimising the simple regret is an interesting and challenging open problem.
 

### 2016-02-05 causal-bandit/icml-2015/code/causal_bandits.py
removes unused code and identif info

@@ -1,12 +1,7 @@
 # -*- coding: utf-8 -*-
-"""
-Created on Wed Feb  3 09:51:47 2016
 
-@author: finn
-"""
 import numpy as np
 from itertools import product
-from scipy.optimize import minimize
 from math import sqrt,log,ceil
 from numpy.random import binomial
 import matplotlib.pyplot as plt
@@ -50,109 +45,6 @@ def calculate_m(qij_sorted):
         if value >= 1.0/(indx+1):
             return indx
     return len(qij_sorted)/2      
-    
-class VeryConfounded(object):
-    def __init__(self,a,b,n,q1,q2,pYgivenW):
-        #self.pV = np.vstack((1.0-q,q)) TODO fix this.
-        self.n = n # number of V variables
-        self.N = self.n + 2 # number of variables total
-        self.K = 2*self.N + 1 #v_1...v_n = 0,v_1...v_n=1,w0=0,w1=0,w0=1,w1=1,do()
-        self.pYgivenW = pYgivenW # 2*2 matrix p(y|0,0),p(y|0,1),p(y|1,0),p(y|1,1)
-        
-        self.pW0givenA = np.full(self.K,(1-q1)*a + q1*q2*(n-2)/(n-1.0))
-        self.pW0givenA[n:] = (1-q1)*a + q1*(q2*(n-2)/(n-1.0)+1/(n-1.0))
-        self.pW0givenA[0] = a
-        self.pW0givenA[n] = q2
-        self.pW0givenA[[-4,-2,-1]] = (1-q1)*a + q1*q2 # for do(w1=0),do(w1=1),do()
-        self.pW0givenA[-3] = 1 # for do(w0 = 1)
-        self.pW0givenA[-5] = 0 # for do(w0 = 0)
-        
-        self.pW1givenA = np.full(self.K,(1-q1)*b)
-        self.pW1givenA[n:] = (1-q1)*b+q1*q2**(n-2)
-        self.pW1givenA[0] = b
-        self.pW1givenA[n] = q2**(n-1)
-        self.pW1givenA[[-5,-3,-1]] = (1-q1)*b # for do(w0=0),do(w0=1),do()
-        self.pW1givenA[-2] = 1 # for do(w1 = 1)
-        self.pW1givenA[-4] = 0 # for do(w1 = 0)
-        
-        self.pW0givenA = np.vstack((1-self.pW0givenA,self.pW0givenA))
[…]

### 2016-03-28 causal-bandit/icml-2015/response.txt
updates

Thank you to all reviewers for your detailed and thoughtful comments

We apologise for the confusing/missing definitions and lack of intuition for Alg2. We will take care to clarify these issues.
We address question on Sec. 4 here as all reviewers had similar questions. Specific comments follow.

mu_I should be \mu_{\hat a^*_T}, which is the true expected payoff for the intervention chosen by the algorithm.

Pa_Y(X) is the realization of the variables in X that are (direct) parents of Y.

P_a{Pa_Y(X)} is the probability of Pa_Y(X) given we select action a

The definition of R_a on L483 should now make mathematical sense. It is used in the definition of Z_a, which is the truncated
importance weighted estimator of the payoff for action $a$ with respect to sampling distribution Q.

R_a will be large if we get an assignment to the parents of Y that is likely given action $a$ but unlikely given we sample from Q. 
The vector eta determines the likelihood of using each intervention (including the trivial "do nothing").  
Then m(eta) can be interpreted as an upper bound on the worst-case (over the actions) variance of the estimators Z_a, which
is justified by the calculations in L523 and is the reason that eta is chosen to minimise m(eta). A nice outcome is that
for the parallel bandit case the value of m(eta) coincides with the value of m(q) given in Section 3 (argument in L631). 
At a high-level eta is chosen to maximise the information gained about all actions simultaneously.

The generalisation of Alg.1 to arbitrary graphs is not straightforward because in order to optimise eta the values of P_a on the parents
of all variables should be estimated and bounded. Evaluating the propagation of the errors is then very challenging and seems
worthy of a more lengthy article.

Rev. 1

Exploiting the causal structure allows us to learn the payoffs of multiple actions at each timestep, which
greatly speeds the learning as demonstrated in Fig.5. Estimating P(Y|X1...XN) and P(X1...XN) would require 
exponentially (in N) many estimators, which is the issue our approach addresses. We will make this key point clearer.

Simple regret is a standard measure in the bandit community that captures the key aspects of the problem. An alternative would be the cumulative regret,
which is equally interesting and sometimes more/less suitable depending on the setting. In both cases it would be possible to study the asymptotics as an
alternative theoretical approach. This is often easier, but less informative from a practical perspective. We would be curious to know about alternative theoretical
frameworks you might suggest.

For Alg1, in practice we would combine the data from both phases in the estimates for the actions with low natural probability. 
This slightly complicates the proofs and does not improve the regret bounds. We will add a remark to clarify this.

Re. contextual bandits, we want to emphasise that those algorithms cannot be applied in our framework because they rely on 
the context being available at decision time, while in our setting the X vector is not observed until after
an intervention is made.

Although the idea of combining causality with bandit algorithms has been considered before, our approach is (to our knowledge) novel and quite general.
We agree that in most real applications there will be some uncertainty in the causal model and this is a very interesting (and challenging) topic for 
future research. However we believe our article is a good first step.

Re. the proofs. We will include more steps and references in the supplementary material.
We will include an exact ref. for the Chernoff bound in L1162. Intuitively the q_i appears because the
variance for a Bernoulli with parameter q is q(1-q) \leq q and so it is not surprising to see this inside the square root by the central
limit theorem.

Rev. 2
Thank you for spotting the typos in eq.(2) and the do() expression in Sec. 4.

You are correct with regard to estimating P(Y|do(X2)) via X3. We should let X1==X2==X3 or replace fig.2 with the simpler graph X1 -> X2 -> Y, X1 -> Y.

Rev. 4
Thank you for the detailed notes on typos. These will be fixed.

We don't know of any algorithms comparable to Alg.1 and the measure of problem difficulty $m$ is also novel.

Although trivial, it is causal theory that tells us the interventional and conditional probabilities are equal in the parallel bandit setting. 
Alg.1/2 exploit this causal structure to beat the standard algorithms. We presented this setting so as to allow comparison between Alg.1/Alg.2. 
We will add additional experiments for the general setting to the supplementary material.

The experiments demonstrate the causal algorithms exploit the additional assumptions to escape the lower bounds that apply to standard simple regret algorithms. 
All other algorithms will be subject to these bounds and thus perform similarly, however we are willing to add additional curves.

### 2016-03-28 causal-bandit/icml-2015/response.txt
updats

@@
 
-The definition of R_a on L483 should now make mathematical sense. It is used in the definition of Z_a, which is the truncated
+The definition of R_a on L483 should now make mathematical sense. It is used in the definition of Z_a (L476), which is the truncated
 importance weighted estimator of the payoff for action $a$ with respect to sampling distribution Q.

### 2016-03-29 causal-bandit/icml-2015/response.txt
add a point on the definition of action, makes a few sentances slightly shorter, changes para on generalizing alg1

@@
 
-We apologise for the confusing/missing definitions and lack of intuition for Alg2. We will take care to clarify these issues.
-We address question on Sec. 4 here as all reviewers had similar questions. Specific comments follow.
+We apologise for the confusing/missing definitions and lack of intuition for Alg2.
+We will take care to clarify these issues.
+We address comments on Sec.
+4 here as all reviewers had similar questions.
+Specific comments follow.
 
@@
 
-The definition of R_a on L483 should now make mathematical sense. It is used in the definition of Z_a (L476), which is the truncated
-importance weighted estimator of the payoff for action $a$ with respect to sampling distribution Q.
+The definition of R_a on L483 should now make mathematical sense.
+It is used in the definition of Z_a (L476), which is the truncated importance weighted estimator of the payoff for action $a$ with respect to sampling distribution Q.
 
@@
 The vector eta determines the likelihood of using each intervention (including the trivial "do nothing").
-Then m(eta) can be interpreted as an upper bound on the worst-case (over the actions) variance of the estimators Z_a, which
-is justified by the calculations in L523 and is the reason that eta is chosen to minimise m(eta). A nice outcome is that
-for the parallel bandit case the value of m(eta) coincides with the value of m(q) given in Section 3 (argument in L631). 
-At a high-level eta is chosen to maximise the information gained about all actions simultaneously.
+Then m(eta) can be interpreted as an upper bound on the worst-case (over the actions) variance of the estimators Z_a, which is justified by the calculations in L523 and is the reason that eta is chosen to minimise m(eta).
+A nice outcome is that for the parallel bandit case the value of m(eta) coincides with the value of m(q) given in Section 3 (argument in L631).
+At a high-level, eta is chosen to maximise the information gained about all actions simultaneously.
 
-The generalisation of Alg.1 to arbitrary graphs is not straightforward because in order to optimise eta the values of P_a on the parents
-of all variables should be estimated and bounded. Evaluating the propagation of the errors is then very challenging and seems
-worthy of a more lengthy article.
+We don't directly generalize alg1 as it's no longer optimal to trade off observing against selecting hard to learn actions.
+Instead we must optimize the assignment of samples to actions so as to leverage all the information a general causal graph provides about the reward for one action given another.
 
@@
 
-Exploiting the causal structure allows us to learn the payoffs of multiple actions at each timestep, which
-greatly speeds the learning as demonstrated in Fig.5. Estimating P(Y|X1...XN) and P(X1...XN) would require 
-exponentially (in N) many estimators, which is the issue our approach addresses. We will make this key point clearer.
+Exploiting the causal structure allows us to learn the payoffs of multiple actions at each timestep, which greatly speeds the learning as demonstrated in Fig5.
+Estimating P(Y|X1...XN) and P(X1...XN) would require exponentially (in N) many estimators, which is the issue our approach addresses.
+We will make this key point clearer.
 
-Simple regret is a standard measure in the bandit community that captures the key aspects of the problem. An alternative would be the cumulative regret,
-which is equally interesting and sometimes more/less suitable depending on the setting. In both cases it would be possible to study the asymptotics as an
-alternative theoretical approach. This is often easier, but less informative from a practical perspective. We would be curious to know about alternative theoretical
-frameworks you might suggest.
-
-For Alg1, in practice we would combine the data from both phases in the estimates for the actions with low natural probability. 
-This slightly complicates the proofs and does not improve the regret bounds. We will add a remark to clarify this.
-
-Re. contextual bandits, we want to emphasise that those algorithms cannot be applied in our framework because they rely on 
-the context being available at decision time, while in our setting the X vector is not observed until after
-an intervention is made.
+Simple regret is a standard measure in the bandit community that captures the key aspects of the problem.
+An alternative would be the cumulative regret, which is equally interesting and sometimes more/less suitable depending on the setting.
+In both cases it would be possible to study the asymptotics as an alternative theoretical approach.
+This is often easier, but less informative from a practical perspective.
+We would be curious to know about alternative theoretical frameworks you might suggest.
 
 Although the idea of combining causality with bandit algorithms has been considered before, our approach is (to our knowledge) novel and quite general.
-We agree that in most real applications there will be some uncertainty in the causal model and this is a very interesting (and challenging) topic for 
-future research. However we believe our article is a good first step.
+We agree that in most real applications there will be some uncertainty in the causal model and this is a very interesting (and challenging) topic for future research.
+However we believe our article is a good first step.
 
-Re. the proofs. We will include more steps and references in the supplementary material.
-We will include an exact ref. for the Chernoff bound in L1162. Intuitively the q_i appears because the
-variance for a Bernoulli with parameter q is q(1-q) \leq q and so it is not surprising to see this inside the square root by the central
-limit theorem.
+Re. Sec2, an action selects a variable and sets its value. We will clarify this.
+
+Re.
+Alg1, in practice we would combine the data from both phases in the estimates for the actions with low natural probability.
+This slightly complicates the proofs and does not improve the regret bounds.
+We will add a remark to clarify this.
+
+Re.
+contextual bandits, we want to emphasise that those algorithms cannot be applied in our framework because they rely on the context being available at decision time, while in our setting the X vector is not observed until after the intervention.
+
+Re.
+proofs.
+We will include more steps in the supplementary material and an exact ref.
+for the Chernoff bound in L1162.
+Intuitively q_i appears because the variance for a Bernoulli with parameter q is q(1-q) \leq q and so it is not surprising to see this inside the square root by the central limit theorem.
 
 Rev. 2
-Thank you for spotting the typos in eq.(2) and the do() expression in Sec. 4.
+Thank you for spotting the errors in eq.(2) and the do() expression in Sec.4.
 
-You are correct with regard to estimating P(Y|do(X2)) via X3. We should let X1==X2==X3 or replace fig.2 with the simpler graph X1 -> X2 -> Y, X1 -> Y.
+You are correct with regard to estimating P(Y|do(X2)) via X3. We should let X1==X2==X3 or replace fig2 with the simpler graph X1->X2->Y, X1->Y.
 
@@
 
-We don't know of any algorithms comparable to Alg.1 and the measure of problem difficulty $m$ is also novel.
+We don't know of any algorithms comparable to Alg1 and the measure of problem difficulty $m$ is also novel.
 
 Although trivial, it is causal theory that tells us the interventional and conditional probabilities are equal in the parallel bandit setting.
-Alg.1/2 exploit this causal structure to beat the standard algorithms. We presented this setting so as to allow comparison between Alg.1/Alg.2. 
+Alg1&2 exploit this causal structure to beat the standard algorithms.
+We presented this setting so as to allow comparison between Alg1&Alg2.
 We will add additional experiments for the general setting to the supplementary material.

### 2016-03-29 causal-bandit/icml-2015/code/causal_bandits.py

@@ -1,7 +1,12 @@
 # -*- coding: utf-8 -*-
+"""
+Created on Wed Feb  3 09:51:47 2016
 
+@author: finn
+"""
 import numpy as np
 from itertools import product
+from scipy.optimize import minimize
 from math import sqrt,log,ceil
 from numpy.random import binomial
 import matplotlib.pyplot as plt
@@ -45,6 +50,109 @@ def calculate_m(qij_sorted):
         if value >= 1.0/(indx+1):
             return indx
     return len(qij_sorted)/2      
+    
+class VeryConfounded(object):
+    def __init__(self,a,b,n,q1,q2,pYgivenW):
+        #self.pV = np.vstack((1.0-q,q)) TODO fix this.
+        self.n = n # number of V variables
+        self.N = self.n + 2 # number of variables total
+        self.K = 2*self.N + 1 #v_1...v_n = 0,v_1...v_n=1,w0=0,w1=0,w0=1,w1=1,do()
+        self.pYgivenW = pYgivenW # 2*2 matrix p(y|0,0),p(y|0,1),p(y|1,0),p(y|1,1)
+        
+        self.pW0givenA = np.full(self.K,(1-q1)*a + q1*q2*(n-2)/(n-1.0))
+        self.pW0givenA[n:] = (1-q1)*a + q1*(q2*(n-2)/(n-1.0)+1/(n-1.0))
+        self.pW0givenA[0] = a
+        self.pW0givenA[n] = q2
+        self.pW0givenA[[-4,-2,-1]] = (1-q1)*a + q1*q2 # for do(w1=0),do(w1=1),do()
+        self.pW0givenA[-3] = 1 # for do(w0 = 1)
+        self.pW0givenA[-5] = 0 # for do(w0 = 0)
+        
+        self.pW1givenA = np.full(self.K,(1-q1)*b)
+        self.pW1givenA[n:] = (1-q1)*b+q1*q2**(n-2)
+        self.pW1givenA[0] = b
+        self.pW1givenA[n] = q2**(n-1)
+        self.pW1givenA[[-5,-3,-1]] = (1-q1)*b # for do(w0=0),do(w0=1),do()
+        self.pW1givenA[-2] = 1 # for do(w1 = 1)
+        self.pW1givenA[-4] = 0 # for do(w1 = 0)
+        
+        self.pW0givenA = np.vstack((1-self.pW0givenA,self.pW0givenA))
[…]

### 2016-04-22 identify/identify.py
initial exploration of identifiability problem

@@ -1,0 +1,399 @@
+# code based on "On the Identification of Causal Effects", Tian & Pearl
+import graphviz as gv
+import subprocess
+import os
+
+def add_value(dictionary,key,value = None):
+    """ adds the value to list of values for relevent key """
+    if key in dictionary:
+        if value is not None:
+            dictionary[key].append(value)
+    else:
+        if value is None:
+            dictionary[key] = []
+        else:
+            dictionary[key]=[value]
+
+def intersection(lst1,lst2):
+    return [x for x in lst1 if x in lst2]
+
+def math_sum(variables,exclude):
+    sum_over  = variables - exclude
+    if len(sum_over) > 0:
+        return "\sum_{%s}" % ",".join(sum_over)
+    return ""
+
+def math_sum2(variables,exclude):
+    """ takes in two sets and returns a tex expression summing over variables \ exclude """
+    sum_over = variables - exclude
+    if len(sum_over) > 0:
+        return "\sum_{%s}"%",".join(sum_over)
+    return ""
+
+def math_condition(left,*right):
+    result = "P("+left
+    first = True
+    for cond in right:
+        print(cond)
+        if len(cond) > 0:
+            if first:
+                result+="|"
+                first = False
+            else:
+                result+=","
+            result+=",".join(cond)
+    result +=")"
+    return result
+            
+    
+def edge_to_string(edge):
+    return edge[0]+"-"+edge[1]
+
+def edge_lists_to_string(dg,ug):
+    return ",".join([edge_to_string(edge) for edge in dg])+":"+",".join([edge_to_string(edge) for edge in ug])
+
+def read_undirected(file):
+    """ reads the output of listg -t (see below). Creates an edge list for each graph
+    Graph 1, order 3.
+  0 : 2;
+  1 : 2;
+  2 : ;
+
+Graph 2, order 3.
+  0 : 1 2;
+  1 : 2;
+  2 : ;
+  """
+    graphs = []
+    edges = []
+    for line in file.readlines():
+        if line.startswith("Graph"): 
+            if len(edges) > 0:
+                graphs.append(edges)
+            edges = []
+        elif ":" in line:
+
+            line = [x.strip() for x in line.strip(";\n").split(":")]
+            source = line[0]
+            dest = line[1].split(" ")
+            for d in dest:
+                if len(d) > 0:
+                    edges.append((source,d))
+    graphs.append(edges)
+    return graphs
+
+
+def read_directed(file):
[…]

### 2016-04-22 causal-bandit/icml-2015/rebuttal.txt
cleaned up structure of causality folder

Algorithm 2
R 1
 Where is R_a (l477) defined? l483
 What does Pa_Y(X) and in particular P_a{Pa_Y(X)} mean? (l483)
 Where is \mu_I (l516) defined?
R2
What is Pa_Y(X) notation, the subrealization for parents of Pa_Y given a realization X of all variables?Did not understand algorithm2. If we are following what happened with algorithm 1, what we want to do is run the system observationally for a while, and learn which combinations of treatments occur frequently via an estimate of P(X1, ..., Xk) (marginal over all treatment variables if there are k of them in the DAG). Then use estimate of P(X1, ..., Xk) to construct a version of m(q) for a distribution that does not marginally factorize, as we expect to be the case for P(X1,..., Xk) in an arbitrary DAG. Finally, use that version of m(q) to drive what interventions to perform.
Where does eta come from, and what relationship does it have to observed treatment occurrence frequencies? It sort of seems like algorithm 2 is really trying to be optimal with respect to this eta distribution, which has no relationship to actually observed frequencies of treatment occurrence, as happens with algorithm 1. Can the authors explain this? I thought the point was to try to solve “a bandit like problem” for causal relationships given observational data.
R3
Two central notations --Pa_Y(X) and mu_I-- are not defined, making it difficult both to understand the meaning of the quantity R_a and to check the subsequent proofs. 
Need intuition for second algorithm. Explanation of R_a

We cannot directly generalise algorithm1 as it is no longer trivial to determine the expected variance of the estimators for each action given observational data and the optimal way to learn about multiple actions may no longer be to observe. We must come up with a way to optimally assign our limited budget of samples, T, to actions, a, so as to get a good estimate for all actions. This assignment must take advantage of the complex ways a general causal graph provides information about the reward for one action given we take a different one.

As a first step to addressing this difficult problem we make the simplifying assumption that the conditional interventional distributions, P{X| a} are known for all variables except the outcome Y.  Let Pa_Y(X) denote a realization of the variables in X that are (direct) parents of Y. Note that P{Y|do(Pa_Y), other variables} = P{Y|do(Pa_Y)}. We can then construct an importance weighted estimate for the reward Y given each action, a, where the target distribution is P{Pa_Y(X)| a} and the sampling distribution Q is a fixed average over all the target distributions for all the allowable actions, weighted by eta.

R_a is the ratio of the probability of observing some realization of the parents of Y given we take action a to observing that realization given we are sampling actions from Q. R_a will be large if we get an assignment to the parents of Y that is likely given action a but unlikely given we sample from Q.  m(eta) is the largest (over a) expected value of R_a. A large expected value of R_a implies it will be difficult to get a good estimate for that action as we expect only a few (highly weighted) samples will contribute.  We therefore select eta* to minimise m(eta) such that our estimates are not too uncertain for any action.  m(eta*) can then be interpreted as a measure of the underlying difficulty of the problem, which we can connect back to m in algorithm 1.

mu_I is the true expected reward for the action the algorithm estimates to be optimal at time T. It should have been denoted mu_a^*_T, consistent with the remainder of the paper.

Reviewer 1 (causality side)

make clear which mathematical terms in section 3 & 4 are central
many concepts, such as causal models (population-level), regret analysis (sample-level), interventions, actions, etc. are mixed in a way that is sometimes hard to follow.
there should be more instructive but still theoretical ways than finite-sample regrets to argue why knowledge of the causal DAG helps. Great to not only have the hard-to-read finite-sample regret analysis, but also some more instructive considerations of when/how the causal approach helps.

obvious that one can construct theoretical scenarios where causal knowledge or modeling helps for bandit-like scenarios
this direction has already been investigated e.g. by Bottou and Bareinboim - although the scenarios differ.
Motivate with real world problem. (+ how can farmer know DAG in advance)
 You make an interesting point in 453-455.
From Sec. 2 it is not quite clear: does an action consists of (1) a decision on which variable to intervene as well as (2) what value to set the chosen variable to? (It gets clear later that its both.) This seems pretty clear, line 210, not sure how we can improve it.
Why don't you use the initial T/2 data points for the final estimation of expected rewards (Alg1)
We do use it for those actions that can be sufficiently estimated from the observational data. We could combine the estimates for the remaining actions with those obtained via intervention, but this would slightly complicate the proof of the regret bound and only change the bound by a constant factor of at most 2.

 Wouldn't it make sense to also try to estimate the form of the mechanism p(Y|X_1, ..., X_N) because this together with the estimated distributions on X_i should give additional information on the expected rewards.
The number of estimates this requires (without some form of assumption on the function) grows exponentially with the number of variables. It also does not provide any further information on the expected rewards if we accept the stated structural assumptions and set of allowed interventions.
Unclear if the problem they address couldn't also be addressed by some adapted version of contextual bandits (which abstract away from the difference between (1) on which variable one intervenes and (2) to which value one sets this variable).
I don't understand the 2nd half of this comment... Contextual bandits extend standard bandits by allowing there to be additional contextual information, C, available prior to an agent selecting an action, a,. The agent must then build estimates P(Y|a, C). In our setting the additional information is received after the agent selects their action.

Why does the q_i appear on the r.h.s. of the greater-or-equal sing in l1162? I couldn't see that from Chernoff. (2) 
This probably just needs a reference. 
Can you write down more steps that show the last inequality in l1176? It's somehow clear that you can bound it by the integral, but it would be much nicer for the reader if he/she could just go through it and not spend so much time for reconstructing steps.
Yes we will add more steps

Reviewer 2
Section 4: The formula for p(Y | do(x2)) is incorrect, should be \sum_x1 p(x1) p(y| x1, x2). The functional better depend on x2!
Yep, this is a typo. Reviewer is correct.

“however we will never observe (X2 = 1, X1 = 0) so we cannot get a good estimate for P {Y |do(X2 = 1)}.” Not obvious, since \sum_x3 p(x3) p(y | x3, x2) is also equal to p(y |do(x2)).
True. We could let X1=X2=X3 deterministically or just use the simpler graph X1 → X2 → Y, X1 → Y

(2) should bound d from above by N, right? 
Yep, something not quite right about (2)

Reviewer 4
It is a pity that the second algorithm was only tested in a setting, where all interventional probabilities actually equal conditional probabilities (P(Y|X=x) = P(Y|do(X=x)) ). Thus, there is de facto no experiment in the paper that actually uses techniques specific to causal theory.
Although trivial, it is causal theory that tells us the interventional and conditional probabilities are equal in the parallel bandit setting and algorithms 1 & 2 both use this causal structure to beat the standard algorithms. We presented this setting so as to allow comparison between alg1 and alg2. We will endeavour to add additional experiments for the general setting to the supplementary material.

Why do you compare your algorithms to only 1 single competitor algorithm? If no valid/strong reason is given, then please add curves for at least one or two other competitors.
The experiments aim to demonstrate the causal algorithms escape the lower bounds that apply to standard simple regret algorithms. All other algorithms will be subject to these bounds and thus perform with similarly, however we can add additional curves.

### 2016-04-22 causal-bandit/icml-2015/rebuttal_final.txt
cleaned up structure of causality folder

--- 2016-04-22 causal-bandit/icml-2015/rebuttal.txt
@@
-Algorithm 2
-R 1
- Where is R_a (l477) defined? l483
- What does Pa_Y(X) and in particular P_a{Pa_Y(X)} mean? (l483)
- Where is \mu_I (l516) defined?
-R2
-What is Pa_Y(X) notation, the subrealization for parents of Pa_Y given a realization X of all variables?Did not understand algorithm2.
-If we are following what happened with algorithm 1, what we want to do is run the system observationally for a while, and learn which combinations of treatments occur frequently via an estimate of P(X1, ..., Xk) (marginal over all treatment variables if there are k of them in the DAG).
-Then use estimate of P(X1, ..., Xk) to construct a version of m(q) for a distribution that does not marginally factorize, as we expect to be the case for P(X1,..., Xk) in an arbitrary DAG.
-Finally, use that version of m(q) to drive what interventions to perform.
-Where does eta come from, and what relationship does it have to observed treatment occurrence frequencies?
-It sort of seems like algorithm 2 is really trying to be optimal with respect to this eta distribution, which has no relationship to actually observed frequencies of treatment occurrence, as happens with algorithm 1.
-Can the authors explain this?
-I thought the point was to try to solve “a bandit like problem” for causal relationships given observational data.
-R3
-Two central notations --Pa_Y(X) and mu_I-- are not defined, making it difficult both to understand the meaning of the quantity R_a and to check the subsequent proofs. 
-Need intuition for second algorithm. Explanation of R_a
+Thank you to all reviewers for your detailed and thoughtful reviews
+
+We apologise for the confusing/missing definitions and lack of intuition for algorithm 2.
+We will certainly clarify this section.
+As you had very similar questions regarding this section, we will address them together here and then provide specific feedback on the remainder of the paper.
 
@@
 
-Reviewer 1 (causality side)
+Reviewer 1
 
-make clear which mathematical terms in section 3 & 4 are central
-many concepts, such as causal models (population-level), regret analysis (sample-level), interventions, actions, etc. are mixed in a way that is sometimes hard to follow.
-there should be more instructive but still theoretical ways than finite-sample regrets to argue why knowledge of the causal DAG helps.
-Great to not only have the hard-to-read finite-sample regret analysis, but also some more instructive considerations of when/how the causal approach helps.
+The advantages provided by incorporating causal structure into a bandit algorithm are that it allows us to constrain the initial search space (offline pruning only briefly mentioned) and extract information about the results of multiple actions at each timestep.
+Both these things allow an agent to agent to hone in on optimal actions more quickly.
+Estimating P(Y|X1...XN) and P(X1...XN) would require (without assumptions assumptions on the reward function) a number of estimators that grows exponentially with the number of variables.
+It is essentially this issue that techniques that assume additional structure (causal or otherwise) attempt to resolve.
+We will try to make this key point clearer.
 
-obvious that one can construct theoretical scenarios where causal knowledge or modeling helps for bandit-like scenarios
-this direction has already been investigated e.g. by Bottou and Bareinboim - although the scenarios differ.
-Motivate with real world problem. (+ how can farmer know DAG in advance)
- You make an interesting point in 453-455.
-From Sec.
-2 it is not quite clear: does an action consists of (1) a decision on which variable to intervene as well as (2) what value to set the chosen variable to?
-(It gets clear later that its both.) This seems pretty clear, line 210, not sure how we can improve it.
-Why don't you use the initial T/2 data points for the final estimation of expected rewards (Alg1)
-We do use it for those actions that can be sufficiently estimated from the observational data.
-We could combine the estimates for the remaining actions with those obtained via intervention, but this would slightly complicate the proof of the regret bound and only change the bound by a constant factor of at most 2.
+We focus on the simple regret as it was simplest to analyse, whilst still being relevant to practical problems, demonstrating interesting characteristics and allowing comparisons with the wider bandit literature.
+Simple regret is only interesting for finite time-scales as asymptotically any unbiased estimator will, by definition, suffer no regret.
 
- Wouldn't it make sense to also try to estimate the form of the mechanism p(Y|X_1, ..., X_N) because this together with the estimated distributions on X_i should give additional information on the expected rewards.
-The number of estimates this requires (without some form of assumption on the function) grows exponentially with the number of variables.
-It also does not provide any further information on the expected rewards if we accept the stated structural assumptions and set of allowed interventions.
-Unclear if the problem they address couldn't also be addressed by some adapted version of contextual bandits (which abstract away from the difference between (1) on which variable one intervenes and (2) to which value one sets this variable).
-I don't understand the 2nd half of this comment...
-Contextual bandits extend standard bandits by allowing there to be additional contextual information, C, available prior to an agent selecting an action, a,.
-The agent must then build estimates P(Y|a, C).
-In our setting the additional information is received after the agent selects their action.
-
-Why does the q_i appear on the r.h.s. of the greater-or-equal sing in l1162? I couldn't see that from Chernoff. (2) 
-This probably just needs a reference. 
-Can you write down more steps that show the last inequality in l1176?
-It's somehow clear that you can bound it by the integral, but it would be much nicer for the reader if he/she could just go through it and not spend so much time for reconstructing steps.
-Yes we will add more steps
+Although the idea of combining causality with bandit algorithms has been considered before, the way in which we connect causal graphs to bandit problems is (to our knowledge) novel and quite general.
+Certainly in most real applications there will be some uncertainty in the causal model and this is a very interesting (and challenging) topic for future research.
+However we believe this work to be an important first step and worthy of publication.
+We are sorry for the difficulty in following the proofs in the supplementary material.
+We will include more steps and references.
 
 Reviewer 2
-Section 4: The formula for p(Y | do(x2)) is incorrect, should be \sum_x1 p(x1) p(y| x1, x2). The functional better depend on x2!
-Yep, this is a typo. Reviewer is correct.
+Thank you for spotting the typos in eqn. (2) and the do() expression in section 4.
 
-“however we will never observe (X2 = 1, X1 = 0) so we cannot get a good estimate for P {Y |do(X2 = 1)}.” Not obvious, since \sum_x3 p(x3) p(y | x3, x2) is also equal to p(y |do(x2)).
-True. We could let X1=X2=X3 deterministically or just use the simpler graph X1 → X2 → Y, X1 → Y
-
-(2) should bound d from above by N, right? 
-Yep, something not quite right about (2)
+You are also correct with regard to estimating P(Y|do(X2)) via X3. We should let X1==X2==X3 or replace figure 2 with the simpler graph X1 -> X2 -> Y, X1 -> Y.
 
 Reviewer 4
-It is a pity that the second algorithm was only tested in a setting, where all interventional probabilities actually equal conditional probabilities (P(Y|X=x) = P(Y|do(X=x)) ).
-Thus, there is de facto no experiment in the paper that actually uses techniques specific to causal theory.
+Thank you for the detailed notes on typos. These will be fixed.
+
 Although trivial, it is causal theory that tells us the interventional and conditional probabilities are equal in the parallel bandit setting and algorithms 1 & 2 both use this causal structure to beat the standard algorithms.
@@
 
-Why do you compare your algorithms to only 1 single competitor algorithm? If no valid/strong reason is given, then please add curves for at least one or two other competitors.
 The experiments aim to demonstrate the causal algorithms escape the lower bounds that apply to standard simple regret algorithms.
-All other algorithms will be subject to these bounds and thus perform with similarly, however we can add additional curves.
+All other algorithms will be subject to these bounds and thus perform similarly, however we can add additional curves.

### 2016-04-22 TPR/TPR.tex
cleaned up structure of causality folder

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}

\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    punktrect/.style={
    rectangle, 
    rounded corners, 
    % fill=black!10,
    draw=black, thick,
    minimum height=1em, 
    text centered},
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,}
}

\author{Finnian Lattimore}
\title{Causal Inference in Machine Learning:Thesis Proposal Review}

\begin{document}
\def\ci{\perp\!\!\!\perp} % from Wikipedia
\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]
\section*{Causal Inference in Machine Learning}

Start with a focus on causal inference and then show how there is a relationship to specifically bandits.

Learning the outcome of an intervention is central to many real world problems. Does de-worming children in poor countries improve health and educational outcomes[worm-wars]? Would increasing the minimum wage benefit lead to higher unemployment? Will offering this customer a discount improve my revenue? These questions are difficult as they require more than identifying a pattern in data. Correlation is not causation. The key challenge is that intervening changes the state of the world and we only get to observe the outcome of the action we select.

Randomized controlled trials have long been seen as the gold standard in addressing such problems.

Selecting actions under uncertainty is central to many real world problems, from robotics to policy development.

Predicting the outcome of actions is central to many real world problems, from robotics to policy development. There are two key approaches to this problem in machine learning: reinforcement learning and causal inference.

Machine learning has made dramatic progress in identifying complex patterns in large datasets as has been demonstrated by recent work on machine translation [], image recognition[] and XXXX[]. However, causal inference poses an additional challenge. Correlation is not causation. Essentially, causal inference can be seen as the problem of inferring a property of the post interventional distribution from data collected prior to taking the intervention. This requires some form of assumption of the way in which the intervention changes the system.

Randomized controlled trials have long been considered the gold standard for inferring causality. However ...

Learning the outcome of actions is central to many real world problems from robotics to policy development.

In machine learning, there are two separate commonly applied approaches. The first is reinforcement learning, where an agent selects and action and then receives feedback on their performance. In general, the state of the system and feedback both depend on action selected. Causal inference aims to predict the outcome of an action from

Both approaches can be seen as extensions to the concept of randomized controlled trials. Causal inference makes use of assumptions to allow the outcome of interventions to be predicted from non-randomized data. Bandit algorithms take account of the sequential nature of many such decision making processes.

The similarities between the problems that these techniques have been developed to address raises the obvious question of if there are problems best addressed by a combination of these approaches and how they can be combined.

Importance of causal inference. 
What is causal inference

Causal inference makes use of assumptions to allow the outcome of interventions to be predicted from observational data.

Both of these approaches have historically been applied to the problem of determining the effectiveness of a medical treatment. Bandit algorithms deal with the sequential nature of the decision making process. Causal inference with the problem that full randomization is not always feasible.

Causal inference techniques arose from the recognition that randomized experimentation is not always feasible, affordable or considered ethical. People frequently dislike randomization in important decision making processes.

Bandit algorithms recognise the sequential nature of many desision making processes.

The difference between predictive and causal models is fundamental. In a predictive model, we care only about the accuracy of our predictions. we implicitly assume that any actions we take on the basis of our predictions do not influence the data presented to our model. For example, predicting which underground pipe will fail on the basis of geological, and pipe characteristics is predictive if the goal is targeting maintenance to minimize failures. On the other hand if the goal is to select new pipe materials, ...

The problem I always think about when thinking about causal inference is selection bias - that is a back door path from treatment to outcome via a hidden confounder.

Areas to explore (maybe throw up more of these in your talk and keep the TPR document more focussed purely on the reinforcement learning side of things).

	Practical applications of causal inference and discovery
		- practical applications of causal inference abound (usually not explicitly recognized as such in economics and social science)
		- very few practical applications or comparisons of causal discovery techniques on real data.

	Causality, Interpretability and Ethics
		-

	Relationship between causal inference and reinforcement learning
		- why bandits
		- how can structur
		- specific problem

	Feature selection for causal inference and discovery
		- learning the graph
		- learning given the graph. Given a graph, select which features to train on to build a causal model. 
		- can we infer anything about the causal structure from the distributions over the features (yes - causal discovery tells us we can)
		- first discover the graph(s) then do inference with each possible graph.
		- with a predictive model - features selection is to avoid over-fitting. You can create new combinations of features as products etc.
		This is finding a good representation. A way of breaking up the information available into variables in such a way as to allow a good model to be found. This is what deep networks are apparently doing. Is that right? Because they can effectively represent a very large number of different ways of combining whatever is initially chosen as a feature. So what is a good way to break up data into features for causal inference/discovery. We don't want to much entanglement. Come up with simple examples on generated data. How you take a feature and encode effects what space of models is explored. The 'variables' as they come in in the data are by no means the only description for that data. Do we encode something as a single variable with more levels or as two binary variables. Eg -

Question on randomized controlled trials - is there an implicit assumption about how many confounding variables there are - does this tie into how much power a study has for a fixed number of participants?

Predicting the outcome of actions is a critical goal in both artificial intelligence and for many

The goal of causal inference is to predict the outcome of an intervention or action from

Data is being collected on everything. Extract meaningful knowledge. Machine Learning.

\begin{itemize}
\item formalize questions, describe problems
\item literature survey demonstrating problem is not already solved
\item initial attempts to solve problem
\end{itemize}
\end{document}

### 2016-04-22 causal-bandit/icml-2015/tmp.tex
cleaned up structure of causality folder

--- 2016-02-05 causal-bandit/icml-2015/general.tex
@@
-\newcommand{\calP}{\mathcal P}
-\newcommand{\x}{\boldsymbol{x}}
-\newcommand{\Ps}{\operatorname{P}}
-
 We now consider the more general problem where the graph structure is known, but arbitrary.
@@
 a truncated importance weighted estimator.
-
-Let $a \in \calA$ be an intervention and define random variable
-\eq{
-Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
-}
-where $B_a \geq 0$ is some constant to be chosen subsequently and
-\todom{Symbol clash with regret}
-\eq{
-R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\,.
-}
-Abusing notation we define $m(\eta)$ by
-\eq{
-m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
-}
-We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
-
-\begin{algorithm}[H]
-\caption{General Algorithm}\label{alg:general}
-\begin{algorithmic}
-\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
-\FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $a_t = a$ with probability $\eta$
-\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
-\ENDFOR
-\STATE For each $a \in \calA$ compute an estimate of its return:
-\eq{
-\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
-}
-\STATE {\bf return} $\hat a^* = \argmax_a \hat \mu_a$
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:general}
-If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by
-\eq{
-B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.
-}
-Then for $C = \sqrt{2} + 4$ we have
-\eq{
-\mu^* - \EE[\mu_I] \leq C\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}\,.
-}
-\end{theorem}
-
-\begin{proof}
-First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
-We abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
-By definition we have $|Z_{at}| \leq B_a$ and 
-\eq{
-\Var_Q[Z_{at}] 
-\leq \EE_Q[Z_{at}^2] 
-\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] 
-\leq m(\eta)\,.
-}
-Checking the expectation we have
-\eq{
-\EE_Q[Z_{at}] 
-&= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] \\
-&= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] \\
-&= \mu_a - \beta_a\,,
-}
-where 
-\eq{
-0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}
-}
-is the negative bias. 
-The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
-\eq{
-\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
-}
-Let $\epsilon_a > 0$ be given by
-\eq{
-\epsilon = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
-}
-Then by the union bound and Bernstein's inequality 
-\eq{
-&\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \\ 
-&\qquad\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
-}
-Assuming this event does not occur and letting $a^* = \argmax_{a \in \calA} \mu_a$ we have
-\eq{
-\mu_I \geq \hat \mu_I - \epsilon_I  
-\geq \hat \mu_{a^*} - \epsilon_I  
-\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
-}
-By the definition of the truncation
-we have
-\eq{
-\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
-}
-and
-\eq{
-\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
-}
-Therefore for $C = \sqrt{2} + 4$ we have
-\eq{
-\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
-}
-as required.
-\end{proof}
-
-\subsection*{Choosing the Sampling Distribution}
-
-Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$. In light of Theorem \ref{thm:general}
-a natural choice of $\eta$ is the minimiser of $m(\eta)$.
-\eq{
-\eta^* 
-&= \argmin_\eta m(\eta) \\
-&= \argmin_\eta \max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
-&= \argmin_\eta \underbrace{\max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
-}
-Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
-Therefore the minimisation problem may be tackled using standard techniques from convex optimisation. 
-An interpretation of $m(\eta^*)$ is the minimum achievable worst-case variance of the importance weighted estimator.
-In the experimental section we present some special cases,
-but for now we give two simple results. The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
-
-\begin{proposition}\label{pro:m-bound}
-$m(\eta^*) \leq |\calA|$.
-\end{proposition}
-
-\begin{proof}
-By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|$. Then
-\eq{
-m(\eta) 
-&= \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
-&\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] \\
-&\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA| \qedhere
-}
-\end{proof}
-
-The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
-Recall in that setting that 
-\eq{
-\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
-}
-For $a = do()$ choose $\eta_a = 1/2$. 
-\todom{Check please!}
-Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
-Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
-It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.
-
-\begin{remark}
-The algorithm in the previous section for the parallel bandit problem did not assume knowledge of the conditional distributions on $X$ and 
-instead only the graph structure. Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
-Finally, we believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
-for future work.
-\end{remark}

### 2016-04-22 figures/graph_diagram_assumed_structure.tex
cleaned up structure of causality folder

--- 2015-10-07 causal-bandit/observe_then_best.tex
@@
 \documentclass{article}
+\usepackage{tikz} 
+\usepackage[utf8]{inputenc}
+\usepackage{amsmath}
+\usepackage{listings}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{tabularx}
+\usepackage{enumitem}
+\usepackage{algorithm}% http://ctan.org/pkg/algorithm
+\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage[margin=0.7in]{geometry}
+\usepackage{tikz}
 
-\usepackage{todonotes}
+\usepackage{graphicx}
+\usetikzlibrary{arrows,positioning} 
+\usepackage{subcaption}
+\thispagestyle{empty}
+\usepackage{multicol,caption}
+\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
 
-%\usepackage[disable]{todonotes}
-\newcommand{\todot}[2][]{\todo[color=red!20!white,#1]{#2}}
-\newcommand{\todof}[2][]{\todo[color=orange!20!white,#1]{#2}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PACKAGES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage{times}
-\usepackage{fullpage}
-\usepackage{latexsym}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage[boxed]{algorithm}
-\usepackage{algpseudocode}
-\usepackage{mathtools}
-\usepackage{accents}
-\usepackage{tikz}
-\usepackage{pgfplots}
-\usepackage{dsfont}
-\usepackage[bf]{caption}
-\usepackage{hyperref}
-\hypersetup{
-    bookmarks=true,         % show bookmarks bar?
-    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
-    pdftoolbar=true,        % show AcrobatÕs toolbar?
-    pdfmenubar=true,        % show AcrobatÕs menu?
-    pdffitwindow=false,     % window fit to page when opened
-    pdfstartview={FitH},    % fits the width of the page to the window
-    pdftitle={My title},    % title
-    pdfauthor={Author},     % author
-    pdfsubject={Subject},   % subject of the document
-    pdfcreator={Creator},   % creator of the document
-    pdfproducer={Producer}, % producer of the document
-    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
-    pdfnewwindow=true,      % links in new window
-    colorlinks=true,       % false: boxed links; true: colored links
-    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
-    citecolor=blue,        % color of links to bibliography
-    filecolor=magenta,      % color of file links
-    urlcolor=cyan           % color of external links
-}
-\usepackage{comment}
-\usepackage{amsthm}
-\usepackage{natbib}
-\usepackage[capitalize]{cleveref}
-\usepackage{graphicx}
-\usepackage{parskip}
-\usepackage{tikz} 
-\usetikzlibrary{arrows,positioning} 
-\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
 \DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MACROS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\defined}{\vcentcolon =}
-\newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}{\mathbb E}
-\newcommand{\Var}{\operatorname{Var}}
-\newcommand{\calF}{\mathcal F}
-\newcommand{\sr}[1]{\stackrel{#1}}
-\newcommand{\set}[1]{\left\{#1\right\}}
-\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
-\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
-\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
-\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
-\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
-\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
-\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
-\newcommand{\Ber}{\operatorname{Bernoulli}}
-\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
 
@@
            draw=black, thick,
-           minimum width=2.5em,
-           minimum height=2.5em,
+           minimum width=2.2em,
+           minimum height=2.2em,
            font=\footnotesize,
            text centered,
-           fill=blue!20!white},
+           },
      latent/.style={
@@
     shading angle=45 }
+
 }
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% THEOREMS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-
 \begin{document}
-\def\ci{\perp\!\!\!\perp}
-
-\begin{comment}
-\section{Random notes}
-Information directed sampling have developed a general framework to construct algorithms that leverege generic structure between arms to minimize Bayesian regret, by explicitly trading off expected regret and information gain at each timestep.
-
-What can be expressed in the partial monitoring framework that cannot be defined by the feedback graph?
-
-We consider a variant of the stochastic multi-armed bandit problem when we have prior knowledge of the causal structure, but not the functional form, of the relationship between variables.
-
-sequential decision process,exploration-exploitation trade off
-
-Points to present. Why we use regret against single best arm - it makes sense in the stochastic setting. 
-The difference between observational and interventional distributions.
-
-terminology for intervention vs observation.
-Recast bandit algorithm in terms of this.
-Show limitations to existing approaches by pointing to places were existing algorithms eg UCB/explore-exploit/exp3 require unbiased estimates.
-Are unbiased estimates actually impossible in our setting?
-
-This paper is about sequentially choosing actions. 
-Modelling sequntial decition processes.
-
-In the classic multi-armed bandit setting, an ag
-
-General bandit part ...
-define multi-armed bandit, and importance.
-Decision making is fundamental.
-Goal is to select an action to maximize a reward (or minimize a loss).
-In many settings, such as drug trials, online ad placement, etc we can choose actions sequentially.
-This introduces an exploration-exploitation trade-off.
-
-The problem of sequentially choosing actions to minimize some loss wide range of applications in XXXX, YYYY. The multi-armed bandit setting, in which .... .
-
-Now the required causal stuff should go here.
-
-Causal structure can be represented as a directed acyclic graph with nodes representing the variables and an edge from $A \rightarrow B$ meaning $A$ causes $B$.
-Deterministic interventions, consisting of setting variables to some value, are modelled by setting the nodes to the desired values and removing any links coming into them, as their value is now no longer set by other variables in the graph according to the original mechanism.
-
-If the causal graph is known, such that there are no hidden variables causing 2 or more of the variables and the probabilities are all > 0 then the distribution of any variable after any intervention can be inferred (asymptotically) from data collected in absence of any interventions.
-\end{comment}
-
-\section*{Introduction}
-
-Problems requiring choosing an action under uncertainty are rife in all areas of human endeavour.
-For many problems, actions may be chosen sequentially, allowing the agent to learn from the outcome of early choices to improve later ones.
-A widely used framework for sequential decision making is the multi-armed bandit.
-In the classic multi-armed bandit setting there is a finite set of available actions, each associated with a distribution over rewards which is unknown but stationary and independent of the reward distribution of other actions.
-At each timestep the agent selects an action and receives a reward sampled iid from the corresponding reward distribution.
-
-An an alternate approach to selecting actions is causal inference.
-Frameworks for causal inference provide a mechanism to specify assumptions that allow observational distributions over variables to be mapped to interventional ones.
-This allows an agent to predict the outcome of an action based on non-experimental data.
-This approach is common in social science, demography, and economics where explicit experimentation may be difficult.
-For example, predicting the effect of changes to childcare subsidies on workforce participation or school choice on student grades.
-
-We take a first step towards unifying these approaches by considering a variant of the stochastic multi-armed bandit problem where we have prior knowledge of the causal structure governing the available actions.
-This structure creates dependencies between the rewards of different arms such that selecting one action can provide information on the reward for other actions.
-
-There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures.
-Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc.
-There is also an active line of research into bandits with feedback defined by a graph.
-Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action [].
-The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.
-
-\section*{Problem Formulation}
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
-As an example, consider a farmer wishing to optimize the yield of her crop.
-She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
-We assume only a single intervention is feasible due to cost and that each of these variables are independent of one-another (this may not always be the case - temperature could be related to rainfall for example).
-After having selected which variable to control, she plants her crops and observes the values of the remaining input variables and the yield.
-This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
-
-\begin{figure}[h]
-\centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\end{figure}
+\def\ci{\perp\!\!\!\perp} % from Wikipedia
 
@@
 
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[r(X)|do(X_i = j)]$. The optimal reward and action are denoted $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
-}
+\begin{figure}[h]
+\centering
+\label{fig:causalStructure}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty}]
 
-The problem can be treated as a classical multi-armed bandit with $K = 2N$ arms. However, this does not utilize the information provided by the causal assumption.
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node,fill=black!20!white, below right=of 2](5){Y};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5);
 
-Now need to expand upon how the assumption gives us extra information.
-
-The difficulties faced due to bias.
-
-To deal with this issue we use a simple explore-exploit algorithm.
-Our algorithm will explore for $h$ time-steps, sampling actions in a way that depends on our prior knowledge of $q$ but is independent of the observed rewards.
-We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
-
-\section*{Results}
-
-Summarize results here and note differences to classic bandit results
-
-\subsection*{Known and Balanced q}
-
-We begin with the simplest case were we assume that $q_i=\frac{1}{2} \forall i$.
-During the exploration phase we sample actions uniformly at random.
-In this case, this is equivalent to purely observing, that is taking no action and allowing all input variables to take their value randomly as  $X_{t,i} \sim \Ber(\frac{1}{2})$.
-
-We will have $n_i \sim Binomial(h,\frac{1}{2})$ observations for each arm $i$ at the end of the exploration stage. Note that this is independent of the number of arms $K$.
-
-Assume we have $K$ bernoulli arms with means ordered from highest to lowest $\mu_1 ... \mu_K$. Let $\Delta = [\Delta_1...\Delta_K]$ be the differences from the optimal reward $\mu_1$.
-
-\subsubsection*{Regret during explore phase}
-Since the probability we play each arm is constant and uniform during the exploration phase, the expected regret is simply proportional to the average sub-optimality $\Delta$.
-\eqn{
-R_1 = h\sum_i P(i)\Delta_i = \frac{h}{K}\sum_i \Delta_i = h E[\Delta] < h
-}
-
-\subsubsection*{Regret during exploit phase}
-The regret during this phase is proportional to the expected sub-optimality of the arm with the highest empirical mean at the end of the explore phase.
-
-\eqn{
-\hat{i^*} = argmax_i [\hat{\mu}_i]
-}
-\eqn{
-R_2 = (T-h)E[\Delta_{\hat{i^*}}] = (T-h)\sum_i P(\hat{\mu}_i \geq \hat{\mu}_j \forall j)\Delta_i 
-}
-
-The difficulty with this approach is that it is hard to get bounds that are tight for all $\Delta$.
-Instead, we will bound the probability that we select an arm with a sub-optimality gap greater than some $D$.
-\eqn{
-R_2 \leq (T-h)\left(P(\Delta_{\hat{i^*}} \leq D) D+P(\Delta_{\hat{i^*}} > D)  \Delta_{max} \right)
-}
-The goal now is to get a bound for $P(\Delta_{\hat{i^*}} > D)$ in terms of Hoeffdings type bounds for each arm.
-
-Suppose $i = \hat{i^*} \implies \hat{\mu}_i > \hat{\mu}_1$.
-If we haven't over-estimated $\mu_i$ too much, $\hat{\mu}_i - \mu_{i} < \frac{D}{2}$, and haven't under-estimated $\mu_1$ too much, $\mu_1 - \hat{\mu_1} < \frac{D}{2}$, then $\Delta_{\hat{i^*}} = \mu_1 - \mu_i < D$
-
-\eqn{
-\label{eqn:probDeltaToLarge}
-P(\Delta_{\hat{i^*}} > D) \leq  P(\mu_1 - \hat{\mu_1} > \frac{D}{2})+ \sum_{i=2}^K P(\hat{\mu}_i - \mu_{i} > \frac{D}{2})
-}
-
-If we used the empirical mean as an estimator for $\mu_i$, the bound will depend on the number of times we actually observed each arm, which will be a random variable drawn from a multinomial distribution.
-Instead we will use an importance weighted estimator.
-
-\eqn{
-\label{eqn:importance_weighted_estimator}
-\hat{\mu}_i = \frac{1}{h}\sum_{t=1}^h \frac{Y_t\ind{\text{arm $i$ active}}}{q_i}
-}
-
-where $q_i = P(\text{arm $i$ active})$
-
-Hoeffdings gives $ P(\hat{\mu}_i - \mu_{i} > \epsilon) \leq e^{-2h\epsilon^2q_i^2}$. In this case we have assumed $q_i = \frac{1}{2} \forall i$. Putting this into equation \ref{eqn:probDeltaToLarge}:
-
-\eqn{
-\label{eqn:balancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/8}
-}
-
-\eqn{
-R_2 \leq (T-h)[(1-K e^{-hD^2/8})D + K e^{-hD^2/8}] < (T-h)[D + K e^{-hD^2/8}]
-}
-
-Let $D = \sqrt{\frac{8}{h}\log(hK)}$
-
-\eqn{
-R_2 \leq (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)
-}
-
-\subsection*{Total Regret}
-
-Putting together the regret from the exploration and exploitation phases,
-
-\eqn{
-R_T & \leq \frac{h}{K}\sum_i \Delta_i + (T-h)\left(\sqrt{\frac{8}{h}\log(hK)} + \frac{1}{h}\right)\\
-& \leq h + T\left(\sqrt{\frac{8}{h}\log(TK)} + \frac{1}{h}\right)
-}
-
-Now if we let $h = T^{2/3}(\log(KT))^{1/3}$,
-
-\eqn {
-R_T \leq 4T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}} + T^{\frac{1}{3}}(log(KT))^{-\frac{1}{3}}
-}
-
-If $T \geq 2$ and $K \geq 2$, the first term dominates and,
-
-\eqn {
-R_T  \leq 5T^{\frac{2}{3}}(log(KT))^{\frac{1}{3}}
-}
-
-The distribution independent lower bound for optimised UCB is $O(\sqrt{TK})$ (see Bubeck sect 2.4.3) so we would expect our algorithm to do better if $K >> T^{\frac{1}{3}}$
-
-\subsection*{Empirical results}
-\begin{figure}[H]
-\centering
-\caption{Comparison of the UCB and causal-explore-exploit for K=20 and T=10000. Note, $K \sim T^{1/3}$ Plot shows average and standard deviation over 10000 trials.}
-\includegraphics[width=.5\textwidth]{explore_exploit}
+\end{tikzpicture}
 \end{figure}
 
-\section{Generalizing to unbalanced $\boldsymbol{q}$}
-
-When some arms have low natural probability we cannot rely on exploring them adequately by pure observation. We need to explicitly play them during the exploration phase.
-
-We now have an additional trade off to make, which is how much should be observe (learning something about at least half the arms each timestep) versus playing the low probability arms.
-
-Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ...
-\leq q_N$.
-Let $m \in [2,N] = \{m:q_m > \frac{1}{m}\}$ Ie if the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
-If the problem is completely unbalanced, $q_1...q_N = 0$ then $m=N$.
-Let $q_{ij} = \P{X_i = j}$
-
-Suppose we observe for the first $h/2$ timesteps. This is at worst half the optimal.
-
-We then have estimates
-
-\eqn {
-\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
-}
-
-We take this as our estimate for those arms for which $q_{ij} > \frac{1}{m}$
-
-Let $Z_{t,ij} = \ind{Y_t=1,X_{t,i}=j} \sim Bernoulli(q_{ij}\mu_{ij})$,
-
-Chernoff's inequality gives
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
-}
-
-We play each of the remaining $m$ arms $\frac{h}{2m}$ times so for them we get
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/4m}
-}
-
-So for all the arms
-
-\eqn{
-P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}) \leq e^{-hD^2/24m}
-}
-
-and
-
-\eqn{
-\label{eqn:unbalancedHoeffdings}
-P(\Delta_{\hat{i^*}} > D) \leq Ke^{-hD^2/24m}
-}
-
-If we let $D = \sqrt{\frac{24m\log(hK)}{h}}$
-
-\eqn{
-R_T &  \leq h + T\left(\sqrt{\frac{24m\log(hK)}{h}} + \frac{1}{h}\right) \\
-& \leq h + T\left(\sqrt{\frac{24m\log(TK)}{h}} + \frac{1}{h}\right) 
-}
-
-Let $h = T^{2/3}m^{1/3}\log(TK)^{1/3}$
-
-\eqn {
-R_T & \leq 6T^{2/3}m^{1/3}log(KT)^{1/3} + T^{1/3}m^{-1/3}log(KT)^{-1/3}\\ 
-& \leq 7T^{2/3}m^{1/3}log(KT)^{1/3}
-}
-
-Again comparing to the UCB bound of $\sqrt{KT}$, we expect the causal algorithm to do better if $K > m^{2/3}T^{1/3}$, or alternatively if $m \leq \frac{K^{3/2}}{\sqrt{T}}$
-
-To get results that degrade to similar order bounds as UCB when the arms are very unbalanced, I will need to drop the explore/exploit strategy.
-
-\subsection{Generalizing to unknown q}
-We now consider the case where $\boldsymbol{q}$, and thus $m$ is not known in advance. We begin by considering the simple regret
-
-\subsubsection{Bounding the Simple Regret}
-
-\begin{itemize}
-\item $N$ is number of variables.
-\item $q_i$ is probability that $X_{i,t} = 1$ for any $t$
-\item $s_i = \min\set{q_i, 1 - q_i}$
-\item $\mu : \set{1,\ldots,N} \times \set{0,1} \to [0,1]$.
-\item $I_t \in \set{1,\ldots,N} \times \set{0,1}$ is the intervention
-\end{itemize}
-
-Let $R_s(h)$ be the simple regret after $h$ time steps, which is
-\eq{
-R_s(h) = \E\left[\mu^* - \mu_{I_h}\right]\,,
-}
-where $\mu^*$ is the mean payoff of the optimal intervention and $I_h$ is the intervention chosen in the $h$th round.
-We propose a two part algorithm that is provably order optimal when its performance is measured in terms of the simple regret.
-
-\begin{algorithm}[H]
-\caption{Simple Regret Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\State {\bf Input:} $h, N$
-\For{$t \in 1,\ldots,h/2$}
-\State Do nothing and observe $X_{i,t}$ for $i \in \set{1,\ldots,N}$ and $r_t$
-\EndFor
-\State Compute for all $i \in \set{1,\ldots,N}$ and $j \in \set{0,1}$:
-\eq{
-\hat \mu_{i,j} = \frac{\sum_{t=1}^{h/2} \ind{X_{i,t} = j} r_t}{\sum_{t=1}^{h/2} \ind{X_{i,t} = j}}\,.
-}
-\State Compute $\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} X_{i,t}$
-\State Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
-\State Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
-\State Compute $\hat m = \min\set{2 \leq i \leq N : \hat s'_{i} \geq 1/i}$
-\State $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
-\State Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$
-
-\For{$(i,j) \in A$}
-\For{$t \in 1,\ldots,h/2\hat m$}
-\State Choose action $X_{i,t} = j$
-\State Observe reward $r_t(X_{i,t}=j)$
-\EndFor
-\State Recompute $\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{ h/2\hat m} r_t(X_{i,t}=j)$ 
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\begin{theorem}\label{thm:simple-regret}
-Define $m = \min\set{2 \leq i \leq N : q_i \geq 1/i}$.
-Then the algorithm given in \cref{alg:simple} satisfies
-\eq{
-R_s(h) \in O\left(\sqrt{\frac{m}{h} \log \left(\frac{NT}{m}\right)}\right)\,.
-}
-\end{theorem}
-
-\begin{lemma}\label{lem:conc1}
-$\displaystyle \P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{h} \log \frac{2}{\delta}}} \leq \delta$.
-\end{lemma}
-
-\begin{proof}
-Let $Z_t = \ind{X_{i,t} = 1} \in \set{0,1}$.
-Then
-\eq{
-\hat q_i = \frac{2}{h} \sum_{t=1}^{h/2} Z_t\,.
-}
-Now $Z_1,\ldots,Z_{h/2}$ is an i.i.d.\ sequence of Bernoulli random variables with with mean $q_i$. The result follows from the Chernoff bound.
-\end{proof}
-
-\begin{lemma}\label{lem:conc2}
-Let $\delta >0$ and $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$ and
-define $\hat m = \min\set{2 \leq i \leq N : \hat s_{(i)} \geq 1/i}$.
-If
-\eq{
-h \geq 24m \log\frac{4N}{\delta}\,,
-}
-then
-\eq{
-\P{\hat m \leq 2m} \geq 1 - \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-It is easy to see that the worst case occurs when 
-\eq{
-q_i = \begin{cases}
-0 & \text{if } i < m \\
-\frac{1}{m} & \text{otherwise}\,.
-\end{cases}
-}
-Now by \cref{lem:conc1} we have with probability at least $1 - \delta$ that 
-\eq{
-(\forall i) \qquad \left| \hat q_i - q_i\right| 
-&\leq \sqrt{\frac{6q_i}{h} \log \frac{2N}{\delta}} \\
-&= \sqrt{\frac{6}{mh} \log\frac{2N}{\delta}}\,. 
-}
-By assumption 
-\eq{
-h \geq 24m \log \frac{4N}{\delta}\,.
-}
-Therefore with probability at least $1 - \delta$ we have
-\eq{
-(\forall i) \qquad \left|\hat q_i - q_i\right| \leq \frac{1}{2m}\,.
-}
-Note that for $i$ with $q_i = 0$ it is easy to see that $\hat q_i = 0$ (the variance is zero). The above guarantee that
-for $i$ with $q_i = 1/m$ we have $\hat q_i \in [1/(2m), 3/(2m)]$ for all $i$ with probability at least $1 - \delta$,
-which implies that $\hat s_i \geq 1/(2m)$ for all $i$ where $q_i = 1/m$.
-Therefore $\hat m \leq 2m$ with probability at least $1 - \delta$. 
-\end{proof}
-
-\begin{proof}[Proof of \cref{thm:simple-regret}]
-
-for $(i,j)\in A$, the algorithm explicitly selects the action, $X_i = j$,  $\frac{h}{2\hat m}$ times.
-\eq {
-\hat \mu_{i,j} = \frac{2\hat m}{h} \sum_{t=1}^{h/2\hat m} r_t(X_i = j)
-}
-
-Via Hoeffding's Inequality
-
-\eq {
-\P{ \left|\hat \mu_{i,j} - \mu_{i,j}\right| > \epsilon } \leq 2\exp{-\frac{h\epsilon^2}{\hat{m}}}
-}
-
-for $(i,j)\notin A$, the algorithm has observed the reward given $X_i = j$ at least $\frac{h}{2\hat m}$ times.
-
-\eq {
-(i,j)\notin A & \implies \hat{s}_i \geq \frac{1}{\hat m} \\
-& \implies \sum_{t=1}^h \ind{X_i = j} \geq \frac{h}{\hat{m}}
-}
-
-Let  $Z_{ij} = \sum_{t=1}^{h/2} \ind{X_i = j}$ and $t'_1 ... t'_{Z_{ij}} = {t:X_{i,t} = j}$
-
-\eq {
-\hat \mu_{i,j} = \frac{1}{Z_{ij}}\sum_{t'=1}^{Z_{ij}}r_{t'}
-}
-
-\eq {
-\P{\left| \hat \mu_{i,j} - \mu_{ij} \right| > \epsilon} = & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{Z_{ij}} \sum_{t'=1}^{Z_{ij}}r_{t'} - \mu_{ij} \right| > \epsilon | Z_{ij} = z} \\
-= & \sum_{z=1}^\infty \P{Z_{ij}=z}\P{\left| \frac{1}{z} \sum_{t'=1}^{z}r_{t'} - \mu_{ij} \right| > \epsilon } \\
-\leq & \P{\left| \frac{2\hat m}{h} \sum_{t'=1}^{h/2\hat m}r_{t'} - \mu_{ij} \right| > \epsilon } \sum_{z=1}^\infty \P{Z_{ij}=z} \\
-\leq & 2\exp{-\frac{h\epsilon^2}{\hat{m}}}\\
-}
-
-Applying the union bound over all $2N$ actions.
-\eq{
-& \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \epsilon} \leq 4N\exp{-\frac{h\epsilon^2}{\hat{m}}} \\
-\implies & \P{\exists (i,j): \left| \hat \mu_{i,j} - \mu_{i,j} \right| > \sqrt{\frac{\hat m}{h}\log{\frac{4N}{\delta}}}} \leq \delta \\
-}
-
-Now by \cref{lem:conc2}, 
-\eq{
-h \geq 24m \log \frac{4N}{\delta} \implies \P{\hat m > 2m} \leq \delta
-}
-
-Therefore if $h \geq 24m \log \frac{4N}{\delta}$, we have with probability at least $1 - 2\delta$ that
-
-\eqn{
-\label{eqn:joint_bound_estimates_m}
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{2m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
-}
-Suppose $h < 24m \log \frac{4N}{\delta}$. Then
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq 1 \leq \sqrt{\frac{2m}{h} \log \frac{4N}{\delta}}\,.
-}
-Therefore
-\eq{
-\left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \qquad \forall h\ \text{ with probability at least } 1-2\delta
-}
-Therefore 
-\eq{
-R_s(h) 
-&\leq 2\delta + \sqrt{\frac{24m}{h} \log \frac{4N}{\delta}} \\
-&\leq \frac{8m}{h} + \sqrt{\frac{24m}{h} \log \left(\frac{Nh}{m}\right)}
-}
-as required.
-\end{proof}
-
-\subsubsection{Bounding the Bandit Regret}
-
-\begin{itemize}
-\item Use $24Nlog(4N/\delta)$ samples to estimate $m$
-\item If $\hat m > \frac{N^{3/2}}{\sqrt{T}}$ then stop and do UCB with remaining rounds.
-\item Else use causal explore-exploit and let the exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$ 
-\end{itemize}
-
-Since we don't know $\boldsymbol{q}$, and thus $m$, we can't set the exploration time $h$ in advance based on $m$ as we did for the known $\boldsymbol{q}$ case.
-Instead we first use $24Nlog(4N/\delta)$ samples to estimate $m$.
-If $\hat m < \frac{N^{3/2}}{\sqrt{T}}$ we will continue with the causal algorithm and let the total exploration time $h = T^{2/3}\hat{m}^{1/3}log(TK)^{1/3}$.
-
-Since $m \leq N$, by \cref{eqn:joint_bound_estimates_m}, we have that with probability at least $1-2\delta$,
-
-\eqn{
-\label{eqn:bounds_hold}
-(\forall i, j) \qquad \left|\hat \mu_{i,j} - \mu_{i,j}\right| \leq \sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}}\, \text { and } \hat{m} \leq 2m
-}
-
-So the regret in this case is bounded by,
-
-\eq {
-R(T) & =  2\delta E\left[R(T)|\text{ not } \cref{eqn:bounds_hold}\right] + (1-2\delta)E[R(T)| \cref{eqn:bounds_hold}]  \\
-& \leq 2\delta T + \left(h + T\sqrt{\frac{\hat m}{h} \log\frac{4N}{\delta}} \right)\\
- &  \leq \frac{4 \hat m^{1/3}T}{h}  + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(\frac{2Nh}{\hat m^{1/3}}\right)}\right), \text { letting } \delta = \frac{2 \hat m^{1/3}}{h} \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + h + T \left( \sqrt{\frac{\hat m}{h} \log \left(TK\right)}\right) \\
-&  \leq 4\left(\frac{T}{log(TK)}\right)^{1/3} + 2T^{2/3}(2m)^{1/3}log(TK)^{1/3}  \\
-&  \leq 8T^{2/3}m^{1/3}log(TK)^{1/3}  \\
-}
-
-Now make sure I have all the cases right - then have a go with an implementation.
-
-\subsubsection{ Older Attempts}
-We now consider the case where $\boldsymbol{q}$ is not known in advance. Assume as before that we observe for $h/2$ timesteps. From the observations gained in this phase we estimate $\boldsymbol{q}$.
-
-\eqn{
-\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}
-}
-
-Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ and construct $\bar{\boldsymbol{q}}$ such that $\bar{q}_1 \leq \bar{q}_2 \leq ...\leq \bar{q}_N$.
-We then estimate $m$ with $\hat{m} = min_i : \bar{q}_i \geq \frac{1}{i}$
-
-We estimate the reward for the apparently common arms, $i \geq \hat{m}$, as:
-
-\eqn{
-\hat{\mu}_i = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\frac{h}{2}\hat{q}_i} = \frac{\sum_{t=1}^{h/2}\ind{Y = 1,X_i = 1}}{\sum_{t=1}^{h/2}\ind{X_i=1}}
-}
-
-We play the remaining arms $\frac{h}{2\hat{m}}$ times and estimate their reward as:
-
-\eqn{
-\hat{\mu}_i = \frac{2\hat{m}}{h}\sum_{t=1}^{h/2\hat{m}}\ind{Y = 1|X_i=1}
-}
-
-We now consider how errors in the estimation of $m$ effect our estimates of the arm rewards.
-For the arms with $i \geq \hat{m}$ we know $\hat{q}_i \geq \frac{1}{\hat{m}}$ and thus our estimate is based on at least $\frac{h}{2\hat{m}}$ observations.
-Similarly we explicitly play the infrequently observed arms, $i < m$,  $\frac{h}{2\hat{m}}$ times.
-Thus our estimates will only be worse than the known $\boldsymbol{q}$ case if $\hat{m} > m$.
-
-For a fixed $m$ the $\boldsymbol{q}$ most likely to lead us to overestimate $m$ is: \todot{Justify this claim.}
-
-\eqn{
-q_i = \begin{cases}
-0 & \text{ if } i < m\\
-\frac{1}{m} & \text{ if } i \geq m
-\end{cases}
-}
-
-We now bound $P(\hat{m} - m > \varphi)$ for this worst case.
-
-For $i < m$,we have $\bar{q}_i = q_i = 0$. For $i \geq m$ \todot{Is this really true, justify}
-\eq{
-P(q_i - \bar{q}_i \geq C) & =  P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})\\
-& \leq  2P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})
-}
-
-Via Bernstein's Inequality:
-\eqn{
-&2P(q_i - \hat{q}_i \geq C | \hat{q}_i \leq \frac{1}{2}) \leq 2\exp(-\frac{hC^2}{4q_i}) := \gamma \\
-\implies &P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(2/\gamma)}{mh}})\leq \gamma
-}
-
-Define $\hat{m}_i = \frac{1}{\hat{q}_i}$.
-
-\eqn{
-\hat{q}_i \geq q_i - C & \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
-& \implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}\\
-& \implies P(\hat{m}_i - m \geq \frac{m^2C}{1-mC}) \leq \gamma
-}
-
-Let $\varphi = \frac{m^2C}{1-mC} \implies C = \frac{\varphi}{m(\varphi+m)} \implies \gamma = 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2}) \implies \varphi = \frac{2m\sqrt{m\log(2/\gamma)}}{\sqrt{h}-2\sqrt{m \log(2/\gamma)}} $
-
-Note this implies that if $h \geq 16m\log(2/\gamma)$ then $\varphi \leq m$
-
-\eqn {
-P(\hat{m}_i - m \geq \varphi) \leq 2\exp(-\frac{h \varphi^2}{4m(\varphi+m)^2})
-}
-
-If $\hat{m}_i - m \geq \varphi$ for at most $\varphi$ of the variables $i \geq m$, then $\hat{m} - m \leq \varphi$
-
-Let $W_i = \ind{\hat{m_i}-m \geq \varphi}$, $E[W_i] \leq \gamma$,$V[W_i] \leq \gamma$ 
-\eqn{
-P(\sum_{i=m}^N W_i \geq (N-m)\gamma + \varepsilon) \leq \exp(-\frac{\varepsilon^2}{2(N-m)\gamma+2\varepsilon/3})
-}
-Letting $(N-m)\gamma + \varepsilon = \varphi \implies \varepsilon = \varphi - (N-m)\gamma$
-\eqn{
-P(\hat{m}-m \geq \varphi) =  P(\sum_{i=m}^N W_i \geq \varphi) &\leq \exp({-\frac{(\varphi-(N-m)\gamma)^2}{2(N-m)\gamma + 2(\varphi-(N-m)\gamma)/3}})\\
-& \leq \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \text{, where } N\gamma < \varphi
-}
-
-If $P(\hat{m}-m \geq \varphi) \leq \zeta$.
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq & P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \leq \varphi )P(\hat{m}-m \leq \varphi) \\
-& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \varphi)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
-}
-
-Putting it together
-
-\eqn {
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \\
-& \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}))^2}{2N\exp(-\frac{h\varphi^2}{4m(\varphi+m)^2}) + 2\varphi/3}})
-}
-
-Assume $h > 16m\log(2N)$ and . Let $\varphi = \epsilon\sqrt{h}$. Assume $\epsilon \geq \frac{m}{\sqrt{h}}$ so as to ensure $\varphi > m$
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon)& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-N\exp(-\frac{h}{16m}))^2}{2N\exp(-\frac{h}{16m}) + 2\varphi/3}})\\
-& \leq \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{(\varphi-\varphi/2)^2}{\varphi + 2\varphi/3}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\varphi}{20}})\\
-& = \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) + \exp({-\frac{3\epsilon \sqrt{h}}{20}}) := \delta\\
-& \leq \begin{cases}
-2 \exp({-\frac{h\epsilon^2}{m+\epsilon\sqrt{h}}}) & \text{ if } h < \frac{9m^2}{289\epsilon^2} \\
-2\exp({-\frac{3\epsilon\sqrt{h}}{20})} & \text{ otherwise } 
-\end{cases}
-}
-
-\eqn{
-\implies \epsilon \leq \begin{cases}
-\frac{1}{\sqrt{h}}\log(2/\delta) + \sqrt{\frac{m}{h}\log(2/\delta)} & \text{ if } h < \frac{9m^2}{289\epsilon^2}\\
-\frac{20}{3\sqrt{h}}\log(2/\delta) & \text{ otherwise }
-\end{cases}\\
-}
-
-\eqn{
-\implies \epsilon \leq \frac{20}{3\sqrt{h}}\log(2/\delta)+ \sqrt{\frac{m}{h}\log(2/\delta)} 
-}
-
-Does this fit with the assumption we made about $\epsilon$???
-
-\pagebreak
-
-The first term will dominate if
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3} \geq \frac{h\epsilon^2}{m+\varphi}
-}
-
-Setting them equal and solving for $\varphi$. Roughly,
-
-\eqn{
-\varphi = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = \frac{1}{2}(\sqrt{4\epsilon^2 h+ m^2} - m)\\
-\implies \epsilon = \sqrt{\frac{\log(1/\delta)(\log(1/\delta)+m)}{h}} \leq \frac{\log(1/\delta)}{\sqrt{h}} + \sqrt{\frac{m}{h}\log(1/\delta)}
-}
-
-Now repeating the above but more exactly ...
-
-\eqn{
-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}  \geq & \frac{\varphi^2-2\varphi N\gamma}{4\varphi/3} \; \text{, if } \varphi > 3N\gamma \\
-= & \frac{3}{4}(\varphi - 2N\gamma)
-}
-
-\eqn{
-\frac{3}{4}(\varphi - 2N\gamma) = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = &\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}\\
-\leq & \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}
-}
-
-If we let $\varphi =\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2} $
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp(-\frac{3}{4}(\varphi - 2N\gamma))\\
-\implies \epsilon = \sqrt{\frac{4\log^2(2/\delta)+(3m+6N\gamma)\log(2/\delta)}{3h}}\\
-\leq \frac{2\log(2/\delta)}{\sqrt{3h}}+\sqrt{\frac{m+2N\gamma}{h}\log(2/\delta)}
-}
-Ok - but I still need to ensure there exists a $\gamma$ such that  $N\gamma < \frac{\varphi}{3}$ and $P(\hat{m}_i - m > \varphi) \leq \gamma$
-
-\eqn{
-P(\hat{m}_i - m > a) \leq \exp(-\frac{h a^2}{4m(m+a)^2})
-}
-
-It feels like I have just gone round and round in circles and pushed the problem to somewhere else ... my expression for $\varphi$ now contains both $\epsilon$ and $\gamma$ ...
-
-\eqn{
-\varphi \geq \frac{1}{2}\sqrt{m^2+4N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}
-}
-
-\pagebreak
-
-\eqn{
- \implies & C = \frac{\varphi}{m(\varphi+m)}  \geq \frac{\varphi}{\alpha m^2} \text{ if } \varphi < (\alpha-1)m\\
-\implies & P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4\log(1/\gamma)}{mh}} \right)\right) \leq \gamma
-}
-
-If we let $\gamma = \frac{e^{-h/m}}{N}$
-
-\eqn{
-P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{\log N}{h}+\frac{1}{m})+2\sqrt{\frac{\log N}{mh}+\frac{1}{m^2}} \right)\right) \leq \gamma \\
-\implies P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{2}{m})+2\sqrt{\frac{2}{m^2}} \right)\right) \leq \gamma \; \text{, if } h > m\log N \\
-\implies P\left(\hat{m}_i - m \geq 6\alpha m \right) \leq \gamma
-}
-
-This doesn't work. We need $\hat{m}_i - m < (\alpha-1)m$ in order to get this expression.
-
-Or we could let $\varphi = \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}$
-
-\eqn{
-P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp({-\frac{h\epsilon^2}{m+\frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}}})
-}
-\eqn{
-\implies \epsilon = & \frac{1}{\sqrt{2}}\sqrt{\frac{3 \log^2(2/\delta)+2m\log(2/\delta)}{h}+\frac{\sqrt{3}}{h^2}\sqrt{h^2\log^2(2/\delta)(3\log^2(2/\delta)+4m\log(2/\delta)+3mN\gamma)}} \\
-\leq & 
-}
-
 \end{document}
-
-\subsection{Option 1: Targeted sampling during exploration phase}
-
-The key fact we were utilizing to draw conclusions about multiple arms during each timestep of the explore phase is that, given our assumed causal structure, $P(Y|do(X_i=j)) = P(Y|X_i=j)$
-
-If we do some form of targeted sampling, where we say opt to select each action $I = do(X_a = b)$ with some probability not equal to $P(X_a=b)$, then we can no longer estimate $P(Y|X_i=j)$ directly from the proportion of successes given $X_i=j$.
-
-\eqn {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- \label{eqn:marginalizedProbability}
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b)
-}
-
-Define:
-\begin{itemize}
-\item  $\mu^{ij} = E[Y|X_i=j] = P(Y|X_i=j)$
-\item $q_{ab} = P(X_a = b)$, the underlying probability that variable $a$ takes the value $b$
-\item $p_{ab} = P(do(X_a=b)|I=a)$, the probability that we set variable $a$ to $b$ given we are acting on variable $a$
-\end{itemize}
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-This estimator is unbiased and has variance bounded by (proofs in section \ref{seq:proof_expect_var}) :
-
-\eqn{
-\label{eqn:variance}
-V[\hat{\mu}_a^{ij}] \leq \begin{cases}
-\frac{1}{p_{ij}} & \text{ if } a = i \\
-\frac{1}{q_{ij}}\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right) & \text{ if } a \neq i
-\end{cases}
-}
-
-At each time step, we select and arm $do(X_a=b)$ and then receive estimates for all the arms $i,j$ with variances given by \ref{eqn:variance}.
-Note that $\left(\frac{q_{i0}^2}{p_{i0}}+\frac{q_{i1}^2}{p_{i1}} \right)$ is minimized to equal $1$ if we set the intervened on variable according to its natural probability, that is we effectively take no action.
-
-Define  $\bar{q}_i = min_j q_{ij}$, the probability of the most unlikely value for each variable.
-We can assume without loss of generality that $\bar{q}_1 \leq ...
-\leq \bar{q}_N$.
-Define $m = max(N,min\{m: \frac{1}{m}\leq \bar{q}_m\})$ We allocate half our exploration timesteps $\frac{h}{2}$ to observing and allocate the remainder equally to play the unbalanced arms with $i \leq m$.
-
-At the end of the exploration phase, we combine the estimators from each timestep as a weighted sum with weights selected to minimize the variance of the combined estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{t=1}^h w_t \hat{\mu}_{a(t)}^{ij}, \text{ where } w_t = \frac{V_t^{-1}}{\sum_s V_s^{-1}}
-}
-
-We can then use Bernstein's inequality to bound the probability of large deviations
-
-\eqn{
-P\left(\sum_{t=1}^h w_t  \hat{\mu}_{a(t)}^{ij} - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\eqn {
-\sum_{t=1}^h\frac{1}{V_t} & =  \begin{cases}
-\frac{h}{2}q_{kb} & \text{ if } q_{kb} \geq \frac{1}{m} \\
-\frac{h}{2}q_{kb} + \frac{h}{2m} & \text{ otherwise }
-\end{cases}\\
-& \geq \frac{h}{2m} \; \forall q_{kb}
-}
-
-\eqn{
-P\left(\sum_{t=1}^n w_t \hat{\mu}_{a(t)}^{ij} - \mu^{ij} \geq \frac{4}{3}\sqrt{\frac{4m\log(1/\delta)}{h}}\right) \leq \delta \; \forall (i,j)
-}
-
-Lower bound?
-
-\pagebreak
-
-\subsection{Proofs}
-\label{seq:proof_expect_var}
-\subsubsection{Expectation and variance of the estimators}
-\eq{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-If $a = i$:
-\eq{
-E[\hat{\mu}_a^{ij}] = \frac{1}{p_{ij}}E[Y\ind{X_i=j}] = \frac{1}{p_{ij}}P(Y,do(X_i=j)) = \frac{1}{p_{ij}}P(Y|X_i=j)P(do(X_i=j)) = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2  = \frac{1}{p_{ij}^2}E[Y\ind{X_i=j}] - (\mu^{ij})^2  = \frac{1}{p_{ij}}\mu^{ij} - (\mu^{ij})^2 = \frac{\mu^{ij}(1-p_{ij}\mu^{ij})}{p_{ij}} \leq \frac{\mu^{ij}}{p_{ij}} \\
-\leq & \frac{1}{p_{ij}}
-}
-
-if $a \neq i$
-
-\eq{
-E[\hat{\mu}_a^{ij}] & = \frac{q_{a0}}{q_{ij}p_{a0}}E[Y\ind{X_i=j,X_a=0}]+\frac{q_{a1}}{q_{ij}p_{a1}}E[Y\ind{X_i=j,X_a=1}] \\
-& = \frac{q_{a0}}{q_{ij}p_{a0}}P(Y|X_i=j,do(X_i=0))q_{ij}p_{i0}+\frac{q_{a1}}{q_{ij}p_{a1}}P(Y|X_i=j,do(X_i=1))q_{ij}p_{i1} \\
-& =P(Y|X_i=j,do(X_i=0))P(X_i=0)+P(Y|X_i=j,do(X_i=1))P(X_i=1)\\
-& = P(Y|X_i=j)
-}
-
-\eq{
-V[\hat{\mu}_a^{ij}] = & E[(\hat{\mu}_a^{ij})^2] - (\mu^{ij})^2 \\
- = & E[\ind{Y=1,X_i = j,X_a=0}](\frac{q_{a0}}{q_{ij}p_{a0}})^2 + E[\ind{Y=1,X_i = j,X_a=1}](\frac{q_{a1}}{q_{ij}p_{a1}})^2 \\
-& +  2 E[\ind{Y=1,X_i = j,X_a=0}\ind{Y=1,X_i = j,X_a=1}\frac{q_{a0}}{q_{ij}p_{a0}}\frac{q_{a1}}{q_{ij}p_{a1}}] -  (\mu^{ij})^2\\
-=& P(Y|X_i=j,X_a=0)q_{ij}p_{a0}(\frac{q_{a0}}{q_{ij}p_{a0}})^2 + P(Y|X_i=j,X_a=1)q_{ij}p_{a1}(\frac{q_{a1}}{q_{ij}p_{a1}})^2] - (\mu^{ij})^2 \\
-=& \frac{1}{q_{ij}}\left(P(Y|X_i=j,X_a=0)\frac{q_{a0}^2}{p_{a0}} + P(Y|X_i=j,X_a=1)\frac{q_{a1}^2}{p_{a1}}\right)- (\mu^{ij})^2  \\
-\leq & \frac{1}{q_{ij}}\left(\frac{q_{a0}^2}{p_{a0}} + \frac{q_{a1}^2}{p_{a1}}\right) \\
-}
-
-\subsubsection{Bernstein Bound}
-\begin{theorem}{Berstein's Inequality}
-Let $X_t \in \{X_1 ... X_n\}$  be independent, bounded random variables with mean $\mu$ and variance $V_t$ satisfying $|X_t - \mu| \leq b \; \forall t$. Let $V= \frac{1}{n}\sum_{t=1}^n V_t$ 
-\eq{
-& P(\sum_{t=1}^n X_t - \mu \geq \epsilon) \leq \exp(-\frac{\epsilon^2}{2nV+2b\epsilon/3})\\
-\implies & P\left(\sum_{t=1}^n X_t - \mu \geq \frac{2b\log(1/\delta)}{3} + \sqrt{2nV\log(1/\delta)}\right) \leq \delta
-}
-\end{theorem}
-Let $Z_t = w_t V_t$ where $w_t = V_t^{-1}/\sum V_t^{-1}$.
-\eq{
-V[Z_t] = \frac{V_t^{-1}}{(\sum V_t^{-1})^2}\\
-V = \frac{1}{n}\sum V[Z_t] = \frac{1}{n\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2b_z\log(1/\delta)}{3} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta, \text{ where } b_z = max_t \frac{V_t^{-1}b}{\sum V^{-1}}
-}
-
-While we are observing, $|\hat{\mu}_{kb}| = |\frac{Y\ind{X_k=b}}{q_{kb}}|\leq \frac{1}{q_{kb}}$ and $V_t^{-1} = q_{kb}$ so $V_t^{-1}b = 1$.
-
-When we are explicitly setting $X_k=b$, $|\hat{\mu}_{kb}| = |Y|\leq 1$ and $V_t^{-1} =1$ so $V_t^{-1}b = 1$
-
-\eq{
-b_z = \frac{1}{\sum V_t^{-1}}
-}
-
-\eq{
-& P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{2\log(1/\delta)}{3\sum V_t^{-1}} + \sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta \\
-\implies &  P\left(\sum_{t=1}^n w_t X_t - \mu \geq \frac{4}{3}\sqrt{\frac{2\log(1/\delta)}{\sum V_t^{-1}}}\right) \leq \delta
-}
-
-\subsection{Notes}
-Based on equation \ref{eqn:marginalizedProbability}, a natural estimator for $\mu^{ij}$, given we acted to set the variable $a$, would be
-
-\eqn{
-\hat{\mu}_a^{ij} = \begin{cases}
-\frac{Y\ind{X_i=j}}{\ind{X_i=j}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{\ind{X_i=j,X_a=0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{\ind{X_i=j,X_a=1}}q_{a1} & \text{ if } a\neq i
-\end{cases}
-}
-
-However, this forces us to deal with random variables (which may be 0) in the denominator.
-We can avoid this issue by using the importance weighted estimator, which replaces these indicators with their expectation.
-
-\eqn{
-\hat{\mu}_a^{ij} & = \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y\ind{X_i=j,X_a=0}}{q_{ij}p_{a0}}q_{a0}+\frac{Y\ind{X_i=j,X_a=1}}{q_{ij}p_{a1}}q_{a1} & \text{ if } a\neq i
-\end{cases} \\
-&= \begin{cases}
-\frac{Y\ind{X_i=j}}{p_{ij}} & \text{ if } a = i \\
-\frac{Y \ind{X_i=j}}{q_{ij}}\left(\ind{X_a=0}\frac{q_{a0}}{p_{a0}} +\ind{X_a=1}\frac{q_{a1}}{p_{a1}} \right) & \text{ if } a\neq i
-\end{cases}
-}
-
-\subsection{Previous attempt}
-
-For each arm, specified by the tuple $i,j$, we now have $K$ estimators $[\hat{\mu}_1^{ij}...\hat{\mu}_K^{ij}]$ which we wish to combine to form a single estimator $\hat{\mu}^{ij}$. We will pwhere
-
-\eqn{
-m_{a,b}^{ij} &= \sum_{s \in \{t:I_t=(a,b)\}} Z_{ab,s}^{ij} \text{ and,}\\
-Z_{ab,s}^{ij} &= \ind{X_{i,s} = j,Y_s=1} \in \{0,1\}
-}
-
-TODO check/show this estimator is unbiasedool them as a weighted average with weights we can optimize based on the $q's$ so as to minimize the variance of the estimator.
-
-\eqn{
-\hat{\mu}^{ij} = \sum_{a=1}^K w_{a} \hat{\mu}_{a}^{ij} \;\;\text{, where } \sum_{a=1}^K w_a = 1
-}
-
-Putting everything together,
-
-\eqn{
-\hat{\mu}^{ij} = \frac{w_i}{\tau_{ij}}\sum_{s \in \{t: I_t = (i,j)\}} Z_{ij,s}^{ij} + \sum_{a \neq i} \left(\frac{w_a}{q_i(j)}\left[ \frac{q_a}{\tau_{a1}}\sum_{s \in \{t:I_t=(a,1)\}} Z_{a1,s}^{ij}+  \frac{1-q_a}{\tau_{a0}}\sum_{s \in \{t:I_t = (a,0)\}} Z_{a0,s}^{ij} \right] \right)
-}
-
-We now need to show $E[\hat{\mu}^{ij}] = \mu^{ij}$ and get a high probability bound for their difference. For the latter, lets try and use McDiarmids Inequality.
-
-McDiarmid's Inequality states:
-If $Z_i \ci Z_j$ and 
-\eqn{
-|\phi(Z_1...Z_i...Z_N) - \phi(Z_1...Z_i^{'}...Z_N)| < c_i \;\; \forall i
-}
-\eqn{
-\label{eqn:McDiarmid}
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
-& P\left(|\phi(\boldsymbol{Z}) - E[\phi(\boldsymbol{Z})]| \geq 
-\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
-\right) \leq \delta
-}
-
-For our problem, $\hat{\mu}^{ij} = \phi(\boldsymbol{Z}^{ij})$. All the $Z's$ are independent as they correspond indicator functions over the results at different timesteps.
-
-\eqn{
-|\phi(...Z_{ij,s}^{ij}...) - \phi(...Z_{ij,s}^{'ij}...)| & \leq \frac{w_i}{\tau{ij}}\;\;\;\; \leftarrow \tau_{ij} \text{ such $Z$'s}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a q_a}{q_i(j)\tau_{a1}} \;\;\;\; \leftarrow \tau_{a1} \text{ such $Z$'s for each $a$}\\
-|\phi(...Z_{a1,s}^{ij}...) - \phi(...Z_{a1,s}^{'ij}...)| & \leq \frac{w_a (1-q_a)}{q_i(j)\tau_{a0}} \;\;\;\; \leftarrow \tau_{a0} \text{ such $Z$'s for each $a$}
-}
-
-\eqn{
-\label{eqn:sum_ci}
-\sum_l c_l^2 = & \frac{w_i^2}{\tau_{ij}} + \sum_{a \neq i}\frac{w_a^2}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}} +\frac{(1-q_a)^2}{\tau_{a0}}  \right)\\
-= & \sum_{a=1}^K w_a^2 f(a) \text{, where} \\
-f(a) =& \begin{cases}
-\frac{1}{q_i(j)^2}\left(\frac{q_a^2}{\tau_{a1}}+  \frac{(1-q_a)^2}{\tau_{a0}}\right)  & a \neq i\\
-\frac{1}{\tau_{ij}} & a = i
-\end{cases} 
-}
-
-We want to select weights to minimize equation \ref{eqn:sum_ci} so as to achieve as tight a bound as possible in equation \ref{eqn:McDiarmid}.
-
-Applying Legrange Multipliers
-
-\eqn {
-w_a & = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
-\label{eqn:min_sum_ci}
-\sum_i c_i^2 & = \frac{1}{\sum_a \frac{1}{f(a)}}
-}
-
-Substitution \ref{eqn:min_sum_ci} back into \ref{eqn:McDiarmid}
-
-\eqn {
-& P\left(|\hat{\mu}^{ij} - \mu^{ij}| \geq \epsilon \right) \leq 2 \exp \left(-2 \epsilon^2 \sum_{a=1}^K \eta_a^{ij} \right)\\
-& \eta_a^{ij} = \begin{cases}
-\frac{\tau_{a1} \tau_{a0} q_i(j)^2}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2} & a \neq i\\
-\tau_{ij} & a = i
-\end{cases} \\
-}
-
-Substituting this into \ref{eqn:probDeltaToLarge}
-
-\eqn{
-\label{eqn:targeted_prob_delta_too_big}
-P(\Delta_{\hat{i^*}} > D) \leq 2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right)
-}
-
-\eqn{
-R_2 \leq (T-h)\left(D+2 \sum_{(i,j)} exp\left({-\frac{D^2}{2} \sum_{a=1}^K \eta_a^{ij}}\right) \right)
-}
-
-Looking at equation \ref{eqn:targeted_prob_delta_too_big}, let $A = \frac{D^2}{2}$ and
-
-\eqn{
-c_a =  \frac{\tau_{a1}\tau_{a0}}{\tau_{a1}(1-q_a)^2+\tau_{a0}q_a^2}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) := \frac{1}{2}P(\Delta_{\hat{i^*}} > D)  =  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \sum_{a \neq i} c_a  )}+\sum_{i=1}^N e^{-A ( \tau_{i0} + (1-q_{i})^2 \sum_{a \neq i} c_a)}
-}
-
-The goal is to find some assignment $\boldsymbol{\tau}(\boldsymbol{q})$ to minimize this subject to the constraint that $\sum_i \tau_{i1} + \sum_i \tau_{i0} = h$
-
-No luck attempting to do this directly (via Lagrange multipliers ...)
-
-Note, letting $\tau_a = \tau_{a1}+\tau_{a0}$ then $c_a$ is maximized to equal $\tau_a$ if $\tau_{a1} = q_a \tau_a$
-
-Consider the case where we have only 2 variables.
-
-\eqn{
-f(\boldsymbol{\tau}) = e^{-A(\tau_{11} + q_1^2 c_2)}+
-e^{-A(\tau_{21} + q_2^2 c_1)}+
-e^{-A(\tau_{10} + (1-q_1)^2 c_2)}+
-e^{-A(\tau_{20} + (1-q_2)^2 c_1)}
-}
-
-If we consider the extreme case where $q_1=q_2=1$, ie the arms $X_1=0$ and $X_2=0$ never occur naturally.
-
-\eqn{
-f(\boldsymbol{\tau}) = 2 e^{-A(\tau_{11} + \tau_{21})}+
-e^{-A(\tau_{10})}+
-e^{-A(\tau_{20})}
-}
-
-Which is minimized subject to the constraint $\tau_{11}+\tau_{10}+\tau_{21}+\tau_{20}=h$ if we let $\tau_{10}=\tau_{20} = \frac{Ah-log(2)}{3A}$.
-Note that as $h$ gets large, we play the arms that do not occur naturally roughly $\frac{1}{3}$ of the time as opposed to $\frac{1}{4}$ if we were to play uniformly.
-We only learn anything about the arms that do not naturally occur when we play them directly.
-For the other arms we learn something all the time (provided we have some level of balance in all the other arms...) so it makes sense to play the arms that don't occur naturally a little bit more to keep the overall bounds tight.
-
-Ok, what happens if we just let $\tau_{ij} = \frac{h}{K}$
-
-\eqn {
-c_a = \frac{h}{K}\left(\frac{1}{(1-q_a)^2+q_a^2} \right) \geq \frac{h}{K}
-}
-
-\eqn {
-f \leq \sum_{i=1}^N \left(e^{-A(\frac{h}{K}+q_i^2\frac{(K-1)}{K}h)}+ e^{-A(\frac{h}{K}+(1-q_i)^2\frac{(K-1)}{K}h)}\right) \\
-\leq \sum_{i=1}^N \left(e^{-Ah}+ e^{-A(\frac{h}{K})}\right)
-}
-
-Basically worst case, we learn about half the arms fast.
-
-More generally, it would seem by targeted sampling we should be able to learn all the arms with reasonable natural probability fast - and the remaining arms at the rate $h/K$ (or better if we play them more often).
-
-Approaches to finding some better sampling function.
-
-\begin{enumerate}
-\item Find something motivated by the two variable example and show that it performs better than the uniform sampling approach.
-\item  Consider some other edge cases 
-\end{enumerate}
-
-Lets consider the case $q_1 = 1, q_2...q_N = \frac{1}{2}$ We will have $\tau_{11},\tau_{10},\tau_{ij} = \tau \; \forall i > 1,j$
-
-\eqn {
-c_a = \begin{cases}
-\tau_{11} & \text{ if } a=1\\
-2 \tau & \text{ otherwise}
-\end{cases}
-}
-
-\eqn{
-f(\boldsymbol{\tau}) & = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\tau + \frac{1}{4}(\tau_{11}+2(N-2)\tau))} \\
-& = e^{-A(\tau_{11}+2(N-1)\tau)} + e^{-A\tau_{10}} + 2 (N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})} \\
-& <  e^{-A\tau_{10}} + (2 (N-1)+1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}\\
-& =  e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{\tau_{11}}{4} + \frac{N\tau}{2})}
-}
-
-and the constraint is $\tau_{11}+\tau_{10}+2(N-1)\tau = h$
-
-If we maximize $(\frac{\tau_{11}}{4} + \frac{N\tau}{2})$ subject to $\tau_{11}+2(N-1)\tau = (h-\tau_{10})$ we find $\tau = \frac{h-\tau_{10}}{2(N-1)}$ and $\tau_{11} = 0$.
-Note, as $N$ gets large the sensitivity to the trade off between $\tau$ and $\tau_{11}$ becomes negligible.
-
-\eqn {
-\frac{\tau_{11}}{4} + \frac{N\tau}{2} = \begin{cases}
-\frac{(h-\tau_{10})N}{4(N-1)} & \text{if } \tau_{11}=0 \\
-\frac{(h-\tau_{10})N}{4(N-1)} - \frac{1}{4(N-1)} &\text{if } \tau_{11}=(h-\tau_{10})
-\end{cases}
-}
-
-\eqn {
-f(\boldsymbol{\tau}) & <e^{-A\tau_{10}} + (2N-1) e^{-A (\frac{1}{4}(h-\tau_{10}))}
-}
-
-Which is minimized if 
-\eqn {
-\tau_{10} = \frac{h}{5} - \frac{4}{5A}log(\frac{2N-1}{4})
-}
-
-Ok - lets look at the reverse case, where $q_1=\frac{1}{2}, q_2...q_N = 1$
-
-\eqn{
-\sum_a c_a = \begin{cases}
-(N-1)\tau_{11} & \text{ if } a = 1 \\
-2\tau+(N-2)\tau_{11} &\text{ otherwise }
-\end{cases}
-}
-
-The constraint is 
-\eqn {
-2\tau+(N-1)\tau_{11} + (N-1)\tau_{10} = h
-}
-
-\eqn{
-f(\boldsymbol{\tau}) = 2e^{-A(\tau+\frac{1}{4}(N-1)\tau_{11})} + (N-1)e^{-A(2\tau+(N-1)\tau_{11})}+(N-1)e^{-A\tau_{10}}
-}
-
-Holding $\tau_{10}$ constant, the 2nd and 3rd terms are constant. The first is maximized by letting $\tau_{11} = 0, \tau = \frac{h - (N-1)\tau_{10}}{2}$
-
-\eqn{
-f(\boldsymbol{\tau}) & = 2e^{-A\tau} + (N-1)e^{-2A\tau}+(N-1)e^{-A\frac{h - 2\tau}{N-1}} \leftarrow \text{ can't minimize analytically} \\
-\label{eqn:case1balancedapprox}
-& \leq (N+1)e^{-A\tau} +(N-1)e^{-A\frac{h - 2\tau}{N-1}}
-}
-
-which is minimized if
-
-\eqn {
-\tau & = \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N(N+1)}{2(N-1)})\right) \\
-\label{eqn:case1balancedapproxsolution}
-& \sim \frac{1}{N+2}\left( h - N + \frac{N}{A} \log(\frac{N}{2})\right)
-}
-
-Equation \ref{eqn:case1balancedapproxsolution} is also the solution that leads to the exponentiated terms in equation \ref{eqn:case1balancedapprox} being equal - which is should intuitively be optimal since the coefficients in front of them are very similar.
-
-Yet another case study - if $q_1...q_N = 1$ So none of the arms $X_i = 0$ occur naturally.
-
-\eqn{
-&c_a  = \tau_1 \; \forall a\\
-&\sum_{i \neq a}c_a  = (N-1)\tau_1 \; \forall a\\
-& N(\tau_0+\tau_1) = h\\
-& f(\boldsymbol{\tau}) = N e^{-A N\tau_1} + N e^{-A\tau_0}
-}
-
-minimized if 
-\eqn{
-\tau_0 = \frac{h}{N+1}-\frac{log(N)}{A(N+1)}
-}
-
-Note as $N$ get large - this equates to $\tau_0 \sim \frac{h}{N}$ ie almost all the actions are allocated to the arms that do not occur naturally.
-
-Another case study $q_1...q_{N_1} = 1,q_{N_1+1}...q_N = \frac{1}{2}$ (should encompass all previous examples ...)
-
-\eqn{
-&c_a  = \begin{cases}
-\tau_{11} &\text{ if } a \leq N_1
- \\
-2\tau & \text{otherwise}
-\end{cases}\\
-&\sum_{i \neq a}c_a  = \begin{cases}
-(N_1-1)\tau_{11} + 2 N_2 \tau & \text{ if } a \leq N \\
-N_1\tau_{11} + 2(N_2-1)\tau & \text{ otherwise }
-\end{cases}\\
-&N_1\tau_{10}+N_1\tau_{11}+2N_2\tau = h\\
-&f(\boldsymbol{\tau}) =  N_1 e^{-A\tau_{10}}+ N_1 e^{-A(N_1\tau_{11}+2N_2\tau)} + 2N_2 e^{-A(\frac{N_1}{4}\tau_{11}+\frac{N_2+1}{2}\tau)}
-}
-
-The 2nd and 3rd terms are minimized by letting $\tau_{11} = 0$ (although the result is only weakly dependnent on the trade off between $\tau_{11}$ and $\tau$ provided $N_2 >> 2$).
-
-\eqn{
-f(\boldsymbol{\tau}) &=  N_1 e^{-A\tau_{10}}+ N_1 e^{-A2N_2\tau} + 2N_2 e^{-A\frac{N_2+1}{2}\tau}\\
-& \leq N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A\frac{N_2+1}{2}\tau}\\
-& = N_1 e^{-A\tau_{10}} + (2N_2 + N_1)e^{-A(\frac{(N_2+1)(h-N_1\tau_{10})}{4N_2})} 
-}
-
-\eqn{
-\tau_{10} = \left(\frac{N_2+1}{(N_2+1)N_1+4N_2}\right)h-\frac{4N_2\log(\frac{(N_2+1)(2N_2+N_2)}{4N_2})}{A((N_2+1)N_1+4N_2))}
-}
-
-It seems like this function can be optimized in a greedy way.
-That is, start with $\boldsymbol{\tau} = \boldsymbol{0}$, and in each timestep select the $\tau_{ij}$ that leads to the greatest reduction in the error bound.
-This appears to yield broadly consistent results to the analytic solution.
-
-\includegraphics[width=.5\textwidth]{greedy_minimization_close}
-\includegraphics[width=.5\textwidth]{greedy_minimization}
-
-Ok - so when we have many unbalanced variables, we assign almost all actions to the arms corresponding to the rare side.
-However, this will still be $< h/N$ which is only a factor of $2$ better than uniform assignment.
-Doing something smarter than uniform assignment will make more difference where we have a only a few unbalanced variables.
-In this case - smart assignment will lead to bounds almost as tight as the perfectly balanced case.
-Uniform assignment may yield solutions that are almost as bad as the worst case.
-(still better than without using the structure at all)
-
-I still need to cover more fractional $q$'s...
-
-What about if I repeated one of the above analysis but let the unbalanced q = .75?
-
-What can I say overall about worst/best cases/any symmetries in the problem??? Do these solutions lead to the terms in the exponentials being similar?
-
-Goal is to choose something that will give us a simple bound and will be ok for the worst case.
-Its still not entirely clear to me what the worst case will be - after all it depends on the way you assign to $\tau$.
-
-What happens if we assume $\sum_{a\neq i} c_a$ is constant ... this is reasonable if $N$ is large
-
-When you look at each term in $c_a$, its maximized if we play the arms according to their natural probabilities.
-If the arm is balanced, this will also minimize the sum of the two terms for doing $x_i = 0$ and $x_i = 1$.
-If the arm is unbalanced, then we have to trade of balancing this sum versus contributing to the sum of $c_a$.
-Consider just how bad the sum of $c_a$ can be even if we focus entirely on balancing the two terms ...
-maybe its already enough.
-
-Fundamentally the relationship between arms is complex. Its not that we get any direct link from one arm to another - rather pairs of actions tell us something about other arms.
-
-Ok, suppose we spend half our actions, $\frac{h}{2}$, observing - that is selecting variables on which to act uniformly and setting them according to their natural probabilities.
-This contributes $\frac{h}{2N}$ to each $c_a$.
-So $\sum_{i\neq a} c_a \sim \frac{h}{2}$ provided $N-1 \sim N$.
-We don't allocate any more pulls to the more likely arm associated with each variable (this is at most a factor of 2 two few pulls).
-We now decide how to spend the remaining budget.
-
-We can assume without loss of generality that $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq ... \leq q_N$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+\sum_{i=1}^N e^{-A (\frac{h}{8})}
-}
-
-Evenly balanced allocation is $\frac{h}{2N}$.
-For those arms for which $q_i^2 \geq \frac{1}{N}$ they will already get this much information during the observe phase so we will not allocate them any more (again - at worst a factor or 2 two few).
-Lets just allocate the remainder evenly.
-So let $m = min\{m:m\leq N,\frac{1}{m} \leq q_m \}$
-
-\eqn{
-f(\boldsymbol{\tau})  \leq  \sum_{i=1}^N e^{-A ( \tau_{i1} + q_{i}^2 \frac{h}{2})}+N e^{-A (\frac{h}{8})}
-}
-
-\pagebreak
-
-\subsection{Option 2: Some measure of information gain}
-
-\subsection{Option 3: Three stage algorithm}
-\begin{enumerate}
-\item Observe randomly
-\item Play low probability arms
-\item Pick best and exploit
-\end{enumerate}

### 2016-04-22 privacy/causal_differential_privacy.tex
cleaned up structure of causality folder

--- 2014-11-26 causal-bandit/ucb_regret.tex
@@
 
-\title{Regret Bounds for UCB}
-\author{Finnian Lattimore}
+\title{Protecting causal effects}
+\author{}
 
@@
 
-Assume for each arm $i \in \{1...K\} $ there is an unknown distribution of rewards $P(X)$ and a convex function, $\psi$, such that:
+Differential privacy has primarily focused on protecting individuals but there is increasing interest in problems relating to hiding certain aggregate properties of a dataset whilst preserving the ability to use it for a specified purpose.
+In this problem we consider if we can release a dataset for predictive purposes but discourage the inference of causal conclusions about relationships between the covariates.
 
-\begin{equation}
-\begin{aligned}
-log(E[e^{\lambda(X-E[X]}]) & \leq \psi(\lambda)\\
-log(E[e^{\lambda(E[X]-X}]) & \leq \psi(\lambda)
-\end{aligned}
-\end{equation}
-This ensures that the moments of the distribution of $X$ are defined. If we select arm $i$ a fixed number of times $s$:
-\eqn {
-\label{eq:hoeff1}
-& P(|\hat \mu_{is} - \mu_{i}| > \epsilon) \leq 2e^{-s\psi^*(\epsilon)} \\
-\label{eq:hoeff2}
- \implies & P(|\hat \mu_{is} - \mu_{i}| > (\psi^{*})^{-1}\frac{\log(\frac{2}{\delta})}{s}) \leq \delta
-}
+Consider a scenario under which the causal graph generating a dataset is considered known - but may contain unmeasured variables.
 
-Assume that at time $t$ we select arm $I_t$ with the highest upper confidence bound:
+Assume we add noise to the dataset via a single point crossover process (see paper).
+The goal is to prevent reliable estimation of causal effects without effecting our ability to predict a particular target variable.
+Some specific questions:
 
-\begin{equation}
-\label{eq:armSelection}
-I_{t} = argmax_{i=1...K}\left[\hat{\mu}_{it} + (\psi^{*})^{-1}\frac{\alpha \log(t)}{T_{it}}\right]
-\end{equation}
+\begin{itemize}
+\item Under what circumstances (graph structures) it is possible to disrupt inference of a particular causal effect
+\item How can we maxiumully disrupt this inference (ie is there a cut that adds more noise for a given amount of shuffling of the data)
+\item What about if we want to disrupt a specific set of causal inference questions
+\item What about if we want to disrupt as many causal queries as possible.
+\end{itemize}
 
-Then if $\alpha > 2$,
-
-\begin{equation}
-R_n \leq \sum_{i:\Delta i > 0}\left(\frac{\alpha \Delta i}{\psi^{*}(\Delta i/2)}log(n)+\frac{\alpha}{\alpha-2}\right)
-\end{equation}
-
-\begin{theorem} If $I_t = i \neq i^*$ at least one of the following statements is true:
-\begin{enumerate}
-\item The estimated UCB on the best arm, $i^*$, is less than or equal to the actual reward for that arm: $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu^*$
-\item The estimated reward for arm $i$ is greater than or equal to the estimated CI higher than the true reward for that arm: $\hat \mu_{it} \geq \mu_i + \hat \epsilon_{it}$
-\item The number of times we have selected arm $i$ in previous timesteps, $T_{it}$, is less than some bound (that grows logarithmically with $n$).
-$T_{it} < \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \;\;\; \color{red} \longleftarrow \text{feels odd that this grows with $n$, not $t$}$
-\end{enumerate}
-\end{theorem}
-\begin{proof}
-Assume statements 1-3 are all false.
-\eq {
-3. & \implies T_{it} > \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \\
-  & \implies \Delta i > 2 (\psi^*)^{-1} \frac{\alpha log(n)}{T_{it}} \geq 2 (\psi^*)^{-1} \frac{\alpha log(t)}{T_{it}} = 2 \hat \epsilon_{it}\\
-  & \implies \Delta i > 2 \hat \epsilon_{it} \\[10pt]
-1. & \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  >  \mu^*  = \mu_i+\Delta i > \mu_i +  2 \hat \epsilon_{it}\\
-& \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \mu_i +  2 \hat \epsilon_{it}\\[10pt]
-2. & \implies \hat \mu_{it} < \mu_i + \hat \epsilon_{it} \\
-   & \implies \hat \mu_{it} + \hat \epsilon_{it} < \mu_i +2 \hat \epsilon_{it} \\
-   & \implies \hat \mu_{it} + \hat \epsilon_{it} <  \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \longleftarrow \text { UCB for arm $i < $ UCB for arm $i^*$, which contradicts $i \neq i^*$}
-}
-
-\end{proof}
-
-If statements (1) and (2) are both false, then statement (3) places a bound on the number of times we can previously have selected the incorrect arm $i$ in order to select it in this timestep.
-We can write the regret in terms of the number of times we select each arm and its sub-optimality:
-
-\eq {
-\bar{R}_n = & n\mu^* - \sum_{t=1}^{n}E[\mu_{I_t}] \\
-= & \sum_{i=1}^{K}\Delta_i E[T_{in}]\\
-= & \sum_{i=1}^{K}\Delta_i E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i\}\right] \;\; \leftarrow \text{Expected number of times selected arm $I_t$ is $i$ }\\
-}
-
-Let $\gamma = \ceil{\frac{\alpha \log(n)}{\psi^*(\Delta i/2)}}$ and suppose we had selected arm $i$ in all timesteps until $\gamma$.
-In the remaining timesteps, we can only select $i$ if statement 3) is false
-
-\eq{
-\implies E\left[T_{in}\right] & \leq \gamma + E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i \text{ and (3) is false} \}\right] \\[3pt]
-& \leq \gamma + E\left[\sum_{t=\gamma+1}^{n} \mathds{1} \{\text{(1) or (2) is true} \}\right] \leftarrow \text{since if (3) is false, (1) or (2) must be true}\\[3pt]
-& \leq \gamma + \sum_{t=\gamma+1}^{n} \left[ \mathds{P}(\text{(1) is true}) + \mathds{P}(\text{(2) is true})  \}\right] \color{red} \leftarrow \text{Bubeck has $=$ here but are (1) and (2) disjoint?} 
-}
-
-\eq {
-P(\text{(1) is true})  = & P(\hat{\mu}_{i^*t}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{t} \right) \leq \mu^*)\\
-\leq & P(\exists s \in \{1...t\} : \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right) \leq \mu^*) \color{red} \leftarrow \text {to get around the problem that t is random}\\
-\leq & \sum_{s=1}^t P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right) \leq \mu^* \right) \color{red} \leftarrow \text{union bound}\\
-\text{ From equation (\ref{eq:hoeff2}) we have:}\\
-& P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{ \log{\frac{1}{\delta}}}{s} \right) \leq \mu^* \right) < \delta\\
-\text{Let } \delta = t^{-\alpha} \implies & P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right)\leq \mu^* \right) < t^{-\alpha}\\
-\implies & P(\text{(1) is true}) \leq \sum_{s=1}^t t^{-\alpha} = t*t^{-\alpha} =  t^{1-\alpha}
-}
-
-Similarly, $P(\text{(2) is true}) \leq t^{1-\alpha} \implies E[T_{in}] \leq \gamma + \sum_{t=\gamma+1}^n 2t^{1-\alpha}$
+The single point crossover process divides the variables into two groups and makes estimates of the form $P(X_1|X_2)$ unreliable if $X_1$ and $X_2$ are in different partitions.
+We can think of the process as drawing a cut through the graph.
 

### 2016-04-22 review/recipe_for_causal_inference.tex
cleaned up structure of causality folder

\documentclass{article}

\usepackage[latin1]{inputenc}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\begin{document}
\pagestyle{empty}

% Define block styles
\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
    text width=10em, text centered, rounded corners, minimum height=3em]
\tikzstyle{line} = [draw, -latex']

\begin{tikzpicture}[node distance = 2cm, auto]
    % Place nodes
    \node [block] (drawgraph) {Use prior knowledge to draw the graph};
    \node [block, below of=drawgraph] (docalc) {Apply the Do Calculus};
    \node [block, below left of=docalc,node distance=3.5cm] (fail) {Non Identifiable};
    \node [block, below right of=docalc, node distance=3.5cm] (success) {Identifiable};
    \node [block, below of=fail](notident){Bugger! You will have to make additional assumptions or do experiments};
     \node [block, below of=success](ident){Great! Apply the formula from the do-caclulus to estimate effect.};
     \node [block, fill=red!10, dashed, right of=ident,node distance=4.5cm](stats){Deal with pesky statistical issues introduced by finite data as you see fit};

    % Draw edges
    \path [line] (drawgraph) -- (docalc);
    \path [line] (docalc) -- (fail);
    \path [line] (docalc) -- (success);
    \path [line] (fail) -- (notident);
    \path [line] (success) -- (ident);
    \path [line] (stats) -- (ident);

\end{tikzpicture}
\end{document}

### 2016-04-22 review/simpsons_christmas.tex
cleaned up structure of causality folder

--- 2016-04-22 privacy/causal_differential_privacy.tex
@@
 
-Differential privacy has primarily focused on protecting individuals but there is increasing interest in problems relating to hiding certain aggregate properties of a dataset whilst preserving the ability to use it for a specified purpose.
-In this problem we consider if we can release a dataset for predictive purposes but discourage the inference of causal conclusions about relationships between the covariates.
-
 Consider a scenario under which the causal graph generating a dataset is considered known - but may contain unmeasured variables.

### 2016-04-22 causal-bandit/symfer_model.py

@@ -1,0 +1,24 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Jan 27 20:30:16 2016
+
+@author: finn
+"""
+import sys
+sys.path.append("/home/finn/programming/pyscripts/symfer-0.2.1")
+import symfer as sym
+
+pA = .6 #Probability that A = 0
+pB = np.array([.5,.8]) #Probability that B = 0 given A=0,A=1
+pC = np.array([.5,.2])
+pY = np.array([[.2,.9],[.6,.8]]) # P(Y|B,C)
+
+A = {'A':['x0','x1']}
+B = {'B':['y0','y1']}
+C = {'C':['z0','z1']}
+Y = {'Y':['y0','y1']}
+factors = {}    # an empty dictionary which we'll fill with factors
+factors['A'] = sym.Multinom([A],[pA,1-pA])                  # P(A)
+factors['B'] = sym.Multinom([B,A],[pB[0],1-pB[0],pB[1],1-pB[1]]) # P(B|A) 
+factors['C'] = sym.Multinom([C,A],[pC[0],1-pC[0],pC[1],1-pC[1]]) # P(C|A)
+factors['Y'] = sym.Multinom([Y,C,B],[pY[0][0],1-pY[0][0],pY[0][1], 1-pY[0][1],pY[1][0],1-pY[1][0],pY[1][1],1-pY[1][1]]) # P(Y|B,C)

### 2016-04-22 identify.py

@@ -1,350 +1,0 @@
-# code based on "On the Identification of Causal Effects", Tian & Pearl
-import matplotlib.pyplot as plt
-def add_value(dictionary,key,value = None):
-    """ adds the value to list of values for relevent key """
-    if key in dictionary:
-        if value is not None:
-            dictionary[key].append(value)
-    else:
-        if value is None:
-            dictionary[key] = []
-        else:
-            dictionary[key]=[value]
-
-def intersection(lst1,lst2):
-    return [x for x in lst1 if x in lst2]
-
-def math_sum(variables,exclude):
-    sum_over  = variables - exclude
-    if len(sum_over) > 0:
-        return "\sum_{%s}" % ",".join(sum_over)
-    return ""
-    
-
-class Graph():
-    def __init__(self, graph = None):
-        if graph:
-            self.children = graph.children.copy()
-            self.parents = graph.parents.copy()
-            self.siblings = graph.siblings.copy()
-        else:
-            self.children = {} # each node in graph points to a list of children
-            self.parents = {} # each node in the graph points to a list of parents
-            self.siblings = {} # each node in the graph points to a list of siblings (ie nodes connected with a bidirectional link
-
-    @classmethod
-    def from_edge_lists(cls,directed_edges,undirected_edges):
-        g = cls()
-        g.load_edge_lists(directed_edges,undirected_edges)
-        return g
-
-    def load_edge_lists(self,directed_edges,undirected_edges):
-        for from_node,to_node in directed_edges:
-            self.add_edge(from_node,to_node)
-            
-        for a,b in undirected_edges:
-            self.add_bidirected_edge(a,b)
-        
-
-    def add_edge(self,from_node, to_node):
-        add_value(self.children,from_node,to_node)
-        add_value(self.children,to_node)
-        add_value(self.parents, to_node, from_node)
-        add_value(self.parents,from_node)
-        add_value(self.siblings,to_node)
-        add_value(self.siblings,from_node)
-        
-        assert(self.children.keys() == self.parents.keys() == self.siblings.keys()) # all nodes should be in all dicts.
-
-    def add_bidirected_edge(self, a,b):
-        add_value(self.siblings,a,b)
-        add_value(self.siblings,b,a)
-        add_value(self.children,a)
[…]

### 2016-04-22 identify2.py

@@ -1,317 +1,0 @@
-# code based on "On the Identification of Causal Effects", Tian & Pearl
-import matplotlib.pyplot as plt
-
-def add_value(dictionary,key,value = None):
-    """ adds the value to list of values for relevent key """
-    if key in dictionary:
-        if value is not None:
-            dictionary[key].append(value)
-    else:
-        if value is None:
-            dictionary[key] = []
-        else:
-            dictionary[key]=[value]
-
-def intersection(lst1,lst2):
-    return [x for x in lst1 if x in lst2]
-
-def math_sum(variables,exclude):
-    sum_over  = variables - exclude
-    if len(sum_over) > 0:
-        return "\sum_{%s}" % ",".join(sum_over)
-    return ""
-    
-
-class Graph():
-    def __init__(self, graph = None):
-        if graph:
-            self.children = graph.children.copy()
-            self.parents = graph.parents.copy()
-            self.siblings = graph.siblings.copy()
-        else:
-            self.children = {} # each node in graph points to a list of children
-            self.parents = {} # each node in the graph points to a list of parents
-            self.siblings = {} # each node in the graph points to a list of siblings (ie nodes connected with a bidirectional link
-
-    @classmethod
-    def from_edge_lists(cls,directed_edges,undirected_edges):
-        g = cls()
-        g.load_edge_lists(directed_edges,undirected_edges)
-        return g
-
-    def load_edge_lists(self,directed_edges,undirected_edges):
-        for from_node,to_node in directed_edges:
-            self.add_edge(from_node,to_node)
-            
-        for a,b in undirected_edges:
-            self.add_bidirected_edge(a,b)
-        
-
-    def add_edge(self,from_node, to_node):
-        add_value(self.children,from_node,to_node)
-        add_value(self.children,to_node)
-        add_value(self.parents, to_node, from_node)
-        add_value(self.parents,from_node)
-        add_value(self.siblings,to_node)
-        add_value(self.siblings,from_node)
-        self.verify_structure()  
-
-    def add_bidirected_edge(self, a,b):
-        add_value(self.siblings,a,b)
-        add_value(self.siblings,b,a)
-        add_value(self.children,a)
-        add_value(self.children,b)
[…]

### 2016-04-22 whybackdoor.py

@@ -1,0 +1,7 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Apr 13 16:46:32 2016
+
+@author: finn
+"""
+

### 2016-04-30 causal-bandit/nips_2016/general.tex
transfers icml paper to nips format

--- 2016-02-05 causal-bandit/icml-2015/general.tex
@@
 For example, given the causual graph in Figure \ref{fig:causalStructure_confounded}, by applying the truncated factorization and marginalizing, we obtain $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1}$.
-Note that the quality of any estimator we build for $P{Y|do(X_2= j)}$ is not just a function of the number of times we observe $X_2 = j$.
+Note that the quality of any estimator we build for $\P{Y|do(X_2= j)}$ is not just a function of the number of times we observe $X_2 = j$.
 
@@
 \end{figure}
-\todof{connect truncated product formula to here by noting that the truncated formula allows us to estimate $P_a$ from observational quantities - which could plausible be available prior to running the experiment, allowing us to assume the $P_a$'s are known.}
 

### 2016-04-30 causal-bandit/nips_2016/introduction.tex
transfers icml paper to nips format

--- 2016-02-05 causal-bandit/icml-2015/introduction.tex
@@
 In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently and there are $2N$ possible interventions.
-We propose a simple causal best arm identification algorithm (Algorithm~\ref{alg:simple}) for this problem and show that up to logarithmic factors it enjoys minimax optimal
+We propose a simple causal best arm identification algorithm for this problem and show that up to logarithmic factors it enjoys minimax optimal
 simple regret guarantees of $\tilde\Theta(\sqrt{m/T})$ where $m$ depends on the causal model and may be much smaller than $N$.
-\todom{Need ref or proof?}
-In contrast, existing best arm identification algorithms suffer $\Omega(\sqrt{N/T})$ simple regret.
+In contrast, existing best arm identification algorithms suffer $\Omega(\sqrt{N/T})$ simple regret (Thm. 4 by \citet{audibert2010best}).
 This shows theoretically the value of our framework over the traditional bandit problem. 
@@
 Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
-\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been intervened on can provide important contextual information.
-\todof{write one sentence to explain how their work is different to ours}
+\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been 
+intervened on can provide important contextual information. Their work differs in many ways. For example, the focus is on the cumulative regret and
+the context is observed before the action is taken and cannot be controlled by the learning agent.
+
+%\todof{write one sentence to explain how their work is different to ours}
 
@@
 It is easy to come up with a counter example in which it does not).
+
+The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{BJQ13}, 
+where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discusses some of the issues 
+encountered in sequential design, but do not give an algorithm or theoretical results for this case.

### 2016-04-30 causal-bandit/nips_2016/nips_2016.tex
transfers icml paper to nips format

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
% \PassOptionsToPackage{numbers, compress}{natbib}
% before loading nips_2016
%
% to avoid loading the natbib package, add option nonatbib:
% \usepackage[nonatbib]{nips_2016}

\usepackage{nips_2016}

% to compile a camera-ready version, add the [final] option, e.g.:
% \usepackage[final]{nips_2016}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\title{Causal Bandits: Learning Good Interventions via Causal Inference}

% The \author macro works with any number of authors. There are two
% commands used to separate the names and addresses of multiple
% authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to
% break the lines. Using \AND forces a line break at that point. So,
% if LaTeX puts 3 of 4 authors names on the first line, and the last
% on the second line, try using \AND instead of \And before the third
% author name.

\author{
  David S.~Hippocampus\thanks{Use footnote for providing further
    information about author (webpage, alternative
    address)---\emph{not} for acknowledging funding agencies.} \\
  Department of Computer Science\\
  Cranberry-Lemon University\\
  Pittsburgh, PA 15213 \\
  \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

\begin{document}
% \nipsfinalcopy is no longer used

\maketitle

\begin{abstract}
  The abstract paragraph should be indented \nicefrac{1}{2}~inch
  (3~picas) on both the left- and right-hand margins. Use 10~point
  type, with a vertical spacing (leading) of 11~points.  The word
  \textbf{Abstract} must be centered, bold, and in point size 12. Two
  line spaces precede the abstract. The abstract must be limited to
  one paragraph.
\end{abstract}

\section{Submission of papers to NIPS 2016}

\textbf{There is a new style file for papers submitted in 2016!}

NIPS requires electronic submissions.  The electronic submission site
is
\begin{center}
  \url{https://cmt.research.microsoft.com/NIPS2016/}
\end{center}

Please read carefully the instructions below and follow them
faithfully.

\subsection{Style}

Papers to be submitted to NIPS 2016 must be prepared according to the
instructions presented here. Papers may only be up to eight pages
long, including figures. Since 2009 an additional ninth page
\emph{containing only acknowledgments and/or cited references} is
allowed. Papers that exceed nine pages will not be reviewed, or in any
other way considered for presentation at the conference.

The margins in 2016 are the same as since 2007, which allow for
$\sim$$15\%$ more words in the paper compared to earlier years.

Authors are required to use the NIPS \LaTeX{} style files obtainable
at the NIPS website as indicated below. Please make sure you use the
current files and not previous versions. Tweaking the style files may
be grounds for rejection.

\subsection{Retrieval of style files}

The style files for NIPS and other conference information are
available on the World Wide Web at
\begin{center}
  \url{http://www.nips.cc/}
\end{center}
The file \verb+nips_2016.pdf+ contains these instructions and
illustrates the various formatting requirements your NIPS paper must
satisfy.

The only supported style file for NIPS 2016 is \verb+nips_2016.sty+,
rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{}
  2.09, Microsoft Word, and RTF are no longer supported!}

The new \LaTeX{} style file contains two optional arguments:
\verb+final+, which creates a camera-ready copy, and \verb+nonatbib+,
which will not load the \verb+natbib+ package for you in case of
package clash.

At submission time, please omit the \verb+final+ option. This will
anonymize your submission and add line numbers to aid review.  Please
do \emph{not} refer to these line numbers in your paper as they will
be removed during generation of camera-ready copies.

The file \verb+nips_2016.tex+ may be used as a ``shell'' for writing
your paper. All you have to do is replace the author, title, abstract,
and text of the paper with your own.

The formatting instructions contained in these style files are
summarized in Sections \ref{gen_inst}, \ref{headings}, and
\ref{others} below.

\section{General formatting instructions}
\label{gen_inst}

The text must be confined within a rectangle 5.5~inches (33~picas)
wide and 9~inches (54~picas) long. The left margin is 1.5~inch
(9~picas).  Use 10~point type with a vertical spacing (leading) of
11~points.  Times New Roman is the preferred typeface throughout, and
will be selected for you by default.  Paragraphs are separated by
\nicefrac{1}{2}~line space (5.5 points), with no indentation.

The paper title should be 17~point, initial caps/lower case, bold,
centered between two horizontal rules. The top rule should be 4~points
thick and the bottom rule should be 1~point thick. Allow
\nicefrac{1}{4}~inch space above and below the title to rules. All
pages should start at 1~inch (6~picas) from the top of the page.

For the final version, authors' names are set in boldface, and each
name is centered above the corresponding address. The lead author's
name is to be listed first (left-most), and the co-authors' names (if
different address) are set to follow. If there is only one co-author,
list both author and co-author side by side.

Please pay special attention to the instructions in Section \ref{others}
regarding figures, tables, acknowledgments, and references.

\section{Headings: first level}
\label{headings}

All headings should be lower case (except for first word and proper
nouns), flush left, and bold.

First-level headings should be in 12-point type.

\subsection{Headings: second level}

Second-level headings should be in 10-point type.

\subsubsection{Headings: third level}

Third-level headings should be in 10-point type.

\paragraph{Paragraphs}

There is also a \verb+\paragraph+ command available, which sets the
heading in bold, flush left, and inline with the text, with the
heading followed by 1\,em of space.

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone.

\subsection{Citations within the text}

The \verb+natbib+ package will be loaded for you by default.
Citations may be author/year or numeric, as long as you maintain
internal consistency.  As to the format of the references themselves,
any style is acceptable as long as it is used consistently.

The documentation for \verb+natbib+ may be found at
\begin{center}
  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
\end{center}
Of note is the command \verb+\citet+, which produces citations
appropriate for use in inline text.  For example,
\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}
produces
\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

If you wish to load the \verb+natbib+ package with options, you may
add the following before loading the \verb+nips_2016+ package:
\begin{verbatim}
   \PassOptionsToPackage{options}{natbib}
\end{verbatim}

If \verb+natbib+ clashes with another package you load, you can add
the optional argument \verb+nonatbib+ when loading the style file:
\begin{verbatim}
   \usepackage[nonatbib]{nips_2016}
\end{verbatim}

As submission is double blind, refer to your own published work in the
third person. That is, use ``In the previous work of Jones et
al.\ [4],'' not ``In our previous work [4].'' If you cite your other
papers that are not widely available (e.g., a journal paper under
review), use anonymous author names in the citation, e.g., an author
of the form ``A.\ Anonymous.''

\subsection{Footnotes}

Footnotes should be used sparingly.  If you do require a footnote,
indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Place the footnotes at the bottom of the
page on which they appear.  Precede the footnote with a horizontal
rule of 2~inches (12~picas).

Note that footnotes are properly typeset \emph{after} punctuation
marks.\footnote{As in this example.}

\subsection{Figures}

All artwork must be neat, clean, and legible. Lines should be dark
enough for purposes of reproduction. The figure number and caption
always appear after the figure. Place one line space before the figure
caption and one line space after the figure. The figure caption should
be lower case (except for first word and proper nouns); figures are
numbered consecutively.

You may use color figures.  However, it is best for the figure
captions and the paper body to be legible if the paper is printed in
either black/white or in color.
\begin{figure}[h]
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

\subsection{Tables}

All tables must be centered, neat, clean and legible.  The table
number and title always appear before the table.  See
Table~\ref{sample-table}.

Place one line space before the table title, one line space after the
table title, and one line space after the table. The table title must
be lower case (except for first word and proper nouns); tables are
numbered consecutively.

Note that publication-quality tables \emph{do not contain vertical
  rules.} We strongly suggest the use of the \verb+booktabs+ package,
which allows for typesetting high-quality, professional tables:
\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}
This package was used to typeset Table~\ref{sample-table}.

\begin{table}[t]
  \caption{Sample table title}
  \label{sample-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule{1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Final instructions}

Do not change any aspects of the formatting parameters in the style
files.  In particular, do not modify the width or length of the
rectangle the text should fit into, and do not change font sizes
(except perhaps in the \textbf{References} section; see below). Please
note that pages should be numbered.

\section{Preparing PDF files}

Please prepare submission files with paper size ``US Letter,'' and
not, for example, ``A4.''

Fonts were the main cause of problems in the past years. Your PDF file
must only contain Type 1 or Embedded TrueType fonts. Here are a few
instructions to achieve this.

\begin{itemize}

\item You should directly generate PDF files using \verb+pdflatex+.

\item You can check which fonts a PDF files uses.  In Acrobat Reader,
  select the menu Files$>$Document Properties$>$Fonts and select Show
  All Fonts. You can also use the program \verb+pdffonts+ which comes
  with \verb+xpdf+ and is available out-of-the-box on most Linux
  machines.

\item The IEEE has recommendations for generating PDF files whose
  fonts are also acceptable for NIPS. Please see
  \url{http://www.emfield.org/icuwb2010/downloads/IEEE-PDF-SpecV32.pdf}

\item \verb+xfig+ "patterned" shapes are implemented with bitmap
  fonts.  Use "solid" shapes instead.

\item The \verb+\bbold+ package almost always uses bitmap fonts.  You
  should use the equivalent AMS Fonts:
\begin{verbatim}
   \usepackage{amsfonts}
\end{verbatim}
followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or
\verb+\mathbb{C}+ for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You
can also use the following workaround for reals, natural and complex:
\begin{verbatim}
   \newcommand{\RR}{I\!\!R} %real numbers
   \newcommand{\Nat}{I\!\!N} %natural numbers
   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
\end{verbatim}
Note that \verb+amsfonts+ is automatically loaded by the
\verb+amssymb+ package.

\end{itemize}

If your file contains type 3 fonts or non embedded TrueType fonts, we
will ask you to fix it.

\subsection{Margins in \LaTeX{}}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+ from the \verb+graphicx+ package. Always
specify the figure width as a multiple of the line width as in the
example below:
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
See Section 4.4 in the graphics bundle documentation
(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})

A number of width problems arise when \LaTeX{} cannot properly
hyphenate a line. Please give LaTeX hyphenation hints using the
\verb+\-+ command when necessary.

\subsubsection*{Acknowledgments}

Use unnumbered third level headings for the acknowledgments. All
acknowledgments go at the end of the paper. Do not include
acknowledgments in the anonymized submission, only in the final paper.

\section*{References}

References follow the acknowledgments. Use unnumbered first-level
heading for the references. Any choice of citation style is acceptable
as long as you are consistent. It is permissible to reduce the font
size to \verb+small+ (9 point) when listing the references. {\bf
  Remember that you can use a ninth page as long as it contains
  \emph{only} cited references.}
\medskip

\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms
for connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and
T.K.\ Leen (eds.), {\it Advances in Neural Information Processing
  Systems 7}, pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS:
  Exploring Realistic Neural Models with the GEneral NEural SImulation
  System.}  New York: TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of
learning and recall at excitatory recurrent synapses and cholinergic
modulation in rat hippocampal region CA3. {\it Journal of
  Neuroscience} {\bf 15}(7):5249-5262.

\end{document}

### 2016-04-30 causal-bandit/nips_2016/setup.tex
transfers icml paper to nips format

--- 2016-02-05 causal-bandit/icml-2015/setup.tex
@@
 \end{equation}
-This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009} or ``best-arm identification'' problem~\citep{Gabillon2012} and
+This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and
 is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefintely.

### 2016-05-01 causal-bandit/nips_2016/introduction.tex
adds typo corrections to first 3 sections

@@
 order to evaluate their effectiveness via feedback from a single, real-valued reward signal.
-We propose a generalisation of the standard model by assuming that, in additional to the reward signal, the learner observes the values of a number of covariates 
+We propose a generalisation of the standard model by assuming that, in addition to the reward signal, the learner observes the values of a number of covariates 
 drawn from a probabilistic causal model~\citep{Pearl2000}.
@@
 In each season the farmer has enough time and money to intervene and control at most one of these variables:
-deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed a through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
+deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
 When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
@@
 As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
-In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables, and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
+In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
 However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.

### 2016-05-01 causal-bandit/nips_2016/setup.tex
adds typo corrections to first 3 sections

@@
 arbitrary but unknown, real-valued function $r$. The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Conversely, any causal bandit problem can reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by 
+Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by 
 treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.

### 2016-05-01 causal-bandit/nips_2016/simple-regret.tex
adds typo corrections to first 3 sections

--- 2016-02-05 causal-bandit/icml-2015/simple-regret.tex
@@
 The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-The fact that only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
+The fact that the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
 Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
@@
 \begin{theorem}\label{thm:lower}
-For all $T$, $\vec{q}$ and all strategies there exists a reward function such that
+For all $T$, $\vec{q}$ and all strategies, there exists a reward function such that
 \eq{

### 2016-05-01 causal-bandit/nips_2016/general.tex
condences figures, minor changes to section 4

@@
 Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
-For example, given the causual graph in Figure \ref{fig:causalStructure_confounded}, by applying the truncated factorization and marginalizing, we obtain $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1}$.
-Note that the quality of any estimator we build for $\P{Y|do(X_2= j)}$ is not just a function of the number of times we observe $X_2 = j$.
 
-\begin{figure}[h]
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor.
+However, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
+For example, given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$.
+Suppose $X_2 = X_1$ deterministically.
+No matter how large $\P{X_2 = 1}$ is, we will never observe $(X_2=1,X_1 = 0)$ and so cannot get a good estimate for $\P{Y|do(X_2=1)}$.
+
+\begin{figure}
   \centering
+  \begin{minipage}[b]{0.4\textwidth}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \node[main node, below left=of 1](2){$X_2$};
-\node[main node, below right=of 1](3){$X_3$};
-\node[main node, below right=of 2](4){Y};
+\node[main node, below right=of 1](4){Y};
  \path[every node/.style={font=\sffamily\small}]
     (1) edge (2)
-    (1) edge (3)
-    (3) edge (4)
+    (1) edge (4)
     (2) edge (4);
@@
 \caption{}\label{fig:causalStructure_confounded}
-\end{figure}
-
-We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor.
-However, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
-In our example, suppose $\P{X_1=\frac{1}{2}}$ and $X_2 = X_1$ deterministically.
-$\P{X_2 = 1} = \frac{1}{2}$, however we will never observe $(X_2=1,X_1 = 0)$ so we cannot get a good estimate for $\P{Y|do(X_2=1)}$.
-
-Additionally, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
-Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$
-for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
-We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
-
-%It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones.
-We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
-%Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
-%for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
-%We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
-
-\begin{figure}[h]
-\centering
+  \end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.4\textwidth}
     \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \label{fig:causalchain}
+  \end{minipage}
 \end{figure}
+
+Additionally, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
+Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$
+for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
+We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
+
+%It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones.
+We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
+%Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
+%for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
+%We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
 

### 2016-05-02 causal-bandit/nips_2016/general.tex
started re-writing section 4

@@
 For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}.
+However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}.
 \eq{
@@
 We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor.
-However, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
+However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
+Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$
+for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
+We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
+
+To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally allocate samples to actions.
+NEED TO EXPLAIN WHY THIS PROBLEM IS HARD.
+
+To address this difficult problem, we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$.
+$\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$ We then assume the conditional interventional distributions $\P{\parents{Y}|a}$ are known.
+ADD MORE JUSTIFICATION FOR THIS ASSUMPTION.
+IE COULD HAPPEN IF WE HAD EXPERIMENTAL DATA COLLECTED WITH A DIFFERENT OBJECTIVE AND IS A NESSESSARY FIRST STEP TO SOLVING THE MORE GENERAL PROBLEM.
+
+%Firstly, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
 For example, given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$.
-Suppose $X_2 = X_1$ deterministically.
-No matter how large $\P{X_2 = 1}$ is, we will never observe $(X_2=1,X_1 = 0)$ and so cannot get a good estimate for $\P{Y|do(X_2=1)}$.
+Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot get a good estimate for $\P{Y|do(X_2=1)}$.
+Secondly it is no longer optimal
 
 \begin{figure}
-  \centering
-  \begin{minipage}[b]{0.4\textwidth}
+    \begin{subfigure}[b]{0.3\textwidth}
         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
 \node[main node](1){$X_1$};
@@
 \end{tikzpicture}
-\caption{}\label{fig:causalStructure_confounded}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[b]{0.4\textwidth}
+        \caption{}
+        \label{fig:causalStructure_confounded}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.69\textwidth}
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
 \node[main node](1){$X_{1}$};
@@
 \end{tikzpicture}
-\caption{A causal chain graph}.
+        \caption{}
         \label{fig:causalchain}
-  \end{minipage}
+    \end{subfigure}
+    \caption{Example causal graphs}\label{fig:animals}
 \end{figure}
 
-Additionally, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
-Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$
-for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
-We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
+Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
+Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
 
-%It is no longer optimal to divide our samples between observing to learn about multiple actions and explicit intervention to learn single ones.
-We need to incorporate the information we can learn about the reward for intervening on one variable from rounds in which we actually set a different variable.
-%Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ 
-%for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
-%We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$.
+Our algorithm will choose $T$ samples from $Q$ IS THIS LANGUAGE CORRECT.
+CAN WE SAY IT LIKE THIS?
+and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
+The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
 
-Let $P_a$ be a shorthand for $\P{.|a}$ and let $\eta$ be a distribution on available interventions $\calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
-Define $Q = \sum_{a \in \calA} \eta_a \Ps_a$ to be the mixture distribution over the interventions with respect to $\eta$.
-Our algorithm will choose $T$ samples from $Q$ and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via
-a truncated importance weighted estimator. The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail
-that is so detrimental to its concentration guarantees.
-
-Let $a \in \calA$ be an intervention and define random variable
+Define the random variable
 \eq{
@@
 \EE_Q[Z_{at}] 
-&= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] \\
-&= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] \\
-&= \mu_a - \beta_a\,,
+= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] 
+= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] 
+= \mu_a - \beta_a\,,
 }
@@
 \eq{
-&\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \\ 
-&\qquad\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
+\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} 
+\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
 }
@@
 \eta^* 
-&= \argmin_\eta m(\eta) \\
+= \argmin_\eta m(\eta) 
 &= \argmin_\eta \max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
@@
 m(\eta) 
-&= \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
-&\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] \\
-&\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA| \qedhere
+= \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
+\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
+\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA| \qedhere
 }

### 2016-05-03 causal-bandit/nips_2016/general.tex
moves proof general upper bound to supplementary materials

@@
 
-\begin{proof}
-First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
-We abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
-By definition we have $|Z_{at}| \leq B_a$ and 
-\eq{
-\Var_Q[Z_{at}] 
-\leq \EE_Q[Z_{at}^2] 
-\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] 
-\leq m(\eta)\,.
-}
-Checking the expectation we have
-\eq{
-\EE_Q[Z_{at}] 
-= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] 
-= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] 
-= \mu_a - \beta_a\,,
-}
-where 
-\eq{
-0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}
-}
-is the negative bias. 
-The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
-\eq{
-\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
-}
-Let $\epsilon_a > 0$ be given by
-\eq{
-\epsilon = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
-}
-Then by the union bound and Bernstein's inequality 
-\eq{
-\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} 
-\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
-}
-Assuming this event does not occur and letting $a^* = \argmax_{a \in \calA} \mu_a$ we have
-\eq{
-\mu_I \geq \hat \mu_I - \epsilon_I  
-\geq \hat \mu_{a^*} - \epsilon_I  
-\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
-}
-By the definition of the truncation
-we have
-\eq{
-\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
-}
-and
-\eq{
-\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
-}
-Therefore for $C = \sqrt{2} + 4$ we have
-\eq{
-\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
-}
-as required.
-\end{proof}
-
 \begin{remark}\label{rem:truncate}

### 2016-05-03 causal-bandit/nips_2016/tech.tex
moves proof general upper bound to supplementary materials

--- 2016-02-05 causal-bandit/icml-2015/tech.tex
@@
 as required.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% GENERAL-GRAPH UPPER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Proof of Theorem \ref{thm:general}}\label{sec:thm:general}
+
+\begin{proof}
+First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
+We abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
+By definition we have $|Z_{at}| \leq B_a$ and 
+\eq{
+\Var_Q[Z_{at}] 
+\leq \EE_Q[Z_{at}^2] 
+\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] 
+\leq m(\eta)\,.
+}
+Checking the expectation we have
+\eq{
+\EE_Q[Z_{at}] 
+= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] 
+= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] 
+= \mu_a - \beta_a\,,
+}
+where 
+\eq{
+0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}
+}
+is the negative bias. 
+The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
+\eq{
+\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
+}
+Let $\epsilon_a > 0$ be given by
+\eq{
+\epsilon = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
+}
+Then by the union bound and Bernstein's inequality 
+\eq{
+\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} 
+\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
+}
+Assuming this event does not occur and letting $a^* = \argmax_{a \in \calA} \mu_a$ we have
+\eq{
+\mu_I \geq \hat \mu_I - \epsilon_I  
+\geq \hat \mu_{a^*} - \epsilon_I  
+\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
+}
+By the definition of the truncation
+we have
+\eq{
+\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
+}
+and
+\eq{
+\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
+}
+Therefore for $C = \sqrt{2} + 4$ we have
+\eq{
+\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
+}
+Therefore
+\eq{
+\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
+}
+as required.
+\end{proof}

### 2016-05-04 causal-bandit/nips_2016/simple-regret.tex
started re-writing section 4
updates definition of m for simple regret

@@
 
-\begin{figure}[h]
-\centering
+\begin{wrapfigure}[11]{r}{0.5\textwidth}
+  \begin{center}
     \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \end{tikzpicture}
+  \end{center}
   \caption{Causal model for the parallel bandits problem.\label{fig:causalStructure}}
-\end{figure}
+\end{wrapfigure}
 
@@
 
-The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-The fact that the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one means that $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
+The algorithm operates as follows.
+For the first $T/2$ rounds it chooses $do()$ to collect observational data.
+As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
 Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
@@
 The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$).
-Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
+
+%Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
+For $\tau \in [1...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \frac{1}{\tau - 1}}$.
 Define
@@
 \label{eq:m-simple}
-m(\vec{q}) = \max \set{ d : |I_{\frac{1}{d}}| \leq d}\,.
+m(\vec{q}) = \min \set{ d : |I_{d}| \leq d}\,.
 % m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
 }
+
 This captures a trade off between identifying the low probability variables and not having too many of them.
@@
 
-\begin{algorithm}[h]
+To identify a high-reward intervention, Algorithm~\ref{alg:simple} break $T$ rounds into halves. 
+In the first half, the non-interventional action $do()$ is played and the rewards observed.
+These are used to estimate the occurrance probabilities $\vec{q}$ and the value of $m(\vec{q})$.
+The remainder of the rounds are then evenly split to estimate the infrequently observed interventions, as determined by the estimate for $m(\vec{q})$.
+Although very simple, the following two theorems show that this algorithm is effectively optimal.
+
+\begin{wrapfigure}[22]{r}{0.6\textwidth}
+\vspace{-10pt}
+\begin{minipage}{.6\textwidth}
+\begin{algorithm}[H]
 \caption{Parallel Bandit Algorithm}\label{alg:simple}
@@
 \end{algorithm}
-
-To identify a high-reward intervention, Algorithm~\ref{alg:simple} break $T$ rounds into halves. 
-In the first half, the non-interventional action $do()$ is played and the rewards observed.
-These are used to estimate the occurrance probabilities $\vec{q}$ and the value of $m(\vec{q})$.
-The remainder of the rounds are then evenly split to estimate the infrequently observed interventions, as determined by the estimate for $m(\vec{q})$.
-Although very simple, the following two theorems show that this algorithm is effectively optimal.
+\end{minipage}
+\end{wrapfigure}
 

### 2016-05-04 causal-bandit/nips_2016/general.tex
adds clarifications

@@
 We now consider the more general problem where the graph structure is known, but arbitrary.
+
+% How do actions map to interventions (we need not only consider singleton interventions).
+%
+
 For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
@@
 However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
-Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$
-for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. 
+Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$.
 We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
@@
 To address this difficult problem, we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$.
-$\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$ We then assume the conditional interventional distributions $\P{\parents{Y}|a}$ are known.
-ADD MORE JUSTIFICATION FOR THIS ASSUMPTION.
-IE COULD HAPPEN IF WE HAD EXPERIMENTAL DATA COLLECTED WITH A DIFFERENT OBJECTIVE AND IS A NESSESSARY FIRST STEP TO SOLVING THE MORE GENERAL PROBLEM.
+$\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$.
+We then assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known.
+These could be estimated from experimental data on the same covariates $\boldsymbol{X}$, but where the outcome of interest differed such that $Y$ was not included, or similarly from observational data subject to identifiability constraints.
 
@@
 
-Our algorithm will choose $T$ samples from $Q$ IS THIS LANGUAGE CORRECT.
-CAN WE SAY IT LIKE THIS?
-and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
-The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
-
-Define the random variable
-\eq{
-Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
-}
-where $B_a \geq 0$ is some constant to be chosen subsequently and
-\eq{
-R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\,.
-}
-Abusing notation we define $m(\eta)$ by
-\eq{
-m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
-}
-We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately 
-coincides with the version for parallel bandits, justifying the name overloading.
-
+\begin{wrapfigure}[14]{r}{0.5\textwidth}
+\vspace{-12pt}
+\begin{minipage}{.5\textwidth}
 \begin{algorithm}[H]
@@
 \eq{
-\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
+\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
 }
@@
 \end{algorithm}
+\end{minipage}
+\end{wrapfigure}
+
+Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
+Let $\parents{Y}(X)$ denote the realization of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
+
+\eq {
+\hat{\mu_a} = \frac{1}{T} \sum_{t=1}^T R_a(X) Y \ind{R_a(X) \leq B_a}\,, 
+}
+
+where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently.
+The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
+
+The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
+\eq{
+m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
+}
+We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
 
@@
 \end{theorem}
-
-\begin{remark}\label{rem:truncate}
-The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
-often possible to choose $B_a$ significantly
-larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
-and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
-to learn from observational data.
-\end{remark}
 
@@
 
+\begin{remark}\label{rem:truncate}
+The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
+often possible to choose $B_a$ significantly
+larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
+and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
+to learn from observational data.
+\end{remark}
+
 \begin{remark}

### 2016-05-04 causal-bandit/nips_2016/setup.tex
adds clarifications

@@
 An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
-The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$ or just $\parents{i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
+The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
 An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.

### 2016-05-04 causal-bandit/nips_2016/simple-regret.tex
adds clarifications

@@
 
-This captures a trade off between identifying the low probability variables and not having too many of them.
+$I_\tau$ represents the set of actions defined as low probability given a threshold $\tau$. As $\tau$ is increased,
+
+This captures a trade off between identifying the low probability actions and not having too many of them. $I_\tau$ represents the actions defined as 'low p
 When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.

### 2016-05-05 causal-bandit/nips_2016/general.tex
changes theorem 3
makes things clearer and more concise

@@
 \P{X_1...X_N|do(X_i=j)} = 
-\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)  
+\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
 } 
-Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
+where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
 
-We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions $a$ for which the observational estimates were poor.
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational estimates were poor.
 However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
-Consider the graph in Figure \ref{fig:causalchain}, where each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$.
-We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, (but not from $do()$).
+Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$.
+We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, but not from $do()$.
 
@@
 
-To address this difficult problem, we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$.
+To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known.
+These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, or similarly from observational data subject to identifiability constraints.
+Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
+Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
+
+%we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$.
 $\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$.
-We then assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known.
-These could be estimated from experimental data on the same covariates $\boldsymbol{X}$, but where the outcome of interest differed such that $Y$ was not included, or similarly from observational data subject to identifiability constraints.
+We then
 
@@
 Secondly it is no longer optimal
-
-\begin{figure}
-    \begin{subfigure}[b]{0.3\textwidth}
-        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_1$};
-\node[main node, below left=of 1](2){$X_2$};
-\node[main node, below right=of 1](4){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (1) edge (4)
-    (2) edge (4);
-\end{tikzpicture}
-        \caption{}
-        \label{fig:causalStructure_confounded}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.69\textwidth}
-         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, right=of 4](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-  	(2) edge (3)
-    (3) edge (4)
-    (4) edge (5);
-\end{tikzpicture}
-        \caption{}
-        \label{fig:causalchain}
-    \end{subfigure}
-    \caption{Example causal graphs}\label{fig:animals}
-\end{figure}
-
-Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$.
-Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
 
@@
 \ENDFOR
-\STATE For each $a \in \calA$ compute an estimate of its return:
-\eq{
+\FOR{$a \in \calA$}
+\STATE
+\eq {
 \hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
 }
+\ENDFOR
 \STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
@@
 }
-We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
+
+We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
 
 \begin{theorem}\label{thm:general}
-If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by
+If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$
+
 \eq{
-B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.
-}
-Then for $C = \sqrt{2} + 4$ we have
-\eq{
-\mu^* - \EE[\mu_I] \leq C\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}\,.
+\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
 }
@@
 
-\subsection*{Choosing the Sampling Distribution}
+The proof is in the supplementary materials.
+Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
+Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$.
+Thus it has broader applicability to the parallel graph than the generic algorithm given here.
+We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
+for future work.
 
-Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$. In light of Theorem \ref{thm:general}
+\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$.
+In light of Theorem \ref{thm:general}
 a natural choice of $\eta$ is the minimiser of $m(\eta)$.
@@
 \eta^* 
-= \argmin_\eta m(\eta) 
-&= \argmin_\eta \max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \\
-&= \argmin_\eta \underbrace{\max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
+= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
 }
@@
 An interpretation of $m(\eta^*)$ is the minimum achievable worst-case variance of the importance weighted estimator.
-In the experimental section we present some special cases,
-but for now we give two simple results. The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
+In the experimental section we present some special cases, but for now we give two simple results.
+The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
 
@@
 \end{remark}
-
-\begin{remark}
-The algorithm in the previous section for the parallel bandit problem did not assume knowledge of the conditional distributions on $X$ and 
-instead only the graph structure. Thus the algorithm given in that setting has broader applicability to that particular graph than the generic algorithm given here.
-Finally, we believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
-for future work.
-\end{remark}

### 2016-05-05 causal-bandit/nips_2016/introduction.tex
makes things clearer and more concise

@@
 %without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
-
-\subsection{Contributions}
-
-We take the first step towards formalising and solving problems such as the one above. 
+\paragraph{Contributions} We take the first step towards formalising and solving problems such as the one above. 
 In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph.
@@
 
-\subsection{Related Work}
-
-As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
+\paragraph{Related Work} As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
 In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.

### 2016-05-05 causal-bandit/nips_2016/setup.tex
makes things clearer and more concise

@@
 The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
-An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
+An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
 The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from $\parents{i}$ to $X_i$ for each $X_i \in \vec{X}$. 
@@
 
-The objective of the learner is to minimise the simple regret
-\begin{equation}
-\label{eq:regret-simple}
-	\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.
-\end{equation}
-This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and
-is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefintely.
+The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.
 
 Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are possible.
-If the non-intervened variables are observed before an intervention is selected our framework 
-reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}. 
-Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of 
-allowable interventions thereby reducing the complexity.
+If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}.
+Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of allowable interventions thereby reducing the complexity.
 
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one 
-edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some 
-arbitrary but unknown, real-valued function $r$. The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by 
-treating each possible intervention as an arm and ignoring all sampled values for the observed variables except for the reward.
-Intuitively though, one would expect to perform better by making use of the extra observations.
+We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
+The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
+Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.
+Intuitively though, one would expect to perform better by making use of the extra structure and observations.
 

### 2016-05-05 causal-bandit/nips_2016/simple-regret.tex
makes things clearer and more concise

--- 2016-05-04 causal-bandit/nips_2016/simple-regret.tex
@@
 
-\begin{wrapfigure}[11]{r}{0.5\textwidth}
-  \begin{center}
-    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty}]
-
+\begin{figure}
+    \begin{subfigure}[b]{0.34\textwidth}
+	\centering    
+          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+%every node/.style={scale=0.6}
  %nodes
@@
 \node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
+\node[main node, below right=of 2](5){$Y$};
+ \path[every node/.style={font=\tiny}]
     (1) edge (5)
@@
     (4) edge (5);
-
 \end{tikzpicture}
-  \end{center}
-  \caption{Causal model for the parallel bandits problem.\label{fig:causalStructure}}
-\end{wrapfigure}
+        \caption{Parallel graph}
+        \label{fig:causalStructure_confounded}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.2\textwidth}
+    \centering
+        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+\node[main node](1){$X_1$};
+\node[main node, below left=of 1](2){$X_2$};
+\node[main node, below right=of 1](4){$Y$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (2)
+    (1) edge (4)
+    (2) edge (4);
+\end{tikzpicture}
+        \caption{Confounded graph}
+        \label{fig:causalStructure_confounded}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, right=of 4](5){$Y$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (2)
+  	(2) edge (3)
+    (3) edge (4)
+    (4) edge (5);
+\end{tikzpicture}
+        \caption{Chain graph}
+        \label{fig:causalchain}
+    \end{subfigure}
+    \caption{Causal Models}\label{fig:animals}
+\end{figure}
 
@@
 In the farming example, this choice of $Y$ models the success or failure of a seasons crop, which depends stochastically on the various environment variables.
-Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
+%Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.
 
@@
 
-\subsection{The Parallel Bandit Algorithm}
-\label{sub:par-bandit-alg}
+\paragraph{The Parallel Bandit Algorithm}
+%\label{sub:par-bandit-alg}
 
@@
 As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
-Thus we can create good estimators of the returns of those actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
-These are the interventions that are easily estimated ``passively'' (\ie, without playing them explicitly). 
-Those actions where $\P{X_i = j}$ is small may not be observed (often) in the first $T/2$ rounds, so
-the estimates of the returns of these actions from the first $T/2$ samples could be quite poor. For the remaining $T/2$ rounds
-the algorithm makes interventions on the actions $a = do(X_i = j)$ for which $\P{X_i = j}$ is small. The precise allocation is
-carefully chosen to minimise the worst-case simple regret. 
-The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables have an infrequently occurring value (\ie, small $q_i$ or $1-q_i$).
-
-%Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
+Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
+The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates their returns could be poor.
+To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions.
+The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$).
 For $\tau \in [1...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \frac{1}{\tau - 1}}$.
 Define
-\eqn{
+\eq{
 \label{eq:m-simple}
@@
 }
+%Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.
 
-This captures a trade off between identifying the low probability variables and not having too many of them.
-When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.
-
-To identify a high-reward intervention, Algorithm~\ref{alg:simple} break $T$ rounds into halves. 
-In the first half, the non-interventional action $do()$ is played and the rewards observed.
-These are used to estimate the occurrance probabilities $\vec{q}$ and the value of $m(\vec{q})$.
-The remainder of the rounds are then evenly split to estimate the infrequently observed interventions, as determined by the estimate for $m(\vec{q})$.
-Although very simple, the following two theorems show that this algorithm is effectively optimal.
-
-\begin{wrapfigure}[22]{r}{0.6\textwidth}
-\vspace{-10pt}
+\begin{wrapfigure}[18]{r}{0.6\textwidth}
+\vspace{-18pt}
 \begin{minipage}{.6\textwidth}
@@
 
+$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret.
+When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
+We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$.
+Although very simple, the following two theorems show that this algorithm is effectively optimal.
+
 \begin{theorem}\label{thm:uq-simple}
@@
 \end{theorem}
-
-Note that algorithms designed for finite-armed bandits would explore interventions more uniformly and achieve a regret of $\Omega(\sqrt{N/T})$.
-Since $m$ is typically much smaller than $N$ we anticipate that the new algorithm will significantly outperform classical bandit algorithms in
-this setting.
-We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}.
 
@@
 \fi
+%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
+By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
+Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\Omega(\sqrt{N/T})$.
+Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting.
+In practice, you would combine the data from both phases to estimate rewards for the low probability actions.
+We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.

### 2016-05-08 privacy/causal_differential_privacy.tex
adds some formalisms

@@
 \usepackage{parskip}
+
+\usetikzlibrary{arrows,positioning,backgrounds} 
+\tikzset{
+    %Define standard arrow tip
+    >=stealth',
+    %Define style for boxes
+    observed/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\tiny,
+           text centered,
+           scale=.8,
+           fill=blue!20!white},
+     latent/.style={
+           circle,
+           rounded corners,
+           draw=black, thick, dashed,
+           minimum width=.5em,
+           minimum height=.5em,
+           font=\footnotesize,
+           text centered,
+           fill=black!10!white
+           },
+     empty/.style={
+           circle,
+           rounded corners,
+           minimum width=.5em,
+           minimum height=.5em,
+           font=\footnotesize,
+           text centered,
+           },
+    % Define arrow style
+    pil/.style={
+           o->,
+           thick,
+           shorten <=2pt,
+           shorten >=2pt,},
+    sh/.style={ shade, shading=axis, left color=red, right color=green,
+    shading angle=45 }  
+}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+\newcommand{\pt}{\mathcal P}
 
@@
 The goal is to prevent reliable estimation of causal effects without effecting our ability to predict a particular target variable.
-Some specific questions:
+Some example questions:
 
 \begin{itemize}
-\item Under what circumstances (graph structures) it is possible to disrupt inference of a particular causal effect
-\item How can we maxiumully disrupt this inference (ie is there a cut that adds more noise for a given amount of shuffling of the data)
+\item Under what circumstances (graph structures) it is possible to disrupt inference of a particular causal effect \textcolor{red}{always provided there is a direct causal relationship between the exposure variable $X$ and outcome variable $Y$, ie $P(Y|do(X))\neq P(Y)$}
+\item How can we maxiumully disrupt this inference (ie is there a cut that adds more noise for a given amount of shuffling of the data) \textcolor{red}{data dependent}
 \item What about if we want to disrupt a specific set of causal inference questions
@@
 
-The single point crossover process divides the variables into two groups and makes estimates of the form $P(X_1|X_2)$ unreliable if $X_1$ and $X_2$ are in different partitions.
-We can think of the process as drawing a cut through the graph.
+\section{Problem Statement and Definitions}
+
+\begin{itemize}
+\item Let $G$ be a causal directed acyclic graph over vertices $V$ and edges $E$.
+\item Let $S = \set{(X_1,Y_1),(X_2,Y_2),...,(X_m,Y_m): X_i,Y_i \in V,X_i \neq Y_i}$ be a set of pairs of vertices in $G$.
+\item Each pair of vertices $(X_i,Y_i)$ in $S$ represents the causal query $P(Y_i|do(X_i))$
+\item A set of variables $Z \subset V$ is an adjustment for a query $(X,Y)$ if $X, Y \notin Z$ and $P(Y|do(X)) = \sum_Z P(Y|X,Z)P(Z)$
+\item An adjustment for a query is minimal if it does not contain another adjustment as a proper subset.
+\item The Single Point Crossover Process partitions the variables $V$ into two disjoint sets, $\pt_1$ and $\pt_2$ such that $\pt_1 \cup \pt_2 = V$.
+Let $\pt_X$ denote the partition containing the variable $X$.
+\end{itemize}
+
+\subsection{Discouraging inference of a set of causal queries via adjustment}
+
+\begin{lemma}
+One of the following holds.
+\begin{enumerate}
+\item $P(Y|do(X)) = P(Y)$
+\item There is no set of variables that satisfy the back-door criterion for identifying $P(Y|do(X))$ 
+\item There is at least one set of variables $Z$ that are a minimal adjustment for $P(Y|do(X))$, $Z$ may be the empty set. 
+\end{enumerate}
+\end{lemma}
+
+For the remainder we assume that all queries we wish to jam are of case 3, because in case 1, there is no causal effect of $X$ on $Y$ to discover and in case 2, the causal effect of $X$ on $Y$ can already not be inferred from the data and no way in which we partition the data will render it identifiable.
+
+\begin{lemma}Let $Z$ be a set of variables that is a minimal adjustment for the causal query $(X,Y)$.
+Inferring $P(Y|do(X))$ via adjusting for $Z$ is jammed if $Y \cup Z \not\subset \pt_X$.
+In other words, if at least one of the variables in $X \cup Y \cup Z$ is in a different a different partition to the others.
+
+\color{red} This stems from the fact that the expression for calculating $P(Y|do(X))$ via adjusting for $Z$ contains the factor $P(Y|X,Z)$.
+I'm assuming the single point crossover process gives us this result.
+We need to a clear definition of 'jammed' in terms of the crossover-process.
+\end{lemma}
+
+\begin{lemma} Let $\boldsymbol{Z}$ be the set of minimal adjustments for $(X,Y)$. The causal query $(X,Y)$ is jammed if inferring $P(Y|do(X))$ via $Z$ is jammed for $\forall Z \in \boldsymbol{Z}$ 
+\end{lemma}
+
+\subsubsection{An algorithm for finding a cut that renders jams a set of causal queries $S$}
+\begin{enumerate}
+\item For each query $(X_i,Y_i)\in S$, find the set of minimal adjustments $\boldsymbol{Z}_i$  
+\item Construct the set of sets $Q = \set{X_i \cup Y_i \cup Z_{ik} : 1 \leq i \leq m, 0 \leq k \leq |\boldsymbol{Z}_i|}$
+\item Find a partition such that $(Q_i \nsubseteq \pt_1)\land (Q_i \nsubseteq \pt_2)\qquad \forall Q_i \in Q$
+\end{enumerate}
+
+For example in figure \ref{fig:minimal_sets_example}, with $S = \set{(A,B),(E,B),(C,B),(E,A)}$
+
+\eq{
+&S_1 = (A,B),\, \boldsymbol{Z}_1 = \set{\set{C, E},\set{D,E} \set{D,F}} \\
+&S_2 = (E,B),\, \boldsymbol{Z}_2 = \set{\set{C,D}} \\
+&S_3 = (C,B),\, \boldsymbol{Z}_3 = \set{} \\
+&S_4 = (E,A),\, \boldsymbol{Z}_4 = \set{C} \\
+&Q = \set{\set{A,B,C,E},\set{A,B,D,E},\set{A,B,D,F},\set{E,B,C,D},\set{C,B},\set{E,A,C}}
+}
+
+We could jam $S$ with any partition of the form $\set{A,B,...},\set{C,E,F,...}$.$S_2,S_3$ and $S_4$ are jammed because their $X$ and $Y$ variables are in different partitions.
+$S_1$ is jammed because, although $A$ and $B$ are in the same partition, for each of the three minimal adjustment sets, at least one variable is not in $\pt_A$.
+$D$ could be placed in either partition.
+This is not the only solution, $\set{A,C,...},\set{B,E,...}$ is another.
+Note than in any solution $B$ and $C$ must be in different partitions as this is the only way to jam $S_3$.
+
+The number of minimal adjustments for a given query can grow exponentially with the number of nodes in the graph.
+There is an algorithm to enumerate them that requires $O(n^3)$ per adjustment \cite{textor2012adjustment} .
+
+\begin{figure}
+\caption{}
+\label{fig:minimal_sets_example}
+\centering
+   \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+%every node/.style={scale=0.6}
+ %nodes
+\node[main node](1){C};
+\node[main node, below right=of 1](3){E};
+\node[main node, above right=of 3](2){D};
+\node[main node, below right=of 3](4){F};
+\node[main node, below right=of 4](5){B};
+\node[hidden, below left=of 1](7){};
+\node[main node, below left=of 7](6){A};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (2) edge (3) edge (6)
+    (6) edge (5)
+    (2) edge (3) edge (5)
+    (3) edge (4) edge (6)
+    (4) edge (5);
+\end{tikzpicture}
+\end{figure}
+
+\subsubsection{Some things we can say}
+
+\begin{itemize}
+\item If $\boldsymbol{X} = \set{X_1 ...
+X_m}$ and $\boldsymbol{Y} = \set{Y_1...Y_m}$ are disjoint sets, then any partition with $\boldsymbol{X} \subseteq \pt_1$ and $\boldsymbol{Y} \subseteq \pt_2$ will jam $S$.
+This holds trivially if $|S|$ = 1.
+\item The hardest case to jam is when $\boldsymbol{Z}_i = {} \forall i$
+\item If $|S| = 2$ we can always jam $S$ as either $\boldsymbol{X}$ and $\boldsymbol{Y}$ will be disjoint or a variable $A$ will appear in both $S_1$ and $S_2$, in which case we can let $\pt_1 = \set{A}$.
+\end{itemize}
+
+\bibliographystyle{apalike}
+\bibliography{/home/finn/phd/Causality-Identify}
 

### 2016-05-08 causal-bandit/nips_2016/simple-regret.tex
minor tweaks

@@
 reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:causalStructure}.
-All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: 
-\eq{
-	\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
-}
+All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
 In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
@@
 
-In each round the learner either purely observes by selecting $do()$ or chooses a single variable to intervene on and the value to set it to. 
-The remainder of the variables are simultaneously set by independently biased coin flips. 
+In each round the learner either purely observes by selecting $do()$ or sets the value of a single variable. The remaining variables are simultaneously set by independently biased coin flips. 
 The value of all variables are then used to determine the distribution of rewards for that round.
@@
 \begin{wrapfigure}[18]{r}{0.6\textwidth}
-\vspace{-18pt}
+\vspace{-25pt}
 \begin{minipage}{.6\textwidth}

### 2016-05-14 causal-bandit/nips_2016/discussion.tex
first draft

--- 2016-02-05 causal-bandit/icml-2015/discussion.tex
@@
 
+\vspace{-0.4cm} 
 \paragraph{Making Better Use of the Reward Signal}
@@
 
+\vspace{-0.4cm} 
 \paragraph{Cumulative Regret}
@@
 A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
-This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
+This is equivalent to a nominally stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
 
@@
 
+\vspace{-0.4cm} 
 \paragraph{Causal Models with Non-Observable Variables}
-If we assume knowledge of the conditional \textit{interventional} distributions $P_a$ our analysis applies unchanged to the case of causal models with 
-non-observable variables. For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not 
-available from which to build estimates of $P_a$. Some of the interventional distributions $P_a$ may be non-identifiable meaning we can not obtain prior estimates for $P_a$ from 
+If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
+non-observable variables. Some of the interventional distributions may be non-identifiable meaning we can not obtain prior estimates for $\Pn{a}{\parents{Y}}$ from 
 even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
@@
 
+% For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
 % In this case, some conditional distributions may be non-identifiable. 
@@
 
+\vspace{-0.4cm} 
 \paragraph{Partially or Completely Unknown Causal Graph}

### 2016-05-14 causal-bandit/nips_2016/experiments.tex
condences figures, minor changes to section 4
first draft

--- 2016-02-05 causal-bandit/icml-2015/experiments.tex
@@
 under a variety of conditions, including where the importance weighted estimator used by Algorithm \ref{alg:general} is not truncated,
-which is justified in this setting by Remark \ref{rem:truncate}. Throughout we use a model in which $Y$ depends only on a single variable (this is unknown to the algorithms). 
-\eq{
-Y_t \sim \begin{cases}
-\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
-\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
-\end{cases}
-}
-
-This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\left(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon\right)$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
+which is justified in this setting by Remark \ref{rem:truncate}.
+Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms).
+$Y_t \sim \bernoulli(\frac{1}{2}+\epsilon)$ if $X_1=1$ and $Y_t \sim \bernoulli(\frac{1}{2}-\epsilon')$ otherwise, where $\epsilon' = q_1\epsilon/(1-q_1)$.
+This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
 We set $\boldsymbol{q} = 0$ for $q_i \leq m$ and $\frac{1}{2}$ otherwise.
 Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
+
+%\eq{
+%Y_t \sim \begin{cases}
+%\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
+%\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
+%\end{cases}
+%}
 
@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160205_2230}
+  \begin{minipage}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160205_2230}
     \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
     \label{fig:simple_vs_m}
+  \end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
+    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
+    \label{fig:simple_vs_T_vary_epsilon}
+  \end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{exp_regret_vs_T_N50_s10000_20160206_0533}
+    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
+    \label{fig:simple_vs_T}
+  \end{minipage}
 \end{figure}
@@
 This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be minimax optimal asymptotically \citep{LMS14}.
-
-\begin{figure}
-\centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
-\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
-\label{fig:simple_vs_T_vary_epsilon}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=.5\textwidth]{exp_regret_vs_T_N50_s10000_20160206_0533}
-\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
-\label{fig:simple_vs_T}
-\end{figure}

### 2016-05-14 causal-bandit/nips_2016/general.tex
minor tweaks
first draft

@@
 We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, but not from $do()$.
+In addition, variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
+Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$.
+Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot get a good estimate for $\P{Y|do(X_2=1)}$.
 
 To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally allocate samples to actions.
-NEED TO EXPLAIN WHY THIS PROBLEM IS HARD.
-
 To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known.
@@
 
-%Firstly, the variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
-For example, given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$.
-Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot get a good estimate for $\P{Y|do(X_2=1)}$.
-Secondly it is no longer optimal
-
-\begin{wrapfigure}[14]{r}{0.5\textwidth}
-\vspace{-12pt}
+\begin{wrapfigure}[12]{r}{0.5\textwidth}
+\vspace{-25pt}
 \begin{minipage}{.5\textwidth}
@@
 \eq {
-\hat{\mu_a} = \frac{1}{T} \sum_{t=1}^T R_a(X) Y \ind{R_a(X) \leq B_a}\,, 
+\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}\,, 
 }
@@
 \eq{
-m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
+m(\eta) = \max_{a \in \calA} \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,,\text{ where } \EEa \text{ is the expectation with respect to } \Pn{a}.
 }
@@
 \eta^* 
-= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EE_{\Ps_a} \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
+= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
 }
@@
 \begin{proposition}\label{pro:m-bound}
-$m(\eta^*) \leq |\calA|$.
+$m(\eta^*) \leq |\calA|$. \textit{Proof.} 
+\textup{By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|\,\forall a$.}
+\eq{
+m(\eta) 
+= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
+\leq \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
+= \max_a \EEa\left[\frac{1}{\eta_a}\right] = |\calA| %\qedhere
+}
 \end{proposition}
 
-\begin{proof}
-By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|$. Then
-\eq{
-m(\eta) 
-= \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq \max_a \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
-\leq \max_a \EE_{\Ps_a}\left[\frac{1}{\eta_a}\right] = |\calA| \qedhere
-}
-\end{proof}
-
-The second observation is that in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
-Recall in that setting that 
-\eq{
-\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
-}
-For $a = do()$ choose $\eta_a = 1/2$. 
-Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
-Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
-It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.
+The second observation is that, in the parallel bandit setting, $m(\eta) \leq 2m(\boldsymbol{q})$.
+This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q}} / (2m(\boldsymbol{q}))$ for the actions corresponding to $do(X_i=j)$, and applying an argument like Proposition~\ref{pro:m-bound}.
+The proof is in section \ref{} of the supplementary materials.
 

### 2016-05-14 causal-bandit/nips_2016/simple-regret.tex
first draft

@@
 The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a 
-reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:causalStructure}.
+reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:parallel}.
 All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
@@
         \caption{Parallel graph}
-        \label{fig:causalStructure_confounded}
+        \label{fig:parallel}
     \end{subfigure}
@@
     \end{subfigure}
-    \caption{Causal Models}\label{fig:animals}
+    \caption{Causal Models}\label{fig:causalmodels}
 \end{figure}

### 2016-05-15 causal-bandit/nips_2016/discussion.tex
fixes typos in discussion and algorithm 2

@@
 
-The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator terms $Z_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $P_a$ for all $a \in \calA$.
+The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator $\hat{\mu}_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $\Pn{a}{\parents{Y}}$ for all $a \in \calA$.
 In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.

### 2016-05-15 causal-bandit/nips_2016/general.tex
fixes typos in discussion and algorithm 2
fixes reference

@@
 \FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $a_t = a$ with probability $\eta$
+\STATE Sample action $a_t$ from $\eta$
 \STATE Do action $a_t$ and observe $X_t$ and $Y_t$
@@
 The second observation is that, in the parallel bandit setting, $m(\eta) \leq 2m(\boldsymbol{q})$.
-This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q}} / (2m(\boldsymbol{q}))$ for the actions corresponding to $do(X_i=j)$, and applying an argument like Proposition~\ref{pro:m-bound}.
-The proof is in section \ref{} of the supplementary materials.
+This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q}} / (2m(\boldsymbol{q}))$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
 

### 2016-05-17 causal-bandit/nips_2016/general.tex
clarifies action and intervention used interchangably fixes typos after propasition 4

@@
 
-The second observation is that, in the parallel bandit setting, $m(\eta) \leq 2m(\boldsymbol{q})$.
-This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q}} / (2m(\boldsymbol{q}))$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
+The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$.
+This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
 

### 2016-05-17 causal-bandit/nips_2016/introduction.tex
clarifies action and intervention used interchangably fixes typos after propasition 4

@@
 Medical drug testing, policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1959}.
-In this framework, single actions (\eg, experiments or interventions) from a pre-determined set are repeatedly performed in 
+In this framework, single actions (also referred to as interventions) from a pre-determined set are repeatedly performed in 
 order to evaluate their effectiveness via feedback from a single, real-valued reward signal.

### 2016-05-17 causal-bandit/nips_2016/m.py
corrects definition of m(q)

@@ -1,0 +1,34 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue May 17 13:42:54 2016
+
+@author: finn
+"""
+
+import numpy as np
+
+def m1(q):
+    s = np.sort(np.minimum(q,1-q))
+    for indx,value in enumerate(s[1:]): # first element can be anything
+        th =1.0/(indx+1) # if indx=0 th=.5
+        if value >= th:
+            return indx+1
+    return len(s)
+    
+def m2(q):
+    s = np.minimum(q,1-q)
+    for t in range(2,len(q)+1):
+        I_t = (s < 1.0/(t)).sum()
+        if I_t <= t:
+            return t
+
+    
+qs = [np.zeros(10),np.ones(10),np.full(10,.5),np.full(10,.1),np.asarray([0,0,0,0,0,0,0,0,0,.2])]
+for q in qs:
+    print(m1(q),m2(q))
+
+for i in range(10000):
+    q = np.random.uniform(size=20)
+    if m1(q) != m2(q):
+        print("Difference",q)
+

### 2016-05-17 causal-bandit/nips_2016/tech.tex
first draft
corrects definition of m(q)
adds statement and proof of the relationship between m(eta) and m(q)

@@
 \end{proof}
+
+\subsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}
+
+\begin{proposition}\label{pro:m-equivelence} In the parallel bandit setting,
+$m(\eta^*) \leq 2m(\boldsymbol{q})$.
+\end{proposition}
+
+\begin{proof}
+
+Recall that in the parallel bandit setting,
+
+\eq{
+\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}
+}
+
+Let:
+
+\eq {
+\eta_a = \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}}\frac{1}{2m(\boldsymbol{q})} \text { for } a \in do(X_i = j)
+}
+
+Let $D =\sum_{a\in do(X_i=j)}\eta_a$. From the definition of $m(\boldsymbol{q})$, 
+\eq {
+\sum_{a\in do(X_i=j)} \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}} \leq m(\boldsymbol{q}) \implies D \leq \frac{1}{2}
+}
+
+Let $\eta_a = \frac{1}{2} + (1-D)$ for $a = do()$ such that $\sum_{a \in \calA}\eta_a = 1$
+
+Recall that,
+
+\eq{
+m(\eta) &
+= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]
+}
+
+We now show that our choice of $\eta$ ensures $\EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \leq 2m(\boldsymbol{q})$ for all actions $a$.
+
+For the actions $a: \eta_a > 0$, ie $do()$ and $do(X_i = j):\P{X_i=j}<\frac{1}{m(\boldsymbol{q})}$,
+\eq{
+\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq \EEa\left[\frac{\Pn{a}{X_1...X_N}}{\eta_a\Pn{a}{X_1...X_N}}\right] = \EEa\left[\frac{1}{\eta_a}\right] \leq 2m(\boldsymbol{q})
+}
+
+For the actions $a :\eta_a = 0$, ie $do(X_i=j):\P{X_i=j}\geq\frac{1}{m(\boldsymbol{q})}$,
+\eq{
+\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq & \EEa\left[\frac{\ind{X_i=j}\prod_{k\neq i}\P{X_k}}{(1/2+D)\prod_k \P{X_k}}\right] \\=& \EEa\left[\frac{\ind{X_i=j}}{(1/2+D)\P{X_i = j}}\right]
+\leq  \EEa\left[\frac{\ind{X_i=j}}{(1/2)(1/m(\boldsymbol{q}))}\right] \leq 2m(\boldsymbol{q})
+}
+
+\end{proof}
+\
+%in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
+%Recall in that setting that 
+%\eq{
+%\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
+%}
+%For $a = do()$ choose $\eta_a = 1/2$. 
+%Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
+%Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
+%It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.

### 2016-05-17 causal-bandit/nips_2016/general.tex
adds ifsup

@@
 We now consider the more general problem where the graph structure is known, but arbitrary.
-
-% How do actions map to interventions (we need not only consider singleton interventions).
-%
-
 For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
@@
 \end{theorem}
-
+\ifsup 
+The proof is in Section \ref{sec:thm:general}.
+\else
 The proof is in the supplementary materials.
-Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
+\fi Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
 Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$.

### 2016-05-17 causal-bandit/nips_2016/causal_bandit.py
adds code

@@ -1,0 +1,350 @@
+# -*- coding: utf-8 -*-
+
+import numpy as np
+from itertools import product
+from math import sqrt,log,ceil
+from numpy.random import binomial
+import matplotlib.pyplot as plt
+from time import time
+from datetime import datetime as dt
+import pickle
+
+
+REGRET_LABEL = "Regret"
+HORIZON_LABEL = "T"
+M_LABEL = "m(q)"
+
+def now_string():
+    return dt.now().strftime('%Y%m%d_%H%M')
+
+def random_eta(n):
+    eta = np.random.random(n)
+    eta = eta/eta.sum()
+    return eta
+    
+def most_unbalanced_q(N,m):
+    q = np.full(N,1.0/m,dtype=float)
+    q[0:m] = 0
+    return q
+    
+def part_balanced_q(N,m):
+    """ all but m of the variables have probability .5"""
+    q = np.full(N,.5,dtype=float)
+    q[0:m] = 0
+    return q
+        
+def argmax_rand(x):
+    """ return the index of the maximum element in the array, ignoring nans. 
+    If there are multiple max valued elements return 1 at random"""
+    max_val = np.nanmax(x)
+    indicies = np.where(x == max_val)[0]
+    return np.random.choice(indicies) 
+
+def calculate_m(qij_sorted):
+    for indx,value in enumerate(qij_sorted):
+        if value >= 1.0/(indx+1):
+            return indx
+    return len(qij_sorted)/2      
+    
+
+class Parallel(object):
+    def __init__(self,q,epsilon):
+        """ actions are do(x_1 = 0)...do(x_N = 0),do(x_1=1)...do(N_1=1), do() """
+        assert q[0] <= .5, "P(x_1 = 1) should be <= .5 to ensure worst case reward distribution can be created"
+        self.q = q
+        self.N = len(q) # number of variables
+        self.K = 2*self.N+1 #number of actions
+        self.pX = np.vstack((1.0-q,q))
+        self.set_epsilon(epsilon)
+        self.parent_vals = None
+
+        
+    def set_epsilon(self,epsilon):
+        assert epsilon <= .5 ,"epsilon cannot exceed .5"
+        self.epsilon = epsilon
+        self.epsilon_minus = self.epsilon*self.q[0]/(1.0-self.q[0]) 
+        self.expected_rewards = np.full(self.K,.5)
+        self.expected_rewards[0] = .5 - self.epsilon_minus
[…]

### 2016-05-19 causal-bandit/nips_2016/discussion.tex
minor cosmetic sentences and comment on assumption

@@
 Algorithm~\ref{alg:general} for general causal bandit problems adaptively estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
-Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\bigo{\sqrt{m(\eta)/T}}$ and that the parameter $m(\eta)$ is always less than $N$.
+Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ and that the parameter $m(\eta)$ is always less than $N$.
 The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.

### 2016-05-19 causal-bandit/nips_2016/experiments.tex
fixes figure fonts
puts experiemental results in single figure
adds missing y-axis label to experimental results
minor cosmetic sentences and comment on assumption

@@
 \begin{figure}
+    \begin{subfigure}[t]{0.3\textwidth}
 		\centering    
-  \begin{minipage}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160205_2230}
+    		\includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160519_1030.pdf}
     		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
         \label{fig:simple_vs_m}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{exp_regret_vs_T_N50_a4_s10000_20160204_1831}
+    \end{subfigure}\hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+    		\centering
+        \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160519_1030.pdf}
     		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
         \label{fig:simple_vs_T_vary_epsilon}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{exp_regret_vs_T_N50_s10000_20160206_0533}
+    \end{subfigure}\hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+    		\centering
+    		\includegraphics[width=\textwidth]{experiment3_N50_s10000_20160519_1030.pdf}
     		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
     		\label{fig:simple_vs_T}
-  \end{minipage}
+    \end{subfigure}
+    \caption{Experimental results}
+    \label{fig:experiments}
 \end{figure}
 
-Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\epsilon = \sqrt{\frac{N}{8T}}$ , is just too small to be learned.
-This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m << N$.
+%\begin{figure}
+%  \centering
+%  \begin{minipage}[b]{0.3\textwidth}
+%    \includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160518_1534.pdf}
+%    \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
+%    \label{fig:simple_vs_m}
+%  \end{minipage}
+%  \hfill
+%  \begin{minipage}[b]{0.3\textwidth}
+%    \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160518_1534.pdf}
+%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
+%    \label{fig:simple_vs_T_vary_epsilon}
+%  \end{minipage}
+%  \hfill
+%  \begin{minipage}[b]{0.3\textwidth}
+%    \includegraphics[width=\textwidth]{experiment3_N50_s10000_20160518_1534.pdf}
+%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
+%    \label{fig:simple_vs_T}
+%  \end{minipage}
+%\end{figure}
+
+Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
+This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
 In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.

### 2016-05-19 causal-bandit/nips_2016/simple-regret.tex
corrects definition of m(q)
minor cosmetic sentences and comment on assumption

@@
 The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$).
-For $\tau \in [1...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \frac{1}{\tau - 1}}$.
+For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$.
 Define
@@
 \label{eq:m-simple}
-m(\vec{q}) = \min \set{ d : |I_{d}| \leq d}\,.
+m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
 % m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
@@
 $I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret.
-When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
+When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
 We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$.
@@
 By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
-Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\Omega(\sqrt{N/T})$.
+Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$.
 Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting.

### 2016-05-19 causal-bandit/nips_2016/introduction.tex
updates

@@
 We propose a simple causal best arm identification algorithm for this problem and show that up to logarithmic factors it enjoys minimax optimal
-simple regret guarantees of $\tilde\Theta(\sqrt{m/T})$ where $m$ depends on the causal model and may be much smaller than $N$.
-In contrast, existing best arm identification algorithms suffer $\Omega(\sqrt{N/T})$ simple regret (Thm. 4 by \citet{audibert2010best}).
+simple regret guarantees of $\smash{\tilde\Theta(\sqrt{m/T})}$ where $m$ depends on the causal model and may be much smaller than $N$.
+In contrast, existing best arm identification algorithms suffer $\smash{\Omega(\sqrt{N/T})}$ simple regret (Thm. 4 by \citet{audibert2010best}).
 This shows theoretically the value of our framework over the traditional bandit problem. 
@@
 In the general casual bandit problem interventions and observations may have a complex relationship. 
-In \S\ref{sec:simple-regret-general} we propose a new algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
-As in the parallel bandit case, the regret guarantee is often significantly better than $O(\sqrt{N/T})$ and never worse, where $N$ is the number of available interventions.
+In \S\ref{sec:simple-regret-general} we propose a new algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent 
+to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
+As in the parallel bandit case, the regret guarantee scales like $\smash{O(\sqrt{m/T})}$ where $m$ depends on the underlying causal structure, with 
+smaller values corresponding to structures that are easier to learn. The value of $m$ is always less than the number of interventions $N$ and in the special
+case of the parallel bandit (where we have lower bounds) the notions are equivalent.
 

### 2016-05-19 causal-bandit/nips_2016/introduction.tex
updates

@@
 However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
-% We discuss the relationship between causal and contextual bandits further in \S\ref{sec:defs} and leave their combination as future work.
 
 There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-Most recently, \citet{Alon2015}, \citet{Kocak2014} and \citet{wu2015online} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
+Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
 As we discuss in \S\ref{sec:discussion}, the problem they consider shares some similarity with ours however their results are for cumulative regret and thus cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
-% Even in the case of cumulative regret we show that an application of their results to the parallel bandit problem with $N$ variables yields a $\bigo{\sqrt{NT}}$ bound.
+Even without this difference, none
+of these algorithms could be applied to our general setting without modification and even in the parallel bandit setting their regret guarantees would not be optimal.
+The partial monitoring approach taken by \cite{wu2015online}
+could be applied in principle (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood 
+of each factor in advance, while our strategy learns this online.
 \citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.

### 2016-05-19 causal-bandit/nips_2016/discussion.tex
updates

--- 2016-05-15 causal-bandit/nips_2016/discussion.tex
@@
-Algorithm~\ref{alg:general} for general causal bandit problems adaptively estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
-Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\bigo{\sqrt{m(\eta)/T}}$ and that the parameter $m(\eta)$ is always less than $N$.
+Algorithm~\ref{alg:general} for general causal bandit problems 
+estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
+Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ where 
+the parameter $m(\eta)$ measures the difficulty of learning the causal model and is always less than $N$.
 The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.

### 2016-05-19 causal-bandit/nips_2016/introduction.tex
updates related work section

@@
 
-\paragraph{Related Work} As alluded to above, causal bandit problems with $N$ possible interventions can be treated as a classical $N$-armed bandit problem by simply ignoring the causal model and extra observations that are revealed each round.
-In the farming example, the farmer could treat the six possible interventions as arms, ignore the observations of the non-intervened variables and apply an existing best-arm identificaion algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
-However, as we show in \S\ref{sec:simple-regret}, this approach ignores the extra information available in the non-intervened variables and subsequently yields sub-optimal performance.
+\paragraph{Related Work} As alluded to above, causal bandit problems can be treated as classical multi-armed bandit problems by simply ignoring the causal model and extra observations and applying an existing best-arm identification algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
+However, as we show in \S\ref{sec:simple-regret}, ignoring the extra information available in the non-intervened variables yields sub-optimal performance.
 
@@
 Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
-As we discuss in \S\ref{sec:discussion}, the problem they consider shares some similarity with ours however their results are for cumulative regret and thus cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
-Even without this difference, none
-of these algorithms could be applied to our general setting without modification and even in the parallel bandit setting their regret guarantees would not be optimal.
-The partial monitoring approach taken by \cite{wu2015online}
-could be applied in principle (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood 
-of each factor in advance, while our strategy learns this online.
-\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
-They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection. 
+As we discuss in \S\ref{sec:discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
+They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
+The partial monitoring approach taken by \cite{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
+\citet{Yu2009} utilize extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
 \citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled.
-The main difference from our present work is our focus on simple regret and exploitation of the more complex information related rewards via the causal graph.
+The main difference from our present work is our focus on simple regret and the more complex information linking rewards for different arms via causal graphs.
 To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.
 
+%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
+They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
 % Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.

### 2016-05-19 causal-bandit/nips_2016/code/causal_bandit.py

@@ -283,6 +283,7 @@ def plot_regret(regret,xvals):
     ax.errorbar(xvals,mean[:,1],yerr=error[:,1], label="Algorithm 1",linestyle="",marker="o",markersize=6)  
     ax.errorbar(xvals,mean[:,0],yerr=error[:,0], label="Algorithm 2",linestyle="",marker="s",markersize=5) 
     ax.errorbar(xvals,mean[:,2],yerr=error[:,2], label="Successive Rejects",linestyle="",marker="D",markersize=5) 
+    ax.set_ylabel(REGRET_LABEL)
     for item in ([ax.title, ax.xaxis.label, ax.yaxis.label] + ax.get_xticklabels() + ax.get_yticklabels()):
         item.set_fontsize(14)
     return fig,ax
@@ -347,7 +348,7 @@ s = time()
 ## Experiment 1
 N = 50
 epsilon = .3
-simulations = 1000
+simulations = 100
 T = 400
 regret,m_vals,T = regret_vs_m(N,epsilon,simulations,T)
 filename = "experiment1_{0}.pickle".format(now_string())
@@ -361,7 +362,7 @@ plot_regret_versus_m(*plot_params)
 
 ## Experiment 2
 N= 50
-simulations = 1000
+simulations = 100
 regret2,T_vals,N,a = regret_vs_T(N,simulations,epsilon=None,Tstep=None,TperK=10)
 filename = "experiment2_{0}.pickle".format(now_string())
 plot_params = (regret2,T_vals,N,a,2)
@@ -374,7 +375,7 @@ plot_regret_versus_T(*plot_params)
 ### Experiment 3
 N = 50
 epsilon = .3
-simulations = 1000
+simulations = 100
 regret3,T_vals,N,a = regret_vs_T(N,simulations,epsilon=epsilon,Tstep=25,TperK=6)
 plot_params = (regret3,T_vals,N,a,3)
 filename = "experiment3_{0}.pickle".format(now_string())

### 2016-05-19 causal-bandit/nips_2016/code/causal_bandit_cached_figs.py

@@ -285,6 +285,7 @@ def plot_regret(regret,xvals):
     ax.errorbar(xvals,mean[:,1],yerr=error[:,1], label="Algorithm 1",linestyle="",marker="o",markersize=6)  
     ax.errorbar(xvals,mean[:,0],yerr=error[:,0], label="Algorithm 2",linestyle="",marker="s",markersize=5) 
     ax.errorbar(xvals,mean[:,2],yerr=error[:,2], label="Successive Rejects",linestyle="",marker="D",markersize=5) 
+    ax.set_ylabel(REGRET_LABEL)
     for item in ([ax.title, ax.xaxis.label, ax.yaxis.label] + ax.get_xticklabels() + ax.get_yticklabels()):
         item.set_fontsize(14)
     return fig,ax
@@ -352,7 +353,7 @@ T = 400
 #with open("experiment1_{0}.pickle".format(now_string()),"wb") as out:
 #    pickle.dump(regret,out)
 
-with open('data/experiment1_20160518_1339.pickle','rb') as f:
+with open('../data/experiment1_20160518_1339.pickle','rb') as f:
     regret1 = pickle.load(f)
     m_vals = range(2,N,2)
     plot_regret_versus_m(regret1,m_vals,T)
@@ -366,7 +367,7 @@ N= 50
 #with open("experiment2_{0}.pickle".format(now_string()),"wb") as out:
 #    pickle.dump(regret2,out)
 
-with open('data/experiment2_20160518_1206.pickle',"rb") as f:
+with open('../data/experiment2_20160518_1206.pickle',"rb") as f:
     regret2 = pickle.load(f)
     model,eta,mq,T_vals,vary_epsilon,a = setup_regret_vs_T(N,epsilon=None,Tstep=None,TperK=10)
     plot_regret_versus_T(regret2,T_vals,N,a,2)
@@ -380,7 +381,7 @@ epsilon = .3
 #with open("experiment3_{0}.pickle".format(now_string()),"wb") as out:
 #    pickle.dump(regret3, out)
 
-with open('data/experiment3_20160518_0117.pickle',"rb") as f:
+with open('../data/experiment3_20160518_0117.pickle',"rb") as f:
     regret3 = pickle.load(f)
     model,eta,mq,T_vals,vary_epsilon,a = setup_regret_vs_T(N,epsilon=epsilon,Tstep=25,TperK=6)
     plot_regret_versus_T(regret3,T_vals,N,a,3)

### 2016-05-19 causal-bandit/nips_2016/causal_bandit.py
moves code to code directory

@@ -1,350 +1,0 @@
-# -*- coding: utf-8 -*-
-
-import numpy as np
-from itertools import product
-from math import sqrt,log,ceil
-from numpy.random import binomial
-import matplotlib.pyplot as plt
-from time import time
-from datetime import datetime as dt
-import pickle
-
-
-REGRET_LABEL = "Regret"
-HORIZON_LABEL = "T"
-M_LABEL = "m(q)"
-
-def now_string():
-    return dt.now().strftime('%Y%m%d_%H%M')
-
-def random_eta(n):
-    eta = np.random.random(n)
-    eta = eta/eta.sum()
-    return eta
-    
-def most_unbalanced_q(N,m):
-    q = np.full(N,1.0/m,dtype=float)
-    q[0:m] = 0
-    return q
-    
-def part_balanced_q(N,m):
-    """ all but m of the variables have probability .5"""
-    q = np.full(N,.5,dtype=float)
-    q[0:m] = 0
-    return q
-        
-def argmax_rand(x):
-    """ return the index of the maximum element in the array, ignoring nans. 
-    If there are multiple max valued elements return 1 at random"""
-    max_val = np.nanmax(x)
-    indicies = np.where(x == max_val)[0]
-    return np.random.choice(indicies) 
-
-def calculate_m(qij_sorted):
-    for indx,value in enumerate(qij_sorted):
-        if value >= 1.0/(indx+1):
-            return indx
-    return len(qij_sorted)/2      
-    
-
-class Parallel(object):
-    def __init__(self,q,epsilon):
-        """ actions are do(x_1 = 0)...do(x_N = 0),do(x_1=1)...do(N_1=1), do() """
-        assert q[0] <= .5, "P(x_1 = 1) should be <= .5 to ensure worst case reward distribution can be created"
-        self.q = q
-        self.N = len(q) # number of variables
-        self.K = 2*self.N+1 #number of actions
-        self.pX = np.vstack((1.0-q,q))
-        self.set_epsilon(epsilon)
-        self.parent_vals = None
-
-        
-    def set_epsilon(self,epsilon):
-        assert epsilon <= .5 ,"epsilon cannot exceed .5"
-        self.epsilon = epsilon
-        self.epsilon_minus = self.epsilon*self.q[0]/(1.0-self.q[0]) 
-        self.expected_rewards = np.full(self.K,.5)
-        self.expected_rewards[0] = .5 - self.epsilon_minus
[…]

### 2016-05-19 causal-bandit/nips_2016/m.py

@@ -1,34 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Tue May 17 13:42:54 2016
-
-@author: finn
-"""
-
-import numpy as np
-
-def m1(q):
-    s = np.sort(np.minimum(q,1-q))
-    for indx,value in enumerate(s[1:]): # first element can be anything
-        th =1.0/(indx+1) # if indx=0 th=.5
-        if value >= th:
-            return indx+1
-    return len(s)
-    
-def m2(q):
-    s = np.minimum(q,1-q)
-    for t in range(2,len(q)+1):
-        I_t = (s < 1.0/(t)).sum()
-        if I_t <= t:
-            return t
-
-    
-qs = [np.zeros(10),np.ones(10),np.full(10,.5),np.full(10,.1),np.asarray([0,0,0,0,0,0,0,0,0,.2])]
-for q in qs:
-    print(m1(q),m2(q))
-
-for i in range(10000):
-    q = np.random.uniform(size=20)
-    if m1(q) != m2(q):
-        print("Difference",q)
-

### 2016-05-19 causal-bandit/nips_2016/general.tex
minor cosmetic sentences and comment on assumption
status as submitted Thursday

@@
 
-We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational estimates were poor.
-However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
-Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$.
-We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, but not from $do()$.
-In addition, variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
-Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$.
-Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot get a good estimate for $\P{Y|do(X_2=1)}$.
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to 
+write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
+estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
+from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
+takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
+simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
+high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
+Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
+get a good estimate for $\P{Y|do(X_2=1)}$.
 
-To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally allocate samples to actions.
-To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known.
-These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, or similarly from observational data subject to identifiability constraints.
+To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
+allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
+are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
+or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
+start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
 Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
 Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
-
-%we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$.
-$\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$.
-We then
 
@@
 The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$.
-This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
+This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.\ifsup
+The proof is in Section \ref{sec:m-equivelence}.
+\else
+The proof is in the supplementary materials.
+\fi
 

### 2016-05-19 causal-bandit/nips_2016/tech.tex
adds definitions to proof of theorem 3
status as submitted Thursday

@@
 are independent and permutations of the variables can be pushed to the reward function.
-The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound.
+The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound (see Exercise 2.10 in \cite{boucheron2013concentration} for this form of the Chernoff bound).
 
@@
 First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
-We abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
+W define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
 By definition we have $|Z_{at}| \leq B_a$ and 
@@
 \leq \EE_Q[Z_{at}^2] 
-\leq \EE_{\Ps_a}\left[\frac{\Ps_a(\parents{Y}(X))}{\Q{\parents{Y}(X)}}\right] 
+\leq \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
 \leq m(\eta)\,.
@@
 \EE_Q[Z_{at}] 
-= \EE_{\Ps_a}\left[Y \ind{R_{at} \leq B_a}\right] 
-= \EE_{\Ps_a} Y - \EE_{\Ps_a} \left[Y\ind{R_a > B_a}\right] 
+= \EEa \left[Y \ind{R_{at} \leq B_a}\right] 
+= \EEa Y - \EEa \left[Y\ind{R_a > B_a}\right] 
 = \mu_a - \beta_a\,,
@@
 \eq{
-0 \leq \beta_a = \EE_{\Ps_a}[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}
+0 \leq \beta_a = \EEa[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}
 }
@@
 \eq{
-\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EE_{\Ps_a}[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
+\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EEa[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
 }
@@
 
-\subsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}
+\subsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}\label{sec:m-equivelence}
 
-\begin{proposition}\label{pro:m-equivelence} In the parallel bandit setting,
+\begin{proposition} In the parallel bandit setting,
 $m(\eta^*) \leq 2m(\boldsymbol{q})$.
@@
 
+Therefore $m(\eta*) \leq m(\eta) \leq 2m(\boldsymbol{q})$ as required.
+
 \end{proof}

### 2016-05-19 causal-bandit/nips_2016/code/m.py

@@ -23,12 +23,13 @@ def m2(q):
             return t
 
     
-qs = [np.zeros(10),np.ones(10),np.full(10,.5),np.full(10,.1),np.asarray([0,0,0,0,0,0,0,0,0,.2])]
-for q in qs:
-    print(m1(q),m2(q))
+#qs = [np.zeros(10),np.ones(10),np.full(10,.5),np.full(10,.1),np.asarray([0,0,0,0,0,0,0,0,0,.2])]
+#for q in qs:
+#    print(m1(q),m2(q))
+#
+#for i in range(10000):
+#    q = np.random.uniform(size=20)
+#    if m1(q) != m2(q):
+#        print("Difference",q)
 
-for i in range(10000):
-    q = np.random.uniform(size=20)
-    if m1(q) != m2(q):
-        print("Difference",q)
-
+print m2(np.asarray([0,0,0,.5,.5]))

### 2016-05-20 causal-bandit/nips_2016/tech.tex
fixes reference for chernof bound

@@
 are independent and permutations of the variables can be pushed to the reward function.
-The proof of Theorem \ref{thm:uq-simple} requires some lemmas, the first of which is an immediate consequence of the Chernoff bound (see Exercise 2.10 in \cite{boucheron2013concentration} for this form of the Chernoff bound).
+The proof of Theorem \ref{thm:uq-simple} requires some lemmas. \
 
@@
 
+\begin{proof}
+By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \cite{hagerup1990guided}),
+
+\eq{
+\P{\left|\hat q_i - q_i\right| \geq \epsilon} \leq 2e^{-\frac{T\epsilon^2}{6q_i}}
+}
+
+Letting $\delta = 2e^{-\frac{T\epsilon^2}{6q_i}}$ and solving for $\epsilon$ completes the proof.
+
+\end{proof}
+
 \begin{lemma}\label{lem:conc2}
-Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE X_i = p$ and $\delta \in [0,1]$.
+Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE[X_i] = p$ and $\delta \in [0,1]$.
 Then 

### 2016-06-08 TPR/TPR_report.tex
cleaned up structure of causality folder
adds initial writup of contextual-bandit and identifiability problems

--- 2016-01-04 TPR_report.tex
@@
 
-\section{Introduction}
+\section{Problem Summary}
 
@@
 
+\part{Literature review}
 \section{Causal Inference}
@@
 
+\part{Approach and expected outcomes}
+
+The goal of my research is to explore a number of questions around learning to make good decisions.
+
+\begin{itemize}
+\item Develop a framework to unify concepts causal inference concepts with reinforcement learning, starting with multi-armed bandits.(DONE - see section \ref{sec:unificaition})
+\item Demonstrate combining concepts from causal inference with multi-armed bandit algorithms can improve on state of the art performance in sequential decision making (DONE - paper submitted for NIPS 2016)
+\item Causal privacy paper
+
+\item something about identifiability
+\end{itemize}
+
 \section{Unifying the frameworks}
+\label{sec:unificaition}
 
@@
 
-\subsection{Incorporating side information induced by casual graph}
-
-As a first step towards more general problems, we develop and evaluate a bandit algorithm that incorporates the side information induced by a specific causal graph.
-
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
+%\subsection{Incorporating side information induced by casual graph}
+%
+%As a first step towards more general problems, we develop and evaluate a bandit algorithm that incorporates the side information induced by a specific causal graph. 
+%
+%Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}.
 This is an expansion of the model shown in figure \ref{fig:unify_frameworks}.
 We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.
-
-As an example, consider a farmer wishing to optimize the yield of her crop.
+%
+%As an example, consider a farmer wishing to optimize the yield of her crop.
 She can invest in a green house to control temperature, a watering system to control soil moisture, fertilizers to set soil nutrients, etc.
@@
 This repeats across many growing seasons, and the goal is to maximize the total cumulative yield.
-
-\begin{figure}[h]
-\centering
-\caption{Assumed Causal Structure}
-\label{fig:causalStructure}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}, hidden/.style={empty},tg/.style={target}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[tg, below right=of 2](5){Y};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-
-\end{tikzpicture}
-\end{figure}
-
-Let $\boldsymbol{q} \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known horizon $T$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{0,\ldots, N}$ and $J_t \in \set{0,1}$, where $I_t = 0$ represents $do()$.
-
-\item Variables not set by the learner take their values probabilistically according to $\boldsymbol{q}$ 
-\eq{
-X_{t,i} = \begin{cases}
-\sim \Ber(q_i) &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner observes $(\boldsymbol{X}_{t},Y_{t})$ and receives reward $Y_t \sim \Ber(r(\boldsymbol{X}_{t}))$ where $r:\set{0,1}^N \to [0,1]$ is unknown and arbitrary. 
-\end{enumerate}
-
-The problem could be treated as a standard bandit problem with $K = 2N + 1$ arms,
-
-\eq {
-\{do(), do(X_1=0), do(X_1 = 1)\;\; ... \;\;do(X_N = 1)\}
-}
-
-yielding a regret $\bigo{\sqrt{TN}}$. However, this doesn't leverage the side information induced by the causal structure:
-
-\eqn {
-\label{eq:observe}
-P(Y_t|do(X_{i,t} = j)) &= P(Y_t|do(),X_{i,t} = j) \\
-\label{eq:estimation_transfer}
-&= P(Y_t|do(X_{l,t}=0),X_{i,t}=j)(1-q_l)+P(Y_t|do(X_{l,t}=1),X_{i,t}=j)q_l 
-}
-
-The hardness of this problem depends strongly on the vector $\boldsymbol{q}$.
+%
+%\begin{figure}[h]
+%\centering
+%\caption{Assumed Causal Structure}
+%\label{fig:causalStructure}
+%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+%  thick,main node/.style={observed}, hidden/.style={empty},tg/.style={target}]
+%
+% %nodes
+%\node[main node](1){$X_{1}$};
+%\node[main node, right=of 1](2){$X_{2}$};
+%\node[hidden, right=of 2](3){$...$};
+%\node[main node, right=of 3](4){$X_{N}$};
+%\node[tg, below right=of 2](5){Y};
+% \path[every node/.style={font=\sffamily\small}]
+%    (1) edge (5)
+%    	(2) edge (5)
+%    (4) edge (5);
+%	
+%\end{tikzpicture}
+%\end{figure}
+%
+%
+%Let $\boldsymbol{q} \in [0,1]^N$ be a fixed vector where $q_i = P(X_i = 1)$. In each time-step $t$ upto a known horizon $T$:
+% 
+%\begin{enumerate}
+%\item The learner chooses an $I_t \in \set{0,\ldots, N}$ and $J_t \in \set{0,1}$, where $I_t = 0$ represents $do()$.
+%
+%\item Variables not set by the learner take their values probabilistically according to $\boldsymbol{q}$ 
+%\eq{
+%X_{t,i} = \begin{cases}
+%\sim \Ber(q_i) &\text{if } i \neq I_t \\
+%J_t & \text{otherwise}\,.
+%\end{cases}
+%}
+%\item The learner observes $(\boldsymbol{X}_{t},Y_{t})$ and receives reward $Y_t \sim \Ber(r(\boldsymbol{X}_{t}))$ where $r:\set{0,1}^N \to [0,1]$ is unknown and arbitrary. 
+%\end{enumerate}
+%
+%The problem could be treated as a standard bandit problem with $K = 2N + 1$ arms, 
+%
+%\eq {
+%\{do(), do(X_1=0), do(X_1 = 1)\;\; ... \;\;do(X_N = 1)\}
+%}
+%
+%yielding a regret $\bigo{\sqrt{TN}}$. However, this doesn't leverage the side information induced by the causal structure: 
+%
+%\eqn {
+%\label{eq:observe}
+%P(Y_t|do(X_{i,t} = j)) &= P(Y_t|do(),X_{i,t} = j) \\
+%\label{eq:estimation_transfer}
+%&= P(Y_t|do(X_{l,t}=0),X_{i,t}=j)(1-q_l)+P(Y_t|do(X_{l,t}=1),X_{i,t}=j)q_l 
+%}
+%
+%The hardness of this problem depends strongly on the vector $\boldsymbol{q}$.
 This is intuitive from a consideration of the extreme cases.
@@
 If $\boldsymbol{q} = \boldsymbol{\frac{1}{2}}$, the observe action, $do()$, effectively becomes a revealing action and we would anticipate obtaining a regret with minimal dependence on the number of variables $N$.
-
-We consider a simple explore-exploit based algorithm, that will explore for $h$ timesteps, sampling actions in a way that depends on $\boldsymbol{q}$.
+%
+%We consider a simple explore-exploit based algorithm, that will explore for $h$ timesteps, sampling actions in a way that depends on $\boldsymbol{q}$.
 We then select the arm with the highest expected reward for the remaining $T-h$ time steps.
 During the exploration phase, we will balance purely observing, $I_t = 0$, which takes advantage of equation \ref{eq:observe} to reveal the reward for frequently occurring arms, with explicit sampling of infrequently occurring arms.
-
-Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ... \leq q_N$. Let:
-
-\eqn {
-m \in [2,N] = \set{m:q_m > \frac{1}{m}}
-}
-
-$m$ is a measure of the number of infrequently occurring arms.
+%
+%Without loss of generality, we can assume $q_i \in [0,\frac{1}{2}]$ and $q_1 \leq q_2 ... \leq q_N$. Let:
+%
+%\eqn {
+%m \in [2,N] = \set{m:q_m > \frac{1}{m}}
+%}
+%
+%$m$ is a measure of the number of infrequently occurring arms.
 If the problem is completely balanced $q_1...q_N = \frac{1}{2}$ then $m = 2$.
@@
 Let $q_{ij} = \P{X_i = j}$ and $\mu_{ij} = P(Y|do(X_i = j))$.
-
-Suppose we select $I_t = 0$ for the first $h/2$ timesteps.
+%
+%
+%Suppose we select $I_t = 0$ for the first $h/2$ timesteps.
 This is at worst half the optimal.
 We use these observations to estimate the rewards for all the frequently occurring arms, $\set{do(X_i = 0) \; \forall i}$, and $\set{X_i = 1:i \geq m}$
-
-\eqn {
-\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
-}
-
-We then explicitly play each of the $m$ infrequently occurring arms $\frac{h}{2m}$ times and estimate their rewards as:
-
-\eqn {
-\hat{\mu}_{ij} = \frac{2m}{h}\sum_{t = 1}^{h/2m}\ind{Y_t = 1|X_i=j}
-}
-
-We then simply play the arm with the highest estimated reward for the remaining $T-h$ timesteps.
+%
+%\eqn {
+%\hat{\mu}_{ij} = \frac{\sum_{t=1}^{h/2}\ind{Y_t=1,X_{t,i}=j}}{\frac{h}{2}q_{ij}}
+%}
+%
+%We then explicitly play each of the $m$ infrequently occurring arms $\frac{h}{2m}$ times and estimate their rewards as:
+%
+%\eqn {
+%\hat{\mu}_{ij} = \frac{2m}{h}\sum_{t = 1}^{h/2m}\ind{Y_t = 1|X_i=j}
+%}
+%
+%We then simply play the arm with the highest estimated reward for the remaining $T-h$ timesteps.
 We can show that we can select an $h$ (based on $T$, $N$ and $m$) that leads to worst case regret $R_T = \bigotilde{T^{2/3}m^{1/3}}$.
@@
 \footnote{ $\bigotilde{.}$ and $\bigthetatilde{.}$ suppress polylogarithmic factors}
-
-In this algorithm, we have only used the side information provided by the $do()$ action about other actions.
+%
+%
+%In this algorithm, we have only used the side information provided by the $do()$ action about other actions.
 Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
@@
 This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$
-
-\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
+%
+%\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
 In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
 \cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
+%
+%More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
 In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
@@
 This more complex side information does not fit within the graph feedback framework.
-
-Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
+%
+%Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
 It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
 Partial monitoring algorithms yield results that are optimal with respect the the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
+%
+%
+%
+%\subsection{Open questions}
+%
+%There are many open questions still to be addressed.
+For the specific setting described above I need to derive lower bounds for the worst case regret so as to determine how close the current approach is to optimal.
+%
+%It is clearly desirable to obtain an algorithm that could be generalized to an arbitrary causal graph.
+This will likely require the development of a bandit algorithm that can leverage more complex forms of side information like that in equation \ref{eq:estimation_transfer}, in addition to an efficient algorithm for eliminating actions that need not be considered at all.
+%
+%Another open class of problem to consider are cases where the causal structure is not (fully) known and also needs to be learnt within the online setting. 
+%
 
-\subsection{Open questions}
-
-There are many open questions still to be addressed.
-For the specific setting described above I need to derive lower bounds for the worst case regret so as to determine how close the current approach is to optimal.
-
-It is clearly desirable to obtain an algorithm that could be generalized to an arbitrary causal graph.
-This will likely require the development of a bandit algorithm that can leverage more complex forms of side information like that in equation \ref{eq:estimation_transfer}, in addition to an efficient algorithm for eliminating actions that need not be considered at all.
-
-Another open class of problem to consider are cases where the causal structure is not (fully) known and also needs to be learnt within the online setting.
-
-\section{Paid work statement}
-I am not currently undertaking or planning to undertake any paid work.
-
-\bibliographystyle{unsrt}% Select the citation style e.g. ieeetr
+\bibliographystyle{apalike}% Select the citation style e.g. ieeetr
 \bibliography{../library}% write the directory to the .bib file

### 2016-06-08 causal-bandit/arXiv/setup.tex
adds initial writup of contextual-bandit and identifiability problems

--- 2016-05-05 causal-bandit/nips_2016/setup.tex
@@
 
-A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed interventions} $\mathcal{A}$.
+A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed actions} $\mathcal{A}$.
 One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
-We denote the expected reward for the intervention $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
+We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
 the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$. 
@@
 We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
-The set of allowed interventions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
+The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
 Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.

### 2016-06-08 causal-bandit/arXiv/tech.tex
adds initial writup of contextual-bandit and identifiability problems

--- 2016-05-20 causal-bandit/nips_2016/tech.tex
@@
 First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
-W define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$ and $R_{at} = R_a(X_t)$.
+We define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$, $R_{at} = R_a(X_t)$ and $\Pn{a}{.} = \Pns{a}{.}$.
 By definition we have $|Z_{at}| \leq B_a$ and 
@@
 \leq \EE_Q[Z_{at}^2] 
-\leq \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
+\leq \EE_Q[R_{at}^2]
+= \EEa[R_{at}]
+= \EEa\left[\frac{\Pns{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
 \leq m(\eta)\,.
@@
 = \EEa \left[Y \ind{R_{at} \leq B_a}\right] 
-= \EEa Y - \EEa \left[Y\ind{R_a > B_a}\right] 
+= \EEa Y - \EEa \left[Y\ind{R_{at} > B_a}\right] 
 = \mu_a - \beta_a\,,
@@
 \eq{
-0 \leq \beta_a = \EEa[Y \ind{R_a > B_a}] \leq \Pn{a}{R_a(X) > B_a}
+0 \leq \beta_a = \EEa[Y \ind{R_{at} > B_a}] \leq \Pns{a}{R_{at} > B_a}
 }
@@
 \eq{
-\beta_a \leq \Pn{a}{R_a(X) > B_a} \leq \frac{\EEa[R_a(X)]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
+\beta_a \leq \Pns{a}{R_{at} > B_a} \leq \frac{\EEa[R_{at}]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
 }
@@
 \eq{
-\epsilon = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
+\epsilon_a = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
 }
@@
 \eq{
-\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} 
-\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[\hat \mu_a]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
+\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} 
+\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
 }
-Assuming this event does not occur and letting $a^* = \argmax_{a \in \calA} \mu_a$ we have
+
+Let $I = \hat{a}^*_T$ be the action selected by the algorithm, $a^* = \argmax_{a \in \calA} \mu_a$ be the true optimal action and recall that $\EE_Q[Z_{at}] = \mu_a - \beta_a$.
+Assuming the above event does not occur we have,
+
 \eq{

### 2016-06-08 contextual-bandit/barianboim_bandits.tex
adds initial writup of contextual-bandit and identifiability problems

\documentclass{article}
\usepackage{tikz} 
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage[margin=0.7in]{geometry}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multicol,caption}

\usetikzlibrary{arrows,positioning} 
\thispagestyle{empty}
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=2.5em,
           minimum height=2.5em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.5em,
           minimum height=2.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
            empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }

}

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\E}[1]{\mathbb E\left[#1\right]}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\begin{document}

\section{Contextual bandit related problems}

The confounded causal bandit problem, as introduced by \cite{Bareinboim2015}, is illustrated in figure \ref{fig:causalStructure}. At each timestep $t$, $u_t$, $\boldsymbol{r_t}$ and $a'_t$ are generated stochastically. Only $a'$ is observed. The algorithm then selects an action $a_t \in \set{1,...,k}$ and observes the corresponding reward $r_{a_t}$. $a' \in \set{1,...,k}$ corresponds to the action that would have been selected if action choice were not decided by the algorithm. This corresponds exactly to an instance of the standard stochastic contextual bandit problem, with context $x_t = a'_t$. See the following definitions from \cite{Langford2008}

\begin{figure}[h]
\centering
\caption{Causal model for confounded bandits}
\label{fig:causalStructure}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]
 %nodes
\node[latent](1){$u$};
\node[latent, below =of 1](2){$\boldsymbol{r}_t = [r_1,...,r_k] = f(u,\epsilon)$};
\node[main node,  right=of 2](3){$a$};
\node[main node,  left=of 2](4){$a' = g(u,\epsilon')$};
\node[main node,  below=of 2](5){$r_a$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (4)
    (2) edge (5)
    	(3) edge (5);
\end{tikzpicture}
\end{figure}

\begin{definition}[Contextual bandit problem]
In a contextual bandits problem, there is a distribution
$P$ over $(x, r_1, ..., r_k)$, where $x$ is context, $a \in \set{1...k}$ is one of the $k$ arms to be pulled,and $r_a \in [0,1]$ is the reward for arm $a$. The problem is a repeated game: on each round, a sample
$(x, r_1, ..., r_k)$ is drawn from $P$, the context $x$ is announced, and then for precisely one arm a chosen
by the player, its reward $r_a$ is revealed.
\end{definition}

\begin{definition}[Contextual bandit algorithm] A contextual bandit algorithm $\mathcal{B}$ determines an arm $a \in \set{1...k}$ to pull at each timestep $t$, based on the previous observation sequence $(x_1, a_1,r_{a,1})$, and the current context $x_t$.
\end{definition}

\begin{definition}[Regret] Assume we have a hypothosis set $\mathcal{H}$ consisting of hypotheses $h: \mathcal{X} \rightarrow \set{1,...,k}$
The regret of an algorithm $\mathcal{B}$ with respect to $h$ is
\eq{
\Delta R(h,\mathcal{B},T) = T \E{r_{h(x)}} - \E{\sum_{t=1}^T r_{\mathcal{B}(x),t}}
}
and the (psudo) regret of $\mathcal{B}$ with respect to the hypothesis space $\mathcal{H}$ is 
\eq{
\Delta R(\mathcal{H},\mathcal{B},T) = \sup_{h \in \mathcal{H}}\Delta R(h,\mathcal{B},T)
}
\end{definition}

Note - that the definition of a contextual bandit problem does not require that the context is a direct cause of the outcome.

\section{Questions}
Two questions arose in our discussion of this problem.
\subsection{Are noisy features more problematic in the bandit setting}
Intuitively adding features that are noisy (ie only weakly related to the target) is more problematic in the bandit setting than in the batch learning setting because in batch learning, we can apply feature selection or regularization and tune it via cross-validation. It is unclear how to do this (or model/parameter selection in general) in the contextual bandit setting.

Some more concrete version of the question. 
\begin{itemize}
\item Assume we are in the confounded causal bandit setting and $Y_t \in \set{0,1}$ Suppose we make no assumptions about $f$ track the rewards separately for each value of $a'_t$. Under what conditions (on $g$, $f$ and $T$) does including the context have lower regret than applying a standard bandit algorithm.

\item What happens if we add context that is entirely uncorrelated with the target. Ie a contextual bandit where the context $\boldsymbol{x} = [x_1,...x_m]$ but $Y = f(x_1...x_d,\epsilon)$ where $d < m$. How does the regret depend on $m$ and $d$ (with or witout assumptions such as linearity on f). This paper may be relevant \cite{abbasi2012online}.

\item How can we quantify dependence without reference to a model? KL divergence between $P(X,Y)$ and $P(Y)$ averaged over $X$? max over possible kernelized fits (as in kernelized conditional independence testing)? For linear models, all captured in the inverse correlation matrix.

\end{itemize}

\subsection{How should we represent the context} 
In the confounded bandit problem, $x$ is discrete with $k$ unique values. We could:
\begin{enumerate}
\item Treat it as a categorical feature
\item Treat each value of $x$ as a task and apply a multi-task algorithm
\item Treat it as a structured output prediction problem (how)?
\end{enumerate}

Its hard to find a quantitative definition of multi-task learning. One variant involves parameterising models for each task in terms of some weights and putting a joint constraint over the weights of all tasks \ref{}. This is equivalent to treating the tasks as a categorical variable and regularizing for some choices of constraint, and regularizer (given conditions on the loss-function).

The feature-mappings for structured output prediction involve the label - so at first pass this seems different -but I haven't properly understood how this approach works and how the problem can be represented as a structured output prediction problem.

\bibliography{Reinforcment-contextual}
\bibliographystyle{apalike}
\end{document}

### 2016-06-08 identify/identify.tex
adds initial writup of contextual-bandit and identifiability problems

--- 2016-02-02 causal-bandit/icml-2015/causal-bandits-icml.tex
@@
 
-\newif\ifsup\suptrue
+\newif\ifsup\supfalse
 
@@
 \newcommand{\theHalgorithm}{\arabic{algorithm}}
-\usepackage{icml2015stylefiles/icml2015} 
 \usepackage{tikz}
@@
 
-\setlength{\marginparwidth}{10ex}%added by Yifan so that the todos can fit in the margin
+%\setlength{\marginparwidth}{10ex}%added by Yifan so that the todos can fit in the margin
 %\usepackage[obeyFinal]{todonotes}
-\usepackage{todonotes}
+\usepackage[disable]{todonotes}
 
@@
 \newcommand{\KL}{\operatorname{KL}}
-\newcommand{\simpleregret}{R_T^{\text{simple}}}
+\newcommand{\simpleregret}{R_T}
 \newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
@@
 \newcommand{\eg}{\textit{e.g.}}
+\newcommand{\latent}{\mathbf{L}}
+\newcommand{\observed}{\mathbf{O}}
+\newcommand{\variables}{\mathbf{V}}
 
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Causal Bandits}
 
 \begin{document}
-
-\twocolumn[
-\icmltitle{Causal Bandits: Learning Good Interventions via Causal Inference}
-% ALTERNATIVES:
-% - Causal Bandits: Sequential Decision Making with Causal Models
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2015
-% package.
-\icmlauthor{Your Name}{email@yourdomain.edu}
-\icmladdress{Your Fantastic Institute,
-            314159 Pi St., Palo Alto, CA 94306 USA}
-\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
-\icmladdress{Their Fantastic Institute,
-            27182 Exp St., Toronto, ON M6H 2T1 CANADA}
-
-% You may provide any keywords that you 
-% find helpful for describing your paper; these are used to populate 
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{causal,bandit}
-
-\vskip 0.3in
-]
-
-\begin{abstract} 
-We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
-Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
-We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
-than algorithms that do not use the additional causal information.
-\end{abstract}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-\label{sec:intro}
-\input{introduction}
+\section*{Identifiability}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PROBLEM SETUP
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Problem Setup}
-\label{sec:defs}
-\input{setup}
+We have three questions relating to the general idea of identifiability in causal systems.
+For all cases, assume we are working with a causal directed acyclic graph.
+Initially, lets only consider deterministic interventions on single variables in a DAG (not as restrictive as it seems since non-deterministic/multiple interventions can still be modelled by adding nodes to the graph).
+More formally, we have a causal directed acyclic graph $G$ over variables $\variables = \observed \cup \latent$ where $\observed$ are observable and $\latent$ are latent.
+Let the set of causal queries $\mathcal{Q}$ be $P(Y|do(X))$ for all possible pairings of variables $X$,$Y$ in $\observed$.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% UPPER BOUND ON SIMPLE REGRET
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Regret Bounds for Parallel Bandit}
-\label{sec:simple-regret}
+\begin{definition}[Graphical identifiability] a query $Q = P(Y|do(X))$ is identifiable iff we can write an expression for it in terms of conditional probabilities of variables in $\observed$. 
+\end{definition}
 
-\input{simple-regret}
+If a query is graphically identifiable then we can asymptotically obtain the interventional distribution of $Y$ given $do(X=x)$ for any value of $x$ in any probability distribution $P$ compatible with $G$ (with some assumptions on probabilities not being 0).
+Any desired metric (eg difference in mean, KL divergence) can be applied to the pre and post-interventional distributions to obtain a summary of the \textit{causal effect}.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% UPPER BOUND ON SIMPLE REGRET GENERAL CASE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Regret Bounds for General Graphs}
-\label{sec:simple-regret-general}
-\input{general}
+\subsection*{Partial identifiability}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% EXPERIMENTS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Experiments}
-\label{sec:experiments}
-\input{experiments}
+The graphical identifiability criterion is boolean. There are two directions we can come at trying to smooth this boundary.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% DISCUSSION
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\begin{enumerate}
+\item Quantify the difficulty in identifying causal effects for problems that are graphically identifiable.
+What is the variance in the estimators at finite time?
+What is the minimal change in the graph that makes it non-identifiable?
 
-\input{discussion}
+\item Quantify how well we can identify causal effects for problems that are not graphically identifiable.
+For example, how tight are the bounds we can get, what is the minimal change in the graph that would make the problem identifiable (eg observe nodes in $\mathcal{U}$ or delete edges),what minimal assumptions must we make about functional relationships between variables to get a good estimate.
+\end{enumerate}
 
-\bibliography{library,library2}
-\bibliographystyle{icml2015}
+\subsection*{Causal Privacy}
 
-\onecolumn
+Closely related to the question of partial identifiability. What changes can we make that take a problem from identifiable to non-identifiable.
 
-\input{tech}
+\subsection*{Characterizing graphical identifiability}
+
+There is an algorithm that determines if a given query in a causal graph is identifiable or not.
+\ref{} There are also a number of simple graphical criteria that can be used to determine if a graph is or is not identifiable (but no single simple necessary and sufficient criteria).
+These are:
+
+\begin{itemize}
+\item The backdoor criteria (sufficient)
+\item The frontdoor criteria (sufficient)
+\item (nesseesary)
+\end{itemize}
+
+In practise, the back-door criterion is used to estimate causal effects in the vast majority of applications. The front door criterion is also used occasionally.
+
+The question is - can we divide the space of causal queries into sets depending on how they can (or cannot) be identified and estimate the size of the each set?
+Are there interesting features of graphs that predict identifiability?
+
+A given query can be identified in more than one way.
+We could try to determine all the ways in which a query could be identified or apply a hierarchy, for example: identifiable by backdoor, identifiable by frontdoor (but not backdoor, identifiable but not via backdoor or frontdoor, not identifiable.
+
+There are three key components to this problem.
+Identifying a space of graphs, traversing the space in an efficient way and algorithms to compute the quantities we are interested in.
+There are a very large number of possible causal graphs, even  if we restrict the number of nodes.
+There are also a number of ways of defining the space of graphs (for example all DAGs versus all MAGs).
+The existing Identify algorithm \ref{} determines if a query is identifiable or not - and returns an expression if it its.
+It does not enumerate all the expressions via which a query could be identified.
+There are algorithms that can determine if a query can be identified via a backdoor criterion \ref{}.
+No such algorithm that I know of for the front-door criterion.
+Ideally we want an algorithm that can traverse the space of graphs and efficiently utilize information from the previous (related query) to compute identifiability of the next query.
+
+Related questions - 
+\begin{itemize}
+\item Are the expressions that the Idenfiy algorithm outputs optimal?
+Ie - do backdoor criterion adjustments with fewer variables get emmitted in preference to backdoor adjustments with larger numbers of variables?
+\item If there are multiple approaches to identify a given query, how do we define which is optimal? What assumptions do we need to make about the data to define optimality? 
+\item How does one efficiently compute causal effects from such expressions \ref{Richardson, etc}?
+\end{itemize}
 

### 2016-06-08 privacy/causal_differential_privacy.tex
adds initial writup of contextual-bandit and identifiability problems

@@
 
-\begin{lemma}
-One of the following holds.
-\begin{enumerate}
-\item $P(Y|do(X)) = P(Y)$
-\item There is no set of variables that satisfy the back-door criterion for identifying $P(Y|do(X))$ 
-\item There is at least one set of variables $Z$ that are a minimal adjustment for $P(Y|do(X))$, $Z$ may be the empty set. 
-\end{enumerate}
+%Causal effects differ from non-causal ones due to bias introduced by confounding variables, which effect both treatment assignment and outcome. For example, ...
+
+Adjusting for nuisance or confounding variables, either by matching or regression, has been a key technique in inferring causal effects at least as far back as \cite{}.
+The widely used back-door criterion \cite{Pearl2000} clarifies which variables it is appropriate to condition on in order to achieve unbiased estimates of causal effects.
+
+\begin{lemma}Where $X$ and $Y$ are singleton variables, $Z$ is a minimal covariate adjustment for identifying the causal effect of $X$ on $Y$ if and only if $Z$ is a minimal set satisfying the backdoor-criterion relative to $X$ and $Y$ \cite{textor2012adjustment}.
 \end{lemma}
 
-For the remainder we assume that all queries we wish to jam are of case 3, because in case 1, there is no causal effect of $X$ on $Y$ to discover and in case 2, the causal effect of $X$ on $Y$ can already not be inferred from the data and no way in which we partition the data will render it identifiable.
+The existence of a set covariate adjustment is sufficient but not necessary for identifiability. A causal query can also be identified
+
+We will consider non-parametric point identifiability as described by \cite{Pearl2000}.
+
+The do-calculus provides a complete framework to determine if and how a causal effect may be  identified from observational data.
+The framework assumes that the data is generated by a directed acyclic graph (DAG) and that the structure of the graph is known (although some variables may not be observed).
+
+is a sufficient criterion for identification of causal effects via matching.
+More recently it has been generalized to provide necessary and sufficient critera for identification via matching for DAGs \cite{}, and MAGs PAGs \cite{}.
+
+More recently, \cite{Pearl2000} developed the do-calculus, which provides a complete framework for determining if a causal query can be identified from observational data.
+
+\begin{definition}
+Identifiability. 
+A query $P(Y|do(X))$ is identifiable in a causal graph $G$ if it can be computed uniquely from any positive probability distribution over the observed variables - that is, we can compute an expression for $P(Y|do(X))$ that contains only factors of the observed distribution.
+\end{definition}
+\begin{enumerate}
+\item $P(Y|do(X)) = P(Y)$. 
+\label{cat:noncausal}
+\item $P(Y|do(X)) = \sum_Z P(Y|X,Z)P(Z)$
+\label{cat:adjust}
+\item $P(Y|do(X))$ is identifiable but not via (1) or (2)
+\label{cat:identify_other}
+\item $P(Y|do(X))$ is not identifiable
+\label{cat:not_identifiable}
+\end{enumerate}
+
+In this paper we focus on jamming queries that fall into case \ref{cat:adjust}.
+Case \ref{cat:noncausal} cannot be jammed as the SPCP does not modify marginal distributions.
+However, it is also unlikely to be sensitive as it holds only if there is no casual relationship between $X$ and $Y$.
+In case \ref{cat:not_identifiable}, the causal effect cannot be identified from this data set so there is no further work required to jam it.
+Case \ref{cat:identify_other} is complex and rarely used in empirical work.
+We leave it for future work.
 
@@
 In other words, if at least one of the variables in $X \cup Y \cup Z$ is in a different a different partition to the others.
+
+Note: this holds only if $Z$ is a \textit{minimal} adjustment.
+If $Z$ is not minimal then we can let $Z = Z' \cup U$ where $Z'$ is a minimal adjustment.
+In this case, we can still estimate the causal effect of $X$ on $Y$ by conditioning on $Z'$.
 
@@
 
+\subsection{Further work}
+
+We have considered only single variable interventions and outcomes.
+We believe the results generalize in a straightforwardly to the case where, for each causal query, $X$ and $Y$ are sets with a minor assumption that there are no causal paths from one intervention variable $X$ to another.
+
+Although overwhelmingly used in practice, adjustment is not necessary for causal identifiability.
+A given causal graph may also be identifiable via the front door criterion \cite{} or via an expression returned by the identify algorithm \cite{}.
+
+What is the psuedo code to jam all mechanisms?
+We need an algorithm to produce all 'minimal' expressions.
+Then each expression will contain a set of factors.
+At least one of those factors must be jammed for every expression.
+
+Note: A given query may be identifiable via multiple different expressions.
+As we must jam all of them, it is not sufficient to have an algorithm that returns \textit{an} expression (if one exists).
+We are not aware of any such algorithms for the general identifiability problem.
+
+Prove that a single query can still always be jammed by separating $X$ and $Y$.
+Approach  - the expression is a result of the do calculus and nessesarily contains a term with both $X$ and $Y$ together unless...
+
 \bibliographystyle{apalike}

### 2016-06-27 identify/partial/linear_guassian_soft_intervention.py
shows instrumental variable setting is fully identifiabile in linear-gaussian setting

@@ -1,0 +1,113 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Fri Jun 10 11:30:22 2016
+
+@author: finn
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from math import sqrt
+from scipy.stats import norm
+import seaborn as sns
+#np.set_printoptions(precision=1)
+
+
+
+    
+
+samples = 100000
+mz,mu = 1,3
+wx0,wy0 = -10,5
+wxz,wxu,wyu,wyx = 2.0,.1,0.5,4.0
+vz,vu,vex,vey = 0.1,3.0,0.2,0.5
+
+mx = wx0+wxz*mz+wxu*mu # theoretical mean of P(x)
+my = wy0+wyu*mu+wyx*mx # theoretical mean of P(y)
+
+z = np.random.normal(mz,sqrt(vz),size=samples)
+u = np.random.normal(mu,sqrt(vu),size=samples)
+ex = np.random.normal(0,sqrt(vex),size=samples)
+ey = np.random.normal(0,sqrt(vey),size=samples)
+x = wx0 + wxz*z+wxu*u+ex
+y = wy0+wyu*u+wyx*x+ey
+
+
+
+# estimate the covariance matrix for z,x,y
+cov = np.cov(np.vstack((z,x,y)))
+cor = np.corrcoef(np.vstack((z,x,y)))
+
+
+czz = vz 
+czx = wxz*czz
+czy = wyx*czx
+cxx = wxz**2*vz+wxu**2*vu+vex
+cxy = wxu*wyu*vu+wyx*cxx
+cyy = wyu**2*vu+2*wyx*wyu*wxu*vu+wyx**2*cxx+vey
+
+theoretical = [czz,czx,czy,cxx,cxy,cyy]
+simulated = [cov[0,0],cov[0,1],cov[0,2],cov[1,1],cov[1,2],cov[2,2]]
+
+for i in range(len(theoretical)):
+    print("{:.2f},{:.2f}".format(theoretical[i],simulated[i]))
+    
+# check the pearson correlation matrix
+sz = sqrt(vz)
+sx = sqrt(cxx)
+sy = sqrt(cyy)
+corzx = wxz*sz/sx
+corzy = wyx*wxz*sz/sy
+corxy = wyx*sx/sy + wxu*wyu*vu/(sx*sy)
+
+theoretical = [corzx,corzy,corxy]
[…]

### 2016-07-18 identify/adjust/adjust.tex
starts trying to define the adjustment problem

--- 2015-03-13 causal-bandit/c-bandit.tex
@@
 \renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+\newcommand{\N}[2]{\mathcal{N}\left({#1}\;,\;{#2}\right)}
 
@@
            draw=black, thick,
-           minimum width=1em,
-           minimum height=1em,
+           minimum width=2.5em,
+           minimum height=2.5em,
            font=\footnotesize,
            text centered,
+           scale=1,
            fill=blue!20!white},
@@
            draw=black, thick, dashed,
+           minimum width=2.5em,
+           minimum height=2.5em,
+           font=\footnotesize,
+           text centered,
+           fill=black!10!white
+           },
+     empty/.style={
+           circle,
+           rounded corners,
            minimum width=.5em,
@@
            text centered,
-           fill=black!10!white
            },
@@
 
-\title{Intervention Bandits}
-\author{Blah blah}
+\title{Partial Identifiability}
 
@@
 \def\ci{\perp\!\!\!\perp}
-\maketitle
 
-\begin{abstract}
-An abstract.
-\end{abstract}
-
-\section{Introduction}
-
-Useful references are: \cite{BC12}.
-
-\section{Notation}
-
-Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$, figure .
-We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.
-
-Let $p \in [0,1]^K$ be a fixed and known vector. In each time-step $t$:
-
-\begin{enumerate}
-\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
-\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
-\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
-\eq{
-\tilde X_{t,i} = \begin{cases}
-X_{t,i} &\text{if } i \neq I_t \\
-J_t & \text{otherwise}\,.
-\end{cases}
-}
-\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
-\end{enumerate}
-The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
-where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
-\eq{
-R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
-}
-
-\section{Estimating $\mu_{i,j}$}
-The most natural way to estimate $\mu_{i,j}$ is to compute an empirical estimate based on samples when that action was taken. This approach would
-lead directly to the UCB algorithm with $2K$ actions and a regret bound that depended linearly on $K$.
-
-In this instance we can significantly
-outperform the standard approach by exploiting the known causal structure of the problem.
-
-\eqn {
-P(Y|do(X_i=j)) &= P(Y|X_i = j)\\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b|X_i = j) \\
- &= \sum_{b}P(Y|X_i = j,X_a = b)P(X_a = b), \; \forall \, a \in \{1...K\}/i \text{ as } X_a \ci X_i\\
- \label{eqn:marginalizedEstimator}
- &= \sum_{b}P(Y|X_i = j,do(X_a = b))P(X_a = b) 
-}
-
-\subsection{Estimators}
-\label{sec:estimators}
-Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
-
-Considering equation \ref{eqn:marginalizedEstimator}, a natural way to construct an empirical estimator,  $\hat{\mu}_a$, for $P(Y|do(X_i=j))$ utilizing the information obtained by intervening on variable $a$ is:
-
-\eqn {
-\hat{\mu}_a &=
-\begin{cases}
-\label{eqn:marginalizedEstimator2}
-\frac{m_{a,1}}{n_{a,1}}p_a+\frac{m_{a,0}}{n_{a,0}}(1-p_a) & \text{if } a \neq{i} \\
-\frac{m_{i,j}}{n_{i,j}} & \text{if } a=i \\
-\end{cases}\\
-}
-where $\frac{m_{a,b}}{n_{a,b}}$ is an empirical estimator for $P(Y|X_i = j,do(X_a = b))$,
-\eqn{
-m_{a,b}   &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b, Y = 1}_s \\
-n_{a,b}    &= \sum_{s=1}^t \ind{X_i = j, I = a,J = b}_s 
-}
-
-This gives $K$ estimators $\{\hat{\mu_1}...\hat{\mu_K}\}$ to be pooled into a single estimator $\hat{\mu}$.
-
-\eqn {
-\label{eqn:pooledEstimators}
-\hat{\mu} &= \sum^{K}_{a=1}w_a\hat{\mu}_a=w_i \frac{m_{i,j}}{n_{i,j}}+ \sum_{a \neq i} w_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right]\\
-}
-
-If $p$ is not known $p_a$ could be replaced with its empirical estimate $\hat{p}_a$
-
-\eqn{
-\hat{p}_a = \frac{\sum_{s=1}^{t} \ind{X_a=1,I \neq a}_s}{\sum_{s=1}^{t} \ind{I \neq a}_s}
-}
-
-Now I want to select weights $w$ to minimize the variance in $\hat{\mu}$ and then get some kind of high probability bound for it.
-However, if I'm not very careful about how I pool things in equation \ref{eqn:pooledEstimators}, the estimator could get very biased, ie estimators $\hat{\mu_a}$ where I have data only for one state of the variable $a$ are not very informative, so should tend to get low weight, but if we choose to sample the second state based on previous payoffs - then we could end up assigning high weights to those estimators that had fluctuated above the true value...
-
-This complicates getting correct uncertainty bounds for a UCB type algorithm/proof.
-
-Consider, what exactly you mean by not unbiased ...
-
-\section{Importance Sampling Approach}
-
-Consider down weighting estimators proportional to the number of times we played them
-
-\section{Observe then exploit}
-
-\subsection{Basic idea}
-
-Basic idea - observe until the uncertainty on all actions is smaller than $\epsilon$, then exploit (or switch to standard UCB). This is only really going to work if $p_a \sim 0.5$.
-
-Bounds on regret: when observing the regret for each timestep is at most 1 (since the reward is $\in [0,1]$).
-When exploiting regret it at most $\epsilon$.
-If the horizon is $n$, the cost of exploiting $< \epsilon n$ and the cost of observing is $O(1/\epsilon^2)$ (this comes from how quickly we converge to within $\epsilon$ for all arms).
-
-\eqn{
-R_n = O(n\epsilon+\frac{1}{\epsilon^2})
-}
-
-Differentiating and selecting $\epsilon = (\frac{2}{n})^{1/3}$ to minimize the regret yields:
-
-\eqn{
-R_n = O(n^{2/3})
-}
-
-For each arm $X_i = j$, we have:
-
-\eqn{
-\hat{\mu}_{ij} = \frac{m_{ij}}{n_{ij}}
-}
-
-where
-
-\eqn{
-m_{i,j}    &= \sum_{s=1}^t \ind{X_i = j, Y = 1} \\
-n_{i,j}    &= \sum_{s=1}^t \ind{X_i = j} \\
-}
-
-and Hoeffding's Inequality gives:
-
-\eqn {
-P\left(|\hat{\mu}_{ij} - \mu_{ij}| > \sqrt{\frac{1}{2n_{ij}}\log{\frac{2}{\delta}}}\right) \leq \delta
-}
-
-\subsection{Alternate Observe and Exploit With Some Decay}
-Keep track of two separate bounds and always choose the narrower of the two.
-
-\subsection{Observe until number of plausibly optional arms $< \alpha$}
-
-Define the set of plausibly optimal arms to be those with an upper confidence bound that is higher than the largest lower confidence bound.
-
-\eqn {
-A_t = \{k : UB(k) \geq max_{k'}(LB(k'))  \}
-}
-
-We then observe until the size of this set is less than some value $\alpha$ before switching to UCB.
-This takes into account that we don't need narrow confidence bounds on the arms with very low expected reward.
-Once we start doing UCB, we keep track of the ucb bound separately and for each arm use whichever bounds are narrower.
-
-\subsubsection{Regret during observe phase}
-
-Let $n$ be the number of timesteps until the size of the set $A_t$ falls below $\alpha$ or we reach the overall horizon $T$
-
-\eqn {
-n = \min(\{t:size(A_t)< \alpha, T\})
-}
-
-Let $Q(k)$ be the expected reward for arm $k$ and $\mu^*$ be the expected reward for the best arm.
-
-\eqn {
-R_n = & \max_{i=1...K} E\left[\sum_{t=1}^n Y_{it} - \sum_{t=1}^n Y_{I_t,t} \right] \\
-= & E[n]\left(\mu^*  - E[Q]\right) 
-}
-
-Without loss of generality, assume the arms are ordered from best to worst.
-
-\eqn{
-& Q(k = 1...K) = [\mu^*, \mu^*-d_2 ... \mu^*-d_k] \\
-&\implies E[Q] = \mu^* - \frac{1}{K}\sum_{k=2}^K d_k \\
-&\implies  R_n = E[n]\left(\frac{1}{K}\sum_{k=2}^K d_k \right)
-}
-
-Whenever the following two statements hold only arms with indexes less than $\alpha$ can be in the set $A_t$:
-
-\eqn {
-& \mu^* - \hat{\mu}^* \leq \frac{d_\alpha}{2} \\
-& \hat{\mu_k} - \mu_k \leq \frac{d_k}{2} \; \forall k \geq \alpha 
-}
-
-\eqn{
-\label{eq:proboverlap}
-\implies \rho(t) \equiv P(\exists k \in A_t: k \geq \alpha)  \leq P(\mu^* - \hat{\mu}^* > \frac{d_\alpha}{2}) + \sum_{k=\alpha}^K P( \hat{\mu_k} - \mu_k > \frac{d_k}{2})
-}
-
-\eqn {
-P(n > t) = \begin{cases} 0 & \text{if } t \geq T \\
-P(\exists k \in A_t: k \geq \alpha) & \text{otherwise}
-\end{cases}
-}
-
-We now concentrate bounding \ref{eq:proboverlap}. For random variables $W_{sij} = P(Y|X_i=j)$ drawn from a Bernoulli distribution with mean $p_{ij}$ the Chernoff inequality gives:
-
-\eqn {
-& P(\frac{1}{t}\sum_s W_{sij} - p_{ij} \geq \gamma) \leq \exp(-\frac{t\gamma^2}{2p_{ij}(1-p_{ij})+2\gamma/3}) \\
-\implies & P(\frac{1}{t}\sum_s \frac{W_{sij}}{p} - \frac{p_{ij}}{p} \geq \epsilon) = P(\hat{\mu}_k - \mu_k \geq \epsilon) \leq \exp(-\frac{tp\epsilon^2}{2+2\epsilon/3}) = \exp(-\frac{t\epsilon^2}{4+4\epsilon/3})
-}
-Where $p_{ij} = P(Y = 1, X_i = j) \leq P(X_i=j) = p = 1/2 \;\;\; \forall(i,j)$
-
-Similarly, 
-\eqn {
-P(\mu^* - \hat{\mu}^* \geq \epsilon) \leq exp(-tp^2\epsilon^2) = exp(-\frac{t\epsilon^2}{4})
-}
-
-\eqn {
-\implies \rho(t) \leq \exp(-\frac{ d_\alpha^2 t}{16}) + \sum_{k=\alpha}^K exp(-\frac{d_k^2 t}{16(1+d_k/6)})
-}
-
-\eqn{ 
-E[n] = & \sum_{t=1}^T P(n \geq t) \leq \sum_{t=1}^T \rho(t) \\
-= & \min \left(T, \frac{1 - e^{-d_\alpha^2 T/16}}{e^{d_\alpha^2 /16} -1 } +  \sum_{k=\alpha}^K \frac{1 - e^{-d_k^2 T/16(1+d_k/6)}}{e^{d_k^2 /16(1+d_k/6)} -1 } \right)
-}
-
-Since the regret is $R_n = E[n]\left(\frac{1}{K}\sum_{k=2}^K d_k \right)$ the worst case is that just enough of the differences $d_k$ are $0$ to make $E[n] = T$ and the remainder are $1$.
-This occurs if $d_\alpha = 0$, In this case the regret is:
-
-\eqn {
-R_n = T \left(\frac{K-(\alpha+1)}{K}\right)
-}
-
-Clearly, this is not quite the right criteria to decide when to switch ...
-\subsubsection{Regret on observing for h rounds before switching to UCB}
-
-How large the regret is if we initialize with a certain number of counts already in place...
-
-Suppose we have $O_i$ observations for arm $i$.
-We will now begin doing standard UCB - storing a separate set of bounds, only each time we choose an action we will pick the narrower of the observational/interventional confidence bounds.
-
-\eqn{
-\hat{\epsilon}_{it} &= \min \left( \sqrt{\frac{\alpha \log(t)}{2 T_{it}}},\sqrt{\frac{log(1/\delta)}{2 O_i}} \right)\\
-\label{eqn:combinedBoundsEpsilon}
-& = \sqrt{\frac{\alpha \log(t)}{2 \max(T_{it},O_i)}} \;\; \text{if we let $\delta = t^{-\alpha}$}
-}
-
-When running UCB, in order to select a non-optimal arm $i$ at timestep $t$, at least one of the following statements must be true:
-
-\begin{enumerate}
-\item $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu_i + \Delta_i$
-\item $\hat{\mu}_{it} + \hat{\epsilon}_{it} \geq \mu_i + 2\hat{\epsilon}_{it}$
-\item $\Delta_i < 2 \hat{\epsilon}_{it}$
-\end{enumerate}
-
-Given our expression for the confidence bound in equation \ref{eqn:combinedBoundsEpsilon} we can write statement (3) as:
-\eqn{
-\max(T_{it},O_i) < \frac{2\alpha \log(t)}{\Delta_i^2}
-}
-
-Let:
-
-\eqn{
-\beta = \frac{2\alpha \log(n)}{\Delta_i^2} 
-}
-
-If $O_i \geq \beta$, statement (3) is guaranteed to be false for all $t$ and we can bound the expected number of times we select $i$ with how many times we expect the bounds given by (1) and (2) to be true.
-
-For arms $i$ such that $O_i \geq \beta$
-
-\eqn{
-&P(\text{1 is true}) < t^{1-\alpha}\\
-&P(\text{2 is true}) < t^{1-\alpha}\\
-& \implies P(I_t = i) < 2t^{1-\alpha}\\
-& \implies R_n(i) \leq \frac{\Delta_i \alpha}{\alpha -2}
-}
-
-For all other arms with $\Delta_i > 0$, the bound is the same as standard UCB
-\eqn{
-R_n(i) \leq  \frac{2\alpha \log(n)}{\Delta_i} + \frac{\Delta_i \alpha}{\alpha -2} 
-}
-
-Assume we observed for $h$ steps. The expected regret from this period is just $h E[\Delta_i]$. Let $T = h+n$ be the total time.
-
-\eqn{
-R_T < \sum_{i=1}^K \Delta_i \left(\frac{h}{K} + \frac{\alpha}{\alpha-2}\right) + \sum_{i:\Delta_i > 0}P(O_i \leq \beta)\frac{2\alpha \log(n)}{\Delta_i}
-}
-
-Using Hoeffding's Inequality:
-
-\eqn {
-P(O_i \leq \beta) \leq \begin{cases}
-2 e^{-2h(\frac{1}{2}-\frac{\beta}{h})^2} & \text{if $h > 2\beta$} \\
-1 & \text{otherwise}
-\end{cases}
-}
-
-Its tricky - selecting the $h$ to switch at based on the inequality can certainly yield a non-optimal answer.
-Maybe I should figure this out via simulation??
-But some theory to tell me what to simulate would certainly help.
-
-\pagebreak
-
-\subsubsection{General rough thoughts}
-
-When observing the confidence interval on all the arms drops approximately with $\sqrt{\frac{1}{n}}$ (provided that the probability of each occurring is the same) - otherwise for each arm it will be weighted by this drop will be weighted by a function of the probability that it occurs.
-When acting, the interval on a single arm we intervene on drops.
-What is the informational value of reducing the bounds on various arms (it is higher for arms that are more likely to be good).
-
-\subsection{Bounding a weighted sum of unbiased estimators with McDiarmid's Ineqaulity}
-
-McDiarmid's Inequality states:
-If $X_i \ci X_j$ and 
-\eqn{
-|\phi(X_1...X_i...X_N) - \phi(X_1...X_i^{'}...X_N)| < c_i \;\; \forall i
-}
-\eqn{
-& P\left(|\phi(\boldsymbol{X}) - E[\phi(\boldsymbol{X})]| \geq \epsilon \right) \leq 2 \exp \left(-\frac{2 \epsilon^2}{\sum_{i}c_i^2} \right)\\
-\label{eqn:McDiarmid}
-& P\left(|\phi(\boldsymbol{X}) - E[\phi(\boldsymbol{X})]| \geq 
-\sqrt{\frac{\sum_{i}c_i^2}{2}\log{\frac{2}{\delta}}} 
-\right) \leq \delta
-}
-
-In our problem, for each variable $V_i$ and value $j$ we have:
-
-\eqn{
-\hat{\mu}_a = \tilde{P}(Y|V_i = j)= \begin{cases}
-\frac{p_a}{n_{a1}}\sum_{s=1}^{n_{a1}}X_{a1,s} + \frac{1-p_a}{n_{a0}}\sum_{s=1}^{n_{a0}}X_{a0,s} & \text{ if $a \neq i$}\\
-\sum_{s=1}^{n_i}X_s & \text{if $a = i$}
-\end{cases}
-}
-
-where $X_{a1,s}  \sim P(Y|V_i = j, do(V_a = 1))$ and $X_{a0,s}  \sim P(Y|V_i = j, do(V_a = 0))$, both $\in [0,1]$. We then pool $\{\hat{\mu}_1 ... \hat{\mu}_K \}$ to get a single estimate.
-
-\eqn {
-\hat{\mu} = \tilde{P}(Y|V_i = j)= \sum_a w_a \hat{\mu}_a \text{, where } \sum_a w_a = 1
-}
-
-Let
-
-\eqn {
-\hat{\mu} = \phi(\boldsymbol{X}) = \sum_{a \neq i} w_a \left[ 
-\frac{p_a}{n_{a1}}\sum_{s=1}^{n_{a1}}X_{a1,s} + \frac{1-p_a}{n_{a0}}\sum_{s=1}^{n_{a0}}X_{a0,s}
-\right] + \frac{w_i}{n_i} \sum_{s=1}^{n_i}X_{i,s}
-}
-
-Note: this treats the number of samples $n_i$,$n_{a1}$, and $n_{a0}$ as fixed.
-Even for fixed actions, the latter two are still random variables that depend on $\boldsymbol{p}$  Will this bound still hold?
-
-\eqn {
-&\phi(...X_{\alpha 1,s}...) - \phi(...X_{\alpha 1,s}^{'}...) \leq w_\alpha \frac{p_\alpha}{n_{\alpha 1}} \\
-&\phi(...X_{\alpha 0,s}...) - \phi(...X_{\alpha 0,s}^{'}...) \leq w_\alpha \frac{1-p_\alpha}{n_{\alpha 0}}\\
-& \phi(...X_i...) - \phi(...X_i^{'}...) \leq \frac{w_i}{n_i}
-}
-
-\eqn{
-\sum_i c_i^2 = & n_i \left(\frac{w_i}{n_i}\right)^2+ \sum_a \left[ n_{a1}\left(w_a \frac{p_a}{n_{a1}}\right)^2 + n_{a0}\left(w_a \frac{1-p_a}{n_{a0}}\right)^2\right]\\
-= & \frac{w_i^2}{n_i}+ \sum_a w_a^2 \left(\frac{p_a^2}{n_{a1}}+  \frac{(1-p_a)^2}{n_{a0}} \right) \\
-\label{eq:tominimize}
-= & \sum_a w_a^2 f(a) \text{, where }\\ 
-f(a) =& \begin{cases}
-\frac{p_a^2}{n_{a1}}+  \frac{(1-p_a)^2}{n_{a0}}  & a \neq i\\
-\frac{1}{n_i} & a = i
-\end{cases}
-}
-
-We want to choose weights $\boldsymbol{w}$ so as to minimize \ref{eq:tominimize} subject to the constraint $\sum_a w_a = 1$.
-
-The minimum (assuming it exists) should occur at a critical point of:
-
-\eqn {
-L(w_1...w_k,\lambda) = \sum_a w_a^2 f(a) + \lambda((\sum_a w_a)-1)
-}
-
-\eqn{
-& \frac{\partial L}{\partial w_a}  = 2 w_a f(a) + \lambda = 0 \\
-& \implies w_a =  \frac{-\lambda}{2 f(a)}\\
-& \frac{\partial L}{\partial \lambda}  = (\sum_a w_a)-1= 0 \\
-& \implies -\frac{\lambda}{2}\sum_a \frac{1}{f(a)} = 1
- \implies \lambda = -\frac{2}{\sum_a \frac{1}{f(a)}}\\
-& \implies w_a = \frac{1}{f(a)\sum_a \frac{1}{f(a)}} \\
-\label{eqn:sumofsquares}
-& \implies \sum_i c_i^2 = \frac{1}{\sum_a \frac{1}{f(a)}} 
-}
-
-Substituting \ref{eqn:sumofsquares} into the McDiarmid inequality \ref{eqn:McDiarmid}:
-
-\eqn{
-P\left(|\hat{\mu}^{ij} - \mu^{ij}| > \sqrt{\frac{1}{2\sum_a \eta_a^{ij}} \log{\frac{2}{\delta}}}\right) \leq \delta
-}
-
-where:
-
-\eqn {
-& \mu^{ij} = P(Y | V_i = j) \\
-& \eta_a^{ij} = \begin{cases}
-\frac{n_{a1}^{ij} n_{a0}^{ij}}{n_{a1}^{ij}(1-p_a)^2+n_{a0}^{ij}p_a^2} & a \neq i\\
-n_{ij} & a = i
-\end{cases} \\
-&n_{al}^{ij} = \sum_{s=1}^t \ind{do(V_a = l), V_i = j} \\
-& n_{ij} = \sum_{s=1}^t \ind{(do(V_i = j)}
-}
-
-All of the above is to get estimates for $P(Y|V_i = j)$ for some fixed $i,j$.
-Suppose we explore for some fixed total number of rounds $h$, the goal is to minimize the worst confidence bound - that doesn't quite make sense as a goal - we just need to be sure that all actions are worse than our estimate of the best with probability $\delta$.
-...
-anyway
-
-We have control of are the number of times we do each action, $n_{al}$, which obviously also influences the random variables $n_{al}^{ij}$, subject to the constraint $\sum n_{al} = h$.
-
-Objective: minimize the expectation of the largest confidence bound after a fixed number of rounds $n$.
-
-Let $\boldsymbol{n} = [n_{11},n_{10},n_{21},n_{20}...n_{k1},n_{k0}]$
-
-\eqn{
-\tilde{\boldsymbol{n}} = \argmin_{\{\boldsymbol{n}:\|\boldsymbol{n}\|=h\}} \max_{i,j} \left( E \left[ \sqrt{\frac{1}{2\sum_a \eta_a^{ij}} \log{\frac{2}{\delta}}} \right] \right)
-}
-
-We can bound the first part of $\eta_a^{ij}$
-
-\eqn {
-\eta_a^{ij} = \frac{n_{a1}^{ij}n_{a0}^{ij}}{n_{a1}^{ij}(1-p_a)^2+n_{a0}^{ij}p_a^2} & \leq \frac{n_{a1}^{ij}n_{a0}^{ij}}{\max(n_{a1}^{ij}(1-p_a)^2+n_{a0}^{ij}p_a^2)} \\
-& = \min \left( \frac{n_{a0}^{ij}}{(1-p_a)^2}, \frac{n_{a1}^{ij}}{p_a^2} \right)
-}
-
-\eqn {
-E \left[ \min \left( \frac{n_{a0}^{ij}}{(1-p_a)^2}, \frac{n_{a1}^{ij}}{p_a^2} \right)  \right] & \leq 
-\min \left( E\left[\frac{n_{a0}^{ij}}{(1-p_a)^2} \right], E\left[\frac{n_{a1}^{ij}}{p_a^2} \right] \right) \\
-& = P(V_i = j) *\min \left( \frac{n_{a0}}{(1-p_a)^2} , \frac{n_{a1}}{p_a^2}  \right) \\
-}
+\subsection*{Adjustment}
 
@@
 	\centering    
-\caption{The effective number of samples $\eta_a$ versus $n_{a1}$, where $p_a = .8$ and the total number of samples,  $ n_{a1}+n_{a0}=100$.
-The effective number of samples is maximized (and equals the total) if we sample each side according to its probability.}
-\includegraphics[width=.5\textwidth]{samplehalves}
+          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+ %nodes
+\node[main node](2){$X$};
+\node[main node, above right=of 2](3){$Z$};
+\node[main node, below right=of 3](4){$Y$};
+\node[latent, above=of 2](5){$\varepsilon_x$};
+\node[latent, above=of 4](6){$\varepsilon_y$};
+ \path[every node/.style={font=\tiny}]
+    (3) edge (2) edge (4)
+    (2) edge (4)
+    (5) edge (2)
+    (6) edge (4);
+\end{tikzpicture}
 \end{figure}
 
-\subsection{Targeted Sampling}
-Sample a equal number of points from each arm - until the error on all arms is smaller than some threshold.
-Then exploit.
-This should do better than observe-exploit, particularly where the probability of some events occurring without intervention is low.
-
-A more sophisticated version:
-
-\subsection{UCB variant}
-Combine estimators according to:
+To estimate the causal effect of $X$ on $Y$, we need to condition on $Z$ so as to block the backdoor path $X \leftarrow Z \rightarrow Y$.
 
 \eq {
-w_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \text{ and } n_{i,j}=\begin{cases} 
-n_{i,j} & \text{if } a = i \\
-\frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
-\end{cases}
+P(Y|do(X=x)) = \sum_{z \in \mathcal{Z}}P(Y|X=x,Z=z)P(Z)
 }
 
-In this case the estimators from section \ref{sec:estimators} are biased - so its a lot harder to prove a regret bound.
-
-\begin{theorem} (Probably False) With probability at least $1 - \delta$ we have that:
-$\displaystyle \left|\hat{\mu}_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
-where $\beta > 0$ is some constant.
-\end{theorem}
-
-\begin{proof}
-First note that $n_{a,b}$ is a random variable that is bounded by $t$ for all $a,b$.
-We use the short-hand $\mu^{a,b}_{i,j} = \E[q(X) | X_i = j, X_a = b]$. Then
-\eq{
-\mu_{i,j} = p_a \mu_{i,j}^{a,1} + (1 - p_a) \mu_{i,j}^{a,0}\,.
-}
-Now we can apply Hoeffding's bound and the union bound to show that
-\eq{
-\P{\left|\frac{m_{a,b}}{n_{a,b}} - \mu_{i,j}^{a,b}\right| \geq \sqrt{\frac{1}{2n_{a,b}} \log \frac{4t}{\delta}}} \leq \frac{\delta}{2}\,.
-}
-Therefore by the union bound
-\eq{
-\P{\left|p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}} - \mu_{i,j}\right| \geq p_a \sqrt{\frac{1}{2n_{a,1}} \log\frac{4t}{\delta}}
-+ (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}} \leq \delta\,
-}
-Now by Jensen's inequality 
-\eq{
-p_a \sqrt{\frac{1}{2n_{a,1}}\log\frac{4t}{\delta}} + (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}
-&\leq \sqrt{\left(\frac{p_a}{2n_{a,1}} + \frac{1-p_a}{2n_{a,0}}\right)\log \frac{4t}{\delta}} \\
-&\leq \sqrt{\max\set{\frac{p_a}{n_{a,1}}, \frac{1-p_a}{n_{a,0}}} \log\frac{4t}{\delta}} \\
-&= \sqrt{{1 \over 2n_a} \log\frac{4t}{\delta}}\,.
-}
-Similarly,
-\eq{
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{4t}{\delta}}} \leq
-\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{2t}{\delta}}} \leq \delta\,.
-}
-\end{proof}
-
-\section{Algorithm}
-
-\begin{algorithm}
-\caption{UCB}
-\begin{algorithmic}[1]
-\State {\bf Input:} Number of variables $K$, vector $p \in [0,1]^K$, horizon $n$
-\For{$t \in 1,\ldots,n$}
-\For{$i \in 1,\ldots,K$}
-\For{$j \in \set{0,1}$}
-\State Compute $\tilde \mu_{i,j} = \hat \mu_{i,j} + \sqrt{\frac{\alpha}{\sum_{a} n_a} \log n}$ 
-\EndFor
-\EndFor
-\State Choose $I_t, J_t = \argmax_{i,j} \tilde \mu_{i,j}$
-\EndFor
-\end{algorithmic}
-\end{algorithm}
-
-\section{Theorems}
-\section{Experiments}
-
-\begin{figure}[h]
-\caption{Comparison of the performance of standard UCB versus causal UCB with $\beta=3$ and $\beta = 5$.
-100 simulations were run for each algorithm up to a horizon of $10^5$ per value of $K$.
-Error bars span the 1st to 3rd quantile of the regret, round points mark the median and triangular points show the mean.
-For standard UCB the regret increases linearly with the number of arms $K$.
-For causal UCB the increase is sub-linear.
-Increasing $\beta$ leads to slower convergence but lower variance.}
-\includegraphics[width=1\textwidth]{regret_vs_K_series_beta.pdf}
-\end{figure}
-
-\begin{figure}[H]
-\caption{The distribution of regret varies with the $\beta$ parameter in the bound in the estimator.
-As beta increases, the mean regret increases but the variance decreases.
-The plot shows the results of running $100$ independent bandits, with $K=256$ and $\epsilon=0.1$, up to a horizon $h=10^5$ for each value of $\beta$.
-}
-\includegraphics[width=0.7\textwidth]{regret_vs_beta_boxplot.pdf}
-\end{figure}
-
-Simululations to compare the performance of standard UCB with our modified algorithm.
-For each number of arms, $100$ bandits of each type were created and run upto to a horizon of $1000$ timesteps.
-The mean regret and its standard error from these simulations is plotted in figure \ref{fig:regretvsarms} The true data was generated from a model where:
+With a binary treatment $X$ we can also write:
 
 \eq{
-&p = [0.5]^{K}\\
-&q(\boldsymbol{X}) = \begin{cases}0.5 & \text{if $X_{1}=0$}\\ 0.6 & \text{otherwise} \end{cases}
+ATE = E(Y|do(X=1)) - E(Y|do(X=0))
 }
 
-\section{Conclusion}
+This assmumes $Z$ is fully observable, ie there are no unobservable variables, $U$ that cause both $X$ and $Y$.
 
-\newpage
-\section{Notes}
+\subsection*{Covariate Shift}
 
-\subsection{Why Hoeffdings bound doesn't hold if $n$ is a random variable}
+Seperate from the problem of adjustment, lets define covariate shift.
 
-Let $\{Z_1,Z_2...Z_n\} \sim Bernoulli(\frac{1}{2}) $ and $X_i = 2Z_i-1 \implies X_i \in \{-1,1\} \text{ and } E[X_i] = 0$.
+\begin{itemize}
+\item We have training data $d_{train} = \set{(x_1,y_1),...,(x_n,y_n)}$ sampled from $P(X,Y) = P(X)P(Y|X)$.
+\item The test data $d_{test} = \set{(x'_1,y'_1),...,(x'_{n'},y'_{n'})}$ is assumed to be sampled from $Q(X)P(Y|X)$, where $Q(X) \neq P(X)$.
+As usual for test data, we observe only the inputs $\set{x'_1,...,x'_{n'}}$
+\item We wish to use the training data to learn a hypothosis/function $h:\mathcal{X} \rightarrow \mathcal{Y}$ that minimises $\sum_{i = 1}^{n'}\mathcal{L}(h(x'_i,y'_i))$, where $\mathcal{L}$ is some loss.
+\end{itemize}
 
-For a fixed $n$, Hoeffdings inequality says: 
-\eqn{
-\label{eq:HoeffdingsRandom}
-& P\left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \epsilon\right) \leq 2e^{-n\epsilon^2} \\
-\implies & P \left(|\frac{1}{n}\sum_{i=1}^{n}X_i| > \sqrt{\frac{\log{4}}{n}}\right) \leq \frac{1}{2} \\
-\implies & P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \leq \frac{1}{2}
-}
+The covariate shift assumption is that $P(Y|X)$ has not changed.
 
-If $n$ is allowed to be dependent on the sequence of values sampled, this inequality no longer holds.
+%This assumption is plausible if $X$ causes $Y$ and we imagine a mechanism that maps $\mathcal{X}$ to $\mathcal{Y}$ that is independent of the input distribution.
+If $Y$ causes $X$ then we would expect that $P(X)$ and $P(Y|X)$ would in general be related, such that a shift in $P(X)$ would lead us to ancipate that $P(Y|X)$ might also have changed.
 
-\begin{proof}
-Choose $n$ based on the sequence of samples seen so far such that:
-\eq {
-n = min \{ n: n > 4 \text{ and } \sum_{i=1}^{n}X_i > \sqrt{n \log \log n}\}
-}
-By the law of iterated logarithms this quantity is finite with probability $\sim 1$.
+Although we have assumed the underlying mapping, $P(Y|X)$, we are trying to learn has not changed between the test and the train data we still need to adapt standard prediction algorithms as otherwise (through a preference for 'simple' functions) we may select a function that fits well where $P(X)$ is high but performs badly where $Q(X)$ is high.
+There are some situations that require no adaptation.
+For example, if the true function is linear and we fit an (unregularized) linear model, then model that is in expectation optimal on the training data is also optimal for the test data.
+Another interesting example is if the model is a Gaussian process \cite{}.
 
-\eq {
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{\log{n}}}\right) \sim 1 \\
-& \implies  P\left( |\sum_{i=1}^{n}X_i| > \sqrt{n \log{4}}\right) \sim 1 \text{, Since } \log{\log{n}} > \log{4} \;\; \forall n > 4 \\
-}
+\subsection*{Covariate shift in the adjustment problem}
 
-Thus Hoeffdings inequality (equation \ref{eq:HoeffdingsRandom}) does not hold in general if $n$ is not independent of the samples $\{X_i\}$ (The bound does not work if you decide to stop sampling as soon as you reach a point where random walk fluctuations take you outside it)
-
-\end{proof}
-
-\bibliographystyle{plainnat}
-\bibliography{c-bandit}
+In the adjustment problem, we have observational training data, $d_{train} = \set{(x_1,z_1,y_1),...(x_n,z_n,y_n)} \sim P(Z,X)P(Y|Z,X)$.
+If we intervene and set $X=x$ then the data will be sampled from $P(Z)\delta(X-x)P(Y|Z,X)$.
+Note that the covariate shift assumption is satisfied in this scenario, $P(Y|Z,X)$ does not change between the train and test settings.
+What has changed is the joint distribtion $P(X,Z)$.
 
 \end{document}
-
-So the regret for the observe stage is bounded by:
-
-\eqn {
-R \leq \frac{2a}{\exp(a^2/(64+4a/3)) - 1}
-}
-
-But this is just the expected number of steps until one arm does not overlap. With more arms its more complicated - as we need the probability that $\alpha$ of them overlap (another union bound?).
-
-\eqn {
-E[n] = \sum_{t=1}^T P(n \geq t) \leq & \sum_{t=1}^\infty \left( \exp(-\frac{t\epsilon^2}{4+4\epsilon/3}) +exp(-\frac{t\epsilon^2}{4}) \right) \\
-= & \frac{1}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} + \frac{1}{\exp(\epsilon^2/4) - 1} \\
-\leq & \frac{2}{\exp(\epsilon^2/(4+4\epsilon/3)) - 1} \\
-= & \frac{2}{\exp(a^2/(64+4a/3)) - 1}
-}
-
-\subsubsection{A very simple model}
-
-So the regret for the observe stage is bounded by:
-
-\eqn {
-R \leq \frac{2a}{\exp(a^2/(64+4a/3)) - 1}
-}
-
-But this is just the expected number of steps until one arm does not overlap. With more arms its more complicated - as we need the probability that $\alpha$ of them overlap (another union bound?).
-Lets try to analyse a very simple model in which the reward depends only on the value of $X_1$.
-
-\eqn{
-&P(Y|X_1=0,X_2,...X_N) = \frac{1}{2}-a\\
-&P(Y|X_1=1,X_2,...X_N) = \frac{1}{2}+a\\
-&P(X_i = j) = \frac{1}{2} \;\;\; \forall (i,j)\\
-&\implies P(Y) = \frac{1}{2} \\
-&\implies R_n = aE[n]
-}
-
-The goal is now to find $E[n]$ (which will be a function of $a$)
-
-Lets start by considering how long on average we need to observe until the lower bound on the optimal arm is higher than the upper bound on a single other arm, $i$, (with expected reward $\frac{1}{2}$)
-
-If the following three statements hold, then there is no overlap $LB^* = \hat{\mu}^* - \epsilon > UB_i = \hat{\mu}_i + \epsilon$.
-
-\eqn {
-\label{eq:overlap1}
-&\mu^* - \hat{\mu}^* \leq \epsilon \\
-\label{eq:overlap2}
-&\hat{\mu}_i - \mu_i \leq \epsilon \\
-\label{eq:overlap3}
-&\epsilon \leq a/4
-}
-
-If \ref{eq:overlap3} is true, we can bound the probability of overlap  been the confidence intervals of the best arm and one of the other arms, $k$, at some timestep $t$ with the union bound of the probability that either \ref{eq:overlap1} or \ref{eq:overlap2} are false FOR ALL ARMS.
-
-\eqn {
-\rho(t) \equiv P(overlap_k) \leq 
-P(\mu^* - \hat{\mu}^* \geq \epsilon) + P(\hat{\mu}_k - \mu_k \geq \epsilon)  \;\;\;\text{ if } \epsilon \leq a/4
-}
-
-Now the estimators and their bounds are independent, as they are composed from sums of independent variables $X_i$ so the probability that $\alpha$ of them have overlapping bounds is given by a binomial distribution.
-
-\eqn {
-P(O > \alpha)  = & P(n \geq t) = 1 - \sum_{i=1}^\alpha \binom{K-1}{i}\rho(t)^{i}(1-\rho(t))^{K-1-i} \\
-\leq & \exp(-(K-1)D(\frac{\alpha}{K-1} || \rho(t))
-}
-
-\eqn {
-P(|\hat{\mu}_{a0} -\mu_{a0}| > \epsilon_{a0}) \leq \delta/2 \;\; \text{ where } 
-\epsilon_{a0}=\sqrt{\frac{1}{2n_{a0}}\log{\frac{4}{\delta}}}\\ 
-P(|\hat{\mu}_{a1} -\mu_{a1}| > \epsilon_{a1}) \leq \delta/2 \;\; \text{ where } 
-\epsilon_{a1}=\sqrt{\frac{1}{2n_{a1}}\log{\frac{4}{\delta}}}\\ 
-}
-
-We should not need $t$ in here since, although $n_{aj}$ is still random, its not dependent on previous results.
-
-Combining these two halves and applying the union bound gives:
-
-\eqn{
-P\left(|\hat{\mu}_a - \mu | > \sqrt{\frac{1}{2} \log{\frac{4}{\delta}}}
-\left(p_a \sqrt{\frac{1}{2n_{a1}}}+(1-p_a) \sqrt{\frac{1}{2n_{a0}}} \right)\right) \leq \delta
-}
-
-I now need to pool these estimators for each $a \in \{1...K\}/I$ but I can't just use a union bound - I need the bound to shrink by combining the multiple estimates.
-
-For some distributions,

### 2016-07-18 identify/adjust/representations.py

@@ -14,6 +14,7 @@ import seaborn as sns
 from sklearn.metrics import mean_squared_error
 from sklearn.cross_validation import train_test_split
 from scipy.stats import beta
+import pandas as pd
 
 class DataModel:
     def __init__(self,mean_x,cov_x,w,b,epsilon_y):
@@ -126,7 +127,7 @@ class Settings:
         self.samples = samples
         self.simulations = simulations
         self.true_ate = true_ate
-        
+        self.params = [num_features,aw,bw,ab,bb,cm,epsilon,samples,simulations,true_ate]
         self.w = beta.rvs(aw, bw, size=num_features)
         self.b = np.append( beta.rvs(ab,bb,size=num_features),true_ate)
         self.x_mean = np.zeros(num_features)
@@ -138,7 +139,7 @@ class Settings:
         model = self.model
         r = np.zeros((self.simulations,3))
         for s in range(self.simulations):
-            model.sample(1000)
+            model.sample(self.samples)
             ate_unadjusted = model.y[(model.t==1)].mean() - model.y[(model.t==0)].mean()
             ate_fit,scores,selected= lasso_then_ridge(model)
             lsq_fit = least_squares(model)
@@ -148,27 +149,34 @@ class Settings:
         delta = (rm[1]-rm[0])/(re[1]+re[0])
         return delta,rm,re
     
-    def print_it(self):
-        print(self.num_features,"(",self.aw,self.bw,")","(",self.ab,self.bb,")",self.cm,self.epsilon,self.samples,self.simulations,self.true_ate)
-        
+    def __str__(self):
+        return "{0}, ({1},{2}), ({3},{4}), {5}, {6}, {7}, {8}, {9}".format(*self.params)
+      
+import time
+s = time.time()
 
 results = []
 settings = []
-for experiment in range(500):
+n_experiments = 10000
+data = np.zeros((n_experiments,1+3*2+10))
+for experiment in range(n_experiments):
     num_features = np.random.choice(range(5,20))
     aw = np.random.choice(range(1,10))
     bw = np.random.choice(range(1,10))
     ab = np.random.choice(range(1,10))
     bb = np.random.choice(range(1,10))
     cm = np.random.choice(range(1,5))
-    epsilon = np.random.choice(range(1,4))
-    context = Settings(num_features,aw,bw,ab,bb,cm,epsilon)
-    context.print_it()
[…]

### 2016-07-26 identify/adjust/adjust.tex
makes data generating processes explicit
more on definitions of causal effect

@@
 \usepackage{algpseudocode}
+\usepackage{subcaption}
 \usepackage{dsfont}
-\usepackage[bf]{caption}
+
 \usepackage{hyperref}
@@
 \newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}[1]{\mathbb E\left[#1\right]}
+\newcommand{\E}[2]{\mathbb E_{#1}\left[#2\right]}
 \newcommand{\Var}{\operatorname{Var}}
@@
 \begin{figure}[H]
+    \begin{subfigure}[b]{0.45\textwidth}
 	\centering    
@@
 \end{tikzpicture}
+        \caption{prior to intervention}
+        \label{fig:before}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+ %nodes
+\node[main node](2){$x$};
+\node[main node, above right=of 2](3){$Z$};
+\node[main node, below right=of 3](4){$Y$};
+\node[latent, above=of 4](6){$\varepsilon_y$};
+ \path[every node/.style={font=\tiny}]
+    (3) edge (4)
+    (2) edge (4)
+    (6) edge (4);
+\end{tikzpicture}
+        \caption{after $do(X=x)$}
+        \label{fig:after}
+    \end{subfigure}
+
+    \caption{Data generating process}\label{fig:causalmodels}
+    \label{fig:adjustment}
 \end{figure}
+
+The idea of causal inference is to use data generated by one process (prior to an intervention) to infer characteristics of another process (the process after intervention).
+Causal graphical model how an intervention changes the process.
+For each node with incoming links there is a (deterministic) function that determines its value.
+We also implicitly assume that these functions and the distributions of all variables without incoming links does not change as a consequence of the intervention.
+The specific case of causal estimation by adjusting for known confounders is shown in figure \ref{fig:adjustment}.
 
@@
 \eq {
-P(Y|do(X=x)) = \sum_{z \in \mathcal{Z}}P(Y|X=x,Z=z)P(Z)
+P(Y|do(X=x))= P(Y) \text{, in graph \ref{fig:after}} = \sum_{z \in \mathcal{Z}}P(Y|X=x,Z=z)P(Z) \text{, in graph \ref{fig:before}}
 }
@@
 \eq{
-ATE = E(Y|do(X=1)) - E(Y|do(X=0))
+ITE &= Y|do(X=1),Z \leftarrow \text{ individual treatment effect, defined only if there's no unobserved noise $\epsilon_y$}\\
+PTE(Z) &= \E{\epsilon_y}{Y|do(X=1),Z}-\E{\epsilon_y}{Y|do(X=0),Z} \leftarrow \text{ personalized treatment effect}\\
+ATE &= \E{Z,\epsilon_y}{Y|do(X=1)} - \E{Z,\epsilon_y}{Y|do(X=0)} = \E{Z}{ITE(Z)} \leftarrow \text{ average treatment effect}
 }
@@
 This assmumes $Z$ is fully observable, ie there are no unobservable variables, $U$ that cause both $X$ and $Y$.
+
+\subsubsection*{Counterfactuals}
+Let $Y = f_y(X,Z,\epsilon_y)$.
+Suppose we have a sample $\set{x_i = 1,z_i,y_i}$ and corresponding unobserved $\set{\epsilon_{x,i},\epsilon_{y,i}}$ from the model in figure \ref{fig:before}.
+The counterfactual $Y_1$, the value of y that would have been observed had $x=1$, is $y_i$ (since $x$ did equal 1).
+$Y_0$ is the value of y that would have been observed had $x=0$.
+Even if we know $f_y$, we cannot in general determine $Y_0$ because we did not observe $\epsilon_y$.
+
+So now we have one thing that we know with certainty and another about which we could say something about in expectation.
+Does defining treatment effects in terms of counterfactuals in this way have lower noise - than in the other way because we are taking expectations only with respect to one of the parts?
 

### 2016-09-18 causal-bandit/nips_2016/code/algorithms.py
work in progress on scaling more general models

@@ -1,0 +1,120 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Sep 13 11:15:43 2016
+
+@author: finn
+"""
+
+import numpy as np
+from math import sqrt, log
+from causal_bandit_utils import random_eta, argmax_rand
+from scipy.optimize import minimize
+
+
+class GeneralCausal(object):
+    
+    def run(self,T,model,eta,m):
+        n = len(eta)
+        self.B = sqrt(m*T/log(2.0*T*n))
+        actions = range(n)
+        print "possible actions",actions
+        u = np.zeros(n)
+        for t in xrange(T):
+            a = np.random.choice(actions,p=eta)
+            x,y = model.sample(a) #x is an array containing values for each variable
+            pa = model.P(x)
+            r = model.R(pa,eta)
+            z = r*y #(r <= self.B)*r*y #Truncation turned off to run parallel bandit experiments
+            u += z
+        self.u = u/float(T)
+        best_action = np.argmax(u)
+        return model.optimal - model.expected_rewards[best_action]   
+    
+        
+    def find_eta(self,model,length_eta = None):
+        if length_eta is None:
+            length_eta = model.K
+        print "in here"
+        eta0 = random_eta(length_eta)
+        res = minimize(model.m, eta0,bounds = [(0.0,1.0)]*length_eta, constraints=({'type':'eq','fun':lambda eta: eta.sum()-1.0}),options={'disp': True},method='SLSQP')
+        assert res.success, " optimization failed to converge"+res.message
+        return res.x,res.fun
+
+#TODO modify so this can be run on non parallel model    
+class ParallelCausal(object):
+    
+    def run(self,T,model):
+        self.trials = np.zeros(model.K)
+        self.success = np.zeros(model.K)
+        h = T/2
+        for t in range(h):
+            x,y = model.sample(model.K) # do nothing
+            xij = np.hstack((1-x,x,1)) # first N actions represent x_i = 0,2nd N x_i=1, last do()
+            self.trials += xij
+            self.success += y*xij
+            
+        infrequent = self.estimate_infrequent(h)
+        n = int(float(h)/len(infrequent))
+        self.trials[infrequent] = n # note could be improved by adding to rather than reseting observation results - does not change worst case. 
+        self.success[infrequent] = model.sample_multiple(infrequent,n)
+        u = np.true_divide(self.success,self.trials)
+        best_action = argmax_rand(u)
[…]

### 2016-09-18 causal-bandit/nips_2016/code/causal_bandit_utils.py

@@ -1,0 +1,41 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Sep 13 11:18:04 2016
+
+@author: finn
+"""
+import numpy as np
+import datetime as dt
+
+def now_string():
+    return dt.now().strftime('%Y%m%d_%H%M')
+
+def random_eta(n):
+    eta = np.random.random(n)
+    eta = eta/eta.sum()
+    return eta
+    
+         
+def argmax_rand(x):
+    """ return the index of the maximum element in the array, ignoring nans. 
+    If there are multiple max valued elements return 1 at random"""
+    max_val = np.nanmax(x)
+    indicies = np.where(x == max_val)[0]
+    return np.random.choice(indicies) 
+
+ 
+
+    
+def prod_all_but_j2(vector):
+    indx = np.range(len(vector))
+    joint = np.prod(vector)
+    joint_on_v = np.true_divide(joint,vector)
+    for j in np.where(np.isnan(joint_on_v))[0]:
+        joint_on_v[j] = np.prod(vector[indx != j])
+    
+    return joint_on_v
+    
+
+    
+
+    

### 2016-09-18 causal-bandit/nips_2016/code/causal_bandits_general.py

@@ -1,0 +1,51 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Feb  3 09:51:47 2016
+
+@author: finn
+"""
+import numpy as np
+import matplotlib.pyplot as plt
+import pickle
+from models import GeneralModel, Parallel, ParallelConfounded
+from algorithms import GeneralCausal,  ParallelCausal, SuccessiveRejects, LilUCB
+
+REGRET_LABEL = "Regret"
+HORIZON_LABEL = "T"
+M_LABEL = "m(q)"
+
+T_vals = range(50,551,100)
+#confounded_parallel = GeneralModel.create_confounded_parallel(3,.1)
+
+#model1 = Parallel.create(10,.1)
+model2 = ParallelConfounded.create(5,.1)
+
+
+alg1 = ParallelCausal()
+alg2 = GeneralCausal()
+sr = SuccessiveRejects()
+
+
+eta,m = alg2.find_eta(model2)
+
+
+regret = np.zeros((2,len(T_vals)))
+for indx,T in enumerate(T_vals):
+    print(T)
+    regret[0,indx] = sr.run(T,model2)
+    regret[1,indx] = alg2.run(T,model2,eta,m)
+        
+
+
+
+
+
+#result_alg1 = alg1.run(T,model)
+
+#result_sr = sr.run(T,model)
+
+
+
+
+
+              

### 2016-09-18 causal-bandit/nips_2016/code/models.py

@@ -1,0 +1,598 @@
+# -*- coding: utf-8 -*-
+"""
+The classes in this file model the enviroment (data generating process) and store the true value of the reward distribution
+over actions. The true reward distribution is not known to the algorithms but is required to calculate regret.
+
+Models should extend the Model class and must implement:
+Attributes
+- expected_rewards: a numpy array containing the true expected reward for each action
+- optimal: max(expected_rewards)
+- K: the size of the action space
+- parent_assignments: a list of numpy arrays where each array is a possible assignment of values to the parents of Y.
+
+Methods
+- P(x): list of length K, returns the probability of the given assignment ,x, to the parents of Y for given each action. 
+- sample(action): samples from the conditional distribution over the parents of Y and Y given the specified action index.
+  returns X (numpy array of length num_parents(Y)), Y (float)
+
+@author: finn
+"""
+import numpy as np
+from itertools import product,chain
+from numpy.random import binomial
+from pgmpy.models import BayesianModel
+from pgmpy.factors.discrete import TabularCPD
+from pgmpy.inference import VariableElimination
+from pgmpy.sampling import BayesianModelSampling
+from scipy.stats import binom
+from scipy.optimize import minimize
+from scipy.optimize import fmin_slsqp
+
+np.set_printoptions(precision=6,suppress=True,linewidth=200)
+def prod_all_but_j(vector):
+    """ returns a vector where the jth term is the product of all the entries except the jth one """
+    zeros = np.where(vector==0)[0]
+    if len(zeros) > 1:
+        return np.zeros(len(vector))
+    if len(zeros) == 1:
+        result = np.zeros(len(vector))
+        j = zeros[0]
+        result[j] = np.prod(vector[np.arange(len(vector)) != j])
+        return result
+
+    joint = np.prod(vector)
+    return np.true_divide(joint,vector)
+
+
+class Model(object):
+    def R(self,pa,eta):
+        """ returns the ratio of the probability of the given assignment under each action to the probability under the eta weighted sum of actions. """
+        Q = (eta*pa).sum()
+        ratio = np.true_divide(pa,Q)
+        ratio[np.isnan(ratio)] = 0 # we get nan when 0/0 but should just be 0 in this case
+        return ratio
+               
+#    def V(self,eta):
[…]

### 2016-09-18 causal-bandit/nips_2016/code/old_experiments.py

@@ -1,0 +1,284 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Sep 13 11:39:19 2016
+
+@author: finn
+"""
+
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Feb  3 09:51:47 2016
+
+@author: finn
+"""
+import numpy as np
+
+from math import sqrt,log,ceil
+
+import matplotlib.pyplot as plt
+from time import time
+
+import pickle
+
+
+REGRET_LABEL = "Regret"
+HORIZON_LABEL = "T"
+M_LABEL = "m(q)"
+
+
+   
+model = PgmpyModel.create_confounded_parallel(3,.1)        
+eta,m_min = find_eta(model)
+              
+def sample_pa(model,p_indx,sims):
+    result = []
+    for action in range(model.K):
+        d = {}
+        result.append(d)
+        for s in xrange(sims):
+            x,y = model.sample(action)
+            p = tuple(x[p_indx])
+            if p in d:
+                d[p] += 1
+            else:
+                d[p] = 1
+        for key,value in d.iteritems():
+            d[key] = value/float(sims)
+    
+    for x in model.parent_vals:
+        px = model.P(x)
+        spx = [dic.get(tuple(x),0.0) for dic in result]
+        print x,px,spx
+        
+            
+    
+#regret,mean,error = experiment1(50,1000,.3,TperK = 6,Tstep = 25)
+    
+def regret_vs_T(N,simulations,epsilon=None,Tstep=None,TperK=10):
+    
+    a = 9.0
+    q = part_balanced_q(N,2) 
+    if epsilon:
+        model = Parallel(q,epsilon)
+        Tmin = 10
+        vary_epsilon = False
+    else:
+        model =  Parallel(q,.5)
+        Tmin = int(ceil(4*model.K/a))
+        vary_epsilon = True
+    
+    Tstep = model.K if not Tstep else Tstep
+    eta,mq = model.analytic_eta()
+     
+    T_vals = range(Tmin,TperK*model.K,Tstep)
+    
+    causal = GeneralCausal()
+    causal_parallel = ParallelCausal()
+    baseline  = SuccessiveRejects()
+    
+    ts = time()   
+    regret = np.zeros((len(T_vals),4,simulations))
+    for s in xrange(simulations):
[…]

### 2016-09-18 causal-bandit/nips_2016/code/optimise_tmp.py

@@ -1,0 +1,14 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Sat Sep 17 15:26:36 2016
+
+@author: finn
+"""
+
+from cvxpy import *
+
+x = Variable(2)
+obj = Minimize(x[0] + norm(x, 1))
+constraints = [x >= 2]
+prob = Problem(obj, constraints)
+prob.solve()

### 2016-09-18 causal-bandit/nips_2016/code/test_models.py

@@ -1,0 +1,30 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Sat Sep 17 11:11:22 2016
+
+@author: finn
+"""
+
+import unittest
+import numpy.testing as np_test
+from models import ParallelConfounded,GeneralModel
+
+class TestModelParallelConfoundedEquivalence(unittest.TestCase):
+    
+    def setUp(self):
+        N = 7
+        N1 = 2
+        pz = .1
+        q = .1,.3,.4,.7
+        epsilon = .2
+        self.model1 = ParallelConfounded.create(N,N1,pz,q,epsilon)          
+        self.model2 = GeneralModel.create_confounded_parallel(N,N1,pz,q,epsilon)
+        
+    def test_P(self):
+        assignments = self.model1.generate_binary_assignments()
+        for x in assignments:
+            diff = self.model1.P(x) - self.model2.P(x)
+            self.assertLess(max(diff),.01)
+            #TODO understand and replace with numpy array almost equal test method
+            
+        

### 2016-09-18 causal-bandit/nips_2016/code/unused.py

@@ -1,0 +1,48 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Sun Sep 18 11:24:25 2016
+
+@author: finn
+"""
+
+class ScaleableParallelConfounded(ParallelConfounded):
+    def __init__(self,q10,q11,q20,q21,pZ,N1,N2,epsilon):
+        ParallelConfounded.__init__(self,q10,q11,q20,q21,pZ,N1,N2,epsilon)
+        self.parent_assignments = list(product(range(N1+1),range(N2+1))) # all combinations of vector of length 2 specifying [num x_i: i<= N1 = 1, num x_i: i > N = 1]
+        self.pa_given_assignment = np.asarray([self.P_of_reparameterized_assignment(n1,n2) for (n1,n2) in self.parent_assignments])
+                
+    def P_of_reparameterized_assignment(self,n1,n2):
+         pa = np.zeros(7)
+         pa[0] = (1-self.pZ)*binom(self.N1-1,self.q10).pmf(n1)*binom(self.N2,self.q20).pmf(n2)+self.pZ*binom(self.N1-1,self.q11).pmf(n1)*binom(self.N2,self.q21).pmf(n2) # do(X1 = 0)
+         pa[1] = (1-self.pZ)*binom(self.N1,self.q10).pmf(n1)*binom(self.N2-1,self.q20).pmf(n2)+self.pZ*binom(self.N1,self.q11).pmf(n1)*binom(self.N2-1,self.q21).pmf(n2) # do(X2 = 0)
+         pa[2] = (1-self.pZ)*binom(self.N1-1,self.q10).pmf(n1-1)*binom(self.N2,self.q20).pmf(n2)+self.pZ*binom(self.N1-1,self.q11).pmf(n1-1)*binom(self.N2,self.q21).pmf(n2) # do(X1=1)
+         pa[3] = (1-self.pZ)*binom(self.N1,self.q10).pmf(n1)*binom(self.N2-1,self.q20).pmf(n2-1)+self.pZ*binom(self.N1,self.q11).pmf(n1)*binom(self.N2-1,self.q21).pmf(n2-1) # do(X2 = 1)
+         pa[4] = binom(self.N1,self.q10).pmf(n1)*binom(self.N2,self.q20).pmf(n2) # do(Z = 0)
+         pa[5] = binom(self.N1,self.q11).pmf(n1)*binom(self.N2,self.q21).pmf(n2) # do(Z = 1)
[…]

### 2016-09-18 causal-bandit/nips_2016/rebuttal.txt

Thank you to all reviewers for your thoughtful comments.

There's a key distiction between context observed before the agent selects an action and feedback after action is selected. If we have context X observed prior to taking an action a, the agent must learn P(Y|X,a). If X is unknown at the point an action is chosen then the algorithm needs to learn P(Y|a). Real problems can involve a mixture of pre and post intervention feedback. In the farmer example the nutrition content might be measurable prior to selecting an action, but the temperature and rainfall are not observed until the end of the season, along with the reward. Incorporating context into this work is an interesting avenue for future research.

We focused our experiments in the parallel setting as it allowed us to compare the performance of both our algorithms. We agree it would be worthwile to include some non-parralel settings and to demonstrate the consequences of applying Alg1 to a problem for which it is miss-specified. We can compare against additional algorithms such as UCB, the key point is that incorporating causal information allows us to avoid the lower bounds which all existing algorithms are subject to.

Certainly in most real applications there will be some uncertainty in the causal model and the conditional distributions will be unknown. This is a very interesting (and challenging) topic for future research. We believe this work to be an important first step and worthy of publication.

Reviewer 1 - only regret was our strong assumptions - not much more to say?

Reviewer 2

The causal information (graph + conditionals) determines R_a for each action and thus how much playing one action reveals about the reward for another. The effect of this information is nicely summerized by the  general version of m, which extends the idea of 'effective number of arms' to any causal bandit setting. Pearl's do-calculus can be used to determine if the conditional interventional distributions we need can be estimated from observational data.

Reviewer 3

Bareinboim et al (2015) consider a specific causal graph where an unobserved confounder U effects both the action an agent would choose naturally X and the outcome Y. The outcome is ultimately a function of U and the action imposed by the algorithm. Their key insight is that X provides information about the unobserved U, so ignoring it could lead to linear regret. Their TS^c algorithm is a form of contextual bandit with X as context. Unlike in our setting, X must be observed prior to the algorithm selecting an action.

Although the causal relationship between P(Y|do(X)) and P(Y|X) is trivial in the parallel setting, Alg1 uses the fact that it holds to be able to update the rewards for actions that were not selected. Existing algorithms ignore causal structure and would have to learn the rewards for all the arms independently.

The probabilities P(Xi) relate to the value variables take when not intervened on. In the parallel bandit, only one variable is set at each timestep. The remaining variables take their values according to P(Xi).

The majority of existing causal inference liturature is in the batch setting. We used the term online to distinguish our work from this setting.

The analysis of Wu et al is for cummulative regret, doesn't apply to general causal graphs and requires stronger assumptions than we do for the parallel case.

Reviewer 4

Our framework supports probabilistic actions by including proxy variables that influence the variables of interest and defining the set of allowed interventions over these proxies. If agents can't take actions, causal inference and bandit problems are both meaningless.

We nessesarily compare against a baseline that does not use the known causal structure as there are no existing algorithms that can exploit it.

We will improve the clarity of the proofs.

Reviewer 5

We did not make use of the reward signal in the main text as it complicates the algorithms without improving the minmax regret and risks obscuring our key results, particularly for readers from the causal inference side.

### 2016-10-13 causal-bandit/nips_2016/additional_experiments.tex
starts adding additional experiments

\begin{figure}
	\centering    
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
%every node/.style={scale=0.6}
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){$Y$};
\node[main node, above right=of 2](6){$Z$};
 \path[every node/.style={font=\tiny}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5)
    (6) edge (1) edge (2) edge (4);
\end{tikzpicture}
        \caption{Confounded graph}
        \label{fig:parallel_confounded} 
\end{figure}

In this section we compare performance of our algorithms with the standard bandit algorithms on the confounded graph in figure \ref{fig:parallel_confounded}. All the variables are binary and the action space, $\calA$, consists of the set of single variable interventions $do(V_i = j)$. We choose this setting as it can be seen as a generalisation of the parallel bandit case and allows for factorizations that enable us to compute the exact reward distribution and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$). As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.

In figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$.

\eq{
P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
}

Unknown to the algorithms, $Y$ depends only on $X_1$ and $X_N$. 
\eq {
P(Y|\boldsymbol{X}) = \begin{cases} 
a & \text{ if } X_1=0,X_N = 0 \\
b & \text{ if } X_1=0,X_N = 1 \\
c & \text{ if } X_1=1,X_N = 0 \\
d & \text{ if } X_1=1,X_N = 1 \\
\end{cases}
}

We then vary $N_1$ from 2 to $N$ to vary $m$. The values for the CPD's have been chosen such that we can analytically calculate $\eta*$. This allows us to just show the dependence on $m$, removing the noise associated with different models producing selecting values for $\eta$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.

In figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.

In figure \ref{fig:simple_vs_T_misspecified} we exclude actions on $Z$ from the set of allowable actions to demonstrate how algorithm 1 can fail in the presence of a hidden confounding variable. Algorithm 1 incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates. 
\eq{
P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
}

These values have been chosen to show the worst case, where the bias results in Algorithm 1 asymptotically selecting a non-optimal action.

\begin{figure}[H]
    \begin{subfigure}[t]{0.3\textwidth}
		\centering    
    		\includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
        \label{fig:simple_vs_m_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
        \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
        \label{fig:simple_vs_T_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
    		\includegraphics[width=\textwidth]{experiment3_N50_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
    		\label{fig:simple_vs_T_misspecified}
    \end{subfigure}
    \caption{Experimental results on the confounded graph}
    \label{fig:experiments}
\end{figure}

### 2016-10-13 causal-bandit/nips_2016/tech.tex
fixes some inconsistancies in notation and adds extra steps for proof of theorem3
starts adding additional experiments

@@
 are independent and permutations of the variables can be pushed to the reward function.
+
+Throughout, let $m$ be shorthand for $m(\vec{q})$ and $\hat{m}$ for $m(\hat{\vec{q}})$
+
 The proof of Theorem \ref{thm:uq-simple} requires some lemmas. \
@@
 \begin{lemma}\label{lem:m_est}
+
 Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
@@
 }
-For arms not in $a$ we have $\hat p_a \geq 1/m(\vec{\hat q}) \geq 1/(2m)$.
+For arms not in $A$ we have $\hat p_a \geq 1/m(\vec{\hat q}) \geq 1/(2m)$.
 Therefore if $a = do(X_i = j)$, then 

### 2016-10-13 causal-bandit/nips_2016/additional_experiments.tex
updates experiment figuresq

@@
 
-Unknown to the algorithms, $Y$ depends only on $X_1$ and $X_N$. 
-\eq {
-P(Y|\boldsymbol{X}) = \begin{cases} 
-a & \text{ if } X_1=0,X_N = 0 \\
-b & \text{ if } X_1=0,X_N = 1 \\
-c & \text{ if } X_1=1,X_N = 0 \\
-d & \text{ if } X_1=1,X_N = 1 \\
-\end{cases}
-}
-
-We then vary $N_1$ from 2 to $N$ to vary $m$.
-The values for the CPD's have been chosen such that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models producing selecting values for $\eta$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.
+As in the parellel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1)=1) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
+We vary $N_1$ from 2 to $N$ to vary $m$.
+The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
+This allows us to just show the dependence on $m$, removing the noise associated with different models producing selecting values for $\eta*$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.
 
@@
 Algorithm 1 incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates.
+To show the worst case outcome, where this bias results in Algorithm 1 asymptotically selecting a sub-optimal action, we need a slightly more complex CPD's.
+
 \eq{
-P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
-P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
+P(X_i = 1|Z = 0) &= 
+\begin{cases} 
+.166 & \text{ if } i \in \set{1,...6} \\
+.2 & \text{ if } i = 6 \\
+.7 & \text { otherwise} 
+ \end{cases}\\
+ P(X_i = 1|Z = 1) &= 
+\begin{cases} 
+.166 & \text{ if } i \in \set{1,...6} \\
+.8 & \text{ if } i = 6 \\
+.3 & \text { otherwise} 
+ \end{cases}\\
 }
 
-These values have been chosen to show the worst case, where the bias results in Algorithm 1 asymptotically selecting a non-optimal action.
+$Y$ depends only on $X_6$ and $X_N$. 
+\eq {
+P(Y|\boldsymbol{X}) = \begin{cases} 
+a & \text{ if } X_6=0,X_N = 0 \\
+b & \text{ if } X_6=0,X_N = 1 \\
+c & \text{ if } X_6=1,X_N = 1 \\
+d & \text{ if } X_6=1,X_N = 0 \\
+\end{cases}
+}
 
@@
 		\centering    
-    		\includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160519_1030.pdf}
+    		\includegraphics[width=\textwidth]{experiment4_20161012_1536.pdf}
     		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
@@
     		\centering
-        \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160519_1030.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
+        \includegraphics[width=\textwidth]{experiment7_20161012_1147.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
         \label{fig:simple_vs_T_general}
@@
     		\includegraphics[width=\textwidth]{experiment3_N50_s10000_20160519_1030.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
+    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=xx$ with no actions setting $Z$}
     		\label{fig:simple_vs_T_misspecified}

### 2016-10-13 causal-bandit/nips_2016/code/experiment1.py
updates experiments 1-3 to run extra algorithms, cpd values for experiments 4 and 6

@@ -6,34 +6,40 @@ Created on Tue Sep 20 16:47:47 2016
 """
 import numpy as np
 from models import Parallel
-from algorithms import GeneralCausal, ParallelCausal, SuccessiveRejects
-from experiment_config import now_string, Experiment
+from algorithms import GeneralCausal, ParallelCausal, SuccessiveRejects,AlphaUCB,ThompsonSampling
+from experiment_config import Experiment
 
 
 def regret_vs_m(algorithms,m_vals,N,T,epsilon,simulations = 10):  
+    models = []    
     regret = np.zeros((len(algorithms),len(m_vals),simulations))
     for m_indx,m in enumerate(m_vals):
         model = Parallel.create(N,m,epsilon)
+        models.append(model)
         print "built model {0}".format(m)
         for s in xrange(simulations):
             for a_indx,algorithm in enumerate(algorithms):
                 regret[a_indx,m_indx,s] = algorithm.run(T,model)
     
-    return regret
+    return regret,models
 
 experiment = Experiment(1)
+experiment.log_code()
           
 # Experiment 1
 N = 50
 epsilon = .3
-simulations = 100
+simulations = 10000
 T = 400
-algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects()]
+algorithms = [GeneralCausal(truncate='None'),ParallelCausal(),SuccessiveRejects(),AlphaUCB(2),ThompsonSampling()]
 m_vals = range(2,N,2)
     
-regret = regret_vs_m(algorithms,m_vals,N,T,epsilon,simulations = simulations)
-finished = now_string()
+regret,models = regret_vs_m(algorithms,m_vals,N,T,epsilon,simulations = simulations)
 
-experiment.plot_regret(regret,m_vals,"m",algorithms,finished)
-experiment.log_code(finished)
-experiment.log_regret(regret,finished)
+
+experiment.plot_regret(regret,m_vals,"m",algorithms,legend_loc="lower right")
+experiment.log_regret(regret,m_vals)
+experiment.log_state(globals())
+
+
+    

### 2016-10-13 causal-bandit/nips_2016/code/experiment2.py

@@ -7,7 +7,7 @@ Created on Wed Sep 21 11:30:34 2016
 import numpy as np
 from math import sqrt,ceil
 from models import Parallel
-from algorithms import GeneralCausal, ParallelCausal, SuccessiveRejects
+from algorithms import GeneralCausal, ParallelCausal, SuccessiveRejects, AlphaUCB, ThompsonSampling
 from experiment_config import now_string, Experiment
 
 
@@ -24,9 +24,12 @@ def regret_vs_T_vary_epsilon(model,algorithms,T_vals,simulations = 10):
                 regret[a_indx,T_indx,s] = algorithm.run(T,model)
         
     return regret
+    
+experiment = Experiment(2)
+experiment.log_code()
 
 N= 50
-simulations = 100
+simulations = 10000
 a = 9.0
 m = 2
 model = Parallel.create(N,m,.1)
@@ -35,15 +38,14 @@ Tmin = int(ceil(4*model.K/a))
 Tmax = 10*model.K
 T_vals = range(Tmin,Tmax,100)
 
-algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects()]
+algorithms = [GeneralCausal(truncate='None'),ParallelCausal(),SuccessiveRejects(),AlphaUCB(2),ThompsonSampling()]
 
 regret = regret_vs_T_vary_epsilon(model,algorithms,T_vals,simulations = simulations)
 finished = now_string()
 
-experiment = Experiment(2)
-experiment.plot_regret(regret,T_vals,"T",algorithms,finished)
-experiment.log_code(finished)
-experiment.log_regret(regret,finished)
+experiment.plot_regret(regret,T_vals,"T",algorithms,legend_loc = None)
+experiment.log_regret(regret,T_vals)
+experiment.log_state(globals())
 
 
 

### 2016-10-13 causal-bandit/nips_2016/code/experiment3.py

@@ -5,7 +5,7 @@ Created on Wed Sep 21 11:52:59 2016
 @author: finn
 """
 import numpy as np
-from algorithms import GeneralCausal,ParallelCausal,SuccessiveRejects
+from algorithms import GeneralCausal,ParallelCausal,SuccessiveRejects, AlphaUCB,ThompsonSampling
 from models import Parallel
 from experiment_config import Experiment, now_string
 
@@ -21,21 +21,23 @@ def regret_vs_T(model,algorithms,T_vals,simulations = 10):
                 
     return regret
            
-                     
-simulations = 100
+experiment = Experiment(3)
+experiment.log_code()
+                  
+simulations = 10000
 N = 50
 m = 2
 epsilon = .3
 model = Parallel.create(N,m,epsilon)
 T_vals = range(10,6*model.K,25)
-algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects()]
+algorithms = [GeneralCausal(truncate='None'),ParallelCausal(),SuccessiveRejects(),AlphaUCB(2),ThompsonSampling()]
 
 regret = regret_vs_T(model,algorithms,T_vals,simulations = simulations)
 finished = now_string()
 
-experiment = Experiment(3)
-experiment.log_code(finished)
-experiment.log_regret(regret,finished)
-experiment.plot_regret(regret,T_vals,"T",algorithms,finished)
+
+experiment.log_regret(regret,T_vals)
+experiment.plot_regret(regret,T_vals,"T",algorithms,legend_loc = None)
+experiment.log_state(globals())
 
 

### 2016-10-13 causal-bandit/nips_2016/code/experiment4.py

@@ -18,9 +18,9 @@ def regret_vs_m_general(algorithms,N1_vals,N,T,pz,pY,q,epsilon,simulations = 100
     regret = np.zeros((len(algorithms),len(N1_vals),simulations))
 
     for m_indx,N1 in enumerate(N1_vals):
-        model = ScaleableParallelConfounded(q,pz,pY,N1,N-N1)
-        #eta = [0,0,1.0/(N1+6.0),0,0,0,1-N1/(N1+6.0)]
-        #model.compute_m()
+        model = ScaleableParallelConfounded(q,pz,pY,N1,N-N1,compute_m = False)
+        eta = [0,0,1.0/(N1+2.0),0,0,0,1-N1/(N1+2.0)]
+        model.compute_m(eta_short = eta)
         # find the v 
         
         print N1,model.m
@@ -42,10 +42,13 @@ N1_vals = range(1,N,3)
 pz = .4
 q = (0.00001,0.00001,.4,.65)
 epsilon = .3
-simulations = 5000
+simulations = 10000
 T = 400
 algorithms = [SuccessiveRejects(),GeneralCausal(),AlphaUCB(2),ThompsonSampling()]
-pY = np.asarray([[.4,.4],[.7,.7]])
+#pY = np.asarray([[.4,.4],[.7,.7]])
+
+epsilon = .3
+pY = ParallelConfounded.pY_epsilon_best(q,pz,epsilon)
 
 m_vals,regret,models = regret_vs_m_general(algorithms,N1_vals,N,T,pz,pY,q,epsilon,simulations = simulations)
 

### 2016-10-13 causal-bandit/nips_2016/code/experiment7.py

@@ -5,7 +5,7 @@ Created on Mon Oct  3 08:19:07 2016
 @author: finn
 """
 
-from models import ScaleableParallelConfounded
+from models import ScaleableParallelConfounded,ParallelConfounded
 from algorithms import SuccessiveRejects,GeneralCausal,AlphaUCB,ThompsonSampling
 from experiment_config import now_string,Experiment
 import numpy as np
@@ -32,16 +32,16 @@ experiment.log_code()
                 
 N = 50
 N1 = 1
-pz = .3
-q = (.1,.9,.2,.7) #q = (.3,.3,.4,.6)
-epsilon = .2
-pY = ScaleableParallelConfounded.pY_epsilon_best(q,pz,epsilon)
+pz = .4
+q = (0.00001,0.00001,.4,.65)
+epsilon = .3
+pY = ParallelConfounded.pY_epsilon_best(q,pz,epsilon)
 
-simulations = 1000
+simulations = 10000
 
 model = ScaleableParallelConfounded(q,pz,pY,N1,N-N1)
 
-T_vals = range(25,451,25)
+T_vals = range(25,626,25)
 
 algorithms = [GeneralCausal(),SuccessiveRejects(),AlphaUCB(2),ThompsonSampling()]
 

### 2016-10-13 causal-bandit/nips_2016/code/models.py

@@ -199,6 +199,12 @@ class Parallel(Model):
                 return indx
         return len(qij_sorted)/2 
         
+    def action_tuple(self,action):
+        """ convert from action id to the tuple (varaible,value) """
+        if action == 2*self.N+1:
+            return ((None,None))
+        return (action % self.N, action/self.N)
+        
     def set_epsilon(self,epsilon):
         assert epsilon <= .5 ,"epsilon cannot exceed .5"
         self.epsilon = epsilon

### 2016-10-13 causal-bandit/nips_2016/code/experiment5.py
updates experiment5 parameters

@@ -33,14 +33,15 @@ def regret_vs_T(model,algorithms,T_vals,simulations = 10):
 experiment = Experiment(5)
 experiment.log_code()
            
-simulations = 5000                 
+simulations = 10000                 
 N =20
 N1 = 1
 q = (.2,.8,.7,.3)
 pz = 0.523878528335 
-pY = np.asarray([[ 0.02 , 0.95  ],[0.9  , 0.03]])
+#pY = np.asarray([[ 0 , 0.95  ],[0.9  , 0.03]])
+pY = np.asarray([[ 0 , 1  ],[1 , 0]])
 
-N0 = 5
+N0 = 6
 N1 = 1
 N2 = 14
 
@@ -51,7 +52,7 @@ pXgivenZ = np.stack((np.vstack((1.0-pXgivenZ0,pXgivenZ0)),np.vstack((1.0-pXgiven
 pYfunc = lambda x: pY[x[N0],x[N-1]]
 model = ParallelConfoundedNoZAction(pz,pXgivenZ,pYfunc)
 
-T_vals = range(25,2026,500)
+T_vals = range(25,401,25)
 
 algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects()]
 

### 2016-10-15 causal-bandit/nips_2016/tech.tex
updates proof of theorem 1

@@
 are independent and permutations of the variables can be pushed to the reward function.
-
-Throughout, let $m$ be shorthand for $m(\vec{q})$ and $\hat{m}$ for $m(\hat{\vec{q}})$
 
@@
 Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE[X_i] = p$ and $\delta \in [0,1]$. 
-Then 
 \eq{
@@
 \begin{proof}
-For $\delta \geq 1/4$ the result is trivial. Otherwise 
-by Hoeffding's bound and the union bound:
+For $\delta \geq 1/4$ the result is trivial. Otherwise, Hoeffding's inequality gives,
+\eq{
+\P{\left|\frac{1}{t}\sum_{s=1}^t{X_s} - p \right| > \epsilon} \leq 2 e^{-2t\epsilon^2}
+} Let $\epsilon = \sqrt{\frac{2}{n_0}\log{\frac{2}{\delta}}}$ and taking the union bound over $t\geq n_0$,
+
 \eq{
 \P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} 
-&\leq \sum_{t = n_0}^\infty \P{\left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \\
-&\leq 2\sum_{t=n_0}^\infty \exp\left(-\frac{t}{n_0} \log \frac{2}{\delta}\right) 
-\leq 4\delta\,. \qedhere
+&\leq 2\sum_{t=n_0}^\infty \exp\left(-\frac{4t}{n_0} \log \frac{2}{\delta}\right)\\
+& = 2\frac{r^{n0}}{1-r} \text{ where } r = (\frac{\delta}{2})^{4/n_0}
 }
 \end{proof}
+THIS WOULD SEEM TO DIVERGE AS $n_0$ goes to $\infty$. HOW DO WE GET $\leq 4\delta\,. \qedhere$
 
@@
 when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
-From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m$ that $q_i \geq q_m \geq 1/m$ and so by Lemma \ref{lem:conc1} we have
+From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m(\vec{q})$ that $q_i \geq q_m \geq 1/m(\vec{q})$ and so by Lemma \ref{lem:conc1} we have
 \eq{
@@
 &\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
-\geq q_i - \sqrt{\frac{q_i}{8m}}
-\geq \frac{1}{2m}\,.
+\geq q_i - \sqrt{\frac{q_i}{8m(\vec{q})}}
+\geq \frac{1}{2m(\vec{q})}\,.
 }
-Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m$.
-For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m$ that
+Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m(\vec{q})$.
+For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m(\vec{q})$ that
 \eq{
@@
 \leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
-\leq \frac{1}{m} \left(1 + \sqrt{\frac{1}{8}}\right)
-\leq \frac{3}{2m}\,.
+\leq \frac{1}{m(\vec{q})} \left(1 + \sqrt{\frac{1}{8}}\right)
+\leq \frac{3}{2m(\vec{q})}\,.
 }
@@
 \begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
-Let $\delta = m = m(\vec{q}) / N$. Then by Lemma \ref{lem:m_est} we have 
-\eq{
-\P{2m/3 \leq m(\vec{\hat q}) \leq 2m} \geq 1 - \delta\,.
-}
 Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then
-for $a \in A$ the algorithm estimates $\mu_a$ from $T/(2m(\vec{\hat q})) \geq T/(4m)$ samples.
+for $a \in A$ the algorithm estimates $\mu_a$ from $T/(2m(\vec{\hat q})) \geq T/(4m(\vec{q}))$ samples.
 Therefore by Hoeffding's inequality and the union bound we have
 \eq{
-\P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{8m}{T} \log\frac{2N}{\delta}}} \leq \delta\,.
+\P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq \delta\,.
 }
-For arms not in $A$ we have $\hat p_a \geq 1/m(\vec{\hat q}) \geq 1/(2m)$.
-Therefore if $a = do(X_i = j)$, then 
+For arms not in $A$ we have $\hat p_a \geq 1/m(\vec{\hat q})$ For $a = do(X_i = j)$, if $a = do(X_i = j)$, then from Lemma \ref{lem:m_est},
 \eq{
-\hat p_a = \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{1}{2m}\,. 
+\hat p_a = \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{1}{2m(\vec{q})}\, \text{ with probability } 1-\delta
 }
-Therefore $\sum_{t=1}^{T/2} \ind{X_{t,i} = j} \geq T/4m$
-and by Lemma \ref{lem:conc2} we have
+
+Let $H$ be the event $\sum_{t=1}^{T/2} \ind{X_{t,i} = j} < T/4m(\vec{q})$ and $G$ be the event $\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}}$.
+
 \eq{
-\P{\sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{T}{4m} \text{ and } \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}}} \leq 4\delta / N\,.
+P(G) & \leq P(H) + P(G|\neg H)\\
+& \leq \delta + 4\delta \longleftarrow \text{ by Lemma \ref{lem:conc2}, letting } n_0 = \frac{T}{4m(\vec{q})}
 }
-Therefore with probability at least $1 - 6\delta$ we have
+
+WHY DO WE NEED THE UNION BOUND AND THUS Lemma \ref{lem:conc2}.
+For each of these arms we select them some $t > n_0$ times.
+This $t$ is a random variable but doesn't depend on the empirical reward estimates for $s \in {n_0 ...
+t}$.
+
+Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 6\delta$ that,
 \eq{
-(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{8m}{T} \log \frac{N}{\delta}} = \epsilon\,.
+(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}} = \epsilon\,.
 }
@@
 \mu^* - \EE[\mu_{\hat a^*_T}] 
-\leq 6\delta + \epsilon 
-\leq \frac{6m}{T} + \sqrt{\frac{32m}{T} \log \frac{NT}{m}}\,, 
+& \leq 6\delta + \epsilon\\
+& \leq \frac{6m}{T} + \sqrt{\frac{16m}{T} \log \frac{NT}{m}}, \longleftarrow \text{ letting } \delta = \frac{m}{T}\\
+& \leq \sqrt{\frac{44m}{T} \log \frac{NT}{m}}\ \longleftarrow \text{ via Jenson's Inequality}
 }

### 2016-10-20 causal-bandit/nips_2016/code/experiment_config.py
sets min and max axis points

@@ -48,6 +48,8 @@ class Experiment(object):
             ax.errorbar(xvals,mu[indx,:],yerr = error[indx,:],label = alg.label,linestyle="",marker = alg.marker,color=alg.color)
         ax.set_xlabel(xlabel)
         ax.set_ylabel(self.REGRET_LABEL)
+        ax.set_ylim(bottom = 0)
+        ax.set_xlim(left = 0,right = max(xvals))
         
         if legend_loc is not None:
             ax.legend(loc = legend_loc,numpoints=1)

### 2016-10-20 causal-bandit/nips_2016/code/experiment5.py
adds alphaUCB and ThompsonSampling, removes legend from plot

@@ -8,8 +8,8 @@ If the resulting bias exceeds epsilon then the Parallel algorithm will never ide
 """
 
 
-from models import ScaleableParallelConfoundedNoZAction,ParallelConfoundedNoZAction
-from algorithms import SuccessiveRejects,GeneralCausal,ParallelCausal,RandomArm,AlphaUCB,ThompsonSampling
+from models import ParallelConfoundedNoZAction
+from algorithms import SuccessiveRejects,GeneralCausal,ParallelCausal,AlphaUCB,ThompsonSampling
 from experiment_config import Experiment
 import numpy as np
 
@@ -54,9 +54,9 @@ model = ParallelConfoundedNoZAction(pz,pXgivenZ,pYfunc)
 
 T_vals = range(25,401,25)
 
-algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects()]
+algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects(),ThompsonSampling(),AlphaUCB(2)]
 
 regret,pulls = regret_vs_T(model,algorithms,T_vals,simulations = simulations)
 
-experiment.plot_regret(regret,T_vals,"T",algorithms,legend_loc = "lower left")
+experiment.plot_regret(regret,T_vals,"T",algorithms,legend_loc = None)
 experiment.log_state(globals())

### 2016-10-20 causal-bandit/nips_2016/additional_experiments.tex
updates experiment5
updates figures to remove type3 fonts
fits main paper back into 8 pages, makes description of additional experiments more concise

@@
-\begin{figure}
+\begin{figure}[H]
 	\centering    
@@
 
-As in the parellel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1)=1) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
+As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1)=1) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
 We vary $N_1$ from 2 to $N$ to vary $m$.
@@
 Algorithm 1 incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates.
-To show the worst case outcome, where this bias results in Algorithm 1 asymptotically selecting a sub-optimal action, we need a slightly more complex CPD's.
+We let $P(Z) = .52$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
 
@@
 \begin{cases} 
-.166 & \text{ if } i \in \set{1,...6} \\
-.2 & \text{ if } i = 6 \\
+.166 & \text{ if } i \in \set{1,..., 6} \\
+.2 & \text{ if } i = 7 \\
 .7 & \text { otherwise} 
@@
 \begin{cases} 
-.166 & \text{ if } i \in \set{1,...6} \\
-.8 & \text{ if } i = 6 \\
+.166 & \text{ if } i \in \set{1,..., 6} \\
+.8 & \text{ if } i = 7 \\
 .3 & \text { otherwise} 
@@
 
-$Y$ depends only on $X_6$ and $X_N$. 
-\eq {
-P(Y|\boldsymbol{X}) = \begin{cases} 
-a & \text{ if } X_6=0,X_N = 0 \\
-b & \text{ if } X_6=0,X_N = 1 \\
-c & \text{ if } X_6=1,X_N = 1 \\
-d & \text{ if } X_6=1,X_N = 0 \\
-\end{cases}
-}
+The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.
+All remaining variables are likely to be identified as balanced and estimated from observational estimates.
+This leads to the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
 
@@
 		\centering    
-    		\includegraphics[width=\textwidth]{experiment4_20161012_1536.pdf}
+    		\includegraphics[width=\textwidth]{experiment4_20161020_1254.pdf}
     		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
@@
     		\centering
-        \includegraphics[width=\textwidth]{experiment7_20161012_1147.pdf}
+        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
     		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
@@
     		\centering
-    		\includegraphics[width=\textwidth]{experiment3_N50_s10000_20160519_1030.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=xx$ with no actions setting $Z$}
+    		\includegraphics[width=\textwidth]{experiment5_20161020_2027.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 20$, $m=4.3$ with no actions setting $Z$}
     		\label{fig:simple_vs_T_misspecified}

### 2016-10-20 causal-bandit/nips_2016/code/experiment5.py

@@ -35,7 +35,6 @@ experiment.log_code()
            
 simulations = 10000                 
 N =20
-N1 = 1
 q = (.2,.8,.7,.3)
 pz = 0.523878528335 
 #pY = np.asarray([[ 0 , 0.95  ],[0.9  , 0.03]])

### 2016-10-22 causal-bandit/nips_2016/simple-regret.tex
starts adding additional experiments
fits main paper back into 8 pages, makes description of additional experiments more concise
fixed page limits

@@
     \caption{Causal Models}\label{fig:causalmodels}
+    \vspace{-.5cm}
 \end{figure}
@@
 Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
-The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates their returns could be poor.
+The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates of their returns could be poor.
 To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions.
@@
 \eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m}\right)}}\,.
+\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.
 }
@@
 \begin{theorem}\label{thm:lower}
-For all $T$, $\vec{q}$ and all strategies, there exists a reward function such that
-\eq{
-\simpleregret 
-\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.
-}
+For all strategies and $T$, $\vec{q}$, there exist rewards such that
+$\displaystyle \simpleregret 
+\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)$.
 \end{theorem}

### 2016-10-23 causal-bandit/nips_2016/additional_experiments.tex
updates additional experiments and proofs

@@
 In this section we compare performance of our algorithms with the standard bandit algorithms on the confounded graph in figure \ref{fig:parallel_confounded}.
-All the variables are binary and the action space, $\calA$, consists of the set of single variable interventions $do(V_i = j)$.
+All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action, $\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}$.
 We choose this setting as it can be seen as a generalisation of the parallel bandit case and allows for factorizations that enable us to compute the exact reward distribution and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
@@
 
-As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1)=1) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
+As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
 We vary $N_1$ from 2 to $N$ to vary $m$.
 The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models producing selecting values for $\eta*$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.
+This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.
 
@@
 
-In figure \ref{fig:simple_vs_T_misspecified} we exclude actions on $Z$ from the set of allowable actions to demonstrate how algorithm 1 can fail in the presence of a hidden confounding variable.
+In figure \ref{fig:simple_vs_T_misspecified} we exclude actions on $Z$ from the set of allowable actions to demonstrate how algorithm 1 can fail in the presence of a confounding variable.
 Algorithm 1 incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates.
-We let $P(Z) = .52$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
+We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
 
@@
 
+In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree.
+It is sub-optimal to act on either $X_7$ or $X_N$, all other actions are optimal.
 The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.
 All remaining variables are likely to be identified as balanced and estimated from observational estimates.
-This leads to the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
+The CPD values have been chosen to demonstrate the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
 
@@
     		\centering
-    		\includegraphics[width=\textwidth]{experiment5_20161020_2027.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 20$, $m=4.3$ with no actions setting $Z$}
+    		\includegraphics[width=\textwidth]{experiment5_20161021_2026.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m=4.3$ with no actions setting $Z$}
     		\label{fig:simple_vs_T_misspecified}
@@
     \caption{Experimental results on the confounded graph}
-    \label{fig:experiments}
+    \label{fig:experiments_confounded}
 \end{figure}

### 2016-10-23 causal-bandit/nips_2016/tech.tex
updates additional experiments and proofs

@@
 \end{proof}
-
-\begin{lemma}\label{lem:conc2}
-Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE[X_i] = p$ and $\delta \in [0,1]$. 
-\eq{
-\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \leq 4\delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-For $\delta \geq 1/4$ the result is trivial. Otherwise, Hoeffding's inequality gives,
-\eq{
-\P{\left|\frac{1}{t}\sum_{s=1}^t{X_s} - p \right| > \epsilon} \leq 2 e^{-2t\epsilon^2}
-} Let $\epsilon = \sqrt{\frac{2}{n_0}\log{\frac{2}{\delta}}}$ and taking the union bound over $t\geq n_0$,
-
-\eq{
-\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} 
-&\leq 2\sum_{t=n_0}^\infty \exp\left(-\frac{4t}{n_0} \log \frac{2}{\delta}\right)\\
-& = 2\frac{r^{n0}}{1-r} \text{ where } r = (\frac{\delta}{2})^{4/n_0}
-}
-\end{proof}
-THIS WOULD SEEM TO DIVERGE AS $n_0$ goes to $\infty$. HOW DO WE GET $\leq 4\delta\,. \qedhere$
 
@@
 \begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
-Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then
-for $a \in A$ the algorithm estimates $\mu_a$ from $T/(2m(\vec{\hat q})) \geq T/(4m(\vec{q}))$ samples.
-Therefore by Hoeffding's inequality and the union bound we have
+Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then,
+for $a \in A$, the algorithm estimates $\mu_a$ from $T_A \doteq T/(2m(\vec{\hat q}))$ samples.
+From lemma \ref{lem:m_est}, $T_A  \geq T/(4m(\vec{q}))$ with probability $(1-\delta)$.
+Let $H$ be the event $T_A < T/(4m(\vec{q}))$ and $G$ be the event $\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}$
+
 \eq{
-\P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq \delta\,.
-}
-For arms not in $A$ we have $\hat p_a \geq 1/m(\vec{\hat q})$ For $a = do(X_i = j)$, if $a = do(X_i = j)$, then from Lemma \ref{lem:m_est},
-\eq{
-\hat p_a = \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{1}{2m(\vec{q})}\, \text{ with probability } 1-\delta
+\P{G} \leq \P{H} + \P{G|\neg H} \leq \delta + \P{G|\neg H}
 }
 
-Let $H$ be the event $\sum_{t=1}^{T/2} \ind{X_{t,i} = j} < T/4m(\vec{q})$ and $G$ be the event $\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}}$.
+Via Hoeffding's inequality and the union bound,
 
 \eq{
-P(G) & \leq P(H) + P(G|\neg H)\\
-& \leq \delta + 4\delta \longleftarrow \text{ by Lemma \ref{lem:conc2}, letting } n_0 = \frac{T}{4m(\vec{q})}
+\P{G|\neg H}&\doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}},\text{ given }T_A \geq T/(4m(\vec{q}))} \leq \delta\\
+ \implies  \P{G} & \doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq 2\delta\,.
+}
+For arms not in $A$,
+
+\eq{
+\hat p_a = & \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq 1/m(\vec{\hat q}), \; \text {by definition of not being in $A$}\\
+ \geq & \frac{1}{2m(\vec{q})},\; \text{ with probability } 1-\delta \\
+\implies  T_a \doteq & \sum_{t=1}^{T/2} \ind{X_i = j} \geq  \frac{T}{4m(\vec{q})},\; \text{ with probability } 1-\delta
 }
 
-WHY DO WE NEED THE UNION BOUND AND THUS Lemma \ref{lem:conc2}.
-For each of these arms we select them some $t > n_0$ times.
-This $t$ is a random variable but doesn't depend on the empirical reward estimates for $s \in {n_0 ...
-t}$.
+Again applying Hoeffding's and the union bound
 
-Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 6\delta$ that,
 \eq{
-(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}} = \epsilon\,.
+\P{\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}}} \leq 2\delta
+}
+
+Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 4\delta$ that,
+\eq{
+(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}} \doteq \epsilon\,.
 }
@@
 \mu^* - \EE[\mu_{\hat a^*_T}] 
-& \leq 6\delta + \epsilon\\
-& \leq \frac{6m}{T} + \sqrt{\frac{16m}{T} \log \frac{NT}{m}}, \longleftarrow \text{ letting } \delta = \frac{m}{T}\\
-& \leq \sqrt{\frac{44m}{T} \log \frac{NT}{m}}\ \longleftarrow \text{ via Jenson's Inequality}
+& \leq 4\delta + \epsilon\\
+& \leq \frac{8m(\vec{q})}{T} + \sqrt{\frac{2m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ letting } \delta = \frac{2m(\vec{q})}{T}\\
+& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jenson's Inequality}
 }

### 2016-10-23 causal-bandit/nips_2016/code/algorithms.py

@@ -59,7 +59,7 @@ class GeneralCausal(object):
   
 class ParallelCausal(object):
     label = "Algorithm 1"
-  
+    
     def run(self,T,model):
         self.trials = np.zeros(model.K)
         self.success = np.zeros(model.K)
@@ -70,10 +70,10 @@ class ParallelCausal(object):
             self.trials += xij
             self.success += y*xij
             
-        infrequent = self.estimate_infrequent(h)
-        n = int(float(h)/len(infrequent))
-        self.trials[infrequent] = n # note could be improved by adding to rather than reseting observation results - does not change worst case. 
-        self.success[infrequent] = model.sample_multiple(infrequent,n)
+        self.infrequent = self.estimate_infrequent(h)
+        n = int(float(h)/len(self.infrequent))
+        self.trials[self.infrequent] = n # note could be improved by adding to rather than reseting observation results - does not change worst case. 
+        self.success[self.infrequent] = model.sample_multiple(self.infrequent,n)
         self.u = np.true_divide(self.success,self.trials)
         self.r = self.u - model.get_costs()
         self.best_action = argmax_rand(self.r)

### 2016-10-23 causal-bandit/nips_2016/code/experiment5.py

@@ -9,7 +9,7 @@ If the resulting bias exceeds epsilon then the Parallel algorithm will never ide
 
 
 from models import ParallelConfoundedNoZAction
-from algorithms import SuccessiveRejects,GeneralCausal,ParallelCausal,AlphaUCB,ThompsonSampling
+from algorithms import SuccessiveRejects,GeneralCausal,ParallelCausal,AlphaUCB,ThompsonSampling,RandomArm
 from experiment_config import Experiment
 import numpy as np
 
@@ -34,28 +34,34 @@ experiment = Experiment(5)
 experiment.log_code()
            
 simulations = 10000                 
-N =20
+
 q = (.2,.8,.7,.3)
-pz = 0.523878528335 
-#pY = np.asarray([[ 0 , 0.95  ],[0.9  , 0.03]])
+pz = 0.6 
 pY = np.asarray([[ 0 , 1  ],[1 , 0]])
 
 N0 = 6
 N1 = 1
 N2 = 14
+N = N0+N1+N2
 
 q10,q11,q20,q21 = q
-pXgivenZ0 = np.hstack((np.full(N0,1.0/N0),np.full(N1,q10),np.full(N2,q20)))    #(np.full(N1,q10),np.full(N2,q20))
+pXgivenZ0 = np.hstack((np.full(N0,1.0/N0),np.full(N1,q10),np.full(N2,q20)))    
 pXgivenZ1 = np.hstack((np.full(N0,1.0/N0),np.full(N1,q11),np.full(N2,q21)))
 pXgivenZ = np.stack((np.vstack((1.0-pXgivenZ0,pXgivenZ0)),np.vstack((1.0-pXgivenZ1,pXgivenZ1))),axis=2) # PXgivenZ[i,j,k] = P(X_j=i|Z=k)
 pYfunc = lambda x: pY[x[N0],x[N-1]]
 model = ParallelConfoundedNoZAction(pz,pXgivenZ,pYfunc)
 
-T_vals = range(25,401,25)
+T_vals = range(10,601,50)
 
 algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects(),ThompsonSampling(),AlphaUCB(2)]
 
 regret,pulls = regret_vs_T(model,algorithms,T_vals,simulations = simulations)
 
 experiment.plot_regret(regret,T_vals,"T",algorithms,legend_loc = None)
-experiment.log_state(globals())
+#experiment.log_state(globals())
+
+print "m",model.m
+
+    
+
+

### 2016-10-23 causal-bandit/nips_2016/additional_experiments.tex
updates plot bounds and figures

@@
 
-In this section we compare performance of our algorithms with the standard bandit algorithms on the confounded graph in figure \ref{fig:parallel_confounded}.
+In this section we compare performance of the general algorithm with the standard bandit algorithms on the confounded graph in figure \ref{fig:parallel_confounded}.
 All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action, $\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}$.
@@
 
-As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ amd $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
+As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
 We vary $N_1$ from 2 to $N$ to vary $m$.
@@
 
-In figure \ref{fig:simple_vs_T_misspecified} we exclude actions on $Z$ from the set of allowable actions to demonstrate how algorithm 1 can fail in the presence of a confounding variable.
-Algorithm 1 incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates.
+In figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1.
+We exclude actions on $Z$ from the set of allowable actions to demonstrate how Algorithm 1 can fail in the presence of a confounding variable, as it incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates.
 We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
@@
 		\centering    
-    		\includegraphics[width=\textwidth]{experiment4_20161020_1254.pdf}
+    		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
     		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
@@
     		\centering
-    		\includegraphics[width=\textwidth]{experiment5_20161021_2026.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m=4.3$ with no actions setting $Z$}
+    		\includegraphics[width=\textwidth]{experiment5_20161023_2118.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m(\eta*)=4.3$ with no actions setting $Z$}
     		\label{fig:simple_vs_T_misspecified}

### 2016-10-23 causal-bandit/nips_2016/code/experiment4.py

@@ -7,7 +7,7 @@ Created on Tue Sep 20 16:48:05 2016
 
 """
 from models import ParallelConfounded,ScaleableParallelConfounded
-from algorithms import SuccessiveRejects,GeneralCausal,AlphaUCB,ThompsonSampling
+from algorithms import SuccessiveRejects,GeneralCausal,AlphaUCB,ThompsonSampling,ParallelCausal
 from experiment_config import Experiment
 import numpy as np
 
@@ -21,8 +21,7 @@ def regret_vs_m_general(algorithms,N1_vals,N,T,pz,pY,q,epsilon,simulations = 100
         model = ScaleableParallelConfounded(q,pz,pY,N1,N-N1,compute_m = False)
         eta = [0,0,1.0/(N1+2.0),0,0,0,1-N1/(N1+2.0)]
         model.compute_m(eta_short = eta)
-        # find the v 
-        
+       
         print N1,model.m
         m_vals.append(model.m)
         models.append(model)
@@ -45,15 +44,15 @@ epsilon = .3
 simulations = 10000
 T = 400
 algorithms = [SuccessiveRejects(),GeneralCausal(),AlphaUCB(2),ThompsonSampling()]
-#pY = np.asarray([[.4,.4],[.7,.7]])
+
 
 epsilon = .3
 pY = ParallelConfounded.pY_epsilon_best(q,pz,epsilon)
 
 m_vals,regret,models = regret_vs_m_general(algorithms,N1_vals,N,T,pz,pY,q,epsilon,simulations = simulations)
 
-experiment.plot_regret(regret,m_vals,"m",algorithms,legend_loc = "lower right")
-experiment.log_state(globals())
+experiment.plot_regret(regret,m_vals,"m",algorithms,legend_loc = "lower right",legend_extra = [ParallelCausal])
+
 
 
 #etas = np.zeros((len(m_vals),7))

### 2016-10-23 causal-bandit/nips_2016/code/experiment5.py

@@ -12,6 +12,7 @@ from models import ParallelConfoundedNoZAction
 from algorithms import SuccessiveRejects,GeneralCausal,ParallelCausal,AlphaUCB,ThompsonSampling,RandomArm
 from experiment_config import Experiment
 import numpy as np
+from itertools import chain
 
 
 
@@ -51,14 +52,14 @@ pXgivenZ = np.stack((np.vstack((1.0-pXgivenZ0,pXgivenZ0)),np.vstack((1.0-pXgiven
 pYfunc = lambda x: pY[x[N0],x[N-1]]
 model = ParallelConfoundedNoZAction(pz,pXgivenZ,pYfunc)
 
-T_vals = range(10,601,50)
+T_vals = list(chain(range(10,100,25),range(85+50,450,50),range(435+100,1036,100)))
 
 algorithms = [GeneralCausal(),ParallelCausal(),SuccessiveRejects(),ThompsonSampling(),AlphaUCB(2)]
 
 regret,pulls = regret_vs_T(model,algorithms,T_vals,simulations = simulations)
 
 experiment.plot_regret(regret,T_vals,"T",algorithms,legend_loc = None)
-#experiment.log_state(globals())
+
 
 print "m",model.m
 

### 2016-10-23 causal-bandit/nips_2016/code/experiment_config.py

@@ -38,7 +38,7 @@ class Experiment(object):
             a.marker = self.markers[indx]
             a.color = self.colors[indx]
               
-    def plot_regret(self,regret,xvals,xlabel,algorithms,legend_loc = "upper right", log = True):
+    def plot_regret(self,regret,xvals,xlabel,algorithms,legend_loc = "upper right", legend_extra = [],log = True):
         s_axis = regret.ndim - 1 # assumes the last axis is over simulations
         simulations = regret.shape[-1]
         mu = regret.mean(s_axis)
@@ -46,9 +46,16 @@ class Experiment(object):
         fig,ax = plt.subplots()
         for indx,alg in enumerate(algorithms):    
             ax.errorbar(xvals,mu[indx,:],yerr = error[indx,:],label = alg.label,linestyle="",marker = alg.marker,color=alg.color)
+        
+        for alg in legend_extra: # trick matplotlib into adding legend for additional algorithms
+            ax.errorbar(-1,-1,yerr=.0001,label = alg.label,color=alg.color, marker = alg.marker,linestyle="")
+
+        upper = np.nanmax(mu+error)
+            
+        
         ax.set_xlabel(xlabel)
         ax.set_ylabel(self.REGRET_LABEL)
-        ax.set_ylim(bottom = 0)
+        ax.set_ylim(bottom = 0,top = upper)
         ax.set_xlim(left = 0,right = max(xvals))
         
         if legend_loc is not None:
@@ -98,7 +105,8 @@ class Experiment(object):
                 print "unable to load value for {0}".format(key)
         db.close()
         return d
-   
+        
+
                 
 
             

### 2016-10-25 causal-bandit/nips_2016/experiments.tex
updates experiment figuresq
updates figures to remove type3 fonts
updates sty file, experiment description adds github link

@@
-We compare Algorithms \ref{alg:simple} and \ref{alg:general} with Successive Elimination on the parallel bandit problem
-under a variety of conditions, including where the importance weighted estimator used by Algorithm \ref{alg:general} is not truncated,
-which is justified in this setting by Remark \ref{rem:truncate}.
+We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \cite{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
+Thomson sampling and UCB are optimized to minimize cumulative regret.
+We apply them in the fixed horizon, best arm identification setting by running them upto horizon $T$ and then selecting the arm with the highest empirical mean.
+The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.
+
 Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms).
@@
 This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
-We set $\boldsymbol{q} = 0$ for $q_i \leq m$ and $\frac{1}{2}$ otherwise.
+We set $q_i = 0$ for $i \leq m$ and $\frac{1}{2}$ otherwise.
 Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
-
-%\eq{
-%Y_t \sim \begin{cases}
-%\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
-%\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
-%\end{cases}
-%}
-
-We compare the performance of the Algorithm 1, which is specific to the parallel problem, but does not require knowledge of $\boldsymbol{q}$, with that of Algorithm 2 and the Successive Reject algorithm of \cite{audibert2010best}.
 For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
+The code is available from \url{<https://github.com/finnhacks42/causal_bandits>}
 
@@
 		\centering    
-    		\includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160519_1030.pdf}
+    		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
     		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
@@
     		\centering
-        \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160519_1030.pdf}
+        \includegraphics[width=\textwidth]{experiment2_20161020_1249.pdf}
     		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
@@
     		\centering
-    		\includegraphics[width=\textwidth]{experiment3_N50_s10000_20160519_1030.pdf}
+    		\includegraphics[width=\textwidth]{experiment3_20161020_1252.pdf}
     		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
@@
 \end{figure}
-
-%\begin{figure}
-%  \centering
-%  \begin{minipage}[b]{0.3\textwidth}
-%    \includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160518_1534.pdf}
-%    \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
-%    \label{fig:simple_vs_m}
-%  \end{minipage}
-%  \hfill
-%  \begin{minipage}[b]{0.3\textwidth}
-%    \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160518_1534.pdf}
-%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
-%    \label{fig:simple_vs_T_vary_epsilon}
-%  \end{minipage}
-%  \hfill
-%  \begin{minipage}[b]{0.3\textwidth}
-%    \includegraphics[width=\textwidth]{experiment3_N50_s10000_20160518_1534.pdf}
-%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
-%    \label{fig:simple_vs_T}
-%  \end{minipage}
-%\end{figure}
 

### 2016-10-26 causal-bandit/nips_2016/additional_experiments.tex
updates to additional experiments section

@@
 
-In this section we compare performance of the general algorithm with the standard bandit algorithms on the confounded graph in figure \ref{fig:parallel_confounded}.
+We now compare the general algorithm with a range of standard bandit algorithms on the confounded graph in Figure \ref{fig:parallel_confounded}.
 All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action, $\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}$.
-We choose this setting as it can be seen as a generalisation of the parallel bandit case and allows for factorizations that enable us to compute the exact reward distribution and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
+We choose this setting because it generalises the parallel bandit, while simultaneously being sufficiently simple that we can compute the exact reward and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
 As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.
 
-In figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$.
-
+In Figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$. For some $2 \leq N_1 \leq N$ we define 
 \eq{
@@
 }
+As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon P(X_1=1) / P(X_1=0)$.
+The value of $N_1$ determines $m$ and ranges between $2$ and $N$.
+The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
+This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
 
-As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon\frac{P(X_1=1)}{P(X_1=0)}$.
-We vary $N_1$ from 2 to $N$ to vary $m$.
-The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and thus same worst case performance) but different performance for a given reward distribution.
-
-In figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.
-
-In figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1.
-We exclude actions on $Z$ from the set of allowable actions to demonstrate how Algorithm 1 can fail in the presence of a confounding variable, as it incorrectly assumes $P(Y|do(X)) = P(Y|X)$ leading to biased estimates.
+In Figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.  
+In Figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1, but exclude actions on $Z$ from the set of allowable actions to demonstrate that Algorithm 1 can fail in the presence of a confounding variable, which occurs because it incorrectly assumes that $P(Y|do(X)) = P(Y|X)$.
 We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
@@
 In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree.
-It is sub-optimal to act on either $X_7$ or $X_N$, all other actions are optimal.
+It is sub-optimal to act on either $X_7$ or $X_N$, while all other actions are optimal.
 The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.

### 2016-10-28 causal-bandit/nips_2016/experiments.tex
mark's changes

@@
 Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
-This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
-In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
+This gap is learnable by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
+In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
 The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

### 2016-10-28 causal-bandit/nips_2016/general.tex
updates additional experiments and proofs
mark's changes

--- 2016-05-17 causal-bandit/nips_2016/general.tex
@@
 For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}.
+However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \citep{Pearl2000}.
 \eq{
@@
 
-We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational estimates were poor.
-However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable from rounds in which we act on a different variable.
-Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$.
-We can learn the reward for all the interventions $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$, but not from $do()$.
-In addition, variance of the observational estimator for $a = do(X_i = j)$ can be high even if $\P{X_i = j}$ is large.
-Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$.
-Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot get a good estimate for $\P{Y|do(X_2=1)}$.
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to 
+write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
+estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
+from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
+takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
+simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
+high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
+Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
+get a good estimate for $\P{Y|do(X_2=1)}$.
 
-To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally allocate samples to actions.
-To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known.
-These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, or similarly from observational data subject to identifiability constraints.
+To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
+allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
+are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
+or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
+start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
 Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
@@
 
-%we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$.
-$\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$.
-We then
-
-\begin{wrapfigure}[12]{r}{0.5\textwidth}
+\begin{wrapfigure}[11]{r}{0.5\textwidth}
 \vspace{-25pt}
@@
 This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
+\ifsup
+The proof is in Section \ref{sec:m-equivelence}.
+\else
+The proof is in the supplementary materials.
+\fi
 

### 2016-11-18 causal-bandit/nips_2016/causal_bandits_poster2.tex
initial poster draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Jacobs Landscape Poster
% LaTeX Template
% Version 1.0 (29/03/13)
%
% Created by:
% Computational Physics and Biophysics Group, Jacobs University
% https://teamwork.jacobs-university.de:8443/confluence/display/CoPandBiG/LaTeX+Poster
% 
% Further modified by:
% Nathaniel Johnston (nathaniel@njohnston.ca)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------
\documentclass[12pt]{beamer}

\usepackage[scale=1.24]{beamerposter} % Use the beamerposter package for laying out the poster
\usepackage{graphicx} % more modern 
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{times}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage[noend]{algorithmic}
%\algrenewcommand\algorithmicindent{2.0em}%
\algsetup{indent=4em}
\usepackage{algorithm}

\usetheme{confposter} % Use the confposter theme supplied with this template

\setbeamercolor{block title}{fg=ngreen,bg=white} % Colors of the block titles
\setbeamercolor{block body}{fg=black,bg=white} % Colors of the body of blocks
\setbeamercolor{block alerted title}{fg=white,bg=dblue!70} % Colors of the highlighted block titles
\setbeamercolor{block alerted body}{fg=black,bg=dblue!10} % Colors of the body of highlighted blocks
% Many more colors are available for use in beamerthemeconfposter.sty

%-----------------------------------------------------------
% Define the column widths and overall poster size
% To set effective sepwid, onecolwid and twocolwid values, first choose how many columns you want and how much separation you want between columns
% In this template, the separation width chosen is 0.024 of the paper width and a 4-column layout
% onecolwid should therefore be (1-(# of columns+1)*sepwid)/# of columns e.g. (1-(4+1)*0.024)/4 = 0.22
% Set twocolwid to be (2*onecolwid)+sepwid = 0.464
% Set threecolwid to be (3*onecolwid)+2*sepwid = 0.708

\newlength{\sepwid}
\newlength{\onecolwid}
\newlength{\twocolwid}
\newlength{\threecolwid}
\setlength{\paperwidth}{48in} % A0 width: 46.8in
\setlength{\paperheight}{36in} % A0 height: 33.1in
\setlength{\sepwid}{0.024\paperwidth} % Separation width (white space) between columns
\setlength{\onecolwid}{0.22\paperwidth} % Width of one column
\setlength{\twocolwid}{0.464\paperwidth} % Width of two columns
\setlength{\threecolwid}{0.708\paperwidth} % Width of three columns
\setlength{\topmargin}{-0.5in} % Reduce the top margin size
%-----------------------------------------------------------

\usepackage{graphicx}  % Required for including images

\usepackage{booktabs} % Top and bottom rules for tables

\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[#1\right]}
\newcommand{\R}{\mathbb R}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
%\newcommand{\ind}[1]{\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
\newcommand{\KL}{\operatorname{KL}}
\newcommand{\simpleregret}{R_T}
\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
\newcommand{\EE}{\mathbb E}
\newcommand{\EEa}{\EE_a}
\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\newcommand{\actions}{\mathcal{A}}
\newcommand{\calA}{\mathcal A}
\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}

\let\temp\epsilon
\let\epsilon\varepsilon
\let\vec\mathbf

\graphicspath{ {figures/} }
\usetikzlibrary{arrows,positioning,backgrounds} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.0em,
           minimum height=2.0em,
           font=\tiny,
           text centered,
           scale=.8,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

%----------------------------------------------------------------------------------------
%	TITLE SECTION 
%----------------------------------------------------------------------------------------

\title{Causal Bandits: Learning Good Interventions via Causal Inference} % Poster title

\author{Finnian Lattimore*, Tor Lattimore** and Mark Reid*} % Author(s)

\institute{*Australian National University and Data61, **University of Indiana Bloomington} % Institution(s)

%----------------------------------------------------------------------------------------

\begin{document}

\addtobeamertemplate{block end}{}{\vspace*{2ex}} % White space under blocks
\addtobeamertemplate{block alerted end}{}{\vspace*{2ex}} % White space under highlighted (alert) blocks

\setlength{\belowcaptionskip}{2ex} % White space under figures
\setlength\belowdisplayshortskip{2ex} % White space under equations

\begin{frame}[t] % The whole poster is enclosed in one beamer frame

\begin{columns}[t] % The whole poster consists of three major columns, the second of which is split into two columns twice - the [t] option aligns each column's content to the top

\begin{column}{\sepwid}\end{column} % Empty spacer column

\begin{column}{\onecolwid} % The first column

%----------------------------------------------------------------------------------------
%	OBJECTIVES
%----------------------------------------------------------------------------------------

%\begin{alertblock}{Objectives}
%
%Lorem ipsum dolor sit amet, consectetur, nunc tellus pulvinar tortor, commodo eleifend risus arcu sed odio:
%\begin{itemize}
%\item Mollis dignissim, magna augue tincidunt dolor, interdum vestibulum urna
%\item Sed aliquet luctus lectus, eget aliquet leo ullamcorper consequat. Vivamus eros sem, iaculis ut euismod non, sollicitudin vel orci.
%\item Nascetur ridiculus mus.  
%\item Euismod non erat. Nam ultricies pellentesque nunc, ultrices volutpat nisl ultrices a.
%\end{itemize}
%
%\end{alertblock}

%----------------------------------------------------------------------------------------
%	INTRODUCTION
%----------------------------------------------------------------------------------------

\begin{block}{Introduction}

We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
than algorithms that do not use the additional causal information.

\end{block}

%------------------------------------------------
\begin{block}{Setup}
\begin{itemize}
\item Learner given causal graph $\mathcal{G}$ over a set of random variables $\mathcal{X}$
\item Actions assign values to subsets of variables in $\mathcal{G}$. The set of allowed actions/arms $\mathcal{A}$ is a subset of all possible assignments.
\item One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
\item The game proceeds over $T$ rounds. In round $t$, the learner takes action $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$, \textbf{\emph{then}} observes reward $Y_t$ and the value of all other variables in $\mathcal{X}$.
\item Simple regret: goal is to minimize $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}$, where $\mu^*$ is the expected reward of the best action and $\hat a^*_T$ is the action the learner estimates to be optimal at time $T$
\end{itemize}

\begin{figure}
    \begin{subfigure}[b]{0.45\textwidth}
	\centering    
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.0em,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
%every node/.style={scale=0.6}
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){$Y$};
 \path[every node/.style={font=\tiny}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5);
\end{tikzpicture}
        \caption{Parallel graph}
        \label{fig:parallel}
    \end{subfigure}
    \begin{subfigure}[b]{0.5\textwidth}
    \centering    
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.0em,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
%every node/.style={scale=0.6}
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){$Y$};
\node[main node, above right=of 2](6){$Z$};
 \path[every node/.style={font=\tiny}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5)
    (6) edge (1) edge (2) edge (4);
\end{tikzpicture}
        \caption{Confounded parallel graph}
        \label{fig:parallel_confounded} 
    \end{subfigure} 
\caption{Causal graphs}
\label{fig:causal graphs} 
\end{figure}

\end{block}

\begin{block}{Parallel Model}

\begin{itemize}
\item Causal graph as in figure \ref{fig:parallel} with all variables binary.
\item $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
\item For this causal model, $P(Y|do(X_i = j)) = P(Y|X_i = j)$
\item $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$
\item $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$ is set of 'unbalanced' variables using a threshold $1/\tau$
\item The difficulty of the problem is captured by:
\eq{
\label{eq:m-simple}
m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
}
\end{itemize}

\end{block}
\end{column} % End of the first column

%----------------------------------------------------------------------------------------
\begin{column}{\onecolwid} % The 2nd
\begin{block}{Parallel Algorithm}
\setbeamercolor*{block body}{bg=white!95!blue,fg=black}
\begin{algorithm}[H]
\caption{ Parallel Bandit Algorithm \textcolor{white}{\Large A}}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} Total rounds $T$ and $N$.
\FOR{$t \in 1,\ldots,T / 2$}
\STATE Perform empty intervention $do()$
\STATE Observe $\vec{X}_t$ and $Y_t$
\ENDFOR
\FOR{$a = do(X_i = x) \in \actions$}
\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
\ENDFOR
\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
\FOR{$a = do(X_i = x) \in A$}
\FOR{$t \in 1,\ldots,T_A$}
\STATE Intervene with $a$ and observe $Y_t$
\ENDFOR
\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
\ENDFOR
\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
\end{algorithmic}
\end{algorithm}

\eq{
&\textbf{upper-bound, } \simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.\\
&\textbf{lower-bound, }\simpleregret \in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.\\
&\textbf{standard lower-bound, }\simpleregret \in \Omega\left(\sqrt{\frac{N}{T}}\right)\,.
}

\end{block}

\begin{block}{General Graphs}
\begin{itemize}
\item Known but arbitrary causal structure.
\item It's no longer optimal to split between $do()$ and explicitly playing actions. 
\item We assume the conditional interventional distributions over the parents of $Y$, $\P{\parents{Y}|a}$, are known.
\item $\eta$ is a distribution on available interventions that allocates samples to actions.
\item We use a (truncated) importance weighted estimator to transfer information across different actions. 
\item The (untruncated) weights for each action, $R_a(X,\eta) = \frac{\Pn{a}{\parents{Y}(X)}}{ \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}}$, depend on how different the distribution of the parents of $Y$ is under $a$ to the distribution given actions sampled from $\eta$.
\item We optimise $\eta$ to minimize the simple regret.
\eq {
\eta^*  = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[R_a(X,\eta)\right]}_{m(\eta)}\,.
} 
\item The difficulty of the problem is captured by $m(\eta*)$

\end{itemize}
\end{block}
\end{column}

\begin{column}{\onecolwid} % The 2nd
\begin{block}{General Algorithm}
\begin{algorithm}[H]
\caption{General Algorithm  \textcolor{white}{\Large A}}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t$ from $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\FOR{$a \in \calA$}
\STATE
\eq {
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t,\eta)  \ind{R_a(X_t,\eta) \leq B_a}
}
\ENDFOR
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

\eq{
\textbf{upper-bound, } \simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2NT\right)}}\,.
}

\end{block}

\begin{block}{Experiments}
\begin{figure}
    \begin{subfigure}[t]{0.3\textwidth}
		\centering    
    		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
        \label{fig:simple_vs_m}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
        \includegraphics[width=\textwidth]{experiment2_20161020_1249.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
        \label{fig:simple_vs_T_vary_epsilon}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
    		\includegraphics[width=\textwidth]{experiment3_20161020_1252.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
    		\label{fig:simple_vs_T}
    \end{subfigure}
    \caption{Experimental results on the parallel graph in figure \ref{fig:parallel}}
    \label{fig:experiments}
\end{figure}

\begin{figure}[H]
    \begin{subfigure}[t]{0.3\textwidth}
		\centering    
    		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
        \label{fig:simple_vs_m_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
        \label{fig:simple_vs_T_general}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
    		\includegraphics[width=\textwidth]{experiment5_20161023_2118.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m(\eta*)=4.3$ with no actions setting $Z$}
    		\label{fig:simple_vs_T_misspecified}
    \end{subfigure}
    \caption{Experimental results on the confounded graph in figure \ref{fig:parallel_confounded}}
    \label{fig:experiments_confounded}
\end{figure}

\end{block}
\end{column}

\begin{column}{\onecolwid} % The third column

%----------------------------------------------------------------------------------------
%	CONCLUSION
%----------------------------------------------------------------------------------------
\begin{block}{Conclusion}
\begin{itemize}
\item For parallel bandits we have an optimal algorithm (upto log factors) and simple measure of the difficulty of the problem $m(\vec{q})$.
\item For general graphs, $m(\eta) < N$ measures the difficulty of the problem. Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
\end{itemize}
\end{block}

\begin{block}{Discussion \& Future work}

\begin{itemize}
\item The parallel bandit problem can be viewed as an instance of a time varying graph feedback problem, however the algorithms developed for this problem are not optimal for the parallel bandit.
\item Our algorithms only use the reward signal to estimate $\hat{\mu}_a$, not to inform which arms to sample. Making better use of the reward signal could substantially improve practical performance (but not minimax regret). 
\item Cumulative Regret

\item Generalize to consider causal bandit problems where the causal graph is partially or completely unknown.

\end{itemize}

\end{block}

%%----------------------------------------------------------------------------------------
%%	ACKNOWLEDGEMENTS
%%----------------------------------------------------------------------------------------
%
%\setbeamercolor{block title}{fg=red,bg=white} % Change the block title color
%
%\begin{block}{Acknowledgements}
%
%\small{\rmfamily{Nam mollis tristique neque eu luctus. Suspendisse rutrum congue nisi sed convallis. Aenean id neque dolor. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.}} \\
%
%\end{block}

%----------------------------------------------------------------------------------------
%	CONTACT INFORMATION
%----------------------------------------------------------------------------------------

\setbeamercolor{block alerted title}{fg=black,bg=norange} % Change the alert block title colors
\setbeamercolor{block alerted body}{fg=black,bg=white} % Change the alert block body colors

%\begin{alertblock}{Contact Information}
%
%\begin{itemize}
%\item Web: \href{http://www.university.edu/smithlab}{http://www.university.edu/smithlab}
%\item Email: \href{mailto:john@smith.com}{john@smith.com}
%\item Phone: +1 (000) 111 1111
%\end{itemize}
%
%\end{alertblock}

%----------------------------------------------------------------------------------------

\end{column} % End of the third column

\end{columns} % End of all the columns in the poster

\end{frame} % End of the enclosing frame

\end{document}

### 2016-11-22 causal-bandit/nips_2016/causal_bandits_poster2.tex
updated poster - now with less text

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Jacobs Landscape Poster
-% LaTeX Template
-% Version 1.0 (29/03/13)
-%
-% Created by:
-% Computational Physics and Biophysics Group, Jacobs University
-% https://teamwork.jacobs-university.de:8443/confluence/display/CoPandBiG/LaTeX+Poster
-% 
-% Further modified by:
-% Nathaniel Johnston (nathaniel@njohnston.ca)
-%
-% This template has been downloaded from:
-% http://www.LaTeXTemplates.com
-%
-% License:
-% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
+
 %
@@
 
-\usepackage[scale=1.24]{beamerposter} % Use the beamerposter package for laying out the poster
+\usepackage[scale=1.3]{beamerposter} % Use the beamerposter package for laying out the poster
 \usepackage{graphicx} % more modern 
@@
 \usepackage[noend]{algorithmic}
-%\algrenewcommand\algorithmicindent{2.0em}%
-\algsetup{indent=4em}
+\algsetup{indent=6em}
 \usepackage{algorithm}
@@
 \usetheme{confposter} % Use the confposter theme supplied with this template
+
+\setbeamerfont{itemize/enumerate subbody}{size=\normalsize} %to set the body size
+\setbeamertemplate{itemize subitem}{\normalsize\raise1.25pt\hbox{\donotcoloroutermaths$\blacktriangleright$}}  %to set the symbol size
+
+\setbeamertemplate{itemize item}{\normalsize\raise1.25pt\hbox{\donotcoloroutermaths$\blacktriangleright$}}  %to set the symbol size
 
@@
 \setlength{\sepwid}{0.024\paperwidth} % Separation width (white space) between columns
-\setlength{\onecolwid}{0.22\paperwidth} % Width of one column
+\setlength{\onecolwid}{0.3\paperwidth} % Width of one column
 \setlength{\twocolwid}{0.464\paperwidth} % Width of two columns
@@
 
-\institute{*Australian National University and Data61, **University of Indiana Bloomington} % Institution(s)
+\institute{*Australian National University and Data61, ** Indiana University Bloomington} % Institution(s)
 
@@
 
-\setlength{\belowcaptionskip}{2ex} % White space under figures
-\setlength\belowdisplayshortskip{2ex} % White space under equations
+\setlength{\belowcaptionskip}{12pt} % White space under figures
+%\setlength\belowdisplayshortskip{2ex} % White space under equations
+
+%\setlength\abovedisplayskip{40pt}
+%\setlength\belowdisplayskip{40pt}
+%\setlength\abovedisplayshortskip{40pt}
+%\setlength\belowdisplayshortskip{40pt}
 
@@
 
-\begin{column}{\onecolwid} % The first column
-
-%----------------------------------------------------------------------------------------
-%	OBJECTIVES
-%----------------------------------------------------------------------------------------
-
-%\begin{alertblock}{Objectives}
-%
-%Lorem ipsum dolor sit amet, consectetur, nunc tellus pulvinar tortor, commodo eleifend risus arcu sed odio:
-%\begin{itemize}
-%\item Mollis dignissim, magna augue tincidunt dolor, interdum vestibulum urna
-%\item Sed aliquet luctus lectus, eget aliquet leo ullamcorper consequat. Vivamus eros sem, iaculis ut euismod non, sollicitudin vel orci.
-%\item Nascetur ridiculus mus.  
-%\item Euismod non erat. Nam ultricies pellentesque nunc, ultrices volutpat nisl ultrices a.
-%\end{itemize}
-%
-%\end{alertblock}
-
-%----------------------------------------------------------------------------------------
-%	INTRODUCTION
-%----------------------------------------------------------------------------------------
+\begin{column}{.8\onecolwid} % The first column
 
@@
 Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
-We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
-than algorithms that do not use the additional causal information.
-
 \end{block}
@@
 \begin{itemize}
+\setlength\itemsep{28pt}
 \item Learner given causal graph $\mathcal{G}$ over a set of random variables $\mathcal{X}$
@@
 \end{itemize}
+
+\end{block}
+
+\begin{block}{Parallel Model}
 
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.5\textwidth}
+   \begin{subfigure}[b]{0.45\textwidth}
 	\centering    
@@
  %nodes
-\node[main node](1){$X_{1}$};
+\node[main node](1){$x$};
 \node[main node, right=of 1](2){$X_{2}$};
@@
 \node[main node, below right=of 2](5){$Y$};
-\node[main node, above right=of 2](6){$Z$};
  \path[every node/.style={font=\tiny}]
@@
     	(2) edge (5)
-    (4) edge (5)
-    (6) edge (1) edge (2) edge (4);
+    (4) edge (5);
 \end{tikzpicture}
-        \caption{Confounded parallel graph}
-        \label{fig:parallel_confounded} 
+        \caption{after $do(X_1=x)$}
+        \label{fig:parallel_post_action}
     \end{subfigure}
-\caption{Causal graphs}
-\label{fig:causal graphs} 
+\caption{Intervention in the Parallel Model}
+\label{fig:parallel_model} 
 \end{figure}
-
-\end{block}
-
-\begin{block}{Parallel Model}
-
+%\vspace{-1cm}
 \begin{itemize}
+\setlength\itemsep{28pt}
 \item Causal graph as in figure \ref{fig:parallel} with all variables binary.
@@
 \item The difficulty of the problem is captured by:
+\vspace*{.5cm}
 \eq{
@@
 %----------------------------------------------------------------------------------------
-\begin{column}{\onecolwid} % The 2nd
+\begin{column}{1.1\onecolwid} % The 2nd
 \begin{block}{Parallel Algorithm}
@@
 }
-
+\vspace*{-1cm}
 \end{block}
@@
 \begin{block}{General Graphs}
+\begin{figure}
+    \begin{subfigure}[b]{0.48\textwidth}
+    \centering    
+          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.0em,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+%every node/.style={scale=0.6}
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){$Y$};
+\node[main node, above right=of 2](6){$Z$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5)
+    (6) edge (1) edge (2) edge (4);
+\end{tikzpicture}
+        \caption{Confounded parallel graph}
+        \label{fig:parallel_confounded} 
+    \end{subfigure} 
+    \begin{subfigure}[b]{0.48\textwidth}
+    \centering    
+          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.0em,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+%every node/.style={scale=0.6}
+ %nodes
+\node[main node](1){$x$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){$Y$};
+\node[main node, above right=of 2](6){$Z$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5)
+    (6) edge (2) edge (4);
+\end{tikzpicture}
+        \caption{after $do(X_1=x)$}
+        \label{fig:parallel_confounded_post_action} 
+    \end{subfigure} 
+\caption{Intervention in confounded parallel graph}
+\label{fig:causal graphs} 
+\end{figure}
 \begin{itemize}
-\item Known but arbitrary causal structure.
+\item Known but arbitrary causal graph.
 \item It's no longer optimal to split between $do()$ and explicitly playing actions. 
 \item We assume the conditional interventional distributions over the parents of $Y$, $\P{\parents{Y}|a}$, are known.
-\item $\eta$ is a distribution on available interventions that allocates samples to actions.
+\item $\eta$ is a distribution over interventions from which we sample actions.
 \item We use a (truncated) importance weighted estimator to transfer information across different actions.
-\item The (untruncated) weights for each action, $R_a(X,\eta) = \frac{\Pn{a}{\parents{Y}(X)}}{ \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}}$, depend on how different the distribution of the parents of $Y$ is under $a$ to the distribution given actions sampled from $\eta$.
-\item We optimise $\eta$ to minimize the simple regret.
 \eq {
-\eta^*  = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[R_a(X,\eta)\right]}_{m(\eta)}\,.
+R_a(X,\eta) = \frac{\Pn{a}{\parents{Y}(X)}}{ \sum_{b \in \calA} \eta_b \P{\parents{Y}|b}} \; \leftarrow \text{untruncated weights}
 } 
-\item The difficulty of the problem is captured by $m(\eta*)$
-
 \end{itemize}
@@
 
-\begin{column}{\onecolwid} % The 2nd
+\begin{column}{1.1\onecolwid} % The 2nd
 \begin{block}{General Algorithm}
@@
 
+\begin{itemize}
+\item $\eta$ optimised to minimize the simple regret.
+\eq {
+\eta^*  = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[R_a(X,\eta)\right]}_{m(\eta)}\,.
+} 
+\item The difficulty of the problem is captured by $m(\eta*)$.
+\begin{itemize}
+\item $m(\eta)$ is convex in $\eta$.
+\item $m(\eta*) \leq  |\calA|$.
+\item In the parallel setting $m(\eta*) \leq 2 m(\boldsymbol{q})$.
+\end{itemize}
+
+\end{itemize}
+
 \end{block}
-
 \begin{block}{Experiments}
@@
     		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
-    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
+    		\caption{Simple regret vs $m(\boldsymbol{q})$. $T=400$, $N = 50$}
         \label{fig:simple_vs_m}
@@
     		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
-    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
+    		\caption{Simple regret vs $m(\eta*)$. $T=400$, $N = 50$}
         \label{fig:simple_vs_m_general}
@@
 \end{block}
+
 \end{column}
 
-\begin{column}{\onecolwid} % The third column
-
-%----------------------------------------------------------------------------------------
-%	CONCLUSION
-%----------------------------------------------------------------------------------------
-\begin{block}{Conclusion}
-\begin{itemize}
-\item For parallel bandits we have an optimal algorithm (upto log factors) and simple measure of the difficulty of the problem $m(\vec{q})$.
-\item For general graphs, $m(\eta) < N$ measures the difficulty of the problem.
+%\begin{column}{\onecolwid} % The 4th column
+%
+%%----------------------------------------------------------------------------------------
+%%	CONCLUSION
+%%----------------------------------------------------------------------------------------
+%\begin{block}{Conclusion}
+%\begin{itemize}
+%\item For parallel bandits we have an optimal algorithm (upto log factors) and simple measure of the difficulty of the problem $m(\vec{q})$.
+%\item For general graphs, $m(\eta) < N$ measures the difficulty of the problem.
 Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
-\end{itemize}
-\end{block}
-
-\begin{block}{Discussion \& Future work}
-
-\begin{itemize}
-\item The parallel bandit problem can be viewed as an instance of a time varying graph feedback problem, however the algorithms developed for this problem are not optimal for the parallel bandit.
-\item Our algorithms only use the reward signal to estimate $\hat{\mu}_a$, not to inform which arms to sample.
+%\end{itemize}
+%\end{block}
+%
+%\begin{block}{Discussion \& Future work}
+%
+%\begin{itemize}
+%\item The parallel bandit problem can be viewed as an instance of a time varying graph feedback problem,
+%\begin{itemize}
+%\item this does not apply to general causal graphs
+%\item algorithms for feedback graphs are not optimal for the parallel bandit.
+%\end{itemize}
+%
+%\item Our algorithms only use the reward signal to estimate $\hat{\mu}_a$, not to inform which arms to sample.
 Making better use of the reward signal could substantially improve practical performance (but not minimax regret).
-\item Cumulative Regret
-
-\item Generalize to consider causal bandit problems where the causal graph is partially or completely unknown.
-
-\end{itemize}
-
-\end{block}
-
-%%----------------------------------------------------------------------------------------
-%%	ACKNOWLEDGEMENTS
-%%----------------------------------------------------------------------------------------
+%\item Cumulative Regret
 %
-%\setbeamercolor{block title}{fg=red,bg=white} % Change the block title color
+%\item Generalize to consider causal bandit problems where the causal graph is partially or completely unknown. 
 %
-%\begin{block}{Acknowledgements}
+%\end{itemize}
 %
-%\small{\rmfamily{Nam mollis tristique neque eu luctus.
+%
+%
+%
+%
+% 
+%
+%\end{block}
+%
+%\begin{block}
+%
+%%\includegraphics{indiana_university_logo} 
+%%\includegraphics[scale=1]{data61_logo.png} 
+%%\includegraphics[scale=1]{ANU_logo.jpg} 
+%
+%
+%
+%
+%\end{block}
+%
+%%%----------------------------------------------------------------------------------------
+%%%	ACKNOWLEDGEMENTS
+%%%----------------------------------------------------------------------------------------
+%%
+%%\setbeamercolor{block title}{fg=red,bg=white} % Change the block title color
+%%
+%%\begin{block}{Acknowledgements}
+%%
+%%\small{\rmfamily{Nam mollis tristique neque eu luctus.
 Suspendisse rutrum congue nisi sed convallis.
@@
 Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.}} \\
+%%
+%%\end{block}
 %
-%\end{block}
-
-%----------------------------------------------------------------------------------------
-%	CONTACT INFORMATION
-%----------------------------------------------------------------------------------------
-
-\setbeamercolor{block alerted title}{fg=black,bg=norange} % Change the alert block title colors
-\setbeamercolor{block alerted body}{fg=black,bg=white} % Change the alert block body colors
-
-%\begin{alertblock}{Contact Information}
+%%----------------------------------------------------------------------------------------
+%%	CONTACT INFORMATION
+%%----------------------------------------------------------------------------------------
 %
-%\begin{itemize}
-%\item Web: \href{http://www.university.edu/smithlab}{http://www.university.edu/smithlab}
-%\item Email: \href{mailto:john@smith.com}{john@smith.com}
-%\item Phone: +1 (000) 111 1111
-%\end{itemize}
+%\setbeamercolor{block alerted title}{fg=black,bg=norange} % Change the alert block title colors
+%\setbeamercolor{block alerted body}{fg=black,bg=white} % Change the alert block body colors
 %
-%\end{alertblock}
-
-%----------------------------------------------------------------------------------------
-
-\end{column} % End of the third column
+%%\begin{alertblock}{Contact Information}
+%%
+%%\begin{itemize}
+%%\item Web: \href{http://www.university.edu/smithlab}{http://www.university.edu/smithlab}
+%%\item Email: \href{mailto:john@smith.com}{john@smith.com}
+%%\item Phone: +1 (000) 111 1111
+%%\end{itemize}
+%%
+%%\end{alertblock}
+%
+%
+%
+%%----------------------------------------------------------------------------------------
+%
+%\end{column} % End of the 4th column
 

### 2017-04-13 causal-bandit/arXiv/causal-bandits.tex
long overdue update

\documentclass{article}
\pdfoutput=1
\usepackage[final]{nips_anon}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

% additional packages
\usepackage{graphicx} % more modern 
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{times}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{wrapfig}
%

\newif\ifsup\suptrue

\graphicspath{ {figures/} }
\usetikzlibrary{arrows,positioning,backgrounds} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\tiny,
           text centered,
           scale=.8,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[#1\right]}
\newcommand{\R}{\mathbb R}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
%\newcommand{\ind}[1]{\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
\newcommand{\KL}{\operatorname{KL}}
\newcommand{\simpleregret}{R_T}
\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
\newcommand{\EE}{\mathbb E}
\newcommand{\EEa}{\EE_a}
\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\newcommand{\actions}{\mathcal{A}}
\newcommand{\calA}{\mathcal A}
\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\let\temp\epsilon
\let\epsilon\varepsilon

\title{Causal Bandits: Learning Good Interventions via Causal Inference}

\author{
  Finnian Lattimore \\
  Australian National University and Data61/NICTA \\
  \texttt{finn.lattimore@gmail.com} \\
   \And
   Tor Lattimore \\
   University of Alberta \\
   \texttt{tor.lattimore@gmail.com} \\
   \And
   Mark D. Reid \\
   Australian National University and Data61/NICTA \\
   \texttt{mark.reid@anu.edu.au} \\
}

\begin{document}

\maketitle

\begin{abstract} 
We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
than algorithms that do not use the additional causal information.
\end{abstract} 
%\begin{tikzpicture}[background rectangle/.style={fill=olive!45}, show background rectangle]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:intro}
Medical drug testing, policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1959}. 
In this framework, single actions (also referred to as interventions) from a pre-determined set are repeatedly performed in 
order to evaluate their effectiveness via feedback from a single, real-valued reward signal.
We propose a generalisation of the standard model by assuming that, in addition to the reward signal, the learner observes the values of a number of covariates 
drawn from a probabilistic causal model~\citep{Pearl2000}.
Causal models are commonly used in disciplines where explicit experimentation may be difficult such as social science, demography and economics.
For example, when predicting the effect of changes to childcare subsidies on workforce participation, or school choice on grades. 
Results from causal inference relate observational distributions to interventional ones, allowing the outcome of an intervention to be predicted without
explicitly performing it.
By exploiting the causal information we show, theoretically and empirically, how non-interventional observations can be used to improve the rate at 
which high-reward actions can be identified.

The type of problem we are concerned with is best illustrated with an example. 
Consider a farmer wishing to optimise the yield of her crop. 
She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
In each season the farmer has enough time and money to intervene and control at most one of these variables:
deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
%without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
\paragraph{Contributions} We take the first step towards formalising and solving problems such as the one above. 
In \S\ref{sec:defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph. 
We show that our causal bandit framework subsumes the classical bandits (no additional observations) and contextual stochastic bandit problems (observations are revealed before an intervention is chosen) before focusing on the case where, like the above example, observations occur \emph{after} each intervention is made.

Our focus is on the simple regret, which measures the difference between the return of the optimal action and that of the action chosen by the algorithm after $T$ rounds.
In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently and there are $2N$ possible interventions.
We propose a simple causal best arm identification algorithm for this problem and show that up to logarithmic factors it enjoys minimax optimal
simple regret guarantees of $\smash{\tilde\Theta(\sqrt{m/T})}$ where $m$ depends on the causal model and may be much smaller than $N$.
In contrast, existing best arm identification algorithms suffer $\smash{\Omega(\sqrt{N/T})}$ simple regret (Thm. 4 by \citet{audibert2010best}).
This shows theoretically the value of our framework over the traditional bandit problem. 
Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in this framework.

In the general casual bandit problem interventions and observations may have a complex relationship. 
In \S\ref{sec:simple-regret-general} we propose a new algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent 
to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
As in the parallel bandit case, the regret guarantee scales like $\smash{O(\sqrt{m/T})}$ where $m$ depends on the underlying causal structure, with 
smaller values corresponding to structures that are easier to learn. The value of $m$ is always less than the number of interventions $N$ and in the special
case of the parallel bandit (where we have lower bounds) the notions are equivalent.

\paragraph{Related Work} As alluded to above, causal bandit problems can be treated as classical multi-armed bandit problems by simply ignoring the causal model and extra observations and applying an existing best-arm identification algorithm with well understood simple regret guarantees \citep{Jamieson2013}. However, as we show in \S\ref{sec:simple-regret}, ignoring the extra information available in the non-intervened variables yields sub-optimal performance.

A well-studied class of bandit problems with side information are ``contextual bandits''~\cite{Langford2008,Agarwal2014}. Our framework bears a superficial similarity to contextual bandit problems since the extra observations on non-intervened variables might be viewed as context for selecting an intervention. 
However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.

There have been several proposals for bandit problems where extra feedback is received after an action is taken.
Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph. As we discuss in \S\ref{sec:discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting. They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}. The partial monitoring approach taken by \cite{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online. \citet{Yu2009} utilize extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection. \citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled. The main difference from our present work is our focus on simple regret and the more complex information linking rewards for different arms via causal graphs. To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.

%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret. They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward. It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}. Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.

Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been 
intervened on can provide important contextual information. Their work differs in many ways. For example, the focus is on the cumulative regret and
the context is observed before the action is taken and cannot be controlled by the learning agent.

%\todof{write one sentence to explain how their work is different to ours}

 \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions. Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model. Combining their handling of  causal induction with our analysis is left as future work.
% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action. Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context. (Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results. It is easy to come up with a counter example in which it does not).

The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{BJQ13}, 
where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discuss some of the issues 
encountered in sequential design, but do not give an algorithm or theoretical results for this case.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PROBLEM SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Setup}
\label{sec:defs}
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
We will assume each variable only takes on a finite number of distinct values.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from $\parents{i}$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed actions} $\mathcal{A}$.
One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$. 
The causal bandit game proceeds over $T$ rounds.
In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations. 
It then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, 
including the \emph{reward} $Y_t \in \{0,1\}$. 
After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T \in \actions$ based on its prior observations.

The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.

Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are possible. If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}. Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of allowable interventions thereby reducing the complexity.

We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$. The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$. Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward. Intuitively though, one would expect to perform better by making use of the extra structure and observations.

\iffalse
\subsection{Exploiting the Causal Model}

From a causal perspective the key insight is that the assumed causal structure implies that
the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
on $X_{t,i} = j$. Formally,
\eqn {
\label{eq:observe}
\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
}
This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (Figure \ref{fig:causalStructure}). It is not the case in general. 
For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$),that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the interventional distribution of $Y_t$ from the observational one.

This enables us to estimate the rewards for multiple interventions simultaneously by observing the values taken by the variables when we select the $do()$ action. Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and need to be explored separately, by explicitly making the intervention.

To exploit the causal structure we must optimally balance the collection of observational data against making interventions to learn about low probability events. The problem is more challenging because the probabilities $P(X_i=j)$ are unknown and must also be estimated.

\begin{definition}
Let $V_{ij} = \frac{1}{P(X_i = j)}$, for $i \in \set{1...N}$ and $j \in {0,1}$. 
Define 
\eqn{
\label{def:m}
m(\boldsymbol{V}) = \min \set{m: \sum_{ij} \ind{V_{ij} \geq m} = m}\,.
}
\end{definition}

\fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% UPPER BOUND ON SIMPLE REGRET
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Regret Bounds for Parallel Bandit}
\label{sec:simple-regret}
In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special 
case of the causal bandit problem which we call the {\it parallel bandit}.
It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro}, including the farming example. 
It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.

The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a 
reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:parallel}.
All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high). 
The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.

\begin{figure}
    \begin{subfigure}[b]{0.34\textwidth}
	\centering    
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
%every node/.style={scale=0.6}
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){$Y$};
 \path[every node/.style={font=\tiny}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5);
\end{tikzpicture}
        \caption{Parallel graph}
        \label{fig:parallel}
    \end{subfigure}
    \begin{subfigure}[b]{0.2\textwidth}
    \centering
        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
\node[main node](1){$X_1$};
\node[main node, below left=of 1](2){$X_2$};
\node[main node, below right=of 1](4){$Y$};
 \path[every node/.style={font=\tiny}]
    (1) edge (2)
    (1) edge (4)
    (2) edge (4);
\end{tikzpicture}
        \caption{Confounded graph}
        \label{fig:causalStructure_confounded}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
    \centering
         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, right=of 4](5){$Y$};
 \path[every node/.style={font=\tiny}]
    (1) edge (2)
  	(2) edge (3)
    (3) edge (4)
    (4) edge (5);
\end{tikzpicture}
        \caption{Chain graph}
        \label{fig:causalchain}
    \end{subfigure}
    \caption{Causal Models}\label{fig:causalmodels}
\end{figure}

In each round the learner either purely observes by selecting $do()$ or sets the value of a single variable. The remaining variables are simultaneously set by independently biased coin flips. 
The value of all variables are then used to determine the distribution of rewards for that round.
Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
%\todom{Should $Y$ be 0-1 valued or real everywhere?} YES!
The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
$r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
In the farming example, this choice of $Y$ models the success or failure of a seasons crop, which depends stochastically on the various environment variables.
%Note that the expected reward of the optimal intervention is at least as large as the expected reward for doing nothing.

%\todom{$i,j$ notation to $a \in \actions$}
%In our analysis below, we will often use the pair $(i,j)$ for $i \in \{1, \ldots, N\}$ and $j \in \{0,1\}$ as a shorthand for the intervention $do(X_i = j)$.
%We will also make use of the expansion of the expected reward for $do(X_i = j)$:
%\eq{
%\mu_{i,j} 
%&= \E{r(X)|do(X_i = j)} \\
%&= \sum_{\boldsymbol{x} \in \set{0,1}^N : x_i = j} r(\boldsymbol{x})  \prod_{k \neq i} q_k^{x_k} (1 - q_k)^{1-x_k}\,.  
%}
%The sum and product use the restrictions $x_i = j$ and $k \ne i$ because the intervention forces $X_i = j$ with probability 1.
%The optimal intervention is denoted $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and the corresponding optimal reward is $\mu^* = \mu_{i^*,j^*}$.

\paragraph{The Parallel Bandit Algorithm}
%\label{sub:par-bandit-alg}

The algorithm operates as follows. For the first $T/2$ rounds it chooses $do()$ to collect observational data. As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$. Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large. The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates of their returns could be poor. To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions. The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$). For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$. Define
\eq{
\label{eq:m-simple}
m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
% m(\vec{q}) = \max \set{i : \sum_{k=1}^N \ind{\min\set{q_k, \,1 - q_k} \leq 1/i} \leq i}\,.
}
%Let $I_1 = \emptyset$ and for $\tau \in [0,1)$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} \leq \tau}$.

\begin{wrapfigure}[18]{r}{0.6\textwidth}
\vspace{-25pt}
\begin{minipage}{.6\textwidth}
\begin{algorithm}[H]
\caption{Parallel Bandit Algorithm}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} Total rounds $T$ and $N$.
\FOR{$t \in 1,\ldots,T / 2$}
\STATE Perform empty intervention $do()$
\STATE Observe $\vec{X}_t$ and $Y_t$
\ENDFOR
\FOR{$a = do(X_i = x) \in \actions$}
\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
\ENDFOR
\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
\FOR{$a = do(X_i = x) \in A$}
\FOR{$t \in 1,\ldots,T_A$}
\STATE Intervene with $a$ and observe $Y_t$
\ENDFOR
\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
\ENDFOR
\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{wrapfigure}

$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret. When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$. We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$. Although very simple, the following two theorems show that this algorithm is effectively optimal.

\begin{theorem}\label{thm:uq-simple}
Algorithm \ref{alg:simple} satisfies
\eq{
\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m}\right)}}\,.
}
\end{theorem}

\begin{theorem}\label{thm:lower}
For all $T$, $\vec{q}$ and all strategies, there exists a reward function such that
\eq{
\simpleregret 
\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.
}
\end{theorem}

\ifsup
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
\else
The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
\fi
%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions. Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$. Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting. In practice, you would combine the data from both phases to estimate rewards for the low probability actions. We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% UPPER BOUND ON SIMPLE REGRET GENERAL CASE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Regret Bounds for General Graphs}
\label{sec:simple-regret-general}
We now consider the more general problem where the graph structure is known, but arbitrary. For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
} 
where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.

We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to 
write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
get a good estimate for $\P{Y|do(X_2=1)}$.

To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.

\begin{wrapfigure}[10]{r}{0.5\textwidth}
\vspace{-30pt}
\begin{minipage}{.5\textwidth}
\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t$ from $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\FOR{$a \in \calA$}
\STATE
\eq {
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
}
\ENDFOR
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{wrapfigure}

Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator. Let $\parents{Y}(X)$ denote the realization of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$

\eq {
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}\,, 
}

where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently. The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.

The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,,\text{ where } \EEa \text{ is the expectation with respect to } \Pn{a}.
}

We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.

\begin{theorem}\label{thm:general}
If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$

\eq{
\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
}
\end{theorem}
\ifsup 
The proof is in Section \ref{sec:thm:general}.
\else
The proof is in the supplementary materials.
\fi Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$. Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$. Thus it has broader applicability to the parallel graph than the generic algorithm given here. We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
for future work.

\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$. In light of Theorem \ref{thm:general}
a natural choice of $\eta$ is the minimiser of $m(\eta)$.
\eq{
\eta^* 
= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
}
Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
Therefore the minimisation problem may be tackled using standard techniques from convex optimisation. An interpretation of $m(\eta^*)$ is the minimum achievable worst-case variance of the importance weighted estimator. In the experimental section we present some special cases, but for now we give two simple results. The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.

\begin{proposition}\label{pro:m-bound}
$m(\eta^*) \leq |\calA|$. \textit{Proof.} 
\textup{By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|\,\forall a$.}
\eq{
m(\eta) 
= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
\leq \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
= \max_a \EEa\left[\frac{1}{\eta_a}\right] = |\calA| %\qedhere
}
\end{proposition}

The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$. This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}. \ifsup 
The proof is in Section \ref{sec:m-equivelence}.
\else
The proof is in the supplementary materials.
\fi

\begin{remark}\label{rem:truncate}
The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
often possible to choose $B_a$ significantly
larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
to learn from observational data.
\end{remark}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% EXPERIMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiments}
\label{sec:experiments}

We compare Algorithms \ref{alg:simple} and \ref{alg:general} with Successive Elimination on the parallel bandit problem
under a variety of conditions, including where the importance weighted estimator used by Algorithm \ref{alg:general} is not truncated,
which is justified in this setting by Remark \ref{rem:truncate}. Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms). $Y_t \sim \bernoulli(\frac{1}{2}+\epsilon)$ if $X_1=1$ and $Y_t \sim \bernoulli(\frac{1}{2}-\epsilon')$ otherwise, where $\epsilon' = q_1\epsilon/(1-q_1)$. This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions. We set $q_i = 0$ for $i \leq m$ and $\frac{1}{2}$ otherwise. Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.

%\eq{
%Y_t \sim \begin{cases}
%\bernoulli(\frac{1}{2}+\epsilon) & \text{if } X_1 = 1 \\
%\bernoulli(\frac{1}{2}-\frac{q_1}{1-q_1}\epsilon) & \text{otherwise}\,.
%\end{cases}
%}

We compare the performance of the Algorithm 1, which is specific to the parallel problem, but does not require knowledge of $\boldsymbol{q}$, with that of Algorithm 2 and the Successive Reject algorithm of \cite{audibert2010best}. For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.

In figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases. The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure. For the causal algorithms it increases approximately with $\sqrt{m}$. As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.

\begin{figure}
    \begin{subfigure}[t]{0.3\textwidth}
		\centering    
    		\includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
        \label{fig:simple_vs_m}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
        \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
        \label{fig:simple_vs_T_vary_epsilon}
    \end{subfigure}\hfill
    \begin{subfigure}[t]{0.3\textwidth}
    		\centering
    		\includegraphics[width=\textwidth]{experiment3_N50_s10000_20160519_1030.pdf}
    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
    		\label{fig:simple_vs_T}
    \end{subfigure}
    \caption{Experimental results}
    \label{fig:experiments}
\end{figure}

%\begin{figure}
%  \centering
%  \begin{minipage}[b]{0.3\textwidth}
%    \includegraphics[width=\textwidth]{exp_regret_vs_m_N50_T400_s10000_20160518_1534.pdf}
%    \caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
%    \label{fig:simple_vs_m}
%  \end{minipage}
%  \hfill
%  \begin{minipage}[b]{0.3\textwidth}
%    \includegraphics[width=\textwidth]{experiment2_N50_a4_s10000_20160518_1534.pdf}
%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
%    \label{fig:simple_vs_T_vary_epsilon}
%  \end{minipage}
%  \hfill
%  \begin{minipage}[b]{0.3\textwidth}
%    \includegraphics[width=\textwidth]{experiment3_N50_s10000_20160518_1534.pdf}
%    \caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
%    \label{fig:simple_vs_T}
%  \end{minipage}
%\end{figure}

Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned. This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$. In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially. The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data. 
This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be minimax optimal asymptotically \citep{LMS14}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DISCUSSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion \& Future Work}
\label{sec:discussion}
Algorithm~\ref{alg:general} for general causal bandit problems 
estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ where 
the parameter $m(\eta)$ measures the difficulty of learning the causal model and is always less than $N$.
The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.

The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator $\hat{\mu}_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $\Pn{a}{\parents{Y}}$ for all $a \in \calA$.
In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.
Despite these extra estimates, Theorem~\ref{thm:lower} shows that this approach is optimal (up to log factors).
Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.

\vspace{-0.4cm} 
\paragraph{Making Better Use of the Reward Signal}
Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
For the parallel bandit the required modifications should be quite straightforward. The idea would be to adapt the algorithm to essentially use successive elimination in
the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability. In the general case a similar modification is also possible
by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping. Note
that these modifications will not improve the minimax regret, which at least for the parallel bandit is already optimal. For this reason we prefer to emphasize 
the main point that causal structure should be exploited when available. Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which
in some cases may be preferred to a sequential design for logistical reasons. This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.

\vspace{-0.4cm} 
\paragraph{Cumulative Regret}
Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information 
to get near-optimal cumulative regret guarantees. They consider a finite-armed bandit model with side information where in reach round
the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
In this way the learner can gain information about actions it does not take with varying levels of accuracy. The reduction follows by substituting
the importance weighted estimators in place of the Gaussian reward. In the case that $\vec{q}$ is known this would lead to a known variance
and the only (insignificant) difference is the Bernoulli noise model. In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
at least asymptotically.

%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
%\todom{Check this!}
%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.

The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen. The feedback graph is distinct from the causal graph. A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$. For this parallel bandit problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions. However, \citet{Alon2015,Kocak2014} give adversarial algorithms, which when applied to the parallel bandit problem obtain the standard bandit regret. A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action. This is equivalent to a nominally stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.

% \cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action. In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case. \cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.

\vspace{-0.4cm} 
\paragraph{Causal Models with Non-Observable Variables}
If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
non-observable variables. Some of the interventional distributions may be non-identifiable meaning we can not obtain prior estimates for $\Pn{a}{\parents{Y}}$ from 
even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
\ref{alg:general} cannot be used. Estimating these quantities while simultaneously minimising the simple regret is an interesting and challenging open problem.

% For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
% In this case, some conditional distributions may be non-identifiable. 
% The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
% We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
% Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.

% More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}. In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate. The quality of the estimate notably depends not only on the number of times that action was selected. For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$. This more complex side information does not fit within the graph feedback framework.

\vspace{-0.4cm} 
\paragraph{Partially or Completely Unknown Causal Graph}
A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
The latter case arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques. 
Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014b} could potentially be incorporated into a causal bandit algorithm.
In particular, \citet{Hu2014b} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational dataset would be much more challenging.

% (Partially known structure)
% Key results $bigo(n)$ singleton or $bigo(log log n)$ multi-variate experiments are required.
% - Focus on minimizing the number of experiments that must be performed. Examples of active versus online learning. The cost of experiments constant or at least known.

% (Unknown Structure)
% If we need to learn the structure, in an online environment. 
% Experiment is much, much more revealing than inference from observational data. We would expect it to dominate.

%\pagebreak
{\small\bibliography{libraryicml}}
\bibliographystyle{apalike}

\ifsup

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PROOF OF SIMPLE REGRET UPPER BOUND
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}

Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$. The assumption is non-restrictive since all variables
are independent and permutations of the variables can be pushed to the reward function.
The proof of Theorem \ref{thm:uq-simple} requires some lemmas. \

\begin{lemma}\label{lem:conc1}
Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
\eq{
\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
}
\end{lemma}

\begin{proof}
By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \cite{hagerup1990guided}),

\eq{
\P{\left|\hat q_i - q_i\right| \geq \epsilon} \leq 2e^{-\frac{T\epsilon^2}{6q_i}}
}

Letting $\delta = 2e^{-\frac{T\epsilon^2}{6q_i}}$ and solving for $\epsilon$ completes the proof.

\end{proof}

\begin{lemma}\label{lem:conc2}
Let $X_1,X_2\ldots,$ be a sequence of random variables with $X_i \in [0,1]$ and $\EE[X_i] = p$ and $\delta \in [0,1]$.
Then 
\eq{
\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \leq 4\delta\,.
}
\end{lemma}

\begin{proof}
For $\delta \geq 1/4$ the result is trivial. Otherwise 
by Hoeffding's bound and the union bound:
\eq{
\P{\exists t \geq n_0 : \left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} 
&\leq \sum_{t = n_0}^\infty \P{\left|\frac{1}{t} \sum_{s=1}^t X_s - p\right| \geq \sqrt{\frac{2}{n_0} \log \frac{2}{\delta}}} \\
&\leq 2\sum_{t=n_0}^\infty \exp\left(-\frac{t}{n_0} \log \frac{2}{\delta}\right) 
\leq 4\delta\,. \qedhere
}
\end{proof}

\begin{lemma}\label{lem:m_est}
Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
\eq{
\P{2m(\vec{q}) / 3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})} \geq 1 - \delta\,.
}
\end{lemma}

\begin{proof}
Let $F$ be the event that there exists and $1 \leq i \leq N$ for which
\eq{
\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}\,.
}
Then by the union bound and Lemma \ref{lem:conc1} we have $\P{F} \leq \delta$. The result will be completed by showing that
when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m$ that $q_i \geq q_m \geq 1/m$ and so by Lemma \ref{lem:conc1} we have
\eq{
\frac{3}{4} 
&\geq \frac{1}{2} + \sqrt{\frac{3}{T} \log \frac{2N}{\delta}} 
\geq q_i + \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}} 
\geq \hat q_i \\
&\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
\geq q_i - \sqrt{\frac{q_i}{8m}}
\geq \frac{1}{2m}\,.
}
Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m$.
For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m$ that
\eq{
\hat q_i 
\leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
\leq \frac{1}{m} \left(1 + \sqrt{\frac{1}{8}}\right)
\leq \frac{3}{2m}\,.
}
Therefore $m(\vec{\hat q}) \geq 2m(\vec{q}) / 3$ as required. 
\end{proof}

\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
Let $\delta = m = m(\vec{q}) / N$. Then by Lemma \ref{lem:m_est} we have 
\eq{
\P{2m/3 \leq m(\vec{\hat q}) \leq 2m} \geq 1 - \delta\,.
}
Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then
for $a \in A$ the algorithm estimates $\mu_a$ from $T/(2m(\vec{\hat q})) \geq T/(4m)$ samples.
Therefore by Hoeffding's inequality and the union bound we have
\eq{
\P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{8m}{T} \log\frac{2N}{\delta}}} \leq \delta\,.
}
For arms not in $a$ we have $\hat p_a \geq 1/m(\vec{\hat q}) \geq 1/(2m)$.
Therefore if $a = do(X_i = j)$, then 
\eq{
\hat p_a = \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{1}{2m}\,. 
}
Therefore $\sum_{t=1}^{T/2} \ind{X_{t,i} = j} \geq T/4m$
and by Lemma \ref{lem:conc2} we have
\eq{
\P{\sum_{t=1}^{T/2} \ind{X_i = j} \geq \frac{T}{4m} \text{ and } \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{8m}{T} \log \frac{2N}{\delta}}} \leq 4\delta / N\,.
}
Therefore with probability at least $1 - 6\delta$ we have
\eq{
(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{8m}{T} \log \frac{N}{\delta}} = \epsilon\,.
}
If this occurs, then 
\eq{
\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
}
Therefore
\eq{
\mu^* - \EE[\mu_{\hat a^*_T}] 
\leq 6\delta + \epsilon 
\leq \frac{6m}{T} + \sqrt{\frac{32m}{T} \log \frac{NT}{m}}\,, 
}
which completes the result.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LOWER BOUND
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}

We follow a relatively standard path by choosing multiple environments that have different optimal arms, but which cannot all be statistically
separated in $T$ rounds.
Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
For each $i$ define reward function $r_i$ by
\eq{
r_0(\boldsymbol{X}) &= \frac{1}{2} &
r_i(\boldsymbol{X}) &= \begin{cases}
\frac{1}{2} + \epsilon & \text{if } X_i = 1 \\
\frac{1}{2} & \text{otherwise}\,,
\end{cases}
}
where $1/4 \geq \epsilon > 0$ is some constant to be chosen later.
We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
environment determined by $\boldsymbol{q}$ and $r_i$. Let $\operatorname{P}_i$ be the corresponding measure
on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \citet{Tsy08} we have
\eq{
\Prz{\hat a^*_T = a^*} + \Pri{\hat a^*_T \neq a^*} \geq \exp\left(-\KL(\operatorname{P}_0, \operatorname{P}_i)\right)\,,
}
where $\KL(\Ps_0, \Ps_i)$ is the KL divergence between measures $\operatorname{P}_0$ and $\operatorname{P}_i$.
Let $T_i(T) = \sum_{t=1}^T \ind{a_t = do(X_i = 1)}$ be the total number of times the learner intervenes on variable $i$ by setting it to $1$.
Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $\Ps_0$ and $\Ps_i$ may be bounded using the telescoping property (chain rule) and
by bounding the local KL divergence by the $\chi$-squared distance as by \citet{Auer1995}. This leads to 
\eq{
\KL(\Ps_0, \Ps_i) 
&\leq 6\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] 
\leq 6\epsilon^2 \left(\EE_0 T_i(T) + q_i T\right) 
\leq 6\epsilon^2 \left(\EE_0 T_i(T) + \frac{T}{m}\right)\,.
}
Define set $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
Then for $i \in A$ and choosing $\epsilon = \min\set{1/4, \sqrt{m/(18T)}}$ we have
\eq{
\KL(\Ps_0, \Ps_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
}
Now $\sum_{i=1}^m \EE_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
Therefore
\eq{
\sum_{i \in A} \Pri{\hat a^*_T \neq a} 
\geq \sum_{i \in A} \exp\left(-\KL(\Ps_0, \Ps_i)\right) - 1
\geq \frac{|A|}{e} - 1 
\geq \frac{m}{2e} - 1\,.
}
Therefore there exists an $i \in A$ such that
$\Pri{\hat a^*_T \neq a^*} \geq \frac{\frac{m}{2e} - 1}{m}$. 
Therefore if $\epsilon < 1/4$ we have
\eq{
R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{m}{18T}}\,.
}
Otherwise $m \geq 18T$ so $\sqrt{m/T} = \Omega(1)$ and
\eq{
R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{1}{4} \frac{\frac{m}{2e} - 1}{2m} \in \Omega(1) 
}
as required.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% GENERAL-GRAPH UPPER BOUND
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of Theorem \ref{thm:general}}\label{sec:thm:general}

\begin{proof}
First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
We define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$, $R_{at} = R_a(X_t)$ and $\Pn{a}{.} = \Pns{a}{.}$.
By definition we have $|Z_{at}| \leq B_a$ and 
\eq{
\Var_Q[Z_{at}] 
\leq \EE_Q[Z_{at}^2] 
\leq \EE_Q[R_{at}^2]
= \EEa[R_{at}]
= \EEa\left[\frac{\Pns{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
\leq m(\eta)\,.
}
Checking the expectation we have
\eq{
\EE_Q[Z_{at}] 
= \EEa \left[Y \ind{R_{at} \leq B_a}\right] 
= \EEa Y - \EEa \left[Y\ind{R_{at} > B_a}\right] 
= \mu_a - \beta_a\,,
}
where 
\eq{
0 \leq \beta_a = \EEa[Y \ind{R_{at} > B_a}] \leq \Pns{a}{R_{at} > B_a}
}
is the negative bias. 
The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
\eq{
\beta_a \leq \Pns{a}{R_{at} > B_a} \leq \frac{\EEa[R_{at}]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
}
Let $\epsilon_a > 0$ be given by
\eq{
\epsilon_a = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
}
Then by the union bound and Bernstein's inequality 
\eq{
\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} 
\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
}

Let $I = \hat{a}^*_T$ be the action selected by the algorithm, $a^* = \argmax_{a \in \calA} \mu_a$ be the true optimal action and recall that $\EE_Q[Z_{at}] = \mu_a - \beta_a$. Assuming the above event does not occur we have,

\eq{
\mu_I \geq \hat \mu_I - \epsilon_I  
\geq \hat \mu_{a^*} - \epsilon_I  
\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
}
By the definition of the truncation
we have
\eq{
\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
}
and
\eq{
\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
}
Therefore for $C = \sqrt{2} + 4$ we have
\eq{
\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
}
Therefore
\eq{
\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
}
as required.
\end{proof}

\subsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}\label{sec:m-equivelence}

\begin{proposition} In the parallel bandit setting,
$m(\eta^*) \leq 2m(\boldsymbol{q})$.
\end{proposition}

\begin{proof}

Recall that in the parallel bandit setting,

\eq{
\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}
}

Let:

\eq {
\eta_a = \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}}\frac{1}{2m(\boldsymbol{q})} \text { for } a \in do(X_i = j)
}

Let $D =\sum_{a\in do(X_i=j)}\eta_a$. From the definition of $m(\boldsymbol{q})$, 
\eq {
\sum_{a\in do(X_i=j)} \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}} \leq m(\boldsymbol{q}) \implies D \leq \frac{1}{2}
}

Let $\eta_a = \frac{1}{2} + (1-D)$ for $a = do()$ such that $\sum_{a \in \calA}\eta_a = 1$

Recall that,

\eq{
m(\eta) &
= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]
}

We now show that our choice of $\eta$ ensures $\EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \leq 2m(\boldsymbol{q})$ for all actions $a$.

For the actions $a: \eta_a > 0$, ie $do()$ and $do(X_i = j):\P{X_i=j}<\frac{1}{m(\boldsymbol{q})}$,
\eq{
\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq \EEa\left[\frac{\Pn{a}{X_1...X_N}}{\eta_a\Pn{a}{X_1...X_N}}\right] = \EEa\left[\frac{1}{\eta_a}\right] \leq 2m(\boldsymbol{q})
}

For the actions $a :\eta_a = 0$, ie $do(X_i=j):\P{X_i=j}\geq\frac{1}{m(\boldsymbol{q})}$,
\eq{
\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq & \EEa\left[\frac{\ind{X_i=j}\prod_{k\neq i}\P{X_k}}{(1/2+D)\prod_k \P{X_k}}\right] \\=& \EEa\left[\frac{\ind{X_i=j}}{(1/2+D)\P{X_i = j}}\right]
\leq  \EEa\left[\frac{\ind{X_i=j}}{(1/2)(1/m(\boldsymbol{q}))}\right] \leq 2m(\boldsymbol{q})
}

Therefore $m(\eta*) \leq m(\eta) \leq 2m(\boldsymbol{q})$ as required.

\end{proof}
\
%in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
%Recall in that setting that 
%\eq{
%\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
%}
%For $a = do()$ choose $\eta_a = 1/2$. 
%Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
%Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
%It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.

\fi

\end{document}

### 2017-04-13 causal-bandit/nips_2016/causal_bandits_poster3.tex
long overdue update

\documentclass[nofooter,landscape,scale=1.4]{poster}
\usepackage[absolute,overlay]{textpos}

\usepackage{ragged2e}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{nicefrac}

\newcommand{\ubar}[1]{\underline{#1\mkern-3mu}\mkern3mu }

\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}

\title{Causal Bandits: Learning Good Interventions via Causal Inference}
\author{\mbox{}\\[1.3cm] Finnian Lattimore \hspace{2cm} Tor Lattimore \hspace{2cm} Mark Reid}
\footer{Contact}
\date{}

\begin{document}
\begin{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{textblock}{25.33}(2,12)
%\setbeamercolor*{block body}{bg=,fg=black}
\begin{block}{Introduction}
\justifying
We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
than algorithms that do not use the additional causal information.

\eq{
&\textbf{upper-bound, } R \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.\\
&\textbf{lower-bound, }R \in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.\\
&\textbf{standard lower-bound, }\simpleregret \in \Omega\left(\sqrt{\frac{N}{T}}\right)\,.
}
\end{block}

\begin{block}{Model}
\begin{itemize}
\item $K$ jobs 
\item Unknown parameters $\nu_1,\ldots,\nu_K$ with $\nu_k \geq 0$
\item At the start of each time step $t$ the learner chooses resource allocation $M_{k,t}$
\item Resources are limited: $\sum_{k=1}^K M_{k,t} \leq 1$
\item Job $k$ succeeds with probability $\min\set{1, M_{k,t} / \nu_k}$
\end{itemize}
\vspace{0.5cm}
\begin{tikzpicture}[thick]
  \begin{axis}[thick,xlabel={$M_{k,t}$},ylabel={Success probability},width=20cm,height=10cm,scaled ticks=false,xmin=0,xmax=2,ymin=0,xtick={0,0.5,1,1.5,2},legend pos=south east] 
    \addplot+[dashed,mark=none,domain=0:2,samples=100,ultra thick] {and(\x<0.5,\x>=0)*2*\x + and(\x>=0.5,\x<=2)*1};
    \addlegendentry{$\nu_1 = 2$};
    \addplot+[mark=none,domain=0:2,samples=100,ultra thick] {and(\x<1,\x>=0)*\x + and(\x>=1,\x<=2)*1};
    \addlegendentry{$\nu_2 = 1$};
    \addplot+[dashed,mark=none,domain=0:2,samples=100,ultra thick] {and(\x<1.5,\x>=0)*\x*2/3 + and(\x>=1.5,\x<=2)*1};
    \addlegendentry{$\nu_3 = \nicefrac{2}{3}$};
  \end{axis}
\end{tikzpicture}
\begin{itemize}
\item All jobs are restarted after every time step
\item Goal is to maximise the expected number of successful jobs up to known time step $n$
\eq{
\sum_{t=1}^n \sum_{k=1}^K \min\set{1, {M_{k,t} \over \nu_k}}
}
\item $M^*_k$ is the unknown optimal allocation (computed by allocating resources to easiest jobs first)
\end{itemize}
\end{block}

\begin{block}{Regret}
\justifying
The expected regret $R_n$ of an algorithm measures the difference between the optimal allocation and the allocation of the algorithm
\eq{
\E\left[\sum_{t=1}^n \sum_{k=1}^K \left( \min\set{1, {M^*_k \over \nu_k}} - \min\set{1, {M_{k,t} \over \nu_k}}\right)\right]
}
An algorithm is learning if the expected regret is sub-linear: $\lim_{n\to\infty} R_n / n = 0$ \\[1cm]

Minimising the regret in this setting is especially challenging since
over-allocation resources to a job provides only limited information, while under-allocating risks severe sub-optimally.
\end{block}

\end{textblock}

\begin{textblock}{25.33}(29.33,12)
\begin{block}{Estimation}
\begin{itemize}
\item Algorithm works by estimating $\nu_k$ 
\item $X_{k,s} = \ind{\text{job } k \text{ completed in time step } s}$
\item If $M_{k,s} \leq \nu_k$, then ${1 \over \nu_k} = \E[X_{k,s} / M_{k,s}]$
\item Small/large $M_{k,s}$ implies large/small variance
\item Suggests a weighted estimator
\eq{
w_{k,t} &\leftarrow {1 \over 1 - {M_{k,t} \over \bar \nu_{k,t-1}}} 
&{1 \over \hat\nu_{k,t}} &\leftarrow {\sum_{s=1}^{t} w_{k,s} X_{k,s} \over \sum_{s=1}^t w_{k,s} M_{k,s}} 
}
\item $\bar \nu_{k,t-1}$ is a high-probability upper bound on $\nu_k$ obtained in previous time step
\item Confidence bounds about $\hat \nu_k$ are obtained via a modified version of Bernstein's inequality
\end{itemize}
\end{block}

\vspace{-1cm}
{
%\setbeamercolor*{block body}{bg=white!95!blue,fg=black}
\begin{block}{Optimistic Algorithm}
\begin{algorithmic}[1]
\State {\bf input: } $n, K$, $\set{\ubar \nu_{k,0}}_{k=1}^K$
\State $\delta \leftarrow (nK)^{-2}$ and $\bar \nu_{k,0} = \infty$ for each $k$
\For{$t \in 1, \ldots, n$}
\State /* Optimistically choose $M_{k,t}$ using $\ubar \nu_{k,t-1}$ */
\State $(\forall k \in 1,\ldots,K)$ initialise $M_{k,t} \leftarrow 0$
\For{$i \in 1, \ldots, K$}
\State $k \leftarrow \argmin\limits_{k : M_{k,t} = 0} \ubar \nu_{k,t-1}$
\State $M_{k,t} \leftarrow \min \set{\ubar \nu_{k,t-1},\; 1 - \sum_{j=1}^K M_{j,t}}$
\EndFor
\State $(\forall k \in 1,\ldots,K)$ observe $X_{k,t}$ 
\State $(\forall k \in 1,\ldots,K)$ compute estimates:
\eq{
w_{k,t} &\leftarrow {1 \over 1 - {M_{k,t} \over \bar \nu_{k,t-1}}} 
&{1 \over \hat\nu_{k,t}} &\leftarrow {\sum_{s=1}^{t} w_{k,s} X_{k,s} \over \sum_{s=1}^t w_{k,s} M_{k,s}} 
}
\State $(\forall k \in 1,\ldots,K)$ update confidence intervals:
\eq{
R_{k,t} &\leftarrow \max_{s \leq t} w_{k,s}  \qquad \hat V^2_{k,t} \leftarrow \sum_{s \leq t} {w_{k,s} M_{k,s} \over \ubar \nu_{k,t-1}} \\
\tilde \epsilon_{k,t} &\leftarrow {f(R_{k,t}, \hat V^2_{k,t},\delta) \over \sum_{s=1}^t w_{k,s} M_{k,s}}    \\
{1 \over \ubar \nu_{k,t}} &\leftarrow \min\set{{1 \over \ubar \nu_{k,t-1}}, {1 \over \hat \nu_{k,t}} + \tilde\epsilon_{k,t}} \\
{1 \over \bar \nu_{k,t}} &\leftarrow \max\set{{1 \over \bar \nu_{k,t-1}}, {1 \over \hat \nu_{k,t}} - \tilde\epsilon_{k,t}} 
}
\EndFor
\Statex 
\Function{$f$}{$R$,$V^2$, $\delta$}
\State $\delta_0 \leftarrow {\delta \over 3(R+1)^2 (V^2+1)^2}$
\State \Return 
${R+1 \over 3} \log{2 \over \delta_0}$
\Statex $\qquad +\;\;\sqrt{2(V^2+1) \log{2 \over \delta_0} + \left({R+1 \over 3}\right)^2 \log^2{2 \over \delta_0}}$
\EndFunction
\end{algorithmic}
\end{block}
}

\vspace{-1cm}
\begin{block}{Bounds on the Regret}
\vspace{-2.5cm}
\eq{
&R_n \in O\Bigg(\ell \log^2 n + \sum_{k=\ell+1}^K \left(\log {1 \over \nu_k \Delta_{\ell,k}}\right)\log n  \\
\nonumber &\;\;+\sum_{k=\ell+2}^K \left(\log{1 \over \nu_k \Delta_{\ell+1,k}}\right) \log n
+ \sum_{k=\ell+1}^K {\log n \over \Delta_{\ell+1,k}} \Bigg),
}
\begin{itemize}
\item $\ell$ is the number of jobs that are fully allocated under the optimal allocation and $\Delta_{i,j} = {1 \over \nu_{i}} - {1 \over \nu_j}$
\item If all jobs can be fully allocated, then bound is independent of the gaps
\item $\Omega(\sqrt{K n})$ problem-independent lower bound
\end{itemize}

\end{block}

\end{textblock}

\begin{textblock}{25.33}(56.66,12)
\begin{block}{Experiments}
\begin{itemize}
\item Code and data available in supplementary material
\item Data points based on mean of 300 samples
\end{itemize}

\vspace{0.5cm}
\begin{minipage}{17cm}
\begin{tikzpicture}[baseline,font=\scriptsize]
  \begin{axis}[xlabel shift=-5pt,xmax=10,ylabel shift=-5pt,xlabel={$\nu_2$},ylabel={$R_n$},xmin=2,ymin=110,height=10cm,width=15cm,mark size=2pt,compat=newest,xtick={2,3,4,5,6,7,8,9,10}]
    \addplot+[only marks] table {exp1.txt};
  \end{axis}
\end{tikzpicture}
\end{minipage}
\begin{minipage}{6cm}
$K = 2$ \\
$n = 10^4$ \\ 
$\nu_1 = 2$ \\ 
$\nu_2 \in [2,10]$
\end{minipage}

\hspace{-1cm}
\begin{minipage}{17cm}
\begin{tikzpicture}[baseline,font=\scriptsize]
  \begin{axis}[xlabel shift=-5pt,ylabel shift=-5pt,xlabel={$n$},ylabel={$\displaystyle {R_n \over \log^2 n}$},compat=newest,mark size=2pt,height=10cm,scaled ticks=false,width=15cm,xtick={0,1000000},xticklabels={0,1e6}] 
    \addplot+[only marks] table {exp2.txt};
  \end{axis}
\end{tikzpicture}
\end{minipage}
\hspace{0.3cm}
\begin{minipage}{6cm}
$K = 2$ \\
$\nu_1 = \nicefrac{4}{10}$ \\
$\nu_2 = \nicefrac{6}{10}$
\end{minipage}

\begin{minipage}{17cm}
\begin{tikzpicture}[font=\scriptsize]
  \begin{axis}[xtick={0.35,0.60,1.05},ytick={2000,5000},yticklabels={2e3,5e3},xlabel shift=-5pt,ylabel shift=-5pt,xlabel={$\nu_2$},ylabel={$R_n$},compat=newest,mark size=2pt,height=10cm,width=15cm] 
    \addplot+[only marks] table {exp3.txt};
  \end{axis}
\end{tikzpicture}
\end{minipage}
\begin{minipage}{6cm}
$K = 2$ \\
$n = 10^5$ \\ 
$\nu_1 = \nicefrac{4}{10}$ \\
$\nu_2 \in [\nicefrac{4}{10},1]$
\end{minipage}

\begin{minipage}{17cm}
\begin{tikzpicture}[font=\scriptsize]
  \begin{axis}[ytick={0,10000},yticklabels={0,1e4},xlabel shift=-5pt,ylabel shift=-5pt,scaled ticks=false,xlabel={$n$},ylabel={$R_n$},compat=newest,mark size=2pt,height=10cm,width=15cm,xtick={0,100000},xticklabels={0,1e5}] 
    \addplot+[only marks,blue] table[x index=0,y index=1] {exp4.txt};
    \addplot+[each nth point=3, filter discard warning=false, unbounded coords=discard,only marks,red] table[x index=0,y index=2] {exp4.txt};
  \end{axis}
\end{tikzpicture}
\end{minipage}
\begin{minipage}{6cm}
$K = 2$ \\
$\nu_1 = \nicefrac{4}{10}$ \\
$\nu_2 = \nicefrac{6}{10}$ \\
\tikz[baseline=-0.5ex]{\draw[line width=0.1cm,blue] (0,0) -- (1,0);} {\small weighted} \\
\tikz[baseline=-0.5ex]{\draw[line width=0.1cm,red] (0,0) -- (1,0);} {\small unweighted} \\
\end{minipage}

\end{block}

\begin{block}{Applications}
\begin{itemize}
\item Cache allocation
\item Load balancing
\item Resource allocation is a common problem
\end{itemize}
\end{block}

\begin{block}{Summary}
\begin{itemize}
\item Introduced resource allocation setting
\item New algorithm is both practical and computationally efficient
\item Nearly-optimal adaptive bounds on the regret
\item Weighted estimators independently interesting
\end{itemize}
\end{block}

\begin{block}{Open Questions}
\begin{itemize}
\item Reduce $\log^2 n$ dependence to $\log n$
\item Algorithm is not anytime
\item Problem-dependent lower bounds
\item Non-parametric models (such as concave, piecewise linear, or monotonic)
\item Adversarial models
\item Variable rewards for each job
\item Resources not replenished each time step
\item Many more!
\end{itemize}
\end{block}

\end{textblock}

\end{frame}

\end{document}

### 2017-04-13 causal-bandit/nips_2016/mem-alloc-poster.tex
long overdue update

--- 2017-04-13 causal-bandit/nips_2016/causal_bandits_poster3.tex
@@
-\documentclass[nofooter,landscape,scale=1.4]{poster}
+\documentclass[nofooter,scale=1.4]{poster}
 \usepackage[absolute,overlay]{textpos}
@@
 
-\title{Causal Bandits: Learning Good Interventions via Causal Inference}
-\author{\mbox{}\\[1.3cm] Finnian Lattimore \hspace{2cm} Tor Lattimore \hspace{2cm} Mark Reid}
+\title{Optimal Resource Allocation with Semi-Bandit Feedback}
+\author{\mbox{}\\[1.3cm] Tor Lattimore \hspace{2cm} Koby Crammer \hspace{2cm} Csaba Szepesv\'ari}
 \footer{Contact}
@@
 \begin{frame}
+
+\begin{textblock}{20}(0,5)
+\includegraphics[height=8cm]{UoA-logo}
+\end{textblock}
+
+\begin{textblock}{20}(68,7)
+\includegraphics[height=5cm]{technion-logo}
+\end{textblock}
 
@@
 \justifying
-We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
-Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
-We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
-than algorithms that do not use the additional causal information.
-
-\eq{
-&\textbf{upper-bound, } R \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.\\
-&\textbf{lower-bound, }R \in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)\,.\\
-&\textbf{standard lower-bound, }\simpleregret \in \Omega\left(\sqrt{\frac{N}{T}}\right)\,.
-}
+We study a sequential resource allocation problem involving a fixed number of recurring jobs.
+%
+At each time step the learner should distribute available resources among the jobs in order to maximise the expected number of completed jobs. 
+%
+Allocating more resources to a given job increases the probability
+that it completes, but with a cut-off. %KC diminishing returns. 
+Specifically, we assume a linear
+model where the probability increases linearly until it equals one, after which
+allocating additional resources is wasteful.
+%
+We assume the difficulty of each job is unknown
+and present the first algorithm for this problem and prove upper and lower bounds on its regret.
+Despite its apparent simplicity, the problem has a rich structure: 
+We show that an appropriate optimistic algorithm can improve its learning speed dramatically 
+beyond the results one normally expects for similar problems
+as the problem becomes resource-laden.
 \end{block}
@@
 {
-%\setbeamercolor*{block body}{bg=white!95!blue,fg=black}
+\setbeamercolor*{block body}{bg=white!95!blue,fg=black}
 \begin{block}{Optimistic Algorithm}

### 2017-04-13 causal-bandit/nips_2016/poster/causal_bandits_poster.tex
long overdue update

--- 2015-01-15 wiml_poster/Causal_Inference_Lattimore.tex
@@
 
-\documentclass[b0,landscape,25pt]{sciposter}
+\documentclass[a0,landscape,25pt,plainsections]{sciposter}
 
@@
 % The widths of these boxes can be easily edited to accommodate your content as you see fit
-\begin{minipage}[t]{0.15\linewidth}%
+%\begin{minipage}[t]{0.15\linewidth}%
+%\centering
+%%\includegraphics[height=12cm]{nicta_logo2.jpg} % Logo or a photo of you, adjust its dimensions here
+%\end{minipage}
+\begin{minipage}[b]{0.99\linewidth}
 \centering
-\includegraphics[height=12cm]{nicta_logo2.jpg} % Logo or a photo of you, adjust its dimensions here
-\end{minipage}
-\begin{minipage}[b]{0.65\linewidth}
-\centering
-\resizebox{\linewidth}{!}{\veryHuge \color{NavyBlue} \textbf{Causal Inference in Machine Learning: A review}}
+\resizebox{\linewidth}{!}{\veryHuge \color{NavyBlue} \textbf{Causal Bandits: Learning Good Interventions via Causal Inference}}
  \color{Black}\\
 
-\huge \textbf{Finnian Lattimore}\\ % Author(s)
-\huge Australian National University\\ % University/organization
-finnlattimore@gmail.com
+\huge Finnian Lattimore, Tor Lattimore and Mark Reid\\ % Author(s)
 
 \end{minipage}
-\begin{minipage}[t]{0.15\linewidth}
-\centering
-\includegraphics[height=12cm]{anu-logo-notext}
-\end{minipage}
+%\begin{minipage}[t]{0.15\linewidth}
+%\centering
+%%\includegraphics[height=12cm]{anu-logo-notext}
+%\end{minipage}
 

### 2017-04-13 causal-bandit/nips_2016/poster/main.tex
long overdue update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Jacobs Landscape Poster
% LaTeX Template
% Version 1.0 (29/03/13)
%
% Created by:
% Computational Physics and Biophysics Group, Jacobs University
% https://teamwork.jacobs-university.de:8443/confluence/display/CoPandBiG/LaTeX+Poster
% 
% Further modified by:
% Nathaniel Johnston (nathaniel@njohnston.ca)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[final]{beamer}

\usepackage[scale=1.24]{beamerposter} % Use the beamerposter package for laying out the poster

\usetheme{confposter} % Use the confposter theme supplied with this template

\setbeamercolor{block title}{fg=ngreen,bg=white} % Colors of the block titles
\setbeamercolor{block body}{fg=black,bg=white} % Colors of the body of blocks
\setbeamercolor{block alerted title}{fg=white,bg=dblue!70} % Colors of the highlighted block titles
\setbeamercolor{block alerted body}{fg=black,bg=dblue!10} % Colors of the body of highlighted blocks
% Many more colors are available for use in beamerthemeconfposter.sty

%-----------------------------------------------------------
% Define the column widths and overall poster size
% To set effective sepwid, onecolwid and twocolwid values, first choose how many columns you want and how much separation you want between columns
% In this template, the separation width chosen is 0.024 of the paper width and a 4-column layout
% onecolwid should therefore be (1-(# of columns+1)*sepwid)/# of columns e.g. (1-(4+1)*0.024)/4 = 0.22
% Set twocolwid to be (2*onecolwid)+sepwid = 0.464
% Set threecolwid to be (3*onecolwid)+2*sepwid = 0.708

\newlength{\sepwid}
\newlength{\onecolwid}
\newlength{\twocolwid}
\newlength{\threecolwid}
\setlength{\paperwidth}{48in} % A0 width: 46.8in
\setlength{\paperheight}{36in} % A0 height: 33.1in
\setlength{\sepwid}{0.024\paperwidth} % Separation width (white space) between columns
\setlength{\onecolwid}{0.22\paperwidth} % Width of one column
\setlength{\twocolwid}{0.464\paperwidth} % Width of two columns
\setlength{\threecolwid}{0.708\paperwidth} % Width of three columns
\setlength{\topmargin}{-0.5in} % Reduce the top margin size
%-----------------------------------------------------------

\usepackage{graphicx}  % Required for including images

\usepackage{booktabs} % Top and bottom rules for tables

%----------------------------------------------------------------------------------------
%	TITLE SECTION 
%----------------------------------------------------------------------------------------

\title{Causal Bandits: Learning Good Interventions via Causal Inference} % Poster title

\author{Finnian Lattimore, Tor Lattimore and Mark Reid} % Author(s)

\institute{Department and University Name} % Institution(s)

%----------------------------------------------------------------------------------------

\begin{document}

\addtobeamertemplate{block end}{}{\vspace*{2ex}} % White space under blocks
\addtobeamertemplate{block alerted end}{}{\vspace*{2ex}} % White space under highlighted (alert) blocks

\setlength{\belowcaptionskip}{2ex} % White space under figures
\setlength\belowdisplayshortskip{2ex} % White space under equations

\begin{frame}[t] % The whole poster is enclosed in one beamer frame

\begin{columns}[t] % The whole poster consists of three major columns, the second of which is split into two columns twice - the [t] option aligns each column's content to the top

\begin{column}{\sepwid}\end{column} % Empty spacer column

\begin{column}{\onecolwid} % The first column

%----------------------------------------------------------------------------------------
%	INTRODUCTION
%----------------------------------------------------------------------------------------

\begin{block}{Introduction}

Lorem ipsum dolor \textbf{sit amet}, consectetur adipiscing elit. Sed commodo molestie porta. Sed ultrices scelerisque sapien ac commodo. Donec ut volutpat elit. Sed laoreet accumsan mattis. Integer sapien tellus, auctor ac blandit eget, sollicitudin vitae lorem. Praesent dictum tempor pulvinar. Suspendisse potenti. Sed tincidunt varius ipsum, et porta nulla suscipit et. Etiam congue bibendum felis, ac dictum augue cursus a. \textbf{Donec} magna eros, iaculis sit amet placerat quis, laoreet id est. In ut orci purus, interdum ornare nibh. Pellentesque pulvinar, nibh ac malesuada accumsan, urna nunc convallis tortor, ac vehicula nulla tellus eget nulla. Nullam lectus tortor, \textit{consequat tempor hendrerit} quis, vestibulum in diam. Maecenas sed diam augue.

This statement requires citation \cite{Smith:2012qr}.

\end{block}

%------------------------------------------------

\begin{figure}
\includegraphics[width=0.8\linewidth]{placeholder.jpg}
\caption{Figure caption}
\end{figure}

%----------------------------------------------------------------------------------------

\end{column} % End of the first column

\begin{column}{\sepwid}\end{column} % Empty spacer column

\begin{column}{\twocolwid} % Begin a column which is two columns wide (column 2)

\begin{columns}[t,totalwidth=\twocolwid] % Split up the two columns wide column

\begin{column}{\onecolwid}\vspace{-.6in} % The first column within column 2 (column 2.1)

%----------------------------------------------------------------------------------------
%	MATERIALS
%----------------------------------------------------------------------------------------

\begin{block}{Materials}

The following materials were required to complete the research:

\begin{itemize}
\item Curabitur pellentesque dignissim
\item Eu facilisis est tempus quis
\item Duis porta consequat lorem
\item Eu facilisis est tempus quis
\end{itemize}

The materials were prepared according to the steps outlined below:

\begin{enumerate}
\item Curabitur pellentesque dignissim
\item Eu facilisis est tempus quis
\item Duis porta consequat lorem
\item Curabitur pellentesque dignissim
\end{enumerate}

\end{block}

%----------------------------------------------------------------------------------------

\end{column} % End of column 2.1

\begin{column}{\onecolwid}\vspace{-.6in} % The second column within column 2 (column 2.2)

%----------------------------------------------------------------------------------------
%	METHODS
%----------------------------------------------------------------------------------------

\begin{block}{Methods}

Lorem ipsum dolor \textbf{sit amet}, consectetur adipiscing elit. Sed laoreet accumsan mattis. Integer sapien tellus, auctor ac blandit eget, sollicitudin vitae lorem. Praesent dictum tempor pulvinar. Suspendisse potenti. Sed tincidunt varius ipsum, et porta nulla suscipit et. Etiam congue bibendum felis, ac dictum augue cursus a. \textbf{Donec} magna eros, iaculis sit amet placerat quis, laoreet id est. In ut orci purus, interdum ornare nibh. Pellentesque pulvinar, nibh ac malesuada accumsan, urna nunc convallis tortor, ac vehicula nulla tellus eget nulla. Nullam lectus tortor, \textit{consequat tempor hendrerit} quis, vestibulum in diam. Maecenas sed diam augue.

\end{block}

%----------------------------------------------------------------------------------------

\end{column} % End of column 2.2

\end{columns} % End of the split of column 2 - any content after this will now take up 2 columns width

%----------------------------------------------------------------------------------------
%	IMPORTANT RESULT
%----------------------------------------------------------------------------------------

\begin{alertblock}{Important Result}

Lorem ipsum dolor \textbf{sit amet}, consectetur adipiscing elit. Sed commodo molestie porta. Sed ultrices scelerisque sapien ac commodo. Donec ut volutpat elit.

\end{alertblock}

%----------------------------------------------------------------------------------------

\begin{columns}[t,totalwidth=\twocolwid] % Split up the two columns wide column again

\begin{column}{\onecolwid} % The first column within column 2 (column 2.1)

%----------------------------------------------------------------------------------------
%	MATHEMATICAL SECTION
%----------------------------------------------------------------------------------------

\begin{block}{Mathematical Section}

Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin. Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin.

\begin{equation}
E = mc^{2}
\label{eqn:Einstein}
\end{equation}

Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin. Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin.

\begin{equation}
\cos^3 \theta =\frac{1}{4}\cos\theta+\frac{3}{4}\cos 3\theta
\label{eq:refname}
\end{equation}

Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin. Nam quis odio enim, in molestie libero. Vivamus cursus mi at nulla elementum sollicitudin.

\begin{equation}
\kappa =\frac{\xi}{E_{\mathrm{max}}} %\mathbb{ZNR}
\end{equation}

\end{block}

%----------------------------------------------------------------------------------------

\end{column} % End of column 2.1

\begin{column}{\onecolwid} % The second column within column 2 (column 2.2)

%----------------------------------------------------------------------------------------
%	RESULTS
%----------------------------------------------------------------------------------------

\begin{block}{Results}

\begin{figure}
\includegraphics[width=0.8\linewidth]{placeholder.jpg}
\caption{Figure caption}
\end{figure}

Nunc tempus venenatis facilisis. Curabitur suscipit consequat eros non porttitor. Sed a massa dolor, id ornare enim:

\begin{table}
\vspace{2ex}
\begin{tabular}{l l l}
\toprule
\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
\midrule
Treatment 1 & 0.0003262 & 0.562 \\
Treatment 2 & 0.0015681 & 0.910 \\
Treatment 3 & 0.0009271 & 0.296 \\
\bottomrule
\end{tabular}
\caption{Table caption}
\end{table}

\end{block}

%----------------------------------------------------------------------------------------

\end{column} % End of column 2.2

\end{columns} % End of the split of column 2

\end{column} % End of the second column

\begin{column}{\sepwid}\end{column} % Empty spacer column

\begin{column}{\onecolwid} % The third column

%----------------------------------------------------------------------------------------
%	CONCLUSION
%----------------------------------------------------------------------------------------

\begin{block}{Conclusion}

Nunc tempus venenatis facilisis. \textbf{Curabitur suscipit} consequat eros non porttitor. Sed a massa dolor, id ornare enim. Fusce quis massa dictum tortor \textbf{tincidunt mattis}. Donec quam est, lobortis quis pretium at, laoreet scelerisque lacus. Nam quis odio enim, in molestie libero. Vivamus cursus mi at \textit{nulla elementum sollicitudin}.

\end{block}

%----------------------------------------------------------------------------------------
%	ADDITIONAL INFORMATION
%----------------------------------------------------------------------------------------

\begin{block}{Additional Information}

Maecenas ultricies feugiat velit non mattis. Fusce tempus arcu id ligula varius dictum. 
\begin{itemize}
\item Curabitur pellentesque dignissim
\item Eu facilisis est tempus quis
\item Duis porta consequat lorem
\end{itemize}

\end{block}

%----------------------------------------------------------------------------------------
%	REFERENCES
%----------------------------------------------------------------------------------------

\begin{block}{References}

\nocite{*} % Insert publications even if they are not cited in the poster
\small{\bibliographystyle{unsrt}
\bibliography{sample}\vspace{0.75in}}

\end{block}

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------

\setbeamercolor{block title}{fg=red,bg=white} % Change the block title color

\begin{block}{Acknowledgements}

\small{\rmfamily{Nam mollis tristique neque eu luctus. Suspendisse rutrum congue nisi sed convallis. Aenean id neque dolor. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.}} \\

\end{block}

%----------------------------------------------------------------------------------------
%	CONTACT INFORMATION
%----------------------------------------------------------------------------------------

\setbeamercolor{block alerted title}{fg=black,bg=norange} % Change the alert block title colors
\setbeamercolor{block alerted body}{fg=black,bg=white} % Change the alert block body colors

\begin{alertblock}{Contact Information}

\begin{itemize}
\item Web: \href{http://www.university.edu/smithlab}{http://www.university.edu/smithlab}
\item Email: \href{mailto:john@smith.com}{john@smith.com}
\item Phone: +1 (000) 111 1111
\end{itemize}

\end{alertblock}

\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{logo.png} & \hfill & \includegraphics[width=0.4\linewidth]{logo.png}
\end{tabular}
\end{center}

%----------------------------------------------------------------------------------------

\end{column} % End of the third column

\end{columns} % End of all the columns in the poster

\end{frame} % End of the enclosing frame

\end{document}

### 2017-04-13 interpretable/interpretability_and_ethics.tex
long overdue update

\documentclass{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{tikz} 
\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
%\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[#1\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\N}[2]{\mathcal{N}\left({#1}\;,\;{#2}\right)}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.5em,
           minimum height=2.5em,
           font=\footnotesize,
           text centered,
           scale=1,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.5em,
           minimum height=2.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\title{Causality, Ethics and Identifiability}
\author{Finnian Lattimore}

\begin{document}
\def\ci{\perp\!\!\!\perp}

\maketitle
\section{What, Why, When Where, How? (Finale Doschi-Velez)}
Interpretability - What Why When Where How (Finale Doschi-Velez)
\begin{itemize}
\item What: To give or provide meaning to; explain [to humans] (dictionary.com)
\item Related: transparent, accountable, trustworthy, fair, actionable, reliable
\end{itemize}

Why,When,Where needed?
\begin{itemize}
\item Science (why) - understand 'nature of the thing' 
\item debugging
\item safety - decisions are sound, what if inputs are wrong
\item miss-matched objectives - eg side-effects of medication not included in objective
\item legal/ethics
\item unknown, unknowns - something we can't properly quantify in model plus humans involved in decision making.
\end{itemize}

How (to evaluate)?
\begin{itemize}
\item need scientific way of measuring interpret-ability
\item indirect measurement in terms of goal - eg interpret-ability in context of application success measured in terms of some specific application outcome - like increased use of model, better patient welfare, etc. 
\item measure interpret-ability in terms of proxies like sparsity, 'size of model', 'additive', non-negativity.
\item Test how well explanation worked by testing how well people can predict the output of a model for  a given input (simulation)
\item Test how well explanation worked by testing how well people can answer counter-factual queries of the form:
 - What changes to input, weight, cost would change [prediction for X]    
\end{itemize}

Where are the unknown unknowns?
\begin{itemize}
\item in the structure of the data (eg science)
\item in reward function (eg missmatched objective)
\item inputs (ie safety checks)
\item internal features (ie debugging)
\end{itemize}
What kind of interpret-ability do we want?
\begin{itemize}
\item Global:What factors important for decision making in general
\item Local: What factors lead to a particular decision.
\end{itemize}

\subsection{Feature Importance}
Instance based approach  - compute conditional expected value of classifier output given each pixel individually. (Empirical approach).

Unification of model interpret-ability SHAP (read this paper for sure).

\subsection{Maya Gupta (Google)}
Don't trust your data
cross-validation - uses train set data - the problem is that train data doesn't represent test data.

Semantic regularisation - add human structural/human knowledge. 
Non-iid, early adopters, clicks data, past vs future,

Empirical loss trusts the data, regularizers say "don't trust the data"
Loss function is not right - also its not a convex relaxation of the right thing. Ie end goal is not more clicks, its users happiness, revenue, etc.

What makes a great regularizer? 
Captures prior knowledge about the world
Parameter free (no tuning because cross-validation is hard)
Will it work whatever the test distribution
Does it help trust and interpret trained model.
These things may only apply to semantically meaningful features.

An example: monotonicity. Gupta et al JMLR 2016.
Could you build a normal model and then apply constraints, knowing that the solution should be close to optimal?

Alternate constraints: fairness, egregious examples - must not get these wrong, low-churn - similar to another model. Satisfying real world with dataset constraints.

\section{Last talk in interpret-ability session}
Data does not arrive as a given - majority of time is spent getting/cleaning/labelling data. Theory Data -> Algorithm -> Model. In reality there are loops and iterations.

Structured Labelling [CHI 2014] MSR
ModelTracker [CHI 2015, VAST 2016] tool for comparing/examining what models are doing. What is a severe false positive/negative? You should be using a cost function - to decide which model to use (not just predictive accuracy). You can click through to data to see where model is failing and fix it.

concept evolution occurs - as you see the data you get a greater idea of how to classify it. Solve by structured labelling. ie tagging - create subgroups.

\section{Panel discussion}
- are there commonalities or does context matter for interpret-ability. I think it depends on some underlying motivations. 
- providing an explanation can result in people trusting an algorithm too much. 
- trying to explain learned policy in MDPs. Additional question is if you control the system using this policy, what happens? Addition of dynamics adds additional questions. 
- Learning a classifier when we should be learning a policy leads to problems (oscillations etc) because of feedback loops.
Example: what kind of explanation does the FDA want? Should be provably robust - works in out of sample clinical trials. Should make sense with respect to existing models of disease.

- in healthcare causality is the goal and interpret-ability is a first step. We don't do causality because problem is just too complex - but it at least gives us a set of hypothesis.

Thomas Deitrich - causality allows generalizability beyond statistical.
Do as much as you can in an explicit causal way then add statistical residual.

Do we always need interpret-ability in real applications? People sometimes say yes for critical situations? Or is demonstrated predictive success enough. Maya thinks that if there is a trade off we are doing it wrong.

Is it a trade off or should interpret-able models ultimately be more generalizable?

If ML works - we have to be able to apply assumptions.

To get insurance against something I don't know - the unknown unknowns. Sample. 
Human explanations are reconstructed and do not relate to the decision making process.

want to avoid system errors (as goal of interpretable models)
- if you know what you want (transparency, accountability, other things) then you should target that rather than interpretability.

-violation of reasonable explaination/ EU rules makes you liable for up to 4\% of global turn-over. Yikes!
- importance of understanding worst case performance of machine learning algorithms. 
- if task is transfer, you don't care about interpretability
- structure may be added to make better predictions (does not have to be more interpretable).

\section{Finn Random ideas}
- what about if you want to be able to see what would happen if you changed your loss function without re-learning the full model.

Another way of measuring the success of interpretable models is to see if adding a human in the loop can improve desired metric. Ie if motivation for interpret-ability is to allow humans to debug then making this possible should improve accuracy.

I should really spend a few hours per week writing my thesis!

A starting point is what is causal inference and when do you need it.

There is a hierarchy of ethical/philosophical difficulty relating to disparate impact.
\begin{enumerate}
\item The disparate impact is due to biased data collection (solution fix the data)
\item The disparate impact is due to historical/short term causative factors (eg race $\rightarrow$ disadvantage)
\item The disparate impact is due to fundamental/long term causative factors (eg women $\rightarrow$ pregnancy)
\end{enumerate}

wifi 92gf92eh2d
92gf92eh2d

Are simple models (like logistic regression) really interpretable? Not in the presence of unobserved confounding where we have an objective miss-match.

A thing to do
\begin{enumerate}
\item enumerate desirable properties of 'fair' algorithms
\item quantify trade-offs between these properties
\item connect these properties for underlying reasons for desiring fair algorithms
\end{enumerate}

What determines which attributes we consider protected? eg why don't we protect things like intelligence?
\begin{itemize}
\item Not a choice
\item Generally not 'fundamentally' relevant 
\item Easily measurable/observable
\end{itemize}

Justifications for interpretability
\begin{itemize}
\item  better integration with human decision making
\item improved generalization
\item 'debuging'
\item ability to incorporate additional constraints without explicitly defining loss function
\item better adherence
\end{itemize}

Fundamentally causal relationships - those which we expect to be invariant (the system cannot change to remove the relationship).

The closer a variable is to being a direct cause - the more stable that relationship should  be over time (as other things can vary) - also the better the relationship should generalize across situations (eg different countries).

The idealised scenario, under which the target variable and optimisation metric truly capture the desired outcome and the training data is sampled iid from the population of interest can almost never be expected to hold.

Claim: Many of the scenarios motivating interpretable models really require casual models. This is an important distinction to make. Although historically, causal modelling is generally based on interpretable models such as linear or logistic regression, there is no fundamental requirement that this be the case. There has been substantial work on estimating causal effects with complex non linear models, eg \cite{}.

Some of the major risks highlighted with the use of big data and machine learning algorithms for critical decision making are the result of apply non causal algorithms to fundamentally causal problems.

Causality mentioned as a motivating factor for interpretable models \cite{Lipton2016}

Interpretable model not consistently defined because they are motivated by a number of different considerations \& applications. We should first ask why we want an interpretable model then construct a definition that is compatible with our goals and finally develop a solution.

\cite{Barocas2014} consider including a variable indicating membership of a protected class in a model as formal discrimination on the basis of disparate treatment, even in the absence of a discriminatory effect.

data mining may require us to re-evaluate why and whether we care about not discriminating \cite{Barocas2014}

role of randomisation and experiment

Protected class membership is a proxy for variables which are "genuinely relevant in making rational and well- informed decisions" \cite{Barocas2014}

Other potential issues with use of ML/Big Data
\begin{itemize}
\item Lack of a 2nd chance. If everyone has the same picture of you, through a federation of many data sources then failure in one domain may lead to rejection and failure in many others. (at an individual level)
\item Overly homogeneous decision making (at a group level). If everyone is using the same data and algorithms, then we could reinforce current stereotypes and remove opportunities for gradual change. For example, suppose theoretically 
\end{itemize}

Problems motivating interpretable models
\begin{itemize}
\item Models are deployed where their use alters the environment, invalidating future predictions
\item Trust that a model will perform well with respect to real (unmeasurable/hard to quantify) objectives
\item Adversarial settings. 
\item Participatory - critical decisions otherwise made by a small technical elite. 
\item Understand, validate, edit trust \cite{Caruana2015}
\item "trust, which is fundamental if one plans to take action based on a prediction" \cite{Ribeiro2016}
\item trust so that a user will use the model.
\item real world data different (to that used to train/validate)
\item evaluation function used doesn't reflect true goal.
\end{itemize}

\cite{Ribeiro2016} Problems with ML: propagation of feedback loops, failing to notice data leaks, over-estimating model accuracy.

\cite{Duvenaud2000} define a causal model as one which can predict the outcome of an intervention.

Strong arguments for building an interpretable model, at least in an early phase to diagnose potential problems. 
Humans can evaluate for which features P(Y|X) is unlikely to change. People can (at least sometimes) identify features that will not lead to good generalization.

It is important to distinguish 'trust' required for no other reason than to make the user use the model versus 'trust' associated with properties of the model that make it more accurate (ie via being more robust to changes in environment). In the former case, one valid solution would be to train the users to accept machine decisions or hire less questioning people. However, if the fundamental issue is a miss-match between the real world problem and the quantity being optimised by the machine learning algorithm, this would be disastrous. More subtlety, solutions that lead users to place a greater trust in the model that do not alleviate any fundamental mis-match are similarly bad.

Note: related to. Shown that the pattern of likes on facebook can significantly predict sensitive aspects.

Properties of causal models
\begin{itemize}
\item may be more easily contrasted or combined with human reasoning.
\item come with theory on transferability
\item are explicitly designed to predict the outcome of an intervention 
\item can remove conflict between transparency and continued effectiveness.
In some situations, making a model transparent is directly in conflict with its predictive accuracy. 
\end{itemize}

Causality is frequently mentioned in conjunction with interpretebility \cite{}, however the relationship between them remains unclear.

Despite recent work in the area \cite{}, there is still not a unified motivation for or definition of interpretability.

The extent to which interpretability is plausible or genuinely desirable is still under debate in the ML community \cite{}.

There is growing concern in the wider community over the use of automated or machine decision systems and a strong desire for transparency and intpretability. A particularly pertinant example is the introduction of legislation in the European union that grants a right to XXX. This is due to come into effect in 2018 and has major ramifications for the application of machine learning. There are components of this relating to explaination and discrimination.

Demonstrate the need to increase the clarity around issues such as algorithmic discrimination and interpretability.

\cite{Lipton2016} raise the interesting point that interpretability can only be important if we care about more than simply high predictive accuracy. In other words a desire for interpretability implies a miss-match between the real world goal and the optimisation problem presented to a machine learning algorithm.

If a model is non-causal making it transparent can lead to changes in people's behaviour that reduce the predictive accuracy of the model. This creates a trade-off between transparency and the utility of the model that remains even without imposing any requirements for model simplicity. If the model is causal, then any changes in the choices people make will be reflected in the predictions, removing this conflict. For example, if students know exactly what factors lead to a high score from an automated essay marking system they are strongly motivated to change their writing to reflect this. If these factors are direct causes of the desired outcome (clear and compelling writing) this is a good outcome. However, if they are merely associated with good writing then we will end up with impressive sounding gobbledygook. As a more serious example, consider an algorithm used to help make parole decisions that looks at prisoners participation in various programs to assesses the risk of recidivism. Suppose the algorithm determines that book clubs are associated with lower risk, whilst sex offender programs are associated with higher risk. If the model is made transparent to prisoners, then we expect to see the participation in sex offender programs plummet and a strong demand for book clubs. Again if the effects are causal, perhaps the book club fundamentally changes peoples outlook on life and the sex offender programs were poorly designed, this is a positive outcome. However it might be that participation in sex offender programs actually lowers the risk of re-offending in those who take them but that the group who currently participate are simply higher risk than the general prison population. Similarly, the majority of the apparent benefit of the book club could be a function of who opts to take part.

Note: the causal attribute needs to be sufficiently measurable. Proxies may still be subject to manipulation.

\end{document}

### 2017-04-13 thesis/thesis.tex
long overdue update

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}

\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
\DeclareUnicodeCharacter{00A0}{ }
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}

\author{Finnian Lattimore}
\title{Learning how to act: making good decisions with machine learning}

\begin{document}
\def\ci{\perp\!\!\!\perp} % from Wikipedia
\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]
\maketitle

\section*{Notes on papers}
from association to causation via regression (Freedman)
very clear description of the goal of causal inference, very critical causal modelling due to the impossibility of empirically assessing the assumptions.

\section*{What is causality and why do we care?}

The meaning of causality remains widely debated within science and philosophy ...

For this thesis I define a causal model to be any model that aims to predict the outcome of an action or intervention in some system. From this viewpoint, most real problems require causal inference. What use is a model if you are not going to use it to guide any decisions? How then can machine learning be integrating itself into so many key aspects of our lives, while causal inferene remains a small subfield mostly of interest to economists? The answer is two-fold. For some problems, the effect of a decision is obvious and does not impact the system on which it was based. In these cases, the causal aspect of the problem can be safely ignored. Secondly it is likely that machine learners and data scientits, bouyed by successes in the former problems and the excitement of a world awash with data, computing power and non-linear models are overlooking the fact that their predictions may no longer hold if people act on them. Consider the following problems, do they require causal inference?

notes: We don't always need to predict the outcome of an action. In somes

The goal may be more limitied than predicting the outcome of an intervention. To decide between actions, we only assumptions or estimates as to which is better.

For example, in the speech recognition, the immediate result of the action may be known and deterministic (Siri does something). The full consequences of that action are not modelled, (we assume the desired outcome is that Siri )

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Machine translation 
\item Image classification
\item Forecasting the weather
\item Playing Go 
\item Identiying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pnumonia.
\item Predicting who will re-offend on release from prison 
\item Predicting which customers will cease to be your customers
\item Demand prediction for inventory control
\item Predicting who will click on an ad

\item Financial trading
\item Recommending movies
\item Online search
\item Self driving cars

\end{itemize}

The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.

Consider speech recognition. You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words. Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a dvd on elocution). A similar arguament could be made for many applications of machine translation and image classification. What about forecasting the weather? If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decicion will not effect if it actually rains. However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system. For this we need a (causal) model that allows us to predict the outcome under various different interventions.

Identifing spam and automated essay marking systems are similar. The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label. Spammers and students will modify their writing in order to optimise thier results. These systems can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided. (What would the nature of the features have to be such that change did not occur? - they would have to be causes of the label.

What about predicting the risk of death in patients with pnumonia? Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective this seems like a straighforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or ommited) in the model, the relationship between those features and the outcome may change if we start using the model to decide whome to admit. (xxx et al) found exactly this effect. Their model learnt that (among other things) people suffering asthma were LESS likely to die from pnemonia. They realised this was because doctors were treating such patints very agressively, thus acutally lowering their risk. There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original admition and treatment protocols. However, using it to decide who to admit could kill. The key is understanding exactly what question you are asking. In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc). For all these problems the assumptions required to treat them as pure prediction problems are;
\begin{enumerate}
\item The treatment is assumed to be effective (at least better than nothing)
\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
\end{enumerate}
It is easy to see how similar issues to the pneumonia example could occur for the customer and parole cases. For example, if the customer retention model is not aware of existing customer loyalty schemes or the most risky criminals are currently only given parole in exceptional circumstances (not visable to the model) so rarely offend. If people are aware of the model they may also try to game it (as in the spam/essay example) so as to recive loyalty reward or early release. TODO talk about the interesting long-term vs short term dilema introduced by the parole example. (maybe best to discuss it further on)

Demand prediction seems relatively straightforward. These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However there is a potential data sensoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is another case with some subleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalize (can in some instances reduce the problem to standard ML)

I hope these examples gave you a feel for the richness and subtlties of causal inference. We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.

I do not see the distinction between explaination and (causal) prediction. Explaination is all about the ability to compress and to generalize. The more a model can do this, the more we view as providing an understanding of the why.

Does this definition include mediation?

Sometimes the causal component is obvious.

Relationship to generalization. Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time. The assumption is there are less places for change to come in.

If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.

The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)? This may be a reason for using structure learning type algorithms even if you are simply doing prediction.

\subsubsection*{The role of assumptions in causal inference}

Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.

Does predictive accuracy indicate a good causal model?

Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.

Generalizing the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parashutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalize). (and given sufficient generalizability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)

Is causal inference possible (are the conclusions ever valid)?

Only from randomized experiment ... if people can do it without recource to constant randomized experiment then an algorithm exits.

The challenges and limitations of infering causal effects from observation data are numerous. However, there are problems we need to solve. The questions is are these problems so serious as to warrent discregarding all non-experimental data. I think answer is a clear no. Need ongoing validation, with assumptions. We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).

\subsection*{Mediation}
\subsection*{Relationship to interpretability}

A desire for interpretability indicates that something has been left out of the loss function.

Are causal/interpretable models more reliable or more likely to generalize well?

One form of interpretability gives people insight into what the features are that the model is relying on.

If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.

Specifically, people can
\begin{itemize}
\item rule out many possible features as highly unlikely to be relevent to a problem
\end{itemize}

People have access to a lot of detailed prior knowledge.

\subsection*{Relationship to transfer learning}

find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time). Causal models predict the outcome of actions. We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike. So we have to make representations such that things are stable.

This is tightly related to generalizability. If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results. We don't include , the color of the doctors shirt, the size of the room, ...

For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales. However, this relationship may be very unstable over time because the ad slots are sold at auction. The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword. However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.

\subsection*{The two approaches to causality}

What is the role of randomization? How do bandits algorithms work despite being only partially randomized? 
What else can you do to improve randomized studies (variance reduction, lower regret).

What are the issues with standard randomized experiments?

\section*{Causal Inference}

\subsection*{Models of causality and intervention}
\subsubsection*{DAGs}

Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.

A complaint leveled against this view point of causality is that the 'surgery' is too precise and that, in the real world, any intervention will effect many variables (eg Cartwright 2007). However,

\subsubsection*{Counterfactuals}

\subsubsection*{Structural Equation Models}

A translator from graphical independence to counter factual statements.

\subsubsection*{Identifiability}

The notion of identifiability is binary. How could it be softened. There are two obvious approaches. 
1) Look at the finite time convergence for identifiable queries.
2) For non identifiable queries, look at what bounds can be achieved or what additional assumptions would be required to make them identifiable.

Question - why does everyone use the backdoor criterion. Is it much more frequent in the world. Easier to estimate?

\subsection*{Defifinitions}
\subsubsection*{Interactions}
\subsubsection*{Causal effects}

\subsection*{Estimation of causal effects}

\subsubsection*{Relationship to covariate shift}

\subsection*{Discovering causal structure}

\subsubsection*{Discovery based on conditional independence}

\subsubsection*{Discovery with functional models}

\section*{Multi-armed Bandits}

\section*{Unifying the frameworks}
\label{sec:unificaition}

\section*{Interesting open questions}
Cycles - a huge issue. Not covered by Pearl,Rubin etc.

Places to look, statistical control theory, etc. any interesting papers along these lines?

\section*{Data for testing causal models}
Simulators. Open competitions. Converting other data sets. Existing data sets.

\bibliographystyle{apalike}% Select the citation style e.g. ieeetr
\bibliography{../library}% write the directory to the .bib file
\end{document}

### 2017-04-13 thesis/thesis_map.tex
long overdue update

--- 2017-04-13 thesis/thesis.tex
@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[11pt,a4paper]{book}
 \usepackage[utf8]{inputenc}
@@
 
-\section*{Notes on papers}
-from association to causation via regression (Freedman)
-very clear description of the goal of causal inference, very critical causal modelling due to the impossibility of empirically assessing the assumptions.
+\section{This Thesis}
+\begin{itemize}
+\item clarify where causal inference fits within machine learing, for which problems it is required and ...
+\end{itemize}
 
-\section*{What is causality and why do we care?}
+\section*{Uncategorized}
+http://www.news.com.au/finance/money/costs/insurance-companies-secrets-spilt/news-story/f6ef17ae73e3a5636343845666532b91 insurance decisions voodoo because of lack of transparancy and absense of obvious causal link.
+Claim: People are more comfortable with desicion making on the basis of factors they believe to be causally relevet.
 
-The meaning of causality remains widely debated within science and philosophy ...
+\chapter*{Introduction (3500 words)}
 
-For this thesis I define a causal model to be any model that aims to predict the outcome of an action or intervention in some system.
-From this viewpoint, most real problems require causal inference.
-What use is a model if you are not going to use it to guide any decisions?
-How then can machine learning be integrating itself into so many key aspects of our lives, while causal inferene remains a small subfield mostly of interest to economists?
-The answer is two-fold.
-For some problems, the effect of a decision is obvious and does not impact the system on which it was based.
-In these cases, the causal aspect of the problem can be safely ignored.
-Secondly it is likely that machine learners and data scientits, bouyed by successes in the former problems and the excitement of a world awash with data, computing power and non-linear models are overlooking the fact that their predictions may no longer hold if people act on them.
-Consider the following problems, do they require causal inference?
+\paragraph*{Motivating summary}
 
-notes: We don't always need to predict the outcome of an action. In somes
+\section*{What is causality and why do we care? (2000 words)}
 
-The goal may be more limitied than predicting the outcome of an intervention. To decide between actions, we only assumptions or estimates as to which is better.
+\subsection*{Defining causality}
+\begin{itemize}
+\item widely debated in science and philosophy (FIND SOME REFERENCES)
+\item what is explaination?
+\item any model that aims to predict the outcome of an action or intervention in a system
+\item I do not see the distinction between explaination and (causal) prediction.
+Explaination is all about the ability to compress and to generalize.
+The more a model can do this, the more we view as providing an understanding of the why.
+\item mediation?
+\end{itemize}
 
-For example, in the speech recognition, the immediate result of the action may be known and deterministic (Siri does something).
-The full consequences of that action are not modelled, (we assume the desired outcome is that Siri )
+\subsection*{Identifying when we have a causal problem}
+
+\subsubsection*{Examples of typical machine learning probems. Are they causal?}
 
@@
 \item Predicting who will click on an ad
-
 \item Financial trading
@@
 \item Self driving cars
+\end{itemize}
+
+\subsubsection*{What aspects of a problem determine if causal inference is required?}
+(When is pure prediction useful?)
+\begin{itemize}
+\item To decide between actions we only need to rank them (not estimate their actual effect). 
+\item The predicted outcome in the absence of an intervention provides a single point.
+We can use this to find which problems are most serious if left alone - and priortise those for modelling changes.
+\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
+\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
 
@@
 
-The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.
+\section*{Approaches to causality (1000 words)}
 
-Consider speech recognition.
-You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words.
-Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a dvd on elocution).
-A similar arguament could be made for many applications of machine translation and image classification.
-What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decicion will not effect if it actually rains.
-However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
-For this we need a (causal) model that allows us to predict the outcome under various different interventions.
+\paragraph*{Two broad approaches} 
+\begin{itemize}
+\item Build a model to map the natural behaviour of the system to what will happen for some action
+\item Take the action and see what happens
+\end{itemize}
 
-Identifing spam and automated essay marking systems are similar.
-The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label.
-Spammers and students will modify their writing in order to optimise thier results.
-These systems can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided.
-(What would the nature of the features have to be such that change did not occur?
-- they would have to be causes of the label.
+\paragraph*{The first is causal inference}
 
-What about predicting the risk of death in patients with pnumonia?
-Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective this seems like a straighforward prediction problem.
-It is not.
-Depending on how the decision to admit was previously made and what features are included (or ommited) in the model, the relationship between those features and the outcome may change if we start using the model to decide whome to admit.
-(xxx et al) found exactly this effect.
-Their model learnt that (among other things) people suffering asthma were LESS likely to die from pnemonia.
-They realised this was because doctors were treating such patints very agressively, thus acutally lowering their risk.
-There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original admition and treatment protocols.
-However, using it to decide who to admit could kill.
-The key is understanding exactly what question you are asking.
-In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.
+\paragraph*{The second is reinforcement learning}
 
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc).
-For all these problems the assumptions required to treat them as pure prediction problems are;
-\begin{enumerate}
-\item The treatment is assumed to be effective (at least better than nothing)
-\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
-\end{enumerate}
-It is easy to see how similar issues to the pneumonia example could occur for the customer and parole cases.
-For example, if the customer retention model is not aware of existing customer loyalty schemes or the most risky criminals are currently only given parole in exceptional circumstances (not visable to the model) so rarely offend.
-If people are aware of the model they may also try to game it (as in the spam/essay example) so as to recive loyalty reward or early release.
-TODO talk about the interesting long-term vs short term dilema introduced by the parole example.
-(maybe best to discuss it further on)
+\paragraph*{Both generalize from randomized experiment} Reinforcment learning to sequential decisions, causal inference to non-experimental conditions
 
-Demand prediction seems relatively straightforward.
-These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
-It seems unlikely that using the model to ensure stock is available will itself change demand.
-However there is a potential data sensoring issue.
-If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
-Changing availability does then change demand.
+\paragraph*{Both approaches involve assumptions} the latter that we can group context and actions.
 
-Playing Go (and other games) is another case with some subleties.
-At every turn, the AI agent has a number of actions available.
-The board state following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win.
-...
-this is interesting maybe come back to it ...
-an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ...
-One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken).
-When can we learn from the actions others have taken?
-When there is no confounding.
-And does this hold with Go?
-Probably because the board state encapsulates everything that should determine what move is played.
-Learning directly from actions and trying to generalize (can in some instances reduce the problem to standard ML)
+\paragraph*{Limitations of causal inference}
 
-I hope these examples gave you a feel for the richness and subtlties of causal inference.
-We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
+\paragraph*{Limitations of experimets} What are the issues with standard randomized experiments?
 
-I do not see the distinction between explaination and (causal) prediction.
-Explaination is all about the ability to compress and to generalize.
-The more a model can do this, the more we view as providing an understanding of the why.
+\chapter*{Causal models (5000 words)}
+\subsection*{Causal graphical models}
+Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.
 
-Does this definition include mediation?
+A complaint leveled against this view point of causality is that the 'surgery' is too precise and that, in the real world, any intervention will effect many variables (eg Cartwright 2007). However,
 
-Sometimes the causal component is obvious.
+\subsection*{Counterfactuals}
+\subsection*{Structural Equation models}
+\subsection*{Comparing and unifying the models}
+\subsubsection{A translator from graphical independence to counter factual statements}
 
-Relationship to generalization.
-Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
+\chapter*{Two key questions (5000 words)}
+\section*{Causal Inference} 
+\paragraph*{You are willing to assume the causal graph}
+\paragraph*{extremely widely appliled} Implicitly accounts 10,000 of studies in phycology, medicine, business, etc.
+
+\subsection*{Identifiability}
+\paragraph*{Under what conditions in is the problem solveable}
+\subsubsection*{The Do Calculus}
+
+\subsubsection*{Open questions within identifiablity}
+
+\subsection*{Estimation}
+\paragraph*{How well do we actually do with finite datasets?}
+
+\section*{Causal Discovery}
+\paragraph*{You want to learn the graph}
+\paragraph*{Equates to the aim of automating scientific discovery}
+\paragraph*{Increadibly hard}
+\subsection*{Discovery with Conditional independence}
+\subsection*{Discovery with Functional Models}
+
+\chapter*{Multi-Armed Bandits (5000 words)}
+
+What is the role of randomization? How do bandits algorithms work despite being only partially randomized? 
+What else can you do to improve randomized studies (variance reduction, lower regret).
+
+\chapter*{Causal Bandits: Unifying the approaches (5000 words)}
+
+\chapter*{Causal Inference \& Machine Learning (5000 words)}
+A more detailed discussion on where causal inference sits within machine learning and what it can offer.
+
+\section*{Different approaches}
+\paragraph{The two cultures}
+
+\subsubsection*{Translating terminology}
+Economist -> ML
+
+\section*{Challenges for the Machine learning approach}
+\subsection*{No cross validation}
+\paragraph*{The challenge of model selection}
+\paragraph*{Does predictive accuracy indicate a good causal model?}
+\subsection*{Less large, real world datasets}
+\paragraph*{The dearth of experimental data}
+\subsubsection*{Data for testing causal models}
+Simulators. Open competitions. Converting other data sets. Existing data sets.
+
+\section*{Relation to other areas of ML}
+\subsection*{Covariate shift}
+\subsection*{Generalizability}
+\paragraph*{Invariants}
+\paragraph{more stable} Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
 The assumption is there are less places for change to come in.
 
-If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.
+\paragraph{change input distribution}If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.
 
-The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)?
+\paragraph{feature selection} The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)?
 This may be a reason for using structure learning type algorithms even if you are simply doing prediction.
-
-\subsubsection*{The role of assumptions in causal inference}
-
-Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.
-
-Does predictive accuracy indicate a good causal model?
-
-Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.
-
-Generalizing the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parashutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalize).
-(and given sufficient generalizability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)
-
-Is causal inference possible (are the conclusions ever valid)?
-
-Only from randomized experiment ... if people can do it without recource to constant randomized experiment then an algorithm exits.
-
-The challenges and limitations of infering causal effects from observation data are numerous.
-However, there are problems we need to solve.
-The questions is are these problems so serious as to warrent discregarding all non-experimental data.
-I think answer is a clear no.
-Need ongoing validation, with assumptions.
-We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).
-
-\subsection*{Mediation}
-\subsection*{Relationship to interpretability}
+\subsection*{Interpretabilty}
+\paragraph*{interpretable models as proxies for causal models} Let the human do the work.
+If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
 
@@
 
-Are causal/interpretable models more reliable or more likely to generalize well?
-
 One form of interpretability gives people insight into what the features are that the model is relying on.
-
-If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
 
@@
 People have access to a lot of detailed prior knowledge.  
-
-\subsection*{Relationship to transfer learning}
-
+\subsection*{Transfer Learning}
 find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time).
@@
 However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.
-
-\subsection*{The two approaches to causality}
-
-What is the role of randomization? How do bandits algorithms work despite being only partially randomized? 
-What else can you do to improve randomized studies (variance reduction, lower regret).
-
-What are the issues with standard randomized experiments?
-
-\section*{Causal Inference}
-
-\subsection*{Models of causality and intervention}
-\subsubsection*{DAGs}
-
-Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.
-
-A complaint leveled against this view point of causality is that the 'surgery' is too precise and that, in the real world, any intervention will effect many variables (eg Cartwright 2007). However,
-
-\subsubsection*{Counterfactuals}
+\subsection*{Fair Machine Learning}
 
 \subsubsection*{Structural Equation Models}
-
-A translator from graphical independence to counter factual statements.
 
@@
 
-\section*{Multi-armed Bandits}
-
-\section*{Unifying the frameworks}
-\label{sec:unificaition}
+\chapter*{Conclusions (1000 words)}
 
@@
 
-\section*{Data for testing causal models}
-Simulators. Open competitions. Converting other data sets. Existing data sets.
-
 \bibliographystyle{apalike}% Select the citation style e.g. ieeetr

### 2017-04-13 wiml_poster/causality_ethics.tex
long overdue update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% a0poster Landscape Poster
% LaTeX Template
% Version 1.0 (22/06/13)
%
% The a0poster class was created by:
% Gerlinde Kettl and Matthias Weiser (tex@kettl.de)
% 
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[a3,portrait,14pt]{sciposter}

\usepackage{multicol,caption} % This is so we can have multiple columns of text side-by-side
\columnsep=20pt % This is the amount of white space between the columns in the poster
\columnseprule=3pt % This is the thickness of the black line between the columns in the poster

\usepackage[svgnames]{xcolor} % Specify colors by their 'svgnames', for a full list of all colors available see here: http://www.latextemplates.com/svgnames-colors

%\usepackage{times} % Use the times font
%\usepackage{palatino} % Uncomment to use the Palatino font

\usepackage{graphicx} % Required for including images
\graphicspath{{figures/}} % Location of the graphics files
\usepackage{booktabs} % Top and bottom rules for table
\usepackage[font=small,labelfont=bf]{caption} % Required for specifying captions to tables and figures
\usepackage{amsfonts, amsmath, amsthm, amssymb} % For math fonts, symbols and environments
%\usepackage{wrapfig} % Allows wrapping text around tables and figures

\usepackage{listings}
\usepackage{amsfonts}
\usepackage{tabularx}
\usepackage{enumitem}

\usepackage{tikz}
\usepackage{graphicx}
\usepackage{subfig}

\usepackage{lipsum}

%\usepackage[boxruled]{algorithm2e}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\newtheorem{Def}{Definition}
\newtheorem{Theorem}{Theorem}

\usetikzlibrary{arrows,positioning}

\usetikzlibrary{arrows,positioning} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=2.5em,
           minimum height=1.5em,
           font=\footnotesize,
           text centered,
           scale=1,
           fill=blue!20!white},
     latent/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.5em,
           minimum height=1.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

\begin{document}
\def\ci{\perp\!\!\!\perp}
%----------------------------------------------------------------------------------------
%	POSTER HEADER 
%----------------------------------------------------------------------------------------

% The header is divided into three boxes:
% The first is 55% wide and houses the title, subtitle, names and university/organization
% The second is 25% wide and houses contact information
% The third is 19% wide and houses a logo for your university/organization or a photo of you
% The widths of these boxes can be easily edited to accommodate your content as you see fit
\begin{minipage}[t]{0.15\linewidth}%
\centering
\includegraphics[height=3cm]{data61} % Logo or a photo of you, adjust its dimensions here
\end{minipage}
\begin{minipage}[b]{0.65\linewidth}
\centering
\resizebox{\linewidth}{!}{\veryHuge \color{NavyBlue} \textbf{Transparancy, Causality \& Fairness}}
 \color{Black}\\

\textbf{Finnian Lattimore}\\ % Author(s)
Australian National University\\ % University/organization
finnlattimore@gmail.com

\end{minipage}
\begin{minipage}[t]{0.15\linewidth}
\centering
\includegraphics[height=3cm]{anu-logo-notext}
\end{minipage}

%----------------------------------------------------------------------------------------

\begin{multicols}{2} % This is how many columns your poster will be broken into, a poster with many figures may benefit from less columns whereas a text-heavy poster benefits from more

%----------------------------------------------------------------------------------------
%	INTRODUCTION
%----------------------------------------------------------------------------------------

%\color{DarkSlateGray} % DarkSlateGray color for the rest of the content

\section*{Introduction}

The rise of machine learning and big data is accompanied by increasing concerns over transparency and fairness \cite{Barocas2014}. A recent EU directive requires automated decision systems used to profile individuals must provide \textit{"meaningful information about the logic involved."}  and \textit{"shall not be based on special categories of personal data"} \cite{Goodman2016}

\section*{Fairness \& Discrimination}
Consider a binary outcome of interest $Y$, protected variable $X$, other covariates $Z$ and a model $f$ that outputs the probability of a positive $Y$ for each individual.    
\begin{enumerate}
\item \textit{disparate treatment} Two people with otherwise identical attributes should be treated the same. $f(x,z) = f(x',z) \; \forall \; x,x',z$. 
\item \textit{disparate impact} The distribution of the outcome should be the same for all values of the protected variable. $\sum_z f(x,z)P(z|x) = C \; \forall \; x$
\end{enumerate}

\textit{Disparate treatment} can be trivially avoided by excluding the protected variable from the model. However, this is deeply unsatisfying given the presence of proxy variables and can increase bias (figure \ref{fig:recidivism})

Avoiding \textit{disparate impact} may be be expensive in terms of predictive accuracy and/or require \textit{disparate treatment}.

\begin{figure}[H]
\caption{Proxies}
\label{fig:recidivism}
	\centering    
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
 %nodes
\node[latent](1){Actual priors};
\node[main node, below left=of 1](2){Recorded Priors};
\node[main node, below right=of 1](3){Re-offend};
\node[latent, above left=of 2](4){Stop and Search};
\node[main node, above=of 4](5){Race};
\node[main node, above=of 1](6){Education};
 \path[every node/.style={font=\tiny}]
    (1) edge (2) edge (3)
    (4) edge (2)
    (5) edge (4) edge (6)
    (6) edge (1);
\end{tikzpicture}
\end{figure}

\section*{Transparancy/Interpretability}
A desire for interpretability implies a miss-match between the real world goal and the optimisation problem presented to a machine learning algorithm \cite{Lipton2016}.
\begin{enumerate}
\item situations where the true objective is hard to measure or quantify. (model can be assessed in multiple different ways for hard to define characteristics such as \textit{fairness}) 
\item integration with human decision making
\item improved generalizability (users can detect and eliminate features that are irrelevant/artefacts of training data)
\item increased \textit{trust} due to the ability to validate the model for problems such as data leaks \& feedback, where acting on model predictions changes the system. 
\item user \textit{trust} leading to increased adherence to the model (irrespective of any actual advantages). 
\end{enumerate}

\section*{Causal Models}
We define a causal model as a model which can be used to predict the outcome of an intervention or change to a system.

\begin{enumerate}
\item Causal models explicitly predict the outcome of an intervention. If taking actions based on the outcome of the model significantly changes the system, a causal model is required.
\item Interpretable models can mitigate some, but not all problems associated with not building a causal model eg \cite{Caruana2015} (figure \ref{fig:pneumonia}).
\item If a model is non-causal making it transparent can lead to changes in people's behaviour that reduce the predictive accuracy of the model: Consider making the details of an automated essay marking system public. 
\item If membership of a protected class, $X$, is not causally related to the outcome of interest, $Y$, then there exits a set of variables, $Z$, such that $X \ci Y |Z$.

\end{enumerate}

\begin{figure}[H]
	\caption{Predicting risk of death in pneumonia patients}
	\label{fig:pneumonia}
	\centering    
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
 %nodes
\node[main node](1){Asthma, etc};
\node[latent, below left=of 1](2){Treament};
\node[main node, below right=of 1](3){Death};
 \path[every node/.style={font=\tiny}]
    (1) edge (2) edge (3)
    (2) edge (3);
\end{tikzpicture}
\end{figure}

%\section*{Questions}
%\begin{itemize}
%\item Does causality play a role in the definition of discrimination?
%\item Under what assumptions do causal models generalize better?
%\end{itemize}

%----------------------------------------------------------------------------------------
%	CONCLUSIONS
%----------------------------------------------------------------------------------------

\color{SaddleBrown} % SaddleBrown color for the conclusions to make them stand out

\color{DarkSlateGray} % Set the color back to DarkSlateGray for the rest of the content

 %----------------------------------------------------------------------------------------
%	REFERENCES
%----------------------------------------------------------------------------------------

\bibliographystyle{plain} % Plain referencing style
\bibliography{/home/finn/phd/Causality-Interpretability_and_Ethics} % Use the example bibliography file sample.bib

\end{multicols}
\end{document}

### 2017-04-13 causal-bandit/nips_2016/causal_bandits.py

@@ -1,0 +1,581 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Feb  3 09:51:47 2016
+
+@author: finn
+"""
+import numpy as np
+from itertools import product
+from scipy.optimize import minimize
+from math import sqrt,log,ceil
+from numpy.random import binomial
+import matplotlib.pyplot as plt
+from time import time
+from datetime import datetime as dt
+
+
+REGRET_LABEL = "Regret"
+HORIZON_LABEL = "T"
+M_LABEL = "m(q)"
+
+def now_string():
+    return dt.now().strftime('%Y%m%d_%H%M')
+
+def random_eta(n):
+    eta = np.random.random(n)
+    eta = eta/eta.sum()
+    return eta
+    
+def most_unbalanced_q(N,m):
+    q = np.full(N,1.0/m,dtype=float)
+    q[0:m] = 0
+    return q
+    
+def part_balanced_q(N,m):
+    """ all but m of the variables have probability .5"""
+    q = np.full(N,.5,dtype=float)
+    q[0:m] = 0
+    return q
+        
+def argmax_rand(x):
+    """ return the index of the maximum element in the array, ignoring nans. 
+    If there are multiple max valued elements return 1 at random"""
+    max_val = np.nanmax(x)
+    indicies = np.where(x == max_val)[0]
+    return np.random.choice(indicies) 
+
+def calculate_m(qij_sorted):
+    for indx,value in enumerate(qij_sorted):
+        if value >= 1.0/(indx+1):
+            return indx
+    return len(qij_sorted)/2      
+    
+class VeryConfounded(object):
+    def __init__(self,a,b,n,q1,q2,pYgivenW):
+        #self.pV = np.vstack((1.0-q,q)) TODO fix this.
+        self.n = n # number of V variables
+        self.N = self.n + 2 # number of variables total
+        self.K = 2*self.N + 1 #v_1...v_n = 0,v_1...v_n=1,w0=0,w1=0,w0=1,w1=1,do()
+        self.pYgivenW = pYgivenW # 2*2 matrix p(y|0,0),p(y|0,1),p(y|1,0),p(y|1,1)
+        
+        self.pW0givenA = np.full(self.K,(1-q1)*a + q1*q2*(n-2)/(n-1.0))
[…]

### 2017-04-13 colinear/colinear.py

@@ -1,0 +1,102 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Feb 23 20:31:57 2017
+
+@author: finn
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+
+from sklearn.model_selection import KFold
+from sklearn.metrics import r2_score
+from sklearn.linear_model import LinearRegression
+import seaborn as sns
+
+np.random.seed(42)
+
+def sample(n_train, n_test):
+    n = n_train+n_test
+    x1 = np.random.normal(size=n)
+    x2 = x1 + np.random.normal(scale = 0.01,size=n)
+    y = x1 + x2 + np.random.normal(scale = 0.5,size=n)
+    X = np.stack((x1,x2)).T
+    X_train, X_test = X[0:n_train,:],X[n_train:,:]
+    y_train, y_test = y[0:n_train],y[n_train:]
+    return X_train,X_test,y_train,y_test
+
+
+
+
+
+n_train = 50
+n_test = 50
+simulations = 1000
+r2 = np.zeros(simulations)
+coef = np.zeros((simulations,3))
+min_c1 = np.inf
+model1 = None
+model2 = None
+bigest_gap = -np.inf
+
+for indx in range(simulations):
+    X_train,X_test,y_train,y_test = sample(n_train,n_test)
+    model = LinearRegression()
+    y_pred = model.fit(X_train,y_train).predict(X_test)
+    error = r2_score(y_pred,y_test)
+    r2[indx] = error
+    coef[indx,0] = error
+    coef[indx,1:] = model.coef_
+    if model.coef_[0] < min_c1:
+        min_c1 = model.coef_[0]
+        model1 = model
+    if model.coef_[0] - min_c1 > bigest_gap:
+        bigest_gap = model.coef_[0] - min_c1
+        model2 = model
+        
+_,X,_,y = sample(1,50)
+y1 = model1.predict(X)
+y2 = model2.predict(X)
+print model1.coef_
+print model2.coef_
+
+data = pd.DataFrame({"r2":r2,"b_1":coef[:,1],"b_2":coef[:,2]})
+d = pd.melt(data)
+sns.boxplot(x = "variable",y="value",data = d)
+
+
+
+fig, ax = plt.subplots(1,3)
+ax[0].set_title("distribution of $\\beta_1$")
[…]

### 2017-04-13 effects/conditioning.py

@@ -1,0 +1,32 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jun 30 16:31:59 2016
+
+@author: finn
+"""
+import numpy as np
+
+samples = 10
+
+
+x_means = [1,2]
+x_var = [.1,4]
+num_features = len(x_means)
+
+x = np.empty((samples,num_features+1))
+for i in range(num_features):
+    x[:,i] = np.random.normal(x_means[i],x_var[i],size=samples)
+x[:,-1] = 1
+
+w = [.1,5,2]
+
+# how much each feature shifts the mean will be important for its selection via the balanced regression approach. (not for LASSO)
+
+t = 
+
+
+
+
+
+
+

### 2017-04-13 effects/untitled0.py

@@ -1,0 +1,7 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jun 30 16:31:59 2016
+
+@author: finn
+"""
+

### 2017-04-13 identify/adjust/pdtest.py

@@ -1,0 +1,9 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Jul 12 11:31:03 2016
+
+@author: finn
+"""
+import pandas as pd
+import numpy as np
+df = pd.DataFrame(np.random.randn(6,4), columns=list('ABCD'))

### 2017-04-13 llp/estimators.py

@@ -184,6 +184,7 @@ class CausalLLMEstimator(object):
                          verbose=0, n_jobs=4)
 
         # TODO figure out how to modify LMM to handle differences in proportions
+        # TODO figure out how validation/cross-validation works inside this llm implementation
         self.lmm = lmm.fit(X, labels, bag_ids, X_val, labels_val, bag_ids_val)
 
 
@@ -295,6 +296,62 @@ def compare_errors(n_features, nrows ,simulations, c_vals = [0,.5,1],test_sample
             scores[0,c_indx,s] = mean_squared_error(ce_true,ce1)
             scores[1,c_indx,s] = mean_squared_error(ce_true,ce2)
     return scores
+
+
+class EMEstimator2(object):
+    """ Implementation assuming no defiers as described by Tiberio """
+    def fit(self,x_treat,x_control,y_treat,y_control, tol = 1e-2, max_iter = 500):
+         m_treat = LogisticRegression().fit(x_treat,y_treat)
+         m_control = LogisticRegression(x_control,y_control)
+
+         xt0,xc1,xt1,xc0 = x_treat[y_treat == 0],x_control[y_control == 1],x_treat[y_treat == 1],x_control[y_control == 0]
+
+         x = np.vstack((
+             xt0, # have label N
+             xc1, # have label A
+             xt1, # have label A with weight w_A
+             xt1, # have label C with weight w_C
+             xc0, # have label N with weight w_N
+             xc0)) # have label C with weight w_C'
+
+         blocks = [xt0.shape[0],xc1.shape[0],xt1.shape[0],xt1.shape[0],xc0.shape[0],xc0.shape[0]]
+
+         w_A = np.true_divide(m_control.predict_proba(xt1),m_treat.predict_proba(xt1))
+         w_N = np.true_divide(1-m_treat.predict_proba(xc0),1-m_control.predict_proba(xc0))
+
+         sample_weights = np.concatenate((np.ones(blocks[0]+blocks[1]),w_A,1-w_A,w_N,1-w_N))
+
+         y = np.repeat([0,1,1,2,0,2],blocks) #N=0,A=1,C=2
+
+         epsilon = np.Inf
+
+         model = LogisticRegression()
+         while (epsilon > tol):
+             model.fit(x,y,sample_weight = sample_weights)
+             p_xt1 = model.predict_proba(xt1)
[…]

### 2017-04-13 llp/llp.py

@@ -8,28 +8,64 @@ import numpy as np
 from simulator import Linear
 from estimators import StandardEstimator, CausalLLMEstimator, EMEstimator
 
-
+import matplotlib.pyplot as plt
 
 
 # create some simulated data
 
-n = 100
-d = 5
-data_model = Linear(d)
-x_treat,x_control,y_treat,y_control = data_model.sample(n)
+#n = 100
+#d = 5
+#data_model = Linear(d)
+#x_treat,x_control,y_treat,y_control = data_model.sample(n)
+#
+#
+#
+#llm = CausalLLMEstimator()
+#llm.fit(x_treat,x_control,y_treat,y_control)
+#
+#logistic = StandardEstimator()
+#logistic.fit(x_treat,x_control,y_treat,y_control)
+#
+#em = EMEstimator()
+#em.fit(x_treat,x_control,y_treat,y_control)
 
+from sklearn import linear_model, datasets
 
+# import some data to play with
+iris = datasets.load_iris()
+X = iris.data[:, :2]  # we only take the first two features.
+Y = iris.target
 
-llm = CausalLLMEstimator()
-llm.fit(x_treat,x_control,y_treat,y_control)
+h = .02  # step size in the mesh
 
-logistic = StandardEstimator()
-logistic.fit(x_treat,x_control,y_treat,y_control)
+logreg = linear_model.LogisticRegression(C=1e5)
 
-em = EMEstimator()
-em.fit(x_treat,x_control,y_treat,y_control)
+# we create an instance of Neighbours Classifier and fit the data.
+logreg.fit(X, Y)
 
+# Plot the decision boundary. For that, we will assign a color to each
+# point in the mesh [x_min, x_max]x[y_min, y_max].
+x_min, x_max = X[:, 0].min() - .5, X[:, 0].max() + .5
+y_min, y_max = X[:, 1].min() - .5, X[:, 1].max() + .5
+xx, yy = np.meshgrid(np.arange(x_min, x_max, h), np.arange(y_min, y_max, h))
+Z = logreg.predict(np.c_[xx.ravel(), yy.ravel()])
 
+# Put the result into a color plot
+Z = Z.reshape(xx.shape)
+plt.figure(1, figsize=(4, 3))
+plt.pcolormesh(xx, yy, Z, cmap=plt.cm.Paired)
+
+# Plot also the training points
[…]

### 2017-04-21 thesis/thesis2.tex
completed thesismap

--- 2017-04-13 thesis/thesis_map.tex
@@
-\documentclass[11pt,a4paper]{book}
+\documentclass[11pt,a4paper,oneside]{book}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{hyperref}
+\usepackage{epigraph}
 \usetikzlibrary{arrows,positioning} 
@@
 
-\section{This Thesis}
+\epigraph{In vain the grave, with retrospective Eye,
+Would from the apparent what conclude the why,
+Infer the Motive from the Deed, and show
+That what we chanced, was what we meant, to do.}{Alexander Pope}
+
+\chapter{Random Notes}
 \begin{itemize}
-\item clarify where causal inference fits within machine learing, for which problems it is required and ...
+\item With enough variables it becomes almost a given that we will be able to classify observations by what period they were collected in (eg detecting covariate shift).
+This means we can conclude the model will fail (when in most cases it will not).
+Is a form of regularization/reparameterization with an penalty on covariate shift the solution?
+Focus detection of covariate shift to those variables that actually influence the prediction.
+
+\item If variables are (conditionally) independent they are unlikely to be directly causally linked. 
 \end{itemize}
-
-\section*{Uncategorized}
-http://www.news.com.au/finance/money/costs/insurance-companies-secrets-spilt/news-story/f6ef17ae73e3a5636343845666532b91 insurance decisions voodoo because of lack of transparancy and absense of obvious causal link.
-Claim: People are more comfortable with desicion making on the basis of factors they believe to be causally relevet.
 
@@
 
-\paragraph*{Motivating summary}
+\section*{Motivation}
+Many of the most important questions in science and in our personal lives are about the outcomes of doing something.
+Will asking people to pay upfront at the doctors reduce long term health expenditure?
+If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.
+
+These are hard questions because they require more than identifying a pattern in data.
+Correlation is not causation.
+Causal inference has proven so difficult using standard methods (such as instrumental variables and propensity scoring) that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.
+
+We now live in a world of data.
+Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth.
+Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose.
+It does not come from the controlled world of a randomised experiment.
+
+Traditional techniques that assume linearity or substantial prior knowledge of causal structure are poorly suited to data sets that may be high dimensional, have non-linear relationships between variables, and where we have limited theory specifying how the variables are related.
+
+Statistical machine learning has been very successful, but generalizing remains very challenging.
+
+How much data do humans get?
+
+\paragraph*{Big data and machine learning are a huge deal}
+
+\paragraph*{Data age} Exciting times, larger data sets. Hype, opportunities for Machine learnig. Algorithms beign incorporated everywhere.
+
+\paragraph*{The end of theory} (it is not) correlation is not enough no matter how many petabytes sit behind it.
+There are two seperate issues with a correlation only one of which is addressed by larger datasets.
+One is noise - it may have happened by chance.
+The other is much more fundamental.
+
+\paragraph*{Large datasets have allowed us to relax assumptions} We can fit more flexible models, rely less on strong parametric assumptions about the form of the curve.
+
+\paragraph*{Theory is what allows you to extrapolate outside the observed domain (Hal Varian)}
+
+\paragraph*{A cautionary tale} Much hard work remains to be done before we can call and end to theory or the for the automation of science.
+
+\paragraph*{There is nothing special about humans.} We do have some limited capacity to experiment - but can (or believe we can) draw inferenes about the likely role of variables we have never directy manipulated ourselves.
+Current ML algorithms are limited.
+If we ignore this, we will pay a price.
 
 \section*{What is causality and why do we care? (2000 words)}
+
+To explain or to predict \cite{Shmueli2010}
+The two cultures \cite{Breiman2001}
 
@@
 \item Self driving cars
+\item Pricing insurance
 \end{itemize}
+
+For image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see.
+In fact this is because we already have one.
+We can be confident that the process will not change.
+If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
+Covariate shift clearly comes in here.
+Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable.
+The mechanism is known but the function may be complex.
 
@@
 \item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
+\item Ethics - ... People's viewpoint on if its ok...
 
@@
 
-\chapter*{Causal models (5000 words)}
+\chapter*{Causal models (3000 words)}
+
+Causal inference aims to infer the outcome of an intervention in a system from data obtained by observing (but not intervening in) the system.
+To do this we need terminology to describe actions and how we anticipate the system should respond to them.
+Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks.
+In this chapter we describe these approaches, examine what problems they allow us to solve and the assumptions they rely on and discuss their differences.
+
+\subsection*{Counterfactuals}
+Counterfactuals (or potential outcomes) are a way of describing distributions under different actions that were developed from the starting point of generalizing randomized experiments.
+
+There are philosophical objections (references) to counterfactuals because of the way the describe alternate universes that were never realized and are not empirically testable (example).
+
+For interventional queries, of the form; what is the is the probaiblity distribution for the variable Y if we intervene to set X = x, and the system is otherwise unchanged?
+Counterfactuals are a short hand.
+They say what is the distribution of Y had X = x (regardless of the value X actually took).
+
 \subsection*{Causal graphical models}
@@
 
-\subsection*{Counterfactuals}
+Provides an explicit mechinism to map knowelede from one setting to another.
+
+A fully observed causal bayesian network allows asymtoptic point estimates of the causal effect of any intervention (assuming positive density).
+
 \subsection*{Structural Equation models}
+
 \subsection*{Comparing and unifying the models}
+Representation equivelent for interventional queries
+This means it is straightforward to take the best elements of work done in any of the frameworks.
+For example, draw a graphical network to determine if a problem is identifiable and which variables we need to adjust for to obtain an unbiased causal estimate.
+Then use propensity scores or ...
+to estimate that effect.
+Alternatively, make parametric assumptions, to make the problem into a structural equation model.
+
+SWIGs \cite{Richardson2013}
+\cite{Dawid2000} Causal inference without counterfactuals
+
+Determinisim inherent in counterfactuals and SEMs.
+
+Suppose the counterfactuals $Y_t$ and $Y_c$ are jointly normal (with equal variance),
+\eq{
+P(Y_t,Y_c) \sim N()
+}
+
+then their difference is also normal,
+
+\eq{
+P(\tau) = N(\mu_t - \mu_c,2\sigma^{2}(1-\rho))
+}
+
+\includegraphics[scale=.6]{figures/counterfactual_nonidentify.pdf}
+
+The variance of $\tau$ is not identifiable, even from experimental data.
+But it seems on the face of it relevent to our decision making.
+If the example above, if $\rho = 1$ then almost everyone benifits slightly from the treatment.
+However if $\rho=0$, there is a wide range, with some people benifiting a lot and others suffering significant harm.
+
+\subsection*{What does a causal model give us? Resolving Simpson's paradox}
+
 \subsubsection{A translator from graphical independence to counter factual statements}
+\subsection*{Defining causal effects} Summarising the difference between two distributions. There is no one answer. \cite{Janzing2013}
 
 \chapter*{Two key questions (5000 words)}
+
+We can rougly categorize the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
+In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
+The goal is then to estimate the causal effect of some action or range of actions.
+WHERE DOES MEDIATION FIT IN?
+THIS IS ALREADY HUGE, and is central to 1000 of papers published each year.
+Causal discovery aims to leverage much broader assumptions to learn the stucture of causal graph from data.
+THIS IS THE AUTOMATION OF SCIENCE.
+
 \section*{Causal Inference} 
@@
 \subsection*{Identifiability}
+\paragraph*{Definition:asymptopic point estimate is possible, without parametric assumptions}
 \paragraph*{Under what conditions in is the problem solveable}
@@
 
-\subsubsection*{Open questions within identifiablity}
-
 \subsection*{Estimation}
+\cite{Imbens2004} Review of non-parametric estimation
 \paragraph*{How well do we actually do with finite datasets?}
 
+\paragraph*{When is regression causal?}
+
+\subsection*{Non identifiable queries}
+
+\subsubsection*{Parametric assumptions}
+eg linearity
+
+\subsubsection*{Bounds}
+\paragraph{Instrumental variables are an example}
+
 \section*{Causal Discovery}
+It is possible to infer some aspects of causal structure with very general assumptions.
+The set of conditional independences in a non-experimental data set indicates some causal structures are more likely than others.
+In addition, there are subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
+These clues are the key to causal discovery algorithms, which attempt to learn causal structure from non-experimental data.
+
 \paragraph*{You want to learn the graph}
@@
 \paragraph*{Increadibly hard}
+\paragraph*{Methods can also be divided into contraint based and search and score}
 \subsection*{Discovery with Conditional independence}
+If variables are (conditionally) independent they are unlikely to be directly causally linked. 
 \subsection*{Discovery with Functional Models}
+\subsection*{Granger causality}
 
-\chapter*{Multi-Armed Bandits (5000 words)}
+\chapter*{The interventionalist viewpoint (5000 words)}
+
+Instead of trying to infer the outcome of an intervention in a system from passive observations we could just try the intervention and see what happens.
+
+Randomization is not a requirement. -> control over treatment assignment is.
+
+Experiments, bandits, reinforcement learning
+
+\paragraph*{internal validity} Are differences in treatment and control groups down to intervention or the result of bias?
+
+\paragraph*{external validity} Are the results of the study applicable to the broader population of interest.
+
+\paragraph*{Fundamental role of randomization}
+
+\paragraph*{Feature selection} Is it even more important here? How do algorithms degrade as irrelevent features are added? In the supervised vs the RL setting? What would be a fair comparison?
+
+Image of randomization breaking any confounding links
+
+Randomization does not ensure target and control group are exactly alike.
+The more features or variables you include, the more likely that there will be a significant difference across at least one variable.
+But the within group variance also increases, the net effect is that it becomes harder to draw a conclusion but not more biased.
+
+\paragraph*{Practical limitations of experimentation} Failure to generalize. Transportability, imperfect compliance. Too many contexts.
 
@@
 
+\section*{Multi armed bandits}
+\paragraph*{Motivation} Why we need to extend from simple randomized experiment.
+
+\subsection*{Regret}
+\paragraph*{Relationship to Markov decision processes}
+\subsection*{The exploration/exploitation trade-off}
+
+\subsection*{Contexual bandits}
+
+\subsection*{Dynamic Systems}
+\begin{itemize}
+\item An explicit model of actions in a partially known system (eg HMM)
+\item Feynman-Kac Lemma; Solving a PDE can be converted to a stochastic process
+\end{itemize}
+
 \chapter*{Causal Bandits: Unifying the approaches (5000 words)}
+
+\subsection*{Learning from log data}
 
@@
 
+\paragraph*{Are models that predict better more likely to be the correct model?}
+For polynomials, if you know the data was generated by a higher order polynomial, you may still be better of predicting it with a lower order one.
+
 \section*{Relation to other areas of ML}
@@
 \subsection*{Interpretabilty}
-\paragraph*{interpretable models as proxies for causal models} Let the human do the work.
+\paragraph*{Interpretable models as proxies for causal models} Let the human do the work.
 If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
+
+\paragraph*{Causal models are more interpretable?}
 
@@
 People have access to a lot of detailed prior knowledge.
+
+http://www.news.com.au/finance/money/costs/insurance-companies-secrets-spilt/news-story/f6ef17ae73e3a5636343845666532b91 insurance decisions voodoo because of lack of transparancy and absense of obvious causal link.
+Claim: People are more comfortable with desicion making on the basis of factors they believe to be causally relevet.
+
 \subsection*{Transfer Learning}
@@
 However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.
+
 \subsection*{Fair Machine Learning}
+\paragraph*{Exchangeability of people} Is there any connection to the concept of imagining you could be born in any position and designing ethics from there?
+\begin{itemize}
+\item what determines what attributes we deam worthy of protection (eg race, gender, sexuality)
+\item what do we protect (and what exceptions do we give)?
+\item Do we think it is bad to select on the basis of protected attributes not directly related to desired outcome?
+\item Do we think it is bad to select in a way that has a disparate impact on a minority/disadvantaged group.
+\end{itemize}
 
-\subsubsection*{Structural Equation Models}
+\subsection*{Causal inference with Bayesian Graphical models}
 
-\subsubsection*{Identifiability}
+\subsubsection*{The likelyhood principle}
+Look at some examples from the stan book.
 
-The notion of identifiability is binary. How could it be softened. There are two obvious approaches. 
-1) Look at the finite time convergence for identifiable queries.
-2) For non identifiable queries, look at what bounds can be achieved or what additional assumptions would be required to make them identifiable.
+Both frequentist and bayesian methods can be used to approach problems of infering the outcome of an intervention.
+However, Bayesian modelling allows you to very naturally encode aditional assumptions and causal prior knowledge.
+This seems especially critical for problems that are not non-parametrically identifiable.
 
-Question - why does everyone use the backdoor criterion. Is it much more frequent in the world. Easier to estimate?
+In the same way that regression may be either causal or non causal, depending on what interpretation and assumptions the researcher places on the equations relating the variables, Bayesian models may or may not be interpretable causally.
 
-\subsection*{Defifinitions}
-\subsubsection*{Interactions}
-\subsubsection*{Causal effects}
-
-\subsection*{Estimation of causal effects}
-
-\subsubsection*{Relationship to covariate shift}
-
-\subsection*{Discovering causal structure}
-
-\subsubsection*{Discovery based on conditional independence}
-
-\subsubsection*{Discovery with functional models}
+Is setting certain priors equivelent to conditioning?
+How must the model be set up to ensure unbiased causal estimates (in what way are the assumptions different to if you are using bayesian estimation techniques for prediction?)
 
@@
 
-\section*{Interesting open questions}
-Cycles - a huge issue. Not covered by Pearl,Rubin etc.
+\section*{Open questions}
+\paragraph{Cycles} - a huge issue. Not covered by Pearl,Rubin etc.
 

### 2017-04-21 causalpy/causalpy/checks.py

@@ -1,0 +1,84 @@
+#############################################################
+model = LinearGaussianBN()
+model.add_var('X',weights = [0])
+model.add_var('Y',['X'], weights = [0,"w_YX"])
+display((model.mu,model.cov))
+
+
+# check
+n = 10000000
+V_x = 2.0
+w_YX = 2.3
+V_y = 3.0
+observed =('Y',3,.001)
+
+conditional = model.observe([observed[0]],[observed[1]])
+display((YgivenX.mu,YgivenX.cov))
+
+# by direct simulation
+X = np.random.normal(0,np.sqrt(V_x),size = n)
+Y = w_YX*X + np.random.normal(0,np.sqrt(V_y),size = n)
+df = pd.DataFrame(columns = ['X','Y'],data = np.vstack((X,Y)).T)
+samples = df.loc[(df[observed[0]] > observed[1]-observed[2]) & (df[observed[0]] < observed[1]+observed[2]), ['X','Y']]
+print(samples['X'].values.mean(),samples['X'].values.var(ddof=0),"based on {0} samples".format(len(samples)))
+
+# by substitution 2
+mu,cov = conditional.set_params({'X':([0],V_x),'Y':([0,w_YX],V_y)}).parameterized_mean_cov()
+display((mu,cov))
+
+##################################################################
+model = LinearGaussianBN()
+model.add_var('Z',weights=[0])
+model.add_var('X',['Z'],weights = [0,"w_XZ"])
+model.add_var('Y',['Z','X'],weights = [0,"w_YZ","w_YX"])
+
+
+# do some basic checking with samples
+n = 1000000
+V_z = 2
+V_x =  2
+w_xz = .8
+V_y = .1
+w_yz = .8
+w_yx = 2.2
+x = (3,.01)
+
+Z = np.random.normal(0,np.sqrt(V_z),size = n)
+X = w_xz*Z + np.random.normal(0,np.sqrt(V_x),size = n)
+Y = w_yz*Z + w_yx*X + np.random.normal(0,np.sqrt(V_y),size = n)
+df = pd.DataFrame(columns = ['Z','X','Y'],data = np.vstack((Z,X,Y)).T)
+samples = df.loc[(df["X"] > x[0]-x[1]) & (df["X"] < x[0]+x[1]), ['Z','X','Y']]
[…]

### 2017-04-21 causalpy/causalpy/models.py

@@ -1,0 +1,238 @@
+import numpy as np
+from sympy import Symbol, Matrix
+
+class LinearGaussianBN(object):
+    """
+    A Bayesian network in which all variables are linear functions of their parents plus additive gaussian noise.
+    The joint distribution over the variables in such a network is a multivariate gaussian. 
+    """
+    def __init__(self):
+        self._variables = []      
+        self._mean = None
+        self._cov = None
+        self._parents = {}
+        self._weights = {}
+        self._variance = {}
+        self._weights_p = {}
+        self._variance_p = {}
+  
+    
+    def copy(self):
+        model = LinearGaussianBN()
+        model._variables = list(self._variables)
+        model._mean = None if self._mean is None else self._mean[:,:] # this is a Matrix, slice = copy
+        model._cov = None if self._cov is None else self._cov[:,:]
+        model._parents = self._parents.copy()
+        model._weights = self._weights.copy()
+        model._variance = self._variance.copy()
+        model._weights_p =self._weights_p.copy()
+        model._variance_p = self._variance_p.copy()
+        return model 
+
+    
+    def _variance_symbol(self,variable):
+        """returns a symbol for the variance of variable string specified."""
+        return Symbol("V_{0}".format(variable))
+    
+    def _weight_symbol(self,variable,parent):
+        if parent is not None:
+            return Symbol("w_"+variable+parent)
+        return Symbol("w_"+variable+"0")
+
+    def add_var(self,variable,parents = None, weights = None, variance = None): #TODO allow specifying weights here  - eg may want to set some to 0.
+        """
+        Add a variable to the network.
+        - variable: a string representing the variable
+        - parents: a list of the names of the parents of this variable (must already be in the network)
+        - weights (optional): the weights of this variable to its parents. The first entry should be the offset
+        """
+        if parents is None:
+            parents = []
+        for p in parents:
+            if p not in self._variables:
+                raise ValueError("Parent {p} is not a variable in the network".format(p = p))
+        if variable in self._variables:
+            raise ValueError("Duplicate variable name, variable {v} already exists in this network".format(v = variable))
+        
+        if weights is None:
+            beta = [self._weight_symbol(variable,v) if v in parents else 0 for v in self._variables]
+            mu = self._weight_symbol(variable,None)
[…]

### 2017-04-21 causalpy/setup.py

@@ -1,0 +1,79 @@
+"""A setuptools based setup module.
+
+See:
+https://packaging.python.org/en/latest/distributing.html
+https://github.com/pypa/sampleproject
+"""
+
+# Always prefer setuptools over distutils
+from setuptools import setup, find_packages
+# To use a consistent encoding
+from codecs import open
+from os import path
+
+here = path.abspath(path.dirname(__file__))
+
+# Get the long description from the README file
+with open(path.join(here, 'README.rst'), encoding='utf-8') as f:
+    long_description = f.read()
+
+setup(
+    name='causalpy',
+
+    # Versions should comply with PEP440.  For a discussion on single-sourcing
+    # the version across setup.py and the project code, see
+    # https://packaging.python.org/en/latest/single_source_version.html
+    version='0.0.1',
+
+    description='A python library for causal inference',
+    long_description=long_description,
+
+    # The project's main homepage.
+    url='https://github.com/',
+
+    # Author details
+    author='Finnian Lattimore',
+    author_email='finnlattimore@gmail.com',
+
+    # Choose your license
+    license='MIT',
+
+    # See https://pypi.python.org/pypi?%3Aaction=list_classifiers
+    classifiers=[
+        # How mature is this project? Common values are
+        #   3 - Alpha
+        #   4 - Beta
+        #   5 - Production/Stable
+        'Development Status :: 3 - Alpha',
+
+        # Indicate who your project is intended for
+        'Intended Audience :: Developers',
+        'Topic :: Statistics :: Causality',
+
+        # Pick your license as you wish (should match "license" above)
+        'License :: OSI Approved :: MIT License',
+
+        # Specify the Python versions you support here. In particular, ensure
+        # that you indicate whether you support Python 2, Python 3 or both.
+        'Programming Language :: Python :: 2.7'
+    ],
+
+    # What does your project relate to?
+    keywords='sample setuptools development',
+
+    # You can just specify the packages manually here if your project is
+    # simple. Or you can use find_packages().
+    packages=find_packages(exclude=['contrib', 'docs', 'tests']),
+
+    # Alternatively, if you want to distribute just a my_module.py, uncomment
+    # this:
+    #   py_modules=["my_module"],
+
+    # List run-time dependencies here.  These will be installed by pip when
+    # your project is installed. For an analysis of "install_requires" vs pip's
+    # requirements files see:
+    # https://packaging.python.org/en/latest/requirements.html
+    install_requires=['numpy'],
+
+
+)

### 2017-04-21 effects/untitled0.py

@@ -1,7 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Thu Jun 30 16:31:59 2016
-
-@author: finn
-"""
-

### 2017-04-29 thesis/causality_tutorial.tex
bootcamp day2 update

--- 2015-10-22 causality_tutorial.tex
@@
 The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo.
-We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome .
+We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
 \subsection{Structural Equation Models}

### 2017-04-29 thesis/thesis2.tex
bootcamp day2 update

@@
 
+Key issue is that we can't observe the joint distribution over $Y_t$ and $Y_c$ even with randomized experiment.
+
 Suppose the counterfactuals $Y_t$ and $Y_c$ are jointly normal (with equal variance),
@@
 
+Note that we can bound this counterfactual distribution based on the variance of the observed interventional distributions.
+
+Can these issues be resolved by considering personalized causal effects as random variables and avoiding counterfactuals?
+
+Counterfactuals make a natural way to talk about causal effects in natural language.
+They could be seen as a short hand for projecting the world forward, holding everything else constant.
+An (impresise) way of expressing that we wish to consider the effect of only a single change.
+
+What assumptions are required for the kind of counterfactual analysis like X would have been higher had B ...?
+
 \subsection*{What does a causal model give us? Resolving Simpson's paradox}
+
+Simpsons paradox
 

### 2017-04-29 thesis/thesis_map.tex
bootcamp day2 update

@@
 \usepackage{algorithm}% http://ctan.org/pkg/algorithm
-\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage[noend]{algorithmic}
 \usepackage{parskip}
@@
 \usepackage{hyperref}
+\usepackage{epigraph}
 \usetikzlibrary{arrows,positioning} 
 \usepackage{subcaption}
+\usepackage[colorinlistoftodos]{todonotes}
+\usepackage{natbib}
+\usepackage{wrapfig}
 
+\graphicspath{ {figures/} }
 \pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
@@
 
+\newcommand{\KL}{\operatorname{KL}}
+\newcommand{\simpleregret}{R_T}
+\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
+\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
+\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
+\newcommand{\EE}{\mathbb E}
+\newcommand{\EEa}{\EE_a}
+\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
+\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
+\newcommand{\actions}{\mathcal{A}}
+\newcommand{\calA}{\mathcal A}
+\newcommand{\etc}{\textit{etc}}
+\newcommand{\ie}{\textit{i.e.}}
+\newcommand{\eg}{\textit{e.g.}}
+\newcommand{\calP}{\mathcal P}
+\newcommand{\x}{\boldsymbol{x}}
+\newcommand{\Ps}{\operatorname{P}}
+\newcommand{\R}{\mathbb R}
+\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
+\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
+
 \newcommand{\defined}{\vcentcolon =}
@@
 \renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+\newcommand{\cf}[2]{{#1}_{#2}}
+\newcommand{\bernoulli}{\operatorname{Bernoulli}}
+\newcommand{\dirac}{\operatorname{Dirac}}
+\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
+\renewcommand{\vec}[1]{\boldsymbol{#1}}
+
+%\theoremstyle{plain}
+
+\newtheorem{theorem}{Theorem}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+
+%\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{assumption}[theorem]{Assumption}
+\newtheorem{remark}[theorem]{Remark}
+\newtheorem{example}[theorem]{Example}
+\let\temp\epsilon
+\let\epsilon\varepsilon
 
@@
 \begin{document}
+\todo{figure out theoremstle is undefined}
 \def\ci{\perp\!\!\!\perp} % from Wikipedia
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{definition}{Definition}[section]
+
 \maketitle
 
-\section{This Thesis}
-\begin{itemize}
-\item clarify where causal inference fits within machine learing, for which problems it is required and ...
-\end{itemize}
+\epigraph{In vain the grave, with retrospective Eye,
+Would from the apparent what conclude the why,
+Infer the Motive from the Deed, and show
+That what we chanced, was what we meant, to do.}{Alexander Pope}
 
-\section*{Uncategorized}
-http://www.news.com.au/finance/money/costs/insurance-companies-secrets-spilt/news-story/f6ef17ae73e3a5636343845666532b91 insurance decisions voodoo because of lack of transparancy and absense of obvious causal link.
-Claim: People are more comfortable with desicion making on the basis of factors they believe to be causally relevet.
+3100 words at the start of boot camp.
+
+\chapter*{Random notes}
+
+Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.
 
@@
 
-\paragraph*{Motivating summary}
+\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.
+
+\paragraph*{This thesis contributes to knoweldge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
+
+\paragraph*{My key research questions are:} 
+\begin{itemize}
+\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
+\item To summarize the key strands of causal inference research from economics and social sciences for the machine learning community
+\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
+\end{itemize}
+
+\section*{Motivation}
+Many of the most important questions in science and in our personal lives are about the outcomes of doing something.
+Will asking people to pay upfront at the doctors reduce long term health expenditure?
+If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.
+
+These are hard questions because they require more than identifying a pattern in data.
+Correlation is not causation.
+Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.
+
+We now live in a world of data.
+Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth.
+Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose.
+It does not come from the controlled world of a randomised experiment.
+
+The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
+From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries.
+It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge datasets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artifcial intelligence exceeds our own and then grows exponentially.
+
+However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings.
+Getting them to generalize to even slightly different problems or datasets remains very challenging.
+Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it.
+The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on.
+As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
 
@@
 
+To explain or to predict \cite{Shmueli2010}
+The two cultures \cite{Breiman2001}
+
 \subsection*{Defining causality}
 \begin{itemize}
-\item widely debated in science and philosophy (FIND SOME REFERENCES)
+\item widely debated in science and philosophy (REFERENCES)
 \item what is explaination?
@@
 \item Self driving cars
+\item Pricing insurance
 \end{itemize}
+
+%For image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see.
+In fact this is because we already have one.
+We can be confident that the process will not change.
+If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
+Covariate shift clearly comes in here.
+Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable.
+The mechanism is known but the function may be complex.
 
@@
 \item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
-
+\item Ethics - ... People's viewpoint on if its ok...
 \end{itemize}
@@
 \section*{Approaches to causality (1000 words)}
+
+There are two broad approaches to deciding how to act.
+Reinforcement learning and causal inference.
+In reinforcement learning we estimate the effect of actions by taking them.
+We assume there is an agent capable of intervening in the system and try to find policies for the agent to follow in selecting actions so as to maximise some kind of reward.
+This is a very powerful and general framework (REFERENCES TO GENERAL AI).
+However, POINT OUT SOME OF THE DIFFICULTIES WITH REINFORCEMENT LEARNING.
+We are frequently presented with large bodies of data that have been collected from a system in which we did not have any control over what actions were taken.
+
+We will call datasets where we do not have control over the decision making process that generated the data observational.
+Versus experimental datasets, where we do have control (experimental datasets do not always have to be randomized -although that is a powerful approach to ensure we have control.
+There is a space in the middle where we have partially controlled the process by which agent select actions.
+Randomized data with imperfect compliance would be an example.
+
+An agent (capable of intervening in the system) chooses an action from those availble The agent making the decision included in the model. This is
+
+Reinforcement learning addresses the problem of learning from explicitly taking actions. There
+is typically some state or environment. An agent chooses an action from those available in the
+current state. The state then evolves stochastically as a function of the selected action and
+the agent receives some feedback or reward that is a function of the new state. This setting
+differs from the standard classification problem in that the agent must learn from feedback on
+the selected action, rather than being presented with the correct action for a given state. A
+common modelling assumption is that the state evolves only as a function of the previous state
+and the action chosen, and given these, is independent of the previous history of states and
+actions. This is known as a Markov decision process or MDP. A particularly well studied and
+understood model is the singe state MDP. In this case, there are a set of actions, each associated
+with a fixed but unknown reward distribution and at each time step our agent selects an action
+and receives corresponding feedback. This is known as the multi-armed bandit problem.
+
+Causal inference makes use of assumptions to allow the outcome of actions to be predicted from
+observational data. The key to causal inference is a framework that can model how actions
+change the state of the world. This framework then allows us to map information collected in
+one setting to another.
+
+Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit
+algorithms deal with the sequential nature of the decision making process, causal inference with
+the problem that full randomisation is not always feasible, affordable or ethical. The similarities
+between the problems that these techniques have been developed to address raises the question
+of if there are problems best addressed by a combination of these approaches and how they
+can be combined. The goal of my thesis is to explore these questions. In the next sections I
+review the key literature in causal inference and bandits. I then present a general approach to
+how causal models might be incorporated into bandit settings and conclude by demonstrating
+an algorithm that leverages causal assumptions to improve performance in a specific bandit
+setting.
+
+There are two key approaches to causal problems.
+The first is to learn the outcome of actions by directly intervening in the system and seeing what happens.
+We then get feedback on how good those actions were.
+This is the approach taken in reinforcement learning.
+ADVANTAGES AND DISADVANTAGES OF THIS APPROACH.
+The second broand approach is causal inference.
+Here
 
@@
 
+Both these fields relate to the problem of making optimal decisions and both can be seen as generalising randomised controlled experiments.
+Causal inference  is the study of how to estimate the effect of an action in the absence of randomisation.
+Reinforcement learning studies how we can do better if the decisions are to be made sequentially.
+
 \paragraph*{Both approaches involve assumptions} the latter that we can group context and actions.
@@
 
-\chapter*{Causal models (5000 words)}
-\subsection*{Causal graphical models}
-Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.
+\chapter*{Causal models (3000 words)}
+
+Causal inference aims to infer the outcome of an intervention in a system from data obtained by observing (but not intervening in) the system.
+To do this we need terminology to describe actions and how we anticipate the system should respond to them.
+Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks.
+In this chapter we describe these approaches.
+We will examine what problems they allow us to solve, what assumptions they rely on and how they differ.
+We will also use them to describe the following very simple example.
+The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
+
+Suppose we have developed a new drug for some illness and wish to determine how effective it is.
+We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group.
+The people in the treatment group get the drug, everyone else gets a placebo pill.
+The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo.
+We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
+
+\subsection*{Causal bayesian networks}
+
+Causal Bayesian Networks are an extension of Bayesian networks.
+A bayesian network is a graphical way of representing how a distriution factorises.
+Any joint probability distribution can be factorized into a product of conditional probabilities.
+There are multiple valid factorizations, corresponding to permutations of variable ordering.
+
+\begin{equation}
+P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
+\end{equation}
+
+We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}).
+If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.
+%
+%\begin{figure}[h]
+%\centering
+%\caption{General Bayesian network over 3 variables}
+%\label{fig:bayesnet}
+%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+%  thick,main node/]
+%
+% %nodes
+%\node[main node](1){$X_{1}$};
+%\node[main node, below left=of 1](2){$X_{2}$};
+%\node[main node, below right=of 1](3){$X_{3}$};
+%
+%
+% \path[every node/.style={font=\sffamily\small}]
+%    (1) edge node {} (2)
+%    	edge node {} (3)
+%    (2) edge node {} (3);
+%	
+%\end{tikzpicture}
+%\end{figure}
+
+The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorized over the nodes and edges in the graph.
+There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges).
+If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
+
+\begin{equation}
+P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
+\end{equation}
+
+A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
+This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
+We need some notation to represent the distribution after an intervention.
+In this thesis we will use the do opperator introduced by Pearl \ref{XXX}, where $\P{Y|do(X=x)}$ represents the distribution of the variable $Y$ after an intervention that sets the variable(s) $X$ to $x$.
+TALK ABOUT HOW THIS CAN REPRESENT MULTIPLE THINGS CONTINUOUS VERSUS DISCRETE DISTRBUTIONS AND WHETHER OR NOT VARIABLES ARE RESOLVED OR AS A FUNCTION.
+It is important to understand that in general, the interventional distribution $P{Y|do(X=x)}$ will differ from the standard conditional distribution $\P{Y|X=x}$.
+
+If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network.
+This is referred to as the truncated product formula \cite{Pearl2000}.
+
+\begin{equation}
+\label{eq:truncatedproduct}
+ P_{do(Z=z)}(X_{1}...X_{N}) =
+  \begin{cases}
+  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
+   0       & \text{otherwise } 
+  \end{cases}
+\end{equation}
+
+This result does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution.
+The truncated product formula will give different results depending on which you choose.
+The result is possible with causal bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
+In fact, from the interventionalist viewpoint causality the truncation product formula defines what it means for a link to be causal.
+
+Returning to our simple example, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$?
+Since the treatment $X$ is assigned  via deliberate randomisation we know there it is not effected by any unobserved variables.
+$X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and desease subtype of the patient.
+
+The causal bayesian network for this scenario is:
+
+\todo{proper picture}
+\eq{
+X \rightarrow Y\\
+U_l \rightarrow Y
+}
+
+This network represents the (causal) factorization  $P(X,Y) = P(X)P(Y|X)$, so from equation (\ref{eq:truncatedproduct}), $P(Y|do(X)) = P(Y|X)$.
+In this example, the interventional distribution is equivelent to the observational one.
+
+Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechinism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
+A key feature of this way of defining an intervention is the number of things that are invariant to the intervention.
+That is, all the conditional distributions over variables that were not intervened are assumed not to be changed by the intervention.
+
+\todo {draw the system as two seperate models}
+
+A causal bayesian network encodes all possible interventions that can be represented with the do notation.
+If you knew of all the terms in the causal factorisation, then you could compute the distribution over any variable in the graph after setting any subset of other vairables (subject to possitivity assumption)
+\todo {state positivity assumption}
+
+A causal network represents much more information than a bayesian network with identical structure.
+GIVE EXAMPLE IN SOMETHING QUANTIFIABLE LIKE BITS.
+This may be surprising since their conditional probability distributions are estiated from data in the same way.
+It is only possible due to the additional assumptions inherent in a causal network.
+
+Causal bayesian networks are bayesian networks so results that apply to bayesian networks carry directly across;the local markov property states that variables are independent of their non-effects given their direct causes.
+Similarly the global markov property and d-separation also hold in causal networks.
+
+\subsubsection*{Limitations of causal bayesian networks}
+A number of critisims have been leveled at this way of modelling causality.
 
@@
 
+What questions can't be phrased using this terminology
+- certain counterfactual queries
+- cyclic/dynamics
+
+Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.
+
+- add variables for soft interventions
+- unroll to deal with cycles
+
+\subsection*{Structural Equation models}
+
+There is a problem with the term non-parametric. It is used in machine learning in a slightly different way to pearl
+
+\subsubsection*{Definition}
+Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
+The mechanism (but not the output) is assumed to be independent of what is fed into it.
+Uncertainties are not inherent but arise from unmeasured variables.
+Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
+More recently, they have been formalized, generalized to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.
+
+Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
+The noise terms are required to be mutually independent.
+If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable.
+Structural equation models can be represented visually as a network.
+Each variable is a node and arrows are drawn from causes to their effects.
+For our example the SEM is:
+
+This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
+This is justifiable on the grounds that $X$ is random.
+The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
+
+\subsubsection*{Example problem}
+We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
+Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$.
+The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}.
+However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$.
+The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomized experiment.
+In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.
+
+For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
+However a SEM is, by definition, more than a set of simultaneous equations.
+By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
+When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
+
+Limitations \& Critisisms 
+- can express some non-measurable things
+	- example of that
+
+- confusion may occur as to whether interpretation is causal
+
+- can they handle cycles?
+
 \subsection*{Counterfactuals}
-\subsection*{Structural Equation models}
+\todo{use cf command to in counterfactual equations}
+
+The Neyman-Rubin model \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
+Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}.
+Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else.
+
+In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{i}^{1}$ versus what would happen if they got the placebo, $y_{i}^{0}$.
+The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
+The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment.
+This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
+
+Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.
+Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
+The distribution of $\cf{Y}{1}$ is the distribution we would see of $Y$ if everyone was treated.
+Similarly $Y^{0}$ represents the potential outcome for the placebo.
+We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
+We can estimate, from an experimental or observational study, the probability that people recover if treated $P(Y|X=1)$ and the probability that they recover if not treated $P(Y|X=0)$.
+Now if $X=0$ then $Y = \cf{Y}{0}$.
+Equivalently stated:
+
+\eq{
+\P{Y^{0}|X=0}&= \P{Y|X=0}\\
+\P{Y^{1}|X=1}&=P{Y|X=1}
+}
+
+If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
+
+\eq{
+\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
+\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
+}
+
+\eq{
+\implies \P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
+}
+
+The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{Rosenbaum1983}.
+They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
+Again this is justified in our example due to the randomization of treatment assignment.
+In general these assumption do not hold.
+If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model.
+that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.
+
+\todo{SUTVA assumption}
+
+There are some complex philosopical objections to counter factuals arising from the way they describe alternate universes that were never realized and are not empirically testable \ref{XXXXX}.
+This can result in some concrete problems.
+Consider the following example based on  \ref{Dawid2000}.
+
+Again we have a drug, where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
+Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal.
+We will assume equal variance for simplicity.
+
+% vector with mean mu_t,mu_c and a covariance matrix
+\eq{
+P(Y_t,Y_c) \sim N()
+}
+
+then their difference is also normal,
+
+\eq{
+P(\tau) = N(\mu_t - \mu_c,2\sigma^{2}(1-\rho))
+}
+
+\includegraphics[scale=.6]{figures/counterfactual_nonidentify.pdf}
+
+Key issue is that we can never observe the joint distribution over $Y_t$ and $Y_c$.
+As a result, the variance of $\tau$ is not identifiable, even from experimental data.
+It seems on the face of it that $\tau$ is relevent to our decision making.
+If the example above, if $\rho = 1$ then almost everyone benifits slightly from the treatment.
+However if $\rho=0$, there is a wide range, with some people benifiting a lot and others suffering significant harm.
+
+Can these issues be resolved? If so, how?
+
+- note that we can bound this counterfactual distribution based on the variance of the observed interventional distributions. If the variance of XXX is small then this may not be an issue.
+
+- does it actually make sense to make a decison on the basis of this counterfactual that is a function of something we could never observe. I
+
+- part of the problem may be due to the deteminisitic way we have phrased individual treatment effects.
+
+Can these issues be resolved by considering personalized causal effects as random variables and avoiding counterfactuals?
+
+A range of similar issues can arise in counterfactuals \ref{XXXX}. 
+- What assumptions are required for the kind of counterfactual analysis like X would have been higher had B ...?
+- briefly discuss the different flavours of counterfactual questions here.
+
+One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
+Rather than taking about the distribution of Y given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $X$.
+This is certainly convienient, if rather impressise.
+
 \subsection*{Comparing and unifying the models}
+
+Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
+This makes it straighforward to combine key results and algorithms developed within any of these frameworks.
+For example, draw a graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
+Then use propensity scores \ref{} estimate the effect.
+If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
+
+If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorization of the joint distribution over its variables.
+In other words, the network is a causal Bayesian network.
+All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
+Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
+This corresponds to dropping a term in the factorization and the truncated product formula (equation \ref{eq:truncatedproduct}).
+Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
+We can also connect this to counterfactuals via:
+
+\begin{equation}
+\begin{aligned}
+&Y^{0} \equiv P(Y|do(X=0)) \\
+&Y^{1} \equiv P(Y|do(X=1))
+\end{aligned}
+\end{equation}
+
+The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
+When discussing the counterfactual model, we actually made the slightly weaker assumption:
+
+\begin{equation}
+\label{eq:weakignore}
+X \ci Y^{0} \text{ and } X \ci Y^{1}
+\end{equation}
+
+It is possible to relax the independence of errors assumption we made for SEMs slightly to correspond exactly the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \cite{Richardson2013}.
+To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of functional models or bayesian networks or as conditional independence statements involving counterfactual variables (ignorability assumptions).
+By non-parametrically, I mean that we are not making any assumptions about the form of the relationships between variables.
+Models that are not non-parametrically identifiable can still be identified given assumptions about the distributions of variables and the functional relationship between them, for example, that the functions are linear or that the noise is additive \cite{Peters2014}.
+This form of assumption fits extremely naturally into the structural equation framework.
+
+However we can also pose causal queries that are not interventional and cannot be phrased in terms of the do-notation.
+The patients in our drug treatment example could be broken down into four groups.
+\todo{point out objections to doing this}.
+The first group will recover whether or not they receive treatment, the second group will recover if treated but not on the placebo, the third group will recover on the placebo and not if treated, and the last group will not recover on treatment or placebo.
+Unfortunately, we don't know which group each person belongs to.
+Drawing this up as a table:
+
+\begin{tabular}{c|c|c|c}
+group & placebo & treatment & probability of group\\
+\hline
+1 & die & die & $\alpha=P(Y^{0}=0,Y^{1}=0)$\\
+2 & die & recover & $\beta=P(Y^{0}=0,Y^{1}=1)$\\
+3 & recover & die & $\gamma=P(Y^{0}=1,Y^{1}=0)$\\
+4 & recover & recover & $\delta=P(Y^{0}=1,Y^{1}=1)$\\
+\end{tabular}
+
+The queries we have been asking thus far are about $P(Y^{0}=1) = \gamma + \delta$ and $P(Y^{1}=1) = \beta + \delta$, but suppose we asked the question; what is the probability that this patient, who was not treated and died, would have recovered if they had been treated?
+We know they are in either group 1 or 2 since they died without treatment, so the answer is $\frac{\beta}{\alpha+\beta}$.
+Can we estimate the $\alpha, \beta, \gamma, \delta$ or in other words, identify the joint distribution over the counterfactuals $P(Y^{0},Y^{1})$ given the interventional distributions, $P(Y^{0})$ and $P(Y^{1})$?
+The answer is no, putting our constraints and unknowns in matrix form:
+
+\begin{equation}
+\left(
+\begin{array}{cccc}
+0&0&1&1\\
+1&1&0&0\\
+0&1&0&1\\
+1&0&1&0\\
+1&1&1&1\\
+\end{array}
+\right)
+\left(
+\begin{array}{c}
+\alpha\\
+\beta\\
+\gamma\\
+\delta\\
+\end{array}
+\right)= 
+\left(
+\begin{array}{c}
+P(Y^{0})\\
+1-P(Y^{0})\\
+P(Y^{1})\\
+1-P(Y^{1})\\
+1\\
+\end{array}
+\right)
+\implies
+\left(
+\begin{array}{c}
+\alpha\\
+\beta\\
+\gamma\\
+\delta\\
+\end{array}
+\right)=
+\left(
+\begin{array}{c}
+1-P(Y^{0})-P(Y^{1})+\delta\\
+P(Y^{1})-\delta\\
+P(Y^{0})-\delta\\
+\delta\\
+\end{array}
+\right)
+\end{equation}
+The value of $\delta$ is not determined so the query is not identifiable.
+However we do get bounds on the terms.
+Since probabilities cannot be negative, $P(Y{1})-P(Y^{0})-1 \leq \delta \leq min(P(Y^{1}),P(Y^{0})$.
+Note, if we made the additional assumption $\gamma=0$; that the drug did not cause anyone to die who would otherwise have survived, then we can determine the joint distribution over counterfactuals.
+Alternatively, if we could assume that after treatment people returned to their initial state after some period of time, (say we were testing a drug for acne) then we could run a crossover study to determine the joint distribution.
+In a crossover study, the participants are randomly assigned to treatment and placebo, results are measured and then the groups are swapped.
+The scientific and philosophical validity of counterfactual queries remains under question \cite{Dawid2000,Dawid2014}, however they are nonetheless widely posed in the form of attribution of causal effects to different pathways and mediation \cite{Pearl2014,Imai2010a,VanderWeele2011}.
+
+There are differences between the models we have considered when it comes to counterfactual queries.
+Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.
+Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
+Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
+
+In practice, differences in focus and approach across different fields eclipse these actual differences in the models.
+The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
+The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment.
+This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}.
+In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field.
+Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
+In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
+
 \subsubsection{A translator from graphical independence to counter factual statements}
 
+\subsubsection*{More terminology}
+With the graphical framework in place, it is useful to define some key terminology used in describing causal models in terms of the graph structure they refer to.
+
+\begin{enumerate}
+\item confounding
+\item exogenous
+\item endogenous
+\item nuisance variables
+\end{enumerate}
+
+\subsection*{What does a causal model give us? Resolving Simpson's paradox}
+
+We will now demonstrate how we can use our new notation and frameworks for causal inference to resolve a facinating paradox. \todo{ insert some history of simpson's paradox here.}
+
+Suppose a doctor has two treatments, A and B, which she offers to patients for blood preasure control.
+She keeps track of which medication her patients choose and whether or not the treatment is successful.
+She obtains the following results:
+
+\todo{insert data table with success by treatment}
+
+It seems that drug A is performing better.
+However, having read the latest liturature that medication effectiveness may differ between men and women, she decides to check how each drug performed by gender and obtained the following data.
+
+\todo{innsert table}
+
+Once the data is broken down by gender, it appears that treament B is better for both men \emph{and} women.
+This is the paradox.
+If the doctor were unaware of the gender, it seems that she should advise treatment A.
+But as soon as the patient walks though the door  then she should offer treatment B, regardless of which gender the patient turns out to be.
+How does causaul modelling reslove this paradox?
+Let's represent the treatment presribed by the variable $T$, the gender of the patient by $Z$ and wheter or not the treatment was successful by $Y$.
+The key to resolving the paradox is that the doctor really cares about the interventional distribution $P{Y|do(T)}$, not the standard conditional distributions $P{Y|T}$.
+The results from table XXX shows that $\P{Y|T=A}$ is higher than $\P{Y|T=B}$.
+If she has to suggest a treatment for a patient without knowing their gender, she should choose the treatment for which $P{Y|do(T)}$ is highest not $P{Y|T}$.
+If she knows thier gender then she should otimise to $P{Y|do(T),Z}$.
+
+The causal network for this problem is shown below.
+In this example, women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they recive than men.
+This is what makes treatment $A$ appear better if you look at the conditional distribution.
+The fact that a patient took treatment $A$ tells you that they are more likely to be female.
+In the terminology defined in section \ref{}, gender is a confounding variable and we must adjust for it to obtain unbiased estimates of the causal effect of the treatment.
+As we saw from the examples in chapter \ref{XXX}.
+Determining when you really care about the interventional as opposed to the conditional distribution is a sublte but very important question that is frequently overlooked.
+
 \chapter*{Two key questions (5000 words)}
-\section*{Causal Inference} 
-\paragraph*{You are willing to assume the causal graph}
-\paragraph*{extremely widely appliled} Implicitly accounts 10,000 of studies in phycology, medicine, business, etc.
+
+We can rougly categorize the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
+In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
+The goal is then to estimate the causal effect of some action or range of actions.
+WHERE DOES MEDIATION FIT IN?
+THIS IS ALREADY HUGE, and is central to 1000 of papers published each year.
+Causal discovery aims to leverage much broader assumptions to learn the stucture of causal graph from data.
+THIS IS THE AUTOMATION OF SCIENCE.
+
+\section*{Causal effect estimation}
+
+Causal effect estimation addresses the problem where we assume the key structure of the graph is known. That is, we assume that we have at a minimum:
+
+\begin{itemize}
+\item the target/outcome variable we care about
+\item the focus/treatment variables on which we are considering intervetions
+\item any variables which act to confound two or more of the variables we have included.
+\end{itemize}
+
+Some of these variables my be latent or unobserved, in that we do not have measurements for them in the available data.
+However, their position in the network is assumed to be known.
+For example if we were trying to estimate ...
+
+Causal effect estimation is one of the most frequently applied methodolgoies within applied sciences.
+It implicit in millions of studies accross medicine, business, economics and social sciences.
+Almost everytime someone runs a regression model the key quantity of interest is a causal effect.
+Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best levereage those assumptions to infere as much information as we can from our data.
+
+\subsection*{The do calculus}
+Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting?
+
+The do calculus arises from d-seperation.
+
+How do we take our causal frameworks and use them to map information from the observational to interventional settings?
 
 \subsection*{Identifiability}
-\paragraph*{Under what conditions in is the problem solveable}
-\subsubsection*{The Do Calculus}
+A natural question to ask is, given the level of assumptions we have made about the graph, is it possible to estimate a causal effect from observational data?
+This is the identifiablity problem.
+Identifiability is an asymptopic property of the graph.
+It asks if we can obtain an unbiased point estimate for the causal effect of interest in the infinite data limit.
+A problem is non-parametrically identifaible if it is identifiable without any additional assumptions on the functional form of the dependencies between variables in the graph.
 
-\subsubsection*{Open questions within identifiablity}
+The question of non-parametric identifiability is solved!
+The do calculus is complete.
+A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observataional quantities via repeated application of the do calculus.
+
+There is an algorithm \cite{Shpitser2012} based on these properties that, for a given network and interventional (do-type) query, can:
+
+a) determine if the query can be translated into an expression involving only distributions over observed variables.
+In other words, determine if the query is identifiable given the assumptions encoded by the network
+
+b) if it is identiable, return the required expression
+
+\todo{insert link to working javascript version}
+
+There are many interesting questions relating to identifiability that remain open.
+What is the minimal (by some metric) additional information that would be required to make a non-identifiable qurey identifiable?
+What if we assume various restrictions on the functional form of the relationships between the variables?
+Clearly  .
+A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work going back to at least ???
+\ref{}
+
+Is identifiablity a good question to ask?
+It seems natural to see if a problem can be solved with infinite data before we try to see how well we can do it with finite data.
+However, we should not give up on non-identifiable problems.
+We may still be able to achive useful bounds on causal effects with finite data even if point estimation is not possible.
+From that perspective we should be cautious about dividing problems into identifiable and not identifiable.
+
+How do we tackle a problem that is not identifiable. We could look for bounds or we could make additional assumptions (or both).
 
 \subsection*{Estimation}
+\subsection*{Defining causal effects} Summarising the difference between two distributions. There is no one answer. \cite{Janzing2013}
+
+\cite{Imbens2004} Review of non-parametric estimation
+\cite{causal estimation is covariate shift}
 \paragraph*{How well do we actually do with finite datasets?}
 
+\paragraph*{When is regression causal?}
+A regression equation is causal if, when we represent the equations as a bayesian causal model or structural equation model, we claim the arrows are causal AND if the variables we include in the regression form a valid back door adjustment set.
+
+\subsection*{Non identifiable queries}
+
+\subsubsection*{Parametric assumptions}
+eg linearity
+
+\subsubsection*{Bounds}
+\paragraph{Instrumental variables are an example}
+
+Put in the picture of an instrumental variable.
+
+Instrumental variables are one of the key approaches used to estimate causal effects within economics and the social sciences.
+However, the setting they are associated with is non-identifiable.
+How is it that they are so useful nonetheless?
+Tight bounds may be just as good as what could be achieved in the identifiable setting, given finite data.
+
 \section*{Causal Discovery}
+We now move to the much more general problem of learning a causal graph from observational data.
+Amazingly, it is possible to infer some aspects of causal structure with very general assumptions.
+The set of conditional independences in a non-experimental data set indicates some causal structures are more likely than others.
+In addition, there are subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
+These clues are the key to causal discovery algorithms, which attempt to learn causal structure from non-experimental data.
+
 \paragraph*{You want to learn the graph}
+In the causal discovery problem we make much much broader assumptions about the structure of the graph.
+Ie that it is acyclic.
+and may or may not have latent variables.
+We do not assume the existance or directions of any links between the variables.
+
+What's more, we wish to do this from non-experimental data (although there is also work on infering causal structure from experimental data) REFERENCES and it would be natural to connect the two (ANY REFERENCES THERE?).
+
+This is a much grander goal than causal inference and arguably, if acheived would equate to the automation of scientific discovery. 
 \paragraph*{Equates to the aim of automating scientific discovery}
 \paragraph*{Increadibly hard}
+It it is amazing that anything can be done here.
+\paragraph*{Methods can also be divided into contraint based and search and score}
 \subsection*{Discovery with Conditional independence}
+If variables are (conditionally) independent they are unlikely to be directly causally linked. 
 \subsection*{Discovery with Functional Models}
+\subsection*{Granger causality}
 
-\chapter*{Multi-Armed Bandits (5000 words)}
+In previous sections we discussed when assumptions about the structure of the variables in a specific problem is sufficient to identify a causal effect.
+This approach relies on having enough prior knowledge or theory about the problem to allow you to, at least partially, specify the causal network.
+In this section, we consider the much harder problem of causal inference where you need to learn the network.
+Causal inference might seem impossible without specific assumptions about the structure of the variables involved but, amazingly, some aspects of causal structure can be determined from much more general assumptions.
+
+In this section we consider variants of the following problem: Assume there is some acyclic causal network $G$ that generated the distribution $P(\boldsymbol{V})$ from which our data has been sampled.
+Our goal is to recover the network from this data.
+
+What can we look for in the distribution that could give us clues as to the structure of the network?
+
+\subsection{Causal discovery with conditional independence tests}
+One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
+We know, (section \ref{sec:dsep}), if $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in $P$ .
+However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
+This requires that we assume the reverse condition:$(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
+This assumption, commonly referred to as  \textbf{\textit{faithfulness}}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
+Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} for $P$.
+
+Faithfulness is an assumption.
+It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
+However, most distributions generated by a causal bayesian network will be faithful to that network.
+For faithfulness to be violated, different causal effects must exactly balance one-another out.
+For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
+If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
+However, this independence is not stable.
+It would disappear under a small perturbation to any of the parameters.
+
+Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$.
+This set can also be referred to as the markov equivalence class compatible with $P$.
+
+\subsubsection{Without hidden common causes}
+The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of bayesian networks we must consider.
+
+We will begin with a brute force algorithm (described as the SGS algorithm in \cite{Sprites} and IC algorithm in \cite{Pearl2000}).
+While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.
+
+\begin{table}[H]
+ \begin{tabularx}{\textwidth}{X}
+ \hline
+\rule{0pt}{2.5ex} 
+ \textbf{The SGS (or IC) Algorithm}\\
+ \hline
+ \rule{0pt}{2.5ex}
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
+\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
+ \begin{enumerate}[itemsep=8pt]
+  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an undirected link to form a complete graph.
+  \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent.
+If such a set (including the empty set) exists then $a$ and $b$ cannot be directly connected in $G$ so delete the link.
+  \label{alg:SGSexponential}
+
+  \item For all pairs of non-linked variables $(\alpha,\beta)$ with a common neighbour, $c$, if $c \notin \boldsymbol{S}_{\alpha,\beta}$, then $c$ must be a collider in the path $\alpha,c,\beta$ so  add arrows to direct the links $\alpha-c$ and $\beta-c$ towards $c$.
+  \label{alg:SGScolliders}  
+  \item Recursively try to orient any edges that remain undirected to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
+  \label{alg:SGSfinal}
+\end{enumerate}\\
+ \hline
+\end{tabularx}
+\end{table}
+
+The SGS algorithm utilizes the fact that a collider structure (f
+igure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
+Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
+Unfortunately the number of conditional independence tests required for step \ref{alg:SGSexponential} grows exponentially (in the worst case) with the number of variables.
+Not only that, but for each edge that is in the true network, the algorithm will always tests all other possible subsets of variables.
+If the assumption that there are no hidden common causes or that the distribution is faithful are violated, step \ref{alg:SGScolliders} of the SGS algorithm can produce double headed arrows.
+
+The PC algorithm \cite{Sprites} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilize the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
+It also checks for low order conditional independence relations before higher order ones.
+This allows it to exploit any sparsity in the true network, leading to much better average case performance  \cite{Sprites} (although the worst case, where the true network is complete, is still exponential).
+With finite data, the order in which the links are considered can change the output (unlike for SGS).
+The effect of wrongly removing a link early on flows through to later conditional independence tests by changing which nodes are considered adjacent.
+
+\begin{table}[H]
+ \begin{tabularx}{\textwidth}{X}
+ \hline
+\rule{0pt}{2.5ex} 
+ \textbf{The PC Algorithm}\\
+ \hline
+ \rule{0pt}{2.5ex}
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
+\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
+ \begin{enumerate}[itemsep=8pt]
+  \item As for SGS
+  \item \textbf{for} each link $a-b$:
+  \begin{itemize}[label={}]
+   \item $n = 0$
+   \item $\boldsymbol{A}_{a,b} = \{A_{1}...A_{j}\}$ be the set of nodes adjacent to $a$ and/or $b$
+   \item \textbf{while} $a$ and $b$ are connected and $n < j$:
+   \item 
+    	\begin{itemize}[label={}]
+    	\item \textbf{if} any subset of size $n$ of $\boldsymbol{A}$ makes $a$ and $b$ conditionally independent:
+    	\item \begin{itemize}[label={}]
+    			delete the link
+    		  \end{itemize}
+
+    	\item $n = n+1$
+    	\end{itemize}
+  \end{itemize}   
+  \item as for SGS
+  \item as for SGS
+\end{enumerate}\\
+ \hline
+\end{tabularx}
+\end{table}
+
+The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution.
+Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects with equation \ref{eq:allvarsobserved}.
+We can then bound the true causal effect by combining the results for the all the networks.
+This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
+An implementation is available in the R package \cite{Kalisch2012}
+
+\subsubsection{With hidden variables}
+There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables.
+With an unknown number of hidden variables there are an infinity many possible structures to search over.
+In addition, the space of causal networks is not closed under marginalization.
+If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalizing over $\boldsymbol{U}$, may not be faithful to any bayesian network (see figure \ref{fig:DAGSnotclosed}).
+
+%\begin{figure}
+%\centering
+%\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
+%\label{fig:DAGSnotclosed}
+% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+% %nodes
+%\node[main node](1){U}; 
+%\node[main node, below left = of 1](2){B};
+%\node[main node, below right = of 1](3){C};
+%\node[main node, above left= of 2](4){A};
+%\node[main node, above right = of 3](5){D};
+%
+% \path[every node/.style={font=\sffamily\small}]
+%    (1) edge node {} (2)
+%    	edge node {} (3)
+%    (4) edge node {} (2)
+%    (5) edge node {} (3);
+%    	
+%\end{tikzpicture}
+%\end{figure}
+
+The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables). See example figure XXX.
+
+\begin{theorem}
+\cite{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
+\end{theorem}
+
+Since we only care about the causal relationships between observed variables, it is sufficient to search over networks where any hidden variables have no parents and directly cause two of the observed variables.
+Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable.
+These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalization and conditioning.
+
+For any DAG with latent (and selection) variables there is a unique MAG \cite{Richardson2002}.
+This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \cite{Sprites}.
+The logic is very similar.
+Certain structures are ruled out by being inconstant with the observed conditional independence relations.
+The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG.
+Assuming there are no selection variables (see \cite xxx), the PAG can contain four types of link:
+
+\begin{enumerate}
+\item $X \rightarrow Y$, meaning $X$ causes $Y$
+\item $X \leftrightarrow Y$, meaning there is a latent variable that causes $X$ and $Y$.
+\item $X\ {\circ} {\rightarrow}\ Y$, either $X$ causes $Y$ or a latent variable causes both.
+\item $X\ {\circ} {-}  {\circ}\ Y$, either $X$ causes $Y$ or $Y$ causes $X$ or a latent variable causes both.
+\end{enumerate}
+
+The circles indicate where it is ambiguous if there should be an arrowhead (ie where there is one in some MAGs and not in others in the equivalence class).
+Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).
+
+%\begin{figure}
+%\centering
+%\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
+%\label{fig:FCIExamples}
+% \begin{subfigure}[t]{0.8\textwidth}
+% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+% %nodes
+%\node[main node](1){U}; 
+%\node[main node, below left = of 1](2){B};
+%\node[main node, below right = of 1](3){C};
+%\node[main node, above left= of 2](4){A};
+%\node[main node, above right = of 3](5){D};
+%
+% \path[every node/.style={font=\sffamily\small}]
+%    (1) edge node {} (2)
+%    	edge node {} (3)
+%    (4) edge node {} (2)
+%    (5) edge node {} (3);
+%    	
+%\end{tikzpicture}
+% \end{subfigure}
+% \begin{subfigure}[t]{0.8\textwidth}
+% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+% %nodes
+%
+%\node[main node](1){A};
+%\node[main node, below right=of 1](2){B};
+%\node[main node, right=of 2](3){C};
+%\node[main node, above right=of 3](4){D};
+%
+% \path[every node/.style={font=\sffamily\small}]
+%    (1) edge[pil] node {} (2)
+%    (2) edge node {} (3)
+%    (3) edge node {} (2)
+%    (4) edge [pil] node {} (3);    	
+%\end{tikzpicture}
+%\end{subfigure}
+% \begin{subfigure}[t]{0.4\textwidth}
+% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+% %nodes
+%
+%\node[main node](1){C};
+%\node[main node, above left=of 1](2){A};
+%\node[main node, above right=of 1](3){B};
+%\node[main node, below=of 1](4){D};
+%
+% \path[every node/.style={font=\sffamily\small}]
+%    (1) edge node {} (4)
+%    (2) edge node {} (1)
+%    (3) edge node {} (1);    	
+%\end{tikzpicture}
+%\end{subfigure}
+% \begin{subfigure}[t]{0.4\textwidth}
+% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+% %nodes
+%
+%\node[main node](1){C};
+%\node[main node, above left=of 1](2){A};
+%\node[main node, above right=of 1](3){B};
+%\node[main node, below=of 1](4){D};
+%
+% \path[every node/.style={font=\sffamily\small}]
+% 	(1) edge node {} (4)
+%    (2) edge [pil] node {} (1)
+%    (3) edge [pil] node {} (1);    	
+%\end{tikzpicture}
+%\end{subfigure}
+%\end{figure}
+
+The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \cite{Zhang2008}.
+More recently \cite{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
+\cite{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
+
+With latent variables we are not using all the information - so we could go further (to nested markov models and inequalities.) \cite{Richardson2012} \cite{Shipster2014}
+
+These are all constraint based methods ... efficient because they stop early, but also may not be robust to errors early on.
+
+A comparison of algorithms
+
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{| c | c | p{4cm} | c | c | c |}
+\hline
+  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
+  \hline
+  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
+  \hline
+  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
+  \hline
+  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
+  \hline
+  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
+  \hline
+  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
+  \hline
+  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
+ \hline
+\end{tabular}
+
+\subsubsection{Doing conditional independence tests}
+The off-diagonal elements of the standardized inverse of the correlation
+matrix are the negatives of the partial correlation coefficients between the corresponding
+variables given the remaining variables (see e.g. Whittaker, 1990). Hence in the linear case, the
+independence graph can be efficiently constructed by placing an edge between A and B if and
+only if the entry in the standardized inverse correlation matrix is non-zero. In the discrete case,
+Fung and Crawford (1990) have recently proposed a fast algorithm for constructing an
+independence graph from discrete data. We have not tested their procedure as a preprocessor for
+the PC algorithm. (COPIED FROM SPRITES)
+
+\cite{Zhang2012} Kernel independence tests
+
+HSIC \cite{Gretton2008}
+
+\subsection{Discovery with functional models}
+All of the algorithms we have considered so far return a Markov equivalence class.
+They cannot distinguish between two models that result in the same set of conditional independence relations.
+Consider the very simple case where we have only two variables and the only possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
+These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
+No algorithm relying purely on conditional independence relations can separate these two cases.
+
+Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$.
+What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from.
+Recall the functional definition of causality (section \ref{sec:SEM}).
+There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \ref{}, post-non-linear additive noise
+
+Linear models with non-gaussian noise \cite{Hoyer2012}
+
+Allow us to make assumptions about the form of the relationships between variables.
+With just two variables:
+
+Additive noise, \cite{Hoyer2009}
+
+post-non-linear additive noise \cite{Zhang2008a},
+
+causal anti-causal,
+
+deterministic functions,
+
+IGCI \cite{Janzing2012a}
+
+learning what causality looks like.
+
+If we have a method to solve this case we can solve the more general case (with no hidden variables) Putting things together to learn the full network.
+
+\subsection{Scoring}
+
+\chapter*{The interventionalist viewpoint (5000 words)}
+
+The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it.
+There is an obvious alternative.
+Instead of trying to infer the outcome of an intervention from passive observations we could just do it and see what happens.
+There are two key differences between observing a system and explicitly intervening in it.
+Firstly, when we intervene, we can choose which actions to take and thus get some control over which distributions we learn about.
+Secondly, if we are explicitly choosing interventions, we have a perfect model of the probability that we select each action given any context, allowing us to control confounding bias.
+
+\subsection*{The role of randomization}
+
+\paragraph*{The interventionalist viewpoint has devloped from the starting point of randomized controlled experiment}
 
@@
 
+Figure showing how randomization breaking any confounding links
+
+Randomization does not ensure target and control group are exactly alike.
+The more features or variables you include, the more likely that there will be a significant difference across at least one variable.
+But the within group variance also increases, the net effect is that it becomes harder to draw a conclusion but not more biased.
+
+\paragraph*{Limitations of randomization}
+
+\subsection*{Limitations of experimentation}
+
+Failure to generalize. Transportability, imperfect compliance. Too many contexts.
+
+\paragraph*{internal validity} Are differences in treatment and control groups down to intervention or the result of bias?
+
+\paragraph*{external validity} Are the results of the study applicable to the broader population of interest.
+
+\section*{Multi armed bandits}
+\paragraph*{Definition}
+
+\paragraph*{What limitations of randomized experiment does the bandit view address}
+
+\paragraph*{The exploration/exploitation trade-off}
+
+\subsection*{Regret} How do we measure the performance of a bandit algorithm?
+
+%$\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$
+
+Expected Regret
+\eqn{
+R_T(\pi) = \E{max_{i \in [k]} \sum_{t=1}^T{X_{t,i}}} - \E{\sum_{t=1}^T{X_{t,A(t)}}}
+}
+
+Psuedo-Regret
+\eqn{
+R_T(\pi) &= max_{i \in [k]}\E{\sum_{t=1}^T{X_{t,i}}} - \E{\sum_{t=1}^T{X_{t,A(t)}}} \\
+&= n\mu_{i^*} - \E{\sum_{t=1}^T{X_{t,A(t)}}}
+}
+
+Simple Regret
+\eqn{
+\mu_{i^*} - \E{\mu_{\hat i^*}}.
+}
+
+\subsection*{Adversarial Bandits}
+What problem do they solve?
+
+\subsection*{Key approaches and results}
+\paragraph*{optimisim in the face of uncertainty}
+
+\subsection*{The need to add structure}
+\paragraph*{The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.} This makes problems with large or infinite actions spaces intractible.
+
+\subsection*{Key extensions}
+
+\subsubsection*{Contexual bandits}
+\paragraph*{Feature selection} Is it even more important here? How do algorithms degrade as irrelevent features are added? In the supervised vs the RL setting? What would be a fair comparison?
+
+\subsubsection*{Markov decision processes}
+\subsubsection*{Dynamic Systems}
+\begin{itemize}
+\item An explicit model of actions in a partially known system (eg HMM)
+\item Feynman-Kac Lemma; Solving a PDE can be converted to a stochastic process
+\end{itemize}
+
 \chapter*{Causal Bandits: Unifying the approaches (5000 words)}
 
-\chapter*{Causal Inference \& Machine Learning (5000 words)}
-A more detailed discussion on where causal inference sits within machine learning and what it can offer.
+\section*{The framework}
+A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
+Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{} for a simple example.
+In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example only the set of single variable interventions.
+The reward could be a general function of the action selected and the final state of the graph.
+However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action.
+In this thesis I examine the case where the causal graph is known.
+We refer to these problems as \emph{causal bandit problems}.
+Extending this work to simultaneously learning the casusl graph is discussed in section \ref{XXX}.
 
-\section*{Different approaches}
-\paragraph{The two cultures}
+\subsubsection*{Definition of casual bandit problem} \todo{make proper definition}
+A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed actions} $\mathcal{A}$.
+One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
 
-\subsubsection*{Translating terminology}
-Economist -> ML
+\begin{figure}[h]
+\caption{A simple causal graphical model and corresponding complete action space. A and B represent binary variables that can be intervened on and Y represents the reward.}
+\label{fig:unify_frameworks}
+\centering
+\begin{subfigure}[c]{0.3\textwidth}
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
+\node[main node](1){A};
+\node[tg, below right=of 1](2){Y};
+\node[main node, above right=of 2](3){B};
+\path[]
+	(1) edge (2)
+	(3) edge (2);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.4\textwidth}
+Actions = \begin{tabular}{|c|}
+	\hline
+  do(A=0,B=0) \\
+  do(A=0,B=1) \\
+  do(A=1,B=0) \\
+  do(A=1,B=1) \\
+  \hline
+  do(A=0) \\
+  do(A=1) \\
+  do(B=0) \\
+  do(B=1) \\
+  do() \\
+  \hline
+\end{tabular}
+\end{subfigure}
+\end{figure}
 
-\section*{Challenges for the Machine learning approach}
-\subsection*{No cross validation}
-\paragraph*{The challenge of model selection}
-\paragraph*{Does predictive accuracy indicate a good causal model?}
-\subsection*{Less large, real world datasets}
-\paragraph*{The dearth of experimental data}
-\subsubsection*{Data for testing causal models}
-Simulators. Open competitions. Converting other data sets. Existing data sets.
+The number of actions or arms grows exponentially with the number of variables in the graph,
+making it important to use algorithms that leverage the graph structure to reduce the
+search space.
+Modelling a problem as a causal graph only makes sense when rewards are generated stochastically - since causal graphs fundamentally model probability distributions over variables.
+Thus the connection is to stochastic bandit problems (although adversarial bandits algorithms may be applied to stochastic problems).
 
-\section*{Relation to other areas of ML}
-\subsection*{Covariate shift}
-\subsection*{Generalizability}
-\paragraph*{Invariants}
-\paragraph{more stable} Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
-The assumption is there are less places for change to come in.
+We now need to specify the feedback model for the causal bandit problem.
+What information is available to the decision making agent before and after they select an action?
+The causal bandit problem takes on characteristics of different bandit settings depending on the assumptions we make about what actions are available to the agent, what variables are obvserved and whether they are observed before or after the action is chosen.
 
-\paragraph{change input distribution}If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.
+Enumerate the settings
+\begin{enumerate}
+\item Bandit feedback: the agent selects an action and then observes only the reward of that action.
+If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
+See figure XXX as an example.
+\item Bandit feedback with side information (context). The agent can view the value of some variables prior to selecting an action. After selecting they observe the reward of the selected node. 
+\item Post action feedback:
+\end{enumerate}
 
-\paragraph{feature selection} The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)?
-This may be a reason for using structure learning type algorithms even if you are simply doing prediction.
-\subsection*{Interpretabilty}
-\paragraph*{interpretable models as proxies for causal models} Let the human do the work.
-If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
+If we receive feedback on additional nodes, the problem can be more interesting.
+In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
+Consider again the model in figure \ref{fig:unify_frameworks}.
+The causal structure implies:
 
-A desire for interpretability indicates that something has been left out of the loss function.
+\eqn {
+P(Y|do(A = 0)) &= P(Y|do(),A = 0) \\
+&= P(Y|do(B=0),A=0)P(B=0)+P(Y|do(B=1),A=0)P(B=1) 
+}
 
-One form of interpretability gives people insight into what the features are that the model is relying on.
+Thus we gain information about the reward for the action $do(A=0)$ from selecting the action $do()$ or $do(B = b)$ and then observing $A = 0$.
 
-Specifically, people can
-\begin{itemize}
-\item rule out many possible features as highly unlikely to be relevent to a problem
-\end{itemize}
+We only get this form of side information for actions that don't specify the value of every variable, ie those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
+Since the reward distribution for actions that set a subset of the variables is the result of marginalizing out other variables, they can only be optimal if they have lower cost.
+So if the cost of all actions is constant (no matter how many variables must be set), then the problem has the same characteristics as if only the reward node were observable.
 
-People have access to a lot of detailed prior knowledge.  
-\subsection*{Transfer Learning}
-find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time).
-Causal models predict the outcome of actions.
-We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike.
-So we have to make representations such that things are stable.
+If the information on the value of additional nodes is available prior to selecting an action the problem resembles a contextual bandit.
+For example if we observe $A = 0$ then, in deciding between the actions $do(B=0)$ and $do(B=1)$, we would want information on $P(Y|A=0,B=0)$ and $P(Y|A=0,B=1)$.
+Note, side information can still arise if we learn the value of some variables prior to selecting an action and some afterwards.
 
-This is tightly related to generalizability.
-If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results.
-We don't include , the color of the doctors shirt, the size of the room, ...
+Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are possible.
+If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}.
+Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of allowable interventions thereby reducing the complexity.
 
-For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales.
-However, this relationship may be very unstable over time because the ad slots are sold at auction.
-The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword.
-However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.
-\subsection*{Fair Machine Learning}
+\section*{Related work}
+\begin{enumerate}
+\item leaning from log data
+\item bandits with imperfect compliance
+\item rigged casino paper
+\end{enumerate}
 
-\subsubsection*{Structural Equation Models}
+\begin{figure}[h]
+\caption{Example causal graph (based on \cite{Koller2009}) where the outcome of interest (reward) is cholesterol level .
+The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
+For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
+\label{fig:cholesterol_graph}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
+\node[node](1){Health concious};
+\node[node, below left=of 1](2){Diet};
+\node[node, below right=of 1](3){Exercise};
+\node[node, below right=of 2](4){Weight};
+\node[tg,below left=of 4](5){Y=Cholesterol};
+\node[node, below =of 5](6){Test result};
+\node[node, above right=of 3](7){Free time};
+\path[]
+	(1) edge (2) edge (3)
+	(2) edge (4) edge (5)
+	(3) edge (4)
+	(5) edge (6)
+	(7) edge (3);
+\end{tikzpicture}
+\end{figure}
 
-\subsubsection*{Identifiability}
+\section*{Causal bandits with post action feedback}
+WHY THIS PROBLEM. This work was presented at NIPS 2016 \ref{}.
+\todo{add relevent stuff from introduction section}
 
-The notion of identifiability is binary. How could it be softened. There are two obvious approaches. 
-1) Look at the finite time convergence for identifiable queries.
-2) For non identifiable queries, look at what bounds can be achieved or what additional assumptions would be required to make them identifiable.
+\subsection*{Notation}
 
-Question - why does everyone use the backdoor criterion. Is it much more frequent in the world. Easier to estimate?
+We will assume each variable only takes on a finite number of distinct values. (The path to relaxing this assumption would be through leverging the work on continuous armed bandits).
 
-\subsection*{Defifinitions}
-\subsubsection*{Interactions}
-\subsubsection*{Causal effects}
+The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
 
-\subsection*{Estimation of causal effects}
+An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
 
-\subsubsection*{Relationship to covariate shift}
+We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
+the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
 
-\subsection*{Discovering causal structure}
+\subsection*{Definition of causal bandit game with post-action feedback}
+\todo{ re-read and incorporate response from reviewer feedback for NIPS}
 
-\subsubsection*{Discovery based on conditional independence}
+The causal bandit game proceeds over $T$ rounds.
+In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations. 
+It then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, 
+including the \emph{reward} $Y_t \in \{0,1\}$. 
+After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T \in \actions$ based on its prior observations.
 
-\subsubsection*{Discovery with functional models}
+The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.
+
+We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
+The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
+Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.
+Intuitively though, one would expect to perform better by making use of the extra structure and observations.
+
+\subsection*{The parallel bandit problem}
+
+In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem which we call the {\it parallel bandit}.
+It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \ref{XXX}, including the farming example.
+It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
+
+The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:parallel}.
+All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
+
+In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high).
+The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.
+
+\begin{figure}
+    \begin{subfigure}[b]{0.34\textwidth}
+	\centering    
+          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+%every node/.style={scale=0.6}
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){$Y$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5);
+\end{tikzpicture}
+        \caption{Parallel graph}
+        \label{fig:parallel}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.2\textwidth}
+    \centering
+        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+\node[main node](1){$X_1$};
+\node[main node, below left=of 1](2){$X_2$};
+\node[main node, below right=of 1](4){$Y$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (2)
+    (1) edge (4)
+    (2) edge (4);
+\end{tikzpicture}
+        \caption{Confounded graph}
+        \label{fig:causalStructure_confounded}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, right=of 4](5){$Y$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (2)
+  	(2) edge (3)
+    (3) edge (4)
+    (4) edge (5);
+\end{tikzpicture}
+        \caption{Chain graph}
+        \label{fig:causalchain}
+    \end{subfigure}
+    \caption{Causal Models}\label{fig:causalmodels}
+    \vspace{-.5cm}
+\end{figure}
+
+In each round the learner either purely observes by selecting $do()$ or sets the value of a single variable.
+The remaining variables are simultaneously set by independently biased coin flips.
+The value of all variables are then used to determine the distribution of rewards for that round.
+Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
+
+The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
+$r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
+In the farming example, this choice of $Y$ models the success or failure of a seasons crop, 
+which depends stochastically on the various environment variables.
+
+\paragraph{The Parallel Bandit Algorithm}
+%\label{sub:par-bandit-alg}
+The algorithm operates as follows.
+For the first $T/2$ rounds it chooses $do()$ to collect observational data.
+As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
+Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
+The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates of their returns could be poor.
+To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions.
+The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$).
+For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$.
+Define
+
+\eq{
+\label{eq:m-simple}
+m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
+}
+
+\begin{algorithm}[H]
+\caption{Parallel Bandit Algorithm}\label{alg:simple}
+\begin{algorithmic}[1]
+\STATE {\bf Input:} Total rounds $T$ and $N$.
+\FOR{$t \in 1,\ldots,T / 2$}
+\STATE Perform empty intervention $do()$
+\STATE Observe $\vec{X}_t$ and $Y_t$
+\ENDFOR
+\FOR{$a = do(X_i = x) \in \actions$}
+\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
+\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
+\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
+\ENDFOR
+\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
+\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
+\FOR{$a = do(X_i = x) \in A$}
+\FOR{$t \in 1,\ldots,T_A$}
+\STATE Intervene with $a$ and observe $Y_t$
+\ENDFOR
+\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
+\ENDFOR
+\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
+\end{algorithmic}
+\end{algorithm}
+
+$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret.
+When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
+We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$.
+Although very simple, the following two theorems show that this algorithm is effectively optimal.
+
+\begin{theorem}\label{thm:uq-simple}
+Algorithm \ref{alg:simple} satisfies
+\eq{
+\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.
+}
+\end{theorem}
+
+\begin{theorem}\label{thm:lower}
+For all strategies and $T$, $\vec{q}$, there exist rewards such that
+$\displaystyle \simpleregret 
+\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)$.
+\end{theorem}
+
+The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
+
+The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
+of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
+
+%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
+By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
+Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$.
+Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting.
+In practice, you would combine the data from both phases to estimate rewards for the low probability actions.
+We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.
+
+\subsection*{General graphs}
+We now consider the more general problem where the graph structure is known, but arbitrary.
+For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
+However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \citep{Pearl2000}.
+\eq{
+\P{X_1...X_N|do(X_i=j)} = 
+\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
+} 
+where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
+
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to 
+write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
+estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
+from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
+takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
+simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
+high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
+Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
+get a good estimate for $\P{Y|do(X_2=1)}$.
+
+To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
+allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
+are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
+or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
+start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
+Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
+Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
+
+\begin{algorithm}[H]
+\caption{General Algorithm}\label{alg:general}
+\begin{algorithmic}
+\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
+\FOR{$t \in \set{1,\ldots,T}$}
+\STATE Sample action $a_t$ from $\eta$
+\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
+\ENDFOR
+\FOR{$a \in \calA$}
+\STATE
+\eq {
+\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
+}
+\ENDFOR
+\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
+\end{algorithmic}
+\end{algorithm}
+
+Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
+Let $\parents{Y}(X)$ denote the realization of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
+
+\eq {
+\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}\,, 
+}
+
+where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently.
+The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
+
+The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
+\eq{
+m(\eta) = \max_{a \in \calA} \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,,\text{ where } \EEa \text{ is the expectation with respect to } \Pn{a}.
+}
+
+We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
+
+\begin{theorem}\label{thm:general}
+If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$
+
+\eq{
+\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
+}
+\end{theorem}
+
+The proof is in Section \ref{sec:thm:general}.
+
+Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
+Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$.
+Thus it has broader applicability to the parallel graph than the generic algorithm given here.
+We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
+for future work.
+
+\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$.
+In light of Theorem \ref{thm:general}
+a natural choice of $\eta$ is the minimiser of $m(\eta)$.
+\eq{
+\eta^* 
+= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
+}
+Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
+Therefore the minimisation problem may be tackled using standard techniques from convex optimisation.
+The quantity $m(\eta^*)$ may be interpreted as the minimum achievable worst-case variance of the importance weighted estimator.
+In the experimental section we present some special cases, but for now we give two simple results.
+The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
+
+\begin{proposition}\label{pro:m-bound}
+$m(\eta^*) \leq |\calA|$. \textit{Proof.} 
+\textup{By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|\,\forall a$.}
+\eq{
+m(\eta) 
+= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
+\leq \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
+= \max_a \EEa\left[\frac{1}{\eta_a}\right] = |\calA| %\qedhere
+}
+\end{proposition}
+
+The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$.
+This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
+The proof is in section XXX.
+
+\begin{remark}\label{rem:truncate}
+The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
+often possible to choose $B_a$ significantly
+larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
+and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
+to learn from observational data.
+\end{remark}
+
+\subsection*{Experiments}
+We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \cite{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
+Thomson sampling and UCB are optimized to minimize cumulative regret.
+We apply them in the fixed horizon, best arm identification setting by running them upto horizon $T$ and then selecting the arm with the highest empirical mean.
+The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.
+
+Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms).
+$Y_t \sim \bernoulli(\frac{1}{2}+\epsilon)$ if $X_1=1$ and $Y_t \sim \bernoulli(\frac{1}{2}-\epsilon')$ otherwise, where $\epsilon' = q_1\epsilon/(1-q_1)$.
+This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
+We set $q_i = 0$ for $i \leq m$ and $\frac{1}{2}$ otherwise.
+Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
+For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
+The code is available from \url{<https://github.com/finnhacks42/causal_bandits>}
+
+In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
+The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure.
+For the causal algorithms it increases approximately with $\sqrt{m}$.
+As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.
+
+\begin{figure}[h]
+    \begin{subfigure}[t]{0.3\textwidth}
+		\centering    
+    		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
+    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
+        \label{fig:simple_vs_m}
+    \end{subfigure}\hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+    		\centering
+        \includegraphics[width=\textwidth]{experiment2_20161020_1249.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
+        \label{fig:simple_vs_T_vary_epsilon}
+    \end{subfigure}\hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+    		\centering
+    		\includegraphics[width=\textwidth]{experiment3_20161020_1252.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
+    		\label{fig:simple_vs_T}
+    \end{subfigure}
+    \caption{Experimental results}
+    \label{fig:experiments}
+\end{figure}
+
+Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
+This gap is learnable by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
+In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
+The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.
+
+For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
+This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be minimax optimal asymptotically \citep{LMS14}.
+
+\subsection*{Additional experiments} \todo {tie section together}
+\begin{figure}[H]
+	\centering    
+          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
+  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+%every node/.style={scale=0.6}
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, right=of 1](2){$X_{2}$};
+\node[hidden, right=of 2](3){$...$};
+\node[main node, right=of 3](4){$X_{N}$};
+\node[main node, below right=of 2](5){$Y$};
+\node[main node, above right=of 2](6){$Z$};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (5)
+    	(2) edge (5)
+    (4) edge (5)
+    (6) edge (1) edge (2) edge (4);
+\end{tikzpicture}
+        \caption{Confounded graph}
+        \label{fig:parallel_confounded} 
+\end{figure}
+
+We now compare the general algorithm with a range of standard bandit algorithms on the confounded graph in Figure \ref{fig:parallel_confounded}.
+All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action, $\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}$.
+We choose this setting because it generalises the parallel bandit, while simultaneously being sufficiently simple that we can compute the exact reward and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
+As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.
+
+In Figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$. For some $2 \leq N_1 \leq N$ we define 
+\eq{
+P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
+P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
+}
+As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon P(X_1=1) / P(X_1=0)$.
+The value of $N_1$ determines $m$ and ranges between $2$ and $N$.
+The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
+This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
+
+In Figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.  
+In Figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1, but exclude actions on $Z$ from the set of allowable actions to demonstrate that Algorithm 1 can fail in the presence of a confounding variable, which occurs because it incorrectly assumes that $P(Y|do(X)) = P(Y|X)$.
+We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
+
+\eq{
+P(X_i = 1|Z = 0) &= 
+\begin{cases} 
+.166 & \text{ if } i \in \set{1,..., 6} \\
+.2 & \text{ if } i = 7 \\
+.7 & \text { otherwise} 
+ \end{cases}\\
+ P(X_i = 1|Z = 1) &= 
+\begin{cases} 
+.166 & \text{ if } i \in \set{1,..., 6} \\
+.8 & \text{ if } i = 7 \\
+.3 & \text { otherwise} 
+ \end{cases}\\
+}
+
+In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree.
+It is sub-optimal to act on either $X_7$ or $X_N$, while all other actions are optimal.
+The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.
+All remaining variables are likely to be identified as balanced and estimated from observational estimates.
+The CPD values have been chosen to demonstrate the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
+
+\begin{figure}[H]
+    \begin{subfigure}[t]{0.3\textwidth}
+		\centering    
+    		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
+    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
+        \label{fig:simple_vs_m_general}
+    \end{subfigure}\hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+    		\centering
+        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
+        \label{fig:simple_vs_T_general}
+    \end{subfigure}\hfill
+    \begin{subfigure}[t]{0.3\textwidth}
+    		\centering
+    		\includegraphics[width=\textwidth]{experiment5_20161023_2118.pdf}
+    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m(\eta*)=4.3$ with no actions setting $Z$}
+    		\label{fig:simple_vs_T_misspecified}
+    \end{subfigure}
+    \caption{Experimental results on the confounded graph}
+    \label{fig:experiments_confounded}
+\end{figure}
+
+\subsection*{Discussion \& Future work}
+
+Algorithm~\ref{alg:general} for general causal bandit problems 
+estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
+Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ where 
+the parameter $m(\eta)$ measures the difficulty of learning the causal model and is always less than $N$.
+The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.
+
+The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator $\hat{\mu}_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $\Pn{a}{\parents{Y}}$ for all $a \in \calA$.
+In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.
+Despite these extra estimates, Theorem~\ref{thm:lower} shows that this approach is optimal (up to log factors).Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
+
+\paragraph{Making Better Use of the Reward Signal}
+Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
+In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
+Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
+This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
+
+For the parallel bandit the required modifications should be quite straightforward.
+The idea would be to adapt the algorithm to essentially use successive elimination in the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability.
+In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
+Note that these modifications will not improve the minimax regret, which at least for the parallel bandit is already optimal.
+For this reason we prefer to emphasize the main point that causal structure should be exploited when available.
+Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
+This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.
+
+\paragraph{Cumulative Regret}
+Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
+In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information
+to get near-optimal cumulative regret guarantees.
+They consider a finite-armed bandit model with side information where in reach round the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
+In this way the learner can gain information about actions it does not take with varying levels of accuracy.
+The reduction follows by substituting the importance weighted estimators in place of the Gaussian reward.
+In the case that $\vec{q}$ is known this would lead to a known variance and the only (insignificant) difference is the Bernoulli noise model.
+In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
+at least asymptotically.
+
+%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
+%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
+%\todom{Check this!}
+%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
+
+The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
+The feedback graph is distinct from the causal graph.
+A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
+For this parallel bandit problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
+However, \citet{Alon2015,Kocak2014} give adversarial algorithms, which when applied to the parallel bandit problem obtain the standard bandit regret.
+A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
+This is equivalent to a nominally stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
+
+\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
+In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
+\cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
+
+\paragraph{Causal Models with Non-Observable Variables}
+If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
+non-observable variables. Some of the interventional distributions may be non-identifiable meaning we can not obtain prior estimates for $\Pn{a}{\parents{Y}}$ from 
+even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
+\ref{alg:general} cannot be used. Estimating these quantities while simultaneously minimising the simple regret is an interesting and challenging open problem.
+
+% For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
+% In this case, some conditional distributions may be non-identifiable. 
+% The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
+% We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
+% Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
+
+% More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
+In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
+The quality of the estimate notably depends not only on the number of times that action was selected.
+For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$.
+This more complex side information does not fit within the graph feedback framework.
+
+\paragraph{Partially or Completely Unknown Causal Graph}
+A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
+The latter case arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques. 
+Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014b} could potentially be incorporated into a causal bandit algorithm.
+In particular, \citet{Hu2014b} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
+Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational dataset would be much more challenging.
+
+\chapter*{Causality in Marketing}
+\section*{Attribution}
+\section*{Econometric Modelling}
 

### 2017-04-30 thesis/thesis_map.tex
bootcamp sprint3

@@
 \usepackage[colorinlistoftodos]{todonotes}
-\usepackage{natbib}
+\usepackage[numbers]{natbib}
 \usepackage{wrapfig}
@@
 \DeclareUnicodeCharacter{00A0}{ }
+
+%\tikzset{
+%    %Define standard arrow tip
+%    >=stealth',
+%    %Define style for boxes
+%    punkt/.style={
+%           circle,
+%           rounded corners,
+%           draw=black, thick,
+%           text width=1em,
+%           minimum height=1em,
+%           text centered},
+%    punktrect/.style={
+%    rectangle, 
+%    rounded corners, 
+%    % fill=black!10,
+%    draw=black, thick,
+%    minimum height=1em, 
+%    text centered},
+%    % Define arrow style
+%    pil/.style={
+%           o->,
+%           thick,
+%           shorten <=2pt,
+%           shorten >=2pt,}
+%}
+
 \tikzset{
@@
     %Define style for boxes
+    punkt/.style={
+           circle,
+           rounded corners,
+           draw=black, thick,
+           text width=1em,
+           minimum height=1em,
+           text centered},
     observed/.style={
@@
 
+There is a connection here to the degree to which the model is invariant.
+A model that is invariant to more things is more explainitory.
+We could look at levels of hierachy.
+There is causal graphical models, which are invarient to interventions on variables.
+Can I tie this to a physical system and give more indepth examples of the nested tier of hieriachies?
+
+I do not see the distinction between explaination and (causal) prediction.
+Explaination is all about the ability to compress and to generalize.
+The more a model can do this, the more we view as providing an understanding of the why.
+
+Sometimes the causal component is obvious.
+
+Relationship to generalization.
+Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
+The assumption is there are less places for change to come in.
+
+If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.
+
+The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)?
+This may be a reason for using structure learning type algorithms even if you are simply doing prediction.
+
+\subsubsection*{The role of assumptions in causal inference}
+
+Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.
+
+Does predictive accuracy indicate a good causal model?
+
+Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.
+
+Generalizing the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parashutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalize).
+(and given sufficient generalizability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)
+
+Is causal inference possible (are the conclusions ever valid)?
+
+Only from randomized experiment ... if people can do it without recource to constant randomized experiment then an algorithm exits.
+
+The challenges and limitations of infering causal effects from observation data are numerous.
+However, there are problems we need to solve.
+The questions is are these problems so serious as to warrent discregarding all non-experimental data.
+I think answer is a clear no.
+Need ongoing validation, with assumptions.
+We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).
+
+\subsection*{Relationship to interpretability}
+
+A desire for interpretability indicates that something has been left out of the loss function.
+
+Are causal/interpretable models more reliable or more likely to generalize well?
+
+One form of interpretability gives people insight into what the features are that the model is relying on.
+
+If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
+
+Specifically, people can
+\begin{itemize}
+\item rule out many possible features as highly unlikely to be relevent to a problem
+\end{itemize}
+
+People have access to a lot of detailed prior knowledge.
+
+\subsection*{Relationship to transfer learning}
+
+find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time).
+Causal models predict the outcome of actions.
+We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike.
+So we have to make representations such that things are stable.
+
+This is tightly related to generalizability.
+If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results.
+We don't include , the color of the doctors shirt, the size of the room, ...
+
+For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales.
+However, this relationship may be very unstable over time because the ad slots are sold at auction.
+The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword.
+However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.
+
 Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.
@@
 The two cultures \cite{Breiman2001}
+
+\paragraph*{There are two reasons why correlation is not causation}.
+The first is related to variance and over-fitting.
+Obversations are noisy.
+With a finite dataset with enough variables we will be able to find some that are completely unrelated but correlate purely by chance.
+\ref{figure - search for random correlation}.
+The second arises from bias, typically introced by an un-observed confounding variable.
+In this case, variables are correlated not by chance.
+We would expect the relationship to hold if we sampled more data.
+However, they are not causally related in that intervening to set one would not likely effect the other.
+EXAMPLE WITH FIGURE.
 
@@
 \subsubsection*{Examples of typical machine learning probems. Are they causal?}
+
+Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to.
+Which of them require casual inference?
+How can we identify characterists of a prolem that make causal modeling important?
 
@@
 
-%For image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see.
-In fact this is because we already have one.
-We can be confident that the process will not change.
+The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.
+
+Consider speech recognition.
+You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words.
+Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a dvd on elocution).
+A similar arguament could be made for many applications of machine translation and image classification.
+
+In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see.
+We can be fairly confident that the process will not change.
 If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
+
+What about forecasting the weather?
+If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decicion will not effect if it actually rains.
+However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
+For this we need a (causal) model that allows us to predict the outcome under various different interventions.
+
+Identifing spam and automated essay marking systems are similar.
+The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label.
+Spammers and students will modify their writing in order to optimise thier results.
+The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided.
+(What would the nature of the features have to be such that change did not occur?
+- they would have to be causes of the label).
+
+What about predicting the risk of death in patients with pnumonia?
+Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics.
+If we assume that in hospital treatment is more effective this seems like a straighforward prediction problem.
+It is not.
+Depending on how the decision to admit was previously made and what features are included (or ommited) in the model, the relationship between those features and the outcome may change if we start using the model to decide whome to admit.
+(xxx et al) found exactly this effect.
+Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pnemonia.
+They realised this was because doctors were treating such patints very agressively, thus acutally lowering their risk.
+There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols.
+However, using it to decide on what basis to admit people could kill.
+The key is understanding exactly what question you are asking.
+In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.
+
+Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc).
+For all these problems the assumptions required to treat them as pure prediction problems are;
+\begin{enumerate}
+\item The treatment is assumed to be effective (at least better than nothing)
+\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
+\end{enumerate}
+
+Demand prediction seems relatively straightforward.
+These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
+It seems unlikely that using the model to ensure stock is available will itself change demand.
+However there is a potential data sensoring issue.
+If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
+Changing availability does then change demand.
+
+Playing Go (and other games) is another case with some subleties.
+At every turn, the AI agent has a number of actions available.
+The board state following each action is deterministic and given by the rules of the game.
+The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win.
+...
+this is interesting maybe come back to it ...
+an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ...
+One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken).
+When can we learn from the actions others have taken?
+When there is no confounding.
+And does this hold with Go?
+Probably because the board state encapsulates everything that should determine what move is played.
+Learning directly from actions and trying to generalize (can in some instances reduce the problem to standard ML)
+
+Having considered these examples we can now identify some general aspects of the problem that determine wheter or not we require a causal model.
+
+\begin{itemize}
+\item Does acting on the predictions of the model change the mapping from features to target?
+(at least if the decision process is open to scrutiny).
+In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
+\item
+
  Covariate shift clearly comes in here.
@@
 The mechanism is known but the function may be complex.
-
-\subsubsection*{What aspects of a problem determine if causal inference is required?}
-(When is pure prediction useful?)
-\begin{itemize}
+Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models?
+Yes, if the way in which features get their values changes, then causal models will be invarient to that in a way that non-causal ones are not.
+
 \item To decide between actions we only need to rank them (not estimate their actual effect). 
@@
 \end{itemize}
+
+I hope these examples gave you a feel for the richness and subtlties of causal inference.
+We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
 
@@
 If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.
-%
-%\begin{figure}[h]
-%\centering
-%\caption{General Bayesian network over 3 variables}
-%\label{fig:bayesnet}
-%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-%  thick,main node/]
-%
-% %nodes
-%\node[main node](1){$X_{1}$};
-%\node[main node, below left=of 1](2){$X_{2}$};
-%\node[main node, below right=of 1](3){$X_{3}$};
-%
-%
-% \path[every node/.style={font=\sffamily\small}]
-%    (1) edge node {} (2)
-%    	edge node {} (3)
-%    (2) edge node {} (3);
-%	
-%\end{tikzpicture}
-%\end{figure}
+
+\begin{figure}[h]
+\centering
+\caption{General Bayesian network over 3 variables}
+\label{fig:bayesnet}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, below left=of 1](2){$X_{2}$};
+\node[main node, below right=of 1](3){$X_{3}$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3);
+
+\end{tikzpicture}
+\end{figure}
 
@@
 GIVE EXAMPLE IN SOMETHING QUANTIFIABLE LIKE BITS.
+TAKE THE GRAPH IN WITH THREE NODES.
+ASSUME EVERYTHING IS BINARY AND.
+FIGURE OUT HOW MANY VARIABLES ARE NEEDED TO DESCRIBE ALL POSSIBLE INTERVENTIONS VS THE BAYESIAN NETWORK.
 This may be surprising since their conditional probability distributions are estiated from data in the same way.
@@
 A number of critisims have been leveled at this way of modelling causality.
-
-A complaint leveled against this view point of causality is that the 'surgery' is too precise and that, in the real world, any intervention will effect many variables (eg Cartwright 2007). However,
+One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will effect many variables in complex and non-determinisitic ways \ref{causality in complex interventions,Cartwright 2007}.
+However, the deterministic $do()$ opperater turns out to be surprisingly effective at modelling more complex interventions.
+One simply has to add additional variables to the graph that model how our interventions may actually take effect.
+For example, our drug treatment case assumed we set the drug that each person recieves.
+But what if some people in fact fail to take the prescribed treatment.
+How can we model this within the framework of deterministic interventions?
+We simply add a node representing what we prescribed to them (which we intervene on), which probabilistically influences what treatment they actually recieve.
+See figure XXXX
+
+Insert figure of randomized treatment with imperfect compliance here.
 
@@
 - cyclic/dynamics
-
-Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.
-
-- add variables for soft interventions
 - unroll to deal with cycles
@@
 \subsection*{Structural Equation models}
-
-There is a problem with the term non-parametric. It is used in machine learning in a slightly different way to pearl
 
@@
 But as soon as the patient walks though the door  then she should offer treatment B, regardless of which gender the patient turns out to be.
+
 How does causaul modelling reslove this paradox?
@@
 
-The causal network for this problem is shown below.
+The causal network for this problem is shown in figure XXX.
 In this example, women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they recive than men.
@@
 However, their position in the network is assumed to be known.
-For example if we were trying to estimate ...
+For example if we were interested in the effect of years of schooling on wages, we might expect that some measure of inherent ability could influence both the number of years of schooling people chose to pursue and the wages they subsequently obtaine on graduating.
+Even if we have no data to directly measure peoples inherent ability, becuase it influences two of the variables were are modelling, we must include it in the graph.
 
@@
 
-\subsection*{The do calculus}
-Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting?
-
-The do calculus arises from d-seperation.
-
-How do we take our causal frameworks and use them to map information from the observational to interventional settings?
+\subsection*{Mapping from observational to interventional distributions}
+Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting to the interventional one.
+Recall that a bayesian network is a way of representing the joint distribution over its variables in a way which highlights the any conditional idependencies between them.
+
+\begin{theorem}
+\textbf{(Local markov condition)} Given a bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decendents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
+\end{theorem}
+
+The set of conditional independence relations given by the local markov condition can enforce additional independences that also hold in all distributions that are compatible with $G$.
+D-separation is an algorithm that extends the local markov property to find these additional independences.
+It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
+
+By leveraging conditional independencies we can effectivly localize the effect of an intervention to a specific part of a larger graph.
+This gives rise to the do-calculus \cite{Pearl2000}.
+The do calculus consists of three rules.
+They derive from the causal information encoded in a causal network and the properties of d-separation and do not require any addition assumptions other than that of specifying the causal network.
+
+\subsubsection*{Rule 1}
+This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention.
+The intervention $do(\boldsymbol{X}=\boldsymbol{x})$ changes a causal network, $G$, in a simple way.
+Variables in $\boldsymbol{X}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{x}$.
+This corresponds to deleting the edges with arrows into variables in $\boldsymbol{X}$  (see figure \ref{fig:mutilatednet}).
+The resulting 'mutilated' network $G_{\overline{\boldsymbol{X}}}$ remains a causal network and d-separation still applies.
+
+If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
+\begin{equation}
+\label{eq:Do1}
+ P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
+\end{equation}
+
+$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
+\begin{equation}
+\label{eq:Do12}
+ P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) 
+\end{equation}
+
+This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.
+
+\begin{figure}[h]
+\centering
+\caption{Intervention in a causal bayesian network}
+\label{fig:mutilatednet}
+\begin{subfigure}[t]{0.3\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:mutilatedOriginal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 2](4){$W$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3)
+    (4) edge node {} (2);
+
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.5\textwidth}
+\centering
+\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
+\label{fig:mutilatedAfter}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$x$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 2](4){$W$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (3)
+    (2) edge node {} (3);
+
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+\subsubsection*{Rule 2}
+Rule 2 states when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening  $do(\boldsymbol{X}=\boldsymbol{x})$ have the same effect on the distribution of the outcome $\boldsymbol{Y}$.
+You can think of this as when correlation is causation.
+It is easiest to understand by explicitly including the intervention process in the graphical model.
+We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on (figure \ref{fig:augmentedNetwork}).
+Let $\epsilon$ be some arbitrary value not in the set of possible values for $X$.
+If $X = \epsilon$ the distribution of $X$ is what it was without intervention.
+Otherwise, if $X=x$, $X$ deterministically takes the value $x$ and is independent of its previous parents, representing the intervention $do(X=x)$.
+We use the notation $G^{\dagger}$ to represent $G$ augmented with these decision nodes.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
+
+\begin{equation}
+\label{eq:Do2}
+P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w})
+\end{equation}
+
+$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X})$ in $G^{\dagger}$:
+\begin{equation}
+\label{eq:Do22}
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x})
+\end{equation}
+
+If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.
+
+\begin{figure}[h]
+\centering
+\caption{Augmentated causal network with intervention}
+\label{fig:augmentedNetwork}
+\begin{subfigure}[t]{0.4\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:augmentedOriginal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left = of 1](4){$U_{1}$};
+\node[main node, above right = of 1](5){$U_{2}$};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(2) edge node {} (3)
+ 	(4) edge node {} (2)
+ 		edge node {} (1)
+ 	(5) edge node {} (3)
+ 		edge node {} (1)
+ 	(1) edge node {} (2)
+ 		edge node {} (3);
+
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.4\textwidth}
+\centering
+\caption{augmented network after the intervention $do(Z=z)$, $G^{\dagger}_{\overline{Z}}$}
+\label{fig:augmentedAfter}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.3cm,
+  thick,main node/.style={punkt}]
+
+\node[main node](1){$z$};
+\node[below left=of 1](2){};
+\node[below right=of 1](3){};
+\node[main node, below left=of 2](4){$X$};
+\node[main node, below right=of 3](5){$Y$};
+\node[above right=of 1](6){};
+\node[above left=of 1](7){};
+\node[main node, above left=of 7](8){$U_{1}$};
+\node[main node, above right=of 6](9){$U_{2}$};
+\node[main node, above left=of 4](10){$\hat{X}$};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(4) edge node {} (5)
+ 	(10) edge node {} (4)
+ 	(8) edge node {} (4)
+ 	(9) edge node {} (5)
+ 	(1) edge node {} (4)
+ 		edge node {} (5);
+
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+\subsubsection*{Rule 3}
+This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
+\begin{equation}
+\label{eq:Do3}
+P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
+\end{equation}
+$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}})$ in $G^{\dagger}$:
+\begin{equation}
+\label{eq:Do32}
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y})
+\end{equation}
+The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ without conditioning on $\boldsymbol{X}$ implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ which starts on an arrow leaving $\boldsymbol{X}$.
+This means there is no causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$ and the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ does not alter the distribution of $P(\boldsymbol{Y})$.
 
@@
 
-a) determine if the query can be translated into an expression involving only distributions over observed variables.
+\begin{enumerate}
+\item determine if the query can be translated into an expression involving only distributions over observed variables.
 In other words, determine if the query is identifiable given the assumptions encoded by the network
-
-b) if it is identiable, return the required expression
+\item if it is identiable, return the required expression 
+\end{enumerate}
 
 \todo{insert link to working javascript version}
+\todo{figure with typical examples}
 
@@
 We now move to the much more general problem of learning a causal graph from observational data.
-Amazingly, it is possible to infer some aspects of causal structure with very general assumptions.
+In this setting we make much broader assumptions about the structure of the graph.
+For example, that it is acyclic or that we have no unmeasured confounding variables.
+We do not assume the existance or directions of any links between the variables.
+Amazingly, it is possible to infer some aspects of causal structure with such general assumptions.
 The set of conditional independences in a non-experimental data set indicates some causal structures are more likely than others.
-In addition, there are subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
-These clues are the key to causal discovery algorithms, which attempt to learn causal structure from non-experimental data.
-
-\paragraph*{You want to learn the graph}
-In the causal discovery problem we make much much broader assumptions about the structure of the graph.
-Ie that it is acyclic.
-and may or may not have latent variables.
-We do not assume the existance or directions of any links between the variables.
-
-What's more, we wish to do this from non-experimental data (although there is also work on infering causal structure from experimental data) REFERENCES and it would be natural to connect the two (ANY REFERENCES THERE?).
-
-This is a much grander goal than causal inference and arguably, if acheived would equate to the automation of scientific discovery. 
-\paragraph*{Equates to the aim of automating scientific discovery}
-\paragraph*{Increadibly hard}
-It it is amazing that anything can be done here.
-\paragraph*{Methods can also be divided into contraint based and search and score}
-\subsection*{Discovery with Conditional independence}
-If variables are (conditionally) independent they are unlikely to be directly causally linked. 
-\subsection*{Discovery with Functional Models}
-\subsection*{Granger causality}
-
-In previous sections we discussed when assumptions about the structure of the variables in a specific problem is sufficient to identify a causal effect.
-This approach relies on having enough prior knowledge or theory about the problem to allow you to, at least partially, specify the causal network.
-In this section, we consider the much harder problem of causal inference where you need to learn the network.
-Causal inference might seem impossible without specific assumptions about the structure of the variables involved but, amazingly, some aspects of causal structure can be determined from much more general assumptions.
-
-In this section we consider variants of the following problem: Assume there is some acyclic causal network $G$ that generated the distribution $P(\boldsymbol{V})$ from which our data has been sampled.
-Our goal is to recover the network from this data.
-
-What can we look for in the distribution that could give us clues as to the structure of the network?
-
-\subsection{Causal discovery with conditional independence tests}
+In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
+These clues are the key to causal discovery algorithms.
+
+There is also work on infering causal structure from experiental data \ref{}. We discuss this further in section XXX.
+
+Causal discovery is a much grander goal than causal effect estimation given a known causal network.
+Arguably, if acheived, it would equate to the automation of scientific discovery.
+We need simply supply our algorithm with a vast collection of variables (regardless of their relevence to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make.
+Unfortunately, causal disovery is very hard.
+Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows XXX exponentially (or super-exponentially)?
+with the number of variables.
+- some intuition here about the strength of the signal we get from causal disovery.
+Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}.
+Talk about the numerical as well as the statisitcal issues here.
+
+In the next sections we briefly survey the key approaches to causal discovery.
+We roughly divide the methods into those based on those that exploit the connection between the conditional indpendences in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
+This is not the only way we could have grouped them.
+There is also a seperation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
+There are also hybrid methods \ref{}
+
+\subsection*{Conditional independence based methods}
 One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
-We know, (section \ref{sec:dsep}), if $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in $P$ .
+For any bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network.
+If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$.
 However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
-This requires that we assume the reverse condition:$(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
-This assumption, commonly referred to as  \textbf{\textit{faithfulness}}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
-Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} for $P$.
+This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
+This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
+Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \cite{Kollner} for $P$.
 
@@
 It would disappear under a small perturbation to any of the parameters.
+\todo{point out violations are solutions to polynomial equations and thus measure 0 \ref{XX}}
 
@@
 
-The SGS algorithm utilizes the fact that a collider structure (f
-igure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
+The SGS algorithm utilizes the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
 Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
@@
 
+\todo{mention failed replication - if published}
+
 \subsubsection{With hidden variables}
 There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables.
-With an unknown number of hidden variables there are an infinity many possible structures to search over.
+With an unknown number of hidden variables there are infinity many possible structures to search over.
 In addition, the space of causal networks is not closed under marginalization.
@@
 
-%\begin{figure}
-%\centering
-%\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
-%\label{fig:DAGSnotclosed}
-% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
-% %nodes
-%\node[main node](1){U}; 
-%\node[main node, below left = of 1](2){B};
-%\node[main node, below right = of 1](3){C};
-%\node[main node, above left= of 2](4){A};
-%\node[main node, above right = of 3](5){D};
-%
-% \path[every node/.style={font=\sffamily\small}]
-%    (1) edge node {} (2)
-%    	edge node {} (3)
-%    (4) edge node {} (2)
-%    (5) edge node {} (3);
-%    	
-%\end{tikzpicture}
-%\end{figure}
+\begin{figure}
+\centering
+\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
+\label{fig:DAGSnotclosed}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){U}; 
+\node[main node, below left = of 1](2){B};
+\node[main node, below right = of 1](3){C};
+\node[main node, above left= of 2](4){A};
+\node[main node, above right = of 3](5){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    (5) edge node {} (3);
+
+\end{tikzpicture}
+\end{figure}
 
@@
 This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \cite{Sprites}.
-The logic is very similar.
-Certain structures are ruled out by being inconstant with the observed conditional independence relations.
+The logic behind the algorithm is very similar.
+Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations.
 The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG.
@@
 
-%\begin{figure}
-%\centering
-%\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
-%\label{fig:FCIExamples}
-% \begin{subfigure}[t]{0.8\textwidth}
-% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
-% %nodes
-%\node[main node](1){U}; 
-%\node[main node, below left = of 1](2){B};
-%\node[main node, below right = of 1](3){C};
-%\node[main node, above left= of 2](4){A};
-%\node[main node, above right = of 3](5){D};
-%
-% \path[every node/.style={font=\sffamily\small}]
-%    (1) edge node {} (2)
-%    	edge node {} (3)
-%    (4) edge node {} (2)
-%    (5) edge node {} (3);
-%    	
-%\end{tikzpicture}
-% \end{subfigure}
-% \begin{subfigure}[t]{0.8\textwidth}
-% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
-% %nodes
-%
-%\node[main node](1){A};
-%\node[main node, below right=of 1](2){B};
-%\node[main node, right=of 2](3){C};
-%\node[main node, above right=of 3](4){D};
-%
-% \path[every node/.style={font=\sffamily\small}]
-%    (1) edge[pil] node {} (2)
-%    (2) edge node {} (3)
-%    (3) edge node {} (2)
-%    (4) edge [pil] node {} (3);    	
-%\end{tikzpicture}
-%\end{subfigure}
-% \begin{subfigure}[t]{0.4\textwidth}
-% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
-% %nodes
-%
-%\node[main node](1){C};
-%\node[main node, above left=of 1](2){A};
-%\node[main node, above right=of 1](3){B};
-%\node[main node, below=of 1](4){D};
-%
-% \path[every node/.style={font=\sffamily\small}]
-%    (1) edge node {} (4)
-%    (2) edge node {} (1)
-%    (3) edge node {} (1);    	
-%\end{tikzpicture}
-%\end{subfigure}
-% \begin{subfigure}[t]{0.4\textwidth}
-% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
-% %nodes
-%
-%\node[main node](1){C};
-%\node[main node, above left=of 1](2){A};
-%\node[main node, above right=of 1](3){B};
-%\node[main node, below=of 1](4){D};
-%
-% \path[every node/.style={font=\sffamily\small}]
-% 	(1) edge node {} (4)
-%    (2) edge [pil] node {} (1)
-%    (3) edge [pil] node {} (1);    	
-%\end{tikzpicture}
-%\end{subfigure}
-%\end{figure}
+\todo{put these figures in a table with true graph input on left and fci output on right}
+\todo{uniform styles used for nodes accross thesis}
+\begin{figure}
+\centering
+\caption{FCI examples: true graph and fci output}
+\label{fig:FCIExamples}
+ \begin{subfigure}[t]{0.8\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){U}; 
+\node[main node, below left = of 1](2){B};
+\node[main node, below right = of 1](3){C};
+\node[main node, above left= of 2](4){A};
+\node[main node, above right = of 3](5){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    (5) edge node {} (3);
+
+\end{tikzpicture}
+ \end{subfigure}
+ \begin{subfigure}[t]{0.8\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+ %nodes
+
+\node[main node](1){A};
+\node[main node, below right=of 1](2){B};
+\node[main node, right=of 2](3){C};
+\node[main node, above right=of 3](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge[pil] node {} (2)
+    (2) edge node {} (3)
+    (3) edge node {} (2)
+    (4) edge [pil] node {} (3);    	
+\end{tikzpicture}
+\end{subfigure}
+ \begin{subfigure}[t]{0.4\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+ %nodes
+
+\node[main node](1){C};
+\node[main node, above left=of 1](2){A};
+\node[main node, above right=of 1](3){B};
+\node[main node, below=of 1](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (4)
+    (2) edge node {} (1)
+    (3) edge node {} (1);    	
+\end{tikzpicture}
+\end{subfigure}
+ \begin{subfigure}[t]{0.4\textwidth}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+ %nodes
+
+\node[main node](1){C};
+\node[main node, above left=of 1](2){A};
+\node[main node, above right=of 1](3){B};
+\node[main node, below=of 1](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(1) edge node {} (4)
+    (2) edge [pil] node {} (1)
+    (3) edge [pil] node {} (1);    	
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
 
@@
 \subsubsection{Doing conditional independence tests}
-The off-diagonal elements of the standardized inverse of the correlation
-matrix are the negatives of the partial correlation coefficients between the corresponding
-variables given the remaining variables (see e.g. Whittaker, 1990). Hence in the linear case, the
-independence graph can be efficiently constructed by placing an edge between A and B if and
-only if the entry in the standardized inverse correlation matrix is non-zero. In the discrete case,
-Fung and Crawford (1990) have recently proposed a fast algorithm for constructing an
-independence graph from discrete data. We have not tested their procedure as a preprocessor for
-the PC algorithm. (COPIED FROM SPRITES)
-
-\cite{Zhang2012} Kernel independence tests
-
-HSIC \cite{Gretton2008}
-
-\subsection{Discovery with functional models}
-All of the algorithms we have considered so far return a Markov equivalence class.
+
+All the algorithms we have discussed in the previous section rely on being able to perform conditional independence tests.
+This is non-trivial with high dimensional data.
+If the functional relationship between the variables is linear with gaussian noise then the network represents a multivaitae normal distribution and a pair of variables $A$ and $B$ are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero.
+Where the functions are non-linear
+
+\begin{itemize}
+\item \cite{Zhang2012} Kernel independence tests
+\item HSIC \cite{Gretton2008}
+\end{itemize}
+
+\subsection*{Discovery with functional models}
+The algorithms we have considered so far return a Markov equivalence class.
 They cannot distinguish between two models that result in the same set of conditional independence relations.
@@
 Recall the functional definition of causality (section \ref{sec:SEM}).
-There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \ref{}, post-non-linear additive noise
-
-Linear models with non-gaussian noise \cite{Hoyer2012}
-
-Allow us to make assumptions about the form of the relationships between variables.
-With just two variables:
-
-Additive noise, \cite{Hoyer2009}
-
-post-non-linear additive noise \cite{Zhang2008a},
+There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a}, linear models with non-gaussian noise \cite{Hoyer2012}
 
@@
 
+%If X → Y, what is the relationship between P(X) and P(Y|X)? 
+What about P(Y) and P(X|Y)?
+
+%In semi supervised learning we are given training points from P(X,Y) and additional points sampled from P(X). The goal is to learn P(Y|X). The extra points from P(X) will not help if X → Y.
+
 deterministic functions,
@@
 
+Indepdendence of input and mechanism. Daniusis et al 2012
+
+Independence of function and input: If $X \rightarrow Y$ and we have a functional causal model $y = f(x,e)$ then the input
+distribution $P(X)$ and function $f$ represent independent mechanisms. We do not expect a change in the input distribution to modify the function.
+
 learning what causality looks like.
-
-If we have a method to solve this case we can solve the more general case (with no hidden variables) Putting things together to learn the full network.
-
-\subsection{Scoring}
+Lopez-Paz et al 2014.
+
+Instead of positing a functional restriction on the relationship between variables and then developing theory to exploit that assumption, \citep{LopezPaz2014} propose learning what the causal relationship looks like from data.
+They assume there will be a difference between the relationship of $\P{X}$, $\P{Y}$ and $\P{X|Y}$ between $X \rightarrow Y$ versus $Y \rightarrow X$.
+Their algorithm requires a data set in which each row is itself a data set consisting of paris of varibles $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$.
+They use a keneral mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-dataset and train an algorithm to learn the direction of causality.
+Unfortunately we do not have a large collection of datasets where the causal direction is known to train such a model.
+\citep{LopezPaz2014} instead use a simulated dataset so thier model will nessesarily be base on the assumptions they make when generating the data.
+Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.
+Their approach performed well in the causal effect pairs challange, correctly identifing the causal direction in ??
+percent of cases (against a benchmark of 50\% from random guessing)\ref{XXX}.
+
+\citep{Peters2014} have extended results from the bivariate case to the multivariate setting (with no hidden variables)?.
+They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable.
+The build on this to develop an algorithm that allows the construction of causal graphs based on the additive assumption??
+
+\subsection*{Granger causality}
+
+Granger causality was developed within economics and is not typically discussed within the causal discovery liturature.
+
+Although it does differ from the previous approaches in the types of assumption it relies upon it none-the-less tackles the same goal
+
+so we shall include it here.
+
+Granger causalty essentially assumes that the future cannot cause the past.
+This may seem an entirely waterproof assumption.
+However it can be tripped up by unmeasured confounding variables.
+The cock crows before the sun rises - so does it cause the sun to rise?
+No.
 
@@
 \subsection*{The role of randomization}
-
-\paragraph*{The interventionalist viewpoint has devloped from the starting point of randomized controlled experiment}
-
-What is the role of randomization? How do bandits algorithms work despite being only partially randomized? 
-What else can you do to improve randomized studies (variance reduction, lower regret).
-
-Figure showing how randomization breaking any confounding links
+Randomized controlled trials are often presented as the gold standard for determining causal effects in many areas of social sciences and economics \ref{XXX}.
+What is it about randomization that makes it so important when it comes to causality?
+The graphical model for a randomized controlled experiment is shown in figure \ref{fig:random_experiment_network}.
+If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
+Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.
+
+\begin{figure}[h]
+\caption{causal network for a randomized experiment}
+\label{fig:random_experiment_network}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
+\node[node](1){Treatment};
+\node[node, above left=of 1](2){Randomiser};
+\node[node, right=of 1](3){Outcome};
+\node[lt, above left=of 3](4){U};
+\path[]
+	(2) edge (1)
+	(1) edge (3)
+	(4) edge (3);
+\end{tikzpicture}
+\end{figure}
+
+\todo {turn this into something}
+This would be a natural place to discuss what happens with finite datasets where there are many other variables that influence the outcome.
+We expect that across at least some of these variables, the target and control group will not be balanced.
+Does this not inject bias into the causal estimates?
+The anser is no, (I think - a better more general proof would be nice).
+The more such attibutes there are, the higher is the variance in the outcome within the target and control group.
+When we perform a confidence test to ascertain whether or not there is a difference ...
+what happens.
 
 Randomization does not ensure target and control group are exactly alike.
-The more features or variables you include, the more likely that there will be a significant difference across at least one variable.
-But the within group variance also increases, the net effect is that it becomes harder to draw a conclusion but not more biased.
-
-\paragraph*{Limitations of randomization}
-
-\subsection*{Limitations of experimentation}
-
-Failure to generalize. Transportability, imperfect compliance. Too many contexts.
-
-\paragraph*{internal validity} Are differences in treatment and control groups down to intervention or the result of bias?
-
-\paragraph*{external validity} Are the results of the study applicable to the broader population of interest.
+The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables in a between the target and control groups in a finite data sample.
+However, the variance in the outcome within both the target and control groups also increases.
+The net result is increased variance (but not bias) in  the estimate of causal effects.
+
+DISCUSS STRATIFIED RANDOM EXPERIMENT
+
+What is the role of randomization? How do bandits algorithms work despite being only partially randomized? What else can you do to improve randomized studies (variance reduction, lower regret).
+
+The benifit provided by randomization in breaking the link between the treatment variable and any confounders should not be understated.
+The possibility of confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (test for confounding).
+This means causal estimates from non-experimental data are always subjec to the criticism that an important confounder may have been overlooked or not properly adjusted for.
+However, randomized experimentation does have some limitiations.
+
+\subsubsection*{Limiations of randomized experiments}
+\label{subsec:limitations_of_experiment}
+\begin{enumerate}
+
+\item The probability that an individual is treated is set in advance and fixed over the course of the experiment.
+As the experiment proceeds we may start to see evidence that one treatment is much more effective than another.
+However (in a pure randomized experiment) we don't use this information to update the treatment that later patients recive.
+The result is that some will unessearily be given a sub-optimal treatment.
+\item Experiments are expensive and difficult to conduct. This means experimental datasets are often much smaller than observational ones, limiting the complexity of models we can explore.
+
+\item Experiments are often conducted on a convinient but unreprasentaive sample of the broader population of interest (for example first year collage students at research universities).
+This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}.
+The results may not carry over to the general population of interest.
+The question of whether an experiment conducted on one population can be mapped to another is refered to as the transportability problem \citep{Bareinboim2013} \todo{check ref} and relies on very similar assumtions and arguments to causal inference and the do-calculus.
+
+\item The idealized notion of an experiement represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomized experiments in practice.
+There may be imperfect compliance: the treatment selected by the randomizer is not always followed.
+Or output censoring: the experimenter is not able to observe the outcome for all units (for example if people drop out).
+If complience or attrition is not random but associated with (potentailly latent) variables that also effect the outcome then the problem of confounding bias returns.
+See figure XXX for a the graphical model of a randomized experiment with imperfect compliance and outcome censoring.
+
+\item It is not always possible or ethical to conduct a randomized experiment \cite{parashutes} \ref{Parachute use to prevent death and major trauma related to gravitational challenge: systematic review of randomised controlled trials}.
+\end{enumerate}
+
+\begin{figure}[h]
+\begin{center}
+\caption{Experiments are not always ethical}
+\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
+\end{center}
+\end{figure}
 
 \section*{Multi armed bandits}
+\todo{is there a way in latex to have a command for refering to each type of thing like figure, subsection, etc such that it can be updated programatically?}
+
+Multi-armed bandits were originally introduce by XXXX to address problem that it is sub-optimal to fix a probability of treatment in advance as we discussed in subsection \ref{subsec:limitations_of_experiment}.
+The problem can be formally stated as follows:
+
 \paragraph*{Definition}
-
-\paragraph*{What limitations of randomized experiment does the bandit view address}
-
-\paragraph*{The exploration/exploitation trade-off}
 
@@
 
-\subsection*{Adversarial Bandits}
-What problem do they solve?
+\todo{connect bandit regret to early stopping of experiment problem}
+
+Scholars in the 1940's found this problem so frustrating that they suggested it be dropped over Germany to waste their time \cite{XXX}.
+
+\paragraph*{The exploration/exploitation trade-off}
 
@@
 
-\subsection*{Key extensions}
+\subsection*{Key related settings}
+
+\subsubsection*{Adversarial Bandits}
+In the classical multi-armed bandit problem, the rewards for each arm are sampled stochastically from a fixed (but unknown) distribution.
+There is a large body of work analysing a more general formulation of the bandit problem that relaxes this stochastic assumption.
+
+\paragraph*{Definition of the Adversarial bandit problem}
+\begin{itemize}
+\item Different notation of regret
+\item Adversarial bandits are not imume to issues of feedback cycles or drift in the reward distribution because a constancy is implicit in the definition of regret. 
+\end{itemize}
 
@@
 
+\subsection*{Proofs}
+
+\chapter*{Causality \& Fairness}
+\begin{itemize}
+\item There is an additional connction between causality and fairness, that arises when you know that there is a bias in the data and wish to correct for it.
+In causality, it may be due to selection bias or a confounding variable.
+In fairness, it may be due to historical disadvanage.
+In both cases we may have theory that tells us the direction or scale of an effect.
+Ie, we know that race itself has no inherent effect on outcomes such as criminality and education.
+If you have a dataset and a collection of such priors, where the data contradicts these priors due to data bias.
+How do you best make decisions based on that data?
+\end{itemize}
+
 \chapter*{Causality in Marketing}
@@
 
+In the discrete case,
+Fung and Crawford (1990) have recently proposed a fast algorithm for constructing an
+independence graph from discrete data. We have not tested their procedure as a preprocessor for
+the PC algorithm. (COPIED FROM SPRITES)
+
 \bibliographystyle{apalike}% Select the citation style e.g. ieeetr

### 2017-04-30 thesis/thesis_map.tex
bootcamp final sprint

@@
 \subsection*{Estimation}
-\subsection*{Defining causal effects} Summarising the difference between two distributions. There is no one answer. \cite{Janzing2013}
-
-\cite{Imbens2004} Review of non-parametric estimation
-\cite{causal estimation is covariate shift}
-\paragraph*{How well do we actually do with finite datasets?}
-
+\subsubsection*{Defining causal effects} 
+So far we have described causal effect estimation in term of idenfitifying the interventional distribution $\P{Y|do(X)}$ from observational data.
+
+From a decision making process, is it the right thing to do to just pick the X that maximises the expectation of $Y$ (assuming that higher $Y$ is better.
+If it is not we can just add in another variable that is a deterministic function of $Y$ to relect the loss function.)
+
+Typically however we are often interested in estimating some function of the difference between $P{Y|do(X)}$ for various values of $X$.
+For example, the average causal effect (ATE) is defined for discrete binary treatment $T$ by:
+
+\eqn{
+ATE = \E{Y|do(T = 1)} - \E{Y|do(T=0)}
+}
+
+If the variables are continouous the average casual effect is
+
+\todo{insert equation for continous case}
+
+The individual causal effect as defined in section XXX cannot be stated in terms of the do notation as it is not with respect to a distribution over anything.
+However, we can define a very similar concept with respect some observed context which I will refer to as the personalized causal effect.
+
+\todo{insert equation for personalised causal effect}
+
+However, the difference in expectation is not the only way of summarizing the difference between two (or more) distributions.
+\citep{Janzing2013} show that there is no single way of summarising the collection of distributions represented by $\P{Y|do(X)}$ that satistifies a number of reasonable axioms.
+
+An obvious alternative to subtracting the expections would be to consider ratio.
+Is there any reason that we don't do this that relates to the arguament in Dawid?
+What about the smoothness of any uncertainties?
+
+\subsubsection*{Estimating the difference when the back door critera is met}
+
+A very substantial number of practial studies fall into the cateogy of causal inference via the backdoor theorem.
+A study is done.
+It is assumed that it is ignoreable.
+That is that we have a a set of variables satifying the back door adjustment criteia.
+The goal is to estimate the average causal effect.
+Since we have assumed the backdoor criteia is met, the query is identifiable.
+Problem solved.
+Except we have finite data ...
+
+What are the approaches for doing so?
+
+There are multiple ways of doing the adjustment.
+
+In order to block the path of spurious information it is sufficient to (perfectly) model either the treatment selection process $\P{T|Z}$ or the response function $\P{Y|T,Z}$.
+
+\paragraph*{matching}
+
+\paragraph*{regression}
 \paragraph*{When is regression causal?}
@@
 
+\paragraph*{propensity scores}
+
+They have advantages and disadvantes depending on the setting. Methods like regression typically impose a parametric model of some description onto the treatment assignment or response function.
+
+- Connection of neighest matching to neighest neighbour.
+
+Once we have made the assumptions required to define the causal graph and identify a valid adjsutment set, is there any difference between the causal effect estimation problem and standard supervised machine learning.
+
+\cite{causal estimation is covariate shift}
+
+However, there is a critial remaining complication.
+In supervised learning we can split our data into training and validation data sets or apply cross validation to select models and model parameters.
+In the causal estimation case we cannot.
+We lack samples from the counterfactual.
+
+The siginificance to this should not be underestimated.
+Cross-validation has allowed applied machine learning to proceed with very atheoretical approach on the basis that we can identify when a model is sucessful.
+With causal effect estimation there is no guarantee that a model which performs well at prediction (even out of sample) will accuratly esitmate the outcome of an intervention.
+
+GIVE A CONCRETE EXAMPLE here.
+One would be a model which looks at a post-treatment variable in order to make predictions.
+But what about if we assume the model structure is correct and known.
+Can we fit parameters that do better on prediction but worse on casual effect estimation?
+
+\cite{Imbens2004} Review of non-parametric estimation
+
+Discuss the recent competition on this question and the results
+
+Discuss the role that the models used to simulate the data may have in determining which models perform the best.
+
 \subsection*{Non identifiable queries}
-
-\subsubsection*{Parametric assumptions}
-eg linearity
-
-\subsubsection*{Bounds}
-\paragraph{Instrumental variables are an example}
-
-Put in the picture of an instrumental variable.
-
-Instrumental variables are one of the key approaches used to estimate causal effects within economics and the social sciences.
-However, the setting they are associated with is non-identifiable.
-How is it that they are so useful nonetheless?
-Tight bounds may be just as good as what could be achieved in the identifiable setting, given finite data.
+In many cases the query of interest will not be identifiable.
+In this setting we are left with a choice between making additional assumptions (for example that variables are linearly related) or aiming to obtain bounds rather than point estimates for the causal effect of interest.
+Or we may do both.
+One of the most widely used methods in this category is the instrumental variable.
+
+Instrumental variables are often conceptualised as natrual experiments.
+The goal is to find a variable that substantially influences the treatment but that is conditionally independent of the remaining variables in the network.
+
+For example, we might want to compare the results for juvenile offenders depending on wheter they spent their time awaiting trial in a custom facility for young offenders as opposed to a unit within an adult jail.
+The process by which children are assinged to different facilities may well be non-random.
+However, it might be that in many cases the decison was based purely on the fact that the juvenile facilty was full.
+We could then use the number of places remaining in the juvenile facility as an instrumental variable.
+
+The causal network for an ideal instrumental variable is shown in figure XXX.
+It is not identifiable (non-parametrically) because.
+If we assume linearity it ?
+\todo{what exact guarentees do we get for instrumental variables}.
+Without a linearity assumption we would want the instrumental variable to be highly predictive of the treatement assignment so as to obtain reasonalble bounds.
+But if we make the assumption that everthing is linear then it would seem that even a weak instrumental variable should (in theory) surfise.
+
+In some cases we might have a choise between adjuting for causal effects via the backdoor theorem or utilizing an instrumental variable to obtain bounds.
+It is not the case that the former method is always superior even though it has better asymtpoptic results.
+For example
+
+DEMONSTRATE a specific example where instrumental variables obtain better estimates than adjustment even when all the criteria for adjustment have been met.
+Start with a case were the adjustment is very high dimensional and the instrumetal variable is strong.
+(the effect of the varibles on the treatment assingment will have to be week (otherwise the instrument is not strong).
+
+\subsection*{How successful overal is causal effect estimation}
+Really know one knows.
+
+There are relatively few studies that have compared the observational study results with those obtained by randomized experiments.
+One of the key examples is the XX job network study.
+Initial analysis suggested that a range of causal estimation technquies all ...
+
+\paragraph*{How can we construct datasets on which to test casual effect estimation techniques}
+One can start from a randomized experiment and deliberately drop some of the data. Draw up the graphical model for this process and the assumptions inherent in it.
 
@@
 
-causal anti-causal,
+There is a connection between casual discovery and semi-supervised learning.\cite{causal anti causal}.
+Suppose we are trying to learn $\P{Y|X}$.
+The goal of semi-supervised learning is to improve our estimate of $\P{Y|X}$ by leveraging additional data sampled from $\P{X}$.
+However, if the true causal model is $X \rightarrow Y$ then their is some function mapping values of $X$ to $Y$ which should be invariant to any changes in the input distribution $\P{X}$.
+Therefor $\P{X}$ should be indepdend of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning.
+However if the true causal model is $Y \rightarrow X$ then variations in the $P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help.
+
+ONE POINT I'M NOT CERTAIN ON HERE.
+might not knowledge of $\P{X}$ help define which regoin of the mapping we need to be most accurate on in order to miniize the loss (ie ananalgous to the covariate shift probelelm).
+\todo {open question}
 
@@
 distribution $P(X)$ and function $f$ represent independent mechanisms. We do not expect a change in the input distribution to modify the function.
-
-learning what causality looks like.
-Lopez-Paz et al 2014.
 
@@
 
-Multi-armed bandits were originally introduce by XXXX to address problem that it is sub-optimal to fix a probability of treatment in advance as we discussed in subsection \ref{subsec:limitations_of_experiment}.
+Multi-armed bandits were originally introduce by XXXX to address the problem that it is sub-optimal to fix a probability of treatment in advance as we discussed in subsection \ref{subsec:limitations_of_experiment}.
 The problem can be formally stated as follows:
@@
 \paragraph*{Definition} 
+In its classic formulation \cite{Robbins1952} the (stochastic) K-armed bandit describes a sequential decision making problem, with $K$ possible actions or arms.
+Each arm $i$ is associated with a fixed but unknown reward distribution $\nu_i$.
+In order to obtain regret bounds, some assumptions are required on the distributions $\nu_i$.
+It sufficient to assume they are sub-gaussian.
+For each timestep $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$.
+The goal of the learner is to maximize the total reward they receive.
+This problem introduces the fundamental exploration-exploitation trade-off.
+The learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
+
+The performance of bandit algorithms is generally described by the (pseudo) regret, $R(T)$.
+This is the difference between the expected reward obtained by the algorithm and the expected reward of selecting the best action in every timestep.
 
@@
 
+An algorithm is learning if it obtains regret that is sublinear in $T$.
+
+Scholars in the 1940's found this problem so frustrating that they suggested it be dropped over Germany to distract the scientists there from the war effort.\cite{XXX}.
+
+Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
+In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
+The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm, $R_s(T) = \mu^* - \E{\mu_i}$ .
+This problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialization or exploitation phase.
+For example, many strategies might be assessed via simulation prior to one being selected and deployed.
+The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}.
+
 \todo{connect bandit regret to early stopping of experiment problem}
-
-Scholars in the 1940's found this problem so frustrating that they suggested it be dropped over Germany to waste their time \cite{XXX}.
-
-\paragraph*{The exploration/exploitation trade-off}
+The problem of when to stop an experiment early can also be phrased as a bandit problem and analysed with the same techniques.
+XXX et al showed that if use an experimental setup with early stopping, the regret you obtain will be at least twice that you could have achieved had you made each decision sequentially using a bandit approach.
+Of course this neglect any costs that might be associated with the complexity in implementation of the two approaches.
+
+\subsection*{The exploration/exploitation trade-off}
+A central aspect to bandit problem is the trade of between exploiting the information we have already obtained with exploring options about which we remain uncertain.
+Suppose you have sampled five of your local restuarants and found one which you really enjoyed.
+How often should you eat there versus exploring new options.
+The degree to which you should explore versus exploit depends on how many step remain to go in the game and what assumptions you make about the distributions from which the rewards are sampled.
+If this is your last meal you might as well just eat the food you enjoy most but if you hope to live for many meals to come then you stand to gain by sampling a wider range of options.
+In many cases, the total number of rounds that you will get to play in the game is unknown.
+If you cannot imagine that it would be possible for a meal to be substantially better than that which you obtain (in expectation) at your current favoriate then there is little point exploring further.
 
 \subsection*{Key approaches and results}
-\paragraph*{optimisim in the face of uncertainty}
+
+The lower bound on the worst case regret for any algorithm (stochastic or adversarial) for the K-armed bandit problem is  $\bigomega{\sqrt{TK}}$ \cite{Auer1995}.
+
+A key algorithm for stochastic bandits, with tractable analysis and strong performance guarantees, is the UCB algorithm \cite{Auer2002}.
+They key to this algorithm is that it keeps track of an upper confidence bound (hence UCB) on the expected reward for each arm and selects the arm with the highest one.
+This balances exploration and exploitation as an arm with a high upper confidence bound must have either a high expected reward or large uncertainty on the expected reward.
+Assume for notational simplicity that $\mu_1 > \mu_2 > ...
+> \mu_K$, such that $\mu^* = \mu_1$, and let $\Delta_i = \mu_i - \mu^*$ be the sub-optimality for each arm.
+The (problem dependent) regret for UCB is bounded by:
+
+\eqn {
+R^{ucb}(T) \in \bigo{\sum_{i=2}^K \frac{1}{\Delta_i}\log(T)}
+}
+
+This bound blows up as differences $\Delta_i \rightarrow 0$, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
+In the worst case, $R^{ucb}(T) = \bigo{\sqrt{TK\log(T)}}$ \cite{Bubeck2012}.
+Subtle modifications to the UCB algorithm can eliminate the logarithmic term in this worst case regret bound.
+This yields $R^{ucb}(T) = \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.
+
+One key principle that has motivated the design of a number \todo{ask Tor are there actually others aside from UCB?} of key bandit algorithms is the idea of \emph{optimism in the face of uncertainty}.
+An algorithm is trying to simultaneously obtain large rewards and extract as much information as possible.
+If an action has a high upper bound on the possible reward then either the expected reward must be high or the algorithm must be very uncertain about it.
+Choosing it should yeild either a good reward or useful information.
+This is the key idea begind the UCB algorithm \citep{xxx}.
+The remaining complexity lies in optimising how we compute confidence intervals to trade of exploration and exploitation as the game proceeds.
+Subtle variations in how the confidence intervals are defined lead to a number of algorithms, which are optimised to slightly different settings.
+SETTINGS NOT QUITE RIGHT WORD.
+\citep{xxx}.
 
 \subsection*{The need to add structure}
-\paragraph*{The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.} This makes problems with large or infinite actions spaces intractible.
+The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.
+This makes problems with large or infinite actions spaces intractible.
+REFERENCE equation showing linearity with arms.
+CHECK ITS TRUE.
+
+The classic multi-armed bandit is a powerful tool for sequential decision making.
+However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalized linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
 
@@
 There is a large body of work analysing a more general formulation of the bandit problem that relaxes this stochastic assumption.
+
+Adversarial bandits are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
+Instead, simultaneously with the learner selecting an action $I_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{g}_t$.
+As in the stochastic setting, the learner then receives reward $g_{I_t,t}$.
+The seminal algorithm for adversarial bandits is Exp-3, which, like UCB, obtains regret $\bigo{\sqrt{TK\log(T)}}$ regret \cite{Auer1995}.
+Optimal algorithms, with $R(T) = \bigo{\sqrt{TK}}$, have also been demonstrated for the adversarial setting \cite{Audibert2009}.
 
@@
 \subsubsection*{Contexual bandits}
-\paragraph*{Feature selection} Is it even more important here? How do algorithms degrade as irrelevent features are added? In the supervised vs the RL setting? What would be a fair comparison?
+In the previous sections each decision was identical.
+If we were imaging treating patients, we assume that we have no additional information about each patient that is relevent to deciding how to treat them.
+However, in many key applications such as serving ads, etc, etc.
+we clearly do have additional information that we wish to take into account.
+In the contextual bandit setting the goal is to learn
+
+\eq {
+\P{Y|a,x}
+}
+
+As opposed to the standard bandit setting where we are attempting to estimate
+
+\eq{
+\P{Y|a}
+}
+
+The formal defintion of a contextual bandit problem is
+
+The simplest way to generalise the standard bandit algorithms to the contextual case (where the context is discrete) is to simply have a seperate standard bandit instance for each possible setting of the context.
+Regretably this approach scales terribly with increasing complexity of the context.
+\todo{quantify this}.
+It cannot be applied at all if the context is continous.
+
+The approach to solving this problem parallels those from supervised learning and continous armed bandits.
+We make assumptions about the smoothness of the problem.
+Values of context that are similar should lead to comparable rewards for a given action.
+
+Regret bounds \todo {insert regret bounds for contextual bandits}
+
+It should be noted that there is an important difference between the approaches we can apply to the contextul bandit versus the supervised learning setting.
+As for the casual effect estimation problem, we cannot utilize cross-validation offline to select model parameters.
+This makes contenxtual bandit algorithms degrade faster than supervised learning ones as we add irrelevent variables.
+\todo{ask Tor if he thinks this is correct}
+
+RUN A SIMULATION COMPARIGN HOW supervised learning and contextual bandits degrade as we add more features. What would be a fair comparison?
 
 \subsubsection*{Markov decision processes}
+In the bandit setting the reward distributions for each arm are assumed to be fixed.
+A much studied generalisation is the markov decision process or MDP.
+In an MDP we assume there the enviroment has some state.
+The reward an agent recieves depends on the state of the enviroment and the action the agent selects.
+The enviroment then evolves stochasitcally depending on the agents action.
+Multi-armed bandits can be considered a single state MDP, where no matter what action the agent selects, the enviroment returns to that state.
+
+Insert a graphical representation of an MDP
+
+What assumptions are required to make MDPs tractable.
+
+What is the basic algorithm (Q-learning)
+
+What are the key results.
+
 \subsubsection*{Dynamic Systems}
@@
 
+\chapter*{List of software programs for causal inference}
+
+\begin{itemize}
+\item Pcalg. A library in R that implements ...
+\item Tetrad.
+\item There has to be a library for the bayesian trees thing
+\item Linear regression (you can do this anywhere)
+
+\end{itemize}
+
 \chapter*{Causality \& Fairness}
+
+As machine learning is incorporated into decision making systems that have fundamental impacts on people’s lives, such as in employment, criminal justice, health and financial services, there are increasing concerns over transparency and fairness.
+
+The European Union’s General Data Protection Regulation \cite{XXXX}, due to come into effect in 2018, requires that people can obtain "meaningful information about the logic involved" in an automated decision process.
+It also stipulates that such decisions should not be based on special categories of personal data (related to ethnicity, political and religions affiliation or sexuality) unless there "suitable measures" to safeguard individual interests.
+A key concern is discrimination against disadvantaged groups.
+
+Discrimination is frequently defined in terms of either ’disparate treatment’- treating otherwise similar individuals differently on the basis of a protected attribute such as race or gender, or ’disparate impact’ - a substantial difference in outcome between groups.
+We consider how the notions of disparate treatment and impact may be formalised and the implications of how this is done for machine learning.
+We examine the overlap between the motivations for interpretable and causal models, especially with respect to assessing fairness.
+We look at how causal models mitigate some of the trade-offs between transparency and predictive accuracy and we examine to
+what extent the causal relationship between an attribute, any protected attributes and the outcome of interest is relevant to assessing the impact on fairness of its inclusion in a machine learning model.
+
 \begin{itemize}
@@
 \chapter*{Causality in Marketing}
+Assessing the impact of marketing spend is becoming increasinly important to industly.
+Billions of dollars \cite{XXX} are spent each year and new marketing channels and oportunities are openening up at an unprecented rate.
+
+There are three key approaches to assessing the impact of marketing in the liturature.
+
+The first is to look for correlations across companies on metrics such as how much they spend on marketing \ref{}, how confident they are about their ability to market effectively \ref{} or how important the marketing team is percieved to be within the orgainisation with indicators of the health of the company such as sales, revenue or market capitalisation.
+Unfortuneatly is is very difficult to avoid potentail major confounders with this approach.
+Are businesses large and succesful because of their capable marketing teams or can they afford to hire capable marketers because they have sufficient profits available to do so.
+
 \section*{Attribution}
+The second is attribution.
+The goal in this case is to attribute each sale to a specific (or set of specfic) advertisements that the customer was exposed to.
+This approach is the primary way of measuring the sucess of digital advertising strategies such as search, display, and online video ads.
+The most frequent model is last-click attribution.
+In which the last ad the customer was presented with before they made their purchase is assumed to have caused that sale.
+A central criticism of attribution is that it is not measuring incremental sales.
+Each sales is assumed to be a consequence of some form of advertising for those customers who saw (however briefly) some form of advertisicment.
+Of course in reality many of these customers may have made a purshcase regardless of the advertising material to which they were exposed.
+
+System designed to optimsie metrics such as click through or conversion rate can do so either by learning how to change the consumers preferences (by serving them the perfect ad at the ideal time) or by learning who was most likely to buy anyway.
+It seems likley that much of the improvemt in click through rates due to the application of sophisticated machine learning algorithms is beign driven by the latter.
+Unfortuenatly, this does nothing for the bottom line of industry (except for those involved in servering ads).
+
 \section*{Econometric Modelling}
+The final key plank businesses apply to estimate how effective thier marketing are econometric or mixed media models. These models
+
+In practise these models are typically timeseries regression models, fitting some measure of sales against marketing spend (by marketing channel, ie TV, radio, search, digitial display etc) and other key variables such as pricing, competitor spend and pricing and indicators of the health of the economy and demand for the product in question.
+
+The models may also include non-linear transformation of the marketing inputs to represent saturation and decay of time of effects.
+They may also consider the interaction between media chanels and other business levers such as pricing.
+
+The objective of these models is to obtain results that are comparable across channels such that the optimal mix of media can be selected.
+
+The key
 

### 2017-05-18 thesis/thesis_map.tex
updates to chapter 2

@@
 \documentclass[11pt,a4paper,oneside]{book}
+\usepackage[table]{xcolor}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{tabularx}
+\usepackage{booktabs}
 \usepackage{enumitem}
@@
 \usepackage{epigraph}
-\usetikzlibrary{arrows,positioning} 
+\usetikzlibrary{arrows,fit,positioning}
 \usepackage{subcaption}
@@
 
+% this is included to get tables with colored rows to not draw over hlines and vlines
+\usepackage{etoolbox}
+
+\makeatletter
+\patchcmd{\@classz}
+  {\CT@row@color}
+  {\oldCT@column@color}
+  {}
+  {}
+\patchcmd{\@classz}
+  {\CT@column@color}
+  {\CT@row@color}
+  {}
+  {}
+\patchcmd{\@classz}
+  {\oldCT@column@color}
+  {\CT@column@color}
+  {}
+  {}
+\makeatother
+
+\usepackage{amsthm}
+
 \graphicspath{ {figures/} }
@@
 \DeclareUnicodeCharacter{00A0}{ }
-
-%\tikzset{
-%    %Define standard arrow tip
-%    >=stealth',
-%    %Define style for boxes
-%    punkt/.style={
-%           circle,
-%           rounded corners,
-%           draw=black, thick,
-%           text width=1em,
-%           minimum height=1em,
-%           text centered},
-%    punktrect/.style={
-%    rectangle, 
-%    rounded corners, 
-%    % fill=black!10,
-%    draw=black, thick,
-%    minimum height=1em, 
-%    text centered},
-%    % Define arrow style
-%    pil/.style={
-%           o->,
-%           thick,
-%           shorten <=2pt,
-%           shorten >=2pt,}
-%}
 
@@
            text centered,
+           fill=black!10!white
            },
@@
            font=\footnotesize,
-           text centered,
-           fill=black!10!white
+           text centered
            },
@@
            text centered,
+           fill=black!10!white
+           },
+    latentrect/.style={
+           rectangle,
+           rounded corners,
+           draw=black, thick, dashed,
+           minimum width=2.2em,
+           minimum height=2.2em,
+           font=\footnotesize,
+           text centered
            },
@@
 
-%\theoremstyle{plain}
-
+\theoremstyle{plain} 
 \newtheorem{theorem}{Theorem}
@@
 
-%\theoremstyle{definition}
+\theoremstyle{definition}
 \newtheorem{definition}[theorem]{Definition}
@@
 
+\begin{figure}[h]
+\centering
+\caption{Causal inference without a framework}
+\label{fig:mutilatednet}
+\begin{subfigure}[t]{0.3\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:mutilatedOriginal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3);
+
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.5\textwidth}
+\centering
+\caption{network after an intervention $G^*$}
+\label{fig:mutilatedAfter}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={punkt}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$x$};
+\node[main node, below right=of 1](3){$Y$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (3)
+    (2) edge node {} (3);
+
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+\begin{figure}[h]
+\caption{}
+\label{fig:causal_inference_without_do}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](2){Treatment};
+\node[node, right=of 2](3){Outcome};
+\node[lt, above=of 3](4){U};
+\path[]
+	(2) edge (3)
+	(4) edge (3);
+
+\end{tikzpicture}
+\end{figure}
+
 There is a connection here to the degree to which the model is invariant.
-A model that is invariant to more things is more explainitory.
-We could look at levels of hierachy.
-There is causal graphical models, which are invarient to interventions on variables.
-Can I tie this to a physical system and give more indepth examples of the nested tier of hieriachies?
-
-I do not see the distinction between explaination and (causal) prediction.
-Explaination is all about the ability to compress and to generalize.
+A model that is invariant to more things is more explanatory.
+We could look at levels of hierarchy.
+There is causal graphical models, which are invariant to interventions on variables.
+Can I tie this to a physical system and give more in depth examples of the nested tier of hierarchies?
+
+I do not see the distinction between explanation and (causal) prediction.
+Explanation is all about the ability to compress and to generalise.
 The more a model can do this, the more we view as providing an understanding of the why.
@@
 
-Relationship to generalization.
+Relationship to generalisation.
 Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
@@
 
-Generalizing the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parashutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalize).
-(and given sufficient generalizability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)
+Generalising the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parachutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalise).
+(and given sufficient generalisability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)
 
@@
 
-Only from randomized experiment ... if people can do it without recource to constant randomized experiment then an algorithm exits.
-
-The challenges and limitations of infering causal effects from observation data are numerous.
+Only from randomised experiment ... if people can do it without recourse to constant randomised experiment then an algorithm exits.
+
+The challenges and limitations of inferring causal effects from observation data are numerous.
 However, there are problems we need to solve.
-The questions is are these problems so serious as to warrent discregarding all non-experimental data.
+The questions is are these problems so serious as to warrant disregarding all non-experimental data.
 I think answer is a clear no.
@@
 
-Are causal/interpretable models more reliable or more likely to generalize well?
+Are causal/interpretable models more reliable or more likely to generalise well?
 
@@
 \begin{itemize}
-\item rule out many possible features as highly unlikely to be relevent to a problem
+\item rule out many possible features as highly unlikely to be relevant to a problem
 \end{itemize}
@@
 
-This is tightly related to generalizability.
+This is tightly related to generalisability.
 If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results.
@@
 
-\chapter*{Introduction (3500 words)}
+\chapter{Introduction}
 
@@
 
-\paragraph*{This thesis contributes to knoweldge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
+\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
 
@@
 \item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
-\item To summarize the key strands of causal inference research from economics and social sciences for the machine learning community
+\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
 \item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
@@
 
-\section*{Motivation}
+\section{Motivation}
 Many of the most important questions in science and in our personal lives are about the outcomes of doing something.
@@
 From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries.
-It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge datasets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artifcial intelligence exceeds our own and then grows exponentially.
+It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.
 
 However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings.
-Getting them to generalize to even slightly different problems or datasets remains very challenging.
+Getting them to generalise to even slightly different problems or data sets remains very challenging.
 Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it.
@@
 
-\section*{What is causality and why do we care? (2000 words)}
+\section{What is causality and why do we care? (2000 words)}
+
+The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}.
+This has led to a reluctance among applied researchers in many fields to refer to cauality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
+However, the magnitude, direction and even existance of an association depends on what other variables we adjust for (or include in a regression).
+Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.
+
+We discuss more detailed definitions of causality in section c
+
+The what-if type questions from the why.
+\cite{Gelman2010} \cite{}.
+Why do whites to better than blacks in school (in America).
+Suggests that reverse causal inference questions are more interesting and motivate most of social science.
+
+I do not find this distinction useful. 
+We can only change the future - history is useful only as far as it tells us something about the future.
+Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one).
+Problems highlighted as intractible in the reverse causal sense are also intractible in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning.
+For example, the war question posed in Gelman.
+
+A distinction between forward causal iference, what happens if we do X and reverse casual inference
 
@@
 The first is related to variance and over-fitting.
-Obversations are noisy.
-With a finite dataset with enough variables we will be able to find some that are completely unrelated but correlate purely by chance.
+Observations are noisy.
+With a finite data set with enough variables we will be able to find some that are completely unrelated but correlate purely by chance.
 \ref{figure - search for random correlation}.
-The second arises from bias, typically introced by an un-observed confounding variable.
+The second arises from bias, typically introduced by an un-observed confounding variable.
 In this case, variables are correlated not by chance.
@@
 
-\subsection*{Defining causality}
+\subsection{Defining causality}
 \begin{itemize}
 \item widely debated in science and philosophy (REFERENCES)
-\item what is explaination?
+\item what is explanation?
 \item any model that aims to predict the outcome of an action or intervention in a system
-\item I do not see the distinction between explaination and (causal) prediction.
-Explaination is all about the ability to compress and to generalize.
+\item I do not see the distinction between explanation and (causal) prediction.
+Explanation is all about the ability to compress and to generalise.
 The more a model can do this, the more we view as providing an understanding of the why.
@@
 \end{itemize}
-
-\subsection*{Identifying when we have a causal problem}
-
-\subsubsection*{Examples of typical machine learning probems. Are they causal?}
+\subsection{Overview of this thesis}
+\subsection{Identifying when we have a causal problem}
+
+\subsubsection{Examples of typical machine learning problems. Are they causal?}
 
@@
 Which of them require casual inference?
-How can we identify characterists of a prolem that make causal modeling important?
+How can we identify characteristics of a problem that make causal modelling important?
 
@@
 \item Playing Go 
-\item Identiying spam emails
+\item Identifying spam emails
 \item Automated essay marking
-\item Predicting the risk of death in patients with pnumonia.
+\item Predicting the risk of death in patients with pneumonia.
 \item Predicting who will re-offend on release from prison 
@@
 You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words.
-Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a dvd on elocution).
-A similar arguament could be made for many applications of machine translation and image classification.
+Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution).
+A similar argument could be made for many applications of machine translation and image classification.
 
@@
 We can be fairly confident that the process will not change.
-If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
+If we develop a discriminating model that is highly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
 
 What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decicion will not effect if it actually rains.
+If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decision will not effect if it actually rains.
 However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
@@
 
-Identifing spam and automated essay marking systems are similar.
+Identifying spam and automated essay marking systems are similar.
 The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label.
-Spammers and students will modify their writing in order to optimise thier results.
+Spammers and students will modify their writing in order to optimise their results.
 The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided.
@@
 
-What about predicting the risk of death in patients with pnumonia?
+What about predicting the risk of death in patients with pneumonia?
 Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective this seems like a straighforward prediction problem.
+If we assume that in hospital treatment is more effective this seems like a straightforward prediction problem.
 It is not.
-Depending on how the decision to admit was previously made and what features are included (or ommited) in the model, the relationship between those features and the outcome may change if we start using the model to decide whome to admit.
+Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if we start using the model to decide whom to admit.
 (xxx et al) found exactly this effect.
-Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pnemonia.
-They realised this was because doctors were treating such patints very agressively, thus acutally lowering their risk.
+Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia.
+They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
 There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols.
@@
 It seems unlikely that using the model to ensure stock is available will itself change demand.
-However there is a potential data sensoring issue.
+However there is a potential data censoring issue.
 If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
@@
 
-Playing Go (and other games) is another case with some subleties.
+Playing Go (and other games) is another case with some subtleties.
 At every turn, the AI agent has a number of actions available.
@@
 Probably because the board state encapsulates everything that should determine what move is played.
-Learning directly from actions and trying to generalize (can in some instances reduce the problem to standard ML)
-
-Having considered these examples we can now identify some general aspects of the problem that determine wheter or not we require a causal model.
+Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)
+
+Having considered these examples we can now identify some general aspects of the problem that determine whether or not we require a causal model.
 
@@
  Covariate shift clearly comes in here.
-Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable.
+Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occurring and that results will be transferable.
 The mechanism is known but the function may be complex.
 Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models?
-Yes, if the way in which features get their values changes, then causal models will be invarient to that in a way that non-causal ones are not.
+Yes, if the way in which features get their values changes, then causal models will be invariant to that in a way that non-causal ones are not.
 
@@
 \item The predicted outcome in the absence of an intervention provides a single point.
-We can use this to find which problems are most serious if left alone - and priortise those for modelling changes.
-\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
+We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes.
+\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
 \item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
-\item Ethics - ... People's viewpoint on if its ok...
+\item Ethics - ... People's viewpoint on if its OK...
 \end{itemize}
 
-I hope these examples gave you a feel for the richness and subtlties of causal inference.
+I hope these examples gave you a feel for the richness and subtitles of causal inference.
 We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
 
-\section*{Approaches to causality (1000 words)}
+\section{Approaches to causality (1000 words)}
 
@@
 
-We will call datasets where we do not have control over the decision making process that generated the data observational.
-Versus experimental datasets, where we do have control (experimental datasets do not always have to be randomized -although that is a powerful approach to ensure we have control.
+We will call data sets where we do not have control over the decision making process that generated the data observational.
+Versus experimental data sets, where we do have control (experimental data sets do not always have to be randomised -although that is a powerful approach to ensure we have control.
 There is a space in the middle where we have partially controlled the process by which agent select actions.
-Randomized data with imperfect compliance would be an example.
-
-An agent (capable of intervening in the system) chooses an action from those availble The agent making the decision included in the model. This is
+Randomised data with imperfect compliance would be an example.
+
+An agent (capable of intervening in the system) chooses an action from those available The agent making the decision included in the model. This is
 
@@
 ADVANTAGES AND DISADVANTAGES OF THIS APPROACH.
-The second broand approach is causal inference.
+The second broad approach is causal inference.
 Here
+
+\todo{insert figure showing data generating process and observed data defining ML}\ref{all of statistics}
 
@@
 
-\paragraph*{Both generalize from randomized experiment} Reinforcment learning to sequential decisions, causal inference to non-experimental conditions
+\paragraph*{Both generalise from randomised experiment} Reinforcement learning to sequential decisions, causal inference to non-experimental conditions
 
@@
 
-\paragraph*{Limitations of experimets} What are the issues with standard randomized experiments?
-
-\chapter*{Causal models (3000 words)}
-
-Causal inference aims to infer the outcome of an intervention in a system from data obtained by observing (but not intervening in) the system.
+\paragraph*{Limitations of experiments} What are the issues with standard randomised experiments?
+
+\chapter{Causal models}
+\label{chap:causal_models}
+
+Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
 To do this we need terminology to describe actions and how we anticipate the system should respond to them.
@@
 We will examine what problems they allow us to solve, what assumptions they rely on and how they differ.
-We will also use them to describe the following very simple example.
+We will also use them to describe the following simplified examples.
 The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
 
+\vspace*{.3cm}
+\begin{example}
+\label{exm:ranomized_experiment}
 Suppose we have developed a new drug for some illness and wish to determine how effective it is.
@@
 We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
-
-\subsection*{Causal bayesian networks}
-
-Causal Bayesian Networks are an extension of Bayesian networks.
-A bayesian network is a graphical way of representing how a distriution factorises.
-Any joint probability distribution can be factorized into a product of conditional probabilities.
-There are multiple valid factorizations, corresponding to permutations of variable ordering.
-
-\begin{equation}
+\end{example}
+
+\vspace*{.3cm}
+\begin{example}
+\label{exm:adjusting}
+Suppose we wish to estimate what would happen to high school graduation rates if we made pre-school available and compulsory for all 4 year olds.
+We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
+We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.
+\end{example}
+
+\section{Causal bayesian networks}
+
+Causal Bayesian networks are an extension of Bayesian networks.
+A Bayesian network is a graphical way of representing how a distribution factorises.
+Any joint probability distribution can be factorised into a product of conditional probabilities.
+There are multiple valid factorisations, corresponding to permutations of variable ordering.
+
+\eqn{
+\label{eqn:cbn:joint_dist}
 P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
-\end{equation}
-
-We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}).
-If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.
+}
+
+We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}.
+If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network.
+There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.
 
@@
 \centering
-\caption{General Bayesian network over 3 variables}
+\caption{A general Bayesian network for the joint distribution over three variables.
+This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
 \label{fig:bayesnet}
@@
 
-The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorized over the nodes and edges in the graph.
+\begin{figure}
+\centering
+\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
+\label{fig:bayesnet2}
+\begin{subfigure}[t]{0.15\textwidth}
+\centering
+\caption{}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, below=of 1](2){$X_{2}$};
+\node[main node, below=of 2](3){$X_{3}$};
+\path[every node/.style={font=\sffamily\small}]
+    (1) edge (2)
+    (2) edge (3);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.15\textwidth}
+\centering
+\caption{}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){$X_{1}$};
+\node[main node, below=of 1](2){$X_{2}$};
+\node[main node, below=of 2](3){$X_{3}$};
+\path[every node/.style={font=\sffamily\small}]
+    (3) edge (2)
+    (2) edge (1);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.3\textwidth}
+\centering
+\caption{}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){$X_{2}$};
+\node[main node, below left=of 1](2){$X_{1}$};
+\node[main node, below right=of 1](3){$X_{3}$};
+\path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.3\textwidth}
+\centering
+\caption{}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
+ %nodes
+\node[main node](1){$X_{2}$};
+\node[main node, above left=of 1](2){$X_{1}$};
+\node[main node, above right=of 1](3){$X_{3}$};
+\path[every node/.style={font=\sffamily\small}]
+    (2) edge (1)
+    (3) edge (1)
+    (2) edge (3);
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph.
 There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges).
@@
 This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-We need some notation to represent the distribution after an intervention.
-In this thesis we will use the do opperator introduced by Pearl \ref{XXX}, where $\P{Y|do(X=x)}$ represents the distribution of the variable $Y$ after an intervention that sets the variable(s) $X$ to $x$.
-TALK ABOUT HOW THIS CAN REPRESENT MULTIPLE THINGS CONTINUOUS VERSUS DISCRETE DISTRBUTIONS AND WHETHER OR NOT VARIABLES ARE RESOLVED OR AS A FUNCTION.
-It is important to understand that in general, the interventional distribution $P{Y|do(X=x)}$ will differ from the standard conditional distribution $\P{Y|X=x}$.
-
-If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network.
-This is referred to as the truncated product formula \cite{Pearl2000}.
+We need some notation to describe interventions and represent distributions over variables in the network after an intervention.
+In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.
+
+\begin{definition}{The do-notation}
+\begin{itemize}
+\item do(X=x) denotes an intervention that sets the random variable(s) $X$ to $x$.
+\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$.
+This notation is somewhat overloaded.
+It may be used represent a probability, a probability distribution/mass function or a family of distribution functions depending on if the variables are discrete or continuous and whether or not we are treating them as fixed.
+For example it could represent
+\begin{itemize}
+\item The probability $\P{Y=1|do(X=x)}$ as a function of $x$
+\item The probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$
+\item The probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$
+\item a familiarly of density/mass function for $Y$ paramaterised by $x$.
+\end{itemize}
+Where the distinction is important not clear from context we will use one of the more specific forms above. 
+\end{itemize}
+\end{definition}
+
+\begin{theorem}[Truncated product formula]
+\label{thm:truncated_prodcut}
+If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorisation given by the network \citep{Pearl2000}.
 
@@
 \label{eq:truncatedproduct}
- P_{do(Z=z)}(X_{1}...X_{N}) =
+\P{X_1...X_N|do(Z=z)} =
   \begin{cases}
-  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
+  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
    0       & \text{otherwise } 
@@
 \end{equation}
-
-This result does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution.
+\end{theorem}
+
+Theorem \ref{thm:truncated_prodcut} does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution.
 The truncated product formula will give different results depending on which you choose.
 The result is possible with causal bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
-In fact, from the interventionalist viewpoint causality the truncation product formula defines what it means for a link to be causal.
-
-Returning to our simple example, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$?
-Since the treatment $X$ is assigned  via deliberate randomisation we know there it is not effected by any unobserved variables.
-$X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and desease subtype of the patient.
-
-The causal bayesian network for this scenario is:
-
-\todo{proper picture}
-\eq{
-X \rightarrow Y\\
-U_l \rightarrow Y
-}
-
-This network represents the (causal) factorization  $P(X,Y) = P(X)P(Y|X)$, so from equation (\ref{eq:truncatedproduct}), $P(Y|do(X)) = P(Y|X)$.
-In this example, the interventional distribution is equivelent to the observational one.
-
-Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechinism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
-A key feature of this way of defining an intervention is the number of things that are invariant to the intervention.
-That is, all the conditional distributions over variables that were not intervened are assumed not to be changed by the intervention.
-
-\todo {draw the system as two seperate models}
-
-A causal bayesian network encodes all possible interventions that can be represented with the do notation.
-If you knew of all the terms in the causal factorisation, then you could compute the distribution over any variable in the graph after setting any subset of other vairables (subject to possitivity assumption)
-\todo {state positivity assumption}
-
-A causal network represents much more information than a bayesian network with identical structure.
-GIVE EXAMPLE IN SOMETHING QUANTIFIABLE LIKE BITS.
-TAKE THE GRAPH IN WITH THREE NODES.
-ASSUME EVERYTHING IS BINARY AND.
-FIGURE OUT HOW MANY VARIABLES ARE NEEDED TO DESCRIBE ALL POSSIBLE INTERVENTIONS VS THE BAYESIAN NETWORK.
-This may be surprising since their conditional probability distributions are estiated from data in the same way.
-It is only possible due to the additional assumptions inherent in a causal network.
-
-Causal bayesian networks are bayesian networks so results that apply to bayesian networks carry directly across;the local markov property states that variables are independent of their non-effects given their direct causes.
-Similarly the global markov property and d-separation also hold in causal networks.
-
-\subsubsection*{Limitations of causal bayesian networks}
-A number of critisims have been leveled at this way of modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will effect many variables in complex and non-determinisitic ways \ref{causality in complex interventions,Cartwright 2007}.
-However, the deterministic $do()$ opperater turns out to be surprisingly effective at modelling more complex interventions.
-One simply has to add additional variables to the graph that model how our interventions may actually take effect.
-For example, our drug treatment case assumed we set the drug that each person recieves.
-But what if some people in fact fail to take the prescribed treatment.
+In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.
+
+Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$?
+The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and disease sub type of the patient.
+Since $X$ is assigned  via deliberate randomisation we know that it is not effected by any latent variables.
+The causal bayesian network for this scenario is shown in figure \ref{fig:causal_network_example} .This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$.
+In this example, the interventional distribution is equivalent to the observational one.
+
+\begin{figure}[h]
+\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
+\label{fig:causal_network_example}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](2){$X \text{ (Treatment)}$};
+\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
+\path[]
+	(2) edge (3);
+\end{tikzpicture}
+\end{figure}
+
+In example \ref{exm:adjusting} we are interested $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool.
+We could compare it to outlawing pre-school $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$.
+It seems reasonalble to assume that attending could pre-school effect the likelyhood of graduating from high school and that your parents socio-economic status effects both pre-school attendance and high-school graduation.
+If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal bayesian network in figure \ref{exm:adjusting}.
+In this case, the interventional distribution is not equivelent to the observational one.
+If better off parents are more likely to send their children to pre-school these children more likely to graduate high school regardless, then simply comparing the high school graduation rates of those who attended preschool with those who did not will overstate the benifit of preschool.
+To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group, $\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}$.
+
+We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph.
+There is a very powerful and general set of rules that specify how we can transform observational distributions into intervetional ones for a given graph structure.
+These rules are referred to as the Do-calculus \cite{Pearl2000}.
+We discuss them further in section \ref{}.
+
+\begin{figure}
+\center
+\caption{Causal bayesian network for example \ref{exm:adjusting}}
+\label{fig:causal_adjust}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]
+
+ %nodes
+\node[node](1){$Z \text{ (SES)}$};
+\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
+\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
+\path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3); 
+\end{tikzpicture} 
+\end{figure}
+
+Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
+The power of defining an intervention in this way stems from the number of things that are invariant between the two processes.
+All the (conditional) distriubtions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
+
+We could represent problems of the type where we try to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and what they have in common, see figure \ref{fig:causal_ordinary_bayes}.
+This does not require any special framework or notation.
+The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks.
+However, without a causal framework, we have to make assumptions about what will be invarient to the intervention specifically for each such problem we encounter.
+For complex problems, it is very difficult to conceptualise what assumptions we expect to hold without the benifit of a causal framework.
+
+\begin{figure}[h]
+\center
+\caption{Causal inference with ordinary bayesian networks.
+The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that the pre-interventional variable $X$ to $x$.
+The assumptions characterised by this plate model correspond to those implied by the causal bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
+As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
+However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
+\label{fig:causal_ordinary_bayes}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
+\tikzstyle{box}=[rectangle, draw=black!100]
+ %nodes
+\node[node](1){$Z$};
+\node[node, below left=of 1](2){$X$};
+\node[node, below right=of 1](3){$Y$};
+\node[node, right=1.5cm of 3](4){$x$};
+\node[lt, above right=of 4](5){$Z'$};
+\node[lt, below right=of 5](6){$Y'$};
+\node[lt, above=2.0cm of 3](9){$\theta_{y}$};
+\node[lt, above=0.5cm of 9](7){$\theta_z$};
+\node[lt, above=2.0cm of 2](8){$\theta_x$};
+\path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3)
+    (4) edge (6)
+    (5) edge (6)
+    (8) edge (2)
+    (7) edge (1) edge (5)
+    (9) edge (3) edge (6);
+\node[rectangle, inner sep=4.0mm, fit= (1) (2) (3),draw=black!100,label=below right:N] {};
+\node[rectangle, inner sep=4.0mm, fit= (5) (6),draw=black!100,label=below right:M] {};
+\end{tikzpicture} 
+\end{figure}
+
+A causal bayesian network represents much more information than a bayesian network with identical structure.
+A causal network encodes all possible interventions that could be specified with the do-notation.
+For example, if the network in figure \ref{fig:causal_adjust} were an ordinary bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
+The equivelent causal bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
+Encoding all this information without the assumptions implicit in the causal bayesian network would require and additional 30 parameters.
+\footnote{After each single variable intervetion we have a distribution over two variables, which can be represented by three parameters.
+After each two variable intervention, we have a distribution over one variables which requires one parameter.
+This takes us to a total of $6*3+12*1 = 30$ additional parameters.}
+
+Causal bayesian networks are bayesian networks so results that apply to bayesian networks carry directly across;the local Markov property states that variables are independent of their non-effects given their direct causes.
+Similarly the global Markov property and d-separation also hold in causal networks.
+
+\subsubsection{Limitations of causal bayesian networks}
+A number of criticisms have been levelled at this approach to modelling causality.
+One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will effect many variables in complex and non-deterministic ways \citep{2009causality,cartwright2007hunting}.
+However, the deterministic $do()$ operator turns out to be surprisingly effective at modelling more complex interventions by augmenting the causal graph with additional variables that model how our interventions may take effect.
+For example, in our drug treatment case we assumed we set the drug that each person receives.
+But what if some people in fact fail to take the prescribed treatment?
 How can we model this within the framework of deterministic interventions?
-We simply add a node representing what we prescribed to them (which we intervene on), which probabilistically influences what treatment they actually recieve.
-See figure XXXX
-
-Insert figure of randomized treatment with imperfect compliance here.
-
-What questions can't be phrased using this terminology
-- certain counterfactual queries
-- cyclic/dynamics
-- unroll to deal with cycles
-
-\subsection*{Structural Equation models}
-
-\subsubsection*{Definition}
-Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Uncertainties are not inherent but arise from unmeasured variables.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalized, generalized to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-For our example the SEM is:
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random.
-The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-
-\subsubsection*{Example problem}
-We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
-Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$.
-The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}.
-However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$.
-The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomized experiment.
-In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.
-
-For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-
-Limitations \& Critisisms 
-- can express some non-measurable things
-	- example of that
-
-- confusion may occur as to whether interpretation is causal
-
-- can they handle cycles?
-
-\subsection*{Counterfactuals}
-\todo{use cf command to in counterfactual equations}
+We add a node representing what we prescribed to them (which we intervene on), which probabilistically influences what treatment they actually receive.
+See figure \ref{fig:randomized_imperfect_compliance}.
+Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could effect both the actual treatment taken and the outcome.
+
+The causal effect may provide useful bounds on the actual quantity of interest.
+
+\begin{figure}[h]
+\caption{Randomised treatment with imperfect compliance}
+\label{fig:randomized_imperfect_compliance}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](1){Prescribed Treatment};
+\node[node, below right=of 1](2){Actual Treatment};
+\node[node, right=of 2](3){Outcome};
+\node[lt, above left=of 3](4){U};
+\path[]
+	(1) edge (2)
+	(2) edge (3) 
+	(4) edge (3) edge (2);
+\end{tikzpicture}
+\end{figure}
+
+Another key issue with causal bayesian networks is that they cannot handle cyclic dependencies between variables.
+Such feedback loops are common in real systems, for example the relationship between supply and demand in economics or preditor and prey in ecology.
+We might regard the underlying causal mechanisms in these examples to be acyclic; the number of preditors at one point in time influences the number of prey in the next and so on.
+However, if our measurements of these variables must be aggregated over timeframes that are longer than the scale at which these interactions occure the result is a cyclic dependency.
+Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
+Such problems have mostly been studied within the dynamic systems liturature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
+\citep{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with strucural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as bayesian causal networks with additional functional assumptions.
+
+\section{Counterfactuals}
 
@@
 
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{Rosenbaum1983}.
+The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignore-ability assumptions \cite{Rosenbaum1983}.
 They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-Again this is justified in our example due to the randomization of treatment assignment.
+Again this is justified in our example due to the randomisation of treatment assignment.
 In general these assumption do not hold.
 If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model.
-that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.
+that influenced both the decision to get treatment and the likelihood of recovery given treatment or placebo.
 
@@
 
-There are some complex philosopical objections to counter factuals arising from the way they describe alternate universes that were never realized and are not empirically testable \ref{XXXXX}.
+There are some complex philosophical objections to counter factuals arising from the way they describe alternate universes that were never realised and are not empirically testable \ref{XXXXX}.
 This can result in some concrete problems.
@@
 As a result, the variance of $\tau$ is not identifiable, even from experimental data.
-It seems on the face of it that $\tau$ is relevent to our decision making.
-If the example above, if $\rho = 1$ then almost everyone benifits slightly from the treatment.
-However if $\rho=0$, there is a wide range, with some people benifiting a lot and others suffering significant harm.
+It seems on the face of it that $\tau$ is relevant to our decision making.
+If the example above, if $\rho = 1$ then almost everyone benefits slightly from the treatment.
+However if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.
 
@@
 
-- does it actually make sense to make a decison on the basis of this counterfactual that is a function of something we could never observe. I
-
-- part of the problem may be due to the deteminisitic way we have phrased individual treatment effects.
-
-Can these issues be resolved by considering personalized causal effects as random variables and avoiding counterfactuals?
+- does it actually make sense to make a decision on the basis of this counterfactual that is a function of something we could never observe. I
+
+- part of the problem may be due to the deterministic way we have phrased individual treatment effects.
+
+Can these issues be resolved by considering personalised causal effects as random variables and avoiding counterfactuals?
 
@@
 Rather than taking about the distribution of Y given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $X$.
-This is certainly convienient, if rather impressise.
-
-\subsection*{Comparing and unifying the models}
+This is certainly convenient, if rather impressive.
+
+\section{Structural Equation models}
+\label{sec:SEM}
+
+Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
+The mechanism (but not the output) is assumed to be independent of what is fed into it.
+Uncertainties are not inherent but arise from unmeasured variables.
+Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
+More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.
+
+Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
+The noise terms are required to be mutually independent.
+If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable.
+Structural equation models can be represented visually as a network.
+Each variable is a node and arrows are drawn from causes to their effects.
+For example \ref{exm:ranomized_experiment} the SEM is:
+
+\begin{figure}[h]
+\caption{SEM for examples \ref{exm:ranomized_experiment}}
+\label{fig:causal_network_example2}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](2){$X = x$};
+\node[node, right=of 2](3){$Y = f(X,\epsilon_y)$};
+\node[lt, above=of 3](4){$\epsilon_y$};
+\path[]
+	(2) edge (3)
+	(4) edge (3);
+\end{tikzpicture}
+\end{figure}
+
+This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
+This is justifiable on the grounds that $X$ is random.
+The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
+Note that the causal graph in figure \ref{fig:causal_network_example2} is identical to that the bayesian network for the same problem, figure \ref{fig:causal_network_example}.
+The latent variable $\epsilon_y$ was not explicitly drawn in figure \ref{fig:causal_network_example} as the noise can be captured within the probabilistic node, however adding it would not change the model.
+
+We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
+Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$.
+The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}.
+However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$.
+The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomised experiment.
+In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.
+
+For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
+However a SEM is, by definition, more than a set of simultaneous equations.
+By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
+When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
+
+Limitations \& Criticisms 
+- can express some non-measurable things
+	- example of that
+
+- confusion may occur as to whether interpretation is causal
+
+- can they handle cycles?
+
+\section{Comparing and unifying the models}
 
 Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-This makes it straighforward to combine key results and algorithms developed within any of these frameworks.
+This makes it straightforward to combine key results and algorithms developed within any of these frameworks.
 For example, draw a graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
@@
 
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorization of the joint distribution over its variables.
+If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
 In other words, the network is a causal Bayesian network.
@@
 Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorization and the truncated product formula (equation \ref{eq:truncatedproduct}).
+This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}).
 Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
@@
 The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment.
+The Neyman-Rubin framework builds on our understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
 This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}.
@@
 
-\subsubsection*{More terminology}
+\subsubsection{More terminology}
 With the graphical framework in place, it is useful to define some key terminology used in describing causal models in terms of the graph structure they refer to.
@@
 
-\subsection*{What does a causal model give us? Resolving Simpson's paradox}
-
-We will now demonstrate how we can use our new notation and frameworks for causal inference to resolve a facinating paradox. \todo{ insert some history of simpson's paradox here.}
-
-Suppose a doctor has two treatments, A and B, which she offers to patients for blood preasure control.
+\section{What does a causal model give us? Resolving Simpson's paradox}
+
+We will now demonstrate how we can use our new notation and frameworks for causal inference to resolve a fascinating paradox.  \todo{ insert some history of Simpson's paradox here.}
+
+\cite{simpson1951interpretation}
+
+Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
 She keeps track of which medication her patients choose and whether or not the treatment is successful.
-She obtains the following results:
-
-\todo{insert data table with success by treatment}
-
-It seems that drug A is performing better.
-However, having read the latest liturature that medication effectiveness may differ between men and women, she decides to check how each drug performed by gender and obtained the following data.
-
-\todo{innsert table}
-
-Once the data is broken down by gender, it appears that treament B is better for both men \emph{and} women.
-This is the paradox.
-If the doctor were unaware of the gender, it seems that she should advise treatment A.
-But as soon as the patient walks though the door  then she should offer treatment B, regardless of which gender the patient turns out to be.
-
-How does causaul modelling reslove this paradox?
-Let's represent the treatment presribed by the variable $T$, the gender of the patient by $Z$ and wheter or not the treatment was successful by $Y$.
-The key to resolving the paradox is that the doctor really cares about the interventional distribution $P{Y|do(T)}$, not the standard conditional distributions $P{Y|T}$.
-The results from table XXX shows that $\P{Y|T=A}$ is higher than $\P{Y|T=B}$.
-If she has to suggest a treatment for a patient without knowing their gender, she should choose the treatment for which $P{Y|do(T)}$ is highest not $P{Y|T}$.
-If she knows thier gender then she should otimise to $P{Y|do(T),Z}$.
-
-The causal network for this problem is shown in figure XXX.
-In this example, women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they recive than men.
-This is what makes treatment $A$ appear better if you look at the conditional distribution.
-The fact that a patient took treatment $A$ tells you that they are more likely to be female.
-In the terminology defined in section \ref{}, gender is a confounding variable and we must adjust for it to obtain unbiased estimates of the causal effect of the treatment.
-As we saw from the examples in chapter \ref{XXX}.
-Determining when you really care about the interventional as opposed to the conditional distribution is a sublte but very important question that is frequently overlooked.
-
-\chapter*{Two key questions (5000 words)}
-
-We can rougly categorize the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
+She obtains the results in table \ref{tab:simpson1}.
+
+\begin{table}[h]
+\label{tab:simpson1}
+\caption{Treatment results}
+\center
+\begin{tabular}{c c c c c}
+Treatment & Success & Fail & Total & Success Rate\\
+\hline
+\rowcolor[gray]{0.9} 
+A &  87 & 13 & 100 & 87\%\\
+\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
+\hline
+\end{tabular}
+\end{table}
+
+It looks like drug A is performing better.
+However, having read the latest literature on how medications effect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
+
+\begin{table}[h]
+\label{tab:simpson2}
+\caption{Treatment results by gender}
+\center
+\begin{tabular}{c c c c c c}
+Gender & Treatment & Success & Fail & Total & Success Rate\\
+\hline
+\rowcolor[gray]{0.9}
+M & A & 12 & 8 & 20 & 60\%\\
+M & B & 56 & 24 & 80 & 70\%\\
+\rowcolor[gray]{0.9}
+F & A & 75 & 5 & 80 & 94\%\\
+F & B & 19 & 1 & 20 & 95\%\\
+\hline
+\end{tabular}
+\end{table}
+
+Much to her surprise, once the data is broken down by gender, Treatment B looks better for both men \emph{and} women.
+Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must reccomend which to subsidise under a national health scheme).
+Should she choose A or B?
+The ambiguity in this question lies at the heart of Simpson's paradox.
+How does causal modelling resolve the paradox?
+The key is that the doctor is trying to choose between \emph{interventions}.
+She wants to know what the success rate will be if she changes her practise to give all the patients one drug, rather than allowing them to choose as currently occurs.
+
+Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
+The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
+Unfortunately, the data in tables tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable us to estimate the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
+We require some assumptions about the causal relationships between the variables.
+In this example, it seems reasonable to conclude that gender may effect the treatment chosen and the outcome.
+If we assume there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
+
+\begin{figure}[h]
+\caption{An example of Simpson's Paradox}
+\label{fig:simpsons_paradox}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](1){Gender};
+\node[node, below left=of 1](2){Treatment};
+\node[node, below right=of 1](3){Outcome};
+\path[]
+	(1) edge (2) edge (3) 
+	(2) edge (3);
+\end{tikzpicture}
+\end{figure}
+
+With this model, we see women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they receive than men.
+Knowing a patient took drug A tells you they are much more likely to be female.
+When we compare the group of people who took A against those who took B, the greature proportion of females in the first group outweighs the higher benifit of drug B leading to an apparent reversal in effectiveness.
+However, when the doctor intervenes to set the treatment each person recieves there will no longer be a link from gender to treatment.
+So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
+Drug B is the better choice.
+
+\eqn{
+\P{Y|do(T)} = \P{Y|T,female}\P{female}+ \P{Y|T,male}\P{male}
+}
+
+Is the solution to Simson's paradox to always to break down the data by as many variables as possible?
+No.
+Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood preassure', 'M' with 'high' and 'F' with 'normal'.
+This is a drug desined to prevent heart desease.
+One pathway to doing so might well be to lower blood preassure.
+Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
+It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
+Now, however table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood preassure control and better overall outcomes.
+In this setting $\P{Y|do(T)} = \P{Y|T}$.
+Drug A is the better choice.
+
+\begin{figure}[h]
+\caption{An example of Simpson's Paradox}
+\label{fig:simpsons_paradox2}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](1){Blood Preassure};
+\node[node, below left=of 1](2){Treatment};
+\node[node, below right=of 1](3){Outcome};
+\path[]
+	(1) edge (3) 
+	(2) edge (3)
+	(2) edge (1);
+\end{tikzpicture}
+\end{figure}
+
+Note that we have not changed the data itself, only the story we tell around it.
+This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data itself but in what assumptions we are willing to make.
+From a purely statisitcal viewpoint there is no paradox.
+The reversal is just a case of INSERT VECTOR ARGUAMENT HERE.
+The 'paradox' only arises when we need to use the data to select an intervention.
+
+I should also point out that there are many other plausible causal graphs for both scenarios above.
+Perhaps income effects drug choice as well as gender, or gender could effect treatment choice and blood preassure control given treatment, etc.
+Causal modeling provides us with a powerful
+
+Human in the loop ML. Why would you need the human?
+
+\todo{mention relationship to ecological fallacy}
+
+\chapter{Two key questions}
+
+We can roughly categorise the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
 In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
@@
 THIS IS ALREADY HUGE, and is central to 1000 of papers published each year.
-Causal discovery aims to leverage much broader assumptions to learn the stucture of causal graph from data.
+Causal discovery aims to leverage much broader assumptions to learn the structure of causal graph from data.
 THIS IS THE AUTOMATION OF SCIENCE.
 
-\section*{Causal effect estimation}
+\section{Causal effect estimation}
 
@@
 \item the target/outcome variable we care about
-\item the focus/treatment variables on which we are considering intervetions
+\item the focus/treatment variables on which we are considering interventions
 \item any variables which act to confound two or more of the variables we have included.
@@
 However, their position in the network is assumed to be known.
-For example if we were interested in the effect of years of schooling on wages, we might expect that some measure of inherent ability could influence both the number of years of schooling people chose to pursue and the wages they subsequently obtaine on graduating.
-Even if we have no data to directly measure peoples inherent ability, becuase it influences two of the variables were are modelling, we must include it in the graph.
-
-Causal effect estimation is one of the most frequently applied methodolgoies within applied sciences.
-It implicit in millions of studies accross medicine, business, economics and social sciences.
-Almost everytime someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best levereage those assumptions to infere as much information as we can from our data.
-
-\subsection*{Mapping from observational to interventional distributions}
+For example if we were interested in the effect of years of schooling on wages, we might expect that some measure of inherent ability could influence both the number of years of schooling people chose to pursue and the wages they subsequently obtain on graduating.
+Even if we have no data to directly measure peoples inherent ability, because it influences two of the variables were are modelling, we must include it in the graph.
+
+Causal effect estimation is one of the most frequently applied methodologies within applied sciences.
+It implicit in millions of studies across medicine, business, economics and social sciences.
+Almost every time someone runs a regression model the key quantity of interest is a causal effect.
+Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
+
+\subsection{Mapping from observational to interventional distributions}
 Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting to the interventional one.
-Recall that a bayesian network is a way of representing the joint distribution over its variables in a way which highlights the any conditional idependencies between them.
+Recall that a bayesian network is a way of representing the joint distribution over its variables in a way which highlights the any conditional independencies between them.
 
 \begin{theorem}
-\textbf{(Local markov condition)} Given a bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decendents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
+\textbf{(Local Markov condition)} Given a bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
 \end{theorem}
 
-The set of conditional independence relations given by the local markov condition can enforce additional independences that also hold in all distributions that are compatible with $G$.
-D-separation is an algorithm that extends the local markov property to find these additional independences.
+The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$.
+D-separation is an algorithm that extends the local Markov property to find these additional independencies.
 It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
 
-By leveraging conditional independencies we can effectivly localize the effect of an intervention to a specific part of a larger graph.
+By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
 This gives rise to the do-calculus \cite{Pearl2000}.
@@
 
-\subsubsection*{Rule 1}
+\subsubsection{Rule 1}
 This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention.
@@
 
-\subsubsection*{Rule 2}
+\subsubsection{Rule 2}
 Rule 2 states when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening  $do(\boldsymbol{X}=\boldsymbol{x})$ have the same effect on the distribution of the outcome $\boldsymbol{Y}$.
@@
 \centering
-\caption{Augmentated causal network with intervention}
+\caption{Augmented causal network with intervention}
 \label{fig:augmentedNetwork}
@@
 
-\subsubsection*{Rule 3}
+\subsubsection{Rule 3}
 This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
@@
 
-\subsection*{Identifiability}
+\subsection{Identifiability}
 A natural question to ask is, given the level of assumptions we have made about the graph, is it possible to estimate a causal effect from observational data?
 This is the identifiablity problem.
-Identifiability is an asymptopic property of the graph.
+Identifiability is an asymptotic property of the graph.
 It asks if we can obtain an unbiased point estimate for the causal effect of interest in the infinite data limit.
-A problem is non-parametrically identifaible if it is identifiable without any additional assumptions on the functional form of the dependencies between variables in the graph.
+A problem is non-parametrically identifiable if it is identifiable without any additional assumptions on the functional form of the dependencies between variables in the graph.
 
@@
 The do calculus is complete.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observataional quantities via repeated application of the do calculus.
+A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do calculus.
 
@@
 In other words, determine if the query is identifiable given the assumptions encoded by the network
-\item if it is identiable, return the required expression 
+\item if it is identifiable, return the required expression 
 \end{enumerate}
@@
 There are many interesting questions relating to identifiability that remain open.
-What is the minimal (by some metric) additional information that would be required to make a non-identifiable qurey identifiable?
+What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
 What if we assume various restrictions on the functional form of the relationships between the variables?
@@
 However, we should not give up on non-identifiable problems.
-We may still be able to achive useful bounds on causal effects with finite data even if point estimation is not possible.
+We may still be able to achieve useful bounds on causal effects with finite data even if point estimation is not possible.
 From that perspective we should be cautious about dividing problems into identifiable and not identifiable.
@@
 
-\subsection*{Estimation}
-\subsubsection*{Defining causal effects} 
-So far we have described causal effect estimation in term of idenfitifying the interventional distribution $\P{Y|do(X)}$ from observational data.
+\subsection{Estimation}
+\subsubsection{Defining causal effects} 
+So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
 
@@
 
-If the variables are continouous the average casual effect is
-
-\todo{insert equation for continous case}
+If the variables are continuous the average casual effect is
+
+\todo{insert equation for continuous case}
 
 The individual causal effect as defined in section XXX cannot be stated in terms of the do notation as it is not with respect to a distribution over anything.
-However, we can define a very similar concept with respect some observed context which I will refer to as the personalized causal effect.
+However, we can define a very similar concept with respect some observed context which I will refer to as the personalised causal effect.
 
@@
 
-However, the difference in expectation is not the only way of summarizing the difference between two (or more) distributions.
-\citep{Janzing2013} show that there is no single way of summarising the collection of distributions represented by $\P{Y|do(X)}$ that satistifies a number of reasonable axioms.
-
-An obvious alternative to subtracting the expections would be to consider ratio.
-Is there any reason that we don't do this that relates to the arguament in Dawid?
+However, the difference in expectation is not the only way of summarising the difference between two (or more) distributions.
+\citep{Janzing2013} show that there is no single way of summarising the collection of distributions represented by $\P{Y|do(X)}$ that satisfies a number of reasonable axioms.
+
+An obvious alternative to subtracting the exceptions would be to consider ratio.
+Is there any reason that we don't do this that relates to the argument in Dawid?
 What about the smoothness of any uncertainties?
 
-\subsubsection*{Estimating the difference when the back door critera is met}
-
-A very substantial number of practial studies fall into the cateogy of causal inference via the backdoor theorem.
+\subsubsection{Estimating the difference when the back door criteria is met}
+
+A very substantial number of practical studies fall into the category of causal inference via the backdoor theorem.
 A study is done.
-It is assumed that it is ignoreable.
-That is that we have a a set of variables satifying the back door adjustment criteia.
+It is assumed that it is ignore-able.
+That is that we have a a set of variables satisfying the back door adjustment criteria.
 The goal is to estimate the average causal effect.
-Since we have assumed the backdoor criteia is met, the query is identifiable.
+Since we have assumed the backdoor criteria is met, the query is identifiable.
 Problem solved.
@@
 
-They have advantages and disadvantes depending on the setting. Methods like regression typically impose a parametric model of some description onto the treatment assignment or response function.
-
-- Connection of neighest matching to neighest neighbour.
-
-Once we have made the assumptions required to define the causal graph and identify a valid adjsutment set, is there any difference between the causal effect estimation problem and standard supervised machine learning.
+They have advantages and disadvantages depending on the setting. Methods like regression typically impose a parametric model of some description onto the treatment assignment or response function.
+
+- Connection of nearest matching to nearest neighbour.
+
+Once we have made the assumptions required to define the causal graph and identify a valid adjustment set, is there any difference between the causal effect estimation problem and standard supervised machine learning.
 
@@
 
-However, there is a critial remaining complication.
+However, there is a critical remaining complication.
 In supervised learning we can split our data into training and validation data sets or apply cross validation to select models and model parameters.
@@
 
-The siginificance to this should not be underestimated.
-Cross-validation has allowed applied machine learning to proceed with very atheoretical approach on the basis that we can identify when a model is sucessful.
-With causal effect estimation there is no guarantee that a model which performs well at prediction (even out of sample) will accuratly esitmate the outcome of an intervention.
+The significance to this should not be underestimated.
+Cross-validation has allowed applied machine learning to proceed with very a theoretical approach on the basis that we can identify when a model is successful.
+With causal effect estimation there is no guarantee that a model which performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
 
@@
 
-\subsection*{Non identifiable queries}
+\subsection{Non identifiable queries}
 In many cases the query of interest will not be identifiable.
@@
 
-Instrumental variables are often conceptualised as natrual experiments.
+Instrumental variables are often conceptualised as natural experiments.
 The goal is to find a variable that substantially influences the treatment but that is conditionally independent of the remaining variables in the network.
 
-For example, we might want to compare the results for juvenile offenders depending on wheter they spent their time awaiting trial in a custom facility for young offenders as opposed to a unit within an adult jail.
-The process by which children are assinged to different facilities may well be non-random.
-However, it might be that in many cases the decison was based purely on the fact that the juvenile facilty was full.
+For example, we might want to compare the results for juvenile offenders depending on whether they spent their time awaiting trial in a custom facility for young offenders as opposed to a unit within an adult jail.
+The process by which children are assigned to different facilities may well be non-random.
+However, it might be that in many cases the decision was based purely on the fact that the juvenile facility was full.
 We could then use the number of places remaining in the juvenile facility as an instrumental variable.
@@
 If we assume linearity it ?
-\todo{what exact guarentees do we get for instrumental variables}.
-Without a linearity assumption we would want the instrumental variable to be highly predictive of the treatement assignment so as to obtain reasonalble bounds.
-But if we make the assumption that everthing is linear then it would seem that even a weak instrumental variable should (in theory) surfise.
-
-In some cases we might have a choise between adjuting for causal effects via the backdoor theorem or utilizing an instrumental variable to obtain bounds.
-It is not the case that the former method is always superior even though it has better asymtpoptic results.
+\todo{what exact guarantees do we get for instrumental variables}.
+Without a linearity assumption we would want the instrumental variable to be highly predictive of the treatment assignment so as to obtain reasonable bounds.
+But if we make the assumption that everything is linear then it would seem that even a weak instrumental variable should (in theory) suffice.
+
+In some cases we might have a choice between adjusting for causal effects via the backdoor theorem or utilising an instrumental variable to obtain bounds.
+It is not the case that the former method is always superior even though it has better asymptotic results.
 For example
@@
 DEMONSTRATE a specific example where instrumental variables obtain better estimates than adjustment even when all the criteria for adjustment have been met.
-Start with a case were the adjustment is very high dimensional and the instrumetal variable is strong.
-(the effect of the varibles on the treatment assingment will have to be week (otherwise the instrument is not strong).
-
-\subsection*{How successful overal is causal effect estimation}
+Start with a case were the adjustment is very high dimensional and the instrumental variable is strong.
+(the effect of the variables on the treatment assignment will have to be week (otherwise the instrument is not strong).
+
+\subsection{How successful overall is causal effect estimation}
 Really know one knows.
 
-There are relatively few studies that have compared the observational study results with those obtained by randomized experiments.
+There are relatively few studies that have compared the observational study results with those obtained by randomised experiments.
 One of the key examples is the XX job network study.
-Initial analysis suggested that a range of causal estimation technquies all ...
-
-\paragraph*{How can we construct datasets on which to test casual effect estimation techniques}
-One can start from a randomized experiment and deliberately drop some of the data. Draw up the graphical model for this process and the assumptions inherent in it.
-
-\section*{Causal Discovery}
+Initial analysis suggested that a range of causal estimation techniques all ...
+
+\paragraph*{How can we construct data sets on which to test casual effect estimation techniques}
+One can start from a randomised experiment and deliberately drop some of the data. Draw up the graphical model for this process and the assumptions inherent in it.
+
+\section{Causal Discovery}
+
+Cyclic models \cite{Hyttinen2012}
+
 We now move to the much more general problem of learning a causal graph from observational data.
@@
 For example, that it is acyclic or that we have no unmeasured confounding variables.
-We do not assume the existance or directions of any links between the variables.
+We do not assume the existence or directions of any links between the variables.
 Amazingly, it is possible to infer some aspects of causal structure with such general assumptions.
-The set of conditional independences in a non-experimental data set indicates some causal structures are more likely than others.
+The set of conditional independence in a non-experimental data set indicates some causal structures are more likely than others.
 In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
@@
 
-There is also work on infering causal structure from experiental data \ref{}. We discuss this further in section XXX.
+There is also work on inferring causal structure from experiential data \ref{}. We discuss this further in section XXX.
 
 Causal discovery is a much grander goal than causal effect estimation given a known causal network.
-Arguably, if acheived, it would equate to the automation of scientific discovery.
-We need simply supply our algorithm with a vast collection of variables (regardless of their relevence to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make.
-Unfortunately, causal disovery is very hard.
+Arguably, if achieved, it would equate to the automation of scientific discovery.
+We need simply supply our algorithm with a vast collection of variables (regardless of their relevance to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make.
+Unfortunately, causal discovery is very hard.
 Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows XXX exponentially (or super-exponentially)?
 with the number of variables.
-- some intuition here about the strength of the signal we get from causal disovery.
+- some intuition here about the strength of the signal we get from causal discovery.
 Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}.
-Talk about the numerical as well as the statisitcal issues here.
+Talk about the numerical as well as the statistical issues here.
 
 In the next sections we briefly survey the key approaches to causal discovery.
-We roughly divide the methods into those based on those that exploit the connection between the conditional indpendences in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
+We roughly divide the methods into those based on those that exploit the connection between the conditional independencies in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
 This is not the only way we could have grouped them.
-There is also a seperation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
+There is also a separation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
 There are also hybrid methods \ref{}
 
-\subsection*{Conditional independence based methods}
+\subsection{Conditional independence based methods}
 One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
@@
 Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$.
-This set can also be referred to as the markov equivalence class compatible with $P$.
+This set can also be referred to as the Markov equivalence class compatible with $P$.
 
@@
 \textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
+\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
  \begin{enumerate}[itemsep=8pt]
-  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an undirected link to form a complete graph.
+  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an un-directed link to form a complete graph.
   \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent.
@@
   \label{alg:SGScolliders}  
-  \item Recursively try to orient any edges that remain undirected to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
+  \item Recursively try to orient any edges that remain un-directed to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
   \label{alg:SGSfinal}
@@
 
-The SGS algorithm utilizes the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
+The SGS algorithm utilises the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
 Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
@@
 
-The PC algorithm \cite{Sprites} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilize the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
+The PC algorithm \cite{Sprites} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilise the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
 It also checks for low order conditional independence relations before higher order ones.
@@
 \textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
+\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
  \begin{enumerate}[itemsep=8pt]
@@
 We can then bound the true causal effect by combining the results for the all the networks.
-This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
+This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularisation techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
 An implementation is available in the R package \cite{Kalisch2012}
@@
 With an unknown number of hidden variables there are infinity many possible structures to search over.
-In addition, the space of causal networks is not closed under marginalization.
-If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalizing over $\boldsymbol{U}$, may not be faithful to any bayesian network (see figure \ref{fig:DAGSnotclosed}).
+In addition, the space of causal networks is not closed under marginalisation.
+If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalising over $\boldsymbol{U}$, may not be faithful to any bayesian network (see figure \ref{fig:DAGSnotclosed}).
 
@@
 \centering
-\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
+\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalising over $U$. }
 \label{fig:DAGSnotclosed}
@@
 Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable.
-These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalization and conditioning.
+These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalisation and conditioning.
 
@@
 
-\todo{put these figures in a table with true graph input on left and fci output on right}
-\todo{uniform styles used for nodes accross thesis}
+\todo{put these figures in a table with true graph input on left and FCI output on right}
+\todo{uniform styles used for nodes across thesis}
 \begin{figure}
 \centering
-\caption{FCI examples: true graph and fci output}
+\caption{FCI examples: true graph and FCI output}
 \label{fig:FCIExamples}
@@
 
-With latent variables we are not using all the information - so we could go further (to nested markov models and inequalities.) \cite{Richardson2012} \cite{Shipster2014}
+With latent variables we are not using all the information - so we could go further (to nested Markov models and inequalities.) \cite{Richardson2012} \cite{Shipster2014}
 
@@
 This is non-trivial with high dimensional data.
-If the functional relationship between the variables is linear with gaussian noise then the network represents a multivaitae normal distribution and a pair of variables $A$ and $B$ are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero.
+If the functional relationship between the variables is linear with Gaussian noise then the network represents a multivariate normal distribution and a pair of variables $A$ and $B$ are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero.
 Where the functions are non-linear
@@
 
-\subsection*{Discovery with functional models}
+\subsection{Discovery with functional models}
 The algorithms we have considered so far return a Markov equivalence class.
@@
 Recall the functional definition of causality (section \ref{sec:SEM}).
-There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a}, linear models with non-gaussian noise \cite{Hoyer2012}
+There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a}, linear models with non-Gaussian noise \cite{Hoyer2012}
 
@@
 However, if the true causal model is $X \rightarrow Y$ then their is some function mapping values of $X$ to $Y$ which should be invariant to any changes in the input distribution $\P{X}$.
-Therefor $\P{X}$ should be indepdend of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning.
+Therefor $\P{X}$ should be independent of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning.
 However if the true causal model is $Y \rightarrow X$ then variations in the $P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help.
@@
 ONE POINT I'M NOT CERTAIN ON HERE.
-might not knowledge of $\P{X}$ help define which regoin of the mapping we need to be most accurate on in order to miniize the loss (ie ananalgous to the covariate shift probelelm).
+might not knowledge of $\P{X}$ help define which region of the mapping we need to be most accurate on in order to minimise the loss (ie analogous to the covariate shift problem).
 \todo {open question}
@@
 
-Indepdendence of input and mechanism. Daniusis et al 2012
+Independence of input and mechanism. Daniusis et al 2012
 
@@
 They assume there will be a difference between the relationship of $\P{X}$, $\P{Y}$ and $\P{X|Y}$ between $X \rightarrow Y$ versus $Y \rightarrow X$.
-Their algorithm requires a data set in which each row is itself a data set consisting of paris of varibles $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$.
-They use a keneral mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-dataset and train an algorithm to learn the direction of causality.
-Unfortunately we do not have a large collection of datasets where the causal direction is known to train such a model.
-\citep{LopezPaz2014} instead use a simulated dataset so thier model will nessesarily be base on the assumptions they make when generating the data.
+Their algorithm requires a data set in which each row is itself a data set consisting of pairs of variables $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$.
+They use a kernel mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-data set and train an algorithm to learn the direction of causality.
+Unfortunately we do not have a large collection of data sets where the causal direction is known to train such a model.
+\citep{LopezPaz2014} instead use a simulated data set so their model will necessarily be base on the assumptions they make when generating the data.
 Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.
-Their approach performed well in the causal effect pairs challange, correctly identifing the causal direction in ??
+Their approach performed well in the causal effect pairs challenge, correctly identifying the causal direction in ??
 percent of cases (against a benchmark of 50\% from random guessing)\ref{XXX}.
 
-\citep{Peters2014} have extended results from the bivariate case to the multivariate setting (with no hidden variables)?.
+\citep{Peters2014} have extended results from the bi-variate case to the multivariate setting (with no hidden variables)?.
 They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable.
@@
 
-\subsection*{Granger causality}
-
-Granger causality was developed within economics and is not typically discussed within the causal discovery liturature.
+\subsection{Granger causality}
+
+Granger causality was developed within economics and is not typically discussed within the causal discovery literature.
 
@@
 
-Granger causalty essentially assumes that the future cannot cause the past.
+Granger casualty essentially assumes that the future cannot cause the past.
 This may seem an entirely waterproof assumption.
@@
 
-\chapter*{The interventionalist viewpoint (5000 words)}
+\chapter{The interventionist viewpoint}
 
@@
 
-\subsection*{The role of randomization}
-Randomized controlled trials are often presented as the gold standard for determining causal effects in many areas of social sciences and economics \ref{XXX}.
-What is it about randomization that makes it so important when it comes to causality?
-The graphical model for a randomized controlled experiment is shown in figure \ref{fig:random_experiment_network}.
+\subsection{The role of randomisation}
+Randomised controlled trials are often presented as the gold standard for determining causal effects in many areas of social sciences and economics \ref{XXX}.
+What is it about randomisation that makes it so important when it comes to causality?
+The graphical model for a randomised controlled experiment is shown in figure \ref{fig:random_experiment_network}.
 If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
@@
 \begin{figure}[h]
-\caption{causal network for a randomized experiment}
+\caption{causal network for a randomised experiment}
 \label{fig:random_experiment_network}
@@
 \todo {turn this into something}
-This would be a natural place to discuss what happens with finite datasets where there are many other variables that influence the outcome.
+This would be a natural place to discuss what happens with finite data sets where there are many other variables that influence the outcome.
 We expect that across at least some of these variables, the target and control group will not be balanced.
 Does this not inject bias into the causal estimates?
-The anser is no, (I think - a better more general proof would be nice).
-The more such attibutes there are, the higher is the variance in the outcome within the target and control group.
+The answer is no, (I think - a better more general proof would be nice).
+The more such attributes there are, the higher is the variance in the outcome within the target and control group.
 When we perform a confidence test to ascertain whether or not there is a difference ...
@@
 
-Randomization does not ensure target and control group are exactly alike.
+Randomisation does not ensure target and control group are exactly alike.
 The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables in a between the target and control groups in a finite data sample.
@@
 
-What is the role of randomization? How do bandits algorithms work despite being only partially randomized? What else can you do to improve randomized studies (variance reduction, lower regret).
-
-The benifit provided by randomization in breaking the link between the treatment variable and any confounders should not be understated.
+What is the role of randomisation? How do bandits algorithms work despite being only partially randomised? What else can you do to improve randomised studies (variance reduction, lower regret).
+
+The benefit provided by randomisation in breaking the link between the treatment variable and any con founders should not be understated.
 The possibility of confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (test for confounding).
-This means causal estimates from non-experimental data are always subjec to the criticism that an important confounder may have been overlooked or not properly adjusted for.
-However, randomized experimentation does have some limitiations.
-
-\subsubsection*{Limiations of randomized experiments}
+This means causal estimates from non-experimental data are always subject to the criticism that an important confounder may have been overlooked or not properly adjusted for.
+However, randomised experimentation does have some limitations.
+
+\subsubsection{Limitations of randomised experiments}
 \label{subsec:limitations_of_experiment}
@@
 As the experiment proceeds we may start to see evidence that one treatment is much more effective than another.
-However (in a pure randomized experiment) we don't use this information to update the treatment that later patients recive.
-The result is that some will unessearily be given a sub-optimal treatment.
-\item Experiments are expensive and difficult to conduct. This means experimental datasets are often much smaller than observational ones, limiting the complexity of models we can explore.
-
-\item Experiments are often conducted on a convinient but unreprasentaive sample of the broader population of interest (for example first year collage students at research universities).
+However (in a pure randomised experiment) we don't use this information to update the treatment that later patients revive.
+The result is that some will unessential be given a sub-optimal treatment.
+\item Experiments are expensive and difficult to conduct. This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
+
+\item Experiments are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year collage students at research universities).
 This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}.
 The results may not carry over to the general population of interest.
-The question of whether an experiment conducted on one population can be mapped to another is refered to as the transportability problem \citep{Bareinboim2013} \todo{check ref} and relies on very similar assumtions and arguments to causal inference and the do-calculus.
-
-\item The idealized notion of an experiement represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomized experiments in practice.
-There may be imperfect compliance: the treatment selected by the randomizer is not always followed.
+The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} \todo{check ref} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
+
+\item The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
+There may be imperfect compliance: the treatment selected by the randomiser is not always followed.
 Or output censoring: the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If complience or attrition is not random but associated with (potentailly latent) variables that also effect the outcome then the problem of confounding bias returns.
-See figure XXX for a the graphical model of a randomized experiment with imperfect compliance and outcome censoring.
-
-\item It is not always possible or ethical to conduct a randomized experiment \cite{parashutes} \ref{Parachute use to prevent death and major trauma related to gravitational challenge: systematic review of randomised controlled trials}.
+If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.
+See figure XXX for a the graphical model of a randomised experiment with imperfect compliance and outcome censoring.
+
+\item It is not always possible or ethical to conduct a randomised experiment \cite{parachutes} \ref{Parachute use to prevent death and major trauma related to gravitational challenge: systematic review of randomised controlled trials}.
 \end{enumerate}
@@
 
-\section*{Multi armed bandits}
-\todo{is there a way in latex to have a command for refering to each type of thing like figure, subsection, etc such that it can be updated programatically?}
+\section{Multi armed bandits}
+\todo{is there a way in latex to have a command for referring to each type of thing like figure, subsection, etc such that it can be updated programatically?}
 
@@
 In order to obtain regret bounds, some assumptions are required on the distributions $\nu_i$.
-It sufficient to assume they are sub-gaussian.
-For each timestep $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$.
-The goal of the learner is to maximize the total reward they receive.
+It sufficient to assume they are sub-Gaussian.
+For each time step $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$.
+The goal of the learner is to maximise the total reward they receive.
 This problem introduces the fundamental exploration-exploitation trade-off.
@@
 The performance of bandit algorithms is generally described by the (pseudo) regret, $R(T)$.
-This is the difference between the expected reward obtained by the algorithm and the expected reward of selecting the best action in every timestep.
-
-\subsection*{Regret} How do we measure the performance of a bandit algorithm?
+This is the difference between the expected reward obtained by the algorithm and the expected reward of selecting the best action in every time step.
+
+\subsection{Regret} How do we measure the performance of a bandit algorithm?
 
@@
 
-Psuedo-Regret
+Pseudo-Regret
 \eqn{
@@
 
-An algorithm is learning if it obtains regret that is sublinear in $T$.
+An algorithm is learning if it obtains regret that is sub-linear in $T$.
 
@@
 The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm, $R_s(T) = \mu^* - \E{\mu_i}$ .
-This problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialization or exploitation phase.
+This problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialisation or exploitation phase.
 For example, many strategies might be assessed via simulation prior to one being selected and deployed.
@@
 
-\subsection*{The exploration/exploitation trade-off}
+\subsection{The exploration/exploitation trade-off}
 A central aspect to bandit problem is the trade of between exploiting the information we have already obtained with exploring options about which we remain uncertain.
-Suppose you have sampled five of your local restuarants and found one which you really enjoyed.
+Suppose you have sampled five of your local restaurants and found one which you really enjoyed.
 How often should you eat there versus exploring new options.
@@
 In many cases, the total number of rounds that you will get to play in the game is unknown.
-If you cannot imagine that it would be possible for a meal to be substantially better than that which you obtain (in expectation) at your current favoriate then there is little point exploring further.
-
-\subsection*{Key approaches and results}
+If you cannot imagine that it would be possible for a meal to be substantially better than that which you obtain (in expectation) at your current favourite then there is little point exploring further.
+
+\subsection{Key approaches and results}
 
@@
 This balances exploration and exploitation as an arm with a high upper confidence bound must have either a high expected reward or large uncertainty on the expected reward.
-Assume for notational simplicity that $\mu_1 > \mu_2 > ...
-> \mu_K$, such that $\mu^* = \mu_1$, and let $\Delta_i = \mu_i - \mu^*$ be the sub-optimality for each arm.
+Assume for rotational simplicity that $\mu_1 > \mu_2 > ...
+> \mu_K$, such that $\mu^* = \mu_1$, and let $\Delta_i = \mu_i - \mu^*$ be the sub-optimal for each arm.
 The (problem dependent) regret for UCB is bounded by:
@@
 If an action has a high upper bound on the possible reward then either the expected reward must be high or the algorithm must be very uncertain about it.
-Choosing it should yeild either a good reward or useful information.
-This is the key idea begind the UCB algorithm \citep{xxx}.
+Choosing it should yield either a good reward or useful information.
+This is the key idea behind the UCB algorithm \citep{xxx}.
 The remaining complexity lies in optimising how we compute confidence intervals to trade of exploration and exploitation as the game proceeds.
@@
 
-\subsection*{The need to add structure}
+\subsection{The need to add structure}
 The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.
-This makes problems with large or infinite actions spaces intractible.
+This makes problems with large or infinite actions spaces intractable.
 REFERENCE equation showing linearity with arms.
@@
 However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalized linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
-
-\subsection*{Key related settings}
-
-\subsubsection*{Adversarial Bandits}
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and Gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
+
+\subsection{Key related settings}
+
+\subsubsection{Adversarial Bandits}
 In the classical multi-armed bandit problem, the rewards for each arm are sampled stochastically from a fixed (but unknown) distribution.
@@
 \item Different notation of regret
-\item Adversarial bandits are not imume to issues of feedback cycles or drift in the reward distribution because a constancy is implicit in the definition of regret. 
+\item Adversarial bandits are not immune to issues of feedback cycles or drift in the reward distribution because a constancy is implicit in the definition of regret. 
 \end{itemize}
 
-\subsubsection*{Contexual bandits}
+\subsubsection{Contextual bandits}
 In the previous sections each decision was identical.
-If we were imaging treating patients, we assume that we have no additional information about each patient that is relevent to deciding how to treat them.
+If we were imaging treating patients, we assume that we have no additional information about each patient that is relevant to deciding how to treat them.
 However, in many key applications such as serving ads, etc, etc.
@@
 
-The formal defintion of a contextual bandit problem is
-
-The simplest way to generalise the standard bandit algorithms to the contextual case (where the context is discrete) is to simply have a seperate standard bandit instance for each possible setting of the context.
-Regretably this approach scales terribly with increasing complexity of the context.
+The formal definition of a contextual bandit problem is
+
+The simplest way to generalise the standard bandit algorithms to the contextual case (where the context is discrete) is to simply have a separate standard bandit instance for each possible setting of the context.
+Regrettably this approach scales terribly with increasing complexity of the context.
 \todo{quantify this}.
-It cannot be applied at all if the context is continous.
-
-The approach to solving this problem parallels those from supervised learning and continous armed bandits.
+It cannot be applied at all if the context is continuous.
+
+The approach to solving this problem parallels those from supervised learning and continuous armed bandits.
 We make assumptions about the smoothness of the problem.
@@
 
-It should be noted that there is an important difference between the approaches we can apply to the contextul bandit versus the supervised learning setting.
-As for the casual effect estimation problem, we cannot utilize cross-validation offline to select model parameters.
-This makes contenxtual bandit algorithms degrade faster than supervised learning ones as we add irrelevent variables.
+It should be noted that there is an important difference between the approaches we can apply to the contextual bandit versus the supervised learning setting.
+As for the casual effect estimation problem, we cannot utilise cross-validation offline to select model parameters.
+This makes contextual bandit algorithms degrade faster than supervised learning ones as we add irrelevant variables.
 \todo{ask Tor if he thinks this is correct}
 
-RUN A SIMULATION COMPARIGN HOW supervised learning and contextual bandits degrade as we add more features. What would be a fair comparison?
-
-\subsubsection*{Markov decision processes}
+RUN A SIMULATION COMPARING HOW supervised learning and contextual bandits degrade as we add more features. What would be a fair comparison?
+
+\subsubsection{Markov decision processes}
 In the bandit setting the reward distributions for each arm are assumed to be fixed.
-A much studied generalisation is the markov decision process or MDP.
-In an MDP we assume there the enviroment has some state.
-The reward an agent recieves depends on the state of the enviroment and the action the agent selects.
-The enviroment then evolves stochasitcally depending on the agents action.
-Multi-armed bandits can be considered a single state MDP, where no matter what action the agent selects, the enviroment returns to that state.
+A much studied generalisation is the Markov decision process or MDP.
+In an MDP we assume there the environment has some state.
+The reward an agent receives depends on the state of the environment and the action the agent selects.
+The environment then evolves stochastically depending on the agents action.
+Multi-armed bandits can be considered a single state MDP, where no matter what action the agent selects, the environment returns to that state.
 
@@
 
-\subsubsection*{Dynamic Systems}
+\subsubsection{Dynamic Systems}
 \begin{itemize}
@@
 
-\chapter*{Causal Bandits: Unifying the approaches (5000 words)}
-
-\section*{The framework}
+\chapter{Causal Bandits: Unifying the approaches}
+
+\section{The framework}
 A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
@@
 We refer to these problems as \emph{causal bandit problems}.
-Extending this work to simultaneously learning the casusl graph is discussed in section \ref{XXX}.
-
-\subsubsection*{Definition of casual bandit problem} \todo{make proper definition}
+Extending this work to simultaneously learning the casual graph is discussed in section \ref{XXX}.
+
+\subsubsection{Definition of casual bandit problem} \todo{make proper definition}
 A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed actions} $\mathcal{A}$.
@@
 What information is available to the decision making agent before and after they select an action?
-The causal bandit problem takes on characteristics of different bandit settings depending on the assumptions we make about what actions are available to the agent, what variables are obvserved and whether they are observed before or after the action is chosen.
+The causal bandit problem takes on characteristics of different bandit settings depending on the assumptions we make about what actions are available to the agent, what variables are observed and whether they are observed before or after the action is chosen.
 
@@
 We only get this form of side information for actions that don't specify the value of every variable, ie those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-Since the reward distribution for actions that set a subset of the variables is the result of marginalizing out other variables, they can only be optimal if they have lower cost.
+Since the reward distribution for actions that set a subset of the variables is the result of marginalising out other variables, they can only be optimal if they have lower cost.
 So if the cost of all actions is constant (no matter how many variables must be set), then the problem has the same characteristics as if only the reward node were observable.
@@
 
-\section*{Related work}
+\section{Related work}
 \begin{enumerate}
@@
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
-\node[node](1){Health concious};
+\node[node](1){Health conscious};
 \node[node, below left=of 1](2){Diet};
@@
 
-\section*{Causal bandits with post action feedback}
+\section{Causal bandits with post action feedback}
 WHY THIS PROBLEM. This work was presented at NIPS 2016 \ref{}.
-\todo{add relevent stuff from introduction section}
-
-\subsection*{Notation}
-
-We will assume each variable only takes on a finite number of distinct values. (The path to relaxing this assumption would be through leverging the work on continuous armed bandits).
+\todo{add relevant stuff from introduction section}
+
+\subsection{Notation}
+
+We will assume each variable only takes on a finite number of distinct values. (The path to relaxing this assumption would be through levering the work on continuous armed bandits).
 
@@
 
-\subsection*{Definition of causal bandit game with post-action feedback}
+\subsection{Definition of causal bandit game with post-action feedback}
 \todo{ re-read and incorporate response from reviewer feedback for NIPS}
@@
 
-The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.
+The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refereed to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.
 
@@
 
-\subsection*{The parallel bandit problem}
+\subsection{The parallel bandit problem}
 
@@
 
-$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret.
+$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimise the worst-case simple regret.
 When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
-We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$.
+We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilises the samples captured during the observational phase to estimate $m(\vec{q})$.
 Although very simple, the following two theorems show that this algorithm is effectively optimal.
@@
 %We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
-By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
+By utilising knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
 Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$.
@@
 
-\subsection*{General graphs}
+\subsection{General graphs}
 We now consider the more general problem where the graph structure is known, but arbitrary.
 For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \citep{Pearl2000}.
+However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorisation formula \citep{Pearl2000}.
 \eq{
@@
 } 
-where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
-
-We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to 
+where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the Dirac delta function.
+
+We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorisation to 
 write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
@@
 Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
-Let $\parents{Y}(X)$ denote the realization of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
+Let $\parents{Y}(X)$ denote the realisation of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
 
@@
 
-The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
+The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimised to minimise the worst-case simple regret. Abusing notation we define $m(\eta)$ by
 \eq{
@@
 
-\subsection*{Experiments}
+\subsection{Experiments}
 We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \cite{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
-Thomson sampling and UCB are optimized to minimize cumulative regret.
+Thomson sampling and UCB are optimised to minimise cumulative regret.
 We apply them in the fixed horizon, best arm identification setting by running them upto horizon $T$ and then selecting the arm with the highest empirical mean.
@@
 Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
-This gap is learnable by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
+This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
 In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
@@
 
-\subsection*{Additional experiments} \todo {tie section together}
+\subsection{Additional experiments} \todo {tie section together}
 \begin{figure}[H]
@@
 
-\subsection*{Discussion \& Future work}
+\subsection{Discussion \& Future work}
 
@@
 Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
-This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
+This would likely improve the poor performance of our causal algorithm relative to the successive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
 
@@
 Note that these modifications will not improve the minimax regret, which at least for the parallel bandit is already optimal.
-For this reason we prefer to emphasize the main point that causal structure should be exploited when available.
+For this reason we prefer to emphasise the main point that causal structure should be exploited when available.
 Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
@@
 
-The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
+The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each time step the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
 The feedback graph is distinct from the causal graph.
@@
 A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
-The latter case arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques. 
+The latter case arises naturally if we assume free access to a large observational data set, from which the Markov equivalence class can be found via causal discovery techniques. 
 Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014b} could potentially be incorporated into a causal bandit algorithm.
 In particular, \citet{Hu2014b} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
-Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational dataset would be much more challenging.
-
-\subsection*{Proofs}
-
-\chapter*{List of software programs for causal inference}
+Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational data set would be much more challenging.
+
+\subsection{Proofs}
+
+\chapter{List of software programs for causal inference}
 
@@
 
-\chapter*{Causality \& Fairness}
-
-As machine learning is incorporated into decision making systems that have fundamental impacts on people’s lives, such as in employment, criminal justice, health and financial services, there are increasing concerns over transparency and fairness.
-
-The European Union’s General Data Protection Regulation \cite{XXXX}, due to come into effect in 2018, requires that people can obtain "meaningful information about the logic involved" in an automated decision process.
+\chapter{Causality \& Interpretability}
+
+\chapter{Causality \& Fairness}
+
+As machine learning is incorporated into decision making systems that have fundamental impacts on people’s lives, such as in employment, criminal justice, health and financial services, there are increasing concerns over transparency and fairness \cite{WMD, etc}.
+Realisation that machine learning algoithms can be inherit biases from the data we feed into them and the choices those building them make about what variables to enclude, etc.
+
+The European Union’s General Data Protection Regulation \cite{Goodman2016}\todo{check}, due to come into effect in 2018, requires that people can obtain "meaningful information about the logic involved" in an automated decision process .
 It also stipulates that such decisions should not be based on special categories of personal data (related to ethnicity, political and religions affiliation or sexuality) unless there "suitable measures" to safeguard individual interests.
@@
 
-Discrimination is frequently defined in terms of either ’disparate treatment’- treating otherwise similar individuals differently on the basis of a protected attribute such as race or gender, or ’disparate impact’ - a substantial difference in outcome between groups.
+Discrimination is frequently defined in terms of either disparate treatment- treating otherwise similar individuals differently on the basis of a protected attribute such as race or gender, or disparate impact - a process that yields a significant difference in outcome between groups.
 We consider how the notions of disparate treatment and impact may be formalised and the implications of how this is done for machine learning.
 We examine the overlap between the motivations for interpretable and causal models, especially with respect to assessing fairness.
-We look at how causal models mitigate some of the trade-offs between transparency and predictive accuracy and we examine to
-what extent the causal relationship between an attribute, any protected attributes and the outcome of interest is relevant to assessing the impact on fairness of its inclusion in a machine learning model.
-
+We look at how causal models mitigate some of the trade-offs between transparency and predictive accuracy and we examine to what extent the causal relationship between an attribute, any protected attributes and the outcome of interest is relevant to assessing the impact on fairness of its inclusion in a machine learning model.
+
+The lack of part-time work in tech could be argued to constitute indirect discrimination against women. 
+A fear that advertising themselves as supporting flexible working options would attract candidates who lacked drive and ambition.
+Cultural bais that favours those driven by monitary ambition against those
+
+Stability - a non causal model can't tell us about a simple do type opperation on  a single variable.
+
+More broadly, we could draw a model representing the system now that could tell us the result of an intervention on a particular variable (such as setting gender), but the system itself could change (for example with customer preferences).
+There is a notion that a true causal model should be invariant for all time.
+
+Connection to Simpson's paradox.
+But what should we condition on?
+\cite{Romie2010} argue that looking at the department level is the correct viewpoint, since this the point at which hiring decions were made.
+However, it could be argued that if one department was attracting a large number of higher quality candidates, the overall size of that deparmtment should be higher and that the results seen at Berkley reflected a bias in favour of male dominated fields.
+
+Different measures of discrimination - is there a difference between groups (after conditioning on xxx)
+
+To understand how we measure and penalise discrimination, we need to take a step back and ask what are the underlying motivations for fairness?
+
+These differences in underlying cause suggest differences in the approach we should take to remedy them.
+
+\section{How does discrimination arise in ML models}
 \begin{itemize}
-\item There is an additional connction between causality and fairness, that arises when you know that there is a bias in the data and wish to correct for it.
+\item bias in historical decisions fed in as training data
+\item selection bias in input data, due to deliberate or implicit discrimination such as stop and search
+\item deliberate manipulation on the part of the person building the model.
+\item The hardest case is historical disadvantage, creating genuine differences between relevent attributes of groups.
+\item in bias in the label - what happens if you hire minorities but then your staff treat them badly and as a result they underperform or leave.  
+\end{itemize}
+
+\section{Defining Descrimination}
+
+Define discrimination in terms of causal effect of protected variable in decision making process.
+
+\cite{Romei2012} note the relevence of causal inference on discrimintation analysis.
+
+“the central question in any employment- discrimination case is whether the employer would have taken the same action had the employee been of 7A recurring problem known as the omitted-variable bias.
+A multidisciplinary survey on discrimination analysis 9
+a different race (age, sex, religion, national origin etc.) and everything else had been the same” (Carson v. Bethlehem)
+
+Let us mathematically define disparate treatment and disparate impact with respect to a statistical model.
+We will focus on discrete variables for notational simplicity.
+Assume we have an outcome of interest $Y$, protected attributes $X$, other covariates $Z$ and a (potentially stochasitic) model $f$ that maps $\{x, z\}$ to $y_f \in Y$.
+For a given model $f$, we have a distribution over the predicted outputs given the inputs, $\P{Y_f|X,Z}$
+\vspace{.2cm}
+\begin{definition}{Disparate Impact}: A model, $f$, that produces a predictive distribution $\P{Y_f|X,Z}$ has disparate impact if the marginal distribution over the predicted outcome, $Y_f$, depends on a protected attribute.
+\eqn{
+\exists \set{x_1,x_2} \subset X: \sum_z{\P{Y_f|Z,x_1}\P{Z|x_1}} \neq \sum_z{\P{Y_f|Z,x_2}\P{Z|x_2}}
+}
+\end{definition}
+
+\begin{definition}{Disparate Treatment}: A model yeilds disparate treatment if people with identical attributes (excuding protected attributes) are treated differently. 
+\eqn{
+\P{Y_f|Z,X} \neq \P{Y_f|Z}
+}
+\end{definition}
+
+Avoiding treating people differently purely on the basis of attributes such as ethnicity and gender and avoiding large differences in important outcomes such as education and income between such groupings both seem like desirable goals.
+Unfortunately, in general, they conflict with one another, see theorem \ref{thm:disparate_conflict}.
+Any variable we might wish to aviod discriminating on will be correlated to some other measurable covariate $Z$, making it impossible to avoid both disparate treatment and disparate impact.
+\vspace{.2cm}
+\begin{theorem} 
+\label{thm:disparate_conflict}
+Disparate impact and disparate treatment conflict.
+Given covariates $Z$ and protected attributes $X$, a model, $f$, cannot be fair with respect to both disparate impact and disparate treatment unless $\P{Z|X} = \P{Z}$.
+\begin{proof} Assume $f$ yeilds no different treatment, then $\P{Y_f|Z,X} = \P{Y_f|Z}$.
+If we additionally assume no disparate impact, $\sum_z{\P{Y_f|Z}\P{Z|x_1}} = \sum_z{\P{Y_f|Z}\P{Z|x_2}} \;\; \forall \set{x_1,x_2}\subset X$.
+This holds if and only if $\P{Z|x_1} = \P{Z|x_2} \;\; \forall \set{x_1,x_2} \implies \P{Z|X} = \P{Z}$
+\end{proof} 
+\end{theorem}
+
+Further issues.
+
+Disparate treatment, with respect to an observed set of variables $Z$, can be trivially avoided by excluding protected attributes from the training data.
+
+Problems:
+\begin{itemize}
+\item Disparate impact and disparate treatment conflict
+\item 
+\end{itemize}
+
+\section{Addressing discrimination}
+
+Ommititting the protected attribute from the model can increase disparate impact, even if the protected attribute is negatively correlated with the outcome.
+
+\textit{Disparate treatment} can be trivially avoided by excluding the protected variable from the model.
+However, this is deeply unsatisfying given the presence of proxy variables and can increase bias (figure \ref{fig:recidivism})
+
+Avoiding \textit{disparate impact} may be be expensive in terms of predictive accuracy and/or require \textit{disparate treatment}.
+
+\begin{figure}[H]
+\caption{Proxies}
+\label{fig:recidivism}
+	\centering    
+          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,
+  thick,main node/.style={observedrect}, lt/.style={latentrect} , hidden/.style={empty},background rectangle/.style={fill=olive!45}]
+ %nodes
+\node[lt](1){Actual priors};
+\node[main node, below left=of 1](2){Recorded Priors};
+\node[main node, below right=of 1](3){Re-offend};
+\node[lt, above left=of 2](4){Stop and Search};
+\node[main node, above=of 4](5){Race};
+\node[main node, above=of 1](6){Education};
+ \path[every node/.style={font=\tiny}]
+    (1) edge (2) edge (3)
+    (4) edge (2)
+    (5) edge (4) edge (6)
+    (6) edge (1);
+\end{tikzpicture}
+\end{figure}
+
+\begin{itemize}
+\item There is an additional connection between causality and fairness, that arises when you know that there is a bias in the data and wish to correct for it.
 In causality, it may be due to selection bias or a confounding variable.
-In fairness, it may be due to historical disadvanage.
+In fairness, it may be due to historical disadvantage.
 In both cases we may have theory that tells us the direction or scale of an effect.
 Ie, we know that race itself has no inherent effect on outcomes such as criminality and education.
-If you have a dataset and a collection of such priors, where the data contradicts these priors due to data bias.
+If you have a data set and a collection of such priors, where the data contradicts these priors due to data bias.
 How do you best make decisions based on that data?
@@
 
-\chapter*{Causality in Marketing}
-Assessing the impact of marketing spend is becoming increasinly important to industly.
-Billions of dollars \cite{XXX} are spent each year and new marketing channels and oportunities are openening up at an unprecented rate.
-
-There are three key approaches to assessing the impact of marketing in the liturature.
-
-The first is to look for correlations across companies on metrics such as how much they spend on marketing \ref{}, how confident they are about their ability to market effectively \ref{} or how important the marketing team is percieved to be within the orgainisation with indicators of the health of the company such as sales, revenue or market capitalisation.
-Unfortuneatly is is very difficult to avoid potentail major confounders with this approach.
-Are businesses large and succesful because of their capable marketing teams or can they afford to hire capable marketers because they have sufficient profits available to do so.
-
-\section*{Attribution}
+human decicion making can evolve.
+
+The vaugeness inherent in writing laws in natural language allows for them to be re-interpreted.
+
+There is a lack of diversity in models.
+(although radomness is in fact inherent in many) - could one leverage this?
+- you would need an estimate of the probability the model would have assigened each instance to a particular group.
+
+Only humans can provide the ethics
+
+\chapter{Causality in Marketing}
+Assessing the impact of marketing spend is becoming increasingly important to industry.
+Billions of dollars \cite{XXX} are spent each year and new marketing channels and opportunities are opening up at an unprecedented rate.
+
+There are three key approaches to assessing the impact of marketing in the literature.
+
+The first is to look for correlations across companies on metrics such as how much they spend on marketing \ref{}, how confident they are about their ability to market effectively \ref{} or how important the marketing team is perceived to be within the organisation with indicators of the health of the company such as sales, revenue or market capitalisation.
+Unfortunately is is very difficult to avoid potential major confounders with this approach.
+Are businesses large and successful because of their capable marketing teams or can they afford to hire capable marketers because they have sufficient profits available to do so.
+
+\section{Attribution}
 The second is attribution.
-The goal in this case is to attribute each sale to a specific (or set of specfic) advertisements that the customer was exposed to.
-This approach is the primary way of measuring the sucess of digital advertising strategies such as search, display, and online video ads.
+The goal in this case is to attribute each sale to a specific (or set of specific) advertisements that the customer was exposed to.
+This approach is the primary way of measuring the success of digital advertising strategies such as search, display, and online video ads.
 The most frequent model is last-click attribution.
@@
 A central criticism of attribution is that it is not measuring incremental sales.
-Each sales is assumed to be a consequence of some form of advertising for those customers who saw (however briefly) some form of advertisicment.
-Of course in reality many of these customers may have made a purshcase regardless of the advertising material to which they were exposed.
-
-System designed to optimsie metrics such as click through or conversion rate can do so either by learning how to change the consumers preferences (by serving them the perfect ad at the ideal time) or by learning who was most likely to buy anyway.
-It seems likley that much of the improvemt in click through rates due to the application of sophisticated machine learning algorithms is beign driven by the latter.
-Unfortuenatly, this does nothing for the bottom line of industry (except for those involved in servering ads).
-
-\section*{Econometric Modelling}
-The final key plank businesses apply to estimate how effective thier marketing are econometric or mixed media models. These models
-
-In practise these models are typically timeseries regression models, fitting some measure of sales against marketing spend (by marketing channel, ie TV, radio, search, digitial display etc) and other key variables such as pricing, competitor spend and pricing and indicators of the health of the economy and demand for the product in question.
+Each sales is assumed to be a consequence of some form of advertising for those customers who saw (however briefly) some form of advertisement.
+Of course in reality many of these customers may have made a purchaser regardless of the advertising material to which they were exposed.
+
+System designed to optimise metrics such as click through or conversion rate can do so either by learning how to change the consumers preferences (by serving them the perfect ad at the ideal time) or by learning who was most likely to buy anyway.
+It seems likely that much of the improve mt in click through rates due to the application of sophisticated machine learning algorithms is being driven by the latter.
+Unfortunately, this does nothing for the bottom line of industry (except for those involved in serving ads).
+
+\section{Econometric Modelling}
+The final key plank businesses apply to estimate how effective their marketing are econometric or mixed media models. These models
+
+In practise these models are typically time series regression models, fitting some measure of sales against marketing spend (by marketing channel, ie TV, radio, search, digital display etc) and other key variables such as pricing, competitor spend and pricing and indicators of the health of the economy and demand for the product in question.
 
 The models may also include non-linear transformation of the marketing inputs to represent saturation and decay of time of effects.
-They may also consider the interaction between media chanels and other business levers such as pricing.
+They may also consider the interaction between media channels and other business levers such as pricing.
 
@@
 
-\chapter*{Conclusions (1000 words)}
-
-\section*{Open questions}
+\chapter{Conclusions (1000 words)}
+
+\section{Open questions}
 \paragraph{Cycles} - a huge issue. Not covered by Pearl,Rubin etc.
@@
 Fung and Crawford (1990) have recently proposed a fast algorithm for constructing an
-independence graph from discrete data. We have not tested their procedure as a preprocessor for
+independence graph from discrete data. We have not tested their procedure as a processor for
 the PC algorithm. (COPIED FROM SPRITES)

### 2017-05-18 causal-bandit/nips_2016/code/understanding_regret.py

@@ -1,0 +1,7 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Apr 25 16:19:31 2017
+
+@author: finn
+"""
+

### 2017-05-18 causalpy/causalpy/bandits/__init__.py

@@ -1,0 +1,2 @@
+from algorithms import *
+from enviroments import *

### 2017-05-18 causalpy/causalpy/bandits/algorithms.py

@@ -7,7 +7,7 @@ Created on Tue Sep 13 11:15:43 2016
 
 import numpy as np
 from math import sqrt, log
-from models import Parallel,ScaleableParallelConfounded
+from enviroments import Parallel
 
 
 def argmax_rand(x):
@@ -128,38 +128,6 @@ class UCB(object):
         self.best_action = argmax_rand(mu)
         return max(model.expected_rewards) - model.expected_rewards[self.best_action]
         
-class LilUCB(UCB):
-    """ 
-    Implementation based on lil’UCB : An Optimal Exploration Algorithm for Multi-Armed Bandits
-    Jamieson et al, COLT 2014.
-    
-    To convert the algorithm from fixed confidence to fixed budget, we use the halving trick.
-    """
-    label = "LilUCB"
-    def __init__(self,epsilon,gamma,beta,delta,sigma = .5):
-        
-        self.e = epsilon
-        self.g = gamma
-        self.b = beta
-        self.d = delta
-        self.c1 = (1+beta)*(1+sqrt(epsilon))
-        self.c2 = 2*(sigma**2)*(1+epsilon)
-        
-    @classmethod
-    def theoretical(cls):
-        """ Returns LilUCB with parameters supported by theory (from Jameison et al) """
-        
-        return cls(epsilon = .01,gamma = 9.0,beta = 1.0)
-    
-    @classmethod
-    def heuristic(cls):
-        """ Returns LilUCB with parameters heuristically shown to perform well (from Jamieson et al)"""
-        
-    def upper_bound(self,t):
-        mu = np.true_divide(self.success,self.trials)
-        ratio = np.true_divide(self.c2*np.log(np.log((1+self.e)*self.trials)/self.delta),self.trials)
-        interval = self.c1*np.sqrt(ratio)
-        return mu+interval
         
 
 class AlphaUCB(UCB):
@@ -256,39 +224,3 @@ class RandomArm(object):
     def run(self,T,model):
         self.best_action = np.random.randint(0,model.K)
         return max(model.expected_rewards) - model.expected_rewards[self.best_action] 
-
-if __name__ == "__main__":  
-    
-    
-    
-    N =5
-    N1 = 1
-    pz = .1
-    pz = .2
-    q = (.1,.9,.2,.8)
-    pY = np.asarray([[.4,.4],[.7,.7]])
-    epsilon=.1
[…]

### 2017-05-18 causalpy/causalpy/bandits/enviroments.py

@@ -39,7 +39,7 @@ def prod_all_but_j(vector):
 
     joint = np.prod(vector)
     return np.true_divide(joint,vector)
-
+    
 class BasicBernoulliModel(object):
     """ A model for generating data for standard bandits without side information """
     
@@ -48,19 +48,24 @@ class BasicBernoulliModel(object):
         expected_rewards: an array of length K containing the expected rewards for each arm.
         """
         self.expected_rewards = expected_rewards
+        self.K = len(self.expected_rewards)
                 
     @classmethod
-    def create_worst_case(cls,K,T):
-        epsilon = np.sqrt(K/(4.0*T))
+    def create_epsilon_best(cls,K,epsilon):
+        """ creates a bandit where all arms have reward .5 except the last, which has .5 + epsilon """
         expected_rewards = np.full(K,.5)
         expected_rewards[-1] = .5 + epsilon # reward of one arm is just better than the others.
         return cls(expected_rewards)
+    
+    @classmethod
+    def create_worst_case(cls,K,T):
+        epsilon = np.sqrt(K/(4.0*T))
+        return cls.create_epsilon_best(K,epsilon)
         
     
     def sample_multiple(self,actions,n):
         """ draws n samples from the reward distributions of the specified actions. """
         return binomial(n,self.expected_rewards[actions])
-             
 
 class Model(object):
                  
@@ -722,8 +727,6 @@ class ScaleableParallelConfoundedNoZAction(ScaleableParallelConfounded):
         return result
 
 
-from itertools import chain
-
 def estimate_px_and_y_from_samples(model,samples):
         expected_y = np.zeros(model.K,dtype=float)
         shape = list(chain([model.K],[2]*model.N))
@@ -740,166 +743,3 @@ def estimate_px_and_y_from_samples(model,samples):
         xcounts = xcounts/samples
         return xcounts,expected_y
         
-if __name__ == "__main__":  
-    import numpy.testing as np_test
-    import time
-    from pgmpy_model import GeneralModel
-    
-    N = 5
-    N1 = 2
-    N2 = N-N1
[…]

### 2017-05-18 causalpy/causalpy/bandits/regret.py

@@ -1,0 +1,15 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Apr 25 16:57:12 2017
+
+@author: finn
+"""
+
+def regret(algorithm,enviroment):
+    """ standard cumulative regret for a particular run of an algorithm. """
+    selected_arms = algorithm.trials # the counts of trials for each arm. 
+    T = selected_arms.sum() 
+    expected_algorithm_reward = enviroment.expected_rewards 
+    best_reward = T*enviroment.expected_rewards.max()
+    
+    

### 2017-05-18 causalpy/causalpy/bayesian_networks/models.py

@@ -1,5 +1,156 @@
 import numpy as np
 from sympy import Symbol, Matrix
+from itertools import product
+
+def set_difference(lst1,lst2):
+    """ returns the elements and indicies of elements in lst1 that are not in lst2"""
+    elements = []
+    indicies = []
+    for indx,item in enumerate(lst1):
+        if item not in lst2:
+            elements.append(item)
+            indicies.append(indx)
+    return elements,indicies
+    
+def to_symbolic(lst):
+    all_numeric = True
+    result = []
+    for item in lst:
+        try:
+            result.append(Symbol(item))
+            all_numeric=False
+        except TypeError:
+            result.append(item)
+    return result,all_numeric
+    
+
+class DiscreteFactor(object):
+    def __init__(self,variables,card,data):
+        """
+        variables: name of variables in factor
+        card: cardinality of variables (in same order as variables list)
+        data: data for the factor
+        """
+        if len(variables) != len(card):
+            raise ValueError("variables and card must have the same length")
+        expected_len_data = np.prod(card)
+        if len(data) != expected_len_data:
+            raise ValueError("length of data array inconsistent with cardinality, should be {0}".format(expected_len_data))
+        self.variables = variables
+        self.card = card
+        self.data,_ = to_symbolic(data)
+        assignments = product(*[range(c) for c in card])
+        var_assignments = [sorted(zip(variables,a),key=lambda x:x[0]) for a in assignments]
+        keys = ["".join([s[0]+str(s[1]) for s in assignment]) for assignment in var_assignments]
+        self.key2index = dict(zip(keys,range(len(data))))
+        
+    def __str__(self):
+        return self.data.__str__()
+
+    def probability_of(self,assignment):
+        """
+        assignment: list of tuples (variable, value)
+        return the probability of this assignment
+        """
+        key = "".join([s[0]+str(s[1]) for s in sorted(assignment,key=lambda x:x[0])])
+        indx = self.key2index[key]
+        return self.data[indx]
+        
+    def multiply(self,other):
+        # TODO check cardinality of matching vars is the same
+        new_vars,indices = set_difference(other.variables,self.variables)
+        new_card = [other.card[i] for i in indices]
+        result_vars = self.variables+new_vars
+        result_card = self.card+new_card
+        l = len(self.variables)
+        other_var_indicies = [i for i,var in enumerate(result_vars) if var in other.variables] 
[…]

### 2017-05-19 thesis/offcuts.tex
draft of chapter 2 complete

--- 2017-04-30 thesis/thesis_map.tex
@@
-\documentclass[11pt,a4paper,oneside]{book}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{dsfont}
-\usepackage{listings}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{tabularx}
-\usepackage{enumitem}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithm
-\usepackage[noend]{algorithmic}
-\usepackage{parskip}
-\usepackage{tikz}
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{epigraph}
-\usetikzlibrary{arrows,positioning} 
-\usepackage{subcaption}
-\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[numbers]{natbib}
-\usepackage{wrapfig}
-
-\graphicspath{ {figures/} }
-\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
-
-\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
-\DeclareUnicodeCharacter{00A0}{ }
-
-%\tikzset{
-%    %Define standard arrow tip
-%    >=stealth',
-%    %Define style for boxes
-%    punkt/.style={
-%           circle,
-%           rounded corners,
-%           draw=black, thick,
-%           text width=1em,
-%           minimum height=1em,
-%           text centered},
-%    punktrect/.style={
-%    rectangle, 
-%    rounded corners, 
-%    % fill=black!10,
-%    draw=black, thick,
-%    minimum height=1em, 
-%    text centered},
-%    % Define arrow style
-%    pil/.style={
-%           o->,
-%           thick,
-%           shorten <=2pt,
-%           shorten >=2pt,}
-%}
-
-\tikzset{
-    %Define standard arrow tip
-    >=stealth',
-    %Define style for boxes
-    punkt/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           text width=1em,
-           minimum height=1em,
-           text centered},
-    observed/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           },
-     latent/.style={
-           circle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-    target/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-    observedrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           },
-     targetrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-     empty/.style={
-           circle,
-           rounded corners,
-           minimum width=.5em,
-           minimum height=.5em,
-           font=\footnotesize,
-           text centered,
-           },
-    % Define arrow style
-    pil/.style={
-           o->,
-           thick,
-           shorten <=2pt,
-           shorten >=2pt,},
-    sh/.style={ shade, shading=axis, left color=red, right color=green,
-    shading angle=45 }   
-}
-
-\newcommand{\KL}{\operatorname{KL}}
-\newcommand{\simpleregret}{R_T}
-\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
-\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
-\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
-\newcommand{\EE}{\mathbb E}
-\newcommand{\EEa}{\EE_a}
-\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
-\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
-\newcommand{\actions}{\mathcal{A}}
-\newcommand{\calA}{\mathcal A}
-\newcommand{\etc}{\textit{etc}}
-\newcommand{\ie}{\textit{i.e.}}
-\newcommand{\eg}{\textit{e.g.}}
-\newcommand{\calP}{\mathcal P}
-\newcommand{\x}{\boldsymbol{x}}
-\newcommand{\Ps}{\operatorname{P}}
-\newcommand{\R}{\mathbb R}
-\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
-\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
-
-\newcommand{\defined}{\vcentcolon =}
-\newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
-\newcommand{\Var}{\operatorname{Var}}
-\newcommand{\calF}{\mathcal F}
-\newcommand{\sr}[1]{\stackrel{#1}}
-\newcommand{\set}[1]{\left\{#1\right\}}
-\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
-\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
-\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
-\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
-\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
-\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
-\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
-\newcommand{\Ber}{\operatorname{Bernoulli}}
-\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
-\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
-\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
-\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
-\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
-
-\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
-\newcommand{\cf}[2]{{#1}_{#2}}
-\newcommand{\bernoulli}{\operatorname{Bernoulli}}
-\newcommand{\dirac}{\operatorname{Dirac}}
-\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
-\renewcommand{\vec}[1]{\boldsymbol{#1}}
-
-%\theoremstyle{plain}
-
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-
-%\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\let\temp\epsilon
-\let\epsilon\varepsilon
-
-\author{Finnian Lattimore}
-\title{Learning how to act: making good decisions with machine learning}
-
-\begin{document}
-\todo{figure out theoremstle is undefined}
-\def\ci{\perp\!\!\!\perp} % from Wikipedia
-
-\maketitle
-
-\epigraph{In vain the grave, with retrospective Eye,
-Would from the apparent what conclude the why,
-Infer the Motive from the Deed, and show
-That what we chanced, was what we meant, to do.}{Alexander Pope}
-
-3100 words at the start of boot camp.
-
-\chapter*{Random notes}
-
-There is a connection here to the degree to which the model is invariant.
-A model that is invariant to more things is more explainitory.
-We could look at levels of hierachy.
-There is causal graphical models, which are invarient to interventions on variables.
-Can I tie this to a physical system and give more indepth examples of the nested tier of hieriachies?
-
-I do not see the distinction between explaination and (causal) prediction.
-Explaination is all about the ability to compress and to generalize.
-The more a model can do this, the more we view as providing an understanding of the why.
-
-Sometimes the causal component is obvious.
-
-Relationship to generalization.
-Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
-The assumption is there are less places for change to come in.
-
-If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.
-
-The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)?
-This may be a reason for using structure learning type algorithms even if you are simply doing prediction.
-
-\subsubsection*{The role of assumptions in causal inference}
-
-Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.
-
-Does predictive accuracy indicate a good causal model?
-
-Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.
-
-Generalizing the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parashutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalize).
-(and given sufficient generalizability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)
-
-Is causal inference possible (are the conclusions ever valid)?
-
-Only from randomized experiment ... if people can do it without recource to constant randomized experiment then an algorithm exits.
-
-The challenges and limitations of infering causal effects from observation data are numerous.
-However, there are problems we need to solve.
-The questions is are these problems so serious as to warrent discregarding all non-experimental data.
-I think answer is a clear no.
-Need ongoing validation, with assumptions.
-We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).
-
-\subsection*{Relationship to interpretability}
-
-A desire for interpretability indicates that something has been left out of the loss function.
-
-Are causal/interpretable models more reliable or more likely to generalize well?
-
-One form of interpretability gives people insight into what the features are that the model is relying on.
-
-If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
-
-Specifically, people can
-\begin{itemize}
-\item rule out many possible features as highly unlikely to be relevent to a problem
-\end{itemize}
-
-People have access to a lot of detailed prior knowledge.
-
-\subsection*{Relationship to transfer learning}
-
-find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time).
-Causal models predict the outcome of actions.
-We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike.
-So we have to make representations such that things are stable.
-
-This is tightly related to generalizability.
-If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results.
-We don't include , the color of the doctors shirt, the size of the room, ...
-
-For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales.
-However, this relationship may be very unstable over time because the ad slots are sold at auction.
-The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword.
-However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.
-
-Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.
-
-\chapter*{Introduction (3500 words)}
-
-\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.
-
-\paragraph*{This thesis contributes to knoweldge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
-
-\paragraph*{My key research questions are:} 
-\begin{itemize}
-\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
-\item To summarize the key strands of causal inference research from economics and social sciences for the machine learning community
-\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
-\end{itemize}
-
-\section*{Motivation}
-Many of the most important questions in science and in our personal lives are about the outcomes of doing something.
-Will asking people to pay upfront at the doctors reduce long term health expenditure?
-If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.
-
-These are hard questions because they require more than identifying a pattern in data.
-Correlation is not causation.
-Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.
-
-We now live in a world of data.
-Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth.
-Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose.
-It does not come from the controlled world of a randomised experiment.
-
-The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
-From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries.
-It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge datasets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artifcial intelligence exceeds our own and then grows exponentially.
-
-However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings.
-Getting them to generalize to even slightly different problems or datasets remains very challenging.
-Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it.
-The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on.
-As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
-
-\section*{What is causality and why do we care? (2000 words)}
-
-To explain or to predict \cite{Shmueli2010}
-The two cultures \cite{Breiman2001}
-
-\paragraph*{There are two reasons why correlation is not causation}.
-The first is related to variance and over-fitting.
-Obversations are noisy.
-With a finite dataset with enough variables we will be able to find some that are completely unrelated but correlate purely by chance.
-\ref{figure - search for random correlation}.
-The second arises from bias, typically introced by an un-observed confounding variable.
-In this case, variables are correlated not by chance.
-We would expect the relationship to hold if we sampled more data.
-However, they are not causally related in that intervening to set one would not likely effect the other.
-EXAMPLE WITH FIGURE.
-
-\subsection*{Defining causality}
-\begin{itemize}
-\item widely debated in science and philosophy (REFERENCES)
-\item what is explaination?
-\item any model that aims to predict the outcome of an action or intervention in a system
-\item I do not see the distinction between explaination and (causal) prediction.
-Explaination is all about the ability to compress and to generalize.
-The more a model can do this, the more we view as providing an understanding of the why.
-\item mediation?
-\end{itemize}
-
-\subsection*{Identifying when we have a causal problem}
-
-\subsubsection*{Examples of typical machine learning probems. Are they causal?}
-
-Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to.
-Which of them require casual inference?
-How can we identify characterists of a prolem that make causal modeling important?
-
-\begin{itemize}
-\item Speech recognition (for systems like Siri or Google)
-\item Machine translation 
-\item Image classification
-\item Forecasting the weather
-\item Playing Go 
-\item Identiying spam emails
-\item Automated essay marking
-\item Predicting the risk of death in patients with pnumonia.
-\item Predicting who will re-offend on release from prison 
-\item Predicting which customers will cease to be your customers
-\item Demand prediction for inventory control
-\item Predicting who will click on an ad
-\item Financial trading
-\item Recommending movies
-\item Online search
-\item Self driving cars
-\item Pricing insurance
-\end{itemize}
-
-The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.
-
-Consider speech recognition.
-You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words.
-Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a dvd on elocution).
-A similar arguament could be made for many applications of machine translation and image classification.
-
-In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see.
-We can be fairly confident that the process will not change.
-If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
-
-What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decicion will not effect if it actually rains.
-However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
-For this we need a (causal) model that allows us to predict the outcome under various different interventions.
-
-Identifing spam and automated essay marking systems are similar.
-The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label.
-Spammers and students will modify their writing in order to optimise thier results.
-The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided.
-(What would the nature of the features have to be such that change did not occur?
-- they would have to be causes of the label).
-
-What about predicting the risk of death in patients with pnumonia?
-Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective this seems like a straighforward prediction problem.
-It is not.
-Depending on how the decision to admit was previously made and what features are included (or ommited) in the model, the relationship between those features and the outcome may change if we start using the model to decide whome to admit.
-(xxx et al) found exactly this effect.
-Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pnemonia.
-They realised this was because doctors were treating such patints very agressively, thus acutally lowering their risk.
-There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols.
-However, using it to decide on what basis to admit people could kill.
-The key is understanding exactly what question you are asking.
-In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.
-
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc).
-For all these problems the assumptions required to treat them as pure prediction problems are;
-\begin{enumerate}
-\item The treatment is assumed to be effective (at least better than nothing)
-\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
-\end{enumerate}
-
-Demand prediction seems relatively straightforward.
-These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
-It seems unlikely that using the model to ensure stock is available will itself change demand.
-However there is a potential data sensoring issue.
-If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
-Changing availability does then change demand.
-
-Playing Go (and other games) is another case with some subleties.
-At every turn, the AI agent has a number of actions available.
-The board state following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win.
-...
-this is interesting maybe come back to it ...
-an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ...
-One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken).
-When can we learn from the actions others have taken?
-When there is no confounding.
-And does this hold with Go?
-Probably because the board state encapsulates everything that should determine what move is played.
-Learning directly from actions and trying to generalize (can in some instances reduce the problem to standard ML)
-
-Having considered these examples we can now identify some general aspects of the problem that determine wheter or not we require a causal model.
-
-\begin{itemize}
-\item Does acting on the predictions of the model change the mapping from features to target?
-(at least if the decision process is open to scrutiny).
-In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
-\item
-
- Covariate shift clearly comes in here.
-Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable.
-The mechanism is known but the function may be complex.
-Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models?
-Yes, if the way in which features get their values changes, then causal models will be invarient to that in a way that non-causal ones are not.
-
-\item To decide between actions we only need to rank them (not estimate their actual effect). 
-\item The predicted outcome in the absence of an intervention provides a single point.
-We can use this to find which problems are most serious if left alone - and priortise those for modelling changes.
-\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
-\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
-\item Ethics - ... People's viewpoint on if its ok...
-\end{itemize}
-
-I hope these examples gave you a feel for the richness and subtlties of causal inference.
-We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
-
-\section*{Approaches to causality (1000 words)}
-
-There are two broad approaches to deciding how to act.
-Reinforcement learning and causal inference.
-In reinforcement learning we estimate the effect of actions by taking them.
-We assume there is an agent capable of intervening in the system and try to find policies for the agent to follow in selecting actions so as to maximise some kind of reward.
-This is a very powerful and general framework (REFERENCES TO GENERAL AI).
-However, POINT OUT SOME OF THE DIFFICULTIES WITH REINFORCEMENT LEARNING.
-We are frequently presented with large bodies of data that have been collected from a system in which we did not have any control over what actions were taken.
-
-We will call datasets where we do not have control over the decision making process that generated the data observational.
-Versus experimental datasets, where we do have control (experimental datasets do not always have to be randomized -although that is a powerful approach to ensure we have control.
-There is a space in the middle where we have partially controlled the process by which agent select actions.
-Randomized data with imperfect compliance would be an example.
-
-An agent (capable of intervening in the system) chooses an action from those availble The agent making the decision included in the model. This is
-
-Reinforcement learning addresses the problem of learning from explicitly taking actions. There
-is typically some state or environment. An agent chooses an action from those available in the
-current state. The state then evolves stochastically as a function of the selected action and
-the agent receives some feedback or reward that is a function of the new state. This setting
-differs from the standard classification problem in that the agent must learn from feedback on
-the selected action, rather than being presented with the correct action for a given state. A
-common modelling assumption is that the state evolves only as a function of the previous state
-and the action chosen, and given these, is independent of the previous history of states and
-actions. This is known as a Markov decision process or MDP. A particularly well studied and
-understood model is the singe state MDP. In this case, there are a set of actions, each associated
-with a fixed but unknown reward distribution and at each time step our agent selects an action
-and receives corresponding feedback. This is known as the multi-armed bandit problem.
-
-Causal inference makes use of assumptions to allow the outcome of actions to be predicted from
-observational data. The key to causal inference is a framework that can model how actions
-change the state of the world. This framework then allows us to map information collected in
-one setting to another.
-
-Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit
-algorithms deal with the sequential nature of the decision making process, causal inference with
-the problem that full randomisation is not always feasible, affordable or ethical. The similarities
-between the problems that these techniques have been developed to address raises the question
-of if there are problems best addressed by a combination of these approaches and how they
-can be combined. The goal of my thesis is to explore these questions. In the next sections I
-review the key literature in causal inference and bandits. I then present a general approach to
-how causal models might be incorporated into bandit settings and conclude by demonstrating
-an algorithm that leverages causal assumptions to improve performance in a specific bandit
-setting.
-
-There are two key approaches to causal problems.
-The first is to learn the outcome of actions by directly intervening in the system and seeing what happens.
-We then get feedback on how good those actions were.
-This is the approach taken in reinforcement learning.
-ADVANTAGES AND DISADVANTAGES OF THIS APPROACH.
-The second broand approach is causal inference.
-Here
-
-\paragraph*{Two broad approaches} 
-\begin{itemize}
-\item Build a model to map the natural behaviour of the system to what will happen for some action
-\item Take the action and see what happens
-\end{itemize}
-
-\paragraph*{The first is causal inference}
-
-\paragraph*{The second is reinforcement learning}
-
-\paragraph*{Both generalize from randomized experiment} Reinforcment learning to sequential decisions, causal inference to non-experimental conditions
-
-Both these fields relate to the problem of making optimal decisions and both can be seen as generalising randomised controlled experiments.
-Causal inference  is the study of how to estimate the effect of an action in the absence of randomisation.
-Reinforcement learning studies how we can do better if the decisions are to be made sequentially.
-
-\paragraph*{Both approaches involve assumptions} the latter that we can group context and actions.
-
-\paragraph*{Limitations of causal inference}
-
-\paragraph*{Limitations of experimets} What are the issues with standard randomized experiments?
-
-\chapter*{Causal models (3000 words)}
-
-Causal inference aims to infer the outcome of an intervention in a system from data obtained by observing (but not intervening in) the system.
-To do this we need terminology to describe actions and how we anticipate the system should respond to them.
-Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks.
-In this chapter we describe these approaches.
-We will examine what problems they allow us to solve, what assumptions they rely on and how they differ.
-We will also use them to describe the following very simple example.
-The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
-
-Suppose we have developed a new drug for some illness and wish to determine how effective it is.
-We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group.
-The people in the treatment group get the drug, everyone else gets a placebo pill.
-The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo.
-We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
-
-\subsection*{Causal bayesian networks}
-
-Causal Bayesian Networks are an extension of Bayesian networks.
-A bayesian network is a graphical way of representing how a distriution factorises.
-Any joint probability distribution can be factorized into a product of conditional probabilities.
-There are multiple valid factorizations, corresponding to permutations of variable ordering.
-
-\begin{equation}
-P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
-\end{equation}
-
-We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}).
-If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.
-
-\begin{figure}[h]
-\centering
-\caption{General Bayesian network over 3 variables}
-\label{fig:bayesnet}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below left=of 1](2){$X_{2}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{figure}
-
-The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorized over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges).
-If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
-
-\begin{equation}
-P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
-\end{equation}
-
-A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
-This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-We need some notation to represent the distribution after an intervention.
-In this thesis we will use the do opperator introduced by Pearl \ref{XXX}, where $\P{Y|do(X=x)}$ represents the distribution of the variable $Y$ after an intervention that sets the variable(s) $X$ to $x$.
-TALK ABOUT HOW THIS CAN REPRESENT MULTIPLE THINGS CONTINUOUS VERSUS DISCRETE DISTRBUTIONS AND WHETHER OR NOT VARIABLES ARE RESOLVED OR AS A FUNCTION.
-It is important to understand that in general, the interventional distribution $P{Y|do(X=x)}$ will differ from the standard conditional distribution $\P{Y|X=x}$.
-
-If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network.
-This is referred to as the truncated product formula \cite{Pearl2000}.
-
-\begin{equation}
-\label{eq:truncatedproduct}
- P_{do(Z=z)}(X_{1}...X_{N}) =
-  \begin{cases}
-  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
-   0       & \text{otherwise } 
-  \end{cases}
-\end{equation}
-
-This result does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution.
-The truncated product formula will give different results depending on which you choose.
-The result is possible with causal bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
-In fact, from the interventionalist viewpoint causality the truncation product formula defines what it means for a link to be causal.
-
-Returning to our simple example, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$?
-Since the treatment $X$ is assigned  via deliberate randomisation we know there it is not effected by any unobserved variables.
-$X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and desease subtype of the patient.
-
-The causal bayesian network for this scenario is:
-
-\todo{proper picture}
-\eq{
-X \rightarrow Y\\
-U_l \rightarrow Y
-}
-
-This network represents the (causal) factorization  $P(X,Y) = P(X)P(Y|X)$, so from equation (\ref{eq:truncatedproduct}), $P(Y|do(X)) = P(Y|X)$.
-In this example, the interventional distribution is equivelent to the observational one.
-
-Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechinism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
-A key feature of this way of defining an intervention is the number of things that are invariant to the intervention.
-That is, all the conditional distributions over variables that were not intervened are assumed not to be changed by the intervention.
-
-\todo {draw the system as two seperate models}
-
-A causal bayesian network encodes all possible interventions that can be represented with the do notation.
-If you knew of all the terms in the causal factorisation, then you could compute the distribution over any variable in the graph after setting any subset of other vairables (subject to possitivity assumption)
-\todo {state positivity assumption}
-
-A causal network represents much more information than a bayesian network with identical structure.
-GIVE EXAMPLE IN SOMETHING QUANTIFIABLE LIKE BITS.
-TAKE THE GRAPH IN WITH THREE NODES.
-ASSUME EVERYTHING IS BINARY AND.
-FIGURE OUT HOW MANY VARIABLES ARE NEEDED TO DESCRIBE ALL POSSIBLE INTERVENTIONS VS THE BAYESIAN NETWORK.
-This may be surprising since their conditional probability distributions are estiated from data in the same way.
-It is only possible due to the additional assumptions inherent in a causal network.
-
-Causal bayesian networks are bayesian networks so results that apply to bayesian networks carry directly across;the local markov property states that variables are independent of their non-effects given their direct causes.
-Similarly the global markov property and d-separation also hold in causal networks.
-
-\subsubsection*{Limitations of causal bayesian networks}
-A number of critisims have been leveled at this way of modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will effect many variables in complex and non-determinisitic ways \ref{causality in complex interventions,Cartwright 2007}.
-However, the deterministic $do()$ opperater turns out to be surprisingly effective at modelling more complex interventions.
-One simply has to add additional variables to the graph that model how our interventions may actually take effect.
-For example, our drug treatment case assumed we set the drug that each person recieves.
-But what if some people in fact fail to take the prescribed treatment.
-How can we model this within the framework of deterministic interventions?
-We simply add a node representing what we prescribed to them (which we intervene on), which probabilistically influences what treatment they actually recieve.
-See figure XXXX
-
-Insert figure of randomized treatment with imperfect compliance here.
-
-What questions can't be phrased using this terminology
-- certain counterfactual queries
-- cyclic/dynamics
-- unroll to deal with cycles
-
-\subsection*{Structural Equation models}
-
-\subsubsection*{Definition}
-Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Uncertainties are not inherent but arise from unmeasured variables.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalized, generalized to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-For our example the SEM is:
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random.
-The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-
-\subsubsection*{Example problem}
-We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
-Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$.
-The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}.
-However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$.
-The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomized experiment.
-In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.
-
-For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-
-Limitations \& Critisisms 
-- can express some non-measurable things
-	- example of that
-
-- confusion may occur as to whether interpretation is causal
-
-- can they handle cycles?
-
-\subsection*{Counterfactuals}
-\todo{use cf command to in counterfactual equations}
-
-The Neyman-Rubin model \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}.
-Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else.
-
-In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{i}^{1}$ versus what would happen if they got the placebo, $y_{i}^{0}$.
-The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment.
-This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
-
-Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.
-Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $\cf{Y}{1}$ is the distribution we would see of $Y$ if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate, from an experimental or observational study, the probability that people recover if treated $P(Y|X=1)$ and the probability that they recover if not treated $P(Y|X=0)$.
-Now if $X=0$ then $Y = \cf{Y}{0}$.
-Equivalently stated:
-
-\eq{
-\P{Y^{0}|X=0}&= \P{Y|X=0}\\
-\P{Y^{1}|X=1}&=P{Y|X=1}
-}
-
-If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
-
-\eq{
-\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
-\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
-}
-
-\eq{
-\implies \P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
-}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{Rosenbaum1983}.
-They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-Again this is justified in our example due to the randomization of treatment assignment.
-In general these assumption do not hold.
-If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model.
-that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.
-
-\todo{SUTVA assumption}
-
-There are some complex philosopical objections to counter factuals arising from the way they describe alternate universes that were never realized and are not empirically testable \ref{XXXXX}.
-This can result in some concrete problems.
-Consider the following example based on  \ref{Dawid2000}.
-
-Again we have a drug, where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
-Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal.
-We will assume equal variance for simplicity.
-
-% vector with mean mu_t,mu_c and a covariance matrix
-\eq{
-P(Y_t,Y_c) \sim N()
-}
-
-then their difference is also normal,
-
-\eq{
-P(\tau) = N(\mu_t - \mu_c,2\sigma^{2}(1-\rho))
-}
-
-\includegraphics[scale=.6]{figures/counterfactual_nonidentify.pdf}
-
-Key issue is that we can never observe the joint distribution over $Y_t$ and $Y_c$.
-As a result, the variance of $\tau$ is not identifiable, even from experimental data.
-It seems on the face of it that $\tau$ is relevent to our decision making.
-If the example above, if $\rho = 1$ then almost everyone benifits slightly from the treatment.
-However if $\rho=0$, there is a wide range, with some people benifiting a lot and others suffering significant harm.
-
-Can these issues be resolved? If so, how?
-
-- note that we can bound this counterfactual distribution based on the variance of the observed interventional distributions. If the variance of XXX is small then this may not be an issue.
-
-- does it actually make sense to make a decison on the basis of this counterfactual that is a function of something we could never observe. I
-
-- part of the problem may be due to the deteminisitic way we have phrased individual treatment effects.
-
-Can these issues be resolved by considering personalized causal effects as random variables and avoiding counterfactuals?
-
-A range of similar issues can arise in counterfactuals \ref{XXXX}. 
-- What assumptions are required for the kind of counterfactual analysis like X would have been higher had B ...?
-- briefly discuss the different flavours of counterfactual questions here.
-
-One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than taking about the distribution of Y given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $X$.
-This is certainly convienient, if rather impressise.
-
-\subsection*{Comparing and unifying the models}
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-This makes it straighforward to combine key results and algorithms developed within any of these frameworks.
-For example, draw a graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores \ref{} estimate the effect.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorization of the joint distribution over its variables.
-In other words, the network is a causal Bayesian network.
-All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
-Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorization and the truncated product formula (equation \ref{eq:truncatedproduct}).
-Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
-We can also connect this to counterfactuals via:
-
-\begin{equation}
-\begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
-\end{aligned}
-\end{equation}
-
-The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
-
-\begin{equation}
-\label{eq:weakignore}
-X \ci Y^{0} \text{ and } X \ci Y^{1}
-\end{equation}
-
-It is possible to relax the independence of errors assumption we made for SEMs slightly to correspond exactly the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \cite{Richardson2013}.
-To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of functional models or bayesian networks or as conditional independence statements involving counterfactual variables (ignorability assumptions).
-By non-parametrically, I mean that we are not making any assumptions about the form of the relationships between variables.
-Models that are not non-parametrically identifiable can still be identified given assumptions about the distributions of variables and the functional relationship between them, for example, that the functions are linear or that the noise is additive \cite{Peters2014}.
-This form of assumption fits extremely naturally into the structural equation framework.
-
 However we can also pose causal queries that are not interventional and cannot be phrased in terms of the do-notation.
@@
 
-There are differences between the models we have considered when it comes to counterfactual queries.
-Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.
-Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
-Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
-
-In practice, differences in focus and approach across different fields eclipse these actual differences in the models.
-The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field.
-Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
-In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
-
 \subsubsection{A translator from graphical independence to counter factual statements}
 
-\subsubsection*{More terminology}
+\subsubsection{More terminology}
 With the graphical framework in place, it is useful to define some key terminology used in describing causal models in terms of the graph structure they refer to.
@@
 
-\subsection*{What does a causal model give us? Resolving Simpson's paradox}
+\section{Clarifying some confusing definitions}
 
-We will now demonstrate how we can use our new notation and frameworks for causal inference to resolve a facinating paradox. \todo{ insert some history of simpson's paradox here.}
+With a causal framework in place, we are now in a position to clarify some confusing concepts that are very prevelent in economics and social sciences.
 
-Suppose a doctor has two treatments, A and B, which she offers to patients for blood preasure control.
-She keeps track of which medication her patients choose and whether or not the treatment is successful.
-She obtains the following results:
+Confounding
+Exogenous
+Endogenous
+Selection bias
+...
 
-\todo{insert data table with success by treatment}
-
-It seems that drug A is performing better.
-However, having read the latest liturature that medication effectiveness may differ between men and women, she decides to check how each drug performed by gender and obtained the following data.
-
-\todo{innsert table}
-
-Once the data is broken down by gender, it appears that treament B is better for both men \emph{and} women.
-This is the paradox.
-If the doctor were unaware of the gender, it seems that she should advise treatment A.
-But as soon as the patient walks though the door  then she should offer treatment B, regardless of which gender the patient turns out to be.
-
-How does causaul modelling reslove this paradox?
-Let's represent the treatment presribed by the variable $T$, the gender of the patient by $Z$ and wheter or not the treatment was successful by $Y$.
-The key to resolving the paradox is that the doctor really cares about the interventional distribution $P{Y|do(T)}$, not the standard conditional distributions $P{Y|T}$.
-The results from table XXX shows that $\P{Y|T=A}$ is higher than $\P{Y|T=B}$.
-If she has to suggest a treatment for a patient without knowing their gender, she should choose the treatment for which $P{Y|do(T)}$ is highest not $P{Y|T}$.
-If she knows thier gender then she should otimise to $P{Y|do(T),Z}$.
-
-The causal network for this problem is shown in figure XXX.
-In this example, women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they recive than men.
-This is what makes treatment $A$ appear better if you look at the conditional distribution.
-The fact that a patient took treatment $A$ tells you that they are more likely to be female.
-In the terminology defined in section \ref{}, gender is a confounding variable and we must adjust for it to obtain unbiased estimates of the causal effect of the treatment.
-As we saw from the examples in chapter \ref{XXX}.
-Determining when you really care about the interventional as opposed to the conditional distribution is a sublte but very important question that is frequently overlooked.
-
-\chapter*{Two key questions (5000 words)}
-
-We can rougly categorize the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
-In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
-The goal is then to estimate the causal effect of some action or range of actions.
-WHERE DOES MEDIATION FIT IN?
-THIS IS ALREADY HUGE, and is central to 1000 of papers published each year.
-Causal discovery aims to leverage much broader assumptions to learn the stucture of causal graph from data.
-THIS IS THE AUTOMATION OF SCIENCE.
-
-\section*{Causal effect estimation}
-
-Causal effect estimation addresses the problem where we assume the key structure of the graph is known. That is, we assume that we have at a minimum:
-
-\begin{itemize}
-\item the target/outcome variable we care about
-\item the focus/treatment variables on which we are considering intervetions
-\item any variables which act to confound two or more of the variables we have included.
-\end{itemize}
-
-Some of these variables my be latent or unobserved, in that we do not have measurements for them in the available data.
-However, their position in the network is assumed to be known.
-For example if we were interested in the effect of years of schooling on wages, we might expect that some measure of inherent ability could influence both the number of years of schooling people chose to pursue and the wages they subsequently obtaine on graduating.
-Even if we have no data to directly measure peoples inherent ability, becuase it influences two of the variables were are modelling, we must include it in the graph.
-
-Causal effect estimation is one of the most frequently applied methodolgoies within applied sciences.
-It implicit in millions of studies accross medicine, business, economics and social sciences.
-Almost everytime someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best levereage those assumptions to infere as much information as we can from our data.
-
-\subsection*{Mapping from observational to interventional distributions}
-Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting to the interventional one.
-Recall that a bayesian network is a way of representing the joint distribution over its variables in a way which highlights the any conditional idependencies between them.
-
-\begin{theorem}
-\textbf{(Local markov condition)} Given a bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decendents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
-\end{theorem}
-
-The set of conditional independence relations given by the local markov condition can enforce additional independences that also hold in all distributions that are compatible with $G$.
-D-separation is an algorithm that extends the local markov property to find these additional independences.
-It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
-
-By leveraging conditional independencies we can effectivly localize the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \cite{Pearl2000}.
-The do calculus consists of three rules.
-They derive from the causal information encoded in a causal network and the properties of d-separation and do not require any addition assumptions other than that of specifying the causal network.
-
-\subsubsection*{Rule 1}
-This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention.
-The intervention $do(\boldsymbol{X}=\boldsymbol{x})$ changes a causal network, $G$, in a simple way.
-Variables in $\boldsymbol{X}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{x}$.
-This corresponds to deleting the edges with arrows into variables in $\boldsymbol{X}$  (see figure \ref{fig:mutilatednet}).
-The resulting 'mutilated' network $G_{\overline{\boldsymbol{X}}}$ remains a causal network and d-separation still applies.
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\begin{equation}
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
-\end{equation}
-
-$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\begin{equation}
-\label{eq:Do12}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) 
-\end{equation}
-
-This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.
+\chapter*{Random notes}
 
@@
 \centering
-\caption{Intervention in a causal bayesian network}
+\caption{Causal inference without a framework}
 \label{fig:mutilatednet}
@@
 \node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
 
@@
     	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2);
+    (2) edge node {} (3);
 
@@
 \centering
-\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
+\caption{network after an intervention $G^*$}
 \label{fig:mutilatedAfter}
@@
 
-\subsubsection*{Rule 2}
-Rule 2 states when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening  $do(\boldsymbol{X}=\boldsymbol{x})$ have the same effect on the distribution of the outcome $\boldsymbol{Y}$.
-You can think of this as when correlation is causation.
-It is easiest to understand by explicitly including the intervention process in the graphical model.
-We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on (figure \ref{fig:augmentedNetwork}).
-Let $\epsilon$ be some arbitrary value not in the set of possible values for $X$.
-If $X = \epsilon$ the distribution of $X$ is what it was without intervention.
-Otherwise, if $X=x$, $X$ deterministically takes the value $x$ and is independent of its previous parents, representing the intervention $do(X=x)$.
-We use the notation $G^{\dagger}$ to represent $G$ augmented with these decision nodes.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
-
-\begin{equation}
-\label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w})
-\end{equation}
-
-$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X})$ in $G^{\dagger}$:
-\begin{equation}
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x})
-\end{equation}
-
-If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.
-
 \begin{figure}[h]
+\caption{}
+\label{fig:causal_inference_without_do}
 \centering
-\caption{Augmentated causal network with intervention}
-\label{fig:augmentedNetwork}
-\begin{subfigure}[t]{0.4\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:augmentedOriginal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left = of 1](4){$U_{1}$};
-\node[main node, above right = of 1](5){$U_{2}$};
-
- \path[every node/.style={font=\sffamily\small}]
- 	(2) edge node {} (3)
- 	(4) edge node {} (2)
- 		edge node {} (1)
- 	(5) edge node {} (3)
- 		edge node {} (1)
- 	(1) edge node {} (2)
- 		edge node {} (3);
-
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-\centering
-\caption{augmented network after the intervention $do(Z=z)$, $G^{\dagger}_{\overline{Z}}$}
-\label{fig:augmentedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.3cm,
-  thick,main node/.style={punkt}]
-
-\node[main node](1){$z$};
-\node[below left=of 1](2){};
-\node[below right=of 1](3){};
-\node[main node, below left=of 2](4){$X$};
-\node[main node, below right=of 3](5){$Y$};
-\node[above right=of 1](6){};
-\node[above left=of 1](7){};
-\node[main node, above left=of 7](8){$U_{1}$};
-\node[main node, above right=of 6](9){$U_{2}$};
-\node[main node, above left=of 4](10){$\hat{X}$};
-
- \path[every node/.style={font=\sffamily\small}]
- 	(4) edge node {} (5)
- 	(10) edge node {} (4)
- 	(8) edge node {} (4)
- 	(9) edge node {} (5)
- 	(1) edge node {} (4)
- 		edge node {} (5);
-
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-\subsubsection*{Rule 3}
-This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
-\begin{equation}
-\label{eq:Do3}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-\end{equation}
-$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}})$ in $G^{\dagger}$:
-\begin{equation}
-\label{eq:Do32}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y})
-\end{equation}
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ without conditioning on $\boldsymbol{X}$ implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ which starts on an arrow leaving $\boldsymbol{X}$.
-This means there is no causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$ and the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ does not alter the distribution of $P(\boldsymbol{Y})$.
-
-\subsection*{Identifiability}
-A natural question to ask is, given the level of assumptions we have made about the graph, is it possible to estimate a causal effect from observational data?
-This is the identifiablity problem.
-Identifiability is an asymptopic property of the graph.
-It asks if we can obtain an unbiased point estimate for the causal effect of interest in the infinite data limit.
-A problem is non-parametrically identifaible if it is identifiable without any additional assumptions on the functional form of the dependencies between variables in the graph.
-
-The question of non-parametric identifiability is solved!
-The do calculus is complete.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observataional quantities via repeated application of the do calculus.
-
-There is an algorithm \cite{Shpitser2012} based on these properties that, for a given network and interventional (do-type) query, can:
-
-\begin{enumerate}
-\item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network
-\item if it is identiable, return the required expression 
-\end{enumerate}
-
-\todo{insert link to working javascript version}
-\todo{figure with typical examples}
-
-There are many interesting questions relating to identifiability that remain open.
-What is the minimal (by some metric) additional information that would be required to make a non-identifiable qurey identifiable?
-What if we assume various restrictions on the functional form of the relationships between the variables?
-Clearly  .
-A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work going back to at least ???
-\ref{}
-
-Is identifiablity a good question to ask?
-It seems natural to see if a problem can be solved with infinite data before we try to see how well we can do it with finite data.
-However, we should not give up on non-identifiable problems.
-We may still be able to achive useful bounds on causal effects with finite data even if point estimation is not possible.
-From that perspective we should be cautious about dividing problems into identifiable and not identifiable.
-
-How do we tackle a problem that is not identifiable. We could look for bounds or we could make additional assumptions (or both).
-
-\subsection*{Estimation}
-\subsection*{Defining causal effects} Summarising the difference between two distributions. There is no one answer. \cite{Janzing2013}
-
-\cite{Imbens2004} Review of non-parametric estimation
-\cite{causal estimation is covariate shift}
-\paragraph*{How well do we actually do with finite datasets?}
-
-\paragraph*{When is regression causal?}
-A regression equation is causal if, when we represent the equations as a bayesian causal model or structural equation model, we claim the arrows are causal AND if the variables we include in the regression form a valid back door adjustment set.
-
-\subsection*{Non identifiable queries}
-
-\subsubsection*{Parametric assumptions}
-eg linearity
-
-\subsubsection*{Bounds}
-\paragraph{Instrumental variables are an example}
-
-Put in the picture of an instrumental variable.
-
-Instrumental variables are one of the key approaches used to estimate causal effects within economics and the social sciences.
-However, the setting they are associated with is non-identifiable.
-How is it that they are so useful nonetheless?
-Tight bounds may be just as good as what could be achieved in the identifiable setting, given finite data.
-
-\section*{Causal Discovery}
-We now move to the much more general problem of learning a causal graph from observational data.
-In this setting we make much broader assumptions about the structure of the graph.
-For example, that it is acyclic or that we have no unmeasured confounding variables.
-We do not assume the existance or directions of any links between the variables.
-Amazingly, it is possible to infer some aspects of causal structure with such general assumptions.
-The set of conditional independences in a non-experimental data set indicates some causal structures are more likely than others.
-In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
-These clues are the key to causal discovery algorithms.
-
-There is also work on infering causal structure from experiental data \ref{}. We discuss this further in section XXX.
-
-Causal discovery is a much grander goal than causal effect estimation given a known causal network.
-Arguably, if acheived, it would equate to the automation of scientific discovery.
-We need simply supply our algorithm with a vast collection of variables (regardless of their relevence to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make.
-Unfortunately, causal disovery is very hard.
-Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows XXX exponentially (or super-exponentially)?
-with the number of variables.
-- some intuition here about the strength of the signal we get from causal disovery.
-Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}.
-Talk about the numerical as well as the statisitcal issues here.
-
-In the next sections we briefly survey the key approaches to causal discovery.
-We roughly divide the methods into those based on those that exploit the connection between the conditional indpendences in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
-This is not the only way we could have grouped them.
-There is also a seperation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
-There are also hybrid methods \ref{}
-
-\subsection*{Conditional independence based methods}
-One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
-For any bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network.
-If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$.
-However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
-This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
-This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
-Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \cite{Kollner} for $P$.
-
-Faithfulness is an assumption.
-It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
-However, most distributions generated by a causal bayesian network will be faithful to that network.
-For faithfulness to be violated, different causal effects must exactly balance one-another out.
-For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
-If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
-However, this independence is not stable.
-It would disappear under a small perturbation to any of the parameters.
-\todo{point out violations are solutions to polynomial equations and thus measure 0 \ref{XX}}
-
-Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$.
-This set can also be referred to as the markov equivalence class compatible with $P$.
-
-\subsubsection{Without hidden common causes}
-The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of bayesian networks we must consider.
-
-We will begin with a brute force algorithm (described as the SGS algorithm in \cite{Sprites} and IC algorithm in \cite{Pearl2000}).
-While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.
-
-\begin{table}[H]
- \begin{tabularx}{\textwidth}{X}
- \hline
-\rule{0pt}{2.5ex} 
- \textbf{The SGS (or IC) Algorithm}\\
- \hline
- \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
- \begin{enumerate}[itemsep=8pt]
-  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an undirected link to form a complete graph.
-  \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent.
-If such a set (including the empty set) exists then $a$ and $b$ cannot be directly connected in $G$ so delete the link.
-  \label{alg:SGSexponential}
-
-  \item For all pairs of non-linked variables $(\alpha,\beta)$ with a common neighbour, $c$, if $c \notin \boldsymbol{S}_{\alpha,\beta}$, then $c$ must be a collider in the path $\alpha,c,\beta$ so  add arrows to direct the links $\alpha-c$ and $\beta-c$ towards $c$.
-  \label{alg:SGScolliders}  
-  \item Recursively try to orient any edges that remain undirected to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
-  \label{alg:SGSfinal}
-\end{enumerate}\\
- \hline
-\end{tabularx}
-\end{table}
-
-The SGS algorithm utilizes the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
-Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
-Unfortunately the number of conditional independence tests required for step \ref{alg:SGSexponential} grows exponentially (in the worst case) with the number of variables.
-Not only that, but for each edge that is in the true network, the algorithm will always tests all other possible subsets of variables.
-If the assumption that there are no hidden common causes or that the distribution is faithful are violated, step \ref{alg:SGScolliders} of the SGS algorithm can produce double headed arrows.
-
-The PC algorithm \cite{Sprites} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilize the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
-It also checks for low order conditional independence relations before higher order ones.
-This allows it to exploit any sparsity in the true network, leading to much better average case performance  \cite{Sprites} (although the worst case, where the true network is complete, is still exponential).
-With finite data, the order in which the links are considered can change the output (unlike for SGS).
-The effect of wrongly removing a link early on flows through to later conditional independence tests by changing which nodes are considered adjacent.
-
-\begin{table}[H]
- \begin{tabularx}{\textwidth}{X}
- \hline
-\rule{0pt}{2.5ex} 
- \textbf{The PC Algorithm}\\
- \hline
- \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the markov equivelence class of $G$\\
- \begin{enumerate}[itemsep=8pt]
-  \item As for SGS
-  \item \textbf{for} each link $a-b$:
-  \begin{itemize}[label={}]
-   \item $n = 0$
-   \item $\boldsymbol{A}_{a,b} = \{A_{1}...A_{j}\}$ be the set of nodes adjacent to $a$ and/or $b$
-   \item \textbf{while} $a$ and $b$ are connected and $n < j$:
-   \item 
-    	\begin{itemize}[label={}]
-    	\item \textbf{if} any subset of size $n$ of $\boldsymbol{A}$ makes $a$ and $b$ conditionally independent:
-    	\item \begin{itemize}[label={}]
-    			delete the link
-    		  \end{itemize}
-
-    	\item $n = n+1$
-    	\end{itemize}
-  \end{itemize}   
-  \item as for SGS
-  \item as for SGS
-\end{enumerate}\\
- \hline
-\end{tabularx}
-\end{table}
-
-The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution.
-Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects with equation \ref{eq:allvarsobserved}.
-We can then bound the true causal effect by combining the results for the all the networks.
-This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
-An implementation is available in the R package \cite{Kalisch2012}
-
-\todo{mention failed replication - if published}
-
-\subsubsection{With hidden variables}
-There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables.
-With an unknown number of hidden variables there are infinity many possible structures to search over.
-In addition, the space of causal networks is not closed under marginalization.
-If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalizing over $\boldsymbol{U}$, may not be faithful to any bayesian network (see figure \ref{fig:DAGSnotclosed}).
-
-\begin{figure}
-\centering
-\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalizing over $U$. }
-\label{fig:DAGSnotclosed}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){U}; 
-\node[main node, below left = of 1](2){B};
-\node[main node, below right = of 1](3){C};
-\node[main node, above left= of 2](4){A};
-\node[main node, above right = of 3](5){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    (5) edge node {} (3);
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](2){Treatment};
+\node[node, right=of 2](3){Outcome};
+\node[lt, above=of 3](4){U};
+\path[]
+	(2) edge (3)
+	(4) edge (3);
 
@@
 
-The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables). See example figure XXX.
+There is a connection here to the degree to which the model is invariant.
+A model that is invariant to more things is more explanatory.
+We could look at levels of hierarchy.
+There is causal graphical models, which are invariant to interventions on variables.
+Can I tie this to a physical system and give more in depth examples of the nested tier of hierarchies?
 
-\begin{theorem}
-\cite{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
-\end{theorem}
+I do not see the distinction between explanation and (causal) prediction.
+Explanation is all about the ability to compress and to generalise.
+The more a model can do this, the more we view as providing an understanding of the why.
 
-Since we only care about the causal relationships between observed variables, it is sufficient to search over networks where any hidden variables have no parents and directly cause two of the observed variables.
-Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable.
-These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalization and conditioning.
+Sometimes the causal component is obvious.
 
-For any DAG with latent (and selection) variables there is a unique MAG \cite{Richardson2002}.
-This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \cite{Sprites}.
-The logic behind the algorithm is very similar.
-Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations.
-The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG.
-Assuming there are no selection variables (see \cite xxx), the PAG can contain four types of link:
+Relationship to generalisation.
+Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
+The assumption is there are less places for change to come in.
 
-\begin{enumerate}
-\item $X \rightarrow Y$, meaning $X$ causes $Y$
-\item $X \leftrightarrow Y$, meaning there is a latent variable that causes $X$ and $Y$.
-\item $X\ {\circ} {\rightarrow}\ Y$, either $X$ causes $Y$ or a latent variable causes both.
-\item $X\ {\circ} {-}  {\circ}\ Y$, either $X$ causes $Y$ or $Y$ causes $X$ or a latent variable causes both.
-\end{enumerate}
+If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.
 
-The circles indicate where it is ambiguous if there should be an arrowhead (ie where there is one in some MAGs and not in others in the equivalence class).
-Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).
+The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)?
+This may be a reason for using structure learning type algorithms even if you are simply doing prediction.
 
-\todo{put these figures in a table with true graph input on left and fci output on right}
-\todo{uniform styles used for nodes accross thesis}
-\begin{figure}
-\centering
-\caption{FCI examples: true graph and fci output}
-\label{fig:FCIExamples}
- \begin{subfigure}[t]{0.8\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
- %nodes
-\node[main node](1){U}; 
-\node[main node, below left = of 1](2){B};
-\node[main node, below right = of 1](3){C};
-\node[main node, above left= of 2](4){A};
-\node[main node, above right = of 3](5){D};
+\subsubsection*{The role of assumptions in causal inference}
 
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    (5) edge node {} (3);
+Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.
 
-\end{tikzpicture}
- \end{subfigure}
- \begin{subfigure}[t]{0.8\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
- %nodes
+Does predictive accuracy indicate a good causal model?
 
-\node[main node](1){A};
-\node[main node, below right=of 1](2){B};
-\node[main node, right=of 2](3){C};
-\node[main node, above right=of 3](4){D};
+Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.
 
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge[pil] node {} (2)
-    (2) edge node {} (3)
-    (3) edge node {} (2)
-    (4) edge [pil] node {} (3);    	
-\end{tikzpicture}
-\end{subfigure}
- \begin{subfigure}[t]{0.4\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
- %nodes
+Generalising the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parachutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalise).
+(and given sufficient generalisability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)
 
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
+Is causal inference possible (are the conclusions ever valid)?
 
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (4)
-    (2) edge node {} (1)
-    (3) edge node {} (1);    	
-\end{tikzpicture}
-\end{subfigure}
- \begin{subfigure}[t]{0.4\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
- %nodes
+Only from randomised experiment ... if people can do it without recourse to constant randomised experiment then an algorithm exits.
 
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
+The challenges and limitations of inferring causal effects from observation data are numerous.
+However, there are problems we need to solve.
+The questions is are these problems so serious as to warrant disregarding all non-experimental data.
+I think answer is a clear no.
+Need ongoing validation, with assumptions.
+We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).
 
- \path[every node/.style={font=\sffamily\small}]
- 	(1) edge node {} (4)
-    (2) edge [pil] node {} (1)
-    (3) edge [pil] node {} (1);    	
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
+\subsection*{Relationship to interpretability}
 
-The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \cite{Zhang2008}.
-More recently \cite{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
-\cite{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
+A desire for interpretability indicates that something has been left out of the loss function.
 
-With latent variables we are not using all the information - so we could go further (to nested markov models and inequalities.) \cite{Richardson2012} \cite{Shipster2014}
+Are causal/interpretable models more reliable or more likely to generalise well?
 
-These are all constraint based methods ... efficient because they stop early, but also may not be robust to errors early on.
+One form of interpretability gives people insight into what the features are that the model is relying on.
 
-A comparison of algorithms
+If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
 
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{| c | c | p{4cm} | c | c | c |}
-\hline
-  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
-  \hline
-  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
-  \hline
-  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
-  \hline
-  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
-  \hline
-  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
-  \hline
-  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
-  \hline
-  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
- \hline
-\end{tabular}
-
-\subsubsection{Doing conditional independence tests}
-
-All the algorithms we have discussed in the previous section rely on being able to perform conditional independence tests.
-This is non-trivial with high dimensional data.
-If the functional relationship between the variables is linear with gaussian noise then the network represents a multivaitae normal distribution and a pair of variables $A$ and $B$ are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero.
-Where the functions are non-linear
-
+Specifically, people can
 \begin{itemize}
-\item \cite{Zhang2012} Kernel independence tests
-\item HSIC \cite{Gretton2008}
+\item rule out many possible features as highly unlikely to be relevant to a problem
 \end{itemize}
 
-\subsection*{Discovery with functional models}
-The algorithms we have considered so far return a Markov equivalence class.
-They cannot distinguish between two models that result in the same set of conditional independence relations.
-Consider the very simple case where we have only two variables and the only possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
-These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
-No algorithm relying purely on conditional independence relations can separate these two cases.
+People have access to a lot of detailed prior knowledge.
 
-Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$.
-What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from.
-Recall the functional definition of causality (section \ref{sec:SEM}).
-There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a}, linear models with non-gaussian noise \cite{Hoyer2012}
+\subsection*{Relationship to transfer learning}
 
-causal anti-causal,
+find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time).
+Causal models predict the outcome of actions.
+We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike.
+So we have to make representations such that things are stable.
 
-%If X → Y, what is the relationship between P(X) and P(Y|X)? 
-What about P(Y) and P(X|Y)?
+This is tightly related to generalisability.
+If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results.
+We don't include , the color of the doctors shirt, the size of the room, ...
 
-%In semi supervised learning we are given training points from P(X,Y) and additional points sampled from P(X). The goal is to learn P(Y|X). The extra points from P(X) will not help if X → Y.
+For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales.
+However, this relationship may be very unstable over time because the ad slots are sold at auction.
+The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword.
+However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.
 
-deterministic functions,
-
-IGCI \cite{Janzing2012a}
-
-Indepdendence of input and mechanism. Daniusis et al 2012
-
-Independence of function and input: If $X \rightarrow Y$ and we have a functional causal model $y = f(x,e)$ then the input
-distribution $P(X)$ and function $f$ represent independent mechanisms. We do not expect a change in the input distribution to modify the function.
-
-learning what causality looks like.
-Lopez-Paz et al 2014.
-
-Instead of positing a functional restriction on the relationship between variables and then developing theory to exploit that assumption, \citep{LopezPaz2014} propose learning what the causal relationship looks like from data.
-They assume there will be a difference between the relationship of $\P{X}$, $\P{Y}$ and $\P{X|Y}$ between $X \rightarrow Y$ versus $Y \rightarrow X$.
-Their algorithm requires a data set in which each row is itself a data set consisting of paris of varibles $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$.
-They use a keneral mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-dataset and train an algorithm to learn the direction of causality.
-Unfortunately we do not have a large collection of datasets where the causal direction is known to train such a model.
-\citep{LopezPaz2014} instead use a simulated dataset so thier model will nessesarily be base on the assumptions they make when generating the data.
-Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.
-Their approach performed well in the causal effect pairs challange, correctly identifing the causal direction in ??
-percent of cases (against a benchmark of 50\% from random guessing)\ref{XXX}.
-
-\citep{Peters2014} have extended results from the bivariate case to the multivariate setting (with no hidden variables)?.
-They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable.
-The build on this to develop an algorithm that allows the construction of causal graphs based on the additive assumption??
-
-\subsection*{Granger causality}
-
-Granger causality was developed within economics and is not typically discussed within the causal discovery liturature.
-
-Although it does differ from the previous approaches in the types of assumption it relies upon it none-the-less tackles the same goal
-
-so we shall include it here.
-
-Granger causalty essentially assumes that the future cannot cause the past.
-This may seem an entirely waterproof assumption.
-However it can be tripped up by unmeasured confounding variables.
-The cock crows before the sun rises - so does it cause the sun to rise?
-No.
-
-\chapter*{The interventionalist viewpoint (5000 words)}
-
-The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it.
-There is an obvious alternative.
-Instead of trying to infer the outcome of an intervention from passive observations we could just do it and see what happens.
-There are two key differences between observing a system and explicitly intervening in it.
-Firstly, when we intervene, we can choose which actions to take and thus get some control over which distributions we learn about.
-Secondly, if we are explicitly choosing interventions, we have a perfect model of the probability that we select each action given any context, allowing us to control confounding bias.
-
-\subsection*{The role of randomization}
-Randomized controlled trials are often presented as the gold standard for determining causal effects in many areas of social sciences and economics \ref{XXX}.
-What is it about randomization that makes it so important when it comes to causality?
-The graphical model for a randomized controlled experiment is shown in figure \ref{fig:random_experiment_network}.
-If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
-Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.
-
-\begin{figure}[h]
-\caption{causal network for a randomized experiment}
-\label{fig:random_experiment_network}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-\todo {turn this into something}
-This would be a natural place to discuss what happens with finite datasets where there are many other variables that influence the outcome.
-We expect that across at least some of these variables, the target and control group will not be balanced.
-Does this not inject bias into the causal estimates?
-The anser is no, (I think - a better more general proof would be nice).
-The more such attibutes there are, the higher is the variance in the outcome within the target and control group.
-When we perform a confidence test to ascertain whether or not there is a difference ...
-what happens.
-
-Randomization does not ensure target and control group are exactly alike.
-The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables in a between the target and control groups in a finite data sample.
-However, the variance in the outcome within both the target and control groups also increases.
-The net result is increased variance (but not bias) in  the estimate of causal effects.
-
-DISCUSS STRATIFIED RANDOM EXPERIMENT
-
-What is the role of randomization? How do bandits algorithms work despite being only partially randomized? What else can you do to improve randomized studies (variance reduction, lower regret).
-
-The benifit provided by randomization in breaking the link between the treatment variable and any confounders should not be understated.
-The possibility of confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (test for confounding).
-This means causal estimates from non-experimental data are always subjec to the criticism that an important confounder may have been overlooked or not properly adjusted for.
-However, randomized experimentation does have some limitiations.
-
-\subsubsection*{Limiations of randomized experiments}
-\label{subsec:limitations_of_experiment}
-\begin{enumerate}
-
-\item The probability that an individual is treated is set in advance and fixed over the course of the experiment.
-As the experiment proceeds we may start to see evidence that one treatment is much more effective than another.
-However (in a pure randomized experiment) we don't use this information to update the treatment that later patients recive.
-The result is that some will unessearily be given a sub-optimal treatment.
-\item Experiments are expensive and difficult to conduct. This means experimental datasets are often much smaller than observational ones, limiting the complexity of models we can explore.
-
-\item Experiments are often conducted on a convinient but unreprasentaive sample of the broader population of interest (for example first year collage students at research universities).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}.
-The results may not carry over to the general population of interest.
-The question of whether an experiment conducted on one population can be mapped to another is refered to as the transportability problem \citep{Bareinboim2013} \todo{check ref} and relies on very similar assumtions and arguments to causal inference and the do-calculus.
-
-\item The idealized notion of an experiement represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomized experiments in practice.
-There may be imperfect compliance: the treatment selected by the randomizer is not always followed.
-Or output censoring: the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If complience or attrition is not random but associated with (potentailly latent) variables that also effect the outcome then the problem of confounding bias returns.
-See figure XXX for a the graphical model of a randomized experiment with imperfect compliance and outcome censoring.
-
-\item It is not always possible or ethical to conduct a randomized experiment \cite{parashutes} \ref{Parachute use to prevent death and major trauma related to gravitational challenge: systematic review of randomised controlled trials}.
-\end{enumerate}
-
-\begin{figure}[h]
-\begin{center}
-\caption{Experiments are not always ethical}
-\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
-\end{center}
-\end{figure}
-
-\section*{Multi armed bandits}
-\todo{is there a way in latex to have a command for refering to each type of thing like figure, subsection, etc such that it can be updated programatically?}
-
-Multi-armed bandits were originally introduce by XXXX to address problem that it is sub-optimal to fix a probability of treatment in advance as we discussed in subsection \ref{subsec:limitations_of_experiment}.
-The problem can be formally stated as follows:
-
-\paragraph*{Definition}
-
-\subsection*{Regret} How do we measure the performance of a bandit algorithm?
-
-%$\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$
-
-Expected Regret
-\eqn{
-R_T(\pi) = \E{max_{i \in [k]} \sum_{t=1}^T{X_{t,i}}} - \E{\sum_{t=1}^T{X_{t,A(t)}}}
-}
-
-Psuedo-Regret
-\eqn{
-R_T(\pi) &= max_{i \in [k]}\E{\sum_{t=1}^T{X_{t,i}}} - \E{\sum_{t=1}^T{X_{t,A(t)}}} \\
-&= n\mu_{i^*} - \E{\sum_{t=1}^T{X_{t,A(t)}}}
-}
-
-Simple Regret
-\eqn{
-\mu_{i^*} - \E{\mu_{\hat i^*}}.
-}
-
-\todo{connect bandit regret to early stopping of experiment problem}
-
-Scholars in the 1940's found this problem so frustrating that they suggested it be dropped over Germany to waste their time \cite{XXX}.
-
-\paragraph*{The exploration/exploitation trade-off}
-
-\subsection*{Key approaches and results}
-\paragraph*{optimisim in the face of uncertainty}
-
-\subsection*{The need to add structure}
-\paragraph*{The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.} This makes problems with large or infinite actions spaces intractible.
-
-\subsection*{Key related settings}
-
-\subsubsection*{Adversarial Bandits}
-In the classical multi-armed bandit problem, the rewards for each arm are sampled stochastically from a fixed (but unknown) distribution.
-There is a large body of work analysing a more general formulation of the bandit problem that relaxes this stochastic assumption.
-
-\paragraph*{Definition of the Adversarial bandit problem}
-\begin{itemize}
-\item Different notation of regret
-\item Adversarial bandits are not imume to issues of feedback cycles or drift in the reward distribution because a constancy is implicit in the definition of regret. 
-\end{itemize}
-
-\subsubsection*{Contexual bandits}
-\paragraph*{Feature selection} Is it even more important here? How do algorithms degrade as irrelevent features are added? In the supervised vs the RL setting? What would be a fair comparison?
-
-\subsubsection*{Markov decision processes}
-\subsubsection*{Dynamic Systems}
-\begin{itemize}
-\item An explicit model of actions in a partially known system (eg HMM)
-\item Feynman-Kac Lemma; Solving a PDE can be converted to a stochastic process
-\end{itemize}
-
-\chapter*{Causal Bandits: Unifying the approaches (5000 words)}
-
-\section*{The framework}
-A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
-Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{} for a simple example.
-In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example only the set of single variable interventions.
-The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action.
-In this thesis I examine the case where the causal graph is known.
-We refer to these problems as \emph{causal bandit problems}.
-Extending this work to simultaneously learning the casusl graph is discussed in section \ref{XXX}.
-
-\subsubsection*{Definition of casual bandit problem} \todo{make proper definition}
-A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed actions} $\mathcal{A}$.
-One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
-
-\begin{figure}[h]
-\caption{A simple causal graphical model and corresponding complete action space. A and B represent binary variables that can be intervened on and Y represents the reward.}
-\label{fig:unify_frameworks}
-\centering
-\begin{subfigure}[c]{0.3\textwidth}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
-\node[main node](1){A};
-\node[tg, below right=of 1](2){Y};
-\node[main node, above right=of 2](3){B};
-\path[]
-	(1) edge (2)
-	(3) edge (2);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-Actions = \begin{tabular}{|c|}
-	\hline
-  do(A=0,B=0) \\
-  do(A=0,B=1) \\
-  do(A=1,B=0) \\
-  do(A=1,B=1) \\
-  \hline
-  do(A=0) \\
-  do(A=1) \\
-  do(B=0) \\
-  do(B=1) \\
-  do() \\
-  \hline
-\end{tabular}
-\end{subfigure}
-\end{figure}
-
-The number of actions or arms grows exponentially with the number of variables in the graph,
-making it important to use algorithms that leverage the graph structure to reduce the
-search space.
-Modelling a problem as a causal graph only makes sense when rewards are generated stochastically - since causal graphs fundamentally model probability distributions over variables.
-Thus the connection is to stochastic bandit problems (although adversarial bandits algorithms may be applied to stochastic problems).
-
-We now need to specify the feedback model for the causal bandit problem.
-What information is available to the decision making agent before and after they select an action?
-The causal bandit problem takes on characteristics of different bandit settings depending on the assumptions we make about what actions are available to the agent, what variables are obvserved and whether they are observed before or after the action is chosen.
-
-Enumerate the settings
-\begin{enumerate}
-\item Bandit feedback: the agent selects an action and then observes only the reward of that action.
-If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
-See figure XXX as an example.
-\item Bandit feedback with side information (context). The agent can view the value of some variables prior to selecting an action. After selecting they observe the reward of the selected node. 
-\item Post action feedback:
-\end{enumerate}
-
-If we receive feedback on additional nodes, the problem can be more interesting.
-In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
-Consider again the model in figure \ref{fig:unify_frameworks}.
-The causal structure implies:
-
-\eqn {
-P(Y|do(A = 0)) &= P(Y|do(),A = 0) \\
-&= P(Y|do(B=0),A=0)P(B=0)+P(Y|do(B=1),A=0)P(B=1) 
-}
-
-Thus we gain information about the reward for the action $do(A=0)$ from selecting the action $do()$ or $do(B = b)$ and then observing $A = 0$.
-
-We only get this form of side information for actions that don't specify the value of every variable, ie those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-Since the reward distribution for actions that set a subset of the variables is the result of marginalizing out other variables, they can only be optimal if they have lower cost.
-So if the cost of all actions is constant (no matter how many variables must be set), then the problem has the same characteristics as if only the reward node were observable.
-
-If the information on the value of additional nodes is available prior to selecting an action the problem resembles a contextual bandit.
-For example if we observe $A = 0$ then, in deciding between the actions $do(B=0)$ and $do(B=1)$, we would want information on $P(Y|A=0,B=0)$ and $P(Y|A=0,B=1)$.
-Note, side information can still arise if we learn the value of some variables prior to selecting an action and some afterwards.
-
-Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are possible.
-If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}.
-Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of allowable interventions thereby reducing the complexity.
-
-\section*{Related work}
-\begin{enumerate}
-\item leaning from log data
-\item bandits with imperfect compliance
-\item rigged casino paper
-\end{enumerate}
-
-\begin{figure}[h]
-\caption{Example causal graph (based on \cite{Koller2009}) where the outcome of interest (reward) is cholesterol level .
-The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
-For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
-\label{fig:cholesterol_graph}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
-\node[node](1){Health concious};
-\node[node, below left=of 1](2){Diet};
-\node[node, below right=of 1](3){Exercise};
-\node[node, below right=of 2](4){Weight};
-\node[tg,below left=of 4](5){Y=Cholesterol};
-\node[node, below =of 5](6){Test result};
-\node[node, above right=of 3](7){Free time};
-\path[]
-	(1) edge (2) edge (3)
-	(2) edge (4) edge (5)
-	(3) edge (4)
-	(5) edge (6)
-	(7) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-\section*{Causal bandits with post action feedback}
-WHY THIS PROBLEM. This work was presented at NIPS 2016 \ref{}.
-\todo{add relevent stuff from introduction section}
-
-\subsection*{Notation}
-
-We will assume each variable only takes on a finite number of distinct values. (The path to relaxing this assumption would be through leverging the work on continuous armed bandits).
-
-The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
-
-An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-
-We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
-the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
-
-\subsection*{Definition of causal bandit game with post-action feedback}
-\todo{ re-read and incorporate response from reviewer feedback for NIPS}
-
-The causal bandit game proceeds over $T$ rounds.
-In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations. 
-It then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, 
-including the \emph{reward} $Y_t \in \{0,1\}$. 
-After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T \in \actions$ based on its prior observations.
-
-The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refered to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.
-
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
-The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.
-Intuitively though, one would expect to perform better by making use of the extra structure and observations.
-
-\subsection*{The parallel bandit problem}
-
-In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem which we call the {\it parallel bandit}.
-It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \ref{XXX}, including the farming example.
-It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
-
-The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:parallel}.
-All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
-
-In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high).
-The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.
-
-\begin{figure}
-    \begin{subfigure}[b]{0.34\textwidth}
-	\centering    
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-\end{tikzpicture}
-        \caption{Parallel graph}
-        \label{fig:parallel}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.2\textwidth}
-    \centering
-        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_1$};
-\node[main node, below left=of 1](2){$X_2$};
-\node[main node, below right=of 1](4){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-    (1) edge (4)
-    (2) edge (4);
-\end{tikzpicture}
-        \caption{Confounded graph}
-        \label{fig:causalStructure_confounded}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.4\textwidth}
-    \centering
-         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, right=of 4](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-  	(2) edge (3)
-    (3) edge (4)
-    (4) edge (5);
-\end{tikzpicture}
-        \caption{Chain graph}
-        \label{fig:causalchain}
-    \end{subfigure}
-    \caption{Causal Models}\label{fig:causalmodels}
-    \vspace{-.5cm}
-\end{figure}
-
-In each round the learner either purely observes by selecting $do()$ or sets the value of a single variable.
-The remaining variables are simultaneously set by independently biased coin flips.
-The value of all variables are then used to determine the distribution of rewards for that round.
-Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
-
-The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
-$r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
-In the farming example, this choice of $Y$ models the success or failure of a seasons crop, 
-which depends stochastically on the various environment variables.
-
-\paragraph{The Parallel Bandit Algorithm}
-%\label{sub:par-bandit-alg}
-The algorithm operates as follows.
-For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
-Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
-The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates of their returns could be poor.
-To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions.
-The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$).
-For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$.
-Define
-
-\eq{
-\label{eq:m-simple}
-m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
-}
-
-\begin{algorithm}[H]
-\caption{Parallel Bandit Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} Total rounds $T$ and $N$.
-\FOR{$t \in 1,\ldots,T / 2$}
-\STATE Perform empty intervention $do()$
-\STATE Observe $\vec{X}_t$ and $Y_t$
-\ENDFOR
-\FOR{$a = do(X_i = x) \in \actions$}
-\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
-\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
-\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
-\ENDFOR
-\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
-\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
-\FOR{$a = do(X_i = x) \in A$}
-\FOR{$t \in 1,\ldots,T_A$}
-\STATE Intervene with $a$ and observe $Y_t$
-\ENDFOR
-\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
-\ENDFOR
-\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
-\end{algorithmic}
-\end{algorithm}
-
-$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimize the worst-case simple regret.
-When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
-We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilizes the samples captured during the observational phase to estimate $m(\vec{q})$.
-Although very simple, the following two theorems show that this algorithm is effectively optimal.
-
-\begin{theorem}\label{thm:uq-simple}
-Algorithm \ref{alg:simple} satisfies
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.
-}
-\end{theorem}
-
-\begin{theorem}\label{thm:lower}
-For all strategies and $T$, $\vec{q}$, there exist rewards such that
-$\displaystyle \simpleregret 
-\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)$.
-\end{theorem}
-
-The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
-
-The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
-of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
-
-%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
-By utilizing knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
-Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$.
-Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting.
-In practice, you would combine the data from both phases to estimate rewards for the low probability actions.
-We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.
-
-\subsection*{General graphs}
-We now consider the more general problem where the graph structure is known, but arbitrary.
-For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \citep{Pearl2000}.
-\eq{
-\P{X_1...X_N|do(X_i=j)} = 
-\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
-} 
-where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.
-
-We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorization to 
-write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
-estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
-from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
-takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
-simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
-high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
-Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
-get a good estimate for $\P{Y|do(X_2=1)}$.
-
-To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
-allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
-are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
-or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
-start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
-Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
-Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
-
-\begin{algorithm}[H]
-\caption{General Algorithm}\label{alg:general}
-\begin{algorithmic}
-\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
-\FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $a_t$ from $\eta$
-\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
-\ENDFOR
-\FOR{$a \in \calA$}
-\STATE
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
-}
-\ENDFOR
-\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
-\end{algorithmic}
-\end{algorithm}
-
-Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
-Let $\parents{Y}(X)$ denote the realization of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
-
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}\,, 
-}
-
-where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently.
-The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
-
-The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
-\eq{
-m(\eta) = \max_{a \in \calA} \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,,\text{ where } \EEa \text{ is the expectation with respect to } \Pn{a}.
-}
-
-We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
-
-\begin{theorem}\label{thm:general}
-If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$
-
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
-}
-\end{theorem}
-
-The proof is in Section \ref{sec:thm:general}.
-
-Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
-Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$.
-Thus it has broader applicability to the parallel graph than the generic algorithm given here.
-We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
-for future work.
-
-\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$.
-In light of Theorem \ref{thm:general}
-a natural choice of $\eta$ is the minimiser of $m(\eta)$.
-\eq{
-\eta^* 
-= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
-}
-Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
-Therefore the minimisation problem may be tackled using standard techniques from convex optimisation.
-The quantity $m(\eta^*)$ may be interpreted as the minimum achievable worst-case variance of the importance weighted estimator.
-In the experimental section we present some special cases, but for now we give two simple results.
-The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
-
-\begin{proposition}\label{pro:m-bound}
-$m(\eta^*) \leq |\calA|$. \textit{Proof.} 
-\textup{By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|\,\forall a$.}
-\eq{
-m(\eta) 
-= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
-= \max_a \EEa\left[\frac{1}{\eta_a}\right] = |\calA| %\qedhere
-}
-\end{proposition}
-
-The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$.
-This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
-The proof is in section XXX.
-
-\begin{remark}\label{rem:truncate}
-The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
-often possible to choose $B_a$ significantly
-larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
-and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
-to learn from observational data.
-\end{remark}
-
-\subsection*{Experiments}
-We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \cite{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
-Thomson sampling and UCB are optimized to minimize cumulative regret.
-We apply them in the fixed horizon, best arm identification setting by running them upto horizon $T$ and then selecting the arm with the highest empirical mean.
-The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.
-
-Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms).
-$Y_t \sim \bernoulli(\frac{1}{2}+\epsilon)$ if $X_1=1$ and $Y_t \sim \bernoulli(\frac{1}{2}-\epsilon')$ otherwise, where $\epsilon' = q_1\epsilon/(1-q_1)$.
-This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
-We set $q_i = 0$ for $i \leq m$ and $\frac{1}{2}$ otherwise.
-Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
-For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
-The code is available from \url{<https://github.com/finnhacks42/causal_bandits>}
-
-In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
-The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure.
-For the causal algorithms it increases approximately with $\sqrt{m}$.
-As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.
-
-\begin{figure}[h]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
-    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment2_20161020_1249.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
-        \label{fig:simple_vs_T_vary_epsilon}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment3_20161020_1252.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
-    		\label{fig:simple_vs_T}
-    \end{subfigure}
-    \caption{Experimental results}
-    \label{fig:experiments}
-\end{figure}
-
-Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
-This gap is learnable by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
-In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
-The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.
-
-For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
-This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be minimax optimal asymptotically \citep{LMS14}.
-
-\subsection*{Additional experiments} \todo {tie section together}
-\begin{figure}[H]
-	\centering    
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
-\node[main node, above right=of 2](6){$Z$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5)
-    (6) edge (1) edge (2) edge (4);
-\end{tikzpicture}
-        \caption{Confounded graph}
-        \label{fig:parallel_confounded} 
-\end{figure}
-
-We now compare the general algorithm with a range of standard bandit algorithms on the confounded graph in Figure \ref{fig:parallel_confounded}.
-All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action, $\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}$.
-We choose this setting because it generalises the parallel bandit, while simultaneously being sufficiently simple that we can compute the exact reward and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
-As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.
-
-In Figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$. For some $2 \leq N_1 \leq N$ we define 
-\eq{
-P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
-P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
-}
-As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon P(X_1=1) / P(X_1=0)$.
-The value of $N_1$ determines $m$ and ranges between $2$ and $N$.
-The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
-
-In Figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.  
-In Figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1, but exclude actions on $Z$ from the set of allowable actions to demonstrate that Algorithm 1 can fail in the presence of a confounding variable, which occurs because it incorrectly assumes that $P(Y|do(X)) = P(Y|X)$.
-We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
-
-\eq{
-P(X_i = 1|Z = 0) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.2 & \text{ if } i = 7 \\
-.7 & \text { otherwise} 
- \end{cases}\\
- P(X_i = 1|Z = 1) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.8 & \text{ if } i = 7 \\
-.3 & \text { otherwise} 
- \end{cases}\\
-}
-
-In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree.
-It is sub-optimal to act on either $X_7$ or $X_N$, while all other actions are optimal.
-The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.
-All remaining variables are likely to be identified as balanced and estimated from observational estimates.
-The CPD values have been chosen to demonstrate the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
-
-\begin{figure}[H]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
-    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
-        \label{fig:simple_vs_T_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment5_20161023_2118.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m(\eta*)=4.3$ with no actions setting $Z$}
-    		\label{fig:simple_vs_T_misspecified}
-    \end{subfigure}
-    \caption{Experimental results on the confounded graph}
-    \label{fig:experiments_confounded}
-\end{figure}
-
-\subsection*{Discussion \& Future work}
-
-Algorithm~\ref{alg:general} for general causal bandit problems 
-estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
-Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ where 
-the parameter $m(\eta)$ measures the difficulty of learning the causal model and is always less than $N$.
-The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.
-
-The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator $\hat{\mu}_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $\Pn{a}{\parents{Y}}$ for all $a \in \calA$.
-In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.
-Despite these extra estimates, Theorem~\ref{thm:lower} shows that this approach is optimal (up to log factors).Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
-
-\paragraph{Making Better Use of the Reward Signal}
-Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
-In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
-Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
-This would likely improve the poor performance of our causal algorithm relative to the sucessive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
-
-For the parallel bandit the required modifications should be quite straightforward.
-The idea would be to adapt the algorithm to essentially use successive elimination in the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability.
-In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
-Note that these modifications will not improve the minimax regret, which at least for the parallel bandit is already optimal.
-For this reason we prefer to emphasize the main point that causal structure should be exploited when available.
-Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
-This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.
-
-\paragraph{Cumulative Regret}
-Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
-In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information
-to get near-optimal cumulative regret guarantees.
-They consider a finite-armed bandit model with side information where in reach round the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
-In this way the learner can gain information about actions it does not take with varying levels of accuracy.
-The reduction follows by substituting the importance weighted estimators in place of the Gaussian reward.
-In the case that $\vec{q}$ is known this would lead to a known variance and the only (insignificant) difference is the Bernoulli noise model.
-In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
-at least asymptotically.
-
-%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
-%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
-%\todom{Check this!}
-%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
-
-The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
-The feedback graph is distinct from the causal graph.
-A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
-For this parallel bandit problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
-However, \citet{Alon2015,Kocak2014} give adversarial algorithms, which when applied to the parallel bandit problem obtain the standard bandit regret.
-A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
-This is equivalent to a nominally stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
-
-\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
-In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
-\cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-\paragraph{Causal Models with Non-Observable Variables}
-If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
-non-observable variables. Some of the interventional distributions may be non-identifiable meaning we can not obtain prior estimates for $\Pn{a}{\parents{Y}}$ from 
-even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
-\ref{alg:general} cannot be used. Estimating these quantities while simultaneously minimising the simple regret is an interesting and challenging open problem.
-
-% For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
-% In this case, some conditional distributions may be non-identifiable. 
-% The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
-% We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
-% Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
-
-% More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
-In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
-The quality of the estimate notably depends not only on the number of times that action was selected.
-For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$.
-This more complex side information does not fit within the graph feedback framework.
-
-\paragraph{Partially or Completely Unknown Causal Graph}
-A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
-The latter case arises naturally if we assume free access to a large observational dataset, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014b} could potentially be incorporated into a causal bandit algorithm.
-In particular, \citet{Hu2014b} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
-Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational dataset would be much more challenging.
-
-\subsection*{Proofs}
-
-\chapter*{Causality \& Fairness}
-\begin{itemize}
-\item There is an additional connction between causality and fairness, that arises when you know that there is a bias in the data and wish to correct for it.
-In causality, it may be due to selection bias or a confounding variable.
-In fairness, it may be due to historical disadvanage.
-In both cases we may have theory that tells us the direction or scale of an effect.
-Ie, we know that race itself has no inherent effect on outcomes such as criminality and education.
-If you have a dataset and a collection of such priors, where the data contradicts these priors due to data bias.
-How do you best make decisions based on that data?
-\end{itemize}
-
-\chapter*{Causality in Marketing}
-\section*{Attribution}
-\section*{Econometric Modelling}
-
-\chapter*{Conclusions (1000 words)}
-
-\section*{Open questions}
-\paragraph{Cycles} - a huge issue. Not covered by Pearl,Rubin etc.
-
-Places to look, statistical control theory, etc. any interesting papers along these lines?
-
-In the discrete case,
-Fung and Crawford (1990) have recently proposed a fast algorithm for constructing an
-independence graph from discrete data. We have not tested their procedure as a preprocessor for
-the PC algorithm. (COPIED FROM SPRITES)
-
-\bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{../library}% write the directory to the .bib file
-\end{document}
+Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.

### 2017-05-19 thesis/thesis_map.tex
draft of chapter 2 complete

@@
 \usepackage{wrapfig}
+\usepackage{centernot}
 
@@
 \renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
-\newcommand{\cf}[2]{{#1}_{#2}}
+\newcommand{\cf}[2]{{#1}^{#2}}
 \newcommand{\bernoulli}{\operatorname{Bernoulli}}
@@
 
+\newcommand{\ci}{\mathrel{\perp\mspace{-10mu}\perp}}
+\newcommand{\nci}{\centernot{\ci}}
+
 \author{Finnian Lattimore}
@@
 \todo{figure out theoremstle is undefined}
-\def\ci{\perp\!\!\!\perp} % from Wikipedia
+%\def\ci{\perp\!\!\!\perp} % from Wikipedia
 
@@
 
-3100 words at the start of boot camp.
-
-\chapter*{Random notes}
-
-\begin{figure}[h]
-\centering
-\caption{Causal inference without a framework}
-\label{fig:mutilatednet}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:mutilatedOriginal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.5\textwidth}
-\centering
-\caption{network after an intervention $G^*$}
-\label{fig:mutilatedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-\begin{figure}[h]
-\caption{}
-\label{fig:causal_inference_without_do}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){Treatment};
-\node[node, right=of 2](3){Outcome};
-\node[lt, above=of 3](4){U};
-\path[]
-	(2) edge (3)
-	(4) edge (3);
-
-\end{tikzpicture}
-\end{figure}
-
-There is a connection here to the degree to which the model is invariant.
-A model that is invariant to more things is more explanatory.
-We could look at levels of hierarchy.
-There is causal graphical models, which are invariant to interventions on variables.
-Can I tie this to a physical system and give more in depth examples of the nested tier of hierarchies?
-
-I do not see the distinction between explanation and (causal) prediction.
-Explanation is all about the ability to compress and to generalise.
-The more a model can do this, the more we view as providing an understanding of the why.
-
-Sometimes the causal component is obvious.
-
-Relationship to generalisation.
-Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time.
-The assumption is there are less places for change to come in.
-
-If a feature is a cause of an outcome then changing the input distribution over that feature won't break the model. If its an effect it could.
-
-The direct causes (and effects) of a variable of interest make up a sufficient set for prediction (is this true)?
-This may be a reason for using structure learning type algorithms even if you are simply doing prediction.
-
-\subsubsection*{The role of assumptions in causal inference}
-
-Fundamental challenges. How do you cross-validate or compare causal inference models? Lack of real world data on which to compare algorithms.
-
-Does predictive accuracy indicate a good causal model?
-
-Assumptions must be recognised. Without assumptions - only description is possible. Recognising the assumption (and associated risk) means that we understand we should still attempt to experiment.
-
-Generalising the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parachutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalise).
-(and given sufficient generalisability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)
-
-Is causal inference possible (are the conclusions ever valid)?
-
-Only from randomised experiment ... if people can do it without recourse to constant randomised experiment then an algorithm exits.
-
-The challenges and limitations of inferring causal effects from observation data are numerous.
-However, there are problems we need to solve.
-The questions is are these problems so serious as to warrant disregarding all non-experimental data.
-I think answer is a clear no.
-Need ongoing validation, with assumptions.
-We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).
-
-\subsection*{Relationship to interpretability}
-
-A desire for interpretability indicates that something has been left out of the loss function.
-
-Are causal/interpretable models more reliable or more likely to generalise well?
-
-One form of interpretability gives people insight into what the features are that the model is relying on.
-
-If we know the training and test data will be sampled from different distributions, knowing what the features that the model is looking at are, allows people use their background understanding of the world to evaluate whether or not those features are likely to be transferable to the test domain.
-
-Specifically, people can
-\begin{itemize}
-\item rule out many possible features as highly unlikely to be relevant to a problem
-\end{itemize}
-
-People have access to a lot of detailed prior knowledge.
-
-\subsection*{Relationship to transfer learning}
-
-find a feature representation in which $P(Y|X)$ is the same in many different domains (or stable over time).
-Causal models predict the outcome of actions.
-We could directly take these actions and learn $P(Y|a,X)$ for every $(a,X)$ but, in reality, no two situations (or actions) are exactly alike.
-So we have to make representations such that things are stable.
-
-This is tightly related to generalisability.
-If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results.
-We don't include , the color of the doctors shirt, the size of the room, ...
-
-For example, in the advertising setting, we want to know how our on expenditure on paid search ads is linked to sales.
-However, this relationship may be very unstable over time because the ad slots are sold at auction.
-The amount we have to pay to obtain a given position for keyword depends crucially on the amount our competitors are bidding for that keyword.
-However, the relationship between displaying the ad at a particular position and the probability that someone clicks it an then makes a purchase may be much more consistent.
-
-Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.
+\tableofcontents
 
@@
 
-\section{What is causality and why do we care? (2000 words)}
+\section{What is causality and why do we care? }
 
@@
 \end{itemize}
-\subsection{Overview of this thesis}
+
 \subsection{Identifying when we have a causal problem}
@@
 
-\section{Approaches to causality (1000 words)}
+\section{Approaches to causality}
 
@@
 We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
-We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.
+We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectivness of early childhood education including the landmark randomised trial, the Perry Preschool project \citep{weikart1970longitudinal}}
 \end{example}
@@
 If better off parents are more likely to send their children to pre-school these children more likely to graduate high school regardless, then simply comparing the high school graduation rates of those who attended preschool with those who did not will overstate the benifit of preschool.
-To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group, $\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}$.
+To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group,
+
+\eqn{
+\label{eqn:backdoor_example}
+\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}
+}
 
@@
 
-In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{i}^{1}$ versus what would happen if they got the placebo, $y_{i}^{0}$.
+In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would happen if they got the placebo, $\cf{y_{i}}{0}$.
 The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
@@
 We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate, from an experimental or observational study, the probability that people recover if treated $P(Y|X=1)$ and the probability that they recover if not treated $P(Y|X=0)$.
-Now if $X=0$ then $Y = \cf{Y}{0}$.
+We can estimate, (from an experimental or observational study);
+
+\eq{
+\P{Y|X=1}& \text{, the probability that those who took the treatment will recover}\\
+\P{Y|X=0}& \text {, the probability that those who were not treated will recover}
+}
+
+Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
+For those who did not take the treatment, the observed outcome is the sames as the outcome \emph{had} they not taken the treatment.
 Equivalently stated:
@@
 \P{Y^{0}|X=0}&= \P{Y|X=0}\\
-\P{Y^{1}|X=1}&=P{Y|X=1}
+\P{Y^{1}|X=1}&=\P{Y|X=1}
 }
@@
 They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-Again this is justified in our example due to the randomisation of treatment assignment.
-In general these assumption do not hold.
-If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model.
-that influenced both the decision to get treatment and the likelihood of recovery given treatment or placebo.
-
-\todo{SUTVA assumption}
-
-There are some complex philosophical objections to counter factuals arising from the way they describe alternate universes that were never realised and are not empirically testable \ref{XXXXX}.
-This can result in some concrete problems.
-Consider the following example based on  \ref{Dawid2000}.
-
+This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
+In general the treatment assignment will not be independent of the potential outcomes.
+In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated highschool had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$.
+Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$.
+A more general form of the ignore-ability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
+
+\vspace*{.3cm}
+\begin{theorem}[Ignore-ability] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
+
+\eqn{
+\label{eqn:counterfactual1}
+\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
+\label{eqn:counterfactual2}
+\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
+}
+\end{theorem}
+
+If we assume that within each socio-economic status level, attendence at pre-school is independent to the likelyhood of graduating high-school had you attended then we can compute the average rate of high-school graduation given a universal pre-school program from equation \ref{eqn:counterfactual1}.
+Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
+
+Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \cite{Rubin1978}.
+This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
+As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
+The later school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those did and who now share the classroom with them.
+SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
+
+There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
+This makes it quite easy to (accidentally) make statements with counterfactuals that can not be tested with empirical data.
+Consider the following example based on \citet{Dawid2000}.
 Again we have a drug, where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
-Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal.
-We will assume equal variance for simplicity.
+Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal with equal variance (for simplicity).
 
 % vector with mean mu_t,mu_c and a covariance matrix
-\eq{
-P(Y_t,Y_c) \sim N()
-}
-
-then their difference is also normal,
-
-\eq{
-P(\tau) = N(\mu_t - \mu_c,2\sigma^{2}(1-\rho))
-}
-
-\includegraphics[scale=.6]{figures/counterfactual_nonidentify.pdf}
-
-Key issue is that we can never observe the joint distribution over $Y_t$ and $Y_c$.
-As a result, the variance of $\tau$ is not identifiable, even from experimental data.
-It seems on the face of it that $\tau$ is relevant to our decision making.
-If the example above, if $\rho = 1$ then almost everyone benefits slightly from the treatment.
-However if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.
-
-Can these issues be resolved? If so, how?
-
-- note that we can bound this counterfactual distribution based on the variance of the observed interventional distributions. If the variance of XXX is small then this may not be an issue.
-
-- does it actually make sense to make a decision on the basis of this counterfactual that is a function of something we could never observe. I
-
-- part of the problem may be due to the deterministic way we have phrased individual treatment effects.
-
-Can these issues be resolved by considering personalised causal effects as random variables and avoiding counterfactuals?
-
-A range of similar issues can arise in counterfactuals \ref{XXXX}. 
-- What assumptions are required for the kind of counterfactual analysis like X would have been higher had B ...?
-- briefly discuss the different flavours of counterfactual questions here.
+\eqn{
+P(\cf{Y}{1},\cf{Y}{0}) \sim N( 
+         \begin{bmatrix}
+           \mu_{1} \\
+           \mu_2 \\
+         \end{bmatrix},
+         \begin{bmatrix}
+           \sigma^2 & \rho \sigma^2 \\
+           \rho \sigma^2 & \sigma^2 \\
+         \end{bmatrix}
+         )
+}
+
+Their difference is also normal. Let $\tau = \cf{Y}{1} - \cf{Y}{0}$,
+
+\eqn{
+P(\tau) = N(\mu_1 - \mu_0,2\sigma^{2}(1-\rho))
+}
+
+\begin{figure}[h]
+\captionsetup[subfigure]{position=b}
+\centering
+\caption{The distribution of individual treatment effects is not identifiable, even from a randomized controlled trial.}
+\label{fig:metaphysical_distribution_difference}
+\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$.
+The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated.
+\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
+\hspace{0.02\textwidth}
+\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
+{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
+\end{figure}
+
+From a (large) randomized controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}.
+These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated retrospectively.
+However, the distribution over the individual causal effects depends on $\rho$, see figure \ref{fig:metaphys_b}.
+The key problem is that we can never observe the joint distribution over $\cf{Y}{1}$ and $\cf{Y}{0}$.
+As a result, $\rho$ and thus the variance of $\tau$ is not identifiable, even from experimental data.
+\citet{Dawid} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects.
+He further points out that the interventional distributions in figure \ref{fig:metaphys_a}, along with a loss function, contain all the information required to decide how to treat a new patient.
+
+This result is unintuative.
+It seems on the face of it that the distribution of individual causal effects is relevant to our decision making.
+If $\rho = 1$ then almost everyone benefits slightly from the treatment whilst if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.
+This confusion can be resolved by thinking about personalised rather than individual causal effects.
+It is entirely possible that potentialy observable characteristics (such as gender, age, genetics, etc) effect how people will respond to the treatment.
+We can partition people into sub-populations on the basis of these characteristics and measure different \emph{personalised} causal effects for each group.
+The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much we stand to gain from further personalisation.
+The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).
 
 One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than taking about the distribution of Y given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $X$.
-This is certainly convenient, if rather impressive.
+Rather than taking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
+This is certainly convenient, if rather imprecise.
+However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrents careful attention.
+We should always be clear what assumptions we are making and have in mind if it is possible (at least in theory) to validate those assumptions.
 
@@
 Each variable is a node and arrows are drawn from causes to their effects.
-For example \ref{exm:ranomized_experiment} the SEM is:
+Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.
 
 \begin{figure}[h]
-\caption{SEM for examples \ref{exm:ranomized_experiment}}
-\label{fig:causal_network_example2}
+\caption{SEM for example \ref{exm:ranomized_experiment}}
+\label{fig:sem_randomized_treatment}
 \centering
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = x$};
-\node[node, right=of 2](3){$Y = f(X,\epsilon_y)$};
+\node[node](2){$X = f_x(\epsilon_x)$};
+\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
 \node[lt, above=of 3](4){$\epsilon_y$};
+\node[lt, above=of 2](5){$\epsilon_x$};
 \path[]
 	(2) edge (3)
-	(4) edge (3);
+	(4) edge (3)
+	(5) edge (2);
 \end{tikzpicture}
@@
 Note that the causal graph in figure \ref{fig:causal_network_example2} is identical to that the bayesian network for the same problem, figure \ref{fig:causal_network_example}.
-The latent variable $\epsilon_y$ was not explicitly drawn in figure \ref{fig:causal_network_example} as the noise can be captured within the probabilistic node, however adding it would not change the model.
-
-We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'?
-Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$.
-The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}.
-However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$.
-The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomised experiment.
-In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.
-
-For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
+The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as are captured by the probabilistic nature of the nodes in a Bayesian network, however adding them would not change the model.
+
+Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$.
+The function $f_y$ and distribution over $\epsilon_y$ does not change.
+This results in the interventional distribution \footnote{We have assumed the variables are discrete only for notational convinience},
+
+\eqn {
+\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
+}
+
+The observational distribution of $Y$ given $X$ is,
+
+\eqn{
+\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
+& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
+}
+
+The interventional distribution is the same as the observational one.
+The same arguament applies to the intervention $do(X=0)$ and so the causal effect is just the difference in observed outcomes as we found via the causal Bayesian network and counterfactual approaches.
+
+The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
+Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
+
+\eqn{
+\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
+\label{eqn:sem:adjusting}
+&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
+}
+Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivelent to the observational distribution, given by;
+
+\eqn{
+\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
+}
+
+\begin{figure}[h]
+\caption{SEM for example \ref{exm:adjusting}}
+\label{fig:sem:preschool}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](2){$X = f_x(Z,\epsilon_x)$};
+\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
+\node[lt, above=of 6](7){$\epsilon_z$};
+\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
+\node[lt, above=of 3](4){$\epsilon_y$};
+\node[lt, above=of 2](5){$\epsilon_x$};
+\path[]
+	(2) edge (3)
+	(4) edge (3)
+	(5) edge (2)
+	(7) edge (6)
+	(6) edge (2) edge (3);
+\end{tikzpicture}
+\end{figure}
+
+Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
+A structural equation model with $N$ variables looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
 However a SEM is, by definition, more than a set of simultaneous equations.
@@
 When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-
-Limitations \& Criticisms 
-- can express some non-measurable things
-	- example of that
-
-- confusion may occur as to whether interpretation is causal
-
-- can they handle cycles?
+The similarity between the notation used to describe and analyse structural equation models and simultaneuos equations, combined with a reluncatnce to make explicit statements about causality has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
 
@@
 Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-This makes it straightforward to combine key results and algorithms developed within any of these frameworks.
-For example, draw a graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores \ref{} estimate the effect.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-
 If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
@@
 It is possible to relax the independence of errors assumption we made for SEMs slightly to correspond exactly the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \cite{Richardson2013}.
-To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of functional models or bayesian networks or as conditional independence statements involving counterfactual variables (ignorability assumptions).
-By non-parametrically, I mean that we are not making any assumptions about the form of the relationships between variables.
-Models that are not non-parametrically identifiable can still be identified given assumptions about the distributions of variables and the functional relationship between them, for example, that the functions are linear or that the noise is additive \cite{Peters2014}.
-This form of assumption fits extremely naturally into the structural equation framework.
-
-However we can also pose causal queries that are not interventional and cannot be phrased in terms of the do-notation.
-The patients in our drug treatment example could be broken down into four groups.
-\todo{point out objections to doing this}.
-The first group will recover whether or not they receive treatment, the second group will recover if treated but not on the placebo, the third group will recover on the placebo and not if treated, and the last group will not recover on treatment or placebo.
-Unfortunately, we don't know which group each person belongs to.
-Drawing this up as a table:
-
-\begin{tabular}{c|c|c|c}
-group & placebo & treatment & probability of group\\
-\hline
-1 & die & die & $\alpha=P(Y^{0}=0,Y^{1}=0)$\\
-2 & die & recover & $\beta=P(Y^{0}=0,Y^{1}=1)$\\
-3 & recover & die & $\gamma=P(Y^{0}=1,Y^{1}=0)$\\
-4 & recover & recover & $\delta=P(Y^{0}=1,Y^{1}=1)$\\
-\end{tabular}
-
-The queries we have been asking thus far are about $P(Y^{0}=1) = \gamma + \delta$ and $P(Y^{1}=1) = \beta + \delta$, but suppose we asked the question; what is the probability that this patient, who was not treated and died, would have recovered if they had been treated?
-We know they are in either group 1 or 2 since they died without treatment, so the answer is $\frac{\beta}{\alpha+\beta}$.
-Can we estimate the $\alpha, \beta, \gamma, \delta$ or in other words, identify the joint distribution over the counterfactuals $P(Y^{0},Y^{1})$ given the interventional distributions, $P(Y^{0})$ and $P(Y^{1})$?
-The answer is no, putting our constraints and unknowns in matrix form:
-
-\begin{equation}
-\left(
-\begin{array}{cccc}
-0&0&1&1\\
-1&1&0&0\\
-0&1&0&1\\
-1&0&1&0\\
-1&1&1&1\\
-\end{array}
-\right)
-\left(
-\begin{array}{c}
-\alpha\\
-\beta\\
-\gamma\\
-\delta\\
-\end{array}
-\right)= 
-\left(
-\begin{array}{c}
-P(Y^{0})\\
-1-P(Y^{0})\\
-P(Y^{1})\\
-1-P(Y^{1})\\
-1\\
-\end{array}
-\right)
-\implies
-\left(
-\begin{array}{c}
-\alpha\\
-\beta\\
-\gamma\\
-\delta\\
-\end{array}
-\right)=
-\left(
-\begin{array}{c}
-1-P(Y^{0})-P(Y^{1})+\delta\\
-P(Y^{1})-\delta\\
-P(Y^{0})-\delta\\
-\delta\\
-\end{array}
-\right)
-\end{equation}
-The value of $\delta$ is not determined so the query is not identifiable.
-However we do get bounds on the terms.
-Since probabilities cannot be negative, $P(Y{1})-P(Y^{0})-1 \leq \delta \leq min(P(Y^{1}),P(Y^{0})$.
-Note, if we made the additional assumption $\gamma=0$; that the drug did not cause anyone to die who would otherwise have survived, then we can determine the joint distribution over counterfactuals.
-Alternatively, if we could assume that after treatment people returned to their initial state after some period of time, (say we were testing a drug for acne) then we could run a crossover study to determine the joint distribution.
-In a crossover study, the participants are randomly assigned to treatment and placebo, results are measured and then the groups are swapped.
-The scientific and philosophical validity of counterfactual queries remains under question \cite{Dawid2000,Dawid2014}, however they are nonetheless widely posed in the form of attribution of causal effects to different pathways and mediation \cite{Pearl2014,Imai2010a,VanderWeele2011}.
-
-There are differences between the models we have considered when it comes to counterfactual queries.
-Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.
-Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made.
-Structural equation models with independent errors allows the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.
-
-In practice, differences in focus and approach across different fields eclipse these actual differences in the models.
-The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations.
+This correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
+For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
+Then use propensity scores\cite{Rosenbaum1983} to estimate the effect.
+If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
+The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
+These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and legal discrimination \citep{Pearl2000}.
+
+In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
+The work on causal graphical models \cite{Pearl2000,Sprites} focuses on asymptopic, non-parametric estimation and rigorous theoretical foundations.
 The Neyman-Rubin framework builds on our understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
@@
 
-\subsubsection{A translator from graphical independence to counter factual statements}
-
-\subsubsection{More terminology}
-With the graphical framework in place, it is useful to define some key terminology used in describing causal models in terms of the graph structure they refer to.
-
-\begin{enumerate}
-\item confounding
-\item exogenous
-\item endogenous
-\item nuisance variables
-\end{enumerate}
-
 \section{What does a causal model give us? Resolving Simpson's paradox}
 
-We will now demonstrate how we can use our new notation and frameworks for causal inference to resolve a fascinating paradox.  \todo{ insert some history of Simpson's paradox here.}
-
-\cite{simpson1951interpretation}
-
+We will now demonstrate how we can apply our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
+The following example is adapted from \citet{Pearl2000}.
 Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
@@
 \begin{table}[h]
+\caption{Treatment results}
 \label{tab:simpson1}
-\caption{Treatment results}
 \center
@@
 \begin{table}[h]
+\caption{Treatment results by gender}
 \label{tab:simpson2}
-\caption{Treatment results by gender}
 \center
@@
 The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
-Unfortunately, the data in tables tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable us to estimate the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
+Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable us to estimate the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
 We require some assumptions about the causal relationships between the variables.
@@
 
-Note that we have not changed the data itself, only the story we tell around it.
-This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data itself but in what assumptions we are willing to make.
+Note that we have not changed the data itself, only the description of the variables that it is associated with.
+This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data but in what assumptions we are willing to make.
 From a purely statisitcal viewpoint there is no paradox.
-The reversal is just a case of INSERT VECTOR ARGUAMENT HERE.
-The 'paradox' only arises when we need to use the data to select an intervention.
-
-I should also point out that there are many other plausible causal graphs for both scenarios above.
+The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
+The paradox only arises when we attempt to use the data to select an intervention and is resolved when we apply a causal approach to do so.
+
+\eqn{
+\label{eqn:simpsons_paradox}
+\exists \set{N_1,...N_4, N_1'...N_4'} \in \mathbb{N} : \; 
+\textcolor{green}{\frac{N_1}{N_2}< \frac{N_1'}{N_2'}}, \;\; 
+\textcolor{blue}{ \frac{N_3}{N_4}<\frac{N_3'}{N_4'}}
+ \text{ and } \frac{N_1+N_3}{N_2+N_4} > \frac{N_1'+N_3'}{N_2'+N_4'}
+}
+
+\begin{figure}
+\centering
+\caption{Simpson's reversal visualized.
+The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
+However, when we sum them, the ratio is steeper for the un-primed variables.}
+\label{fig:simpsons_vector}
+\includegraphics[scale=.4]{figures/simpson_vector}
+\end{figure}
+
+There are many other plausible causal graphs for both scenarios above.
 Perhaps income effects drug choice as well as gender, or gender could effect treatment choice and blood preassure control given treatment, etc.
-Causal modeling provides us with a powerful
-
-Human in the loop ML. Why would you need the human?
-
-\todo{mention relationship to ecological fallacy}
+Causal modeling provides us with a powerful tool to specify such assumptions and determine how to estimate causal effects for a given model see section \ref{sec:causal effect estimation}.
 
@@
 \section{Causal effect estimation} 
+\label{sec:causal effect estimation}
 

### 2017-06-22 thesis/offcuts.tex
completes draft of chapter3 working on chapers 4 and 5

@@
 Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.
+
+Note that the consistency of the estimator in this case does not guarantee that it is optimal for a finite dataset.
+For example, the expected generalisation error may be reduced by adding a regularizer that reduces the variance at a cost of introducing some bias.
+
+Point out that in some cases we are only considering the treatment for a particular group - in which case we should focus attention on good estimates for the causal effect within that group.
+Causal effect of treatment on the treated is a special case of that.
+
+\eqn{
+ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
+}
+
+\eqn{
+\hat{\mu}_{do(x)} = \left(\frac{\sum_{i=1}^n \ind{x_i = x,\vb{z_i} = \vb{z}}y_i}{\sum_{i=1}^n \ind{x_i = x,\vb{z_i} = \vb{z}}}\right)
+}
+
+What about research into efficient estimators for identifiable queries where the back door criterion is not satisfied. - eg front door criterion (probably little research).
+
+%\subsection{Non identifiable queries}
+In many cases the query of interest will not be identifiable.
+In this setting we are left with a choice between making additional assumptions (for example that variables are linearly related) or aiming to obtain bounds rather than point estimates for the causal effect of interest.
+Or we may do both.
+One of the most widely used methods in this category is the instrumental variable.
+
+Instrumental variables are often conceptualised as natural experiments.
+The goal is to find a variable that substantially influences the treatment but that is conditionally independent of the remaining variables in the network.
+
+For example, we might want to compare the results for juvenile offenders depending on whether they spent their time awaiting trial in a custom facility for young offenders as opposed to a unit within an adult jail.
+The process by which children are assigned to different facilities may well be non-random.
+However, it might be that in many cases the decision was based purely on the fact that the juvenile facility was full.
+We could then use the number of places remaining in the juvenile facility as an instrumental variable.
+
+The causal network for an ideal instrumental variable is shown in figure XXX.
+It is not identifiable (non-parametrically) because.
+If we assume linearity it ?
+\todo{what exact guarantees do we get for instrumental variables}.
+Without a linearity assumption we would want the instrumental variable to be highly predictive of the treatment assignment so as to obtain reasonable bounds.
+But if we make the assumption that everything is linear then it would seem that even a weak instrumental variable should (in theory) suffice.
+
+In some cases we might have a choice between adjusting for causal effects via the backdoor theorem or utilising an instrumental variable to obtain bounds.
+It is not the case that the former method is always superior even though it has better asymptotic results.
+For example
+
+DEMONSTRATE a specific example where instrumental variables obtain better estimates than adjustment even when all the criteria for adjustment have been met.
+Start with a case were the adjustment is very high dimensional and the instrumental variable is strong.
+(the effect of the variables on the treatment assignment will have to be week (otherwise the instrument is not strong).
+
+\subsection{Does causal effect estimation work}
+Really know one knows.
+
+There are relatively few studies that have compared the observational study results with those obtained by randomised experiments.
+One of the key examples is the XX job network study.
+Initial analysis suggested that a range of causal estimation techniques all ...
+
+A Cochrane meta-analysis of studies comparing observational with experimental results found ... \citep{Anglemyer2014}.
+
+\paragraph*{How can we construct data sets on which to test casual effect estimation techniques}
+One can start from a randomised experiment and deliberately drop some of the data. Draw up the graphical model for this process and the assumptions inherent in it.
+
+Modelling a problem as a causal graph only makes sense when rewards are generated stochastically - since causal graphs fundamentally model probability distributions over variables.
+Thus the connection is to stochastic bandit problems (although adversarial bandits algorithms may be applied to stochastic problems).

### 2017-06-25 thesis/offcuts.tex
working on chapter 4

@@
 Thus the connection is to stochastic bandit problems (although adversarial bandits algorithms may be applied to stochastic problems).
+
+\subsection{The exploration/exploitation trade-off}
+A central aspect to bandit problem is the trade of between exploiting the information we have already obtained with exploring options about which we remain uncertain.
+Suppose you have sampled five of your local restaurants and found one which you really enjoyed.
+How often should you eat there versus exploring new options.
+The degree to which you should explore versus exploit depends on how many step remain to go in the game and what assumptions you make about the distributions from which the rewards are sampled.
+If this is your last meal you might as well just eat the food you enjoy most but if you hope to live for many meals to come then you stand to gain by sampling a wider range of options.
+In many cases, the total number of rounds that you will get to play in the game is unknown.
+If you cannot imagine that it would be possible for a meal to be substantially better than that which you obtain (in expectation) at your current favourite then there is little point exploring further.
+
+\eqn{
+\rawregret = \max_{i=1,...,k} \left(\sum_{t=1}^T Y_{t,i} \right) - \sum_{t=1}^T Y_{t,a_t}
+}
+
+To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by selecting, in every time-step, the the single arm that yields the best reward in hindsight.
+
+The term regret is heavily overloaded in the bandit liturature.
+In addition to the quanity in equation \ref{eqn:raw_regret} it may refer to the difference between the reward obtained by the algorithm and the reward that would have been obtained by selecting, in every time-step, the the single arm that yields the best reward in hindsight \citep{} or too the expectation .
+
+\vspace{0.5cm}
+\begin{definition}[Expected Regret]
+\eqn{
+\label{eqn:expected_regret_definition}
+\expectedregret(\pi) = \E{max_{i \in [k]} \sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+}
+\end{definition}
+
+\vspace{0.5cm}
+\begin{definition}[Stochasitic k-armed bandit problem]
+The multi-armed bandit problem proceeds over $T$ rounds. In each round $t$, 
+\begin{enumerate}
+\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
+\item the world stochasitcally generates rewards for each action $i \in {1,...,k}$, $Y_{t,i} \sim \Pns{i}{y}$
+\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\end{enumerate}
+At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$.
+We denote the expected reward for the action $i$, $\E{ \Pns{i}{y}}$, by $\mu_i$ and the action with the highest expected reward by $i^*$.
+\end{definition}

### 2017-06-25 thesis/thesis_map.tex
working on chapter 4

@@
            draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
+           minimum width=2.3em,
+           minimum height=2.3em,
            font=\footnotesize,
@@
 \newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
-\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
+\newcommand{\Q}[1]{\operatorname{Q}\left(#1\right)}
 \newcommand{\EE}{\mathbb E}
@@
 \newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
-
+\newcommand{\indep}[1]{\mathcal{I}\left(#1\right)}
+\newcommand{\vb}[1]{\boldsymbol{#1}}
+\newcommand{\card}[1]{\left\vert{#1}\right\vert}
+\newcommand{\actionspace}{\mathcal{A}}
+
+\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
 \newcommand{\defined}{\vcentcolon =}
@@
 \newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
+
 \newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
@@
 
-\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
+\renewcommand{\P}[1]{\operatorname{P}\left(#1\right)}
 \newcommand{\cf}[2]{{#1}^{#2}}
@@
 \renewcommand{\vec}[1]{\boldsymbol{#1}}
+\newcommand{\regret}{\bar{R}_{T}} %psuedo-regret
+\newcommand{\expectedregret}{\mathbb{E}R_{T}} % expected regret
+\newcommand{\rawregret}{R_{T}} % plain regret 
+\newcommand{\quotes}[1]{``#1''}
 
@@
 
+\chapter*{Todo}
+\begin{enumerate}
+\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
+\end{enumerate}
+
 \chapter{Introduction}
@@
 The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
-From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries.
+From health care, to entertainment and self driving cars; machine learning algorithms will transform many industries.
 It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.
@@
 The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}.
-This has led to a reluctance among applied researchers in many fields to refer to cauality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
-However, the magnitude, direction and even existance of an association depends on what other variables we adjust for (or include in a regression).
+This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
+However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression).
 Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.
@@
 Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one).
-Problems highlighted as intractible in the reverse causal sense are also intractible in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning.
+Problems highlighted as intractable in the reverse causal sense are also intractable in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning.
 For example, the war question posed in Gelman.
 
-A distinction between forward causal iference, what happens if we do X and reverse casual inference
+A distinction between forward causal inference, what happens if we do X and reverse casual inference
 
@@
 
-\subsection{Identifying when we have a causal problem}
+\section{Identifying when we have a causal problem}
 
@@
 They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
-There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols.
+There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addition and treatment protocols.
 However, using it to decide on what basis to admit people could kill.
@@
 \item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
-\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
+\item Does acting on the result of the prediction change the predictive distribution p(y|x)? IE change people's behaviour.
 \item Ethics - ... People's viewpoint on if its OK...
@@
 Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
-To do this we need terminology to describe actions and how we anticipate the system should respond to them.
-Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks.
-In this chapter we describe these approaches.
-We will examine what problems they allow us to solve, what assumptions they rely on and how they differ.
+To do this we need terminology to describe actions and how we anticipate the system will respond to them.
+Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
+In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ.
 We will also use them to describe the following simplified examples.
-The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.
+The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.
 
@@
 \label{exm:ranomized_experiment}
-Suppose we have developed a new drug for some illness and wish to determine how effective it is.
-We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group.
-The people in the treatment group get the drug, everyone else gets a placebo pill.
-The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo.
+Suppose a pharmaceutical company wants to assess the effectiveness of a new drug on recovery from a given illness.
+This is typically tested by taking a large group of representative patients and randomly assigning half of them to a treatment group (who receive the drug) and the other half to a control group (who receive a placebo).
+The goal is to determine the clinical impacts of the drug by comparing the differences between the outcomes for the two groups (in this case, simplified to only two outcomes - recovery or non-recovery).
 We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
@@
 \label{exm:adjusting}
-Suppose we wish to estimate what would happen to high school graduation rates if we made pre-school available and compulsory for all 4 year olds.
-We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
-We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectivness of early childhood education including the landmark randomised trial, the Perry Preschool project \citep{weikart1970longitudinal}}
+Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all 4 year olds.
+We have a large cross-sectional data set on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
+We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
 \end{example}
 
-\section{Causal bayesian networks}
+\section{Causal Bayesian networks}
 
@@
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
+  thick,main node/.style={observed}]
 
@@
 \centering
-\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
+\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2|X_1}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
 \label{fig:bayesnet2}
@@
 \caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
  %nodes
@@
 The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges).
+There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true $G$ can have extra edges).
 If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
 
-\begin{equation}
-P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
-\end{equation}
-
-A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
+\eqn{
+\P{X_{1},...,X_{N}} = \prod_{i = 1...N}\P{X_{i}|parents_{X_{i}}}
+}
+
+A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
 This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
@@
 \begin{itemize}
-\item do(X=x) denotes an intervention that sets the random variable(s) $X$ to $x$.
+\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
 \item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$.
 This notation is somewhat overloaded.
-It may be used represent a probability, a probability distribution/mass function or a family of distribution functions depending on if the variables are discrete or continuous and whether or not we are treating them as fixed.
+It may be used represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
 For example it could represent
 \begin{itemize}
-\item The probability $\P{Y=1|do(X=x)}$ as a function of $x$
-\item The probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$
-\item The probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$
-\item a familiarly of density/mass function for $Y$ paramaterised by $x$.
+\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
+\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
+\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
+\item a family of density/mass function for $Y$ parameterised by $x$.
 \end{itemize}
-Where the distinction is important not clear from context we will use one of the more specific forms above. 
+Where the distinction is important and not clear from context we will use one of the more specific forms above. 
 \end{itemize}
@@
 
-\begin{theorem}[Truncated product formula]
+\begin{theorem}[Truncated product formula \citep{Pearl2000}]
 \label{thm:truncated_prodcut}
-If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorisation given by the network \citep{Pearl2000}.
+If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.
 
@@
   \begin{cases}
-  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
+  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistent with $Z=z$}  \\
    0       & \text{otherwise } 
@@
 
-Theorem \ref{thm:truncated_prodcut} does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution.
-The truncated product formula will give different results depending on which you choose.
-The result is possible with causal bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
+Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution.
+The truncated product formula will give different results depending on the selected network.
+The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
 In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.
@@
 Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$?
-The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and disease sub type of the patient.
-Since $X$ is assigned  via deliberate randomisation we know that it is not effected by any latent variables.
-The causal bayesian network for this scenario is shown in figure \ref{fig:causal_network_example} .This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$.
+The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient.
+Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables.
+The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}.
+This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$.
 In this example, the interventional distribution is equivalent to the observational one.
@@
 
-In example \ref{exm:adjusting} we are interested $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool.
-We could compare it to outlawing pre-school $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$.
-It seems reasonalble to assume that attending could pre-school effect the likelyhood of graduating from high school and that your parents socio-economic status effects both pre-school attendance and high-school graduation.
-If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal bayesian network in figure \ref{exm:adjusting}.
-In this case, the interventional distribution is not equivelent to the observational one.
-If better off parents are more likely to send their children to pre-school these children more likely to graduate high school regardless, then simply comparing the high school graduation rates of those who attended preschool with those who did not will overstate the benifit of preschool.
-To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group,
+In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool.
+We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$.
+It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation.
+If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}.
+In this case, the interventional distribution is not equivalent to the observational one.
+If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool.
+To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,
 
@@
 We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph.
-There is a very powerful and general set of rules that specify how we can transform observational distributions into intervetional ones for a given graph structure.
+There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure.
 These rules are referred to as the Do-calculus \cite{Pearl2000}.
-We discuss them further in section \ref{}.
+We discuss them further in section \ref{sec:do_calculus}.
 
@@
 \center
-\caption{Causal bayesian network for example \ref{exm:adjusting}}
+\caption{Causal Bayesian network for example \ref{exm:adjusting}}
 \label{fig:causal_adjust}
@@
 The power of defining an intervention in this way stems from the number of things that are invariant between the two processes.
-All the (conditional) distriubtions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
+All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
 
@@
 The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks.
-However, without a causal framework, we have to make assumptions about what will be invarient to the intervention specifically for each such problem we encounter.
-For complex problems, it is very difficult to conceptualise what assumptions we expect to hold without the benifit of a causal framework.
+However, without a causal framework, we have to make assumptions about what will be invariant to the intervention specifically for each such problem we encounter.
+For complex problems, it is very difficult to conceptualise the assumptions we expect to hold without the benefit of a causal framework.
 
@@
 \center
-\caption{Causal inference with ordinary bayesian networks.
-The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that the pre-interventional variable $X$ to $x$.
-The assumptions characterised by this plate model correspond to those implied by the causal bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
+\caption{Causal inference with ordinary Bayesian networks.
+The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$.
+The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
 As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
@@
 
-A causal bayesian network represents much more information than a bayesian network with identical structure.
+A causal Bayesian network represents much more information than a Bayesian network with identical structure.
 A causal network encodes all possible interventions that could be specified with the do-notation.
-For example, if the network in figure \ref{fig:causal_adjust} were an ordinary bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
-The equivelent causal bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal bayesian network would require and additional 30 parameters.
-\footnote{After each single variable intervetion we have a distribution over two variables, which can be represented by three parameters.
+For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
+The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
+Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
 After each two variable intervention, we have a distribution over one variables which requires one parameter.
-This takes us to a total of $6*3+12*1 = 30$ additional parameters.}
-
-Causal bayesian networks are bayesian networks so results that apply to bayesian networks carry directly across;the local Markov property states that variables are independent of their non-effects given their direct causes.
+This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
+
+Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across; the local Markov property states that variables are independent of their non-effects given their direct causes.
 Similarly the global Markov property and d-separation also hold in causal networks.
 
-\subsubsection{Limitations of causal bayesian networks}
+\subsubsection{Limitations of causal Bayesian networks}
 A number of criticisms have been levelled at this approach to modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will effect many variables in complex and non-deterministic ways \citep{2009causality,cartwright2007hunting}.
-However, the deterministic $do()$ operator turns out to be surprisingly effective at modelling more complex interventions by augmenting the causal graph with additional variables that model how our interventions may take effect.
-For example, in our drug treatment case we assumed we set the drug that each person receives.
-But what if some people in fact fail to take the prescribed treatment?
-How can we model this within the framework of deterministic interventions?
-We add a node representing what we prescribed to them (which we intervene on), which probabilistically influences what treatment they actually receive.
-See figure \ref{fig:randomized_imperfect_compliance}.
-Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could effect both the actual treatment taken and the outcome.
-
-The causal effect may provide useful bounds on the actual quantity of interest.
+One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}.
+However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions.
+For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
+But what if some people failed to take the prescribed treatment.
+We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
+Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
 
@@
 
-Another key issue with causal bayesian networks is that they cannot handle cyclic dependencies between variables.
-Such feedback loops are common in real systems, for example the relationship between supply and demand in economics or preditor and prey in ecology.
-We might regard the underlying causal mechanisms in these examples to be acyclic; the number of preditors at one point in time influences the number of prey in the next and so on.
-However, if our measurements of these variables must be aggregated over timeframes that are longer than the scale at which these interactions occure the result is a cyclic dependency.
+Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables.
+Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology.
+We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
+However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
 Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
-Such problems have mostly been studied within the dynamic systems liturature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
-\citep{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with strucural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as bayesian causal networks with additional functional assumptions.
+Such problems have mostly been studied within the dynamic systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
+\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
 
 \section{Counterfactuals}
-
-The Neyman-Rubin model \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}.
-Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else.
-
-In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would happen if they got the placebo, $\cf{y_{i}}{0}$.
+\label{sec:counterfactuals}
+
+The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
+Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
+Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
+
+In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would have happened if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would have happened if they got the placebo, $\cf{y_{i}}{0}$.
 The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment.
+The problem can be resolved if, instead of people, there are units that can be assumed to be identical or that will revert exactly to their initial state some time after treatment.
 This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
 
-Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.
+Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo.
 Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $\cf{Y}{1}$ is the distribution we would see of $Y$ if everyone was treated.
+The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated.
 Similarly $Y^{0}$ represents the potential outcome for the placebo.
-We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate, (from an experimental or observational study);
-
-\eq{
-\P{Y|X=1}& \text{, the probability that those who took the treatment will recover}\\
-\P{Y|X=0}& \text {, the probability that those who were not treated will recover}
-}
+The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
+We can estimate (from an experimental or observational study):
+\begin{itemize}
+\item $\P{Y|X=1}$, the probability that those who took the treatment will recover
+\item $\P{Y|X=0}$, the probability that those who were not treated will recover
+\end{itemize}
 
@@
 
-\eq{
-\implies \P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
-}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignore-ability assumptions \cite{Rosenbaum1983}.
-They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated.
+This implies the counterfactual distributions are equivelent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
+
+\eq{
+\P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
+}
+
+The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorebility assumptions \citep{Rosenbaum1983}.
+They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
 This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
 In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated highschool had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$.
+In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$.
 Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$.
-A more general form of the ignore-ability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
+A more general form of the ignorebility assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
 
 \vspace*{.3cm}
-\begin{theorem}[Ignore-ability] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
+\begin{theorem}[Ignorability] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
 
@@
 
-If we assume that within each socio-economic status level, attendence at pre-school is independent to the likelyhood of graduating high-school had you attended then we can compute the average rate of high-school graduation given a universal pre-school program from equation \ref{eqn:counterfactual1}.
+Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
 Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
 
-Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \cite{Rubin1978}.
+Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}.
 This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
 As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
-The later school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those did and who now share the classroom with them.
+The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
 SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
@@
 There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-This makes it quite easy to (accidentally) make statements with counterfactuals that can not be tested with empirical data.
+This makes it quite easy to (accidentally) make statements about counterfactuals that cannot be tested with empirical data.
 Consider the following example based on \citet{Dawid2000}.
-Again we have a drug, where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
+Again we have a drug where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
 Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal with equal variance (for simplicity).
@@
 \centering
-\caption{The distribution of individual treatment effects is not identifiable, even from a randomized controlled trial.}
+\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
 \label{fig:metaphysical_distribution_difference}
@@
 
-From a (large) randomized controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}.
-These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated retrospectively.
+From a (large) randomised controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}.
+These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated respectively.
 However, the distribution over the individual causal effects depends on $\rho$, see figure \ref{fig:metaphys_b}.
@@
 
-This result is unintuative.
+This result is unintuitive.
 It seems on the face of it that the distribution of individual causal effects is relevant to our decision making.
@@
 This confusion can be resolved by thinking about personalised rather than individual causal effects.
-It is entirely possible that potentialy observable characteristics (such as gender, age, genetics, etc) effect how people will respond to the treatment.
+It is entirely possible that potentially observable characteristics (such as gender, age, genetics, etc) affect how people will respond to the treatment.
 We can partition people into sub-populations on the basis of these characteristics and measure different \emph{personalised} causal effects for each group.
-The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much we stand to gain from further personalisation.
+The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much can be gained from further personalisation.
 The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).
@@
 One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than taking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
+Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
 This is certainly convenient, if rather imprecise.
-However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrents careful attention.
-We should always be clear what assumptions we are making and have in mind if it is possible (at least in theory) to validate those assumptions.
+However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention.
+It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).
 
@@
 Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.
+More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \citep{Pearl2000}.
 
@@
 The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable.
+If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent variable.
 Structural equation models can be represented visually as a network.
@@
 The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-Note that the causal graph in figure \ref{fig:causal_network_example2} is identical to that the bayesian network for the same problem, figure \ref{fig:causal_network_example}.
-The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as are captured by the probabilistic nature of the nodes in a Bayesian network, however adding them would not change the model.
+Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
+The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
 
@@
 The function $f_y$ and distribution over $\epsilon_y$ does not change.
-This results in the interventional distribution \footnote{We have assumed the variables are discrete only for notational convinience},
+This results in the interventional distribution \footnote{We have assumed the variables are discrete only for notational convenience},
 
@@
 The interventional distribution is the same as the observational one.
-The same arguament applies to the intervention $do(X=0)$ and so the causal effect is just the difference in observed outcomes as we found via the causal Bayesian network and counterfactual approaches.
+The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
 
@@
 }
-Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivelent to the observational distribution, given by;
+Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivalent to the observational distribution given by:
 
@@
 Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
-A structural equation model with $N$ variables looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
+A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
 However a SEM is, by definition, more than a set of simultaneous equations.
@@
 When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-The similarity between the notation used to describe and analyse structural equation models and simultaneuos equations, combined with a reluncatnce to make explicit statements about causality has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
+The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
 
@@
 
-It is possible to relax the independence of errors assumption we made for SEMs slightly to correspond exactly the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \cite{Richardson2013}.
-This correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
+It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}.
+The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
 For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores\cite{Rosenbaum1983} to estimate the effect.
+Then use propensity scores \citep{Rosenbaum1983} to estimate the effect.
 If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
 The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
-These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and legal discrimination \citep{Pearl2000}.
+These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
 
 In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
-The work on causal graphical models \cite{Pearl2000,Sprites} focuses on asymptopic, non-parametric estimation and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on our understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field.
-Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}.
+The work on causal graphical models \citep{Pearl2000,Sprites} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
+The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
+This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
+In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field.
+Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
 In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
@@
 
-We will now demonstrate how we can apply our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
+We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
 The following example is adapted from \citet{Pearl2000}.
@@
 
-It looks like drug A is performing better.
-However, having read the latest literature on how medications effect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
+Drug A appears to perform better.
+However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
 
@@
 
-Much to her surprise, once the data is broken down by gender, Treatment B looks better for both men \emph{and} women.
-Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must reccomend which to subsidise under a national health scheme).
+Once the data is broken down by gender, drug B looks better for both men \emph{and} women.
+Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must recommend which to subsidise under a national health scheme).
 Should she choose A or B?
@@
 The key is that the doctor is trying to choose between \emph{interventions}.
-She wants to know what the success rate will be if she changes her practise to give all the patients one drug, rather than allowing them to choose as currently occurs.
+She wants to know what the success rate will be if she changes her practice to give all the patients one drug, rather than allowing them to choose as currently occurs.
 
@@
 The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
-Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable us to estimate the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
-We require some assumptions about the causal relationships between the variables.
-In this example, it seems reasonable to conclude that gender may effect the treatment chosen and the outcome.
-If we assume there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
+Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
+Some assumptions about the causal relationships between the variables are required.
+In this example, it seems reasonable to conclude that gender may affect the treatment chosen and the outcome.
+Assuming there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
 
@@
 
-With this model, we see women are more likely to choose treatment $A$ and also more likely to recover regardless of what treatment they receive than men.
-Knowing a patient took drug A tells you they are much more likely to be female.
-When we compare the group of people who took A against those who took B, the greature proportion of females in the first group outweighs the higher benifit of drug B leading to an apparent reversal in effectiveness.
-However, when the doctor intervenes to set the treatment each person recieves there will no longer be a link from gender to treatment.
+With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
+Knowing a patient took drug A indicates they are more likely to be female.
+When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
+However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
 So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
@@
 
-Is the solution to Simson's paradox to always to break down the data by as many variables as possible?
+Is the solution to Simpson's paradox to always to break down the data by as many variables as possible?
 No.
-Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood preassure', 'M' with 'high' and 'F' with 'normal'.
-This is a drug desined to prevent heart desease.
-One pathway to doing so might well be to lower blood preassure.
+Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
+This is a drug designed to prevent heart disease.
+One pathway to doing so might well be to lower blood pressure.
 Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
 It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
-Now, however table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood preassure control and better overall outcomes.
-In this setting $\P{Y|do(T)} = \P{Y|T}$.
-Drug A is the better choice.
+Now, however, table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood pressure control and better overall outcomes.
+In this setting $\P{Y|do(T)} = \P{Y|T}$ and drug A is the better choice.
 
@@
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Blood Preassure};
+\node[node](1){Blood Pressure};
 \node[node, below left=of 1](2){Treatment};
@@
 Note that we have not changed the data itself, only the description of the variables that it is associated with.
-This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data but in what assumptions we are willing to make.
-From a purely statisitcal viewpoint there is no paradox.
+This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
+From a purely statistical viewpoint there is no paradox.
 The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
@@
 \centering
-\caption{Simpson's reversal visualized.
+\caption{Simpson's reversal visualised.
 The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
@@
 There are many other plausible causal graphs for both scenarios above.
-Perhaps income effects drug choice as well as gender, or gender could effect treatment choice and blood preassure control given treatment, etc.
-Causal modeling provides us with a powerful tool to specify such assumptions and determine how to estimate causal effects for a given model see section \ref{sec:causal effect estimation}.
+Perhaps income affects drug choice as well as gender, or gender might affect treatment choice and blood pressure control given treatment, etc.
+Causal modelling provides a powerful tool to specify such assumptions and to determine how to estimate causal effects for a given model see section \ref{sec:causal effect estimation}.
 
 \chapter{Two key questions}
+\label{chap:two_key_questions}
 
@@
 In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
-The goal is then to estimate the causal effect of some action or range of actions.
-WHERE DOES MEDIATION FIT IN?
-THIS IS ALREADY HUGE, and is central to 1000 of papers published each year.
-Causal discovery aims to leverage much broader assumptions to learn the structure of causal graph from data.
-THIS IS THE AUTOMATION OF SCIENCE.
+The goal is then to estimate the effect of an intervention or range of interventions in the system.
+Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}.
+Almost every time someone runs a regression model the key quantity of interest is a causal effect.
+Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
+
+Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
+This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
+Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
+An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
 
@@
 
-Causal effect estimation addresses the problem where we assume the key structure of the graph is known. That is, we assume that we have at a minimum:
+Estimating causal effects from observational data comes down to determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
+We did this on an ad-hoc basis to resolve the examples discussed in chapter \ref{chap:causal_models}.
+In this chapter we describe a principled approach to mapping observational quantities to interventional ones and discuss some of the key issues involved in estimating such expressions from finite sample data.
+We assume the basic structure of the graph is known.
+That is, we assume that we can draw a network containing (at a minimum):
 
 \begin{itemize}
-\item the target/outcome variable we care about
-\item the focus/treatment variables on which we are considering interventions
-\item any variables which act to confound two or more of the variables we have included.
+\item the target/outcome variable we care about,
+\item the focus/treatment variables on which we are considering interventions,
+\item any variables which act to confound two or more of the other variables we have included, and
+\item any links between variables we have included. 
 \end{itemize}
 
-Some of these variables my be latent or unobserved, in that we do not have measurements for them in the available data.
-However, their position in the network is assumed to be known.
-For example if we were interested in the effect of years of schooling on wages, we might expect that some measure of inherent ability could influence both the number of years of schooling people chose to pursue and the wages they subsequently obtain on graduating.
-Even if we have no data to directly measure peoples inherent ability, because it influences two of the variables were are modelling, we must include it in the graph.
-
-Causal effect estimation is one of the most frequently applied methodologies within applied sciences.
-It implicit in millions of studies across medicine, business, economics and social sciences.
-Almost every time someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
-
-\subsection{Mapping from observational to interventional distributions}
-Once we have drawn the causal network for our problem, how can we use that to map information from the observational setting to the interventional one.
-Recall that a bayesian network is a way of representing the joint distribution over its variables in a way which highlights the any conditional independencies between them.
-
+Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
+For example, consider estimating the impact of schooling on wages.
+Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
+Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
+
+How can the structure of the causal graph be leveraged to compute interventional distributions from observational ones?
+Given the graph corresponding to the observational distribution, the graph after any intervention can be obtained by removing any links into variables directly set by the intervention.
+The joint interventional distribution is the product of the factors associated with the interventional graph, as given by the truncated product formula \ref{eq:truncatedproduct}.
+If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
+However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
+
+The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
+By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
+This gives rise to the do-calculus \cite{Pearl2000}.
+The do calculus consists of three rules.
+They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.
+
+\subsection{Independence in Bayesian networks: D-separation}
+\label{sec:dsep}
+
+Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
+Therefor, in this section, we briefly review the key properties of Bayesian networks.
+A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
+Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
+
+\vspace{.3cm}
 \begin{theorem}
-\textbf{(Local Markov condition)} Given a bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
+\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
 \end{theorem}
@@
 
-By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \cite{Pearl2000}.
-The do calculus consists of three rules.
-They derive from the causal information encoded in a causal network and the properties of d-separation and do not require any addition assumptions other than that of specifying the causal network.
-
-\subsubsection{Rule 1}
-This rule describes which variables, on which we have not intervened, effect the distribution of the outcome given some intervention.
-The intervention $do(\boldsymbol{X}=\boldsymbol{x})$ changes a causal network, $G$, in a simple way.
-Variables in $\boldsymbol{X}$ are no longer determined by their parents but instead take on fixed values specified by $\boldsymbol{x}$.
-This corresponds to deleting the edges with arrows into variables in $\boldsymbol{X}$  (see figure \ref{fig:mutilatednet}).
-The resulting 'mutilated' network $G_{\overline{\boldsymbol{X}}}$ remains a causal network and d-separation still applies.
+The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
+From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
+Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
+In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
+In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
+
+The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
+The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
+As an example, consider a scholarship available to female or disadvantaged students.
+Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship.
+There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
+However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
+\begin{figure}[h]
+\centering
+\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
+\label{fig:dseptriples}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y| Z$}
+\label{fig:dseptripple1}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$X$};
+\node[main node, below =of 1](2){$Z$};
+\node[main node, below =of 2](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    (2) edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|Z$}
+\label{fig:dseptripple2}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$Y$};
+\node[main node, below =of 1](2){$Z$};
+\node[main node, below =of 2](3){$X$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    (2) edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|Z$}
+\label{fig:dseptripple3}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$Z$};
+\node[main node, below left =of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+        edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|\{\}$}
+\label{fig:dseptripple4}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$Z$};
+\node[main node, above left =of 1](2){$X$};
+\node[main node, above right =of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (1)
+    (3) edge node {} (1);	
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
+
+\vspace{.3cm}
+\begin{definition}[unblocked path]
+\label{def:unblocked}
+A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
+It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
+\end{definition}
+
+\vspace{.3cm}
+\begin{definition}[d-separation]
+\label{def:dsep}
+The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, there are no unblocked paths from any $X \in \boldsymbol{X}$ to any $Y \in \boldsymbol{Y}$ after conditioning on $\boldsymbol{Z}$.
+\end{definition}
+
+\vspace{.3cm}
+\begin{theorem}[d-separation and conditional independence]
+\label{thrm:dsep_independence}
+If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in a Bayesian network $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
+Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected (not d-separated) given $\boldsymbol{Z}$ then it is possible to construct a distribution $P'$ that factorises over $G$ in which they are dependent.
+\end{theorem}
+
+Theorem \ref{thrm:dsep_independence} says that independencies implied by d-separation on a graph hold in every distribution that can be factored over that graph and that if ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in \emph{all} distributions that can be factored over $G$ then they are d-separated in $G$.
+If we denote the independencies implied by d-separation in a graph by $\indep{G}$ and the set of independencies in a distribution by $\indep{P}$ then $\indep{G} \subseteq \indep{P}$.
+
+If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
+However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
+A particular case in which this occurs is when there are deterministic relationships between variables.
+If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
+This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
+This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
+D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
+
+\section{The Do Calculus} 
+\label{sec:do_calculus}
+
+The do-calculus is a set of three rules \citep{Pearl1995} that can be applied to simplify the expression for an interventional distribution.
+If by repeated application of the do-calculus, along with standard probability transformations, we can obtain an expression containing only observational quantities then we can use it to estimate the interventional distribution from observational data.
+Let $\boldsymbol{X}$,$\boldsymbol{Y}$,$\boldsymbol{Z}$ and $\boldsymbol{W}$ be disjoint sets of variables in a causal graph $G$.
+We denote the graph $G$ after the an intervention $do(\boldsymbol{X})$, which has the effect of removing all edges into variables in $\boldsymbol{X}$, as $G_{\overline{\boldsymbol{X}}}$
+
+\subsubsection{Rule 1: (adding or removing evidence)}
+
+Rule 1 allows us to remove (or insert) observational evidence from the right hand side of a conditional interventional distribution.
+It follows directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still applies after an intervention.
 
 If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\begin{equation}
+\eqn{
 \label{eq:Do1}
  P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
-\end{equation}
-
-$\implies$ if $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\begin{equation}
-\label{eq:Do12}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) 
-\end{equation}
-
-This stems directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still holds in the mutilated network.
+}
 
@@
 \centering
-\caption{Intervention in a causal bayesian network}
+\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
 \label{fig:mutilatednet}
@@
 
-\subsubsection{Rule 2}
-Rule 2 states when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening  $do(\boldsymbol{X}=\boldsymbol{x})$ have the same effect on the distribution of the outcome $\boldsymbol{Y}$.
-You can think of this as when correlation is causation.
-It is easiest to understand by explicitly including the intervention process in the graphical model.
-We can depict the possibility of intervention by adding a new decision node $\hat{X}$ as a parent of each $X$ in the set of nodes we are intervening on (figure \ref{fig:augmentedNetwork}).
-Let $\epsilon$ be some arbitrary value not in the set of possible values for $X$.
-If $X = \epsilon$ the distribution of $X$ is what it was without intervention.
-Otherwise, if $X=x$, $X$ deterministically takes the value $x$ and is independent of its previous parents, representing the intervention $do(X=x)$.
-We use the notation $G^{\dagger}$ to represent $G$ augmented with these decision nodes.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X},\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
-
-\begin{equation}
+\subsubsection{Rule 2: (exchanging actions with observations)}
+Rule 2 describes when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening, $do(\boldsymbol{X}=\boldsymbol{x})$, have the same effect on the distribution over $\boldsymbol{Y}$.
+Let $G_{\underline{\boldsymbol{X}}}$ denote the causal graph $G$ with all edges \emph{leaving} $X$ removed.
+
+If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W})$ in $G_{\underline{\boldsymbol{X}}}$:
+\eqn{
+\label{eq:Do22}
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W})
+}
+
+The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$.
+Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
+However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
+Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
+If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths.
+This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).
+
+Equation \ref{eq:Do22} does not cover cases where acting on one set of variables allows us to replace acting on another set with conditioning (see figure \ref{fig:rule2_2}).
+The general form of rule 2 is given in equation \ref{eq:Do2}.
+
+If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W},\boldsymbol{Z})$ in $G_{\underline{\boldsymbol{X}}\overline{\boldsymbol{Z}}}$:
+\eqn{
 \label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W}=\boldsymbol{w})
-\end{equation}
-
-$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}}|\boldsymbol{X})$ in $G^{\dagger}$:
-\begin{equation}
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x})
-\end{equation}
-
-If the outcome does not depend on how the decision to assign the interventional variables was made, then the interventional distribution equals the observational one.
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),
+do(\boldsymbol{Z}=\boldsymbol{z}),
+\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W})
+}
 
@@
 \centering
-\caption{Augmented causal network with intervention}
-\label{fig:augmentedNetwork}
-\begin{subfigure}[t]{0.4\textwidth}
+\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
+In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
+If we condition on $W$, we block this indirect path.}
+\label{fig:rule2}
+\begin{subfigure}[t]{0.29\textwidth}
 \centering
 \caption{original network, $G$}
-\label{fig:augmentedOriginal}
+\label{fig:rule2_orig}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={punkt}]
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$W$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)  edge  (3)
+    (2) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\overline{X}}$}
+\label{fig:rule2_after}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$W$};
+\node[main node, below left=of 1](2){$x$};
+\node[main node, below right=of 1](3){$Y$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1)	edge (3)
+    (2) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\underline{X}}$}
+\label{fig:rule2_indirect}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$W$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)
+    	edge  (3);
+
+\end{tikzpicture} 
+\end{subfigure}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\caption{An example of applying equation \ref{eq:Do2}.
+In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
+Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
+Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
+}
+\label{fig:rule2_2}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:rule22_orig}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
  %nodes
@@
 \node[main node, below right=of 1](3){$Y$};
-\node[main node, above left = of 1](4){$U_{1}$};
-\node[main node, above right = of 1](5){$U_{2}$};
-
+\node[main node, above left=of 1](4){$U$};
+\node[main node, above right=of 1](5){$V$};
  \path[every node/.style={font=\sffamily\small}]
- 	(2) edge node {} (3)
- 	(4) edge node {} (2)
- 		edge node {} (1)
- 	(5) edge node {} (3)
- 		edge node {} (1)
- 	(1) edge node {} (2)
- 		edge node {} (3);
-
+    (1) edge  (2)
+    (2) edge (3)
+    (4) edge (1) edge (2)
+    (5) edge (1) edge (3);
 \end{tikzpicture} 
 \end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-\centering
-\caption{augmented network after the intervention $do(Z=z)$, $G^{\dagger}_{\overline{Z}}$}
-\label{fig:augmentedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.3cm,
-  thick,main node/.style={punkt}]
-
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{after $do(Z)$, $G_{\overline{Z}}$}
+\label{fig:rule22_after}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
 \node[main node](1){$z$};
-\node[below left=of 1](2){};
-\node[below right=of 1](3){};
-\node[main node, below left=of 2](4){$X$};
-\node[main node, below right=of 3](5){$Y$};
-\node[above right=of 1](6){};
-\node[above left=of 1](7){};
-\node[main node, above left=of 7](8){$U_{1}$};
-\node[main node, above right=of 6](9){$U_{2}$};
-\node[main node, above left=of 4](10){$\hat{X}$};
-
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 1](4){$U$};
+\node[main node, above right=of 1](5){$V$};
  \path[every node/.style={font=\sffamily\small}]
- 	(4) edge node {} (5)
- 	(10) edge node {} (4)
- 	(8) edge node {} (4)
- 	(9) edge node {} (5)
- 	(1) edge node {} (4)
- 		edge node {} (5);
-
+    (1) edge  (2)
+    (2) edge (3)
+    (4) edge (2)
+    (5) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\underline{X}\overline{Z}}$}
+\label{fig:rule22_indirect}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 1](4){$U$};
+\node[main node, above right=of 1](5){$V$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)
+    (4) edge (2)
+    (5) edge (3);
 \end{tikzpicture} 
@@
 
-\subsubsection{Rule 3}
+\subsubsection{Rule 3: (adding or removing actions)}
 This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\dagger}_{\boldsymbol{\overline{Z}}}$:
-\begin{equation}
+A simple case of rule 3 is given in equation \ref{eq:Do3_1}.
+If $\boldsymbol{Y}$ is independent of $\boldsymbol{{X}}$ in $G$ after removing links entering $\boldsymbol{X}$ then can be no direct path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and any intervention on $\boldsymbol{X}$ will not affect $\boldsymbol{Y}$.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{{X}})$ in $G_{\boldsymbol{\overline{X}}}$:
+
+\eqn{
+\label{eq:Do3_1}
+\P{\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})} = \P{\boldsymbol{Y}}
+}
+
+The general case of rule 3 is easier to state by explicitly representing the intervention in the graphical model.
+Let $G^{\boldsymbol{\hat{X}}}$ denote the graph $G$ after adding a variable $\hat{X_i}$ as a parent of each variable $X_i \in \boldsymbol{X}$ (see figure \ref{fig:rule3_augmented}).
+The variable $\hat{X_i}$ can be thought of as representing the mechanism by which $X_i$ takes its value, either by being set via intervention or as a stochastic function of its other parents \citep{Koller2009}.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\boldsymbol{\hat{X}}}_{\boldsymbol{\overline{Z}}}$:
+\eqn{
 \label{eq:Do3}
 P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-\end{equation}
-$\implies$ if  $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}})$ in $G^{\dagger}$:
-\begin{equation}
-\label{eq:Do32}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})) = P(\boldsymbol{Y})
-\end{equation}
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ without conditioning on $\boldsymbol{X}$ implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ which starts on an arrow leaving $\boldsymbol{X}$.
-This means there is no causal effect of $\boldsymbol{X}$ on $\boldsymbol{Y}$ and the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ does not alter the distribution of $P(\boldsymbol{Y})$.
+}
+
+The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
+These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
+
+\begin{figure}[h]
+\centering
+\caption{Example application of equation \ref{eq:Do3}.
+$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
+We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
+\label{fig:rule3}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:rule3_original}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, below right=of 2](4){$W$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (4)
+    (4) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{augmented graph $G^{\hat{X}}$}
+\label{fig:rule3_augmented}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, below right=of 2](4){$W$};
+\node[main node, above left=of 2](5){$\hat{X}$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (4)
+    (4) edge (3)
+    (5) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\overline{X}}$}
+\label{fig:rule3_after}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, below right=of 2](4){$W$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (3)
+    (2) edge (4)
+    (4) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\end{figure}
 
 \subsection{Identifiability}
-A natural question to ask is, given the level of assumptions we have made about the graph, is it possible to estimate a causal effect from observational data?
+A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
 This is the identifiablity problem.
-Identifiability is an asymptotic property of the graph.
-It asks if we can obtain an unbiased point estimate for the causal effect of interest in the infinite data limit.
-A problem is non-parametrically identifiable if it is identifiable without any additional assumptions on the functional form of the dependencies between variables in the graph.
-
-The question of non-parametric identifiability is solved!
-The do calculus is complete.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do calculus.
-
-There is an algorithm \cite{Shpitser2012} based on these properties that, for a given network and interventional (do-type) query, can:
+It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
+A query is non-parametrically identifiable if it is identifiable without  assumptions about the functional form of the dependencies between variables in the graph.
+
+\vspace*{.3cm}
+\begin{definition}[Non-parametric identifiablity]
+Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
+A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
+\end{definition}
+
+The question of non-parametric identifiability is solved.
+The do calculus is complete \cite{Shpitser2006b,Huang2006}.
+A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
+There is a polynomial time algorithm \cite{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
 
@@
 \item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network
-\item if it is identifiable, return the required expression 
+In other words, determine if the query is identifiable given the assumptions encoded by the network, and
+\item if it is identifiable, return the required expression. 
 \end{enumerate}
 
-\todo{insert link to working javascript version}
-\todo{figure with typical examples}
-
-There are many interesting questions relating to identifiability that remain open.
+Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
+I have created a javascript implementation of the identifiability algorithm \cite{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
+
+\begin{figure}[h]
+\centering
+\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
+\label{fig:identify}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{}
+\label{fig:id1}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3)
+    (4) edge (1) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{}
+\label{fig:id2}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$X$};
+\node[ob, right=of 1](2){$Z$};
+\node[ob, right=of 2](3){$Y$};
+\node[lt,above=of 2](4){$U$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2)
+    (2) edge (3)
+    (4) edge (1) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{}
+\label{fig:id3}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U_3$};
+\node[lt,above=of 2](5){$U_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (4) edge (1) edge (3)
+    (5) edge (2) edge (1);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\vspace*{.5cm}
+\caption{}
+\label{fig:nid1}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[lt](1){$U$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\vspace*{.5cm}
+\caption{}
+\label{fig:nid2_instrumental_variable}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[lt](1){$U$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[ob, above left=of 2](4){$Z$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3)
+    (4) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\vspace*{.5cm}
+\caption{}
+\label{fig:nid3}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U_3$};
+\node[lt,above=of 2](5){$U_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (4) edge (1) edge (3)
+    (5) edge (2) edge (1)
+    (1) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\end{figure}
+
+Many interesting questions relating to identifiability remain open.
 What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
 What if we assume various restrictions on the functional form of the relationships between the variables?
-Clearly  .
-A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work going back to at least ???
-\ref{}
-
-Is identifiablity a good question to ask?
-It seems natural to see if a problem can be solved with infinite data before we try to see how well we can do it with finite data.
-However, we should not give up on non-identifiable problems.
-We may still be able to achieve useful bounds on causal effects with finite data even if point estimation is not possible.
-From that perspective we should be cautious about dividing problems into identifiable and not identifiable.
-
-How do we tackle a problem that is not identifiable. We could look for bounds or we could make additional assumptions (or both).
-
-\subsection{Estimation}
-\subsubsection{Defining causal effects} 
+Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
+A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
+
+Although identifiability is a natural and important question to ask, it does not partition causal questions into solvable and unsolvable.
+Estimators for identifiable queries can be slow to converge and we may be able to obtain useful bounds on causal effects in cases where point estimates are not identified.
+
+\section{Estimation}
+\subsection{Defining causal effects} 
 So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
-
-From a decision making process, is it the right thing to do to just pick the X that maximises the expectation of $Y$ (assuming that higher $Y$ is better.
-If it is not we can just add in another variable that is a deterministic function of $Y$ to relect the loss function.)
-
-Typically however we are often interested in estimating some function of the difference between $P{Y|do(X)}$ for various values of $X$.
-For example, the average causal effect (ATE) is defined for discrete binary treatment $T$ by:
-
-\eqn{
-ATE = \E{Y|do(T = 1)} - \E{Y|do(T=0)}
-}
-
-If the variables are continuous the average casual effect is
-
-\todo{insert equation for continuous case}
-
-The individual causal effect as defined in section XXX cannot be stated in terms of the do notation as it is not with respect to a distribution over anything.
-However, we can define a very similar concept with respect some observed context which I will refer to as the personalised causal effect.
-
-\todo{insert equation for personalised causal effect}
-
-However, the difference in expectation is not the only way of summarising the difference between two (or more) distributions.
-\citep{Janzing2013} show that there is no single way of summarising the collection of distributions represented by $\P{Y|do(X)}$ that satisfies a number of reasonable axioms.
-
-An obvious alternative to subtracting the exceptions would be to consider ratio.
-Is there any reason that we don't do this that relates to the argument in Dawid?
-What about the smoothness of any uncertainties?
-
-\subsubsection{Estimating the difference when the back door criteria is met}
-
-A very substantial number of practical studies fall into the category of causal inference via the backdoor theorem.
-A study is done.
-It is assumed that it is ignore-able.
-That is that we have a a set of variables satisfying the back door adjustment criteria.
-The goal is to estimate the average causal effect.
-Since we have assumed the backdoor criteria is met, the query is identifiable.
-Problem solved.
-Except we have finite data ...
-
-What are the approaches for doing so?
-
-There are multiple ways of doing the adjustment.
-
-In order to block the path of spurious information it is sufficient to (perfectly) model either the treatment selection process $\P{T|Z}$ or the response function $\P{Y|T,Z}$.
-
-\paragraph*{matching}
-
-\paragraph*{regression}
-\paragraph*{When is regression causal?}
-A regression equation is causal if, when we represent the equations as a bayesian causal model or structural equation model, we claim the arrows are causal AND if the variables we include in the regression form a valid back door adjustment set.
-
-\paragraph*{propensity scores}
-
-They have advantages and disadvantages depending on the setting. Methods like regression typically impose a parametric model of some description onto the treatment assignment or response function.
-
-- Connection of nearest matching to nearest neighbour.
-
-Once we have made the assumptions required to define the causal graph and identify a valid adjustment set, is there any difference between the causal effect estimation problem and standard supervised machine learning.
-
-\cite{causal estimation is covariate shift}
-
-However, there is a critical remaining complication.
-In supervised learning we can split our data into training and validation data sets or apply cross validation to select models and model parameters.
-In the causal estimation case we cannot.
-We lack samples from the counterfactual.
-
-The significance to this should not be underestimated.
-Cross-validation has allowed applied machine learning to proceed with very a theoretical approach on the basis that we can identify when a model is successful.
-With causal effect estimation there is no guarantee that a model which performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
-
-GIVE A CONCRETE EXAMPLE here.
-One would be a model which looks at a post-treatment variable in order to make predictions.
-But what about if we assume the model structure is correct and known.
-Can we fit parameters that do better on prediction but worse on casual effect estimation?
-
-\cite{Imbens2004} Review of non-parametric estimation
-
-Discuss the recent competition on this question and the results
-
-Discuss the role that the models used to simulate the data may have in determining which models perform the best.
-
-\subsection{Non identifiable queries}
-In many cases the query of interest will not be identifiable.
-In this setting we are left with a choice between making additional assumptions (for example that variables are linearly related) or aiming to obtain bounds rather than point estimates for the causal effect of interest.
-Or we may do both.
-One of the most widely used methods in this category is the instrumental variable.
-
-Instrumental variables are often conceptualised as natural experiments.
-The goal is to find a variable that substantially influences the treatment but that is conditionally independent of the remaining variables in the network.
-
-For example, we might want to compare the results for juvenile offenders depending on whether they spent their time awaiting trial in a custom facility for young offenders as opposed to a unit within an adult jail.
-The process by which children are assigned to different facilities may well be non-random.
-However, it might be that in many cases the decision was based purely on the fact that the juvenile facility was full.
-We could then use the number of places remaining in the juvenile facility as an instrumental variable.
-
-The causal network for an ideal instrumental variable is shown in figure XXX.
-It is not identifiable (non-parametrically) because.
-If we assume linearity it ?
-\todo{what exact guarantees do we get for instrumental variables}.
-Without a linearity assumption we would want the instrumental variable to be highly predictive of the treatment assignment so as to obtain reasonable bounds.
-But if we make the assumption that everything is linear then it would seem that even a weak instrumental variable should (in theory) suffice.
-
-In some cases we might have a choice between adjusting for causal effects via the backdoor theorem or utilising an instrumental variable to obtain bounds.
-It is not the case that the former method is always superior even though it has better asymptotic results.
-For example
-
-DEMONSTRATE a specific example where instrumental variables obtain better estimates than adjustment even when all the criteria for adjustment have been met.
-Start with a case were the adjustment is very high dimensional and the instrumental variable is strong.
-(the effect of the variables on the treatment assignment will have to be week (otherwise the instrument is not strong).
-
-\subsection{How successful overall is causal effect estimation}
-Really know one knows.
-
-There are relatively few studies that have compared the observational study results with those obtained by randomised experiments.
-One of the key examples is the XX job network study.
-Initial analysis suggested that a range of causal estimation techniques all ...
-
-\paragraph*{How can we construct data sets on which to test casual effect estimation techniques}
-One can start from a randomised experiment and deliberately drop some of the data. Draw up the graphical model for this process and the assumptions inherent in it.
+This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set.
+From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.
+
+\eqn{
+\label{eqn:utility_max}
+x* =\argmax_x \Esub{y\sim\P{Y|do(X=x)}}{{\cal{U}}(y)}
+}
+
+Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
+There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
+For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
+
+\eqn{
+\label{eqn:ACE}
+ACE = \E{Y|do(X = 1)} - \E{Y|do(X=0)}
+}
+
+Assuming the expectations in equation \ref{eqn:ACE} are well defined, the ACE captures the shift in the mean outcome that arises from varying $X$.
+It does not capture changes in variance or higher moments of the distribution.
+The ACE can be generalised to non-discrete interventions by considering the effect on the expectation of $Y$ of an infinitesimal change in $x$.
+If $X$ is linearly related to $Y$ then the ACE is constant and equivalent to the corresponding coefficient in the linear structural equation model.
+
+\eqn{
+\label{eqn:ACE_continuous}
+ACE(x) = \frac{d}{dx}\E{Y|do(X=x)}
+}
+
+The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}.
+Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
+However we can personalise the average causal effect by stating it with respect to some observed context.
+I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
+
+\eqn{
+PCE(z) = \E{Y|do(X=1),z} - \E{Y|do(X=0),z}
+}
+
+In some cases we may be interested in the average causal effect for some sub-group of the population.
+A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
+This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.
+
+\eqn{
+ATT = \Esub{z\sim \P{Z|x=1}}{Y|do(X = 1)} - \Esub{z\sim \P{Z|x=1}}{Y|do(X=0)}
+}
+
+Causal effects can also be written in terms of counterfactuals.
+The ACE is $\E{\cf{Y}{1} - \cf{Y}{0}}$.
+We could estimate the ratio of expectations $\frac{\E{\cf{Y}{1}}}{\E{\cf{Y}{0}}}$ instead of the difference.
+However, the quantity   $\E{\frac{\cf{Y}{1}}{\cf{Y}{0}}}$ depends on the joint distribution over the counterfactual variables $(\cf{Y}{1},\cf{Y}{0})$ and thus cannot be computed from the interventional distribution.
+
+Another way of conceptualising causal effects is as a property indicating the strength of the causal link between two variables.
+This notion is complex to formalise when the relationship between variables is non-linear.
+Suppose $Y = X \oplus Z$ with $P(Z=1)=\frac{1}{2}$, the interventional distributions over $X$ are identical after marginalising out $Z$.
+\citet{Janzing2013} propose a number of postulates that a notion of causal strength could satisfy, demonstrate why previous measures fail these postulates and propose an alternative based on information flow.
+
+\subsection{Estimating causal effects by adjusting for confounding variables}
+\label{sec:causal_effect_under_ignorability}
+Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors.
+Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds.
+Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds.
+This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}.
+The setting is also referred to as unconfounded.
+
+\vspace*{0.3cm}
+\begin{theorem}[The backdoor criterion]\citep{Pearl2000}
+Let $\boldsymbol{X}$, $\boldsymbol{Z}$ and $\boldsymbol{Y}$ be disjoint sets of vertices in a causal graph $G$.
+If $\boldsymbol{Z}$ blocks (see Definition \ref{def:unblocked}) for every path from $X_i$ to $Y_j$ that contains a link into $X_i$, for every pair $(X_i \in \boldsymbol{X},Y_j \in \boldsymbol{Y})$, and no node in $\boldsymbol{Z}$ is a decedent of a node in $\boldsymbol{X}$ then the backdoor criterion is satisfied and;
+
+\eqn{
+\label{eqn:backdoor}
+\P{\boldsymbol{y}|do(\boldsymbol{x})} =\sum_{\boldsymbol{z}} \P{\boldsymbol{y}|\boldsymbol{x},\boldsymbol{z}}\P{\boldsymbol{z}}
+}
+\end{theorem}
+
+The backdoor criterion derives from rule 2 of the do-calculus.
+Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
+There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
+The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
+Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}.
+Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
+For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
+
+\begin{figure}
+\centering
+\caption{}
+\label{fig:adjustment}
+\begin{subfigure}[t]{0.48\textwidth}
+\centering
+\caption{There can be multiple valid adjustment sets. $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.}
+\label{fig:adjust_multiple}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](2){$X$};
+\node[ob, right=of 2](3){$Y$};
+\node[ob,above=of 3](4){$Z_2$};
+\node[ob,above=of 2](5){$Z_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (5) edge (2) edge (4)
+    (4) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.48\textwidth}
+\centering
+\caption{Conditioning on $Z$ opens the backdoor path $X-U_1-Z-U_2-Y$}
+\label{fig:m-graph}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U_3$};
+\node[lt,above=of 2](5){$U_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (4) edge (1) edge (3)
+    (5) edge (2) edge (1);
+\end{tikzpicture} 
+\end{subfigure}
+\end{figure}
+
+Given that a set of variables $\boldsymbol{Z}$ satisfies the backdoor criterion (or equivalently the conditional ignorability assumption), the interventional distribution is asymptotically identifiable and can be estimated from equation \ref{eqn:backdoor}.
+The expected value of $Y$ after the intervention $do(X=x)$ is given by equation \ref{eqn:expected_post_intervention} and the average causal effect for a binary intervention $x \in \set{0,1}$ is given by equation \ref{eqn:ace_ignoreable}.
+
+\eqn{
+\label{eqn:expected_post_intervention}
+\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}}
+}
+
+\eqn{
+\label{eqn:ace_ignoreable}
+ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
+}
+
+Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata.
+However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
+There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
+The key approaches are based on matching on covariates, propensity score methods and regression.
+We now examine these approaches from a machine learning perceptive.
+
+In standard supervised learning, we have a training set $(\vb{x_1},y_1),...,(\vb{x_n},y_n)$ assumed to be sampled i.i.d from an unknown distribution $\P{\vb{x},y} =\P{\vb{x}}\P{y|\vb{x}}$.
+The goal is to select a hypothesis $h \in \mathcal{H} : \mathcal{X} \rightarrow \mathcal{Y}$ such that, on unseen data $ \sim \P{\vb{x},y}$, $h(\vb{x})$ is close (by some metric) in expectation to $y$.
+In other words we wish to minimise the generalisation error $E_{out}(h)$,
+
+\eqn{
+E_{out}(h) = \Esub{(\vb{x},y)\sim \P{\vb{x}}\P{y|\vb{x}}}{L(h(\vb{x}),y)}
+}
+
+We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
+We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
+
+\eqn{
+E_{in}(h) = \frac{1}{n}\sum_{i=1}^n L(h(\vb{x}_i),y_i)
+}
+
+The VC-dimension of the hypothesis space provides (typically loose) bounds on the probability that $E_{out} >> E_{in}$.
+However, in practice, the generalisation error is usually estimated empirically from a hold-out set of the sample that was not used to train the model, or via cross-validation.
+
+In the causal effect estimation under ignorability, we have training data $(\vb{x}_1,\vb{z}_1,y_1),...,(\vb{x}_n,\vb{z}_n,y_n)$ sampled i.i.d from $\P{\vb{z}}\P{\vb{x}|\vb{z}}\P{y|\vb{x},\vb{z}}$.
+Estimating $\E{Y|do(\vb{X}=\vb{x})}$ corresponds to selecting a hypothesis $h \in \mathcal{H}:\mathcal{X} \times \mathcal{Z} \rightarrow \mathcal{Y}$ that minimises;
+
+\eqn{
+\label{eqn:causal_covariate_shift}
+E_{out} &= \Esub{(\vb{x},\vb{z},y)\sim \delta(\vb{x}-\vb{x'})\P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},\\
+&= \Esub{(\vb{z},y)\sim \P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},
+}
+
+\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
+If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
+\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
+Take the case where we have a binary intervention $x \in \set{0,1}$.
+Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$.
+More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.
+
+\subsubsection{Regression}
+
+The regression approach is to learn a function that is a good approximation to the output surface $\E{Y|X,Z}$.
+Let $f_1(z)= \E{Y|X=1,Z=z}$.
+The expectation of $Y$ after the intervention $X=1$ is then obtained by taking the expectation with respect to $Z$, $\E{Y|do(X=1)} = \Esub{z \sim \P{Z}}{\E{Y|X=1,z}}$.
+We can learn a parametric regression model $\hat{f}_1(z)$ via empirical risk minimisation.
+
+\eqn{
+\hat{f}_1(z) = h_1(z;\hat{\theta}_{obs}) \text{, where } \hat{\theta}_{obs} = \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)  \right]
+}
+
+This estimator is consistent with respect to the observational distribution.
+As the sample size tends to infinity, $\hat{\theta}_{obs}$ approaches the parameter within the hypothesis space that minimises the expected loss given data sampled from the observational distribution.
+
+\eqn{
+\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z|x=1}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
+}
+
+If the model is correctly specified such that $f_1(z) = h_1(z;\theta^*)$ for some  $\theta^* \in \Theta$ then the empirical risk minimisation estimate is consistent with respect to the loss over any distribution of $Z$  \citep{Sugiyama2007}, including the interventional one.
+
+\eqn{
+\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \theta^* = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
+}
+
+The average causal effect can then be estimated by,
+
+\eqn{
+\label{eqn:cef_empirical_regression}
+\hat{\tau}_{reg} = \sum_{i=1}^n \left(\hat{f}_1(z_i) - \hat{f}_0(z_i)\right)
+}
+
+Regression thus has a direct causal interpretation if the parametric model is correctly specified and the covariates included form a valid backdoor adjustment set for the treatment variable of interest in the corresponding structural equation model.
+
+\subsubsection{Propensity scores}
+If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
+The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
+
+\begin{figure}[h]
+\centering
+\caption{
+Parametric regression may yeild poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
+In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
+If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$.
+We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points).
+In both cases the linear model is a good fit to the data.
+However, the resulting estimate of the causal effect is very poor for the lower values of $z$.
+}
+\label{fig:covariate_shift_adjust}
+\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
+\end{figure}
+
+A general approach to estimating the expectation of some function $f(\cdot)$ with respect to data from some distribution $\P{\cdot}$, when we have data sampled from a different distribution $\Q{\cdot}$ is importance sampling \citep{horvitz1952generalization,Koller2009}.
+\eqn{
+\label{eqn:importance_sampling}
+\Esub{\vb{v} \sim \P{\vb{v}}}{f(v)} = \Esub{\vb{v} \sim \Q{\vb{v}}}{f(v)\frac{\P{\vb{v}}}{\Q{\vb{v}}}}
+}
+
+This importance weighting approach can be applied to the covariate shift/average causal effect problem by weighting the terms in the empirical risk minimisation estimator \citep{Sugiyama2007}.
+
+\eqn{
+\label{eqn:importance_weighted_erm}
+\hat{\theta}_{iw} &= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{\P{z_i}\delta(x_i-1)}{\P{z_i}\P{x_i=1|z_i}}  \right] \\
+&= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{1}{e(z_i)}  \right],
+}
+
+where $e(\vb{z})$ is the propensity score, defined by \citep{Rosenbaum1983};
+\eqn{
+e(\vb{z})\equiv \P{x=1|\vb{z}}
+}
+
+The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
+Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
+
+The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$.
+\citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$.
+This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$.
+Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
+In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
+However in general, it must be estimated from data.
+Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and nueral networks can also be applied   \citep{Austin2011}.
+\citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
+
+\eqn{
+\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}} = \Esub{z\sim \P{\vb{Z|x}}}{\E{Y|x,\vb{z}}\frac{1}{e(\vb{z})}}
+}
+
+\eqn{
+\label{eqn:simple_inverse_propensity}
+\hat{\tau}_{ip} = \frac{1}{n}\sum_{i=1}^n \left(\frac{\ind{x_i=1}y_i}{e(\vb{z}_i)} -\frac{\ind{x_i=0}y_i}{1-e(\vb{z}_i)}\right)
+}
+
+\subsubsection{Matching}
+There is a straightforward connection between matching and regression for causal effect estimation.
+If $h \in \mathcal{H} \implies h + a \in \mathcal{H}$ for any constant $a$ and $\hat{f}$ is selected by minimising empirical risk with an $L_2$ loss then $\frac{1}{n}\sum_{i=1}^n \ind{x_i=1}\hat{f}_1(z_i) = \frac{1}{n}\sum_{i=1}^n \ind{x_i=1}y_i$ \footnote{\citep{Imbens2004} state this holds for most implementations}, and equation \ref{eqn:cef_empirical_regression} can be re-written as:
+
+\eqn{
+\label{eqn:cef_regression_missing}
+\hat{\tau}_{reg} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \hat{f}_0(z_i)\right) +\ind{x_i=0}\left(\hat{f}_1(z_i) - y_i \right)\right]
+}
+
+This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
+For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
+Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
+\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
+
+\eqn{
+\hat{\tau}_{match} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \frac{1}{k}\sum_{j \in J_k(i)}y_j\right) +\ind{x_i=0}\left(\frac{1}{k}\sum_{j \in J_k(i)}y_j - y_i \right)\right]
+}
+
+This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
+There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{cochran1973controlling,rosenbaum1985constructing}.
+Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
+An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.
+
+The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
+A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
+
+The significance of this should not be underestimated.
+Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
+With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
+\citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem.
+There is relatively little theory on model selection for estimating the propensity score.
+To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion.
+It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}.
+With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
+I have found no examples of such an approach.
+
+The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding exeriments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
+Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
+The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
 
@@
 
-Cyclic models \cite{Hyttinen2012}
+%Cyclic models \cite{Hyttinen2012}
 
@@
 
-There is also work on inferring causal structure from experiential data \ref{}. We discuss this further in section XXX.
+%There is also work on inferring causal structure from experiential data \ref{}. We discuss this further in section XXX.
 
@@
 Unfortunately, causal discovery is very hard.
-Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows XXX exponentially (or super-exponentially)?
-with the number of variables.
-- some intuition here about the strength of the signal we get from causal discovery.
+Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows exponentially with the square of the number of variables .
+
+% - some intuition here about the strength of the signal we get from causal discovery.
 Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}.
@@
 We roughly divide the methods into those based on those that exploit the connection between the conditional independencies in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
-This is not the only way we could have grouped them.
+
+%This is not the only way we could have grouped them.
 There is also a separation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
@@
 One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
-For any bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network.
+For any Bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network.
 If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$.
@@
 This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
-This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
-Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \cite{Kollner} for $P$.
+This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
+Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \cite{pearl1988probabilistic} for $P$.
 
@@
 It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
-However, most distributions generated by a causal bayesian network will be faithful to that network.
+However, most distributions generated by a causal Bayesian network will be faithful to that network.
 For faithfulness to be violated, different causal effects must exactly balance one-another out.
@@
 It would disappear under a small perturbation to any of the parameters.
-\todo{point out violations are solutions to polynomial equations and thus measure 0 \ref{XX}}
-
-Given the faithfulness assumption, our causal discovery problem reduces to finding the set of bayesian networks that have exactly the dependency structure as we observe in $P$.
+In discrete systems, violations correspond to the solutions to polynomial equations over values in the CPD tables and thus are a space of measure zero with respect to all possible distributions associated with the graph \citep{Koller2009}.
+
+Given the faithfulness assumption, our causal discovery problem reduces to finding the set of Bayesian networks that have exactly the dependency structure as we observe in $P$.
 This set can also be referred to as the Markov equivalence class compatible with $P$.
@@
 \subsubsection{Without hidden common causes}
-The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of bayesian networks we must consider.
-
-We will begin with a brute force algorithm (described as the SGS algorithm in \cite{Sprites} and IC algorithm in \cite{Pearl2000}).
+The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of Bayesian networks we must consider.
+This assumption is referred to as causal sufficiency \citep{Sprites2000}.
+
+We will begin with a brute force algorithm (described as the SGS algorithm in \citet{Sprites2000} and IC algorithm in \citet{Pearl2000}).
 While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.
@@
  \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
 \textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
@@
 
-The PC algorithm \cite{Sprites} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilise the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
+The PC algorithm \citet{Sprites2000} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilise the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
 It also checks for low order conditional independence relations before higher order ones.
-This allows it to exploit any sparsity in the true network, leading to much better average case performance  \cite{Sprites} (although the worst case, where the true network is complete, is still exponential).
+This allows it to exploit any sparsity in the true network, leading to much better average case performance  \citep{Sprites2000} (although the worst case, where the true network is complete, is still exponential).
 With finite data, the order in which the links are considered can change the output (unlike for SGS).
@@
  \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) bayesian network $G$\\
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
 \textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
@@
 The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution.
-Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects with equation \ref{eq:allvarsobserved}.
+Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects from the truncated product formula \ref{eq:truncatedproduct}.
 We can then bound the true causal effect by combining the results for the all the networks.
-This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularisation techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}.
-An implementation is available in the R package \cite{Kalisch2012}
-
-\todo{mention failed replication - if published}
+This procedure is the IDA algorithm \citep{Maathuis2009} and has been found to outperform standard regularisation techniques at finding causal effects in a high-dimensional yeast gene expression data set \citep{Maathuis2010}.
+An implementation is available in the R package \citep{Kalisch2012}
 
@@
 In addition, the space of causal networks is not closed under marginalisation.
-If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalising over $\boldsymbol{U}$, may not be faithful to any bayesian network (see figure \ref{fig:DAGSnotclosed}).
+If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalising over $\boldsymbol{U}$, may not be faithful to any Bayesian network (see figure \ref{fig:DAGSnotclosed}).
+The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables).
 
@@
 \label{fig:DAGSnotclosed}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={punkt}]
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={observed},lt/.style={latent}]
  %nodes
-\node[main node](1){U}; 
+\node[lt](1){U}; 
 \node[main node, below left = of 1](2){B};
@@
 
-The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables). See example figure XXX.
-
 \begin{theorem}
@@
 Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations.
-The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG.
-Assuming there are no selection variables (see \cite xxx), the PAG can contain four types of link:
+The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \cite{Spirtes1995}.
+Assuming there are no selection variables, the PAG can contain four types of link:
 
@@
 
-The circles indicate where it is ambiguous if there should be an arrowhead (ie where there is one in some MAGs and not in others in the equivalence class).
+The circles indicate where it is ambiguous if there should be an arrowhead (IE where there is one in some MAGs and not in others in the equivalence class).
 Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).
 
-\todo{put these figures in a table with true graph input on left and FCI output on right}
-\todo{uniform styles used for nodes across thesis}
 \begin{figure}
@@
 \label{fig:FCIExamples}
- \begin{subfigure}[t]{0.8\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+ \begin{subfigure}[t]{0.4\textwidth}
+ \centering
+ \caption{True DAG}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed},lt/.style={latent}]
  %nodes
-\node[main node](1){U}; 
+\node[lt](1){U}; 
 \node[main node, below left = of 1](2){B};
@@
  \end{subfigure}
- \begin{subfigure}[t]{0.8\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={punkt}]
+ \begin{subfigure}[t]{0.4\textwidth}
+  \centering
+  \caption{Partial ancestral graph}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed}]
  %nodes
@@
 \end{subfigure}
+\par\bigskip 
  \begin{subfigure}[t]{0.4\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+ \centering
+ \caption{True DAG}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm, thick,main node/.style={observed},lt/.style={latent}]
  %nodes
@@
  \begin{subfigure}[t]{0.4\textwidth}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={punkt}]
+  \centering
+  \caption{Partial ancestral graph}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={observed}]
  %nodes
@@
 
-With latent variables we are not using all the information - so we could go further (to nested Markov models and inequalities.) \cite{Richardson2012} \cite{Shipster2014}
-
-These are all constraint based methods ... efficient because they stop early, but also may not be robust to errors early on.
-
-A comparison of algorithms
-
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{| c | c | p{4cm} | c | c | c |}
-\hline
-  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
-  \hline
-  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
-  \hline
-  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
-  \hline
-  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
-  \hline
-  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
-  \hline
-  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
-  \hline
-  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
- \hline
-\end{tabular}
-
-\subsubsection{Doing conditional independence tests}
-
-All the algorithms we have discussed in the previous section rely on being able to perform conditional independence tests.
+Latent variables can create constraints on the marginal distribution over the observed variables that cannot be expressed in term of conditional independencies.
+These generalised constraints can be expressed and leveraged within nested Markov models \citep{Richardson2012,Shipster2014}
+
+%\renewcommand{\arraystretch}{1.5}
+%\begin{tabular}{| c | c | p{4cm} | c | c | c |}
+%\hline
+%  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
+%  \hline
+%  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
+%  \hline
+%  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
+%  \hline
+%  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
+%  \hline
+%  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
+%  \hline
+%  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
+%  \hline
+%  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
+% \hline
+%\end{tabular}
+
+All the algorithms discussed in this section rely on being able to perform conditional independence tests.
 This is non-trivial with high dimensional data.
-If the functional relationship between the variables is linear with Gaussian noise then the network represents a multivariate normal distribution and a pair of variables $A$ and $B$ are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero.
-Where the functions are non-linear
-
-\begin{itemize}
-\item \cite{Zhang2012} Kernel independence tests
-\item HSIC \cite{Gretton2008}
-\end{itemize}
+If the functional relationship between the variables is linear with Gaussian noise then the network represents a multivariate normal distribution and a pair of variables are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero \citep{Koller2009}.
+Where the functions are non-linear one can apply kernelised independence tests \citep{Gretton2008,Zhang2012}
 
@@
 They cannot distinguish between two models that result in the same set of conditional independence relations.
-Consider the very simple case where we have only two variables and the only possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
+Consider the very simple case where  there are only two variables and the possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
 These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
@@
 Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$.
-What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from.
+What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from?
 Recall the functional definition of causality (section \ref{sec:SEM}).
-There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertible functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a}, linear models with non-Gaussian noise \cite{Hoyer2012}
-
-There is a connection between casual discovery and semi-supervised learning.\cite{causal anti causal}.
+There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a} or linear models with non-Gaussian noise \cite{Hoyer2012}.
+
+The causal direction can also be identified via a connection between casual discovery and semi-supervised learning \citep{Janzing2012}.
 Suppose we are trying to learn $\P{Y|X}$.
@@
 Therefor $\P{X}$ should be independent of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning.
-However if the true causal model is $Y \rightarrow X$ then variations in the $P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help.
-
-ONE POINT I'M NOT CERTAIN ON HERE.
-might not knowledge of $\P{X}$ help define which region of the mapping we need to be most accurate on in order to minimise the loss (ie analogous to the covariate shift problem).
-\todo {open question}
-
-%If X → Y, what is the relationship between P(X) and P(Y|X)? 
-What about P(Y) and P(X|Y)?
-
-%In semi supervised learning we are given training points from P(X,Y) and additional points sampled from P(X). The goal is to learn P(Y|X). The extra points from P(X) will not help if X → Y.
-
-deterministic functions,
-
-IGCI \cite{Janzing2012a}
-
-Independence of input and mechanism. Daniusis et al 2012
-
-Independence of function and input: If $X \rightarrow Y$ and we have a functional causal model $y = f(x,e)$ then the input
-distribution $P(X)$ and function $f$ represent independent mechanisms. We do not expect a change in the input distribution to modify the function.
+However if the true causal model is $Y \rightarrow X$ then variations in the $\P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help.
+This assumption of independence of mechanism and input can also allow the identification of the causal direction in SEMs even where the functions are deterministic and invertable \citep{Daniusis2010}.
+\citet{Janzing2012a} leverage an information geometric viewpoint of the independence of mechanism and input to infer the causal direction between a pair of associated variables.
 
@@
 Unfortunately we do not have a large collection of data sets where the causal direction is known to train such a model.
-\citep{LopezPaz2014} instead use a simulated data set so their model will necessarily be base on the assumptions they make when generating the data.
+\citet{LopezPaz2014} instead use a simulated data set so their model will necessarily be based on the assumptions they make when generating the data.
 Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.
-Their approach performed well in the causal effect pairs challenge, correctly identifying the causal direction in ??
-percent of cases (against a benchmark of 50\% from random guessing)\ref{XXX}.
-
-\citep{Peters2014} have extended results from the bi-variate case to the multivariate setting (with no hidden variables)?.
+
+\citep{Peters2014} have extended results from the bi-variate case to the multivariate setting.
 They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable.
-The build on this to develop an algorithm that allows the construction of causal graphs based on the additive assumption??
-
-\subsection{Granger causality}
-
-Granger causality was developed within economics and is not typically discussed within the causal discovery literature.
-
-Although it does differ from the previous approaches in the types of assumption it relies upon it none-the-less tackles the same goal
-
-so we shall include it here.
-
-Granger casualty essentially assumes that the future cannot cause the past.
-This may seem an entirely waterproof assumption.
-However it can be tripped up by unmeasured confounding variables.
-The cock crows before the sun rises - so does it cause the sun to rise?
-No.
+They build on this to develop an algorithm that allows the construction of causal graphs based on the additive noise assumption.
 
@@
 There is an obvious alternative.
-Instead of trying to infer the outcome of an intervention from passive observations we could just do it and see what happens.
-There are two key differences between observing a system and explicitly intervening in it.
-Firstly, when we intervene, we can choose which actions to take and thus get some control over which distributions we learn about.
-Secondly, if we are explicitly choosing interventions, we have a perfect model of the probability that we select each action given any context, allowing us to control confounding bias.
-
-\subsection{The role of randomisation}
-Randomised controlled trials are often presented as the gold standard for determining causal effects in many areas of social sciences and economics \ref{XXX}.
+Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens.
+There are three key differences between observing a system and explicitly intervening in it.
+Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from.
+Selecting data points optimally for learning is the focus of the optimal experimental design liturature within statistics \citep{pukelsheim2006optimal} and the active learning liturature in machine learning \citep{settles2010active}.
+Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias.
+Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning.
+For example, in a drug trial , assigning people a sub-optimal treatment has real world costs.
+This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain.
+This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.
+
+\section{Randomised experiments}
+Randomised controlled trials are often presented as the gold standard for determining causal effects.
 What is it about randomisation that makes it so important when it comes to causality?
@@
 
-\begin{figure}[h]
+\begin{figure}
 \caption{causal network for a randomised experiment}
@@
 
-\todo {turn this into something}
-This would be a natural place to discuss what happens with finite data sets where there are many other variables that influence the outcome.
-We expect that across at least some of these variables, the target and control group will not be balanced.
-Does this not inject bias into the causal estimates?
-The answer is no, (I think - a better more general proof would be nice).
-The more such attributes there are, the higher is the variance in the outcome within the target and control group.
-When we perform a confidence test to ascertain whether or not there is a difference ...
-what happens.
-
 Randomisation does not ensure target and control group are exactly alike.
@@
 
-DISCUSS STRATIFIED RANDOM EXPERIMENT
-
-What is the role of randomisation? How do bandits algorithms work despite being only partially randomised? What else can you do to improve randomised studies (variance reduction, lower regret).
-
-The benefit provided by randomisation in breaking the link between the treatment variable and any con founders should not be understated.
-The possibility of confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (test for confounding).
+Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates belived to influence the outcome of interest.
+This introduces a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, nessessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
+The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
+See \citet{imbens2015causal} for a discussion of the trade-offs between statified versus completely random experiments.
+
+\begin{figure}
+\caption{causal network for a stratified randomised experiment}
+\label{fig:random_experiment_network_stratified}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
+\node[node](1){Treatment};
+\node[node, above left=of 1](2){Randomiser};
+\node[node, above right=of 1](5){Covariates};
+\node[node, below right=of 5](3){Outcome};
+\node[lt, above=of 3](4){U};
+
+\path[]
+	(2) edge (1)
+	(1) edge (3)
+	(4) edge (3)
+	(5) edge (3) edge (1);
+\end{tikzpicture}
+\end{figure}
+
+The benefit provided by randomisation in breaking the link between the treatment variable and any latent confounders should not be understated.
+The possibility of unobserved confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (there is no test for confounding).
 This means causal estimates from non-experimental data are always subject to the criticism that an important confounder may have been overlooked or not properly adjusted for.
-However, randomised experimentation does have some limitations.
-
-\subsubsection{Limitations of randomised experiments}
+However, randomised experiments do have some limitations.
+
+\subsection{Limitations of randomised experiments}
 \label{subsec:limitations_of_experiment}
-\begin{enumerate}
-
-\item The probability that an individual is treated is set in advance and fixed over the course of the experiment.
-As the experiment proceeds we may start to see evidence that one treatment is much more effective than another.
-However (in a pure randomised experiment) we don't use this information to update the treatment that later patients revive.
-The result is that some will unessential be given a sub-optimal treatment.
-\item Experiments are expensive and difficult to conduct. This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
-
-\item Experiments are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year collage students at research universities).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}.
-The results may not carry over to the general population of interest.
-The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} \todo{check ref} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
-
-\item The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
-There may be imperfect compliance: the treatment selected by the randomiser is not always followed.
-Or output censoring: the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.
-See figure XXX for a the graphical model of a randomised experiment with imperfect compliance and outcome censoring.
-
-\item It is not always possible or ethical to conduct a randomised experiment \cite{parachutes} \ref{Parachute use to prevent death and major trauma related to gravitational challenge: systematic review of randomised controlled trials}.
-\end{enumerate}
-
-\begin{figure}[h]
+
+The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
+There may be imperfect compliance, the treatment selected by the randomiser is not always followed, or output censoring, the experimenter is not able to observe the outcome for all units (for example if people drop out).
+If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.\footnote{Non-compliance is a problem if the goal is to estimate the causal effect of the treatment on the outcome but not if the goal is to estimate the causal effect of prescribing the treatment.
+The latter makes sense in a context where the process by which people decide whether or not to take the treatment they have been prescribed is likely to be the same if the treatment were made available more generally beyond experimental trial.} See figure \ref{fig:random_experiment_network_imperfect_compliance} for a the graphical model of a randomised experiment with imperfect compliance.
+
+\begin{figure}
+\caption{causal network for a randomised experiment with imperfect compliance}
+\label{fig:random_experiment_network_imperfect_compliance}
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
+\node[node](1){Actual Treatment};
+\node[node,above left=of 1](5){Assigned Treatment};
+\node[node, above =of 5](2){Randomiser};
+\node[node, right=of 1](3){Outcome};
+\node[lt, above left=of 3](4){U};
+\path[]
+	(2) edge (5)
+	(5) edge (1)
+	(1) edge (3)
+	(4) edge (3) edge (1);
+\end{tikzpicture}
+\end{figure}
+
+It is not always possible or ethical to conduct a randomized controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parashute use for the reduction of the mortality and morbity associated with falls from large hights \ref{fig:parashute}.
+When experimentation is possible, it is frequently difficult or expensive.
+This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
+In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities).
+This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest.
+The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
+
+\begin{figure}
 \begin{center}
-\caption{Experiments are not always ethical}
+\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parashutes for the prevention of morbitity and mortality associated with falls from large heights.}
+\label{fig:parashute}
 \includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
@@
 
+Finally, non-adaptive randomised experiments are not optimal from either an active or reinforcement learning perspective.
+As an experiment proceeds information is obtained about the expectation and variance of each intervention (or treatment).
+Fixed experimental designs cannot make use of this information to select which intervention to try next.
+This results in both poorer estimates for a fixed number of experimental samples and more sub-optimal actions during the course of the experiment.
+
 \section{Multi armed bandits}
-\todo{is there a way in latex to have a command for referring to each type of thing like figure, subsection, etc such that it can be updated programatically?}
-
-Multi-armed bandits were originally introduce by XXXX to address the problem that it is sub-optimal to fix a probability of treatment in advance as we discussed in subsection \ref{subsec:limitations_of_experiment}.
-The problem can be formally stated as follows:
-
-\paragraph*{Definition} 
-In its classic formulation \cite{Robbins1952} the (stochastic) K-armed bandit describes a sequential decision making problem, with $K$ possible actions or arms.
-Each arm $i$ is associated with a fixed but unknown reward distribution $\nu_i$.
-In order to obtain regret bounds, some assumptions are required on the distributions $\nu_i$.
-It sufficient to assume they are sub-Gaussian.
-For each time step $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$.
+\label{sec:multi-armed-bandits}
+
+Multi-armed bandits address the problem of desiging experiments that can adapt as samples are observed.
+Their introduction is generally attributed to \citet{Thompson1993}.
+In its classic formulation \cite{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
+Each arm $i$ is associated with a fixed but unknown reward distribution.
+\footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
+It sufficient (but not nessessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
 The goal of the learner is to maximise the total reward they receive.
-This problem introduces the fundamental exploration-exploitation trade-off.
-The learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
-
-The performance of bandit algorithms is generally described by the (pseudo) regret, $R(T)$.
-This is the difference between the expected reward obtained by the algorithm and the expected reward of selecting the best action in every time step.
-
-\subsection{Regret} How do we measure the performance of a bandit algorithm?
-
-%$\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$
-
-Expected Regret
-\eqn{
-R_T(\pi) = \E{max_{i \in [k]} \sum_{t=1}^T{X_{t,i}}} - \E{\sum_{t=1}^T{X_{t,A(t)}}}
-}
-
-Pseudo-Regret
-\eqn{
-R_T(\pi) &= max_{i \in [k]}\E{\sum_{t=1}^T{X_{t,i}}} - \E{\sum_{t=1}^T{X_{t,A(t)}}} \\
-&= n\mu_{i^*} - \E{\sum_{t=1}^T{X_{t,A(t)}}}
-}
-
-Simple Regret
-\eqn{
-\mu_{i^*} - \E{\mu_{\hat i^*}}.
-}
-
-An algorithm is learning if it obtains regret that is sub-linear in $T$.
-
-Scholars in the 1940's found this problem so frustrating that they suggested it be dropped over Germany to distract the scientists there from the war effort.\cite{XXX}.
-
-Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
-In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
-The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm, $R_s(T) = \mu^* - \E{\mu_i}$ .
-This problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialisation or exploitation phase.
-For example, many strategies might be assessed via simulation prior to one being selected and deployed.
-The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}.
-
-\todo{connect bandit regret to early stopping of experiment problem}
-The problem of when to stop an experiment early can also be phrased as a bandit problem and analysed with the same techniques.
-XXX et al showed that if use an experimental setup with early stopping, the regret you obtain will be at least twice that you could have achieved had you made each decision sequentially using a bandit approach.
-Of course this neglect any costs that might be associated with the complexity in implementation of the two approaches.
-
-\subsection{The exploration/exploitation trade-off}
-A central aspect to bandit problem is the trade of between exploiting the information we have already obtained with exploring options about which we remain uncertain.
-Suppose you have sampled five of your local restaurants and found one which you really enjoyed.
-How often should you eat there versus exploring new options.
-The degree to which you should explore versus exploit depends on how many step remain to go in the game and what assumptions you make about the distributions from which the rewards are sampled.
-If this is your last meal you might as well just eat the food you enjoy most but if you hope to live for many meals to come then you stand to gain by sampling a wider range of options.
-In many cases, the total number of rounds that you will get to play in the game is unknown.
-If you cannot imagine that it would be possible for a meal to be substantially better than that which you obtain (in expectation) at your current favourite then there is little point exploring further.
-
-\subsection{Key approaches and results}
+This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
+
+\vspace{0.5cm}
+\begin{definition}[Stochasitic k-armed bandit problem]
+Let $\actionspace = \set{1,...k}$ be the set of available actions (or bandit arms) and $\P{\vec{Y}} = \P{Y_1,...,Y_k}$ be the joint distribution over the rewards for each action.
+The multi-armed bandit problem proceeds over $T$ rounds.
+In each round $t$,
+\begin{enumerate}
+\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
+\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{Y}}$
+\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\end{enumerate}
+At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$.
+We denote the expected reward for the action $i$, $\E{ \P{Y_i}}$, by $\mu_i$ and the action with the highest expected reward by $i^*$.
+\end{definition}
+
+The total reward a bandit algorithm/policy can expect to acheive depends on the distributions from which the rewards for each action are sampled.
+To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
+This difference is known as the (cummulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning liturature.
+There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
+In addition, researches often refer to the expected regret as \quotes{the regret}.},
+\eqn{
+\label{eqn:raw_regret}
+\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
+}
+
+Both the rewards and the actions selected by the algorithm are random variables.
+The majority of work in the bandit liturature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Audibert2007}.
+The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is refered to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
+A stochastic bandit algorithm is learning if it obtains psuedo-regret that is sub-linear in $T$.
+
+\vspace{0.5cm}
+\begin{definition}[Pseudo-Regret]
+\eqn{
+\label{eqn:pseudo_regret}
+\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}} \\
+&= n\mu_{i^*} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+}
+\end{definition}
+
+\subsection{Approaches, Regret bounds and results}
+
+The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators.
+The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards.
+As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
+The expectation and variance guarentees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
+This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
+Instead, we focus on computing upper and lower bounds on the regret.
+
+\begin{figure}[h]
+\centering
+\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples.
+This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim bernoulli(0.2)$.
+In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$.
+$\E{\hat{\mu}} = 0.439$.
+The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping.
+Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded.
+This has some interesting potential real world implications.
+Early stopping of clinical trials is controversial.
+A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
+\label{fig:early_stopping}
+\includegraphics[scale=1]{extreme_early_stopping}
+\end{figure}
+
+Talk about problem dependent and worst case bounds
+
+Lay out what they are for the stochastic bandit setting.
+
+\begin{itemize}
+\item problem dependent versus worst-case bounds
+\item upper and lower bounds
+\end{itemize}
+Here we can talk about the different kinds of bounds that are possible to achieve and what has been done so far (for stochastic k-armed bandits)
 
@@
 
-\subsection{The need to add structure}
-The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.
-This makes problems with large or infinite actions spaces intractable.
-REFERENCE equation showing linearity with arms.
-CHECK ITS TRUE.
-
-The classic multi-armed bandit is a powerful tool for sequential decision making.
-However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and Gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
+The bandit learning algorithm is sometimes compared to the standard supervised learning in terms of the feedback recieved.
+One way of viewing the standard supervised learning setting is that the label for each instance corresponds to the correct or optimal choice for that setting.
+However, this does not have to be the case.
+In the bandit setting the algorithm observes only the reward for the selected action.
+How are standard supervised learning problems translated to bandit problems for testing algorithms?
+(is it contextual bandit problems?)
+
+Bandit algorithms that require the horizon $T$ to be known in advance are refered to as \emph{fixed horizon} algorithms, as opposed to \emph{anytime} algorithms that do not rely on knowing the horizon.
+A fixed horizon algorithm may be converted to an anytime algorithm via the so called "doubling trick" \citep{}.
+The algorithm starts with some fixed horizon and increases it at a fixed rate everytime the current horizon is reached until the problem terminates.
+\todo{what can be said about the optimality of this?}
+
+Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
+In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
+The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.
+
+\vspace{0.5cm}
+\begin{definition}[Simple Regret]
+\eqn{
+\label{eqn:simple_regret}
+\mu_{i^*} - \E{\mu_{\hat i^*}}.
+}
+\end{definition}
+
+The best arm identification problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialisation or exploitation phase.
+For example, many strategies might be assessed via simulation prior to one being selected and deployed.
+The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}.
+
+\todo{connect bandit regret to early stopping of experiment problem}
+The problem of when to stop an experiment early can also be phrased as a bandit problem and analysed with the same techniques.
+XXX et al showed that if use an experimental setup with early stopping, the regret you obtain will be at least twice that you could have achieved had you made each decision sequentially using a bandit approach.
+Of course this neglect any costs that might be associated with the complexity in implementation of the two approaches.
 
@@
 \subsubsection{Contextual bandits}
+Contextual bandits are a generalisation of multi-armed bandits.
+
+\vspace{0.5cm}
+\begin{definition}[Contextual bandit problem]
+Let $\P{\vec{Y},\vec{X}}$ be the joint distribution over the rewards for each action and some context $\vec{x} \sim \mathcal{X}$. In each round $t \in \set{1,...T}$, 
+\begin{enumerate}
+\item the world stochastically generates the vector of rewards for each action and the context, $([Y_{t,1},...,Y_{t,k}],\vec{x}_t) \sim \P{\vec{Y},\vec{X}}$ and reveals $\vec{x}_t$ to the learner
+\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps.
+\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\end{enumerate}
+\end{definition}
+
+What are the reductions of contextual bandits to other problem (ie k armed-bandits?)
+
 In the previous sections each decision was identical.
@@
 
+\subsubsection{Learning from logged bandit data}
+
 \subsubsection{Markov decision processes}
+
+A key limitation of the bandit setting is that
+
+This is reasonable in settings where we have many similar but independent units or individuals and the goal is to identify which action is optimal (in a given context). However
+
+The reward of an action is not immediately known. - a long sequence of steps. Delayed reward.
+
 In the bandit setting the reward distributions for each arm are assumed to be fixed.
@@
 
+\subsection{The need to add structure}
+The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.
+This makes problems with large or infinite actions spaces intractable.
+REFERENCE equation showing linearity with arms.
+CHECK ITS TRUE.
+
+The classic multi-armed bandit is a powerful tool for sequential decision making.
+However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and Gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
+
+\section{Questions}
+\begin{enumerate}
+\item When does the expected regret make sense as the thing we actually would wish to optimise?
+Is optimising the expected regret any different to optimising the psuedo-regret?
+Who is the definition of regret down to?
+Does this definition of regret come from learning with expert advise?
+\item Is there a reference for the doubling trick? Is it an automatic way of converting a fixed horizon algorithm to an anytime one?
+\item Why is the dependence on the number of arms $K$ in the problem dependent bounds but $\sqrt{K}$ in the worst case bounds?
+\item rewards in the multi-armed bandit setting can be sampled from some joint distribution?
+\item Is Learning from logged feedback data just covariate shift too?
+\item In contextual bandits, does it ever make sense that $Y$ causes $X$?
+If the reward is some function of the action taken and the context, then the full vector of counterfactual outcomes $Y$ is never instantiated.
+Are there any implications of this for inference?
+\item Why does anything solving a regression problem work in the anti-causal direction - it would seem to break the fundamental underlying assumption of additive noise?
+\end{enumerate}
+
 \chapter{Causal Bandits: Unifying the approaches}
@@
 \section{The framework}
+\label{sec:intro_causal_bandit}
 A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
-Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{} for a simple example.
-In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example only the set of single variable interventions.
+Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{fig:unify_frameworks} for a simple example.
+In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example the set of single variable interventions.
 The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node minus the cost of the selected action.
-In this thesis I examine the case where the causal graph is known.
+However for simplicity, we will consider the reward to be the value of a single specified node.
 We refer to these problems as \emph{causal bandit problems}.
-Extending this work to simultaneously learning the casual graph is discussed in section \ref{XXX}.
-
-\subsubsection{Definition of casual bandit problem} \todo{make proper definition}
-A learner for a casual bandit problem is given the casual model's graph $\mathcal{G}$ and a set of \emph{allowed actions} $\mathcal{A}$.
-One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
-
-\begin{figure}[h]
-\caption{A simple causal graphical model and corresponding complete action space. A and B represent binary variables that can be intervened on and Y represents the reward.}
+In this thesis I focus on the case where the causal graph is known.
+Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.
+
+\begin{figure}
+\caption{A simple causal graphical model and corresponding complete action space. $W$ and $Z$ represent binary variables that can be intervened on and $Y$ represents the reward.}
 \label{fig:unify_frameworks}
@@
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
-\node[main node](1){A};
+\node[main node](1){W};
 \node[tg, below right=of 1](2){Y};
-\node[main node, above right=of 2](3){B};
+\node[main node, above right=of 2](3){X};
 \path[]
@@
 \begin{subfigure}[t]{0.4\textwidth}
-Actions = \begin{tabular}{|c|}
+$\actionspace$ = \begin{tabular}{|c|}
 	\hline
-  do(A=0,B=0) \\
-  do(A=0,B=1) \\
-  do(A=1,B=0) \\
-  do(A=1,B=1) \\
+  $do(W=0,Z=0)$ \\
+  $do(W=0,Z=1)$ \\
+  $do(W=1,Z=0)$ \\
+  $do(W=1,Z=1)$ \\
   \hline
-  do(A=0) \\
-  do(A=1) \\
-  do(B=0) \\
-  do(B=1) \\
-  do() \\
+  $do(W=0)$ \\
+  $do(W=1)$ \\
+  $do(Z=0)$ \\
+  $do(Z=1)$ \\
+  $do()$ \\
   \hline
@@
 
-The number of actions or arms grows exponentially with the number of variables in the graph,
-making it important to use algorithms that leverage the graph structure to reduce the
-search space.
-Modelling a problem as a causal graph only makes sense when rewards are generated stochastically - since causal graphs fundamentally model probability distributions over variables.
-Thus the connection is to stochastic bandit problems (although adversarial bandits algorithms may be applied to stochastic problems).
-
-We now need to specify the feedback model for the causal bandit problem.
-What information is available to the decision making agent before and after they select an action?
-The causal bandit problem takes on characteristics of different bandit settings depending on the assumptions we make about what actions are available to the agent, what variables are observed and whether they are observed before or after the action is chosen.
-
-Enumerate the settings
+The type of problem we are concerned with is best illustrated with an example.
+Consider a farmer wishing to optimise the yield of her crop.
+She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
+In each season the farmer has enough time and money to intervene and control at most one of these
+variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
+When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
+How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
+
+We will assume each variable only takes on a finite number of distinct values.
+(The path to relaxing this assumption would be through levering the work on continuous armed bandits).
+The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
+An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
+We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and
+the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
+
+\vspace*{.5cm}
+\begin{definition}[Causal bandit problem]
+A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
+Each action $a \in \actionspace$ assings a value to a subset of the variables in $\mathcal{X}$.
+One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
+
+The causal bandit game proceeds over $T$ rounds. In each round $t$, the learner:
 \begin{enumerate}
-\item Bandit feedback: the agent selects an action and then observes only the reward of that action.
-If feedback is received only on the reward node then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions.
-See figure XXX as an example.
-\item Bandit feedback with side information (context). The agent can view the value of some variables prior to selecting an action. After selecting they observe the reward of the selected node. 
-\item Post action feedback:
+\item \emph{observes} the value of a subset of the variables $\vec{X}^c_t$,
+\item \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations, and finally
+\item \emph{observes} sampled values for another subset of variables $\vec{X}^o_t$ drawn from $\P{\vec{X}^o_t | do(\vec{X}_t = \vec{x}_t)}$ including the \emph{reward} $Y_t \in \{0,1\}$.
 \end{enumerate}
 
-If we receive feedback on additional nodes, the problem can be more interesting.
+We refer to the set of variables that can be observed prior to selecting an action $\vec{X}^c$ as contextual variables and the set of variables observed after the action is chosen, $\vec{X}^o$, as post-action feedback variables.
+Note that $\vec{X}^c$ and $\vec{X}^o$ need not be disjoint.
+A variable may be observed both prior to and after the agent selects an action, and the action may change its value.
+The objective of the learner is to minimise either the simple (equation \ref{}) or cummulative regret (equation \ref{}).
+
+\end{definition}
+
+The causal bandit problem takes on characteristics of different bandit settings depending on the actionspace $\actionspace$, which variables are observable prior to selecting an action and on which variables we recieve post-action feedback.
+If feedback is recieved only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
+If we recieve post-action feedback on additional nodes the problem can be more interesting.
 In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
@@
 
-\eqn {
-P(Y|do(A = 0)) &= P(Y|do(),A = 0) \\
-&= P(Y|do(B=0),A=0)P(B=0)+P(Y|do(B=1),A=0)P(B=1) 
-}
-
-Thus we gain information about the reward for the action $do(A=0)$ from selecting the action $do()$ or $do(B = b)$ and then observing $A = 0$.
-
-We only get this form of side information for actions that don't specify the value of every variable, ie those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-Since the reward distribution for actions that set a subset of the variables is the result of marginalising out other variables, they can only be optimal if they have lower cost.
-So if the cost of all actions is constant (no matter how many variables must be set), then the problem has the same characteristics as if only the reward node were observable.
-
-If the information on the value of additional nodes is available prior to selecting an action the problem resembles a contextual bandit.
-For example if we observe $A = 0$ then, in deciding between the actions $do(B=0)$ and $do(B=1)$, we would want information on $P(Y|A=0,B=0)$ and $P(Y|A=0,B=1)$.
-Note, side information can still arise if we learn the value of some variables prior to selecting an action and some afterwards.
-
-Although we will focus on the intervene-then-observe ordering of events within each round, other scenarios are possible.
-If the non-intervened variables are observed before an intervention is selected our framework reduces to stochastic contextual bandits, which are already reasonably well understood~\citep{Agarwal2014}.
-Even if no observations are made during the rounds, the causal model may still allow offline pruning of the set of allowable interventions thereby reducing the complexity.
-
-\section{Related work}
-\begin{enumerate}
-\item leaning from log data
-\item bandits with imperfect compliance
-\item rigged casino paper
-\end{enumerate}
-
-\begin{figure}[h]
+\eq{
+P(Y|do(W = 0)) &= P(Y|do(),W = 0) \\
+&= P(Y|do(X=0),W=0)P(X=0)+P(Y|do(X=1),W=0)P(X=1) 
+}
+
+\begin{figure}
 \caption{Example causal graph (based on \cite{Koller2009}) where the outcome of interest (reward) is cholesterol level .
@@
 
-\section{Causal bandits with post action feedback}
-WHY THIS PROBLEM. This work was presented at NIPS 2016 \ref{}.
-\todo{add relevant stuff from introduction section}
-
-\subsection{Notation}
-
-We will assume each variable only takes on a finite number of distinct values. (The path to relaxing this assumption would be through levering the work on continuous armed bandits).
-
-The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
-
-An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-
-We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and 
-the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
-
-\subsection{Definition of causal bandit game with post-action feedback}
-\todo{ re-read and incorporate response from reviewer feedback for NIPS}
-
-The causal bandit game proceeds over $T$ rounds.
-In round $t$, the learner \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations. 
-It then \emph{observes} sampled values for all non-intervened variables $\vec{X}^c_t$ drawn from $\P{\vec{X}^c_t | do(\vec{X}_t = \vec{x}_t)}$, 
-including the \emph{reward} $Y_t \in \{0,1\}$. 
-After $T$ observations the learner outputs an estimate of the optimal action $\hat a^*_T \in \actions$ based on its prior observations.
-
-The objective of the learner is to minimise the simple regret $\simpleregret = \mu^* - \E{\mu_{\hat a^*_T}}.$ This is sometimes refereed to as a ``pure exploration''~\citep{Bubeck2009a} or ``best-arm identification'' problem~\citep{Gabillon2012a} and is most appropriate when, as in drug and policy testing, the learner has a fixed experimental budget after which its policy will be fixed indefinitely.
+Thus we gain information about the reward for the action $do(W=0)$ from selecting the action $do()$ or $do(X = x)$ and then observing $W = 0$.
+We only get this form of side information for actions that don't specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
+If additional variables are only observed before an intervention is selected the causal bandit problem reduces to stochastic contextual bandits, which are already reasonably well understood \citep{Agarwal2014}.
 
@@
 Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.
-Intuitively though, one would expect to perform better by making use of the extra structure and observations.
+However, the number of actions or arms grows exponentially with the number of variables in the graph making it important to develop algorithms that leverage the graph structure and additional observations.
+
+\section{Causal bandits with post action feedback}
+
+We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, and where the goal of the learner is to minimise the simple regret.
+I presented this work at NIPS 2016 \citep{NIPS2016_6195}.
+
+\paragraph{Related Work} As alluded to above, causal bandit problems can be treated as classical multi-armed bandit problems by simply ignoring the causal model and extra observations and applying an existing best-arm identification algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
+However, as we show in \S\ref{sec:simple-regret}, ignoring the extra information available in the non-intervened variables yields sub-optimal performance.
+
+A well-studied class of bandit problems with side information are ``contextual bandits''~\cite{Langford2008,Agarwal2014}.
+Our framework bears a superficial similarity to contextual bandit problems since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
+However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
+
+There have been several proposals for bandit problems where extra feedback is received after an action is taken.
+Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
+As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
+They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
+The partial monitoring approach taken by \cite{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
+\citet{Yu2009} utilize extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
+\citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled.
+The main difference from our present work is our focus on simple regret and the more complex information linking rewards for different arms via causal graphs.
+To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.
+
+%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
+They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
+% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
+It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
+Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
+
+Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
+\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been 
+intervened on can provide important contextual information. Their work differs in many ways. For example, the focus is on the cumulative regret and
+the context is observed before the action is taken and cannot be controlled by the learning agent.
+
+%\todof{write one sentence to explain how their work is different to ours}
+
+ \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions.
+Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
+Combining their handling of  causal induction with our analysis is left as future work.
+% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
+Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
+(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
+It is easy to come up with a counter example in which it does not).
+
+The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{Bottou2013a}, 
+where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discuss some of the issues 
+encountered in sequential design, but do not give an algorithm or theoretical results for this case.
 
 \subsection{The parallel bandit problem}
-
+\label{sec:simple-regret}
 In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem which we call the {\it parallel bandit}.
-It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \ref{XXX}, including the farming example.
+It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro_causal_bandit}, including the farming example.
 It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
@@
 \begin{figure}
-    \begin{subfigure}[b]{0.34\textwidth}
-	\centering    
+  \caption{Causal Models}\label{fig:causalmodels}
+    \begin{subfigure}[t]{0.34\textwidth}
+	\centering   
+	 \caption{Parallel graph}\label{fig:parallel} 
           \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
@@
 \end{tikzpicture} 
-        \caption{Parallel graph}
-        \label{fig:parallel}
 \end{subfigure}
-    \begin{subfigure}[b]{0.2\textwidth}
-    \centering
+    \begin{subfigure}[t]{0.25\textwidth}
+    \centering
+     \caption{Confounded graph}
+        \label{fig:causalStructure_confounded}
         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
@@
 \end{tikzpicture}
-        \caption{Confounded graph}
-        \label{fig:causalStructure_confounded}
     \end{subfigure}
-    \begin{subfigure}[b]{0.4\textwidth}
-    \centering
+    \begin{subfigure}[t]{0.4\textwidth}
+    \centering
+     \caption{Chain graph}
+        \label{fig:causalchain}
          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
@@
 \end{tikzpicture}
-        \caption{Chain graph}
-        \label{fig:causalchain}
     \end{subfigure}
-    \caption{Causal Models}\label{fig:causalmodels}
-    \vspace{-.5cm}
 \end{figure}
@@
 
-The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
-
 The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
-of $\hat p_a$ and $\hat m$ about their true values and may be found in the supplementary material.
+of $\hat p_a$ and $\hat m$ about their true values and may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
 
@@
 \subsection{General graphs}
+\label{sec:simple-regret-general}
 We now consider the more general problem where the graph structure is known, but arbitrary.
@@
 
-We could naively generalize our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorisation to 
+We could naively generalise our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorisation to 
 write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
@@
 Thus it has broader applicability to the parallel graph than the generic algorithm given here.
-We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to minimax optimal, but leave lower bounds
+We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to mini-max optimal, but leave lower bounds
 for future work.
@@
 This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
-The proof is in section XXX.
+The proof is in section \ref{sec:m-equivelence}.
 
@@
 larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
-and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{BJQ13} who also use importance weighted estimators
+and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{Bottou2013a} who also use importance weighted estimators
 to learn from observational data.
@@
 Thomson sampling and UCB are optimised to minimise cumulative regret.
-We apply them in the fixed horizon, best arm identification setting by running them upto horizon $T$ and then selecting the arm with the highest empirical mean.
+We apply them in the fixed horizon, best arm identification setting by running them up to horizon $T$ and then selecting the arm with the highest empirical mean.
 The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.
@@
 For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
-This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be minimax optimal asymptotically \citep{LMS14}.
-
-\subsection{Additional experiments} \todo {tie section together}
+This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be mini-max optimal asymptotically \citep{LMS14}.
+
 \begin{figure}[H]
@@
 We now compare the general algorithm with a range of standard bandit algorithms on the confounded graph in Figure \ref{fig:parallel_confounded}.
-All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action, $\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}$.
+All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action,
+
+\eq{
+\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}
+}
+
 We choose this setting because it generalises the parallel bandit, while simultaneously being sufficiently simple that we can compute the exact reward and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
@@
 \subsection{Discussion \& Future work}
-
+\label{sec:causal_bandit_discussion}
 Algorithm~\ref{alg:general} for general causal bandit problems 
@@
 In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
-Note that these modifications will not improve the minimax regret, which at least for the parallel bandit is already optimal.
+Note that these modifications will not improve the mini-max regret, which at least for the parallel bandit is already optimal.
 For this reason we prefer to emphasise the main point that causal structure should be exploited when available.
@@
 The latter case arises naturally if we assume free access to a large observational data set, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014b} could potentially be incorporated into a causal bandit algorithm.
-In particular, \citet{Hu2014b} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
+Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
+In particular, \citet{Hu2014} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
 Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational data set would be much more challenging.
@@
 \subsection{Proofs}
-
-\chapter{List of software programs for causal inference}
+\subsubsection{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}
+
+Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$. The assumption is non-restrictive since all variables
+are independent and permutations of the variables can be pushed to the reward function.
+
+The proof of Theorem \ref{thm:uq-simple} requires some lemmas. \
+
+\begin{lemma}\label{lem:conc1}
+Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
+\eq{
+\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
+}
+\end{lemma}
+
+\begin{proof}
+By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \cite{hagerup1990guided}),
+
+\eq{
+\P{\left|\hat q_i - q_i\right| \geq \epsilon} \leq 2e^{-\frac{T\epsilon^2}{6q_i}}
+}
+
+Letting $\delta = 2e^{-\frac{T\epsilon^2}{6q_i}}$ and solving for $\epsilon$ completes the proof.
+
+\end{proof}
+
+\begin{lemma}\label{lem:m_est}
+
+Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
+\eq{
+\P{2m(\vec{q}) / 3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})} \geq 1 - \delta\,.
+}
+\end{lemma}
+
+\begin{proof}
+Let $F$ be the event that there exists and $1 \leq i \leq N$ for which
+\eq{
+\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}\,.
+}
+Then by the union bound and Lemma \ref{lem:conc1} we have $\P{F} \leq \delta$. The result will be completed by showing that
+when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
+From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m(\vec{q})$ that $q_i \geq q_m \geq 1/m(\vec{q})$ and so by Lemma \ref{lem:conc1} we have
+\eq{
+\frac{3}{4} 
+&\geq \frac{1}{2} + \sqrt{\frac{3}{T} \log \frac{2N}{\delta}} 
+\geq q_i + \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}} 
+\geq \hat q_i \\
+&\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
+\geq q_i - \sqrt{\frac{q_i}{8m(\vec{q})}}
+\geq \frac{1}{2m(\vec{q})}\,.
+}
+Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m(\vec{q})$.
+For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m(\vec{q})$ that
+\eq{
+\hat q_i 
+\leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
+\leq \frac{1}{m(\vec{q})} \left(1 + \sqrt{\frac{1}{8}}\right)
+\leq \frac{3}{2m(\vec{q})}\,.
+}
+Therefore $m(\vec{\hat q}) \geq 2m(\vec{q}) / 3$ as required. 
+\end{proof}
+
+\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
+Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then,
+for $a \in A$, the algorithm estimates $\mu_a$ from $T_A \doteq T/(2m(\vec{\hat q}))$ samples.
+From lemma \ref{lem:m_est}, $T_A  \geq T/(4m(\vec{q}))$ with probability $(1-\delta)$.
+Let $H$ be the event $T_A < T/(4m(\vec{q}))$ and $G$ be the event $\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}$
+
+\eq{
+\P{G} \leq \P{H} + \P{G|\neg H} \leq \delta + \P{G|\neg H}
+}
+
+Via Hoeffding's inequality and the union bound,
+
+\eq{
+\P{G|\neg H}&\doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}},\text{ given }T_A \geq T/(4m(\vec{q}))} \leq \delta\\
+ \implies  \P{G} & \doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq 2\delta\,.
+}
+For arms not in $A$,
+
+\eq{
+\hat p_a = & \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq 1/m(\vec{\hat q}), \; \text {by definition of not being in $A$}\\
+ \geq & \frac{1}{2m(\vec{q})},\; \text{ with probability } 1-\delta \\
+\implies  T_a \doteq & \sum_{t=1}^{T/2} \ind{X_i = j} \geq  \frac{T}{4m(\vec{q})},\; \text{ with probability } 1-\delta
+}
+
+Again applying Hoeffding's and the union bound
+
+\eq{
+\P{\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}}} \leq 2\delta
+}
+
+Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 4\delta$ that,
+\eq{
+(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}} \doteq \epsilon\,.
+}
+If this occurs, then 
+\eq{
+\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
+}
+Therefore
+\eq{
+\mu^* - \EE[\mu_{\hat a^*_T}] 
+& \leq 4\delta + \epsilon\\
+& \leq \frac{8m(\vec{q})}{T} + \sqrt{\frac{2m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ letting } \delta = \frac{2m(\vec{q})}{T}\\
+& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jenson's Inequality}
+}
+which completes the result.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LOWER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}
+
+We follow a relatively standard path by choosing multiple environments that have different optimal arms, but which cannot all be statistically
+separated in $T$ rounds.
+Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
+For each $i$ define reward function $r_i$ by
+\eq{
+r_0(\boldsymbol{X}) &= \frac{1}{2} &
+r_i(\boldsymbol{X}) &= \begin{cases}
+\frac{1}{2} + \epsilon & \text{if } X_i = 1 \\
+\frac{1}{2} & \text{otherwise}\,,
+\end{cases}
+}
+where $1/4 \geq \epsilon > 0$ is some constant to be chosen later.
+We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
+environment determined by $\boldsymbol{q}$ and $r_i$. Let $\operatorname{P}_i$ be the corresponding measure
+on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \citet{Tsy08} we have
+\eq{
+\Prz{\hat a^*_T = a^*} + \Pri{\hat a^*_T \neq a^*} \geq \exp\left(-\KL(\operatorname{P}_0, \operatorname{P}_i)\right)\,,
+}
+where $\KL(\Ps_0, \Ps_i)$ is the KL divergence between measures $\operatorname{P}_0$ and $\operatorname{P}_i$.
+Let $T_i(T) = \sum_{t=1}^T \ind{a_t = do(X_i = 1)}$ be the total number of times the learner intervenes on variable $i$ by setting it to $1$.
+Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $\Ps_0$ and $\Ps_i$ may be bounded using the telescoping property (chain rule) and
+by bounding the local KL divergence by the $\chi$-squared distance as by \citet{Auer1995}. This leads to 
+\eq{
+\KL(\Ps_0, \Ps_i) 
+&\leq 6\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] 
+\leq 6\epsilon^2 \left(\EE_0 T_i(T) + q_i T\right) 
+\leq 6\epsilon^2 \left(\EE_0 T_i(T) + \frac{T}{m}\right)\,.
+}
+Define set $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
+Then for $i \in A$ and choosing $\epsilon = \min\set{1/4, \sqrt{m/(18T)}}$ we have
+\eq{
+\KL(\Ps_0, \Ps_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
+}
+Now $\sum_{i=1}^m \EE_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
+Therefore
+\eq{
+\sum_{i \in A} \Pri{\hat a^*_T \neq a} 
+\geq \sum_{i \in A} \exp\left(-\KL(\Ps_0, \Ps_i)\right) - 1
+\geq \frac{|A|}{e} - 1 
+\geq \frac{m}{2e} - 1\,.
+}
+Therefore there exists an $i \in A$ such that
+$\Pri{\hat a^*_T \neq a^*} \geq \frac{\frac{m}{2e} - 1}{m}$. 
+Therefore if $\epsilon < 1/4$ we have
+\eq{
+R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{m}{18T}}\,.
+}
+Otherwise $m \geq 18T$ so $\sqrt{m/T} = \Omega(1)$ and
+\eq{
+R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{1}{4} \frac{\frac{m}{2e} - 1}{2m} \in \Omega(1) 
+}
+as required.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% GENERAL-GRAPH UPPER BOUND
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Proof of Theorem \ref{thm:general}}\label{sec:thm:general}
+
+\begin{proof}
+First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
+We define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$, $R_{at} = R_a(X_t)$ and $\Pn{a}{.} = \Pns{a}{.}$.
+By definition we have $|Z_{at}| \leq B_a$ and 
+\eq{
+\Var_Q[Z_{at}] 
+\leq \EE_Q[Z_{at}^2] 
+\leq \EE_Q[R_{at}^2]
+= \EEa[R_{at}]
+= \EEa\left[\frac{\Pns{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
+\leq m(\eta)\,.
+}
+Checking the expectation we have
+\eq{
+\EE_Q[Z_{at}] 
+= \EEa \left[Y \ind{R_{at} \leq B_a}\right] 
+= \EEa Y - \EEa \left[Y\ind{R_{at} > B_a}\right] 
+= \mu_a - \beta_a\,,
+}
+where 
+\eq{
+0 \leq \beta_a = \EEa[Y \ind{R_{at} > B_a}] \leq \Pns{a}{R_{at} > B_a}
+}
+is the negative bias. 
+The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
+\eq{
+\beta_a \leq \Pns{a}{R_{at} > B_a} \leq \frac{\EEa[R_{at}]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
+}
+Let $\epsilon_a > 0$ be given by
+\eq{
+\epsilon_a = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
+}
+Then by the union bound and Bernstein's inequality 
+\eq{
+\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} 
+\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
+}
+
+Let $I = \hat{a}^*_T$ be the action selected by the algorithm, $a^* = \argmax_{a \in \calA} \mu_a$ be the true optimal action and recall that $\EE_Q[Z_{at}] = \mu_a - \beta_a$.
+Assuming the above event does not occur we have,
+
+\eq{
+\mu_I \geq \hat \mu_I - \epsilon_I  
+\geq \hat \mu_{a^*} - \epsilon_I  
+\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
+}
+By the definition of the truncation
+we have
+\eq{
+\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
+}
+and
+\eq{
+\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
+}
+Therefore for $C = \sqrt{2} + 4$ we have
+\eq{
+\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
+}
+Therefore
+\eq{
+\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
+}
+as required.
+\end{proof}
+
+\subsubsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}\label{sec:m-equivelence}
+
+\begin{proposition} In the parallel bandit setting,
+$m(\eta^*) \leq 2m(\boldsymbol{q})$.
+\end{proposition}
+
+\begin{proof}
+
+Recall that in the parallel bandit setting,
+
+\eq{
+\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}
+}
+
+Let:
+
+\eq {
+\eta_a = \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}}\frac{1}{2m(\boldsymbol{q})} \text { for } a \in do(X_i = j)
+}
+
+Let $D =\sum_{a\in do(X_i=j)}\eta_a$. From the definition of $m(\boldsymbol{q})$, 
+\eq {
+\sum_{a\in do(X_i=j)} \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}} \leq m(\boldsymbol{q}) \implies D \leq \frac{1}{2}
+}
+
+Let $\eta_a = \frac{1}{2} + (1-D)$ for $a = do()$ such that $\sum_{a \in \calA}\eta_a = 1$
+
+Recall that,
+
+\eq{
+m(\eta) &
+= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]
+}
+
+We now show that our choice of $\eta$ ensures $\EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \leq 2m(\boldsymbol{q})$ for all actions $a$.
+
+For the actions $a: \eta_a > 0$, ie $do()$ and $do(X_i = j):\P{X_i=j}<\frac{1}{m(\boldsymbol{q})}$,
+\eq{
+\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq \EEa\left[\frac{\Pn{a}{X_1...X_N}}{\eta_a\Pn{a}{X_1...X_N}}\right] = \EEa\left[\frac{1}{\eta_a}\right] \leq 2m(\boldsymbol{q})
+}
+
+For the actions $a :\eta_a = 0$, ie $do(X_i=j):\P{X_i=j}\geq\frac{1}{m(\boldsymbol{q})}$,
+\eq{
+\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq & \EEa\left[\frac{\ind{X_i=j}\prod_{k\neq i}\P{X_k}}{(1/2+D)\prod_k \P{X_k}}\right] \\=& \EEa\left[\frac{\ind{X_i=j}}{(1/2+D)\P{X_i = j}}\right]
+\leq  \EEa\left[\frac{\ind{X_i=j}}{(1/2)(1/m(\boldsymbol{q}))}\right] \leq 2m(\boldsymbol{q})
+}
+
+Therefore $m(\eta*) \leq m(\eta) \leq 2m(\boldsymbol{q})$ as required.
+
+\end{proof}
+\
+%in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
+%Recall in that setting that 
+%\eq{
+%\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
+%}
+%For $a = do()$ choose $\eta_a = 1/2$. 
+%Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
+%Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
+%It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.
+
+\chapter{More stuff}
+%\chapter{List of software programs for causal inference}
 
@@
 \item Tetrad.
-\item There has to be a library for the bayesian trees thing
+\item There has to be a library for the Bayesian trees thing
 \item Linear regression (you can do this anywhere)
@@
 
-\chapter{Causality \& Interpretability}
-
-\chapter{Causality \& Fairness}
+Notorious paper Automated Inferences of Criminality using Face Images.
+- What possible actions could we take?
+Ban people with most risky 'face types' from critical jobs?
+But dependent on causal structure, conditioning on the fact that people are in a position to apply for such jobs could reverse the dependency.
+Ones facial appearance is a readily observable characteristic and there is evidence that people implicity make assumptions on the basis of it.
+Consider trials where people with 'more criminal' faces got longer sentances for similar crimes.
+%\chapter{Causality \& Interpretability}
+
+%\chapter{Causality \& Fairness}
+
+Generalisation error weights towards regions where density of x is high (it does not care about minorities).
 
 As machine learning is incorporated into decision making systems that have fundamental impacts on people’s lives, such as in employment, criminal justice, health and financial services, there are increasing concerns over transparency and fairness \cite{WMD, etc}.
-Realisation that machine learning algoithms can be inherit biases from the data we feed into them and the choices those building them make about what variables to enclude, etc.
+Realisation that machine learning algorithms can be inherit biases from the data we feed into them and the choices those building them make about what variables to include, etc.
 
@@
 A fear that advertising themselves as supporting flexible working options would attract candidates who lacked drive and ambition.
-Cultural bais that favours those driven by monitary ambition against those
-
-Stability - a non causal model can't tell us about a simple do type opperation on  a single variable.
+Cultural bias that favours those driven by monetary ambition against those
+
+Stability - a non causal model can't tell us about a simple do type operation on  a single variable.
 
@@
 But what should we condition on?
-\cite{Romie2010} argue that looking at the department level is the correct viewpoint, since this the point at which hiring decions were made.
-However, it could be argued that if one department was attracting a large number of higher quality candidates, the overall size of that deparmtment should be higher and that the results seen at Berkley reflected a bias in favour of male dominated fields.
+\cite{Romie2010} argue that looking at the department level is the correct viewpoint, since this the point at which hiring decisions were made.
+However, it could be argued that if one department was attracting a large number of higher quality candidates, the overall size of that department should be higher and that the results seen at Berkeley reflected a bias in favour of male dominated fields.
 
@@
 
-\section{How does discrimination arise in ML models}
+%\section{How does discrimination arise in ML models}
 \begin{itemize}
@@
 \item deliberate manipulation on the part of the person building the model.
-\item The hardest case is historical disadvantage, creating genuine differences between relevent attributes of groups.
-\item in bias in the label - what happens if you hire minorities but then your staff treat them badly and as a result they underperform or leave.  
+\item The hardest case is historical disadvantage, creating genuine differences between relevant attributes of groups.
+\item in bias in the label - what happens if you hire minorities but then your staff treat them badly and as a result they under-perform or leave.  
 \end{itemize}
 
-\section{Defining Descrimination}
+%\section{Defining Descrimination}
 
@@
 
-\cite{Romei2012} note the relevence of causal inference on discrimintation analysis.
+\cite{Romei2012} note the relevance of causal inference on discrimination analysis.
 
@@
 We will focus on discrete variables for notational simplicity.
-Assume we have an outcome of interest $Y$, protected attributes $X$, other covariates $Z$ and a (potentially stochasitic) model $f$ that maps $\{x, z\}$ to $y_f \in Y$.
+Assume we have an outcome of interest $Y$, protected attributes $X$, other covariates $Z$ and a (potentially stochastic) model $f$ that maps $\{x, z\}$ to $y_f \in Y$.
 For a given model $f$, we have a distribution over the predicted outputs given the inputs, $\P{Y_f|X,Z}$
@@
 
-\begin{definition}{Disparate Treatment}: A model yeilds disparate treatment if people with identical attributes (excuding protected attributes) are treated differently. 
+\begin{definition}{Disparate Treatment}: A model yields disparate treatment if people with identical attributes (excluding protected attributes) are treated differently. 
 \eqn{
@@
 Unfortunately, in general, they conflict with one another, see theorem \ref{thm:disparate_conflict}.
-Any variable we might wish to aviod discriminating on will be correlated to some other measurable covariate $Z$, making it impossible to avoid both disparate treatment and disparate impact.
+Any variable we might wish to avoid discriminating on will be correlated to some other measurable covariate $Z$, making it impossible to avoid both disparate treatment and disparate impact.
 \vspace{.2cm}
@@
 Given covariates $Z$ and protected attributes $X$, a model, $f$, cannot be fair with respect to both disparate impact and disparate treatment unless $\P{Z|X} = \P{Z}$.
-\begin{proof} Assume $f$ yeilds no different treatment, then $\P{Y_f|Z,X} = \P{Y_f|Z}$.
+\begin{proof} Assume $f$ yields no different treatment, then $\P{Y_f|Z,X} = \P{Y_f|Z}$.
 If we additionally assume no disparate impact, $\sum_z{\P{Y_f|Z}\P{Z|x_1}} = \sum_z{\P{Y_f|Z}\P{Z|x_2}} \;\; \forall \set{x_1,x_2}\subset X$.
@@
 
-\section{Addressing discrimination}
-
-Ommititting the protected attribute from the model can increase disparate impact, even if the protected attribute is negatively correlated with the outcome.
+%\section{Addressing discrimination}
+
+Omitting the protected attribute from the model can increase disparate impact, even if the protected attribute is negatively correlated with the outcome.
 
@@
 In both cases we may have theory that tells us the direction or scale of an effect.
-Ie, we know that race itself has no inherent effect on outcomes such as criminality and education.
+IE, we know that race itself has no inherent effect on outcomes such as criminality and education.
 If you have a data set and a collection of such priors, where the data contradicts these priors due to data bias.
@@
 
-human decicion making can evolve.
-
-The vaugeness inherent in writing laws in natural language allows for them to be re-interpreted.
+human decision making can evolve.
+
+The vagueness inherent in writing laws in natural language allows for them to be re-interpreted.
 
 There is a lack of diversity in models.
-(although radomness is in fact inherent in many) - could one leverage this?
-- you would need an estimate of the probability the model would have assigened each instance to a particular group.
+(although randomness is in fact inherent in many) - could one leverage this?
+- you would need an estimate of the probability the model would have assigned each instance to a particular group.
 
@@
 
-\chapter{Causality in Marketing}
+%\chapter*{Random Questions}
+\begin{itemize}
+\item Why not use ratios instead of differences in defining causal effects (is it anything to do with statement in Dawid? What about sensitivity of uncertainty in estimators.
+\end{itemize}
+
+%\chapter{Causality in Marketing}
 Assessing the impact of marketing spend is becoming increasingly important to industry.
@@
 
-\section{Attribution}
+%\section{Attribution}
 The second is attribution.
@@
 
-\section{Econometric Modelling}
+%\section{Econometric Modelling}
 The final key plank businesses apply to estimate how effective their marketing are econometric or mixed media models. These models
 
-In practise these models are typically time series regression models, fitting some measure of sales against marketing spend (by marketing channel, ie TV, radio, search, digital display etc) and other key variables such as pricing, competitor spend and pricing and indicators of the health of the economy and demand for the product in question.
+In practise these models are typically time series regression models, fitting some measure of sales against marketing spend (by marketing channel, IE TV, radio, search, digital display etc) and other key variables such as pricing, competitor spend and pricing and indicators of the health of the economy and demand for the product in question.
 
@@
 
-\chapter{Conclusions (1000 words)}
-
-\section{Open questions}
+\chapter{Conclusion}
+
+%\section{Open questions}
 \paragraph{Cycles} - a huge issue. Not covered by Pearl,Rubin etc.
@@
 \bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{../library}% write the directory to the .bib file
+\bibliography{library}% write the directory to the .bib file
 \end{document}

### 2017-07-01 thesis/offcuts.tex
draft of section on adversarial bandits done

@@
 \chapter*{Random notes}
+
+\section*{Causality and marketing}
+\begin{itemize}
+\item If your sales are entirely online, can you attribute all sales to marketing because; start with 0 sales.
+Spend some dollars on marketing, which leads to customers.
+Additional sales can occur because of customer experience, word of mouth or new advertising, but regardless these additional sales are a consequence of the advertising spend at time 0.
+(what about natural search)?
+\item Do we have any idea how people found out about us (when they purchase our brands?), ie from search, from a friend, etc
+\item The previous arguament doesn't apply when there are physical stores - since people may find out about the product without any advertising in that case. 
+\item Does buying visits through paid search increase the ranking one acheives in natural search. 
+\item would be interesting to analyse natural search position as a function of market share. 
+\end{itemize}
 
@@
 \end{definition}
+
+\todo{connect bandit regret to early stopping of experiment problem}
+The problem of when to stop an experiment early can also be phrased as a bandit problem and analysed with the same techniques.
+XXX et al showed that if use an experimental setup with early stopping, the regret you obtain will be at least twice that you could have achieved had you made each decision sequentially using a bandit approach.
+Of course this neglect any costs that might be associated with the complexity in implementation of the two approaches.
+
+The bandit learning algorithm is sometimes compared to the standard supervised learning in terms of the feedback recieved.
+One way of viewing the standard supervised learning setting is that the label for each instance corresponds to the correct or optimal choice for that setting.
+However, this does not have to be the case.
+In the bandit setting the algorithm observes only the reward for the selected action.
+How are standard supervised learning problems translated to bandit problems for testing algorithms?
+(is it contextual bandit problems?)
+
+Bandit algorithms that require the horizon $T$ to be known in advance are refered to as \emph{fixed horizon} algorithms, as opposed to \emph{anytime} algorithms that do not rely on knowing the horizon.
+A fixed horizon algorithm may be converted to an anytime algorithm via the so called "doubling trick" \citep{}.
+The algorithm starts with some fixed horizon and increases it at a fixed rate everytime the current horizon is reached until the problem terminates.
+\todo{what can be said about the optimality of this?}
+
+In the classical multi-armed bandit problem, the rewards for each arm are sampled stochastically from a fixed (but unknown) distribution.
+There is a large body of work analysing a more general formulation of the bandit problem that relaxes this stochastic assumption.
+
+This lower bound says that for any algorithm for the k-armed bandit problem, there exists a set of reward distributions such that the expected is $\bigomega{\sqrt{kT}}$.
+
+The expectation is with respect the random sequence of actions.

### 2017-07-01 thesis/thesis_map.tex
draft of section on adversarial bandits done

@@
 \newcommand{\KL}{\operatorname{KL}}
-\newcommand{\simpleregret}{R_T}
+
 \newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
@@
 \newcommand{\rawregret}{R_{T}} % plain regret
+\newcommand{\simpleregret}{R_T}
 \newcommand{\quotes}[1]{``#1''}
@@
 \item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
+\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (accross viewpoints)
+\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
+When is a problem causal, what bounds does predictive accuracy give on causal models?
+To what degree is predictive accuracy a metric on which causal models can be judged?
+\item Benchmark estimation algorithms
+\item clarify causal interpretation of dirichlet models
+\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software benchmarking solution?
+\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption.
+For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability.
 \end{enumerate}
@@
 Both the rewards and the actions selected by the algorithm are random variables.
-The majority of work in the bandit liturature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Audibert2007}.
+The majority of work in the bandit liturature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}.
 The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is refered to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
@@
 
-\subsection{Approaches, Regret bounds and results}
+The regret is invariant to adding a constant to the expected rewards for all actions.
+However, it still depends on key characteristics of the reward distributions for each action.
+Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e bernoulli bandits, gaussian bandits) or that the rewards are bounded in some range.
+Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
+
+\todo{consider mentioning Thompson sampling and Bayesian bandits}
+
+\subsection{Stochastic bandits: Approaches and results}
 
@@
 This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
-Instead, we focus on computing upper and lower bounds on the regret.
+Instead, the focus is on computing bounds on the expected regret.
 
@@
 
-Talk about problem dependent and worst case bounds
-
-Lay out what they are for the stochastic bandit setting.
-
-\begin{itemize}
-\item problem dependent versus worst-case bounds
-\item upper and lower bounds
-\end{itemize}
-Here we can talk about the different kinds of bounds that are possible to achieve and what has been done so far (for stochastic k-armed bandits)
-
-The lower bound on the worst case regret for any algorithm (stochastic or adversarial) for the K-armed bandit problem is  $\bigomega{\sqrt{TK}}$ \cite{Auer1995}.
-
-A key algorithm for stochastic bandits, with tractable analysis and strong performance guarantees, is the UCB algorithm \cite{Auer2002}.
-They key to this algorithm is that it keeps track of an upper confidence bound (hence UCB) on the expected reward for each arm and selects the arm with the highest one.
-This balances exploration and exploitation as an arm with a high upper confidence bound must have either a high expected reward or large uncertainty on the expected reward.
-Assume for rotational simplicity that $\mu_1 > \mu_2 > ...
-> \mu_K$, such that $\mu^* = \mu_1$, and let $\Delta_i = \mu_i - \mu^*$ be the sub-optimal for each arm.
-The (problem dependent) regret for UCB is bounded by:
+There are a few key hueristic principles that are used to guide the development of bandit algorithms.
+The first is \emph{optimism in the face of uncertainty}.
+Applied to stochastic bandits, the optimism in the face of uncertainty princple suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
+The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
+Thus selecting it yields either a good reward or useful information.
+
+\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the expontial family.
+They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptopically and proved this is asymptopically efficient.
+However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm.
+\citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$.
+Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds.
+We now assume the rewards are bounded in $[0,1]$.
+The algorithm and regret bounds can be generalised to subgaussian reward distributions, see \citet{Bubeck2012}.
+
+\begin{algorithm}[H]
+\caption{UCB-1}\label{alg:ucb-1}
+\begin{algorithmic}[1]
+\STATE {\bf Input:} horizon $T$.
+\STATE Play each arm once. 
+\FOR{$t \in 1,\ldots,T$}
+\STATE Count the number of times each arm has been selected previously $n_{t,i} = \sum_{s=1}^{t-1} \ind{a_s = i}$
+\STATE Calculate the mean reward for each arm $\hat{\mu}_{t,i} = \frac{1}{n_{t,i}} \sum_{s=1}^{t-1}\ind{a_s = i}Y_t$
+\STATE Select arm $a_t \in \argmax_{i = \set{1,...,k}}\left(\hat{\mu}_{t,i}+\sqrt{\frac{2 \log{t}}{n_{t,i}}}\right)$
+\ENDFOR
+\end{algorithmic}
+\end{algorithm}
+
+Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
+The problem dependent psuedo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
 
 \eqn {
-R^{ucb}(T) \in \bigo{\sum_{i=2}^K \frac{1}{\Delta_i}\log(T)}
-}
-
-This bound blows up as differences $\Delta_i \rightarrow 0$, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
-In the worst case, $R^{ucb}(T) = \bigo{\sqrt{TK\log(T)}}$ \cite{Bubeck2012}.
-Subtle modifications to the UCB algorithm can eliminate the logarithmic term in this worst case regret bound.
-This yields $R^{ucb}(T) = \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.
-
-One key principle that has motivated the design of a number \todo{ask Tor are there actually others aside from UCB?} of key bandit algorithms is the idea of \emph{optimism in the face of uncertainty}.
-An algorithm is trying to simultaneously obtain large rewards and extract as much information as possible.
-If an action has a high upper bound on the possible reward then either the expected reward must be high or the algorithm must be very uncertain about it.
-Choosing it should yield either a good reward or useful information.
-This is the key idea behind the UCB algorithm \citep{xxx}.
-The remaining complexity lies in optimising how we compute confidence intervals to trade of exploration and exploitation as the game proceeds.
-Subtle variations in how the confidence intervals are defined lead to a number of algorithms, which are optimised to slightly different settings.
-SETTINGS NOT QUITE RIGHT WORD.
-\citep{xxx}.
-
-The bandit learning algorithm is sometimes compared to the standard supervised learning in terms of the feedback recieved.
-One way of viewing the standard supervised learning setting is that the label for each instance corresponds to the correct or optimal choice for that setting.
-However, this does not have to be the case.
-In the bandit setting the algorithm observes only the reward for the selected action.
-How are standard supervised learning problems translated to bandit problems for testing algorithms?
-(is it contextual bandit problems?)
-
-Bandit algorithms that require the horizon $T$ to be known in advance are refered to as \emph{fixed horizon} algorithms, as opposed to \emph{anytime} algorithms that do not rely on knowing the horizon.
-A fixed horizon algorithm may be converted to an anytime algorithm via the so called "doubling trick" \citep{}.
-The algorithm starts with some fixed horizon and increases it at a fixed rate everytime the current horizon is reached until the problem terminates.
-\todo{what can be said about the optimality of this?}
-
-Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
+\label{eqn:ucb_problem_dependent_regret}
+\regret \leq & \sum_{i:\Delta_i > 0}\left(\frac{8\log(T)}{\Delta_i}+2 \right)
+}
+
+Somewhat unintuitively, the regret increases as the value of the arms gets closer together.
+This is because it becomes harder for the algorithm to identify the optimal arm.
+As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
+The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
+The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which ia maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
+By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.
+
+\eqn{
+\label{eqn:ucb_worst_case_regret}
+\regret \in \bigo{\sqrt{kT\log(T)}}
+}
+
+\begin{figure}
+\caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
+The solid curve shows the mean (cummulative) regret, over a 1000 simulations for a 2-armed, bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
+The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
+\label{fig:ucb_regret_vs_delta}
+\includegraphics[scale=1]{ucb_regret_vs_delta}
+\end{figure}
+
+The form of the dependence on the number of arms $k$ and horizon $T$ differs between the problem dependent and worst case regret.
+The problem dependent regret grows linearly with the number of arms, $k$, and logarithmically with $T$.
+The difference stems from the fact the problem dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
+\citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by,
+
+\eqn{
+\label{eqn:k-arm-bandit-lower-bound}
+\regret \in \bigomega{\sqrt{kT}}
+}
+
+Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
+This yeilds regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.
+
+\begin{figure}
+\caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$.
+The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $bernoulli([.3,.7])$ over 1000 simulations.
+The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
+\label{fig:ucb_upperbound_vs_actual_performance}
+\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
+\end{figure}
+
+\subsection{Pure-exploration problems}
+Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
 In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
@@
 \label{eqn:simple_regret}
-\mu_{i^*} - \E{\mu_{\hat i^*}}.
+\simpleregret=\mu_{i^*} - \E{\mu_{\hat i^*}}.
 }
@@
 For example, many strategies might be assessed via simulation prior to one being selected and deployed.
-The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}.
-
-\todo{connect bandit regret to early stopping of experiment problem}
-The problem of when to stop an experiment early can also be phrased as a bandit problem and analysed with the same techniques.
-XXX et al showed that if use an experimental setup with early stopping, the regret you obtain will be at least twice that you could have achieved had you made each decision sequentially using a bandit approach.
-Of course this neglect any costs that might be associated with the complexity in implementation of the two approaches.
-
-\subsection{Key related settings}
-
-\subsubsection{Adversarial Bandits}
-In the classical multi-armed bandit problem, the rewards for each arm are sampled stochastically from a fixed (but unknown) distribution.
-There is a large body of work analysing a more general formulation of the bandit problem that relaxes this stochastic assumption.
+The worst case simple regret for a k-armed bandit is lower bounded by equation \ref{eqn:worst_case_simple_regret_lower_bound} (\citep{Bubeck2009}).
+
+\eqn{
+\label{eqn:worst_case_simple_regret_lower_bound}
+\simpleregret \in \bigo{\sqrt{K/T}}
+}
+
+\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
+
+Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds. The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period.
+
+\subsection{Adversarial Bandits}
 
 Adversarial bandits are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
-Instead, simultaneously with the learner selecting an action $I_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{g}_t$.
-As in the stochastic setting, the learner then receives reward $g_{I_t,t}$.
-The seminal algorithm for adversarial bandits is Exp-3, which, like UCB, obtains regret $\bigo{\sqrt{TK\log(T)}}$ regret \cite{Auer1995}.
-Optimal algorithms, with $R(T) = \bigo{\sqrt{TK}}$, have also been demonstrated for the adversarial setting \cite{Audibert2009}.
-
-\paragraph*{Definition of the Adversarial bandit problem}
-\begin{itemize}
-\item Different notation of regret
-\item Adversarial bandits are not immune to issues of feedback cycles or drift in the reward distribution because a constancy is implicit in the definition of regret. 
-\end{itemize}
-
-\subsubsection{Contextual bandits}
+Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$.
+As in the stochastic setting, the learner then receives reward only for the selected action.
+
+\vspace{0.5cm}
+\begin{definition}[Adversarial k-armed bandit problem]
+Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
+\begin{enumerate}
+\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
+\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
+\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\end{enumerate} 
+\end{definition}
+
+Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivous}, as opposed to \emph{non-oblivous} adversaries, which can generate rewards as a function of the history of the game.
+In the case of oblivous adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
+
+The pseudo-regret for adversarial bandits is defined, in the same way as for stochastic bandits, as the difference between the reward obtained by playing the single arm with the highest expected reward in every round and the expected reward obtained by the algorithm.
+
+\eqn{
+\label{eqn:pseudo_regret-adversarial}
+\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+}
+
+The policy (or algorithm) used by the learner is avaiable to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
+This implies the adversary can ensure that any learner with a deterministic policy suffers regret $\bigo{T}$ by forecasting their entire sequence of actions.
+For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forcasts what action the learner will play in round 2, given they recieved a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
+This implies adversarial bandit policies must be sufficiently random to avoid such exploitation \footnote{The UCB algorithm, defined by algorithm \ref{alg:ucb-1}, is deterministic if the order in which arms are played during the first $k$ rounds is fixed and the method for selecting which arm to play when multiple-arms have the same upper-confidence bound is not-random (for example, select the arm one with the lowest index $i$).}
+
+The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case psuedo-regret of $\bigo{\sqrt{TK\log(T)}}$  \cite{Auer1995}.
+Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \cite{Audibert2009}.
+The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
+However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nontheless obtain good problem dependent regret in more favourable enviroments \citep{bubeck2012bothworlds,Auer2016}.
+
+Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
+However, psuedo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, is not a good measure of an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
+In settings were the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the psuedo-regret do not reflect how sub-optimal the algorithm may be.
+Adversarial bandit algorithms may perform better in non-stationary settings than standard stochastic policies to the extend that they explore more (to avoid the adversary simulating their behaviour), however it is preferable to develop algorithms specifically for non-stationary settings (subject to assumptions about how rapidly or frequently rewards can change), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
+
+\begin{figure}
+  \caption{The pseudo-regret can be negative if rewards are non-stationary.
+This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
+Figure (a) shows the expected rewards of each arm, and the proportion of time that arm-1 is played, as a function of time.
+The single best-arm is arm-1 as it has the highest expected reward (averaged over $t$).
+An oracle that selects arm-1 in every round obtains an expected reward of $5,000$.
+However, despite not being designed to do so, the UCB-algorithm can adapt to the changing reward distribution to obtain costistently higher rewards.
+The distribution of regret over the 1000 simulations is shown in figure (b).}\label{fig:non-stationary-bandit}
+    \begin{subfigure}[t]{0.49\textwidth}
+	\centering   
+	 \caption{Behaviour over time}\label{fig:non-stationary-bandit-time} 
+         \includegraphics[scale=.5]{non-stationary-bandit-time}
+\end{subfigure}
+    \begin{subfigure}[t]{0.49\textwidth}
+    \centering
+     \caption{Cummulative regret distribution}
+        \label{fig:non-stationary-bandit-regret}
+        \includegraphics[scale=.5]{non-stationary-bandit-regret}
+    \end{subfigure}
+\end{figure}
+
+\subsection{Contextual bandits}
 Contextual bandits are a generalisation of multi-armed bandits.
@@
 
-\subsubsection{Markov decision processes}
+\subsection{Markov decision processes}
 
@@
 
-This is reasonable in settings where we have many similar but independent units or individuals and the goal is to identify which action is optimal (in a given context). However
+This is reasonable in settings where we have many similar but independent units or individuals and the goal is to identify which action is optimal (in a given context), as occurs in standard experimental settings as well.
+
+Relation to SUTVA assumption, selecing a action for one 'unit' does not effect the reward distributions for other units.
 
@@
 \item Why does anything solving a regression problem work in the anti-causal direction - it would seem to break the fundamental underlying assumption of additive noise?
+\item is there a benchmark library of implementations of key bandit algorithms?
+\item Bubeck 2012 states that if the adversary is oblivious, the psuedo-regret coincides with the standard regret, 'which is always the ultimate quantity of interest'.
+Why, they are not the same for stochastic bandits are they?
+\item what is the worst-case regret of UCB in an adversarial setting?
+\item is there a concept of problem dependent regret for adversarial bandits? Has there been analysis of the problem dependent regret for key adversarial algorithms such as EXP-3?
+\item adversarial regret bounds for Thompson sampling, UCB with random arm selection from upper-bound set?
+\item do continuous armed bandits implie that there are arms with any arbitrary gap in reward (including worst case gap for that T) and thus you always end up with something like worst case regret?)
+\item in what real world setting do adversarial bandits potentially make sense?
 \end{enumerate}

### 2017-07-01 causalpy/causalpy/bandits/algorithms.py

@@ -94,13 +94,18 @@ class ThompsonSampling(object):
     def run(self,T,model):
         self.trials = np.full(model.K,2,dtype=int)
         self.success = np.full(model.K,1,dtype=int)
+        self.rewards = np.zeros(T)
+        self.arms = np.zeros(T)
         
         for t in xrange(T):
             fails = self.trials - self.success
             theta = np.random.beta(self.success,fails)
             arm = argmax_rand(theta)
             self.trials[arm] +=1
-            self.success[arm]+= model.sample_multiple(arm,1)
+            reward = model.sample_multiple(arm,1)
+            self.success[arm]+= reward
+            self.rewards[t] = reward
+            self.arms[t] = arm
         
         mu = np.true_divide(self.success,self.trials)
         self.best_action = argmax_rand(mu)
@@ -113,25 +118,45 @@ class UCB(object):
     def run(self,T,model):
         if T <= model.K: # result is not defined if the horizon is shorter than the number of actions
             self.best_action = None
+            self.rewards = None
             return np.nan
+            
+        self.rewards = np.zeros(T)
+        self.arms = np.zeros(T)
         
         actions = range(0,model.K)
         self.trials = np.ones(model.K)
         self.success = model.sample_multiple(actions,1)
+        self.arms[0:model.K] = np.asarray(actions)
+        self.rewards[0:model.K] = self.success
         
+        best_reward = max(model.expected_rewards)
+       
         for t in range(model.K,T):
             arm = argmax_rand(self.upper_bound(t))
             self.trials[arm] += 1
-            self.success[arm] +=model.sample_multiple(arm,1)
+            reward = model.sample_multiple(arm,1)
+            self.success[arm] += reward
+            self.arms[t] = arm
+            self.rewards[t] = reward
+            #self.regret[t] = best_reward - model.expected_rewards[arm]
         
         mu = np.true_divide(self.success,self.trials)
         self.best_action = argmax_rand(mu)
-        return max(model.expected_rewards) - model.expected_rewards[self.best_action]
+        return best_reward - model.expected_rewards[self.best_action]
+    
+    
+        
+    
+        
         
         
 
 class AlphaUCB(UCB):
-    """ Implementation based on ... """
+    """ 
+    Implementation based on ... 
+    UCB-1 corresponds to setting alpha = 4
+    """
     label = "UCB"
     
     def __init__(self,alpha):

### 2017-07-01 causalpy/causalpy/bandits/enviroments.py

@@ -40,6 +40,27 @@ def prod_all_but_j(vector):
     joint = np.prod(vector)
     return np.true_divide(joint,vector)
     
+class TimeVaryingBernoulliModel(object):
+    """ 
+    A model that shifts the rewards for each arm linearly,
+    as a function of the time-step".
+    """
+    def __init__(self,initial_expected_rewards,final_expected_rewards,T):
+        self.K = len(initial_expected_rewards)
+        self.t = 0
+        self.slope = (final_expected_rewards - initial_expected_rewards)/float(T)
+        self.intercept = initial_expected_rewards
+        self.expected_rewards = self.expected_rewards_t(0)
+    
+    def expected_rewards_t(self,t):
+        return self.intercept + self.slope*t
+        
+    def sample_multiple(self,actions,n):
+        p = self.expected_rewards[actions]
+        self.t += 1
+        self.expected_rewards = self.expected_rewards_t(self.t)
+        return binomial(n,p)
+    
 class BasicBernoulliModel(object):
     """ A model for generating data for standard bandits without side information """
     
@@ -47,7 +68,7 @@ class BasicBernoulliModel(object):
         """
         expected_rewards: an array of length K containing the expected rewards for each arm.
         """
-        self.expected_rewards = expected_rewards
+        self.expected_rewards = np.asarray(expected_rewards)
         self.K = len(self.expected_rewards)
                 
     @classmethod

### 2017-07-01 causalpy/examples/two_armed_bernoulli_bandit_usb.py

@@ -1,0 +1,41 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jun 29 14:39:57 2017
+
+@author: finn
+"""
+
+from causalpy.bandits import BasicBernoulliModel, AlphaUCB
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+simulations = 1000
+T = 10
+
+delta = .3
+expected_rewards = [.5-delta/2.0,.5+delta/2.0]
+enviroment = BasicBernoulliModel(expected_rewards)
+algorithm = AlphaUCB(4.0)
+
+algorithm.run(T,enviroment)
+regret = algorithm.get_cumulative_regret()
+
+#for i,T in enumerate(T_vals):
+#    enviroment = BasicBernoulliModel.create_epsilon_best(K,.1)
+#    best_reward = np.empty(simulations)
+#    expected_best_is_best = 0
+#    for s in xrange(simulations):
+#        rewards = enviroment.sample_multiple(np.arange(0,K),T)
+#        best_arm = rewards.argmax()
+#        best = rewards[best_arm]
+#        best_reward[s] = best
+#        expected_best_is_best += int(best_arm == (K-1))
+#        
+#    
+#    max_expected_reward = enviroment.expected_rewards.max()*T  
+#    average_max_reward = best_reward.mean()
+#    result[i,0] = max_expected_reward
+#    result[i,1] = average_max_reward
+#    result[i,2] = expected_best_is_best/float(s)
+    

### 2017-07-06 thesis/offcuts.tex
mostly complets contextual bandits section

@@
 The expectation is with respect the random sequence of actions.
+
+.
+Although, in this case, the contextual bandit algorithm should eventually converge to the selecting the same optimal arm as a standard multi-armed bandit algorithm that ignored the context the regret incured will be higher.
+
+Note that we can't simply apply a (online) supervised learning algorithm because the samples we recieve depend on the choice of action.
+
+The regret bounds how well we can do relative to the best hypothosis in the space.
+Increasing the size of the hypothosis space increases the upper bound on the regret, as the algorithm has a larger space to explore.
+Connected to the bias variance trade-off.
+
+In a finite experiment, we can achieve lower 'regret' by allowing the algorithm to search only over a simpler hypothosis space.
+
+The results we have stated only apply to finite hypthosis classes.
+However the XXX algorithms can be generalised to continuous hypothosis classes to obtain regret bounds in terms of VC-dimension \citep{}.
+
+It should be noted that there is an important difference between the approaches we can apply to the contextual bandit versus the supervised learning setting.
+As for the casual effect estimation problem, we cannot utilise cross-validation offline to select model parameters.
+This makes contextual bandit algorithms degrade faster than supervised learning ones as we add irrelevant variables.
+\todo{ask Tor if he thinks this is correct}
+
+RUN A SIMULATION COMPARING HOW supervised learning and contextual bandits degrade as we add more features. What would be a fair comparison?

### 2017-07-06 thesis/thesis_map.tex
mostly complets contextual bandits section

@@
 
-\chapter*{Todo}
+\chapter*{To-do}
 \begin{enumerate}
 \item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
-\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (accross viewpoints)
+\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
 \item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
@@
 \item Benchmark estimation algorithms
-\item clarify causal interpretation of dirichlet models
-\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software benchmarking solution?
+\item clarify causal interpretation of Dirichlet models
+\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench-marking solution?
 \item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption.
@@
 
-This implies the counterfactual distributions are equivelent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
+This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
 
@@
 
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorebility assumptions \citep{Rosenbaum1983}.
+The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}.
 They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
@@
 Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$.
-A more general form of the ignorebility assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
+A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
 
@@
 \caption{
-Parametric regression may yeild poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
+Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
 In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
@@
 However in general, it must be estimated from data.
-Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and nueral networks can also be applied   \citep{Austin2011}.
+Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
 \citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
@@
 
-The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding exeriments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
+The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
 Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
@@
 Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from.
-Selecting data points optimally for learning is the focus of the optimal experimental design liturature within statistics \citep{pukelsheim2006optimal} and the active learning liturature in machine learning \citep{settles2010active}.
+Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}.
 Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias.
@@
 
-Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates belived to influence the outcome of interest.
-This introduces a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, nessessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
+Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest.
+This introduces a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
 The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
-See \citet{imbens2015causal} for a discussion of the trade-offs between statified versus completely random experiments.
+See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.
+
+IF THE PROPENSITY SCORE IS THE SAME FOR ALL COVARIATES IT DOESN'T SEEM ONE MUST ADJUST.
+\todo{double check}
 
@@
 
-It is not always possible or ethical to conduct a randomized controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parashute use for the reduction of the mortality and morbity associated with falls from large hights \ref{fig:parashute}.
+It is not always possible or ethical to conduct a randomised controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
 When experimentation is possible, it is frequently difficult or expensive.
@@
 \begin{center}
-\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parashutes for the prevention of morbitity and mortality associated with falls from large heights.}
+\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parachutes for the prevention of morbidity and mortality associated with falls from large heights.}
 \label{fig:parashute}
@@
 
-Multi-armed bandits address the problem of desiging experiments that can adapt as samples are observed.
+Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed.
 Their introduction is generally attributed to \citet{Thompson1993}.
@@
 \footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not nessessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
+It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
 The goal of the learner is to maximise the total reward they receive.
@@
 \item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{Y}}$
-\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
 \end{enumerate}
@@
 
-The total reward a bandit algorithm/policy can expect to acheive depends on the distributions from which the rewards for each action are sampled.
+The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
 To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
-This difference is known as the (cummulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning liturature.
+This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
 There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.},
+In addition, researches often refer to the expected regret as \quotes{the regret}.}.
 \eqn{
@@
 Both the rewards and the actions selected by the algorithm are random variables.
-The majority of work in the bandit liturature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}.
-The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is refered to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
-A stochastic bandit algorithm is learning if it obtains psuedo-regret that is sub-linear in $T$.
+The majority of work in the bandit literature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}.
+The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is referred to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
+A stochastic bandit algorithm is learning if it obtains pseudo-regret that is sub-linear in $T$.
 
@@
 However, it still depends on key characteristics of the reward distributions for each action.
-Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e bernoulli bandits, gaussian bandits) or that the rewards are bounded in some range.
+Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range.
 Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
 
-\todo{consider mentioning Thompson sampling and Bayesian bandits}
-
 \subsection{Stochastic bandits: Approaches and results}
@@
 The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators.
-The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards.
+The action selected by an algorithm at a given time-step can depend on the history of previous actions and rewards.
 As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
-The expectation and variance guarentees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
+The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
 This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
@@
 \caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples.
-This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim bernoulli(0.2)$.
+This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim Bernoulli(0.2)$.
 In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$.
@@
 
-There are a few key hueristic principles that are used to guide the development of bandit algorithms.
-The first is \emph{optimism in the face of uncertainty}.
-Applied to stochastic bandits, the optimism in the face of uncertainty princple suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
+There are a few key principles that are used to guide the development of bandit algorithms.
+The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
+A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
+This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.
+
+Another key approach is \emph{optimism in the face of uncertainty}.
+Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
 The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
@@
 
-\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the expontial family.
-They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptopically and proved this is asymptopically efficient.
-However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm.
+\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family.
+They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient.
+However, their algorithm is complex and memory intensive to compute as, at each time-step, it relies on the entire sequence of rewards for each arm.
 \citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$.
@@
 Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
-The problem dependent psuedo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
+The problem dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
 
@@
 The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
-The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which ia maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
+The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every time-step, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which ia maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
 By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.
@@
 \caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
-The solid curve shows the mean (cummulative) regret, over a 1000 simulations for a 2-armed, bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
+The solid curve shows the mean (cumulative) regret, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
 The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
@@
 Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
-This yeilds regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.
+This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.
 
@@
 
+Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal.
+This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}.
+Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
+However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
+
 \subsection{Pure-exploration problems}
@@
 
-\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
+\todo{discuss the key ways in which the simple regret problem differs from cumulative regret}
 
@@
 
-Adversarial bandits are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
+Adversarial bandits, described by \citet{Auer1995}, are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
 Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$.
@@
 \begin{definition}[Adversarial k-armed bandit problem]
+\label{def:adversarial-k-armed-bandit}
 Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
@@
 \item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
 \end{enumerate} 
@@
 
-Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivous}, as opposed to \emph{non-oblivous} adversaries, which can generate rewards as a function of the history of the game.
-In the case of oblivous adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
+Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
+In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
 
@@
 
-The policy (or algorithm) used by the learner is avaiable to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
+The policy (or algorithm) used by the learner is available to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
 This implies the adversary can ensure that any learner with a deterministic policy suffers regret $\bigo{T}$ by forecasting their entire sequence of actions.
-For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forcasts what action the learner will play in round 2, given they recieved a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
+For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forecasts what action the learner will play in round 2, given they received a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
 This implies adversarial bandit policies must be sufficiently random to avoid such exploitation \footnote{The UCB algorithm, defined by algorithm \ref{alg:ucb-1}, is deterministic if the order in which arms are played during the first $k$ rounds is fixed and the method for selecting which arm to play when multiple-arms have the same upper-confidence bound is not-random (for example, select the arm one with the lowest index $i$).}
 
-The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case psuedo-regret of $\bigo{\sqrt{TK\log(T)}}$  \cite{Auer1995}.
+The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case pseudo-regret of $\bigo{\sqrt{TK\log(T)}}$  \cite{Auer1995}.
 Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \cite{Audibert2009}.
 The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
-However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nontheless obtain good problem dependent regret in more favourable enviroments \citep{bubeck2012bothworlds,Auer2016}.
+However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nonetheless obtain good problem dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
 
 Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, psuedo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, is not a good measure of an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings were the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the psuedo-regret do not reflect how sub-optimal the algorithm may be.
+However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, is not a good measure of an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
+In settings were the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
 Adversarial bandit algorithms may perform better in non-stationary settings than standard stochastic policies to the extend that they explore more (to avoid the adversary simulating their behaviour), however it is preferable to develop algorithms specifically for non-stationary settings (subject to assumptions about how rapidly or frequently rewards can change), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
 
-\begin{figure}
+\begin{figure}[h]
   \caption{The pseudo-regret can be negative if rewards are non-stationary.
-This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
+This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed Bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
 Figure (a) shows the expected rewards of each arm, and the proportion of time that arm-1 is played, as a function of time.
@@
 An oracle that selects arm-1 in every round obtains an expected reward of $5,000$.
-However, despite not being designed to do so, the UCB-algorithm can adapt to the changing reward distribution to obtain costistently higher rewards.
+However, despite not being designed to do so, the UCB-algorithm can adapt to the changing reward distribution to obtain consistently higher rewards.
 The distribution of regret over the 1000 simulations is shown in figure (b).}\label{fig:non-stationary-bandit}
@@
     \centering
-     \caption{Cummulative regret distribution}
+     \caption{Cumulative regret distribution}
         \label{fig:non-stationary-bandit-regret}
@@
 \subsection{Contextual bandits}
-Contextual bandits are a generalisation of multi-armed bandits.
+\label{sec:contextual-bandits}
+
+In the standard multi-armed bandit setting, each decision is identical and the goal is to learn a single best action.
+However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
+The best treatment to offer an individual patient could depend on their age, gender, disease sub type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
+Similarly, decisions on which ad or content to display on a web page or which product to recommend can be \emph{personalised} based on the previous behaviour of the user.
+A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
+Contextual bandits are a generalisation of multi-armed bandits that make use of this additional contextual information.
+The term contextual bandit was coined by \citet{Langford2008}, however close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
 
 \vspace{0.5cm}
-\begin{definition}[Contextual bandit problem]
+\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analogously to definition \ref{def:adversarial-k-armed-bandit}}]
+\label{def:contextual-bandit-problem}
 Let $\P{\vec{Y},\vec{X}}$ be the joint distribution over the rewards for each action and some context $\vec{x} \sim \mathcal{X}$. In each round $t \in \set{1,...T}$, 
@@
 \item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps.
-\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
 \end{enumerate}
@@
 
-What are the reductions of contextual bandits to other problem (ie k armed-bandits?)
-
-In the previous sections each decision was identical.
-If we were imaging treating patients, we assume that we have no additional information about each patient that is relevant to deciding how to treat them.
-However, in many key applications such as serving ads, etc, etc.
-we clearly do have additional information that we wish to take into account.
-In the contextual bandit setting the goal is to learn
-
-\eq {
-\P{Y|a,x}
-}
-
-As opposed to the standard bandit setting where we are attempting to estimate
-
-\eq{
-\P{Y|a}
-}
-
-The formal definition of a contextual bandit problem is
-
-The simplest way to generalise the standard bandit algorithms to the contextual case (where the context is discrete) is to simply have a separate standard bandit instance for each possible setting of the context.
-Regrettably this approach scales terribly with increasing complexity of the context.
-\todo{quantify this}.
-It cannot be applied at all if the context is continuous.
-
-The approach to solving this problem parallels those from supervised learning and continuous armed bandits.
-We make assumptions about the smoothness of the problem.
-Values of context that are similar should lead to comparable rewards for a given action.
-
-Regret bounds \todo {insert regret bounds for contextual bandits}
-
-It should be noted that there is an important difference between the approaches we can apply to the contextual bandit versus the supervised learning setting.
-As for the casual effect estimation problem, we cannot utilise cross-validation offline to select model parameters.
-This makes contextual bandit algorithms degrade faster than supervised learning ones as we add irrelevant variables.
-\todo{ask Tor if he thinks this is correct}
-
-RUN A SIMULATION COMPARING HOW supervised learning and contextual bandits degrade as we add more features. What would be a fair comparison?
+Standard multi-armed bandits learn to select the action $a$ that, with high probability, maximises $\E{Y|a}$.
+Contextual bandits learn to select actions that maximise $\E{Y|\vec{x},a}$.
+The reward for contextual bandits should  be compared to an oracle that acts optimally based on the context.
+To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothesis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
+The pseudo-regret is the difference between the expected regret obtained by an oracle that selects actions based on the single best hypothesis or policy $h$ at each time-step, and the expected reward obtained by the algorithm.
+
+\eqn{
+\label{eqn:regret_contextual}
+\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{Y_{t,h(\vec{x}_t)}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+}
+
+If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a separate standard bandit instance for each value of the context.
+This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothesis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occurred such that $\sum_{c=1}^N T_c = T$.
+The regret is maximised if $T_c = T/N$ resulting in total regret $\bigo{N\sqrt{kT/N}}$.}.
+This is optimal with respect to this class of hypothesis.
+However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the in the relationships between different contexts and actions.
+As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
+For example, with $n$ binary variables, $N = 2^n$}.
+We expect some form of smoothness; values of context that are similar should lead to comparable rewards for a given action.
+We need algorithms that can leverage such assumptions.
+
+An alternate reduction to the standard bandit problem, which allows us to constrain the hypothesis space to explore, is to treat each hypothesis $h$ as a bandit arm \citep{Langford2008}.
+At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
+Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothesis class considered, limiting our ability to learn any complex mappings from context to actions.
+The key problem with this approach is each sample is used to update our knowledge about only one hypothesis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothesis simultaneously.
+
+Suppose that, at each time-step $t$, after selecting an action, the learner received the reward for chosen action but observed the full vector of rewards $[Y_{t,1},...,Y_{t,k}]$.
+This is known as the full information setting.
+In this case, the learner can simulate running each hypothesis over the history to compute the reward it would have obtained and use the hypothesis with the best empirical reward to select the next action.
+This is the \emph{follow the leader} algorithm, which obtains optimal regret $\bigo{\sqrt{T log(\card{\mathcal{H}})}}$ for the full-information problem \citep{cesa2006prediction}.
+Unfortunately, in the contextual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
+As in causal effect estimation, we can view this as a missing data problem.
+However, the data is missing not at random because which component of the reward is observed depends on the action selected which in turn is a function of the previous history of actions and rewards.
+
+The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly separating exploration from exploitation.
+Epoch-greedy is an explore-exploit algorithm.
+It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothesis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
+The hypothesis with the highest empirical reward is then used to select actions for the remaining time-steps.
+The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
+
+The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
+However, it involves maintaining a list of weights for each hypothesis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
+The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
+
+Both Epoch-greedy and ILOVECONBANDITS involve solving problems of the form,
+
+\eqn{
+\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{x}_t) = a_t}
+}
+
+This expression equates to identifying the best empirical policy based on previous data.
+The algorithms assume the existence of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
+The computational tractability of these algorithms on large (or infinite) hypothesis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classification problem \citep{Dudik2011a}.
+
+Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalised versions of the UCB or Thompson sampling algorithms.
+For linear pay off models, both approaches yield algorithms with strong regret guarantees, Lin-UCB \citep{li2010contextual} and generalised Thompson sampling for linear pay-offs \citep{Agrawal2013}.
+
+It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
+\citet{Bareinboim2015} demonstrate that policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information.
+This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
+However, the context should be relevant, such that $\P{\vec{Y}|\vec{X}} \neq \P{\vec{Y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
+
+\begin{figure}[h]
+\centering
+\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
+$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}_a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
+Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
+To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $a_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},a,Y)_1^{t-1}$.
+}
+\label{fig:contextual-bandit-causal-graph}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$\vec{X}$ causes $\vec{Y}$}
+\label{fig:context-graph-causal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$\vec{Y}$};
+\node[main node, below=of 1](2){$Y$};
+\node[main node, above right=of 2](3){$a$};
+\node[main node, left=of 1](4){$\vec{X}$};
+ \path[every node/.style={font=\sffamily\small}]
+   (4) edge (1)
+   (3) edge (2)
+   (1) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$\vec{Y}$ causes $\vec{X}$}
+\label{fig:context-graph-anti-causal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$\vec{Y}$};
+\node[main node, below=of 1](2){$Y$};
+\node[main node, above right=of 2](3){$a$};
+\node[main node, left=of 1](4){$\vec{X}$};
+ \path[every node/.style={font=\sffamily\small}]
+   (1) edge (4)
+   (3) edge (2)
+   (1) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$\vec{X}$ and $\vec{Y}$ are confounded}
+\label{fig:context-graph-confounded}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$\vec{Y}$};
+\node[main node, below=of 1](2){$Y$};
+\node[main node, above right=of 2](3){$a$};
+\node[main node, left=of 1](4){$\vec{X}$};
+\node[lt, below=of 4](5){$U$};
+ \path[every node/.style={font=\sffamily\small}]
+   (3) edge (2)
+   (1) edge (2)
+   (5) edge (4) edge (1);
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
 
 \subsubsection{Learning from logged bandit data}
+Another topic of interest within the contextual bandit community that is deeply related to causal effect estimation from observational data is learning from logged bandit feedback data \citep{}.
+
+If no assumptions are made about the behaviour of the hypothesis that generated the data.... does it reduce to causal effect estimation?
+
+However, in the applications considered in these setting such as ad serving, the details of the hypothesis generating the data are often known.
 
@@
 
+\subsubsection{Limitations of bandit problems}
+\begin{itemize}
+\item Reward needs to be observed immediately after action is taken (or at least before the next action must be selected).
+\item State assumed not to change
+
+\end{itemize}
+
+Reward known immediately. Action chosen for one individual does not effect reward for the next (SUTVA).
+
 A key limitation of the bandit setting is that
@@
 
-Relation to SUTVA assumption, selecing a action for one 'unit' does not effect the reward distributions for other units.
+Relation to SUTVA assumption, selecting a action for one 'unit' does not effect the reward distributions for other units.
 
@@
 
+causal bandits capture a very natural form of structure that can arise between different arms.
+
 The classic multi-armed bandit is a powerful tool for sequential decision making.
@@
 \begin{enumerate}
+\item In the $\epsilon$-greedy approach, does one compute best action based only on exploration steps?
+\item Why does the naive reduction appear to get sub-optimal regret?
 \item When does the expected regret make sense as the thing we actually would wish to optimise?
-Is optimising the expected regret any different to optimising the psuedo-regret?
+Is optimising the expected regret any different to optimising the pseudo-regret?
 Who is the definition of regret down to?
@@
 \item Is there a reference for the doubling trick? Is it an automatic way of converting a fixed horizon algorithm to an anytime one?
-\item Why is the dependence on the number of arms $K$ in the problem dependent bounds but $\sqrt{K}$ in the worst case bounds?
-\item rewards in the multi-armed bandit setting can be sampled from some joint distribution?
 \item Is Learning from logged feedback data just covariate shift too?
@@
 \item is there a benchmark library of implementations of key bandit algorithms?
-\item Bubeck 2012 states that if the adversary is oblivious, the psuedo-regret coincides with the standard regret, 'which is always the ultimate quantity of interest'.
+\item Bubeck 2012 states that if the adversary is oblivious, the pseudo-regret coincides with the standard regret, 'which is always the ultimate quantity of interest'.
 Why, they are not the same for stochastic bandits are they?
@@
 \item adversarial regret bounds for Thompson sampling, UCB with random arm selection from upper-bound set?
-\item do continuous armed bandits implie that there are arms with any arbitrary gap in reward (including worst case gap for that T) and thus you always end up with something like worst case regret?)
+\item do continuous armed bandits implies that there are arms with any arbitrary gap in reward (including worst case gap for that T) and thus you always end up with something like worst case regret?)
 \item in what real world setting do adversarial bandits potentially make sense?
+\item Does supervised learning really 'reveal the value of every action'.
 \end{enumerate}
@@
 In each season the farmer has enough time and money to intervene and control at most one of these
-variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
+variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry.
 When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
@@
 A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
-Each action $a \in \actionspace$ assings a value to a subset of the variables in $\mathcal{X}$.
+Each action $a \in \actionspace$ assigns a value to a subset of the variables in $\mathcal{X}$.
 One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
@@
 A variable may be observed both prior to and after the agent selects an action, and the action may change its value.
-The objective of the learner is to minimise either the simple (equation \ref{}) or cummulative regret (equation \ref{}).
+The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
 
@@
 
-The causal bandit problem takes on characteristics of different bandit settings depending on the actionspace $\actionspace$, which variables are observable prior to selecting an action and on which variables we recieve post-action feedback.
-If feedback is recieved only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
-If we recieve post-action feedback on additional nodes the problem can be more interesting.
+The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
+If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
+If we receive post-action feedback on additional nodes the problem can be more interesting.
 In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
@@
 There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
+Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on un played actions are revealed according to a feedback graph.
 As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
@@
 The partial monitoring approach taken by \cite{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
-\citet{Yu2009} utilize extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
+\citet{Yu2009} utilise extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
 \citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled.
@@
 But dependent on causal structure, conditioning on the fact that people are in a position to apply for such jobs could reverse the dependency.
-Ones facial appearance is a readily observable characteristic and there is evidence that people implicity make assumptions on the basis of it.
+Ones facial appearance is a readily observable characteristic and there is evidence that people implicitly make assumptions on the basis of it.
 Consider trials where people with 'more criminal' faces got longer sentances for similar crimes.

### 2017-07-06 thesis/thesis_map.tex
adds to chapter 4

--- 2017-07-01 thesis/thesis_map.tex
@@
 
+IF THE PROPENSITY SCORE IS THE SAME FOR ALL COVARIATES IT DOESN'T SEEM ONE MUST ADJUST.
+\todo{double check}
+
 \begin{figure}
@@
 There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.},
+In addition, researches often refer to the expected regret as \quotes{the regret}.}.
 \eqn{
@@
 Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
-
-\todo{consider mentioning Thompson sampling and Bayesian bandits}
 
@@
 
-There are a few key hueristic principles that are used to guide the development of bandit algorithms.
-The first is \emph{optimism in the face of uncertainty}.
+There are a few key principles that are used to guide the development of bandit algorithms.
+The simplest is to explicitly seperate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
+A common example in practice is unform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
+This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.
+
+Another key approach is \emph{optimism in the face of uncertainty}.
 Applied to stochastic bandits, the optimism in the face of uncertainty princple suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
@@
 
+Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal.
+This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}.
+Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
+However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivelent to UCB.
+
 \subsection{Pure-exploration problems}
@@
 
-The pseudo-regret for adversarial bandits is defined, in the same way as for stochastic bandits, as the difference between the reward obtained by playing the single arm with the highest expected reward in every round and the expected reward obtained by the algorithm.
+For oblivious adversarial bandits we can define regret analagously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also refered to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
+We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
 
@@
 \label{eqn:pseudo_regret-adversarial}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+\regret(\pi) &= \max_{i \in \set{1,...,k}}{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
 }
@@
 
-\begin{figure}
+\begin{figure}[h]
   \caption{The pseudo-regret can be negative if rewards are non-stationary.
@@
 \subsection{Contextual bandits}
-Contextual bandits are a generalisation of multi-armed bandits.
+\label{sec:contextual-bandits}
+
+In the standard multi-armed bandit setting, each decision is identical and the goal is to learn a single best action.
+However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
+The best treatment to offer an individual patient could depend on their age, gender, disease subtype or genetics and will not always align with the treatment that is best on average (or for the majority of people).
+Similarly, decisions on which ad or content to display on a webpage or which product to reccomend can be \emph{personalised} based on the previous behaviour of the user.
+A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
+Contextual bandits are a generalisation of multi-armed bandits that make use of this additional contextual information.
+The term contextual bandit was coined by \citet{Langford2008}, however close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
 
 \vspace{0.5cm}
-\begin{definition}[Contextual bandit problem]
+\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analagously to definition \ref{def:adversarial-k-armed-bandit}}]
+\label{def:contextual-bandit-problem}
 Let $\P{\vec{Y},\vec{X}}$ be the joint distribution over the rewards for each action and some context $\vec{x} \sim \mathcal{X}$. In each round $t \in \set{1,...T}$, 
@@
 
-What are the reductions of contextual bandits to other problem (ie k armed-bandits?)
-
-In the previous sections each decision was identical.
-If we were imaging treating patients, we assume that we have no additional information about each patient that is relevant to deciding how to treat them.
-However, in many key applications such as serving ads, etc, etc.
-we clearly do have additional information that we wish to take into account.
-In the contextual bandit setting the goal is to learn
-
-\eq {
-\P{Y|a,x}
-}
-
-As opposed to the standard bandit setting where we are attempting to estimate
-
-\eq{
-\P{Y|a}
-}
-
-The formal definition of a contextual bandit problem is
-
-The simplest way to generalise the standard bandit algorithms to the contextual case (where the context is discrete) is to simply have a separate standard bandit instance for each possible setting of the context.
-Regrettably this approach scales terribly with increasing complexity of the context.
-\todo{quantify this}.
-It cannot be applied at all if the context is continuous.
-
-The approach to solving this problem parallels those from supervised learning and continuous armed bandits.
-We make assumptions about the smoothness of the problem.
-Values of context that are similar should lead to comparable rewards for a given action.
-
-Regret bounds \todo {insert regret bounds for contextual bandits}
-
-It should be noted that there is an important difference between the approaches we can apply to the contextual bandit versus the supervised learning setting.
-As for the casual effect estimation problem, we cannot utilise cross-validation offline to select model parameters.
-This makes contextual bandit algorithms degrade faster than supervised learning ones as we add irrelevant variables.
-\todo{ask Tor if he thinks this is correct}
-
-RUN A SIMULATION COMPARING HOW supervised learning and contextual bandits degrade as we add more features. What would be a fair comparison?
+Standard multi-armed bandits learn to select the action $a$ that, with high probability, maximises $\E{Y|a}$.
+Contextual bandits learn to select actions that maximise $\E{Y|\vec{x},a}$.
+The reward for contextual bandits should  be compared to an oracle that acts optimally based on the context.
+To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothosis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
+The pseudo-regret is the difference betweent the expected regret obtained by an oracle that selects actions based on the single best hypothosis or policy $h$ at each timestep, and the expected reward obtained by the algorithm.
+
+\eqn{
+\label{eqn:regret_contextual}
+\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{Y_{t,h(\vec{x}_t)}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+}
+
+If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a seperate standard bandit instance for each value of the context.
+This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothosis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occured such that $\sum_{c=1}^N T_c = T$.
+The regret is maximised if $T_c = T/N$ resulting in total regret $\bigo{N\sqrt{kT/N}}$.}.
+This is optimal with respect to this class of hypothosis.
+However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any stucture in the in the relationships between different contexts and actions.
+As in the supervisied learning setting, the existance of some form of low-dimensional structure is key to learning in realistic problems, where the context is continous or high-dimensional \footnote{Even if the context is genuintly discrete, $N$ grows exponentially with the number of variables.
+For example, with $n$ binary variables, $N = 2^n$}.
+We expect some form of smoothness; values of context that are similar should lead to comparable rewards for a given action.
+We need algorithms that can leverage such assumptions.
+
+An alternate reduction to the standard bandit problem, which allows us to constrain the hypothosis space to explore, is to treat each hypothosis $h$ as a bandit arm \citep{Langford2008}.
+At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypthosis, take action $h(\vec{x})$ and observe the associated reward.
+Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothosis class considered, limiting our ability to learn any complex mappings from context to actions.
+The key problem with this approach is each sample is used to update our knoweledge about only one hypothosis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothosis simultaneously.
+
+Suppose that, at each timestep $t$, after selecting an action, the learner recieved the reward for chosen action but observed the full vector of rewards $[Y_{t,1},...,Y_{t,k}]$.
+This is known as the full information setting.
+In this case, the learner can simulate running each hypothosis over the history to compute the reward it would have obtained and use the hypothosis with the best empirical reward to select the next action.
+This is the \emph{follow the leader} algorithm, which obtains optimal regret $\bigo{\sqrt{T log(\card{\mathcal{H}})}}$ for the full-information problem \citep{cesa2006prediction}.
+Unfortunately, in the contexual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
+As in causal effect estimation, we can view this as a missing data problem.
+However, the data is missing not at random because which component of the reward is observed depends on the action selected which in turn is a function of the previous history of actions and rewards.
+
+The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly seperating exploration from exploitation.
+Epoch-greedy is an explore-explot algorithm.
+It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothosis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
+The hypothosis with the highest empirical reward is then used to select actions for the remaining timesteps.
+The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
+
+The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
+However, it involves maintaining a list of weights for each hypothosis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothosis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothosis spaces in a straighforward way.
+The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
+
+Both Epoch-greedy and ILOVECONBANDITS involve solving problems of the form,
+
+\eqn{
+\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{x}_t) = a_t}
+}
+
+This expression equates to identifying the best emprical policy based on previous data.
+The algorithms assume the existance of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
+The computational tractability of these algorithms on large (or infinite) hypothosis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classiciation problem \citep{Dudik2011a}.
+
+Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalized versions of the UCB or Thompson sampling algorithms.
+For linear pay off models, both approaches yield algorithms with strong regret guarantees, Lin-UCB \citep{li2010contextual} and generalised thompson sampling for linear pay-offs \citep{Agrawal2013}.
+
+It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
+\citet{Bareinboim2015} demonstrate that policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can acheive lower regret than those that ignore this information.
+This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
+However, the context should be relevent, such that $\P{\vec{Y}|\vec{X}} \neq \P{\vec{Y}}$, otherwise including it is the equivelent to adding irrelevent featurs to a supervised learning problem.
+
+\begin{figure}[h]
+\centering
+\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
+$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}_a$ is the reward recieved by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
+Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
+To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $a_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},a,Y)_1^{t-1}$.
+}
+\label{fig:contextual-bandit-causal-graph}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$\vec{X}$ causes $\vec{Y}$}
+\label{fig:context-graph-causal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$\vec{Y}$};
+\node[main node, below=of 1](2){$Y$};
+\node[main node, above right=of 2](3){$a$};
+\node[main node, left=of 1](4){$\vec{X}$};
+ \path[every node/.style={font=\sffamily\small}]
+   (4) edge (1)
+   (3) edge (2)
+   (1) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$\vec{Y}$ causes $\vec{X}$}
+\label{fig:context-graph-anti-causal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$\vec{Y}$};
+\node[main node, below=of 1](2){$Y$};
+\node[main node, above right=of 2](3){$a$};
+\node[main node, left=of 1](4){$\vec{X}$};
+ \path[every node/.style={font=\sffamily\small}]
+   (1) edge (4)
+   (3) edge (2)
+   (1) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$\vec{X}$ and $\vec{Y}$ are counfounded}
+\label{fig:context-graph-confounded}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$\vec{Y}$};
+\node[main node, below=of 1](2){$Y$};
+\node[main node, above right=of 2](3){$a$};
+\node[main node, left=of 1](4){$\vec{X}$};
+\node[lt, below=of 4](5){$U$};
+ \path[every node/.style={font=\sffamily\small}]
+   (3) edge (2)
+   (1) edge (2)
+   (5) edge (4) edge (1);
+\end{tikzpicture}
+\end{subfigure}
+\end{figure}
 
 \subsubsection{Learning from logged bandit data}
+Another topic of interest within the contextual bandit community that is deeply related to causal effect estimation from observational data is learning from logged bandit feedback data \citep{}.
+
+If no assumptions are made about the behaviour of the hypothosis that generated the data.... does it reduce to causal effect estimation?
+
+However, in the applications considered in these setting such as ad serving, the details of the hypothosis generating the data are often known.
 
 \subsection{Markov decision processes}
+
+\subsubsection{Limitations of bandit problems}
+\begin{itemize}
+\item Reward needs to be observed immediately after action is taken (or at least before the next action must be selected).
+\item State assumed not to change
+
+\end{itemize}
+
+Reward known immediately. Action chosen for one individual does not effect reward for the next (SUTVA).
 
@@
 \begin{enumerate}
+\item In the $\epsilon$-greedy approach, does one compute best action based only on exploration steps?
+\item Why does the naive reduction appear to get sub-optimal regret?
 \item When does the expected regret make sense as the thing we actually would wish to optimise?
@@
 \item in what real world setting do adversarial bandits potentially make sense?
+\item Does supervised learning really 'reveal the value of every action'.
 \end{enumerate}

### 2017-07-08 thesis/offcuts.tex
finishes off contextual bandits and off-policy evaluation section drafts

@@
 RUN A SIMULATION COMPARING HOW supervised learning and contextual bandits degrade as we add more features. What would be a fair comparison?
+
+\subsection{Markov decision processes}
+
+\subsubsection{Limitations of bandit problems}
+\begin{itemize}
+\item Reward needs to be observed immediately after action is taken (or at least before the next action must be selected).
+\item State assumed not to change
+
+\end{itemize}
+
+Reward known immediately. Action chosen for one individual does not effect reward for the next (SUTVA).
+
+A key limitation of the bandit setting is that
+
+This is reasonable in settings where we have many similar but independent units or individuals and the goal is to identify which action is optimal (in a given context), as occurs in standard experimental settings as well.
+
+Relation to SUTVA assumption, selecing a action for one 'unit' does not effect the reward distributions for other units.
+
+The reward of an action is not immediately known. - a long sequence of steps. Delayed reward.
+
+In the bandit setting the reward distributions for each arm are assumed to be fixed.
+A much studied generalisation is the Markov decision process or MDP.
+In an MDP we assume there the environment has some state.
+The reward an agent receives depends on the state of the environment and the action the agent selects.
+The environment then evolves stochastically depending on the agents action.
+Multi-armed bandits can be considered a single state MDP, where no matter what action the agent selects, the environment returns to that state.
+
+Insert a graphical representation of an MDP
+
+What assumptions are required to make MDPs tractable.
+
+What is the basic algorithm (Q-learning)
+
+What are the key results.
+
+\subsubsection{Dynamic Systems}
+\begin{itemize}
+\item An explicit model of actions in a partially known system (eg HMM)
+\item Feynman-Kac Lemma; Solving a PDE can be converted to a stochastic process
+\end{itemize}
+
+\subsection{The need to add structure}
+The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.
+This makes problems with large or infinite actions spaces intractable.
+REFERENCE equation showing linearity with arms.
+CHECK ITS TRUE.
+
+causal bandits capture a very natural form of structure that can arise between different arms.
+
+The classic multi-armed bandit is a powerful tool for sequential decision making.
+However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and Gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
+
+\section{Questions}
+\begin{enumerate}
+\item In the $\epsilon$-greedy approach, does one compute best action based only on exploration steps?
+\item Why does the naive reduction appear to get sub-optimal regret?
+\item When does the expected regret make sense as the thing we actually would wish to optimise?
+Is optimising the expected regret any different to optimising the psuedo-regret?
+Who is the definition of regret down to?
+Does this definition of regret come from learning with expert advise?
+\item Is there a reference for the doubling trick? Is it an automatic way of converting a fixed horizon algorithm to an anytime one?
+\item Is Learning from logged feedback data just covariate shift too?
+\item In contextual bandits, does it ever make sense that $Y$ causes $X$?
+If the reward is some function of the action taken and the context, then the full vector of counterfactual outcomes $Y$ is never instantiated.
+Are there any implications of this for inference?
+\item Why does anything solving a regression problem work in the anti-causal direction - it would seem to break the fundamental underlying assumption of additive noise?
+\item is there a benchmark library of implementations of key bandit algorithms?
+\item Bubeck 2012 states that if the adversary is oblivious, the psuedo-regret coincides with the standard regret, 'which is always the ultimate quantity of interest'.
+Why, they are not the same for stochastic bandits are they?
+\item what is the worst-case regret of UCB in an adversarial setting?
+\item is there a concept of problem dependent regret for adversarial bandits? Has there been analysis of the problem dependent regret for key adversarial algorithms such as EXP-3?
+\item adversarial regret bounds for Thompson sampling, UCB with random arm selection from upper-bound set?
+\item do continuous armed bandits implie that there are arms with any arbitrary gap in reward (including worst case gap for that T) and thus you always end up with something like worst case regret?)
+\item in what real world setting do adversarial bandits potentially make sense?
+\item Does supervised learning really 'reveal the value of every action'.
+\end{enumerate}

### 2017-07-08 thesis/thesis_map.tex
finishes off contextual bandits and off-policy evaluation section drafts

@@
 \newcommand{\cf}[2]{{#1}^{#2}}
+\newcommand{\cfb}[2]{{#1}_t^{#2}} % used to represent reward
 \newcommand{\bernoulli}{\operatorname{Bernoulli}}
@@
 \begin{definition}[Stochasitic k-armed bandit problem]
-Let $\actionspace = \set{1,...k}$ be the set of available actions (or bandit arms) and $\P{\vec{Y}} = \P{Y_1,...,Y_k}$ be the joint distribution over the rewards for each action.
+Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{y_1,...,y_k}$ be a joint distribution over the rewards for each action.
 The multi-armed bandit problem proceeds over $T$ rounds.
@@
 \item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{Y}}$
+\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{y}}$
 \item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
@@
 At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$.
-We denote the expected reward for the action $i$, $\E{ \P{Y_i}}$, by $\mu_i$ and the action with the highest expected reward by $i^*$.
+We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
 \end{definition}
@@
 The difference stems from the fact the problem dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
-\citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by,
-
-\eqn{
-\label{eqn:k-arm-bandit-lower-bound}
-\regret \in \bigomega{\sqrt{kT}}
-}
+\citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by $\regret \in \bigomega{\sqrt{kT}}$
 
@@
 \label{def:contextual-bandit-problem}
-Let $\P{\vec{Y},\vec{X}}$ be the joint distribution over the rewards for each action and some context $\vec{x} \sim \mathcal{X}$. In each round $t \in \set{1,...T}$, 
+Let $\P{\vec{x},\vec{y}}$ be the joint distribution over the rewards for each action and some context $\vec{X} \in \mathcal{X}$. In each round $t \in \set{1,...T}$, 
 \begin{enumerate}
-\item the world stochastically generates the vector of rewards for each action and the context, $([Y_{t,1},...,Y_{t,k}],\vec{x}_t) \sim \P{\vec{Y},\vec{X}}$ and reveals $\vec{x}_t$ to the learner
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps.
-\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\item the world stochastically generates the vector of rewards for each action and the context, $(\vec{X}_t,[\cfb{Y}{1},...,\cfb{Y}{k}]) \sim \P{\vec{x},\vec{y}}$ and reveals $\vec{X}_t$ to the learner
+\item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps,
+\item the learner observes and recieves (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
 \end{enumerate}
@@
 \label{eqn:regret_contextual}
-\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{Y_{t,h(\vec{x}_t)}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{\cfb{Y}{h(\vec{X}_t)}}} - \E{\sum_{t=1}^T{\cfb{Y}{A_t}}}
 }
@@
 
-Suppose that, at each timestep $t$, after selecting an action, the learner recieved the reward for chosen action but observed the full vector of rewards $[Y_{t,1},...,Y_{t,k}]$.
+Suppose that, at each timestep $t$, after selecting an action, the learner recieved the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
 This is known as the full information setting.
@@
 \eqn{
-\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{x}_t) = a_t}
+\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{X}_t) = A_t}
 }
@@
 Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalized versions of the UCB or Thompson sampling algorithms.
-For linear pay off models, both approaches yield algorithms with strong regret guarantees, Lin-UCB \citep{li2010contextual} and generalised thompson sampling for linear pay-offs \citep{Agrawal2013}.
+For linear pay-off models, both approaches yield algorithms with strong regret guarantees, \emph{Lin-UCB} \citep{li2010contextual} and \emph{Generalised Thompson Sampling} \citep{Agrawal2013}.
 
@@
 This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
-However, the context should be relevent, such that $\P{\vec{Y}|\vec{X}} \neq \P{\vec{Y}}$, otherwise including it is the equivelent to adding irrelevent featurs to a supervised learning problem.
+However, the context should be relevent, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivelent to adding irrelevent featurs to a supervised learning problem.
 
@@
 \caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
-$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}_a$ is the reward recieved by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
+$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward recieved by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
 Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
-To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $a_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},a,Y)_1^{t-1}$.
+To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
 }
@@
 \node[main node, below=of 1](2){$Y$};
-\node[main node, above right=of 2](3){$a$};
+\node[main node, above right=of 2](3){$A$};
 \node[main node, left=of 1](4){$\vec{X}$};
@@
 \subsubsection{Learning from logged bandit data}
-Another topic of interest within the contextual bandit community that is deeply related to causal effect estimation from observational data is learning from logged bandit feedback data \citep{}.
-
-If no assumptions are made about the behaviour of the hypothosis that generated the data.... does it reduce to causal effect estimation?
-
-However, in the applications considered in these setting such as ad serving, the details of the hypothosis generating the data are often known.
-
-\subsection{Markov decision processes}
-
-\subsubsection{Limitations of bandit problems}
-\begin{itemize}
-\item Reward needs to be observed immediately after action is taken (or at least before the next action must be selected).
-\item State assumed not to change
-
-\end{itemize}
-
-Reward known immediately. Action chosen for one individual does not effect reward for the next (SUTVA).
-
-A key limitation of the bandit setting is that
-
-This is reasonable in settings where we have many similar but independent units or individuals and the goal is to identify which action is optimal (in a given context), as occurs in standard experimental settings as well.
-
-Relation to SUTVA assumption, selecing a action for one 'unit' does not effect the reward distributions for other units.
-
-The reward of an action is not immediately known. - a long sequence of steps. Delayed reward.
-
-In the bandit setting the reward distributions for each arm are assumed to be fixed.
-A much studied generalisation is the Markov decision process or MDP.
-In an MDP we assume there the environment has some state.
-The reward an agent receives depends on the state of the environment and the action the agent selects.
-The environment then evolves stochastically depending on the agents action.
-Multi-armed bandits can be considered a single state MDP, where no matter what action the agent selects, the environment returns to that state.
-
-Insert a graphical representation of an MDP
-
-What assumptions are required to make MDPs tractable.
-
-What is the basic algorithm (Q-learning)
-
-What are the key results.
-
-\subsubsection{Dynamic Systems}
-\begin{itemize}
-\item An explicit model of actions in a partially known system (eg HMM)
-\item Feynman-Kac Lemma; Solving a PDE can be converted to a stochastic process
-\end{itemize}
-
-\subsection{The need to add structure}
-The regret for a bandit problem grows linearly with the number of (sub-optimal) actions.
-This makes problems with large or infinite actions spaces intractable.
-REFERENCE equation showing linearity with arms.
-CHECK ITS TRUE.
-
-causal bandits capture a very natural form of structure that can arise between different arms.
-
-The classic multi-armed bandit is a powerful tool for sequential decision making.
-However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and Gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.
-
-\section{Questions}
-\begin{enumerate}
-\item In the $\epsilon$-greedy approach, does one compute best action based only on exploration steps?
-\item Why does the naive reduction appear to get sub-optimal regret?
-\item When does the expected regret make sense as the thing we actually would wish to optimise?
-Is optimising the expected regret any different to optimising the psuedo-regret?
-Who is the definition of regret down to?
-Does this definition of regret come from learning with expert advise?
-\item Is there a reference for the doubling trick? Is it an automatic way of converting a fixed horizon algorithm to an anytime one?
-\item Is Learning from logged feedback data just covariate shift too?
-\item In contextual bandits, does it ever make sense that $Y$ causes $X$?
-If the reward is some function of the action taken and the context, then the full vector of counterfactual outcomes $Y$ is never instantiated.
-Are there any implications of this for inference?
-\item Why does anything solving a regression problem work in the anti-causal direction - it would seem to break the fundamental underlying assumption of additive noise?
-\item is there a benchmark library of implementations of key bandit algorithms?
-\item Bubeck 2012 states that if the adversary is oblivious, the psuedo-regret coincides with the standard regret, 'which is always the ultimate quantity of interest'.
-Why, they are not the same for stochastic bandits are they?
-\item what is the worst-case regret of UCB in an adversarial setting?
-\item is there a concept of problem dependent regret for adversarial bandits? Has there been analysis of the problem dependent regret for key adversarial algorithms such as EXP-3?
-\item adversarial regret bounds for Thompson sampling, UCB with random arm selection from upper-bound set?
-\item do continuous armed bandits implie that there are arms with any arbitrary gap in reward (including worst case gap for that T) and thus you always end up with something like worst case regret?)
-\item in what real world setting do adversarial bandits potentially make sense?
-\item Does supervised learning really 'reveal the value of every action'.
-\end{enumerate}
+Another topic of interest within the contextual bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
+In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit enviroment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
+The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
+
+This problem differs from the contextual bandit problem in that the learner is not interacting with the enviroment.
+As a result, there is no exploration-exploitation trade-off to be made.
+However, the problem does not reduce to supervised learning, because the label, $y$ is not the desired .
+In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
+The majority of the liturature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
+\citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original estimator $\pi$ did not depend on $\vec{X}$.
+
+If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, discussed in section \ref{sec:causal_effect_under_ignorability}.
+The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ statisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
+The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivelent to $do(A=a)$.
+However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see \citet{Pearl2000}, section 4.2.
+In this case, letting $\Pns{\pi'}{a|\vec{x}}$ denote the distribution over actions under policy $\pi'$ given context $\vec{x}$, the expected (per round) reward obtained by $\pi'$ is given by,
+
+\eqn{
+\E{y|\pi'} = \E{y|do(a \sim \pi'(\vec{x}))} = \Esub{(\vec{x},a) \sim \P{\vec{x}}\Pns{\pi'}{a|\vec{x}}}{y|\vec{x},a}
+}
+
+As in estimating average causal effects under ignorability, we have a covariate shift problem, with training data sampled from $\P{\vec{x}}\Pns{\pi}{a|\vec{x}}\P{y|\vec{x},a}$ but generalisation error measured with respect to $\P{\vec{x}}\Pns{\pi'}{a|\vec{x}}\P{y|\vec{x},a}$.
+A difference in practice, is that in the applications frequently considered under learning from logged feedback data, such as ad serving or recommender systems, there may be substantial information available about $\pi$, in the best case, $\Pns{\pi}{a|\vec{x}}$ is known.
+This makes estimators utilizing inverse propensity weighting, including doubly robust estimators as in \citet{Dudik2011}, more attractive.
+
+\begin{figure}
+\centering
+\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$}
+\label{fig:off-policy-evaluation}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+\node[main node](1){$\vec{X}$};
+\node[main node, below right=of 1](2){$Y$};
+\node[main node,  below left=of 1](3){$A$};
+ \path[every node/.style={font=\sffamily\small}]
+   (3) edge (2)
+   (1) edge (2) edge (3);
+\end{tikzpicture} 
+\end{figure}
+
+\citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the emprical best because the variance of the estimators for some policies may be much higher than for others.
 

### 2017-07-09 thesis/thesis_map.tex
rough draft of chapter 4

@@
 Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates belived to influence the outcome of interest.
-This introduces a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, nessessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
+If we stratify in such a way that the probability an instance recieves a given treatement is indepdenent of its covariates, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can still estiamte the average causal effects directly from the differences in outcome accross treatments.
+However, more complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, nessessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
 The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
@@
 
-IF THE PROPENSITY SCORE IS THE SAME FOR ALL COVARIATES IT DOESN'T SEEM ONE MUST ADJUST.
-\todo{double check}
-
 \begin{figure}
-\caption{causal network for a stratified randomised experiment}
+\caption{causal network for a stratified randomised experiment if the probability an individual is assigned a given treatment depends on some covariates.}
 \label{fig:random_experiment_network_stratified}
@@
 
-\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
+%\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
 
@@
 It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
-\citet{Bareinboim2015} demonstrate that policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can acheive lower regret than those that ignore this information.
+However, the context should be relevent, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivelent to adding irrelevent featurs to a supervised learning problem.
+\citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can acheive lower regret than those that ignore this information.
 This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
-However, the context should be relevent, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivelent to adding irrelevent featurs to a supervised learning problem.
 
@@
 
-\subsubsection{Learning from logged bandit data}
-Another topic of interest within the contextual bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
+\subsection{Learning from logged bandit data}
+Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
 In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit enviroment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
@@
 \citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the emprical best because the variance of the estimators for some policies may be much higher than for others.
+
+\subsection{Adding structure to actions}
+
+The classic multi-armed bandit is a powerful tool for sequential decision making.
+However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \citep{filippi2010parametric}, dependent bandits \citep{Pandey2007}, X-armed bandits \citep{Bubeck2010} and Gaussian process bandits \citep{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback\citep{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \citep{Piccolboni2001,Bartok2014}.
+
+In the next chapter, I propose a very natural connection between causal graphs and bandit problems and show it induces a novel form of additional feedback and structure between arms that cannot be can exploited by any of these previous approaches.
 
@@
 
-A well-studied class of bandit problems with side information are ``contextual bandits''~\cite{Langford2008,Agarwal2014}.
-Our framework bears a superficial similarity to contextual bandit problems since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
+Our framework bears a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
 However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.

### 2017-07-09 thesis/thesis_map.tex
spelling, citation fixes

@@
 \item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
-\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (accross viewpoints)
+\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
 \item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
@@
 \item Benchmark estimation algorithms
-\item clarify causal interpretation of dirichlet models
-\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software benchmarking solution?
+\item clarify causal interpretation of Dirichlet models
+\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
 \item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption.
@@
 
-The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}.
+The notion of causality has been widely debated in science and philosophy \citep{} but is still widely viewed as poorly defined \citep{}.
 This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
@@
 The what-if type questions from the why.
-\cite{Gelman2010} \cite{}.
+\citep{Gelman2010} \citep{}.
 Why do whites to better than blacks in school (in America).
@@
 
-To explain or to predict \cite{Shmueli2010}
-The two cultures \cite{Breiman2001}
+To explain or to predict \citep{Shmueli2010}
+The two cultures \citep{Breiman2001}
 
@@
 There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure.
-These rules are referred to as the Do-calculus \cite{Pearl2000}.
+These rules are referred to as the Do-calculus \citep{Pearl2000}.
 We discuss them further in section \ref{sec:do_calculus}.
@@
 By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \cite{Pearl2000}.
+This gives rise to the do-calculus \citep{Pearl2000}.
 The do calculus consists of three rules.
@@
 The question of non-parametric identifiability is solved.
-The do calculus is complete \cite{Shpitser2006b,Huang2006}.
+The do calculus is complete \citep{Shpitser2006b,Huang2006}.
 A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
-There is a polynomial time algorithm \cite{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
+There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
 
@@
 Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
-I have created a javascript implementation of the identifiability algorithm \cite{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
+I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
 
@@
 
-%Cyclic models \cite{Hyttinen2012}
+%Cyclic models \citep{Hyttinen2012}
 
@@
 This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
-Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \cite{pearl1988probabilistic} for $P$.
+Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \citep{pearl1988probabilistic} for $P$.
 
@@
 \begin{theorem}
-\cite{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
+\citet{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
 \end{theorem}
@@
 
-For any DAG with latent (and selection) variables there is a unique MAG \cite{Richardson2002}.
-This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \cite{Sprites}.
+For any DAG with latent (and selection) variables there is a unique MAG \citep{Richardson2002}.
+This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites}.
 The logic behind the algorithm is very similar.
 Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations.
-The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \cite{Spirtes1995}.
+The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \citep{Spirtes1995}.
 Assuming there are no selection variables, the PAG can contain four types of link:
@@
 
-The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \cite{Zhang2008}.
-More recently \cite{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
-\cite{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
+The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \citep{Zhang2008}.
+More recently \citet{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
+\citet{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
 
@@
 Recall the functional definition of causality (section \ref{sec:SEM}).
-There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \cite{Hoyer2009}, post-non-linear additive noise \cite{Zhang2008a} or linear models with non-Gaussian noise \cite{Hoyer2012}.
+There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \citep{Hoyer2009}, post-non-linear additive noise \citep{Zhang2008a} or linear models with non-Gaussian noise \citep{Hoyer2012}.
 
@@
 Randomisation does not ensure target and control group are exactly alike.
-The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables in a between the target and control groups in a finite data sample.
-However, the variance in the outcome within both the target and control groups also increases.
+The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables between the target and control groups in a finite data sample.
+However, the variance in the outcome, within both the target and control groups, also increases.
 The net result is increased variance (but not bias) in  the estimate of causal effects.
 
-Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates belived to influence the outcome of interest.
-If we stratify in such a way that the probability an instance recieves a given treatement is indepdenent of its covariates, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can still estiamte the average causal effects directly from the differences in outcome accross treatments.
-However, more complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, nessessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
+Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest.
+If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can estimate the average causal effects directly from the differences in outcome across treatments.
+More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
 The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
-See \citet{imbens2015causal} for a discussion of the trade-offs between statified versus completely random experiments.
+See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.
 
@@
 
-Multi-armed bandits address the problem of desiging experiments that can adapt as samples are observed.
+Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed.
 Their introduction is generally attributed to \citet{Thompson1993}.
-In its classic formulation \cite{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
+In its classic formulation \citep{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
 Each arm $i$ is associated with a fixed but unknown reward distribution.
 \footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not nessessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
+It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
 The goal of the learner is to maximise the total reward they receive.
@@
 \item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{y}}$
-\item the learner observes and recieves (only) the reward for the selected action $Y_{t,a_t}$ 
+\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
 \end{enumerate}
@@
 
-The total reward a bandit algorithm/policy can expect to acheive depends on the distributions from which the rewards for each action are sampled.
+The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
 To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
-This difference is known as the (cummulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning liturature.
+This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
 There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
@@
 However, it still depends on key characteristics of the reward distributions for each action.
-Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e bernoulli bandits, gaussian bandits) or that the rewards are bounded in some range.
+Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range.
 Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
@@
 As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
-The expectation and variance guarentees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
+The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
 This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
@@
 There are a few key principles that are used to guide the development of bandit algorithms.
-The simplest is to explicitly seperate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
-A common example in practice is unform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
+The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
+A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
 This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.
@@
 Another key approach is \emph{optimism in the face of uncertainty}.
-Applied to stochastic bandits, the optimism in the face of uncertainty princple suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
+Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
 The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
@@
 
-\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the expontial family.
-They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptopically and proved this is asymptopically efficient.
+\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family.
+They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient.
 However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm.
@@
 We now assume the rewards are bounded in $[0,1]$.
-The algorithm and regret bounds can be generalised to subgaussian reward distributions, see \citet{Bubeck2012}.
+The algorithm and regret bounds can be generalised to submission reward distributions, see \citet{Bubeck2012}.
 
@@
 The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
-The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which ia maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
+The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which is maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
 By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.
@@
 Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
-This yeilds regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \cite{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \cite{Lattimore2015}.
+This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \citep{Lattimore2015}.
 
@@
 Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
-However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivelent to UCB.
+However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
 
 \subsection{Pure-exploration problems}
-Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
+Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
 In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
@@
 
-Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivous}, as opposed to \emph{non-oblivous} adversaries, which can generate rewards as a function of the history of the game.
-In the case of oblivous adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
-
-For oblivious adversarial bandits we can define regret analagously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also refered to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
+Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
+In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
+
+For oblivious adversarial bandits we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
 We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
@@
 
-The policy (or algorithm) used by the learner is avaiable to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
+The policy (or algorithm) used by the learner is available to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
 This implies the adversary can ensure that any learner with a deterministic policy suffers regret $\bigo{T}$ by forecasting their entire sequence of actions.
-For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forcasts what action the learner will play in round 2, given they recieved a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
+For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forecasts what action the learner will play in round 2, given they received a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
 This implies adversarial bandit policies must be sufficiently random to avoid such exploitation \footnote{The UCB algorithm, defined by algorithm \ref{alg:ucb-1}, is deterministic if the order in which arms are played during the first $k$ rounds is fixed and the method for selecting which arm to play when multiple-arms have the same upper-confidence bound is not-random (for example, select the arm one with the lowest index $i$).}
 
-The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case psuedo-regret of $\bigo{\sqrt{TK\log(T)}}$  \cite{Auer1995}.
-Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \cite{Audibert2009}.
+The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case pseudo-regret of $\bigo{\sqrt{TK\log(T)}}$  \citep{Auer1995}.
+Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \citep{Audibert2009}.
 The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
-However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nontheless obtain good problem dependent regret in more favourable enviroments \citep{bubeck2012bothworlds,Auer2016}.
+However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nonetheless obtain good problem dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
 
 Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, psuedo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, is not a good measure of an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings were the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the psuedo-regret do not reflect how sub-optimal the algorithm may be.
+However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, is not a good measure of an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
+In settings were the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
 Adversarial bandit algorithms may perform better in non-stationary settings than standard stochastic policies to the extend that they explore more (to avoid the adversary simulating their behaviour), however it is preferable to develop algorithms specifically for non-stationary settings (subject to assumptions about how rapidly or frequently rewards can change), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
@@
   \caption{The pseudo-regret can be negative if rewards are non-stationary.
-This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
+This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed Bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
 Figure (a) shows the expected rewards of each arm, and the proportion of time that arm-1 is played, as a function of time.
@@
 However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
-The best treatment to offer an individual patient could depend on their age, gender, disease subtype or genetics and will not always align with the treatment that is best on average (or for the majority of people).
-Similarly, decisions on which ad or content to display on a webpage or which product to reccomend can be \emph{personalised} based on the previous behaviour of the user.
+The best treatment to offer an individual patient could depend on their age, gender, disease sub-type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
+Similarly, decisions on which ad or content to display on a webpage or which product to recommend can be \emph{personalised} based on the previous behaviour of the user.
 A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
@@
 \vspace{0.5cm}
-\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analagously to definition \ref{def:adversarial-k-armed-bandit}}]
+\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analogously to definition \ref{def:adversarial-k-armed-bandit}}]
 \label{def:contextual-bandit-problem}
@@
 \item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps,
-\item the learner observes and recieves (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
+\item the learner observes and receives (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
 \end{enumerate}
@@
 The reward for contextual bandits should  be compared to an oracle that acts optimally based on the context.
-To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothosis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
-The pseudo-regret is the difference betweent the expected regret obtained by an oracle that selects actions based on the single best hypothosis or policy $h$ at each timestep, and the expected reward obtained by the algorithm.
+To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothesis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
+The pseudo-regret is the difference between the expected regret obtained by an oracle that selects actions based on the single best hypothesis or policy $h$ at each timestep, and the expected reward obtained by the algorithm.
 
@@
 
-If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a seperate standard bandit instance for each value of the context.
-This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothosis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occured such that $\sum_{c=1}^N T_c = T$.
+If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a separate standard bandit instance for each value of the context.
+This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothesis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occurred such that $\sum_{c=1}^N T_c = T$.
 The regret is maximised if $T_c = T/N$ resulting in total regret $\bigo{N\sqrt{kT/N}}$.}.
-This is optimal with respect to this class of hypothosis.
-However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any stucture in the in the relationships between different contexts and actions.
-As in the supervisied learning setting, the existance of some form of low-dimensional structure is key to learning in realistic problems, where the context is continous or high-dimensional \footnote{Even if the context is genuintly discrete, $N$ grows exponentially with the number of variables.
+This is optimal with respect to this class of hypothesis.
+However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the in the relationships between different contexts and actions.
+As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
 For example, with $n$ binary variables, $N = 2^n$}.
@@
 
-An alternate reduction to the standard bandit problem, which allows us to constrain the hypothosis space to explore, is to treat each hypothosis $h$ as a bandit arm \citep{Langford2008}.
-At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypthosis, take action $h(\vec{x})$ and observe the associated reward.
-Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothosis class considered, limiting our ability to learn any complex mappings from context to actions.
-The key problem with this approach is each sample is used to update our knoweledge about only one hypothosis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothosis simultaneously.
-
-Suppose that, at each timestep $t$, after selecting an action, the learner recieved the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
+An alternate reduction to the standard bandit problem, which allows us to constrain the hypothesis space to explore, is to treat each hypothesis $h$ as a bandit arm \citep{Langford2008}.
+At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
+Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothesis class considered, limiting our ability to learn any complex mappings from context to actions.
+The key problem with this approach is each sample is used to update our knowledge about only one hypothesis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothesis simultaneously.
+
+Suppose that, at each timestep $t$, after selecting an action, the learner received the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
 This is known as the full information setting.
-In this case, the learner can simulate running each hypothosis over the history to compute the reward it would have obtained and use the hypothosis with the best empirical reward to select the next action.
+In this case, the learner can simulate running each hypothesis over the history to compute the reward it would have obtained and use the hypothesis with the best empirical reward to select the next action.
 This is the \emph{follow the leader} algorithm, which obtains optimal regret $\bigo{\sqrt{T log(\card{\mathcal{H}})}}$ for the full-information problem \citep{cesa2006prediction}.
-Unfortunately, in the contexual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
+Unfortunately, in the contextual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
 As in causal effect estimation, we can view this as a missing data problem.
@@
 
-The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly seperating exploration from exploitation.
-Epoch-greedy is an explore-explot algorithm.
-It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothosis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
-The hypothosis with the highest empirical reward is then used to select actions for the remaining timesteps.
+The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly separating exploration from exploitation.
+Epoch-greedy is an explore-exploit algorithm.
+It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothesis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
+The hypothesis with the highest empirical reward is then used to select actions for the remaining time steps.
 The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
@@
 The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
-However, it involves maintaining a list of weights for each hypothosis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothosis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothosis spaces in a straighforward way.
+However, it involves maintaining a list of weights for each hypothesis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
 The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
@@
 
-This expression equates to identifying the best emprical policy based on previous data.
-The algorithms assume the existance of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
-The computational tractability of these algorithms on large (or infinite) hypothosis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classiciation problem \citep{Dudik2011a}.
-
-Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalized versions of the UCB or Thompson sampling algorithms.
+This expression equates to identifying the best empirical policy based on previous data.
+The algorithms assume the existence of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
+The computational tractability of these algorithms on large (or infinite) hypothesis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classification problem \citep{Dudik2011a}.
+
+Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalised versions of the UCB or Thompson sampling algorithms.
 For linear pay-off models, both approaches yield algorithms with strong regret guarantees, \emph{Lin-UCB} \citep{li2010contextual} and \emph{Generalised Thompson Sampling} \citep{Agrawal2013}.
@@
 It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
-However, the context should be relevent, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivelent to adding irrelevent featurs to a supervised learning problem.
-\citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can acheive lower regret than those that ignore this information.
+However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
+\citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information.
 This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
@@
 \caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
-$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward recieved by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
+$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
 Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
@@
 \centering
-\caption{$\vec{X}$ and $\vec{Y}$ are counfounded}
+\caption{$\vec{X}$ and $\vec{Y}$ are confounded}
 \label{fig:context-graph-confounded}
@@
 Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
-In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit enviroment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
+In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
 The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
 
-This problem differs from the contextual bandit problem in that the learner is not interacting with the enviroment.
+This problem differs from the contextual bandit problem in that the learner is not interacting with the environment.
 As a result, there is no exploration-exploitation trade-off to be made.
@@
 In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
-The majority of the liturature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
+The majority of the literature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
 \citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original estimator $\pi$ did not depend on $\vec{X}$.
@@
 If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, discussed in section \ref{sec:causal_effect_under_ignorability}.
-The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ statisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
-The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivelent to $do(A=a)$.
+The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ satisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
+The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivalent to $do(A=a)$.
 However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see \citet{Pearl2000}, section 4.2.
@@
 A difference in practice, is that in the applications frequently considered under learning from logged feedback data, such as ad serving or recommender systems, there may be substantial information available about $\pi$, in the best case, $\Pns{\pi}{a|\vec{x}}$ is known.
-This makes estimators utilizing inverse propensity weighting, including doubly robust estimators as in \citet{Dudik2011}, more attractive.
+This makes estimators utilising inverse propensity weighting, including doubly robust estimators as in \citet{Dudik2011}, more attractive.
 
@@
 
-\citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the emprical best because the variance of the estimators for some policies may be much higher than for others.
+\citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the empirical best because the variance of the estimators for some policies may be much higher than for others.
 
@@
 \begin{figure}
-\caption{Example causal graph (based on \cite{Koller2009}) where the outcome of interest (reward) is cholesterol level .
+\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level .
 The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
@@
 There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on unplayed actions are revealed according to a feedback graph.
+Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on un-played actions are revealed according to a feedback graph.
 As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
 They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
-The partial monitoring approach taken by \cite{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
-\citet{Yu2009} utilize extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
+The partial monitoring approach taken by \citet{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
+\citet{Yu2009} utilise extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
 \citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled.
@@
 % Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \cite{Bartok2014}.
+It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \citep{Bartok2014}.
 Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
@@
 \subsection{Experiments}
-We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \cite{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
+We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
 Thomson sampling and UCB are optimised to minimise cumulative regret.
@@
 
-\cite{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
+\citet{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
 In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
-\cite{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
+\citep{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
 
@@
 The latter case arises naturally if we assume free access to a large observational data set, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\cite{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
+Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\citep{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
 In particular, \citet{Hu2014} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
@@
 \begin{proof}
-By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \cite{hagerup1990guided}),
+By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \citet{hagerup1990guided}),
 
@@
 But dependent on causal structure, conditioning on the fact that people are in a position to apply for such jobs could reverse the dependency.
-Ones facial appearance is a readily observable characteristic and there is evidence that people implicity make assumptions on the basis of it.
-Consider trials where people with 'more criminal' faces got longer sentances for similar crimes.
+Ones facial appearance is a readily observable characteristic and there is evidence that people implicitly make assumptions on the basis of it.
+Consider trials where people with 'more criminal' faces got longer sentences for similar crimes.
 %\chapter{Causality \& Interpretability}
@@
 
-As machine learning is incorporated into decision making systems that have fundamental impacts on people’s lives, such as in employment, criminal justice, health and financial services, there are increasing concerns over transparency and fairness \cite{WMD, etc}.
+As machine learning is incorporated into decision making systems that have fundamental impacts on people’s lives, such as in employment, criminal justice, health and financial services, there are increasing concerns over transparency and fairness \citep{WMD, etc}.
 Realisation that machine learning algorithms can be inherit biases from the data we feed into them and the choices those building them make about what variables to include, etc.
 
-The European Union’s General Data Protection Regulation \cite{Goodman2016}\todo{check}, due to come into effect in 2018, requires that people can obtain "meaningful information about the logic involved" in an automated decision process .
+The European Union’s General Data Protection Regulation \citep{Goodman2016}\todo{check}, due to come into effect in 2018, requires that people can obtain "meaningful information about the logic involved" in an automated decision process .
 It also stipulates that such decisions should not be based on special categories of personal data (related to ethnicity, political and religions affiliation or sexuality) unless there "suitable measures" to safeguard individual interests.
@@
 But what should we condition on?
-\cite{Romie2010} argue that looking at the department level is the correct viewpoint, since this the point at which hiring decisions were made.
+\citep{Romie2010} argue that looking at the department level is the correct viewpoint, since this the point at which hiring decisions were made.
 However, it could be argued that if one department was attracting a large number of higher quality candidates, the overall size of that department should be higher and that the results seen at Berkeley reflected a bias in favour of male dominated fields.
@@
 
-\cite{Romei2012} note the relevance of causal inference on discrimination analysis.
+\citep{Romei2012} note the relevance of causal inference on discrimination analysis.
 
@@
 Assessing the impact of marketing spend is becoming increasingly important to industry.
-Billions of dollars \cite{XXX} are spent each year and new marketing channels and opportunities are opening up at an unprecedented rate.
+Billions of dollars \citep{XXX} are spent each year and new marketing channels and opportunities are opening up at an unprecedented rate.
 

### 2017-07-09 thesis/offcuts_chapter6.tex
minor edits

--- 2017-07-09 thesis/thesis_map.tex
@@
-\documentclass[11pt,a4paper,oneside]{book}
-\usepackage[table]{xcolor}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{dsfont}
-\usepackage{listings}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{tabularx}
-\usepackage{booktabs}
-\usepackage{enumitem}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithm
-\usepackage[noend]{algorithmic}
-\usepackage{parskip}
-\usepackage{tikz}
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{epigraph}
-\usetikzlibrary{arrows,fit,positioning}
-\usepackage{subcaption}
-\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[numbers]{natbib}
-\usepackage{wrapfig}
-\usepackage{centernot}
-
-% this is included to get tables with colored rows to not draw over hlines and vlines
-\usepackage{etoolbox}
-
-\makeatletter
-\patchcmd{\@classz}
-  {\CT@row@color}
-  {\oldCT@column@color}
-  {}
-  {}
-\patchcmd{\@classz}
-  {\CT@column@color}
-  {\CT@row@color}
-  {}
-  {}
-\patchcmd{\@classz}
-  {\oldCT@column@color}
-  {\CT@column@color}
-  {}
-  {}
-\makeatother
-
-\usepackage{amsthm}
-
-\graphicspath{ {figures/} }
-\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
-
-\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
-\DeclareUnicodeCharacter{00A0}{ }
-
-\tikzset{
-    %Define standard arrow tip
-    >=stealth',
-    %Define style for boxes
-    punkt/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           text width=1em,
-           minimum height=1em,
-           text centered},
-    observed/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.3em,
-           minimum height=2.3em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-     latent/.style={
-           circle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered
-           },
-    target/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-    observedrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-    latentrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered
-           },
-     targetrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-     empty/.style={
-           circle,
-           rounded corners,
-           minimum width=.5em,
-           minimum height=.5em,
-           font=\footnotesize,
-           text centered,
-           },
-    % Define arrow style
-    pil/.style={
-           o->,
-           thick,
-           shorten <=2pt,
-           shorten >=2pt,},
-    sh/.style={ shade, shading=axis, left color=red, right color=green,
-    shading angle=45 }   
-}
-
-\newcommand{\KL}{\operatorname{KL}}
-
-\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
-\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
-\newcommand{\Q}[1]{\operatorname{Q}\left(#1\right)}
-\newcommand{\EE}{\mathbb E}
-\newcommand{\EEa}{\EE_a}
-\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
-\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
-\newcommand{\actions}{\mathcal{A}}
-\newcommand{\calA}{\mathcal A}
-\newcommand{\etc}{\textit{etc}}
-\newcommand{\ie}{\textit{i.e.}}
-\newcommand{\eg}{\textit{e.g.}}
-\newcommand{\calP}{\mathcal P}
-\newcommand{\x}{\boldsymbol{x}}
-\newcommand{\Ps}{\operatorname{P}}
-\newcommand{\R}{\mathbb R}
-\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
-\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
-\newcommand{\indep}[1]{\mathcal{I}\left(#1\right)}
-\newcommand{\vb}[1]{\boldsymbol{#1}}
-\newcommand{\card}[1]{\left\vert{#1}\right\vert}
-\newcommand{\actionspace}{\mathcal{A}}
-
-\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
-\newcommand{\defined}{\vcentcolon =}
-\newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
-\newcommand{\Var}{\operatorname{Var}}
-\newcommand{\calF}{\mathcal F}
-\newcommand{\sr}[1]{\stackrel{#1}}
-\newcommand{\set}[1]{\left\{#1\right\}}
-\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
-
-\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
-\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
-\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
-\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
-\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
-\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
-\newcommand{\Ber}{\operatorname{Bernoulli}}
-\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
-\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
-\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
-\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
-\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
-
-\renewcommand{\P}[1]{\operatorname{P}\left(#1\right)}
-\newcommand{\cf}[2]{{#1}^{#2}}
-\newcommand{\cfb}[2]{{#1}_t^{#2}} % used to represent reward
-\newcommand{\bernoulli}{\operatorname{Bernoulli}}
-\newcommand{\dirac}{\operatorname{Dirac}}
-\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
-\renewcommand{\vec}[1]{\boldsymbol{#1}}
-\newcommand{\regret}{\bar{R}_{T}} %psuedo-regret
-\newcommand{\expectedregret}{\mathbb{E}R_{T}} % expected regret
-\newcommand{\rawregret}{R_{T}} % plain regret
-\newcommand{\simpleregret}{R_T}
-\newcommand{\quotes}[1]{``#1''}
-
-\theoremstyle{plain} 
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\let\temp\epsilon
-\let\epsilon\varepsilon
-
-\newcommand{\ci}{\mathrel{\perp\mspace{-10mu}\perp}}
-\newcommand{\nci}{\centernot{\ci}}
-
-\author{Finnian Lattimore}
-\title{Learning how to act: making good decisions with machine learning}
-
-\begin{document}
-\todo{figure out theoremstle is undefined}
-%\def\ci{\perp\!\!\!\perp} % from Wikipedia
-
-\maketitle
-
-\epigraph{In vain the grave, with retrospective Eye,
-Would from the apparent what conclude the why,
-Infer the Motive from the Deed, and show
-That what we chanced, was what we meant, to do.}{Alexander Pope}
-
-\tableofcontents
-
-\chapter*{Todo}
-\begin{enumerate}
-\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
-\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
-\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
-When is a problem causal, what bounds does predictive accuracy give on causal models?
-To what degree is predictive accuracy a metric on which causal models can be judged?
-\item Benchmark estimation algorithms
-\item clarify causal interpretation of Dirichlet models
-\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
-\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption.
-For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability.
-\end{enumerate}
-
-\chapter{Introduction}
-
-\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.
-
-\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
-
-\paragraph*{My key research questions are:} 
-\begin{itemize}
-\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
-\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
-\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
-\end{itemize}
-
-\section{Motivation}
-Many of the most important questions in science and in our personal lives are about the outcomes of doing something.
-Will asking people to pay upfront at the doctors reduce long term health expenditure?
-If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.
-
-These are hard questions because they require more than identifying a pattern in data.
-Correlation is not causation.
-Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.
-
-We now live in a world of data.
-Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth.
-Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose.
-It does not come from the controlled world of a randomised experiment.
-
-The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
-From health care, to entertainment and self driving cars; machine learning algorithms will transform many industries.
-It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.
-
-However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings.
-Getting them to generalise to even slightly different problems or data sets remains very challenging.
-Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it.
-The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on.
-As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
-
-\section{What is causality and why do we care? }
-
-The notion of causality has been widely debated in science and philosophy \citep{} but is still widely viewed as poorly defined \citep{}.
-This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
-However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression).
-Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.
-
-We discuss more detailed definitions of causality in section c
-
-The what-if type questions from the why.
-\citep{Gelman2010} \citep{}.
-Why do whites to better than blacks in school (in America).
-Suggests that reverse causal inference questions are more interesting and motivate most of social science.
-
-I do not find this distinction useful. 
-We can only change the future - history is useful only as far as it tells us something about the future.
-Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one).
-Problems highlighted as intractable in the reverse causal sense are also intractable in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning.
-For example, the war question posed in Gelman.
-
-A distinction between forward causal inference, what happens if we do X and reverse casual inference
-
-To explain or to predict \citep{Shmueli2010}
-The two cultures \citep{Breiman2001}
-
-\paragraph*{There are two reasons why correlation is not causation}.
-The first is related to variance and over-fitting.
-Observations are noisy.
-With a finite data set with enough variables we will be able to find some that are completely unrelated but correlate purely by chance.
-\ref{figure - search for random correlation}.
-The second arises from bias, typically introduced by an un-observed confounding variable.
-In this case, variables are correlated not by chance.
-We would expect the relationship to hold if we sampled more data.
-However, they are not causally related in that intervening to set one would not likely effect the other.
-EXAMPLE WITH FIGURE.
-
-\subsection{Defining causality}
-\begin{itemize}
-\item widely debated in science and philosophy (REFERENCES)
-\item what is explanation?
-\item any model that aims to predict the outcome of an action or intervention in a system
-\item I do not see the distinction between explanation and (causal) prediction.
-Explanation is all about the ability to compress and to generalise.
-The more a model can do this, the more we view as providing an understanding of the why.
-\item mediation?
-\end{itemize}
-
-\section{Identifying when we have a causal problem}
-
-\subsubsection{Examples of typical machine learning problems. Are they causal?}
-
-Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to.
-Which of them require casual inference?
-How can we identify characteristics of a problem that make causal modelling important?
-
-\begin{itemize}
-\item Speech recognition (for systems like Siri or Google)
-\item Machine translation 
-\item Image classification
-\item Forecasting the weather
-\item Playing Go 
-\item Identifying spam emails
-\item Automated essay marking
-\item Predicting the risk of death in patients with pneumonia.
-\item Predicting who will re-offend on release from prison 
-\item Predicting which customers will cease to be your customers
-\item Demand prediction for inventory control
-\item Predicting who will click on an ad
-\item Financial trading
-\item Recommending movies
-\item Online search
-\item Self driving cars
-\item Pricing insurance
-\end{itemize}
-
-The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.
-
-Consider speech recognition.
-You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words.
-Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution).
-A similar argument could be made for many applications of machine translation and image classification.
-
-In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see.
-We can be fairly confident that the process will not change.
-If we develop a discriminating model that is highly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
-
-What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decision will not effect if it actually rains.
-However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
-For this we need a (causal) model that allows us to predict the outcome under various different interventions.
-
-Identifying spam and automated essay marking systems are similar.
-The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label.
-Spammers and students will modify their writing in order to optimise their results.
-The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided.
-(What would the nature of the features have to be such that change did not occur?
-- they would have to be causes of the label).
-
-What about predicting the risk of death in patients with pneumonia?
-Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective this seems like a straightforward prediction problem.
-It is not.
-Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if we start using the model to decide whom to admit.
-(xxx et al) found exactly this effect.
-Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia.
-They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
-There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addition and treatment protocols.
-However, using it to decide on what basis to admit people could kill.
-The key is understanding exactly what question you are asking.
-In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.
-
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc).
-For all these problems the assumptions required to treat them as pure prediction problems are;
-\begin{enumerate}
-\item The treatment is assumed to be effective (at least better than nothing)
-\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
-\end{enumerate}
-
-Demand prediction seems relatively straightforward.
-These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
-It seems unlikely that using the model to ensure stock is available will itself change demand.
-However there is a potential data censoring issue.
-If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
-Changing availability does then change demand.
-
-Playing Go (and other games) is another case with some subtleties.
-At every turn, the AI agent has a number of actions available.
-The board state following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win.
-...
-this is interesting maybe come back to it ...
-an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ...
-One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken).
-When can we learn from the actions others have taken?
-When there is no confounding.
-And does this hold with Go?
-Probably because the board state encapsulates everything that should determine what move is played.
-Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)
-
-Having considered these examples we can now identify some general aspects of the problem that determine whether or not we require a causal model.
-
-\begin{itemize}
-\item Does acting on the predictions of the model change the mapping from features to target?
-(at least if the decision process is open to scrutiny).
-In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
-\item
-
- Covariate shift clearly comes in here.
-Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occurring and that results will be transferable.
-The mechanism is known but the function may be complex.
-Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models?
-Yes, if the way in which features get their values changes, then causal models will be invariant to that in a way that non-causal ones are not.
-
-\item To decide between actions we only need to rank them (not estimate their actual effect). 
-\item The predicted outcome in the absence of an intervention provides a single point.
-We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes.
-\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
-\item Does acting on the result of the prediction change the predictive distribution p(y|x)? IE change people's behaviour.
-\item Ethics - ... People's viewpoint on if its OK...
-\end{itemize}
-
-I hope these examples gave you a feel for the richness and subtitles of causal inference.
-We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
-
-\section{Approaches to causality}
-
-There are two broad approaches to deciding how to act.
-Reinforcement learning and causal inference.
-In reinforcement learning we estimate the effect of actions by taking them.
-We assume there is an agent capable of intervening in the system and try to find policies for the agent to follow in selecting actions so as to maximise some kind of reward.
-This is a very powerful and general framework (REFERENCES TO GENERAL AI).
-However, POINT OUT SOME OF THE DIFFICULTIES WITH REINFORCEMENT LEARNING.
-We are frequently presented with large bodies of data that have been collected from a system in which we did not have any control over what actions were taken.
-
-We will call data sets where we do not have control over the decision making process that generated the data observational.
-Versus experimental data sets, where we do have control (experimental data sets do not always have to be randomised -although that is a powerful approach to ensure we have control.
-There is a space in the middle where we have partially controlled the process by which agent select actions.
-Randomised data with imperfect compliance would be an example.
-
-An agent (capable of intervening in the system) chooses an action from those available The agent making the decision included in the model. This is
-
-Reinforcement learning addresses the problem of learning from explicitly taking actions. There
-is typically some state or environment. An agent chooses an action from those available in the
-current state. The state then evolves stochastically as a function of the selected action and
-the agent receives some feedback or reward that is a function of the new state. This setting
-differs from the standard classification problem in that the agent must learn from feedback on
-the selected action, rather than being presented with the correct action for a given state. A
-common modelling assumption is that the state evolves only as a function of the previous state
-and the action chosen, and given these, is independent of the previous history of states and
-actions. This is known as a Markov decision process or MDP. A particularly well studied and
-understood model is the singe state MDP. In this case, there are a set of actions, each associated
-with a fixed but unknown reward distribution and at each time step our agent selects an action
-and receives corresponding feedback. This is known as the multi-armed bandit problem.
-
-Causal inference makes use of assumptions to allow the outcome of actions to be predicted from
-observational data. The key to causal inference is a framework that can model how actions
-change the state of the world. This framework then allows us to map information collected in
-one setting to another.
-
-Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit
-algorithms deal with the sequential nature of the decision making process, causal inference with
-the problem that full randomisation is not always feasible, affordable or ethical. The similarities
-between the problems that these techniques have been developed to address raises the question
-of if there are problems best addressed by a combination of these approaches and how they
-can be combined. The goal of my thesis is to explore these questions. In the next sections I
-review the key literature in causal inference and bandits. I then present a general approach to
-how causal models might be incorporated into bandit settings and conclude by demonstrating
-an algorithm that leverages causal assumptions to improve performance in a specific bandit
-setting.
-
-There are two key approaches to causal problems.
-The first is to learn the outcome of actions by directly intervening in the system and seeing what happens.
-We then get feedback on how good those actions were.
-This is the approach taken in reinforcement learning.
-ADVANTAGES AND DISADVANTAGES OF THIS APPROACH.
-The second broad approach is causal inference.
-Here
-
-\todo{insert figure showing data generating process and observed data defining ML}\ref{all of statistics}
-
-\paragraph*{Two broad approaches} 
-\begin{itemize}
-\item Build a model to map the natural behaviour of the system to what will happen for some action
-\item Take the action and see what happens
-\end{itemize}
-
-\paragraph*{The first is causal inference}
-
-\paragraph*{The second is reinforcement learning}
-
-\paragraph*{Both generalise from randomised experiment} Reinforcement learning to sequential decisions, causal inference to non-experimental conditions
-
-Both these fields relate to the problem of making optimal decisions and both can be seen as generalising randomised controlled experiments.
-Causal inference  is the study of how to estimate the effect of an action in the absence of randomisation.
-Reinforcement learning studies how we can do better if the decisions are to be made sequentially.
-
-\paragraph*{Both approaches involve assumptions} the latter that we can group context and actions.
-
-\paragraph*{Limitations of causal inference}
-
-\paragraph*{Limitations of experiments} What are the issues with standard randomised experiments?
-
-\chapter{Causal models}
-\label{chap:causal_models}
-
-Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
-To do this we need terminology to describe actions and how we anticipate the system will respond to them.
-Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
-In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ.
-We will also use them to describe the following simplified examples.
-The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:ranomized_experiment}
-Suppose a pharmaceutical company wants to assess the effectiveness of a new drug on recovery from a given illness.
-This is typically tested by taking a large group of representative patients and randomly assigning half of them to a treatment group (who receive the drug) and the other half to a control group (who receive a placebo).
-The goal is to determine the clinical impacts of the drug by comparing the differences between the outcomes for the two groups (in this case, simplified to only two outcomes - recovery or non-recovery).
-We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
-\end{example}
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:adjusting}
-Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all 4 year olds.
-We have a large cross-sectional data set on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
-We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
-\end{example}
-
-\section{Causal Bayesian networks}
-
-Causal Bayesian networks are an extension of Bayesian networks.
-A Bayesian network is a graphical way of representing how a distribution factorises.
-Any joint probability distribution can be factorised into a product of conditional probabilities.
-There are multiple valid factorisations, corresponding to permutations of variable ordering.
-
-\eqn{
-\label{eqn:cbn:joint_dist}
-P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
-}
-
-We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}.
-If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network.
-There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.
-
-\begin{figure}[h]
-\centering
-\caption{A general Bayesian network for the joint distribution over three variables.
-This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
-\label{fig:bayesnet}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below left=of 1](2){$X_{2}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{figure}
-
-\begin{figure}
-\centering
-\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2|X_1}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
-\label{fig:bayesnet2}
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (3) edge (2)
-    (2) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, below left=of 1](2){$X_{1}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, above left=of 1](2){$X_{1}$};
-\node[main node, above right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (2) edge (1)
-    (3) edge (1)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true $G$ can have extra edges).
-If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
-
-\eqn{
-\P{X_{1},...,X_{N}} = \prod_{i = 1...N}\P{X_{i}|parents_{X_{i}}}
-}
-
-A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
-This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-We need some notation to describe interventions and represent distributions over variables in the network after an intervention.
-In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.
-
-\begin{definition}{The do-notation}
-\begin{itemize}
-\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
-\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$.
-This notation is somewhat overloaded.
-It may be used represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
-For example it could represent
-\begin{itemize}
-\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
-\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
-\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
-\item a family of density/mass function for $Y$ parameterised by $x$.
-\end{itemize}
-Where the distinction is important and not clear from context we will use one of the more specific forms above. 
-\end{itemize}
-\end{definition}
-
-\begin{theorem}[Truncated product formula \citep{Pearl2000}]
-\label{thm:truncated_prodcut}
-If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.
-
-\begin{equation}
-\label{eq:truncatedproduct}
-\P{X_1...X_N|do(Z=z)} =
-  \begin{cases}
-  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistent with $Z=z$}  \\
-   0       & \text{otherwise } 
-  \end{cases}
-\end{equation}
-\end{theorem}
-
-Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution.
-The truncated product formula will give different results depending on the selected network.
-The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
-In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.
-
-Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$?
-The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient.
-Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables.
-The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}.
-This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$.
-In this example, the interventional distribution is equivalent to the observational one.
-
-\begin{figure}[h]
-\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
-\label{fig:causal_network_example}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X \text{ (Treatment)}$};
-\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
-\path[]
-	(2) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool.
-We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$.
-It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation.
-If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}.
-In this case, the interventional distribution is not equivalent to the observational one.
-If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool.
-To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,
-
-\eqn{
-\label{eqn:backdoor_example}
-\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}
-}
-
-We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph.
-There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure.
-These rules are referred to as the Do-calculus \citep{Pearl2000}.
-We discuss them further in section \ref{sec:do_calculus}.
-
-\begin{figure}
-\center
-\caption{Causal Bayesian network for example \ref{exm:adjusting}}
-\label{fig:causal_adjust}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]
-
- %nodes
-\node[node](1){$Z \text{ (SES)}$};
-\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
-\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3); 
-\end{tikzpicture} 
-\end{figure}
-
-Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
-The power of defining an intervention in this way stems from the number of things that are invariant between the two processes.
-All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
-
-We could represent problems of the type where we try to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and what they have in common, see figure \ref{fig:causal_ordinary_bayes}.
-This does not require any special framework or notation.
-The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks.
-However, without a causal framework, we have to make assumptions about what will be invariant to the intervention specifically for each such problem we encounter.
-For complex problems, it is very difficult to conceptualise the assumptions we expect to hold without the benefit of a causal framework.
-
-\begin{figure}[h]
-\center
-\caption{Causal inference with ordinary Bayesian networks.
-The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$.
-The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
-As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
-However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
-\label{fig:causal_ordinary_bayes}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
-\tikzstyle{box}=[rectangle, draw=black!100]
- %nodes
-\node[node](1){$Z$};
-\node[node, below left=of 1](2){$X$};
-\node[node, below right=of 1](3){$Y$};
-\node[node, right=1.5cm of 3](4){$x$};
-\node[lt, above right=of 4](5){$Z'$};
-\node[lt, below right=of 5](6){$Y'$};
-\node[lt, above=2.0cm of 3](9){$\theta_{y}$};
-\node[lt, above=0.5cm of 9](7){$\theta_z$};
-\node[lt, above=2.0cm of 2](8){$\theta_x$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (6)
-    (5) edge (6)
-    (8) edge (2)
-    (7) edge (1) edge (5)
-    (9) edge (3) edge (6);
-\node[rectangle, inner sep=4.0mm, fit= (1) (2) (3),draw=black!100,label=below right:N] {};
-\node[rectangle, inner sep=4.0mm, fit= (5) (6),draw=black!100,label=below right:M] {};
-\end{tikzpicture} 
-\end{figure}
-
-A causal Bayesian network represents much more information than a Bayesian network with identical structure.
-A causal network encodes all possible interventions that could be specified with the do-notation.
-For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
-The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
-After each two variable intervention, we have a distribution over one variables which requires one parameter.
-This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
-
-Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across; the local Markov property states that variables are independent of their non-effects given their direct causes.
-Similarly the global Markov property and d-separation also hold in causal networks.
-
-\subsubsection{Limitations of causal Bayesian networks}
-A number of criticisms have been levelled at this approach to modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}.
-However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions.
-For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
-But what if some people failed to take the prescribed treatment.
-We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
-Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
-
-\begin{figure}[h]
-\caption{Randomised treatment with imperfect compliance}
-\label{fig:randomized_imperfect_compliance}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Prescribed Treatment};
-\node[node, below right=of 1](2){Actual Treatment};
-\node[node, right=of 2](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(1) edge (2)
-	(2) edge (3) 
-	(4) edge (3) edge (2);
-\end{tikzpicture}
-\end{figure}
-
-Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables.
-Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology.
-We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
-However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
-Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
-Such problems have mostly been studied within the dynamic systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
-\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
-
-\section{Counterfactuals}
-\label{sec:counterfactuals}
-
-The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
-Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
-
-In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would have happened if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would have happened if they got the placebo, $\cf{y_{i}}{0}$.
-The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, there are units that can be assumed to be identical or that will revert exactly to their initial state some time after treatment.
-This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
-
-Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo.
-Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate (from an experimental or observational study):
-\begin{itemize}
-\item $\P{Y|X=1}$, the probability that those who took the treatment will recover
-\item $\P{Y|X=0}$, the probability that those who were not treated will recover
-\end{itemize}
-
-Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
-For those who did not take the treatment, the observed outcome is the sames as the outcome \emph{had} they not taken the treatment.
-Equivalently stated:
-
-\eq{
-\P{Y^{0}|X=0}&= \P{Y|X=0}\\
-\P{Y^{1}|X=1}&=\P{Y|X=1}
-}
-
-If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
-
-\eq{
-\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
-\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
-}
-
-This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
-
-\eq{
-\P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
-}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}.
-They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
-In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$.
-Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$.
-A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
-
-\vspace*{.3cm}
-\begin{theorem}[Ignorability] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
-
-\eqn{
-\label{eqn:counterfactual1}
-\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
-\label{eqn:counterfactual2}
-\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
-}
-\end{theorem}
-
-Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
-Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
-
-Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}.
-This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
-As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
-The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
-SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
-
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-This makes it quite easy to (accidentally) make statements about counterfactuals that cannot be tested with empirical data.
-Consider the following example based on \citet{Dawid2000}.
-Again we have a drug where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
-Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal with equal variance (for simplicity).
-
-% vector with mean mu_t,mu_c and a covariance matrix
-\eqn{
-P(\cf{Y}{1},\cf{Y}{0}) \sim N( 
-         \begin{bmatrix}
-           \mu_{1} \\
-           \mu_2 \\
-         \end{bmatrix},
-         \begin{bmatrix}
-           \sigma^2 & \rho \sigma^2 \\
-           \rho \sigma^2 & \sigma^2 \\
-         \end{bmatrix}
-         )
-}
-
-Their difference is also normal. Let $\tau = \cf{Y}{1} - \cf{Y}{0}$,
-
-\eqn{
-P(\tau) = N(\mu_1 - \mu_0,2\sigma^{2}(1-\rho))
-}
-
-\begin{figure}[h]
-\captionsetup[subfigure]{position=b}
-\centering
-\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
-\label{fig:metaphysical_distribution_difference}
-\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$.
-The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated.
-\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
-\hspace{0.02\textwidth}
-\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
-{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
-\end{figure}
-
-From a (large) randomised controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}.
-These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated respectively.
-However, the distribution over the individual causal effects depends on $\rho$, see figure \ref{fig:metaphys_b}.
-The key problem is that we can never observe the joint distribution over $\cf{Y}{1}$ and $\cf{Y}{0}$.
-As a result, $\rho$ and thus the variance of $\tau$ is not identifiable, even from experimental data.
-\citet{Dawid} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects.
-He further points out that the interventional distributions in figure \ref{fig:metaphys_a}, along with a loss function, contain all the information required to decide how to treat a new patient.
-
-This result is unintuitive.
-It seems on the face of it that the distribution of individual causal effects is relevant to our decision making.
-If $\rho = 1$ then almost everyone benefits slightly from the treatment whilst if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.
-This confusion can be resolved by thinking about personalised rather than individual causal effects.
-It is entirely possible that potentially observable characteristics (such as gender, age, genetics, etc) affect how people will respond to the treatment.
-We can partition people into sub-populations on the basis of these characteristics and measure different \emph{personalised} causal effects for each group.
-The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much can be gained from further personalisation.
-The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).
-
-One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
-This is certainly convenient, if rather imprecise.
-However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention.
-It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).
-
-\section{Structural Equation models}
-\label{sec:SEM}
-
-Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Uncertainties are not inherent but arise from unmeasured variables.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \citep{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.
-
-\begin{figure}[h]
-\caption{SEM for example \ref{exm:ranomized_experiment}}
-\label{fig:sem_randomized_treatment}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(\epsilon_x)$};
-\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2);
-\end{tikzpicture}
-\end{figure}
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random.
-The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
-The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
-
-Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$.
-The function $f_y$ and distribution over $\epsilon_y$ does not change.
-This results in the interventional distribution \footnote{We have assumed the variables are discrete only for notational convenience},
-
-\eqn {
-\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
-}
-
-The observational distribution of $Y$ given $X$ is,
-
-\eqn{
-\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
-& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
-}
-
-The interventional distribution is the same as the observational one.
-The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
-
-The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
-Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
-
-\eqn{
-\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
-\label{eqn:sem:adjusting}
-&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
-}
-Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivalent to the observational distribution given by:
-
-\eqn{
-\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
-}
-
-\begin{figure}[h]
-\caption{SEM for example \ref{exm:adjusting}}
-\label{fig:sem:preschool}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(Z,\epsilon_x)$};
-\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
-\node[lt, above=of 6](7){$\epsilon_z$};
-\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2)
-	(7) edge (6)
-	(6) edge (2) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
-A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
-
-\section{Comparing and unifying the models}
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
-In other words, the network is a causal Bayesian network.
-All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
-Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}).
-Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
-We can also connect this to counterfactuals via:
-
-\begin{equation}
-\begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
-\end{aligned}
-\end{equation}
-
-The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
-
-\begin{equation}
-\label{eq:weakignore}
-X \ci Y^{0} \text{ and } X \ci Y^{1}
-\end{equation}
-
-It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}.
-The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
-For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores \citep{Rosenbaum1983} to estimate the effect.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
-These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
-
-In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
-The work on causal graphical models \citep{Pearl2000,Sprites} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field.
-Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
-In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
-
-\section{What does a causal model give us? Resolving Simpson's paradox}
-
-We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
-The following example is adapted from \citet{Pearl2000}.
-Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
-She keeps track of which medication her patients choose and whether or not the treatment is successful.
-She obtains the results in table \ref{tab:simpson1}.
-
-\begin{table}[h]
-\caption{Treatment results}
-\label{tab:simpson1}
-\center
-\begin{tabular}{c c c c c}
-Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9} 
-A &  87 & 13 & 100 & 87\%\\
-\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Drug A appears to perform better.
-However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
-
-\begin{table}[h]
-\caption{Treatment results by gender}
-\label{tab:simpson2}
-\center
-\begin{tabular}{c c c c c c}
-Gender & Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9}
-M & A & 12 & 8 & 20 & 60\%\\
-M & B & 56 & 24 & 80 & 70\%\\
-\rowcolor[gray]{0.9}
-F & A & 75 & 5 & 80 & 94\%\\
-F & B & 19 & 1 & 20 & 95\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Once the data is broken down by gender, drug B looks better for both men \emph{and} women.
-Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must recommend which to subsidise under a national health scheme).
-Should she choose A or B?
-The ambiguity in this question lies at the heart of Simpson's paradox.
-How does causal modelling resolve the paradox?
-The key is that the doctor is trying to choose between \emph{interventions}.
-She wants to know what the success rate will be if she changes her practice to give all the patients one drug, rather than allowing them to choose as currently occurs.
-
-Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
-The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
-Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
-Some assumptions about the causal relationships between the variables are required.
-In this example, it seems reasonable to conclude that gender may affect the treatment chosen and the outcome.
-Assuming there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
-
-\begin{figure}[h]
-\caption{An example of Simpson's Paradox}
-\label{fig:simpsons_paradox}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Gender};
-\node[node, below left=of 1](2){Treatment};
-\node[node, below right=of 1](3){Outcome};
-\path[]
-	(1) edge (2) edge (3) 
-	(2) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
-Knowing a patient took drug A indicates they are more likely to be female.
-When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
-However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
-So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
-Drug B is the better choice.
-
-\eqn{
-\P{Y|do(T)} = \P{Y|T,female}\P{female}+ \P{Y|T,male}\P{male}
-}
-
-Is the solution to Simpson's paradox to always to break down the data by as many variables as possible?
-No.
-Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
-This is a drug designed to prevent heart disease.
-One pathway to doing so might well be to lower blood pressure.
-Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
-It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
-Now, however, table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood pressure control and better overall outcomes.
-In this setting $\P{Y|do(T)} = \P{Y|T}$ and drug A is the better choice.
-
-\begin{figure}[h]
-\caption{An example of Simpson's Paradox}
-\label{fig:simpsons_paradox2}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Blood Pressure};
-\node[node, below left=of 1](2){Treatment};
-\node[node, below right=of 1](3){Outcome};
-\path[]
-	(1) edge (3) 
-	(2) edge (3)
-	(2) edge (1);
-\end{tikzpicture}
-\end{figure}
-
-Note that we have not changed the data itself, only the description of the variables that it is associated with.
-This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
-From a purely statistical viewpoint there is no paradox.
-The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
-The paradox only arises when we attempt to use the data to select an intervention and is resolved when we apply a causal approach to do so.
-
-\eqn{
-\label{eqn:simpsons_paradox}
-\exists \set{N_1,...N_4, N_1'...N_4'} \in \mathbb{N} : \; 
-\textcolor{green}{\frac{N_1}{N_2}< \frac{N_1'}{N_2'}}, \;\; 
-\textcolor{blue}{ \frac{N_3}{N_4}<\frac{N_3'}{N_4'}}
- \text{ and } \frac{N_1+N_3}{N_2+N_4} > \frac{N_1'+N_3'}{N_2'+N_4'}
-}
-
-\begin{figure}
-\centering
-\caption{Simpson's reversal visualised.
-The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
-However, when we sum them, the ratio is steeper for the un-primed variables.}
-\label{fig:simpsons_vector}
-\includegraphics[scale=.4]{figures/simpson_vector}
-\end{figure}
-
-There are many other plausible causal graphs for both scenarios above.
-Perhaps income affects drug choice as well as gender, or gender might affect treatment choice and blood pressure control given treatment, etc.
-Causal modelling provides a powerful tool to specify such assumptions and to determine how to estimate causal effects for a given model see section \ref{sec:causal effect estimation}.
-
-\chapter{Two key questions}
-\label{chap:two_key_questions}
-
-We can roughly categorise the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
-In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
-The goal is then to estimate the effect of an intervention or range of interventions in the system.
-Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}.
-Almost every time someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
-
-Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
-This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
-Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
-An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
-
-\section{Causal effect estimation} 
-\label{sec:causal effect estimation}
-
-Estimating causal effects from observational data comes down to determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
-We did this on an ad-hoc basis to resolve the examples discussed in chapter \ref{chap:causal_models}.
-In this chapter we describe a principled approach to mapping observational quantities to interventional ones and discuss some of the key issues involved in estimating such expressions from finite sample data.
-We assume the basic structure of the graph is known.
-That is, we assume that we can draw a network containing (at a minimum):
-
-\begin{itemize}
-\item the target/outcome variable we care about,
-\item the focus/treatment variables on which we are considering interventions,
-\item any variables which act to confound two or more of the other variables we have included, and
-\item any links between variables we have included. 
-\end{itemize}
-
-Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
-For example, consider estimating the impact of schooling on wages.
-Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
-Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
-
-How can the structure of the causal graph be leveraged to compute interventional distributions from observational ones?
-Given the graph corresponding to the observational distribution, the graph after any intervention can be obtained by removing any links into variables directly set by the intervention.
-The joint interventional distribution is the product of the factors associated with the interventional graph, as given by the truncated product formula \ref{eq:truncatedproduct}.
-If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
-However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
-
-The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
-By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \citep{Pearl2000}.
-The do calculus consists of three rules.
-They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.
-
-\subsection{Independence in Bayesian networks: D-separation}
-\label{sec:dsep}
-
-Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
-Therefor, in this section, we briefly review the key properties of Bayesian networks.
-A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
-Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
-
-\vspace{.3cm}
-\begin{theorem}
-\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
-\end{theorem}
-
-The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$.
-D-separation is an algorithm that extends the local Markov property to find these additional independencies.
-It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
-
-The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
-From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
-Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
-In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
-In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
-
-The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
-The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
-As an example, consider a scholarship available to female or disadvantaged students.
-Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship.
-There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
-However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
-\begin{figure}[h]
-\centering
-\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
-\label{fig:dseptriples}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y| Z$}
-\label{fig:dseptripple1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$X$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Y$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$X$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, below left =of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-        edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|\{\}$}
-\label{fig:dseptripple4}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, above left =of 1](2){$X$};
-\node[main node, above right =of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (1)
-    (3) edge node {} (1);	
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-\vspace{.3cm}
-\begin{definition}[unblocked path]
-\label{def:unblocked}
-A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
-It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
-\end{definition}
-
-\vspace{.3cm}
-\begin{definition}[d-separation]
-\label{def:dsep}
-The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, there are no unblocked paths from any $X \in \boldsymbol{X}$ to any $Y \in \boldsymbol{Y}$ after conditioning on $\boldsymbol{Z}$.
-\end{definition}
-
-\vspace{.3cm}
-\begin{theorem}[d-separation and conditional independence]
-\label{thrm:dsep_independence}
-If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in a Bayesian network $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
-Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected (not d-separated) given $\boldsymbol{Z}$ then it is possible to construct a distribution $P'$ that factorises over $G$ in which they are dependent.
-\end{theorem}
-
-Theorem \ref{thrm:dsep_independence} says that independencies implied by d-separation on a graph hold in every distribution that can be factored over that graph and that if ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in \emph{all} distributions that can be factored over $G$ then they are d-separated in $G$.
-If we denote the independencies implied by d-separation in a graph by $\indep{G}$ and the set of independencies in a distribution by $\indep{P}$ then $\indep{G} \subseteq \indep{P}$.
-
-If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
-However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
-A particular case in which this occurs is when there are deterministic relationships between variables.
-If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
-This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
-This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
-D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
-
-\section{The Do Calculus} 
-\label{sec:do_calculus}
-
-The do-calculus is a set of three rules \citep{Pearl1995} that can be applied to simplify the expression for an interventional distribution.
-If by repeated application of the do-calculus, along with standard probability transformations, we can obtain an expression containing only observational quantities then we can use it to estimate the interventional distribution from observational data.
-Let $\boldsymbol{X}$,$\boldsymbol{Y}$,$\boldsymbol{Z}$ and $\boldsymbol{W}$ be disjoint sets of variables in a causal graph $G$.
-We denote the graph $G$ after the an intervention $do(\boldsymbol{X})$, which has the effect of removing all edges into variables in $\boldsymbol{X}$, as $G_{\overline{\boldsymbol{X}}}$
-
-\subsubsection{Rule 1: (adding or removing evidence)}
-
-Rule 1 allows us to remove (or insert) observational evidence from the right hand side of a conditional interventional distribution.
-It follows directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still applies after an intervention.
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
-}
-
-\begin{figure}[h]
-\centering
-\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
-\label{fig:mutilatednet}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:mutilatedOriginal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2);
-
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.5\textwidth}
-\centering
-\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
-\label{fig:mutilatedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-\subsubsection{Rule 2: (exchanging actions with observations)}
-Rule 2 describes when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening, $do(\boldsymbol{X}=\boldsymbol{x})$, have the same effect on the distribution over $\boldsymbol{Y}$.
-Let $G_{\underline{\boldsymbol{X}}}$ denote the causal graph $G$ with all edges \emph{leaving} $X$ removed.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W})$ in $G_{\underline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W})
-}
-
-The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$.
-Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
-However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
-Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
-If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths.
-This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).
-
-Equation \ref{eq:Do22} does not cover cases where acting on one set of variables allows us to replace acting on another set with conditioning (see figure \ref{fig:rule2_2}).
-The general form of rule 2 is given in equation \ref{eq:Do2}.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W},\boldsymbol{Z})$ in $G_{\underline{\boldsymbol{X}}\overline{\boldsymbol{Z}}}$:
-\eqn{
-\label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),
-do(\boldsymbol{Z}=\boldsymbol{z}),
-\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W})
-}
-
-\begin{figure}[h]
-\centering
-\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
-In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
-If we condition on $W$, we block this indirect path.}
-\label{fig:rule2}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule2_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)  edge  (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule2_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1)	edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}}$}
-\label{fig:rule2_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    	edge  (3);
-
-\end{tikzpicture} 
-\end{subfigure}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\caption{An example of applying equation \ref{eq:Do2}.
-In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
-Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
-Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
-}
-\label{fig:rule2_2}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule22_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (1) edge (2)
-    (5) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{after $do(Z)$, $G_{\overline{Z}}$}
-\label{fig:rule22_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}\overline{Z}}$}
-\label{fig:rule22_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\end{figure}
-
-\subsubsection{Rule 3: (adding or removing actions)}
-This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-A simple case of rule 3 is given in equation \ref{eq:Do3_1}.
-If $\boldsymbol{Y}$ is independent of $\boldsymbol{{X}}$ in $G$ after removing links entering $\boldsymbol{X}$ then can be no direct path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and any intervention on $\boldsymbol{X}$ will not affect $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{{X}})$ in $G_{\boldsymbol{\overline{X}}}$:
-
-\eqn{
-\label{eq:Do3_1}
-\P{\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})} = \P{\boldsymbol{Y}}
-}
-
-The general case of rule 3 is easier to state by explicitly representing the intervention in the graphical model.
-Let $G^{\boldsymbol{\hat{X}}}$ denote the graph $G$ after adding a variable $\hat{X_i}$ as a parent of each variable $X_i \in \boldsymbol{X}$ (see figure \ref{fig:rule3_augmented}).
-The variable $\hat{X_i}$ can be thought of as representing the mechanism by which $X_i$ takes its value, either by being set via intervention or as a stochastic function of its other parents \citep{Koller2009}.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\boldsymbol{\hat{X}}}_{\boldsymbol{\overline{Z}}}$:
-\eqn{
-\label{eq:Do3}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-}
-
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
-These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
-
-\begin{figure}[h]
-\centering
-\caption{Example application of equation \ref{eq:Do3}.
-$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
-We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
-\label{fig:rule3}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule3_original}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{augmented graph $G^{\hat{X}}$}
-\label{fig:rule3_augmented}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
-\node[main node, above left=of 2](5){$\hat{X}$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3)
-    (5) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule3_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\end{figure}
-
-\subsection{Identifiability}
-A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
-This is the identifiablity problem.
-It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
-A query is non-parametrically identifiable if it is identifiable without  assumptions about the functional form of the dependencies between variables in the graph.
-
-\vspace*{.3cm}
-\begin{definition}[Non-parametric identifiablity]
-Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
-A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
-\end{definition}
-
-The question of non-parametric identifiability is solved.
-The do calculus is complete \citep{Shpitser2006b,Huang2006}.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
-There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
-
-\begin{enumerate}
-\item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network, and
-\item if it is identifiable, return the required expression. 
-\end{enumerate}
-
-Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
-I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
-
-\begin{figure}[h]
-\centering
-\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
-\label{fig:identify}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$X$};
-\node[ob, right=of 1](2){$Z$};
-\node[ob, right=of 2](3){$Y$};
-\node[lt,above=of 2](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid2_instrumental_variable}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[ob, above left=of 2](4){$Z$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1)
-    (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\end{figure}
-
-Many interesting questions relating to identifiability remain open.
-What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
-What if we assume various restrictions on the functional form of the relationships between the variables?
-Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
-A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
-
-Although identifiability is a natural and important question to ask, it does not partition causal questions into solvable and unsolvable.
-Estimators for identifiable queries can be slow to converge and we may be able to obtain useful bounds on causal effects in cases where point estimates are not identified.
-
-\section{Estimation}
-\subsection{Defining causal effects} 
-So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
-This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set.
-From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.
-
-\eqn{
-\label{eqn:utility_max}
-x* =\argmax_x \Esub{y\sim\P{Y|do(X=x)}}{{\cal{U}}(y)}
-}
-
-Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
-There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
-For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
-
-\eqn{
-\label{eqn:ACE}
-ACE = \E{Y|do(X = 1)} - \E{Y|do(X=0)}
-}
-
-Assuming the expectations in equation \ref{eqn:ACE} are well defined, the ACE captures the shift in the mean outcome that arises from varying $X$.
-It does not capture changes in variance or higher moments of the distribution.
-The ACE can be generalised to non-discrete interventions by considering the effect on the expectation of $Y$ of an infinitesimal change in $x$.
-If $X$ is linearly related to $Y$ then the ACE is constant and equivalent to the corresponding coefficient in the linear structural equation model.
-
-\eqn{
-\label{eqn:ACE_continuous}
-ACE(x) = \frac{d}{dx}\E{Y|do(X=x)}
-}
-
-The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}.
-Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
-However we can personalise the average causal effect by stating it with respect to some observed context.
-I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
-
-\eqn{
-PCE(z) = \E{Y|do(X=1),z} - \E{Y|do(X=0),z}
-}
-
-In some cases we may be interested in the average causal effect for some sub-group of the population.
-A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
-This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.
-
-\eqn{
-ATT = \Esub{z\sim \P{Z|x=1}}{Y|do(X = 1)} - \Esub{z\sim \P{Z|x=1}}{Y|do(X=0)}
-}
-
-Causal effects can also be written in terms of counterfactuals.
-The ACE is $\E{\cf{Y}{1} - \cf{Y}{0}}$.
-We could estimate the ratio of expectations $\frac{\E{\cf{Y}{1}}}{\E{\cf{Y}{0}}}$ instead of the difference.
-However, the quantity   $\E{\frac{\cf{Y}{1}}{\cf{Y}{0}}}$ depends on the joint distribution over the counterfactual variables $(\cf{Y}{1},\cf{Y}{0})$ and thus cannot be computed from the interventional distribution.
-
-Another way of conceptualising causal effects is as a property indicating the strength of the causal link between two variables.
-This notion is complex to formalise when the relationship between variables is non-linear.
-Suppose $Y = X \oplus Z$ with $P(Z=1)=\frac{1}{2}$, the interventional distributions over $X$ are identical after marginalising out $Z$.
-\citet{Janzing2013} propose a number of postulates that a notion of causal strength could satisfy, demonstrate why previous measures fail these postulates and propose an alternative based on information flow.
-
-\subsection{Estimating causal effects by adjusting for confounding variables}
-\label{sec:causal_effect_under_ignorability}
-Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors.
-Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds.
-Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds.
-This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}.
-The setting is also referred to as unconfounded.
-
-\vspace*{0.3cm}
-\begin{theorem}[The backdoor criterion]\citep{Pearl2000}
-Let $\boldsymbol{X}$, $\boldsymbol{Z}$ and $\boldsymbol{Y}$ be disjoint sets of vertices in a causal graph $G$.
-If $\boldsymbol{Z}$ blocks (see Definition \ref{def:unblocked}) for every path from $X_i$ to $Y_j$ that contains a link into $X_i$, for every pair $(X_i \in \boldsymbol{X},Y_j \in \boldsymbol{Y})$, and no node in $\boldsymbol{Z}$ is a decedent of a node in $\boldsymbol{X}$ then the backdoor criterion is satisfied and;
-
-\eqn{
-\label{eqn:backdoor}
-\P{\boldsymbol{y}|do(\boldsymbol{x})} =\sum_{\boldsymbol{z}} \P{\boldsymbol{y}|\boldsymbol{x},\boldsymbol{z}}\P{\boldsymbol{z}}
-}
-\end{theorem}
-
-The backdoor criterion derives from rule 2 of the do-calculus.
-Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
-There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
-The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
-Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}.
-Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
-For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
-
-\begin{figure}
-\centering
-\caption{}
-\label{fig:adjustment}
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{There can be multiple valid adjustment sets. $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.}
-\label{fig:adjust_multiple}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](2){$X$};
-\node[ob, right=of 2](3){$Y$};
-\node[ob,above=of 3](4){$Z_2$};
-\node[ob,above=of 2](5){$Z_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (5) edge (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{Conditioning on $Z$ opens the backdoor path $X-U_1-Z-U_2-Y$}
-\label{fig:m-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\end{figure}
-
-Given that a set of variables $\boldsymbol{Z}$ satisfies the backdoor criterion (or equivalently the conditional ignorability assumption), the interventional distribution is asymptotically identifiable and can be estimated from equation \ref{eqn:backdoor}.
-The expected value of $Y$ after the intervention $do(X=x)$ is given by equation \ref{eqn:expected_post_intervention} and the average causal effect for a binary intervention $x \in \set{0,1}$ is given by equation \ref{eqn:ace_ignoreable}.
-
-\eqn{
-\label{eqn:expected_post_intervention}
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}}
-}
-
-\eqn{
-\label{eqn:ace_ignoreable}
-ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
-}
-
-Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata.
-However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
-There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
-The key approaches are based on matching on covariates, propensity score methods and regression.
-We now examine these approaches from a machine learning perceptive.
-
-In standard supervised learning, we have a training set $(\vb{x_1},y_1),...,(\vb{x_n},y_n)$ assumed to be sampled i.i.d from an unknown distribution $\P{\vb{x},y} =\P{\vb{x}}\P{y|\vb{x}}$.
-The goal is to select a hypothesis $h \in \mathcal{H} : \mathcal{X} \rightarrow \mathcal{Y}$ such that, on unseen data $ \sim \P{\vb{x},y}$, $h(\vb{x})$ is close (by some metric) in expectation to $y$.
-In other words we wish to minimise the generalisation error $E_{out}(h)$,
-
-\eqn{
-E_{out}(h) = \Esub{(\vb{x},y)\sim \P{\vb{x}}\P{y|\vb{x}}}{L(h(\vb{x}),y)}
-}
-
-We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
-We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
-
-\eqn{
-E_{in}(h) = \frac{1}{n}\sum_{i=1}^n L(h(\vb{x}_i),y_i)
-}
-
-The VC-dimension of the hypothesis space provides (typically loose) bounds on the probability that $E_{out} >> E_{in}$.
-However, in practice, the generalisation error is usually estimated empirically from a hold-out set of the sample that was not used to train the model, or via cross-validation.
-
-In the causal effect estimation under ignorability, we have training data $(\vb{x}_1,\vb{z}_1,y_1),...,(\vb{x}_n,\vb{z}_n,y_n)$ sampled i.i.d from $\P{\vb{z}}\P{\vb{x}|\vb{z}}\P{y|\vb{x},\vb{z}}$.
-Estimating $\E{Y|do(\vb{X}=\vb{x})}$ corresponds to selecting a hypothesis $h \in \mathcal{H}:\mathcal{X} \times \mathcal{Z} \rightarrow \mathcal{Y}$ that minimises;
-
-\eqn{
-\label{eqn:causal_covariate_shift}
-E_{out} &= \Esub{(\vb{x},\vb{z},y)\sim \delta(\vb{x}-\vb{x'})\P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},\\
-&= \Esub{(\vb{z},y)\sim \P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},
-}
-
-\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
-If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
-\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
-Take the case where we have a binary intervention $x \in \set{0,1}$.
-Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$.
-More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.
-
-\subsubsection{Regression}
-
-The regression approach is to learn a function that is a good approximation to the output surface $\E{Y|X,Z}$.
-Let $f_1(z)= \E{Y|X=1,Z=z}$.
-The expectation of $Y$ after the intervention $X=1$ is then obtained by taking the expectation with respect to $Z$, $\E{Y|do(X=1)} = \Esub{z \sim \P{Z}}{\E{Y|X=1,z}}$.
-We can learn a parametric regression model $\hat{f}_1(z)$ via empirical risk minimisation.
-
-\eqn{
-\hat{f}_1(z) = h_1(z;\hat{\theta}_{obs}) \text{, where } \hat{\theta}_{obs} = \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)  \right]
-}
-
-This estimator is consistent with respect to the observational distribution.
-As the sample size tends to infinity, $\hat{\theta}_{obs}$ approaches the parameter within the hypothesis space that minimises the expected loss given data sampled from the observational distribution.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z|x=1}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-If the model is correctly specified such that $f_1(z) = h_1(z;\theta^*)$ for some  $\theta^* \in \Theta$ then the empirical risk minimisation estimate is consistent with respect to the loss over any distribution of $Z$  \citep{Sugiyama2007}, including the interventional one.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \theta^* = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-The average causal effect can then be estimated by,
-
-\eqn{
-\label{eqn:cef_empirical_regression}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left(\hat{f}_1(z_i) - \hat{f}_0(z_i)\right)
-}
-
-Regression thus has a direct causal interpretation if the parametric model is correctly specified and the covariates included form a valid backdoor adjustment set for the treatment variable of interest in the corresponding structural equation model.
-
-\subsubsection{Propensity scores}
-If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
-The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
-
-\begin{figure}[h]
-\centering
-\caption{
-Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
-In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
-If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$.
-We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points).
-In both cases the linear model is a good fit to the data.
-However, the resulting estimate of the causal effect is very poor for the lower values of $z$.
-}
-\label{fig:covariate_shift_adjust}
-\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
-\end{figure}
-
-A general approach to estimating the expectation of some function $f(\cdot)$ with respect to data from some distribution $\P{\cdot}$, when we have data sampled from a different distribution $\Q{\cdot}$ is importance sampling \citep{horvitz1952generalization,Koller2009}.
-\eqn{
-\label{eqn:importance_sampling}
-\Esub{\vb{v} \sim \P{\vb{v}}}{f(v)} = \Esub{\vb{v} \sim \Q{\vb{v}}}{f(v)\frac{\P{\vb{v}}}{\Q{\vb{v}}}}
-}
-
-This importance weighting approach can be applied to the covariate shift/average causal effect problem by weighting the terms in the empirical risk minimisation estimator \citep{Sugiyama2007}.
-
-\eqn{
-\label{eqn:importance_weighted_erm}
-\hat{\theta}_{iw} &= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{\P{z_i}\delta(x_i-1)}{\P{z_i}\P{x_i=1|z_i}}  \right] \\
-&= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{1}{e(z_i)}  \right],
-}
-
-where $e(\vb{z})$ is the propensity score, defined by \citep{Rosenbaum1983};
-\eqn{
-e(\vb{z})\equiv \P{x=1|\vb{z}}
-}
-
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
-Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
-
-The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$.
-\citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$.
-This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$.
-Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
-In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
-However in general, it must be estimated from data.
-Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
-\citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
-
-\eqn{
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}} = \Esub{z\sim \P{\vb{Z|x}}}{\E{Y|x,\vb{z}}\frac{1}{e(\vb{z})}}
-}
-
-\eqn{
-\label{eqn:simple_inverse_propensity}
-\hat{\tau}_{ip} = \frac{1}{n}\sum_{i=1}^n \left(\frac{\ind{x_i=1}y_i}{e(\vb{z}_i)} -\frac{\ind{x_i=0}y_i}{1-e(\vb{z}_i)}\right)
-}
-
-\subsubsection{Matching}
-There is a straightforward connection between matching and regression for causal effect estimation.
-If $h \in \mathcal{H} \implies h + a \in \mathcal{H}$ for any constant $a$ and $\hat{f}$ is selected by minimising empirical risk with an $L_2$ loss then $\frac{1}{n}\sum_{i=1}^n \ind{x_i=1}\hat{f}_1(z_i) = \frac{1}{n}\sum_{i=1}^n \ind{x_i=1}y_i$ \footnote{\citep{Imbens2004} state this holds for most implementations}, and equation \ref{eqn:cef_empirical_regression} can be re-written as:
-
-\eqn{
-\label{eqn:cef_regression_missing}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \hat{f}_0(z_i)\right) +\ind{x_i=0}\left(\hat{f}_1(z_i) - y_i \right)\right]
-}
-
-This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
-For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
-Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
-\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
-
-\eqn{
-\hat{\tau}_{match} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \frac{1}{k}\sum_{j \in J_k(i)}y_j\right) +\ind{x_i=0}\left(\frac{1}{k}\sum_{j \in J_k(i)}y_j - y_i \right)\right]
-}
-
-This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
-There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{cochran1973controlling,rosenbaum1985constructing}.
-Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
-An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.
-
-The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
-A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
-
-The significance of this should not be underestimated.
-Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
-With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
-\citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem.
-There is relatively little theory on model selection for estimating the propensity score.
-To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion.
-It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}.
-With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
-I have found no examples of such an approach.
-
-The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
-Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
-The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
-
-\section{Causal Discovery}
-
-%Cyclic models \citep{Hyttinen2012}
-
-We now move to the much more general problem of learning a causal graph from observational data.
-In this setting we make much broader assumptions about the structure of the graph.
-For example, that it is acyclic or that we have no unmeasured confounding variables.
-We do not assume the existence or directions of any links between the variables.
-Amazingly, it is possible to infer some aspects of causal structure with such general assumptions.
-The set of conditional independence in a non-experimental data set indicates some causal structures are more likely than others.
-In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
-These clues are the key to causal discovery algorithms.
-
-%There is also work on inferring causal structure from experiential data \ref{}. We discuss this further in section XXX.
-
-Causal discovery is a much grander goal than causal effect estimation given a known causal network.
-Arguably, if achieved, it would equate to the automation of scientific discovery.
-We need simply supply our algorithm with a vast collection of variables (regardless of their relevance to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make.
-Unfortunately, causal discovery is very hard.
-Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows exponentially with the square of the number of variables .
-
-% - some intuition here about the strength of the signal we get from causal discovery.
-Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}.
-Talk about the numerical as well as the statistical issues here.
-
-In the next sections we briefly survey the key approaches to causal discovery.
-We roughly divide the methods into those based on those that exploit the connection between the conditional independencies in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
-
-%This is not the only way we could have grouped them.
-There is also a separation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
-There are also hybrid methods \ref{}
-
-\subsection{Conditional independence based methods}
-One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
-For any Bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network.
-If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$.
-However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
-This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
-This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
-Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \citep{pearl1988probabilistic} for $P$.
-
-Faithfulness is an assumption.
-It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
-However, most distributions generated by a causal Bayesian network will be faithful to that network.
-For faithfulness to be violated, different causal effects must exactly balance one-another out.
-For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
-If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
-However, this independence is not stable.
-It would disappear under a small perturbation to any of the parameters.
-In discrete systems, violations correspond to the solutions to polynomial equations over values in the CPD tables and thus are a space of measure zero with respect to all possible distributions associated with the graph \citep{Koller2009}.
-
-Given the faithfulness assumption, our causal discovery problem reduces to finding the set of Bayesian networks that have exactly the dependency structure as we observe in $P$.
-This set can also be referred to as the Markov equivalence class compatible with $P$.
-
-\subsubsection{Without hidden common causes}
-The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of Bayesian networks we must consider.
-This assumption is referred to as causal sufficiency \citep{Sprites2000}.
-
-We will begin with a brute force algorithm (described as the SGS algorithm in \citet{Sprites2000} and IC algorithm in \citet{Pearl2000}).
-While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.
-
-\begin{table}[H]
- \begin{tabularx}{\textwidth}{X}
- \hline
-\rule{0pt}{2.5ex} 
- \textbf{The SGS (or IC) Algorithm}\\
- \hline
- \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
- \begin{enumerate}[itemsep=8pt]
-  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an un-directed link to form a complete graph.
-  \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent.
-If such a set (including the empty set) exists then $a$ and $b$ cannot be directly connected in $G$ so delete the link.
-  \label{alg:SGSexponential}
-
-  \item For all pairs of non-linked variables $(\alpha,\beta)$ with a common neighbour, $c$, if $c \notin \boldsymbol{S}_{\alpha,\beta}$, then $c$ must be a collider in the path $\alpha,c,\beta$ so  add arrows to direct the links $\alpha-c$ and $\beta-c$ towards $c$.
-  \label{alg:SGScolliders}  
-  \item Recursively try to orient any edges that remain un-directed to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
-  \label{alg:SGSfinal}
-\end{enumerate}\\
- \hline
-\end{tabularx}
-\end{table}
-
-The SGS algorithm utilises the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
-Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
-Unfortunately the number of conditional independence tests required for step \ref{alg:SGSexponential} grows exponentially (in the worst case) with the number of variables.
-Not only that, but for each edge that is in the true network, the algorithm will always tests all other possible subsets of variables.
-If the assumption that there are no hidden common causes or that the distribution is faithful are violated, step \ref{alg:SGScolliders} of the SGS algorithm can produce double headed arrows.
-
-The PC algorithm \citet{Sprites2000} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilise the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
-It also checks for low order conditional independence relations before higher order ones.
-This allows it to exploit any sparsity in the true network, leading to much better average case performance  \citep{Sprites2000} (although the worst case, where the true network is complete, is still exponential).
-With finite data, the order in which the links are considered can change the output (unlike for SGS).
-The effect of wrongly removing a link early on flows through to later conditional independence tests by changing which nodes are considered adjacent.
-
-\begin{table}[H]
- \begin{tabularx}{\textwidth}{X}
- \hline
-\rule{0pt}{2.5ex} 
- \textbf{The PC Algorithm}\\
- \hline
- \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
- \begin{enumerate}[itemsep=8pt]
-  \item As for SGS
-  \item \textbf{for} each link $a-b$:
-  \begin{itemize}[label={}]
-   \item $n = 0$
-   \item $\boldsymbol{A}_{a,b} = \{A_{1}...A_{j}\}$ be the set of nodes adjacent to $a$ and/or $b$
-   \item \textbf{while} $a$ and $b$ are connected and $n < j$:
-   \item 
-    	\begin{itemize}[label={}]
-    	\item \textbf{if} any subset of size $n$ of $\boldsymbol{A}$ makes $a$ and $b$ conditionally independent:
-    	\item \begin{itemize}[label={}]
-    			delete the link
-    		  \end{itemize}
-
-    	\item $n = n+1$
-    	\end{itemize}
-  \end{itemize}   
-  \item as for SGS
-  \item as for SGS
-\end{enumerate}\\
- \hline
-\end{tabularx}
-\end{table}
-
-The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution.
-Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects from the truncated product formula \ref{eq:truncatedproduct}.
-We can then bound the true causal effect by combining the results for the all the networks.
-This procedure is the IDA algorithm \citep{Maathuis2009} and has been found to outperform standard regularisation techniques at finding causal effects in a high-dimensional yeast gene expression data set \citep{Maathuis2010}.
-An implementation is available in the R package \citep{Kalisch2012}
-
-\subsubsection{With hidden variables}
-There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables.
-With an unknown number of hidden variables there are infinity many possible structures to search over.
-In addition, the space of causal networks is not closed under marginalisation.
-If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalising over $\boldsymbol{U}$, may not be faithful to any Bayesian network (see figure \ref{fig:DAGSnotclosed}).
-The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables).
-
-\begin{figure}
-\centering
-\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalising over $U$. }
-\label{fig:DAGSnotclosed}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){U}; 
-\node[main node, below left = of 1](2){B};
-\node[main node, below right = of 1](3){C};
-\node[main node, above left= of 2](4){A};
-\node[main node, above right = of 3](5){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    (5) edge node {} (3);
-
-\end{tikzpicture}
-\end{figure}
-
-\begin{theorem}
-\citet{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
-\end{theorem}
-
-Since we only care about the causal relationships between observed variables, it is sufficient to search over networks where any hidden variables have no parents and directly cause two of the observed variables.
-Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable.
-These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalisation and conditioning.
-
-For any DAG with latent (and selection) variables there is a unique MAG \citep{Richardson2002}.
-This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites}.
-The logic behind the algorithm is very similar.
-Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations.
-The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \citep{Spirtes1995}.
-Assuming there are no selection variables, the PAG can contain four types of link:
-
-\begin{enumerate}
-\item $X \rightarrow Y$, meaning $X$ causes $Y$
-\item $X \leftrightarrow Y$, meaning there is a latent variable that causes $X$ and $Y$.
-\item $X\ {\circ} {\rightarrow}\ Y$, either $X$ causes $Y$ or a latent variable causes both.
-\item $X\ {\circ} {-}  {\circ}\ Y$, either $X$ causes $Y$ or $Y$ causes $X$ or a latent variable causes both.
-\end{enumerate}
-
-The circles indicate where it is ambiguous if there should be an arrowhead (IE where there is one in some MAGs and not in others in the equivalence class).
-Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).
-
-\begin{figure}
-\centering
-\caption{FCI examples: true graph and FCI output}
-\label{fig:FCIExamples}
- \begin{subfigure}[t]{0.4\textwidth}
- \centering
- \caption{True DAG}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){U}; 
-\node[main node, below left = of 1](2){B};
-\node[main node, below right = of 1](3){C};
-\node[main node, above left= of 2](4){A};
-\node[main node, above right = of 3](5){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    (5) edge node {} (3);
-
-\end{tikzpicture}
- \end{subfigure}
- \begin{subfigure}[t]{0.4\textwidth}
-  \centering
-  \caption{Partial ancestral graph}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed}]
- %nodes
-
-\node[main node](1){A};
-\node[main node, below right=of 1](2){B};
-\node[main node, right=of 2](3){C};
-\node[main node, above right=of 3](4){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge[pil] node {} (2)
-    (2) edge node {} (3)
-    (3) edge node {} (2)
-    (4) edge [pil] node {} (3);    	
-\end{tikzpicture}
-\end{subfigure}
-\par\bigskip 
- \begin{subfigure}[t]{0.4\textwidth}
- \centering
- \caption{True DAG}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm, thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (4)
-    (2) edge node {} (1)
-    (3) edge node {} (1);    	
-\end{tikzpicture}
-\end{subfigure}
- \begin{subfigure}[t]{0.4\textwidth}
-  \centering
-  \caption{Partial ancestral graph}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={observed}]
- %nodes
-
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
-
- \path[every node/.style={font=\sffamily\small}]
- 	(1) edge node {} (4)
-    (2) edge [pil] node {} (1)
-    (3) edge [pil] node {} (1);    	
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \citep{Zhang2008}.
-More recently \citet{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
-\citet{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
-
-Latent variables can create constraints on the marginal distribution over the observed variables that cannot be expressed in term of conditional independencies.
-These generalised constraints can be expressed and leveraged within nested Markov models \citep{Richardson2012,Shipster2014}
-
-%\renewcommand{\arraystretch}{1.5}
-%\begin{tabular}{| c | c | p{4cm} | c | c | c |}
-%\hline
-%  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
-%  \hline
-%  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
-%  \hline
-%  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
-%  \hline
-%  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
-%  \hline
-%  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
-%  \hline
-%  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
-%  \hline
-%  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
-% \hline
-%\end{tabular}
-
-All the algorithms discussed in this section rely on being able to perform conditional independence tests.
-This is non-trivial with high dimensional data.
-If the functional relationship between the variables is linear with Gaussian noise then the network represents a multivariate normal distribution and a pair of variables are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero \citep{Koller2009}.
-Where the functions are non-linear one can apply kernelised independence tests \citep{Gretton2008,Zhang2012}
-
-\subsection{Discovery with functional models}
-The algorithms we have considered so far return a Markov equivalence class.
-They cannot distinguish between two models that result in the same set of conditional independence relations.
-Consider the very simple case where  there are only two variables and the possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
-These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
-No algorithm relying purely on conditional independence relations can separate these two cases.
-
-Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$.
-What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from?
-Recall the functional definition of causality (section \ref{sec:SEM}).
-There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \citep{Hoyer2009}, post-non-linear additive noise \citep{Zhang2008a} or linear models with non-Gaussian noise \citep{Hoyer2012}.
-
-The causal direction can also be identified via a connection between casual discovery and semi-supervised learning \citep{Janzing2012}.
-Suppose we are trying to learn $\P{Y|X}$.
-The goal of semi-supervised learning is to improve our estimate of $\P{Y|X}$ by leveraging additional data sampled from $\P{X}$.
-However, if the true causal model is $X \rightarrow Y$ then their is some function mapping values of $X$ to $Y$ which should be invariant to any changes in the input distribution $\P{X}$.
-Therefor $\P{X}$ should be independent of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning.
-However if the true causal model is $Y \rightarrow X$ then variations in the $\P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help.
-This assumption of independence of mechanism and input can also allow the identification of the causal direction in SEMs even where the functions are deterministic and invertable \citep{Daniusis2010}.
-\citet{Janzing2012a} leverage an information geometric viewpoint of the independence of mechanism and input to infer the causal direction between a pair of associated variables.
-
-Instead of positing a functional restriction on the relationship between variables and then developing theory to exploit that assumption, \citep{LopezPaz2014} propose learning what the causal relationship looks like from data.
-They assume there will be a difference between the relationship of $\P{X}$, $\P{Y}$ and $\P{X|Y}$ between $X \rightarrow Y$ versus $Y \rightarrow X$.
-Their algorithm requires a data set in which each row is itself a data set consisting of pairs of variables $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$.
-They use a kernel mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-data set and train an algorithm to learn the direction of causality.
-Unfortunately we do not have a large collection of data sets where the causal direction is known to train such a model.
-\citet{LopezPaz2014} instead use a simulated data set so their model will necessarily be based on the assumptions they make when generating the data.
-Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.
-
-\citep{Peters2014} have extended results from the bi-variate case to the multivariate setting.
-They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable.
-They build on this to develop an algorithm that allows the construction of causal graphs based on the additive noise assumption.
-
-\chapter{The interventionist viewpoint}
-
-The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it.
-There is an obvious alternative.
-Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens.
-There are three key differences between observing a system and explicitly intervening in it.
-Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from.
-Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}.
-Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias.
-Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning.
-For example, in a drug trial , assigning people a sub-optimal treatment has real world costs.
-This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain.
-This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.
-
-\section{Randomised experiments}
-Randomised controlled trials are often presented as the gold standard for determining causal effects.
-What is it about randomisation that makes it so important when it comes to causality?
-The graphical model for a randomised controlled experiment is shown in figure \ref{fig:random_experiment_network}.
-If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
-Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.
-
-\begin{figure}
-\caption{causal network for a randomised experiment}
-\label{fig:random_experiment_network}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-Randomisation does not ensure target and control group are exactly alike.
-The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables between the target and control groups in a finite data sample.
-However, the variance in the outcome, within both the target and control groups, also increases.
-The net result is increased variance (but not bias) in  the estimate of causal effects.
-
-Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest.
-If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can estimate the average causal effects directly from the differences in outcome across treatments.
-More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
-The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
-See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.
-
-\begin{figure}
-\caption{causal network for a stratified randomised experiment if the probability an individual is assigned a given treatment depends on some covariates.}
-\label{fig:random_experiment_network_stratified}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, above right=of 1](5){Covariates};
-\node[node, below right=of 5](3){Outcome};
-\node[lt, above=of 3](4){U};
-
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3)
-	(5) edge (3) edge (1);
-\end{tikzpicture}
-\end{figure}
-
-The benefit provided by randomisation in breaking the link between the treatment variable and any latent confounders should not be understated.
-The possibility of unobserved confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (there is no test for confounding).
-This means causal estimates from non-experimental data are always subject to the criticism that an important confounder may have been overlooked or not properly adjusted for.
-However, randomised experiments do have some limitations.
-
-\subsection{Limitations of randomised experiments}
-\label{subsec:limitations_of_experiment}
-
-The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
-There may be imperfect compliance, the treatment selected by the randomiser is not always followed, or output censoring, the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.\footnote{Non-compliance is a problem if the goal is to estimate the causal effect of the treatment on the outcome but not if the goal is to estimate the causal effect of prescribing the treatment.
-The latter makes sense in a context where the process by which people decide whether or not to take the treatment they have been prescribed is likely to be the same if the treatment were made available more generally beyond experimental trial.} See figure \ref{fig:random_experiment_network_imperfect_compliance} for a the graphical model of a randomised experiment with imperfect compliance.
-
-\begin{figure}
-\caption{causal network for a randomised experiment with imperfect compliance}
-\label{fig:random_experiment_network_imperfect_compliance}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Actual Treatment};
-\node[node,above left=of 1](5){Assigned Treatment};
-\node[node, above =of 5](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (5)
-	(5) edge (1)
-	(1) edge (3)
-	(4) edge (3) edge (1);
-\end{tikzpicture}
-\end{figure}
-
-It is not always possible or ethical to conduct a randomised controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
-When experimentation is possible, it is frequently difficult or expensive.
-This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
-In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest.
-The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
-
-\begin{figure}
-\begin{center}
-\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parachutes for the prevention of morbidity and mortality associated with falls from large heights.}
-\label{fig:parashute}
-\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
-\end{center}
-\end{figure}
-
-Finally, non-adaptive randomised experiments are not optimal from either an active or reinforcement learning perspective.
-As an experiment proceeds information is obtained about the expectation and variance of each intervention (or treatment).
-Fixed experimental designs cannot make use of this information to select which intervention to try next.
-This results in both poorer estimates for a fixed number of experimental samples and more sub-optimal actions during the course of the experiment.
-
-\section{Multi armed bandits}
-\label{sec:multi-armed-bandits}
-
-Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed.
-Their introduction is generally attributed to \citet{Thompson1993}.
-In its classic formulation \citep{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
-Each arm $i$ is associated with a fixed but unknown reward distribution.
-\footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
-The goal of the learner is to maximise the total reward they receive.
-This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
-
-\vspace{0.5cm}
-\begin{definition}[Stochasitic k-armed bandit problem]
-Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{y_1,...,y_k}$ be a joint distribution over the rewards for each action.
-The multi-armed bandit problem proceeds over $T$ rounds.
-In each round $t$,
-\begin{enumerate}
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{y}}$
-\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
-\end{enumerate}
-At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$.
-We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
-\end{definition}
-
-The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
-To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
-This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
-There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.}.
-\eqn{
-\label{eqn:raw_regret}
-\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
-}
-
-Both the rewards and the actions selected by the algorithm are random variables.
-The majority of work in the bandit literature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}.
-The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is referred to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
-A stochastic bandit algorithm is learning if it obtains pseudo-regret that is sub-linear in $T$.
-
-\vspace{0.5cm}
-\begin{definition}[Pseudo-Regret]
-\eqn{
-\label{eqn:pseudo_regret}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}} \\
-&= n\mu_{i^*} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
-}
-\end{definition}
-
-The regret is invariant to adding a constant to the expected rewards for all actions.
-However, it still depends on key characteristics of the reward distributions for each action.
-Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range.
-Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
-
-\subsection{Stochastic bandits: Approaches and results}
-
-The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators.
-The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards.
-As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
-The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
-This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
-Instead, the focus is on computing bounds on the expected regret.
-
-\begin{figure}[h]
-\centering
-\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples.
-This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim Bernoulli(0.2)$.
-In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$.
-$\E{\hat{\mu}} = 0.439$.
-The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping.
-Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded.
-This has some interesting potential real world implications.
-Early stopping of clinical trials is controversial.
-A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
-\label{fig:early_stopping}
-\includegraphics[scale=1]{extreme_early_stopping}
-\end{figure}
-
-There are a few key principles that are used to guide the development of bandit algorithms.
-The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
-A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
-This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.
-
-Another key approach is \emph{optimism in the face of uncertainty}.
-Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
-The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
-Thus selecting it yields either a good reward or useful information.
-
-\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family.
-They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient.
-However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm.
-\citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$.
-Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds.
-We now assume the rewards are bounded in $[0,1]$.
-The algorithm and regret bounds can be generalised to submission reward distributions, see \citet{Bubeck2012}.
-
-\begin{algorithm}[H]
-\caption{UCB-1}\label{alg:ucb-1}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} horizon $T$.
-\STATE Play each arm once. 
-\FOR{$t \in 1,\ldots,T$}
-\STATE Count the number of times each arm has been selected previously $n_{t,i} = \sum_{s=1}^{t-1} \ind{a_s = i}$
-\STATE Calculate the mean reward for each arm $\hat{\mu}_{t,i} = \frac{1}{n_{t,i}} \sum_{s=1}^{t-1}\ind{a_s = i}Y_t$
-\STATE Select arm $a_t \in \argmax_{i = \set{1,...,k}}\left(\hat{\mu}_{t,i}+\sqrt{\frac{2 \log{t}}{n_{t,i}}}\right)$
-\ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
-The problem dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
-
-\eqn {
-\label{eqn:ucb_problem_dependent_regret}
-\regret \leq & \sum_{i:\Delta_i > 0}\left(\frac{8\log(T)}{\Delta_i}+2 \right)
-}
-
-Somewhat unintuitively, the regret increases as the value of the arms gets closer together.
-This is because it becomes harder for the algorithm to identify the optimal arm.
-As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
-The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
-The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which is maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
-By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.
-
-\eqn{
-\label{eqn:ucb_worst_case_regret}
-\regret \in \bigo{\sqrt{kT\log(T)}}
-}
-
-\begin{figure}
-\caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
-The solid curve shows the mean (cumulative) regret, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
-The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
-\label{fig:ucb_regret_vs_delta}
-\includegraphics[scale=1]{ucb_regret_vs_delta}
-\end{figure}
-
-The form of the dependence on the number of arms $k$ and horizon $T$ differs between the problem dependent and worst case regret.
-The problem dependent regret grows linearly with the number of arms, $k$, and logarithmically with $T$.
-The difference stems from the fact the problem dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
-\citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by $\regret \in \bigomega{\sqrt{kT}}$
-
-Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
-This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \citep{Lattimore2015}.
-
-\begin{figure}
-\caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$.
-The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $bernoulli([.3,.7])$ over 1000 simulations.
-The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
-\label{fig:ucb_upperbound_vs_actual_performance}
-\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
-\end{figure}
-
-Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal.
-This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}.
-Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
-However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
-
-\subsection{Pure-exploration problems}
-Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
-In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
-The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.
-
-\vspace{0.5cm}
-\begin{definition}[Simple Regret]
-\eqn{
-\label{eqn:simple_regret}
-\simpleregret=\mu_{i^*} - \E{\mu_{\hat i^*}}.
-}
-\end{definition}
-
-The best arm identification problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialisation or exploitation phase.
-For example, many strategies might be assessed via simulation prior to one being selected and deployed.
-The worst case simple regret for a k-armed bandit is lower bounded by equation \ref{eqn:worst_case_simple_regret_lower_bound} (\citep{Bubeck2009}).
-
-\eqn{
-\label{eqn:worst_case_simple_regret_lower_bound}
-\simpleregret \in \bigo{\sqrt{K/T}}
-}
-
-%\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
-
-Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds. The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period.
-
-\subsection{Adversarial Bandits}
-
-Adversarial bandits, described by \citet{Auer1995}, are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
-Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$.
-As in the stochastic setting, the learner then receives reward only for the selected action.
-
-\vspace{0.5cm}
-\begin{definition}[Adversarial k-armed bandit problem]
-\label{def:adversarial-k-armed-bandit}
-Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
-\begin{enumerate}
-\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
-\end{enumerate} 
-\end{definition}
-
-Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
-In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
-
-For oblivious adversarial bandits we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
-We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
-
-\eqn{
-\label{eqn:pseudo_regret-adversarial}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
-}
-
-The policy (or algorithm) used by the learner is available to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
-This implies the adversary can ensure that any learner with a deterministic policy suffers regret $\bigo{T}$ by forecasting their entire sequence of actions.
-For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forecasts what action the learner will play in round 2, given they received a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
-This implies adversarial bandit policies must be sufficiently random to avoid such exploitation \footnote{The UCB algorithm, defined by algorithm \ref{alg:ucb-1}, is deterministic if the order in which arms are played during the first $k$ rounds is fixed and the method for selecting which arm to play when multiple-arms have the same upper-confidence bound is not-random (for example, select the arm one with the lowest index $i$).}
-
-The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case pseudo-regret of $\bigo{\sqrt{TK\log(T)}}$  \citep{Auer1995}.
-Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \citep{Audibert2009}.
-The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
-However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nonetheless obtain good problem dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
-
-Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, is not a good measure of an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings were the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
-Adversarial bandit algorithms may perform better in non-stationary settings than standard stochastic policies to the extend that they explore more (to avoid the adversary simulating their behaviour), however it is preferable to develop algorithms specifically for non-stationary settings (subject to assumptions about how rapidly or frequently rewards can change), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
-
-\begin{figure}[h]
-  \caption{The pseudo-regret can be negative if rewards are non-stationary.
-This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed Bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
-Figure (a) shows the expected rewards of each arm, and the proportion of time that arm-1 is played, as a function of time.
-The single best-arm is arm-1 as it has the highest expected reward (averaged over $t$).
-An oracle that selects arm-1 in every round obtains an expected reward of $5,000$.
-However, despite not being designed to do so, the UCB-algorithm can adapt to the changing reward distribution to obtain consistently higher rewards.
-The distribution of regret over the 1000 simulations is shown in figure (b).}\label{fig:non-stationary-bandit}
-    \begin{subfigure}[t]{0.49\textwidth}
-	\centering   
-	 \caption{Behaviour over time}\label{fig:non-stationary-bandit-time} 
-         \includegraphics[scale=.5]{non-stationary-bandit-time}
-\end{subfigure}
-    \begin{subfigure}[t]{0.49\textwidth}
-    \centering
-     \caption{Cumulative regret distribution}
-        \label{fig:non-stationary-bandit-regret}
-        \includegraphics[scale=.5]{non-stationary-bandit-regret}
-    \end{subfigure}
-\end{figure}
-
-\subsection{Contextual bandits}
-\label{sec:contextual-bandits}
-
-In the standard multi-armed bandit setting, each decision is identical and the goal is to learn a single best action.
-However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
-The best treatment to offer an individual patient could depend on their age, gender, disease sub-type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
-Similarly, decisions on which ad or content to display on a webpage or which product to recommend can be \emph{personalised} based on the previous behaviour of the user.
-A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
-Contextual bandits are a generalisation of multi-armed bandits that make use of this additional contextual information.
-The term contextual bandit was coined by \citet{Langford2008}, however close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
-
-\vspace{0.5cm}
-\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analogously to definition \ref{def:adversarial-k-armed-bandit}}]
-\label{def:contextual-bandit-problem}
-Let $\P{\vec{x},\vec{y}}$ be the joint distribution over the rewards for each action and some context $\vec{X} \in \mathcal{X}$. In each round $t \in \set{1,...T}$, 
-\begin{enumerate}
-\item the world stochastically generates the vector of rewards for each action and the context, $(\vec{X}_t,[\cfb{Y}{1},...,\cfb{Y}{k}]) \sim \P{\vec{x},\vec{y}}$ and reveals $\vec{X}_t$ to the learner
-\item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps,
-\item the learner observes and receives (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
-\end{enumerate}
-\end{definition}
-
-Standard multi-armed bandits learn to select the action $a$ that, with high probability, maximises $\E{Y|a}$.
-Contextual bandits learn to select actions that maximise $\E{Y|\vec{x},a}$.
-The reward for contextual bandits should  be compared to an oracle that acts optimally based on the context.
-To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothesis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
-The pseudo-regret is the difference between the expected regret obtained by an oracle that selects actions based on the single best hypothesis or policy $h$ at each timestep, and the expected reward obtained by the algorithm.
-
-\eqn{
-\label{eqn:regret_contextual}
-\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{\cfb{Y}{h(\vec{X}_t)}}} - \E{\sum_{t=1}^T{\cfb{Y}{A_t}}}
-}
-
-If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a separate standard bandit instance for each value of the context.
-This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothesis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occurred such that $\sum_{c=1}^N T_c = T$.
-The regret is maximised if $T_c = T/N$ resulting in total regret $\bigo{N\sqrt{kT/N}}$.}.
-This is optimal with respect to this class of hypothesis.
-However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the in the relationships between different contexts and actions.
-As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
-For example, with $n$ binary variables, $N = 2^n$}.
-We expect some form of smoothness; values of context that are similar should lead to comparable rewards for a given action.
-We need algorithms that can leverage such assumptions.
-
-An alternate reduction to the standard bandit problem, which allows us to constrain the hypothesis space to explore, is to treat each hypothesis $h$ as a bandit arm \citep{Langford2008}.
-At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
-Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothesis class considered, limiting our ability to learn any complex mappings from context to actions.
-The key problem with this approach is each sample is used to update our knowledge about only one hypothesis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothesis simultaneously.
-
-Suppose that, at each timestep $t$, after selecting an action, the learner received the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
-This is known as the full information setting.
-In this case, the learner can simulate running each hypothesis over the history to compute the reward it would have obtained and use the hypothesis with the best empirical reward to select the next action.
-This is the \emph{follow the leader} algorithm, which obtains optimal regret $\bigo{\sqrt{T log(\card{\mathcal{H}})}}$ for the full-information problem \citep{cesa2006prediction}.
-Unfortunately, in the contextual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
-As in causal effect estimation, we can view this as a missing data problem.
-However, the data is missing not at random because which component of the reward is observed depends on the action selected which in turn is a function of the previous history of actions and rewards.
-
-The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly separating exploration from exploitation.
-Epoch-greedy is an explore-exploit algorithm.
-It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothesis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
-The hypothesis with the highest empirical reward is then used to select actions for the remaining time steps.
-The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
-
-The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
-However, it involves maintaining a list of weights for each hypothesis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
-The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
-
-Both Epoch-greedy and ILOVECONBANDITS involve solving problems of the form,
-
-\eqn{
-\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{X}_t) = A_t}
-}
-
-This expression equates to identifying the best empirical policy based on previous data.
-The algorithms assume the existence of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
-The computational tractability of these algorithms on large (or infinite) hypothesis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classification problem \citep{Dudik2011a}.
-
-Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalised versions of the UCB or Thompson sampling algorithms.
-For linear pay-off models, both approaches yield algorithms with strong regret guarantees, \emph{Lin-UCB} \citep{li2010contextual} and \emph{Generalised Thompson Sampling} \citep{Agrawal2013}.
-
-It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
-However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
-\citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information.
-This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
-
-\begin{figure}[h]
-\centering
-\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
-$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
-Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
-To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
-}
-\label{fig:contextual-bandit-causal-graph}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ causes $\vec{Y}$}
-\label{fig:context-graph-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$Y$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (4) edge (1)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{Y}$ causes $\vec{X}$}
-\label{fig:context-graph-anti-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$Y$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (4)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ and $\vec{Y}$ are confounded}
-\label{fig:context-graph-confounded}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$Y$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
-\node[lt, below=of 4](5){$U$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2)
-   (5) edge (4) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\end{figure}
-
-\subsection{Learning from logged bandit data}
-Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
-In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
-The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
-
-This problem differs from the contextual bandit problem in that the learner is not interacting with the environment.
-As a result, there is no exploration-exploitation trade-off to be made.
-However, the problem does not reduce to supervised learning, because the label, $y$ is not the desired .
-In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
-The majority of the literature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
-\citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original estimator $\pi$ did not depend on $\vec{X}$.
-
-If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, discussed in section \ref{sec:causal_effect_under_ignorability}.
-The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ satisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
-The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivalent to $do(A=a)$.
-However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see \citet{Pearl2000}, section 4.2.
-In this case, letting $\Pns{\pi'}{a|\vec{x}}$ denote the distribution over actions under policy $\pi'$ given context $\vec{x}$, the expected (per round) reward obtained by $\pi'$ is given by,
-
-\eqn{
-\E{y|\pi'} = \E{y|do(a \sim \pi'(\vec{x}))} = \Esub{(\vec{x},a) \sim \P{\vec{x}}\Pns{\pi'}{a|\vec{x}}}{y|\vec{x},a}
-}
-
-As in estimating average causal effects under ignorability, we have a covariate shift problem, with training data sampled from $\P{\vec{x}}\Pns{\pi}{a|\vec{x}}\P{y|\vec{x},a}$ but generalisation error measured with respect to $\P{\vec{x}}\Pns{\pi'}{a|\vec{x}}\P{y|\vec{x},a}$.
-A difference in practice, is that in the applications frequently considered under learning from logged feedback data, such as ad serving or recommender systems, there may be substantial information available about $\pi$, in the best case, $\Pns{\pi}{a|\vec{x}}$ is known.
-This makes estimators utilising inverse propensity weighting, including doubly robust estimators as in \citet{Dudik2011}, more attractive.
-
-\begin{figure}
-\centering
-\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$}
-\label{fig:off-policy-evaluation}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
-\node[main node](1){$\vec{X}$};
-\node[main node, below right=of 1](2){$Y$};
-\node[main node,  below left=of 1](3){$A$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2) edge (3);
-\end{tikzpicture} 
-\end{figure}
-
-\citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the empirical best because the variance of the estimators for some policies may be much higher than for others.
-
-\subsection{Adding structure to actions}
-
-The classic multi-armed bandit is a powerful tool for sequential decision making.
-However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \citep{filippi2010parametric}, dependent bandits \citep{Pandey2007}, X-armed bandits \citep{Bubeck2010} and Gaussian process bandits \citep{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback\citep{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \citep{Piccolboni2001,Bartok2014}.
-
-In the next chapter, I propose a very natural connection between causal graphs and bandit problems and show it induces a novel form of additional feedback and structure between arms that cannot be can exploited by any of these previous approaches.
-
-\chapter{Causal Bandits: Unifying the approaches}
-
-\section{The framework}
-\label{sec:intro_causal_bandit}
-A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
-Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{fig:unify_frameworks} for a simple example.
-In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example the set of single variable interventions.
-The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node.
-We refer to these problems as \emph{causal bandit problems}.
-In this thesis I focus on the case where the causal graph is known.
-Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.
-
-\begin{figure}
-\caption{A simple causal graphical model and corresponding complete action space. $W$ and $Z$ represent binary variables that can be intervened on and $Y$ represents the reward.}
-\label{fig:unify_frameworks}
-\centering
-\begin{subfigure}[c]{0.3\textwidth}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
-\node[main node](1){W};
-\node[tg, below right=of 1](2){Y};
-\node[main node, above right=of 2](3){X};
-\path[]
-	(1) edge (2)
-	(3) edge (2);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-$\actionspace$ = \begin{tabular}{|c|}
-	\hline
-  $do(W=0,Z=0)$ \\
-  $do(W=0,Z=1)$ \\
-  $do(W=1,Z=0)$ \\
-  $do(W=1,Z=1)$ \\
-  \hline
-  $do(W=0)$ \\
-  $do(W=1)$ \\
-  $do(Z=0)$ \\
-  $do(Z=1)$ \\
-  $do()$ \\
-  \hline
-\end{tabular}
-\end{subfigure}
-\end{figure}
-
-The type of problem we are concerned with is best illustrated with an example.
-Consider a farmer wishing to optimise the yield of her crop.
-She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
-In each season the farmer has enough time and money to intervene and control at most one of these
-variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry.
-When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
-How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
-
-We will assume each variable only takes on a finite number of distinct values.
-(The path to relaxing this assumption would be through levering the work on continuous armed bandits).
-The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
-An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and
-the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
-
-\vspace*{.5cm}
-\begin{definition}[Causal bandit problem]
-A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
-Each action $a \in \actionspace$ assigns a value to a subset of the variables in $\mathcal{X}$.
-One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
-
-The causal bandit game proceeds over $T$ rounds. In each round $t$, the learner:
-\begin{enumerate}
-\item \emph{observes} the value of a subset of the variables $\vec{X}^c_t$,
-\item \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations, and finally
-\item \emph{observes} sampled values for another subset of variables $\vec{X}^o_t$ drawn from $\P{\vec{X}^o_t | do(\vec{X}_t = \vec{x}_t)}$ including the \emph{reward} $Y_t \in \{0,1\}$.
-\end{enumerate}
-
-We refer to the set of variables that can be observed prior to selecting an action $\vec{X}^c$ as contextual variables and the set of variables observed after the action is chosen, $\vec{X}^o$, as post-action feedback variables.
-Note that $\vec{X}^c$ and $\vec{X}^o$ need not be disjoint.
-A variable may be observed both prior to and after the agent selects an action, and the action may change its value.
-The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
-
-\end{definition}
-
-The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
-If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
-If we receive post-action feedback on additional nodes the problem can be more interesting.
-In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
-Consider again the model in figure \ref{fig:unify_frameworks}.
-The causal structure implies:
-
-\eq{
-P(Y|do(W = 0)) &= P(Y|do(),W = 0) \\
-&= P(Y|do(X=0),W=0)P(X=0)+P(Y|do(X=1),W=0)P(X=1) 
-}
-
-\begin{figure}
-\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level .
-The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
-For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
-\label{fig:cholesterol_graph}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
-\node[node](1){Health conscious};
-\node[node, below left=of 1](2){Diet};
-\node[node, below right=of 1](3){Exercise};
-\node[node, below right=of 2](4){Weight};
-\node[tg,below left=of 4](5){Y=Cholesterol};
-\node[node, below =of 5](6){Test result};
-\node[node, above right=of 3](7){Free time};
-\path[]
-	(1) edge (2) edge (3)
-	(2) edge (4) edge (5)
-	(3) edge (4)
-	(5) edge (6)
-	(7) edge (3);
-\end{tikzpicture}
-\end{figure}
-
-Thus we gain information about the reward for the action $do(W=0)$ from selecting the action $do()$ or $do(X = x)$ and then observing $W = 0$.
-We only get this form of side information for actions that don't specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-If additional variables are only observed before an intervention is selected the causal bandit problem reduces to stochastic contextual bandits, which are already reasonably well understood \citep{Agarwal2014}.
-
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
-The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.
-However, the number of actions or arms grows exponentially with the number of variables in the graph making it important to develop algorithms that leverage the graph structure and additional observations.
-
-\section{Causal bandits with post action feedback}
-
-We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, and where the goal of the learner is to minimise the simple regret.
-I presented this work at NIPS 2016 \citep{NIPS2016_6195}.
-
-\paragraph{Related Work} As alluded to above, causal bandit problems can be treated as classical multi-armed bandit problems by simply ignoring the causal model and extra observations and applying an existing best-arm identification algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
-However, as we show in \S\ref{sec:simple-regret}, ignoring the extra information available in the non-intervened variables yields sub-optimal performance.
-
-Our framework bears a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
-However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
-
-There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on un-played actions are revealed according to a feedback graph.
-As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
-They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
-The partial monitoring approach taken by \citet{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
-\citet{Yu2009} utilise extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
-\citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled.
-The main difference from our present work is our focus on simple regret and the more complex information linking rewards for different arms via causal graphs.
-To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.
-
-%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
-They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
-% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \citep{Bartok2014}.
-Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
-
-Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
-\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been 
-intervened on can provide important contextual information. Their work differs in many ways. For example, the focus is on the cumulative regret and
-the context is observed before the action is taken and cannot be controlled by the learning agent.
-
-%\todof{write one sentence to explain how their work is different to ours}
-
- \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions.
-Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
-Combining their handling of  causal induction with our analysis is left as future work.
-% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
-Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
-(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
-It is easy to come up with a counter example in which it does not).
-
-The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{Bottou2013a}, 
-where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discuss some of the issues 
-encountered in sequential design, but do not give an algorithm or theoretical results for this case.
-
-\subsection{The parallel bandit problem}
-\label{sec:simple-regret}
-In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem which we call the {\it parallel bandit}.
-It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro_causal_bandit}, including the farming example.
-It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
-
-The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:parallel}.
-All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
-
-In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high).
-The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.
-
-\begin{figure}
-  \caption{Causal Models}\label{fig:causalmodels}
-    \begin{subfigure}[t]{0.34\textwidth}
-	\centering   
-	 \caption{Parallel graph}\label{fig:parallel} 
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-\end{tikzpicture} 
-\end{subfigure}
-    \begin{subfigure}[t]{0.25\textwidth}
-    \centering
-     \caption{Confounded graph}
-        \label{fig:causalStructure_confounded}
-        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_1$};
-\node[main node, below left=of 1](2){$X_2$};
-\node[main node, below right=of 1](4){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-    (1) edge (4)
-    (2) edge (4);
-\end{tikzpicture}
-    \end{subfigure}
-    \begin{subfigure}[t]{0.4\textwidth}
-    \centering
-     \caption{Chain graph}
-        \label{fig:causalchain}
-         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, right=of 4](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-  	(2) edge (3)
-    (3) edge (4)
-    (4) edge (5);
-\end{tikzpicture}
-    \end{subfigure}
-\end{figure}
-
-In each round the learner either purely observes by selecting $do()$ or sets the value of a single variable.
-The remaining variables are simultaneously set by independently biased coin flips.
-The value of all variables are then used to determine the distribution of rewards for that round.
-Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
-
-The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
-$r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
-In the farming example, this choice of $Y$ models the success or failure of a seasons crop, 
-which depends stochastically on the various environment variables.
-
-\paragraph{The Parallel Bandit Algorithm}
-%\label{sub:par-bandit-alg}
-The algorithm operates as follows.
-For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
-Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
-The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates of their returns could be poor.
-To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions.
-The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$).
-For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$.
-Define
-
-\eq{
-\label{eq:m-simple}
-m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
-}
-
-\begin{algorithm}[H]
-\caption{Parallel Bandit Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} Total rounds $T$ and $N$.
-\FOR{$t \in 1,\ldots,T / 2$}
-\STATE Perform empty intervention $do()$
-\STATE Observe $\vec{X}_t$ and $Y_t$
-\ENDFOR
-\FOR{$a = do(X_i = x) \in \actions$}
-\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
-\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
-\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
-\ENDFOR
-\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
-\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
-\FOR{$a = do(X_i = x) \in A$}
-\FOR{$t \in 1,\ldots,T_A$}
-\STATE Intervene with $a$ and observe $Y_t$
-\ENDFOR
-\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
-\ENDFOR
-\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
-\end{algorithmic}
-\end{algorithm}
-
-$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimise the worst-case simple regret.
-When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
-We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilises the samples captured during the observational phase to estimate $m(\vec{q})$.
-Although very simple, the following two theorems show that this algorithm is effectively optimal.
-
-\begin{theorem}\label{thm:uq-simple}
-Algorithm \ref{alg:simple} satisfies
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.
-}
-\end{theorem}
-
-\begin{theorem}\label{thm:lower}
-For all strategies and $T$, $\vec{q}$, there exist rewards such that
-$\displaystyle \simpleregret 
-\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)$.
-\end{theorem}
-
-The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
-of $\hat p_a$ and $\hat m$ about their true values and may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
-
-%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
-By utilising knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
-Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$.
-Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting.
-In practice, you would combine the data from both phases to estimate rewards for the low probability actions.
-We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.
-
-\subsection{General graphs}
-\label{sec:simple-regret-general}
-We now consider the more general problem where the graph structure is known, but arbitrary.
-For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorisation formula \citep{Pearl2000}.
-\eq{
-\P{X_1...X_N|do(X_i=j)} = 
-\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
-} 
-where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the Dirac delta function.
-
-We could naively generalise our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorisation to 
-write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
-estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
-from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
-takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
-simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
-high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
-Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
-get a good estimate for $\P{Y|do(X_2=1)}$.
-
-To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
-allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
-are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
-or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
-start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
-Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
-Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
-
-\begin{algorithm}[H]
-\caption{General Algorithm}\label{alg:general}
-\begin{algorithmic}
-\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
-\FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $a_t$ from $\eta$
-\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
-\ENDFOR
-\FOR{$a \in \calA$}
-\STATE
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
-}
-\ENDFOR
-\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
-\end{algorithmic}
-\end{algorithm}
-
-Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
-Let $\parents{Y}(X)$ denote the realisation of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
-
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}\,, 
-}
-
-where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently.
-The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
-
-The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimised to minimise the worst-case simple regret. Abusing notation we define $m(\eta)$ by
-\eq{
-m(\eta) = \max_{a \in \calA} \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,,\text{ where } \EEa \text{ is the expectation with respect to } \Pn{a}.
-}
-
-We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
-
-\begin{theorem}\label{thm:general}
-If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$
-
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
-}
-\end{theorem}
-
-The proof is in Section \ref{sec:thm:general}.
-
-Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
-Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$.
-Thus it has broader applicability to the parallel graph than the generic algorithm given here.
-We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to mini-max optimal, but leave lower bounds
-for future work.
-
-\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$.
-In light of Theorem \ref{thm:general}
-a natural choice of $\eta$ is the minimiser of $m(\eta)$.
-\eq{
-\eta^* 
-= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
-}
-Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
-Therefore the minimisation problem may be tackled using standard techniques from convex optimisation.
-The quantity $m(\eta^*)$ may be interpreted as the minimum achievable worst-case variance of the importance weighted estimator.
-In the experimental section we present some special cases, but for now we give two simple results.
-The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
-
-\begin{proposition}\label{pro:m-bound}
-$m(\eta^*) \leq |\calA|$. \textit{Proof.} 
-\textup{By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|\,\forall a$.}
-\eq{
-m(\eta) 
-= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
-= \max_a \EEa\left[\frac{1}{\eta_a}\right] = |\calA| %\qedhere
-}
-\end{proposition}
-
-The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$.
-This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
-The proof is in section \ref{sec:m-equivelence}.
-
-\begin{remark}\label{rem:truncate}
-The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
-often possible to choose $B_a$ significantly
-larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
-and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{Bottou2013a} who also use importance weighted estimators
-to learn from observational data.
-\end{remark}
-
-\subsection{Experiments}
-We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
-Thomson sampling and UCB are optimised to minimise cumulative regret.
-We apply them in the fixed horizon, best arm identification setting by running them up to horizon $T$ and then selecting the arm with the highest empirical mean.
-The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.
-
-Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms).
-$Y_t \sim \bernoulli(\frac{1}{2}+\epsilon)$ if $X_1=1$ and $Y_t \sim \bernoulli(\frac{1}{2}-\epsilon')$ otherwise, where $\epsilon' = q_1\epsilon/(1-q_1)$.
-This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
-We set $q_i = 0$ for $i \leq m$ and $\frac{1}{2}$ otherwise.
-Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
-For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
-The code is available from \url{<https://github.com/finnhacks42/causal_bandits>}
-
-In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
-The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure.
-For the causal algorithms it increases approximately with $\sqrt{m}$.
-As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.
-
-\begin{figure}[h]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
-    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment2_20161020_1249.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
-        \label{fig:simple_vs_T_vary_epsilon}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment3_20161020_1252.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
-    		\label{fig:simple_vs_T}
-    \end{subfigure}
-    \caption{Experimental results}
-    \label{fig:experiments}
-\end{figure}
-
-Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
-This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
-In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
-The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.
-
-For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
-This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be mini-max optimal asymptotically \citep{LMS14}.
-
-\begin{figure}[H]
-	\centering    
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
-\node[main node, above right=of 2](6){$Z$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5)
-    (6) edge (1) edge (2) edge (4);
-\end{tikzpicture}
-        \caption{Confounded graph}
-        \label{fig:parallel_confounded} 
-\end{figure}
-
-We now compare the general algorithm with a range of standard bandit algorithms on the confounded graph in Figure \ref{fig:parallel_confounded}.
-All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action,
-
-\eq{
-\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}
-}
-
-We choose this setting because it generalises the parallel bandit, while simultaneously being sufficiently simple that we can compute the exact reward and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
-As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.
-
-In Figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$. For some $2 \leq N_1 \leq N$ we define 
-\eq{
-P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
-P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
-}
-As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon P(X_1=1) / P(X_1=0)$.
-The value of $N_1$ determines $m$ and ranges between $2$ and $N$.
-The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
-
-In Figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.  
-In Figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1, but exclude actions on $Z$ from the set of allowable actions to demonstrate that Algorithm 1 can fail in the presence of a confounding variable, which occurs because it incorrectly assumes that $P(Y|do(X)) = P(Y|X)$.
-We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
-
-\eq{
-P(X_i = 1|Z = 0) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.2 & \text{ if } i = 7 \\
-.7 & \text { otherwise} 
- \end{cases}\\
- P(X_i = 1|Z = 1) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.8 & \text{ if } i = 7 \\
-.3 & \text { otherwise} 
- \end{cases}\\
-}
-
-In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree.
-It is sub-optimal to act on either $X_7$ or $X_N$, while all other actions are optimal.
-The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.
-All remaining variables are likely to be identified as balanced and estimated from observational estimates.
-The CPD values have been chosen to demonstrate the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
-
-\begin{figure}[H]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
-    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
-        \label{fig:simple_vs_T_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment5_20161023_2118.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m(\eta*)=4.3$ with no actions setting $Z$}
-    		\label{fig:simple_vs_T_misspecified}
-    \end{subfigure}
-    \caption{Experimental results on the confounded graph}
-    \label{fig:experiments_confounded}
-\end{figure}
-
-\subsection{Discussion \& Future work}
-\label{sec:causal_bandit_discussion}
-Algorithm~\ref{alg:general} for general causal bandit problems 
-estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
-Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ where 
-the parameter $m(\eta)$ measures the difficulty of learning the causal model and is always less than $N$.
-The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.
-
-The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator $\hat{\mu}_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $\Pn{a}{\parents{Y}}$ for all $a \in \calA$.
-In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.
-Despite these extra estimates, Theorem~\ref{thm:lower} shows that this approach is optimal (up to log factors).Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
-
-\paragraph{Making Better Use of the Reward Signal}
-Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
-In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
-Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
-This would likely improve the poor performance of our causal algorithm relative to the successive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
-
-For the parallel bandit the required modifications should be quite straightforward.
-The idea would be to adapt the algorithm to essentially use successive elimination in the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability.
-In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
-Note that these modifications will not improve the mini-max regret, which at least for the parallel bandit is already optimal.
-For this reason we prefer to emphasise the main point that causal structure should be exploited when available.
-Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
-This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.
-
-\paragraph{Cumulative Regret}
-Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
-In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information
-to get near-optimal cumulative regret guarantees.
-They consider a finite-armed bandit model with side information where in reach round the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
-In this way the learner can gain information about actions it does not take with varying levels of accuracy.
-The reduction follows by substituting the importance weighted estimators in place of the Gaussian reward.
-In the case that $\vec{q}$ is known this would lead to a known variance and the only (insignificant) difference is the Bernoulli noise model.
-In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
-at least asymptotically.
-
-%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
-%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
-%\todom{Check this!}
-%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
-
-The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each time step the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
-The feedback graph is distinct from the causal graph.
-A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
-For this parallel bandit problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
-However, \citet{Alon2015,Kocak2014} give adversarial algorithms, which when applied to the parallel bandit problem obtain the standard bandit regret.
-A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
-This is equivalent to a nominally stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
-
-\citet{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
-In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
-\citep{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-\paragraph{Causal Models with Non-Observable Variables}
-If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
-non-observable variables. Some of the interventional distributions may be non-identifiable meaning we can not obtain prior estimates for $\Pn{a}{\parents{Y}}$ from 
-even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
-\ref{alg:general} cannot be used. Estimating these quantities while simultaneously minimising the simple regret is an interesting and challenging open problem.
-
-% For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
-% In this case, some conditional distributions may be non-identifiable. 
-% The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
-% We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
-% Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
-
-% More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
-In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
-The quality of the estimate notably depends not only on the number of times that action was selected.
-For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$.
-This more complex side information does not fit within the graph feedback framework.
-
-\paragraph{Partially or Completely Unknown Causal Graph}
-A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
-The latter case arises naturally if we assume free access to a large observational data set, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\citep{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
-In particular, \citet{Hu2014} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
-Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational data set would be much more challenging.
-
-\subsection{Proofs}
-\subsubsection{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}
-
-Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$. The assumption is non-restrictive since all variables
-are independent and permutations of the variables can be pushed to the reward function.
-
-The proof of Theorem \ref{thm:uq-simple} requires some lemmas. \
-
-\begin{lemma}\label{lem:conc1}
-Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
-\eq{
-\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \citet{hagerup1990guided}),
-
-\eq{
-\P{\left|\hat q_i - q_i\right| \geq \epsilon} \leq 2e^{-\frac{T\epsilon^2}{6q_i}}
-}
-
-Letting $\delta = 2e^{-\frac{T\epsilon^2}{6q_i}}$ and solving for $\epsilon$ completes the proof.
-
-\end{proof}
-
-\begin{lemma}\label{lem:m_est}
-
-Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
-\eq{
-\P{2m(\vec{q}) / 3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})} \geq 1 - \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-Let $F$ be the event that there exists and $1 \leq i \leq N$ for which
-\eq{
-\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}\,.
-}
-Then by the union bound and Lemma \ref{lem:conc1} we have $\P{F} \leq \delta$. The result will be completed by showing that
-when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
-From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m(\vec{q})$ that $q_i \geq q_m \geq 1/m(\vec{q})$ and so by Lemma \ref{lem:conc1} we have
-\eq{
-\frac{3}{4} 
-&\geq \frac{1}{2} + \sqrt{\frac{3}{T} \log \frac{2N}{\delta}} 
-\geq q_i + \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}} 
-\geq \hat q_i \\
-&\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
-\geq q_i - \sqrt{\frac{q_i}{8m(\vec{q})}}
-\geq \frac{1}{2m(\vec{q})}\,.
-}
-Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m(\vec{q})$.
-For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m(\vec{q})$ that
-\eq{
-\hat q_i 
-\leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
-\leq \frac{1}{m(\vec{q})} \left(1 + \sqrt{\frac{1}{8}}\right)
-\leq \frac{3}{2m(\vec{q})}\,.
-}
-Therefore $m(\vec{\hat q}) \geq 2m(\vec{q}) / 3$ as required. 
-\end{proof}
-
-\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
-Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then,
-for $a \in A$, the algorithm estimates $\mu_a$ from $T_A \doteq T/(2m(\vec{\hat q}))$ samples.
-From lemma \ref{lem:m_est}, $T_A  \geq T/(4m(\vec{q}))$ with probability $(1-\delta)$.
-Let $H$ be the event $T_A < T/(4m(\vec{q}))$ and $G$ be the event $\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}$
-
-\eq{
-\P{G} \leq \P{H} + \P{G|\neg H} \leq \delta + \P{G|\neg H}
-}
-
-Via Hoeffding's inequality and the union bound,
-
-\eq{
-\P{G|\neg H}&\doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}},\text{ given }T_A \geq T/(4m(\vec{q}))} \leq \delta\\
- \implies  \P{G} & \doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq 2\delta\,.
-}
-For arms not in $A$,
-
-\eq{
-\hat p_a = & \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq 1/m(\vec{\hat q}), \; \text {by definition of not being in $A$}\\
- \geq & \frac{1}{2m(\vec{q})},\; \text{ with probability } 1-\delta \\
-\implies  T_a \doteq & \sum_{t=1}^{T/2} \ind{X_i = j} \geq  \frac{T}{4m(\vec{q})},\; \text{ with probability } 1-\delta
-}
-
-Again applying Hoeffding's and the union bound
-
-\eq{
-\P{\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}}} \leq 2\delta
-}
-
-Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 4\delta$ that,
-\eq{
-(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}} \doteq \epsilon\,.
-}
-If this occurs, then 
-\eq{
-\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_{\hat a^*_T}] 
-& \leq 4\delta + \epsilon\\
-& \leq \frac{8m(\vec{q})}{T} + \sqrt{\frac{2m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ letting } \delta = \frac{2m(\vec{q})}{T}\\
-& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jenson's Inequality}
-}
-which completes the result.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LOWER BOUND
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}
-
-We follow a relatively standard path by choosing multiple environments that have different optimal arms, but which cannot all be statistically
-separated in $T$ rounds.
-Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
-For each $i$ define reward function $r_i$ by
-\eq{
-r_0(\boldsymbol{X}) &= \frac{1}{2} &
-r_i(\boldsymbol{X}) &= \begin{cases}
-\frac{1}{2} + \epsilon & \text{if } X_i = 1 \\
-\frac{1}{2} & \text{otherwise}\,,
-\end{cases}
-}
-where $1/4 \geq \epsilon > 0$ is some constant to be chosen later.
-We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
-environment determined by $\boldsymbol{q}$ and $r_i$. Let $\operatorname{P}_i$ be the corresponding measure
-on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \citet{Tsy08} we have
-\eq{
-\Prz{\hat a^*_T = a^*} + \Pri{\hat a^*_T \neq a^*} \geq \exp\left(-\KL(\operatorname{P}_0, \operatorname{P}_i)\right)\,,
-}
-where $\KL(\Ps_0, \Ps_i)$ is the KL divergence between measures $\operatorname{P}_0$ and $\operatorname{P}_i$.
-Let $T_i(T) = \sum_{t=1}^T \ind{a_t = do(X_i = 1)}$ be the total number of times the learner intervenes on variable $i$ by setting it to $1$.
-Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $\Ps_0$ and $\Ps_i$ may be bounded using the telescoping property (chain rule) and
-by bounding the local KL divergence by the $\chi$-squared distance as by \citet{Auer1995}. This leads to 
-\eq{
-\KL(\Ps_0, \Ps_i) 
-&\leq 6\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] 
-\leq 6\epsilon^2 \left(\EE_0 T_i(T) + q_i T\right) 
-\leq 6\epsilon^2 \left(\EE_0 T_i(T) + \frac{T}{m}\right)\,.
-}
-Define set $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
-Then for $i \in A$ and choosing $\epsilon = \min\set{1/4, \sqrt{m/(18T)}}$ we have
-\eq{
-\KL(\Ps_0, \Ps_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
-}
-Now $\sum_{i=1}^m \EE_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
-Therefore
-\eq{
-\sum_{i \in A} \Pri{\hat a^*_T \neq a} 
-\geq \sum_{i \in A} \exp\left(-\KL(\Ps_0, \Ps_i)\right) - 1
-\geq \frac{|A|}{e} - 1 
-\geq \frac{m}{2e} - 1\,.
-}
-Therefore there exists an $i \in A$ such that
-$\Pri{\hat a^*_T \neq a^*} \geq \frac{\frac{m}{2e} - 1}{m}$. 
-Therefore if $\epsilon < 1/4$ we have
-\eq{
-R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{m}{18T}}\,.
-}
-Otherwise $m \geq 18T$ so $\sqrt{m/T} = \Omega(1)$ and
-\eq{
-R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{1}{4} \frac{\frac{m}{2e} - 1}{2m} \in \Omega(1) 
-}
-as required.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% GENERAL-GRAPH UPPER BOUND
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Proof of Theorem \ref{thm:general}}\label{sec:thm:general}
-
-\begin{proof}
-First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
-We define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$, $R_{at} = R_a(X_t)$ and $\Pn{a}{.} = \Pns{a}{.}$.
-By definition we have $|Z_{at}| \leq B_a$ and 
-\eq{
-\Var_Q[Z_{at}] 
-\leq \EE_Q[Z_{at}^2] 
-\leq \EE_Q[R_{at}^2]
-= \EEa[R_{at}]
-= \EEa\left[\frac{\Pns{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq m(\eta)\,.
-}
-Checking the expectation we have
-\eq{
-\EE_Q[Z_{at}] 
-= \EEa \left[Y \ind{R_{at} \leq B_a}\right] 
-= \EEa Y - \EEa \left[Y\ind{R_{at} > B_a}\right] 
-= \mu_a - \beta_a\,,
-}
-where 
-\eq{
-0 \leq \beta_a = \EEa[Y \ind{R_{at} > B_a}] \leq \Pns{a}{R_{at} > B_a}
-}
-is the negative bias. 
-The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
-\eq{
-\beta_a \leq \Pns{a}{R_{at} > B_a} \leq \frac{\EEa[R_{at}]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
-}
-Let $\epsilon_a > 0$ be given by
-\eq{
-\epsilon_a = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
-}
-Then by the union bound and Bernstein's inequality 
-\eq{
-\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} 
-\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
-}
-
-Let $I = \hat{a}^*_T$ be the action selected by the algorithm, $a^* = \argmax_{a \in \calA} \mu_a$ be the true optimal action and recall that $\EE_Q[Z_{at}] = \mu_a - \beta_a$.
-Assuming the above event does not occur we have,
-
-\eq{
-\mu_I \geq \hat \mu_I - \epsilon_I  
-\geq \hat \mu_{a^*} - \epsilon_I  
-\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
-}
-By the definition of the truncation
-we have
-\eq{
-\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
-}
-and
-\eq{
-\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
-}
-Therefore for $C = \sqrt{2} + 4$ we have
-\eq{
-\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
-}
-as required.
-\end{proof}
-
-\subsubsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}\label{sec:m-equivelence}
-
-\begin{proposition} In the parallel bandit setting,
-$m(\eta^*) \leq 2m(\boldsymbol{q})$.
-\end{proposition}
-
-\begin{proof}
-
-Recall that in the parallel bandit setting,
-
-\eq{
-\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}
-}
-
-Let:
-
-\eq {
-\eta_a = \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}}\frac{1}{2m(\boldsymbol{q})} \text { for } a \in do(X_i = j)
-}
-
-Let $D =\sum_{a\in do(X_i=j)}\eta_a$. From the definition of $m(\boldsymbol{q})$, 
-\eq {
-\sum_{a\in do(X_i=j)} \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}} \leq m(\boldsymbol{q}) \implies D \leq \frac{1}{2}
-}
-
-Let $\eta_a = \frac{1}{2} + (1-D)$ for $a = do()$ such that $\sum_{a \in \calA}\eta_a = 1$
-
-Recall that,
-
-\eq{
-m(\eta) &
-= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]
-}
-
-We now show that our choice of $\eta$ ensures $\EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \leq 2m(\boldsymbol{q})$ for all actions $a$.
-
-For the actions $a: \eta_a > 0$, ie $do()$ and $do(X_i = j):\P{X_i=j}<\frac{1}{m(\boldsymbol{q})}$,
-\eq{
-\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq \EEa\left[\frac{\Pn{a}{X_1...X_N}}{\eta_a\Pn{a}{X_1...X_N}}\right] = \EEa\left[\frac{1}{\eta_a}\right] \leq 2m(\boldsymbol{q})
-}
-
-For the actions $a :\eta_a = 0$, ie $do(X_i=j):\P{X_i=j}\geq\frac{1}{m(\boldsymbol{q})}$,
-\eq{
-\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq & \EEa\left[\frac{\ind{X_i=j}\prod_{k\neq i}\P{X_k}}{(1/2+D)\prod_k \P{X_k}}\right] \\=& \EEa\left[\frac{\ind{X_i=j}}{(1/2+D)\P{X_i = j}}\right]
-\leq  \EEa\left[\frac{\ind{X_i=j}}{(1/2)(1/m(\boldsymbol{q}))}\right] \leq 2m(\boldsymbol{q})
-}
-
-Therefore $m(\eta*) \leq m(\eta) \leq 2m(\boldsymbol{q})$ as required.
-
-\end{proof}
-\
-%in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
-%Recall in that setting that 
-%\eq{
-%\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
-%}
-%For $a = do()$ choose $\eta_a = 1/2$. 
-%Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
-%Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
-%It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.
-
 \chapter{More stuff}
@@
 the PC algorithm. (COPIED FROM SPRITES)
-
-\bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{library}% write the directory to the .bib file
-\end{document}

### 2017-07-09 thesis/thesis_map.tex
minor edits

@@
 \usepackage[colorinlistoftodos]{todonotes}
-\usepackage[numbers]{natbib}
+\usepackage[numbers]{natbib} %numbers
 \usepackage{wrapfig}
@@
 As a result, $\rho$ and thus the variance of $\tau$ is not identifiable, even from experimental data.
-\citet{Dawid} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects.
+\citet{Dawid2000} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects.
 He further points out that the interventional distributions in figure \ref{fig:metaphys_a}, along with a loss function, contain all the information required to decide how to treat a new patient.
@@
 In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
-The work on causal graphical models \citep{Pearl2000,Sprites} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
+The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
 The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
@@
 This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
-There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{cochran1973controlling,rosenbaum1985constructing}.
+There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
 Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
@@
 For any DAG with latent (and selection) variables there is a unique MAG \citep{Richardson2002}.
-This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites}.
+This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites2000}.
 The logic behind the algorithm is very similar.
@@
 
-The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{Bottou2013a}, 
+The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{Bottou2013}, 
 where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discuss some of the issues 
@@
 larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
-and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{Bottou2013a} who also use importance weighted estimators
+and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{Bottou2013} who also use importance weighted estimators
 to learn from observational data.
@@
 \subsection{Experiments}
-We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{audibert2010best}, Thompson Sampling and UCB under a variety of conditions.
+We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{Audibert2010}, Thompson Sampling and UCB under a variety of conditions.
 Thomson sampling and UCB are optimised to minimise cumulative regret.
@@
 For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
-This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be mini-max optimal asymptotically \citep{LMS14}.
+This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be mini-max optimal asymptotically \citep{li2014minimax}.
 
@@
 The latter case arises naturally if we assume free access to a large observational data set, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\citep{Eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
+Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\citep{eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
 In particular, \citet{Hu2014} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
@@
 \begin{proof}
-By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \citet{hagerup1990guided}),
+By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \citet{Hagerup1990}),
 
@@
 
-\chapter{More stuff}
-%\chapter{List of software programs for causal inference}
-
-\begin{itemize}
-\item Pcalg. A library in R that implements ...
-\item Tetrad.
-\item There has to be a library for the Bayesian trees thing
-\item Linear regression (you can do this anywhere)
-
-\end{itemize}
-
-Notorious paper Automated Inferences of Criminality using Face Images.
-- What possible actions could we take?
-Ban people with most risky 'face types' from critical jobs?
-But dependent on causal structure, conditioning on the fact that people are in a position to apply for such jobs could reverse the dependency.
-Ones facial appearance is a readily observable characteristic and there is evidence that people implicitly make assumptions on the basis of it.
-Consider trials where people with 'more criminal' faces got longer sentences for similar crimes.
-%\chapter{Causality \& Interpretability}
-
-%\chapter{Causality \& Fairness}
-
-Generalisation error weights towards regions where density of x is high (it does not care about minorities).
-
-As machine learning is incorporated into decision making systems that have fundamental impacts on people’s lives, such as in employment, criminal justice, health and financial services, there are increasing concerns over transparency and fairness \citep{WMD, etc}.
-Realisation that machine learning algorithms can be inherit biases from the data we feed into them and the choices those building them make about what variables to include, etc.
-
-The European Union’s General Data Protection Regulation \citep{Goodman2016}\todo{check}, due to come into effect in 2018, requires that people can obtain "meaningful information about the logic involved" in an automated decision process .
-It also stipulates that such decisions should not be based on special categories of personal data (related to ethnicity, political and religions affiliation or sexuality) unless there "suitable measures" to safeguard individual interests.
-A key concern is discrimination against disadvantaged groups.
-
-Discrimination is frequently defined in terms of either disparate treatment- treating otherwise similar individuals differently on the basis of a protected attribute such as race or gender, or disparate impact - a process that yields a significant difference in outcome between groups.
-We consider how the notions of disparate treatment and impact may be formalised and the implications of how this is done for machine learning.
-We examine the overlap between the motivations for interpretable and causal models, especially with respect to assessing fairness.
-We look at how causal models mitigate some of the trade-offs between transparency and predictive accuracy and we examine to what extent the causal relationship between an attribute, any protected attributes and the outcome of interest is relevant to assessing the impact on fairness of its inclusion in a machine learning model.
-
-The lack of part-time work in tech could be argued to constitute indirect discrimination against women. 
-A fear that advertising themselves as supporting flexible working options would attract candidates who lacked drive and ambition.
-Cultural bias that favours those driven by monetary ambition against those
-
-Stability - a non causal model can't tell us about a simple do type operation on  a single variable.
-
-More broadly, we could draw a model representing the system now that could tell us the result of an intervention on a particular variable (such as setting gender), but the system itself could change (for example with customer preferences).
-There is a notion that a true causal model should be invariant for all time.
-
-Connection to Simpson's paradox.
-But what should we condition on?
-\citep{Romie2010} argue that looking at the department level is the correct viewpoint, since this the point at which hiring decisions were made.
-However, it could be argued that if one department was attracting a large number of higher quality candidates, the overall size of that department should be higher and that the results seen at Berkeley reflected a bias in favour of male dominated fields.
-
-Different measures of discrimination - is there a difference between groups (after conditioning on xxx)
-
-To understand how we measure and penalise discrimination, we need to take a step back and ask what are the underlying motivations for fairness?
-
-These differences in underlying cause suggest differences in the approach we should take to remedy them.
-
-%\section{How does discrimination arise in ML models}
-\begin{itemize}
-\item bias in historical decisions fed in as training data
-\item selection bias in input data, due to deliberate or implicit discrimination such as stop and search
-\item deliberate manipulation on the part of the person building the model.
-\item The hardest case is historical disadvantage, creating genuine differences between relevant attributes of groups.
-\item in bias in the label - what happens if you hire minorities but then your staff treat them badly and as a result they under-perform or leave.  
-\end{itemize}
-
-%\section{Defining Descrimination}
-
-Define discrimination in terms of causal effect of protected variable in decision making process.
-
-\citep{Romei2012} note the relevance of causal inference on discrimination analysis.
-
-“the central question in any employment- discrimination case is whether the employer would have taken the same action had the employee been of 7A recurring problem known as the omitted-variable bias.
-A multidisciplinary survey on discrimination analysis 9
-a different race (age, sex, religion, national origin etc.) and everything else had been the same” (Carson v. Bethlehem)
-
-Let us mathematically define disparate treatment and disparate impact with respect to a statistical model.
-We will focus on discrete variables for notational simplicity.
-Assume we have an outcome of interest $Y$, protected attributes $X$, other covariates $Z$ and a (potentially stochastic) model $f$ that maps $\{x, z\}$ to $y_f \in Y$.
-For a given model $f$, we have a distribution over the predicted outputs given the inputs, $\P{Y_f|X,Z}$
-\vspace{.2cm}
-\begin{definition}{Disparate Impact}: A model, $f$, that produces a predictive distribution $\P{Y_f|X,Z}$ has disparate impact if the marginal distribution over the predicted outcome, $Y_f$, depends on a protected attribute.
-\eqn{
-\exists \set{x_1,x_2} \subset X: \sum_z{\P{Y_f|Z,x_1}\P{Z|x_1}} \neq \sum_z{\P{Y_f|Z,x_2}\P{Z|x_2}}
-}
-\end{definition}
-
-\begin{definition}{Disparate Treatment}: A model yields disparate treatment if people with identical attributes (excluding protected attributes) are treated differently. 
-\eqn{
-\P{Y_f|Z,X} \neq \P{Y_f|Z}
-}
-\end{definition}
-
-Avoiding treating people differently purely on the basis of attributes such as ethnicity and gender and avoiding large differences in important outcomes such as education and income between such groupings both seem like desirable goals.
-Unfortunately, in general, they conflict with one another, see theorem \ref{thm:disparate_conflict}.
-Any variable we might wish to avoid discriminating on will be correlated to some other measurable covariate $Z$, making it impossible to avoid both disparate treatment and disparate impact.
-\vspace{.2cm}
-\begin{theorem} 
-\label{thm:disparate_conflict}
-Disparate impact and disparate treatment conflict.
-Given covariates $Z$ and protected attributes $X$, a model, $f$, cannot be fair with respect to both disparate impact and disparate treatment unless $\P{Z|X} = \P{Z}$.
-\begin{proof} Assume $f$ yields no different treatment, then $\P{Y_f|Z,X} = \P{Y_f|Z}$.
-If we additionally assume no disparate impact, $\sum_z{\P{Y_f|Z}\P{Z|x_1}} = \sum_z{\P{Y_f|Z}\P{Z|x_2}} \;\; \forall \set{x_1,x_2}\subset X$.
-This holds if and only if $\P{Z|x_1} = \P{Z|x_2} \;\; \forall \set{x_1,x_2} \implies \P{Z|X} = \P{Z}$
-\end{proof} 
-\end{theorem}
-
-Further issues.
-
-Disparate treatment, with respect to an observed set of variables $Z$, can be trivially avoided by excluding protected attributes from the training data.
-
-Problems:
-\begin{itemize}
-\item Disparate impact and disparate treatment conflict
-\item 
-\end{itemize}
-
-%\section{Addressing discrimination}
-
-Omitting the protected attribute from the model can increase disparate impact, even if the protected attribute is negatively correlated with the outcome.
-
-\textit{Disparate treatment} can be trivially avoided by excluding the protected variable from the model.
-However, this is deeply unsatisfying given the presence of proxy variables and can increase bias (figure \ref{fig:recidivism})
-
-Avoiding \textit{disparate impact} may be be expensive in terms of predictive accuracy and/or require \textit{disparate treatment}.
-
-\begin{figure}[H]
-\caption{Proxies}
-\label{fig:recidivism}
-	\centering    
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,
-  thick,main node/.style={observedrect}, lt/.style={latentrect} , hidden/.style={empty},background rectangle/.style={fill=olive!45}]
- %nodes
-\node[lt](1){Actual priors};
-\node[main node, below left=of 1](2){Recorded Priors};
-\node[main node, below right=of 1](3){Re-offend};
-\node[lt, above left=of 2](4){Stop and Search};
-\node[main node, above=of 4](5){Race};
-\node[main node, above=of 1](6){Education};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2) edge (3)
-    (4) edge (2)
-    (5) edge (4) edge (6)
-    (6) edge (1);
-\end{tikzpicture}
-\end{figure}
-
-\begin{itemize}
-\item There is an additional connection between causality and fairness, that arises when you know that there is a bias in the data and wish to correct for it.
-In causality, it may be due to selection bias or a confounding variable.
-In fairness, it may be due to historical disadvantage.
-In both cases we may have theory that tells us the direction or scale of an effect.
-IE, we know that race itself has no inherent effect on outcomes such as criminality and education.
-If you have a data set and a collection of such priors, where the data contradicts these priors due to data bias.
-How do you best make decisions based on that data?
-\end{itemize}
-
-human decision making can evolve.
-
-The vagueness inherent in writing laws in natural language allows for them to be re-interpreted.
-
-There is a lack of diversity in models.
-(although randomness is in fact inherent in many) - could one leverage this?
-- you would need an estimate of the probability the model would have assigned each instance to a particular group.
-
-Only humans can provide the ethics
-
-%\chapter*{Random Questions}
-\begin{itemize}
-\item Why not use ratios instead of differences in defining causal effects (is it anything to do with statement in Dawid? What about sensitivity of uncertainty in estimators.
-\end{itemize}
-
-%\chapter{Causality in Marketing}
-Assessing the impact of marketing spend is becoming increasingly important to industry.
-Billions of dollars \citep{XXX} are spent each year and new marketing channels and opportunities are opening up at an unprecedented rate.
-
-There are three key approaches to assessing the impact of marketing in the literature.
-
-The first is to look for correlations across companies on metrics such as how much they spend on marketing \ref{}, how confident they are about their ability to market effectively \ref{} or how important the marketing team is perceived to be within the organisation with indicators of the health of the company such as sales, revenue or market capitalisation.
-Unfortunately is is very difficult to avoid potential major confounders with this approach.
-Are businesses large and successful because of their capable marketing teams or can they afford to hire capable marketers because they have sufficient profits available to do so.
-
-%\section{Attribution}
-The second is attribution.
-The goal in this case is to attribute each sale to a specific (or set of specific) advertisements that the customer was exposed to.
-This approach is the primary way of measuring the success of digital advertising strategies such as search, display, and online video ads.
-The most frequent model is last-click attribution.
-In which the last ad the customer was presented with before they made their purchase is assumed to have caused that sale.
-A central criticism of attribution is that it is not measuring incremental sales.
-Each sales is assumed to be a consequence of some form of advertising for those customers who saw (however briefly) some form of advertisement.
-Of course in reality many of these customers may have made a purchaser regardless of the advertising material to which they were exposed.
-
-System designed to optimise metrics such as click through or conversion rate can do so either by learning how to change the consumers preferences (by serving them the perfect ad at the ideal time) or by learning who was most likely to buy anyway.
-It seems likely that much of the improve mt in click through rates due to the application of sophisticated machine learning algorithms is being driven by the latter.
-Unfortunately, this does nothing for the bottom line of industry (except for those involved in serving ads).
-
-%\section{Econometric Modelling}
-The final key plank businesses apply to estimate how effective their marketing are econometric or mixed media models. These models
-
-In practise these models are typically time series regression models, fitting some measure of sales against marketing spend (by marketing channel, IE TV, radio, search, digital display etc) and other key variables such as pricing, competitor spend and pricing and indicators of the health of the economy and demand for the product in question.
-
-The models may also include non-linear transformation of the marketing inputs to represent saturation and decay of time of effects.
-They may also consider the interaction between media channels and other business levers such as pricing.
-
-The objective of these models is to obtain results that are comparable across channels such that the optimal mix of media can be selected.
-
-The key
-
-\chapter{Conclusion}
-
-%\section{Open questions}
-\paragraph{Cycles} - a huge issue. Not covered by Pearl,Rubin etc.
-
-Places to look, statistical control theory, etc. any interesting papers along these lines?
-
-In the discrete case,
-Fung and Crawford (1990) have recently proposed a fast algorithm for constructing an
-independence graph from discrete data. We have not tested their procedure as a processor for
-the PC algorithm. (COPIED FROM SPRITES)
-
 \bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{library}% write the directory to the .bib file
+\bibliography{library2}% write the directory to the .bib file
 \end{document}

### 2017-07-13 thesis/thesis_map.tex
starts on chapter 1

@@
 \usepackage[utf8]{inputenc}
+\usepackage{textcomp}
 \usepackage{amsmath}
@@
 
-\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.
-
-\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
-
-\paragraph*{My key research questions are:} 
-\begin{itemize}
-\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
-\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
-\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
-\end{itemize}
-
 \section{Motivation}
@@
 From health care, to entertainment and self driving cars; machine learning algorithms will transform many industries.
-It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.
-
-However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings.
+It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory \citep{anderson2008end} and that we may be only a short step from \quotes{The Singularity}, where artificial intelligence exceeds our own and then grows exponentially.
+
+However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings.
 Getting them to generalise to even slightly different problems or data sets remains very challenging.
@@
 
-The notion of causality has been widely debated in science and philosophy \citep{} but is still widely viewed as poorly defined \citep{}.
+The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined.
 This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
 However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression).
-Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.
+Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense the \emph{implications} of the reported associations.
 
@@
 With a finite data set with enough variables we will be able to find some that are completely unrelated but correlate purely by chance.
-\ref{figure - search for random correlation}.
 The second arises from bias, typically introduced by an un-observed confounding variable.
@@
 \begin{itemize}
-\item widely debated in science and philosophy (REFERENCES)
 \item what is explanation?
@@
 
-\section{Identifying when we have a causal problem}
+\section{What problems require causal inference?}
 
@@
 \item Speech recognition (for systems like Siri or Google)
-\item Machine translation 
 \item Image classification
 \item Forecasting the weather
-\item Playing Go 
 \item Identifying spam emails
@@
 \item Predicting who will re-offend on release from prison 
-\item Predicting which customers will cease to be your customers
+\item Customer churn modelling
 \item Demand prediction for inventory control
-\item Predicting who will click on an ad
-\item Financial trading
-\item Recommending movies
-\item Online search
-\item Self driving cars
-\item Pricing insurance
+\item Playing Go 
 \end{itemize}
 
-The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.
+Asking if these problem require causal inference is somewhat disingenuous as they are not posed in sufficient detail to determine if causality is an important consideration.
+In particular, I failed to specify how any (machine learning) model we might build would be used, and what actions could be taken in response to its predictions.
+This is a trivial but significant observation.
+
+It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model.
 
 Consider speech recognition.
-You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words.
+You say something, which causes sound waves, which are converted converted to a digital signal that Siri maps to words.
 Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution).
@@
 
-In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see.
-We can be fairly confident that the process will not change.
-If we develop a discriminating model that is highly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
+In image classification we do not particularly care about building a strong model for exactly how the thing that was photographed translates to an array of pixels, provided we can be fairly confident that the process will not change.
+If we develop a discriminative model that is highly accurate at classifying cats from dogs, we do not need to understand its internal workings (assuming we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
 
 What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decision will not effect if it actually rains.
+If you are using a short term forecast to decide whether to pack an umbrella  causality can be ignored - your decision will not effect if it actually rains.
 However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
@@
 
-Identifying spam and automated essay marking systems are similar.
-The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label.
+Identifying spam and automated essay marking both involve  processing text to determine an underlying (complex) attribute such as its topic or quality.
+In both cases, there is inherent competition between the algorithm and the people generating the text.
+As a result, decisions made by the algorithm are likely to change the relationship between the features it relies on and the true label.
 Spammers and students will modify their writing in order to optimise their results.
-The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided.
-(What would the nature of the features have to be such that change did not occur?
-- they would have to be causes of the label).
+A standard supervised learning approach can only work if the resulting change in the mapping from features to label is sufficiently gradual.
+There are two key ways of ensuring this.
+The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm.
+The second is to ensure features are causes of the outcome and thus cannot be manipulated independtly of the label.
+
+This example also highlights a connection between causal models and transparancy in machine learning.
+If we are using a non-causal model to make decision effecting people, there will be a trade-off between the performance and transparancy of the model; not because the requirement for transparancy restricts us to simple models but because revealing how the model works allows people to change their behaviour to game it.
 
 What about predicting the risk of death in patients with pneumonia?
-Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective this seems like a straightforward prediction problem.
+Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics.
+If we assume that in hospital treatment is more effective for serious cases, this seems like a straightforward prediction problem.
 It is not.
-Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if we start using the model to decide whom to admit.
-(xxx et al) found exactly this effect.
-Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia.
+Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admition decisons.
+\citet{Caruana2015} found exactly this effect in a real data set.
+The model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia.
 They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
-There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addition and treatment protocols.
-However, using it to decide on what basis to admit people could kill.
-The key is understanding exactly what question you are asking.
-In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.
-
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc).
+The issue is not with model, it performed very well at the task for which it was trained, which is to predict wo would be likely to die under the original admition and treatment protocols.
+However, using it to decide how to \emph{change} these protocols could kill.
+The actual question of interest in this case is what happens to patients with charachteristics $X$ if we assign them treatment according to decision rule $Z$.
+
+Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
 For all these problems the assumptions required to treat them as pure prediction problems are;
@@
 
-Demand prediction seems relatively straightforward.
+Demand prediction seems like a relatively straightforward prediction problem.
 These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
 It seems unlikely that using the model to ensure stock is available will itself change demand.
-However there is a potential data censoring issue.
+However, depending on the way demand is measured, there is a potential data censoring issue.
 If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
@@
 
+Optimising hard to the wrong metric.
+
+Mention transferability viewpoint
+
 I hope these examples gave you a feel for the richness and subtitles of causal inference.
 We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
+
+\section{An overview of this thesis and its contributions}
+
+\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.
+
+\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
+
+\paragraph*{My key research questions are:} 
+\begin{itemize}
+\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
+\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
+\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
+\end{itemize}
 
@@
 
+\chapter{Conclusion}
+
 \bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{library2}% write the directory to the .bib file
+\bibliography{library}% write the directory to the .bib file
 \end{document}

### 2017-07-16 thesis/thesis_map.tex
more work on intro, adds Tiberios suggestions

@@
 For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability.
+\item Does defining causality in term of the truncated product formula ensure that a causal Bayesian network is unique?
+It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition.
 \end{enumerate}
@@
 Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
-For all these problems the assumptions required to treat them as pure prediction problems are;
+For all these problems the assumptions required to treat them as pure prediction problems are
 \begin{enumerate}
@@
 
-Playing Go (and other games) is another case with some subtleties.
+Playing Go (and other games) is a case with some subtleties.
 At every turn, the AI agent has a number of actions available.
@@
 \label{exm:adjusting}
-Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all 4 year olds.
-We have a large cross-sectional data set on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
+Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds.
+We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
 We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
@@
 \centering
-\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2|X_1}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
+\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$.
+Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution.
+Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
 \label{fig:bayesnet2}
@@
 The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true $G$ can have extra edges).
+There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true: $G$ can have extra edges).
 If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
@@
 
+\vspace{0.5cm}
 \begin{definition}{The do-notation}
@@
 This notation is somewhat overloaded.
-It may be used represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
+It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
 For example it could represent
@@
 
+\vspace{0.5cm}
 \begin{theorem}[Truncated product formula \citep{Pearl2000}]
@@
 If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.
+Let $\parents{X_i}$ denote the parents of the variable $X_i$ in $G$.
 
@@
 \label{eq:truncatedproduct}
-\P{X_1...X_N|do(Z=z)} =
-  \begin{cases}
-  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistent with $Z=z$}  \\
-   0       & \text{otherwise } 
-  \end{cases}
+\P{X_1...X_N|do(Z=z)} = \ind{Z = z}
+  \prod_{X_i \notin Z}\P{X_{i}|\parents{X_i}}
 \end{equation}
@@
 The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
-In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.
+In fact, from the interventionist viewpoint of causality, the truncated product formula defines what it means for a link to be causal.
 
@@
 
-Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across; the local Markov property states that variables are independent of their non-effects given their direct causes.
-Similarly the global Markov property and d-separation also hold in causal networks.
+Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes.
+The global Markov property and d-separation also hold in causal networks.
+D-speration, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
+We include a brief introduction in section \ref{sec:dsep}.
 
@@
 For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
-But what if some people failed to take the prescribed treatment.
+But, what if some people failed to take the prescribed treatment?
 We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
@@
 Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
-Such problems have mostly been studied within the dynamic systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
+Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
 \citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
@@
 \begin{itemize}
-\item $\P{Y|X=1}$, the probability that those who took the treatment will recover
-\item $\P{Y|X=0}$, the probability that those who were not treated will recover
+\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
+\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
 \end{itemize}
@@
 Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
-For those who did not take the treatment, the observed outcome is the sames as the outcome \emph{had} they not taken the treatment.
+For those who did not take the treatment, the observed outcome is the same as the outcome \emph{had} they not taken the treatment.
 Equivalently stated:
@@
 In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, $X \nci \cf{Y}{0}$.
-Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, $X \nci \cf{Y}{1}$.
+In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e.
+$X \nci \cf{Y}{0}$.
+Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e.
+$X \nci \cf{Y}{1}$.
 A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
@@
 \vspace*{.3cm}
-\begin{theorem}[Ignorability] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
+\begin{theorem}[Ignorability \citep{Rosenbaum1983, Pearl2000}] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
 
@@
 The function $f_y$ and distribution over $\epsilon_y$ does not change.
-This results in the interventional distribution \footnote{We have assumed the variables are discrete only for notational convenience},
+This results in the interventional distribution, \footnote{We have assumed the variables are discrete only for notational convenience}
 
@@
 \begin{figure}[h]
-\caption{An example of Simpson's Paradox}
+\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
 \label{fig:simpsons_paradox}
@@
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Gender};
-\node[node, below left=of 1](2){Treatment};
-\node[node, below right=of 1](3){Outcome};
+\node[node](1){Z = Gender};
+\node[node, below left=of 1](2){T = Treatment};
+\node[node, below right=of 1](3){Y = Outcome};
 \path[]
@@
 \begin{figure}[h]
-\caption{An example of Simpson's Paradox}
+\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
 \label{fig:simpsons_paradox2}
@@
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Blood Pressure};
-\node[node, below left=of 1](2){Treatment};
-\node[node, below right=of 1](3){Outcome};
+\node[node](1){Z = Blood Pressure};
+\node[node, below left=of 1](2){T = Treatment};
+\node[node, below right=of 1](3){Y = Outcome};
 \path[]

### 2017-07-20 thesis/offcuts.tex
starts on chapter 1
working on intro

@@
+\begin{itemize}
+\item Speech recognition (for systems like Siri or Google)
+\item Machine translation 
+\item Image classification
+\item Forecasting the weather
+\item Playing Go 
+\item Identifying spam emails
+\item Automated essay marking
+\item Predicting the risk of death in patients with pneumonia.
+\item Predicting who will re-offend on release from prison 
+\item Predicting which customers will cease to be your customers
+\item Demand prediction for inventory control
+\item Predicting who will click on an ad
+\item Financial trading
+\item Recommending movies
+\item Online search
+\item Self driving cars
+\item Pricing insurance
+\end{itemize}
+
 However we can also pose causal queries that are not interventional and cannot be phrased in terms of the do-notation.
@@
 \end{enumerate}
+
+connection to bias variance trade-off. Increasing the assumptions we impose via a model introduces bias but decreases variance in the result.
+
+For example, in Australia we might detect that driving on unpaved roads is associated with poorer health outcomes.
+It is plausible that there are some negative health impacts of bumpy roads, but more likely the correlation occurs because unpaved roads and poor access to health services are both a \emph{consequence} of living in regional areas.
+
+Problems highlighted as intractable in the reverse causal sense are also intractable in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning.
+For example, the war question posed in Gelman.
+
+Observing a system, without the ability to intervene or experiment on it. Fundamental challenge because larger samples are not enough.
+
+In this case, the observed correlation is not an artifact of random noise in a sample of data. We would expect the association to continue to hold if sampled new data.

### 2017-07-20 thesis/thesis_map.tex
working on intro

@@
 
-\section{What is causality and why do we care? }
+\section{What is causality? }
 
@@
 However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression).
-Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense the \emph{implications} of the reported associations.
-
-We discuss more detailed definitions of causality in section c
-
-The what-if type questions from the why.
-\citep{Gelman2010} \citep{}.
-Why do whites to better than blacks in school (in America).
-Suggests that reverse causal inference questions are more interesting and motivate most of social science.
-
-I do not find this distinction useful. 
-We can only change the future - history is useful only as far as it tells us something about the future.
-Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one).
-Problems highlighted as intractable in the reverse causal sense are also intractable in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning.
-For example, the war question posed in Gelman.
-
-A distinction between forward causal inference, what happens if we do X and reverse casual inference
-
-To explain or to predict \citep{Shmueli2010}
-The two cultures \citep{Breiman2001}
-
-\paragraph*{There are two reasons why correlation is not causation}.
-The first is related to variance and over-fitting.
-Observations are noisy.
-With a finite data set with enough variables we will be able to find some that are completely unrelated but correlate purely by chance.
-The second arises from bias, typically introduced by an un-observed confounding variable.
-In this case, variables are correlated not by chance.
-We would expect the relationship to hold if we sampled more data.
-However, they are not causally related in that intervening to set one would not likely effect the other.
-EXAMPLE WITH FIGURE.
-
-\subsection{Defining causality}
-\begin{itemize}
-\item what is explanation?
-\item any model that aims to predict the outcome of an action or intervention in a system
-\item I do not see the distinction between explanation and (causal) prediction.
-Explanation is all about the ability to compress and to generalise.
-The more a model can do this, the more we view as providing an understanding of the why.
-\item mediation?
-\end{itemize}
-
-\section{What problems require causal inference?}
-
-\subsubsection{Examples of typical machine learning problems. Are they causal?}
-
-Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to.
-Which of them require casual inference?
-How can we identify characteristics of a problem that make causal modelling important?
+Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via \quotes{common sense} the implications of the reported associations.
+
+There are two ways in which an association detected in a dataset may be non-causal.
+The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample.
+Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated.
+For example, based on data from the years 1999 to 2009, the age of miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}.
+We would not expect this relationship to hold if we were to look at a new sample of data.
+This form of spurious correlation also has serious repercussions.
+It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published.
+As a consequnce, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
+This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}.
+This issue can be ameliorated by getting more data and by seperating learning models from evaluating their performance, for example by evaluating models on a stict hold-out set or on the extent to which their results can be replicated.
+
+However, a strong association, observed in multiple indpendent studies may still not be causal.
+The correlation can arise because both variables are consequences of some other, unmeasured factor.
+For example, the reading ability of children under twelve is strongy correlated with their height, because older children are taller and can read better.
+However height is not a cause of reading ability because if we were to intervene to increase a child's height, for example by giving them growth hormones, we would not expect dramatic improvements in their reading.
+Similarly, extra tutoring in reading will not make a child grow taller.
+This problem is fundamentally different to the issue of spurious correlations arising by chance in finite datasets.
+Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to seperate causation from correlation.
+
+The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables.
+In this thesis, I take an interventionalist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal.
+This viewpoint captures the types of questions that motivate this thesis.
+How can we change the way we do things to obtain better outcomes.
+
+Causality is often liked to explaination; understanding how and why things happen.
+I view explaination in terms of compression and generalisabilty: how much information about the world a model can capture.
+This creates a hierachy in the degree to which models are explainitory, rather than a simple binary distinction.
+A standard predictive model encodes all the information we need to predict some output given inputs provided the system generating the data does not change.
+A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed.
+More detailed causal models could predict the outcome for a wide range of combiations of interventions conditional on a range of contexts.
+By considering conditional interventions within our definition of causal questions we also capture mediation: the study of through which pathways one variable causes another \cite{vanderweele2015explanation}.
+Finally, a model that can distil how elements interact to mathematical equatons like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
+
+\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs.
+The former aims to identify the effect of a known cause.
+The latter can be viewed as identiying causes of an effect.
+They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}.
+However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined.
+The goal of identifying the causes of an effect can be formalised within the graphical causal model framework.
+Solving this problem is certainly much more challanging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables.
+These practical difficulties may well be overwealming, particularly in fields such as social science and economics where datasets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively.
+However, this does not mean that the problem of identifying causes of effects is ill-posed in principle.
+It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.
+
+There has traditionally been a large gap between researches in machine learning who focus on prediction, use largly non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models.
+However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of for the training and evaluation models.
+In the next section we emphasise the subtley that can exist in determining if a problem is causal by examiming some typical examples.
+
+\section{What makes a causal inference problem?}
+
+Machine learning is in the midst of a boom, driven by the availability of large datasets and the computation resources to process them.
+Machine learning techniques are being applied a huge range of problems, in both industry and achademia.
+The following examles are inteded to capture the bredth of problems that machine learning algorithms are actively beeing applied to.
+Which, if any, of these problems require causal inference?
 
@@
 
-Asking if these problem require causal inference is somewhat disingenuous as they are not posed in sufficient detail to determine if causality is an important consideration.
-In particular, I failed to specify how any (machine learning) model we might build would be used, and what actions could be taken in response to its predictions.
-This is a trivial but significant observation.
-
-It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model.
+The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important considertion.
+In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.
 
@@
 \citet{Caruana2015} found exactly this effect in a real data set.
-The model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia.
+The model learnt that people suffering asthma were \emph{less} likely to die from pneumonia.
 They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
@@
 Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
-For all these problems the assumptions required to treat them as pure prediction problems are
-\begin{enumerate}
-\item The treatment is assumed to be effective (at least better than nothing)
-\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
-\end{enumerate}
+Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.
 
@@
 The board state following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win.
+The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem.
+There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}.
+Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$.
+This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players.
+Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a compbination of these approaches \citep{silver2016mastering}.
+The supervised learning was enhanced by having the agent play (varients) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
+
+The problem of playing go is clearly causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However,
+
 ...
@@
 One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken).
+
 When can we learn from the actions others have taken?
@@
 
+This is a trivial but significant observation. It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model. 
 \begin{itemize}
@@
 
+How can we characterise which problems are predictive and which are causal?
+
 Mention transferability viewpoint
@@
 We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
+
+Always, unless you can demonstrate otherwise.
 
@@
 \section{Approaches to causality}
+This stuff gets built into explaining how the thesis fits together.
 
@@
 
+\section{Granger causality}
+
+Mentioned in the category of causal frameworks,
+
+Granger causality is less of a framework for defining causality and on which a wide range of algorithms and methods can be constructed, but a method for infering causal structure from data.
+As such is discussion really belongs in section \ref{sec:causal-discovery}.
+
+TODO move this to a paragraph at the end of the comparing and unifying section. \todo{}
+
 \section{Comparing and unifying the models}
@@
 \section{Causal Discovery}
+\label{sec:causal-discovery}
 
@@
 \end{document}
+
+\chapter{Causation vs Prediction}
+
+The two cultures \citep{Breiman2001}
+ - wrong because suggets we can just use the machine learning approach for everything.
+Does not address the problems that a) structural assumptions required for causal inference.
+b) we cannot use cross-vaidation to assess the validity of a causal model (really b is the whole point).
+
+To explain or to predict \citep{Shmueli2010}
+- wrong because they assume all the parts of explanatory modelling are required - so black box models cannot be used.
+
+Predict the relationship between X and Y as a way to stratify to reduce the variance in an experiment.
+
+They consider a different definition of explanatory models,
+
+Definition as backwards.
+Explanatory models focus on theory because some theory (or equivelently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
+
+Causal modelling frameworks allow us to seperate assumptions about the structure of the relationships between variables from assumptions about the functional form.
+
+What can be data driven and what cannot? (for a given set of assumptions).
+
+Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the languate of EoP but are built on a hybrid of both approaches.

### 2017-07-21 thesis/offcuts.tex
working in intro

@@
 In this case, the observed correlation is not an artifact of random noise in a sample of data. We would expect the association to continue to hold if sampled new data.
+
+ Covariate shift clearly comes in here.
+Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occurring and that results will be transferable.
+The mechanism is known but the function may be complex.
+Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models?
+Yes, if the way in which features get their values changes, then causal models will be invariant to that in a way that non-causal ones are not.

### 2017-07-21 thesis/thesis_map.tex
working in intro

@@
 
-The problem of playing go is clearly causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However,
-
-...
-this is interesting maybe come back to it ...
-an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ...
-One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken).
-
-When can we learn from the actions others have taken?
-When there is no confounding.
-And does this hold with Go?
-Probably because the board state encapsulates everything that should determine what move is played.
-Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)
-
-Having considered these examples we can now identify some general aspects of the problem that determine whether or not we require a causal model.
-
-This is a trivial but significant observation. It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model. 
+The problem of playing go is causal from the interventionalist perspective.
+The agent wishes to learn the probability of a win given an action they take.
+However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach.
+The actions the agent has to explore are the sames ones as human players explored to generate the trainging data, and both have the same objective - to win the game.
+In addition, the state of the board encapsulates all the information relevent to selecting a move.
+These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.
+
+Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference. A predictive model may be sufficient if;
+
 \begin{itemize}
@@
 \item 
-
- Covariate shift clearly comes in here.
-Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occurring and that results will be transferable.
-The mechanism is known but the function may be complex.
-Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models?
-Yes, if the way in which features get their values changes, then causal models will be invariant to that in a way that non-causal ones are not.
+This is more likely be be the case if the actions are recorded in the training data (unlike in the pneumonia example), and the algorithm has the same objective and information on which to base its decisions as generated the training data.
 
@@
 
+We can also view causal inference problems as a particular form of the transfer learning problem. We have training data obtained from one system, but we wish te estimate
+
+It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model.
+
 Optimising hard to the wrong metric.
 
-How can we characterise which problems are predictive and which are causal?
-
-Mention transferability viewpoint
-
+In conclusion, it is rare that anyone would build a model without indending that some kind of decion be made on the basis of its results.
+For this reason we assume that a causal model is required unless we can justify otherwise.
 I hope these examples gave you a feel for the richness and subtitles of causal inference.
 We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
-
-Always, unless you can demonstrate otherwise.
 

### 2017-07-28 thesis/offcuts_chapter6.tex
adds draft introduction, moves captions to below figures

@@
 the PC algorithm. (COPIED FROM SPRITES)
+
+We can also view causal inference problems as a particular form of transfer learning.
+We have training data obtained from one system, and wish to use it to estimate characteristics or make predictions about another, in this case the original system after some intervention.
+To do this we must make some assumptions about the relationship between the two systems.
+If they are arbitrarily different then information about the first one is not transferable to the second.
+
+I hope these examples give a feel for the richness and subtitles of causal problems.
+causal inference.
+We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
+
+We will call data sets where we do not have control over the decision making process that generated the data observational.
+Versus experimental data sets, where we do have control (experimental data sets do not always have to be randomised - although that is a powerful approach to ensure we have control.
+There is a space in the middle where we have partially controlled the process by which agent select actions.
+Randomised data with imperfect compliance would be an example.

### 2017-08-03 thesis/thesis_map.tex
updated prior

@@
 
+\section*{Questions}
+\begin{itemize}
+\item Are there any problematic counterfactuals being used in defining bandit problems
+\item in the Schölkopf causal fairness paper - is there an issue with the use of twin networks. Could it be fixed with SWIGs?
+\item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
+\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomized experimental design?
+\item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
+\item What does conditioning mean in the Bayesian sense?
+- can imposing a prior which is true with respect to that variable, increase bias due to confounding?
+(In the same way that in the pnemonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.
+
+\item Does the way we define supervised learning and loss fuctions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?
+
+\end{itemize}
+
 \chapter{Introduction}
@@
 
-However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings.
+However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within narrow problem settings.
 Getting them to generalise to even slightly different problems or data sets remains very challenging.
@@
 
-\section{What makes a causal inference problem?}
+\section{What makes a problem causal?}
 
@@
 The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm.
-The second is to ensure features are causes of the outcome and thus cannot be manipulated independtly of the label.
+The second is to use a model in which the features are related to the outcome in such a way that they cannot be manipulated independently.
 
@@
 
-Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference. A predictive model may be sufficient if;
-
-\begin{itemize}
-\item Does acting on the predictions of the model change the mapping from features to target?
-(at least if the decision process is open to scrutiny).
-In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
-\item 
-This is more likely be be the case if the actions are recorded in the training data (unlike in the pneumonia example), and the algorithm has the same objective and information on which to base its decisions as generated the training data.
-
-\item To decide between actions we only need to rank them (not estimate their actual effect). 
-\item The predicted outcome in the absence of an intervention provides a single point.
-We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes.
-\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
-\item Does acting on the result of the prediction change the predictive distribution p(y|x)? IE change people's behaviour.
-\item Ethics - ... People's viewpoint on if its OK...
-\end{itemize}
-
-We can also view causal inference problems as a particular form of the transfer learning problem. We have training data obtained from one system, but we wish te estimate
-
-It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model.
-
-Optimising hard to the wrong metric.
-
-In conclusion, it is rare that anyone would build a model without indending that some kind of decion be made on the basis of its results.
-For this reason we assume that a causal model is required unless we can justify otherwise.
-I hope these examples gave you a feel for the richness and subtitles of causal inference.
-We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.
+Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference.
+A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes.
+The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people.
+Think about problems like credit scoring and parole decisons.
+There are strong ethical grounds for demanding transparancy, but if the goals of society and the individuals are not perfectly aligned and there is any posibility that people can manipulate features independtly of the outcome, there will be a conflict between model accuracy and transparancy.
+It is rare to build a model without any intent to make some kind of decion on the basis of its results.
+Thus, I argue we should assume a causal model is required until we can justify otherwise.
 
@@
 
-\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.
-
-\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
-
-\paragraph*{My key research questions are:} 
-\begin{itemize}
-\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
-\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
-\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
-\end{itemize}
-
-\section{Approaches to causality}
-This stuff gets built into explaining how the thesis fits together.
-
-There are two broad approaches to deciding how to act.
-Reinforcement learning and causal inference.
-In reinforcement learning we estimate the effect of actions by taking them.
-We assume there is an agent capable of intervening in the system and try to find policies for the agent to follow in selecting actions so as to maximise some kind of reward.
-This is a very powerful and general framework (REFERENCES TO GENERAL AI).
-However, POINT OUT SOME OF THE DIFFICULTIES WITH REINFORCEMENT LEARNING.
-We are frequently presented with large bodies of data that have been collected from a system in which we did not have any control over what actions were taken.
-
-We will call data sets where we do not have control over the decision making process that generated the data observational.
-Versus experimental data sets, where we do have control (experimental data sets do not always have to be randomised -although that is a powerful approach to ensure we have control.
-There is a space in the middle where we have partially controlled the process by which agent select actions.
-Randomised data with imperfect compliance would be an example.
-
-An agent (capable of intervening in the system) chooses an action from those available The agent making the decision included in the model. This is
-
-Reinforcement learning addresses the problem of learning from explicitly taking actions. There
-is typically some state or environment. An agent chooses an action from those available in the
-current state. The state then evolves stochastically as a function of the selected action and
-the agent receives some feedback or reward that is a function of the new state. This setting
-differs from the standard classification problem in that the agent must learn from feedback on
-the selected action, rather than being presented with the correct action for a given state. A
-common modelling assumption is that the state evolves only as a function of the previous state
-and the action chosen, and given these, is independent of the previous history of states and
-actions. This is known as a Markov decision process or MDP. A particularly well studied and
-understood model is the singe state MDP. In this case, there are a set of actions, each associated
-with a fixed but unknown reward distribution and at each time step our agent selects an action
-and receives corresponding feedback. This is known as the multi-armed bandit problem.
-
-Causal inference makes use of assumptions to allow the outcome of actions to be predicted from
-observational data. The key to causal inference is a framework that can model how actions
-change the state of the world. This framework then allows us to map information collected in
-one setting to another.
-
-Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit
-algorithms deal with the sequential nature of the decision making process, causal inference with
-the problem that full randomisation is not always feasible, affordable or ethical. The similarities
-between the problems that these techniques have been developed to address raises the question
-of if there are problems best addressed by a combination of these approaches and how they
-can be combined. The goal of my thesis is to explore these questions. In the next sections I
-review the key literature in causal inference and bandits. I then present a general approach to
-how causal models might be incorporated into bandit settings and conclude by demonstrating
-an algorithm that leverages causal assumptions to improve performance in a specific bandit
-setting.
-
-There are two key approaches to causal problems.
-The first is to learn the outcome of actions by directly intervening in the system and seeing what happens.
-We then get feedback on how good those actions were.
-This is the approach taken in reinforcement learning.
-ADVANTAGES AND DISADVANTAGES OF THIS APPROACH.
-The second broad approach is causal inference.
-Here
-
-\todo{insert figure showing data generating process and observed data defining ML}\ref{all of statistics}
-
-\paragraph*{Two broad approaches} 
-\begin{itemize}
-\item Build a model to map the natural behaviour of the system to what will happen for some action
-\item Take the action and see what happens
-\end{itemize}
-
-\paragraph*{The first is causal inference}
-
-\paragraph*{The second is reinforcement learning}
-
-\paragraph*{Both generalise from randomised experiment} Reinforcement learning to sequential decisions, causal inference to non-experimental conditions
-
-Both these fields relate to the problem of making optimal decisions and both can be seen as generalising randomised controlled experiments.
-Causal inference  is the study of how to estimate the effect of an action in the absence of randomisation.
-Reinforcement learning studies how we can do better if the decisions are to be made sequentially.
-
-\paragraph*{Both approaches involve assumptions} the latter that we can group context and actions.
-
-\paragraph*{Limitations of causal inference}
-
-\paragraph*{Limitations of experiments} What are the issues with standard randomised experiments?
+Causal questions are everywhere.
+Statistics, economics, social science, artificial intelligence and machine learning.
+Techniques and for addressing such problems have developed in parralel within many diciplines.
+These techniques can usefully be categorised into two broad approaches, reinforcement learning and observational causal inference.
+
+In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them.
+We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal.
+A particular strand of research within reinforcement learning are multi-armed bandit problems.
+They describe settings in which there are a set of available actions, the agent repeately decides which to select and then observes the outcome of the chosen action.
+They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.
+
+The approach of learning the outcome of an action by taking it has plays a key role in advancing our knoweledge of the world.
+However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen.
+Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.
+
+Observational causal inference can be viewed as a form of transfer learning.
+The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the orignial system.
+This is impossible without some assumptions about how the the orignal system and the system after intervention are related to one-another.
+The key to observational inference is a framework that models how actions change the state of the world and thus allows us to map information collected in one setting to another.
+
+Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials.
+Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem randomisation is not always feasible, affordable or ethical.
+The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
+
+The goals of this thesis are: to clarify the distinction between prediction and causal inference and highlight when and how we can leverage advances in prediction to improve causal inference and to draw together the key approaches to solving causal problems, from both the interventionalist and observational viewpoints.
+In chapter two, I describe the existing frameworks for formalising causality developed withing economics, statistics and machine learning and discuss how they relate to one another.
+Chapter three reviews the key approaches to causal inference from observational data.
+Chapter four highlights the interventionalist viewpoint, covering traditional randomised experiments and the closely related multi-armed bandit problems and highlights connections between the algorithms applied to bandit problems and those for estimating causal effects from observational data.
+
+In chapter five, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
+The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
+I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
+
+Chapters two to four predominantly review the existing liturature with the goal of making connections between the different viewpoints and approaches.
+They do not contain novel technical results but present the material in a unified way.
+Chapter five is a novel contribution.
+It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit liturature and describes an algorithm that can leverage this structure to make better decisons.
 
@@
 
-Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
+Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
 To do this we need terminology to describe actions and how we anticipate the system will respond to them.
@@
 \centering
-\caption{A general Bayesian network for the joint distribution over three variables.
-This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
-\label{fig:bayesnet}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \end{tikzpicture}
+\caption{A general Bayesian network for the joint distribution over three variables.
+This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
+\label{fig:bayesnet}
 \end{figure}
@@
 \centering
-\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$.
-Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution.
-Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
-\label{fig:bayesnet2}
 \begin{subfigure}[t]{0.15\textwidth}
@@
 \end{subfigure}
+\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$.
+Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution.
+Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
+\label{fig:bayesnet2}
 \end{figure}
@@
 \begin{figure}[h]
-\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
-\label{fig:causal_network_example}
 \centering
@@
 \end{tikzpicture}
+\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
+\label{fig:causal_network_example}
 \end{figure}
@@
 \center
-\caption{Causal Bayesian network for example \ref{exm:adjusting}}
-\label{fig:causal_adjust}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]
@@
 \end{tikzpicture}
+\caption{Causal Bayesian network for example \ref{exm:adjusting}}
+\label{fig:causal_adjust} 
 \end{figure}
@@
 \center
-\caption{Causal inference with ordinary Bayesian networks.
-The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$.
-The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
-As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
-However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
-\label{fig:causal_ordinary_bayes}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
@@
 \end{tikzpicture} 
+\caption{Causal inference with ordinary Bayesian networks.
+The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$.
+The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
+As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
+However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
+\label{fig:causal_ordinary_bayes}
 \end{figure}
@@
 \begin{figure}[h]
-\caption{Randomised treatment with imperfect compliance}
-\label{fig:randomized_imperfect_compliance}
 \centering
@@
 \end{tikzpicture}
+\caption{Randomised treatment with imperfect compliance}
+\label{fig:randomized_imperfect_compliance}
 \end{figure}
@@
 \centering
-\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
-\label{fig:metaphysical_distribution_difference}
 \subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$.
@@
 {\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
+\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
+\label{fig:metaphysical_distribution_difference}
 \end{figure}
@@
 \begin{figure}[h]
-\caption{SEM for example \ref{exm:ranomized_experiment}}
-\label{fig:sem_randomized_treatment}
 \centering
@@
 \end{tikzpicture}
+\caption{SEM for example \ref{exm:ranomized_experiment}}
+\label{fig:sem_randomized_treatment}
 \end{figure}
@@
 \begin{figure}[h]
-\caption{SEM for example \ref{exm:adjusting}}
-\label{fig:sem:preschool}
 \centering
@@
 \end{tikzpicture}
+\caption{SEM for example \ref{exm:adjusting}}
+\label{fig:sem:preschool}
 \end{figure}
@@
 \begin{figure}[h]
-\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
-\label{fig:simpsons_paradox}
 \centering
@@
 \end{tikzpicture}
+\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
+\label{fig:simpsons_paradox}
 \end{figure}
@@
 \begin{figure}[h]
-\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
-\label{fig:simpsons_paradox2}
 \centering
@@
 \end{tikzpicture}
+\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
+\label{fig:simpsons_paradox2}
 \end{figure}
@@
 \centering
+\includegraphics[scale=.4]{figures/simpson_vector}
 \caption{Simpson's reversal visualised.
@@
 \label{fig:simpsons_vector}
-\includegraphics[scale=.4]{figures/simpson_vector}
 \end{figure}
@@
 \centering
-\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
-\label{fig:dseptriples}
 \begin{subfigure}[t]{0.2\textwidth}
@@
 \end{subfigure}
+\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
+\label{fig:dseptriples}
 \end{figure}
@@
 \centering
-\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
-\label{fig:mutilatednet}
 \begin{subfigure}[t]{0.3\textwidth}
@@
 \end{subfigure}
+\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
+\label{fig:mutilatednet}
 \end{figure}
@@
 \centering
-\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
-In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
-If we condition on $W$, we block this indirect path.}
-\label{fig:rule2}
 \begin{subfigure}[t]{0.29\textwidth}
@@
 \end{subfigure}
+\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
+In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
+If we condition on $W$, we block this indirect path.}
+\label{fig:rule2}
 \end{figure}
@@
 \centering
-\caption{An example of applying equation \ref{eq:Do2}.
-In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
-Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
-Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
-}
-\label{fig:rule2_2}
 \begin{subfigure}[t]{0.29\textwidth}
@@
 \end{subfigure}
+\caption{An example of applying equation \ref{eq:Do2}.
+In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
+Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
+Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
+}
+\label{fig:rule2_2}
 \end{figure}
@@
 \centering
-\caption{Example application of equation \ref{eq:Do3}.
-$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
-We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
-\label{fig:rule3}
 \begin{subfigure}[t]{0.29\textwidth}
@@
 \end{subfigure}
+\caption{Example application of equation \ref{eq:Do3}.
+$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
+We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
+\label{fig:rule3}
 \end{figure}
@@
 \centering
-\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
-\label{fig:identify}
 \begin{subfigure}[t]{0.29\textwidth}
@@
 \end{subfigure}
+\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
+\label{fig:identify}
 \end{figure}
@@
 \centering
-\caption{}
-\label{fig:adjustment}
 \begin{subfigure}[t]{0.48\textwidth}
@@
 \end{subfigure}
+\caption{}
+\label{fig:adjustment}
 \end{figure}
+
+\todo{fix this caption}
 
@@
 \centering
+\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
 \caption{
@@
 \label{fig:covariate_shift_adjust}
-\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
 \end{figure}
@@
 \centering
-\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalising over $U$. }
-\label{fig:DAGSnotclosed}
  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={observed},lt/.style={latent}]
@@
 \end{tikzpicture}
+\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalising over $U$. }
+\label{fig:DAGSnotclosed}
 \end{figure}
@@
 \centering
-\caption{FCI examples: true graph and FCI output}
-\label{fig:FCIExamples}
  \begin{subfigure}[t]{0.4\textwidth}
@@
 \end{subfigure}
+\caption{FCI examples: true graph and FCI output}
+\label{fig:FCIExamples}
 \end{figure}
@@
 \begin{figure}
-\caption{causal network for a randomised experiment}
-\label{fig:random_experiment_network}
 \centering
@@
 \end{tikzpicture}
+\caption{causal network for a randomised experiment}
+\label{fig:random_experiment_network}
 \end{figure}
@@
 \begin{figure}
-\caption{causal network for a stratified randomised experiment if the probability an individual is assigned a given treatment depends on some covariates.}
-\label{fig:random_experiment_network_stratified}
 \centering
@@
 \end{tikzpicture}
+\caption{causal network for a stratified randomised experiment if the probability an individual is assigned a given treatment depends on some covariates.}
+\label{fig:random_experiment_network_stratified}
 \end{figure}
@@
 \begin{figure}
-\caption{causal network for a randomised experiment with imperfect compliance}
-\label{fig:random_experiment_network_imperfect_compliance}
 \centering
@@
 \end{tikzpicture}
+\caption{causal network for a randomised experiment with imperfect compliance}
+\label{fig:random_experiment_network_imperfect_compliance}
 \end{figure}
@@
 \begin{center}
+\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
+\end{center}
 \caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parachutes for the prevention of morbidity and mortality associated with falls from large heights.}
 \label{fig:parashute}
-\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
-\end{center}
 \end{figure}
@@
 \centering
+\includegraphics[scale=1]{extreme_early_stopping}
 \caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples.
@@
 \label{fig:early_stopping}
-\includegraphics[scale=1]{extreme_early_stopping}
 \end{figure}
@@
 \begin{figure}
+\includegraphics[scale=1]{ucb_regret_vs_delta}
 \caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
@@
 \label{fig:ucb_regret_vs_delta}
-\includegraphics[scale=1]{ucb_regret_vs_delta}
 \end{figure}
@@
 \begin{figure}
+\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
 \caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$.
@@
 \label{fig:ucb_upperbound_vs_actual_performance}
-\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
 \end{figure}
@@
 \begin{figure}[h]
+    \begin{subfigure}[t]{0.49\textwidth}
+	\centering   
+	 \caption{Behaviour over time}\label{fig:non-stationary-bandit-time} 
+         \includegraphics[scale=.5]{non-stationary-bandit-time}
+\end{subfigure}
+    \begin{subfigure}[t]{0.49\textwidth}
+    \centering
+     \caption{Cumulative regret distribution}
+        \label{fig:non-stationary-bandit-regret}
+        \includegraphics[scale=.5]{non-stationary-bandit-regret}
+    \end{subfigure}
       \caption{The pseudo-regret can be negative if rewards are non-stationary.
@@
 The distribution of regret over the 1000 simulations is shown in figure (b).}\label{fig:non-stationary-bandit}
-    \begin{subfigure}[t]{0.49\textwidth}
-	\centering   
-	 \caption{Behaviour over time}\label{fig:non-stationary-bandit-time} 
-         \includegraphics[scale=.5]{non-stationary-bandit-time}
-\end{subfigure}
-    \begin{subfigure}[t]{0.49\textwidth}
-    \centering
-     \caption{Cumulative regret distribution}
-        \label{fig:non-stationary-bandit-regret}
-        \includegraphics[scale=.5]{non-stationary-bandit-regret}
-    \end{subfigure}
 \end{figure}
@@
 \centering
-\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
-$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
-Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
-To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
-}
-\label{fig:contextual-bandit-causal-graph}
 \begin{subfigure}[t]{0.29\textwidth}
@@
 \end{subfigure}
+\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
+$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
+Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
+To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
+}
+\label{fig:contextual-bandit-causal-graph}
 \end{figure}
@@
 \centering
-\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$}
-\label{fig:off-policy-evaluation}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
@@
 \end{tikzpicture} 
+\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$}
+\label{fig:off-policy-evaluation}
 \end{figure}
@@
 \begin{figure}
-\caption{A simple causal graphical model and corresponding complete action space. $W$ and $Z$ represent binary variables that can be intervened on and $Y$ represents the reward.}
-\label{fig:unify_frameworks}
 \centering
@@
 \end{subfigure}
+\caption{A simple causal graphical model and corresponding complete action space. $W$ and $Z$ represent binary variables that can be intervened on and $Y$ represents the reward.}
+\label{fig:unify_frameworks}
 \end{figure}
@@
 \begin{figure}
-\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level .
-The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
-For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
-\label{fig:cholesterol_graph}
 \centering
@@
 \end{tikzpicture}
+\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level .
+The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
+For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
+\label{fig:cholesterol_graph}
 \end{figure}
@@
 \begin{figure}
-  \caption{Causal Models}\label{fig:causalmodels}
     \begin{subfigure}[t]{0.34\textwidth}
@@
     \end{subfigure}
+      \caption{Causal Models}\label{fig:causalmodels}\label{fig:causal_models_causal_bandit}
 \end{figure}
@@
 \end{proof}
-\
-%in the parallel bandit setting the $m(\eta)$ given in this section approximately coincides with the $m(\vec{q})$ in Eq.\ \ref{eq:m-simple}.
-%Recall in that setting that 
-%\eq{
-%\actions = \set{do()} \cup \set{do(X_i = j) : 1 \leq i \leq N \text{ and } j \in \set{0,1}}\,.
-%}
-%For $a = do()$ choose $\eta_a = 1/2$. 
-%Let $m = m(\vec{q})$ and note that, by definition, there are at most $m$ pairs $(i,j)$ such that $\P{X_i = j} \leq 1/m$.
-%Thus, for $a = do(X_i = j)$ letting $\eta_a \propto \ind{\P{X_i = j} \leq 1/m} / (2m)$ guarantees $\eta_a \ge \frac{1}{2m}$ when $\eta_a \ne 0$.
-%It is then easy to check that $m(\eta) \leq 2m$ using an argument like that for Proposition~\ref{pro:m-bound}.
 
 \chapter{Conclusion}
+
+\chapter*{Careful with that prior}
+
+Suppose we have a data generated by the linear guassian Bayesian network below.
+The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$.
+The variable $U$ is latent.
+
+\eqn{
+\P{U} &= N(0 , v_u) \\
+\label{eqn:careful_with_priors}
+\P{Z|U} &= N(w_{zu}U , v_z) \\
+\P{X|Z} &= N(w_{xz}Z , v_x) \\
+\P{Y|U,Z,X} & = N(w_{yu}U +w_{yz}Z + w_{yx}X , v_y)
+}
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[lt, right=of 1](4){$U$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2)
+    	edge (3)
+    (2) edge (3)
+    (4) edge (1) edge (3);
+
+\end{tikzpicture} 
+\caption{ }
+\label{fig:careful_with_priors}
+\end{figure}
+
+As each variable is a linear function of its parents, with gaussian noise, resulting in a joint distribution $\P{U,Z,X,Y}$ that is multivariate normal.
+Marginalising out $U$ and conditioning on $X$ and $Z$ yields,
+
+\eqn{
+Y \sim N(w_{yx} X + \beta Z, \epsilon) 
+}
+
+where
+
+\eqn{
+\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and}\\
+\epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
+}
+
+The causal effect of $X$ on $Y$ is identifiable (even without the linear gaussian assumptions) as $Z$ satisfies the backdoor criterion and the expectation of the coefficient for $X$ in a standard OLS regression of $Y$ against $X$ and $Z$ is $w_{yx}$.
+However, the causal effect of $Z$ on $Y$ is not identifiable due the presense of the unobserved confounder $U$.
+The coefficient $\beta$ captures both the causal relationship between $Z$ and $Y$, $w_{yz}$ and the indirect relationship through $U$, which we cannot seperate without observing $U$.
+
+If we do a Bayesian regression of $X$ and $Z$ against $Y$ and use intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ we are likely to select something centered around $w_{yz}$, but this is not correct and can introduce bias into our estimate of the coefficent for $w_{yx}$.
+The prior for $\beta$ should be based on our belief about how we expect $Z$ to be \emph{associated} with $Y$ after marginalising out any confounding variables.
+
+This result is counter intuative.
+When fitting a model to estimate the affect of marketing on sales, it is natural to put a prior on the coefficient for price that is strongly negative because we believe increasing price decreases sales.
+However, our prior should be on the expected association between price and sales, holding fixed all the other variables in the model, not the causal relationship between price and sales.
+An alternative is to explicitly include $U$ as a latent variable in the model.
+However, this comes at a computational cost with little benifit for estimation, unless we have good knoweledge of the relationships between $U$, $Z$ and $Y$.
+More fundamentally, at whatever level we decide to stop adding latent variables and specify a prior, we can introduce this form of bias if priors are selected naivily on the basis of causal intuition.
+
+\section{Example}
+
+\eqn{
+U &\sim N(0 , 1)\\
+Z &\sim N(2U , 0.09)\\
+X &\sim N(0.5Z , 1)\\
+Y &\sim N(3U + -Z + 0.5X , 0.25)
+}
+
+\begin{figure}[h]
+\includegraphics[scale=.3]{prior_no_prior}
+\caption{no prior, posterior on $w_{yx}$ centered around $w_{yx}$}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[scale=.3]{prior_bad_prior}
+\caption{Prior around $w_{yz}$, $\beta \sim N(w_{yz},\sigma = 0.5)$. The posterior on $w_{yx}$ is biased away from its true value}
+\end{figure}
 

### 2017-08-03 causalpy/causalpy/bayesian_networks/models.py

@@ -192,8 +192,8 @@ class LinearGaussianBN(object):
         return self._weights[variable]
     
     def _variance_symbol(self,variable):
-        """returns a symbol for the variance of variable string specified."""
-        return Symbol("V_{0}".format(variable))
+        """returns a symbol for the marginal variance of variable string specified."""
+        return Symbol("e_{0}".format(variable))
     
     def _weight_symbol(self,variable,parent):
         if parent is not None:

### 2017-08-24 thesis/offcuts_chapter6.tex
begins restructure to two intro chapters

@@
+They also consider the impact of different assumptions about the relationship between the suggested action $A$ and actual action $X$, which are not captured by structure of the graph.
+
+\section{Causal Discovery}
+\label{sec:causal-discovery}
+
+%Cyclic models \citep{Hyttinen2012}
+
+We now move to the much more general problem of learning a causal graph from observational data.
+In this setting we make much broader assumptions about the structure of the graph.
+For example, that it is acyclic or that we have no unmeasured confounding variables.
+We do not assume the existence or directions of any links between the variables.
+Amazingly, it is possible to infer some aspects of causal structure with such general assumptions.
+The set of conditional independence in a non-experimental data set indicates some causal structures are more likely than others.
+In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
+These clues are the key to causal discovery algorithms.
+
+%There is also work on inferring causal structure from experiential data \ref{}. We discuss this further in section XXX.
+
+Causal discovery is a much grander goal than causal effect estimation given a known causal network.
+Arguably, if achieved, it would equate to the automation of scientific discovery.
+We need simply supply our algorithm with a vast collection of variables (regardless of their relevance to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make.
+Unfortunately, causal discovery is very hard.
+Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows exponentially with the square of the number of variables .
+
+% - some intuition here about the strength of the signal we get from causal discovery.
+Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}.
+Talk about the numerical as well as the statistical issues here.
+
+In the next sections we briefly survey the key approaches to causal discovery.
+We roughly divide the methods into those based on those that exploit the connection between the conditional independencies in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
+
+%This is not the only way we could have grouped them.
+There is also a separation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
+There are also hybrid methods \ref{}
+
+\subsection{Conditional independence based methods}
+One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
+For any Bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network.
+If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$.
+However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
+This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
+This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
+Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \citep{pearl1988probabilistic} for $P$.
+
+Faithfulness is an assumption.
+It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
+However, most distributions generated by a causal Bayesian network will be faithful to that network.
+For faithfulness to be violated, different causal effects must exactly balance one-another out.
+For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
+If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
+However, this independence is not stable.
+It would disappear under a small perturbation to any of the parameters.
+In discrete systems, violations correspond to the solutions to polynomial equations over values in the CPD tables and thus are a space of measure zero with respect to all possible distributions associated with the graph \citep{Koller2009}.
+
+Given the faithfulness assumption, our causal discovery problem reduces to finding the set of Bayesian networks that have exactly the dependency structure as we observe in $P$.
+This set can also be referred to as the Markov equivalence class compatible with $P$.
+
+\subsubsection{Without hidden common causes}
+The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of Bayesian networks we must consider.
+This assumption is referred to as causal sufficiency \citep{Sprites2000}.
+
+We will begin with a brute force algorithm (described as the SGS algorithm in \citet{Sprites2000} and IC algorithm in \citet{Pearl2000}).
+While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.
+
+\begin{table}[H]
+ \begin{tabularx}{\textwidth}{X}
+ \hline
+\rule{0pt}{2.5ex} 
+ \textbf{The SGS (or IC) Algorithm}\\
+ \hline
+ \rule{0pt}{2.5ex}
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
+\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
+ \begin{enumerate}[itemsep=8pt]
+  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an un-directed link to form a complete graph.
+  \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent.
+If such a set (including the empty set) exists then $a$ and $b$ cannot be directly connected in $G$ so delete the link.
+  \label{alg:SGSexponential}
+
+  \item For all pairs of non-linked variables $(\alpha,\beta)$ with a common neighbour, $c$, if $c \notin \boldsymbol{S}_{\alpha,\beta}$, then $c$ must be a collider in the path $\alpha,c,\beta$ so  add arrows to direct the links $\alpha-c$ and $\beta-c$ towards $c$.
+  \label{alg:SGScolliders}  
+  \item Recursively try to orient any edges that remain un-directed to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
+  \label{alg:SGSfinal}
+\end{enumerate}\\
+ \hline
+\end{tabularx}
+\end{table}
+
+The SGS algorithm utilises the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
+Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
+Unfortunately the number of conditional independence tests required for step \ref{alg:SGSexponential} grows exponentially (in the worst case) with the number of variables.
+Not only that, but for each edge that is in the true network, the algorithm will always tests all other possible subsets of variables.
+If the assumption that there are no hidden common causes or that the distribution is faithful are violated, step \ref{alg:SGScolliders} of the SGS algorithm can produce double headed arrows.
+
+The PC algorithm \citet{Sprites2000} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilise the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
+It also checks for low order conditional independence relations before higher order ones.
+This allows it to exploit any sparsity in the true network, leading to much better average case performance  \citep{Sprites2000} (although the worst case, where the true network is complete, is still exponential).
+With finite data, the order in which the links are considered can change the output (unlike for SGS).
+The effect of wrongly removing a link early on flows through to later conditional independence tests by changing which nodes are considered adjacent.
+
+\begin{table}[H]
+ \begin{tabularx}{\textwidth}{X}
+ \hline
+\rule{0pt}{2.5ex} 
+ \textbf{The PC Algorithm}\\
+ \hline
+ \rule{0pt}{2.5ex}
+\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
+\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
+ \begin{enumerate}[itemsep=8pt]
+  \item As for SGS
+  \item \textbf{for} each link $a-b$:
+  \begin{itemize}[label={}]
+   \item $n = 0$
+   \item $\boldsymbol{A}_{a,b} = \{A_{1}...A_{j}\}$ be the set of nodes adjacent to $a$ and/or $b$
+   \item \textbf{while} $a$ and $b$ are connected and $n < j$:
+   \item 
+    	\begin{itemize}[label={}]
+    	\item \textbf{if} any subset of size $n$ of $\boldsymbol{A}$ makes $a$ and $b$ conditionally independent:
+    	\item \begin{itemize}[label={}]
+    			delete the link
+    		  \end{itemize}
+
+    	\item $n = n+1$
+    	\end{itemize}
+  \end{itemize}   
+  \item as for SGS
+  \item as for SGS
+\end{enumerate}\\
+ \hline
+\end{tabularx}
+\end{table}
+
+The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution.
+Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects from the truncated product formula \ref{eq:truncatedproduct}.
+We can then bound the true causal effect by combining the results for the all the networks.
+This procedure is the IDA algorithm \citep{Maathuis2009} and has been found to outperform standard regularisation techniques at finding causal effects in a high-dimensional yeast gene expression data set \citep{Maathuis2010}.
+An implementation is available in the R package \citep{Kalisch2012}
+
+\subsubsection{With hidden variables}
+There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables.
+With an unknown number of hidden variables there are infinity many possible structures to search over.
+In addition, the space of causal networks is not closed under marginalisation.
+If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalising over $\boldsymbol{U}$, may not be faithful to any Bayesian network (see figure \ref{fig:DAGSnotclosed}).
+The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables).
+
+\begin{figure}
+\centering
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){U}; 
+\node[main node, below left = of 1](2){B};
+\node[main node, below right = of 1](3){C};
+\node[main node, above left= of 2](4){A};
+\node[main node, above right = of 3](5){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    (5) edge node {} (3);
+
+\end{tikzpicture}
+\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalising over $U$. }
+\label{fig:DAGSnotclosed}
+\end{figure}
+
+\begin{theorem}
+\citet{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
+\end{theorem}
+
+Since we only care about the causal relationships between observed variables, it is sufficient to search over networks where any hidden variables have no parents and directly cause two of the observed variables.
+Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable.
+These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalisation and conditioning.
+
+For any DAG with latent (and selection) variables there is a unique MAG \citep{Richardson2002}.
+This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites2000}.
+The logic behind the algorithm is very similar.
+Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations.
+The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \citep{Spirtes1995}.
+Assuming there are no selection variables, the PAG can contain four types of link:
+
+\begin{enumerate}
+\item $X \rightarrow Y$, meaning $X$ causes $Y$
+\item $X \leftrightarrow Y$, meaning there is a latent variable that causes $X$ and $Y$.
+\item $X\ {\circ} {\rightarrow}\ Y$, either $X$ causes $Y$ or a latent variable causes both.
+\item $X\ {\circ} {-}  {\circ}\ Y$, either $X$ causes $Y$ or $Y$ causes $X$ or a latent variable causes both.
+\end{enumerate}
+
+The circles indicate where it is ambiguous if there should be an arrowhead (IE where there is one in some MAGs and not in others in the equivalence class).
+Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).
+
+\begin{figure}
+\centering
+ \begin{subfigure}[t]{0.4\textwidth}
+ \centering
+ \caption{True DAG}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){U}; 
+\node[main node, below left = of 1](2){B};
+\node[main node, below right = of 1](3){C};
+\node[main node, above left= of 2](4){A};
+\node[main node, above right = of 3](5){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (4) edge node {} (2)
+    (5) edge node {} (3);
+
+\end{tikzpicture}
+ \end{subfigure}
+ \begin{subfigure}[t]{0.4\textwidth}
+  \centering
+  \caption{Partial ancestral graph}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed}]
+ %nodes
+
+\node[main node](1){A};
+\node[main node, below right=of 1](2){B};
+\node[main node, right=of 2](3){C};
+\node[main node, above right=of 3](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge[pil] node {} (2)
+    (2) edge node {} (3)
+    (3) edge node {} (2)
+    (4) edge [pil] node {} (3);    	
+\end{tikzpicture}
+\end{subfigure}
+\par\bigskip 
+ \begin{subfigure}[t]{0.4\textwidth}
+ \centering
+ \caption{True DAG}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm, thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+
+\node[main node](1){C};
+\node[main node, above left=of 1](2){A};
+\node[main node, above right=of 1](3){B};
+\node[main node, below=of 1](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (4)
+    (2) edge node {} (1)
+    (3) edge node {} (1);    	
+\end{tikzpicture}
+\end{subfigure}
+ \begin{subfigure}[t]{0.4\textwidth}
+  \centering
+  \caption{Partial ancestral graph}
+ \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={observed}]
+ %nodes
+
+\node[main node](1){C};
+\node[main node, above left=of 1](2){A};
+\node[main node, above right=of 1](3){B};
+\node[main node, below=of 1](4){D};
+
+ \path[every node/.style={font=\sffamily\small}]
+ 	(1) edge node {} (4)
+    (2) edge [pil] node {} (1)
+    (3) edge [pil] node {} (1);    	
+\end{tikzpicture}
+\end{subfigure}
+\caption{FCI examples: true graph and FCI output}
+\label{fig:FCIExamples}
+\end{figure}
+
+The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \citep{Zhang2008}.
+More recently \citet{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
+\citet{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
+
+Latent variables can create constraints on the marginal distribution over the observed variables that cannot be expressed in term of conditional independencies.
+These generalised constraints can be expressed and leveraged within nested Markov models \citep{Richardson2012,Shipster2014}
+
+%\renewcommand{\arraystretch}{1.5}
+%\begin{tabular}{| c | c | p{4cm} | c | c | c |}
+%\hline
+%  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
+%  \hline
+%  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
+%  \hline
+%  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
+%  \hline
+%  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
+%  \hline
+%  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
+%  \hline
+%  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
+%  \hline
+%  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
+% \hline
+%\end{tabular}
+
+All the algorithms discussed in this section rely on being able to perform conditional independence tests.
+This is non-trivial with high dimensional data.
+If the functional relationship between the variables is linear with Gaussian noise then the network represents a multivariate normal distribution and a pair of variables are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero \citep{Koller2009}.
+Where the functions are non-linear one can apply kernelised independence tests \citep{Gretton2008,Zhang2012}
+
+\subsection{Discovery with functional models}
+The algorithms we have considered so far return a Markov equivalence class.
+They cannot distinguish between two models that result in the same set of conditional independence relations.
+Consider the very simple case where  there are only two variables and the possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
+These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
+No algorithm relying purely on conditional independence relations can separate these two cases.
+
+Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$.
+What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from?
+Recall the functional definition of causality (section \ref{sec:SEM}).
+There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \citep{Hoyer2009}, post-non-linear additive noise \citep{Zhang2008a} or linear models with non-Gaussian noise \citep{Hoyer2012}.
+
+The causal direction can also be identified via a connection between casual discovery and semi-supervised learning \citep{Janzing2012}.
+Suppose we are trying to learn $\P{Y|X}$.
+The goal of semi-supervised learning is to improve our estimate of $\P{Y|X}$ by leveraging additional data sampled from $\P{X}$.
+However, if the true causal model is $X \rightarrow Y$ then their is some function mapping values of $X$ to $Y$ which should be invariant to any changes in the input distribution $\P{X}$.
+Therefor $\P{X}$ should be independent of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning.
+However if the true causal model is $Y \rightarrow X$ then variations in the $\P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help.
+This assumption of independence of mechanism and input can also allow the identification of the causal direction in SEMs even where the functions are deterministic and invertable \citep{Daniusis2010}.
+\citet{Janzing2012a} leverage an information geometric viewpoint of the independence of mechanism and input to infer the causal direction between a pair of associated variables.
+
+Instead of positing a functional restriction on the relationship between variables and then developing theory to exploit that assumption, \citep{LopezPaz2014} propose learning what the causal relationship looks like from data.
+They assume there will be a difference between the relationship of $\P{X}$, $\P{Y}$ and $\P{X|Y}$ between $X \rightarrow Y$ versus $Y \rightarrow X$.
+Their algorithm requires a data set in which each row is itself a data set consisting of pairs of variables $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$.
+They use a kernel mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-data set and train an algorithm to learn the direction of causality.
+Unfortunately we do not have a large collection of data sets where the causal direction is known to train such a model.
+\citet{LopezPaz2014} instead use a simulated data set so their model will necessarily be based on the assumptions they make when generating the data.
+Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.
+
+\citep{Peters2014} have extended results from the bi-variate case to the multivariate setting.
+They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable.
+They build on this to develop an algorithm that allows the construction of causal graphs based on the additive noise assumption.
+
 \chapter{More stuff}

### 2017-08-24 thesis/thesis_map.tex
begins restructure to two intro chapters

@@
 
+\newcommand{\V}[1]{\mathbb V\left[#1\right]}
 \newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
@@
 
+\item What are the nessessary and sufficient conditioins for decision making?
+How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function).
+Ie when is the mean sufficient?
+What about the mean and the variance?
+- when can we take one set of results and compute something about a different loss function?
+
 \item Does the way we define supervised learning and loss fuctions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?
@@
 
+-------------------
+
+In chapter \ref{chap:interventionalist_view} I describe the interventionalist viewpoint on learning the outcome of actions.
+
+I review the key problems and results within the bandit liturature
+
+highlight some interesting connections between existing strands of work on bandits and causal inference.
+
+-------------------
+
 Causal questions are everywhere.
@@
 
-\chapter{Causal models}
-\label{chap:causal_models}
+\chapter{Causal inference from observational data}
+\label{chap:causality}
+
+\section{Causal models}
+\label{sec:causal_models}
 
@@
 
+Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes.
+The global Markov property and d-separation also hold in causal networks.
+D-speration, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
+We include a brief introduction in section \ref{sec:dsep}.
+
+\subsection{Can't everything be encoded in a Bayesian model?}
 Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
@@
 
-Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes.
-The global Markov property and d-separation also hold in causal networks.
-D-speration, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
-We include a brief introduction in section \ref{sec:dsep}.
-
-\subsubsection{Limitations of causal Bayesian networks}
+\subsection{Limitations of causal Bayesian networks}
 A number of criticisms have been levelled at this approach to modelling causality.
@@
 
-Their difference is also normal. Let $\tau = \cf{Y}{1} - \cf{Y}{0}$,
+The distribution of \emph{individual treatment effects}, $\tau = \cf{Y}{1} - \cf{Y}{0}$ is also normal,
 
@@
 The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
-
-\section{Granger causality}
-
-Mentioned in the category of causal frameworks,
-
-Granger causality is less of a framework for defining causality and on which a wide range of algorithms and methods can be constructed, but a method for infering causal structure from data.
-As such is discussion really belongs in section \ref{sec:causal-discovery}.
-
-TODO move this to a paragraph at the end of the comparing and unifying section. \todo{}
 
@@
 
-\chapter{Two key questions}
-\label{chap:two_key_questions}
-
+\section{Causal effect estimation} 
+\label{sec:causal effect estimation}
 We can roughly categorise the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
@@
 An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
-
-\section{Causal effect estimation} 
-\label{sec:causal effect estimation}
+\todo{note that I focus on problems where causal structure known, reference rewiew on structure learning}
 
@@
 
-\section{Causal Discovery}
-\label{sec:causal-discovery}
-
-%Cyclic models \citep{Hyttinen2012}
-
-We now move to the much more general problem of learning a causal graph from observational data.
-In this setting we make much broader assumptions about the structure of the graph.
-For example, that it is acyclic or that we have no unmeasured confounding variables.
-We do not assume the existence or directions of any links between the variables.
-Amazingly, it is possible to infer some aspects of causal structure with such general assumptions.
-The set of conditional independence in a non-experimental data set indicates some causal structures are more likely than others.
-In addition, there can be subtle asymmetries in the relationship between the joint distribution of cause and effect and the distributions of cause given effect and effect given cause.
-These clues are the key to causal discovery algorithms.
-
-%There is also work on inferring causal structure from experiential data \ref{}. We discuss this further in section XXX.
-
-Causal discovery is a much grander goal than causal effect estimation given a known causal network.
-Arguably, if achieved, it would equate to the automation of scientific discovery.
-We need simply supply our algorithm with a vast collection of variables (regardless of their relevance to the problem) and it would learn the causal structure and from that allow us to estimate the effects of any intervention we cared to make.
-Unfortunately, causal discovery is very hard.
-Even with the assumption that the causal graph is acyclic and there are no latent variables, the number of possible graphs grows exponentially with the square of the number of variables .
-
-% - some intuition here about the strength of the signal we get from causal discovery.
-Small initial errors in greedy approaches can cascade leading to unstable inference and large losses \ref{sprites?}.
-Talk about the numerical as well as the statistical issues here.
-
-In the next sections we briefly survey the key approaches to causal discovery.
-We roughly divide the methods into those based on those that exploit the connection between the conditional independencies in a joint distribution and the structure of a causal model and those that leverage assumptions about the functional form of the relationships between cause and effect.
-
-%This is not the only way we could have grouped them.
-There is also a separation between constraint based algorithms, that XXXX and those that search and score multiple graph structures before selecting the best.
-There are also hybrid methods \ref{}
-
-\subsection{Conditional independence based methods}
-One general approach is to look for clues about the structure of the network in the conditional independence relations in the distribution.
-For any Bayesian network, $G$, (causal or not) we can read of conditional independencies in the joint distribution from the structure of the network.
-If a set of variables $Z$ d-separates $X$ and $Y$ in $G$ then $(X \ci Y|Z)$ in the distribution $P$.
-However, we want to work in the other direction, from conditional independence in the distribution to the structure of the network.
-This requires that we assume the reverse condition: $(X \ci Y|Z)$ in $P$ must imply $Z$ d-separates $X$ and $Y$ in $G$.
-This assumption, commonly referred to as  \textbf{\textit{faithfulness}} \ref{Sprites2000}, says there are no additional independence relations that are satisfied in $P$ but not in all distributions $\boldsymbol{P'}$ that are compatible with $G$.
-Stating that $P$ is faithful to $G$ is equivalent to $G$ is a \textbf{\textit{perfect map}} \citep{pearl1988probabilistic} for $P$.
-
-Faithfulness is an assumption.
-It does not always hold and we cannot verify it from the observational data we wish to use for causal inference.
-However, most distributions generated by a causal Bayesian network will be faithful to that network.
-For faithfulness to be violated, different causal effects must exactly balance one-another out.
-For example, consider a simple binary variable model of chocolate consumption, income and obesity (figure).
-If the coefficients in the conditional probability tables are just right then the direct effect of chocolate on obesity will exactly balance the indirect effect through income and obesity will appear independent of chocolate consumption.
-However, this independence is not stable.
-It would disappear under a small perturbation to any of the parameters.
-In discrete systems, violations correspond to the solutions to polynomial equations over values in the CPD tables and thus are a space of measure zero with respect to all possible distributions associated with the graph \citep{Koller2009}.
-
-Given the faithfulness assumption, our causal discovery problem reduces to finding the set of Bayesian networks that have exactly the dependency structure as we observe in $P$.
-This set can also be referred to as the Markov equivalence class compatible with $P$.
-
-\subsubsection{Without hidden common causes}
-The strong assumption that there are no hidden variables that cause two or more variables in $\boldsymbol{V}$ significantly reduces the 'search space' of Bayesian networks we must consider.
-This assumption is referred to as causal sufficiency \citep{Sprites2000}.
-
-We will begin with a brute force algorithm (described as the SGS algorithm in \citet{Sprites2000} and IC algorithm in \citet{Pearl2000}).
-While it is impractical for all but the smallest of networks, it demonstrates key concepts that also underlie the more useful and complex algorithms we will discuss later.
-
-\begin{table}[H]
- \begin{tabularx}{\textwidth}{X}
- \hline
-\rule{0pt}{2.5ex} 
- \textbf{The SGS (or IC) Algorithm}\\
- \hline
- \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
- \begin{enumerate}[itemsep=8pt]
-  \item Join all pairs of vertices $(a,b) \in \boldsymbol{V}$ with an un-directed link to form a complete graph.
-  \item For each link $a-b$ search for a set $\boldsymbol{S}_{a,b} \subseteq V \setminus \{a,b\}$ that renders $a$ and $b$ conditionally independent.
-If such a set (including the empty set) exists then $a$ and $b$ cannot be directly connected in $G$ so delete the link.
-  \label{alg:SGSexponential}
-
-  \item For all pairs of non-linked variables $(\alpha,\beta)$ with a common neighbour, $c$, if $c \notin \boldsymbol{S}_{\alpha,\beta}$, then $c$ must be a collider in the path $\alpha,c,\beta$ so  add arrows to direct the links $\alpha-c$ and $\beta-c$ towards $c$.
-  \label{alg:SGScolliders}  
-  \item Recursively try to orient any edges that remain un-directed to avoid creating cycles (because they are not there by assumption) and additional colliders (because any colliders were found in step \ref{alg:SGScolliders}).
-  \label{alg:SGSfinal}
-\end{enumerate}\\
- \hline
-\end{tabularx}
-\end{table}
-
-The SGS algorithm utilises the fact that a collider structure (figure \ref{fig:dseptripple4}) induces a distinct conditional independence relation.
-Assuming you have a consistent conditional independence test, it converges to return a partially directed network that represents the Markov equivalence class for the generating causal model.
-Unfortunately the number of conditional independence tests required for step \ref{alg:SGSexponential} grows exponentially (in the worst case) with the number of variables.
-Not only that, but for each edge that is in the true network, the algorithm will always tests all other possible subsets of variables.
-If the assumption that there are no hidden common causes or that the distribution is faithful are violated, step \ref{alg:SGScolliders} of the SGS algorithm can produce double headed arrows.
-
-The PC algorithm \citet{Sprites2000} modifies step \ref{alg:SGSexponential} of the SGS algorithm to utilise the fact that if two variables $(a,b)$ are conditionally independent given some set, they will also be conditionally independent given a set that contains only variables adjacent to $a$ or $b$.
-It also checks for low order conditional independence relations before higher order ones.
-This allows it to exploit any sparsity in the true network, leading to much better average case performance  \citep{Sprites2000} (although the worst case, where the true network is complete, is still exponential).
-With finite data, the order in which the links are considered can change the output (unlike for SGS).
-The effect of wrongly removing a link early on flows through to later conditional independence tests by changing which nodes are considered adjacent.
-
-\begin{table}[H]
- \begin{tabularx}{\textwidth}{X}
- \hline
-\rule{0pt}{2.5ex} 
- \textbf{The PC Algorithm}\\
- \hline
- \rule{0pt}{2.5ex}
-\textbf{Input:} A distribution $P$, over variables $\boldsymbol{V} = \{V_{1}...V_{k}\}$, that was generated by and is faithful to an (unknown) Bayesian network $G$\\
-\textbf{Output:} A partially directed network that represents the Markov equivalence class of $G$\\
- \begin{enumerate}[itemsep=8pt]
-  \item As for SGS
-  \item \textbf{for} each link $a-b$:
-  \begin{itemize}[label={}]
-   \item $n = 0$
-   \item $\boldsymbol{A}_{a,b} = \{A_{1}...A_{j}\}$ be the set of nodes adjacent to $a$ and/or $b$
-   \item \textbf{while} $a$ and $b$ are connected and $n < j$:
-   \item 
-    	\begin{itemize}[label={}]
-    	\item \textbf{if} any subset of size $n$ of $\boldsymbol{A}$ makes $a$ and $b$ conditionally independent:
-    	\item \begin{itemize}[label={}]
-    			delete the link
-    		  \end{itemize}
-
-    	\item $n = n+1$
-    	\end{itemize}
-  \end{itemize}   
-  \item as for SGS
-  \item as for SGS
-\end{enumerate}\\
- \hline
-\end{tabularx}
-\end{table}
-
-The PC algorithm also returns a set of Markov equivalent networks consistent with the distribution.
-Since we have assumed there are no hidden variables, for any single graph in this set we can calculate causal effects from the truncated product formula \ref{eq:truncatedproduct}.
-We can then bound the true causal effect by combining the results for the all the networks.
-This procedure is the IDA algorithm \citep{Maathuis2009} and has been found to outperform standard regularisation techniques at finding causal effects in a high-dimensional yeast gene expression data set \citep{Maathuis2010}.
-An implementation is available in the R package \citep{Kalisch2012}
-
-\subsubsection{With hidden variables}
-There are an number of difficulties in extending the approach of the last section to deal with the case where there are latent variables.
-With an unknown number of hidden variables there are infinity many possible structures to search over.
-In addition, the space of causal networks is not closed under marginalisation.
-If we have a distribution that $P'(\boldsymbol{O},\boldsymbol{U})$ generated by and is faithful to a network $G$ the distribution $P(\boldsymbol{O})$, that results from marginalising over $\boldsymbol{U}$, may not be faithful to any Bayesian network (see figure \ref{fig:DAGSnotclosed}).
-The key to constraining the space of possible models is that many latent structures are equivalent (under transforms of the hidden variables).
-
-\begin{figure}
-\centering
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){U}; 
-\node[main node, below left = of 1](2){B};
-\node[main node, below right = of 1](3){C};
-\node[main node, above left= of 2](4){A};
-\node[main node, above right = of 3](5){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    (5) edge node {} (3);
-
-\end{tikzpicture}
-\caption{A distribution faithful to this DAG is not faithful to any DAG over the variables $\{A,B,C,D\}$ after marginalising over $U$. }
-\label{fig:DAGSnotclosed}
-\end{figure}
-
-\begin{theorem}
-\citet{Verma1993} For every latent structure there is a dependency equivalent structure such that every latent (unobserved) variable is a root node with exactly two children .
-\end{theorem}
-
-Since we only care about the causal relationships between observed variables, it is sufficient to search over networks where any hidden variables have no parents and directly cause two of the observed variables.
-Instead of representing hidden variables explicitly we can capture the necessary independence relations with a more general graphical model that supports bi-directed edges that play the role of a hidden confounding variable.
-These models, referred to as maximal ancestral graphs (MAGs) are closed under marginalisation and conditioning.
-
-For any DAG with latent (and selection) variables there is a unique MAG \citep{Richardson2002}.
-This makes it possible to extend the PC algorithm to latent structures, resulting in the FCI algorithm \citep{Sprites2000}.
-The logic behind the algorithm is very similar.
-Certain structures are ruled out as a consequence of being inconstant with the observed conditional independence relations.
-The output is an equivalence class of MAGs, which can be represented graphically as a partial ancestral graph PAG \citep{Spirtes1995}.
-Assuming there are no selection variables, the PAG can contain four types of link:
-
-\begin{enumerate}
-\item $X \rightarrow Y$, meaning $X$ causes $Y$
-\item $X \leftrightarrow Y$, meaning there is a latent variable that causes $X$ and $Y$.
-\item $X\ {\circ} {\rightarrow}\ Y$, either $X$ causes $Y$ or a latent variable causes both.
-\item $X\ {\circ} {-}  {\circ}\ Y$, either $X$ causes $Y$ or $Y$ causes $X$ or a latent variable causes both.
-\end{enumerate}
-
-The circles indicate where it is ambiguous if there should be an arrowhead (IE where there is one in some MAGs and not in others in the equivalence class).
-Counter-intuitively it is sometimes possible to  rule out or confirm the existence of a confounding variable and fully determine the causal type of a link (see examples in figure \ref{fig:FCIExamples}).
-
-\begin{figure}
-\centering
- \begin{subfigure}[t]{0.4\textwidth}
- \centering
- \caption{True DAG}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){U}; 
-\node[main node, below left = of 1](2){B};
-\node[main node, below right = of 1](3){C};
-\node[main node, above left= of 2](4){A};
-\node[main node, above right = of 3](5){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (4) edge node {} (2)
-    (5) edge node {} (3);
-
-\end{tikzpicture}
- \end{subfigure}
- \begin{subfigure}[t]{0.4\textwidth}
-  \centering
-  \caption{Partial ancestral graph}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.7cm,  thick,main node/.style={observed}]
- %nodes
-
-\node[main node](1){A};
-\node[main node, below right=of 1](2){B};
-\node[main node, right=of 2](3){C};
-\node[main node, above right=of 3](4){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge[pil] node {} (2)
-    (2) edge node {} (3)
-    (3) edge node {} (2)
-    (4) edge [pil] node {} (3);    	
-\end{tikzpicture}
-\end{subfigure}
-\par\bigskip 
- \begin{subfigure}[t]{0.4\textwidth}
- \centering
- \caption{True DAG}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm, thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (4)
-    (2) edge node {} (1)
-    (3) edge node {} (1);    	
-\end{tikzpicture}
-\end{subfigure}
- \begin{subfigure}[t]{0.4\textwidth}
-  \centering
-  \caption{Partial ancestral graph}
- \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=0.7cm,  thick,main node/.style={observed}]
- %nodes
-
-\node[main node](1){C};
-\node[main node, above left=of 1](2){A};
-\node[main node, above right=of 1](3){B};
-\node[main node, below=of 1](4){D};
-
- \path[every node/.style={font=\sffamily\small}]
- 	(1) edge node {} (4)
-    (2) edge [pil] node {} (1)
-    (3) edge [pil] node {} (1);    	
-\end{tikzpicture}
-\end{subfigure}
-\caption{FCI examples: true graph and FCI output}
-\label{fig:FCIExamples}
-\end{figure}
-
-The FCI algorithm can be made complete such that it discovers all aspects of the true causal structure that are identifiable from the conditional independence relations of a distribution over observed variables and the faithfulness assumption \citep{Zhang2008}.
-More recently \citet{Colombo2012} have proposed the RFCI algorithm, which in some cases returns more ambiguous links than FCI but is substantially faster.
-\citet{Claassen2013} point out that the problem of learning sparse causal networks from data is not NP-hard and propose the FCI+ algorithm, that requires $O(N^{2(k+2)})$ conditional independence tests, where $k$ is the maximum node degree over the observed variables.
-
-Latent variables can create constraints on the marginal distribution over the observed variables that cannot be expressed in term of conditional independencies.
-These generalised constraints can be expressed and leveraged within nested Markov models \citep{Richardson2012,Shipster2014}
-
-%\renewcommand{\arraystretch}{1.5}
-%\begin{tabular}{| c | c | p{4cm} | c | c | c |}
-%\hline
-%  \textbf {Alg.} &\textbf{ Method }& \textbf{Scales (num.vars) }& $\sim $ \textbf {Vars} & \textbf {Latent } & \textbf {Reference} \\
-%  \hline
-%  IC/SGS & Constraint based & Exponential & 10 & No & Pearl(2000)/Sprites(2000)\\
-%  \hline
-%  PC & Constraint based & Worst case exponential, polynomial for sparse graphs & 5000 & No & Sprites(2000) \\ 
-%  \hline
-%  FCI & Constraint based & Worst case exponential, polynomial variant FCI+ for sparse graphs & 30 & Yes & Sprites(2000) \\
-%  \hline
-%  RFCI & Constraint based & ? & 500 & Yes & Colombo(2012) \\
-%  \hline
-%  GES & Search \& Score  & Worst case exponential & 50 & No & Chickering(2002) \\
-%  \hline
-%  MMHC & Hybrid & ? & 5000 & No & Tsamardinos(2006) \\
-% \hline
-%\end{tabular}
-
-All the algorithms discussed in this section rely on being able to perform conditional independence tests.
-This is non-trivial with high dimensional data.
-If the functional relationship between the variables is linear with Gaussian noise then the network represents a multivariate normal distribution and a pair of variables are conditionally independent if and only if the corresponding entry in the inverse correlation matrix is non-zero \citep{Koller2009}.
-Where the functions are non-linear one can apply kernelised independence tests \citep{Gretton2008,Zhang2012}
-
-\subsection{Discovery with functional models}
-The algorithms we have considered so far return a Markov equivalence class.
-They cannot distinguish between two models that result in the same set of conditional independence relations.
-Consider the very simple case where  there are only two variables and the possible causal structures are $X \rightarrow Y$ or $Y \rightarrow X$.
-These models have the same dependency structure but in one case $P(Y|do(X)) = P(Y|X)$ and in the other $P(Y|do(X)) = P(Y)$.
-No algorithm relying purely on conditional independence relations can separate these two cases.
-
-Let us focus only on the two variable case $X \rightarrow Y$ or $Y \rightarrow X$.
-What possible clues could there be in the distribution $P(X,Y)$ that could indicate which causal model it was generated from?
-Recall the functional definition of causality (section \ref{sec:SEM}).
-There are a number of assumptions about the form of the functions that can allow us to identify the causal direction: non-invertable functions, additive noise \citep{Hoyer2009}, post-non-linear additive noise \citep{Zhang2008a} or linear models with non-Gaussian noise \citep{Hoyer2012}.
-
-The causal direction can also be identified via a connection between casual discovery and semi-supervised learning \citep{Janzing2012}.
-Suppose we are trying to learn $\P{Y|X}$.
-The goal of semi-supervised learning is to improve our estimate of $\P{Y|X}$ by leveraging additional data sampled from $\P{X}$.
-However, if the true causal model is $X \rightarrow Y$ then their is some function mapping values of $X$ to $Y$ which should be invariant to any changes in the input distribution $\P{X}$.
-Therefor $\P{X}$ should be independent of $\P{Y|X}$ and semi-supervised learning should not perform any better than standard supervised learning.
-However if the true causal model is $Y \rightarrow X$ then variations in the $\P{X}$ can result from both the input distribution over $Y$ and the mapping from $X$ to $Y$ and semi-supervised learning could help.
-This assumption of independence of mechanism and input can also allow the identification of the causal direction in SEMs even where the functions are deterministic and invertable \citep{Daniusis2010}.
-\citet{Janzing2012a} leverage an information geometric viewpoint of the independence of mechanism and input to infer the causal direction between a pair of associated variables.
-
-Instead of positing a functional restriction on the relationship between variables and then developing theory to exploit that assumption, \citep{LopezPaz2014} propose learning what the causal relationship looks like from data.
-They assume there will be a difference between the relationship of $\P{X}$, $\P{Y}$ and $\P{X|Y}$ between $X \rightarrow Y$ versus $Y \rightarrow X$.
-Their algorithm requires a data set in which each row is itself a data set consisting of pairs of variables $(x_i,y_i)$ with a label indicating the direction of causality between $X$ and $Y$.
-They use a kernel mean embedding to represent the distributions $\P{X}$, $\P{Y}$ and $\P{X|Y}$ as features for each individual sub-data set and train an algorithm to learn the direction of causality.
-Unfortunately we do not have a large collection of data sets where the causal direction is known to train such a model.
-\citet{LopezPaz2014} instead use a simulated data set so their model will necessarily be based on the assumptions they make when generating the data.
-Nonetheless this approach makes it possible to rapidly construct a model from a wide range of possible assumptions, without doing a lot of theory to design a specific algorithm optimised to that setting.
-
-\citep{Peters2014} have extended results from the bi-variate case to the multivariate setting.
-They show that if we can come up with a condition that guarantees identifiability for the bi-variate case, we can extend that result to get the conditions under which the multivariate case is identifiable.
-They build on this to develop an algorithm that allows the construction of causal graphs based on the additive noise assumption.
-
-\chapter{The interventionist viewpoint}
+\chapter{Learning from interventions}
+\label{chap:interventionalist_view}
 
@@
 \begin{definition}[Stochasitic k-armed bandit problem]
-Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{y_1,...,y_k}$ be a joint distribution over the rewards for each action.
+Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action.
 The multi-armed bandit problem proceeds over $T$ rounds.
@@
 \item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the world stochastically generates the rewards for each action, $[Y_{t,1},...,Y_{t,k}] \sim \P{\vec{y}}$
-\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
+\item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
+\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
 \end{enumerate}
-At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T Y_{t,a_t}$.
+
+At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T \cfb{Y}{a_t}$.
 We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
+Note we have used counterfactual notation (see ) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have recieved if it selected action $i$ at timestep $t$.
 \end{definition}
@@
 \label{eqn:raw_regret}
-\rawregret = \sum_{t=1}^T Y_{t,i^*} - \sum_{t=1}^T Y_{t,a_t}
+\rawregret = \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
 }
@@
 \label{eqn:pseudo_regret}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}} \\
-&= n\mu_{i^*} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}} \\
+&= n\mu_{i^*} - \E{\sum_{t=1}^T{\cfb{Y}{i}}}
 }
@@
 Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
+
+\subsection{The counterfactual nature of regret}
+\todo{Fix this bit up}
+
+Note that we have used the conterfactual notation to represent the rewards given each action.
+The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep.
+The psuedo-regret (equation \ref{eqn:pseudo_regret}) remains unchanged as both it and the learner's actions depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
+However, the regret as defined in equation \ref{eqn:raw_regret} is no longer properly defined as it relies on both $\cfb{Y}{a_t}$ and $\cfb{Y}{i^*}$ being defined at each timestep.
+The variance (but not expectation) of $\rawregret$ depends on the covariance between counterfactual variables.
+
+ The learner is only able to observe properties of the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
+ The stochastic bandit problem can be defined without recourse to counterfactual variables by
+
+ The algorithm and definition of regret depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
 
@@
 Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, is not a good measure of an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings were the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
-Adversarial bandit algorithms may perform better in non-stationary settings than standard stochastic policies to the extend that they explore more (to avoid the adversary simulating their behaviour), however it is preferable to develop algorithms specifically for non-stationary settings (subject to assumptions about how rapidly or frequently rewards can change), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
+However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, does not fully capture an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
+In settings where the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
+An example of a setting that lies between stochastic and adversarial bandit problems is the non-stationary setting, in which the rewards are generated stochastically from a distribution that varies over time.
+Adversarial bandit algorithms may perform better in such settings than standard stochastic policies to the extent that they explore more (to avoid the adversary simulating their behaviour) and thus adapt quicker to changes in the reward distribution.
+Adversarial algorithms also have stronger worst case regret guarantees, since even the weak regret for stochastic bandits is not guaranteed to be sublinear in such settings.
+However, if we have constraints on how rapidly or frequently the reward distributions can change over time it is better to use algorithms specificly developed to exploit such information (and compare them against a stronger notion of regret), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
+
+from the information available to the agent, the joint distribution over counterfactuals cannot be inferred, unless the assignment mechanism in noiseless - so it does not make sense to define regret that depends on it.
 
@@
 
+\subsubsection{The causal structure of contextual bandits}
+
+\todo{think carefully about the graphs in this sections}
 It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
@@
 \node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$Y$};
+\node[main node, below=of 1](2){$\vec{Y}^a$};
 \node[main node, above right=of 2](3){$A$};
@@
 A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
-Each action $a \in \actionspace$ assigns a value to a subset of the variables in $\mathcal{X}$.
-One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable} and takes on values in $\{0, 1\}$.
+Each action $a \in \actionspace$ assigns a value to a subset of the variables in $\mathcal{X}$ and incurs a known cost $C(a)$ on the learner.
+One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable}.
 
@@
 \item \emph{observes} the value of a subset of the variables $\vec{X}^c_t$,
-\item \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations, and finally
-\item \emph{observes} sampled values for another subset of variables $\vec{X}^o_t$ drawn from $\P{\vec{X}^o_t | do(\vec{X}_t = \vec{x}_t)}$ including the \emph{reward} $Y_t \in \{0,1\}$.
+\item \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations and rewards, 
+\item \emph{observes} values for another subset of variables $\vec{X}^o_t$ drawn from $\P{\vec{X}^o_t |\vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$,
+\item \emph{obtains reward} $r_t = Y_t - C(a_t)$, where $Y_t$ is sampled from $\P{Y_t| \vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$
 \end{enumerate} 
+\end{definition}
 
@@
 A variable may be observed both prior to and after the agent selects an action, and the action may change its value.
+The notation $\P{\cdot|\vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$ denotes distributions conditional on having observed $\vec{X}^c_t$ and \emph{then} intervened to set $\vec{X}_t = \vec{x}_t$.
+The values of variables in $\vec{X}^c_t$ that are non-decendents of $\vec{X}_t$ remain unchanged by the intervention.
 The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
 
-\end{definition}
-
-The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
+The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and correspondign costs, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
+If we can (cheaply) intervene on all of the parents of $Y$ simultaneusly then any context or alternative actions are irrelevent and the problem reduces to a standard multi-armed bandit problem.
+This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.
+
+\vspace{0.5cm}
+\begin{theorem}
+\label{theorem:cb-intervene-all-parents}
+Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$.
+If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and the problem reduces to a standard multi-armed bandit (over actions in in $\mathcal{A'}$).
+\begin{proof}
+for any action $a \in \mathcal{A}$, 
+\eq{
+ \E{Y|\vec{X}^c_t,a} &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\vec{X}^c_t,a,\parents{Y}}} \\
+ &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\parents{Y}}}\\
+ &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|do(\parents{Y})}}\\
+ &\leq \max_{\parents{Y}}\E{Y|do(\parents{Y})} = \E{Y|a'} \text{ for some $a' \in \mathcal{A'}$}
+}
+\end{proof}
+\end{theorem}
+
 If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
@@
 We only get this form of side information for actions that don't specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-If additional variables are only observed before an intervention is selected the causal bandit problem reduces to stochastic contextual bandits, which are already reasonably well understood \citep{Agarwal2014}.
-
-We note that classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
+If additional variables are only observed before an intervention is selected the causal bandit problem can be treated as a stochastic contextual bandit problem, which are already reasonably well understood \citep{Agarwal2014}.
+
+Two other problems that sit in the space between causal inference and bandit problems, bandits with unobserved confounders \cite{Bareinboim2015} and compliance aware bandits \citep{DellaPenna2016}, can also be viewed as specific causal bandit problems.
+In the work on bandits with unobserved confounders it is assumed that the reward given each action may depend on some latent variable $U$ which we cannot observe directly.
+However, prior to selecting an action, we can observe  $I$, the action that would have been selected under an alternate (stationary) policy, which may depend on $U$, see figure \ref{fig:confounded-causal-bandit-graph}.
+In this case the set of contextual variables $\vec{X}^c = I$, the set of post-action feedback variables $\vec{X}^o = \set{Y}$ and the action space consists of all possible assigments of values to a single node $X$, $\actionspace = do(X=x)$.
+This setting reduces to a contextual bandit problem in our causal bandit framework.
+However in thier work on bandits with unobserved confounders \citet{Forney2017} also leverage the fact that $I$ represents the action that would have been selected under an alternate policy to fuse data collected under the previous (observational) policy with data collected under the new policy.
+This information is not encoded in the causal bandit graph in figure \ref{fig:confounded-causal-bandit-graph}, as $I$ could be any variable that is influenced by the unobserved context $U$, and thus cannot be exploited by a standard contextual bandit algorithm.
+
+Compliance aware bandits describe situations in which the action recommended by the bandit algorithm is not always followed.
+For example a patient may refuse to take a treatment or an advertiser may have complex rules about how many ads a given customer can recieve which prevends some of the suggestions from the ad recomendation engine from being followed.
+After an action is selected the algorithm can observe the action that was actually taken in addition to the reward.
+\citet{DellaPenna2016} analyse this setting with binary treatments both with and without the presense of a latent confounding variable $U$, see figure \ref{fig:imperfect-compliance-causal-bandit-graph}.
+In this case, there are no contextual variables and the action space is again the set of assignments to a single variable but there is post-action feedback, which reveals the value of the action that was actually taken.\footnote{There are some interesting variants of the compliance aware bandit setting that, to my knowledge, have not been analysed.
+The first is if the confounding variable $U$ is observable, either as context or post-action feedback.
+The second is if we extend the allowable action set to include acting directly on $X$, albeit at a higher cost than acting on $A$.
+It is also worth noting the connection between this setting and instrumental variables \cite{balke1997bounds}.
+By making some functional assumptions about the relationships between the variables, we can use $A$ as an instrumental variable to bound or estimate the (casual) effect of $X$ on $Y$.
+The estimation will be somewhat complicated in the bandit setting because the action chosen at each timestep is dependent on the previous sequence actions and rewards.}
+
+\begin{figure}[h]
+\centering
+\begin{subfigure}[t]{0.49\textwidth}
+\centering
+\caption{Bandits with unobserved confounders:$\vec{X}^c=\set{I}$, $\vec{X}^o = \set{Y}$, $\actionspace = do(X=x)$}
+\label{fig:confounded-causal-bandit-graph}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){${U}$};
+\node[main node, below right =of 1](2){${Y}$};
+\node[main node, below left=of 1](3){$I$};
+\node[main node, above right=of 2](4){$X$};
+ \path[every node/.style={font=\sffamily\small}]
+   (1) edge (2) edge (3)
+   (4) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.49\textwidth}
+\centering
+\caption{Compliance aware bandits: $\vec{X}^c=\set{}$, $\vec{X}^o = \set{T,Y}$, $\actionspace = do(X=x)$ }
+\label{fig:imperfect-compliance-causal-bandit-graph}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$U$};
+\node[main node, below left=of 1](2){$T$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 2](4){$X$};
+ \path[every node/.style={font=\sffamily\small}]
+   (1) edge (2) edge (3)
+   (2) edge (3)
+   (4) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\caption{}
+\label{fig:causal-bandit-version-confounded-compliance}
+\end{figure}
+
+The classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
 The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
@@
 
-We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, and where the goal of the learner is to minimise the simple regret.
+We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, the cost of all allowable actions is equal and the goal of the learner is to minimise the simple regret.
 I presented this work at NIPS 2016 \citep{NIPS2016_6195}.
@@
 In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
-Note that these modifications will not improve the mini-max regret, which at least for the parallel bandit is already optimal.
-For this reason we prefer to emphasise the main point that causal structure should be exploited when available.
+This has now been done by \citet{Sen2017a}, leading to problem dependent regret bounds for causal bandit problems.
+Note that these modifications do not improve the mini-max regret, which at least for the parallel bandit is already optimal.
+For this reason we focused on emphasising the point that causal structure can and should be exploited when available.
 Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
@@
 
-\chapter*{Careful with that prior}
-
-Suppose we have a data generated by the linear guassian Bayesian network below.
-The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$.
-The variable $U$ is latent.
-
-\eqn{
-\P{U} &= N(0 , v_u) \\
-\label{eqn:careful_with_priors}
-\P{Z|U} &= N(w_{zu}U , v_z) \\
-\P{X|Z} &= N(w_{xz}Z , v_x) \\
-\P{Y|U,Z,X} & = N(w_{yu}U +w_{yz}Z + w_{yx}X , v_y)
-}
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
-
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[lt, right=of 1](4){$U$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    	edge (3)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-
-\end{tikzpicture} 
-\caption{ }
-\label{fig:careful_with_priors}
-\end{figure}
-
-As each variable is a linear function of its parents, with gaussian noise, resulting in a joint distribution $\P{U,Z,X,Y}$ that is multivariate normal.
-Marginalising out $U$ and conditioning on $X$ and $Z$ yields,
-
-\eqn{
-Y \sim N(w_{yx} X + \beta Z, \epsilon) 
-}
-
-where
-
-\eqn{
-\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and}\\
-\epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
-}
-
-The causal effect of $X$ on $Y$ is identifiable (even without the linear gaussian assumptions) as $Z$ satisfies the backdoor criterion and the expectation of the coefficient for $X$ in a standard OLS regression of $Y$ against $X$ and $Z$ is $w_{yx}$.
-However, the causal effect of $Z$ on $Y$ is not identifiable due the presense of the unobserved confounder $U$.
-The coefficient $\beta$ captures both the causal relationship between $Z$ and $Y$, $w_{yz}$ and the indirect relationship through $U$, which we cannot seperate without observing $U$.
-
-If we do a Bayesian regression of $X$ and $Z$ against $Y$ and use intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ we are likely to select something centered around $w_{yz}$, but this is not correct and can introduce bias into our estimate of the coefficent for $w_{yx}$.
-The prior for $\beta$ should be based on our belief about how we expect $Z$ to be \emph{associated} with $Y$ after marginalising out any confounding variables.
-
-This result is counter intuative.
-When fitting a model to estimate the affect of marketing on sales, it is natural to put a prior on the coefficient for price that is strongly negative because we believe increasing price decreases sales.
-However, our prior should be on the expected association between price and sales, holding fixed all the other variables in the model, not the causal relationship between price and sales.
-An alternative is to explicitly include $U$ as a latent variable in the model.
-However, this comes at a computational cost with little benifit for estimation, unless we have good knoweledge of the relationships between $U$, $Z$ and $Y$.
-More fundamentally, at whatever level we decide to stop adding latent variables and specify a prior, we can introduce this form of bias if priors are selected naivily on the basis of causal intuition.
-
-\section{Example}
-
-\eqn{
-U &\sim N(0 , 1)\\
-Z &\sim N(2U , 0.09)\\
-X &\sim N(0.5Z , 1)\\
-Y &\sim N(3U + -Z + 0.5X , 0.25)
-}
-
-\begin{figure}[h]
-\includegraphics[scale=.3]{prior_no_prior}
-\caption{no prior, posterior on $w_{yx}$ centered around $w_{yx}$}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[scale=.3]{prior_bad_prior}
-\caption{Prior around $w_{yz}$, $\beta \sim N(w_{yz},\sigma = 0.5)$. The posterior on $w_{yx}$ is biased away from its true value}
-\end{figure}
+\chapter*{Regret is a goat}
+
+Consider a stochastic, two-armed bernoulli bandit with the (general) joint distribution of the (conterfactual) rewards $\P{\cf{Y}{1},\cf{Y}{2}}$ given in table \ref{table:joint_bernoulli_counterfactuals}.
+Suppose without loss of generality that arm 1 is the optimal arm.
+
+\begin{table}[h]
+\caption{General joint distribution over two bernoulli random variables.}
+\label{table:joint_bernoulli_counterfactuals}
+\begin{center}
+\begin{tabular}{ c | c | c }
+$\cf{Y}{1}$ & $\cf{Y}{2}$ & $\P{\cf{Y}{1},\cf{Y}{2}}$ \\
+  \hline			
+  1 & 1 & $a$ \\
+  1 & 0 & $p-a$ \\
+  0 & 1 & $q-a$ \\
+  0 & 0 & $1+a - (p+q)$ \\
+  \hline  
+\end{tabular}
+\end{center}
+\end{table}
+
+Recall that the regret is defined by,
+
+\eqn{
+\rawregret = \sum_{t=1}^T (\cfb{Y}{i^*} - \cfb{Y}{a_t})
+}
+
+Consider an algorithm that always selects arm 2.
+The variance (but not expectation) of the regret incured by this algorithm depends on the covariance between the counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{2}$, which cannot estimated from data.
+
+\eqn{
+\E{\rawregret} = \sum_{t=1}^T(\E{\cf{Y}{1}}-\E{\cf{Y}{2}}) = T(p-q)
+}
+
+\eqn{
+\V{\rawregret} &= \E{\left(\sum_{t=1}^T (\cfb{Y}{1} - \cfb{Y}{2})\right)^2} - \left(\sum_{t=1}^T(\E{\cf{Y}{1}}-\E{\cf{Y}{2}})\right)^2\\
+&= \sum_{t=1}^T \E{(\cfb{Y}{1} - \cfb{Y}{2})^2 - (\E{\cfb{Y}{1}}-\E{\cfb{Y}{2}})^2}\\
+&= \sum_{t=1}^T\left(\V{\cfb{Y}{1}}+\V{\cfb{Y}{2}} - 2 Cov(\cfb{Y}{1},\cfb{Y}{2})\right)\\ &= T\left(p(1-p)+q(1-q) - 2(a-pq)\right)
+}
 

### 2017-08-26 causal-bandit/arXiv-extended/causal-bandits-extended.tex
creates extended version of nip paper for arXiv, restructuring work on thesis

--- 2017-08-24 thesis/thesis_map.tex
@@
-\documentclass[11pt,a4paper,oneside]{book}
+\documentclass{article}
+
+%\usepackage[final]{nips_anon}
+
 \usepackage[table]{xcolor}
@@
 
-\newcommand{\V}[1]{\mathbb V\left[#1\right]}
 \newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
@@
 
-\author{Finnian Lattimore}
-\title{Learning how to act: making good decisions with machine learning}
+\title{Causal Bandits: Learning Good Interventions via Causal Inference}
+
+% The \author macro works with any number of authors. There are two
+% commands used to separate the names and addresses of multiple
+% authors: \And and \AND.
+%
+% Using \And between authors leaves it to LaTeX to determine where to
+% break the lines. Using \AND forces a line break at that point. So,
+% if LaTeX puts 3 of 4 authors names on the first line, and the last
+% on the second line, try using \AND instead of \And before the third
+% author name.
+\author{
+  Finnian Lattimore \\
+  Australian National University and Data61/NICTA \\
+  \texttt{finn.lattimore@gmail.com} \\
+   \and
+   Tor Lattimore \\
+   University of Alberta \\
+  \texttt{tor.lattimore@gmail.com} \\
+   \and
+   Mark D. Reid \\
+   Australian National University and Data61/NICTA \\
+   \texttt{mark.reid@anu.edu.au} \\
+}
 
 \begin{document}
-\todo{figure out theoremstle is undefined}
-%\def\ci{\perp\!\!\!\perp} % from Wikipedia
 
@@
 
-\epigraph{In vain the grave, with retrospective Eye,
-Would from the apparent what conclude the why,
-Infer the Motive from the Deed, and show
-That what we chanced, was what we meant, to do.}{Alexander Pope}
-
-\tableofcontents
-
-\chapter*{Todo}
-\begin{enumerate}
-\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
-\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
-\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
-When is a problem causal, what bounds does predictive accuracy give on causal models?
-To what degree is predictive accuracy a metric on which causal models can be judged?
-\item Benchmark estimation algorithms
-\item clarify causal interpretation of Dirichlet models
-\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
-\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption.
-For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability.
-\item Does defining causality in term of the truncated product formula ensure that a causal Bayesian network is unique?
-It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition.
-\end{enumerate}
-
-\section*{Questions}
-\begin{itemize}
-\item Are there any problematic counterfactuals being used in defining bandit problems
-\item in the Schölkopf causal fairness paper - is there an issue with the use of twin networks. Could it be fixed with SWIGs?
-\item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
-\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomized experimental design?
-\item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
-\item What does conditioning mean in the Bayesian sense?
-- can imposing a prior which is true with respect to that variable, increase bias due to confounding?
-(In the same way that in the pnemonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.
-
-\item What are the nessessary and sufficient conditioins for decision making?
-How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function).
-Ie when is the mean sufficient?
-What about the mean and the variance?
-- when can we take one set of results and compute something about a different loss function?
-
-\item Does the way we define supervised learning and loss fuctions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?
-
-\end{itemize}
-
-\chapter{Introduction}
-
-\section{Motivation}
-Many of the most important questions in science and in our personal lives are about the outcomes of doing something.
-Will asking people to pay upfront at the doctors reduce long term health expenditure?
-If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.
-
-These are hard questions because they require more than identifying a pattern in data.
-Correlation is not causation.
-Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.
-
-We now live in a world of data.
-Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth.
-Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose.
-It does not come from the controlled world of a randomised experiment.
-
-The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
-From health care, to entertainment and self driving cars; machine learning algorithms will transform many industries.
-It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory \citep{anderson2008end} and that we may be only a short step from \quotes{The Singularity}, where artificial intelligence exceeds our own and then grows exponentially.
-
-However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within narrow problem settings.
-Getting them to generalise to even slightly different problems or data sets remains very challenging.
-Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it.
-The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on.
-As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
-
-\section{What is causality? }
-
-The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined.
-This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
-However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression).
-Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via \quotes{common sense} the implications of the reported associations.
-
-There are two ways in which an association detected in a dataset may be non-causal.
-The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample.
-Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated.
-For example, based on data from the years 1999 to 2009, the age of miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}.
-We would not expect this relationship to hold if we were to look at a new sample of data.
-This form of spurious correlation also has serious repercussions.
-It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published.
-As a consequnce, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
-This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}.
-This issue can be ameliorated by getting more data and by seperating learning models from evaluating their performance, for example by evaluating models on a stict hold-out set or on the extent to which their results can be replicated.
-
-However, a strong association, observed in multiple indpendent studies may still not be causal.
-The correlation can arise because both variables are consequences of some other, unmeasured factor.
-For example, the reading ability of children under twelve is strongy correlated with their height, because older children are taller and can read better.
-However height is not a cause of reading ability because if we were to intervene to increase a child's height, for example by giving them growth hormones, we would not expect dramatic improvements in their reading.
-Similarly, extra tutoring in reading will not make a child grow taller.
-This problem is fundamentally different to the issue of spurious correlations arising by chance in finite datasets.
-Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to seperate causation from correlation.
-
-The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables.
-In this thesis, I take an interventionalist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal.
-This viewpoint captures the types of questions that motivate this thesis.
-How can we change the way we do things to obtain better outcomes.
-
-Causality is often liked to explaination; understanding how and why things happen.
-I view explaination in terms of compression and generalisabilty: how much information about the world a model can capture.
-This creates a hierachy in the degree to which models are explainitory, rather than a simple binary distinction.
-A standard predictive model encodes all the information we need to predict some output given inputs provided the system generating the data does not change.
-A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed.
-More detailed causal models could predict the outcome for a wide range of combiations of interventions conditional on a range of contexts.
-By considering conditional interventions within our definition of causal questions we also capture mediation: the study of through which pathways one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact to mathematical equatons like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
-
-\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs.
-The former aims to identify the effect of a known cause.
-The latter can be viewed as identiying causes of an effect.
-They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}.
-However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined.
-The goal of identifying the causes of an effect can be formalised within the graphical causal model framework.
-Solving this problem is certainly much more challanging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables.
-These practical difficulties may well be overwealming, particularly in fields such as social science and economics where datasets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively.
-However, this does not mean that the problem of identifying causes of effects is ill-posed in principle.
-It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.
-
-There has traditionally been a large gap between researches in machine learning who focus on prediction, use largly non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models.
-However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of for the training and evaluation models.
-In the next section we emphasise the subtley that can exist in determining if a problem is causal by examiming some typical examples.
-
-\section{What makes a problem causal?}
-
-Machine learning is in the midst of a boom, driven by the availability of large datasets and the computation resources to process them.
-Machine learning techniques are being applied a huge range of problems, in both industry and achademia.
-The following examles are inteded to capture the bredth of problems that machine learning algorithms are actively beeing applied to.
-Which, if any, of these problems require causal inference?
-
-\begin{itemize}
-\item Speech recognition (for systems like Siri or Google)
-\item Image classification
-\item Forecasting the weather
-\item Identifying spam emails
-\item Automated essay marking
-\item Predicting the risk of death in patients with pneumonia.
-\item Predicting who will re-offend on release from prison 
-\item Customer churn modelling
-\item Demand prediction for inventory control
-\item Playing Go 
-\end{itemize}
-
-The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important considertion.
-In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.
-
-Consider speech recognition.
-You say something, which causes sound waves, which are converted converted to a digital signal that Siri maps to words.
-Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution).
-A similar argument could be made for many applications of machine translation and image classification.
-
-In image classification we do not particularly care about building a strong model for exactly how the thing that was photographed translates to an array of pixels, provided we can be fairly confident that the process will not change.
-If we develop a discriminative model that is highly accurate at classifying cats from dogs, we do not need to understand its internal workings (assuming we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
-
-What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella  causality can be ignored - your decision will not effect if it actually rains.
-However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
-For this we need a (causal) model that allows us to predict the outcome under various different interventions.
-
-Identifying spam and automated essay marking both involve  processing text to determine an underlying (complex) attribute such as its topic or quality.
-In both cases, there is inherent competition between the algorithm and the people generating the text.
-As a result, decisions made by the algorithm are likely to change the relationship between the features it relies on and the true label.
-Spammers and students will modify their writing in order to optimise their results.
-A standard supervised learning approach can only work if the resulting change in the mapping from features to label is sufficiently gradual.
-There are two key ways of ensuring this.
-The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm.
-The second is to use a model in which the features are related to the outcome in such a way that they cannot be manipulated independently.
-
-This example also highlights a connection between causal models and transparancy in machine learning.
-If we are using a non-causal model to make decision effecting people, there will be a trade-off between the performance and transparancy of the model; not because the requirement for transparancy restricts us to simple models but because revealing how the model works allows people to change their behaviour to game it.
-
-What about predicting the risk of death in patients with pneumonia?
-Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective for serious cases, this seems like a straightforward prediction problem.
-It is not.
-Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admition decisons.
-\citet{Caruana2015} found exactly this effect in a real data set.
-The model learnt that people suffering asthma were \emph{less} likely to die from pneumonia.
-They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
-The issue is not with model, it performed very well at the task for which it was trained, which is to predict wo would be likely to die under the original admition and treatment protocols.
-However, using it to decide how to \emph{change} these protocols could kill.
-The actual question of interest in this case is what happens to patients with charachteristics $X$ if we assign them treatment according to decision rule $Z$.
-
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
-Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.
-
-Demand prediction seems like a relatively straightforward prediction problem.
-These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
-It seems unlikely that using the model to ensure stock is available will itself change demand.
-However, depending on the way demand is measured, there is a potential data censoring issue.
-If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
-Changing availability does then change demand.
-
-Playing Go (and other games) is a case with some subtleties.
-At every turn, the AI agent has a number of actions available.
-The board state following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem.
-There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}.
-Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$.
-This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players.
-Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a compbination of these approaches \citep{silver2016mastering}.
-The supervised learning was enhanced by having the agent play (varients) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
-
-The problem of playing go is causal from the interventionalist perspective.
-The agent wishes to learn the probability of a win given an action they take.
-However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach.
-The actions the agent has to explore are the sames ones as human players explored to generate the trainging data, and both have the same objective - to win the game.
-In addition, the state of the board encapsulates all the information relevent to selecting a move.
-These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.
-
-Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference.
-A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes.
-The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people.
-Think about problems like credit scoring and parole decisons.
-There are strong ethical grounds for demanding transparancy, but if the goals of society and the individuals are not perfectly aligned and there is any posibility that people can manipulate features independtly of the outcome, there will be a conflict between model accuracy and transparancy.
-It is rare to build a model without any intent to make some kind of decion on the basis of its results.
-Thus, I argue we should assume a causal model is required until we can justify otherwise.
-
-\section{An overview of this thesis and its contributions}
-
--------------------
-
-In chapter \ref{chap:interventionalist_view} I describe the interventionalist viewpoint on learning the outcome of actions.
-
-I review the key problems and results within the bandit liturature
-
-highlight some interesting connections between existing strands of work on bandits and causal inference.
-
--------------------
-
-Causal questions are everywhere.
-Statistics, economics, social science, artificial intelligence and machine learning.
-Techniques and for addressing such problems have developed in parralel within many diciplines.
-These techniques can usefully be categorised into two broad approaches, reinforcement learning and observational causal inference.
-
-In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them.
-We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal.
-A particular strand of research within reinforcement learning are multi-armed bandit problems.
-They describe settings in which there are a set of available actions, the agent repeately decides which to select and then observes the outcome of the chosen action.
-They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.
-
-The approach of learning the outcome of an action by taking it has plays a key role in advancing our knoweledge of the world.
-However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen.
-Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.
-
-Observational causal inference can be viewed as a form of transfer learning.
-The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the orignial system.
-This is impossible without some assumptions about how the the orignal system and the system after intervention are related to one-another.
-The key to observational inference is a framework that models how actions change the state of the world and thus allows us to map information collected in one setting to another.
-
-Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials.
-Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem randomisation is not always feasible, affordable or ethical.
-The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
-
-The goals of this thesis are: to clarify the distinction between prediction and causal inference and highlight when and how we can leverage advances in prediction to improve causal inference and to draw together the key approaches to solving causal problems, from both the interventionalist and observational viewpoints.
-In chapter two, I describe the existing frameworks for formalising causality developed withing economics, statistics and machine learning and discuss how they relate to one another.
-Chapter three reviews the key approaches to causal inference from observational data.
-Chapter four highlights the interventionalist viewpoint, covering traditional randomised experiments and the closely related multi-armed bandit problems and highlights connections between the algorithms applied to bandit problems and those for estimating causal effects from observational data.
-
-In chapter five, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
-The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
-I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
-
-Chapters two to four predominantly review the existing liturature with the goal of making connections between the different viewpoints and approaches.
-They do not contain novel technical results but present the material in a unified way.
-Chapter five is a novel contribution.
-It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit liturature and describes an algorithm that can leverage this structure to make better decisons.
-
-\chapter{Causal inference from observational data}
-\label{chap:causality}
-
-\section{Causal models}
-\label{sec:causal_models}
-
-Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
-To do this we need terminology to describe actions and how we anticipate the system will respond to them.
-Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
-In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ.
-We will also use them to describe the following simplified examples.
-The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:ranomized_experiment}
-Suppose a pharmaceutical company wants to assess the effectiveness of a new drug on recovery from a given illness.
-This is typically tested by taking a large group of representative patients and randomly assigning half of them to a treatment group (who receive the drug) and the other half to a control group (who receive a placebo).
-The goal is to determine the clinical impacts of the drug by comparing the differences between the outcomes for the two groups (in this case, simplified to only two outcomes - recovery or non-recovery).
-We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
-\end{example}
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:adjusting}
-Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds.
-We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
-We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
-\end{example}
-
-\section{Causal Bayesian networks}
-
-Causal Bayesian networks are an extension of Bayesian networks.
-A Bayesian network is a graphical way of representing how a distribution factorises.
-Any joint probability distribution can be factorised into a product of conditional probabilities.
-There are multiple valid factorisations, corresponding to permutations of variable ordering.
-
-\eqn{
-\label{eqn:cbn:joint_dist}
-P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
-}
-
-We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}.
-If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network.
-There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below left=of 1](2){$X_{2}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\caption{A general Bayesian network for the joint distribution over three variables.
-This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
-\label{fig:bayesnet}
-\end{figure}
-
-\begin{figure}
-\centering
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (3) edge (2)
-    (2) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, below left=of 1](2){$X_{1}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, above left=of 1](2){$X_{1}$};
-\node[main node, above right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (2) edge (1)
-    (3) edge (1)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$.
-Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution.
-Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
-\label{fig:bayesnet2}
-\end{figure}
-
-The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true: $G$ can have extra edges).
-If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
-
-\eqn{
-\P{X_{1},...,X_{N}} = \prod_{i = 1...N}\P{X_{i}|parents_{X_{i}}}
-}
-
-A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
-This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-We need some notation to describe interventions and represent distributions over variables in the network after an intervention.
-In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.
-
-\vspace{0.5cm}
-\begin{definition}{The do-notation}
-\begin{itemize}
-\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
-\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$.
-This notation is somewhat overloaded.
-It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
-For example it could represent
-\begin{itemize}
-\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
-\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
-\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
-\item a family of density/mass function for $Y$ parameterised by $x$.
-\end{itemize}
-Where the distinction is important and not clear from context we will use one of the more specific forms above. 
-\end{itemize}
-\end{definition}
-
-\vspace{0.5cm}
-\begin{theorem}[Truncated product formula \citep{Pearl2000}]
-\label{thm:truncated_prodcut}
-If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.
-Let $\parents{X_i}$ denote the parents of the variable $X_i$ in $G$.
-
-\begin{equation}
-\label{eq:truncatedproduct}
-\P{X_1...X_N|do(Z=z)} = \ind{Z = z}
-  \prod_{X_i \notin Z}\P{X_{i}|\parents{X_i}}
-\end{equation}
-\end{theorem}
-
-Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution.
-The truncated product formula will give different results depending on the selected network.
-The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
-In fact, from the interventionist viewpoint of causality, the truncated product formula defines what it means for a link to be causal.
-
-Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$?
-The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient.
-Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables.
-The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}.
-This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$.
-In this example, the interventional distribution is equivalent to the observational one.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X \text{ (Treatment)}$};
-\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
-\path[]
-	(2) edge (3);
-\end{tikzpicture}
-\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
-\label{fig:causal_network_example}
-\end{figure}
-
-In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool.
-We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$.
-It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation.
-If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}.
-In this case, the interventional distribution is not equivalent to the observational one.
-If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool.
-To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,
-
-\eqn{
-\label{eqn:backdoor_example}
-\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}
-}
-
-We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph.
-There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure.
-These rules are referred to as the Do-calculus \citep{Pearl2000}.
-We discuss them further in section \ref{sec:do_calculus}.
-
-\begin{figure}
-\center
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]
-
- %nodes
-\node[node](1){$Z \text{ (SES)}$};
-\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
-\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3); 
-\end{tikzpicture}
-\caption{Causal Bayesian network for example \ref{exm:adjusting}}
-\label{fig:causal_adjust} 
-\end{figure}
-
-Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes.
-The global Markov property and d-separation also hold in causal networks.
-D-speration, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
-We include a brief introduction in section \ref{sec:dsep}.
-
-\subsection{Can't everything be encoded in a Bayesian model?}
-Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
-The power of defining an intervention in this way stems from the number of things that are invariant between the two processes.
-All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
-
-We could represent problems of the type where we try to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and what they have in common, see figure \ref{fig:causal_ordinary_bayes}.
-This does not require any special framework or notation.
-The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks.
-However, without a causal framework, we have to make assumptions about what will be invariant to the intervention specifically for each such problem we encounter.
-For complex problems, it is very difficult to conceptualise the assumptions we expect to hold without the benefit of a causal framework.
-
-\begin{figure}[h]
-\center
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
-\tikzstyle{box}=[rectangle, draw=black!100]
- %nodes
-\node[node](1){$Z$};
-\node[node, below left=of 1](2){$X$};
-\node[node, below right=of 1](3){$Y$};
-\node[node, right=1.5cm of 3](4){$x$};
-\node[lt, above right=of 4](5){$Z'$};
-\node[lt, below right=of 5](6){$Y'$};
-\node[lt, above=2.0cm of 3](9){$\theta_{y}$};
-\node[lt, above=0.5cm of 9](7){$\theta_z$};
-\node[lt, above=2.0cm of 2](8){$\theta_x$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (6)
-    (5) edge (6)
-    (8) edge (2)
-    (7) edge (1) edge (5)
-    (9) edge (3) edge (6);
-\node[rectangle, inner sep=4.0mm, fit= (1) (2) (3),draw=black!100,label=below right:N] {};
-\node[rectangle, inner sep=4.0mm, fit= (5) (6),draw=black!100,label=below right:M] {};
-\end{tikzpicture} 
-\caption{Causal inference with ordinary Bayesian networks.
-The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$.
-The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
-As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
-However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
-\label{fig:causal_ordinary_bayes}
-\end{figure}
-
-A causal Bayesian network represents much more information than a Bayesian network with identical structure.
-A causal network encodes all possible interventions that could be specified with the do-notation.
-For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
-The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
-After each two variable intervention, we have a distribution over one variables which requires one parameter.
-This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
-
-\subsection{Limitations of causal Bayesian networks}
-A number of criticisms have been levelled at this approach to modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}.
-However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions.
-For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
-But, what if some people failed to take the prescribed treatment?
-We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
-Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Prescribed Treatment};
-\node[node, below right=of 1](2){Actual Treatment};
-\node[node, right=of 2](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(1) edge (2)
-	(2) edge (3) 
-	(4) edge (3) edge (2);
-\end{tikzpicture}
-\caption{Randomised treatment with imperfect compliance}
-\label{fig:randomized_imperfect_compliance}
-\end{figure}
-
-Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables.
-Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology.
-We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
-However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
-Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
-Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
-\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
-
-\section{Counterfactuals}
-\label{sec:counterfactuals}
-
-The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
-Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
-
-In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would have happened if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would have happened if they got the placebo, $\cf{y_{i}}{0}$.
-The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, there are units that can be assumed to be identical or that will revert exactly to their initial state some time after treatment.
-This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
-
-Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo.
-Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate (from an experimental or observational study):
-\begin{itemize}
-\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
-\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
-\end{itemize}
-
-Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
-For those who did not take the treatment, the observed outcome is the same as the outcome \emph{had} they not taken the treatment.
-Equivalently stated:
-
-\eq{
-\P{Y^{0}|X=0}&= \P{Y|X=0}\\
-\P{Y^{1}|X=1}&=\P{Y|X=1}
-}
-
-If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
-
-\eq{
-\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
-\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
-}
-
-This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
-
-\eq{
-\P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
-}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}.
-They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
-In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e.
-$X \nci \cf{Y}{0}$.
-Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e.
-$X \nci \cf{Y}{1}$.
-A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
-
-\vspace*{.3cm}
-\begin{theorem}[Ignorability \citep{Rosenbaum1983, Pearl2000}] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
-
-\eqn{
-\label{eqn:counterfactual1}
-\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
-\label{eqn:counterfactual2}
-\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
-}
-\end{theorem}
-
-Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
-Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
-
-Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}.
-This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
-As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
-The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
-SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
-
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-This makes it quite easy to (accidentally) make statements about counterfactuals that cannot be tested with empirical data.
-Consider the following example based on \citet{Dawid2000}.
-Again we have a drug where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
-Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal with equal variance (for simplicity).
-
-% vector with mean mu_t,mu_c and a covariance matrix
-\eqn{
-P(\cf{Y}{1},\cf{Y}{0}) \sim N( 
-         \begin{bmatrix}
-           \mu_{1} \\
-           \mu_2 \\
-         \end{bmatrix},
-         \begin{bmatrix}
-           \sigma^2 & \rho \sigma^2 \\
-           \rho \sigma^2 & \sigma^2 \\
-         \end{bmatrix}
-         )
-}
-
-The distribution of \emph{individual treatment effects}, $\tau = \cf{Y}{1} - \cf{Y}{0}$ is also normal,
-
-\eqn{
-P(\tau) = N(\mu_1 - \mu_0,2\sigma^{2}(1-\rho))
-}
-
-\begin{figure}[h]
-\captionsetup[subfigure]{position=b}
-\centering
-\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$.
-The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated.
-\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
-\hspace{0.02\textwidth}
-\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
-{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
-\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
-\label{fig:metaphysical_distribution_difference}
-\end{figure}
-
-From a (large) randomised controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}.
-These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated respectively.
-However, the distribution over the individual causal effects depends on $\rho$, see figure \ref{fig:metaphys_b}.
-The key problem is that we can never observe the joint distribution over $\cf{Y}{1}$ and $\cf{Y}{0}$.
-As a result, $\rho$ and thus the variance of $\tau$ is not identifiable, even from experimental data.
-\citet{Dawid2000} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects.
-He further points out that the interventional distributions in figure \ref{fig:metaphys_a}, along with a loss function, contain all the information required to decide how to treat a new patient.
-
-This result is unintuitive.
-It seems on the face of it that the distribution of individual causal effects is relevant to our decision making.
-If $\rho = 1$ then almost everyone benefits slightly from the treatment whilst if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.
-This confusion can be resolved by thinking about personalised rather than individual causal effects.
-It is entirely possible that potentially observable characteristics (such as gender, age, genetics, etc) affect how people will respond to the treatment.
-We can partition people into sub-populations on the basis of these characteristics and measure different \emph{personalised} causal effects for each group.
-The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much can be gained from further personalisation.
-The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).
-
-One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
-This is certainly convenient, if rather imprecise.
-However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention.
-It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).
-
-\section{Structural Equation models}
-\label{sec:SEM}
-
-Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Uncertainties are not inherent but arise from unmeasured variables.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \citep{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(\epsilon_x)$};
-\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:ranomized_experiment}}
-\label{fig:sem_randomized_treatment}
-\end{figure}
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random.
-The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
-The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
-
-Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$.
-The function $f_y$ and distribution over $\epsilon_y$ does not change.
-This results in the interventional distribution, \footnote{We have assumed the variables are discrete only for notational convenience}
-
-\eqn {
-\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
-}
-
-The observational distribution of $Y$ given $X$ is,
-
-\eqn{
-\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
-& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
-}
-
-The interventional distribution is the same as the observational one.
-The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
-
-The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
-Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
-
-\eqn{
-\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
-\label{eqn:sem:adjusting}
-&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
-}
-Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivalent to the observational distribution given by:
-
-\eqn{
-\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
-}
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(Z,\epsilon_x)$};
-\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
-\node[lt, above=of 6](7){$\epsilon_z$};
-\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2)
-	(7) edge (6)
-	(6) edge (2) edge (3);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:adjusting}}
-\label{fig:sem:preschool}
-\end{figure}
-
-Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
-A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
-
-\section{Comparing and unifying the models}
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
-In other words, the network is a causal Bayesian network.
-All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
-Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}).
-Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
-We can also connect this to counterfactuals via:
-
-\begin{equation}
-\begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
-\end{aligned}
-\end{equation}
-
-The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
-
-\begin{equation}
-\label{eq:weakignore}
-X \ci Y^{0} \text{ and } X \ci Y^{1}
-\end{equation}
-
-It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}.
-The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
-For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores \citep{Rosenbaum1983} to estimate the effect.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
-These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
-
-In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
-The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field.
-Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
-In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
-
-\section{What does a causal model give us? Resolving Simpson's paradox}
-
-We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
-The following example is adapted from \citet{Pearl2000}.
-Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
-She keeps track of which medication her patients choose and whether or not the treatment is successful.
-She obtains the results in table \ref{tab:simpson1}.
-
-\begin{table}[h]
-\caption{Treatment results}
-\label{tab:simpson1}
-\center
-\begin{tabular}{c c c c c}
-Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9} 
-A &  87 & 13 & 100 & 87\%\\
-\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Drug A appears to perform better.
-However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
-
-\begin{table}[h]
-\caption{Treatment results by gender}
-\label{tab:simpson2}
-\center
-\begin{tabular}{c c c c c c}
-Gender & Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9}
-M & A & 12 & 8 & 20 & 60\%\\
-M & B & 56 & 24 & 80 & 70\%\\
-\rowcolor[gray]{0.9}
-F & A & 75 & 5 & 80 & 94\%\\
-F & B & 19 & 1 & 20 & 95\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Once the data is broken down by gender, drug B looks better for both men \emph{and} women.
-Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must recommend which to subsidise under a national health scheme).
-Should she choose A or B?
-The ambiguity in this question lies at the heart of Simpson's paradox.
-How does causal modelling resolve the paradox?
-The key is that the doctor is trying to choose between \emph{interventions}.
-She wants to know what the success rate will be if she changes her practice to give all the patients one drug, rather than allowing them to choose as currently occurs.
-
-Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
-The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
-Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
-Some assumptions about the causal relationships between the variables are required.
-In this example, it seems reasonable to conclude that gender may affect the treatment chosen and the outcome.
-Assuming there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Gender};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (2) edge (3) 
-	(2) edge (3);
-\end{tikzpicture}
-\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
-\label{fig:simpsons_paradox}
-\end{figure}
-
-With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
-Knowing a patient took drug A indicates they are more likely to be female.
-When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
-However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
-So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
-Drug B is the better choice.
-
-\eqn{
-\P{Y|do(T)} = \P{Y|T,female}\P{female}+ \P{Y|T,male}\P{male}
-}
-
-Is the solution to Simpson's paradox to always to break down the data by as many variables as possible?
-No.
-Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
-This is a drug designed to prevent heart disease.
-One pathway to doing so might well be to lower blood pressure.
-Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
-It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
-Now, however, table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood pressure control and better overall outcomes.
-In this setting $\P{Y|do(T)} = \P{Y|T}$ and drug A is the better choice.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Blood Pressure};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (3) 
-	(2) edge (3)
-	(2) edge (1);
-\end{tikzpicture}
-\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
-\label{fig:simpsons_paradox2}
-\end{figure}
-
-Note that we have not changed the data itself, only the description of the variables that it is associated with.
-This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
-From a purely statistical viewpoint there is no paradox.
-The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
-The paradox only arises when we attempt to use the data to select an intervention and is resolved when we apply a causal approach to do so.
-
-\eqn{
-\label{eqn:simpsons_paradox}
-\exists \set{N_1,...N_4, N_1'...N_4'} \in \mathbb{N} : \; 
-\textcolor{green}{\frac{N_1}{N_2}< \frac{N_1'}{N_2'}}, \;\; 
-\textcolor{blue}{ \frac{N_3}{N_4}<\frac{N_3'}{N_4'}}
- \text{ and } \frac{N_1+N_3}{N_2+N_4} > \frac{N_1'+N_3'}{N_2'+N_4'}
-}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.4]{figures/simpson_vector}
-\caption{Simpson's reversal visualised.
-The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
-However, when we sum them, the ratio is steeper for the un-primed variables.}
-\label{fig:simpsons_vector}
-\end{figure}
-
-There are many other plausible causal graphs for both scenarios above.
-Perhaps income affects drug choice as well as gender, or gender might affect treatment choice and blood pressure control given treatment, etc.
-Causal modelling provides a powerful tool to specify such assumptions and to determine how to estimate causal effects for a given model see section \ref{sec:causal effect estimation}.
-
-\section{Causal effect estimation} 
-\label{sec:causal effect estimation}
-We can roughly categorise the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
-In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
-The goal is then to estimate the effect of an intervention or range of interventions in the system.
-Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}.
-Almost every time someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
-
-Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
-This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
-Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
-An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
-\todo{note that I focus on problems where causal structure known, reference rewiew on structure learning}
-
-Estimating causal effects from observational data comes down to determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
-We did this on an ad-hoc basis to resolve the examples discussed in chapter \ref{chap:causal_models}.
-In this chapter we describe a principled approach to mapping observational quantities to interventional ones and discuss some of the key issues involved in estimating such expressions from finite sample data.
-We assume the basic structure of the graph is known.
-That is, we assume that we can draw a network containing (at a minimum):
-
-\begin{itemize}
-\item the target/outcome variable we care about,
-\item the focus/treatment variables on which we are considering interventions,
-\item any variables which act to confound two or more of the other variables we have included, and
-\item any links between variables we have included. 
-\end{itemize}
-
-Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
-For example, consider estimating the impact of schooling on wages.
-Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
-Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
-
-How can the structure of the causal graph be leveraged to compute interventional distributions from observational ones?
-Given the graph corresponding to the observational distribution, the graph after any intervention can be obtained by removing any links into variables directly set by the intervention.
-The joint interventional distribution is the product of the factors associated with the interventional graph, as given by the truncated product formula \ref{eq:truncatedproduct}.
-If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
-However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
-
-The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
-By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \citep{Pearl2000}.
-The do calculus consists of three rules.
-They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.
-
-\subsection{Independence in Bayesian networks: D-separation}
-\label{sec:dsep}
-
-Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
-Therefor, in this section, we briefly review the key properties of Bayesian networks.
-A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
-Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
-
-\vspace{.3cm}
-\begin{theorem}
-\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
-\end{theorem}
-
-The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$.
-D-separation is an algorithm that extends the local Markov property to find these additional independencies.
-It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
-
-The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
-From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
-Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
-In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
-In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
-
-The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
-The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
-As an example, consider a scholarship available to female or disadvantaged students.
-Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship.
-There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
-However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y| Z$}
-\label{fig:dseptripple1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$X$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Y$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$X$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, below left =of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-        edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|\{\}$}
-\label{fig:dseptripple4}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, above left =of 1](2){$X$};
-\node[main node, above right =of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (1)
-    (3) edge node {} (1);	
-\end{tikzpicture}
-\end{subfigure}
-\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
-\label{fig:dseptriples}
-\end{figure}
-
-\vspace{.3cm}
-\begin{definition}[unblocked path]
-\label{def:unblocked}
-A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
-It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
-\end{definition}
-
-\vspace{.3cm}
-\begin{definition}[d-separation]
-\label{def:dsep}
-The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, there are no unblocked paths from any $X \in \boldsymbol{X}$ to any $Y \in \boldsymbol{Y}$ after conditioning on $\boldsymbol{Z}$.
-\end{definition}
-
-\vspace{.3cm}
-\begin{theorem}[d-separation and conditional independence]
-\label{thrm:dsep_independence}
-If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in a Bayesian network $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
-Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected (not d-separated) given $\boldsymbol{Z}$ then it is possible to construct a distribution $P'$ that factorises over $G$ in which they are dependent.
-\end{theorem}
-
-Theorem \ref{thrm:dsep_independence} says that independencies implied by d-separation on a graph hold in every distribution that can be factored over that graph and that if ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in \emph{all} distributions that can be factored over $G$ then they are d-separated in $G$.
-If we denote the independencies implied by d-separation in a graph by $\indep{G}$ and the set of independencies in a distribution by $\indep{P}$ then $\indep{G} \subseteq \indep{P}$.
-
-If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
-However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
-A particular case in which this occurs is when there are deterministic relationships between variables.
-If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
-This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
-This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
-D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
-
-\section{The Do Calculus} 
-\label{sec:do_calculus}
-
-The do-calculus is a set of three rules \citep{Pearl1995} that can be applied to simplify the expression for an interventional distribution.
-If by repeated application of the do-calculus, along with standard probability transformations, we can obtain an expression containing only observational quantities then we can use it to estimate the interventional distribution from observational data.
-Let $\boldsymbol{X}$,$\boldsymbol{Y}$,$\boldsymbol{Z}$ and $\boldsymbol{W}$ be disjoint sets of variables in a causal graph $G$.
-We denote the graph $G$ after the an intervention $do(\boldsymbol{X})$, which has the effect of removing all edges into variables in $\boldsymbol{X}$, as $G_{\overline{\boldsymbol{X}}}$
-
-\subsubsection{Rule 1: (adding or removing evidence)}
-
-Rule 1 allows us to remove (or insert) observational evidence from the right hand side of a conditional interventional distribution.
-It follows directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still applies after an intervention.
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:mutilatedOriginal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2);
-
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.5\textwidth}
-\centering
-\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
-\label{fig:mutilatedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{subfigure}
-\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
-\label{fig:mutilatednet}
-\end{figure}
-
-\subsubsection{Rule 2: (exchanging actions with observations)}
-Rule 2 describes when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening, $do(\boldsymbol{X}=\boldsymbol{x})$, have the same effect on the distribution over $\boldsymbol{Y}$.
-Let $G_{\underline{\boldsymbol{X}}}$ denote the causal graph $G$ with all edges \emph{leaving} $X$ removed.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W})$ in $G_{\underline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W})
-}
-
-The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$.
-Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
-However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
-Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
-If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths.
-This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).
-
-Equation \ref{eq:Do22} does not cover cases where acting on one set of variables allows us to replace acting on another set with conditioning (see figure \ref{fig:rule2_2}).
-The general form of rule 2 is given in equation \ref{eq:Do2}.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W},\boldsymbol{Z})$ in $G_{\underline{\boldsymbol{X}}\overline{\boldsymbol{Z}}}$:
-\eqn{
-\label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),
-do(\boldsymbol{Z}=\boldsymbol{z}),
-\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W})
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule2_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)  edge  (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule2_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1)	edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}}$}
-\label{fig:rule2_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    	edge  (3);
-
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
-In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
-If we condition on $W$, we block this indirect path.}
-\label{fig:rule2}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule22_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (1) edge (2)
-    (5) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{after $do(Z)$, $G_{\overline{Z}}$}
-\label{fig:rule22_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}\overline{Z}}$}
-\label{fig:rule22_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of applying equation \ref{eq:Do2}.
-In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
-Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
-Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
-}
-\label{fig:rule2_2}
-\end{figure}
-
-\subsubsection{Rule 3: (adding or removing actions)}
-This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-A simple case of rule 3 is given in equation \ref{eq:Do3_1}.
-If $\boldsymbol{Y}$ is independent of $\boldsymbol{{X}}$ in $G$ after removing links entering $\boldsymbol{X}$ then can be no direct path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and any intervention on $\boldsymbol{X}$ will not affect $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{{X}})$ in $G_{\boldsymbol{\overline{X}}}$:
-
-\eqn{
-\label{eq:Do3_1}
-\P{\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})} = \P{\boldsymbol{Y}}
-}
-
-The general case of rule 3 is easier to state by explicitly representing the intervention in the graphical model.
-Let $G^{\boldsymbol{\hat{X}}}$ denote the graph $G$ after adding a variable $\hat{X_i}$ as a parent of each variable $X_i \in \boldsymbol{X}$ (see figure \ref{fig:rule3_augmented}).
-The variable $\hat{X_i}$ can be thought of as representing the mechanism by which $X_i$ takes its value, either by being set via intervention or as a stochastic function of its other parents \citep{Koller2009}.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\boldsymbol{\hat{X}}}_{\boldsymbol{\overline{Z}}}$:
-\eqn{
-\label{eq:Do3}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-}
-
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
-These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule3_original}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{augmented graph $G^{\hat{X}}$}
-\label{fig:rule3_augmented}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
-\node[main node, above left=of 2](5){$\hat{X}$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3)
-    (5) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule3_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Example application of equation \ref{eq:Do3}.
-$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
-We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
-\label{fig:rule3}
-\end{figure}
-
-\subsection{Identifiability}
-A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
-This is the identifiablity problem.
-It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
-A query is non-parametrically identifiable if it is identifiable without  assumptions about the functional form of the dependencies between variables in the graph.
-
-\vspace*{.3cm}
-\begin{definition}[Non-parametric identifiablity]
-Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
-A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
-\end{definition}
-
-The question of non-parametric identifiability is solved.
-The do calculus is complete \citep{Shpitser2006b,Huang2006}.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
-There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
-
-\begin{enumerate}
-\item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network, and
-\item if it is identifiable, return the required expression. 
-\end{enumerate}
-
-Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
-I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$X$};
-\node[ob, right=of 1](2){$Z$};
-\node[ob, right=of 2](3){$Y$};
-\node[lt,above=of 2](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid2_instrumental_variable}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[ob, above left=of 2](4){$Z$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1)
-    (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
-\label{fig:identify}
-\end{figure}
-
-Many interesting questions relating to identifiability remain open.
-What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
-What if we assume various restrictions on the functional form of the relationships between the variables?
-Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
-A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
-
-Although identifiability is a natural and important question to ask, it does not partition causal questions into solvable and unsolvable.
-Estimators for identifiable queries can be slow to converge and we may be able to obtain useful bounds on causal effects in cases where point estimates are not identified.
-
-\section{Estimation}
-\subsection{Defining causal effects} 
-So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
-This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set.
-From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.
-
-\eqn{
-\label{eqn:utility_max}
-x* =\argmax_x \Esub{y\sim\P{Y|do(X=x)}}{{\cal{U}}(y)}
-}
-
-Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
-There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
-For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
-
-\eqn{
-\label{eqn:ACE}
-ACE = \E{Y|do(X = 1)} - \E{Y|do(X=0)}
-}
-
-Assuming the expectations in equation \ref{eqn:ACE} are well defined, the ACE captures the shift in the mean outcome that arises from varying $X$.
-It does not capture changes in variance or higher moments of the distribution.
-The ACE can be generalised to non-discrete interventions by considering the effect on the expectation of $Y$ of an infinitesimal change in $x$.
-If $X$ is linearly related to $Y$ then the ACE is constant and equivalent to the corresponding coefficient in the linear structural equation model.
-
-\eqn{
-\label{eqn:ACE_continuous}
-ACE(x) = \frac{d}{dx}\E{Y|do(X=x)}
-}
-
-The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}.
-Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
-However we can personalise the average causal effect by stating it with respect to some observed context.
-I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
-
-\eqn{
-PCE(z) = \E{Y|do(X=1),z} - \E{Y|do(X=0),z}
-}
-
-In some cases we may be interested in the average causal effect for some sub-group of the population.
-A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
-This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.
-
-\eqn{
-ATT = \Esub{z\sim \P{Z|x=1}}{Y|do(X = 1)} - \Esub{z\sim \P{Z|x=1}}{Y|do(X=0)}
-}
-
-Causal effects can also be written in terms of counterfactuals.
-The ACE is $\E{\cf{Y}{1} - \cf{Y}{0}}$.
-We could estimate the ratio of expectations $\frac{\E{\cf{Y}{1}}}{\E{\cf{Y}{0}}}$ instead of the difference.
-However, the quantity   $\E{\frac{\cf{Y}{1}}{\cf{Y}{0}}}$ depends on the joint distribution over the counterfactual variables $(\cf{Y}{1},\cf{Y}{0})$ and thus cannot be computed from the interventional distribution.
-
-Another way of conceptualising causal effects is as a property indicating the strength of the causal link between two variables.
-This notion is complex to formalise when the relationship between variables is non-linear.
-Suppose $Y = X \oplus Z$ with $P(Z=1)=\frac{1}{2}$, the interventional distributions over $X$ are identical after marginalising out $Z$.
-\citet{Janzing2013} propose a number of postulates that a notion of causal strength could satisfy, demonstrate why previous measures fail these postulates and propose an alternative based on information flow.
-
-\subsection{Estimating causal effects by adjusting for confounding variables}
-\label{sec:causal_effect_under_ignorability}
-Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors.
-Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds.
-Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds.
-This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}.
-The setting is also referred to as unconfounded.
-
-\vspace*{0.3cm}
-\begin{theorem}[The backdoor criterion]\citep{Pearl2000}
-Let $\boldsymbol{X}$, $\boldsymbol{Z}$ and $\boldsymbol{Y}$ be disjoint sets of vertices in a causal graph $G$.
-If $\boldsymbol{Z}$ blocks (see Definition \ref{def:unblocked}) for every path from $X_i$ to $Y_j$ that contains a link into $X_i$, for every pair $(X_i \in \boldsymbol{X},Y_j \in \boldsymbol{Y})$, and no node in $\boldsymbol{Z}$ is a decedent of a node in $\boldsymbol{X}$ then the backdoor criterion is satisfied and;
-
-\eqn{
-\label{eqn:backdoor}
-\P{\boldsymbol{y}|do(\boldsymbol{x})} =\sum_{\boldsymbol{z}} \P{\boldsymbol{y}|\boldsymbol{x},\boldsymbol{z}}\P{\boldsymbol{z}}
-}
-\end{theorem}
-
-The backdoor criterion derives from rule 2 of the do-calculus.
-Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
-There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
-The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
-Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}.
-Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
-For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
-
-\begin{figure}
-\centering
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{There can be multiple valid adjustment sets. $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.}
-\label{fig:adjust_multiple}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](2){$X$};
-\node[ob, right=of 2](3){$Y$};
-\node[ob,above=of 3](4){$Z_2$};
-\node[ob,above=of 2](5){$Z_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (5) edge (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{Conditioning on $Z$ opens the backdoor path $X-U_1-Z-U_2-Y$}
-\label{fig:m-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{}
-\label{fig:adjustment}
-\end{figure}
-
-\todo{fix this caption}
-
-Given that a set of variables $\boldsymbol{Z}$ satisfies the backdoor criterion (or equivalently the conditional ignorability assumption), the interventional distribution is asymptotically identifiable and can be estimated from equation \ref{eqn:backdoor}.
-The expected value of $Y$ after the intervention $do(X=x)$ is given by equation \ref{eqn:expected_post_intervention} and the average causal effect for a binary intervention $x \in \set{0,1}$ is given by equation \ref{eqn:ace_ignoreable}.
-
-\eqn{
-\label{eqn:expected_post_intervention}
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}}
-}
-
-\eqn{
-\label{eqn:ace_ignoreable}
-ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
-}
-
-Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata.
-However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
-There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
-The key approaches are based on matching on covariates, propensity score methods and regression.
-We now examine these approaches from a machine learning perceptive.
-
-In standard supervised learning, we have a training set $(\vb{x_1},y_1),...,(\vb{x_n},y_n)$ assumed to be sampled i.i.d from an unknown distribution $\P{\vb{x},y} =\P{\vb{x}}\P{y|\vb{x}}$.
-The goal is to select a hypothesis $h \in \mathcal{H} : \mathcal{X} \rightarrow \mathcal{Y}$ such that, on unseen data $ \sim \P{\vb{x},y}$, $h(\vb{x})$ is close (by some metric) in expectation to $y$.
-In other words we wish to minimise the generalisation error $E_{out}(h)$,
-
-\eqn{
-E_{out}(h) = \Esub{(\vb{x},y)\sim \P{\vb{x}}\P{y|\vb{x}}}{L(h(\vb{x}),y)}
-}
-
-We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
-We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
-
-\eqn{
-E_{in}(h) = \frac{1}{n}\sum_{i=1}^n L(h(\vb{x}_i),y_i)
-}
-
-The VC-dimension of the hypothesis space provides (typically loose) bounds on the probability that $E_{out} >> E_{in}$.
-However, in practice, the generalisation error is usually estimated empirically from a hold-out set of the sample that was not used to train the model, or via cross-validation.
-
-In the causal effect estimation under ignorability, we have training data $(\vb{x}_1,\vb{z}_1,y_1),...,(\vb{x}_n,\vb{z}_n,y_n)$ sampled i.i.d from $\P{\vb{z}}\P{\vb{x}|\vb{z}}\P{y|\vb{x},\vb{z}}$.
-Estimating $\E{Y|do(\vb{X}=\vb{x})}$ corresponds to selecting a hypothesis $h \in \mathcal{H}:\mathcal{X} \times \mathcal{Z} \rightarrow \mathcal{Y}$ that minimises;
-
-\eqn{
-\label{eqn:causal_covariate_shift}
-E_{out} &= \Esub{(\vb{x},\vb{z},y)\sim \delta(\vb{x}-\vb{x'})\P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},\\
-&= \Esub{(\vb{z},y)\sim \P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},
-}
-
-\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
-If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
-\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
-Take the case where we have a binary intervention $x \in \set{0,1}$.
-Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$.
-More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.
-
-\subsubsection{Regression}
-
-The regression approach is to learn a function that is a good approximation to the output surface $\E{Y|X,Z}$.
-Let $f_1(z)= \E{Y|X=1,Z=z}$.
-The expectation of $Y$ after the intervention $X=1$ is then obtained by taking the expectation with respect to $Z$, $\E{Y|do(X=1)} = \Esub{z \sim \P{Z}}{\E{Y|X=1,z}}$.
-We can learn a parametric regression model $\hat{f}_1(z)$ via empirical risk minimisation.
-
-\eqn{
-\hat{f}_1(z) = h_1(z;\hat{\theta}_{obs}) \text{, where } \hat{\theta}_{obs} = \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)  \right]
-}
-
-This estimator is consistent with respect to the observational distribution.
-As the sample size tends to infinity, $\hat{\theta}_{obs}$ approaches the parameter within the hypothesis space that minimises the expected loss given data sampled from the observational distribution.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z|x=1}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-If the model is correctly specified such that $f_1(z) = h_1(z;\theta^*)$ for some  $\theta^* \in \Theta$ then the empirical risk minimisation estimate is consistent with respect to the loss over any distribution of $Z$  \citep{Sugiyama2007}, including the interventional one.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \theta^* = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-The average causal effect can then be estimated by,
-
-\eqn{
-\label{eqn:cef_empirical_regression}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left(\hat{f}_1(z_i) - \hat{f}_0(z_i)\right)
-}
-
-Regression thus has a direct causal interpretation if the parametric model is correctly specified and the covariates included form a valid backdoor adjustment set for the treatment variable of interest in the corresponding structural equation model.
-
-\subsubsection{Propensity scores}
-If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
-The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
-\caption{
-Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
-In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
-If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$.
-We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points).
-In both cases the linear model is a good fit to the data.
-However, the resulting estimate of the causal effect is very poor for the lower values of $z$.
-}
-\label{fig:covariate_shift_adjust}
-\end{figure}
-
-A general approach to estimating the expectation of some function $f(\cdot)$ with respect to data from some distribution $\P{\cdot}$, when we have data sampled from a different distribution $\Q{\cdot}$ is importance sampling \citep{horvitz1952generalization,Koller2009}.
-\eqn{
-\label{eqn:importance_sampling}
-\Esub{\vb{v} \sim \P{\vb{v}}}{f(v)} = \Esub{\vb{v} \sim \Q{\vb{v}}}{f(v)\frac{\P{\vb{v}}}{\Q{\vb{v}}}}
-}
-
-This importance weighting approach can be applied to the covariate shift/average causal effect problem by weighting the terms in the empirical risk minimisation estimator \citep{Sugiyama2007}.
-
-\eqn{
-\label{eqn:importance_weighted_erm}
-\hat{\theta}_{iw} &= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{\P{z_i}\delta(x_i-1)}{\P{z_i}\P{x_i=1|z_i}}  \right] \\
-&= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{1}{e(z_i)}  \right],
-}
-
-where $e(\vb{z})$ is the propensity score, defined by \citep{Rosenbaum1983};
-\eqn{
-e(\vb{z})\equiv \P{x=1|\vb{z}}
-}
-
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
-Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
-
-The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$.
-\citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$.
-This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$.
-Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
-In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
-However in general, it must be estimated from data.
-Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
-\citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
-
-\eqn{
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}} = \Esub{z\sim \P{\vb{Z|x}}}{\E{Y|x,\vb{z}}\frac{1}{e(\vb{z})}}
-}
-
-\eqn{
-\label{eqn:simple_inverse_propensity}
-\hat{\tau}_{ip} = \frac{1}{n}\sum_{i=1}^n \left(\frac{\ind{x_i=1}y_i}{e(\vb{z}_i)} -\frac{\ind{x_i=0}y_i}{1-e(\vb{z}_i)}\right)
-}
-
-\subsubsection{Matching}
-There is a straightforward connection between matching and regression for causal effect estimation.
-If $h \in \mathcal{H} \implies h + a \in \mathcal{H}$ for any constant $a$ and $\hat{f}$ is selected by minimising empirical risk with an $L_2$ loss then $\frac{1}{n}\sum_{i=1}^n \ind{x_i=1}\hat{f}_1(z_i) = \frac{1}{n}\sum_{i=1}^n \ind{x_i=1}y_i$ \footnote{\citep{Imbens2004} state this holds for most implementations}, and equation \ref{eqn:cef_empirical_regression} can be re-written as:
-
-\eqn{
-\label{eqn:cef_regression_missing}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \hat{f}_0(z_i)\right) +\ind{x_i=0}\left(\hat{f}_1(z_i) - y_i \right)\right]
-}
-
-This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
-For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
-Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
-\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
-
-\eqn{
-\hat{\tau}_{match} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \frac{1}{k}\sum_{j \in J_k(i)}y_j\right) +\ind{x_i=0}\left(\frac{1}{k}\sum_{j \in J_k(i)}y_j - y_i \right)\right]
-}
-
-This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
-There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
-Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
-An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.
-
-The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
-A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
-
-The significance of this should not be underestimated.
-Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
-With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
-\citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem.
-There is relatively little theory on model selection for estimating the propensity score.
-To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion.
-It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}.
-With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
-I have found no examples of such an approach.
-
-The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
-Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
-The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
-
-\chapter{Learning from interventions}
-\label{chap:interventionalist_view}
-
-The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it.
-There is an obvious alternative.
-Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens.
-There are three key differences between observing a system and explicitly intervening in it.
-Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from.
-Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}.
-Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias.
-Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning.
-For example, in a drug trial , assigning people a sub-optimal treatment has real world costs.
-This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain.
-This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.
-
-\section{Randomised experiments}
-Randomised controlled trials are often presented as the gold standard for determining causal effects.
-What is it about randomisation that makes it so important when it comes to causality?
-The graphical model for a randomised controlled experiment is shown in figure \ref{fig:random_experiment_network}.
-If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
-Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3);
-\end{tikzpicture}
-\caption{causal network for a randomised experiment}
-\label{fig:random_experiment_network}
-\end{figure}
-
-Randomisation does not ensure target and control group are exactly alike.
-The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables between the target and control groups in a finite data sample.
-However, the variance in the outcome, within both the target and control groups, also increases.
-The net result is increased variance (but not bias) in  the estimate of causal effects.
-
-Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest.
-If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can estimate the average causal effects directly from the differences in outcome across treatments.
-More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
-The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
-See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, above right=of 1](5){Covariates};
-\node[node, below right=of 5](3){Outcome};
-\node[lt, above=of 3](4){U};
-
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3)
-	(5) edge (3) edge (1);
-\end{tikzpicture}
-\caption{causal network for a stratified randomised experiment if the probability an individual is assigned a given treatment depends on some covariates.}
-\label{fig:random_experiment_network_stratified}
-\end{figure}
-
-The benefit provided by randomisation in breaking the link between the treatment variable and any latent confounders should not be understated.
-The possibility of unobserved confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (there is no test for confounding).
-This means causal estimates from non-experimental data are always subject to the criticism that an important confounder may have been overlooked or not properly adjusted for.
-However, randomised experiments do have some limitations.
-
-\subsection{Limitations of randomised experiments}
-\label{subsec:limitations_of_experiment}
-
-The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
-There may be imperfect compliance, the treatment selected by the randomiser is not always followed, or output censoring, the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.\footnote{Non-compliance is a problem if the goal is to estimate the causal effect of the treatment on the outcome but not if the goal is to estimate the causal effect of prescribing the treatment.
-The latter makes sense in a context where the process by which people decide whether or not to take the treatment they have been prescribed is likely to be the same if the treatment were made available more generally beyond experimental trial.} See figure \ref{fig:random_experiment_network_imperfect_compliance} for a the graphical model of a randomised experiment with imperfect compliance.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Actual Treatment};
-\node[node,above left=of 1](5){Assigned Treatment};
-\node[node, above =of 5](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (5)
-	(5) edge (1)
-	(1) edge (3)
-	(4) edge (3) edge (1);
-\end{tikzpicture}
-\caption{causal network for a randomised experiment with imperfect compliance}
-\label{fig:random_experiment_network_imperfect_compliance}
-\end{figure}
-
-It is not always possible or ethical to conduct a randomised controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
-When experimentation is possible, it is frequently difficult or expensive.
-This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
-In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest.
-The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
-\end{center}
-\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parachutes for the prevention of morbidity and mortality associated with falls from large heights.}
-\label{fig:parashute}
-\end{figure}
-
-Finally, non-adaptive randomised experiments are not optimal from either an active or reinforcement learning perspective.
-As an experiment proceeds information is obtained about the expectation and variance of each intervention (or treatment).
-Fixed experimental designs cannot make use of this information to select which intervention to try next.
-This results in both poorer estimates for a fixed number of experimental samples and more sub-optimal actions during the course of the experiment.
-
-\section{Multi armed bandits}
-\label{sec:multi-armed-bandits}
-
-Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed.
-Their introduction is generally attributed to \citet{Thompson1993}.
-In its classic formulation \citep{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
-Each arm $i$ is associated with a fixed but unknown reward distribution.
-\footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
-The goal of the learner is to maximise the total reward they receive.
-This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
-
-\vspace{0.5cm}
-\begin{definition}[Stochasitic k-armed bandit problem]
-Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action.
-The multi-armed bandit problem proceeds over $T$ rounds.
-In each round $t$,
-\begin{enumerate}
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
-\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
-\end{enumerate}
-
-At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T \cfb{Y}{a_t}$.
-We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
-Note we have used counterfactual notation (see ) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have recieved if it selected action $i$ at timestep $t$.
-\end{definition}
-
-The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
-To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
-This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
-There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.}.
-\eqn{
-\label{eqn:raw_regret}
-\rawregret = \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
-}
-
-Both the rewards and the actions selected by the algorithm are random variables.
-The majority of work in the bandit literature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}.
-The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is referred to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
-A stochastic bandit algorithm is learning if it obtains pseudo-regret that is sub-linear in $T$.
-
-\vspace{0.5cm}
-\begin{definition}[Pseudo-Regret]
-\eqn{
-\label{eqn:pseudo_regret}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}} \\
-&= n\mu_{i^*} - \E{\sum_{t=1}^T{\cfb{Y}{i}}}
-}
-\end{definition}
-
-The regret is invariant to adding a constant to the expected rewards for all actions.
-However, it still depends on key characteristics of the reward distributions for each action.
-Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range.
-Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
-
-\subsection{The counterfactual nature of regret}
-\todo{Fix this bit up}
-
-Note that we have used the conterfactual notation to represent the rewards given each action.
-The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep.
-The psuedo-regret (equation \ref{eqn:pseudo_regret}) remains unchanged as both it and the learner's actions depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
-However, the regret as defined in equation \ref{eqn:raw_regret} is no longer properly defined as it relies on both $\cfb{Y}{a_t}$ and $\cfb{Y}{i^*}$ being defined at each timestep.
-The variance (but not expectation) of $\rawregret$ depends on the covariance between counterfactual variables.
-
- The learner is only able to observe properties of the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
- The stochastic bandit problem can be defined without recourse to counterfactual variables by
-
- The algorithm and definition of regret depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
-
-\subsection{Stochastic bandits: Approaches and results}
-
-The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators.
-The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards.
-As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
-The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
-This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
-Instead, the focus is on computing bounds on the expected regret.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=1]{extreme_early_stopping}
-\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples.
-This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim Bernoulli(0.2)$.
-In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$.
-$\E{\hat{\mu}} = 0.439$.
-The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping.
-Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded.
-This has some interesting potential real world implications.
-Early stopping of clinical trials is controversial.
-A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
-\label{fig:early_stopping}
-\end{figure}
-
-There are a few key principles that are used to guide the development of bandit algorithms.
-The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
-A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
-This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.
-
-Another key approach is \emph{optimism in the face of uncertainty}.
-Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
-The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
-Thus selecting it yields either a good reward or useful information.
-
-\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family.
-They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient.
-However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm.
-\citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$.
-Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds.
-We now assume the rewards are bounded in $[0,1]$.
-The algorithm and regret bounds can be generalised to submission reward distributions, see \citet{Bubeck2012}.
-
-\begin{algorithm}[H]
-\caption{UCB-1}\label{alg:ucb-1}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} horizon $T$.
-\STATE Play each arm once. 
-\FOR{$t \in 1,\ldots,T$}
-\STATE Count the number of times each arm has been selected previously $n_{t,i} = \sum_{s=1}^{t-1} \ind{a_s = i}$
-\STATE Calculate the mean reward for each arm $\hat{\mu}_{t,i} = \frac{1}{n_{t,i}} \sum_{s=1}^{t-1}\ind{a_s = i}Y_t$
-\STATE Select arm $a_t \in \argmax_{i = \set{1,...,k}}\left(\hat{\mu}_{t,i}+\sqrt{\frac{2 \log{t}}{n_{t,i}}}\right)$
-\ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
-The problem dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
-
-\eqn {
-\label{eqn:ucb_problem_dependent_regret}
-\regret \leq & \sum_{i:\Delta_i > 0}\left(\frac{8\log(T)}{\Delta_i}+2 \right)
-}
-
-Somewhat unintuitively, the regret increases as the value of the arms gets closer together.
-This is because it becomes harder for the algorithm to identify the optimal arm.
-As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
-The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
-The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which is maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
-By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.
-
-\eqn{
-\label{eqn:ucb_worst_case_regret}
-\regret \in \bigo{\sqrt{kT\log(T)}}
-}
-
-\begin{figure}
-\includegraphics[scale=1]{ucb_regret_vs_delta}
-\caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
-The solid curve shows the mean (cumulative) regret, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
-The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
-\label{fig:ucb_regret_vs_delta}
-\end{figure}
-
-The form of the dependence on the number of arms $k$ and horizon $T$ differs between the problem dependent and worst case regret.
-The problem dependent regret grows linearly with the number of arms, $k$, and logarithmically with $T$.
-The difference stems from the fact the problem dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
-\citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by $\regret \in \bigomega{\sqrt{kT}}$
-
-Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
-This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \citep{Lattimore2015}.
-
-\begin{figure}
-\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
-\caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$.
-The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $bernoulli([.3,.7])$ over 1000 simulations.
-The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
-\label{fig:ucb_upperbound_vs_actual_performance}
-\end{figure}
-
-Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal.
-This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}.
-Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
-However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
-
-\subsection{Pure-exploration problems}
-Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
-In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
-The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.
-
-\vspace{0.5cm}
-\begin{definition}[Simple Regret]
-\eqn{
-\label{eqn:simple_regret}
-\simpleregret=\mu_{i^*} - \E{\mu_{\hat i^*}}.
-}
-\end{definition}
-
-The best arm identification problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialisation or exploitation phase.
-For example, many strategies might be assessed via simulation prior to one being selected and deployed.
-The worst case simple regret for a k-armed bandit is lower bounded by equation \ref{eqn:worst_case_simple_regret_lower_bound} (\citep{Bubeck2009}).
-
-\eqn{
-\label{eqn:worst_case_simple_regret_lower_bound}
-\simpleregret \in \bigo{\sqrt{K/T}}
-}
-
-%\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
-
-Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds. The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period.
-
-\subsection{Adversarial Bandits}
-
-Adversarial bandits, described by \citet{Auer1995}, are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
-Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$.
-As in the stochastic setting, the learner then receives reward only for the selected action.
-
-\vspace{0.5cm}
-\begin{definition}[Adversarial k-armed bandit problem]
-\label{def:adversarial-k-armed-bandit}
-Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
-\begin{enumerate}
-\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
-\end{enumerate} 
-\end{definition}
-
-Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
-In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
-
-For oblivious adversarial bandits we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
-We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
-
-\eqn{
-\label{eqn:pseudo_regret-adversarial}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
-}
-
-The policy (or algorithm) used by the learner is available to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
-This implies the adversary can ensure that any learner with a deterministic policy suffers regret $\bigo{T}$ by forecasting their entire sequence of actions.
-For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forecasts what action the learner will play in round 2, given they received a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
-This implies adversarial bandit policies must be sufficiently random to avoid such exploitation \footnote{The UCB algorithm, defined by algorithm \ref{alg:ucb-1}, is deterministic if the order in which arms are played during the first $k$ rounds is fixed and the method for selecting which arm to play when multiple-arms have the same upper-confidence bound is not-random (for example, select the arm one with the lowest index $i$).}
-
-The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case pseudo-regret of $\bigo{\sqrt{TK\log(T)}}$  \citep{Auer1995}.
-Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \citep{Audibert2009}.
-The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
-However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nonetheless obtain good problem dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
-
-Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, does not fully capture an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings where the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
-An example of a setting that lies between stochastic and adversarial bandit problems is the non-stationary setting, in which the rewards are generated stochastically from a distribution that varies over time.
-Adversarial bandit algorithms may perform better in such settings than standard stochastic policies to the extent that they explore more (to avoid the adversary simulating their behaviour) and thus adapt quicker to changes in the reward distribution.
-Adversarial algorithms also have stronger worst case regret guarantees, since even the weak regret for stochastic bandits is not guaranteed to be sublinear in such settings.
-However, if we have constraints on how rapidly or frequently the reward distributions can change over time it is better to use algorithms specificly developed to exploit such information (and compare them against a stronger notion of regret), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
-
-from the information available to the agent, the joint distribution over counterfactuals cannot be inferred, unless the assignment mechanism in noiseless - so it does not make sense to define regret that depends on it.
-
-\begin{figure}[h]
-    \begin{subfigure}[t]{0.49\textwidth}
-	\centering   
-	 \caption{Behaviour over time}\label{fig:non-stationary-bandit-time} 
-         \includegraphics[scale=.5]{non-stationary-bandit-time}
-\end{subfigure}
-    \begin{subfigure}[t]{0.49\textwidth}
-    \centering
-     \caption{Cumulative regret distribution}
-        \label{fig:non-stationary-bandit-regret}
-        \includegraphics[scale=.5]{non-stationary-bandit-regret}
-    \end{subfigure}
-      \caption{The pseudo-regret can be negative if rewards are non-stationary.
-This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed Bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
-Figure (a) shows the expected rewards of each arm, and the proportion of time that arm-1 is played, as a function of time.
-The single best-arm is arm-1 as it has the highest expected reward (averaged over $t$).
-An oracle that selects arm-1 in every round obtains an expected reward of $5,000$.
-However, despite not being designed to do so, the UCB-algorithm can adapt to the changing reward distribution to obtain consistently higher rewards.
-The distribution of regret over the 1000 simulations is shown in figure (b).}\label{fig:non-stationary-bandit}
-\end{figure}
-
-\subsection{Contextual bandits}
-\label{sec:contextual-bandits}
-
-In the standard multi-armed bandit setting, each decision is identical and the goal is to learn a single best action.
-However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
-The best treatment to offer an individual patient could depend on their age, gender, disease sub-type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
-Similarly, decisions on which ad or content to display on a webpage or which product to recommend can be \emph{personalised} based on the previous behaviour of the user.
-A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
-Contextual bandits are a generalisation of multi-armed bandits that make use of this additional contextual information.
-The term contextual bandit was coined by \citet{Langford2008}, however close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
-
-\vspace{0.5cm}
-\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analogously to definition \ref{def:adversarial-k-armed-bandit}}]
-\label{def:contextual-bandit-problem}
-Let $\P{\vec{x},\vec{y}}$ be the joint distribution over the rewards for each action and some context $\vec{X} \in \mathcal{X}$. In each round $t \in \set{1,...T}$, 
-\begin{enumerate}
-\item the world stochastically generates the vector of rewards for each action and the context, $(\vec{X}_t,[\cfb{Y}{1},...,\cfb{Y}{k}]) \sim \P{\vec{x},\vec{y}}$ and reveals $\vec{X}_t$ to the learner
-\item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps,
-\item the learner observes and receives (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
-\end{enumerate}
-\end{definition}
-
-Standard multi-armed bandits learn to select the action $a$ that, with high probability, maximises $\E{Y|a}$.
-Contextual bandits learn to select actions that maximise $\E{Y|\vec{x},a}$.
-The reward for contextual bandits should  be compared to an oracle that acts optimally based on the context.
-To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothesis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
-The pseudo-regret is the difference between the expected regret obtained by an oracle that selects actions based on the single best hypothesis or policy $h$ at each timestep, and the expected reward obtained by the algorithm.
-
-\eqn{
-\label{eqn:regret_contextual}
-\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{\cfb{Y}{h(\vec{X}_t)}}} - \E{\sum_{t=1}^T{\cfb{Y}{A_t}}}
-}
-
-If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a separate standard bandit instance for each value of the context.
-This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothesis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occurred such that $\sum_{c=1}^N T_c = T$.
-The regret is maximised if $T_c = T/N$ resulting in total regret $\bigo{N\sqrt{kT/N}}$.}.
-This is optimal with respect to this class of hypothesis.
-However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the in the relationships between different contexts and actions.
-As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
-For example, with $n$ binary variables, $N = 2^n$}.
-We expect some form of smoothness; values of context that are similar should lead to comparable rewards for a given action.
-We need algorithms that can leverage such assumptions.
-
-An alternate reduction to the standard bandit problem, which allows us to constrain the hypothesis space to explore, is to treat each hypothesis $h$ as a bandit arm \citep{Langford2008}.
-At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
-Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothesis class considered, limiting our ability to learn any complex mappings from context to actions.
-The key problem with this approach is each sample is used to update our knowledge about only one hypothesis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothesis simultaneously.
-
-Suppose that, at each timestep $t$, after selecting an action, the learner received the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
-This is known as the full information setting.
-In this case, the learner can simulate running each hypothesis over the history to compute the reward it would have obtained and use the hypothesis with the best empirical reward to select the next action.
-This is the \emph{follow the leader} algorithm, which obtains optimal regret $\bigo{\sqrt{T log(\card{\mathcal{H}})}}$ for the full-information problem \citep{cesa2006prediction}.
-Unfortunately, in the contextual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
-As in causal effect estimation, we can view this as a missing data problem.
-However, the data is missing not at random because which component of the reward is observed depends on the action selected which in turn is a function of the previous history of actions and rewards.
-
-The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly separating exploration from exploitation.
-Epoch-greedy is an explore-exploit algorithm.
-It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothesis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
-The hypothesis with the highest empirical reward is then used to select actions for the remaining time steps.
-The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
-
-The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
-However, it involves maintaining a list of weights for each hypothesis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
-The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
-
-Both Epoch-greedy and ILOVECONBANDITS involve solving problems of the form,
-
-\eqn{
-\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{X}_t) = A_t}
-}
-
-This expression equates to identifying the best empirical policy based on previous data.
-The algorithms assume the existence of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
-The computational tractability of these algorithms on large (or infinite) hypothesis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classification problem \citep{Dudik2011a}.
-
-Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalised versions of the UCB or Thompson sampling algorithms.
-For linear pay-off models, both approaches yield algorithms with strong regret guarantees, \emph{Lin-UCB} \citep{li2010contextual} and \emph{Generalised Thompson Sampling} \citep{Agrawal2013}.
-
-\subsubsection{The causal structure of contextual bandits}
-
-\todo{think carefully about the graphs in this sections}
-It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
-However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
-\citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information.
-This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ causes $\vec{Y}$}
-\label{fig:context-graph-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (4) edge (1)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{Y}$ causes $\vec{X}$}
-\label{fig:context-graph-anti-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (4)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ and $\vec{Y}$ are confounded}
-\label{fig:context-graph-confounded}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
-\node[lt, below=of 4](5){$U$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2)
-   (5) edge (4) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
-$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
-Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
-To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
-}
-\label{fig:contextual-bandit-causal-graph}
-\end{figure}
-
-\subsection{Learning from logged bandit data}
-Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
-In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
-The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
-
-This problem differs from the contextual bandit problem in that the learner is not interacting with the environment.
-As a result, there is no exploration-exploitation trade-off to be made.
-However, the problem does not reduce to supervised learning, because the label, $y$ is not the desired .
-In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
-The majority of the literature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
-\citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original estimator $\pi$ did not depend on $\vec{X}$.
-
-If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, discussed in section \ref{sec:causal_effect_under_ignorability}.
-The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ satisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
-The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivalent to $do(A=a)$.
-However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see \citet{Pearl2000}, section 4.2.
-In this case, letting $\Pns{\pi'}{a|\vec{x}}$ denote the distribution over actions under policy $\pi'$ given context $\vec{x}$, the expected (per round) reward obtained by $\pi'$ is given by,
-
-\eqn{
-\E{y|\pi'} = \E{y|do(a \sim \pi'(\vec{x}))} = \Esub{(\vec{x},a) \sim \P{\vec{x}}\Pns{\pi'}{a|\vec{x}}}{y|\vec{x},a}
-}
-
-As in estimating average causal effects under ignorability, we have a covariate shift problem, with training data sampled from $\P{\vec{x}}\Pns{\pi}{a|\vec{x}}\P{y|\vec{x},a}$ but generalisation error measured with respect to $\P{\vec{x}}\Pns{\pi'}{a|\vec{x}}\P{y|\vec{x},a}$.
-A difference in practice, is that in the applications frequently considered under learning from logged feedback data, such as ad serving or recommender systems, there may be substantial information available about $\pi$, in the best case, $\Pns{\pi}{a|\vec{x}}$ is known.
-This makes estimators utilising inverse propensity weighting, including doubly robust estimators as in \citet{Dudik2011}, more attractive.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
-\node[main node](1){$\vec{X}$};
-\node[main node, below right=of 1](2){$Y$};
-\node[main node,  below left=of 1](3){$A$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2) edge (3);
-\end{tikzpicture} 
-\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$}
-\label{fig:off-policy-evaluation}
-\end{figure}
-
-\citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the empirical best because the variance of the estimators for some policies may be much higher than for others.
-
-\subsection{Adding structure to actions}
-
-The classic multi-armed bandit is a powerful tool for sequential decision making.
-However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \citep{filippi2010parametric}, dependent bandits \citep{Pandey2007}, X-armed bandits \citep{Bubeck2010} and Gaussian process bandits \citep{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback\citep{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \citep{Piccolboni2001,Bartok2014}.
-
-In the next chapter, I propose a very natural connection between causal graphs and bandit problems and show it induces a novel form of additional feedback and structure between arms that cannot be can exploited by any of these previous approaches.
-
-\chapter{Causal Bandits: Unifying the approaches}
-
-\section{The framework}
+\begin{abstract} 
+We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
+Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
+We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
+than algorithms that do not use the additional causal information. This is an extended version of our paper published in NIPS 2016, \citet{NIPS2016_6195}.
+\end{abstract}
+
+\section{Introduction}
 \label{sec:intro_causal_bandit}
+
+Medical drug testing, policy setting, and other scientific processes are commonly framed and analysed in the language of sequential experimental design and, in special cases, as bandit problems~\citep{Robbins1952,Chernoff1959}.
+In this framework, single actions (also referred to as interventions) from a pre-determined set are repeatedly performed in 
+order to evaluate their effectiveness via feedback from a single, real-valued reward signal.
+We propose a generalisation of the standard model by assuming that, in addition to the reward signal, the learner observes the values of a number of covariates 
+drawn from a probabilistic causal model~\citep{Pearl2000}.
+Causal models are commonly used in disciplines where explicit experimentation may be difficult such as social science, demography and economics.
+For example, when predicting the effect of changes to childcare subsidies on workforce participation, or school choice on grades. 
+Results from causal inference relate observational distributions to interventional ones, allowing the outcome of an intervention to be predicted without
+explicitly performing it.
+By exploiting the causal information we show, theoretically and empirically, how non-interventional observations can be used to improve the rate at 
+which high-reward actions can be identified.
+
+The type of problem we are concerned with is best illustrated with an example. 
+Consider a farmer wishing to optimise the yield of her crop. 
+She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
+In each season the farmer has enough time and money to intervene and control at most one of these variables:
+deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertilizer; and irrigation or rain-proof covers will keep the soil wet or dry.
+When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
+How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
+%without sacrificing too much crop yield (relative to always choosing the best intervention) in the process?
+
+\paragraph{Contributions} We take the first step towards formalising and solving problems such as the one above. 
+In \S\ref{sec:setup-defs} we formally introduce \emph{causal bandit problems} in which interventions are treated as arms in a bandit problem but their influence on the reward --- along with any other observations --- is assumed to conform to a known causal graph.
+We show that our causal bandit framework subsumes the classical bandits (no additional observations) and contextual stochastic bandit problems (observations are revealed before an intervention is chosen) before focusing on the case where, like the above example, observations occur \emph{after} each intervention is made.
+
+Our focus is on the simple regret, which measures the difference between the return of the optimal action and that of the action chosen by the algorithm after $T$ rounds.
+In \S\ref{sec:simple-regret} we analyse a specific family of causal bandit problems that we call \emph{parallel bandit} problems in which $N$ factors affect the reward independently and there are $2N$ possible interventions.
+We propose a simple causal best arm identification algorithm for this problem and show that up to logarithmic factors it enjoys minimax optimal
+simple regret guarantees of $\smash{\tilde\Theta(\sqrt{m/T})}$ where $m$ depends on the causal model and may be much smaller than $N$.
+In contrast, existing best arm identification algorithms suffer $\smash{\Omega(\sqrt{N/T})}$ simple regret (Thm. 4 by \citet{audibert2010best}).
+This shows theoretically the value of our framework over the traditional bandit problem. 
+Experiments in \S\ref{sec:experiments} further demonstrate the value of causal models in this framework.
+
+In the general casual bandit problem interventions and observations may have a complex relationship. 
+In \S\ref{sec:simple-regret-general} we propose a new algorithm inspired by importance-sampling that a) enjoys sub-linear regret equivalent 
+to the optimal rate in the parallel bandit setting and b) captures many of the intricacies of sharing information in a causal graph in the general case.
+As in the parallel bandit case, the regret guarantee scales like $\smash{O(\sqrt{m/T})}$ where $m$ depends on the underlying causal structure, with 
+smaller values corresponding to structures that are easier to learn. The value of $m$ is always less than the number of interventions $N$ and in the special
+case of the parallel bandit (where we have lower bounds) the notions are equivalent.
+
+\section{Problem Setting}
+\label{sec:setup-defs}
+
 A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
@@
 We refer to these problems as \emph{causal bandit problems}.
-In this thesis I focus on the case where the causal graph is known.
+In this paper we focus on the case where the causal graph is known.
 Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.
@@
 
-The type of problem we are concerned with is best illustrated with an example.
-Consider a farmer wishing to optimise the yield of her crop.
-She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
-In each season the farmer has enough time and money to intervene and control at most one of these
-variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry.
-When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
-How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
-
 We will assume each variable only takes on a finite number of distinct values.
@@
 The values of variables in $\vec{X}^c_t$ that are non-decendents of $\vec{X}_t$ remain unchanged by the intervention.
-The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
+The objective of the learner is to minimise either the simple or cumulative regret.
 
@@
 
-\section{Causal bandits with post action feedback}
-
-We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, the cost of all allowable actions is equal and the goal of the learner is to minimise the simple regret.
-I presented this work at NIPS 2016 \citep{NIPS2016_6195}.
-
 \paragraph{Related Work} As alluded to above, causal bandit problems can be treated as classical multi-armed bandit problems by simply ignoring the causal model and extra observations and applying an existing best-arm identification algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
@@
 
-Our framework bears a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
+Our framework bears a superficial similarity to contextual bandit problems, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention. 
 However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
@@
 Combining their handling of  causal induction with our analysis is left as future work.
-% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
-Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
-(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
-It is easy to come up with a counter example in which it does not).
 
@@
 encountered in sequential design, but do not give an algorithm or theoretical results for this case.
+
+\section{Causal bandits with post action feedback}
+
+We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, the cost of all allowable actions is equal and the goal of the learner is to minimise the simple regret.
 
@@
 In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
-This has now been done by \citet{Sen2017a}, leading to problem dependent regret bounds for causal bandit problems.
+This has now been done by \citet{Sen2017a}, leading to the first problem dependent regret bounds for causal bandit problems.
 Note that these modifications do not improve the mini-max regret, which at least for the parallel bandit is already optimal.
-For this reason we focused on emphasising the point that causal structure can and should be exploited when available.
+For this reason we focused on emphasising the main point that causal structure can and should be exploited when available.
 Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
@@
 
-\chapter{Conclusion}
-
-\chapter*{Regret is a goat}
-
-Consider a stochastic, two-armed bernoulli bandit with the (general) joint distribution of the (conterfactual) rewards $\P{\cf{Y}{1},\cf{Y}{2}}$ given in table \ref{table:joint_bernoulli_counterfactuals}.
-Suppose without loss of generality that arm 1 is the optimal arm.
-
-\begin{table}[h]
-\caption{General joint distribution over two bernoulli random variables.}
-\label{table:joint_bernoulli_counterfactuals}
-\begin{center}
-\begin{tabular}{ c | c | c }
-$\cf{Y}{1}$ & $\cf{Y}{2}$ & $\P{\cf{Y}{1},\cf{Y}{2}}$ \\
-  \hline			
-  1 & 1 & $a$ \\
-  1 & 0 & $p-a$ \\
-  0 & 1 & $q-a$ \\
-  0 & 0 & $1+a - (p+q)$ \\
-  \hline  
-\end{tabular}
-\end{center}
-\end{table}
-
-Recall that the regret is defined by,
-
-\eqn{
-\rawregret = \sum_{t=1}^T (\cfb{Y}{i^*} - \cfb{Y}{a_t})
-}
-
-Consider an algorithm that always selects arm 2.
-The variance (but not expectation) of the regret incured by this algorithm depends on the covariance between the counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{2}$, which cannot estimated from data.
-
-\eqn{
-\E{\rawregret} = \sum_{t=1}^T(\E{\cf{Y}{1}}-\E{\cf{Y}{2}}) = T(p-q)
-}
-
-\eqn{
-\V{\rawregret} &= \E{\left(\sum_{t=1}^T (\cfb{Y}{1} - \cfb{Y}{2})\right)^2} - \left(\sum_{t=1}^T(\E{\cf{Y}{1}}-\E{\cf{Y}{2}})\right)^2\\
-&= \sum_{t=1}^T \E{(\cfb{Y}{1} - \cfb{Y}{2})^2 - (\E{\cfb{Y}{1}}-\E{\cfb{Y}{2}})^2}\\
-&= \sum_{t=1}^T\left(\V{\cfb{Y}{1}}+\V{\cfb{Y}{2}} - 2 Cov(\cfb{Y}{1},\cfb{Y}{2})\right)\\ &= T\left(p(1-p)+q(1-q) - 2(a-pq)\right)
-}
-
-\bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{library}% write the directory to the .bib file
+{\small\bibliography{library}}
+\bibliographystyle{apalike}
 \end{document}
-
-\chapter{Causation vs Prediction}
-
-The two cultures \citep{Breiman2001}
- - wrong because suggets we can just use the machine learning approach for everything.
-Does not address the problems that a) structural assumptions required for causal inference.
-b) we cannot use cross-vaidation to assess the validity of a causal model (really b is the whole point).
-
-To explain or to predict \citep{Shmueli2010}
-- wrong because they assume all the parts of explanatory modelling are required - so black box models cannot be used.
-
-Predict the relationship between X and Y as a way to stratify to reduce the variance in an experiment.
-
-They consider a different definition of explanatory models,
-
-Definition as backwards.
-Explanatory models focus on theory because some theory (or equivelently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
-
-Causal modelling frameworks allow us to seperate assumptions about the structure of the relationships between variables from assumptions about the functional form.
-
-What can be data driven and what cannot? (for a given set of assumptions).
-
-Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the languate of EoP but are built on a hybrid of both approaches.

### 2017-08-26 thesis/offcuts_chapter6.tex
creates extended version of nip paper for arXiv, restructuring work on thesis

@@
+Observational causal inference, the problem of estimating the impact of some intervention in a system from observational data, has been studied in depth, particularly within economics and the social sciences for
+
+The problem of estimating the impact of an intervention in a system from observational data has been studied in depth, particularly in economics and social sciences, for many years.
+
 They also consider the impact of different assumptions about the relationship between the suggested action $A$ and actual action $X$, which are not captured by structure of the graph.

### 2017-08-26 thesis/thesis_map.tex
creates extended version of nip paper for arXiv, restructuring work on thesis

@@
 It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.
+\todo{reference a review paper on causal discovery here}
 
@@
 
-\chapter{Causal inference from observational data}
+\chapter{Learning from observational data}
 \label{chap:causality}
+
+The goal of causal inference is to learn the effect of taking an action.
+We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world.
+We are generating and storing data on almost every aspect of our lives at an unprecedented rate.
+As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase.
+However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question.
+An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severly limted.
+This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it.
+This is the problem of observational causal inference.
 
@@
 
+As previusly mentioned, observational causal inference is
+
 Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
-To do this we need terminology to describe actions and how we anticipate the system will respond to them.
+As previuosly mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention.
+Doing this requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate the system will respond to them.
 Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
+
+counterfactuals
+
+MOTIVATE WHY I AM DESCRIBING THEM ALL SUMMARISE WHAT EACH OFFERS.
+
 In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ.
@@
 
-\section{Causal Bayesian networks}
+\subsection{Causal Bayesian networks}
 
@@
 
-\subsection{Can't everything be encoded in a Bayesian model?}
+\subsubsection{Can't everything be encoded in a Bayesian model?}
 Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
@@
 
-\subsection{Limitations of causal Bayesian networks}
+\subsubsection{Limitations of causal Bayesian networks}
 A number of criticisms have been levelled at this approach to modelling causality.
@@
 
-\section{Counterfactuals}
+\subsection{Counterfactuals}
 \label{sec:counterfactuals}
@@
 
-\section{Structural Equation models}
+\subsection{Structural Equation models}
 \label{sec:SEM}
@@
 
-\section{Comparing and unifying the models}
+\subsection{Comparing and unifying the models}
 
@@
 
-\section{The Do Calculus} 
+\subsection{The Do Calculus} 
 \label{sec:do_calculus}
@@
 
-\section{Estimation}
-\subsection{Defining causal effects} 
+\subsection{Estimation}
+\subsubsection{Defining causal effects} 
 So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
@@
 
-\subsection{Estimating causal effects by adjusting for confounding variables}
+\subsubsection{Estimating causal effects by adjusting for confounding variables}
 \label{sec:causal_effect_under_ignorability}

### 2017-08-27 thesis/offcuts_chapter6.tex
rough restructure complete

@@
+\section{careful prior}
+
+\subsubsection{Example}
+
+\eqn{
+U &\sim N(0 , 1)\\
+Z &\sim N(2U , 0.09)\\
+X &\sim N(0.5Z , 1)\\
+Y &\sim N(3U + -Z + 0.5X , 0.25)
+}
+
+\begin{figure}[h]
+\includegraphics[scale=.3]{prior_no_prior}
+\caption{no prior, posterior on $w_{yx}$ centered around $w_{yx}$}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[scale=.3]{prior_bad_prior}
+\caption{Prior around $w_{yz}$, $\beta \sim N(w_{yz},\sigma = 0.5)$. The posterior on $w_{yx}$ is biased away from its true value}
+\end{figure}
+
 Observational causal inference, the problem of estimating the impact of some intervention in a system from observational data, has been studied in depth, particularly within economics and the social sciences for
@@
 Randomised data with imperfect compliance would be an example.
+
+However, although each framework has it own set of advantages, it turns out that they are functionally equivelent for a wide class of problems - those relating to estimating the outcome of an intervention.
+
+\section{Philosophical objections to counterfactuls}
+There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
+This makes it quite easy to (accidentally) make statements about counterfactuals that cannot be tested with empirical data.
+Consider the following example based on \citet{Dawid2000}.
+Again we have a drug where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
+Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal with equal variance (for simplicity).
+
+% vector with mean mu_t,mu_c and a covariance matrix
+\eqn{
+P(\cf{Y}{1},\cf{Y}{0}) \sim N( 
+         \begin{bmatrix}
+           \mu_{1} \\
+           \mu_2 \\
+         \end{bmatrix},
+         \begin{bmatrix}
+           \sigma^2 & \rho \sigma^2 \\
+           \rho \sigma^2 & \sigma^2 \\
+         \end{bmatrix}
+         )
+}
+
+The distribution of \emph{individual treatment effects}, $\tau = \cf{Y}{1} - \cf{Y}{0}$ is also normal,
+
+\eqn{
+P(\tau) = N(\mu_1 - \mu_0,2\sigma^{2}(1-\rho))
+}
+
+\begin{figure}[h]
+\captionsetup[subfigure]{position=b}
+\centering
+\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$.
+The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated.
+\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
+\hspace{0.02\textwidth}
+\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
+{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
+\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
+\label{fig:metaphysical_distribution_difference}
+\end{figure}
+
+From a (large) randomised controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}.
+These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated respectively.
+However, the distribution over the individual causal effects depends on $\rho$, see figure \ref{fig:metaphys_b}.
+The key problem is that we can never observe the joint distribution over $\cf{Y}{1}$ and $\cf{Y}{0}$.
+As a result, $\rho$ and thus the variance of $\tau$ is not identifiable, even from experimental data.
+\citet{Dawid2000} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects.
+He further points out that the interventional distributions in figure \ref{fig:metaphys_a}, along with a loss function, contain all the information required to decide how to treat a new patient.
+
+This result is unintuitive.
+It seems on the face of it that the distribution of individual causal effects is relevant to our decision making.
+If $\rho = 1$ then almost everyone benefits slightly from the treatment whilst if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.
+This confusion can be resolved by thinking about personalised rather than individual causal effects.
+It is entirely possible that potentially observable characteristics (such as gender, age, genetics, etc) affect how people will respond to the treatment.
+We can partition people into sub-populations on the basis of these characteristics and measure different \emph{personalised} causal effects for each group.
+The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much can be gained from further personalisation.
+The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).

### 2017-08-27 thesis/thesis_map.tex
rough restructure complete

@@
 
+\setcounter{secnumdepth}{4}
+
 \begin{document}
@@
 \item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
-\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomized experimental design?
+\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomised experimental design?
 \item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
@@
 - can imposing a prior which is true with respect to that variable, increase bias due to confounding?
-(In the same way that in the pnemonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.
-
-\item What are the nessessary and sufficient conditioins for decision making?
+(In the same way that in the pneumonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.
+
+\item What are the necessary and sufficient conditions for decision making?
 How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function).
@@
 
-\item Does the way we define supervised learning and loss fuctions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?
+\item Does the way we define supervised learning and loss functions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?
 
@@
 
-There are two ways in which an association detected in a dataset may be non-causal.
+There are two ways in which an association detected in a data set may be non-causal.
 The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample.
@@
 It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published.
-As a consequnce, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
+As a consequence, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
 This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}.
-This issue can be ameliorated by getting more data and by seperating learning models from evaluating their performance, for example by evaluating models on a stict hold-out set or on the extent to which their results can be replicated.
-
-However, a strong association, observed in multiple indpendent studies may still not be causal.
+This issue can be ameliorated by getting more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.
+
+However, a strong association, observed in multiple independent studies may still not be causal.
 The correlation can arise because both variables are consequences of some other, unmeasured factor.
-For example, the reading ability of children under twelve is strongy correlated with their height, because older children are taller and can read better.
+For example, the reading ability of children under twelve is strongly correlated with their height, because older children are taller and can read better.
 However height is not a cause of reading ability because if we were to intervene to increase a child's height, for example by giving them growth hormones, we would not expect dramatic improvements in their reading.
 Similarly, extra tutoring in reading will not make a child grow taller.
-This problem is fundamentally different to the issue of spurious correlations arising by chance in finite datasets.
-Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to seperate causation from correlation.
+This problem is fundamentally different to the issue of spurious correlations arising by chance in finite data sets.
+Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to separate causation from correlation.
 
 The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables.
-In this thesis, I take an interventionalist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal.
+In this thesis, I take an interventionist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal.
 This viewpoint captures the types of questions that motivate this thesis.
@@
 
-Causality is often liked to explaination; understanding how and why things happen.
-I view explaination in terms of compression and generalisabilty: how much information about the world a model can capture.
-This creates a hierachy in the degree to which models are explainitory, rather than a simple binary distinction.
+Causality is often liked to explanation; understanding how and why things happen.
+I view explanation in terms of compression and generalisabilty: how much information about the world a model can capture.
+This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction.
 A standard predictive model encodes all the information we need to predict some output given inputs provided the system generating the data does not change.
 A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed.
-More detailed causal models could predict the outcome for a wide range of combiations of interventions conditional on a range of contexts.
+More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts.
 By considering conditional interventions within our definition of causal questions we also capture mediation: the study of through which pathways one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact to mathematical equatons like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
+Finally, a model that can distil how elements interact to mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
 
@@
 The former aims to identify the effect of a known cause.
-The latter can be viewed as identiying causes of an effect.
+The latter can be viewed as identifying causes of an effect.
 They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}.
@@
 The goal of identifying the causes of an effect can be formalised within the graphical causal model framework.
-Solving this problem is certainly much more challanging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables.
-These practical difficulties may well be overwealming, particularly in fields such as social science and economics where datasets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively.
+Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables.
+These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively.
 However, this does not mean that the problem of identifying causes of effects is ill-posed in principle.
-It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.
+It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature which we review briefly in section \ref{sec:causal-discovery}.
 \todo{reference a review paper on causal discovery here}
 
-There has traditionally been a large gap between researches in machine learning who focus on prediction, use largly non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models.
+There has traditionally been a large gap between researches in machine learning who focus on prediction, use largely non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models.
 However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of for the training and evaluation models.
-In the next section we emphasise the subtley that can exist in determining if a problem is causal by examiming some typical examples.
+In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.
 
@@
 
-Machine learning is in the midst of a boom, driven by the availability of large datasets and the computation resources to process them.
-Machine learning techniques are being applied a huge range of problems, in both industry and achademia.
-The following examles are inteded to capture the bredth of problems that machine learning algorithms are actively beeing applied to.
+Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them.
+Machine learning techniques are being applied a huge range of problems, in both industry and academia.
+The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to.
 Which, if any, of these problems require causal inference?
@@
 
-The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important considertion.
+The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important consideration.
 In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.
@@
 
-This example also highlights a connection between causal models and transparancy in machine learning.
-If we are using a non-causal model to make decision effecting people, there will be a trade-off between the performance and transparancy of the model; not because the requirement for transparancy restricts us to simple models but because revealing how the model works allows people to change their behaviour to game it.
+This example also highlights a connection between causal models and transparency in machine learning.
+If we are using a non-causal model to make decision effecting people, there will be a trade-off between the performance and transparency of the model; not because the requirement for transparency restricts us to simple models but because revealing how the model works allows people to change their behaviour to game it.
 
@@
 It is not.
-Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admition decisons.
+Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admission decisions.
 \citet{Caruana2015} found exactly this effect in a real data set.
@@
 They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
-The issue is not with model, it performed very well at the task for which it was trained, which is to predict wo would be likely to die under the original admition and treatment protocols.
+The issue is not with model, it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols.
 However, using it to decide how to \emph{change} these protocols could kill.
-The actual question of interest in this case is what happens to patients with charachteristics $X$ if we assign them treatment according to decision rule $Z$.
+The actual question of interest in this case is what happens to patients with characteristics $X$ if we assign them treatment according to decision rule $Z$.
 
@@
 This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players.
-Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a compbination of these approaches \citep{silver2016mastering}.
-The supervised learning was enhanced by having the agent play (varients) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
+Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}.
+The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
 
@@
 However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach.
-The actions the agent has to explore are the sames ones as human players explored to generate the trainging data, and both have the same objective - to win the game.
-In addition, the state of the board encapsulates all the information relevent to selecting a move.
+The actions the agent has to explore are the sames ones as human players explored to generate the training data, and both have the same objective - to win the game.
+In addition, the state of the board encapsulates all the information relevant to selecting a move.
 These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.
 
-Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference.
+Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predictive) inference.
 A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes.
 The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people.
-Think about problems like credit scoring and parole decisons.
-There are strong ethical grounds for demanding transparancy, but if the goals of society and the individuals are not perfectly aligned and there is any posibility that people can manipulate features independtly of the outcome, there will be a conflict between model accuracy and transparancy.
-It is rare to build a model without any intent to make some kind of decion on the basis of its results.
+Think about problems like credit scoring and parole decisions.
+There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency.
+It is rare to build a model without any intent to make some kind of decision on the basis of its results.
 Thus, I argue we should assume a causal model is required until we can justify otherwise.
@@
 
--------------------
-
-In chapter \ref{chap:interventionalist_view} I describe the interventionalist viewpoint on learning the outcome of actions.
-
-I review the key problems and results within the bandit liturature
-
-highlight some interesting connections between existing strands of work on bandits and causal inference.
-
--------------------
-
-Causal questions are everywhere.
-Statistics, economics, social science, artificial intelligence and machine learning.
-Techniques and for addressing such problems have developed in parralel within many diciplines.
-These techniques can usefully be categorised into two broad approaches, reinforcement learning and observational causal inference.
+As we have shown, problems involving causality are ubiquitous in many fields.
+As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning.
+Although the focus and terminology can differ substantially between fields these techniques all address the underlying goal of estimating the effect of or optimally selecting interventions in some system.
+Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.
 
@@
 A particular strand of research within reinforcement learning are multi-armed bandit problems.
-They describe settings in which there are a set of available actions, the agent repeately decides which to select and then observes the outcome of the chosen action.
+They describe settings in which there are a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action.
 They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.
 
-The approach of learning the outcome of an action by taking it has plays a key role in advancing our knoweledge of the world.
+The approach of learning the outcome of an action by taking it has plays a key role in advancing our knowledge of the world.
 However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen.
 Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.
-
 Observational causal inference can be viewed as a form of transfer learning.
-The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the orignial system.
-This is impossible without some assumptions about how the the orignal system and the system after intervention are related to one-another.
-The key to observational inference is a framework that models how actions change the state of the world and thus allows us to map information collected in one setting to another.
+The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system.
+This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another.
+The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.
 
@@
 
-The goals of this thesis are: to clarify the distinction between prediction and causal inference and highlight when and how we can leverage advances in prediction to improve causal inference and to draw together the key approaches to solving causal problems, from both the interventionalist and observational viewpoints.
-In chapter two, I describe the existing frameworks for formalising causality developed withing economics, statistics and machine learning and discuss how they relate to one another.
-Chapter three reviews the key approaches to causal inference from observational data.
-Chapter four highlights the interventionalist viewpoint, covering traditional randomised experiments and the closely related multi-armed bandit problems and highlights connections between the algorithms applied to bandit problems and those for estimating causal effects from observational data.
-
-In chapter five, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
+The primary goal of this thesis is to draw together the key approaches to solving causal problems from both the interventionalist and observational viewpoints.
+
+In chapter \ref{chap:causality} I describe key frameworks for causal inference from observational data and discuss how the relate to one-anther and review how they can be leveraged to answer causal queries.
+Chapter \ref{chap:interventionalist_view} covers the interventionalist viewpoint, including traditional randomised experiments and the closely related multi-armed bandit problems.
+It describes the key problems and results within the bandit literature and highlights some interesting connections between existing strands of work on bandits and causal inference.
+In chapter \ref{chap:causal-bandit}, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
 The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
@@
 
-Chapters two to four predominantly review the existing liturature with the goal of making connections between the different viewpoints and approaches.
+Chapters \ref{chap:causality} and \ref{chap:interventionalist_view} predominantly review the existing literature with the goal of highlighting connections between the different viewpoints and approaches.
 They do not contain novel technical results but present the material in a unified way.
-Chapter five is a novel contribution.
-It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit liturature and describes an algorithm that can leverage this structure to make better decisons.
+Chapter \ref{chap:causal-bandit} is a novel contribution.
+It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit literature and describes an algorithm that can leverage this structure to make better decisions.
 
@@
 However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question.
-An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severly limted.
+An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited.
 This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it.
@@
 
-As previusly mentioned, observational causal inference is
-
 Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
-As previuosly mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention.
-Doing this requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate the system will respond to them.
+As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention.
+Mapping properties from one system to another requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate a system will respond to them.
 Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
 
-counterfactuals
-
-MOTIVATE WHY I AM DESCRIBING THEM ALL SUMMARISE WHAT EACH OFFERS.
-
-In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ.
-We will also use them to describe the following simplified examples.
-The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.
+Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings.
+They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another.
+Counterfactual assertions can be expressed very naturally in human languages and are quite prevalent in  every day conversations;\quotes{if I had worked harder I would have got better grades}, \quotes{she would have been much sicker if she hadn't taken antibiotics}.
+Structural equation models have been developed and applied primarily within economics and related disciplines and can be seen as an attempt to capture key aspects of the behaviour of people with mathematics.
+Questions around designing policies or interventions play a central role in economics.
+Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised) framework and associated set of methods for estimating causal effects.
+Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks.
+They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of conditions under causal effects can be inferred from observational data.
+
+An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}.
+However, the literature on causal inference techniques remains spit between the different frameworks.
+Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scores, are described using the counterfactual framework.
+Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions are often based around structural equation models.
+This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.
+
+In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they he problems they allow us to solve, the assumptions they rely on and how they differ.
+We hope that by describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them.
+However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers.
+In order to demonstrate the notation and formalisms each framework provides we will use them to describe the following simple examples.
 
@@
 \caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$.
-Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution.
+Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical structure implies exactly the same set of independencies exhibited by the distribution.
 Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
@@
 
+A causal Bayesian network represents much more information than a Bayesian network with identical structure.
+A causal network encodes all possible interventions that could be specified with the do-notation.
+For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
+The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
+Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
+After each two variable intervention, we have a distribution over one variables which requires one parameter.
+This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
+
 Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes.
 The global Markov property and d-separation also hold in causal networks.
-D-speration, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
-We include a brief introduction in section \ref{sec:dsep}.
-
-\subsubsection{Can't everything be encoded in a Bayesian model?}
+D-separation, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
+We include a brief review of D-separation in section \ref{sec:dsep}.
+
+\subsubsection{Limitations of causal Bayesian networks}
+A number of criticisms have been levelled at this approach to modelling causality.
+One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}.
+However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions.
+For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
+But, what if some people failed to take the prescribed treatment?
+We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
+Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](1){Prescribed Treatment};
+\node[node, below right=of 1](2){Actual Treatment};
+\node[node, right=of 2](3){Outcome};
+\node[lt, above left=of 3](4){U};
+\path[]
+	(1) edge (2)
+	(2) edge (3) 
+	(4) edge (3) edge (2);
+\end{tikzpicture}
+\caption{Randomised treatment with imperfect compliance}
+\label{fig:randomized_imperfect_compliance}
+\end{figure}
+
+Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables.
+Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology.
+We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
+However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
+Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
+Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
+\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
+
+\subsection{Counterfactuals}
+\label{sec:counterfactuals}
+
+The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
+Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
+Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
+
+In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would have happened if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would have happened if they got the placebo, $\cf{y_{i}}{0}$.
+The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
+The problem can be resolved if, instead of people, there are units that can be assumed to be identical or that will revert exactly to their initial state some time after treatment.
+This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
+
+Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo.
+Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
+The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated.
+Similarly $Y^{0}$ represents the potential outcome for the placebo.
+The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
+We can estimate (from an experimental or observational study):
+\begin{itemize}
+\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
+\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
+\end{itemize}
+
+Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
+For those who did not take the treatment, the observed outcome is the same as the outcome \emph{had} they not taken the treatment.
+Equivalently stated:
+
+\eq{
+\P{Y^{0}|X=0}&= \P{Y|X=0}\\
+\P{Y^{1}|X=1}&=\P{Y|X=1}
+}
+
+If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
+
+\eq{
+\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
+\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
+}
+
+This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
+
+\eq{
+\P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
+}
+
+The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}.
+They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
+This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
+In general the treatment assignment will not be independent of the potential outcomes.
+In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e.
+$X \nci \cf{Y}{0}$.
+Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e.
+$X \nci \cf{Y}{1}$.
+A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
+
+\vspace*{.3cm}
+\begin{theorem}[Ignorability \citep{Rosenbaum1983, Pearl2000}] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
+
+\eqn{
+\label{eqn:counterfactual1}
+\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
+\label{eqn:counterfactual2}
+\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
+}
+\end{theorem}
+
+Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
+Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
+
+Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}.
+This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
+As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
+The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
+SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
+
+There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
+In particular, statements involving joint distributions over counterfactual variables may not be able to be validated empirically \citet{Dawid2000}.
+One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
+Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
+This is certainly convenient, if rather imprecise.
+However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention.
+It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).
+
+\subsection{Structural Equation models}
+\label{sec:SEM}
+
+Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
+The mechanism (but not the output) is assumed to be independent of what is fed into it.
+Uncertainties are not inherent but arise from unmeasured variables.
+Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
+More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \citep{Pearl2000}.
+
+Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
+The noise terms are required to be mutually independent.
+If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent variable.
+Structural equation models can be represented visually as a network.
+Each variable is a node and arrows are drawn from causes to their effects.
+Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](2){$X = f_x(\epsilon_x)$};
+\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
+\node[lt, above=of 3](4){$\epsilon_y$};
+\node[lt, above=of 2](5){$\epsilon_x$};
+\path[]
+	(2) edge (3)
+	(4) edge (3)
+	(5) edge (2);
+\end{tikzpicture}
+\caption{SEM for example \ref{exm:ranomized_experiment}}
+\label{fig:sem_randomized_treatment}
+\end{figure}
+
+This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
+This is justifiable on the grounds that $X$ is random.
+The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
+Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
+The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
+
+Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$.
+The function $f_y$ and distribution over $\epsilon_y$ does not change.
+This results in the interventional distribution, \footnote{We have assumed the variables are discrete only for notational convenience}
+
+\eqn {
+\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
+}
+
+The observational distribution of $Y$ given $X$ is,
+
+\eqn{
+\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
+& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
+}
+
+The interventional distribution is the same as the observational one.
+The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
+
+The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
+Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
+
+\eqn{
+\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
+\label{eqn:sem:adjusting}
+&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
+}
+Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivalent to the observational distribution given by:
+
+\eqn{
+\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
+}
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](2){$X = f_x(Z,\epsilon_x)$};
+\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
+\node[lt, above=of 6](7){$\epsilon_z$};
+\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
+\node[lt, above=of 3](4){$\epsilon_y$};
+\node[lt, above=of 2](5){$\epsilon_x$};
+\path[]
+	(2) edge (3)
+	(4) edge (3)
+	(5) edge (2)
+	(7) edge (6)
+	(6) edge (2) edge (3);
+\end{tikzpicture}
+\caption{SEM for example \ref{exm:adjusting}}
+\label{fig:sem:preschool}
+\end{figure}
+
+Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
+A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
+However a SEM is, by definition, more than a set of simultaneous equations.
+By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
+When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
+The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
+
+\subsection{Comparing and unifying the models}
+\label{sec:unifying_causal_models}
+
+Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
+If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
+In other words, the network is a causal Bayesian network.
+All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
+Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
+This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}).
+Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
+We can also connect this to counterfactuals via:
+
+\begin{equation}
+\begin{aligned}
+&Y^{0} \equiv P(Y|do(X=0)) \\
+&Y^{1} \equiv P(Y|do(X=1))
+\end{aligned}
+\end{equation}
+
+The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
+When discussing the counterfactual model, we actually made the slightly weaker assumption:
+
+\begin{equation}
+\label{eq:weakignore}
+X \ci Y^{0} \text{ and } X \ci Y^{1}
+\end{equation}
+
+It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}.
+The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
+For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
+Then use propensity scores \citep{Rosenbaum1983} to estimate the effect.
+If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
+The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
+These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
+
+In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
+The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
+The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
+This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
+In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field.
+Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
+In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
+
+\subsection{What does a causal model give us? Resolving Simpson's paradox}
+
+We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
+The following example is adapted from \citet{Pearl2000}.
+Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
+She keeps track of which medication her patients choose and whether or not the treatment is successful.
+She obtains the results in table \ref{tab:simpson1}.
+
+\begin{table}[h]
+\caption{Treatment results}
+\label{tab:simpson1}
+\center
+\begin{tabular}{c c c c c}
+Treatment & Success & Fail & Total & Success Rate\\
+\hline
+\rowcolor[gray]{0.9} 
+A &  87 & 13 & 100 & 87\%\\
+\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
+\hline
+\end{tabular}
+\end{table}
+
+Drug A appears to perform better.
+However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
+
+\begin{table}[h]
+\caption{Treatment results by gender}
+\label{tab:simpson2}
+\center
+\begin{tabular}{c c c c c c}
+Gender & Treatment & Success & Fail & Total & Success Rate\\
+\hline
+\rowcolor[gray]{0.9}
+M & A & 12 & 8 & 20 & 60\%\\
+M & B & 56 & 24 & 80 & 70\%\\
+\rowcolor[gray]{0.9}
+F & A & 75 & 5 & 80 & 94\%\\
+F & B & 19 & 1 & 20 & 95\%\\
+\hline
+\end{tabular}
+\end{table}
+
+Once the data is broken down by gender, drug B looks better for both men \emph{and} women.
+Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must recommend which to subsidise under a national health scheme).
+Should she choose A or B?
+The ambiguity in this question lies at the heart of Simpson's paradox.
+How does causal modelling resolve the paradox?
+The key is that the doctor is trying to choose between \emph{interventions}.
+She wants to know what the success rate will be if she changes her practice to give all the patients one drug, rather than allowing them to choose as currently occurs.
+
+Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
+The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
+Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
+Some assumptions about the causal relationships between the variables are required.
+In this example, it seems reasonable to conclude that gender may affect the treatment chosen and the outcome.
+Assuming there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](1){Z = Gender};
+\node[node, below left=of 1](2){T = Treatment};
+\node[node, below right=of 1](3){Y = Outcome};
+\path[]
+	(1) edge (2) edge (3) 
+	(2) edge (3);
+\end{tikzpicture}
+\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
+\label{fig:simpsons_paradox}
+\end{figure}
+
+With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
+Knowing a patient took drug A indicates they are more likely to be female.
+When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
+However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
+So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
+Drug B is the better choice.
+
+\eqn{
+\P{Y|do(T)} = \P{Y|T,female}\P{female}+ \P{Y|T,male}\P{male}
+}
+
+Is the solution to Simpson's paradox to always to break down the data by as many variables as possible?
+No.
+Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
+This is a drug designed to prevent heart disease.
+One pathway to doing so might well be to lower blood pressure.
+Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
+It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
+Now, however, table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood pressure control and better overall outcomes.
+In this setting $\P{Y|do(T)} = \P{Y|T}$ and drug A is the better choice.
+
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
+\node[node](1){Z = Blood Pressure};
+\node[node, below left=of 1](2){T = Treatment};
+\node[node, below right=of 1](3){Y = Outcome};
+\path[]
+	(1) edge (3) 
+	(2) edge (3)
+	(2) edge (1);
+\end{tikzpicture}
+\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
+\label{fig:simpsons_paradox2}
+\end{figure}
+
+Note that we have not changed the data itself, only the description of the variables that it is associated with.
+This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
+From a purely statistical viewpoint there is no paradox.
+The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
+The paradox only arises when we attempt to use the data to select an intervention and is resolved when we apply a causal approach to do so.
+
+\eqn{
+\label{eqn:simpsons_paradox}
+\exists \set{N_1,...N_4, N_1'...N_4'} \in \mathbb{N} : \; 
+\textcolor{green}{\frac{N_1}{N_2}< \frac{N_1'}{N_2'}}, \;\; 
+\textcolor{blue}{ \frac{N_3}{N_4}<\frac{N_3'}{N_4'}}
+ \text{ and } \frac{N_1+N_3}{N_2+N_4} > \frac{N_1'+N_3'}{N_2'+N_4'}
+}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.4]{figures/simpson_vector}
+\caption{Simpson's reversal visualised.
+The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
+However, when we sum them, the ratio is steeper for the un-primed variables.}
+\label{fig:simpsons_vector}
+\end{figure}
+
+There are many other plausible causal graphs for both scenarios above.
+Perhaps income affects drug choice as well as gender, or gender might affect treatment choice and blood pressure control given treatment, etc.
+Causal modelling provides a powerful tool to specify such assumptions and to determine how to estimate causal effects for a given model as we discuss in the next section.
+
+\section{Answering Causal Questions} 
+\label{sec:causal effect estimation}
+We can roughly categorise the problems studied within causal inference from observational data into two groups, causal effect estimation and causal discovery.
+In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
+The goal is then to estimate the effect of an intervention or range of interventions in the system.
+Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}.
+Almost every time someone runs a regression model the key quantity of interest is a causal effect.
+Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
+
+Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
+This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
+Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
+An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
+In this thesis I have focused on problems where the structure of the causal graph is known.
+Extending my work to problems where the causal structure is unknown, leveraging the work on causal discovery, is a rich and fascinating line of potential future work, which I discuss very briefly in section \ref{sec:causal_bandit_discussion}.
+
+\subsection{Mapping from observational to interventional distributions}
+
+A central component of estimating causal effects from observational data is determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
+We did this on an ad-hoc basis to resolve the examples discussed in section \ref{sec:causal_models}.
+In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:estimation_from_finite_data}, discuss the key issues involved in estimating such expressions from finite sample data.
+We assume the basic structure of the graph is known.
+That is, we assume that we can draw a network containing (at a minimum):
+
+\begin{itemize}
+\item the target/outcome variable we care about,
+\item the focus/treatment variables on which we are considering interventions,
+\item any variables which act to confound two or more of the other variables we have included, and
+\item any links between variables we have included. 
+\end{itemize}
+
+Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
+For example, consider estimating the impact of schooling on wages.
+Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
+Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
+
+How can the structure of the causal graph be leveraged to compute interventional distributions from observational ones?
+Given the graph corresponding to the observational distribution, the graph after any intervention can be obtained by removing any links into variables directly set by the intervention.
+The joint interventional distribution is the product of the factors associated with the interventional graph, as given by the truncated product formula \ref{eq:truncatedproduct}.
+If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
+However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
+
+The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
+By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
+This gives rise to the do-calculus \citep{Pearl2000}.
+The do calculus consists of three rules.
+They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.
+
+\subsubsection{Independence in Bayesian networks: D-separation}
+\label{sec:dsep}
+
+Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
+Therefor, in this section, we briefly review the key properties of Bayesian networks.
+A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
+Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
+
+\vspace{.3cm}
+\begin{theorem}
+\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
+\end{theorem}
+
+The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$.
+D-separation is an algorithm that extends the local Markov property to find these additional independencies.
+It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
+
+The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
+From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
+Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
+In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
+In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
+
+The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
+The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
+As an example, consider a scholarship available to female or disadvantaged students.
+Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship.
+There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
+However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
+\begin{figure}[h]
+\centering
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y| Z$}
+\label{fig:dseptripple1}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$X$};
+\node[main node, below =of 1](2){$Z$};
+\node[main node, below =of 2](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    (2) edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|Z$}
+\label{fig:dseptripple2}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$Y$};
+\node[main node, below =of 1](2){$Z$};
+\node[main node, below =of 2](3){$X$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    (2) edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|Z$}
+\label{fig:dseptripple3}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$Z$};
+\node[main node, below left =of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+        edge node {} (3);	
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}[t]{0.2\textwidth}
+\centering
+\caption{$X \ci Y|\{\}$}
+\label{fig:dseptripple4}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
+  thick,main node/.style={observed}]
+\node[main node](1){$Z$};
+\node[main node, above left =of 1](2){$X$};
+\node[main node, above right =of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge node {} (1)
+    (3) edge node {} (1);	
+\end{tikzpicture}
+\end{subfigure}
+\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
+\label{fig:dseptriples}
+\end{figure}
+
+\vspace{.3cm}
+\begin{definition}[unblocked path]
+\label{def:unblocked}
+A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
+It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
+\end{definition}
+
+\vspace{.3cm}
+\begin{definition}[d-separation]
+\label{def:dsep}
+The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, there are no unblocked paths from any $X \in \boldsymbol{X}$ to any $Y \in \boldsymbol{Y}$ after conditioning on $\boldsymbol{Z}$.
+\end{definition}
+
+\vspace{.3cm}
+\begin{theorem}[d-separation and conditional independence]
+\label{thrm:dsep_independence}
+If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in a Bayesian network $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
+Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected (not d-separated) given $\boldsymbol{Z}$ then it is possible to construct a distribution $P'$ that factorises over $G$ in which they are dependent.
+\end{theorem}
+
+Theorem \ref{thrm:dsep_independence} says that independencies implied by d-separation on a graph hold in every distribution that can be factored over that graph and that if ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in \emph{all} distributions that can be factored over $G$ then they are d-separated in $G$.
+If we denote the independencies implied by d-separation in a graph by $\indep{G}$ and the set of independencies in a distribution by $\indep{P}$ then $\indep{G} \subseteq \indep{P}$.
+
+If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
+However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
+A particular case in which this occurs is when there are deterministic relationships between variables.
+If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
+This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
+This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
+D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
+
+\subsubsection{The Do Calculus} 
+\label{sec:do_calculus}
+
+The do-calculus is a set of three rules \citep{Pearl1995} that can be applied to simplify the expression for an interventional distribution.
+If by repeated application of the do-calculus, along with standard probability transformations, we can obtain an expression containing only observational quantities then we can use it to estimate the interventional distribution from observational data.
+Let $\boldsymbol{X}$,$\boldsymbol{Y}$,$\boldsymbol{Z}$ and $\boldsymbol{W}$ be disjoint sets of variables in a causal graph $G$.
+We denote the graph $G$ after the an intervention $do(\boldsymbol{X})$, which has the effect of removing all edges into variables in $\boldsymbol{X}$, as $G_{\overline{\boldsymbol{X}}}$
+
+\paragraph{Rule 1: (adding or removing evidence)}
+
+Rule 1 allows us to remove (or insert) observational evidence from the right hand side of a conditional interventional distribution.
+It follows directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still applies after an intervention.
+
+If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
+\eqn{
+\label{eq:Do1}
+ P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
+}
+
+\begin{figure}[h]
+\centering
+\begin{subfigure}[t]{0.3\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:mutilatedOriginal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 2](4){$W$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (2)
+    	edge node {} (3)
+    (2) edge node {} (3)
+    (4) edge node {} (2);
+
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.5\textwidth}
+\centering
+\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
+\label{fig:mutilatedAfter}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$U$};
+\node[main node, below left=of 1](2){$x$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 2](4){$W$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge node {} (3)
+    (2) edge node {} (3);
+
+\end{tikzpicture}
+\end{subfigure}
+\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
+\label{fig:mutilatednet}
+\end{figure}
+
+\paragraph{Rule 2: (exchanging actions with observations)}
+Rule 2 describes when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening, $do(\boldsymbol{X}=\boldsymbol{x})$, have the same effect on the distribution over $\boldsymbol{Y}$.
+Let $G_{\underline{\boldsymbol{X}}}$ denote the causal graph $G$ with all edges \emph{leaving} $X$ removed.
+
+If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W})$ in $G_{\underline{\boldsymbol{X}}}$:
+\eqn{
+\label{eq:Do22}
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W})
+}
+
+The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$.
+Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
+However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
+Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
+If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths.
+This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).
+
+Equation \ref{eq:Do22} does not cover cases where acting on one set of variables allows us to replace acting on another set with conditioning (see figure \ref{fig:rule2_2}).
+The general form of rule 2 is given in equation \ref{eq:Do2}.
+
+If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W},\boldsymbol{Z})$ in $G_{\underline{\boldsymbol{X}}\overline{\boldsymbol{Z}}}$:
+\eqn{
+\label{eq:Do2}
+P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),
+do(\boldsymbol{Z}=\boldsymbol{z}),
+\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W})
+}
+
+\begin{figure}[h]
+\centering
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:rule2_orig}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$W$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)  edge  (3)
+    (2) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\overline{X}}$}
+\label{fig:rule2_after}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$W$};
+\node[main node, below left=of 1](2){$x$};
+\node[main node, below right=of 1](3){$Y$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1)	edge (3)
+    (2) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\underline{X}}$}
+\label{fig:rule2_indirect}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$W$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)
+    	edge  (3);
+
+\end{tikzpicture} 
+\end{subfigure}
+\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
+In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
+If we condition on $W$, we block this indirect path.}
+\label{fig:rule2}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:rule22_orig}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 1](4){$U$};
+\node[main node, above right=of 1](5){$V$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)
+    (2) edge (3)
+    (4) edge (1) edge (2)
+    (5) edge (1) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{after $do(Z)$, $G_{\overline{Z}}$}
+\label{fig:rule22_after}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 1](4){$U$};
+\node[main node, above right=of 1](5){$V$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)
+    (2) edge (3)
+    (4) edge (2)
+    (5) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\underline{X}\overline{Z}}$}
+\label{fig:rule22_indirect}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+
+ %nodes
+\node[main node](1){$z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, above left=of 1](4){$U$};
+\node[main node, above right=of 1](5){$V$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge  (2)
+    (4) edge (2)
+    (5) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\caption{An example of applying equation \ref{eq:Do2}.
+In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
+Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
+Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
+}
+\label{fig:rule2_2}
+\end{figure}
+
+\paragraph{Rule 3: (adding or removing actions)}
+This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
+A simple case of rule 3 is given in equation \ref{eq:Do3_1}.
+If $\boldsymbol{Y}$ is independent of $\boldsymbol{{X}}$ in $G$ after removing links entering $\boldsymbol{X}$ then can be no direct path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and any intervention on $\boldsymbol{X}$ will not affect $\boldsymbol{Y}$.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{{X}})$ in $G_{\boldsymbol{\overline{X}}}$:
+
+\eqn{
+\label{eq:Do3_1}
+\P{\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})} = \P{\boldsymbol{Y}}
+}
+
+The general case of rule 3 is easier to state by explicitly representing the intervention in the graphical model.
+Let $G^{\boldsymbol{\hat{X}}}$ denote the graph $G$ after adding a variable $\hat{X_i}$ as a parent of each variable $X_i \in \boldsymbol{X}$ (see figure \ref{fig:rule3_augmented}).
+The variable $\hat{X_i}$ can be thought of as representing the mechanism by which $X_i$ takes its value, either by being set via intervention or as a stochastic function of its other parents \citep{Koller2009}.
+
+if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\boldsymbol{\hat{X}}}_{\boldsymbol{\overline{Z}}}$:
+\eqn{
+\label{eq:Do3}
+P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
+}
+
+The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
+These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
+
+\begin{figure}[h]
+\centering
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{original network, $G$}
+\label{fig:rule3_original}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, below right=of 2](4){$W$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (4)
+    (4) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{augmented graph $G^{\hat{X}}$}
+\label{fig:rule3_augmented}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, below right=of 2](4){$W$};
+\node[main node, above left=of 2](5){$\hat{X}$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (4)
+    (4) edge (3)
+    (5) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$G_{\overline{X}}$}
+\label{fig:rule3_after}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed}]
+ %nodes
+\node[main node](1){$Z$};
+\node[main node, below left=of 1](2){$X$};
+\node[main node, below right=of 1](3){$Y$};
+\node[main node, below right=of 2](4){$W$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (3)
+    (2) edge (4)
+    (4) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\caption{Example application of equation \ref{eq:Do3}.
+$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
+We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
+\label{fig:rule3}
+\end{figure}
+
+\subsubsection{Identifiability}
+A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
+This is the identifiablity problem.
+It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
+A query is non-parametrically identifiable if it is identifiable without  assumptions about the functional form of the dependencies between variables in the graph.
+
+\vspace*{.3cm}
+\begin{definition}[Non-parametric identifiablity]
+Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
+A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
+\end{definition}
+
+The question of non-parametric identifiability is solved.
+The do calculus is complete \citep{Shpitser2006b,Huang2006}.
+A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
+There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
+
+\begin{enumerate}
+\item determine if the query can be translated into an expression involving only distributions over observed variables.
+In other words, determine if the query is identifiable given the assumptions encoded by the network, and
+\item if it is identifiable, return the required expression. 
+\end{enumerate}
+
+Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
+I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
+
+\begin{figure}[h]
+\centering
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{}
+\label{fig:id1}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3)
+    (4) edge (1) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{}
+\label{fig:id2}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$X$};
+\node[ob, right=of 1](2){$Z$};
+\node[ob, right=of 2](3){$Y$};
+\node[lt,above=of 2](4){$U$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2)
+    (2) edge (3)
+    (4) edge (1) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{}
+\label{fig:id3}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U_3$};
+\node[lt,above=of 2](5){$U_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (4) edge (1) edge (3)
+    (5) edge (2) edge (1);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\vspace*{.5cm}
+\caption{}
+\label{fig:nid1}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[lt](1){$U$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\vspace*{.5cm}
+\caption{}
+\label{fig:nid2_instrumental_variable}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[lt](1){$U$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[ob, above left=of 2](4){$Z$};
+ \path[every node/.style={font=\sffamily\small}]
+    (1) edge (2) edge (3)
+    (2) edge (3)
+    (4) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\vspace*{.5cm}
+\caption{}
+\label{fig:nid3}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U_3$};
+\node[lt,above=of 2](5){$U_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (4) edge (1) edge (3)
+    (5) edge (2) edge (1)
+    (1) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
+\label{fig:identify}
+\end{figure}
+
+Many interesting questions relating to identifiability remain open.
+What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
+What if we assume various restrictions on the functional form of the relationships between the variables?
+Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
+A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
+
+Although identifiability is a natural and important question to ask, it does not partition causal questions into solvable and unsolvable.
+Estimators for identifiable queries can be slow to converge and we may be able to obtain useful bounds on causal effects in cases where point estimates are not identified.
+
+\subsection{Defining causal effects} 
+So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
+This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set.
+From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.
+
+\eqn{
+\label{eqn:utility_max}
+x* =\argmax_x \Esub{y\sim\P{Y|do(X=x)}}{{\cal{U}}(y)}
+}
+
+Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
+There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
+For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
+
+\eqn{
+\label{eqn:ACE}
+ACE = \E{Y|do(X = 1)} - \E{Y|do(X=0)}
+}
+
+Assuming the expectations in equation \ref{eqn:ACE} are well defined, the ACE captures the shift in the mean outcome that arises from varying $X$.
+It does not capture changes in variance or higher moments of the distribution.
+The ACE can be generalised to non-discrete interventions by considering the effect on the expectation of $Y$ of an infinitesimal change in $x$.
+If $X$ is linearly related to $Y$ then the ACE is constant and equivalent to the corresponding coefficient in the linear structural equation model.
+
+\eqn{
+\label{eqn:ACE_continuous}
+ACE(x) = \frac{d}{dx}\E{Y|do(X=x)}
+}
+
+The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}.
+Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
+However we can personalise the average causal effect by stating it with respect to some observed context.
+I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
+
+\eqn{
+PCE(z) = \E{Y|do(X=1),z} - \E{Y|do(X=0),z}
+}
+
+In some cases we may be interested in the average causal effect for some sub-group of the population.
+A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
+This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.
+
+\eqn{
+ATT = \Esub{z\sim \P{Z|x=1}}{Y|do(X = 1)} - \Esub{z\sim \P{Z|x=1}}{Y|do(X=0)}
+}
+
+Causal effects can also be written in terms of counterfactuals.
+The ACE is $\E{\cf{Y}{1} - \cf{Y}{0}}$.
+We could estimate the ratio of expectations $\frac{\E{\cf{Y}{1}}}{\E{\cf{Y}{0}}}$ instead of the difference.
+However, the quantity   $\E{\frac{\cf{Y}{1}}{\cf{Y}{0}}}$ depends on the joint distribution over the counterfactual variables $(\cf{Y}{1},\cf{Y}{0})$ and thus cannot be computed from the interventional distribution.
+
+Another way of conceptualising causal effects is as a property indicating the strength of the causal link between two variables.
+This notion is complex to formalise when the relationship between variables is non-linear.
+Suppose $Y = X \oplus Z$ with $P(Z=1)=\frac{1}{2}$, the interventional distributions over $X$ are identical after marginalising out $Z$.
+\citet{Janzing2013} propose a number of postulates that a notion of causal strength could satisfy, demonstrate why previous measures fail these postulates and propose an alternative based on information flow.
+
+\subsection{Estimating causal effects by adjusting for confounding variables}
+\label{sec:causal_effect_under_ignorability}
+Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors.
+Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds.
+Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds.
+This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}.
+The setting is also referred to as unconfounded.
+
+\vspace*{0.3cm}
+\begin{theorem}[The backdoor criterion]\citep{Pearl2000}
+Let $\boldsymbol{X}$, $\boldsymbol{Z}$ and $\boldsymbol{Y}$ be disjoint sets of vertices in a causal graph $G$.
+If $\boldsymbol{Z}$ blocks (see Definition \ref{def:unblocked}) for every path from $X_i$ to $Y_j$ that contains a link into $X_i$, for every pair $(X_i \in \boldsymbol{X},Y_j \in \boldsymbol{Y})$, and no node in $\boldsymbol{Z}$ is a decedent of a node in $\boldsymbol{X}$ then the backdoor criterion is satisfied and;
+
+\eqn{
+\label{eqn:backdoor}
+\P{\boldsymbol{y}|do(\boldsymbol{x})} =\sum_{\boldsymbol{z}} \P{\boldsymbol{y}|\boldsymbol{x},\boldsymbol{z}}\P{\boldsymbol{z}}
+}
+\end{theorem}
+
+The backdoor criterion derives from rule 2 of the do-calculus.
+Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
+There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
+The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
+Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}.
+Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
+For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
+
+\begin{figure}
+\centering
+\begin{subfigure}[t]{0.48\textwidth}
+\centering
+\caption{There can be multiple valid adjustment sets. $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.}
+\label{fig:adjust_multiple}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](2){$X$};
+\node[ob, right=of 2](3){$Y$};
+\node[ob,above=of 3](4){$Z_2$};
+\node[ob,above=of 2](5){$Z_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (5) edge (2) edge (4)
+    (4) edge (3);
+\end{tikzpicture} 
+\end{subfigure}
+\begin{subfigure}[t]{0.48\textwidth}
+\centering
+\caption{Conditioning on $Z$ opens the backdoor path $X-U_1-Z-U_2-Y$}
+\label{fig:m-graph}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,lt/.style={latent},ob/.style={observed}]
+\node[ob](1){$Z$};
+\node[ob, below left=of 1](2){$X$};
+\node[ob, below right=of 1](3){$Y$};
+\node[lt,above=of 3](4){$U_3$};
+\node[lt,above=of 2](5){$U_1$};
+ \path[every node/.style={font=\sffamily\small}]
+    (2) edge (3)
+    (4) edge (1) edge (3)
+    (5) edge (2) edge (1);
+\end{tikzpicture} 
+\end{subfigure}
+\caption{}
+\label{fig:adjustment}
+\end{figure}
+
+\todo{fix this caption}
+
+Given that a set of variables $\boldsymbol{Z}$ satisfies the backdoor criterion (or equivalently the conditional ignorability assumption), the interventional distribution is asymptotically identifiable and can be estimated from equation \ref{eqn:backdoor}.
+The expected value of $Y$ after the intervention $do(X=x)$ is given by equation \ref{eqn:expected_post_intervention} and the average causal effect for a binary intervention $x \in \set{0,1}$ is given by equation \ref{eqn:ace_ignoreable}.
+
+\eqn{
+\label{eqn:expected_post_intervention}
+\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}}
+}
+
+\eqn{
+\label{eqn:ace_ignoreable}
+ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
+}
+
+Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata.
+However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
+There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
+The key approaches are based on matching on covariates, propensity score methods and regression.
+We now examine these approaches from a machine learning perceptive.
+
+In standard supervised learning, we have a training set $(\vb{x_1},y_1),...,(\vb{x_n},y_n)$ assumed to be sampled i.i.d from an unknown distribution $\P{\vb{x},y} =\P{\vb{x}}\P{y|\vb{x}}$.
+The goal is to select a hypothesis $h \in \mathcal{H} : \mathcal{X} \rightarrow \mathcal{Y}$ such that, on unseen data $ \sim \P{\vb{x},y}$, $h(\vb{x})$ is close (by some metric) in expectation to $y$.
+In other words we wish to minimise the generalisation error $E_{out}(h)$,
+
+\eqn{
+E_{out}(h) = \Esub{(\vb{x},y)\sim \P{\vb{x}}\P{y|\vb{x}}}{L(h(\vb{x}),y)}
+}
+
+We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
+We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
+
+\eqn{
+E_{in}(h) = \frac{1}{n}\sum_{i=1}^n L(h(\vb{x}_i),y_i)
+}
+
+The VC-dimension of the hypothesis space provides (typically loose) bounds on the probability that $E_{out} >> E_{in}$.
+However, in practice, the generalisation error is usually estimated empirically from a hold-out set of the sample that was not used to train the model, or via cross-validation.
+
+In the causal effect estimation under ignorability, we have training data $(\vb{x}_1,\vb{z}_1,y_1),...,(\vb{x}_n,\vb{z}_n,y_n)$ sampled i.i.d from $\P{\vb{z}}\P{\vb{x}|\vb{z}}\P{y|\vb{x},\vb{z}}$.
+Estimating $\E{Y|do(\vb{X}=\vb{x})}$ corresponds to selecting a hypothesis $h \in \mathcal{H}:\mathcal{X} \times \mathcal{Z} \rightarrow \mathcal{Y}$ that minimises;
+
+\eqn{
+\label{eqn:causal_covariate_shift}
+E_{out} &= \Esub{(\vb{x},\vb{z},y)\sim \delta(\vb{x}-\vb{x'})\P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},\\
+&= \Esub{(\vb{z},y)\sim \P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},
+}
+
+\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
+If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
+\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
+Take the case where we have a binary intervention $x \in \set{0,1}$.
+Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$.
+More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.
+
+\subsubsection{Regression}
+
+The regression approach is to learn a function that is a good approximation to the output surface $\E{Y|X,Z}$.
+Let $f_1(z)= \E{Y|X=1,Z=z}$.
+The expectation of $Y$ after the intervention $X=1$ is then obtained by taking the expectation with respect to $Z$, $\E{Y|do(X=1)} = \Esub{z \sim \P{Z}}{\E{Y|X=1,z}}$.
+We can learn a parametric regression model $\hat{f}_1(z)$ via empirical risk minimisation.
+
+\eqn{
+\hat{f}_1(z) = h_1(z;\hat{\theta}_{obs}) \text{, where } \hat{\theta}_{obs} = \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)  \right]
+}
+
+This estimator is consistent with respect to the observational distribution.
+As the sample size tends to infinity, $\hat{\theta}_{obs}$ approaches the parameter within the hypothesis space that minimises the expected loss given data sampled from the observational distribution.
+
+\eqn{
+\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z|x=1}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
+}
+
+If the model is correctly specified such that $f_1(z) = h_1(z;\theta^*)$ for some  $\theta^* \in \Theta$ then the empirical risk minimisation estimate is consistent with respect to the loss over any distribution of $Z$  \citep{Sugiyama2007}, including the interventional one.
+
+\eqn{
+\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \theta^* = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
+}
+
+The average causal effect can then be estimated by,
+
+\eqn{
+\label{eqn:cef_empirical_regression}
+\hat{\tau}_{reg} = \sum_{i=1}^n \left(\hat{f}_1(z_i) - \hat{f}_0(z_i)\right)
+}
+
+Regression thus has a direct causal interpretation if the parametric model is correctly specified and the covariates included form a valid backdoor adjustment set for the treatment variable of interest in the corresponding structural equation model.
+
+\subsubsection{Propensity scores}
+If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
+The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
+\caption{
+Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
+In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
+If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$.
+We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points).
+In both cases the linear model is a good fit to the data.
+However, the resulting estimate of the causal effect is very poor for the lower values of $z$.
+}
+\label{fig:covariate_shift_adjust}
+\end{figure}
+
+A general approach to estimating the expectation of some function $f(\cdot)$ with respect to data from some distribution $\P{\cdot}$, when we have data sampled from a different distribution $\Q{\cdot}$ is importance sampling \citep{horvitz1952generalization,Koller2009}.
+\eqn{
+\label{eqn:importance_sampling}
+\Esub{\vb{v} \sim \P{\vb{v}}}{f(v)} = \Esub{\vb{v} \sim \Q{\vb{v}}}{f(v)\frac{\P{\vb{v}}}{\Q{\vb{v}}}}
+}
+
+This importance weighting approach can be applied to the covariate shift/average causal effect problem by weighting the terms in the empirical risk minimisation estimator \citep{Sugiyama2007}.
+
+\eqn{
+\label{eqn:importance_weighted_erm}
+\hat{\theta}_{iw} &= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{\P{z_i}\delta(x_i-1)}{\P{z_i}\P{x_i=1|z_i}}  \right] \\
+&= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{1}{e(z_i)}  \right],
+}
+
+where $e(\vb{z})$ is the propensity score, defined by \citep{Rosenbaum1983};
+\eqn{
+e(\vb{z})\equiv \P{x=1|\vb{z}}
+}
+
+The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
+Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
+
+The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$.
+\citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$.
+This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$.
+Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
+In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
+However in general, it must be estimated from data.
+Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
+\citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
+
+\eqn{
+\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}} = \Esub{z\sim \P{\vb{Z|x}}}{\E{Y|x,\vb{z}}\frac{1}{e(\vb{z})}}
+}
+
+\eqn{
+\label{eqn:simple_inverse_propensity}
+\hat{\tau}_{ip} = \frac{1}{n}\sum_{i=1}^n \left(\frac{\ind{x_i=1}y_i}{e(\vb{z}_i)} -\frac{\ind{x_i=0}y_i}{1-e(\vb{z}_i)}\right)
+}
+
+\subsubsection{Matching}
+There is a straightforward connection between matching and regression for causal effect estimation.
+If $h \in \mathcal{H} \implies h + a \in \mathcal{H}$ for any constant $a$ and $\hat{f}$ is selected by minimising empirical risk with an $L_2$ loss then $\frac{1}{n}\sum_{i=1}^n \ind{x_i=1}\hat{f}_1(z_i) = \frac{1}{n}\sum_{i=1}^n \ind{x_i=1}y_i$ \footnote{\citep{Imbens2004} state this holds for most implementations}, and equation \ref{eqn:cef_empirical_regression} can be re-written as:
+
+\eqn{
+\label{eqn:cef_regression_missing}
+\hat{\tau}_{reg} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \hat{f}_0(z_i)\right) +\ind{x_i=0}\left(\hat{f}_1(z_i) - y_i \right)\right]
+}
+
+This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
+For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
+Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
+\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
+
+\eqn{
+\hat{\tau}_{match} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \frac{1}{k}\sum_{j \in J_k(i)}y_j\right) +\ind{x_i=0}\left(\frac{1}{k}\sum_{j \in J_k(i)}y_j - y_i \right)\right]
+}
+
+This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
+There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
+Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
+An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.
+
+The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
+A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
+
+The significance of this should not be underestimated.
+Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
+With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
+\citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem.
+There is relatively little theory on model selection for estimating the propensity score.
+To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion.
+It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}.
+With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
+I have found no examples of such an approach.
+
+The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
+Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
+The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
+
+\section{Bayesian reasoning and causality}
+\label{sec:bayesian_causality}
+The Bayesian approach to modelling encapsulates a very general approach to combining prior knowledge and assumptions with data to draw inferences.
+You can specify a complex model, capturing  your assumptions about the underlying mechanisms of your system, describe what you are uncertain about with parameters (and priors over their values) and then update your beliefs about the values of these parameters based on the observed data.
+If you believe the model to be a mechanistic representation of the way world works then you can use it to infer the consequences of interventions.
+
+Why then do we need a special framework for causal inference; can't we simply encode all the assumptions required to infer the outcome of an intervention in a given system in a Bayesian model?
+The short answer is we could.
+However, causal graphical models represent a useful level of abstraction for many problems.
 Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
-The power of defining an intervention in this way stems from the number of things that are invariant between the two processes.
+We do not need to specify the functional relationships between the variables or priors over their distributions.
+The power of defining an intervention in this way stems from the number of things that are assumed invariant between the two processes.
 All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
+The causal framework also provides, via the do-calculus, a rigorous method to draw a boundary around which variables we must measure and model to identify causal effects.
 
@@
 
-A causal Bayesian network represents much more information than a Bayesian network with identical structure.
-A causal network encodes all possible interventions that could be specified with the do-notation.
-For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
-The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
-After each two variable intervention, we have a distribution over one variables which requires one parameter.
-This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
-
-\subsubsection{Limitations of causal Bayesian networks}
-A number of criticisms have been levelled at this approach to modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}.
-However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions.
-For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
-But, what if some people failed to take the prescribed treatment?
-We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
-Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
+We can also combine causal graphical models with Bayesian approaches to estimation.
+For example, we can use a causal graphical model to identify a set of variables that form a valid adjustment set and then use Bayesian regression to estimate posterior distribution over a given causal effect.
+However, in doing so we need to be careful about the way in which we specify priors as I highlight in the next section.
+
+\subsection{Careful with that prior}
+
+Suppose we have a data generated by the linear Gaussian Bayesian network below.
+The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$.
+The variable $U$ is latent.
+
+\eqn{
+\P{U} &= N(0 , v_u) \\
+\label{eqn:careful_with_priors}
+\P{Z|U} &= N(w_{zu}U , v_z) \\
+\P{X|Z} &= N(w_{xz}Z , v_x) \\
+\P{Y|U,Z,X} & = N(w_{yu}U +w_{yz}Z + w_{yx}X , v_y)
+}
 
@@
 \centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Prescribed Treatment};
-\node[node, below right=of 1](2){Actual Treatment};
-\node[node, right=of 2](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(1) edge (2)
-	(2) edge (3) 
-	(4) edge (3) edge (2);
-\end{tikzpicture}
-\caption{Randomised treatment with imperfect compliance}
-\label{fig:randomized_imperfect_compliance}
-\end{figure}
-
-Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables.
-Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology.
-We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
-However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
-Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
-Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
-\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
-
-\subsection{Counterfactuals}
-\label{sec:counterfactuals}
-
-The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
-Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
-
-In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would have happened if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would have happened if they got the placebo, $\cf{y_{i}}{0}$.
-The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, there are units that can be assumed to be identical or that will revert exactly to their initial state some time after treatment.
-This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
-
-Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo.
-Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate (from an experimental or observational study):
-\begin{itemize}
-\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
-\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
-\end{itemize}
-
-Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
-For those who did not take the treatment, the observed outcome is the same as the outcome \emph{had} they not taken the treatment.
-Equivalently stated:
-
-\eq{
-\P{Y^{0}|X=0}&= \P{Y|X=0}\\
-\P{Y^{1}|X=1}&=\P{Y|X=1}
-}
-
-If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
-
-\eq{
-\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
-\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
-}
-
-This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
-
-\eq{
-\P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
-}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}.
-They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
-In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e.
-$X \nci \cf{Y}{0}$.
-Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e.
-$X \nci \cf{Y}{1}$.
-A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
-
-\vspace*{.3cm}
-\begin{theorem}[Ignorability \citep{Rosenbaum1983, Pearl2000}] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
-
-\eqn{
-\label{eqn:counterfactual1}
-\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
-\label{eqn:counterfactual2}
-\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
-}
-\end{theorem}
-
-Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
-Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
-
-Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}.
-This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
-As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
-The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
-SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
-
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-This makes it quite easy to (accidentally) make statements about counterfactuals that cannot be tested with empirical data.
-Consider the following example based on \citet{Dawid2000}.
-Again we have a drug where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$.
-Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal with equal variance (for simplicity).
-
-% vector with mean mu_t,mu_c and a covariance matrix
-\eqn{
-P(\cf{Y}{1},\cf{Y}{0}) \sim N( 
-         \begin{bmatrix}
-           \mu_{1} \\
-           \mu_2 \\
-         \end{bmatrix},
-         \begin{bmatrix}
-           \sigma^2 & \rho \sigma^2 \\
-           \rho \sigma^2 & \sigma^2 \\
-         \end{bmatrix}
-         )
-}
-
-The distribution of \emph{individual treatment effects}, $\tau = \cf{Y}{1} - \cf{Y}{0}$ is also normal,
-
-\eqn{
-P(\tau) = N(\mu_1 - \mu_0,2\sigma^{2}(1-\rho))
-}
-
-\begin{figure}[h]
-\captionsetup[subfigure]{position=b}
-\centering
-\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$.
-The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated.
-\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
-\hspace{0.02\textwidth}
-\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
-{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
-\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
-\label{fig:metaphysical_distribution_difference}
-\end{figure}
-
-From a (large) randomised controlled trial we can estimate the marginal distributions over the counterfactual variables, see figure \ref{fig:metaphys_a}.
-These represent the distributions we would expect over the outcome $Y$ if everyone were treated or not treated respectively.
-However, the distribution over the individual causal effects depends on $\rho$, see figure \ref{fig:metaphys_b}.
-The key problem is that we can never observe the joint distribution over $\cf{Y}{1}$ and $\cf{Y}{0}$.
-As a result, $\rho$ and thus the variance of $\tau$ is not identifiable, even from experimental data.
-\citet{Dawid2000} argues that we should avoid using counterfactuals as they are defined in terms of (metaphysical) individual causal effects.
-He further points out that the interventional distributions in figure \ref{fig:metaphys_a}, along with a loss function, contain all the information required to decide how to treat a new patient.
-
-This result is unintuitive.
-It seems on the face of it that the distribution of individual causal effects is relevant to our decision making.
-If $\rho = 1$ then almost everyone benefits slightly from the treatment whilst if $\rho=0$, there is a wide range, with some people benefiting a lot and others suffering significant harm.
-This confusion can be resolved by thinking about personalised rather than individual causal effects.
-It is entirely possible that potentially observable characteristics (such as gender, age, genetics, etc) affect how people will respond to the treatment.
-We can partition people into sub-populations on the basis of these characteristics and measure different \emph{personalised} causal effects for each group.
-The variance of the potential outcome distributions $f(\cf{Y}{1})$ and $f(\cf{Y}{0})$ provides bounds on how much can be gained from further personalisation.
-The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).
-
-One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
-This is certainly convenient, if rather imprecise.
-However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention.
-It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).
-
-\subsection{Structural Equation models}
-\label{sec:SEM}
-
-Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Uncertainties are not inherent but arise from unmeasured variables.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \citep{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(\epsilon_x)$};
-\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:ranomized_experiment}}
-\label{fig:sem_randomized_treatment}
-\end{figure}
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random.
-The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
-The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
-
-Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$.
-The function $f_y$ and distribution over $\epsilon_y$ does not change.
-This results in the interventional distribution, \footnote{We have assumed the variables are discrete only for notational convenience}
-
-\eqn {
-\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
-}
-
-The observational distribution of $Y$ given $X$ is,
-
-\eqn{
-\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
-& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
-}
-
-The interventional distribution is the same as the observational one.
-The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
-
-The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
-Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
-
-\eqn{
-\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
-\label{eqn:sem:adjusting}
-&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
-}
-Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivalent to the observational distribution given by:
-
-\eqn{
-\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
-}
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(Z,\epsilon_x)$};
-\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
-\node[lt, above=of 6](7){$\epsilon_z$};
-\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2)
-	(7) edge (6)
-	(6) edge (2) edge (3);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:adjusting}}
-\label{fig:sem:preschool}
-\end{figure}
-
-Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
-A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
-
-\subsection{Comparing and unifying the models}
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
-In other words, the network is a causal Bayesian network.
-All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
-Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}).
-Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
-We can also connect this to counterfactuals via:
-
-\begin{equation}
-\begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
-\end{aligned}
-\end{equation}
-
-The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
-
-\begin{equation}
-\label{eq:weakignore}
-X \ci Y^{0} \text{ and } X \ci Y^{1}
-\end{equation}
-
-It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}.
-The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
-For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores \citep{Rosenbaum1983} to estimate the effect.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
-These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
-
-In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
-The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field.
-Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
-In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
-
-\section{What does a causal model give us? Resolving Simpson's paradox}
-
-We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
-The following example is adapted from \citet{Pearl2000}.
-Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
-She keeps track of which medication her patients choose and whether or not the treatment is successful.
-She obtains the results in table \ref{tab:simpson1}.
-
-\begin{table}[h]
-\caption{Treatment results}
-\label{tab:simpson1}
-\center
-\begin{tabular}{c c c c c}
-Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9} 
-A &  87 & 13 & 100 & 87\%\\
-\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Drug A appears to perform better.
-However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
-
-\begin{table}[h]
-\caption{Treatment results by gender}
-\label{tab:simpson2}
-\center
-\begin{tabular}{c c c c c c}
-Gender & Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9}
-M & A & 12 & 8 & 20 & 60\%\\
-M & B & 56 & 24 & 80 & 70\%\\
-\rowcolor[gray]{0.9}
-F & A & 75 & 5 & 80 & 94\%\\
-F & B & 19 & 1 & 20 & 95\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Once the data is broken down by gender, drug B looks better for both men \emph{and} women.
-Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must recommend which to subsidise under a national health scheme).
-Should she choose A or B?
-The ambiguity in this question lies at the heart of Simpson's paradox.
-How does causal modelling resolve the paradox?
-The key is that the doctor is trying to choose between \emph{interventions}.
-She wants to know what the success rate will be if she changes her practice to give all the patients one drug, rather than allowing them to choose as currently occurs.
-
-Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
-The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
-Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
-Some assumptions about the causal relationships between the variables are required.
-In this example, it seems reasonable to conclude that gender may affect the treatment chosen and the outcome.
-Assuming there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Gender};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (2) edge (3) 
-	(2) edge (3);
-\end{tikzpicture}
-\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
-\label{fig:simpsons_paradox}
-\end{figure}
-
-With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
-Knowing a patient took drug A indicates they are more likely to be female.
-When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
-However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
-So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
-Drug B is the better choice.
-
-\eqn{
-\P{Y|do(T)} = \P{Y|T,female}\P{female}+ \P{Y|T,male}\P{male}
-}
-
-Is the solution to Simpson's paradox to always to break down the data by as many variables as possible?
-No.
-Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
-This is a drug designed to prevent heart disease.
-One pathway to doing so might well be to lower blood pressure.
-Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
-It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
-Now, however, table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood pressure control and better overall outcomes.
-In this setting $\P{Y|do(T)} = \P{Y|T}$ and drug A is the better choice.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Blood Pressure};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (3) 
-	(2) edge (3)
-	(2) edge (1);
-\end{tikzpicture}
-\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
-\label{fig:simpsons_paradox2}
-\end{figure}
-
-Note that we have not changed the data itself, only the description of the variables that it is associated with.
-This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
-From a purely statistical viewpoint there is no paradox.
-The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
-The paradox only arises when we attempt to use the data to select an intervention and is resolved when we apply a causal approach to do so.
-
-\eqn{
-\label{eqn:simpsons_paradox}
-\exists \set{N_1,...N_4, N_1'...N_4'} \in \mathbb{N} : \; 
-\textcolor{green}{\frac{N_1}{N_2}< \frac{N_1'}{N_2'}}, \;\; 
-\textcolor{blue}{ \frac{N_3}{N_4}<\frac{N_3'}{N_4'}}
- \text{ and } \frac{N_1+N_3}{N_2+N_4} > \frac{N_1'+N_3'}{N_2'+N_4'}
-}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.4]{figures/simpson_vector}
-\caption{Simpson's reversal visualised.
-The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
-However, when we sum them, the ratio is steeper for the un-primed variables.}
-\label{fig:simpsons_vector}
-\end{figure}
-
-There are many other plausible causal graphs for both scenarios above.
-Perhaps income affects drug choice as well as gender, or gender might affect treatment choice and blood pressure control given treatment, etc.
-Causal modelling provides a powerful tool to specify such assumptions and to determine how to estimate causal effects for a given model see section \ref{sec:causal effect estimation}.
-
-\section{Causal effect estimation} 
-\label{sec:causal effect estimation}
-We can roughly categorise the problems studied within causal inference into two groups, causal effect estimation and causal discovery.
-In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
-The goal is then to estimate the effect of an intervention or range of interventions in the system.
-Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}.
-Almost every time someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
-
-Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
-This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
-Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
-An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
-\todo{note that I focus on problems where causal structure known, reference rewiew on structure learning}
-
-Estimating causal effects from observational data comes down to determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
-We did this on an ad-hoc basis to resolve the examples discussed in chapter \ref{chap:causal_models}.
-In this chapter we describe a principled approach to mapping observational quantities to interventional ones and discuss some of the key issues involved in estimating such expressions from finite sample data.
-We assume the basic structure of the graph is known.
-That is, we assume that we can draw a network containing (at a minimum):
-
-\begin{itemize}
-\item the target/outcome variable we care about,
-\item the focus/treatment variables on which we are considering interventions,
-\item any variables which act to confound two or more of the other variables we have included, and
-\item any links between variables we have included. 
-\end{itemize}
-
-Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
-For example, consider estimating the impact of schooling on wages.
-Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
-Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
-
-How can the structure of the causal graph be leveraged to compute interventional distributions from observational ones?
-Given the graph corresponding to the observational distribution, the graph after any intervention can be obtained by removing any links into variables directly set by the intervention.
-The joint interventional distribution is the product of the factors associated with the interventional graph, as given by the truncated product formula \ref{eq:truncatedproduct}.
-If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
-However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
-
-The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
-By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \citep{Pearl2000}.
-The do calculus consists of three rules.
-They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.
-
-\subsection{Independence in Bayesian networks: D-separation}
-\label{sec:dsep}
-
-Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
-Therefor, in this section, we briefly review the key properties of Bayesian networks.
-A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
-Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
-
-\vspace{.3cm}
-\begin{theorem}
-\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
-\end{theorem}
-
-The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$.
-D-separation is an algorithm that extends the local Markov property to find these additional independencies.
-It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
-
-The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
-From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
-Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
-In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
-In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
-
-The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
-The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
-As an example, consider a scholarship available to female or disadvantaged students.
-Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship.
-There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
-However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y| Z$}
-\label{fig:dseptripple1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$X$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Y$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$X$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, below left =of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-        edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|\{\}$}
-\label{fig:dseptripple4}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, above left =of 1](2){$X$};
-\node[main node, above right =of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (1)
-    (3) edge node {} (1);	
-\end{tikzpicture}
-\end{subfigure}
-\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
-\label{fig:dseptriples}
-\end{figure}
-
-\vspace{.3cm}
-\begin{definition}[unblocked path]
-\label{def:unblocked}
-A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
-It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
-\end{definition}
-
-\vspace{.3cm}
-\begin{definition}[d-separation]
-\label{def:dsep}
-The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, there are no unblocked paths from any $X \in \boldsymbol{X}$ to any $Y \in \boldsymbol{Y}$ after conditioning on $\boldsymbol{Z}$.
-\end{definition}
-
-\vspace{.3cm}
-\begin{theorem}[d-separation and conditional independence]
-\label{thrm:dsep_independence}
-If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in a Bayesian network $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
-Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected (not d-separated) given $\boldsymbol{Z}$ then it is possible to construct a distribution $P'$ that factorises over $G$ in which they are dependent.
-\end{theorem}
-
-Theorem \ref{thrm:dsep_independence} says that independencies implied by d-separation on a graph hold in every distribution that can be factored over that graph and that if ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in \emph{all} distributions that can be factored over $G$ then they are d-separated in $G$.
-If we denote the independencies implied by d-separation in a graph by $\indep{G}$ and the set of independencies in a distribution by $\indep{P}$ then $\indep{G} \subseteq \indep{P}$.
-
-If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
-However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
-A particular case in which this occurs is when there are deterministic relationships between variables.
-If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
-This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
-This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
-D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
-
-\subsection{The Do Calculus} 
-\label{sec:do_calculus}
-
-The do-calculus is a set of three rules \citep{Pearl1995} that can be applied to simplify the expression for an interventional distribution.
-If by repeated application of the do-calculus, along with standard probability transformations, we can obtain an expression containing only observational quantities then we can use it to estimate the interventional distribution from observational data.
-Let $\boldsymbol{X}$,$\boldsymbol{Y}$,$\boldsymbol{Z}$ and $\boldsymbol{W}$ be disjoint sets of variables in a causal graph $G$.
-We denote the graph $G$ after the an intervention $do(\boldsymbol{X})$, which has the effect of removing all edges into variables in $\boldsymbol{X}$, as $G_{\overline{\boldsymbol{X}}}$
-
-\subsubsection{Rule 1: (adding or removing evidence)}
-
-Rule 1 allows us to remove (or insert) observational evidence from the right hand side of a conditional interventional distribution.
-It follows directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still applies after an intervention.
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:mutilatedOriginal}
 \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2);
-
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.5\textwidth}
-\centering
-\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
-\label{fig:mutilatedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{subfigure}
-\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
-\label{fig:mutilatednet}
-\end{figure}
-
-\subsubsection{Rule 2: (exchanging actions with observations)}
-Rule 2 describes when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening, $do(\boldsymbol{X}=\boldsymbol{x})$, have the same effect on the distribution over $\boldsymbol{Y}$.
-Let $G_{\underline{\boldsymbol{X}}}$ denote the causal graph $G$ with all edges \emph{leaving} $X$ removed.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W})$ in $G_{\underline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W})
-}
-
-The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$.
-Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
-However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
-Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
-If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths.
-This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).
-
-Equation \ref{eq:Do22} does not cover cases where acting on one set of variables allows us to replace acting on another set with conditioning (see figure \ref{fig:rule2_2}).
-The general form of rule 2 is given in equation \ref{eq:Do2}.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W},\boldsymbol{Z})$ in $G_{\underline{\boldsymbol{X}}\overline{\boldsymbol{Z}}}$:
-\eqn{
-\label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),
-do(\boldsymbol{Z}=\boldsymbol{z}),
-\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W})
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule2_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)  edge  (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule2_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1)	edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}}$}
-\label{fig:rule2_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    	edge  (3);
-
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
-In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
-If we condition on $W$, we block this indirect path.}
-\label{fig:rule2}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule22_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
+  thick,main node/.style={observed},lt/.style={latent}]
 
@@
 \node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
+\node[lt, right=of 1](4){$U$};
+
  \path[every node/.style={font=\sffamily\small}]
     (1) edge (2)
-    (2) edge (3)
-    (4) edge (1) edge (2)
-    (5) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{after $do(Z)$, $G_{\overline{Z}}$}
-\label{fig:rule22_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}\overline{Z}}$}
-\label{fig:rule22_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of applying equation \ref{eq:Do2}.
-In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
-Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
-Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
-}
-\label{fig:rule2_2}
-\end{figure}
-
-\subsubsection{Rule 3: (adding or removing actions)}
-This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-A simple case of rule 3 is given in equation \ref{eq:Do3_1}.
-If $\boldsymbol{Y}$ is independent of $\boldsymbol{{X}}$ in $G$ after removing links entering $\boldsymbol{X}$ then can be no direct path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and any intervention on $\boldsymbol{X}$ will not affect $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{{X}})$ in $G_{\boldsymbol{\overline{X}}}$:
-
-\eqn{
-\label{eq:Do3_1}
-\P{\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})} = \P{\boldsymbol{Y}}
-}
-
-The general case of rule 3 is easier to state by explicitly representing the intervention in the graphical model.
-Let $G^{\boldsymbol{\hat{X}}}$ denote the graph $G$ after adding a variable $\hat{X_i}$ as a parent of each variable $X_i \in \boldsymbol{X}$ (see figure \ref{fig:rule3_augmented}).
-The variable $\hat{X_i}$ can be thought of as representing the mechanism by which $X_i$ takes its value, either by being set via intervention or as a stochastic function of its other parents \citep{Koller2009}.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\boldsymbol{\hat{X}}}_{\boldsymbol{\overline{Z}}}$:
-\eqn{
-\label{eq:Do3}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-}
-
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
-These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule3_original}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{augmented graph $G^{\hat{X}}$}
-\label{fig:rule3_augmented}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
-\node[main node, above left=of 2](5){$\hat{X}$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3)
-    (5) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule3_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Example application of equation \ref{eq:Do3}.
-$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
-We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
-\label{fig:rule3}
-\end{figure}
-
-\subsection{Identifiability}
-A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
-This is the identifiablity problem.
-It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
-A query is non-parametrically identifiable if it is identifiable without  assumptions about the functional form of the dependencies between variables in the graph.
-
-\vspace*{.3cm}
-\begin{definition}[Non-parametric identifiablity]
-Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
-A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
-\end{definition}
-
-The question of non-parametric identifiability is solved.
-The do calculus is complete \citep{Shpitser2006b,Huang2006}.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
-There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
-
-\begin{enumerate}
-\item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network, and
-\item if it is identifiable, return the required expression. 
-\end{enumerate}
-
-Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
-I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
+    	edge (3)
     (2) edge (3)
     (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$X$};
-\node[ob, right=of 1](2){$Z$};
-\node[ob, right=of 2](3){$Y$};
-\node[lt,above=of 2](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid2_instrumental_variable}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[ob, above left=of 2](4){$Z$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1)
-    (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
-\label{fig:identify}
+
+\end{tikzpicture} 
+\caption{ }
+\label{fig:careful_with_priors}
 \end{figure}
 
-Many interesting questions relating to identifiability remain open.
-What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
-What if we assume various restrictions on the functional form of the relationships between the variables?
-Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
-A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
-
-Although identifiability is a natural and important question to ask, it does not partition causal questions into solvable and unsolvable.
-Estimators for identifiable queries can be slow to converge and we may be able to obtain useful bounds on causal effects in cases where point estimates are not identified.
-
-\subsection{Estimation}
-\subsubsection{Defining causal effects} 
-So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
-This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set.
-From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.
-
-\eqn{
-\label{eqn:utility_max}
-x* =\argmax_x \Esub{y\sim\P{Y|do(X=x)}}{{\cal{U}}(y)}
-}
-
-Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
-There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
-For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
-
-\eqn{
-\label{eqn:ACE}
-ACE = \E{Y|do(X = 1)} - \E{Y|do(X=0)}
-}
-
-Assuming the expectations in equation \ref{eqn:ACE} are well defined, the ACE captures the shift in the mean outcome that arises from varying $X$.
-It does not capture changes in variance or higher moments of the distribution.
-The ACE can be generalised to non-discrete interventions by considering the effect on the expectation of $Y$ of an infinitesimal change in $x$.
-If $X$ is linearly related to $Y$ then the ACE is constant and equivalent to the corresponding coefficient in the linear structural equation model.
-
-\eqn{
-\label{eqn:ACE_continuous}
-ACE(x) = \frac{d}{dx}\E{Y|do(X=x)}
-}
-
-The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}.
-Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
-However we can personalise the average causal effect by stating it with respect to some observed context.
-I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
-
-\eqn{
-PCE(z) = \E{Y|do(X=1),z} - \E{Y|do(X=0),z}
-}
-
-In some cases we may be interested in the average causal effect for some sub-group of the population.
-A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
-This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.
-
-\eqn{
-ATT = \Esub{z\sim \P{Z|x=1}}{Y|do(X = 1)} - \Esub{z\sim \P{Z|x=1}}{Y|do(X=0)}
-}
-
-Causal effects can also be written in terms of counterfactuals.
-The ACE is $\E{\cf{Y}{1} - \cf{Y}{0}}$.
-We could estimate the ratio of expectations $\frac{\E{\cf{Y}{1}}}{\E{\cf{Y}{0}}}$ instead of the difference.
-However, the quantity   $\E{\frac{\cf{Y}{1}}{\cf{Y}{0}}}$ depends on the joint distribution over the counterfactual variables $(\cf{Y}{1},\cf{Y}{0})$ and thus cannot be computed from the interventional distribution.
-
-Another way of conceptualising causal effects is as a property indicating the strength of the causal link between two variables.
-This notion is complex to formalise when the relationship between variables is non-linear.
-Suppose $Y = X \oplus Z$ with $P(Z=1)=\frac{1}{2}$, the interventional distributions over $X$ are identical after marginalising out $Z$.
-\citet{Janzing2013} propose a number of postulates that a notion of causal strength could satisfy, demonstrate why previous measures fail these postulates and propose an alternative based on information flow.
-
-\subsubsection{Estimating causal effects by adjusting for confounding variables}
-\label{sec:causal_effect_under_ignorability}
-Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors.
-Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds.
-Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds.
-This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}.
-The setting is also referred to as unconfounded.
-
-\vspace*{0.3cm}
-\begin{theorem}[The backdoor criterion]\citep{Pearl2000}
-Let $\boldsymbol{X}$, $\boldsymbol{Z}$ and $\boldsymbol{Y}$ be disjoint sets of vertices in a causal graph $G$.
-If $\boldsymbol{Z}$ blocks (see Definition \ref{def:unblocked}) for every path from $X_i$ to $Y_j$ that contains a link into $X_i$, for every pair $(X_i \in \boldsymbol{X},Y_j \in \boldsymbol{Y})$, and no node in $\boldsymbol{Z}$ is a decedent of a node in $\boldsymbol{X}$ then the backdoor criterion is satisfied and;
-
-\eqn{
-\label{eqn:backdoor}
-\P{\boldsymbol{y}|do(\boldsymbol{x})} =\sum_{\boldsymbol{z}} \P{\boldsymbol{y}|\boldsymbol{x},\boldsymbol{z}}\P{\boldsymbol{z}}
-}
-\end{theorem}
-
-The backdoor criterion derives from rule 2 of the do-calculus.
-Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
-There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
-The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
-Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}.
-Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
-For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
-
-\begin{figure}
-\centering
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{There can be multiple valid adjustment sets. $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.}
-\label{fig:adjust_multiple}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](2){$X$};
-\node[ob, right=of 2](3){$Y$};
-\node[ob,above=of 3](4){$Z_2$};
-\node[ob,above=of 2](5){$Z_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (5) edge (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{Conditioning on $Z$ opens the backdoor path $X-U_1-Z-U_2-Y$}
-\label{fig:m-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{}
-\label{fig:adjustment}
+As each variable is a linear function of its parents, with Gaussian noise, resulting in a joint distribution $\P{U,Z,X,Y}$ that is multivariate normal.
+Marginalising out $U$ and conditioning on $X$ and $Z$ yields,
+
+\eqn{
+Y \sim N(w_{yx} X + \beta Z, \epsilon) 
+}
+
+where
+
+\eqn{
+\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and}\\
+\epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
+}
+
+The causal effect of $X$ on $Y$ is identifiable (even without the linear Gaussian assumptions) as $Z$ satisfies the backdoor criterion and the expectation of the coefficient for $X$ in a standard OLS regression of $Y$ against $X$ and $Z$ is $w_{yx}$.
+However, the causal effect of $Z$ on $Y$ is not identifiable due the presence of the unobserved confounder $U$.
+The coefficient $\beta$ captures both the causal relationship between $Z$ and $Y$, $w_{yz}$ and the indirect relationship through $U$, which we cannot separate without observing $U$.
+
+If we do a Bayesian regression of $X$ and $Z$ against $Y$ and use intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ we are likely to select something centered around $w_{yz}$, but this is not correct and can introduce bias into our estimate of the coefficient for $w_{yx}$, as we demonstrate in figure \ref{fig:careful-prior-example}.
+The prior for $\beta$ should be based on our belief about how we expect $Z$ to be \emph{associated} with $Y$ after marginalising out any confounding variables.
+
+\begin{figure}[h]
+\begin{subfigure}[t]{0.99\textwidth}
+     \caption{No prior on $\beta$, the posterior on $w_{yx}$ is centered around its true value}
+    \includegraphics[scale=1]{prior_no_prior}
+\end{subfigure}
+\begin{subfigure}[t]{0.99\textwidth}
+     \caption{Prior on $\beta$ centered around $w_{yz}$, the causal effect of $Z$ on $Y$, $Prior(\beta)= N(w_{yz},\sigma = 0.5)$.
+The posterior on $w_{yx}$ is biased away from its true value (as is the posterior on $\beta$ but this is not the key quantity of interest).}
+    \includegraphics[scale=1]{prior_bad_prior.pdf}
+\end{subfigure}
+\caption{An example demonstrating how selecting a prior for a nuisance parameter centered around the causal effect of that parameter on the outcome, rather than its association with the outcome, can bias estimates for actual parameter of interest.
+We sample $n=1000$ data points from the joint distribution defined by $P(U)P(Z|U)P(X|Z)P(Y|U,Z,X) \sim N(0,1)N(2U,0.09)N(0.5Z,1)N(3U -Z + 0.5X , 0.25)$ and fit the model $Y \sim N(w_yx X+\beta Z,\epsilon)$ in Stan and plot the posterior distributions over $w_yx$ and $\beta$.
+Figure (a) shows the results with no prior (equivalently an improper prior) for both $w_yx$ and $\beta$.
+Figure (b) shows the results when we place a Gaussian prior centered around the $w_yz$ on the distribution for $\beta$.}\label{fig:careful-prior-example}
 \end{figure}
 
-\todo{fix this caption}
-
-Given that a set of variables $\boldsymbol{Z}$ satisfies the backdoor criterion (or equivalently the conditional ignorability assumption), the interventional distribution is asymptotically identifiable and can be estimated from equation \ref{eqn:backdoor}.
-The expected value of $Y$ after the intervention $do(X=x)$ is given by equation \ref{eqn:expected_post_intervention} and the average causal effect for a binary intervention $x \in \set{0,1}$ is given by equation \ref{eqn:ace_ignoreable}.
-
-\eqn{
-\label{eqn:expected_post_intervention}
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}}
-}
-
-\eqn{
-\label{eqn:ace_ignoreable}
-ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
-}
-
-Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata.
-However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
-There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
-The key approaches are based on matching on covariates, propensity score methods and regression.
-We now examine these approaches from a machine learning perceptive.
-
-In standard supervised learning, we have a training set $(\vb{x_1},y_1),...,(\vb{x_n},y_n)$ assumed to be sampled i.i.d from an unknown distribution $\P{\vb{x},y} =\P{\vb{x}}\P{y|\vb{x}}$.
-The goal is to select a hypothesis $h \in \mathcal{H} : \mathcal{X} \rightarrow \mathcal{Y}$ such that, on unseen data $ \sim \P{\vb{x},y}$, $h(\vb{x})$ is close (by some metric) in expectation to $y$.
-In other words we wish to minimise the generalisation error $E_{out}(h)$,
-
-\eqn{
-E_{out}(h) = \Esub{(\vb{x},y)\sim \P{\vb{x}}\P{y|\vb{x}}}{L(h(\vb{x}),y)}
-}
-
-We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
-We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
-
-\eqn{
-E_{in}(h) = \frac{1}{n}\sum_{i=1}^n L(h(\vb{x}_i),y_i)
-}
-
-The VC-dimension of the hypothesis space provides (typically loose) bounds on the probability that $E_{out} >> E_{in}$.
-However, in practice, the generalisation error is usually estimated empirically from a hold-out set of the sample that was not used to train the model, or via cross-validation.
-
-In the causal effect estimation under ignorability, we have training data $(\vb{x}_1,\vb{z}_1,y_1),...,(\vb{x}_n,\vb{z}_n,y_n)$ sampled i.i.d from $\P{\vb{z}}\P{\vb{x}|\vb{z}}\P{y|\vb{x},\vb{z}}$.
-Estimating $\E{Y|do(\vb{X}=\vb{x})}$ corresponds to selecting a hypothesis $h \in \mathcal{H}:\mathcal{X} \times \mathcal{Z} \rightarrow \mathcal{Y}$ that minimises;
-
-\eqn{
-\label{eqn:causal_covariate_shift}
-E_{out} &= \Esub{(\vb{x},\vb{z},y)\sim \delta(\vb{x}-\vb{x'})\P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},\\
-&= \Esub{(\vb{z},y)\sim \P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},
-}
-
-\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
-If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
-\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
-Take the case where we have a binary intervention $x \in \set{0,1}$.
-Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$.
-More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.
-
-\subsubsection{Regression}
-
-The regression approach is to learn a function that is a good approximation to the output surface $\E{Y|X,Z}$.
-Let $f_1(z)= \E{Y|X=1,Z=z}$.
-The expectation of $Y$ after the intervention $X=1$ is then obtained by taking the expectation with respect to $Z$, $\E{Y|do(X=1)} = \Esub{z \sim \P{Z}}{\E{Y|X=1,z}}$.
-We can learn a parametric regression model $\hat{f}_1(z)$ via empirical risk minimisation.
-
-\eqn{
-\hat{f}_1(z) = h_1(z;\hat{\theta}_{obs}) \text{, where } \hat{\theta}_{obs} = \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)  \right]
-}
-
-This estimator is consistent with respect to the observational distribution.
-As the sample size tends to infinity, $\hat{\theta}_{obs}$ approaches the parameter within the hypothesis space that minimises the expected loss given data sampled from the observational distribution.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z|x=1}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-If the model is correctly specified such that $f_1(z) = h_1(z;\theta^*)$ for some  $\theta^* \in \Theta$ then the empirical risk minimisation estimate is consistent with respect to the loss over any distribution of $Z$  \citep{Sugiyama2007}, including the interventional one.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \theta^* = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-The average causal effect can then be estimated by,
-
-\eqn{
-\label{eqn:cef_empirical_regression}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left(\hat{f}_1(z_i) - \hat{f}_0(z_i)\right)
-}
-
-Regression thus has a direct causal interpretation if the parametric model is correctly specified and the covariates included form a valid backdoor adjustment set for the treatment variable of interest in the corresponding structural equation model.
-
-\subsubsection{Propensity scores}
-If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
-The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
-\caption{
-Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
-In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
-If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$.
-We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points).
-In both cases the linear model is a good fit to the data.
-However, the resulting estimate of the causal effect is very poor for the lower values of $z$.
-}
-\label{fig:covariate_shift_adjust}
-\end{figure}
-
-A general approach to estimating the expectation of some function $f(\cdot)$ with respect to data from some distribution $\P{\cdot}$, when we have data sampled from a different distribution $\Q{\cdot}$ is importance sampling \citep{horvitz1952generalization,Koller2009}.
-\eqn{
-\label{eqn:importance_sampling}
-\Esub{\vb{v} \sim \P{\vb{v}}}{f(v)} = \Esub{\vb{v} \sim \Q{\vb{v}}}{f(v)\frac{\P{\vb{v}}}{\Q{\vb{v}}}}
-}
-
-This importance weighting approach can be applied to the covariate shift/average causal effect problem by weighting the terms in the empirical risk minimisation estimator \citep{Sugiyama2007}.
-
-\eqn{
-\label{eqn:importance_weighted_erm}
-\hat{\theta}_{iw} &= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{\P{z_i}\delta(x_i-1)}{\P{z_i}\P{x_i=1|z_i}}  \right] \\
-&= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{1}{e(z_i)}  \right],
-}
-
-where $e(\vb{z})$ is the propensity score, defined by \citep{Rosenbaum1983};
-\eqn{
-e(\vb{z})\equiv \P{x=1|\vb{z}}
-}
-
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
-Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
-
-The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$.
-\citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$.
-This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$.
-Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
-In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
-However in general, it must be estimated from data.
-Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
-\citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
-
-\eqn{
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}} = \Esub{z\sim \P{\vb{Z|x}}}{\E{Y|x,\vb{z}}\frac{1}{e(\vb{z})}}
-}
-
-\eqn{
-\label{eqn:simple_inverse_propensity}
-\hat{\tau}_{ip} = \frac{1}{n}\sum_{i=1}^n \left(\frac{\ind{x_i=1}y_i}{e(\vb{z}_i)} -\frac{\ind{x_i=0}y_i}{1-e(\vb{z}_i)}\right)
-}
-
-\subsubsection{Matching}
-There is a straightforward connection between matching and regression for causal effect estimation.
-If $h \in \mathcal{H} \implies h + a \in \mathcal{H}$ for any constant $a$ and $\hat{f}$ is selected by minimising empirical risk with an $L_2$ loss then $\frac{1}{n}\sum_{i=1}^n \ind{x_i=1}\hat{f}_1(z_i) = \frac{1}{n}\sum_{i=1}^n \ind{x_i=1}y_i$ \footnote{\citep{Imbens2004} state this holds for most implementations}, and equation \ref{eqn:cef_empirical_regression} can be re-written as:
-
-\eqn{
-\label{eqn:cef_regression_missing}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \hat{f}_0(z_i)\right) +\ind{x_i=0}\left(\hat{f}_1(z_i) - y_i \right)\right]
-}
-
-This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
-For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
-Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
-\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
-
-\eqn{
-\hat{\tau}_{match} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \frac{1}{k}\sum_{j \in J_k(i)}y_j\right) +\ind{x_i=0}\left(\frac{1}{k}\sum_{j \in J_k(i)}y_j - y_i \right)\right]
-}
-
-This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
-There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
-Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
-An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.
-
-The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
-A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
-
-The significance of this should not be underestimated.
-Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
-With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
-\citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem.
-There is relatively little theory on model selection for estimating the propensity score.
-To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion.
-It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}.
-With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
-I have found no examples of such an approach.
-
-The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
-Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
-The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
+This result is counter intuitive.
+When fitting a model to estimate the affect of marketing on sales, it is natural to put a prior on the coefficient for price that is strongly negative because we believe increasing price decreases sales.
+However, our prior should be on the expected association between price and sales, holding fixed all the other variables in the model, not the causal relationship between price and sales.
+An alternative is to explicitly include $U$ as a latent variable in the model.
+However, this comes at a computational cost with little benefit for estimation, unless we have good knowledge of the relationships between $U$, $Z$ and $Y$.
+More fundamentally, at whatever level we decide to stop adding latent variables and specify a prior, we can introduce this form of bias if priors are selected naively on the basis of causal intuition and the issue is not limited to linear-Gaussian models.
+I believe this is an important insight for the many applications of Bayesian modelling that make use of human elicited priors, since people tend to think causally.
+The solutions, either explicitly modelling latent variables to higher levels or allowing for them in setting priors, have the effect of broadening the priors over nuisance parameters in the model.
 
@@
 We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
-Note we have used counterfactual notation (see ) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have recieved if it selected action $i$ at timestep $t$.
+Note we have used counterfactual notation (see ) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have received if it selected action $i$ at timestep $t$.
 \end{definition}
@@
 
-Note that we have used the conterfactual notation to represent the rewards given each action.
+Note that we have used the counterfactual notation to represent the rewards given each action.
 The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep.
-The psuedo-regret (equation \ref{eqn:pseudo_regret}) remains unchanged as both it and the learner's actions depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
+The pseudo-regret (equation \ref{eqn:pseudo_regret}) remains unchanged as both it and the learner's actions depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
 However, the regret as defined in equation \ref{eqn:raw_regret} is no longer properly defined as it relies on both $\cfb{Y}{a_t}$ and $\cfb{Y}{i^*}$ being defined at each timestep.
@@
 Adversarial bandit algorithms may perform better in such settings than standard stochastic policies to the extent that they explore more (to avoid the adversary simulating their behaviour) and thus adapt quicker to changes in the reward distribution.
-Adversarial algorithms also have stronger worst case regret guarantees, since even the weak regret for stochastic bandits is not guaranteed to be sublinear in such settings.
-However, if we have constraints on how rapidly or frequently the reward distributions can change over time it is better to use algorithms specificly developed to exploit such information (and compare them against a stronger notion of regret), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
+Adversarial algorithms also have stronger worst case regret guarantees, since even the weak regret for stochastic bandits is not guaranteed to be sub-linear in such settings.
+However, if we have constraints on how rapidly or frequently the reward distributions can change over time it is better to use algorithms specifically developed to exploit such information (and compare them against a stronger notion of regret), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
 
@@
 \chapter{Causal Bandits: Unifying the approaches}
-
+\label{chap:causal-bandit}
+\todo{write an intro to this chapter}
 \section{The framework}
@@
 The notation $\P{\cdot|\vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$ denotes distributions conditional on having observed $\vec{X}^c_t$ and \emph{then} intervened to set $\vec{X}_t = \vec{x}_t$.
-The values of variables in $\vec{X}^c_t$ that are non-decendents of $\vec{X}_t$ remain unchanged by the intervention.
+The values of variables in $\vec{X}^c_t$ that are non-descendents of $\vec{X}_t$ remain unchanged by the intervention.
 The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
 
-The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and correspondign costs, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
-If we can (cheaply) intervene on all of the parents of $Y$ simultaneusly then any context or alternative actions are irrelevent and the problem reduces to a standard multi-armed bandit problem.
+The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
+If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant and the problem reduces to a standard multi-armed bandit problem.
 This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.
@@
 However, prior to selecting an action, we can observe  $I$, the action that would have been selected under an alternate (stationary) policy, which may depend on $U$, see figure \ref{fig:confounded-causal-bandit-graph}.
-In this case the set of contextual variables $\vec{X}^c = I$, the set of post-action feedback variables $\vec{X}^o = \set{Y}$ and the action space consists of all possible assigments of values to a single node $X$, $\actionspace = do(X=x)$.
+In this case the set of contextual variables $\vec{X}^c = I$, the set of post-action feedback variables $\vec{X}^o = \set{Y}$ and the action space consists of all possible assignments of values to a single node $X$, $\actionspace = do(X=x)$.
 This setting reduces to a contextual bandit problem in our causal bandit framework.
-However in thier work on bandits with unobserved confounders \citet{Forney2017} also leverage the fact that $I$ represents the action that would have been selected under an alternate policy to fuse data collected under the previous (observational) policy with data collected under the new policy.
+However in their work on bandits with unobserved confounders \citet{Forney2017} also leverage the fact that $I$ represents the action that would have been selected under an alternate policy to fuse data collected under the previous (observational) policy with data collected under the new policy.
 This information is not encoded in the causal bandit graph in figure \ref{fig:confounded-causal-bandit-graph}, as $I$ could be any variable that is influenced by the unobserved context $U$, and thus cannot be exploited by a standard contextual bandit algorithm.
@@
 Compliance aware bandits describe situations in which the action recommended by the bandit algorithm is not always followed.
-For example a patient may refuse to take a treatment or an advertiser may have complex rules about how many ads a given customer can recieve which prevends some of the suggestions from the ad recomendation engine from being followed.
+For example a patient may refuse to take a treatment or an advertiser may have complex rules about how many ads a given customer can receive which prevents some of the suggestions from the ad recommendation engine from being followed.
 After an action is selected the algorithm can observe the action that was actually taken in addition to the reward.
-\citet{DellaPenna2016} analyse this setting with binary treatments both with and without the presense of a latent confounding variable $U$, see figure \ref{fig:imperfect-compliance-causal-bandit-graph}.
+\citet{DellaPenna2016} analyse this setting with binary treatments both with and without the presence of a latent confounding variable $U$, see figure \ref{fig:imperfect-compliance-causal-bandit-graph}.
 In this case, there are no contextual variables and the action space is again the set of assignments to a single variable but there is post-action feedback, which reveals the value of the action that was actually taken.\footnote{There are some interesting variants of the compliance aware bandit setting that, to my knowledge, have not been analysed.
@@
 
-Consider a stochastic, two-armed bernoulli bandit with the (general) joint distribution of the (conterfactual) rewards $\P{\cf{Y}{1},\cf{Y}{2}}$ given in table \ref{table:joint_bernoulli_counterfactuals}.
+Consider a stochastic, two-armed Bernoulli bandit with the (general) joint distribution of the (counterfactual) rewards $\P{\cf{Y}{1},\cf{Y}{2}}$ given in table \ref{table:joint_bernoulli_counterfactuals}.
 Suppose without loss of generality that arm 1 is the optimal arm.
@@
 \begin{table}[h]
-\caption{General joint distribution over two bernoulli random variables.}
+\caption{General joint distribution over two Bernoulli random variables.}
 \label{table:joint_bernoulli_counterfactuals}
@@
 Consider an algorithm that always selects arm 2.
-The variance (but not expectation) of the regret incured by this algorithm depends on the covariance between the counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{2}$, which cannot estimated from data.
+The variance (but not expectation) of the regret incurred by this algorithm depends on the covariance between the counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{2}$, which cannot estimated from data.
 
@@
 The two cultures \citep{Breiman2001}
- - wrong because suggets we can just use the machine learning approach for everything.
+ - wrong because suggests we can just use the machine learning approach for everything.
 Does not address the problems that a) structural assumptions required for causal inference.
-b) we cannot use cross-vaidation to assess the validity of a causal model (really b is the whole point).
+b) we cannot use cross-validation to assess the validity of a causal model (really b is the whole point).
 
@@
 Definition as backwards.
-Explanatory models focus on theory because some theory (or equivelently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
-
-Causal modelling frameworks allow us to seperate assumptions about the structure of the relationships between variables from assumptions about the functional form.
+Explanatory models focus on theory because some theory (or equivalently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
+
+Causal modelling frameworks allow us to separate assumptions about the structure of the relationships between variables from assumptions about the functional form.
 
@@
 
-Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the languate of EoP but are built on a hybrid of both approaches.
+Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the language of EoP but are built on a hybrid of both approaches.

### 2017-08-31 thesis/offcuts_chapter6.tex
restructure mostly complete

@@
+\chapter*{Todo}
+\begin{enumerate}
+\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
+\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
+\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
+When is a problem causal, what bounds does predictive accuracy give on causal models?
+To what degree is predictive accuracy a metric on which causal models can be judged?
+\item Benchmark estimation algorithms
+\item clarify causal interpretation of Dirichlet models
+\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
+\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption.
+For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability.
+\item Does defining causality in term of the truncated product formula ensure that a causal Bayesian network is unique?
+It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition.
+\end{enumerate}
+
+\section*{Questions}
+\begin{itemize}
+\item Are there any problematic counterfactuals being used in defining bandit problems
+\item in the Schölkopf causal fairness paper - is there an issue with the use of twin networks. Could it be fixed with SWIGs?
+\item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
+\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomised experimental design?
+\item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
+\item What does conditioning mean in the Bayesian sense?
+- can imposing a prior which is true with respect to that variable, increase bias due to confounding?
+(In the same way that in the pneumonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.
+
+\item What are the necessary and sufficient conditions for decision making?
+How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function).
+Ie when is the mean sufficient?
+What about the mean and the variance?
+- when can we take one set of results and compute something about a different loss function?
+
+\item Does the way we define supervised learning and loss functions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?
+
+\end{itemize}
+
 \section{careful prior}
@@
 The metaphysical nature of individual causal effects only arises when we are at the point where the only remaining variation is due to inherent randomness (or variables that we could not even in principle measure).
+
+\section{intro}
+Chapters \ref{chap:causality} and \ref{chap:interventionalist_view} predominantly review the existing literature with the goal of highlighting connections between the different viewpoints and approaches.
+They do not contain novel technical results but present the material in a unified way.
+Chapter \ref{chap:causal-bandit} is a novel contribution.
+It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit literature and describes an algorithm that can leverage this structure to make better decisions.
+
+There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
+This makes it easy to make statements using counterfactuals that cannot be confirmed empirically (even with infinite experimental data).
+We now demonstrate with a simple example that the variance of the cummulative regret depends on unobservable properties of the joint distribution over counterfactual rewards.
+
+Take a stochastic, two-armed, Gaussain bandit with the joint distribution of the (counterfactual) rewards $\P{\cfb{Y}{1},\cfb{Y}{2}}$ given by equation \ref{eqn:joint-gaussain-reward}.
+Suppose without loss of generality that arm 1 is the optimal arm, such that $\mu_1 > \mu_2$.

### 2017-08-31 thesis/thesis_map.tex
restructure mostly complete

@@
 
-\setcounter{secnumdepth}{4}
+\setcounter{secnumdepth}{3}
 
 \begin{document}
-\todo{figure out theoremstle is undefined}
+
 %\def\ci{\perp\!\!\!\perp} % from Wikipedia
@@
 \tableofcontents
-
-\chapter*{Todo}
-\begin{enumerate}
-\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
-\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
-\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
-When is a problem causal, what bounds does predictive accuracy give on causal models?
-To what degree is predictive accuracy a metric on which causal models can be judged?
-\item Benchmark estimation algorithms
-\item clarify causal interpretation of Dirichlet models
-\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
-\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption.
-For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability.
-\item Does defining causality in term of the truncated product formula ensure that a causal Bayesian network is unique?
-It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition.
-\end{enumerate}
-
-\section*{Questions}
-\begin{itemize}
-\item Are there any problematic counterfactuals being used in defining bandit problems
-\item in the Schölkopf causal fairness paper - is there an issue with the use of twin networks. Could it be fixed with SWIGs?
-\item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
-\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomised experimental design?
-\item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
-\item What does conditioning mean in the Bayesian sense?
-- can imposing a prior which is true with respect to that variable, increase bias due to confounding?
-(In the same way that in the pneumonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.
-
-\item What are the necessary and sufficient conditions for decision making?
-How much about a the distribution of rewards for different arms must I know in order to select one (under different assumptions about the loss function).
-Ie when is the mean sufficient?
-What about the mean and the variance?
-- when can we take one set of results and compute something about a different loss function?
-
-\item Does the way we define supervised learning and loss functions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?
-
-\end{itemize}
 
@@
 As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning.
-Although the focus and terminology can differ substantially between fields these techniques all address the underlying goal of estimating the effect of or optimally selecting interventions in some system.
+Although the focus and terminology can differ substantially between fields these techniques all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system.
 Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.
@@
 
-The primary goal of this thesis is to draw together the key approaches to solving causal problems from both the interventionalist and observational viewpoints.
-
-In chapter \ref{chap:causality} I describe key frameworks for causal inference from observational data and discuss how the relate to one-anther and review how they can be leveraged to answer causal queries.
+The goal of this thesis is to draw together the key approaches to solving causal problems from both the interventionalist and observational viewpoints.
+In chapter \ref{chap:causality} I describe key frameworks for causal inference from observational data, discuss how the relate to one-anther and review how they can be leveraged to answer causal queries.
 Chapter \ref{chap:interventionalist_view} covers the interventionalist viewpoint, including traditional randomised experiments and the closely related multi-armed bandit problems.
@@
 I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
-
-Chapters \ref{chap:causality} and \ref{chap:interventionalist_view} predominantly review the existing literature with the goal of highlighting connections between the different viewpoints and approaches.
-They do not contain novel technical results but present the material in a unified way.
-Chapter \ref{chap:causal-bandit} is a novel contribution.
-It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit literature and describes an algorithm that can leverage this structure to make better decisions.
 
@@
 
-In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they he problems they allow us to solve, the assumptions they rely on and how they differ.
+In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ.
 We hope that by describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them.
@@
 We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
-Note we have used counterfactual notation (see ) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have received if it selected action $i$ at timestep $t$.
+Note we have used counterfactual notation (see section \ref{sec:counterfactuals}) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have received had it selected action $i$ at timestep $t$.
+I discuss the (potentially) counterfactual nature of regret further in section \ref{sec:counterfactual-nature-regret}.
 \end{definition}
@@
 \regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}} \\
-&= n\mu_{i^*} - \E{\sum_{t=1}^T{\cfb{Y}{i}}}
+&= T\mu_{i^*} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}}
 }
@@
 Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
-
-\subsection{The counterfactual nature of regret}
-\todo{Fix this bit up}
-
-Note that we have used the counterfactual notation to represent the rewards given each action.
-The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep.
-The pseudo-regret (equation \ref{eqn:pseudo_regret}) remains unchanged as both it and the learner's actions depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
-However, the regret as defined in equation \ref{eqn:raw_regret} is no longer properly defined as it relies on both $\cfb{Y}{a_t}$ and $\cfb{Y}{i^*}$ being defined at each timestep.
-The variance (but not expectation) of $\rawregret$ depends on the covariance between counterfactual variables.
-
- The learner is only able to observe properties of the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
- The stochastic bandit problem can be defined without recourse to counterfactual variables by
-
- The algorithm and definition of regret depend only on the marginal distributions over $\cfb{Y}{i}$ for each $i \in \set{1,...k}$.
 
@@
 \begin{enumerate}
-\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[Y_{t,1},...,Y_{t,k}]$. 
+\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[\cfb{Y}{1},...,\cfb{Y}{k}]$. 
 \item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the learner observes and receives (only) the reward for the selected action $Y_{t,a_t}$ 
+\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
 \end{enumerate} 
@@
 \label{eqn:pseudo_regret-adversarial}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}{\sum_{t=1}^T{Y_{t,i}}} - \E{\sum_{t=1}^T{Y_{t,a_t}}}
+\regret(\pi) &= \max_{i \in \set{1,...,k}}{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}}
 }
@@
 
-\todo{think carefully about the graphs in this sections}
 It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
@@
 This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
+It is even possible to have (useful) context $\vec{X}$ that is a consequence of $\vec{Y}$ as in figure \ref{fig:context-graph-anti-causal} and example \ref{exm:anti-casual-context}.
 
@@
    (4) edge (1)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{Y}$ causes $\vec{X}$}
-\label{fig:context-graph-anti-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (4)
    (3) edge (2)
@@
 \end{subfigure}
+\begin{subfigure}[t]{0.29\textwidth}
+\centering
+\caption{$\vec{Y}$ causes $\vec{X}$}
+\label{fig:context-graph-anti-causal}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+  thick,main node/.style={observed},lt/.style={latent}]
+ %nodes
+\node[lt](1){$\vec{Y}$};
+\node[main node, below=of 1](2){$\vec{Y}^a$};
+\node[main node, above right=of 2](3){$A$};
+\node[main node, left=of 1](4){$\vec{X}$};
+ \path[every node/.style={font=\sffamily\small}]
+   (1) edge (4)
+   (3) edge (2)
+   (1) edge (2);
+\end{tikzpicture} 
+\end{subfigure}
 \caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
@@
 Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
+The causal structures in figures \ref{fig:context-graph-causal} and \ref{fig:context-graph-confounded} also make sense if the elements of $\vec{Y}$ are genuine counterfactuals, that is only the reward of the selected action is ever realised.
+The context $\vec{X}$ can only be useful in causal graphs of the form in figure \ref{fig:context-graph-anti-causal} if the components of $\vec{Y}$ are realised prior to the agent selecting an action, i.e when the components of $\vec{Y}$ are not truly counterfactual, it just that the agent is limited to only observing the outcome of the action they choose.
 To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
@@
 \end{figure}
+
+\vspace{0.5cm}
+\begin{example}
+\label{exm:anti-casual-context}
+Imagine a modified game of roulette.
+As in standard roulette, it consists of a spinning wheel with a ball that can land in one of 37 numbered pockets.
+However, unlike in the standard game, where punters place bets before the wheel is spun, in this variant the wheel is spun first.
+Five seconds before it comes to rest the face of the wheel darkens to hide the position of the ball from the players.
+They then place bets after it has stopped (and are each allowed only to check if the position they have bet on was correct).
+At the point in time the players are selecting their actions, the rewards for each possible choice are fully determined (all zero except for the pocket containing the ball).
+Suppose a canny gambler realises the ball makes a tiny sound, from rocking back and forth, if it is close to the left or right side of the wheel such that the pocket holding it is close to horizontal.
+The existence (or not) of this sound is a consequence of the position of the ball and thus the reward state and the gambler can use it as context to improve the likelyhood of obtaining a reward.
+\end{example}
+
+It is difficult to come up with realistic examples in which the full sequence of rewards is generated but only the one associated with the selected action is observable to the learner.
+In the typically cited applications of contextual bandit algorithms, such as selecting how to treat patients or which ad to serve only the reward of the selected action is ever realised, and the alternate rewards are counterfactual variables.
 
@@
 
+\section{The counterfactual nature of regret}
+\label{sec:counterfactual-nature-regret}
+
+We conclude this chapter on bandits with a note on the counterfactual nature of regret.
+I have not seen this discussed with respect to bandit regret elsewhere, however the issues raised are very closely related to the work by \citet{Dawid2000}, on problems associated with the use of counterfactual variables for (observational) causal inference.
+
+It is possible, as we demonstrated in example \ref{exm:anti-casual-context}, to construct bandit problems in which the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$ is actually generated at each time-step (and, at least in principle, is observable to someone although not the agent).
+However, in most real applications for bandit problems, only the outcome associated with the selected action is ever realised and so the remaining components of the reward vector are counterfactual, justifying our use of the counterfactual notation to denote them.
+There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
+This makes it easy to make statements using counterfactuals that cannot be confirmed empirically (even with infinite experimental data).
+We now show with a simple example that the standard definition of regret is such a fundamentally counterfactual quantity.
+Recall that the cumulative regret is defined by,
+
+\eqn{
+\label{eqn:raw_regret2}
+\rawregret &= \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
+\\&= \sum_{t=1}^T \left(\cfb{Y}{i^*} - \cfb{Y}{a_t}\right)
+}
+
+Take a stochastic, two-armed, Gaussian bandit with the joint distribution of the (counterfactual) rewards $\P{\cfb{Y}{1},\cfb{Y}{2}}$ given by equation \ref{eqn:joint-gaussain-reward}.
+Suppose without loss of generality that arm 1 is the optimal arm, such that $\mu_1 > \mu_2$.
+
+% vector with mean mu_t,mu_c and a covariance matrix
+\eqn{
+\label{eqn:joint-gaussain-reward}
+P(\cfb{Y}{1},\cfb{Y}{2}) \sim N( 
+         \begin{bmatrix}
+           \mu_{1} \\
+           \mu_2 \\
+         \end{bmatrix},
+         \begin{bmatrix}
+           \sigma^2 & \rho \sigma^2 \\
+           \rho \sigma^2 & \sigma^2 \\
+         \end{bmatrix}
+         )
+}
+
+Consider an algorithm that always selects arm 2.
+The distribution over the difference of jointly normal random variables is also normal, letting, $\tau_t = \cfb{Y}{1} - \cfb{Y}{2}$ yields $\P{\tau_t} = N(\mu_1 - \mu_2,2\sigma^{2}(1-\rho))$.
+Thus the distribution over the regret for this algorithm is given by,
+
+\eqn{
+\rawregret \sim N\left(T(\mu_1 - \mu_2), 2\sigma^2 T(1-\rho)\right)
+}
+
+The parameters of the marginal distributions, $\P{\cfb{Y}{1}}$ and $\P{\cfb{Y}{2}}$, can be estimated directly by simply sampling from each arm.
+However, the covariance $\rho$ cannot, because we never simultaneously observe the rewards for both arms.
+As a result, even if we have full knowledge of the reward distributions for each arm, we cannot in general compute the distribution over regret without un-testable assumptions about the covariance between counterfactuals, see figure \ref{fig:metaphysical_distribution_difference}.
+
+\begin{figure}[h]
+\captionsetup[subfigure]{position=b}
+\centering
+\subcaptionbox{Marginal distributions over the rewards for each action for $\mu_1=1$, $\mu_2=0$ and $\sigma = 1$.
+\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
+\hspace{0.02\textwidth}
+\subcaptionbox{The distribution over $R_T$ for $T=10$ for two different values of $\rho$ \label{fig:metaphys_b}}
+{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
+\caption{The distribution over the regret as defined by equation \ref{eqn:raw_regret2} depends on unobservable properties of the joint distribution over counterfactual rewards.
+The same (marginal) distributions over the rewards can correspond to quite different regret distributions.}
+\label{fig:metaphysical_distribution_difference}
+\end{figure}
+
+This result is somewhat perturbing.
+The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep and the behaviour of standard bandit algorithms depends only on the marginal reward distributions for each arm.
+The expectation of the regret as defined by equation \ref{eqn:pseudo_regret} also remains unchanged as both it's definition and the learner's actions depend only on the marginal distributions.
+It seems therefor unfortunate that we should have to assume, for example, that the rewards for alternate actions are independent of one-another in order to be able to analyse the variance of the regret.
+Particularly as this assumption is likely to be violated in many realistic bandit problems.
+For example, a given user may be more (or less) likely to buy something no matter which ad they are served leading to a positive correlation between counterfactual rewards, or an illness might have two (unobservable) sub-types with each medication (bandit arm) effective only against one, resulting in a negative correlation between the counterfactual rewards.
+
+We could focus only on analysing the expected regret, since this depends only on the marginal distributions.
+However, there are many real problems for which we do care how tightly concentrated the regret is around its expectation.
+For example, if we are risk averse, we may prefer an algorithm with slightly higher expected regret but a lower probability of suffering extremely large regret.
+This raises the question, can we construct an alternate definition of regret that allows us to capture how consistently bandit algorithms behave, but which does not depend on any properties of the joint distribution over counterfactual rewards.
+A natural candidate would be,
+
+\eqn{
+R_T = \sum_{i=1}^k N_i(T)\Delta_i, 
+}
+
+where $N_i(T)$ is the number of times arm $i$ was played up to timestep $T$ and $\Delta_i$ is the degree to which arm $i$ is sub-optimal, $\Delta_i = \mu^* - \mu_i$.
+The expectation of this variant of regret is the same as for the standard version defined in equation \ref{eqn:raw_regret2}.
+It depends only on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions.
+Furthermore, this quantity has already been analysed in existing work on the concentration of bandit regret, \citep{Audibert2007,Audibert2009exploration} as a more tractable proxy to the standard regret.
+In conclusion, when selecting measures of bandit performance, it is worth noting whether or not they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.
+
 \chapter{Causal Bandits: Unifying the approaches}
@@
 \todo{write an intro to this chapter}
-\section{The framework}
+
+As we have demonstrated in the previous two chapters both observational causal inference and experimental approaches a
+
+\section{A framework to connect bandit problems with observational causal inference}
 \label{sec:intro_causal_bandit}
@@
 
-\chapter*{Regret is a goat}
-
-Consider a stochastic, two-armed Bernoulli bandit with the (general) joint distribution of the (counterfactual) rewards $\P{\cf{Y}{1},\cf{Y}{2}}$ given in table \ref{table:joint_bernoulli_counterfactuals}.
-Suppose without loss of generality that arm 1 is the optimal arm.
-
-\begin{table}[h]
-\caption{General joint distribution over two Bernoulli random variables.}
-\label{table:joint_bernoulli_counterfactuals}
-\begin{center}
-\begin{tabular}{ c | c | c }
-$\cf{Y}{1}$ & $\cf{Y}{2}$ & $\P{\cf{Y}{1},\cf{Y}{2}}$ \\
-  \hline			
-  1 & 1 & $a$ \\
-  1 & 0 & $p-a$ \\
-  0 & 1 & $q-a$ \\
-  0 & 0 & $1+a - (p+q)$ \\
-  \hline  
-\end{tabular}
-\end{center}
-\end{table}
-
-Recall that the regret is defined by,
-
-\eqn{
-\rawregret = \sum_{t=1}^T (\cfb{Y}{i^*} - \cfb{Y}{a_t})
-}
-
-Consider an algorithm that always selects arm 2.
-The variance (but not expectation) of the regret incurred by this algorithm depends on the covariance between the counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{2}$, which cannot estimated from data.
-
-\eqn{
-\E{\rawregret} = \sum_{t=1}^T(\E{\cf{Y}{1}}-\E{\cf{Y}{2}}) = T(p-q)
-}
-
-\eqn{
-\V{\rawregret} &= \E{\left(\sum_{t=1}^T (\cfb{Y}{1} - \cfb{Y}{2})\right)^2} - \left(\sum_{t=1}^T(\E{\cf{Y}{1}}-\E{\cf{Y}{2}})\right)^2\\
-&= \sum_{t=1}^T \E{(\cfb{Y}{1} - \cfb{Y}{2})^2 - (\E{\cfb{Y}{1}}-\E{\cfb{Y}{2}})^2}\\
-&= \sum_{t=1}^T\left(\V{\cfb{Y}{1}}+\V{\cfb{Y}{2}} - 2 Cov(\cfb{Y}{1},\cfb{Y}{2})\right)\\ &= T\left(p(1-p)+q(1-q) - 2(a-pq)\right)
-}
-
 \bibliographystyle{apalike}% Select the citation style e.g. ieeetr

### 2017-08-31 thesis/thesis_map.tex
adds intro to causal-bandit chapter

@@
 \label{chap:causal-bandit}
-\todo{write an intro to this chapter}
-
-As we have demonstrated in the previous two chapters both observational causal inference and experimental approaches a
+
+As we have seen in the previous two chapters, critical insights and methods for learning how to act have emerged from both the observational and experimental approaches to causality.
+From observational inference, we have the do-calculus, which allows us to map information between observational and interventional distributions and we have a range of estimators designed for these types of mapping problems including propensity scores and doubly robust methods.
+The bandit community has developed carefully tuned algorithms to balance exploration and exploitation, estimation approaches for identifying and selecting the optimal action (rather than estimating the rewards for all actions) and techniques to bound the bias of estimators due to the adaptive nature of decision making processes.
+
+However, despite the commonality of the underlying problem, there has been relatively little work on intersection between observational causal inference and adaptive decision making processes.
+Although some ideas, such as doubly robust estimators and inverse propensity score weighting have transferred across (from causal inference to off-policy evaluation), there are many more opportunities for fruitful work in this space.
+In most key real world decision making processes, from algorithms recommending drug treatments to robots deciding how to open doors, we have access to both potentially huge observational data sets as well as the ability to intervene in the system at some level.
+It is critical that we develop methods that allow us to incorporate the causal knowledge we have about the world, along with data sets collected under a range of different conditions, into our current decision making process.
+In this chapter I introduce a very general framework that connects causal graphical models with multi-armed bandit problems and demonstrate how it can be leveraged to make better decisions.
 

### 2017-09-02 thesis/offcuts_chapter6.tex
adds intro to causal-bandit chapter
adds abstract, improves contributions section

@@
+\section*{Questions for Tor}
+- How do I fix the page numbering in my thesis?
+- How does one derive the regret decompostion
+- Any suggestions on reviewers for my thesis (or people who should not be reviewers)
+- Who else do I need to cite in my bandits chapter?
+
 \chapter*{Todo}
 \begin{enumerate}
+\item Check what I actually have to do to submit
 \item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
@@
 It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition.
+\item Statistisian/Econometricians and computer science seem to use different definitions of optimality (bandits vs efficient estimators). Clarify this - would make a good blog post I think. 
 \end{enumerate}
@@
 \end{itemize}
-
-\section{careful prior}
-
-\subsubsection{Example}
-
-\eqn{
-U &\sim N(0 , 1)\\
-Z &\sim N(2U , 0.09)\\
-X &\sim N(0.5Z , 1)\\
-Y &\sim N(3U + -Z + 0.5X , 0.25)
-}
-
-\begin{figure}[h]
-\includegraphics[scale=.3]{prior_no_prior}
-\caption{no prior, posterior on $w_{yx}$ centered around $w_{yx}$}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[scale=.3]{prior_bad_prior}
-\caption{Prior around $w_{yz}$, $\beta \sim N(w_{yz},\sigma = 0.5)$. The posterior on $w_{yx}$ is biased away from its true value}
-\end{figure}
 
@@
 \section{intro}
+\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.
+
+We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
+Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
+We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
+than algorithms that do not use the additional causal information.
+
 Chapters \ref{chap:causality} and \ref{chap:interventionalist_view} predominantly review the existing literature with the goal of highlighting connections between the different viewpoints and approaches.
@@
 It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit literature and describes an algorithm that can leverage this structure to make better decisions.
+
+\paragraph*{My key research questions are:} 
+\begin{itemize}
+\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
+\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
+\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
+\end{itemize}
+
+A secondary goal is to summarise the key strands of causal inference research from economics and social sciences for the machine learning community and highlight connections and productive areas for future research.
+
+I describe the key existing frameworks for
+
+Demonstrate connections between existing lines of work within the bandit and casusal inference communities.
+
+To summarise the key strands of causal inference research from economics and social sciences for the machine learning community.
+\section{Random}
 

### 2017-09-02 thesis/thesis_map.tex
adds abstract, improves contributions section

@@
-\documentclass[11pt,a4paper,oneside]{book}
+\documentclass[11pt,a4paper,twoside]{report}
 \usepackage[table]{xcolor}
@@
 
-\maketitle
+\begin{titlepage}
+    \begin{center}
+        \vspace*{1cm}
+
+        \Huge
+        {Learning how to act: making good decisions with machine learning}
+
+        \vspace{2cm}
+        \Large
+        {Finnian Lattimore}
+
+        \vfill
+        \Large
+        A thesis submitted for the degree of Doctor of Philosophy
+
+         The Australian National University
+
+        \vspace{1cm}
+
+        September 2017
+
+    \end{center}
+\end{titlepage}
 
@@
 That what we chanced, was what we meant, to do.}{Alexander Pope}
+
+\clearpage
+\vspace*{\fill}
+\begin{center}
+\begin{minipage}{.9\textwidth}
+Except where acknowledged in the customary manner, the material 
+presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any 
+university.
+
+\vspace{40mm}  % vertical space
+
+\hspace{80mm}\rule{40mm}{.15mm}\par   % horizontal space, line, start new line
+\hspace{80mm} Finnian Lattimore\par
+\hspace{80mm} September, 2017
+\end{minipage}
+\end{center}
+\vfill % equivalent to \vspace{\fill}
+\clearpage
+
+\begin{abstract}
+
+This thesis is about machine learning and statistical approaches to decision making.
+How can we learn from data to anticipate the consequence of, and optimally select, interventions or actions?
+Such problems, from deciding which medication to prescribe to patients, who should be released on bail or how much to charge for insurance, to designing policies to reduce unemployment or improve school attendance are ubiqutes and have far reaching impacts on our lives.
+There are two fundamental approaches to learning how to act: reinforcment learning, in which an agent directly intervenes in a system and learns from the outcome, and observational causal inference, whereby we seek to infer the outcome of an intervention from observing the system.
+
+The goal of this thesis to to connect and unify these key approaches.
+I introduce causal bandit problems: a framework which combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcment learning problems that are simple enough to admit formal analysis.
+I show that knoweledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms.
+I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.
+
+\end{abstract}
+
+%\chapter*{Acknowledgements}
 
@@
 \begin{itemize}
-\item Speech recognition (for systems like Siri or Google)
+\item Speech recognition (for systems like Siri or Google Assistant)
 \item Image classification
@@
 
-\section{An overview of this thesis and its contributions}
+\section{Observe or intervene: two very different approaches to causal problems}
 
@@
 As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning.
-Although the focus and terminology can differ substantially between fields these techniques all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system.
+Although the focus and terminology can differ substantially between fields these techniques, all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system.
 Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.
@@
 
-The goal of this thesis is to draw together the key approaches to solving causal problems from both the interventionalist and observational viewpoints.
-In chapter \ref{chap:causality} I describe key frameworks for causal inference from observational data, discuss how the relate to one-anther and review how they can be leveraged to answer causal queries.
-Chapter \ref{chap:interventionalist_view} covers the interventionalist viewpoint, including traditional randomised experiments and the closely related multi-armed bandit problems.
-It describes the key problems and results within the bandit literature and highlights some interesting connections between existing strands of work on bandits and causal inference.
-In chapter \ref{chap:causal-bandit}, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
-The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
+\section{This thesis and its contributions}
+
+I view causal problems as one of the greatest arising challenges for machine learning.
+They incorporate a large set of problems, of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence.
+For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry we need to understand how to leverage our improved approaches to prediction to tackle causal problems.
+
+\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints.
+My major contribution is a framework which unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
+This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
 I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
+My other contributions are primarily connective rather than technical.
+I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias.
+I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data and covariate shift.
+I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.
+
+\paragraph{Thesis overview} This thesis is divided into three key chapters, learning from observational data, learning from interventional data and unifying the approaches.
+Chapter \ref{chap:causality} covers learning to act from observational data.
+In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems.
+Section \ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \ref{sec:defining-causal-effects} and \ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation.
+Section \ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.
+
+Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems.
+In \S\ref{sec:multi-armed-bandits} I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}).
+I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}.
+In \S\ref{sec:off-policy-evaluation} I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of of causal effect estimation from observational data.
+Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.
+
+In chapter \ref{chap:causal-bandit} I introduce causal bandit problems: a framework which unifies causal graphical models with multi-armed bandit problems.
+Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model.
+I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.
+
+In \S\ref{sec:causal-bandit-post-action-feedback} I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected.
+I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms.
+In \S\ref{sec:simple-regret} I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}.
+I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning.
+In \S\ref{sec:simple-regret-general} I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}.
+Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems.
+Finally in \S\ref{sec:causal_bandit_discussion} I discuss extensions and potential future work.
 
@@
 
+\todo{is there a mechanical method to go from one framework to another}
+
+\todo{does it matter if variables are discrete or continuous}
+
 Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
@@
 \subsection{Defining causal effects} 
+\label{sec:defining-causal-effects}
 So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
@@
 \section{Randomised experiments}
+\label{sec:randomized_experiment}
 Randomised controlled trials are often presented as the gold standard for determining causal effects.
@@
 \subsection{Stochastic bandits: Approaches and results}
+\label{sec:stochastic-bandit}
 
@@
 \subsection{Pure-exploration problems}
+\label{sec:pure-exploitation-problems}
 Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
@@
 \subsection{Adversarial Bandits}
+\label{sec:adversarial-bandits}
 
@@
 \subsubsection{The causal structure of contextual bandits}
+\label{sec:causal-structure-contextual-bandit}
 
@@
 \subsection{Learning from logged bandit data}
+\label{sec:off-policy-evaluation}
 Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
@@
 \subsection{Adding structure to actions}
+\label{sec:adding-structure-bandit-actions}
 
@@
 \section{Causal bandits with post action feedback}
+\label{sec:causal-bandit-post-action-feedback}
 
@@
 \subsection{Experiments}
+\label{sec:causal-bandit-experiments}
 We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{Audibert2010}, Thompson Sampling and UCB under a variety of conditions.
@@
 
+Causal questions are so important
+
+Summarise the amazing stuff I have done
+
+Many interesting future directions for research
+
 \bibliographystyle{apalike}% Select the citation style e.g. ieeetr

### 2017-09-04 thesis/thesis_map.tex
minor fixes

@@
 How can we learn from data to anticipate the consequence of, and optimally select, interventions or actions?
-Such problems, from deciding which medication to prescribe to patients, who should be released on bail or how much to charge for insurance, to designing policies to reduce unemployment or improve school attendance are ubiqutes and have far reaching impacts on our lives.
-There are two fundamental approaches to learning how to act: reinforcment learning, in which an agent directly intervenes in a system and learns from the outcome, and observational causal inference, whereby we seek to infer the outcome of an intervention from observing the system.
-
-The goal of this thesis to to connect and unify these key approaches.
-I introduce causal bandit problems: a framework which combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcment learning problems that are simple enough to admit formal analysis.
-I show that knoweledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms.
+Problems such as deciding which medication to prescribe to patients, who should be released on bail, and how much to charge for insurance are ubiquitous, and have far reaching impacts on our lives.
+There are two fundamental approaches to learning how to act: reinforcement learning, in which an agent directly intervenes in a system and learns from the outcome, and observational causal inference, whereby we seek to infer the outcome of an intervention from observing the system.
+
+The goal of this thesis to connect and unify these key approaches.
+I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis.
+I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms.
 I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.
@@
 \section{Motivation}
-Many of the most important questions in science and in our personal lives are about the outcomes of doing something.
+Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something.
 Will asking people to pay upfront at the doctors reduce long term health expenditure?
-If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.
+If we developed a drug to suppress particular genes, could we cure multiple sclerosis and would delaying teen-aged pregnancies improve the outcome for their children.
 
@@
 Correlation is not causation.
-Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.
+Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, despite the fact that the variables involved are susceptible to human intuition and understanding.
 
 We now live in a world of data.
-Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth.
-Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose.
+Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and where the US Institute of Health is considering if all infants should be genetically sequenced at birth.
+Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by-product of some other purpose.
 It does not come from the controlled world of a randomised experiment.
@@
 The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
-From health care, to entertainment and self driving cars; machine learning algorithms will transform many industries.
+From health care, to entertainment and self-driving cars, machine learning algorithms will transform many industries.
 It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory \citep{anderson2008end} and that we may be only a short step from \quotes{The Singularity}, where artificial intelligence exceeds our own and then grows exponentially.
@@
 Getting them to generalise to even slightly different problems or data sets remains very challenging.
-Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it.
+Deciding how we should act or what policies we should implement requires predictions about how a system will behave if we change it.
 The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on.
-As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
+As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live in, it is critical to ensure we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
 
@@
 The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined.
-This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
-However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression).
-Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via \quotes{common sense} the implications of the reported associations.
+This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
+However, the magnitude, direction and even existence of an association depends which variables are controlled for (or included in a regression).
+Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.
 
@@
 Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated.
-For example, based on data from the years 1999 to 2009, the age of miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}.
-We would not expect this relationship to hold if we were to look at a new sample of data.
+For example, based on data from the years 1999 to 2009, the age of Miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}.
+We would not expect this relationship to hold in a new sample of data.
 This form of spurious correlation also has serious repercussions.
 It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published.
-As a consequence, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
+Consequently, the likelihood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
 This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}.
-This issue can be ameliorated by getting more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.
+This issue can be ameliorated by obtaining more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.
 
@@
 For example, the reading ability of children under twelve is strongly correlated with their height, because older children are taller and can read better.
-However height is not a cause of reading ability because if we were to intervene to increase a child's height, for example by giving them growth hormones, we would not expect dramatic improvements in their reading.
-Similarly, extra tutoring in reading will not make a child grow taller.
+However height is not a cause of reading ability because interventions to increase a child's height, for example by giving them growth hormones, would not be expected to improve their reading.
+Similarly, extra lessons in reading will not make a child grow taller.
 This problem is fundamentally different to the issue of spurious correlations arising by chance in finite data sets.
@@
 This viewpoint captures the types of questions that motivate this thesis.
-How can we change the way we do things to obtain better outcomes.
+How can we change the way we do things to obtain better outcomes?
 
 Causality is often liked to explanation; understanding how and why things happen.
-I view explanation in terms of compression and generalisabilty: how much information about the world a model can capture.
+I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture.
 This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction.
-A standard predictive model encodes all the information we need to predict some output given inputs provided the system generating the data does not change.
+A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change.
 A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed.
 More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts.
-By considering conditional interventions within our definition of causal questions we also capture mediation: the study of through which pathways one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact to mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
-
-\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs.
+By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}.
+Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
+
+\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{what if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs.
 The former aims to identify the effect of a known cause.
 The latter can be viewed as identifying causes of an effect.
-They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}.
-However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined.
+They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, that we describe in section \ref{sec:causal_models}.
+However, they state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined.
 The goal of identifying the causes of an effect can be formalised within the graphical causal model framework.
-Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables.
-These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively.
+Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on a given outcome, since it requires us to test or infer the effect of interventions on many different variables.
+These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are often relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "what if" questions are hard to resolve conclusively.
 However, this does not mean that the problem of identifying causes of effects is ill-posed in principle.
-It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature which we review briefly in section \ref{sec:causal-discovery}.
+It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see for example \citep{}.
 \todo{reference a review paper on causal discovery here}
 
-There has traditionally been a large gap between researches in machine learning who focus on prediction, use largely non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models.
-However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of for the training and evaluation models.
-In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.
+There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models.
+However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models.
+In the next section we emphasise the subtley that can exist in determining if a problem is causal by examining some typical examples.
 
@@
 Consider speech recognition.
-You say something, which causes sound waves, which are converted converted to a digital signal that Siri maps to words.
+You say something, which causes sound waves, which are converted to a digital signal that Siri maps to words.
 Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution).
@@
 What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella  causality can be ignored - your decision will not effect if it actually rains.
-However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system.
+If you are using a short term forecast to decide whether to pack an umbrella, causality can be ignored.
+Your decision will not affect if it actually rains.
+However, longer term climate forecasts might (theoretically) lead to action on emissions which would then change the weather system.
 For this we need a (causal) model that allows us to predict the outcome under various different interventions.
@@
 This example also highlights a connection between causal models and transparency in machine learning.
-If we are using a non-causal model to make decision effecting people, there will be a trade-off between the performance and transparency of the model; not because the requirement for transparency restricts us to simple models but because revealing how the model works allows people to change their behaviour to game it.
+If we are using a non-causal model to make decisions affecting people, there will be a trade-off between the performance and transparency of the model; not because the requirement for transparency restricts us to simple models, but because revealing how the model works allows people to change their behaviour to game it.
 
@@
 Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective for serious cases, this seems like a straightforward prediction problem.
+If we assume that in hospital treatment is more effective for serious cases, this appears to be straightforward prediction.
 It is not.
@@
 \citet{Caruana2015} found exactly this effect in a real data set.
-The model learnt that people suffering asthma were \emph{less} likely to die from pneumonia.
-They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk.
-The issue is not with model, it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols.
+The model learned that people suffering asthma were \emph{less} likely to die from pneumonia.
+This was because doctors treated such patients very aggressively, thus actually lowering their risk.
+The issue is not with model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols.
 However, using it to decide how to \emph{change} these protocols could kill.
-The actual question of interest in this case is what happens to patients with characteristics $X$ if we assign them treatment according to decision rule $Z$.
+The actual question of interest in this case is what happens to patients with characteristics $X$ they are assigned treatment according to decision rule (or policy) $\pi(X)$.
 
@@
 
-\todo{is there a mechanical method to go from one framework to another}
-
-\todo{does it matter if variables are discrete or continuous}
+Clarify Cheng's Questions - is there a mechanical method to go from one framework to another, does it matter if variables are discrete or continuous. \todo{}
 
@@
 We did this on an ad-hoc basis to resolve the examples discussed in section \ref{sec:causal_models}.
-In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:estimation_from_finite_data}, discuss the key issues involved in estimating such expressions from finite sample data.
+In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:causal_effect_under_ignorability}, discuss the key issues involved in estimating such expressions from finite sample data.
 We assume the basic structure of the graph is known.
@@
 \centering
-\caption{There can be multiple valid adjustment sets. $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.}
+\caption{Multiple valid adjustment sets.}
 \label{fig:adjust_multiple}
@@
 \centering
-\caption{Conditioning on $Z$ opens the backdoor path $X-U_1-Z-U_2-Y$}
+\caption{Don't condition on all observable variables}
 \label{fig:m-graph}
@@
 \end{subfigure}
-\caption{}
+\caption{Identifying an optimal adjustment set is not always intuatively obvious.
+There may be multiple valid adjustment sets.
+In sub-figure (a) $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.
+Adding additional variables (even pre-treatment variables) to a valid adjustment set can result in an invalid set as in sub-figure (b).
+In this case the empty set is a valid adjustment set for the causal effect of $X$ on $Y$, however adding $Z$ would open a backdoor path $X-U_1-Z-U_2-Y$}
 \label{fig:adjustment}
 \end{figure}
-
-\todo{fix this caption}
 
@@
 As a result, there is no exploration-exploitation trade-off to be made.
-However, the problem does not reduce to supervised learning, because the label, $y$ is not the desired .
+However, the problem does not reduce to supervised learning, because of the bandit nature of the feedback - only the reward of the selected action is observed.
 In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
@@
 \node[main node,  below left=of 1](3){$A$};
+\node[lt, above=of 2](4){$U$};
  \path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
-   (1) edge (2) edge (3);
-\end{tikzpicture} 
-\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$}
+   (1) edge (2) edge (3)
+   (4) edge (1) edge (2);
+\end{tikzpicture} 
+\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$.
+The outcome $Y$ may depend on $\vec{X}$ directly, indirectly through a latent variable $U$ or both.}
 \label{fig:off-policy-evaluation}
@@
 
-Causal questions are so important
-
-Summarise the amazing stuff I have done
-
-Many interesting future directions for research
+TODO - write conclusion \todo{}
+
+Here is the problem and why its important
+
+Here is how what I have done contributes to solving thes problems
+
+What remains to be done - directions for future work
+
+- Relaxing assumption that interventional distributions known
+- Algorithms for causal bandits in which both contextual and post-action feedback is available.
+- Off-policy evaluation for general causal bandits.
+
+Other approaches to identifying causal effects, such instrumental variables, the \quotes{front door} approach could also be applied to off-policy evaluation in some settings.
+
+Ideas developed within the off-policy selection liturature are also relevent classic causal problems within economics etc.
+ - optimal action selection (rather than evaluation)
+ - quantification of finite time properties of estimators (as opposed to asymptopic efficiency). eg \cite{li2014minimax}
+
+Warm start for (contextual) bandit problem based on potentially biased estimates. (We can detect the presense of a confounder with experimental data).
+
+Alternatives to cross-validation.
 

### 2017-09-07 thesis/offcuts.tex
adds minor fixes up to start of 2211

@@
 We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).
+
+\subsection{Fairness and Ethics}
+Causal graphical models are not enough.
+A causal graph that is correct in the sense that it accurately predicts the outcome of an intervention, does not capture how that intervention could turn out in an alternate world in which implicit bias and discrimination did not exist.
+As is well understood in economics, we can get a local minima in which individuals and companies are acting rationally (for example by not hiring women) but the consequence for society as a whole is worse.
 

### 2017-09-07 thesis/offcuts_chapter6.tex
adds minor fixes up to start of 2211

@@
 \begin{enumerate}
+\item Write a talk - include slide on causal inference without do
 \item Check what I actually have to do to submit
@@
 Suppose without loss of generality that arm 1 is the optimal arm, such that $\mu_1 > \mu_2$.
+
+\section{Conclusion}
+
+Ideas developed within the off-policy selection liturature are also relevent classic causal problems within economics etc.
+ - optimal action selection (rather than evaluation)
+ - quantification of finite time properties of estimators (as opposed to asymptopic efficiency). eg \cite{}
+
+Most of the underlying applications do involve decisions which are in fact made sequentially (or in sequential batches).
+Even if a fully online system is impractical, the underlying concepts of the exploration/exploitation trade-off
+
+That is a principled, yet very broadly applicable strategy to select model parameters in the absense of strong theory
+
+While theoretical bounds are extremely important and much work remains to be done in this area, particularly on finite time bounds, it is very difficult to obtain tig
+
+We do not have sufficent theory to give tight finite time bounds on the quality of complex estimation technqiues for causal estimation (or supervise

### 2017-09-07 thesis/thesis_map.tex
adds minor fixes up to start of 2211

@@
 I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis.
-I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms.
+I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be exploited by existing algorithms.
 I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.
@@
 However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models.
-In the next section we emphasise the subtley that can exist in determining if a problem is causal by examining some typical examples.
+In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.
 
@@
 
+Predicting recidivism among paroled prisoners or customer churn also fit within the class of problems where the goal is to identify a group for which a problem will occur in order to target treatment (additional support and monitoring for people on parole, loyalty rewards to improve customer retention, hospitalisation for the severely ill).
+Predictive models can be applied to such problems where the most effective treatment is known for a given target group, and where deciding who to treat on the basis of the model predictions will not change the relationship between the features and outcome.
+
 Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
@@
 Demand prediction seems like a relatively straightforward prediction problem.
-These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
+Models use features such as location, pricing, marketing, time of year and weather, to forecast the demand for a product.
 It seems unlikely that using the model to ensure stock is available will itself change demand.
@@
 At every turn, the AI agent has a number of actions available.
-The board state following each action is deterministic and given by the rules of the game.
+The state of the board following each action is deterministic and given by the rules of the game.
 The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem.
-There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}.
-Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$.
-This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players.
+There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win.
+It must identify higher level features of the board state that are associated with winning.}.
+Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of a player selecting action $a$, given board state $s$.
+This allows the agent to estimate the likelihood of winning from a given starting state by simulating many times the remainder of the game, drawing actions from $\P{a|s}$ for both players.
 Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}.
-The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
+The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelihood an opponent will play a given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
 
@@
 A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes.
-The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people.
+The second requirement is particularly difficult to satisfy when an algorithm is making important decisions affecting individual people.
 Think about problems like credit scoring and parole decisions.
 There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency.
-It is rare to build a model without any intent to make some kind of decision on the basis of its results.
+It is rare to build a model without any intent to make some kind of decision based on its results.
 Thus, I argue we should assume a causal model is required until we can justify otherwise.
@@
 
-The approach of learning the outcome of an action by taking it has plays a key role in advancing our knowledge of the world.
+The approach of learning the outcome of an action by taking it plays a key role in advancing our knowledge of the world.
 However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen.
-Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.
+Estimating the effect of an action from such observational data sets is the problem addressed by observational causal inference.
 Observational causal inference can be viewed as a form of transfer learning.
@@
 Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials.
-Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem randomisation is not always feasible, affordable or ethical.
-The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
+Bandit algorithms deal with the sequential nature of the decision making process, and causal inference with problems where randomisation is not feasible, affordable or ethical.
+The similarities between the problems addressed by these techniques raises the question of whether there are problems best addressed by a combination of these approaches, and if so, how they can be combined.
 
@@
 
-I view causal problems as one of the greatest arising challenges for machine learning.
-They incorporate a large set of problems, of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence.
-For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry we need to understand how to leverage our improved approaches to prediction to tackle causal problems.
+I view causal problems as one of the greatest current challenges for machine learning.
+They incorporate a large set of problems of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence.
+For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry, we need to understand how to leverage our improved approaches to prediction to tackle causal problems.
 
 \paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints.
-My major contribution is a framework which unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
+My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
 This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
@@
 I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias.
-I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data and covariate shift.
+I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift.
 I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.
@@
 Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems.
-In \S\ref{sec:multi-armed-bandits} I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}).
+In \S\ref{sec:multi-armed-bandits}, I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}).
 I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}.
-In \S\ref{sec:off-policy-evaluation} I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of of causal effect estimation from observational data.
+In \S\ref{sec:off-policy-evaluation}, I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of causal effect estimation from observational data.
 Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.
 
-In chapter \ref{chap:causal-bandit} I introduce causal bandit problems: a framework which unifies causal graphical models with multi-armed bandit problems.
+In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems: a framework that unifies causal graphical models with multi-armed bandit problems.
 Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model.
@@
 
-In \S\ref{sec:causal-bandit-post-action-feedback} I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected.
+In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected.
 I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms.
-In \S\ref{sec:simple-regret} I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}.
+In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}.
 I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning.
-In \S\ref{sec:simple-regret-general} I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}.
+In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}.
 Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems.
-Finally in \S\ref{sec:causal_bandit_discussion} I discuss extensions and potential future work.
+Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.
 
@@
 An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited.
-This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it.
+This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system from observing it.
 This is the problem of observational causal inference.
@@
 They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another.
-Counterfactual assertions can be expressed very naturally in human languages and are quite prevalent in  every day conversations;\quotes{if I had worked harder I would have got better grades}, \quotes{she would have been much sicker if she hadn't taken antibiotics}.
-Structural equation models have been developed and applied primarily within economics and related disciplines and can be seen as an attempt to capture key aspects of the behaviour of people with mathematics.
+Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}.
+Structural equation models have been developed and applied primarily within economics and related disciplines.
+They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics.
 Questions around designing policies or interventions play a central role in economics.
-Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised) framework and associated set of methods for estimating causal effects.
+Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised \citep{Pearl2000}) framework and associated set of methods for estimating causal effects.
 Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks.
-They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of conditions under causal effects can be inferred from observational data.
+They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.
 
@@
 However, the literature on causal inference techniques remains spit between the different frameworks.
-Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scores, are described using the counterfactual framework.
-Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions are often based around structural equation models.
+Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework.
+Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models.
 This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.
 
-In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ.
-We hope that by describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them.
+In the next sections, we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ.
+By describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them.
 However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers.
-In order to demonstrate the notation and formalisms each framework provides we will use them to describe the following simple examples.
+In order to demonstrate the notation and formalisms each framework provides, we will use them to describe the following simple examples.
 
@@
 A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
-This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
+This means an intervention to change the value of $X_{i}$ can be expected to affect $X_{j}$, but interventions on $X_{j}$ will not affect $X_{i}$.
 We need some notation to describe interventions and represent distributions over variables in the network after an intervention.
-In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.
+In this thesis, I use the do operator introduced by Pearl \citep{Pearl2000}.
 
@@
 It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
-For example it could represent
+For example, it could represent
 \begin{itemize}
@@
 The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient.
-Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables.
+Since $X$ is assigned via deliberate randomisation, it cannot be affected by any latent variables.
 The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}.
@@
 It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation.
-If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}.
+If we assume that socio-economic status is the only such variable (nothing else affects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}.
 In this case, the interventional distribution is not equivalent to the observational one.
@@
 A causal network encodes all possible interventions that could be specified with the do-notation.
-For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
+For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by seven parameters.
 The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
+Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional thirty parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
 After each two variable intervention, we have a distribution over one variables which requires one parameter.
@@
 But, what if some people failed to take the prescribed treatment?
-We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
+We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive (figure \ref{fig:randomized_imperfect_compliance}).
 Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
@@
 We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
-However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
-Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
+However, if our measurements of these variables must be aggregated over time periods that are longer than the scale at which these interactions occur, the result is a cyclical dependency.
+Even were we able to measure on shorter timescales, there might then not be sufficient data on each variable for inference.
 Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
-\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
+\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and that can be seen as Bayesian causal networks with additional functional assumptions.
 
@@
 
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
+There are objections to counterfactuals arising from the way they describe alternate universes that were never realised.
 In particular, statements involving joint distributions over counterfactual variables may not be able to be validated empirically \citet{Dawid2000}.
@@
 The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
+Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that of the Bayesian network for the same problem (figure \ref{fig:causal_network_example}).
 The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
@@
 The interventional distribution is the same as the observational one.
-The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
+The same argument applies to the intervention $do(X=0)$, and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
 
@@
 
-Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
+Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise, which is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
 A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
 However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
+By declaring it to be structural, we are saying that it represents \emph{causal} assumptions about the relationships between variables.
+When visualised as a network, the absence of an arrow between two variables encodes the assumption that one does not cause the other.
 The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
@@
 The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
+When discussing the counterfactual model, we made the slightly weaker assumption:
 
@@
 The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
+The Neyman-Rubin framework builds on the understanding of randomised experiments and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
 This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
@@
 
-We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
+We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934}, and popularised by \citet{simpson1951interpretation}.
 The following example is adapted from \citet{Pearl2000}.
 Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
-She keeps track of which medication her patients choose and whether or not the treatment is successful.
+She keeps track of the medication her patients choose and whether or not the treatment is successful.
 She obtains the results in table \ref{tab:simpson1}.
@@
 Drug A appears to perform better.
-However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
+However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group, and obtains the data in table \ref{tab:simpson2}.
 
@@
 Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
-The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
+The doctor is concerned with $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
 Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
@@
 \end{tikzpicture}
-\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
+\caption{An example of causal network that can give rise to Simpson's Paradox. In this case, we should select treatment on the basis of the gender-specific results.}
 \label{fig:simpsons_paradox}
@@
 
-With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
+With this model, women are more likely to choose drug A and are also more likely to recover than men, regardless of the treatment they receive.
 Knowing a patient took drug A indicates they are more likely to be female.
-When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
-However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
-So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
+When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B, leading to an apparent reversal in effectiveness.
+However, when the doctor intervenes to set the treatment each person receives, there will no longer be a link from gender to treatment.
+So, in this case she should choose the drug to prescribe from the gender-specific table (and weight by the proportion of the population that belongs to each gender).
 Drug B is the better choice.
@@
 No.
-Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
+Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2}, but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
 This is a drug designed to prevent heart disease.
@@
 This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
-From a purely statistical viewpoint there is no paradox.
+From a purely statistical viewpoint, there is no paradox.
 The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
@@
 \caption{Simpson's reversal visualised.
-The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
+The ratios involving $N_i'$ are steeper than those involving $N_i$ for both the blue and green vectors.
 However, when we sum them, the ratio is steeper for the un-primed variables.}
@@
 Almost every time someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
+Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, the assumptions required to do so, and how to best leverage those assumptions to infer as much information as possible from the available data.
 
 Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
-This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
+This is critical in fields where there is abundant data, but limited theoretical knowledge on how variables are related to one another.
 Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
 An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
-In this thesis I have focused on problems where the structure of the causal graph is known.
-Extending my work to problems where the causal structure is unknown, leveraging the work on causal discovery, is a rich and fascinating line of potential future work, which I discuss very briefly in section \ref{sec:causal_bandit_discussion}.
+In this thesis, I have focused on problems where the structure of the causal graph is known.
+Extending my work to problems where the causal structure is unknown, leveraging the work on causal discovery, is a rich and fascinating line of potential future work, which I discuss briefly in section \ref{sec:causal_bandit_discussion}.
 
@@
 
-Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
+Some of these variables may be latent, in that the available data does not record their value, however their position in the network is assumed to be known.
 For example, consider estimating the impact of schooling on wages.
 Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
-Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
+Even if we have no data to directly assess people's inherent ability we must include ability in the graph because it influences two of the variables we are modelling.
 
@@
 If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
-However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
+However, if there are latent variables, the joint interventional distribution will contain terms that cannot be estimated from the observed data.
 
 The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
-By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
+By leveraging conditional independencies, we can effectively localise the effect of an intervention to a specific part of a larger graph.
 This gives rise to the do-calculus \citep{Pearl2000}.
@@
 This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
-This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
+This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connectedness to assert lack of independence.
 D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
@@
 A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
-This is the identifiablity problem.
+This is the identifiability problem.
 It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
@@
 \end{subfigure}
-\caption{Identifying an optimal adjustment set is not always intuatively obvious.
+\caption{Identifying an optimal adjustment set is not always intuitively obvious.
 There may be multiple valid adjustment sets.
@@
 
-Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal.
+Finally, there is the heuristic principle of playing each arm with probability proportional to the likelihood that it is optimal.
 This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}.
@@
 Suppose a canny gambler realises the ball makes a tiny sound, from rocking back and forth, if it is close to the left or right side of the wheel such that the pocket holding it is close to horizontal.
-The existence (or not) of this sound is a consequence of the position of the ball and thus the reward state and the gambler can use it as context to improve the likelyhood of obtaining a reward.
+The existence (or not) of this sound is a consequence of the position of the ball and thus the reward state and the gambler can use it as context to improve the likelihood of obtaining a reward.
 \end{example}
@@
 
+We now formalise the definition of causal bandit problems.
 We will assume each variable only takes on a finite number of distinct values.
 (The path to relaxing this assumption would be through levering the work on continuous armed bandits).
-The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
-An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and
-the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
 
@@
 A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
-Each action $a \in \actionspace$ assigns a value to a subset of the variables in $\mathcal{X}$ and incurs a known cost $C(a)$ on the learner.
+Each action $a \in \actionspace$, denoted $do(\vec{X}=\vec{x})$, assigns values $\vec{x}$ to a corresponding set of variables $\vec{X} \subset \mathcal{X}$ and incurs a known cost $C(a)$ on the learner.
 One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable}.
@@
 The values of variables in $\vec{X}^c_t$ that are non-descendents of $\vec{X}_t$ remain unchanged by the intervention.
-The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
+The objective of the learner is to minimise either the simple (equation \ref{eqn:simple_regret}) or cumulative regret (equation \ref{eqn:pseudo_regret}).
+
+The empty intervention (where no variable is set) is denoted $do()$.
+The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
+We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
 
@@
 If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant and the problem reduces to a standard multi-armed bandit problem.
-This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.
-
-\vspace{0.5cm}
-\begin{theorem}
-\label{theorem:cb-intervene-all-parents}
-Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$.
-If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and the problem reduces to a standard multi-armed bandit (over actions in in $\mathcal{A'}$).
-\begin{proof}
-for any action $a \in \mathcal{A}$, 
-\eq{
- \E{Y|\vec{X}^c_t,a} &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\vec{X}^c_t,a,\parents{Y}}} \\
- &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\parents{Y}}}\\
- &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|do(\parents{Y})}}\\
- &\leq \max_{\parents{Y}}\E{Y|do(\parents{Y})} = \E{Y|a'} \text{ for some $a' \in \mathcal{A'}$}
-}
-\end{proof}
-\end{theorem}
+
+%This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.
 
@@
 
+\begin{remark} If the actionspace $\mathcal{A}$ is unconstrained, that is consists of all possible assignments of values to variables, (and all actions have equal cost) the optimal action will set the value of all the parents of $Y$ and Algorithm \ref{alg:general} cannot do better than uniform exploration over these arms.
+In this case, after we use the causal structure to eliminate irrelevent actions prior to taking any samples, the problem of selecting within the remaining actions can be treated as a standard multi-armed bandit problem.
+\end{remark}
+
+\vspace{0.5cm}
+\begin{theorem}
+\label{theorem:cb-intervene-all-parents}
+Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$.
+If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and the problem reduces to a standard multi-armed bandit (over actions in in $\mathcal{A'}$).
+\begin{proof}
+for any action $a \in \mathcal{A}$, 
+\eq{
+ \E{Y|\vec{X}^c_t,a} &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\vec{X}^c_t,a,\parents{Y}}} \\
+ &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\parents{Y}}}\\
+ &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|do(\parents{Y})}}\\
+ &\leq \max_{\parents{Y}}\E{Y|do(\parents{Y})} = \E{Y|a'} \text{ for some $a' \in \mathcal{A'}$}
+}
+This proves the optimal action $a^* \in \mathcal{A'}$
+
+We now consider using importance weighted estimators from Algorithm \ref{alg:general} to estimate the rewards for all actions in $ \mathcal{A'}$. The optimal sampling weights $\eta$ are given by,
+
+\eq{
+\eta^*  = \argmin_\eta \max_{a \in \mathcal{A'}} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]\,.
+}
+
+Note that we now only have to obtain estimates for actions $a \in \mathcal{A'}$, since we know the others to be sub-optimal, so the $max$ is only over these actions.
+However we $b$ still sums over all possible actions in the demoninator of the importance sampling estimator, to allow for the possiblity that playing sub-optimal actions allows more efficient estimation of the optimal actions.
+We now prove that this is not the case and that, in this specific setting, Algorithm \ref{alg:general} cannot do better than uniform sampling over the actions $a \in \mathcal{A'}$.
+Now each action $a \in \mathcal{A'}$ consists of a given assignment $\vec{x}_a$ to $\parents{Y}$.
+
+\eq{
+a =& do(\parents{Y} = \vec{x}_a) \implies \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right] = \frac{1}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}=\vec{x}_a}}\\
+& \implies \eta^* = \argmax_\eta \left[\min_{a \in \mathcal{A'}} \sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}=\vec{x}_a} \right] 
+}
+
+Let $N_a$ denote $\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}=\vec{x}_a}$, which can be viewed as an effective number of samples for action $a$.
+Choosing $\eta_b = \ind{b \in \mathcal{A'}}\frac{1}{|\mathcal{A'}|}$, corresponding to uniform exploration over the optimal arms only, yields $N_a = \frac{1}{|\mathcal{A'}|}$ for all $a$.
+To do better, we would need to find weights $\eta$ such that $N_a > \frac{1}{|\mathcal{A'}|}$ for all $a$.
+However,
+
+\eq{
+N_a > \frac{1}{|\mathcal{A'}|} \forall{a} &\implies \sum_{a\in \mathcal{A'}}N_a > 1\\ 
+&\implies \sum_{b \in \calA}\eta_b\sum_{a\in \mathcal{A'}}  \Pn{b}{\parents{Y}=\vec{x}_a} > 1
+\implies \sum_{b \in \calA}\eta_b > 1
+}
+This violates the fact that $\eta$ is a distribution over the actions, and thus must have weights that sum to 1, thus completing the proof.
+\end{proof}
+\end{theorem}
+
 \subsection{Experiments}
@@
 
-TODO - write conclusion \todo{}
-
-Here is the problem and why its important
-
-Here is how what I have done contributes to solving thes problems
-
-What remains to be done - directions for future work
-
-- Relaxing assumption that interventional distributions known
-- Algorithms for causal bandits in which both contextual and post-action feedback is available.
-- Off-policy evaluation for general causal bandits.
-
-Other approaches to identifying causal effects, such instrumental variables, the \quotes{front door} approach could also be applied to off-policy evaluation in some settings.
-
-Ideas developed within the off-policy selection liturature are also relevent classic causal problems within economics etc.
- - optimal action selection (rather than evaluation)
- - quantification of finite time properties of estimators (as opposed to asymptopic efficiency). eg \cite{li2014minimax}
-
-Warm start for (contextual) bandit problem based on potentially biased estimates. (We can detect the presense of a confounder with experimental data).
-
-Alternatives to cross-validation.
+The underlying motivation behind much applied statistical and machine learning work is to guide us to make better decisions.
+In many cases, the actions that we take in response to the model will change the system from which the data was generated.
+It is critical that we are able to recognise the causal nature of such problems and appropriately model the decision making part of the process.
+If machine learning is to be as transformative in fields such as economics, medicine and social science as it has been for image recognition, voice processing and machine translation, we must develop methods to estimate the effect of, and optimally select, interventions that are as effective as those we have for pattern recognition.
+We need to bridge the gap between the theory driven models of economics and science and the black box prediction approach that has been so successful in machine learning.
+This will involve clarifying what information is (currently) best encoded by theory and what can be successfully inferred from data and developing methods that can incorporate the theory or structure required to allow models to generalise from one setting to another whilst retaining the flexibility to capture complex patterns in empirical data.
+
+A better understanding of causality is also relevant for the discussions around transparency and ethics in machine learning, particularly with respect to the European Union's new General Data Protection Regulation, which requires that automated decision systems that significantly affect individuals provide provide \textit{"meaningful information about the logic involved."} \cite{Goodman2016}.
+The recognition that there is a fundamental trade-off between accuracy and transparency, unless we can build perfect causal models, when the interests of individuals and society diverge has implications for they way we design and regulate systems that have the potential to have major impacts on people's lives.
+We must develop approaches to ensuring machine learning decisions are reasonable and ethical that allow effected individuals recourse to dispute or improve outcomes but do not undermine the ability of the system to function.
+
+The observational and interventional viewpoints on learning to act contribute complementary components to a general approach.
+Observational causal inference provides, through the do-calculus, a formal means to map information from observational to interventional settings, as well as between different interventions.
+Bandit algorithms,
+
+I have developed a framework that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this framework encode some key existing problems in the literature.
+I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.
+
+This work represents an important first step towards a unified approach to causal inference and optimal decision making.
+There is much exciting work remaining to be done.
+Although the causal bandit framework can capture contextual information as well as post-action feedback, I have formally analysed and developed algorithms only for the latter.
+Additionally, to make the problem tractable, I made the (major) assumption that the interventional distribution over the parents of the outcome was known.
+This can be relaxed to assuming the interventional distribution over some Markov blanket with respect to the outcome is known.
+Information can then be shared between actions outside the blanket in the same way as in algorithm \ref{alg:general}, whilst actions inside could be learned explicitly.
+Relaxing this assumption entirely is a much more challenging problem.
+However, as is demonstrated by the specific example of the parallel causal bandit problem, it is possible to develop algorithms that require only the casual structure of the graph and yield substantially lower regret.
+
+Another interesting line of research is the question of off-policy evaluation for causal-bandit problems.
+As in the online case, knowledge of the causal structure between variables in the graph provides additional information about the reward for the actions that were not selected at each timestep.
+The problem differs from typical observational causal inference in several ways; the focus is on identifying an optimal policy, rather than unbiased estimation of all policies, the goal is to explore the value of interventions on a range of different variables, rather than the optimal setting of a single variable, and the data will, in general, be non-stationary in a rather special way due to the adaptive nature of policy that generated it.
+Existing work on off-policy evaluation focuses (at least implicitly) on estimating causal effects by adjusting for all variables that simultaneously effect both the action selection and outcome.
+If additional information is available about the causal relationships between variables involved, other approaches to identifying causal effects such as instrumental variables or the \quotes{front door method} \citep{Pearl2000} could be also be applied to off-policy evaluation.
+
+Insights from the bandit literature can also be applied to more classical causal inference problems.
+In particular, estimators that are geared towards optimal action selection (rather than evaluation of all actions) and approaches to quantify the finite time properties, as opposed to asymptotic efficiency, of estimators, for example \citep{li2014minimax}.
+
+An important line of research, that is relevant to both the observational and interventional approaches to causal inference, is developing methodologies for model evaluation and selection that provide something equivalent to what cross-validation does for supervised learning.
+
+My hope is that in the next years, combining the reinforcement learning approach to decision making, with causal graphical models and causal effect estimation techniques developed within statistics and economics, will revolutionise our ability to make good data driven decisions.
 
@@
 \chapter{Causation vs Prediction}
+
+Mixing continuous experimentation at an individual level with batch approaches.
+There may be fixed costs associated with \quotes{enabling} certain actions.
+For example, social programs may have some fixed overheads, no matter how many people are enrolled in them, etc.
+Reuse data obtained to optimise for one outcome to learn about optimising for another.
+We don't have a single outcome or loss-measure that fully incorporates everything we care about for the majority of problems.
 

### 2017-09-10 thesis/offcuts_chapter6.tex
fixed grammer issues

@@
 \section{Philosophical objections to counterfactuls}
+from the information available to the agent, the joint distribution over counterfactuals cannot be inferred, unless the assignment mechanism in noiseless - so it does not make sense to define regret that depends on it.
+
 There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.

### 2017-09-10 thesis/thesis_map.tex
fixed grammer issues

@@
 I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
-My other contributions are primarily connective rather than technical.
+I make a number of additional connective contributions that are not encompassed by my work on causal bandit problems.
 I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias.
@@
 Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
-Therefor, in this section, we briefly review the key properties of Bayesian networks.
+Therefore, in this section, we briefly review the key properties of Bayesian networks.
 A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
-Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
+Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way that highlights conditional independencies between them.
 
@@
 D-separation is an algorithm that extends the local Markov property to find these additional independencies.
-It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
-
-The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
-From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
+It provides a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
+
+The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ provides no additional information about $X$.
+From a graphical perspective, you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
 Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
 In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
-In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
+In figure (d), the path is unblocked if we condition on $Z$ and blocked otherwise.
 
 The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
-The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
+The counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
 As an example, consider a scholarship available to female or disadvantaged students.
@@
 There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
-However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
+However, if we know a student is receiving a scholarship, then learning that they are male  increases the probability that they are disadvantaged.
 \begin{figure}[h]
@@
 If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
-However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
-A particular case in which this occurs is when there are deterministic relationships between variables.
-If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
-This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
-This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connectedness to assert lack of independence.
+However, it is possible to construct distributions that do not have a perfect map; that is, they contain conditional independencies that cannot be represented by d-separation.
+The presence of deterministic relationships between variables is one instance where this occurs.
+If there is a Bayesian network $G$ with some deterministic nodes, then we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
+This does not conflict with theorem \ref{thrm:dsep_independence}, as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
+This subtlety led to confusion in the independencies that hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connectedness to assert lack of independence.
 D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
@@
 Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
-However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
+However, setting $X$ tells us nothing about its parents and therefore affects $Y$ only via direct paths out of $X$.
 Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
@@
 \caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
-In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
+In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$, which also influences $Y$.
 If we condition on $W$, we block this indirect path.}
@@
 The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
-These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
+These are the only paths by which intervening in $\boldsymbol{X}$ can affect $\boldsymbol{Y}$.
 
@@
 The do calculus is complete \citep{Shpitser2006b,Huang2006}.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
+A problem is identifiable if and only if the interventional distribution of interest can be transformed into terms containing only observational quantities via repeated application of the do-calculus.
 There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
@@
 \item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network, and
+In other words, it can determine if the query is identifiable given the assumptions encoded by the network, and
 \item if it is identifiable, return the required expression. 
@@
 
-Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
+Figure \ref{fig:identify} shows some examples of identifiable and non-identifiable queries.
 I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
@@
 What if we assume various restrictions on the functional form of the relationships between the variables?
-Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
+Some queries that are not identifiable non-parametrically can be identified by additional assumptions, such as linearity.
 A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
@@
 Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
-There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
-For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
+There are several ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
+For a binary treatment variable $X$, the average causal effect, ACE \footnote{Also referred to as the average treatment effect (ATE)} is defined as:
 
@@
 Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
-However we can personalise the average causal effect by stating it with respect to some observed context.
-I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
+However, we can personalise the average causal effect by stating it with respect to some observed context.
+I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE).
+However, that term is also used for the sample rather than population effect.}
 
@@
 
-In some cases we may be interested in the average causal effect for some sub-group of the population.
+In some cases, the average causal effect for some sub-group of the population is of prime interest.
 A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
@@
 The backdoor criterion derives from rule 2 of the do-calculus.
-Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
-There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
+Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set that satisfies the backdoor criterion.
+There may be more than one valid adjustment set, (figure \ref{fig:adjust_multiple}).
 The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
@@
 Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
-For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
+For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether estimates of the causal effect of smoking on neonatal mortality should adjust for birth weight.
 
@@
 Adding additional variables (even pre-treatment variables) to a valid adjustment set can result in an invalid set as in sub-figure (b).
-In this case the empty set is a valid adjustment set for the causal effect of $X$ on $Y$, however adding $Z$ would open a backdoor path $X-U_1-Z-U_2-Y$}
+In this case the empty set is a valid adjustment set for the causal effect of $X$ on $Y$.
+However adding $Z$ would open a backdoor path $X-U_1-Z-U_2-Y$}
 \label{fig:adjustment}
@@
 However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
-There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
+There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability (see \citet{Imbens2004} for a comprehensive review).
 The key approaches are based on matching on covariates, propensity score methods and regression.
@@
 
-We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
+We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, as we only have access to a sample.
 We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
@@
 \citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
-If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
+If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$, but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
 \footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
@@
 
-The average causal effect can then be estimated by,
+The average causal effect can then be estimated by:
 
@@
 The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
-Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
+Doubly robust methods are asymptotically unbiased so long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
 
@@
 Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
-In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
-However in general, it must be estimated from data.
-Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
+In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013}, the propensity score may be known.
+However, in general, it must be estimated from data.
+Frequently this is undertaken with a simple parametric model, such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
 \citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
@@
 This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
-There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
+There are many variants of matching estimators using different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
 Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
@@
 
-The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
+The practical performance of the estimation approaches discussed in this section depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
 A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
@@
 The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
-Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
-The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
+Unfortunately, there are a relatively small number of examples where comparable observational and experimental data are available.
+The results are mixed with later studies finding generally better alignment of results, but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data.
 
@@
 The Bayesian approach to modelling encapsulates a very general approach to combining prior knowledge and assumptions with data to draw inferences.
-You can specify a complex model, capturing  your assumptions about the underlying mechanisms of your system, describe what you are uncertain about with parameters (and priors over their values) and then update your beliefs about the values of these parameters based on the observed data.
-If you believe the model to be a mechanistic representation of the way world works then you can use it to infer the consequences of interventions.
-
-Why then do we need a special framework for causal inference; can't we simply encode all the assumptions required to infer the outcome of an intervention in a given system in a Bayesian model?
+You can specify a complex model that captures your assumptions about the underlying mechanisms of your system, describe uncertainty about the parameters (and priors over their values), and then update your beliefs about the values of these parameters based on the observed data.
+If you believe the model to be a mechanistic representation of the way the world works then it can be used to infer the consequences of interventions.
+
+Why then do we need a special framework for causal inference?
+Why not encode all the assumptions required to infer the outcome of an intervention in a given system in a Bayesian model?
 The short answer is we could.
@@
 
-We could represent problems of the type where we try to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and what they have in common, see figure \ref{fig:causal_ordinary_bayes}.
+We can represent problems where the goal is to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and their common features (figure \ref{fig:causal_ordinary_bayes}).
 This does not require any special framework or notation.
 The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks.
-However, without a causal framework, we have to make assumptions about what will be invariant to the intervention specifically for each such problem we encounter.
+However, without a causal framework, we have to make assumptions about what will be invariant to the intervention for each such problem encountered.
 For complex problems, it is very difficult to conceptualise the assumptions we expect to hold without the benefit of a causal framework.
@@
 As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
-However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
+However, this would then entail a complex transformation relating the parameters between the two plates, rather than a simple invariance.}
 \label{fig:causal_ordinary_bayes}
@@
 
-We can also combine causal graphical models with Bayesian approaches to estimation.
+Causal graphical models can also be combined with Bayesian approaches to estimation.
 For example, we can use a causal graphical model to identify a set of variables that form a valid adjustment set and then use Bayesian regression to estimate posterior distribution over a given causal effect.
-However, in doing so we need to be careful about the way in which we specify priors as I highlight in the next section.
+However, as I highlight in the next section, the way priors are specified is critical.
 
@@
 
-Suppose we have a data generated by the linear Gaussian Bayesian network below.
+Suppose data is generated by the linear Gaussian Bayesian network below.
 The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$.
@@
 
-If we do a Bayesian regression of $X$ and $Z$ against $Y$ and use intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ we are likely to select something centered around $w_{yz}$, but this is not correct and can introduce bias into our estimate of the coefficient for $w_{yx}$, as we demonstrate in figure \ref{fig:careful-prior-example}.
-The prior for $\beta$ should be based on our belief about how we expect $Z$ to be \emph{associated} with $Y$ after marginalising out any confounding variables.
+Undertaking a Bayesian regression of $X$ and $Z$ against $Y$ and using intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ will lead to selection of something centered around $w_{yz}$.
+However, this is incorrect and can introduce bias into our estimate of the coefficient for $w_{yx}$, as demonstrated in figure \ref{fig:careful-prior-example}.
+The prior for $\beta$ should be based on the belief about how $Z$ is expected to be \emph{associated} with $Y$, after marginalising out any confounding variables.
 
@@
 This result is counter intuitive.
-When fitting a model to estimate the affect of marketing on sales, it is natural to put a prior on the coefficient for price that is strongly negative because we believe increasing price decreases sales.
+When fitting a model to estimate the effect of marketing on sales, it is natural to put a prior on the coefficient for price that is strongly negative because we believe increasing price decreases sales.
 However, our prior should be on the expected association between price and sales, holding fixed all the other variables in the model, not the causal relationship between price and sales.
@@
 However, this comes at a computational cost with little benefit for estimation, unless we have good knowledge of the relationships between $U$, $Z$ and $Y$.
-More fundamentally, at whatever level we decide to stop adding latent variables and specify a prior, we can introduce this form of bias if priors are selected naively on the basis of causal intuition and the issue is not limited to linear-Gaussian models.
-I believe this is an important insight for the many applications of Bayesian modelling that make use of human elicited priors, since people tend to think causally.
+More fundamentally, at whatever level we decide to stop adding latent variables and specify a prior, we can introduce this form of bias if priors are selected naively on the basis of causal intuition.
+The issue is not limited to linear-Gaussian models.
+This is an important insight for the many applications of Bayesian modelling that make use of human-elicited priors, since people tend to think causally.
 The solutions, either explicitly modelling latent variables to higher levels or allowing for them in setting priors, have the effect of broadening the priors over nuisance parameters in the model.
@@
 
-The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it.
+The previous sections focused on aspects of the problem of estimating the likely effect of an intervention from data gathered prior to making the intervention.
 There is an obvious alternative.
-Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens.
+Instead of trying to infer the outcome of an intervention from passive observations, one can intervene and see what happens.
 There are three key differences between observing a system and explicitly intervening in it.
-Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from.
+First, we determine the nature of the intervention and thereby control the data points used to estimate causal effects.
 Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}.
@@
 If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
-Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.
+Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that affect both treatment and outcome.
 
@@
 Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest.
-If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can estimate the average causal effects directly from the differences in outcome across treatments.
-More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
+If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds.
+We can then estimate the average causal effects directly from the differences in outcome across treatments.
+More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, (figure \ref{fig:random_experiment_network_stratified}).
+This necessitates that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability (\S\ref{sec:causal_effect_under_ignorability}).
 The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
@@
 The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
-There may be imperfect compliance, the treatment selected by the randomiser is not always followed, or output censoring, the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.\footnote{Non-compliance is a problem if the goal is to estimate the causal effect of the treatment on the outcome but not if the goal is to estimate the causal effect of prescribing the treatment.
-The latter makes sense in a context where the process by which people decide whether or not to take the treatment they have been prescribed is likely to be the same if the treatment were made available more generally beyond experimental trial.} See figure \ref{fig:random_experiment_network_imperfect_compliance} for a the graphical model of a randomised experiment with imperfect compliance.
+There may be imperfect compliance so that the treatment selected by the randomiser is not always followed, or output censoring in which the experimenter is not able to observe the outcome for all units (for example if people drop out).
+If compliance or attrition is not random, but associated with (potentially latent) variables that also affect the outcome, then the problem of confounding bias returns.\footnote{Non-compliance is a problem if the goal is to estimate the causal effect of the treatment on the outcome but not if the goal is to estimate the causal effect of prescribing the treatment.
+The latter makes sense in a context where the process by which people decide whether to take the treatment they have been prescribed is likely to be the same if the treatment were made available more generally beyond the experimental trial.} See figure \ref{fig:random_experiment_network_imperfect_compliance} for a graphical model of a randomised experiment with imperfect compliance.
 
@@
 
-It is not always possible or ethical to conduct a randomised controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
+It is not always possible or ethical to conduct a randomised controlled trial, as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
 When experimentation is possible, it is frequently difficult or expensive.
-This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
-In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest.
+This means experimental data sets are often much smaller than observational ones, limiting the complexity of models that can be explored.
+In addition, they are often conducted on a convenient, but unrepresentative, sample of the broader population of interest (for example first year university students).
+This can result in estimates with high \emph{internal validity} \citep{Campbell1963}, in that they should replicate well in a similar population, but very low \emph{external validity} in that the results may not carry over to the general population of interest.
 The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
@@
 Finally, non-adaptive randomised experiments are not optimal from either an active or reinforcement learning perspective.
-As an experiment proceeds information is obtained about the expectation and variance of each intervention (or treatment).
+As an experiment proceeds, information is obtained about the expectation and variance of each intervention (or treatment).
 Fixed experimental designs cannot make use of this information to select which intervention to try next.
@@
 \footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
+It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$, the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
 The goal of the learner is to maximise the total reward they receive.
-This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
+This problem captures the exploration-exploitation trade-off.
+The learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
 
@@
 
-At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T \cfb{Y}{a_t}$.
+At the end of the game, the total reward obtained by the learner is $\sum_{t=1}^T \cfb{Y}{a_t}$.
 We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
@@
 The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
-To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
+To account for this, the performance of bandit algorithms is quantified by the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
 This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
@@
 However, it still depends on key characteristics of the reward distributions for each action.
-Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range.
-Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
+Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits), or that the rewards are bounded in some range.
+Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem-dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
 
@@
 As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
-The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
+The expectation and variance guarantees of standard estimators do not hold in this setting (see figure \ref{fig:early_stopping} for a concrete example).
 This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
@@
 Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded.
-This has some interesting potential real world implications.
+This has some interesting real world implications.
 Early stopping of clinical trials is controversial.
@@
 There are a few key principles that are used to guide the development of bandit algorithms.
-The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
+The simplest is to explicitly separate exploration from exploitation, and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
 A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
@@
 Another key approach is \emph{optimism in the face of uncertainty}.
-Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
+Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
 The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
@@
 Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
-The problem dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
+The problem-dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
 
@@
 
-The form of the dependence on the number of arms $k$ and horizon $T$ differs between the problem dependent and worst case regret.
-The problem dependent regret grows linearly with the number of arms, $k$, and logarithmically with $T$.
-The difference stems from the fact the problem dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
+The form of the dependence on the number of arms $k$ and horizon $T$ differs between the problem-dependent and worst case regret.
+The problem-dependent regret grows linearly with the number of arms, $k$, and logarithmically with $T$.
+The difference stems from the fact the problem-dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
 \citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by $\regret \in \bigomega{\sqrt{kT}}$
@@
 Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
-This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \citep{Lattimore2015}.
+This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem-dependent bound of the form achieved by UCB \citep{Lattimore2015}.
 
@@
 Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
-However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
+However, it is complex to analyse.
+\citet{kaufmann2012thompson} demonstrate that it obtains optimal problem-dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
 
@@
 \label{sec:pure-exploitation-problems}
-Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
+Another problem that has attracted recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
 In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
@@
 
-For oblivious adversarial bandits we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
-We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
+For oblivious adversarial bandits, we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
+We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because the sequence of rewards is fixed.
+However the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
 
@@
 Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \citep{Audibert2009}.
-The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
-However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nonetheless obtain good problem dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
+The focus, for adversarial bandits, is on analysing the worst case regret because the problem-dependent regret is not well defined without additional assumptions.
+However, there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting, but can nonetheless obtain good problem-dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
 
@@
 However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, does not fully capture an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings where the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
+In settings where the rewards change over time, the pseudo-regret can be negative (see figure \ref{fig:non-stationary-bandit}) so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
 An example of a setting that lies between stochastic and adversarial bandit problems is the non-stationary setting, in which the rewards are generated stochastically from a distribution that varies over time.
@@
 Adversarial algorithms also have stronger worst case regret guarantees, since even the weak regret for stochastic bandits is not guaranteed to be sub-linear in such settings.
-However, if we have constraints on how rapidly or frequently the reward distributions can change over time it is better to use algorithms specifically developed to exploit such information (and compare them against a stronger notion of regret), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
-
-from the information available to the agent, the joint distribution over counterfactuals cannot be inferred, unless the assignment mechanism in noiseless - so it does not make sense to define regret that depends on it.
+However, if there are constraints on how rapidly or frequently the reward distributions can change over time, it is better to use algorithms specifically developed to exploit such information and compare them against a stronger notion of regret (see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}).
 
@@
 In the standard multi-armed bandit setting, each decision is identical and the goal is to learn a single best action.
-However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
+However, in most real life (sequential) decision making processes, the optimal action depends on some context.
 The best treatment to offer an individual patient could depend on their age, gender, disease sub-type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
-Similarly, decisions on which ad or content to display on a webpage or which product to recommend can be \emph{personalised} based on the previous behaviour of the user.
+Similarly, decisions on which ad or content to display on a webpage, or which product to recommend, can be \emph{personalised} based on the previous behaviour of the user.
 A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
+
 Contextual bandits are a generalisation of multi-armed bandits that make use of this additional contextual information.
-The term contextual bandit was coined by \citet{Langford2008}, however close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
+The term contextual bandit was coined by \citet{Langford2008}.
+However, close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
 
@@
 This is optimal with respect to this class of hypothesis.
-However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the in the relationships between different contexts and actions.
+However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the relationships between different contexts and actions.
 As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
 For example, with $n$ binary variables, $N = 2^n$}.
-We expect some form of smoothness; values of context that are similar should lead to comparable rewards for a given action.
+We expect some form of smoothness; that is, values of context that are similar should lead to comparable rewards for a given action.
 We need algorithms that can leverage such assumptions.
@@
 As in causal effect estimation, we can view this as a missing data problem.
-However, the data is missing not at random because which component of the reward is observed depends on the action selected which in turn is a function of the previous history of actions and rewards.
-
-The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly separating exploration from exploitation.
+However, the data is missing not at random because the component of the reward that is observed depends on the action selected, which in turn is a function of the previous history of actions and rewards.
+
+The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a \quotes{data missing at random} problem by explicitly separating exploration from exploitation.
 Epoch-greedy is an explore-exploit algorithm.
@@
 The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
-However, it involves maintaining a list of weights for each hypothesis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
+However, it involves maintaining a list of weights for each hypothesis $h$, resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
 The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
@@
 
-This expression equates to identifying the best empirical policy based on previous data.
+This expression equates to identifying the empirically best policy based on previous data.
 The algorithms assume the existence of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
@@
 
-It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
+The definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$ (see figure \ref{fig:contextual-bandit-causal-graph}).
 However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
@@
 As in standard roulette, it consists of a spinning wheel with a ball that can land in one of 37 numbered pockets.
-However, unlike in the standard game, where punters place bets before the wheel is spun, in this variant the wheel is spun first.
+However, unlike the standard game where punters place bets before the wheel is spun, in this variant the wheel is spun first.
 Five seconds before it comes to rest the face of the wheel darkens to hide the position of the ball from the players.
@@
 
-It is difficult to come up with realistic examples in which the full sequence of rewards is generated but only the one associated with the selected action is observable to the learner.
-In the typically cited applications of contextual bandit algorithms, such as selecting how to treat patients or which ad to serve only the reward of the selected action is ever realised, and the alternate rewards are counterfactual variables.
+It is difficult to come up with realistic examples in which the full sequence of rewards is generated, but only the one associated with the selected action is observable to the learner.
+In the typically cited applications of contextual bandit algorithms, such as selecting how to treat patients or which ad to use, only the reward of the selected action is ever realised, and the alternate rewards are counterfactual variables.
 
@@
 As a result, there is no exploration-exploitation trade-off to be made.
-However, the problem does not reduce to supervised learning, because of the bandit nature of the feedback - only the reward of the selected action is observed.
+However, the problem does not reduce to supervised learning because of the bandit nature of the feedback; only the reward of the selected action is observed.
 In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
@@
 
-If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, discussed in section \ref{sec:causal_effect_under_ignorability}.
-The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ satisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
+If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, as we discussed in section \ref{sec:causal_effect_under_ignorability}.
+The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that affect the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ satisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
 The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivalent to $do(A=a)$.
-However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see \citet{Pearl2000}, section 4.2.
+However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see section 4.2.
+of \citet{Pearl2000}.
 In this case, letting $\Pns{\pi'}{a|\vec{x}}$ denote the distribution over actions under policy $\pi'$ given context $\vec{x}$, the expected (per round) reward obtained by $\pi'$ is given by,
@@
 
-We conclude this chapter on bandits with a note on the counterfactual nature of regret.
-I have not seen this discussed with respect to bandit regret elsewhere, however the issues raised are very closely related to the work by \citet{Dawid2000}, on problems associated with the use of counterfactual variables for (observational) causal inference.
+We conclude this chapter on bandits with a note on the counterfactual nature of regret, which has not been discussed with respect to bandit regret.
+However, the issues raised are very closely related to the work by \citet{Dawid2000} on problems associated with the use of counterfactual variables for (observational) causal inference.
 
 It is possible, as we demonstrated in example \ref{exm:anti-casual-context}, to construct bandit problems in which the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$ is actually generated at each time-step (and, at least in principle, is observable to someone although not the agent).
-However, in most real applications for bandit problems, only the outcome associated with the selected action is ever realised and so the remaining components of the reward vector are counterfactual, justifying our use of the counterfactual notation to denote them.
+However, in most real applications for bandit problems, only the outcome associated with the selected action is ever realised and so the remaining components of the reward vector are counterfactual, justifying the use of the counterfactual notation to denote them.
 There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
 This makes it easy to make statements using counterfactuals that cannot be confirmed empirically (even with infinite experimental data).
-We now show with a simple example that the standard definition of regret is such a fundamentally counterfactual quantity.
+We now show with a simple example that the standard definition of regret is a fundamentally counterfactual quantity.
 Recall that the cumulative regret is defined by,
@@
 However, the covariance $\rho$ cannot, because we never simultaneously observe the rewards for both arms.
-As a result, even if we have full knowledge of the reward distributions for each arm, we cannot in general compute the distribution over regret without un-testable assumptions about the covariance between counterfactuals, see figure \ref{fig:metaphysical_distribution_difference}.
+As a result, even with full knowledge of the reward distributions for each arm, the distribution over regret cannot, in general, be computed without un-testable assumptions about the covariance between counterfactuals (see figure \ref{fig:metaphysical_distribution_difference}).
 
@@
 This result is somewhat perturbing.
-The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep and the behaviour of standard bandit algorithms depends only on the marginal reward distributions for each arm.
-The expectation of the regret as defined by equation \ref{eqn:pseudo_regret} also remains unchanged as both it's definition and the learner's actions depend only on the marginal distributions.
-It seems therefor unfortunate that we should have to assume, for example, that the rewards for alternate actions are independent of one-another in order to be able to analyse the variance of the regret.
-Particularly as this assumption is likely to be violated in many realistic bandit problems.
-For example, a given user may be more (or less) likely to buy something no matter which ad they are served leading to a positive correlation between counterfactual rewards, or an illness might have two (unobservable) sub-types with each medication (bandit arm) effective only against one, resulting in a negative correlation between the counterfactual rewards.
+The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep, and the behaviour of standard bandit algorithms depends only on the marginal reward distributions for each arm.
+The expectation of the regret as defined by equation \ref{eqn:pseudo_regret} also remains unchanged as both its definition and the learner's actions depend only on the marginal distributions.
+It seems therefor unfortunate that we should have to assume, for example, that the rewards for alternate actions are independent of one-another to be able to analyse the variance of the regret.
+This is particularly so, as this assumption is likely to be violated in many realistic bandit problems.
+For example, a given user may be more (or less) likely to buy something no matter which advertisement they are served, leading to a positive correlation between counterfactual rewards.
+Equally, an illness might have two (unobservable) sub-types, with each medication (bandit arm) effective only against one, resulting in a negative correlation between the counterfactual rewards.
 
@@
 For example, if we are risk averse, we may prefer an algorithm with slightly higher expected regret but a lower probability of suffering extremely large regret.
-This raises the question, can we construct an alternate definition of regret that allows us to capture how consistently bandit algorithms behave, but which does not depend on any properties of the joint distribution over counterfactual rewards.
-A natural candidate would be,
+This raises the question, is it possible to construct an alternate definition of regret that can capture how consistently bandit algorithms behave, but that does not depend on any properties of the joint distribution over counterfactual rewards?
+A natural candidate would be:
 
@@
 Furthermore, this quantity has already been analysed in existing work on the concentration of bandit regret, \citep{Audibert2007,Audibert2009exploration} as a more tractable proxy to the standard regret.
-In conclusion, when selecting measures of bandit performance, it is worth noting whether or not they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.
+In conclusion, when selecting measures of bandit performance, it is worth noting whether they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.
 
@@
 
-However, despite the commonality of the underlying problem, there has been relatively little work on intersection between observational causal inference and adaptive decision making processes.
+However, despite the commonality of the underlying problem, there has been relatively little work on the intersection between observational causal inference and adaptive decision making processes.
 Although some ideas, such as doubly robust estimators and inverse propensity score weighting have transferred across (from causal inference to off-policy evaluation), there are many more opportunities for fruitful work in this space.
-In most key real world decision making processes, from algorithms recommending drug treatments to robots deciding how to open doors, we have access to both potentially huge observational data sets as well as the ability to intervene in the system at some level.
+In most key real world decision making processes, from algorithms recommending drug treatments to robots deciding how to open doors, we have access to both potentially huge observational data sets, as well as the ability to intervene in the system at some level.
 It is critical that we develop methods that allow us to incorporate the causal knowledge we have about the world, along with data sets collected under a range of different conditions, into our current decision making process.
-In this chapter I introduce a very general framework that connects causal graphical models with multi-armed bandit problems and demonstrate how it can be leveraged to make better decisions.
+In this chapter, I introduce a very general framework that connects causal graphical models with multi-armed bandit problems and demonstrate how it can be leveraged to make better decisions.
 
@@
 The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node.
+However, for simplicity, we will consider the reward to be the value of a single specified node.
 We refer to these problems as \emph{causal bandit problems}.
-In this thesis I focus on the case where the causal graph is known.
+In this thesis, I focus on the case where the causal graph is known.
 Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.
@@
 variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry.
-When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
+Where there is no intervenion, the temperature, soil, and moisture vary naturally from season to season due to weather conditions.
+These are all observed along with the final crop yield at the end of each season.
 How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
@@
 
-The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
-If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant and the problem reduces to a standard multi-armed bandit problem.
+The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action, and on which variables we receive post-action feedback.
+If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant, and the problem reduces to a standard multi-armed bandit problem.
 
@@
 Thus we gain information about the reward for the action $do(W=0)$ from selecting the action $do()$ or $do(X = x)$ and then observing $W = 0$.
-We only get this form of side information for actions that don't specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-If additional variables are only observed before an intervention is selected the causal bandit problem can be treated as a stochastic contextual bandit problem, which are already reasonably well understood \citep{Agarwal2014}.
-
-Two other problems that sit in the space between causal inference and bandit problems, bandits with unobserved confounders \cite{Bareinboim2015} and compliance aware bandits \citep{DellaPenna2016}, can also be viewed as specific causal bandit problems.
-In the work on bandits with unobserved confounders it is assumed that the reward given each action may depend on some latent variable $U$ which we cannot observe directly.
+We only get this form of side information for actions that do not specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
+If additional variables are only observed before an intervention is selected, the causal bandit problem can be treated as a stochastic contextual bandit problem, which is already reasonably well understood \citep{Agarwal2014}.
+
+Two other problems that sit in the space between causal inference and bandit problems - bandits with unobserved confounders \cite{Bareinboim2015} and compliance aware bandits  \citep{DellaPenna2016} - can also be viewed as specific causal bandit problems.
+In the work on bandits with unobserved confounders, it is assumed that the reward given each action may depend on some latent variable $U$, which we cannot observe directly.
 However, prior to selecting an action, we can observe  $I$, the action that would have been selected under an alternate (stationary) policy, which may depend on $U$, see figure \ref{fig:confounded-causal-bandit-graph}.
-In this case the set of contextual variables $\vec{X}^c = I$, the set of post-action feedback variables $\vec{X}^o = \set{Y}$ and the action space consists of all possible assignments of values to a single node $X$, $\actionspace = do(X=x)$.
+In this case, the set of contextual variables $\vec{X}^c = I$, the set of post-action feedback variables $\vec{X}^o = \set{Y}$ and the action space consists of all possible assignments of values to a single node $X$, $\actionspace = do(X=x)$.
 This setting reduces to a contextual bandit problem in our causal bandit framework.
-However in their work on bandits with unobserved confounders \citet{Forney2017} also leverage the fact that $I$ represents the action that would have been selected under an alternate policy to fuse data collected under the previous (observational) policy with data collected under the new policy.
+However in their work on bandits with unobserved confounders, \citet{Forney2017} also leverage the fact that $I$ represents the action that would have been selected under an alternate policy to fuse data collected under the previous (observational) policy with data collected under the new policy.
 This information is not encoded in the causal bandit graph in figure \ref{fig:confounded-causal-bandit-graph}, as $I$ could be any variable that is influenced by the unobserved context $U$, and thus cannot be exploited by a standard contextual bandit algorithm.
 
-Compliance aware bandits describe situations in which the action recommended by the bandit algorithm is not always followed.
-For example a patient may refuse to take a treatment or an advertiser may have complex rules about how many ads a given customer can receive which prevents some of the suggestions from the ad recommendation engine from being followed.
-After an action is selected the algorithm can observe the action that was actually taken in addition to the reward.
+Compliance-aware bandits describe situations in which the action recommended by the bandit algorithm is not always followed.
+For example, a patient may refuse to take a treatment or an advertiser may have complex rules about how many advertisements a given customer can receive, which prevents some of the suggestions from the ad recommendation engine from being followed.
+After an action is selected, the algorithm can observe the action that was actually taken, in addition to the reward.
 \citet{DellaPenna2016} analyse this setting with binary treatments both with and without the presence of a latent confounding variable $U$, see figure \ref{fig:imperfect-compliance-causal-bandit-graph}.
@@
 By making some functional assumptions about the relationships between the variables, we can use $A$ as an instrumental variable to bound or estimate the (casual) effect of $X$ on $Y$.
-The estimation will be somewhat complicated in the bandit setting because the action chosen at each timestep is dependent on the previous sequence actions and rewards.}
+The estimation will be complicated in the bandit setting because the action chosen at each timestep is dependent on the previous sequence of actions and rewards.}
 
@@
 Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on un-played actions are revealed according to a feedback graph.
-As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
+As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework.
+However, the regret bounds are not optimal in our setting.
 They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
-The partial monitoring approach taken by \citet{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
+The partial monitoring approach taken by \citet{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy require knowledge about the likelihood of each factor in advance, while our strategy learns this online.
 \citet{Yu2009} utilise extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
@@
 
-Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
-\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been 
-intervened on can provide important contextual information. Their work differs in many ways. For example, the focus is on the cumulative regret and
-the context is observed before the action is taken and cannot be controlled by the learning agent.
+Two pieces of recent work also consider applying ideas from causal inference to bandit problems.\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been
+intervened on can provide important contextual information.
+Their work differs from this thesis in many ways.
+For example, the focus is on the cumulative regret, and the context is observed before the action is taken and cannot be controlled by the learning agent.
 
@@
 
- \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions.
+\citet{Ortega2014thompson} present an analysis and extension of Thompson sampling, assuming actions are causal interventions.
 Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
@@
 \label{sec:simple-regret}
-In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem which we call the {\it parallel bandit}.
+In this section, we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem, which we call the {\it parallel bandit}.
 It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro_causal_bandit}, including the farming example.
-It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
+It also suffices to witness the regret gap between algorithms that make use of causal models and those that do not.
 
@@
 
-In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high).
+In the farming example above, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high).
 The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.
@@
 \begin{remark} If the actionspace $\mathcal{A}$ is unconstrained, that is consists of all possible assignments of values to variables, (and all actions have equal cost) the optimal action will set the value of all the parents of $Y$ and Algorithm \ref{alg:general} cannot do better than uniform exploration over these arms.
-In this case, after we use the causal structure to eliminate irrelevent actions prior to taking any samples, the problem of selecting within the remaining actions can be treated as a standard multi-armed bandit problem.
+In this case, after we use the causal structure to eliminate irrelevant actions prior to taking any samples, the problem of selecting within the remaining actions can be treated as a standard multi-armed bandit problem.
 \end{remark}
@@
 Note that we now only have to obtain estimates for actions $a \in \mathcal{A'}$, since we know the others to be sub-optimal, so the $max$ is only over these actions.
-However we $b$ still sums over all possible actions in the demoninator of the importance sampling estimator, to allow for the possiblity that playing sub-optimal actions allows more efficient estimation of the optimal actions.
+However we $b$ still sums over all possible actions in the denominator of the importance sampling estimator, to allow for the possibility that playing sub-optimal actions allows more efficient estimation of the optimal actions.
 We now prove that this is not the case and that, in this specific setting, Algorithm \ref{alg:general} cannot do better than uniform sampling over the actions $a \in \mathcal{A'}$.
@@
 Observational causal inference provides, through the do-calculus, a formal means to map information from observational to interventional settings, as well as between different interventions.
-Bandit algorithms,
+Bandit algorithms, capture the sequential nature of decision making processes and provide techniques to carefully balance exploration and exploitation.
 

### 2017-09-10 thesis/abstract.tex
fixes title pages

--- 2017-09-04 thesis/thesis_map.tex
@@
-\documentclass[11pt,a4paper,twoside]{report}
-\usepackage[table]{xcolor}
-\usepackage[utf8]{inputenc}
-\usepackage{textcomp}
-\usepackage{amsmath}
-\usepackage{dsfont}
-\usepackage{listings}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{tabularx}
-\usepackage{booktabs}
-\usepackage{enumitem}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithm
-\usepackage[noend]{algorithmic}
-\usepackage{parskip}
-\usepackage{tikz}
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{epigraph}
-\usetikzlibrary{arrows,fit,positioning}
-\usepackage{subcaption}
-\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[numbers]{natbib} %numbers
-\usepackage{wrapfig}
-\usepackage{centernot}
-
-% this is included to get tables with colored rows to not draw over hlines and vlines
-\usepackage{etoolbox}
-
-\makeatletter
-\patchcmd{\@classz}
-  {\CT@row@color}
-  {\oldCT@column@color}
-  {}
-  {}
-\patchcmd{\@classz}
-  {\CT@column@color}
-  {\CT@row@color}
-  {}
-  {}
-\patchcmd{\@classz}
-  {\oldCT@column@color}
-  {\CT@column@color}
-  {}
-  {}
-\makeatother
-
-\usepackage{amsthm}
-
-\graphicspath{ {figures/} }
-\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
-
-\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
-\DeclareUnicodeCharacter{00A0}{ }
-
-\tikzset{
-    %Define standard arrow tip
-    >=stealth',
-    %Define style for boxes
-    punkt/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           text width=1em,
-           minimum height=1em,
-           text centered},
-    observed/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.3em,
-           minimum height=2.3em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-     latent/.style={
-           circle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered
-           },
-    target/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-    observedrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-    latentrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered
-           },
-     targetrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-     empty/.style={
-           circle,
-           rounded corners,
-           minimum width=.5em,
-           minimum height=.5em,
-           font=\footnotesize,
-           text centered,
-           },
-    % Define arrow style
-    pil/.style={
-           o->,
-           thick,
-           shorten <=2pt,
-           shorten >=2pt,},
-    sh/.style={ shade, shading=axis, left color=red, right color=green,
-    shading angle=45 }   
-}
-
-\newcommand{\KL}{\operatorname{KL}}
-
-\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
-\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
-\newcommand{\Q}[1]{\operatorname{Q}\left(#1\right)}
-\newcommand{\EE}{\mathbb E}
-\newcommand{\EEa}{\EE_a}
-\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
-\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
-\newcommand{\actions}{\mathcal{A}}
-\newcommand{\calA}{\mathcal A}
-\newcommand{\etc}{\textit{etc}}
-\newcommand{\ie}{\textit{i.e.}}
-\newcommand{\eg}{\textit{e.g.}}
-\newcommand{\calP}{\mathcal P}
-\newcommand{\x}{\boldsymbol{x}}
-\newcommand{\Ps}{\operatorname{P}}
-\newcommand{\R}{\mathbb R}
-\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
-\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
-\newcommand{\indep}[1]{\mathcal{I}\left(#1\right)}
-\newcommand{\vb}[1]{\boldsymbol{#1}}
-\newcommand{\card}[1]{\left\vert{#1}\right\vert}
-\newcommand{\actionspace}{\mathcal{A}}
-
-\newcommand{\V}[1]{\mathbb V\left[#1\right]}
-\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
-\newcommand{\defined}{\vcentcolon =}
-\newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
-\newcommand{\Var}{\operatorname{Var}}
-\newcommand{\calF}{\mathcal F}
-\newcommand{\sr}[1]{\stackrel{#1}}
-\newcommand{\set}[1]{\left\{#1\right\}}
-\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
-
-\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
-\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
-\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
-\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
-\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
-\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
-\newcommand{\Ber}{\operatorname{Bernoulli}}
-\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
-\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
-\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
-\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
-\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
-
-\renewcommand{\P}[1]{\operatorname{P}\left(#1\right)}
-\newcommand{\cf}[2]{{#1}^{#2}}
-\newcommand{\cfb}[2]{{#1}_t^{#2}} % used to represent reward
-\newcommand{\bernoulli}{\operatorname{Bernoulli}}
-\newcommand{\dirac}{\operatorname{Dirac}}
-\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
-\renewcommand{\vec}[1]{\boldsymbol{#1}}
-\newcommand{\regret}{\bar{R}_{T}} %psuedo-regret
-\newcommand{\expectedregret}{\mathbb{E}R_{T}} % expected regret
-\newcommand{\rawregret}{R_{T}} % plain regret
-\newcommand{\simpleregret}{R_T}
-\newcommand{\quotes}[1]{``#1''}
-
-\theoremstyle{plain} 
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\let\temp\epsilon
-\let\epsilon\varepsilon
-
-\newcommand{\ci}{\mathrel{\perp\mspace{-10mu}\perp}}
-\newcommand{\nci}{\centernot{\ci}}
-
-\author{Finnian Lattimore}
-\title{Learning how to act: making good decisions with machine learning}
-
-\setcounter{secnumdepth}{3}
-
-\begin{document}
-
-%\def\ci{\perp\!\!\!\perp} % from Wikipedia
-
-\begin{titlepage}
-    \begin{center}
-        \vspace*{1cm}
-
-        \Huge
-        {Learning how to act: making good decisions with machine learning}
-
-        \vspace{2cm}
-        \Large
-        {Finnian Lattimore}
-
-        \vfill
-        \Large
-        A thesis submitted for the degree of Doctor of Philosophy
-
-         The Australian National University
-
-        \vspace{1cm}
-
-        September 2017
-
-    \end{center}
-\end{titlepage}
-
-\epigraph{In vain the grave, with retrospective Eye,
-Would from the apparent what conclude the why,
-Infer the Motive from the Deed, and show
-That what we chanced, was what we meant, to do.}{Alexander Pope}
-
-\clearpage
-\vspace*{\fill}
-\begin{center}
-\begin{minipage}{.9\textwidth}
-Except where acknowledged in the customary manner, the material 
-presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any 
-university.
-
-\vspace{40mm}  % vertical space
-
-\hspace{80mm}\rule{40mm}{.15mm}\par   % horizontal space, line, start new line
-\hspace{80mm} Finnian Lattimore\par
-\hspace{80mm} September, 2017
-\end{minipage}
-\end{center}
-\vfill % equivalent to \vspace{\fill}
-\clearpage
-
-\begin{abstract}
+\chapter*{Abstract}
 
@@
 I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis.
-I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms.
+I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be exploited by existing algorithms.
 I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.
-
-\end{abstract}
-
-%\chapter*{Acknowledgements}
-
-\tableofcontents
-
-\chapter{Introduction}
-
-\section{Motivation}
-Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something.
-Will asking people to pay upfront at the doctors reduce long term health expenditure?
-If we developed a drug to suppress particular genes, could we cure multiple sclerosis and would delaying teen-aged pregnancies improve the outcome for their children.
-
-These are hard questions because they require more than identifying a pattern in data.
-Correlation is not causation.
-Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, despite the fact that the variables involved are susceptible to human intuition and understanding.
-
-We now live in a world of data.
-Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and where the US Institute of Health is considering if all infants should be genetically sequenced at birth.
-Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by-product of some other purpose.
-It does not come from the controlled world of a randomised experiment.
-
-The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
-From health care, to entertainment and self-driving cars, machine learning algorithms will transform many industries.
-It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory \citep{anderson2008end} and that we may be only a short step from \quotes{The Singularity}, where artificial intelligence exceeds our own and then grows exponentially.
-
-However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within narrow problem settings.
-Getting them to generalise to even slightly different problems or data sets remains very challenging.
-Deciding how we should act or what policies we should implement requires predictions about how a system will behave if we change it.
-The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on.
-As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live in, it is critical to ensure we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
-
-\section{What is causality? }
-
-The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined.
-This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
-However, the magnitude, direction and even existence of an association depends which variables are controlled for (or included in a regression).
-Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.
-
-There are two ways in which an association detected in a data set may be non-causal.
-The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample.
-Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated.
-For example, based on data from the years 1999 to 2009, the age of Miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}.
-We would not expect this relationship to hold in a new sample of data.
-This form of spurious correlation also has serious repercussions.
-It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published.
-Consequently, the likelihood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
-This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}.
-This issue can be ameliorated by obtaining more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.
-
-However, a strong association, observed in multiple independent studies may still not be causal.
-The correlation can arise because both variables are consequences of some other, unmeasured factor.
-For example, the reading ability of children under twelve is strongly correlated with their height, because older children are taller and can read better.
-However height is not a cause of reading ability because interventions to increase a child's height, for example by giving them growth hormones, would not be expected to improve their reading.
-Similarly, extra lessons in reading will not make a child grow taller.
-This problem is fundamentally different to the issue of spurious correlations arising by chance in finite data sets.
-Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to separate causation from correlation.
-
-The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables.
-In this thesis, I take an interventionist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal.
-This viewpoint captures the types of questions that motivate this thesis.
-How can we change the way we do things to obtain better outcomes?
-
-Causality is often liked to explanation; understanding how and why things happen.
-I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture.
-This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction.
-A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change.
-A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed.
-More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts.
-By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
-
-\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{what if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs.
-The former aims to identify the effect of a known cause.
-The latter can be viewed as identifying causes of an effect.
-They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, that we describe in section \ref{sec:causal_models}.
-However, they state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined.
-The goal of identifying the causes of an effect can be formalised within the graphical causal model framework.
-Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on a given outcome, since it requires us to test or infer the effect of interventions on many different variables.
-These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are often relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "what if" questions are hard to resolve conclusively.
-However, this does not mean that the problem of identifying causes of effects is ill-posed in principle.
-It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see for example \citep{}.
-\todo{reference a review paper on causal discovery here}
-
-There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models.
-However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models.
-In the next section we emphasise the subtley that can exist in determining if a problem is causal by examining some typical examples.
-
-\section{What makes a problem causal?}
-
-Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them.
-Machine learning techniques are being applied a huge range of problems, in both industry and academia.
-The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to.
-Which, if any, of these problems require causal inference?
-
-\begin{itemize}
-\item Speech recognition (for systems like Siri or Google Assistant)
-\item Image classification
-\item Forecasting the weather
-\item Identifying spam emails
-\item Automated essay marking
-\item Predicting the risk of death in patients with pneumonia.
-\item Predicting who will re-offend on release from prison 
-\item Customer churn modelling
-\item Demand prediction for inventory control
-\item Playing Go 
-\end{itemize}
-
-The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important consideration.
-In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.
-
-Consider speech recognition.
-You say something, which causes sound waves, which are converted to a digital signal that Siri maps to words.
-Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution).
-A similar argument could be made for many applications of machine translation and image classification.
-
-In image classification we do not particularly care about building a strong model for exactly how the thing that was photographed translates to an array of pixels, provided we can be fairly confident that the process will not change.
-If we develop a discriminative model that is highly accurate at classifying cats from dogs, we do not need to understand its internal workings (assuming we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
-
-What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella, causality can be ignored.
-Your decision will not affect if it actually rains.
-However, longer term climate forecasts might (theoretically) lead to action on emissions which would then change the weather system.
-For this we need a (causal) model that allows us to predict the outcome under various different interventions.
-
-Identifying spam and automated essay marking both involve processing text to determine an underlying (complex) attribute such as its topic or quality.
-In both cases, there is inherent competition between the algorithm and the people generating the text.
-As a result, decisions made by the algorithm are likely to change the relationship between the features it relies on and the true label.
-Spammers and students will modify their writing in order to optimise their results.
-A standard supervised learning approach can only work if the resulting change in the mapping from features to label is sufficiently gradual.
-There are two key ways of ensuring this.
-The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm.
-The second is to use a model in which the features are related to the outcome in such a way that they cannot be manipulated independently.
-
-This example also highlights a connection between causal models and transparency in machine learning.
-If we are using a non-causal model to make decisions affecting people, there will be a trade-off between the performance and transparency of the model; not because the requirement for transparency restricts us to simple models, but because revealing how the model works allows people to change their behaviour to game it.
-
-What about predicting the risk of death in patients with pneumonia?
-Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective for serious cases, this appears to be straightforward prediction.
-It is not.
-Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admission decisions.
-\citet{Caruana2015} found exactly this effect in a real data set.
-The model learned that people suffering asthma were \emph{less} likely to die from pneumonia.
-This was because doctors treated such patients very aggressively, thus actually lowering their risk.
-The issue is not with model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols.
-However, using it to decide how to \emph{change} these protocols could kill.
-The actual question of interest in this case is what happens to patients with characteristics $X$ they are assigned treatment according to decision rule (or policy) $\pi(X)$.
-
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
-Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.
-
-Demand prediction seems like a relatively straightforward prediction problem.
-These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
-It seems unlikely that using the model to ensure stock is available will itself change demand.
-However, depending on the way demand is measured, there is a potential data censoring issue.
-If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
-Changing availability does then change demand.
-
-Playing Go (and other games) is a case with some subtleties.
-At every turn, the AI agent has a number of actions available.
-The board state following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem.
-There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}.
-Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$.
-This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players.
-Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}.
-The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
-
-The problem of playing go is causal from the interventionalist perspective.
-The agent wishes to learn the probability of a win given an action they take.
-However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach.
-The actions the agent has to explore are the sames ones as human players explored to generate the training data, and both have the same objective - to win the game.
-In addition, the state of the board encapsulates all the information relevant to selecting a move.
-These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.
-
-Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predictive) inference.
-A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes.
-The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people.
-Think about problems like credit scoring and parole decisions.
-There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency.
-It is rare to build a model without any intent to make some kind of decision on the basis of its results.
-Thus, I argue we should assume a causal model is required until we can justify otherwise.
-
-\section{Observe or intervene: two very different approaches to causal problems}
-
-As we have shown, problems involving causality are ubiquitous in many fields.
-As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning.
-Although the focus and terminology can differ substantially between fields these techniques, all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system.
-Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.
-
-In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them.
-We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal.
-A particular strand of research within reinforcement learning are multi-armed bandit problems.
-They describe settings in which there are a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action.
-They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.
-
-The approach of learning the outcome of an action by taking it has plays a key role in advancing our knowledge of the world.
-However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen.
-Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.
-Observational causal inference can be viewed as a form of transfer learning.
-The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system.
-This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another.
-The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.
-
-Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials.
-Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem randomisation is not always feasible, affordable or ethical.
-The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
-
-\section{This thesis and its contributions}
-
-I view causal problems as one of the greatest arising challenges for machine learning.
-They incorporate a large set of problems, of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence.
-For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry we need to understand how to leverage our improved approaches to prediction to tackle causal problems.
-
-\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints.
-My major contribution is a framework which unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
-This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
-I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
-My other contributions are primarily connective rather than technical.
-I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias.
-I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data and covariate shift.
-I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.
-
-\paragraph{Thesis overview} This thesis is divided into three key chapters, learning from observational data, learning from interventional data and unifying the approaches.
-Chapter \ref{chap:causality} covers learning to act from observational data.
-In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems.
-Section \ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \ref{sec:defining-causal-effects} and \ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation.
-Section \ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.
-
-Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems.
-In \S\ref{sec:multi-armed-bandits} I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}).
-I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}.
-In \S\ref{sec:off-policy-evaluation} I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of of causal effect estimation from observational data.
-Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.
-
-In chapter \ref{chap:causal-bandit} I introduce causal bandit problems: a framework which unifies causal graphical models with multi-armed bandit problems.
-Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model.
-I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.
-
-In \S\ref{sec:causal-bandit-post-action-feedback} I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected.
-I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms.
-In \S\ref{sec:simple-regret} I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}.
-I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning.
-In \S\ref{sec:simple-regret-general} I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}.
-Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems.
-Finally in \S\ref{sec:causal_bandit_discussion} I discuss extensions and potential future work.
-
-\chapter{Learning from observational data}
-\label{chap:causality}
-
-The goal of causal inference is to learn the effect of taking an action.
-We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world.
-We are generating and storing data on almost every aspect of our lives at an unprecedented rate.
-As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase.
-However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question.
-An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited.
-This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it.
-This is the problem of observational causal inference.
-
-\section{Causal models}
-\label{sec:causal_models}
-
-Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
-As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention.
-Mapping properties from one system to another requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate a system will respond to them.
-Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
-
-Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings.
-They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another.
-Counterfactual assertions can be expressed very naturally in human languages and are quite prevalent in  every day conversations;\quotes{if I had worked harder I would have got better grades}, \quotes{she would have been much sicker if she hadn't taken antibiotics}.
-Structural equation models have been developed and applied primarily within economics and related disciplines and can be seen as an attempt to capture key aspects of the behaviour of people with mathematics.
-Questions around designing policies or interventions play a central role in economics.
-Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised) framework and associated set of methods for estimating causal effects.
-Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks.
-They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of conditions under causal effects can be inferred from observational data.
-
-An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}.
-However, the literature on causal inference techniques remains spit between the different frameworks.
-Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scores, are described using the counterfactual framework.
-Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions are often based around structural equation models.
-This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.
-
-In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ.
-We hope that by describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them.
-However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers.
-In order to demonstrate the notation and formalisms each framework provides we will use them to describe the following simple examples.
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:ranomized_experiment}
-Suppose a pharmaceutical company wants to assess the effectiveness of a new drug on recovery from a given illness.
-This is typically tested by taking a large group of representative patients and randomly assigning half of them to a treatment group (who receive the drug) and the other half to a control group (who receive a placebo).
-The goal is to determine the clinical impacts of the drug by comparing the differences between the outcomes for the two groups (in this case, simplified to only two outcomes - recovery or non-recovery).
-We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
-\end{example}
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:adjusting}
-Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds.
-We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
-We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
-\end{example}
-
-\subsection{Causal Bayesian networks}
-
-Causal Bayesian networks are an extension of Bayesian networks.
-A Bayesian network is a graphical way of representing how a distribution factorises.
-Any joint probability distribution can be factorised into a product of conditional probabilities.
-There are multiple valid factorisations, corresponding to permutations of variable ordering.
-
-\eqn{
-\label{eqn:cbn:joint_dist}
-P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
-}
-
-We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}.
-If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network.
-There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below left=of 1](2){$X_{2}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\caption{A general Bayesian network for the joint distribution over three variables.
-This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
-\label{fig:bayesnet}
-\end{figure}
-
-\begin{figure}
-\centering
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (3) edge (2)
-    (2) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, below left=of 1](2){$X_{1}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, above left=of 1](2){$X_{1}$};
-\node[main node, above right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (2) edge (1)
-    (3) edge (1)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$.
-Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical structure implies exactly the same set of independencies exhibited by the distribution.
-Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
-\label{fig:bayesnet2}
-\end{figure}
-
-The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true: $G$ can have extra edges).
-If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
-
-\eqn{
-\P{X_{1},...,X_{N}} = \prod_{i = 1...N}\P{X_{i}|parents_{X_{i}}}
-}
-
-A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
-This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-We need some notation to describe interventions and represent distributions over variables in the network after an intervention.
-In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.
-
-\vspace{0.5cm}
-\begin{definition}{The do-notation}
-\begin{itemize}
-\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
-\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$.
-This notation is somewhat overloaded.
-It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
-For example it could represent
-\begin{itemize}
-\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
-\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
-\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
-\item a family of density/mass function for $Y$ parameterised by $x$.
-\end{itemize}
-Where the distinction is important and not clear from context we will use one of the more specific forms above. 
-\end{itemize}
-\end{definition}
-
-\vspace{0.5cm}
-\begin{theorem}[Truncated product formula \citep{Pearl2000}]
-\label{thm:truncated_prodcut}
-If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.
-Let $\parents{X_i}$ denote the parents of the variable $X_i$ in $G$.
-
-\begin{equation}
-\label{eq:truncatedproduct}
-\P{X_1...X_N|do(Z=z)} = \ind{Z = z}
-  \prod_{X_i \notin Z}\P{X_{i}|\parents{X_i}}
-\end{equation}
-\end{theorem}
-
-Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution.
-The truncated product formula will give different results depending on the selected network.
-The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
-In fact, from the interventionist viewpoint of causality, the truncated product formula defines what it means for a link to be causal.
-
-Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$?
-The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient.
-Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables.
-The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}.
-This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$.
-In this example, the interventional distribution is equivalent to the observational one.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X \text{ (Treatment)}$};
-\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
-\path[]
-	(2) edge (3);
-\end{tikzpicture}
-\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
-\label{fig:causal_network_example}
-\end{figure}
-
-In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool.
-We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$.
-It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation.
-If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}.
-In this case, the interventional distribution is not equivalent to the observational one.
-If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool.
-To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,
-
-\eqn{
-\label{eqn:backdoor_example}
-\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}
-}
-
-We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph.
-There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure.
-These rules are referred to as the Do-calculus \citep{Pearl2000}.
-We discuss them further in section \ref{sec:do_calculus}.
-
-\begin{figure}
-\center
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]
-
- %nodes
-\node[node](1){$Z \text{ (SES)}$};
-\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
-\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3); 
-\end{tikzpicture}
-\caption{Causal Bayesian network for example \ref{exm:adjusting}}
-\label{fig:causal_adjust} 
-\end{figure}
-
-A causal Bayesian network represents much more information than a Bayesian network with identical structure.
-A causal network encodes all possible interventions that could be specified with the do-notation.
-For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
-The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
-After each two variable intervention, we have a distribution over one variables which requires one parameter.
-This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
-
-Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes.
-The global Markov property and d-separation also hold in causal networks.
-D-separation, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
-We include a brief review of D-separation in section \ref{sec:dsep}.
-
-\subsubsection{Limitations of causal Bayesian networks}
-A number of criticisms have been levelled at this approach to modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}.
-However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions.
-For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
-But, what if some people failed to take the prescribed treatment?
-We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
-Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Prescribed Treatment};
-\node[node, below right=of 1](2){Actual Treatment};
-\node[node, right=of 2](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(1) edge (2)
-	(2) edge (3) 
-	(4) edge (3) edge (2);
-\end{tikzpicture}
-\caption{Randomised treatment with imperfect compliance}
-\label{fig:randomized_imperfect_compliance}
-\end{figure}
-
-Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables.
-Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology.
-We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
-However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
-Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
-Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
-\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
-
-\subsection{Counterfactuals}
-\label{sec:counterfactuals}
-
-The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
-Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
-
-In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would have happened if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would have happened if they got the placebo, $\cf{y_{i}}{0}$.
-The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, there are units that can be assumed to be identical or that will revert exactly to their initial state some time after treatment.
-This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
-
-Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo.
-Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate (from an experimental or observational study):
-\begin{itemize}
-\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
-\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
-\end{itemize}
-
-Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
-For those who did not take the treatment, the observed outcome is the same as the outcome \emph{had} they not taken the treatment.
-Equivalently stated:
-
-\eq{
-\P{Y^{0}|X=0}&= \P{Y|X=0}\\
-\P{Y^{1}|X=1}&=\P{Y|X=1}
-}
-
-If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
-
-\eq{
-\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
-\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
-}
-
-This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
-
-\eq{
-\P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
-}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}.
-They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
-In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e.
-$X \nci \cf{Y}{0}$.
-Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e.
-$X \nci \cf{Y}{1}$.
-A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
-
-\vspace*{.3cm}
-\begin{theorem}[Ignorability \citep{Rosenbaum1983, Pearl2000}] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
-
-\eqn{
-\label{eqn:counterfactual1}
-\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
-\label{eqn:counterfactual2}
-\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
-}
-\end{theorem}
-
-Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
-Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
-
-Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}.
-This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
-As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
-The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
-SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
-
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-In particular, statements involving joint distributions over counterfactual variables may not be able to be validated empirically \citet{Dawid2000}.
-One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
-This is certainly convenient, if rather imprecise.
-However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention.
-It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).
-
-\subsection{Structural Equation models}
-\label{sec:SEM}
-
-Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Uncertainties are not inherent but arise from unmeasured variables.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \citep{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(\epsilon_x)$};
-\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:ranomized_experiment}}
-\label{fig:sem_randomized_treatment}
-\end{figure}
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random.
-The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
-The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
-
-Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$.
-The function $f_y$ and distribution over $\epsilon_y$ does not change.
-This results in the interventional distribution, \footnote{We have assumed the variables are discrete only for notational convenience}
-
-\eqn {
-\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
-}
-
-The observational distribution of $Y$ given $X$ is,
-
-\eqn{
-\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
-& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
-}
-
-The interventional distribution is the same as the observational one.
-The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
-
-The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
-Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
-
-\eqn{
-\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
-\label{eqn:sem:adjusting}
-&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
-}
-Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivalent to the observational distribution given by:
-
-\eqn{
-\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
-}
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(Z,\epsilon_x)$};
-\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
-\node[lt, above=of 6](7){$\epsilon_z$};
-\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2)
-	(7) edge (6)
-	(6) edge (2) edge (3);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:adjusting}}
-\label{fig:sem:preschool}
-\end{figure}
-
-Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
-A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
-
-\subsection{Comparing and unifying the models}
-\label{sec:unifying_causal_models}
-
-Clarify Cheng's Questions - is there a mechanical method to go from one framework to another, does it matter if variables are discrete or continuous. \todo{}
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
-In other words, the network is a causal Bayesian network.
-All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
-Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}).
-Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
-We can also connect this to counterfactuals via:
-
-\begin{equation}
-\begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
-\end{aligned}
-\end{equation}
-
-The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
-
-\begin{equation}
-\label{eq:weakignore}
-X \ci Y^{0} \text{ and } X \ci Y^{1}
-\end{equation}
-
-It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}.
-The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
-For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores \citep{Rosenbaum1983} to estimate the effect.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
-These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
-
-In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
-The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field.
-Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
-In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
-
-\subsection{What does a causal model give us? Resolving Simpson's paradox}
-
-We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
-The following example is adapted from \citet{Pearl2000}.
-Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
-She keeps track of which medication her patients choose and whether or not the treatment is successful.
-She obtains the results in table \ref{tab:simpson1}.
-
-\begin{table}[h]
-\caption{Treatment results}
-\label{tab:simpson1}
-\center
-\begin{tabular}{c c c c c}
-Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9} 
-A &  87 & 13 & 100 & 87\%\\
-\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Drug A appears to perform better.
-However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
-
-\begin{table}[h]
-\caption{Treatment results by gender}
-\label{tab:simpson2}
-\center
-\begin{tabular}{c c c c c c}
-Gender & Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9}
-M & A & 12 & 8 & 20 & 60\%\\
-M & B & 56 & 24 & 80 & 70\%\\
-\rowcolor[gray]{0.9}
-F & A & 75 & 5 & 80 & 94\%\\
-F & B & 19 & 1 & 20 & 95\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Once the data is broken down by gender, drug B looks better for both men \emph{and} women.
-Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must recommend which to subsidise under a national health scheme).
-Should she choose A or B?
-The ambiguity in this question lies at the heart of Simpson's paradox.
-How does causal modelling resolve the paradox?
-The key is that the doctor is trying to choose between \emph{interventions}.
-She wants to know what the success rate will be if she changes her practice to give all the patients one drug, rather than allowing them to choose as currently occurs.
-
-Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
-The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
-Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
-Some assumptions about the causal relationships between the variables are required.
-In this example, it seems reasonable to conclude that gender may affect the treatment chosen and the outcome.
-Assuming there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Gender};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (2) edge (3) 
-	(2) edge (3);
-\end{tikzpicture}
-\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
-\label{fig:simpsons_paradox}
-\end{figure}
-
-With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
-Knowing a patient took drug A indicates they are more likely to be female.
-When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
-However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
-So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
-Drug B is the better choice.
-
-\eqn{
-\P{Y|do(T)} = \P{Y|T,female}\P{female}+ \P{Y|T,male}\P{male}
-}
-
-Is the solution to Simpson's paradox to always to break down the data by as many variables as possible?
-No.
-Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
-This is a drug designed to prevent heart disease.
-One pathway to doing so might well be to lower blood pressure.
-Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
-It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
-Now, however, table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood pressure control and better overall outcomes.
-In this setting $\P{Y|do(T)} = \P{Y|T}$ and drug A is the better choice.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Blood Pressure};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (3) 
-	(2) edge (3)
-	(2) edge (1);
-\end{tikzpicture}
-\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
-\label{fig:simpsons_paradox2}
-\end{figure}
-
-Note that we have not changed the data itself, only the description of the variables that it is associated with.
-This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
-From a purely statistical viewpoint there is no paradox.
-The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
-The paradox only arises when we attempt to use the data to select an intervention and is resolved when we apply a causal approach to do so.
-
-\eqn{
-\label{eqn:simpsons_paradox}
-\exists \set{N_1,...N_4, N_1'...N_4'} \in \mathbb{N} : \; 
-\textcolor{green}{\frac{N_1}{N_2}< \frac{N_1'}{N_2'}}, \;\; 
-\textcolor{blue}{ \frac{N_3}{N_4}<\frac{N_3'}{N_4'}}
- \text{ and } \frac{N_1+N_3}{N_2+N_4} > \frac{N_1'+N_3'}{N_2'+N_4'}
-}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.4]{figures/simpson_vector}
-\caption{Simpson's reversal visualised.
-The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
-However, when we sum them, the ratio is steeper for the un-primed variables.}
-\label{fig:simpsons_vector}
-\end{figure}
-
-There are many other plausible causal graphs for both scenarios above.
-Perhaps income affects drug choice as well as gender, or gender might affect treatment choice and blood pressure control given treatment, etc.
-Causal modelling provides a powerful tool to specify such assumptions and to determine how to estimate causal effects for a given model as we discuss in the next section.
-
-\section{Answering Causal Questions} 
-\label{sec:causal effect estimation}
-We can roughly categorise the problems studied within causal inference from observational data into two groups, causal effect estimation and causal discovery.
-In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
-The goal is then to estimate the effect of an intervention or range of interventions in the system.
-Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}.
-Almost every time someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
-
-Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
-This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
-Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
-An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
-In this thesis I have focused on problems where the structure of the causal graph is known.
-Extending my work to problems where the causal structure is unknown, leveraging the work on causal discovery, is a rich and fascinating line of potential future work, which I discuss very briefly in section \ref{sec:causal_bandit_discussion}.
-
-\subsection{Mapping from observational to interventional distributions}
-
-A central component of estimating causal effects from observational data is determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
-We did this on an ad-hoc basis to resolve the examples discussed in section \ref{sec:causal_models}.
-In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:causal_effect_under_ignorability}, discuss the key issues involved in estimating such expressions from finite sample data.
-We assume the basic structure of the graph is known.
-That is, we assume that we can draw a network containing (at a minimum):
-
-\begin{itemize}
-\item the target/outcome variable we care about,
-\item the focus/treatment variables on which we are considering interventions,
-\item any variables which act to confound two or more of the other variables we have included, and
-\item any links between variables we have included. 
-\end{itemize}
-
-Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
-For example, consider estimating the impact of schooling on wages.
-Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
-Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
-
-How can the structure of the causal graph be leveraged to compute interventional distributions from observational ones?
-Given the graph corresponding to the observational distribution, the graph after any intervention can be obtained by removing any links into variables directly set by the intervention.
-The joint interventional distribution is the product of the factors associated with the interventional graph, as given by the truncated product formula \ref{eq:truncatedproduct}.
-If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
-However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
-
-The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
-By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \citep{Pearl2000}.
-The do calculus consists of three rules.
-They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.
-
-\subsubsection{Independence in Bayesian networks: D-separation}
-\label{sec:dsep}
-
-Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
-Therefor, in this section, we briefly review the key properties of Bayesian networks.
-A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
-Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
-
-\vspace{.3cm}
-\begin{theorem}
-\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
-\end{theorem}
-
-The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$.
-D-separation is an algorithm that extends the local Markov property to find these additional independencies.
-It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
-
-The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
-From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
-Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
-In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
-In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
-
-The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
-The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
-As an example, consider a scholarship available to female or disadvantaged students.
-Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship.
-There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
-However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y| Z$}
-\label{fig:dseptripple1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$X$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Y$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$X$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, below left =of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-        edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|\{\}$}
-\label{fig:dseptripple4}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, above left =of 1](2){$X$};
-\node[main node, above right =of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (1)
-    (3) edge node {} (1);	
-\end{tikzpicture}
-\end{subfigure}
-\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
-\label{fig:dseptriples}
-\end{figure}
-
-\vspace{.3cm}
-\begin{definition}[unblocked path]
-\label{def:unblocked}
-A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
-It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
-\end{definition}
-
-\vspace{.3cm}
-\begin{definition}[d-separation]
-\label{def:dsep}
-The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, there are no unblocked paths from any $X \in \boldsymbol{X}$ to any $Y \in \boldsymbol{Y}$ after conditioning on $\boldsymbol{Z}$.
-\end{definition}
-
-\vspace{.3cm}
-\begin{theorem}[d-separation and conditional independence]
-\label{thrm:dsep_independence}
-If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in a Bayesian network $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
-Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected (not d-separated) given $\boldsymbol{Z}$ then it is possible to construct a distribution $P'$ that factorises over $G$ in which they are dependent.
-\end{theorem}
-
-Theorem \ref{thrm:dsep_independence} says that independencies implied by d-separation on a graph hold in every distribution that can be factored over that graph and that if ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in \emph{all} distributions that can be factored over $G$ then they are d-separated in $G$.
-If we denote the independencies implied by d-separation in a graph by $\indep{G}$ and the set of independencies in a distribution by $\indep{P}$ then $\indep{G} \subseteq \indep{P}$.
-
-If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
-However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
-A particular case in which this occurs is when there are deterministic relationships between variables.
-If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
-This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
-This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
-D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
-
-\subsubsection{The Do Calculus} 
-\label{sec:do_calculus}
-
-The do-calculus is a set of three rules \citep{Pearl1995} that can be applied to simplify the expression for an interventional distribution.
-If by repeated application of the do-calculus, along with standard probability transformations, we can obtain an expression containing only observational quantities then we can use it to estimate the interventional distribution from observational data.
-Let $\boldsymbol{X}$,$\boldsymbol{Y}$,$\boldsymbol{Z}$ and $\boldsymbol{W}$ be disjoint sets of variables in a causal graph $G$.
-We denote the graph $G$ after the an intervention $do(\boldsymbol{X})$, which has the effect of removing all edges into variables in $\boldsymbol{X}$, as $G_{\overline{\boldsymbol{X}}}$
-
-\paragraph{Rule 1: (adding or removing evidence)}
-
-Rule 1 allows us to remove (or insert) observational evidence from the right hand side of a conditional interventional distribution.
-It follows directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still applies after an intervention.
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:mutilatedOriginal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2);
-
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.5\textwidth}
-\centering
-\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
-\label{fig:mutilatedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{subfigure}
-\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
-\label{fig:mutilatednet}
-\end{figure}
-
-\paragraph{Rule 2: (exchanging actions with observations)}
-Rule 2 describes when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening, $do(\boldsymbol{X}=\boldsymbol{x})$, have the same effect on the distribution over $\boldsymbol{Y}$.
-Let $G_{\underline{\boldsymbol{X}}}$ denote the causal graph $G$ with all edges \emph{leaving} $X$ removed.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W})$ in $G_{\underline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W})
-}
-
-The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$.
-Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
-However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
-Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
-If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths.
-This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).
-
-Equation \ref{eq:Do22} does not cover cases where acting on one set of variables allows us to replace acting on another set with conditioning (see figure \ref{fig:rule2_2}).
-The general form of rule 2 is given in equation \ref{eq:Do2}.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W},\boldsymbol{Z})$ in $G_{\underline{\boldsymbol{X}}\overline{\boldsymbol{Z}}}$:
-\eqn{
-\label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),
-do(\boldsymbol{Z}=\boldsymbol{z}),
-\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W})
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule2_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)  edge  (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule2_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1)	edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}}$}
-\label{fig:rule2_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    	edge  (3);
-
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
-In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
-If we condition on $W$, we block this indirect path.}
-\label{fig:rule2}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule22_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (1) edge (2)
-    (5) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{after $do(Z)$, $G_{\overline{Z}}$}
-\label{fig:rule22_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}\overline{Z}}$}
-\label{fig:rule22_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of applying equation \ref{eq:Do2}.
-In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
-Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
-Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
-}
-\label{fig:rule2_2}
-\end{figure}
-
-\paragraph{Rule 3: (adding or removing actions)}
-This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-A simple case of rule 3 is given in equation \ref{eq:Do3_1}.
-If $\boldsymbol{Y}$ is independent of $\boldsymbol{{X}}$ in $G$ after removing links entering $\boldsymbol{X}$ then can be no direct path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and any intervention on $\boldsymbol{X}$ will not affect $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{{X}})$ in $G_{\boldsymbol{\overline{X}}}$:
-
-\eqn{
-\label{eq:Do3_1}
-\P{\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})} = \P{\boldsymbol{Y}}
-}
-
-The general case of rule 3 is easier to state by explicitly representing the intervention in the graphical model.
-Let $G^{\boldsymbol{\hat{X}}}$ denote the graph $G$ after adding a variable $\hat{X_i}$ as a parent of each variable $X_i \in \boldsymbol{X}$ (see figure \ref{fig:rule3_augmented}).
-The variable $\hat{X_i}$ can be thought of as representing the mechanism by which $X_i$ takes its value, either by being set via intervention or as a stochastic function of its other parents \citep{Koller2009}.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\boldsymbol{\hat{X}}}_{\boldsymbol{\overline{Z}}}$:
-\eqn{
-\label{eq:Do3}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-}
-
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
-These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule3_original}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{augmented graph $G^{\hat{X}}$}
-\label{fig:rule3_augmented}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
-\node[main node, above left=of 2](5){$\hat{X}$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3)
-    (5) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule3_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Example application of equation \ref{eq:Do3}.
-$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
-We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
-\label{fig:rule3}
-\end{figure}
-
-\subsubsection{Identifiability}
-A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
-This is the identifiablity problem.
-It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
-A query is non-parametrically identifiable if it is identifiable without  assumptions about the functional form of the dependencies between variables in the graph.
-
-\vspace*{.3cm}
-\begin{definition}[Non-parametric identifiablity]
-Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
-A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
-\end{definition}
-
-The question of non-parametric identifiability is solved.
-The do calculus is complete \citep{Shpitser2006b,Huang2006}.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
-There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
-
-\begin{enumerate}
-\item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network, and
-\item if it is identifiable, return the required expression. 
-\end{enumerate}
-
-Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
-I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$X$};
-\node[ob, right=of 1](2){$Z$};
-\node[ob, right=of 2](3){$Y$};
-\node[lt,above=of 2](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid2_instrumental_variable}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[ob, above left=of 2](4){$Z$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1)
-    (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
-\label{fig:identify}
-\end{figure}
-
-Many interesting questions relating to identifiability remain open.
-What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
-What if we assume various restrictions on the functional form of the relationships between the variables?
-Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
-A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
-
-Although identifiability is a natural and important question to ask, it does not partition causal questions into solvable and unsolvable.
-Estimators for identifiable queries can be slow to converge and we may be able to obtain useful bounds on causal effects in cases where point estimates are not identified.
-
-\subsection{Defining causal effects} 
-\label{sec:defining-causal-effects}
-So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
-This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set.
-From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.
-
-\eqn{
-\label{eqn:utility_max}
-x* =\argmax_x \Esub{y\sim\P{Y|do(X=x)}}{{\cal{U}}(y)}
-}
-
-Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
-There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
-For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
-
-\eqn{
-\label{eqn:ACE}
-ACE = \E{Y|do(X = 1)} - \E{Y|do(X=0)}
-}
-
-Assuming the expectations in equation \ref{eqn:ACE} are well defined, the ACE captures the shift in the mean outcome that arises from varying $X$.
-It does not capture changes in variance or higher moments of the distribution.
-The ACE can be generalised to non-discrete interventions by considering the effect on the expectation of $Y$ of an infinitesimal change in $x$.
-If $X$ is linearly related to $Y$ then the ACE is constant and equivalent to the corresponding coefficient in the linear structural equation model.
-
-\eqn{
-\label{eqn:ACE_continuous}
-ACE(x) = \frac{d}{dx}\E{Y|do(X=x)}
-}
-
-The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}.
-Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
-However we can personalise the average causal effect by stating it with respect to some observed context.
-I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
-
-\eqn{
-PCE(z) = \E{Y|do(X=1),z} - \E{Y|do(X=0),z}
-}
-
-In some cases we may be interested in the average causal effect for some sub-group of the population.
-A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
-This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.
-
-\eqn{
-ATT = \Esub{z\sim \P{Z|x=1}}{Y|do(X = 1)} - \Esub{z\sim \P{Z|x=1}}{Y|do(X=0)}
-}
-
-Causal effects can also be written in terms of counterfactuals.
-The ACE is $\E{\cf{Y}{1} - \cf{Y}{0}}$.
-We could estimate the ratio of expectations $\frac{\E{\cf{Y}{1}}}{\E{\cf{Y}{0}}}$ instead of the difference.
-However, the quantity   $\E{\frac{\cf{Y}{1}}{\cf{Y}{0}}}$ depends on the joint distribution over the counterfactual variables $(\cf{Y}{1},\cf{Y}{0})$ and thus cannot be computed from the interventional distribution.
-
-Another way of conceptualising causal effects is as a property indicating the strength of the causal link between two variables.
-This notion is complex to formalise when the relationship between variables is non-linear.
-Suppose $Y = X \oplus Z$ with $P(Z=1)=\frac{1}{2}$, the interventional distributions over $X$ are identical after marginalising out $Z$.
-\citet{Janzing2013} propose a number of postulates that a notion of causal strength could satisfy, demonstrate why previous measures fail these postulates and propose an alternative based on information flow.
-
-\subsection{Estimating causal effects by adjusting for confounding variables}
-\label{sec:causal_effect_under_ignorability}
-Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors.
-Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds.
-Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds.
-This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}.
-The setting is also referred to as unconfounded.
-
-\vspace*{0.3cm}
-\begin{theorem}[The backdoor criterion]\citep{Pearl2000}
-Let $\boldsymbol{X}$, $\boldsymbol{Z}$ and $\boldsymbol{Y}$ be disjoint sets of vertices in a causal graph $G$.
-If $\boldsymbol{Z}$ blocks (see Definition \ref{def:unblocked}) for every path from $X_i$ to $Y_j$ that contains a link into $X_i$, for every pair $(X_i \in \boldsymbol{X},Y_j \in \boldsymbol{Y})$, and no node in $\boldsymbol{Z}$ is a decedent of a node in $\boldsymbol{X}$ then the backdoor criterion is satisfied and;
-
-\eqn{
-\label{eqn:backdoor}
-\P{\boldsymbol{y}|do(\boldsymbol{x})} =\sum_{\boldsymbol{z}} \P{\boldsymbol{y}|\boldsymbol{x},\boldsymbol{z}}\P{\boldsymbol{z}}
-}
-\end{theorem}
-
-The backdoor criterion derives from rule 2 of the do-calculus.
-Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
-There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
-The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
-Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}.
-Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
-For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
-
-\begin{figure}
-\centering
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{Multiple valid adjustment sets.}
-\label{fig:adjust_multiple}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](2){$X$};
-\node[ob, right=of 2](3){$Y$};
-\node[ob,above=of 3](4){$Z_2$};
-\node[ob,above=of 2](5){$Z_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (5) edge (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{Don't condition on all observable variables}
-\label{fig:m-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Identifying an optimal adjustment set is not always intuatively obvious.
-There may be multiple valid adjustment sets.
-In sub-figure (a) $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.
-Adding additional variables (even pre-treatment variables) to a valid adjustment set can result in an invalid set as in sub-figure (b).
-In this case the empty set is a valid adjustment set for the causal effect of $X$ on $Y$, however adding $Z$ would open a backdoor path $X-U_1-Z-U_2-Y$}
-\label{fig:adjustment}
-\end{figure}
-
-Given that a set of variables $\boldsymbol{Z}$ satisfies the backdoor criterion (or equivalently the conditional ignorability assumption), the interventional distribution is asymptotically identifiable and can be estimated from equation \ref{eqn:backdoor}.
-The expected value of $Y$ after the intervention $do(X=x)$ is given by equation \ref{eqn:expected_post_intervention} and the average causal effect for a binary intervention $x \in \set{0,1}$ is given by equation \ref{eqn:ace_ignoreable}.
-
-\eqn{
-\label{eqn:expected_post_intervention}
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}}
-}
-
-\eqn{
-\label{eqn:ace_ignoreable}
-ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
-}
-
-Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata.
-However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
-There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
-The key approaches are based on matching on covariates, propensity score methods and regression.
-We now examine these approaches from a machine learning perceptive.
-
-In standard supervised learning, we have a training set $(\vb{x_1},y_1),...,(\vb{x_n},y_n)$ assumed to be sampled i.i.d from an unknown distribution $\P{\vb{x},y} =\P{\vb{x}}\P{y|\vb{x}}$.
-The goal is to select a hypothesis $h \in \mathcal{H} : \mathcal{X} \rightarrow \mathcal{Y}$ such that, on unseen data $ \sim \P{\vb{x},y}$, $h(\vb{x})$ is close (by some metric) in expectation to $y$.
-In other words we wish to minimise the generalisation error $E_{out}(h)$,
-
-\eqn{
-E_{out}(h) = \Esub{(\vb{x},y)\sim \P{\vb{x}}\P{y|\vb{x}}}{L(h(\vb{x}),y)}
-}
-
-We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
-We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
-
-\eqn{
-E_{in}(h) = \frac{1}{n}\sum_{i=1}^n L(h(\vb{x}_i),y_i)
-}
-
-The VC-dimension of the hypothesis space provides (typically loose) bounds on the probability that $E_{out} >> E_{in}$.
-However, in practice, the generalisation error is usually estimated empirically from a hold-out set of the sample that was not used to train the model, or via cross-validation.
-
-In the causal effect estimation under ignorability, we have training data $(\vb{x}_1,\vb{z}_1,y_1),...,(\vb{x}_n,\vb{z}_n,y_n)$ sampled i.i.d from $\P{\vb{z}}\P{\vb{x}|\vb{z}}\P{y|\vb{x},\vb{z}}$.
-Estimating $\E{Y|do(\vb{X}=\vb{x})}$ corresponds to selecting a hypothesis $h \in \mathcal{H}:\mathcal{X} \times \mathcal{Z} \rightarrow \mathcal{Y}$ that minimises;
-
-\eqn{
-\label{eqn:causal_covariate_shift}
-E_{out} &= \Esub{(\vb{x},\vb{z},y)\sim \delta(\vb{x}-\vb{x'})\P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},\\
-&= \Esub{(\vb{z},y)\sim \P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},
-}
-
-\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
-If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
-\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
-Take the case where we have a binary intervention $x \in \set{0,1}$.
-Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$.
-More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.
-
-\subsubsection{Regression}
-
-The regression approach is to learn a function that is a good approximation to the output surface $\E{Y|X,Z}$.
-Let $f_1(z)= \E{Y|X=1,Z=z}$.
-The expectation of $Y$ after the intervention $X=1$ is then obtained by taking the expectation with respect to $Z$, $\E{Y|do(X=1)} = \Esub{z \sim \P{Z}}{\E{Y|X=1,z}}$.
-We can learn a parametric regression model $\hat{f}_1(z)$ via empirical risk minimisation.
-
-\eqn{
-\hat{f}_1(z) = h_1(z;\hat{\theta}_{obs}) \text{, where } \hat{\theta}_{obs} = \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)  \right]
-}
-
-This estimator is consistent with respect to the observational distribution.
-As the sample size tends to infinity, $\hat{\theta}_{obs}$ approaches the parameter within the hypothesis space that minimises the expected loss given data sampled from the observational distribution.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z|x=1}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-If the model is correctly specified such that $f_1(z) = h_1(z;\theta^*)$ for some  $\theta^* \in \Theta$ then the empirical risk minimisation estimate is consistent with respect to the loss over any distribution of $Z$  \citep{Sugiyama2007}, including the interventional one.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \theta^* = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-The average causal effect can then be estimated by,
-
-\eqn{
-\label{eqn:cef_empirical_regression}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left(\hat{f}_1(z_i) - \hat{f}_0(z_i)\right)
-}
-
-Regression thus has a direct causal interpretation if the parametric model is correctly specified and the covariates included form a valid backdoor adjustment set for the treatment variable of interest in the corresponding structural equation model.
-
-\subsubsection{Propensity scores}
-If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
-The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
-\caption{
-Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
-In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
-If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$.
-We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points).
-In both cases the linear model is a good fit to the data.
-However, the resulting estimate of the causal effect is very poor for the lower values of $z$.
-}
-\label{fig:covariate_shift_adjust}
-\end{figure}
-
-A general approach to estimating the expectation of some function $f(\cdot)$ with respect to data from some distribution $\P{\cdot}$, when we have data sampled from a different distribution $\Q{\cdot}$ is importance sampling \citep{horvitz1952generalization,Koller2009}.
-\eqn{
-\label{eqn:importance_sampling}
-\Esub{\vb{v} \sim \P{\vb{v}}}{f(v)} = \Esub{\vb{v} \sim \Q{\vb{v}}}{f(v)\frac{\P{\vb{v}}}{\Q{\vb{v}}}}
-}
-
-This importance weighting approach can be applied to the covariate shift/average causal effect problem by weighting the terms in the empirical risk minimisation estimator \citep{Sugiyama2007}.
-
-\eqn{
-\label{eqn:importance_weighted_erm}
-\hat{\theta}_{iw} &= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{\P{z_i}\delta(x_i-1)}{\P{z_i}\P{x_i=1|z_i}}  \right] \\
-&= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{1}{e(z_i)}  \right],
-}
-
-where $e(\vb{z})$ is the propensity score, defined by \citep{Rosenbaum1983};
-\eqn{
-e(\vb{z})\equiv \P{x=1|\vb{z}}
-}
-
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
-Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
-
-The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$.
-\citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$.
-This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$.
-Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
-In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
-However in general, it must be estimated from data.
-Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
-\citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
-
-\eqn{
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}} = \Esub{z\sim \P{\vb{Z|x}}}{\E{Y|x,\vb{z}}\frac{1}{e(\vb{z})}}
-}
-
-\eqn{
-\label{eqn:simple_inverse_propensity}
-\hat{\tau}_{ip} = \frac{1}{n}\sum_{i=1}^n \left(\frac{\ind{x_i=1}y_i}{e(\vb{z}_i)} -\frac{\ind{x_i=0}y_i}{1-e(\vb{z}_i)}\right)
-}
-
-\subsubsection{Matching}
-There is a straightforward connection between matching and regression for causal effect estimation.
-If $h \in \mathcal{H} \implies h + a \in \mathcal{H}$ for any constant $a$ and $\hat{f}$ is selected by minimising empirical risk with an $L_2$ loss then $\frac{1}{n}\sum_{i=1}^n \ind{x_i=1}\hat{f}_1(z_i) = \frac{1}{n}\sum_{i=1}^n \ind{x_i=1}y_i$ \footnote{\citep{Imbens2004} state this holds for most implementations}, and equation \ref{eqn:cef_empirical_regression} can be re-written as:
-
-\eqn{
-\label{eqn:cef_regression_missing}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \hat{f}_0(z_i)\right) +\ind{x_i=0}\left(\hat{f}_1(z_i) - y_i \right)\right]
-}
-
-This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
-For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
-Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
-\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
-
-\eqn{
-\hat{\tau}_{match} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \frac{1}{k}\sum_{j \in J_k(i)}y_j\right) +\ind{x_i=0}\left(\frac{1}{k}\sum_{j \in J_k(i)}y_j - y_i \right)\right]
-}
-
-This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
-There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
-Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
-An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.
-
-The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
-A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
-
-The significance of this should not be underestimated.
-Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
-With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
-\citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem.
-There is relatively little theory on model selection for estimating the propensity score.
-To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion.
-It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}.
-With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
-I have found no examples of such an approach.
-
-The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
-Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
-The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
-
-\section{Bayesian reasoning and causality}
-\label{sec:bayesian_causality}
-The Bayesian approach to modelling encapsulates a very general approach to combining prior knowledge and assumptions with data to draw inferences.
-You can specify a complex model, capturing  your assumptions about the underlying mechanisms of your system, describe what you are uncertain about with parameters (and priors over their values) and then update your beliefs about the values of these parameters based on the observed data.
-If you believe the model to be a mechanistic representation of the way world works then you can use it to infer the consequences of interventions.
-
-Why then do we need a special framework for causal inference; can't we simply encode all the assumptions required to infer the outcome of an intervention in a given system in a Bayesian model?
-The short answer is we could.
-However, causal graphical models represent a useful level of abstraction for many problems.
-Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
-We do not need to specify the functional relationships between the variables or priors over their distributions.
-The power of defining an intervention in this way stems from the number of things that are assumed invariant between the two processes.
-All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
-The causal framework also provides, via the do-calculus, a rigorous method to draw a boundary around which variables we must measure and model to identify causal effects.
-
-We could represent problems of the type where we try to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and what they have in common, see figure \ref{fig:causal_ordinary_bayes}.
-This does not require any special framework or notation.
-The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks.
-However, without a causal framework, we have to make assumptions about what will be invariant to the intervention specifically for each such problem we encounter.
-For complex problems, it is very difficult to conceptualise the assumptions we expect to hold without the benefit of a causal framework.
-
-\begin{figure}[h]
-\center
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
-\tikzstyle{box}=[rectangle, draw=black!100]
- %nodes
-\node[node](1){$Z$};
-\node[node, below left=of 1](2){$X$};
-\node[node, below right=of 1](3){$Y$};
-\node[node, right=1.5cm of 3](4){$x$};
-\node[lt, above right=of 4](5){$Z'$};
-\node[lt, below right=of 5](6){$Y'$};
-\node[lt, above=2.0cm of 3](9){$\theta_{y}$};
-\node[lt, above=0.5cm of 9](7){$\theta_z$};
-\node[lt, above=2.0cm of 2](8){$\theta_x$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (6)
-    (5) edge (6)
-    (8) edge (2)
-    (7) edge (1) edge (5)
-    (9) edge (3) edge (6);
-\node[rectangle, inner sep=4.0mm, fit= (1) (2) (3),draw=black!100,label=below right:N] {};
-\node[rectangle, inner sep=4.0mm, fit= (5) (6),draw=black!100,label=below right:M] {};
-\end{tikzpicture} 
-\caption{Causal inference with ordinary Bayesian networks.
-The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$.
-The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
-As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
-However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
-\label{fig:causal_ordinary_bayes}
-\end{figure}
-
-We can also combine causal graphical models with Bayesian approaches to estimation.
-For example, we can use a causal graphical model to identify a set of variables that form a valid adjustment set and then use Bayesian regression to estimate posterior distribution over a given causal effect.
-However, in doing so we need to be careful about the way in which we specify priors as I highlight in the next section.
-
-\subsection{Careful with that prior}
-
-Suppose we have a data generated by the linear Gaussian Bayesian network below.
-The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$.
-The variable $U$ is latent.
-
-\eqn{
-\P{U} &= N(0 , v_u) \\
-\label{eqn:careful_with_priors}
-\P{Z|U} &= N(w_{zu}U , v_z) \\
-\P{X|Z} &= N(w_{xz}Z , v_x) \\
-\P{Y|U,Z,X} & = N(w_{yu}U +w_{yz}Z + w_{yx}X , v_y)
-}
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
-
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[lt, right=of 1](4){$U$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    	edge (3)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-
-\end{tikzpicture} 
-\caption{ }
-\label{fig:careful_with_priors}
-\end{figure}
-
-As each variable is a linear function of its parents, with Gaussian noise, resulting in a joint distribution $\P{U,Z,X,Y}$ that is multivariate normal.
-Marginalising out $U$ and conditioning on $X$ and $Z$ yields,
-
-\eqn{
-Y \sim N(w_{yx} X + \beta Z, \epsilon) 
-}
-
-where
-
-\eqn{
-\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and}\\
-\epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
-}
-
-The causal effect of $X$ on $Y$ is identifiable (even without the linear Gaussian assumptions) as $Z$ satisfies the backdoor criterion and the expectation of the coefficient for $X$ in a standard OLS regression of $Y$ against $X$ and $Z$ is $w_{yx}$.
-However, the causal effect of $Z$ on $Y$ is not identifiable due the presence of the unobserved confounder $U$.
-The coefficient $\beta$ captures both the causal relationship between $Z$ and $Y$, $w_{yz}$ and the indirect relationship through $U$, which we cannot separate without observing $U$.
-
-If we do a Bayesian regression of $X$ and $Z$ against $Y$ and use intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ we are likely to select something centered around $w_{yz}$, but this is not correct and can introduce bias into our estimate of the coefficient for $w_{yx}$, as we demonstrate in figure \ref{fig:careful-prior-example}.
-The prior for $\beta$ should be based on our belief about how we expect $Z$ to be \emph{associated} with $Y$ after marginalising out any confounding variables.
-
-\begin{figure}[h]
-\begin{subfigure}[t]{0.99\textwidth}
-     \caption{No prior on $\beta$, the posterior on $w_{yx}$ is centered around its true value}
-    \includegraphics[scale=1]{prior_no_prior}
-\end{subfigure}
-\begin{subfigure}[t]{0.99\textwidth}
-     \caption{Prior on $\beta$ centered around $w_{yz}$, the causal effect of $Z$ on $Y$, $Prior(\beta)= N(w_{yz},\sigma = 0.5)$.
-The posterior on $w_{yx}$ is biased away from its true value (as is the posterior on $\beta$ but this is not the key quantity of interest).}
-    \includegraphics[scale=1]{prior_bad_prior.pdf}
-\end{subfigure}
-\caption{An example demonstrating how selecting a prior for a nuisance parameter centered around the causal effect of that parameter on the outcome, rather than its association with the outcome, can bias estimates for actual parameter of interest.
-We sample $n=1000$ data points from the joint distribution defined by $P(U)P(Z|U)P(X|Z)P(Y|U,Z,X) \sim N(0,1)N(2U,0.09)N(0.5Z,1)N(3U -Z + 0.5X , 0.25)$ and fit the model $Y \sim N(w_yx X+\beta Z,\epsilon)$ in Stan and plot the posterior distributions over $w_yx$ and $\beta$.
-Figure (a) shows the results with no prior (equivalently an improper prior) for both $w_yx$ and $\beta$.
-Figure (b) shows the results when we place a Gaussian prior centered around the $w_yz$ on the distribution for $\beta$.}\label{fig:careful-prior-example}
-\end{figure}
-
-This result is counter intuitive.
-When fitting a model to estimate the affect of marketing on sales, it is natural to put a prior on the coefficient for price that is strongly negative because we believe increasing price decreases sales.
-However, our prior should be on the expected association between price and sales, holding fixed all the other variables in the model, not the causal relationship between price and sales.
-An alternative is to explicitly include $U$ as a latent variable in the model.
-However, this comes at a computational cost with little benefit for estimation, unless we have good knowledge of the relationships between $U$, $Z$ and $Y$.
-More fundamentally, at whatever level we decide to stop adding latent variables and specify a prior, we can introduce this form of bias if priors are selected naively on the basis of causal intuition and the issue is not limited to linear-Gaussian models.
-I believe this is an important insight for the many applications of Bayesian modelling that make use of human elicited priors, since people tend to think causally.
-The solutions, either explicitly modelling latent variables to higher levels or allowing for them in setting priors, have the effect of broadening the priors over nuisance parameters in the model.
-
-\chapter{Learning from interventions}
-\label{chap:interventionalist_view}
-
-The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it.
-There is an obvious alternative.
-Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens.
-There are three key differences between observing a system and explicitly intervening in it.
-Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from.
-Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}.
-Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias.
-Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning.
-For example, in a drug trial , assigning people a sub-optimal treatment has real world costs.
-This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain.
-This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.
-
-\section{Randomised experiments}
-\label{sec:randomized_experiment}
-Randomised controlled trials are often presented as the gold standard for determining causal effects.
-What is it about randomisation that makes it so important when it comes to causality?
-The graphical model for a randomised controlled experiment is shown in figure \ref{fig:random_experiment_network}.
-If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
-Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3);
-\end{tikzpicture}
-\caption{causal network for a randomised experiment}
-\label{fig:random_experiment_network}
-\end{figure}
-
-Randomisation does not ensure target and control group are exactly alike.
-The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables between the target and control groups in a finite data sample.
-However, the variance in the outcome, within both the target and control groups, also increases.
-The net result is increased variance (but not bias) in  the estimate of causal effects.
-
-Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest.
-If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can estimate the average causal effects directly from the differences in outcome across treatments.
-More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
-The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
-See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, above right=of 1](5){Covariates};
-\node[node, below right=of 5](3){Outcome};
-\node[lt, above=of 3](4){U};
-
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3)
-	(5) edge (3) edge (1);
-\end{tikzpicture}
-\caption{causal network for a stratified randomised experiment if the probability an individual is assigned a given treatment depends on some covariates.}
-\label{fig:random_experiment_network_stratified}
-\end{figure}
-
-The benefit provided by randomisation in breaking the link between the treatment variable and any latent confounders should not be understated.
-The possibility of unobserved confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (there is no test for confounding).
-This means causal estimates from non-experimental data are always subject to the criticism that an important confounder may have been overlooked or not properly adjusted for.
-However, randomised experiments do have some limitations.
-
-\subsection{Limitations of randomised experiments}
-\label{subsec:limitations_of_experiment}
-
-The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
-There may be imperfect compliance, the treatment selected by the randomiser is not always followed, or output censoring, the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.\footnote{Non-compliance is a problem if the goal is to estimate the causal effect of the treatment on the outcome but not if the goal is to estimate the causal effect of prescribing the treatment.
-The latter makes sense in a context where the process by which people decide whether or not to take the treatment they have been prescribed is likely to be the same if the treatment were made available more generally beyond experimental trial.} See figure \ref{fig:random_experiment_network_imperfect_compliance} for a the graphical model of a randomised experiment with imperfect compliance.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Actual Treatment};
-\node[node,above left=of 1](5){Assigned Treatment};
-\node[node, above =of 5](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (5)
-	(5) edge (1)
-	(1) edge (3)
-	(4) edge (3) edge (1);
-\end{tikzpicture}
-\caption{causal network for a randomised experiment with imperfect compliance}
-\label{fig:random_experiment_network_imperfect_compliance}
-\end{figure}
-
-It is not always possible or ethical to conduct a randomised controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
-When experimentation is possible, it is frequently difficult or expensive.
-This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
-In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest.
-The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
-\end{center}
-\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parachutes for the prevention of morbidity and mortality associated with falls from large heights.}
-\label{fig:parashute}
-\end{figure}
-
-Finally, non-adaptive randomised experiments are not optimal from either an active or reinforcement learning perspective.
-As an experiment proceeds information is obtained about the expectation and variance of each intervention (or treatment).
-Fixed experimental designs cannot make use of this information to select which intervention to try next.
-This results in both poorer estimates for a fixed number of experimental samples and more sub-optimal actions during the course of the experiment.
-
-\section{Multi armed bandits}
-\label{sec:multi-armed-bandits}
-
-Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed.
-Their introduction is generally attributed to \citet{Thompson1993}.
-In its classic formulation \citep{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
-Each arm $i$ is associated with a fixed but unknown reward distribution.
-\footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
-The goal of the learner is to maximise the total reward they receive.
-This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
-
-\vspace{0.5cm}
-\begin{definition}[Stochasitic k-armed bandit problem]
-Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action.
-The multi-armed bandit problem proceeds over $T$ rounds.
-In each round $t$,
-\begin{enumerate}
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
-\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
-\end{enumerate}
-
-At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T \cfb{Y}{a_t}$.
-We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
-Note we have used counterfactual notation (see section \ref{sec:counterfactuals}) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have received had it selected action $i$ at timestep $t$.
-I discuss the (potentially) counterfactual nature of regret further in section \ref{sec:counterfactual-nature-regret}.
-\end{definition}
-
-The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
-To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
-This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
-There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.}.
-\eqn{
-\label{eqn:raw_regret}
-\rawregret = \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
-}
-
-Both the rewards and the actions selected by the algorithm are random variables.
-The majority of work in the bandit literature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}.
-The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is referred to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
-A stochastic bandit algorithm is learning if it obtains pseudo-regret that is sub-linear in $T$.
-
-\vspace{0.5cm}
-\begin{definition}[Pseudo-Regret]
-\eqn{
-\label{eqn:pseudo_regret}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}} \\
-&= T\mu_{i^*} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}}
-}
-\end{definition}
-
-The regret is invariant to adding a constant to the expected rewards for all actions.
-However, it still depends on key characteristics of the reward distributions for each action.
-Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range.
-Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
-
-\subsection{Stochastic bandits: Approaches and results}
-\label{sec:stochastic-bandit}
-
-The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators.
-The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards.
-As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
-The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
-This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
-Instead, the focus is on computing bounds on the expected regret.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=1]{extreme_early_stopping}
-\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples.
-This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim Bernoulli(0.2)$.
-In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$.
-$\E{\hat{\mu}} = 0.439$.
-The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping.
-Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded.
-This has some interesting potential real world implications.
-Early stopping of clinical trials is controversial.
-A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
-\label{fig:early_stopping}
-\end{figure}
-
-There are a few key principles that are used to guide the development of bandit algorithms.
-The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
-A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
-This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.
-
-Another key approach is \emph{optimism in the face of uncertainty}.
-Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
-The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
-Thus selecting it yields either a good reward or useful information.
-
-\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family.
-They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient.
-However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm.
-\citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$.
-Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds.
-We now assume the rewards are bounded in $[0,1]$.
-The algorithm and regret bounds can be generalised to submission reward distributions, see \citet{Bubeck2012}.
-
-\begin{algorithm}[H]
-\caption{UCB-1}\label{alg:ucb-1}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} horizon $T$.
-\STATE Play each arm once. 
-\FOR{$t \in 1,\ldots,T$}
-\STATE Count the number of times each arm has been selected previously $n_{t,i} = \sum_{s=1}^{t-1} \ind{a_s = i}$
-\STATE Calculate the mean reward for each arm $\hat{\mu}_{t,i} = \frac{1}{n_{t,i}} \sum_{s=1}^{t-1}\ind{a_s = i}Y_t$
-\STATE Select arm $a_t \in \argmax_{i = \set{1,...,k}}\left(\hat{\mu}_{t,i}+\sqrt{\frac{2 \log{t}}{n_{t,i}}}\right)$
-\ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
-The problem dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
-
-\eqn {
-\label{eqn:ucb_problem_dependent_regret}
-\regret \leq & \sum_{i:\Delta_i > 0}\left(\frac{8\log(T)}{\Delta_i}+2 \right)
-}
-
-Somewhat unintuitively, the regret increases as the value of the arms gets closer together.
-This is because it becomes harder for the algorithm to identify the optimal arm.
-As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
-The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
-The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which is maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
-By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.
-
-\eqn{
-\label{eqn:ucb_worst_case_regret}
-\regret \in \bigo{\sqrt{kT\log(T)}}
-}
-
-\begin{figure}
-\includegraphics[scale=1]{ucb_regret_vs_delta}
-\caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
-The solid curve shows the mean (cumulative) regret, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
-The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
-\label{fig:ucb_regret_vs_delta}
-\end{figure}
-
-The form of the dependence on the number of arms $k$ and horizon $T$ differs between the problem dependent and worst case regret.
-The problem dependent regret grows linearly with the number of arms, $k$, and logarithmically with $T$.
-The difference stems from the fact the problem dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
-\citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by $\regret \in \bigomega{\sqrt{kT}}$
-
-Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
-This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \citep{Lattimore2015}.
-
-\begin{figure}
-\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
-\caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$.
-The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $bernoulli([.3,.7])$ over 1000 simulations.
-The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
-\label{fig:ucb_upperbound_vs_actual_performance}
-\end{figure}
-
-Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal.
-This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}.
-Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
-However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
-
-\subsection{Pure-exploration problems}
-\label{sec:pure-exploitation-problems}
-Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
-In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
-The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.
-
-\vspace{0.5cm}
-\begin{definition}[Simple Regret]
-\eqn{
-\label{eqn:simple_regret}
-\simpleregret=\mu_{i^*} - \E{\mu_{\hat i^*}}.
-}
-\end{definition}
-
-The best arm identification problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialisation or exploitation phase.
-For example, many strategies might be assessed via simulation prior to one being selected and deployed.
-The worst case simple regret for a k-armed bandit is lower bounded by equation \ref{eqn:worst_case_simple_regret_lower_bound} (\citep{Bubeck2009}).
-
-\eqn{
-\label{eqn:worst_case_simple_regret_lower_bound}
-\simpleregret \in \bigo{\sqrt{K/T}}
-}
-
-%\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
-
-Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds. The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period.
-
-\subsection{Adversarial Bandits}
-\label{sec:adversarial-bandits}
-
-Adversarial bandits, described by \citet{Auer1995}, are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
-Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$.
-As in the stochastic setting, the learner then receives reward only for the selected action.
-
-\vspace{0.5cm}
-\begin{definition}[Adversarial k-armed bandit problem]
-\label{def:adversarial-k-armed-bandit}
-Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
-\begin{enumerate}
-\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[\cfb{Y}{1},...,\cfb{Y}{k}]$. 
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
-\end{enumerate} 
-\end{definition}
-
-Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
-In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
-
-For oblivious adversarial bandits we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
-We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
-
-\eqn{
-\label{eqn:pseudo_regret-adversarial}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}}
-}
-
-The policy (or algorithm) used by the learner is available to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
-This implies the adversary can ensure that any learner with a deterministic policy suffers regret $\bigo{T}$ by forecasting their entire sequence of actions.
-For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forecasts what action the learner will play in round 2, given they received a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
-This implies adversarial bandit policies must be sufficiently random to avoid such exploitation \footnote{The UCB algorithm, defined by algorithm \ref{alg:ucb-1}, is deterministic if the order in which arms are played during the first $k$ rounds is fixed and the method for selecting which arm to play when multiple-arms have the same upper-confidence bound is not-random (for example, select the arm one with the lowest index $i$).}
-
-The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case pseudo-regret of $\bigo{\sqrt{TK\log(T)}}$  \citep{Auer1995}.
-Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \citep{Audibert2009}.
-The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
-However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nonetheless obtain good problem dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
-
-Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, does not fully capture an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings where the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
-An example of a setting that lies between stochastic and adversarial bandit problems is the non-stationary setting, in which the rewards are generated stochastically from a distribution that varies over time.
-Adversarial bandit algorithms may perform better in such settings than standard stochastic policies to the extent that they explore more (to avoid the adversary simulating their behaviour) and thus adapt quicker to changes in the reward distribution.
-Adversarial algorithms also have stronger worst case regret guarantees, since even the weak regret for stochastic bandits is not guaranteed to be sub-linear in such settings.
-However, if we have constraints on how rapidly or frequently the reward distributions can change over time it is better to use algorithms specifically developed to exploit such information (and compare them against a stronger notion of regret), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
-
-from the information available to the agent, the joint distribution over counterfactuals cannot be inferred, unless the assignment mechanism in noiseless - so it does not make sense to define regret that depends on it.
-
-\begin{figure}[h]
-    \begin{subfigure}[t]{0.49\textwidth}
-	\centering   
-	 \caption{Behaviour over time}\label{fig:non-stationary-bandit-time} 
-         \includegraphics[scale=.5]{non-stationary-bandit-time}
-\end{subfigure}
-    \begin{subfigure}[t]{0.49\textwidth}
-    \centering
-     \caption{Cumulative regret distribution}
-        \label{fig:non-stationary-bandit-regret}
-        \includegraphics[scale=.5]{non-stationary-bandit-regret}
-    \end{subfigure}
-      \caption{The pseudo-regret can be negative if rewards are non-stationary.
-This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed Bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
-Figure (a) shows the expected rewards of each arm, and the proportion of time that arm-1 is played, as a function of time.
-The single best-arm is arm-1 as it has the highest expected reward (averaged over $t$).
-An oracle that selects arm-1 in every round obtains an expected reward of $5,000$.
-However, despite not being designed to do so, the UCB-algorithm can adapt to the changing reward distribution to obtain consistently higher rewards.
-The distribution of regret over the 1000 simulations is shown in figure (b).}\label{fig:non-stationary-bandit}
-\end{figure}
-
-\subsection{Contextual bandits}
-\label{sec:contextual-bandits}
-
-In the standard multi-armed bandit setting, each decision is identical and the goal is to learn a single best action.
-However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
-The best treatment to offer an individual patient could depend on their age, gender, disease sub-type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
-Similarly, decisions on which ad or content to display on a webpage or which product to recommend can be \emph{personalised} based on the previous behaviour of the user.
-A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
-Contextual bandits are a generalisation of multi-armed bandits that make use of this additional contextual information.
-The term contextual bandit was coined by \citet{Langford2008}, however close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
-
-\vspace{0.5cm}
-\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analogously to definition \ref{def:adversarial-k-armed-bandit}}]
-\label{def:contextual-bandit-problem}
-Let $\P{\vec{x},\vec{y}}$ be the joint distribution over the rewards for each action and some context $\vec{X} \in \mathcal{X}$. In each round $t \in \set{1,...T}$, 
-\begin{enumerate}
-\item the world stochastically generates the vector of rewards for each action and the context, $(\vec{X}_t,[\cfb{Y}{1},...,\cfb{Y}{k}]) \sim \P{\vec{x},\vec{y}}$ and reveals $\vec{X}_t$ to the learner
-\item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps,
-\item the learner observes and receives (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
-\end{enumerate}
-\end{definition}
-
-Standard multi-armed bandits learn to select the action $a$ that, with high probability, maximises $\E{Y|a}$.
-Contextual bandits learn to select actions that maximise $\E{Y|\vec{x},a}$.
-The reward for contextual bandits should  be compared to an oracle that acts optimally based on the context.
-To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothesis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
-The pseudo-regret is the difference between the expected regret obtained by an oracle that selects actions based on the single best hypothesis or policy $h$ at each timestep, and the expected reward obtained by the algorithm.
-
-\eqn{
-\label{eqn:regret_contextual}
-\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{\cfb{Y}{h(\vec{X}_t)}}} - \E{\sum_{t=1}^T{\cfb{Y}{A_t}}}
-}
-
-If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a separate standard bandit instance for each value of the context.
-This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothesis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occurred such that $\sum_{c=1}^N T_c = T$.
-The regret is maximised if $T_c = T/N$ resulting in total regret $\bigo{N\sqrt{kT/N}}$.}.
-This is optimal with respect to this class of hypothesis.
-However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the in the relationships between different contexts and actions.
-As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
-For example, with $n$ binary variables, $N = 2^n$}.
-We expect some form of smoothness; values of context that are similar should lead to comparable rewards for a given action.
-We need algorithms that can leverage such assumptions.
-
-An alternate reduction to the standard bandit problem, which allows us to constrain the hypothesis space to explore, is to treat each hypothesis $h$ as a bandit arm \citep{Langford2008}.
-At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
-Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothesis class considered, limiting our ability to learn any complex mappings from context to actions.
-The key problem with this approach is each sample is used to update our knowledge about only one hypothesis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothesis simultaneously.
-
-Suppose that, at each timestep $t$, after selecting an action, the learner received the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
-This is known as the full information setting.
-In this case, the learner can simulate running each hypothesis over the history to compute the reward it would have obtained and use the hypothesis with the best empirical reward to select the next action.
-This is the \emph{follow the leader} algorithm, which obtains optimal regret $\bigo{\sqrt{T log(\card{\mathcal{H}})}}$ for the full-information problem \citep{cesa2006prediction}.
-Unfortunately, in the contextual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
-As in causal effect estimation, we can view this as a missing data problem.
-However, the data is missing not at random because which component of the reward is observed depends on the action selected which in turn is a function of the previous history of actions and rewards.
-
-The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly separating exploration from exploitation.
-Epoch-greedy is an explore-exploit algorithm.
-It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothesis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
-The hypothesis with the highest empirical reward is then used to select actions for the remaining time steps.
-The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
-
-The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
-However, it involves maintaining a list of weights for each hypothesis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
-The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
-
-Both Epoch-greedy and ILOVECONBANDITS involve solving problems of the form,
-
-\eqn{
-\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{X}_t) = A_t}
-}
-
-This expression equates to identifying the best empirical policy based on previous data.
-The algorithms assume the existence of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
-The computational tractability of these algorithms on large (or infinite) hypothesis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classification problem \citep{Dudik2011a}.
-
-Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalised versions of the UCB or Thompson sampling algorithms.
-For linear pay-off models, both approaches yield algorithms with strong regret guarantees, \emph{Lin-UCB} \citep{li2010contextual} and \emph{Generalised Thompson Sampling} \citep{Agrawal2013}.
-
-\subsubsection{The causal structure of contextual bandits}
-\label{sec:causal-structure-contextual-bandit}
-
-It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
-However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
-\citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information.
-This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
-It is even possible to have (useful) context $\vec{X}$ that is a consequence of $\vec{Y}$ as in figure \ref{fig:context-graph-anti-causal} and example \ref{exm:anti-casual-context}.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ causes $\vec{Y}$}
-\label{fig:context-graph-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (4) edge (1)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ and $\vec{Y}$ are confounded}
-\label{fig:context-graph-confounded}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
-\node[lt, below=of 4](5){$U$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2)
-   (5) edge (4) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{Y}$ causes $\vec{X}$}
-\label{fig:context-graph-anti-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (4)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
-$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
-Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
-The causal structures in figures \ref{fig:context-graph-causal} and \ref{fig:context-graph-confounded} also make sense if the elements of $\vec{Y}$ are genuine counterfactuals, that is only the reward of the selected action is ever realised.
-The context $\vec{X}$ can only be useful in causal graphs of the form in figure \ref{fig:context-graph-anti-causal} if the components of $\vec{Y}$ are realised prior to the agent selecting an action, i.e when the components of $\vec{Y}$ are not truly counterfactual, it just that the agent is limited to only observing the outcome of the action they choose.
-To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
-}
-\label{fig:contextual-bandit-causal-graph}
-\end{figure}
-
-\vspace{0.5cm}
-\begin{example}
-\label{exm:anti-casual-context}
-Imagine a modified game of roulette.
-As in standard roulette, it consists of a spinning wheel with a ball that can land in one of 37 numbered pockets.
-However, unlike in the standard game, where punters place bets before the wheel is spun, in this variant the wheel is spun first.
-Five seconds before it comes to rest the face of the wheel darkens to hide the position of the ball from the players.
-They then place bets after it has stopped (and are each allowed only to check if the position they have bet on was correct).
-At the point in time the players are selecting their actions, the rewards for each possible choice are fully determined (all zero except for the pocket containing the ball).
-Suppose a canny gambler realises the ball makes a tiny sound, from rocking back and forth, if it is close to the left or right side of the wheel such that the pocket holding it is close to horizontal.
-The existence (or not) of this sound is a consequence of the position of the ball and thus the reward state and the gambler can use it as context to improve the likelyhood of obtaining a reward.
-\end{example}
-
-It is difficult to come up with realistic examples in which the full sequence of rewards is generated but only the one associated with the selected action is observable to the learner.
-In the typically cited applications of contextual bandit algorithms, such as selecting how to treat patients or which ad to serve only the reward of the selected action is ever realised, and the alternate rewards are counterfactual variables.
-
-\subsection{Learning from logged bandit data}
-\label{sec:off-policy-evaluation}
-Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
-In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
-The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
-
-This problem differs from the contextual bandit problem in that the learner is not interacting with the environment.
-As a result, there is no exploration-exploitation trade-off to be made.
-However, the problem does not reduce to supervised learning, because of the bandit nature of the feedback - only the reward of the selected action is observed.
-In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
-The majority of the literature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
-\citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original estimator $\pi$ did not depend on $\vec{X}$.
-
-If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, discussed in section \ref{sec:causal_effect_under_ignorability}.
-The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ satisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
-The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivalent to $do(A=a)$.
-However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see \citet{Pearl2000}, section 4.2.
-In this case, letting $\Pns{\pi'}{a|\vec{x}}$ denote the distribution over actions under policy $\pi'$ given context $\vec{x}$, the expected (per round) reward obtained by $\pi'$ is given by,
-
-\eqn{
-\E{y|\pi'} = \E{y|do(a \sim \pi'(\vec{x}))} = \Esub{(\vec{x},a) \sim \P{\vec{x}}\Pns{\pi'}{a|\vec{x}}}{y|\vec{x},a}
-}
-
-As in estimating average causal effects under ignorability, we have a covariate shift problem, with training data sampled from $\P{\vec{x}}\Pns{\pi}{a|\vec{x}}\P{y|\vec{x},a}$ but generalisation error measured with respect to $\P{\vec{x}}\Pns{\pi'}{a|\vec{x}}\P{y|\vec{x},a}$.
-A difference in practice, is that in the applications frequently considered under learning from logged feedback data, such as ad serving or recommender systems, there may be substantial information available about $\pi$, in the best case, $\Pns{\pi}{a|\vec{x}}$ is known.
-This makes estimators utilising inverse propensity weighting, including doubly robust estimators as in \citet{Dudik2011}, more attractive.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
-\node[main node](1){$\vec{X}$};
-\node[main node, below right=of 1](2){$Y$};
-\node[main node,  below left=of 1](3){$A$};
-\node[lt, above=of 2](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2) edge (3)
-   (4) edge (1) edge (2);
-\end{tikzpicture} 
-\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$.
-The outcome $Y$ may depend on $\vec{X}$ directly, indirectly through a latent variable $U$ or both.}
-\label{fig:off-policy-evaluation}
-\end{figure}
-
-\citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the empirical best because the variance of the estimators for some policies may be much higher than for others.
-
-\subsection{Adding structure to actions}
-\label{sec:adding-structure-bandit-actions}
-
-The classic multi-armed bandit is a powerful tool for sequential decision making.
-However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \citep{filippi2010parametric}, dependent bandits \citep{Pandey2007}, X-armed bandits \citep{Bubeck2010} and Gaussian process bandits \citep{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback\citep{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \citep{Piccolboni2001,Bartok2014}.
-
-In the next chapter, I propose a very natural connection between causal graphs and bandit problems and show it induces a novel form of additional feedback and structure between arms that cannot be can exploited by any of these previous approaches.
-
-\section{The counterfactual nature of regret}
-\label{sec:counterfactual-nature-regret}
-
-We conclude this chapter on bandits with a note on the counterfactual nature of regret.
-I have not seen this discussed with respect to bandit regret elsewhere, however the issues raised are very closely related to the work by \citet{Dawid2000}, on problems associated with the use of counterfactual variables for (observational) causal inference.
-
-It is possible, as we demonstrated in example \ref{exm:anti-casual-context}, to construct bandit problems in which the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$ is actually generated at each time-step (and, at least in principle, is observable to someone although not the agent).
-However, in most real applications for bandit problems, only the outcome associated with the selected action is ever realised and so the remaining components of the reward vector are counterfactual, justifying our use of the counterfactual notation to denote them.
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-This makes it easy to make statements using counterfactuals that cannot be confirmed empirically (even with infinite experimental data).
-We now show with a simple example that the standard definition of regret is such a fundamentally counterfactual quantity.
-Recall that the cumulative regret is defined by,
-
-\eqn{
-\label{eqn:raw_regret2}
-\rawregret &= \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
-\\&= \sum_{t=1}^T \left(\cfb{Y}{i^*} - \cfb{Y}{a_t}\right)
-}
-
-Take a stochastic, two-armed, Gaussian bandit with the joint distribution of the (counterfactual) rewards $\P{\cfb{Y}{1},\cfb{Y}{2}}$ given by equation \ref{eqn:joint-gaussain-reward}.
-Suppose without loss of generality that arm 1 is the optimal arm, such that $\mu_1 > \mu_2$.
-
-% vector with mean mu_t,mu_c and a covariance matrix
-\eqn{
-\label{eqn:joint-gaussain-reward}
-P(\cfb{Y}{1},\cfb{Y}{2}) \sim N( 
-         \begin{bmatrix}
-           \mu_{1} \\
-           \mu_2 \\
-         \end{bmatrix},
-         \begin{bmatrix}
-           \sigma^2 & \rho \sigma^2 \\
-           \rho \sigma^2 & \sigma^2 \\
-         \end{bmatrix}
-         )
-}
-
-Consider an algorithm that always selects arm 2.
-The distribution over the difference of jointly normal random variables is also normal, letting, $\tau_t = \cfb{Y}{1} - \cfb{Y}{2}$ yields $\P{\tau_t} = N(\mu_1 - \mu_2,2\sigma^{2}(1-\rho))$.
-Thus the distribution over the regret for this algorithm is given by,
-
-\eqn{
-\rawregret \sim N\left(T(\mu_1 - \mu_2), 2\sigma^2 T(1-\rho)\right)
-}
-
-The parameters of the marginal distributions, $\P{\cfb{Y}{1}}$ and $\P{\cfb{Y}{2}}$, can be estimated directly by simply sampling from each arm.
-However, the covariance $\rho$ cannot, because we never simultaneously observe the rewards for both arms.
-As a result, even if we have full knowledge of the reward distributions for each arm, we cannot in general compute the distribution over regret without un-testable assumptions about the covariance between counterfactuals, see figure \ref{fig:metaphysical_distribution_difference}.
-
-\begin{figure}[h]
-\captionsetup[subfigure]{position=b}
-\centering
-\subcaptionbox{Marginal distributions over the rewards for each action for $\mu_1=1$, $\mu_2=0$ and $\sigma = 1$.
-\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
-\hspace{0.02\textwidth}
-\subcaptionbox{The distribution over $R_T$ for $T=10$ for two different values of $\rho$ \label{fig:metaphys_b}}
-{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
-\caption{The distribution over the regret as defined by equation \ref{eqn:raw_regret2} depends on unobservable properties of the joint distribution over counterfactual rewards.
-The same (marginal) distributions over the rewards can correspond to quite different regret distributions.}
-\label{fig:metaphysical_distribution_difference}
-\end{figure}
-
-This result is somewhat perturbing.
-The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep and the behaviour of standard bandit algorithms depends only on the marginal reward distributions for each arm.
-The expectation of the regret as defined by equation \ref{eqn:pseudo_regret} also remains unchanged as both it's definition and the learner's actions depend only on the marginal distributions.
-It seems therefor unfortunate that we should have to assume, for example, that the rewards for alternate actions are independent of one-another in order to be able to analyse the variance of the regret.
-Particularly as this assumption is likely to be violated in many realistic bandit problems.
-For example, a given user may be more (or less) likely to buy something no matter which ad they are served leading to a positive correlation between counterfactual rewards, or an illness might have two (unobservable) sub-types with each medication (bandit arm) effective only against one, resulting in a negative correlation between the counterfactual rewards.
-
-We could focus only on analysing the expected regret, since this depends only on the marginal distributions.
-However, there are many real problems for which we do care how tightly concentrated the regret is around its expectation.
-For example, if we are risk averse, we may prefer an algorithm with slightly higher expected regret but a lower probability of suffering extremely large regret.
-This raises the question, can we construct an alternate definition of regret that allows us to capture how consistently bandit algorithms behave, but which does not depend on any properties of the joint distribution over counterfactual rewards.
-A natural candidate would be,
-
-\eqn{
-R_T = \sum_{i=1}^k N_i(T)\Delta_i, 
-}
-
-where $N_i(T)$ is the number of times arm $i$ was played up to timestep $T$ and $\Delta_i$ is the degree to which arm $i$ is sub-optimal, $\Delta_i = \mu^* - \mu_i$.
-The expectation of this variant of regret is the same as for the standard version defined in equation \ref{eqn:raw_regret2}.
-It depends only on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions.
-Furthermore, this quantity has already been analysed in existing work on the concentration of bandit regret, \citep{Audibert2007,Audibert2009exploration} as a more tractable proxy to the standard regret.
-In conclusion, when selecting measures of bandit performance, it is worth noting whether or not they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.
-
-\chapter{Causal Bandits: Unifying the approaches}
-\label{chap:causal-bandit}
-
-As we have seen in the previous two chapters, critical insights and methods for learning how to act have emerged from both the observational and experimental approaches to causality.
-From observational inference, we have the do-calculus, which allows us to map information between observational and interventional distributions and we have a range of estimators designed for these types of mapping problems including propensity scores and doubly robust methods.
-The bandit community has developed carefully tuned algorithms to balance exploration and exploitation, estimation approaches for identifying and selecting the optimal action (rather than estimating the rewards for all actions) and techniques to bound the bias of estimators due to the adaptive nature of decision making processes.
-
-However, despite the commonality of the underlying problem, there has been relatively little work on intersection between observational causal inference and adaptive decision making processes.
-Although some ideas, such as doubly robust estimators and inverse propensity score weighting have transferred across (from causal inference to off-policy evaluation), there are many more opportunities for fruitful work in this space.
-In most key real world decision making processes, from algorithms recommending drug treatments to robots deciding how to open doors, we have access to both potentially huge observational data sets as well as the ability to intervene in the system at some level.
-It is critical that we develop methods that allow us to incorporate the causal knowledge we have about the world, along with data sets collected under a range of different conditions, into our current decision making process.
-In this chapter I introduce a very general framework that connects causal graphical models with multi-armed bandit problems and demonstrate how it can be leveraged to make better decisions.
-
-\section{A framework to connect bandit problems with observational causal inference}
-\label{sec:intro_causal_bandit}
-A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
-Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{fig:unify_frameworks} for a simple example.
-In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example the set of single variable interventions.
-The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node.
-We refer to these problems as \emph{causal bandit problems}.
-In this thesis I focus on the case where the causal graph is known.
-Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.
-
-\begin{figure}
-\centering
-\begin{subfigure}[c]{0.3\textwidth}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
-\node[main node](1){W};
-\node[tg, below right=of 1](2){Y};
-\node[main node, above right=of 2](3){X};
-\path[]
-	(1) edge (2)
-	(3) edge (2);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-$\actionspace$ = \begin{tabular}{|c|}
-	\hline
-  $do(W=0,Z=0)$ \\
-  $do(W=0,Z=1)$ \\
-  $do(W=1,Z=0)$ \\
-  $do(W=1,Z=1)$ \\
-  \hline
-  $do(W=0)$ \\
-  $do(W=1)$ \\
-  $do(Z=0)$ \\
-  $do(Z=1)$ \\
-  $do()$ \\
-  \hline
-\end{tabular}
-\end{subfigure}
-\caption{A simple causal graphical model and corresponding complete action space. $W$ and $Z$ represent binary variables that can be intervened on and $Y$ represents the reward.}
-\label{fig:unify_frameworks}
-\end{figure}
-
-The type of problem we are concerned with is best illustrated with an example.
-Consider a farmer wishing to optimise the yield of her crop.
-She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
-In each season the farmer has enough time and money to intervene and control at most one of these
-variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry.
-When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
-How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
-
-We will assume each variable only takes on a finite number of distinct values.
-(The path to relaxing this assumption would be through levering the work on continuous armed bandits).
-The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
-An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and
-the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
-
-\vspace*{.5cm}
-\begin{definition}[Causal bandit problem]
-A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
-Each action $a \in \actionspace$ assigns a value to a subset of the variables in $\mathcal{X}$ and incurs a known cost $C(a)$ on the learner.
-One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable}.
-
-The causal bandit game proceeds over $T$ rounds. In each round $t$, the learner:
-\begin{enumerate}
-\item \emph{observes} the value of a subset of the variables $\vec{X}^c_t$,
-\item \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations and rewards, 
-\item \emph{observes} values for another subset of variables $\vec{X}^o_t$ drawn from $\P{\vec{X}^o_t |\vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$,
-\item \emph{obtains reward} $r_t = Y_t - C(a_t)$, where $Y_t$ is sampled from $\P{Y_t| \vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$
-\end{enumerate} 
-\end{definition}
-
-We refer to the set of variables that can be observed prior to selecting an action $\vec{X}^c$ as contextual variables and the set of variables observed after the action is chosen, $\vec{X}^o$, as post-action feedback variables.
-Note that $\vec{X}^c$ and $\vec{X}^o$ need not be disjoint.
-A variable may be observed both prior to and after the agent selects an action, and the action may change its value.
-The notation $\P{\cdot|\vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$ denotes distributions conditional on having observed $\vec{X}^c_t$ and \emph{then} intervened to set $\vec{X}_t = \vec{x}_t$.
-The values of variables in $\vec{X}^c_t$ that are non-descendents of $\vec{X}_t$ remain unchanged by the intervention.
-The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
-
-The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
-If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant and the problem reduces to a standard multi-armed bandit problem.
-This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.
-
-\vspace{0.5cm}
-\begin{theorem}
-\label{theorem:cb-intervene-all-parents}
-Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$.
-If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and the problem reduces to a standard multi-armed bandit (over actions in in $\mathcal{A'}$).
-\begin{proof}
-for any action $a \in \mathcal{A}$, 
-\eq{
- \E{Y|\vec{X}^c_t,a} &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\vec{X}^c_t,a,\parents{Y}}} \\
- &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\parents{Y}}}\\
- &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|do(\parents{Y})}}\\
- &\leq \max_{\parents{Y}}\E{Y|do(\parents{Y})} = \E{Y|a'} \text{ for some $a' \in \mathcal{A'}$}
-}
-\end{proof}
-\end{theorem}
-
-If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
-If we receive post-action feedback on additional nodes the problem can be more interesting.
-In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
-Consider again the model in figure \ref{fig:unify_frameworks}.
-The causal structure implies:
-
-\eq{
-P(Y|do(W = 0)) &= P(Y|do(),W = 0) \\
-&= P(Y|do(X=0),W=0)P(X=0)+P(Y|do(X=1),W=0)P(X=1) 
-}
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
-\node[node](1){Health conscious};
-\node[node, below left=of 1](2){Diet};
-\node[node, below right=of 1](3){Exercise};
-\node[node, below right=of 2](4){Weight};
-\node[tg,below left=of 4](5){Y=Cholesterol};
-\node[node, below =of 5](6){Test result};
-\node[node, above right=of 3](7){Free time};
-\path[]
-	(1) edge (2) edge (3)
-	(2) edge (4) edge (5)
-	(3) edge (4)
-	(5) edge (6)
-	(7) edge (3);
-\end{tikzpicture}
-\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level .
-The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
-For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
-\label{fig:cholesterol_graph}
-\end{figure}
-
-Thus we gain information about the reward for the action $do(W=0)$ from selecting the action $do()$ or $do(X = x)$ and then observing $W = 0$.
-We only get this form of side information for actions that don't specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-If additional variables are only observed before an intervention is selected the causal bandit problem can be treated as a stochastic contextual bandit problem, which are already reasonably well understood \citep{Agarwal2014}.
-
-Two other problems that sit in the space between causal inference and bandit problems, bandits with unobserved confounders \cite{Bareinboim2015} and compliance aware bandits \citep{DellaPenna2016}, can also be viewed as specific causal bandit problems.
-In the work on bandits with unobserved confounders it is assumed that the reward given each action may depend on some latent variable $U$ which we cannot observe directly.
-However, prior to selecting an action, we can observe  $I$, the action that would have been selected under an alternate (stationary) policy, which may depend on $U$, see figure \ref{fig:confounded-causal-bandit-graph}.
-In this case the set of contextual variables $\vec{X}^c = I$, the set of post-action feedback variables $\vec{X}^o = \set{Y}$ and the action space consists of all possible assignments of values to a single node $X$, $\actionspace = do(X=x)$.
-This setting reduces to a contextual bandit problem in our causal bandit framework.
-However in their work on bandits with unobserved confounders \citet{Forney2017} also leverage the fact that $I$ represents the action that would have been selected under an alternate policy to fuse data collected under the previous (observational) policy with data collected under the new policy.
-This information is not encoded in the causal bandit graph in figure \ref{fig:confounded-causal-bandit-graph}, as $I$ could be any variable that is influenced by the unobserved context $U$, and thus cannot be exploited by a standard contextual bandit algorithm.
-
-Compliance aware bandits describe situations in which the action recommended by the bandit algorithm is not always followed.
-For example a patient may refuse to take a treatment or an advertiser may have complex rules about how many ads a given customer can receive which prevents some of the suggestions from the ad recommendation engine from being followed.
-After an action is selected the algorithm can observe the action that was actually taken in addition to the reward.
-\citet{DellaPenna2016} analyse this setting with binary treatments both with and without the presence of a latent confounding variable $U$, see figure \ref{fig:imperfect-compliance-causal-bandit-graph}.
-In this case, there are no contextual variables and the action space is again the set of assignments to a single variable but there is post-action feedback, which reveals the value of the action that was actually taken.\footnote{There are some interesting variants of the compliance aware bandit setting that, to my knowledge, have not been analysed.
-The first is if the confounding variable $U$ is observable, either as context or post-action feedback.
-The second is if we extend the allowable action set to include acting directly on $X$, albeit at a higher cost than acting on $A$.
-It is also worth noting the connection between this setting and instrumental variables \cite{balke1997bounds}.
-By making some functional assumptions about the relationships between the variables, we can use $A$ as an instrumental variable to bound or estimate the (casual) effect of $X$ on $Y$.
-The estimation will be somewhat complicated in the bandit setting because the action chosen at each timestep is dependent on the previous sequence actions and rewards.}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.49\textwidth}
-\centering
-\caption{Bandits with unobserved confounders:$\vec{X}^c=\set{I}$, $\vec{X}^o = \set{Y}$, $\actionspace = do(X=x)$}
-\label{fig:confounded-causal-bandit-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){${U}$};
-\node[main node, below right =of 1](2){${Y}$};
-\node[main node, below left=of 1](3){$I$};
-\node[main node, above right=of 2](4){$X$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (2) edge (3)
-   (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.49\textwidth}
-\centering
-\caption{Compliance aware bandits: $\vec{X}^c=\set{}$, $\vec{X}^o = \set{T,Y}$, $\actionspace = do(X=x)$ }
-\label{fig:imperfect-compliance-causal-bandit-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$U$};
-\node[main node, below left=of 1](2){$T$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$X$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (2) edge (3)
-   (2) edge (3)
-   (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{}
-\label{fig:causal-bandit-version-confounded-compliance}
-\end{figure}
-
-The classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
-The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.
-However, the number of actions or arms grows exponentially with the number of variables in the graph making it important to develop algorithms that leverage the graph structure and additional observations.
-
-\section{Causal bandits with post action feedback}
-\label{sec:causal-bandit-post-action-feedback}
-
-We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, the cost of all allowable actions is equal and the goal of the learner is to minimise the simple regret.
-I presented this work at NIPS 2016 \citep{NIPS2016_6195}.
-
-\paragraph{Related Work} As alluded to above, causal bandit problems can be treated as classical multi-armed bandit problems by simply ignoring the causal model and extra observations and applying an existing best-arm identification algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
-However, as we show in \S\ref{sec:simple-regret}, ignoring the extra information available in the non-intervened variables yields sub-optimal performance.
-
-Our framework bears a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
-However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
-
-There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on un-played actions are revealed according to a feedback graph.
-As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
-They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
-The partial monitoring approach taken by \citet{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
-\citet{Yu2009} utilise extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
-\citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled.
-The main difference from our present work is our focus on simple regret and the more complex information linking rewards for different arms via causal graphs.
-To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.
-
-%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
-They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
-% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \citep{Bartok2014}.
-Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
-
-Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
-\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been 
-intervened on can provide important contextual information. Their work differs in many ways. For example, the focus is on the cumulative regret and
-the context is observed before the action is taken and cannot be controlled by the learning agent.
-
-%\todof{write one sentence to explain how their work is different to ours}
-
- \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions.
-Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
-Combining their handling of  causal induction with our analysis is left as future work.
-% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
-Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
-(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
-It is easy to come up with a counter example in which it does not).
-
-The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{Bottou2013}, 
-where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discuss some of the issues 
-encountered in sequential design, but do not give an algorithm or theoretical results for this case.
-
-\subsection{The parallel bandit problem}
-\label{sec:simple-regret}
-In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem which we call the {\it parallel bandit}.
-It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro_causal_bandit}, including the farming example.
-It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
-
-The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:parallel}.
-All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
-
-In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high).
-The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.
-
-\begin{figure}
-    \begin{subfigure}[t]{0.34\textwidth}
-	\centering   
-	 \caption{Parallel graph}\label{fig:parallel} 
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-\end{tikzpicture} 
-\end{subfigure}
-    \begin{subfigure}[t]{0.25\textwidth}
-    \centering
-     \caption{Confounded graph}
-        \label{fig:causalStructure_confounded}
-        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_1$};
-\node[main node, below left=of 1](2){$X_2$};
-\node[main node, below right=of 1](4){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-    (1) edge (4)
-    (2) edge (4);
-\end{tikzpicture}
-    \end{subfigure}
-    \begin{subfigure}[t]{0.4\textwidth}
-    \centering
-     \caption{Chain graph}
-        \label{fig:causalchain}
-         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, right=of 4](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-  	(2) edge (3)
-    (3) edge (4)
-    (4) edge (5);
-\end{tikzpicture}
-    \end{subfigure}
-      \caption{Causal Models}\label{fig:causalmodels}\label{fig:causal_models_causal_bandit}
-\end{figure}
-
-In each round the learner either purely observes by selecting $do()$ or sets the value of a single variable.
-The remaining variables are simultaneously set by independently biased coin flips.
-The value of all variables are then used to determine the distribution of rewards for that round.
-Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
-
-The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
-$r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
-In the farming example, this choice of $Y$ models the success or failure of a seasons crop, 
-which depends stochastically on the various environment variables.
-
-\paragraph{The Parallel Bandit Algorithm}
-%\label{sub:par-bandit-alg}
-The algorithm operates as follows.
-For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
-Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
-The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates of their returns could be poor.
-To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions.
-The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$).
-For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$.
-Define
-
-\eq{
-\label{eq:m-simple}
-m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
-}
-
-\begin{algorithm}[H]
-\caption{Parallel Bandit Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} Total rounds $T$ and $N$.
-\FOR{$t \in 1,\ldots,T / 2$}
-\STATE Perform empty intervention $do()$
-\STATE Observe $\vec{X}_t$ and $Y_t$
-\ENDFOR
-\FOR{$a = do(X_i = x) \in \actions$}
-\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
-\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
-\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
-\ENDFOR
-\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
-\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
-\FOR{$a = do(X_i = x) \in A$}
-\FOR{$t \in 1,\ldots,T_A$}
-\STATE Intervene with $a$ and observe $Y_t$
-\ENDFOR
-\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
-\ENDFOR
-\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
-\end{algorithmic}
-\end{algorithm}
-
-$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimise the worst-case simple regret.
-When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
-We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilises the samples captured during the observational phase to estimate $m(\vec{q})$.
-Although very simple, the following two theorems show that this algorithm is effectively optimal.
-
-\begin{theorem}\label{thm:uq-simple}
-Algorithm \ref{alg:simple} satisfies
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.
-}
-\end{theorem}
-
-\begin{theorem}\label{thm:lower}
-For all strategies and $T$, $\vec{q}$, there exist rewards such that
-$\displaystyle \simpleregret 
-\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)$.
-\end{theorem}
-
-The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
-of $\hat p_a$ and $\hat m$ about their true values and may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
-
-%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
-By utilising knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
-Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$.
-Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting.
-In practice, you would combine the data from both phases to estimate rewards for the low probability actions.
-We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.
-
-\subsection{General graphs}
-\label{sec:simple-regret-general}
-We now consider the more general problem where the graph structure is known, but arbitrary.
-For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorisation formula \citep{Pearl2000}.
-\eq{
-\P{X_1...X_N|do(X_i=j)} = 
-\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
-} 
-where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the Dirac delta function.
-
-We could naively generalise our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorisation to 
-write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
-estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
-from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
-takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
-simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
-high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
-Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
-get a good estimate for $\P{Y|do(X_2=1)}$.
-
-To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
-allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
-are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
-or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
-start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
-Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
-Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
-
-\begin{algorithm}[H]
-\caption{General Algorithm}\label{alg:general}
-\begin{algorithmic}
-\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
-\FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $a_t$ from $\eta$
-\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
-\ENDFOR
-\FOR{$a \in \calA$}
-\STATE
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
-}
-\ENDFOR
-\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
-\end{algorithmic}
-\end{algorithm}
-
-Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
-Let $\parents{Y}(X)$ denote the realisation of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
-
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}\,, 
-}
-
-where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently.
-The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
-
-The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimised to minimise the worst-case simple regret. Abusing notation we define $m(\eta)$ by
-\eq{
-m(\eta) = \max_{a \in \calA} \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,,\text{ where } \EEa \text{ is the expectation with respect to } \Pn{a}.
-}
-
-We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
-
-\begin{theorem}\label{thm:general}
-If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$
-
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
-}
-\end{theorem}
-
-The proof is in Section \ref{sec:thm:general}.
-
-Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
-Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$.
-Thus it has broader applicability to the parallel graph than the generic algorithm given here.
-We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to mini-max optimal, but leave lower bounds
-for future work.
-
-\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$.
-In light of Theorem \ref{thm:general}
-a natural choice of $\eta$ is the minimiser of $m(\eta)$.
-\eq{
-\eta^* 
-= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
-}
-Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
-Therefore the minimisation problem may be tackled using standard techniques from convex optimisation.
-The quantity $m(\eta^*)$ may be interpreted as the minimum achievable worst-case variance of the importance weighted estimator.
-In the experimental section we present some special cases, but for now we give two simple results.
-The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
-
-\begin{proposition}\label{pro:m-bound}
-$m(\eta^*) \leq |\calA|$. \textit{Proof.} 
-\textup{By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|\,\forall a$.}
-\eq{
-m(\eta) 
-= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
-= \max_a \EEa\left[\frac{1}{\eta_a}\right] = |\calA| %\qedhere
-}
-\end{proposition}
-
-The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$.
-This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
-The proof is in section \ref{sec:m-equivelence}.
-
-\begin{remark}\label{rem:truncate}
-The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
-often possible to choose $B_a$ significantly
-larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
-and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{Bottou2013} who also use importance weighted estimators
-to learn from observational data.
-\end{remark}
-
-\subsection{Experiments}
-\label{sec:causal-bandit-experiments}
-We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{Audibert2010}, Thompson Sampling and UCB under a variety of conditions.
-Thomson sampling and UCB are optimised to minimise cumulative regret.
-We apply them in the fixed horizon, best arm identification setting by running them up to horizon $T$ and then selecting the arm with the highest empirical mean.
-The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.
-
-Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms).
-$Y_t \sim \bernoulli(\frac{1}{2}+\epsilon)$ if $X_1=1$ and $Y_t \sim \bernoulli(\frac{1}{2}-\epsilon')$ otherwise, where $\epsilon' = q_1\epsilon/(1-q_1)$.
-This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
-We set $q_i = 0$ for $i \leq m$ and $\frac{1}{2}$ otherwise.
-Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
-For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
-The code is available from \url{<https://github.com/finnhacks42/causal_bandits>}
-
-In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
-The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure.
-For the causal algorithms it increases approximately with $\sqrt{m}$.
-As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.
-
-\begin{figure}[h]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
-    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment2_20161020_1249.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
-        \label{fig:simple_vs_T_vary_epsilon}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment3_20161020_1252.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
-    		\label{fig:simple_vs_T}
-    \end{subfigure}
-    \caption{Experimental results}
-    \label{fig:experiments}
-\end{figure}
-
-Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
-This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
-In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
-The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.
-
-For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
-This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be mini-max optimal asymptotically \citep{li2014minimax}.
-
-\begin{figure}[H]
-	\centering    
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
-\node[main node, above right=of 2](6){$Z$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5)
-    (6) edge (1) edge (2) edge (4);
-\end{tikzpicture}
-        \caption{Confounded graph}
-        \label{fig:parallel_confounded} 
-\end{figure}
-
-We now compare the general algorithm with a range of standard bandit algorithms on the confounded graph in Figure \ref{fig:parallel_confounded}.
-All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action,
-
-\eq{
-\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}
-}
-
-We choose this setting because it generalises the parallel bandit, while simultaneously being sufficiently simple that we can compute the exact reward and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
-As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.
-
-In Figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$. For some $2 \leq N_1 \leq N$ we define 
-\eq{
-P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
-P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
-}
-As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon P(X_1=1) / P(X_1=0)$.
-The value of $N_1$ determines $m$ and ranges between $2$ and $N$.
-The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
-
-In Figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.  
-In Figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1, but exclude actions on $Z$ from the set of allowable actions to demonstrate that Algorithm 1 can fail in the presence of a confounding variable, which occurs because it incorrectly assumes that $P(Y|do(X)) = P(Y|X)$.
-We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
-
-\eq{
-P(X_i = 1|Z = 0) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.2 & \text{ if } i = 7 \\
-.7 & \text { otherwise} 
- \end{cases}\\
- P(X_i = 1|Z = 1) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.8 & \text{ if } i = 7 \\
-.3 & \text { otherwise} 
- \end{cases}\\
-}
-
-In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree.
-It is sub-optimal to act on either $X_7$ or $X_N$, while all other actions are optimal.
-The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.
-All remaining variables are likely to be identified as balanced and estimated from observational estimates.
-The CPD values have been chosen to demonstrate the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
-
-\begin{figure}[H]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
-    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
-        \label{fig:simple_vs_T_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment5_20161023_2118.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m(\eta*)=4.3$ with no actions setting $Z$}
-    		\label{fig:simple_vs_T_misspecified}
-    \end{subfigure}
-    \caption{Experimental results on the confounded graph}
-    \label{fig:experiments_confounded}
-\end{figure}
-
-\subsection{Discussion \& Future work}
-\label{sec:causal_bandit_discussion}
-Algorithm~\ref{alg:general} for general causal bandit problems 
-estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
-Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ where 
-the parameter $m(\eta)$ measures the difficulty of learning the causal model and is always less than $N$.
-The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.
-
-The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator $\hat{\mu}_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $\Pn{a}{\parents{Y}}$ for all $a \in \calA$.
-In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.
-Despite these extra estimates, Theorem~\ref{thm:lower} shows that this approach is optimal (up to log factors).Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
-
-\paragraph{Making Better Use of the Reward Signal}
-Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
-In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
-Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
-This would likely improve the poor performance of our causal algorithm relative to the successive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
-
-For the parallel bandit the required modifications should be quite straightforward.
-The idea would be to adapt the algorithm to essentially use successive elimination in the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability.
-In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
-This has now been done by \citet{Sen2017a}, leading to problem dependent regret bounds for causal bandit problems.
-Note that these modifications do not improve the mini-max regret, which at least for the parallel bandit is already optimal.
-For this reason we focused on emphasising the point that causal structure can and should be exploited when available.
-Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
-This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.
-
-\paragraph{Cumulative Regret}
-Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
-In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information
-to get near-optimal cumulative regret guarantees.
-They consider a finite-armed bandit model with side information where in reach round the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
-In this way the learner can gain information about actions it does not take with varying levels of accuracy.
-The reduction follows by substituting the importance weighted estimators in place of the Gaussian reward.
-In the case that $\vec{q}$ is known this would lead to a known variance and the only (insignificant) difference is the Bernoulli noise model.
-In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
-at least asymptotically.
-
-%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
-%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
-%\todom{Check this!}
-%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
-
-The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each time step the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
-The feedback graph is distinct from the causal graph.
-A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
-For this parallel bandit problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
-However, \citet{Alon2015,Kocak2014} give adversarial algorithms, which when applied to the parallel bandit problem obtain the standard bandit regret.
-A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
-This is equivalent to a nominally stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
-
-\citet{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
-In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
-\citep{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-\paragraph{Causal Models with Non-Observable Variables}
-If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
-non-observable variables. Some of the interventional distributions may be non-identifiable meaning we can not obtain prior estimates for $\Pn{a}{\parents{Y}}$ from 
-even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
-\ref{alg:general} cannot be used. Estimating these quantities while simultaneously minimising the simple regret is an interesting and challenging open problem.
-
-% For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
-% In this case, some conditional distributions may be non-identifiable. 
-% The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
-% We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
-% Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
-
-% More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
-In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
-The quality of the estimate notably depends not only on the number of times that action was selected.
-For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$.
-This more complex side information does not fit within the graph feedback framework.
-
-\paragraph{Partially or Completely Unknown Causal Graph}
-A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
-The latter case arises naturally if we assume free access to a large observational data set, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\citep{eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
-In particular, \citet{Hu2014} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
-Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational data set would be much more challenging.
-
-\subsection{Proofs}
-\subsubsection{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}
-
-Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$. The assumption is non-restrictive since all variables
-are independent and permutations of the variables can be pushed to the reward function.
-
-The proof of Theorem \ref{thm:uq-simple} requires some lemmas. \
-
-\begin{lemma}\label{lem:conc1}
-Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
-\eq{
-\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \citet{Hagerup1990}),
-
-\eq{
-\P{\left|\hat q_i - q_i\right| \geq \epsilon} \leq 2e^{-\frac{T\epsilon^2}{6q_i}}
-}
-
-Letting $\delta = 2e^{-\frac{T\epsilon^2}{6q_i}}$ and solving for $\epsilon$ completes the proof.
-
-\end{proof}
-
-\begin{lemma}\label{lem:m_est}
-
-Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
-\eq{
-\P{2m(\vec{q}) / 3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})} \geq 1 - \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-Let $F$ be the event that there exists and $1 \leq i \leq N$ for which
-\eq{
-\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}\,.
-}
-Then by the union bound and Lemma \ref{lem:conc1} we have $\P{F} \leq \delta$. The result will be completed by showing that
-when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
-From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m(\vec{q})$ that $q_i \geq q_m \geq 1/m(\vec{q})$ and so by Lemma \ref{lem:conc1} we have
-\eq{
-\frac{3}{4} 
-&\geq \frac{1}{2} + \sqrt{\frac{3}{T} \log \frac{2N}{\delta}} 
-\geq q_i + \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}} 
-\geq \hat q_i \\
-&\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
-\geq q_i - \sqrt{\frac{q_i}{8m(\vec{q})}}
-\geq \frac{1}{2m(\vec{q})}\,.
-}
-Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m(\vec{q})$.
-For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m(\vec{q})$ that
-\eq{
-\hat q_i 
-\leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
-\leq \frac{1}{m(\vec{q})} \left(1 + \sqrt{\frac{1}{8}}\right)
-\leq \frac{3}{2m(\vec{q})}\,.
-}
-Therefore $m(\vec{\hat q}) \geq 2m(\vec{q}) / 3$ as required. 
-\end{proof}
-
-\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
-Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then,
-for $a \in A$, the algorithm estimates $\mu_a$ from $T_A \doteq T/(2m(\vec{\hat q}))$ samples.
-From lemma \ref{lem:m_est}, $T_A  \geq T/(4m(\vec{q}))$ with probability $(1-\delta)$.
-Let $H$ be the event $T_A < T/(4m(\vec{q}))$ and $G$ be the event $\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}$
-
-\eq{
-\P{G} \leq \P{H} + \P{G|\neg H} \leq \delta + \P{G|\neg H}
-}
-
-Via Hoeffding's inequality and the union bound,
-
-\eq{
-\P{G|\neg H}&\doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}},\text{ given }T_A \geq T/(4m(\vec{q}))} \leq \delta\\
- \implies  \P{G} & \doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq 2\delta\,.
-}
-For arms not in $A$,
-
-\eq{
-\hat p_a = & \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq 1/m(\vec{\hat q}), \; \text {by definition of not being in $A$}\\
- \geq & \frac{1}{2m(\vec{q})},\; \text{ with probability } 1-\delta \\
-\implies  T_a \doteq & \sum_{t=1}^{T/2} \ind{X_i = j} \geq  \frac{T}{4m(\vec{q})},\; \text{ with probability } 1-\delta
-}
-
-Again applying Hoeffding's and the union bound
-
-\eq{
-\P{\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}}} \leq 2\delta
-}
-
-Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 4\delta$ that,
-\eq{
-(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}} \doteq \epsilon\,.
-}
-If this occurs, then 
-\eq{
-\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_{\hat a^*_T}] 
-& \leq 4\delta + \epsilon\\
-& \leq \frac{8m(\vec{q})}{T} + \sqrt{\frac{2m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ letting } \delta = \frac{2m(\vec{q})}{T}\\
-& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jenson's Inequality}
-}
-which completes the result.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LOWER BOUND
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}
-
-We follow a relatively standard path by choosing multiple environments that have different optimal arms, but which cannot all be statistically
-separated in $T$ rounds.
-Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
-For each $i$ define reward function $r_i$ by
-\eq{
-r_0(\boldsymbol{X}) &= \frac{1}{2} &
-r_i(\boldsymbol{X}) &= \begin{cases}
-\frac{1}{2} + \epsilon & \text{if } X_i = 1 \\
-\frac{1}{2} & \text{otherwise}\,,
-\end{cases}
-}
-where $1/4 \geq \epsilon > 0$ is some constant to be chosen later.
-We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
-environment determined by $\boldsymbol{q}$ and $r_i$. Let $\operatorname{P}_i$ be the corresponding measure
-on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \citet{Tsy08} we have
-\eq{
-\Prz{\hat a^*_T = a^*} + \Pri{\hat a^*_T \neq a^*} \geq \exp\left(-\KL(\operatorname{P}_0, \operatorname{P}_i)\right)\,,
-}
-where $\KL(\Ps_0, \Ps_i)$ is the KL divergence between measures $\operatorname{P}_0$ and $\operatorname{P}_i$.
-Let $T_i(T) = \sum_{t=1}^T \ind{a_t = do(X_i = 1)}$ be the total number of times the learner intervenes on variable $i$ by setting it to $1$.
-Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $\Ps_0$ and $\Ps_i$ may be bounded using the telescoping property (chain rule) and
-by bounding the local KL divergence by the $\chi$-squared distance as by \citet{Auer1995}. This leads to 
-\eq{
-\KL(\Ps_0, \Ps_i) 
-&\leq 6\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] 
-\leq 6\epsilon^2 \left(\EE_0 T_i(T) + q_i T\right) 
-\leq 6\epsilon^2 \left(\EE_0 T_i(T) + \frac{T}{m}\right)\,.
-}
-Define set $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
-Then for $i \in A$ and choosing $\epsilon = \min\set{1/4, \sqrt{m/(18T)}}$ we have
-\eq{
-\KL(\Ps_0, \Ps_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
-}
-Now $\sum_{i=1}^m \EE_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
-Therefore
-\eq{
-\sum_{i \in A} \Pri{\hat a^*_T \neq a} 
-\geq \sum_{i \in A} \exp\left(-\KL(\Ps_0, \Ps_i)\right) - 1
-\geq \frac{|A|}{e} - 1 
-\geq \frac{m}{2e} - 1\,.
-}
-Therefore there exists an $i \in A$ such that
-$\Pri{\hat a^*_T \neq a^*} \geq \frac{\frac{m}{2e} - 1}{m}$. 
-Therefore if $\epsilon < 1/4$ we have
-\eq{
-R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{m}{18T}}\,.
-}
-Otherwise $m \geq 18T$ so $\sqrt{m/T} = \Omega(1)$ and
-\eq{
-R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{1}{4} \frac{\frac{m}{2e} - 1}{2m} \in \Omega(1) 
-}
-as required.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% GENERAL-GRAPH UPPER BOUND
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Proof of Theorem \ref{thm:general}}\label{sec:thm:general}
-
-\begin{proof}
-First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
-We define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$, $R_{at} = R_a(X_t)$ and $\Pn{a}{.} = \Pns{a}{.}$.
-By definition we have $|Z_{at}| \leq B_a$ and 
-\eq{
-\Var_Q[Z_{at}] 
-\leq \EE_Q[Z_{at}^2] 
-\leq \EE_Q[R_{at}^2]
-= \EEa[R_{at}]
-= \EEa\left[\frac{\Pns{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq m(\eta)\,.
-}
-Checking the expectation we have
-\eq{
-\EE_Q[Z_{at}] 
-= \EEa \left[Y \ind{R_{at} \leq B_a}\right] 
-= \EEa Y - \EEa \left[Y\ind{R_{at} > B_a}\right] 
-= \mu_a - \beta_a\,,
-}
-where 
-\eq{
-0 \leq \beta_a = \EEa[Y \ind{R_{at} > B_a}] \leq \Pns{a}{R_{at} > B_a}
-}
-is the negative bias. 
-The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
-\eq{
-\beta_a \leq \Pns{a}{R_{at} > B_a} \leq \frac{\EEa[R_{at}]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
-}
-Let $\epsilon_a > 0$ be given by
-\eq{
-\epsilon_a = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
-}
-Then by the union bound and Bernstein's inequality 
-\eq{
-\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} 
-\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
-}
-
-Let $I = \hat{a}^*_T$ be the action selected by the algorithm, $a^* = \argmax_{a \in \calA} \mu_a$ be the true optimal action and recall that $\EE_Q[Z_{at}] = \mu_a - \beta_a$.
-Assuming the above event does not occur we have,
-
-\eq{
-\mu_I \geq \hat \mu_I - \epsilon_I  
-\geq \hat \mu_{a^*} - \epsilon_I  
-\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
-}
-By the definition of the truncation
-we have
-\eq{
-\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
-}
-and
-\eq{
-\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
-}
-Therefore for $C = \sqrt{2} + 4$ we have
-\eq{
-\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
-}
-as required.
-\end{proof}
-
-\subsubsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}\label{sec:m-equivelence}
-
-\begin{proposition} In the parallel bandit setting,
-$m(\eta^*) \leq 2m(\boldsymbol{q})$.
-\end{proposition}
-
-\begin{proof}
-
-Recall that in the parallel bandit setting,
-
-\eq{
-\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}
-}
-
-Let:
-
-\eq {
-\eta_a = \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}}\frac{1}{2m(\boldsymbol{q})} \text { for } a \in do(X_i = j)
-}
-
-Let $D =\sum_{a\in do(X_i=j)}\eta_a$. From the definition of $m(\boldsymbol{q})$, 
-\eq {
-\sum_{a\in do(X_i=j)} \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}} \leq m(\boldsymbol{q}) \implies D \leq \frac{1}{2}
-}
-
-Let $\eta_a = \frac{1}{2} + (1-D)$ for $a = do()$ such that $\sum_{a \in \calA}\eta_a = 1$
-
-Recall that,
-
-\eq{
-m(\eta) &
-= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]
-}
-
-We now show that our choice of $\eta$ ensures $\EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \leq 2m(\boldsymbol{q})$ for all actions $a$.
-
-For the actions $a: \eta_a > 0$, ie $do()$ and $do(X_i = j):\P{X_i=j}<\frac{1}{m(\boldsymbol{q})}$,
-\eq{
-\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq \EEa\left[\frac{\Pn{a}{X_1...X_N}}{\eta_a\Pn{a}{X_1...X_N}}\right] = \EEa\left[\frac{1}{\eta_a}\right] \leq 2m(\boldsymbol{q})
-}
-
-For the actions $a :\eta_a = 0$, ie $do(X_i=j):\P{X_i=j}\geq\frac{1}{m(\boldsymbol{q})}$,
-\eq{
-\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq & \EEa\left[\frac{\ind{X_i=j}\prod_{k\neq i}\P{X_k}}{(1/2+D)\prod_k \P{X_k}}\right] \\=& \EEa\left[\frac{\ind{X_i=j}}{(1/2+D)\P{X_i = j}}\right]
-\leq  \EEa\left[\frac{\ind{X_i=j}}{(1/2)(1/m(\boldsymbol{q}))}\right] \leq 2m(\boldsymbol{q})
-}
-
-Therefore $m(\eta*) \leq m(\eta) \leq 2m(\boldsymbol{q})$ as required.
-
-\end{proof}
-
-\chapter{Conclusion}
-
-TODO - write conclusion \todo{}
-
-Here is the problem and why its important
-
-Here is how what I have done contributes to solving thes problems
-
-What remains to be done - directions for future work
-
-- Relaxing assumption that interventional distributions known
-- Algorithms for causal bandits in which both contextual and post-action feedback is available.
-- Off-policy evaluation for general causal bandits.
-
-Other approaches to identifying causal effects, such instrumental variables, the \quotes{front door} approach could also be applied to off-policy evaluation in some settings.
-
-Ideas developed within the off-policy selection liturature are also relevent classic causal problems within economics etc.
- - optimal action selection (rather than evaluation)
- - quantification of finite time properties of estimators (as opposed to asymptopic efficiency). eg \cite{li2014minimax}
-
-Warm start for (contextual) bandit problem based on potentially biased estimates. (We can detect the presense of a confounder with experimental data).
-
-Alternatives to cross-validation.
-
-\bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{library}% write the directory to the .bib file
-\end{document}
-
-\chapter{Causation vs Prediction}
-
-The two cultures \citep{Breiman2001}
- - wrong because suggests we can just use the machine learning approach for everything.
-Does not address the problems that a) structural assumptions required for causal inference.
-b) we cannot use cross-validation to assess the validity of a causal model (really b is the whole point).
-
-To explain or to predict \citep{Shmueli2010}
-- wrong because they assume all the parts of explanatory modelling are required - so black box models cannot be used.
-
-Predict the relationship between X and Y as a way to stratify to reduce the variance in an experiment.
-
-They consider a different definition of explanatory models,
-
-Definition as backwards.
-Explanatory models focus on theory because some theory (or equivalently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
-
-Causal modelling frameworks allow us to separate assumptions about the structure of the relationships between variables from assumptions about the functional form.
-
-What can be data driven and what cannot? (for a given set of assumptions).
-
-Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the language of EoP but are built on a hybrid of both approaches.

### 2017-09-10 thesis/original_statement.tex
fixes title pages

--- 2017-09-04 thesis/thesis_map.tex
@@
-\documentclass[11pt,a4paper,twoside]{report}
-\usepackage[table]{xcolor}
-\usepackage[utf8]{inputenc}
-\usepackage{textcomp}
-\usepackage{amsmath}
-\usepackage{dsfont}
-\usepackage{listings}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{tabularx}
-\usepackage{booktabs}
-\usepackage{enumitem}
-\usepackage{algorithm}% http://ctan.org/pkg/algorithm
-\usepackage[noend]{algorithmic}
-\usepackage{parskip}
-\usepackage{tikz}
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage{epigraph}
-\usetikzlibrary{arrows,fit,positioning}
-\usepackage{subcaption}
-\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[numbers]{natbib} %numbers
-\usepackage{wrapfig}
-\usepackage{centernot}
-
-% this is included to get tables with colored rows to not draw over hlines and vlines
-\usepackage{etoolbox}
-
-\makeatletter
-\patchcmd{\@classz}
-  {\CT@row@color}
-  {\oldCT@column@color}
-  {}
-  {}
-\patchcmd{\@classz}
-  {\CT@column@color}
-  {\CT@row@color}
-  {}
-  {}
-\patchcmd{\@classz}
-  {\oldCT@column@color}
-  {\CT@column@color}
-  {}
-  {}
-\makeatother
-
-\usepackage{amsthm}
-
-\graphicspath{ {figures/} }
-\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
-
-\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
-\DeclareUnicodeCharacter{00A0}{ }
-
-\tikzset{
-    %Define standard arrow tip
-    >=stealth',
-    %Define style for boxes
-    punkt/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           text width=1em,
-           minimum height=1em,
-           text centered},
-    observed/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.3em,
-           minimum height=2.3em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-     latent/.style={
-           circle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered
-           },
-    target/.style={
-           circle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-    observedrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!10!white
-           },
-    latentrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick, dashed,
-           minimum width=2.2em,
-           minimum height=2.2em,
-           font=\footnotesize,
-           text centered
-           },
-     targetrect/.style={
-           rectangle,
-           rounded corners,
-           draw=black, thick,
-           minimum width=6em,
-           minimum height=2em,
-           font=\footnotesize,
-           text centered,
-           fill=black!20!white,
-           },
-     empty/.style={
-           circle,
-           rounded corners,
-           minimum width=.5em,
-           minimum height=.5em,
-           font=\footnotesize,
-           text centered,
-           },
-    % Define arrow style
-    pil/.style={
-           o->,
-           thick,
-           shorten <=2pt,
-           shorten >=2pt,},
-    sh/.style={ shade, shading=axis, left color=red, right color=green,
-    shading angle=45 }   
-}
-
-\newcommand{\KL}{\operatorname{KL}}
-
-\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
-\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
-\newcommand{\Q}[1]{\operatorname{Q}\left(#1\right)}
-\newcommand{\EE}{\mathbb E}
-\newcommand{\EEa}{\EE_a}
-\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
-\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
-\newcommand{\actions}{\mathcal{A}}
-\newcommand{\calA}{\mathcal A}
-\newcommand{\etc}{\textit{etc}}
-\newcommand{\ie}{\textit{i.e.}}
-\newcommand{\eg}{\textit{e.g.}}
-\newcommand{\calP}{\mathcal P}
-\newcommand{\x}{\boldsymbol{x}}
-\newcommand{\Ps}{\operatorname{P}}
-\newcommand{\R}{\mathbb R}
-\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
-\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}
-\newcommand{\indep}[1]{\mathcal{I}\left(#1\right)}
-\newcommand{\vb}[1]{\boldsymbol{#1}}
-\newcommand{\card}[1]{\left\vert{#1}\right\vert}
-\newcommand{\actionspace}{\mathcal{A}}
-
-\newcommand{\V}[1]{\mathbb V\left[#1\right]}
-\newcommand{\Esub}[2]{\mathbb E_{#1}\left[{#2}\right]}
-\newcommand{\defined}{\vcentcolon =}
-\newcommand{\rdefined}{=\vcentcolon}
-\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
-\newcommand{\Var}{\operatorname{Var}}
-\newcommand{\calF}{\mathcal F}
-\newcommand{\sr}[1]{\stackrel{#1}}
-\newcommand{\set}[1]{\left\{#1\right\}}
-\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
-
-\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
-\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
-\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
-\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
-\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
-\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
-\newcommand{\Ber}{\operatorname{Bernoulli}}
-\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
-\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
-\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
-\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
-\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}
-
-\renewcommand{\P}[1]{\operatorname{P}\left(#1\right)}
-\newcommand{\cf}[2]{{#1}^{#2}}
-\newcommand{\cfb}[2]{{#1}_t^{#2}} % used to represent reward
-\newcommand{\bernoulli}{\operatorname{Bernoulli}}
-\newcommand{\dirac}{\operatorname{Dirac}}
-\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
-\renewcommand{\vec}[1]{\boldsymbol{#1}}
-\newcommand{\regret}{\bar{R}_{T}} %psuedo-regret
-\newcommand{\expectedregret}{\mathbb{E}R_{T}} % expected regret
-\newcommand{\rawregret}{R_{T}} % plain regret
-\newcommand{\simpleregret}{R_T}
-\newcommand{\quotes}[1]{``#1''}
-
-\theoremstyle{plain} 
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{assumption}[theorem]{Assumption}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{example}[theorem]{Example}
-\let\temp\epsilon
-\let\epsilon\varepsilon
-
-\newcommand{\ci}{\mathrel{\perp\mspace{-10mu}\perp}}
-\newcommand{\nci}{\centernot{\ci}}
-
-\author{Finnian Lattimore}
-\title{Learning how to act: making good decisions with machine learning}
-
-\setcounter{secnumdepth}{3}
-
-\begin{document}
-
-%\def\ci{\perp\!\!\!\perp} % from Wikipedia
-
-\begin{titlepage}
-    \begin{center}
-        \vspace*{1cm}
-
-        \Huge
-        {Learning how to act: making good decisions with machine learning}
-
-        \vspace{2cm}
-        \Large
-        {Finnian Lattimore}
-
-        \vfill
-        \Large
-        A thesis submitted for the degree of Doctor of Philosophy
-
-         The Australian National University
-
-        \vspace{1cm}
-
-        September 2017
-
-    \end{center}
-\end{titlepage}
-
-\epigraph{In vain the grave, with retrospective Eye,
-Would from the apparent what conclude the why,
-Infer the Motive from the Deed, and show
-That what we chanced, was what we meant, to do.}{Alexander Pope}
-
 \clearpage
@@
 \clearpage
-
-\begin{abstract}
-
-This thesis is about machine learning and statistical approaches to decision making.
-How can we learn from data to anticipate the consequence of, and optimally select, interventions or actions?
-Problems such as deciding which medication to prescribe to patients, who should be released on bail, and how much to charge for insurance are ubiquitous, and have far reaching impacts on our lives.
-There are two fundamental approaches to learning how to act: reinforcement learning, in which an agent directly intervenes in a system and learns from the outcome, and observational causal inference, whereby we seek to infer the outcome of an intervention from observing the system.
-
-The goal of this thesis to connect and unify these key approaches.
-I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis.
-I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms.
-I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.
-
-\end{abstract}
-
-%\chapter*{Acknowledgements}
-
-\tableofcontents
-
-\chapter{Introduction}
-
-\section{Motivation}
-Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something.
-Will asking people to pay upfront at the doctors reduce long term health expenditure?
-If we developed a drug to suppress particular genes, could we cure multiple sclerosis and would delaying teen-aged pregnancies improve the outcome for their children.
-
-These are hard questions because they require more than identifying a pattern in data.
-Correlation is not causation.
-Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, despite the fact that the variables involved are susceptible to human intuition and understanding.
-
-We now live in a world of data.
-Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and where the US Institute of Health is considering if all infants should be genetically sequenced at birth.
-Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by-product of some other purpose.
-It does not come from the controlled world of a randomised experiment.
-
-The rise of big data sets and powerful computers has seen an explosion in the application of machine learning.
-From health care, to entertainment and self-driving cars, machine learning algorithms will transform many industries.
-It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory \citep{anderson2008end} and that we may be only a short step from \quotes{The Singularity}, where artificial intelligence exceeds our own and then grows exponentially.
-
-However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within narrow problem settings.
-Getting them to generalise to even slightly different problems or data sets remains very challenging.
-Deciding how we should act or what policies we should implement requires predictions about how a system will behave if we change it.
-The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on.
-As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live in, it is critical to ensure we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.
-
-\section{What is causality? }
-
-The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined.
-This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
-However, the magnitude, direction and even existence of an association depends which variables are controlled for (or included in a regression).
-Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.
-
-There are two ways in which an association detected in a data set may be non-causal.
-The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample.
-Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated.
-For example, based on data from the years 1999 to 2009, the age of Miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}.
-We would not expect this relationship to hold in a new sample of data.
-This form of spurious correlation also has serious repercussions.
-It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published.
-Consequently, the likelihood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}.
-This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}.
-This issue can be ameliorated by obtaining more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.
-
-However, a strong association, observed in multiple independent studies may still not be causal.
-The correlation can arise because both variables are consequences of some other, unmeasured factor.
-For example, the reading ability of children under twelve is strongly correlated with their height, because older children are taller and can read better.
-However height is not a cause of reading ability because interventions to increase a child's height, for example by giving them growth hormones, would not be expected to improve their reading.
-Similarly, extra lessons in reading will not make a child grow taller.
-This problem is fundamentally different to the issue of spurious correlations arising by chance in finite data sets.
-Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to separate causation from correlation.
-
-The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables.
-In this thesis, I take an interventionist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal.
-This viewpoint captures the types of questions that motivate this thesis.
-How can we change the way we do things to obtain better outcomes?
-
-Causality is often liked to explanation; understanding how and why things happen.
-I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture.
-This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction.
-A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change.
-A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed.
-More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts.
-By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
-
-\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{what if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs.
-The former aims to identify the effect of a known cause.
-The latter can be viewed as identifying causes of an effect.
-They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, that we describe in section \ref{sec:causal_models}.
-However, they state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined.
-The goal of identifying the causes of an effect can be formalised within the graphical causal model framework.
-Solving this problem is certainly much more challenging than identifying the effect of a specific intervention on a given outcome, since it requires us to test or infer the effect of interventions on many different variables.
-These practical difficulties may well be overwhelming, particularly in fields such as social science and economics where data sets are often relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "what if" questions are hard to resolve conclusively.
-However, this does not mean that the problem of identifying causes of effects is ill-posed in principle.
-It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see for example \citep{}.
-\todo{reference a review paper on causal discovery here}
-
-There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models.
-However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models.
-In the next section we emphasise the subtley that can exist in determining if a problem is causal by examining some typical examples.
-
-\section{What makes a problem causal?}
-
-Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them.
-Machine learning techniques are being applied a huge range of problems, in both industry and academia.
-The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to.
-Which, if any, of these problems require causal inference?
-
-\begin{itemize}
-\item Speech recognition (for systems like Siri or Google Assistant)
-\item Image classification
-\item Forecasting the weather
-\item Identifying spam emails
-\item Automated essay marking
-\item Predicting the risk of death in patients with pneumonia.
-\item Predicting who will re-offend on release from prison 
-\item Customer churn modelling
-\item Demand prediction for inventory control
-\item Playing Go 
-\end{itemize}
-
-The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important consideration.
-In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.
-
-Consider speech recognition.
-You say something, which causes sound waves, which are converted to a digital signal that Siri maps to words.
-Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution).
-A similar argument could be made for many applications of machine translation and image classification.
-
-In image classification we do not particularly care about building a strong model for exactly how the thing that was photographed translates to an array of pixels, provided we can be fairly confident that the process will not change.
-If we develop a discriminative model that is highly accurate at classifying cats from dogs, we do not need to understand its internal workings (assuming we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).
-
-What about forecasting the weather?
-If you are using a short term forecast to decide whether to pack an umbrella, causality can be ignored.
-Your decision will not affect if it actually rains.
-However, longer term climate forecasts might (theoretically) lead to action on emissions which would then change the weather system.
-For this we need a (causal) model that allows us to predict the outcome under various different interventions.
-
-Identifying spam and automated essay marking both involve processing text to determine an underlying (complex) attribute such as its topic or quality.
-In both cases, there is inherent competition between the algorithm and the people generating the text.
-As a result, decisions made by the algorithm are likely to change the relationship between the features it relies on and the true label.
-Spammers and students will modify their writing in order to optimise their results.
-A standard supervised learning approach can only work if the resulting change in the mapping from features to label is sufficiently gradual.
-There are two key ways of ensuring this.
-The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm.
-The second is to use a model in which the features are related to the outcome in such a way that they cannot be manipulated independently.
-
-This example also highlights a connection between causal models and transparency in machine learning.
-If we are using a non-causal model to make decisions affecting people, there will be a trade-off between the performance and transparency of the model; not because the requirement for transparency restricts us to simple models, but because revealing how the model works allows people to change their behaviour to game it.
-
-What about predicting the risk of death in patients with pneumonia?
-Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics.
-If we assume that in hospital treatment is more effective for serious cases, this appears to be straightforward prediction.
-It is not.
-Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admission decisions.
-\citet{Caruana2015} found exactly this effect in a real data set.
-The model learned that people suffering asthma were \emph{less} likely to die from pneumonia.
-This was because doctors treated such patients very aggressively, thus actually lowering their risk.
-The issue is not with model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols.
-However, using it to decide how to \emph{change} these protocols could kill.
-The actual question of interest in this case is what happens to patients with characteristics $X$ they are assigned treatment according to decision rule (or policy) $\pi(X)$.
-
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
-Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.
-
-Demand prediction seems like a relatively straightforward prediction problem.
-These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product.
-It seems unlikely that using the model to ensure stock is available will itself change demand.
-However, depending on the way demand is measured, there is a potential data censoring issue.
-If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero.
-Changing availability does then change demand.
-
-Playing Go (and other games) is a case with some subtleties.
-At every turn, the AI agent has a number of actions available.
-The board state following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem.
-There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}.
-Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$.
-This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players.
-Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a combination of these approaches \citep{silver2016mastering}.
-The supervised learning was enhanced by having the agent play (variants) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.
-
-The problem of playing go is causal from the interventionalist perspective.
-The agent wishes to learn the probability of a win given an action they take.
-However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach.
-The actions the agent has to explore are the sames ones as human players explored to generate the training data, and both have the same objective - to win the game.
-In addition, the state of the board encapsulates all the information relevant to selecting a move.
-These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.
-
-Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predictive) inference.
-A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes.
-The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people.
-Think about problems like credit scoring and parole decisions.
-There are strong ethical grounds for demanding transparency, but if the goals of society and the individuals are not perfectly aligned and there is any possibility that people can manipulate features independently of the outcome, there will be a conflict between model accuracy and transparency.
-It is rare to build a model without any intent to make some kind of decision on the basis of its results.
-Thus, I argue we should assume a causal model is required until we can justify otherwise.
-
-\section{Observe or intervene: two very different approaches to causal problems}
-
-As we have shown, problems involving causality are ubiquitous in many fields.
-As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning.
-Although the focus and terminology can differ substantially between fields these techniques, all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system.
-Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.
-
-In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them.
-We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal.
-A particular strand of research within reinforcement learning are multi-armed bandit problems.
-They describe settings in which there are a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action.
-They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.
-
-The approach of learning the outcome of an action by taking it has plays a key role in advancing our knowledge of the world.
-However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen.
-Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.
-Observational causal inference can be viewed as a form of transfer learning.
-The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system.
-This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another.
-The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.
-
-Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials.
-Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem randomisation is not always feasible, affordable or ethical.
-The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.
-
-\section{This thesis and its contributions}
-
-I view causal problems as one of the greatest arising challenges for machine learning.
-They incorporate a large set of problems, of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence.
-For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry we need to understand how to leverage our improved approaches to prediction to tackle causal problems.
-
-\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints.
-My major contribution is a framework which unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
-This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
-I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
-My other contributions are primarily connective rather than technical.
-I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias.
-I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data and covariate shift.
-I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.
-
-\paragraph{Thesis overview} This thesis is divided into three key chapters, learning from observational data, learning from interventional data and unifying the approaches.
-Chapter \ref{chap:causality} covers learning to act from observational data.
-In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems.
-Section \ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \ref{sec:defining-causal-effects} and \ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation.
-Section \ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.
-
-Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems.
-In \S\ref{sec:multi-armed-bandits} I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}).
-I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}.
-In \S\ref{sec:off-policy-evaluation} I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of of causal effect estimation from observational data.
-Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.
-
-In chapter \ref{chap:causal-bandit} I introduce causal bandit problems: a framework which unifies causal graphical models with multi-armed bandit problems.
-Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model.
-I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.
-
-In \S\ref{sec:causal-bandit-post-action-feedback} I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected.
-I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms.
-In \S\ref{sec:simple-regret} I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}.
-I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning.
-In \S\ref{sec:simple-regret-general} I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}.
-Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems.
-Finally in \S\ref{sec:causal_bandit_discussion} I discuss extensions and potential future work.
-
-\chapter{Learning from observational data}
-\label{chap:causality}
-
-The goal of causal inference is to learn the effect of taking an action.
-We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world.
-We are generating and storing data on almost every aspect of our lives at an unprecedented rate.
-As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase.
-However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question.
-An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited.
-This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it.
-This is the problem of observational causal inference.
-
-\section{Causal models}
-\label{sec:causal_models}
-
-Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
-As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention.
-Mapping properties from one system to another requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate a system will respond to them.
-Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
-
-Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings.
-They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another.
-Counterfactual assertions can be expressed very naturally in human languages and are quite prevalent in  every day conversations;\quotes{if I had worked harder I would have got better grades}, \quotes{she would have been much sicker if she hadn't taken antibiotics}.
-Structural equation models have been developed and applied primarily within economics and related disciplines and can be seen as an attempt to capture key aspects of the behaviour of people with mathematics.
-Questions around designing policies or interventions play a central role in economics.
-Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised) framework and associated set of methods for estimating causal effects.
-Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks.
-They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of conditions under causal effects can be inferred from observational data.
-
-An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}.
-However, the literature on causal inference techniques remains spit between the different frameworks.
-Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scores, are described using the counterfactual framework.
-Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions are often based around structural equation models.
-This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.
-
-In the next sections we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ.
-We hope that by describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them.
-However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers.
-In order to demonstrate the notation and formalisms each framework provides we will use them to describe the following simple examples.
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:ranomized_experiment}
-Suppose a pharmaceutical company wants to assess the effectiveness of a new drug on recovery from a given illness.
-This is typically tested by taking a large group of representative patients and randomly assigning half of them to a treatment group (who receive the drug) and the other half to a control group (who receive a placebo).
-The goal is to determine the clinical impacts of the drug by comparing the differences between the outcomes for the two groups (in this case, simplified to only two outcomes - recovery or non-recovery).
-We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.
-\end{example}
-
-\vspace*{.3cm}
-\begin{example}
-\label{exm:adjusting}
-Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds.
-We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
-We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
-\end{example}
-
-\subsection{Causal Bayesian networks}
-
-Causal Bayesian networks are an extension of Bayesian networks.
-A Bayesian network is a graphical way of representing how a distribution factorises.
-Any joint probability distribution can be factorised into a product of conditional probabilities.
-There are multiple valid factorisations, corresponding to permutations of variable ordering.
-
-\eqn{
-\label{eqn:cbn:joint_dist}
-P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
-}
-
-We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}.
-If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network.
-There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below left=of 1](2){$X_{2}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\caption{A general Bayesian network for the joint distribution over three variables.
-This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
-\label{fig:bayesnet}
-\end{figure}
-
-\begin{figure}
-\centering
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.15\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, below=of 1](2){$X_{2}$};
-\node[main node, below=of 2](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (3) edge (2)
-    (2) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, below left=of 1](2){$X_{1}$};
-\node[main node, below right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$X_{2}$};
-\node[main node, above left=of 1](2){$X_{1}$};
-\node[main node, above right=of 1](3){$X_{3}$};
-\path[every node/.style={font=\sffamily\small}]
-    (2) edge (1)
-    (3) edge (1)
-    (2) edge (3);
-\end{tikzpicture}
-\end{subfigure}
-\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$.
-Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical structure implies exactly the same set of independencies exhibited by the distribution.
-Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
-\label{fig:bayesnet2}
-\end{figure}
-
-The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph.
-There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$, (the converse is not true: $G$ can have extra edges).
-If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;
-
-\eqn{
-\P{X_{1},...,X_{N}} = \prod_{i = 1...N}\P{X_{i}|parents_{X_{i}}}
-}
-
-A causal Bayesian network is a Bayesian network in which a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$.
-This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change.
-We need some notation to describe interventions and represent distributions over variables in the network after an intervention.
-In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.
-
-\vspace{0.5cm}
-\begin{definition}{The do-notation}
-\begin{itemize}
-\item $do(X=x)$ denotes an intervention that sets the random variable(s) $X$ to $x$.
-\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$.
-This notation is somewhat overloaded.
-It may be used to represent a probability distribution/mass function or a family of distribution functions depending on whether the variables are discrete or continuous and whether or not we are treating them as fixed.
-For example it could represent
-\begin{itemize}
-\item the probability $\P{Y=1|do(X=x)}$ as a function of $x$,
-\item the probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$,
-\item the probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$,
-\item a family of density/mass function for $Y$ parameterised by $x$.
-\end{itemize}
-Where the distinction is important and not clear from context we will use one of the more specific forms above. 
-\end{itemize}
-\end{definition}
-
-\vspace{0.5cm}
-\begin{theorem}[Truncated product formula \citep{Pearl2000}]
-\label{thm:truncated_prodcut}
-If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by dropping the terms for each of the variables in $Z$ from the factorisation given by the network.
-Let $\parents{X_i}$ denote the parents of the variable $X_i$ in $G$.
-
-\begin{equation}
-\label{eq:truncatedproduct}
-\P{X_1...X_N|do(Z=z)} = \ind{Z = z}
-  \prod_{X_i \notin Z}\P{X_{i}|\parents{X_i}}
-\end{equation}
-\end{theorem}
-
-Theorem \ref{thm:truncated_prodcut} does not hold for standard Bayesian networks because there are multiple valid networks for the same distribution.
-The truncated product formula will give different results depending on the selected network.
-The result is possible with causal Bayesian networks because it follows directly from the assumption that the direction of the link indicates causality.
-In fact, from the interventionist viewpoint of causality, the truncated product formula defines what it means for a link to be causal.
-
-Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$?
-The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient.
-Since $X$ is assigned via deliberate randomisation we know that it is not affected by any latent variables.
-The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}.
-This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$.
-In this example, the interventional distribution is equivalent to the observational one.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X \text{ (Treatment)}$};
-\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
-\path[]
-	(2) edge (3);
-\end{tikzpicture}
-\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
-\label{fig:causal_network_example}
-\end{figure}
-
-In example \ref{exm:adjusting} we are interested in $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool.
-We could compare it to outlawing preschool $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$.
-It seems reasonable to assume that preschool attendance affects the likelihood of high school graduation \footnote{The effect does not have to be homogeneous, it may depend non-linearly on characteristics of the child, family and school.} and that parental socio-economic status would affect \emph{both} the likelihood of preschool attendance and high school graduation.
-If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal Bayesian network in figure \ref{fig:causal_adjust}.
-In this case, the interventional distribution is not equivalent to the observational one.
-If parents with high socio-economic status are more likely to send their children to preschool and these children are more likely to graduate high school regardless, comparing the graduation rates of those who attended preschool with those who did not will overstate the benefit of preschool.
-To obtain the interventional distribution we have to estimate the impact of preschool on high school graduation for each socio-economic level separately and then weight the results by the proportion of the population in that group,
-
-\eqn{
-\label{eqn:backdoor_example}
-\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}
-}
-
-We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph.
-There is a very powerful and general set of rules that specifies how we can transform observational distributions into interventional ones for a given graph structure.
-These rules are referred to as the Do-calculus \citep{Pearl2000}.
-We discuss them further in section \ref{sec:do_calculus}.
-
-\begin{figure}
-\center
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]
-
- %nodes
-\node[node](1){$Z \text{ (SES)}$};
-\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
-\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3); 
-\end{tikzpicture}
-\caption{Causal Bayesian network for example \ref{exm:adjusting}}
-\label{fig:causal_adjust} 
-\end{figure}
-
-A causal Bayesian network represents much more information than a Bayesian network with identical structure.
-A causal network encodes all possible interventions that could be specified with the do-notation.
-For example, if the network in figure \ref{fig:causal_adjust} were an ordinary Bayesian network and all the variables were binary, the associated distribution could be described by 7 parameters.
-The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional 30 parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
-After each two variable intervention, we have a distribution over one variables which requires one parameter.
-This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
-
-Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes.
-The global Markov property and d-separation also hold in causal networks.
-D-separation, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
-We include a brief review of D-separation in section \ref{sec:dsep}.
-
-\subsubsection{Limitations of causal Bayesian networks}
-A number of criticisms have been levelled at this approach to modelling causality.
-One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}.
-However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions.
-For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter.
-But, what if some people failed to take the prescribed treatment?
-We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}.
-Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Prescribed Treatment};
-\node[node, below right=of 1](2){Actual Treatment};
-\node[node, right=of 2](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(1) edge (2)
-	(2) edge (3) 
-	(4) edge (3) edge (2);
-\end{tikzpicture}
-\caption{Randomised treatment with imperfect compliance}
-\label{fig:randomized_imperfect_compliance}
-\end{figure}
-
-Another key issue with causal Bayesian networks is that they cannot handle cyclic dependencies between variables.
-Such feedback loops are common in real-life systems, for example the relationship between supply and demand in economics or predator and prey in ecology.
-We might regard the underlying causal mechanisms in these examples to be acyclic; the number of predators at one time influences the number of prey in the next period and so on.
-However, if our measurements of these variables must be aggregated over time frames that are longer than the scale at which these interactions occur the result is a cyclical dependency.
-Even were we able to measure on shorter timescales, we might then not have sufficient data on each variable for inference.
-Such problems have mostly been studied within the dynamical systems literature, typically focusing on understanding the stationary or equilibrium state of the system and making very specific assumptions about functional form in order to make problems tractable.
-\citet{Poole2013} compare the equilibrium approach to reasoning about cyclic problems with structural equation models, which we discuss in section \ref{sec:SEM} and which can be seen as Bayesian causal networks with additional functional assumptions.
-
-\subsection{Counterfactuals}
-\label{sec:counterfactuals}
-
-The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals.
-Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
-Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
-
-In example \ref{exm:ranomized_experiment}, the causal effect of the drug relative to placebo for person $i$ is the difference between what would have happened if they were given the drug, denoted $\cf{y_{i}}{1}$ versus what would have happened if they got the placebo, $\cf{y_{i}}{0}$.
-The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated.
-The problem can be resolved if, instead of people, there are units that can be assumed to be identical or that will revert exactly to their initial state some time after treatment.
-This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.
-
-Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo.
-Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated.
-The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated.
-Similarly $Y^{0}$ represents the potential outcome for the placebo.
-The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
-We can estimate (from an experimental or observational study):
-\begin{itemize}
-\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
-\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
-\end{itemize}
-
-Now, for those who took the treatment, the outcome \emph{had} they taken the treatment $\cf{Y}{1}$ is the same as the observed outcome.
-For those who did not take the treatment, the observed outcome is the same as the outcome \emph{had} they not taken the treatment.
-Equivalently stated:
-
-\eq{
-\P{Y^{0}|X=0}&= \P{Y|X=0}\\
-\P{Y^{1}|X=1}&=\P{Y|X=1}
-}
-
-If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:
-
-\eq{
-\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
-\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
-}
-
-This implies the counterfactual distributions are equivalent to the corresponding conditional distributions and, for a binary outcome $Y$, the causal effect is,
-
-\eq{
-\P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
-}
-
-The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}.
-They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated.
-This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment.
-In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e.
-$X \nci \cf{Y}{0}$.
-Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e.
-$X \nci \cf{Y}{1}$.
-A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
-
-\vspace*{.3cm}
-\begin{theorem}[Ignorability \citep{Rosenbaum1983, Pearl2000}] If $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$,
-
-\eqn{
-\label{eqn:counterfactual1}
-\P{\cf{Y}{1}} &= \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}  \\
-\label{eqn:counterfactual2}
-\P{\cf{Y}{0}} &= \sum_{z \in Z}\P{Y|X=0,Z}\P{Z}
-}
-\end{theorem}
-
-Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
-Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
-
-Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}.
-This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
-As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
-The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
-SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.
-
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-In particular, statements involving joint distributions over counterfactual variables may not be able to be validated empirically \citet{Dawid2000}.
-One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation.
-Rather than talking about the distribution of $Y$ given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $x$.
-This is certainly convenient, if rather imprecise.
-However, the ease with which we can make statements with counterfactuals that cannot be tested with empirical data warrants careful attention.
-It is important to be clear what assumptions are being made and whether or not they could be validated (at least in theory).
-
-\subsection{Structural Equation models}
-\label{sec:SEM}
-
-Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input.
-The mechanism (but not the output) is assumed to be independent of what is fed into it.
-Uncertainties are not inherent but arise from unmeasured variables.
-Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}.
-More recently, they have been formalised, generalised to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \citep{Pearl2000}.
-
-Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables.
-The noise terms are required to be mutually independent.
-If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent variable.
-Structural equation models can be represented visually as a network.
-Each variable is a node and arrows are drawn from causes to their effects.
-Figure \ref{fig:sem_randomized_treatment} illustrates the SEM for example \ref{exm:ranomized_experiment}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(\epsilon_x)$};
-\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:ranomized_experiment}}
-\label{fig:sem_randomized_treatment}
-\end{figure}
-
-This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$.
-This is justifiable on the grounds that $X$ is random.
-The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).
-Note that the causal graph in figure \ref{fig:sem_randomized_treatment} is identical to that the Bayesian network for the same problem, figure \ref{fig:causal_network_example}.
-The latent variables $\epsilon_x$ and $\epsilon_y$ are not explicitly drawn in figure \ref{fig:causal_network_example} as they are captured by the probabilistic nature of the nodes in a Bayesian network.
-
-Taking the \emph{action} $X=1$ corresponds to replacing the equation $X=f_x(\epsilon_x)$ with $X=1$.
-The function $f_y$ and distribution over $\epsilon_y$ does not change.
-This results in the interventional distribution, \footnote{We have assumed the variables are discrete only for notational convenience}
-
-\eqn {
-\P{Y=y|do(X=1)} = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y}
-}
-
-The observational distribution of $Y$ given $X$ is,
-
-\eqn{
-\P{Y=y|X=1} &= \sum_{\epsilon_x}\sum_{\epsilon_y}\P{\epsilon_x|X=1}\P{\epsilon_y|\epsilon_x}\ind{f_y(1,\epsilon_y)=y} \\
-& = \sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,\epsilon_y)=y} \text{, as } \epsilon_x \ci \epsilon_y
-}
-
-The interventional distribution is the same as the observational one.
-The same argument applies to the intervention $do(X=0)$ and so the causal effect is simply the difference in observed outcomes as found via the causal Bayesian network and counterfactual approaches.
-
-The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
-Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
-
-\eqn{
-\P{Y=y|do(X=1)} &= \sum_{z}\sum_{\epsilon_y}\P{z}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y} \\
-\label{eqn:sem:adjusting}
-&=\sum_{z}\P{z}\underbrace{\sum_{\epsilon_y}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}}_{\P{Y=y|X=1,Z=z}}
-}
-Equation \ref{eqn:sem:adjusting} corresponds to equations \ref{eqn:backdoor_example} and \ref{eqn:counterfactual1}. It is not equivalent to the observational distribution given by:
-
-\eqn{
-\P{Y=y|X=1} = \sum_{z}\sum_{\epsilon_y}\P{z|X=1}\P{\epsilon_y}\ind{f_y(1,z,\epsilon_y)=y}
-}
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](2){$X = f_x(Z,\epsilon_x)$};
-\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
-\node[lt, above=of 6](7){$\epsilon_z$};
-\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
-\node[lt, above=of 3](4){$\epsilon_y$};
-\node[lt, above=of 2](5){$\epsilon_x$};
-\path[]
-	(2) edge (3)
-	(4) edge (3)
-	(5) edge (2)
-	(7) edge (6)
-	(6) edge (2) edge (3);
-\end{tikzpicture}
-\caption{SEM for example \ref{exm:adjusting}}
-\label{fig:sem:preschool}
-\end{figure}
-
-Structural equation models are generally applied with strong constraints on the functional form of the relationship between the variables and noise is typically assumed to be additive, $X_i = f_i(\cdot)+\epsilon_i$.
-A structural equation model with $N$ variables resembles a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation.
-However a SEM is, by definition, more than a set of simultaneous equations.
-By declaring it to be structural we are saying that it represents assumptions about the relationships between variables.
-When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.
-The similarity between the notation used to describe and analyse structural equation models and simultaneous equations, combined with a reluctance to make explicit statements about causality, has led to some confusion in the interpretation of SEMs \citep{heckman2015causal,Pearl2000}.
-
-\subsection{Comparing and unifying the models}
-\label{sec:unifying_causal_models}
-
-Clarify Cheng's Questions - is there a mechanical method to go from one framework to another, does it matter if variables are discrete or continuous. \todo{}
-
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
-If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
-In other words, the network is a causal Bayesian network.
-All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.
-Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant.
-This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}).
-Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks.
-We can also connect this to counterfactuals via:
-
-\begin{equation}
-\begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
-\end{aligned}
-\end{equation}
-
-The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals.
-When discussing the counterfactual model, we actually made the slightly weaker assumption:
-
-\begin{equation}
-\label{eq:weakignore}
-X \ci Y^{0} \text{ and } X \ci Y^{1}
-\end{equation}
-
-It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}.
-The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks.
-For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate.
-Then use propensity scores \citep{Rosenbaum1983} to estimate the effect.
-If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.
-The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation.
-These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
-
-In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks.
-The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations.
-The Neyman-Rubin framework builds on the understanding of randomised experiment and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
-This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
-In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field.
-Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
-In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
-
-\subsection{What does a causal model give us? Resolving Simpson's paradox}
-
-We will now demonstrate our new notation and frameworks for causal inference to resolve a fascinating paradox, noted by \citet{yule1903notes}, demonstrated in real data by \citet{Cohen1934} and popularised by \citet{simpson1951interpretation}.
-The following example is adapted from \citet{Pearl2000}.
-Suppose a doctor has two treatments, A and B, which she offers to patients to prevent heart disease.
-She keeps track of which medication her patients choose and whether or not the treatment is successful.
-She obtains the results in table \ref{tab:simpson1}.
-
-\begin{table}[h]
-\caption{Treatment results}
-\label{tab:simpson1}
-\center
-\begin{tabular}{c c c c c}
-Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9} 
-A &  87 & 13 & 100 & 87\%\\
-\rowcolor{white} B & 75 & 25 & 100 & 75\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Drug A appears to perform better.
-However, having read the latest literature on how medications affect men and women differently, she decides to break down her results by gender to see how well the drugs perform for each group and obtains the data in table \ref{tab:simpson2}.
-
-\begin{table}[h]
-\caption{Treatment results by gender}
-\label{tab:simpson2}
-\center
-\begin{tabular}{c c c c c c}
-Gender & Treatment & Success & Fail & Total & Success Rate\\
-\hline
-\rowcolor[gray]{0.9}
-M & A & 12 & 8 & 20 & 60\%\\
-M & B & 56 & 24 & 80 & 70\%\\
-\rowcolor[gray]{0.9}
-F & A & 75 & 5 & 80 & 94\%\\
-F & B & 19 & 1 & 20 & 95\%\\
-\hline
-\end{tabular}
-\end{table}
-
-Once the data is broken down by gender, drug B looks better for both men \emph{and} women.
-Suppose the doctor must choose only one drug to prescribe to all her patients in future (perhaps she must recommend which to subsidise under a national health scheme).
-Should she choose A or B?
-The ambiguity in this question lies at the heart of Simpson's paradox.
-How does causal modelling resolve the paradox?
-The key is that the doctor is trying to choose between \emph{interventions}.
-She wants to know what the success rate will be if she changes her practice to give all the patients one drug, rather than allowing them to choose as currently occurs.
-
-Let's represent the treatment by the variable $T$, the gender of the patient by $Z$ and whether or not the treatment was successful by $Y$.
-The doctor cares about $\P{Y|do(T)}$, not the standard conditional distributions $\P{Y|T}$.
-Unfortunately, the data in tables \ref{tab:simpson1} and \ref{tab:simpson2} is insufficient to enable estimation of the interventional distribution $\P{Y|do(T)}$ or determine if $do(T=A)$ is better or worse than $do(T=B)$.
-Some assumptions about the causal relationships between the variables are required.
-In this example, it seems reasonable to conclude that gender may affect the treatment chosen and the outcome.
-Assuming there are no other such confounding variables (for example income) then we obtain the causal network in figure \ref{fig:simpsons_paradox}.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Gender};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (2) edge (3) 
-	(2) edge (3);
-\end{tikzpicture}
-\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
-\label{fig:simpsons_paradox}
-\end{figure}
-
-With this model, women are more likely to choose drug A and are also more likely to recover than men regardless of the treatment they receive.
-Knowing a patient took drug A indicates they are more likely to be female.
-When we compare the group of people who took A against those who took B, the effect of the higher proportion of females in the first group conceals the greater benefit of drug B leading to an apparent reversal in effectiveness.
-However, when the doctor intervenes to set the treatment each person receives there will no longer be a link from gender to treatment.
-So in this case she should choose which drug to prescribe from the gender specific table (and weight by the proportion of the population that belongs to each gender).
-Drug B is the better choice.
-
-\eqn{
-\P{Y|do(T)} = \P{Y|T,female}\P{female}+ \P{Y|T,male}\P{male}
-}
-
-Is the solution to Simpson's paradox to always to break down the data by as many variables as possible?
-No.
-Suppose we have the identical data as in \ref{tab:simpson1} and \ref{tab:simpson2} but replace the column name 'gender' with 'blood pressure', 'M' with 'high' and 'F' with 'normal'.
-This is a drug designed to prevent heart disease.
-One pathway to doing so might well be to lower blood pressure.
-Figure \ref{fig:simpsons_paradox2} shows a plausible causal graph for this setting.
-It differs from the graph in figure \ref{fig:simpsons_paradox} only in the direction of a single link.
-Now, however, table \ref{tab:simpson2} tells us that people who took treatment $A$ had better blood pressure control and better overall outcomes.
-In this setting $\P{Y|do(T)} = \P{Y|T}$ and drug A is the better choice.
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
-\node[node](1){Z = Blood Pressure};
-\node[node, below left=of 1](2){T = Treatment};
-\node[node, below right=of 1](3){Y = Outcome};
-\path[]
-	(1) edge (3) 
-	(2) edge (3)
-	(2) edge (1);
-\end{tikzpicture}
-\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
-\label{fig:simpsons_paradox2}
-\end{figure}
-
-Note that we have not changed the data itself, only the description of the variables that it is associated with.
-This illustrates that the resolution to Simpson's paradox lies fundamentally not in the data, but in the assumptions we are willing to make.
-From a purely statistical viewpoint there is no paradox.
-The reversal just stems from the mathematical property of ratios expressed in equation \ref{eqn:simpsons_paradox} and represented graphically in figure \ref{fig:simpsons_vector}.
-The paradox only arises when we attempt to use the data to select an intervention and is resolved when we apply a causal approach to do so.
-
-\eqn{
-\label{eqn:simpsons_paradox}
-\exists \set{N_1,...N_4, N_1'...N_4'} \in \mathbb{N} : \; 
-\textcolor{green}{\frac{N_1}{N_2}< \frac{N_1'}{N_2'}}, \;\; 
-\textcolor{blue}{ \frac{N_3}{N_4}<\frac{N_3'}{N_4'}}
- \text{ and } \frac{N_1+N_3}{N_2+N_4} > \frac{N_1'+N_3'}{N_2'+N_4'}
-}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.4]{figures/simpson_vector}
-\caption{Simpson's reversal visualised.
-The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors.
-However, when we sum them, the ratio is steeper for the un-primed variables.}
-\label{fig:simpsons_vector}
-\end{figure}
-
-There are many other plausible causal graphs for both scenarios above.
-Perhaps income affects drug choice as well as gender, or gender might affect treatment choice and blood pressure control given treatment, etc.
-Causal modelling provides a powerful tool to specify such assumptions and to determine how to estimate causal effects for a given model as we discuss in the next section.
-
-\section{Answering Causal Questions} 
-\label{sec:causal effect estimation}
-We can roughly categorise the problems studied within causal inference from observational data into two groups, causal effect estimation and causal discovery.
-In causal effect estimation we assume (at least implicitly) that key aspects of the causal graph are known.
-The goal is then to estimate the effect of an intervention or range of interventions in the system.
-Causal effect estimation is implicit in countless studies in economics, social science and epidemiology of everything from the effect of education on earnings \citep{card1999causal}, diet on cancer \citep{bingham2004diet} and breastfeeding on intelligence \citep{jain2002good} to the effect of pet ownership on survival after a heart attack \citep{friedmann1995pet}.
-Almost every time someone runs a regression model the key quantity of interest is a causal effect.
-Given how it underlies so much of our scientific progress, there is a enormous potential in properly understanding when we can draw causal conclusions, exactly what assumptions are required to do so and how we can best leverage those assumptions to infer as much information as we can from our data.
-
-Causal discovery aims to leverage much broader assumptions to learn the structure of causal graphs from data.
-This is critical in fields where we are generating a lot of data but have limited theoretical knowledge from which to draw on to determine how variables are related to one another.
-Causal discovery algorithms are being applied in bioinformatics \citep{Bay2002,Sachs2005,Ram2006,Alekseyenko2011,Statnikov2012,Gao2015,Sokolova2015,Taruttis2015}, medical imaging \citep{Ramsey2010} and climate science \citep{Uphoff2013}.
-An effective and generalisable approach for causal discovery would be a major step towards the automation of the scientific endeavour.
-In this thesis I have focused on problems where the structure of the causal graph is known.
-Extending my work to problems where the causal structure is unknown, leveraging the work on causal discovery, is a rich and fascinating line of potential future work, which I discuss very briefly in section \ref{sec:causal_bandit_discussion}.
-
-\subsection{Mapping from observational to interventional distributions}
-
-A central component of estimating causal effects from observational data is determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
-We did this on an ad-hoc basis to resolve the examples discussed in section \ref{sec:causal_models}.
-In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:causal_effect_under_ignorability}, discuss the key issues involved in estimating such expressions from finite sample data.
-We assume the basic structure of the graph is known.
-That is, we assume that we can draw a network containing (at a minimum):
-
-\begin{itemize}
-\item the target/outcome variable we care about,
-\item the focus/treatment variables on which we are considering interventions,
-\item any variables which act to confound two or more of the other variables we have included, and
-\item any links between variables we have included. 
-\end{itemize}
-
-Some of these variables may be latent in that the available data does not record their value, however their position in the network is assumed to be known.
-For example, consider estimating the impact of schooling on wages.
-Some measure of inherent ability could influence both the number of years of schooling people choose to pursue and the wages they receive.
-Even if we have no data to directly assess people's inherent ability we must include it in the graph because it influences two of the variables we are modelling.
-
-How can the structure of the causal graph be leveraged to compute interventional distributions from observational ones?
-Given the graph corresponding to the observational distribution, the graph after any intervention can be obtained by removing any links into variables directly set by the intervention.
-The joint interventional distribution is the product of the factors associated with the interventional graph, as given by the truncated product formula \ref{eq:truncatedproduct}.
-If there are no latent variables the interventional distribution of interest can be obtained by marginalising over the joint (interventional) distribution.
-However, if there are latent variables the joint interventional distribution will contain terms that cannot be estimated from the observed data.
-
-The key to estimating causal effects in the presence of latent variables lies in combining the assumption of how an intervention changes the graph, encoded by the truncated product formula, with information the graph structure provides about conditional independencies between variables.
-By leveraging conditional independencies we can effectively localise the effect of an intervention to a specific part of a larger graph.
-This gives rise to the do-calculus \citep{Pearl2000}.
-The do calculus consists of three rules.
-They are derived from the causal information encoded in a causal network and the properties of d-separation and do not require any additional assumptions other than that of specifying the causal network.
-
-\subsubsection{Independence in Bayesian networks: D-separation}
-\label{sec:dsep}
-
-Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks.
-Therefor, in this section, we briefly review the key properties of Bayesian networks.
-A more thorough introduction (including proofs) can be found in \citep{Koller2009}.
-Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.
-
-\vspace{.3cm}
-\begin{theorem}
-\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
-\end{theorem}
-
-The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$.
-D-separation is an algorithm that extends the local Markov property to find these additional independencies.
-It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.
-
-The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$.
-From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network.
-Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$.
-In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise.
-In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.
-
-The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure.
-The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}.
-As an example, consider a scholarship available to female or disadvantaged students.
-Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship.
-There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent.
-However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y| Z$}
-\label{fig:dseptripple1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$X$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Y$};
-\node[main node, below =of 1](2){$Z$};
-\node[main node, below =of 2](3){$X$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    (2) edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|Z$}
-\label{fig:dseptripple3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, below left =of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-        edge node {} (3);	
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.2\textwidth}
-\centering
-\caption{$X \ci Y|\{\}$}
-\label{fig:dseptripple4}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
-  thick,main node/.style={observed}]
-\node[main node](1){$Z$};
-\node[main node, above left =of 1](2){$X$};
-\node[main node, above right =of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge node {} (1)
-    (3) edge node {} (1);	
-\end{tikzpicture}
-\end{subfigure}
-\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
-\label{fig:dseptriples}
-\end{figure}
-
-\vspace{.3cm}
-\begin{definition}[unblocked path]
-\label{def:unblocked}
-A path from $X$ to $Y$ is a sequence of edges linking adjacent nodes starting at $X$ and finishing at $Y$,  $(X,V_{1},V_{2}...V_{k},Y)$.
-It is unblocked if every triple, $X-V_{1}-V_{2},V_{1}-V_{2}-V{3},...,V_{k-1}-V_{k}-Y$ in the path is unblocked (each triple will belong to one of the cases in figure \ref{fig:dseptriples})
-\end{definition}
-
-\vspace{.3cm}
-\begin{definition}[d-separation]
-\label{def:dsep}
-The variables $\boldsymbol{X}$ are d-separated from $\boldsymbol{Y}$ given $\boldsymbol{Z}$ in the network $G$ if, there are no unblocked paths from any $X \in \boldsymbol{X}$ to any $Y \in \boldsymbol{Y}$ after conditioning on $\boldsymbol{Z}$.
-\end{definition}
-
-\vspace{.3cm}
-\begin{theorem}[d-separation and conditional independence]
-\label{thrm:dsep_independence}
-If a set of variables $\boldsymbol{Z}$ d-separates $\boldsymbol{X}$ and $\boldsymbol{Y}$ in a Bayesian network $G$ then ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in all distributions $P$ compatible with $G$.
-Conversely, if $\boldsymbol{X}$ and $\boldsymbol{Y}$ are d-connected (not d-separated) given $\boldsymbol{Z}$ then it is possible to construct a distribution $P'$ that factorises over $G$ in which they are dependent.
-\end{theorem}
-
-Theorem \ref{thrm:dsep_independence} says that independencies implied by d-separation on a graph hold in every distribution that can be factored over that graph and that if ${(\boldsymbol{X} \ci \boldsymbol{Y}|\boldsymbol{Z})}$ in \emph{all} distributions that can be factored over $G$ then they are d-separated in $G$.
-If we denote the independencies implied by d-separation in a graph by $\indep{G}$ and the set of independencies in a distribution by $\indep{P}$ then $\indep{G} \subseteq \indep{P}$.
-
-If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$.
-However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation.
-A particular case in which this occurs is when there are deterministic relationships between variables.
-If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent.
-This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables.
-This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connecteness to assert lack of independence.
-D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.
-
-\subsubsection{The Do Calculus} 
-\label{sec:do_calculus}
-
-The do-calculus is a set of three rules \citep{Pearl1995} that can be applied to simplify the expression for an interventional distribution.
-If by repeated application of the do-calculus, along with standard probability transformations, we can obtain an expression containing only observational quantities then we can use it to estimate the interventional distribution from observational data.
-Let $\boldsymbol{X}$,$\boldsymbol{Y}$,$\boldsymbol{Z}$ and $\boldsymbol{W}$ be disjoint sets of variables in a causal graph $G$.
-We denote the graph $G$ after the an intervention $do(\boldsymbol{X})$, which has the effect of removing all edges into variables in $\boldsymbol{X}$, as $G_{\overline{\boldsymbol{X}}}$
-
-\paragraph{Rule 1: (adding or removing evidence)}
-
-Rule 1 allows us to remove (or insert) observational evidence from the right hand side of a conditional interventional distribution.
-It follows directly from the fact that the relationship between d-separation in a network and independence in the corresponding probability distribution still applies after an intervention.
-
-If $(\boldsymbol{Y} \ci \boldsymbol{W} | \boldsymbol{Z},\boldsymbol{X})$ in $G_{\overline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do1}
- P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z},\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{Z}=\boldsymbol{z}) 
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.3\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:mutilatedOriginal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (2)
-    	edge node {} (3)
-    (2) edge node {} (3)
-    (4) edge node {} (2);
-
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.5\textwidth}
-\centering
-\caption{network after the intervention $do(X=x)$, $G_{\overline{X}}$}
-\label{fig:mutilatedAfter}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$U$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$W$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge node {} (3)
-    (2) edge node {} (3);
-
-\end{tikzpicture}
-\end{subfigure}
-\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
-\label{fig:mutilatednet}
-\end{figure}
-
-\paragraph{Rule 2: (exchanging actions with observations)}
-Rule 2 describes when conditioning on $\boldsymbol{X} = \boldsymbol{x}$ and intervening, $do(\boldsymbol{X}=\boldsymbol{x})$, have the same effect on the distribution over $\boldsymbol{Y}$.
-Let $G_{\underline{\boldsymbol{X}}}$ denote the causal graph $G$ with all edges \emph{leaving} $X$ removed.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W})$ in $G_{\underline{\boldsymbol{X}}}$:
-\eqn{
-\label{eq:Do22}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},\boldsymbol{W})
-}
-
-The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$.
-Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$.
-However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$.
-Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$.
-If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths.
-This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).
-
-Equation \ref{eq:Do22} does not cover cases where acting on one set of variables allows us to replace acting on another set with conditioning (see figure \ref{fig:rule2_2}).
-The general form of rule 2 is given in equation \ref{eq:Do2}.
-
-If  $(\boldsymbol{Y} \ci \boldsymbol{X}|\boldsymbol{W},\boldsymbol{Z})$ in $G_{\underline{\boldsymbol{X}}\overline{\boldsymbol{Z}}}$:
-\eqn{
-\label{eq:Do2}
-P(\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x}),
-do(\boldsymbol{Z}=\boldsymbol{z}),
-\boldsymbol{W}) = P(\boldsymbol{Y}|\boldsymbol{X}=\boldsymbol{x},do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W})
-}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule2_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)  edge  (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule2_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$x$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1)	edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}}$}
-\label{fig:rule2_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$W$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    	edge  (3);
-
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$.
-In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$.
-If we condition on $W$, we block this indirect path.}
-\label{fig:rule2}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule22_orig}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (1) edge (2)
-    (5) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{after $do(Z)$, $G_{\overline{Z}}$}
-\label{fig:rule22_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (2) edge (3)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\underline{X}\overline{Z}}$}
-\label{fig:rule22_indirect}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
-
- %nodes
-\node[main node](1){$z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 1](4){$U$};
-\node[main node, above right=of 1](5){$V$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge  (2)
-    (4) edge (2)
-    (5) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{An example of applying equation \ref{eq:Do2}.
-In this case $(Y \ci X|Z) \text{ in } G_{\underline{X}\overline{Z}}\implies \P{Y|do(X=x),do(Z=z)} = \P{Y|X=x,do(Z=z)}$.
-Observing, rather than intervening, on $Z$ would not have allowed us to exchange $do(X=x)$ for $X=x$.
-Conditioning on $Z$ does block the indirect path $X-Z-V-Y$ but opens $X-U-Z-V-Y$.
-}
-\label{fig:rule2_2}
-\end{figure}
-
-\paragraph{Rule 3: (adding or removing actions)}
-This rule describes cases where the intervention $do(\boldsymbol{X}=\boldsymbol{x})$ has no effect on the distribution of the outcome $\boldsymbol{Y}$.
-A simple case of rule 3 is given in equation \ref{eq:Do3_1}.
-If $\boldsymbol{Y}$ is independent of $\boldsymbol{{X}}$ in $G$ after removing links entering $\boldsymbol{X}$ then can be no direct path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ and any intervention on $\boldsymbol{X}$ will not affect $\boldsymbol{Y}$.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{{X}})$ in $G_{\boldsymbol{\overline{X}}}$:
-
-\eqn{
-\label{eq:Do3_1}
-\P{\boldsymbol{Y}|do(\boldsymbol{X}=\boldsymbol{x})} = \P{\boldsymbol{Y}}
-}
-
-The general case of rule 3 is easier to state by explicitly representing the intervention in the graphical model.
-Let $G^{\boldsymbol{\hat{X}}}$ denote the graph $G$ after adding a variable $\hat{X_i}$ as a parent of each variable $X_i \in \boldsymbol{X}$ (see figure \ref{fig:rule3_augmented}).
-The variable $\hat{X_i}$ can be thought of as representing the mechanism by which $X_i$ takes its value, either by being set via intervention or as a stochastic function of its other parents \citep{Koller2009}.
-
-if $(\boldsymbol{Y} \ci \boldsymbol{\hat{X}} |\boldsymbol{Z},\boldsymbol{W})$ in $G^{\boldsymbol{\hat{X}}}_{\boldsymbol{\overline{Z}}}$:
-\eqn{
-\label{eq:Do3}
-P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),do(\boldsymbol{X}=\boldsymbol{x}),\boldsymbol{W}=\boldsymbol{w}) = P(\boldsymbol{Y}|do(\boldsymbol{Z}=\boldsymbol{z}),\boldsymbol{W}=\boldsymbol{w})
-}
-
-The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$.
-These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{original network, $G$}
-\label{fig:rule3_original}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{augmented graph $G^{\hat{X}}$}
-\label{fig:rule3_augmented}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
-\node[main node, above left=of 2](5){$\hat{X}$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (4)
-    (4) edge (3)
-    (5) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$G_{\overline{X}}$}
-\label{fig:rule3_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed}]
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, below right=of 2](4){$W$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (3)
-    (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Example application of equation \ref{eq:Do3}.
-$(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$.
-We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
-\label{fig:rule3}
-\end{figure}
-
-\subsubsection{Identifiability}
-A natural question to ask is, given a set of assumptions about the causal graph, is it possible to estimate a given interventional distribution from observational data?
-This is the identifiablity problem.
-It asks if we can obtain an unbiased point estimate for the causal query of interest in the infinite data limit.
-A query is non-parametrically identifiable if it is identifiable without  assumptions about the functional form of the dependencies between variables in the graph.
-
-\vspace*{.3cm}
-\begin{definition}[Non-parametric identifiablity]
-Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
-A causal query of the form $\P{\boldsymbol{Y}|do(\boldsymbol{X}),\boldsymbol{W}}$, where $\boldsymbol{Y}$,$\boldsymbol{X}$ and $\boldsymbol{W}$ are disjoint subsets of $\boldsymbol{V}$, is non-parametrically identifiable if it is uniquely determined by $\P{\cdot}$ and $G$.
-\end{definition}
-
-The question of non-parametric identifiability is solved.
-The do calculus is complete \citep{Shpitser2006b,Huang2006}.
-A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus.
-There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:
-
-\begin{enumerate}
-\item determine if the query can be translated into an expression involving only distributions over observed variables.
-In other words, determine if the query is identifiable given the assumptions encoded by the network, and
-\item if it is identifiable, return the required expression. 
-\end{enumerate}
-
-Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries.
-I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$X$};
-\node[ob, right=of 1](2){$Z$};
-\node[ob, right=of 2](3){$Y$};
-\node[lt,above=of 2](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{}
-\label{fig:id3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid2_instrumental_variable}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[lt](1){$U$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[ob, above left=of 2](4){$Z$};
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\vspace*{.5cm}
-\caption{}
-\label{fig:nid3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1)
-    (1) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
-\label{fig:identify}
-\end{figure}
-
-Many interesting questions relating to identifiability remain open.
-What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable?
-What if we assume various restrictions on the functional form of the relationships between the variables?
-Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity.
-A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.
-
-Although identifiability is a natural and important question to ask, it does not partition causal questions into solvable and unsolvable.
-Estimators for identifiable queries can be slow to converge and we may be able to obtain useful bounds on causal effects in cases where point estimates are not identified.
-
-\subsection{Defining causal effects} 
-\label{sec:defining-causal-effects}
-So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data.
-This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set.
-From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.
-
-\eqn{
-\label{eqn:utility_max}
-x* =\argmax_x \Esub{y\sim\P{Y|do(X=x)}}{{\cal{U}}(y)}
-}
-
-Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function.
-There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions.
-For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:
-
-\eqn{
-\label{eqn:ACE}
-ACE = \E{Y|do(X = 1)} - \E{Y|do(X=0)}
-}
-
-Assuming the expectations in equation \ref{eqn:ACE} are well defined, the ACE captures the shift in the mean outcome that arises from varying $X$.
-It does not capture changes in variance or higher moments of the distribution.
-The ACE can be generalised to non-discrete interventions by considering the effect on the expectation of $Y$ of an infinitesimal change in $x$.
-If $X$ is linearly related to $Y$ then the ACE is constant and equivalent to the corresponding coefficient in the linear structural equation model.
-
-\eqn{
-\label{eqn:ACE_continuous}
-ACE(x) = \frac{d}{dx}\E{Y|do(X=x)}
-}
-
-The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}.
-Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution.
-However we can personalise the average causal effect by stating it with respect to some observed context.
-I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}
-
-\eqn{
-PCE(z) = \E{Y|do(X=1),z} - \E{Y|do(X=0),z}
-}
-
-In some cases we may be interested in the average causal effect for some sub-group of the population.
-A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
-This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.
-
-\eqn{
-ATT = \Esub{z\sim \P{Z|x=1}}{Y|do(X = 1)} - \Esub{z\sim \P{Z|x=1}}{Y|do(X=0)}
-}
-
-Causal effects can also be written in terms of counterfactuals.
-The ACE is $\E{\cf{Y}{1} - \cf{Y}{0}}$.
-We could estimate the ratio of expectations $\frac{\E{\cf{Y}{1}}}{\E{\cf{Y}{0}}}$ instead of the difference.
-However, the quantity   $\E{\frac{\cf{Y}{1}}{\cf{Y}{0}}}$ depends on the joint distribution over the counterfactual variables $(\cf{Y}{1},\cf{Y}{0})$ and thus cannot be computed from the interventional distribution.
-
-Another way of conceptualising causal effects is as a property indicating the strength of the causal link between two variables.
-This notion is complex to formalise when the relationship between variables is non-linear.
-Suppose $Y = X \oplus Z$ with $P(Z=1)=\frac{1}{2}$, the interventional distributions over $X$ are identical after marginalising out $Z$.
-\citet{Janzing2013} propose a number of postulates that a notion of causal strength could satisfy, demonstrate why previous measures fail these postulates and propose an alternative based on information flow.
-
-\subsection{Estimating causal effects by adjusting for confounding variables}
-\label{sec:causal_effect_under_ignorability}
-Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors.
-Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds.
-Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds.
-This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}.
-The setting is also referred to as unconfounded.
-
-\vspace*{0.3cm}
-\begin{theorem}[The backdoor criterion]\citep{Pearl2000}
-Let $\boldsymbol{X}$, $\boldsymbol{Z}$ and $\boldsymbol{Y}$ be disjoint sets of vertices in a causal graph $G$.
-If $\boldsymbol{Z}$ blocks (see Definition \ref{def:unblocked}) for every path from $X_i$ to $Y_j$ that contains a link into $X_i$, for every pair $(X_i \in \boldsymbol{X},Y_j \in \boldsymbol{Y})$, and no node in $\boldsymbol{Z}$ is a decedent of a node in $\boldsymbol{X}$ then the backdoor criterion is satisfied and;
-
-\eqn{
-\label{eqn:backdoor}
-\P{\boldsymbol{y}|do(\boldsymbol{x})} =\sum_{\boldsymbol{z}} \P{\boldsymbol{y}|\boldsymbol{x},\boldsymbol{z}}\P{\boldsymbol{z}}
-}
-\end{theorem}
-
-The backdoor criterion derives from rule 2 of the do-calculus.
-Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion.
-There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}.
-The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}.
-Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}.
-Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for.
-For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.
-
-\begin{figure}
-\centering
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{Multiple valid adjustment sets.}
-\label{fig:adjust_multiple}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](2){$X$};
-\node[ob, right=of 2](3){$Y$};
-\node[ob,above=of 3](4){$Z_2$};
-\node[ob,above=of 2](5){$Z_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (5) edge (2) edge (4)
-    (4) edge (3);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.48\textwidth}
-\centering
-\caption{Don't condition on all observable variables}
-\label{fig:m-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,lt/.style={latent},ob/.style={observed}]
-\node[ob](1){$Z$};
-\node[ob, below left=of 1](2){$X$};
-\node[ob, below right=of 1](3){$Y$};
-\node[lt,above=of 3](4){$U_3$};
-\node[lt,above=of 2](5){$U_1$};
- \path[every node/.style={font=\sffamily\small}]
-    (2) edge (3)
-    (4) edge (1) edge (3)
-    (5) edge (2) edge (1);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Identifying an optimal adjustment set is not always intuatively obvious.
-There may be multiple valid adjustment sets.
-In sub-figure (a) $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$.
-Adding additional variables (even pre-treatment variables) to a valid adjustment set can result in an invalid set as in sub-figure (b).
-In this case the empty set is a valid adjustment set for the causal effect of $X$ on $Y$, however adding $Z$ would open a backdoor path $X-U_1-Z-U_2-Y$}
-\label{fig:adjustment}
-\end{figure}
-
-Given that a set of variables $\boldsymbol{Z}$ satisfies the backdoor criterion (or equivalently the conditional ignorability assumption), the interventional distribution is asymptotically identifiable and can be estimated from equation \ref{eqn:backdoor}.
-The expected value of $Y$ after the intervention $do(X=x)$ is given by equation \ref{eqn:expected_post_intervention} and the average causal effect for a binary intervention $x \in \set{0,1}$ is given by equation \ref{eqn:ace_ignoreable}.
-
-\eqn{
-\label{eqn:expected_post_intervention}
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}}
-}
-
-\eqn{
-\label{eqn:ace_ignoreable}
-ACE = \Esub{z\sim \P{\vb{Z}}}{\E{Y|1,\vb{z}}-\E{Y|0,\vb{z}}}
-}
-
-Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata.
-However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates.
-There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review.
-The key approaches are based on matching on covariates, propensity score methods and regression.
-We now examine these approaches from a machine learning perceptive.
-
-In standard supervised learning, we have a training set $(\vb{x_1},y_1),...,(\vb{x_n},y_n)$ assumed to be sampled i.i.d from an unknown distribution $\P{\vb{x},y} =\P{\vb{x}}\P{y|\vb{x}}$.
-The goal is to select a hypothesis $h \in \mathcal{H} : \mathcal{X} \rightarrow \mathcal{Y}$ such that, on unseen data $ \sim \P{\vb{x},y}$, $h(\vb{x})$ is close (by some metric) in expectation to $y$.
-In other words we wish to minimise the generalisation error $E_{out}(h)$,
-
-\eqn{
-E_{out}(h) = \Esub{(\vb{x},y)\sim \P{\vb{x}}\P{y|\vb{x}}}{L(h(\vb{x}),y)}
-}
-
-We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample.
-We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.
-
-\eqn{
-E_{in}(h) = \frac{1}{n}\sum_{i=1}^n L(h(\vb{x}_i),y_i)
-}
-
-The VC-dimension of the hypothesis space provides (typically loose) bounds on the probability that $E_{out} >> E_{in}$.
-However, in practice, the generalisation error is usually estimated empirically from a hold-out set of the sample that was not used to train the model, or via cross-validation.
-
-In the causal effect estimation under ignorability, we have training data $(\vb{x}_1,\vb{z}_1,y_1),...,(\vb{x}_n,\vb{z}_n,y_n)$ sampled i.i.d from $\P{\vb{z}}\P{\vb{x}|\vb{z}}\P{y|\vb{x},\vb{z}}$.
-Estimating $\E{Y|do(\vb{X}=\vb{x})}$ corresponds to selecting a hypothesis $h \in \mathcal{H}:\mathcal{X} \times \mathcal{Z} \rightarrow \mathcal{Y}$ that minimises;
-
-\eqn{
-\label{eqn:causal_covariate_shift}
-E_{out} &= \Esub{(\vb{x},\vb{z},y)\sim \delta(\vb{x}-\vb{x'})\P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},\\
-&= \Esub{(\vb{z},y)\sim \P{\vb{z}}\P{y|\vb{x},\vb{z}}}{L_2(h(\vb{x},\vb{z}),y)},
-}
-
-\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
-If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
-\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
-Take the case where we have a binary intervention $x \in \set{0,1}$.
-Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$.
-More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.
-
-\subsubsection{Regression}
-
-The regression approach is to learn a function that is a good approximation to the output surface $\E{Y|X,Z}$.
-Let $f_1(z)= \E{Y|X=1,Z=z}$.
-The expectation of $Y$ after the intervention $X=1$ is then obtained by taking the expectation with respect to $Z$, $\E{Y|do(X=1)} = \Esub{z \sim \P{Z}}{\E{Y|X=1,z}}$.
-We can learn a parametric regression model $\hat{f}_1(z)$ via empirical risk minimisation.
-
-\eqn{
-\hat{f}_1(z) = h_1(z;\hat{\theta}_{obs}) \text{, where } \hat{\theta}_{obs} = \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)  \right]
-}
-
-This estimator is consistent with respect to the observational distribution.
-As the sample size tends to infinity, $\hat{\theta}_{obs}$ approaches the parameter within the hypothesis space that minimises the expected loss given data sampled from the observational distribution.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z|x=1}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-If the model is correctly specified such that $f_1(z) = h_1(z;\theta^*)$ for some  $\theta^* \in \Theta$ then the empirical risk minimisation estimate is consistent with respect to the loss over any distribution of $Z$  \citep{Sugiyama2007}, including the interventional one.
-
-\eqn{
-\lim_{n \rightarrow \infty}\hat{\theta}_{obs} = \theta^* = \argmin_{\theta \in \Theta} \Esub{(z,y) \sim \P{z}\P{y|x=1,z}}{L\left(h_1(z;\theta),y \right)}
-}
-
-The average causal effect can then be estimated by,
-
-\eqn{
-\label{eqn:cef_empirical_regression}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left(\hat{f}_1(z_i) - \hat{f}_0(z_i)\right)
-}
-
-Regression thus has a direct causal interpretation if the parametric model is correctly specified and the covariates included form a valid backdoor adjustment set for the treatment variable of interest in the corresponding structural equation model.
-
-\subsubsection{Propensity scores}
-If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
-The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
-\caption{
-Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
-In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
-If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$.
-We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points).
-In both cases the linear model is a good fit to the data.
-However, the resulting estimate of the causal effect is very poor for the lower values of $z$.
-}
-\label{fig:covariate_shift_adjust}
-\end{figure}
-
-A general approach to estimating the expectation of some function $f(\cdot)$ with respect to data from some distribution $\P{\cdot}$, when we have data sampled from a different distribution $\Q{\cdot}$ is importance sampling \citep{horvitz1952generalization,Koller2009}.
-\eqn{
-\label{eqn:importance_sampling}
-\Esub{\vb{v} \sim \P{\vb{v}}}{f(v)} = \Esub{\vb{v} \sim \Q{\vb{v}}}{f(v)\frac{\P{\vb{v}}}{\Q{\vb{v}}}}
-}
-
-This importance weighting approach can be applied to the covariate shift/average causal effect problem by weighting the terms in the empirical risk minimisation estimator \citep{Sugiyama2007}.
-
-\eqn{
-\label{eqn:importance_weighted_erm}
-\hat{\theta}_{iw} &= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{\P{z_i}\delta(x_i-1)}{\P{z_i}\P{x_i=1|z_i}}  \right] \\
-&= \argmin_{\theta \in \Theta}\left[\frac{1}{n}\sum_{i=1}^n \ind{x_i=1} L\left(h_1(z_i;\theta),y_i\right)\frac{1}{e(z_i)}  \right],
-}
-
-where $e(\vb{z})$ is the propensity score, defined by \citep{Rosenbaum1983};
-\eqn{
-e(\vb{z})\equiv \P{x=1|\vb{z}}
-}
-
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
-Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
-
-The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$.
-\citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$.
-This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$.
-Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}.
-In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known.
-However in general, it must be estimated from data.
-Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}.
-\citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.
-
-\eqn{
-\E{Y|do(X=x)} = \Esub{z\sim \P{\vb{Z}}}{\E{Y|x,\vb{z}}} = \Esub{z\sim \P{\vb{Z|x}}}{\E{Y|x,\vb{z}}\frac{1}{e(\vb{z})}}
-}
-
-\eqn{
-\label{eqn:simple_inverse_propensity}
-\hat{\tau}_{ip} = \frac{1}{n}\sum_{i=1}^n \left(\frac{\ind{x_i=1}y_i}{e(\vb{z}_i)} -\frac{\ind{x_i=0}y_i}{1-e(\vb{z}_i)}\right)
-}
-
-\subsubsection{Matching}
-There is a straightforward connection between matching and regression for causal effect estimation.
-If $h \in \mathcal{H} \implies h + a \in \mathcal{H}$ for any constant $a$ and $\hat{f}$ is selected by minimising empirical risk with an $L_2$ loss then $\frac{1}{n}\sum_{i=1}^n \ind{x_i=1}\hat{f}_1(z_i) = \frac{1}{n}\sum_{i=1}^n \ind{x_i=1}y_i$ \footnote{\citep{Imbens2004} state this holds for most implementations}, and equation \ref{eqn:cef_empirical_regression} can be re-written as:
-
-\eqn{
-\label{eqn:cef_regression_missing}
-\hat{\tau}_{reg} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \hat{f}_0(z_i)\right) +\ind{x_i=0}\left(\hat{f}_1(z_i) - y_i \right)\right]
-}
-
-This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
-For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
-Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
-\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
-
-\eqn{
-\hat{\tau}_{match} = \sum_{i=1}^n \left[\ind{x_i=1}\left(y_i - \frac{1}{k}\sum_{j \in J_k(i)}y_j\right) +\ind{x_i=0}\left(\frac{1}{k}\sum_{j \in J_k(i)}y_j - y_i \right)\right]
-}
-
-This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression.
-There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}.
-Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}.
-An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.
-
-The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions.
-A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.
-
-The significance of this should not be underestimated.
-Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
-With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
-\citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem.
-There is relatively little theory on model selection for estimating the propensity score.
-To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion.
-It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}.
-With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
-I have found no examples of such an approach.
-
-The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}.
-Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available.
-The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.
-
-\section{Bayesian reasoning and causality}
-\label{sec:bayesian_causality}
-The Bayesian approach to modelling encapsulates a very general approach to combining prior knowledge and assumptions with data to draw inferences.
-You can specify a complex model, capturing  your assumptions about the underlying mechanisms of your system, describe what you are uncertain about with parameters (and priors over their values) and then update your beliefs about the values of these parameters based on the observed data.
-If you believe the model to be a mechanistic representation of the way world works then you can use it to infer the consequences of interventions.
-
-Why then do we need a special framework for causal inference; can't we simply encode all the assumptions required to infer the outcome of an intervention in a given system in a Bayesian model?
-The short answer is we could.
-However, causal graphical models represent a useful level of abstraction for many problems.
-Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention.
-We do not need to specify the functional relationships between the variables or priors over their distributions.
-The power of defining an intervention in this way stems from the number of things that are assumed invariant between the two processes.
-All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.
-The causal framework also provides, via the do-calculus, a rigorous method to draw a boundary around which variables we must measure and model to identify causal effects.
-
-We could represent problems of the type where we try to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and what they have in common, see figure \ref{fig:causal_ordinary_bayes}.
-This does not require any special framework or notation.
-The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks.
-However, without a causal framework, we have to make assumptions about what will be invariant to the intervention specifically for each such problem we encounter.
-For complex problems, it is very difficult to conceptualise the assumptions we expect to hold without the benefit of a causal framework.
-
-\begin{figure}[h]
-\center
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
-\tikzstyle{box}=[rectangle, draw=black!100]
- %nodes
-\node[node](1){$Z$};
-\node[node, below left=of 1](2){$X$};
-\node[node, below right=of 1](3){$Y$};
-\node[node, right=1.5cm of 3](4){$x$};
-\node[lt, above right=of 4](5){$Z'$};
-\node[lt, below right=of 5](6){$Y'$};
-\node[lt, above=2.0cm of 3](9){$\theta_{y}$};
-\node[lt, above=0.5cm of 9](7){$\theta_z$};
-\node[lt, above=2.0cm of 2](8){$\theta_x$};
-\path[every node/.style={font=\sffamily\small}]
-    (1) edge (2) edge (3)
-    (2) edge (3)
-    (4) edge (6)
-    (5) edge (6)
-    (8) edge (2)
-    (7) edge (1) edge (5)
-    (9) edge (3) edge (6);
-\node[rectangle, inner sep=4.0mm, fit= (1) (2) (3),draw=black!100,label=below right:N] {};
-\node[rectangle, inner sep=4.0mm, fit= (5) (6),draw=black!100,label=below right:M] {};
-\end{tikzpicture} 
-\caption{Causal inference with ordinary Bayesian networks.
-The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$.
-The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$.
-As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate.
-However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
-\label{fig:causal_ordinary_bayes}
-\end{figure}
-
-We can also combine causal graphical models with Bayesian approaches to estimation.
-For example, we can use a causal graphical model to identify a set of variables that form a valid adjustment set and then use Bayesian regression to estimate posterior distribution over a given causal effect.
-However, in doing so we need to be careful about the way in which we specify priors as I highlight in the next section.
-
-\subsection{Careful with that prior}
-
-Suppose we have a data generated by the linear Gaussian Bayesian network below.
-The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$.
-The variable $U$ is latent.
-
-\eqn{
-\P{U} &= N(0 , v_u) \\
-\label{eqn:careful_with_priors}
-\P{Z|U} &= N(w_{zu}U , v_z) \\
-\P{X|Z} &= N(w_{xz}Z , v_x) \\
-\P{Y|U,Z,X} & = N(w_{yu}U +w_{yz}Z + w_{yx}X , v_y)
-}
-
-\begin{figure}[h]
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
-
- %nodes
-\node[main node](1){$Z$};
-\node[main node, below left=of 1](2){$X$};
-\node[main node, below right=of 1](3){$Y$};
-\node[lt, right=of 1](4){$U$};
-
- \path[every node/.style={font=\sffamily\small}]
-    (1) edge (2)
-    	edge (3)
-    (2) edge (3)
-    (4) edge (1) edge (3);
-
-\end{tikzpicture} 
-\caption{ }
-\label{fig:careful_with_priors}
-\end{figure}
-
-As each variable is a linear function of its parents, with Gaussian noise, resulting in a joint distribution $\P{U,Z,X,Y}$ that is multivariate normal.
-Marginalising out $U$ and conditioning on $X$ and $Z$ yields,
-
-\eqn{
-Y \sim N(w_{yx} X + \beta Z, \epsilon) 
-}
-
-where
-
-\eqn{
-\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and}\\
-\epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
-}
-
-The causal effect of $X$ on $Y$ is identifiable (even without the linear Gaussian assumptions) as $Z$ satisfies the backdoor criterion and the expectation of the coefficient for $X$ in a standard OLS regression of $Y$ against $X$ and $Z$ is $w_{yx}$.
-However, the causal effect of $Z$ on $Y$ is not identifiable due the presence of the unobserved confounder $U$.
-The coefficient $\beta$ captures both the causal relationship between $Z$ and $Y$, $w_{yz}$ and the indirect relationship through $U$, which we cannot separate without observing $U$.
-
-If we do a Bayesian regression of $X$ and $Z$ against $Y$ and use intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ we are likely to select something centered around $w_{yz}$, but this is not correct and can introduce bias into our estimate of the coefficient for $w_{yx}$, as we demonstrate in figure \ref{fig:careful-prior-example}.
-The prior for $\beta$ should be based on our belief about how we expect $Z$ to be \emph{associated} with $Y$ after marginalising out any confounding variables.
-
-\begin{figure}[h]
-\begin{subfigure}[t]{0.99\textwidth}
-     \caption{No prior on $\beta$, the posterior on $w_{yx}$ is centered around its true value}
-    \includegraphics[scale=1]{prior_no_prior}
-\end{subfigure}
-\begin{subfigure}[t]{0.99\textwidth}
-     \caption{Prior on $\beta$ centered around $w_{yz}$, the causal effect of $Z$ on $Y$, $Prior(\beta)= N(w_{yz},\sigma = 0.5)$.
-The posterior on $w_{yx}$ is biased away from its true value (as is the posterior on $\beta$ but this is not the key quantity of interest).}
-    \includegraphics[scale=1]{prior_bad_prior.pdf}
-\end{subfigure}
-\caption{An example demonstrating how selecting a prior for a nuisance parameter centered around the causal effect of that parameter on the outcome, rather than its association with the outcome, can bias estimates for actual parameter of interest.
-We sample $n=1000$ data points from the joint distribution defined by $P(U)P(Z|U)P(X|Z)P(Y|U,Z,X) \sim N(0,1)N(2U,0.09)N(0.5Z,1)N(3U -Z + 0.5X , 0.25)$ and fit the model $Y \sim N(w_yx X+\beta Z,\epsilon)$ in Stan and plot the posterior distributions over $w_yx$ and $\beta$.
-Figure (a) shows the results with no prior (equivalently an improper prior) for both $w_yx$ and $\beta$.
-Figure (b) shows the results when we place a Gaussian prior centered around the $w_yz$ on the distribution for $\beta$.}\label{fig:careful-prior-example}
-\end{figure}
-
-This result is counter intuitive.
-When fitting a model to estimate the affect of marketing on sales, it is natural to put a prior on the coefficient for price that is strongly negative because we believe increasing price decreases sales.
-However, our prior should be on the expected association between price and sales, holding fixed all the other variables in the model, not the causal relationship between price and sales.
-An alternative is to explicitly include $U$ as a latent variable in the model.
-However, this comes at a computational cost with little benefit for estimation, unless we have good knowledge of the relationships between $U$, $Z$ and $Y$.
-More fundamentally, at whatever level we decide to stop adding latent variables and specify a prior, we can introduce this form of bias if priors are selected naively on the basis of causal intuition and the issue is not limited to linear-Gaussian models.
-I believe this is an important insight for the many applications of Bayesian modelling that make use of human elicited priors, since people tend to think causally.
-The solutions, either explicitly modelling latent variables to higher levels or allowing for them in setting priors, have the effect of broadening the priors over nuisance parameters in the model.
-
-\chapter{Learning from interventions}
-\label{chap:interventionalist_view}
-
-The previous sections all focus on aspects of the question; how can we estimate the effect of an intervention in a system from data collected prior to taking it.
-There is an obvious alternative.
-Instead of trying to infer the outcome of an intervention from passive observations one can intervene and see what happens.
-There are three key differences between observing a system and explicitly intervening in it.
-Firstly, when we intervene we choose which actions to take and thus have control over which data points we obtain to learn from.
-Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}.
-Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias.
-Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning.
-For example, in a drug trial , assigning people a sub-optimal treatment has real world costs.
-This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain.
-This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.
-
-\section{Randomised experiments}
-\label{sec:randomized_experiment}
-Randomised controlled trials are often presented as the gold standard for determining causal effects.
-What is it about randomisation that makes it so important when it comes to causality?
-The graphical model for a randomised controlled experiment is shown in figure \ref{fig:random_experiment_network}.
-If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process.
-Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3);
-\end{tikzpicture}
-\caption{causal network for a randomised experiment}
-\label{fig:random_experiment_network}
-\end{figure}
-
-Randomisation does not ensure target and control group are exactly alike.
-The more other features (observed or latent) influence the outcome, the more likely it is that there will be a significant difference in the joint distribution of these variables between the target and control groups in a finite data sample.
-However, the variance in the outcome, within both the target and control groups, also increases.
-The net result is increased variance (but not bias) in  the estimate of causal effects.
-
-Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest.
-If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds and we can estimate the average causal effects directly from the differences in outcome across treatments.
-More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, see figure \ref{fig:random_experiment_network_stratified}, necessitating that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability \S\ref{sec:causal_effect_under_ignorability}.
-The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
-See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Treatment};
-\node[node, above left=of 1](2){Randomiser};
-\node[node, above right=of 1](5){Covariates};
-\node[node, below right=of 5](3){Outcome};
-\node[lt, above=of 3](4){U};
-
-\path[]
-	(2) edge (1)
-	(1) edge (3)
-	(4) edge (3)
-	(5) edge (3) edge (1);
-\end{tikzpicture}
-\caption{causal network for a stratified randomised experiment if the probability an individual is assigned a given treatment depends on some covariates.}
-\label{fig:random_experiment_network_stratified}
-\end{figure}
-
-The benefit provided by randomisation in breaking the link between the treatment variable and any latent confounders should not be understated.
-The possibility of unobserved confounders cannot be empirically ruled out from observational data \citep{Pearl2000} (there is no test for confounding).
-This means causal estimates from non-experimental data are always subject to the criticism that an important confounder may have been overlooked or not properly adjusted for.
-However, randomised experiments do have some limitations.
-
-\subsection{Limitations of randomised experiments}
-\label{subsec:limitations_of_experiment}
-
-The idealised notion of an experiment represented by figure \ref{fig:random_experiment_network} does not capture the complexities of randomised experiments in practice.
-There may be imperfect compliance, the treatment selected by the randomiser is not always followed, or output censoring, the experimenter is not able to observe the outcome for all units (for example if people drop out).
-If compliance or attrition is not random but associated with (potentially latent) variables that also effect the outcome then the problem of confounding bias returns.\footnote{Non-compliance is a problem if the goal is to estimate the causal effect of the treatment on the outcome but not if the goal is to estimate the causal effect of prescribing the treatment.
-The latter makes sense in a context where the process by which people decide whether or not to take the treatment they have been prescribed is likely to be the same if the treatment were made available more generally beyond experimental trial.} See figure \ref{fig:random_experiment_network_imperfect_compliance} for a the graphical model of a randomised experiment with imperfect compliance.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick, node/.style={observedrect}, lt/.style={latent}]
-\node[node](1){Actual Treatment};
-\node[node,above left=of 1](5){Assigned Treatment};
-\node[node, above =of 5](2){Randomiser};
-\node[node, right=of 1](3){Outcome};
-\node[lt, above left=of 3](4){U};
-\path[]
-	(2) edge (5)
-	(5) edge (1)
-	(1) edge (3)
-	(4) edge (3) edge (1);
-\end{tikzpicture}
-\caption{causal network for a randomised experiment with imperfect compliance}
-\label{fig:random_experiment_network_imperfect_compliance}
-\end{figure}
-
-It is not always possible or ethical to conduct a randomised controlled trial as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
-When experimentation is possible, it is frequently difficult or expensive.
-This means experimental data sets are often much smaller than observational ones, limiting the complexity of models we can explore.
-In addition, they are often conducted on a convenient but unrepresentative sample of the broader population of interest (for example first year students at research universities).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963} in that they should replicate well in a similar population, but very low \emph{external validity}; the results may not carry over to the general population of interest.
-The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.4]{double_blind_parashute_trial_cartoon.png}
-\end{center}
-\caption{Experiments are not always ethical; an illustration of a randomised cross-over trial of parachutes for the prevention of morbidity and mortality associated with falls from large heights.}
-\label{fig:parashute}
-\end{figure}
-
-Finally, non-adaptive randomised experiments are not optimal from either an active or reinforcement learning perspective.
-As an experiment proceeds information is obtained about the expectation and variance of each intervention (or treatment).
-Fixed experimental designs cannot make use of this information to select which intervention to try next.
-This results in both poorer estimates for a fixed number of experimental samples and more sub-optimal actions during the course of the experiment.
-
-\section{Multi armed bandits}
-\label{sec:multi-armed-bandits}
-
-Multi-armed bandits address the problem of designing experiments that can adapt as samples are observed.
-Their introduction is generally attributed to \citet{Thompson1993}.
-In its classic formulation \citep{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
-Each arm $i$ is associated with a fixed but unknown reward distribution.
-\footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$ the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
-The goal of the learner is to maximise the total reward they receive.
-This problem captures the exploration-exploitation trade-off, the learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.
-
-\vspace{0.5cm}
-\begin{definition}[Stochasitic k-armed bandit problem]
-Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action.
-The multi-armed bandit problem proceeds over $T$ rounds.
-In each round $t$,
-\begin{enumerate}
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
-\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
-\end{enumerate}
-
-At the end of the game the total reward obtained by the learner is $\sum_{t=1}^T \cfb{Y}{a_t}$.
-We denote the expected reward for the action $i$ by $\mu_i$ and the action with the highest expected reward by $i^*$.
-Note we have used counterfactual notation (see section \ref{sec:counterfactuals}) to denote the rewards for each action, $\cfb{Y}{i}$ is the reward the algorithm would have received had it selected action $i$ at timestep $t$.
-I discuss the (potentially) counterfactual nature of regret further in section \ref{sec:counterfactual-nature-regret}.
-\end{definition}
-
-The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
-To account for this, the performance of bandit algorithms is quantified by the the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
-This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
-There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.}.
-\eqn{
-\label{eqn:raw_regret}
-\rawregret = \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
-}
-
-Both the rewards and the actions selected by the algorithm are random variables.
-The majority of work in the bandit literature focuses on analysing and optimising some form of the expected regret, however there has been some work that also considers the concentration of the regret \citep{Auer1995,Audibert2007,Audibert2009exploration}.
-The expectation of the regret, as defined by equation \ref{eqn:raw_regret}, is referred to as the pseudo-regret \citep{Bubeck2012} and is given by equation \ref{eqn:pseudo_regret}.
-A stochastic bandit algorithm is learning if it obtains pseudo-regret that is sub-linear in $T$.
-
-\vspace{0.5cm}
-\begin{definition}[Pseudo-Regret]
-\eqn{
-\label{eqn:pseudo_regret}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}\E{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}} \\
-&= T\mu_{i^*} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}}
-}
-\end{definition}
-
-The regret is invariant to adding a constant to the expected rewards for all actions.
-However, it still depends on key characteristics of the reward distributions for each action.
-Bandit algorithms are designed given assumptions about the form of the distributions, such as that they come from a given family (i.e Bernoulli bandits, Gaussian bandits) or that the rewards are bounded in some range.
-Given these assumptions, the performance of the algorithm is characterised in two ways; by the \emph{problem dependent regret}, which typically depends on how far each arm is from optimal and by the \emph{worst case regret}, which is the maximum regret over all possible configurations of the reward distributions (for a given horizon $T$ and number of arms $k$).
-
-\subsection{Stochastic bandits: Approaches and results}
-\label{sec:stochastic-bandit}
-
-The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators.
-The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards.
-As a result, the probability that each action is selected evolves over time, the actions are not sampled i.i.d from a fixed distribution and the number of times each action is selected is a random variable.
-The expectation and variance guarantees of standard estimators do not hold in this setting, see figure \ref{fig:early_stopping} for a concrete example.
-This makes it very difficult to  obtain an analytical expression for the expected regret for a given algorithm and problem.
-Instead, the focus is on computing bounds on the expected regret.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=1]{extreme_early_stopping}
-\caption{Standard empirical estimators can be biased if the number of samples, $n$, is not fixed in advance, but is a random variable that depends on the values of previous samples.
-This example plots the distribution (over $10^6$ simulations) of $\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$, where $X_i \sim Bernoulli(0.2)$.
-In each simulation, we stop taking samples if the average value of $X_i$ up to that point exceeds a threshold of $0.3$ or $n$ reaches $100$.
-$\E{\hat{\mu}} = 0.439$.
-The estimator is substantially biased above $\E{X_i} = 0.2$ by the early stopping.
-Note that excluding experiments that were stopped early creates a bias in the opposite direction, $\E{\hat{\mu}|n=100} = 0.185$, as trials that obtained positive results early are excluded.
-This has some interesting potential real world implications.
-Early stopping of clinical trials is controversial.
-A researcher conducting a meta-analysis who wished to avoid (rather than bound) bias due to early stopping would have to exclude not only those trials which were stopped early but those which \emph{could} have been stopped early.}
-\label{fig:early_stopping}
-\end{figure}
-
-There are a few key principles that are used to guide the development of bandit algorithms.
-The simplest is to explicitly separate exploration from exploitation and base estimation of the expected rewards of each arm only on the data generated during exploration steps.
-A common example in practice is uniform exploration (or A/B testing) for some fixed period followed by selecting the action found to be best during the exploration phase.
-This results in simpler analysis, particularly if the number of exploration steps is fixed in advance, however it is sub-optimal, even if the exploration period is adaptive \citep{NIPS2016_6179}.
-
-Another key approach is \emph{optimism in the face of uncertainty}.
-Applied to stochastic bandits, the optimism in the face of uncertainty principle suggests computing a plausible upper-bound for the expected reward of each arm, and selecting the arm with the highest upper bound.
-The optimism principle encourages exploitation and exploration because a high upper bound on the expected reward for an action implies either the expected reward or the uncertainty about the reward for that action is high.
-Thus selecting it yields either a good reward or useful information.
-
-\citet{Lai1985} leveraged the optimism in the face of uncertainty principle to develop an algorithm for specific families of reward distributions, including the exponential family.
-They showed that, for a given bandit problem, the pseudo-regret increased with $\bigo{log(T)}$ asymptotically and proved this is asymptotically efficient.
-However, their algorithm is complex and memory intensive to compute as, at each timestep, it relies on the entire sequence of rewards for each arm.
-\citet{Agrawal1995} developed a simpler algorithm that computed upper bounds based only on the mean of previous samples for each arm, whist retaining the logarithmic dependence on $T$.
-Finally, \citet{Auer2002} developed the UCB-1 algorithm, see algorithm \ref{alg:ucb-1}, which requires only that the reward distributions are bounded, and proved finite-time regret bounds.
-We now assume the rewards are bounded in $[0,1]$.
-The algorithm and regret bounds can be generalised to submission reward distributions, see \citet{Bubeck2012}.
-
-\begin{algorithm}[H]
-\caption{UCB-1}\label{alg:ucb-1}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} horizon $T$.
-\STATE Play each arm once. 
-\FOR{$t \in 1,\ldots,T$}
-\STATE Count the number of times each arm has been selected previously $n_{t,i} = \sum_{s=1}^{t-1} \ind{a_s = i}$
-\STATE Calculate the mean reward for each arm $\hat{\mu}_{t,i} = \frac{1}{n_{t,i}} \sum_{s=1}^{t-1}\ind{a_s = i}Y_t$
-\STATE Select arm $a_t \in \argmax_{i = \set{1,...,k}}\left(\hat{\mu}_{t,i}+\sqrt{\frac{2 \log{t}}{n_{t,i}}}\right)$
-\ENDFOR
-\end{algorithmic}
-\end{algorithm}
-
-Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
-The problem dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
-
-\eqn {
-\label{eqn:ucb_problem_dependent_regret}
-\regret \leq & \sum_{i:\Delta_i > 0}\left(\frac{8\log(T)}{\Delta_i}+2 \right)
-}
-
-Somewhat unintuitively, the regret increases as the value of the arms gets closer together.
-This is because it becomes harder for the algorithm to identify the optimal arm.
-As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
-The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
-The regret cannot exceed what would be obtained by selecting the a sub-optimal arm in every timestep, $T\Delta$, so the worst case regret is bounded by the minimum of equation \ref{eqn:ucb_problem_dependent_regret} and $T\Delta$ which is maximised when they are equal, see figure \ref{fig:ucb_regret_vs_delta}.
-By solving this equality for $\Delta$ one can show the worst case regret is bounded by equation \ref{eqn:ucb_worst_case_regret}, see \citet{Bubeck2012}.
-
-\eqn{
-\label{eqn:ucb_worst_case_regret}
-\regret \in \bigo{\sqrt{kT\log(T)}}
-}
-
-\begin{figure}
-\includegraphics[scale=1]{ucb_regret_vs_delta}
-\caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
-The solid curve shows the mean (cumulative) regret, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
-The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
-\label{fig:ucb_regret_vs_delta}
-\end{figure}
-
-The form of the dependence on the number of arms $k$ and horizon $T$ differs between the problem dependent and worst case regret.
-The problem dependent regret grows linearly with the number of arms, $k$, and logarithmically with $T$.
-The difference stems from the fact the problem dependent regret defines how the regret grows for a given set of reward distributions as $T$ increases, whereas in the worst case regret, the gap between expected rewards is varied as a function of $T$.
-\citet{Auer1995} show that the worst case regret for the k-armed bandit problem is lower bounded by $\regret \in \bigomega{\sqrt{kT}}$
-
-Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
-This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem dependent bound of the form achieved by UCB \citep{Lattimore2015}.
-
-\begin{figure}
-\includegraphics[scale=1]{upper_bound_vs_actual_average_regret}
-\caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$.
-The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $bernoulli([.3,.7])$ over 1000 simulations.
-The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
-\label{fig:ucb_upperbound_vs_actual_performance}
-\end{figure}
-
-Finally, there is the heuristic principle of playing each arm with probability proportional to the likelyhood that it is optimal.
-This approach is generally called Thompson sampling as it was the method proposed in the original bandit paper by \citet{Thompson1993}.
-Thompson sampling has strong empirical performance, \citep{chapelle2011empirical}.
-However, it is complex to analyse, \citet{kaufmann2012thompson} demonstrate that it obtains optimal problem dependent bounds, \citet{agrawal2013further} show that it obtains worst case regret of $\bigo{\sqrt{kT\log(T)}}$, equivalent to UCB.
-
-\subsection{Pure-exploration problems}
-\label{sec:pure-exploitation-problems}
-Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}.
-In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$.
-The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.
-
-\vspace{0.5cm}
-\begin{definition}[Simple Regret]
-\eqn{
-\label{eqn:simple_regret}
-\simpleregret=\mu_{i^*} - \E{\mu_{\hat i^*}}.
-}
-\end{definition}
-
-The best arm identification problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialisation or exploitation phase.
-For example, many strategies might be assessed via simulation prior to one being selected and deployed.
-The worst case simple regret for a k-armed bandit is lower bounded by equation \ref{eqn:worst_case_simple_regret_lower_bound} (\citep{Bubeck2009}).
-
-\eqn{
-\label{eqn:worst_case_simple_regret_lower_bound}
-\simpleregret \in \bigo{\sqrt{K/T}}
-}
-
-%\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
-
-Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds. The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period.
-
-\subsection{Adversarial Bandits}
-\label{sec:adversarial-bandits}
-
-Adversarial bandits, described by \citet{Auer1995}, are an alternate, widely studied, setting that relaxes the assumption that rewards are generated stochastically.
-Instead, simultaneously with the learner selecting an action $a_t$, a potentially malicious adversary selects the reward vector $\boldsymbol{Y}_t$.
-As in the stochastic setting, the learner then receives reward only for the selected action.
-
-\vspace{0.5cm}
-\begin{definition}[Adversarial k-armed bandit problem]
-\label{def:adversarial-k-armed-bandit}
-Let $\actionspace = \set{1,...k}$ be the set of available actions.  In each round $t \in 1,...,T$, 
-\begin{enumerate}
-\item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[\cfb{Y}{1},...,\cfb{Y}{k}]$. 
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
-\item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
-\end{enumerate} 
-\end{definition}
-
-Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
-In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
-
-For oblivious adversarial bandits we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
-We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because sequence of rewards is fixed, however the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
-
-\eqn{
-\label{eqn:pseudo_regret-adversarial}
-\regret(\pi) &= \max_{i \in \set{1,...,k}}{\sum_{t=1}^T{\cfb{Y}{i}}} - \E{\sum_{t=1}^T{\cfb{Y}{a_t}}}
-}
-
-The policy (or algorithm) used by the learner is available to the adversary before the game begins, and there are no limitations placed on the amount of computation the adversary can perform in selecting the reward sequences.
-This implies the adversary can ensure that any learner with a deterministic policy suffers regret $\bigo{T}$ by forecasting their entire sequence of actions.
-For example, if the learner will play $a_1 = 1$ in the first round, then the adversary sets the reward $\vec{Y_1} = [0,1,1,...1]$, forecasts what action the learner will play in round 2, given they received a reward of 0 in round 1, and again generates the reward vector such that the action the learner will select obtains no reward, and all other actions obtain the maximum reward.
-This implies adversarial bandit policies must be sufficiently random to avoid such exploitation \footnote{The UCB algorithm, defined by algorithm \ref{alg:ucb-1}, is deterministic if the order in which arms are played during the first $k$ rounds is fixed and the method for selecting which arm to play when multiple-arms have the same upper-confidence bound is not-random (for example, select the arm one with the lowest index $i$).}
-
-The seminal algorithm for adversarial bandits is Exp-3 \citep{Auer2002}, which, like UCB, obtains worst case pseudo-regret of $\bigo{\sqrt{TK\log(T)}}$  \citep{Auer1995}.
-Optimal algorithms, with $\regret = \bigo{\sqrt{TK}}$, have also been demonstrated for the oblivious adversarial setting \citep{Audibert2009}.
-The focus, for adversarial bandits, is on analysing the worst case regret because the problem dependent regret is not well defined without additional assumptions.
-However there has been recent work on developing algorithms that are optimised for both the adversarial and stochastic settings, in that they are sufficiently cautious to avoid linear regret in the adversarial setting but can nonetheless obtain good problem dependent regret in more favourable environments \citep{bubeck2012bothworlds,Auer2016}.
-
-Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, does not fully capture an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings where the rewards change over time, the pseudo-regret can be negative, see figure \ref{fig:non-stationary-bandit}, so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
-An example of a setting that lies between stochastic and adversarial bandit problems is the non-stationary setting, in which the rewards are generated stochastically from a distribution that varies over time.
-Adversarial bandit algorithms may perform better in such settings than standard stochastic policies to the extent that they explore more (to avoid the adversary simulating their behaviour) and thus adapt quicker to changes in the reward distribution.
-Adversarial algorithms also have stronger worst case regret guarantees, since even the weak regret for stochastic bandits is not guaranteed to be sub-linear in such settings.
-However, if we have constraints on how rapidly or frequently the reward distributions can change over time it is better to use algorithms specifically developed to exploit such information (and compare them against a stronger notion of regret), see for example \citep{garivier2008upper,Garivier2011a,Besbes2014}.
-
-from the information available to the agent, the joint distribution over counterfactuals cannot be inferred, unless the assignment mechanism in noiseless - so it does not make sense to define regret that depends on it.
-
-\begin{figure}[h]
-    \begin{subfigure}[t]{0.49\textwidth}
-	\centering   
-	 \caption{Behaviour over time}\label{fig:non-stationary-bandit-time} 
-         \includegraphics[scale=.5]{non-stationary-bandit-time}
-\end{subfigure}
-    \begin{subfigure}[t]{0.49\textwidth}
-    \centering
-     \caption{Cumulative regret distribution}
-        \label{fig:non-stationary-bandit-regret}
-        \includegraphics[scale=.5]{non-stationary-bandit-regret}
-    \end{subfigure}
-      \caption{The pseudo-regret can be negative if rewards are non-stationary.
-This example shows the results of 1000 simulations of running the UCB-1 algorithm on a 2-armed Bernoulli bandit problem where the expected rewards change linearly over time, up to a horizon $T=10,000$.
-Figure (a) shows the expected rewards of each arm, and the proportion of time that arm-1 is played, as a function of time.
-The single best-arm is arm-1 as it has the highest expected reward (averaged over $t$).
-An oracle that selects arm-1 in every round obtains an expected reward of $5,000$.
-However, despite not being designed to do so, the UCB-algorithm can adapt to the changing reward distribution to obtain consistently higher rewards.
-The distribution of regret over the 1000 simulations is shown in figure (b).}\label{fig:non-stationary-bandit}
-\end{figure}
-
-\subsection{Contextual bandits}
-\label{sec:contextual-bandits}
-
-In the standard multi-armed bandit setting, each decision is identical and the goal is to learn a single best action.
-However, in most real life (sequential) decision making processes, which action is optimal depends on some context.
-The best treatment to offer an individual patient could depend on their age, gender, disease sub-type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
-Similarly, decisions on which ad or content to display on a webpage or which product to recommend can be \emph{personalised} based on the previous behaviour of the user.
-A movie recommender system that learned a single \quotes{best} movie for everyone would not be very useful.
-Contextual bandits are a generalisation of multi-armed bandits that make use of this additional contextual information.
-The term contextual bandit was coined by \citet{Langford2008}, however close variants of the underlying problem have also been posed under the names; \quotes{associative reinforcement learning} \citep{kaelbling1994associative}, \quotes{bandits with concomitant variables}\citep{woodroofe1979one} and  \quotes{bandit problems with side information} \citep{wang2005bandit}.
-
-\vspace{0.5cm}
-\begin{definition}[Stochastic Contextual Bandit \footnote{Contextual bandits can also be defined in the adversarial setting analogously to definition \ref{def:adversarial-k-armed-bandit}}]
-\label{def:contextual-bandit-problem}
-Let $\P{\vec{x},\vec{y}}$ be the joint distribution over the rewards for each action and some context $\vec{X} \in \mathcal{X}$. In each round $t \in \set{1,...T}$, 
-\begin{enumerate}
-\item the world stochastically generates the vector of rewards for each action and the context, $(\vec{X}_t,[\cfb{Y}{1},...,\cfb{Y}{k}]) \sim \P{\vec{x},\vec{y}}$ and reveals $\vec{X}_t$ to the learner
-\item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps,
-\item the learner observes and receives (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
-\end{enumerate}
-\end{definition}
-
-Standard multi-armed bandits learn to select the action $a$ that, with high probability, maximises $\E{Y|a}$.
-Contextual bandits learn to select actions that maximise $\E{Y|\vec{x},a}$.
-The reward for contextual bandits should  be compared to an oracle that acts optimally based on the context.
-To achieve this, even when the context is continuous, the regret is defined with respect to a class of hypothesis that map from context to action, $h \in \mathcal{H}: \mathcal{X} \rightarrow \set{1,...,k}$.
-The pseudo-regret is the difference between the expected regret obtained by an oracle that selects actions based on the single best hypothesis or policy $h$ at each timestep, and the expected reward obtained by the algorithm.
-
-\eqn{
-\label{eqn:regret_contextual}
-\regret = \max_{h \in \mathcal{H}}\E{\sum_{t=1}^T{\cfb{Y}{h(\vec{X}_t)}}} - \E{\sum_{t=1}^T{\cfb{Y}{A_t}}}
-}
-
-If the context is discrete, $\mathcal{X} = \set{1,...,N}$, the contextual bandit problem can be reduced to the standard multi-armed bandit problem by creating a separate standard bandit instance for each value of the context.
-This approach results in a worst case regret of $\bigo{\sqrt{NkT}}$, with respect to the hypothesis class $\mathcal{H} = \mathcal{X} \times \actionspace$, consisting of all possible mappings from context to action\footnote{This follows from the fact that we have $N$ standard bandit instances, each suffering regret $\bigo{\sqrt{kT_c}}$, where $T_c$ is the number of times context $c$ occurred such that $\sum_{c=1}^N T_c = T$.
-The regret is maximised if $T_c = T/N$ resulting in total regret $\bigo{N\sqrt{kT/N}}$.}.
-This is optimal with respect to this class of hypothesis.
-However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the in the relationships between different contexts and actions.
-As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
-For example, with $n$ binary variables, $N = 2^n$}.
-We expect some form of smoothness; values of context that are similar should lead to comparable rewards for a given action.
-We need algorithms that can leverage such assumptions.
-
-An alternate reduction to the standard bandit problem, which allows us to constrain the hypothesis space to explore, is to treat each hypothesis $h$ as a bandit arm \citep{Langford2008}.
-At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
-Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothesis class considered, limiting our ability to learn any complex mappings from context to actions.
-The key problem with this approach is each sample is used to update our knowledge about only one hypothesis, as opposed to the supervised learning setting, where each data point is (implicitly) used to compute the loss for every hypothesis simultaneously.
-
-Suppose that, at each timestep $t$, after selecting an action, the learner received the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
-This is known as the full information setting.
-In this case, the learner can simulate running each hypothesis over the history to compute the reward it would have obtained and use the hypothesis with the best empirical reward to select the next action.
-This is the \emph{follow the leader} algorithm, which obtains optimal regret $\bigo{\sqrt{T log(\card{\mathcal{H}})}}$ for the full-information problem \citep{cesa2006prediction}.
-Unfortunately, in the contextual bandit problem, the (counterfactual) rewards associated with alternate action choices are not observed.
-As in causal effect estimation, we can view this as a missing data problem.
-However, the data is missing not at random because which component of the reward is observed depends on the action selected which in turn is a function of the previous history of actions and rewards.
-
-The Epoch-greedy algorithm, \citep{Langford2008}, addresses these issues by transforming the contextual bandit problem into a data missing at random problem by explicitly separating exploration from exploitation.
-Epoch-greedy is an explore-exploit algorithm.
-It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothesis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
-The hypothesis with the highest empirical reward is then used to select actions for the remaining time steps.
-The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
-
-The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
-However, it involves maintaining a list of weights for each hypothesis $h$ resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
-The ILOVECONBANDITS algorithm combines the best of both worlds to obtain a computationally efficient algorithm with (almost) optimal regret \citep{Agarwal2014}
-
-Both Epoch-greedy and ILOVECONBANDITS involve solving problems of the form,
-
-\eqn{
-\argmax_{h \in \mathcal{H}} \sum_{t=1}^\tau Y_t \ind{h(\vec{X}_t) = A_t}
-}
-
-This expression equates to identifying the best empirical policy based on previous data.
-The algorithms assume the existence of an oracle that can solve this problem and report complexity in terms of the number of calls required to the oracle.
-The computational tractability of these algorithms on large (or infinite) hypothesis spaces stems from the fact that this problem (also known as the argmax-oracle), can be reduced to solving a cost sensitive classification problem \citep{Dudik2011a}.
-
-Finally, if we have a parametric model for the relationship between context, action and reward that allows (efficient) computation of the posterior or confidence bounds on the reward for each arm given context, we can develop generalised versions of the UCB or Thompson sampling algorithms.
-For linear pay-off models, both approaches yield algorithms with strong regret guarantees, \emph{Lin-UCB} \citep{li2010contextual} and \emph{Generalised Thompson Sampling} \citep{Agrawal2013}.
-
-\subsubsection{The causal structure of contextual bandits}
-\label{sec:causal-structure-contextual-bandit}
-
-It is worth noting that definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$, see figure \ref{fig:contextual-bandit-causal-graph}.
-However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
-\citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information.
-This is an example of the case represented in figure \ref{fig:context-graph-confounded}.
-It is even possible to have (useful) context $\vec{X}$ that is a consequence of $\vec{Y}$ as in figure \ref{fig:context-graph-anti-causal} and example \ref{exm:anti-casual-context}.
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ causes $\vec{Y}$}
-\label{fig:context-graph-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (4) edge (1)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{X}$ and $\vec{Y}$ are confounded}
-\label{fig:context-graph-confounded}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
-\node[lt, below=of 4](5){$U$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2)
-   (5) edge (4) edge (1);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.29\textwidth}
-\centering
-\caption{$\vec{Y}$ causes $\vec{X}$}
-\label{fig:context-graph-anti-causal}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$\vec{Y}$};
-\node[main node, below=of 1](2){$\vec{Y}^a$};
-\node[main node, above right=of 2](3){$A$};
-\node[main node, left=of 1](4){$\vec{X}$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (4)
-   (3) edge (2)
-   (1) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{Several potential causal graphical models for the contextual bandit problem (if actions are selected at random).
-$\vec{X}$ and $\vec{Y}$ represent the context and reward vectors respectively, and $Y = \vec{Y}^a$ is the reward received by the learner, which is a deterministic function of $\vec{Y}$ and $a$.
-Regardless of the causal structure, observing the context $\vec{X}$ provides information about the vector of rewards $\vec{Y}$.
-The causal structures in figures \ref{fig:context-graph-causal} and \ref{fig:context-graph-confounded} also make sense if the elements of $\vec{Y}$ are genuine counterfactuals, that is only the reward of the selected action is ever realised.
-The context $\vec{X}$ can only be useful in causal graphs of the form in figure \ref{fig:context-graph-anti-causal} if the components of $\vec{Y}$ are realised prior to the agent selecting an action, i.e when the components of $\vec{Y}$ are not truly counterfactual, it just that the agent is limited to only observing the outcome of the action they choose.
-To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
-}
-\label{fig:contextual-bandit-causal-graph}
-\end{figure}
-
-\vspace{0.5cm}
-\begin{example}
-\label{exm:anti-casual-context}
-Imagine a modified game of roulette.
-As in standard roulette, it consists of a spinning wheel with a ball that can land in one of 37 numbered pockets.
-However, unlike in the standard game, where punters place bets before the wheel is spun, in this variant the wheel is spun first.
-Five seconds before it comes to rest the face of the wheel darkens to hide the position of the ball from the players.
-They then place bets after it has stopped (and are each allowed only to check if the position they have bet on was correct).
-At the point in time the players are selecting their actions, the rewards for each possible choice are fully determined (all zero except for the pocket containing the ball).
-Suppose a canny gambler realises the ball makes a tiny sound, from rocking back and forth, if it is close to the left or right side of the wheel such that the pocket holding it is close to horizontal.
-The existence (or not) of this sound is a consequence of the position of the ball and thus the reward state and the gambler can use it as context to improve the likelyhood of obtaining a reward.
-\end{example}
-
-It is difficult to come up with realistic examples in which the full sequence of rewards is generated but only the one associated with the selected action is observable to the learner.
-In the typically cited applications of contextual bandit algorithms, such as selecting how to treat patients or which ad to serve only the reward of the selected action is ever realised, and the alternate rewards are counterfactual variables.
-
-\subsection{Learning from logged bandit data}
-\label{sec:off-policy-evaluation}
-Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
-In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
-The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
-
-This problem differs from the contextual bandit problem in that the learner is not interacting with the environment.
-As a result, there is no exploration-exploitation trade-off to be made.
-However, the problem does not reduce to supervised learning, because of the bandit nature of the feedback - only the reward of the selected action is observed.
-In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
-The majority of the literature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
-\citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original estimator $\pi$ did not depend on $\vec{X}$.
-
-If the original policy is assumed to be stationary, the problem of evaluating an alternate policy $\pi'$ is almost identical to that of causal effect estimation under ignorability, discussed in section \ref{sec:causal_effect_under_ignorability}.
-The causal structure can be represented in figure \ref{fig:off-policy-evaluation} There is an (implicit) assumption that all variables that impact the choice of action by $\pi$ are included in $\vec{X}$, ensuring that $\vec{X}$ satisfies the backdoor criterion with respect to identifying the causal effect of $do(A = a)$ on the observed reward $Y$, for any action $a \in \set{1,...,k}$.
-The only difference is that the goal is to evaluate alternate policies $\pi'$ that may be stochastic and depend on $\vec{x}$, as opposed to only policies of the form $\pi'(x) = a$, equivalent to $do(A=a)$.
-However, the identification of such stochastic, conditional policies can be reduced to the identification of $\P{y|do(A=a),\vec{x}}$, see \citet{Pearl2000}, section 4.2.
-In this case, letting $\Pns{\pi'}{a|\vec{x}}$ denote the distribution over actions under policy $\pi'$ given context $\vec{x}$, the expected (per round) reward obtained by $\pi'$ is given by,
-
-\eqn{
-\E{y|\pi'} = \E{y|do(a \sim \pi'(\vec{x}))} = \Esub{(\vec{x},a) \sim \P{\vec{x}}\Pns{\pi'}{a|\vec{x}}}{y|\vec{x},a}
-}
-
-As in estimating average causal effects under ignorability, we have a covariate shift problem, with training data sampled from $\P{\vec{x}}\Pns{\pi}{a|\vec{x}}\P{y|\vec{x},a}$ but generalisation error measured with respect to $\P{\vec{x}}\Pns{\pi'}{a|\vec{x}}\P{y|\vec{x},a}$.
-A difference in practice, is that in the applications frequently considered under learning from logged feedback data, such as ad serving or recommender systems, there may be substantial information available about $\pi$, in the best case, $\Pns{\pi}{a|\vec{x}}$ is known.
-This makes estimators utilising inverse propensity weighting, including doubly robust estimators as in \citet{Dudik2011}, more attractive.
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
-\node[main node](1){$\vec{X}$};
-\node[main node, below right=of 1](2){$Y$};
-\node[main node,  below left=of 1](3){$A$};
-\node[lt, above=of 2](4){$U$};
- \path[every node/.style={font=\sffamily\small}]
-   (3) edge (2)
-   (1) edge (2) edge (3)
-   (4) edge (1) edge (2);
-\end{tikzpicture} 
-\caption{Causal graphical model for learning from logged feedback data under the assumption the original policy $\pi$ for selecting actions was stationary and dependent only on some observed context $\vec{X}$, $a \sim \pi(\vec{x})$.
-The outcome $Y$ may depend on $\vec{X}$ directly, indirectly through a latent variable $U$ or both.}
-\label{fig:off-policy-evaluation}
-\end{figure}
-
-\citet{Swaminathan2015} point out that the problem of identifying the optimal policy (subject to some risk minimisation goal) is not as simple as estimating the expected reward associated with each policy in some space and selecting the empirical best because the variance of the estimators for some policies may be much higher than for others.
-
-\subsection{Adding structure to actions}
-\label{sec:adding-structure-bandit-actions}
-
-The classic multi-armed bandit is a powerful tool for sequential decision making.
-However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \citep{filippi2010parametric}, dependent bandits \citep{Pandey2007}, X-armed bandits \citep{Bubeck2010} and Gaussian process bandits \citep{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback\citep{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \citep{Piccolboni2001,Bartok2014}.
-
-In the next chapter, I propose a very natural connection between causal graphs and bandit problems and show it induces a novel form of additional feedback and structure between arms that cannot be can exploited by any of these previous approaches.
-
-\section{The counterfactual nature of regret}
-\label{sec:counterfactual-nature-regret}
-
-We conclude this chapter on bandits with a note on the counterfactual nature of regret.
-I have not seen this discussed with respect to bandit regret elsewhere, however the issues raised are very closely related to the work by \citet{Dawid2000}, on problems associated with the use of counterfactual variables for (observational) causal inference.
-
-It is possible, as we demonstrated in example \ref{exm:anti-casual-context}, to construct bandit problems in which the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$ is actually generated at each time-step (and, at least in principle, is observable to someone although not the agent).
-However, in most real applications for bandit problems, only the outcome associated with the selected action is ever realised and so the remaining components of the reward vector are counterfactual, justifying our use of the counterfactual notation to denote them.
-There are complex philosophical objections to counterfactuals arising from the way they describe alternate universes that were never realised.
-This makes it easy to make statements using counterfactuals that cannot be confirmed empirically (even with infinite experimental data).
-We now show with a simple example that the standard definition of regret is such a fundamentally counterfactual quantity.
-Recall that the cumulative regret is defined by,
-
-\eqn{
-\label{eqn:raw_regret2}
-\rawregret &= \sum_{t=1}^T \cfb{Y}{i^*} - \sum_{t=1}^T \cfb{Y}{a_t}
-\\&= \sum_{t=1}^T \left(\cfb{Y}{i^*} - \cfb{Y}{a_t}\right)
-}
-
-Take a stochastic, two-armed, Gaussian bandit with the joint distribution of the (counterfactual) rewards $\P{\cfb{Y}{1},\cfb{Y}{2}}$ given by equation \ref{eqn:joint-gaussain-reward}.
-Suppose without loss of generality that arm 1 is the optimal arm, such that $\mu_1 > \mu_2$.
-
-% vector with mean mu_t,mu_c and a covariance matrix
-\eqn{
-\label{eqn:joint-gaussain-reward}
-P(\cfb{Y}{1},\cfb{Y}{2}) \sim N( 
-         \begin{bmatrix}
-           \mu_{1} \\
-           \mu_2 \\
-         \end{bmatrix},
-         \begin{bmatrix}
-           \sigma^2 & \rho \sigma^2 \\
-           \rho \sigma^2 & \sigma^2 \\
-         \end{bmatrix}
-         )
-}
-
-Consider an algorithm that always selects arm 2.
-The distribution over the difference of jointly normal random variables is also normal, letting, $\tau_t = \cfb{Y}{1} - \cfb{Y}{2}$ yields $\P{\tau_t} = N(\mu_1 - \mu_2,2\sigma^{2}(1-\rho))$.
-Thus the distribution over the regret for this algorithm is given by,
-
-\eqn{
-\rawregret \sim N\left(T(\mu_1 - \mu_2), 2\sigma^2 T(1-\rho)\right)
-}
-
-The parameters of the marginal distributions, $\P{\cfb{Y}{1}}$ and $\P{\cfb{Y}{2}}$, can be estimated directly by simply sampling from each arm.
-However, the covariance $\rho$ cannot, because we never simultaneously observe the rewards for both arms.
-As a result, even if we have full knowledge of the reward distributions for each arm, we cannot in general compute the distribution over regret without un-testable assumptions about the covariance between counterfactuals, see figure \ref{fig:metaphysical_distribution_difference}.
-
-\begin{figure}[h]
-\captionsetup[subfigure]{position=b}
-\centering
-\subcaptionbox{Marginal distributions over the rewards for each action for $\mu_1=1$, $\mu_2=0$ and $\sigma = 1$.
-\label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
-\hspace{0.02\textwidth}
-\subcaptionbox{The distribution over $R_T$ for $T=10$ for two different values of $\rho$ \label{fig:metaphys_b}}
-{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
-\caption{The distribution over the regret as defined by equation \ref{eqn:raw_regret2} depends on unobservable properties of the joint distribution over counterfactual rewards.
-The same (marginal) distributions over the rewards can correspond to quite different regret distributions.}
-\label{fig:metaphysical_distribution_difference}
-\end{figure}
-
-This result is somewhat perturbing.
-The stochastic bandit problem can be defined without recourse to counterfactual variables by having the world stochastically generate only a reward for the selected action at each timestep and the behaviour of standard bandit algorithms depends only on the marginal reward distributions for each arm.
-The expectation of the regret as defined by equation \ref{eqn:pseudo_regret} also remains unchanged as both it's definition and the learner's actions depend only on the marginal distributions.
-It seems therefor unfortunate that we should have to assume, for example, that the rewards for alternate actions are independent of one-another in order to be able to analyse the variance of the regret.
-Particularly as this assumption is likely to be violated in many realistic bandit problems.
-For example, a given user may be more (or less) likely to buy something no matter which ad they are served leading to a positive correlation between counterfactual rewards, or an illness might have two (unobservable) sub-types with each medication (bandit arm) effective only against one, resulting in a negative correlation between the counterfactual rewards.
-
-We could focus only on analysing the expected regret, since this depends only on the marginal distributions.
-However, there are many real problems for which we do care how tightly concentrated the regret is around its expectation.
-For example, if we are risk averse, we may prefer an algorithm with slightly higher expected regret but a lower probability of suffering extremely large regret.
-This raises the question, can we construct an alternate definition of regret that allows us to capture how consistently bandit algorithms behave, but which does not depend on any properties of the joint distribution over counterfactual rewards.
-A natural candidate would be,
-
-\eqn{
-R_T = \sum_{i=1}^k N_i(T)\Delta_i, 
-}
-
-where $N_i(T)$ is the number of times arm $i$ was played up to timestep $T$ and $\Delta_i$ is the degree to which arm $i$ is sub-optimal, $\Delta_i = \mu^* - \mu_i$.
-The expectation of this variant of regret is the same as for the standard version defined in equation \ref{eqn:raw_regret2}.
-It depends only on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions.
-Furthermore, this quantity has already been analysed in existing work on the concentration of bandit regret, \citep{Audibert2007,Audibert2009exploration} as a more tractable proxy to the standard regret.
-In conclusion, when selecting measures of bandit performance, it is worth noting whether or not they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.
-
-\chapter{Causal Bandits: Unifying the approaches}
-\label{chap:causal-bandit}
-
-As we have seen in the previous two chapters, critical insights and methods for learning how to act have emerged from both the observational and experimental approaches to causality.
-From observational inference, we have the do-calculus, which allows us to map information between observational and interventional distributions and we have a range of estimators designed for these types of mapping problems including propensity scores and doubly robust methods.
-The bandit community has developed carefully tuned algorithms to balance exploration and exploitation, estimation approaches for identifying and selecting the optimal action (rather than estimating the rewards for all actions) and techniques to bound the bias of estimators due to the adaptive nature of decision making processes.
-
-However, despite the commonality of the underlying problem, there has been relatively little work on intersection between observational causal inference and adaptive decision making processes.
-Although some ideas, such as doubly robust estimators and inverse propensity score weighting have transferred across (from causal inference to off-policy evaluation), there are many more opportunities for fruitful work in this space.
-In most key real world decision making processes, from algorithms recommending drug treatments to robots deciding how to open doors, we have access to both potentially huge observational data sets as well as the ability to intervene in the system at some level.
-It is critical that we develop methods that allow us to incorporate the causal knowledge we have about the world, along with data sets collected under a range of different conditions, into our current decision making process.
-In this chapter I introduce a very general framework that connects causal graphical models with multi-armed bandit problems and demonstrate how it can be leveraged to make better decisions.
-
-\section{A framework to connect bandit problems with observational causal inference}
-\label{sec:intro_causal_bandit}
-A natural way to connect the causal framework with the bandit setting is to model the action space as interventions on variables in a causal directed acyclic graph.
-Each possible assignment of variables to values is a potential action (or bandit arm), see figure \ref{fig:unify_frameworks} for a simple example.
-In some settings, it makes sense to restrict the action space available to the agent to a subset of all the possible actions, for example the set of single variable interventions.
-The reward could be a general function of the action selected and the final state of the graph.
-However for simplicity, we will consider the reward to be the value of a single specified node.
-We refer to these problems as \emph{causal bandit problems}.
-In this thesis I focus on the case where the causal graph is known.
-Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.
-
-\begin{figure}
-\centering
-\begin{subfigure}[c]{0.3\textwidth}
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=3em,thick,main node/.style={observed}, tg/.style={target}]
-\node[main node](1){W};
-\node[tg, below right=of 1](2){Y};
-\node[main node, above right=of 2](3){X};
-\path[]
-	(1) edge (2)
-	(3) edge (2);
-\end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}[t]{0.4\textwidth}
-$\actionspace$ = \begin{tabular}{|c|}
-	\hline
-  $do(W=0,Z=0)$ \\
-  $do(W=0,Z=1)$ \\
-  $do(W=1,Z=0)$ \\
-  $do(W=1,Z=1)$ \\
-  \hline
-  $do(W=0)$ \\
-  $do(W=1)$ \\
-  $do(Z=0)$ \\
-  $do(Z=1)$ \\
-  $do()$ \\
-  \hline
-\end{tabular}
-\end{subfigure}
-\caption{A simple causal graphical model and corresponding complete action space. $W$ and $Z$ represent binary variables that can be intervened on and $Y$ represents the reward.}
-\label{fig:unify_frameworks}
-\end{figure}
-
-The type of problem we are concerned with is best illustrated with an example.
-Consider a farmer wishing to optimise the yield of her crop.
-She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown.
-In each season the farmer has enough time and money to intervene and control at most one of these
-variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry.
-When not intervened upon, the temperature, soil, and moisture vary naturally from season to season due to weather conditions and these are all observed along with the final crop yield at the end of each season.
-How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?
-
-We will assume each variable only takes on a finite number of distinct values.
-(The path to relaxing this assumption would be through levering the work on continuous armed bandits).
-The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
-An \emph{intervention or action (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
-We denote the expected reward for the action $a = do(\vec{X} = \vec{x})$ by $\mu_{a} := \E{Y | do(\vec{X} = \vec{x})}$ and
-the optimal expected reward by $\mu^* := \max_{a\in\actions} \mu_{a}$.
-
-\vspace*{.5cm}
-\begin{definition}[Causal bandit problem]
-A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
-Each action $a \in \actionspace$ assigns a value to a subset of the variables in $\mathcal{X}$ and incurs a known cost $C(a)$ on the learner.
-One variable $Y \in \mathcal{X}$ is designated as the \emph{reward variable}.
-
-The causal bandit game proceeds over $T$ rounds. In each round $t$, the learner:
-\begin{enumerate}
-\item \emph{observes} the value of a subset of the variables $\vec{X}^c_t$,
-\item \emph{intervenes} by choosing $a_t = do(\vec{X}_t = \vec{x}_t) \in \mathcal{A}$ based on previous observations and rewards, 
-\item \emph{observes} values for another subset of variables $\vec{X}^o_t$ drawn from $\P{\vec{X}^o_t |\vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$,
-\item \emph{obtains reward} $r_t = Y_t - C(a_t)$, where $Y_t$ is sampled from $\P{Y_t| \vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$
-\end{enumerate} 
-\end{definition}
-
-We refer to the set of variables that can be observed prior to selecting an action $\vec{X}^c$ as contextual variables and the set of variables observed after the action is chosen, $\vec{X}^o$, as post-action feedback variables.
-Note that $\vec{X}^c$ and $\vec{X}^o$ need not be disjoint.
-A variable may be observed both prior to and after the agent selects an action, and the action may change its value.
-The notation $\P{\cdot|\vec{X}^c_t, do(\vec{X}_t = \vec{x}_t)}$ denotes distributions conditional on having observed $\vec{X}^c_t$ and \emph{then} intervened to set $\vec{X}_t = \vec{x}_t$.
-The values of variables in $\vec{X}^c_t$ that are non-descendents of $\vec{X}_t$ remain unchanged by the intervention.
-The objective of the learner is to minimise either the simple (equation \ref{}) or cumulative regret (equation \ref{}).
-
-The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action and on which variables we receive post-action feedback.
-If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant and the problem reduces to a standard multi-armed bandit problem.
-This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.
-
-\vspace{0.5cm}
-\begin{theorem}
-\label{theorem:cb-intervene-all-parents}
-Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$.
-If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and the problem reduces to a standard multi-armed bandit (over actions in in $\mathcal{A'}$).
-\begin{proof}
-for any action $a \in \mathcal{A}$, 
-\eq{
- \E{Y|\vec{X}^c_t,a} &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\vec{X}^c_t,a,\parents{Y}}} \\
- &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\parents{Y}}}\\
- &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|do(\parents{Y})}}\\
- &\leq \max_{\parents{Y}}\E{Y|do(\parents{Y})} = \E{Y|a'} \text{ for some $a' \in \mathcal{A'}$}
-}
-\end{proof}
-\end{theorem}
-
-If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
-If we receive post-action feedback on additional nodes the problem can be more interesting.
-In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
-Consider again the model in figure \ref{fig:unify_frameworks}.
-The causal structure implies:
-
-\eq{
-P(Y|do(W = 0)) &= P(Y|do(),W = 0) \\
-&= P(Y|do(X=0),W=0)P(X=0)+P(Y|do(X=1),W=0)P(X=1) 
-}
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},tg/.style={targetrect}]
-\node[node](1){Health conscious};
-\node[node, below left=of 1](2){Diet};
-\node[node, below right=of 1](3){Exercise};
-\node[node, below right=of 2](4){Weight};
-\node[tg,below left=of 4](5){Y=Cholesterol};
-\node[node, below =of 5](6){Test result};
-\node[node, above right=of 3](7){Free time};
-\path[]
-	(1) edge (2) edge (3)
-	(2) edge (4) edge (5)
-	(3) edge (4)
-	(5) edge (6)
-	(7) edge (3);
-\end{tikzpicture}
-\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level .
-The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
-For example, no actions involving 'Test Result' need to be considered and interventions on 'Diet' do not need to be considered in conjunction with any other variables.}
-\label{fig:cholesterol_graph}
-\end{figure}
-
-Thus we gain information about the reward for the action $do(W=0)$ from selecting the action $do()$ or $do(X = x)$ and then observing $W = 0$.
-We only get this form of side information for actions that don't specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-If additional variables are only observed before an intervention is selected the causal bandit problem can be treated as a stochastic contextual bandit problem, which are already reasonably well understood \citep{Agarwal2014}.
-
-Two other problems that sit in the space between causal inference and bandit problems, bandits with unobserved confounders \cite{Bareinboim2015} and compliance aware bandits \citep{DellaPenna2016}, can also be viewed as specific causal bandit problems.
-In the work on bandits with unobserved confounders it is assumed that the reward given each action may depend on some latent variable $U$ which we cannot observe directly.
-However, prior to selecting an action, we can observe  $I$, the action that would have been selected under an alternate (stationary) policy, which may depend on $U$, see figure \ref{fig:confounded-causal-bandit-graph}.
-In this case the set of contextual variables $\vec{X}^c = I$, the set of post-action feedback variables $\vec{X}^o = \set{Y}$ and the action space consists of all possible assignments of values to a single node $X$, $\actionspace = do(X=x)$.
-This setting reduces to a contextual bandit problem in our causal bandit framework.
-However in their work on bandits with unobserved confounders \citet{Forney2017} also leverage the fact that $I$ represents the action that would have been selected under an alternate policy to fuse data collected under the previous (observational) policy with data collected under the new policy.
-This information is not encoded in the causal bandit graph in figure \ref{fig:confounded-causal-bandit-graph}, as $I$ could be any variable that is influenced by the unobserved context $U$, and thus cannot be exploited by a standard contextual bandit algorithm.
-
-Compliance aware bandits describe situations in which the action recommended by the bandit algorithm is not always followed.
-For example a patient may refuse to take a treatment or an advertiser may have complex rules about how many ads a given customer can receive which prevents some of the suggestions from the ad recommendation engine from being followed.
-After an action is selected the algorithm can observe the action that was actually taken in addition to the reward.
-\citet{DellaPenna2016} analyse this setting with binary treatments both with and without the presence of a latent confounding variable $U$, see figure \ref{fig:imperfect-compliance-causal-bandit-graph}.
-In this case, there are no contextual variables and the action space is again the set of assignments to a single variable but there is post-action feedback, which reveals the value of the action that was actually taken.\footnote{There are some interesting variants of the compliance aware bandit setting that, to my knowledge, have not been analysed.
-The first is if the confounding variable $U$ is observable, either as context or post-action feedback.
-The second is if we extend the allowable action set to include acting directly on $X$, albeit at a higher cost than acting on $A$.
-It is also worth noting the connection between this setting and instrumental variables \cite{balke1997bounds}.
-By making some functional assumptions about the relationships between the variables, we can use $A$ as an instrumental variable to bound or estimate the (casual) effect of $X$ on $Y$.
-The estimation will be somewhat complicated in the bandit setting because the action chosen at each timestep is dependent on the previous sequence actions and rewards.}
-
-\begin{figure}[h]
-\centering
-\begin{subfigure}[t]{0.49\textwidth}
-\centering
-\caption{Bandits with unobserved confounders:$\vec{X}^c=\set{I}$, $\vec{X}^o = \set{Y}$, $\actionspace = do(X=x)$}
-\label{fig:confounded-causal-bandit-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){${U}$};
-\node[main node, below right =of 1](2){${Y}$};
-\node[main node, below left=of 1](3){$I$};
-\node[main node, above right=of 2](4){$X$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (2) edge (3)
-   (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\begin{subfigure}[t]{0.49\textwidth}
-\centering
-\caption{Compliance aware bandits: $\vec{X}^c=\set{}$, $\vec{X}^o = \set{T,Y}$, $\actionspace = do(X=x)$ }
-\label{fig:imperfect-compliance-causal-bandit-graph}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
-  thick,main node/.style={observed},lt/.style={latent}]
- %nodes
-\node[lt](1){$U$};
-\node[main node, below left=of 1](2){$T$};
-\node[main node, below right=of 1](3){$Y$};
-\node[main node, above left=of 2](4){$X$};
- \path[every node/.style={font=\sffamily\small}]
-   (1) edge (2) edge (3)
-   (2) edge (3)
-   (4) edge (2);
-\end{tikzpicture} 
-\end{subfigure}
-\caption{}
-\label{fig:causal-bandit-version-confounded-compliance}
-\end{figure}
-
-The classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
-The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
-Conversely, any causal bandit problem can be reduced to a classical stochastic $|\mathcal{A}|$-armed bandit problem by treating each possible intervention as an independent arm and ignoring all sampled values for the observed variables except for the reward.
-However, the number of actions or arms grows exponentially with the number of variables in the graph making it important to develop algorithms that leverage the graph structure and additional observations.
-
-\section{Causal bandits with post action feedback}
-\label{sec:causal-bandit-post-action-feedback}
-
-We now focus on causal bandit problems with post-action feedback, in which the value of all the variables are observed after an intervention is selected, the cost of all allowable actions is equal and the goal of the learner is to minimise the simple regret.
-I presented this work at NIPS 2016 \citep{NIPS2016_6195}.
-
-\paragraph{Related Work} As alluded to above, causal bandit problems can be treated as classical multi-armed bandit problems by simply ignoring the causal model and extra observations and applying an existing best-arm identification algorithm with well understood simple regret guarantees \citep{Jamieson2013}.
-However, as we show in \S\ref{sec:simple-regret}, ignoring the extra information available in the non-intervened variables yields sub-optimal performance.
-
-Our framework bears a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
-However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
-
-There have been several proposals for bandit problems where extra feedback is received after an action is taken.
-Most recently, \citet{Alon2015}, \citet{Kocak2014} have considered very general models related to partial monitoring games~\citep{Bartok2014} where rewards on un-played actions are revealed according to a feedback graph.
-As we discuss in \S\ref{sec:causal_bandit_discussion}, the parallel bandit problem can be captured in this framework, however the regret bounds are not optimal in our setting.
-They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
-The partial monitoring approach taken by \citet{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy would need to know the likelihood of each factor in advance, while our strategy learns this online.
-\citet{Yu2009} utilise extra observations to detect changes in the reward distribution, whereas we assume fixed reward distributions and use extra observations to improve arm selection.
-\citet{Avner2012} analyse bandit problems where the choice of arm to pull and arm to receive feedback on are decoupled.
-The main difference from our present work is our focus on simple regret and the more complex information linking rewards for different arms via causal graphs.
-To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.
-
-%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
-They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
-% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
-It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \citep{Bartok2014}.
-Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
-
-Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
-\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been 
-intervened on can provide important contextual information. Their work differs in many ways. For example, the focus is on the cumulative regret and
-the context is observed before the action is taken and cannot be controlled by the learning agent.
-
-%\todof{write one sentence to explain how their work is different to ours}
-
- \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling assuming actions are causal interventions.
-Their focus is on causal induction (\ie, learning an unknown causal model) instead of exploiting a known causal model.
-Combining their handling of  causal induction with our analysis is left as future work.
-% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
-Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
-(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
-It is easy to come up with a counter example in which it does not).
-
-The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{Bottou2013}, 
-where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discuss some of the issues 
-encountered in sequential design, but do not give an algorithm or theoretical results for this case.
-
-\subsection{The parallel bandit problem}
-\label{sec:simple-regret}
-In this section we propose and analyse an algorithm for achieving the optimal regret in a natural special case of the causal bandit problem which we call the {\it parallel bandit}.
-It is simple enough to admit a thorough analysis but rich enough to model the type of problem discussed in \S\ref{sec:intro_causal_bandit}, including the farming example.
-It also suffices to witness the regret gap between algorithms that make use of causal models and those which do not.
-
-The causal model for this class of problems has $N$ binary variables $\{ X_1, \ldots, X_N \}$ where each $X_i \in \{0,1\}$ are independent causes of a reward variable $Y \in \set{0,1}$, as shown in Figure~\ref{fig:parallel}.
-All variables are observable and the set of allowable actions are all size 0 and size 1 interventions: $\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}$
-
-In the farming example from the introduction, $X_1$ might represent temperature (\eg, $X_1=0$ for low and $X_1=1$ for high).
-The interventions $do(X_1 = 0)$ and $do(X_1 = 1)$ indicate the use of shades or heat lamps to keep the temperature low or high, respectively.
-
-\begin{figure}
-    \begin{subfigure}[t]{0.34\textwidth}
-	\centering   
-	 \caption{Parallel graph}\label{fig:parallel} 
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5);
-\end{tikzpicture} 
-\end{subfigure}
-    \begin{subfigure}[t]{0.25\textwidth}
-    \centering
-     \caption{Confounded graph}
-        \label{fig:causalStructure_confounded}
-        \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_1$};
-\node[main node, below left=of 1](2){$X_2$};
-\node[main node, below right=of 1](4){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-    (1) edge (4)
-    (2) edge (4);
-\end{tikzpicture}
-    \end{subfigure}
-    \begin{subfigure}[t]{0.4\textwidth}
-    \centering
-     \caption{Chain graph}
-        \label{fig:causalchain}
-         \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, right=of 4](5){$Y$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (2)
-  	(2) edge (3)
-    (3) edge (4)
-    (4) edge (5);
-\end{tikzpicture}
-    \end{subfigure}
-      \caption{Causal Models}\label{fig:causalmodels}\label{fig:causal_models_causal_bandit}
-\end{figure}
-
-In each round the learner either purely observes by selecting $do()$ or sets the value of a single variable.
-The remaining variables are simultaneously set by independently biased coin flips.
-The value of all variables are then used to determine the distribution of rewards for that round.
-Formally, when not intervened upon we assume that each $X_i \sim \bernoulli(q_i)$ where $\vec{q} = (q_1, \ldots, q_N) \in [0,1]^N$ so that $q_i = \P{X_i = 1}$.
-
-The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
-$r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
-In the farming example, this choice of $Y$ models the success or failure of a seasons crop, 
-which depends stochastically on the various environment variables.
-
-\paragraph{The Parallel Bandit Algorithm}
-%\label{sub:par-bandit-alg}
-The algorithm operates as follows.
-For the first $T/2$ rounds it chooses $do()$ to collect observational data.
-As the only link from each $X_1,\ldots,X_N$ to $Y$ is a direct, causal one, $\P{Y|do(X_i=j)}=\P{Y|X_i=j}$.
-Thus we can create good estimators for the returns of the actions $do(X_i = j)$ for which $\P{X_i = j}$ is large.
-The actions for which $\P{X_i = j}$ is small may not be observed (often) so  estimates of their returns could be poor.
-To address this, the remaining $T/2$ rounds are evenly split to estimate the rewards for these infrequently observed actions.
-The difficulty of the problem depends on $\vec{q}$ and, in particular, how many of the variables are unbalanced (\ie, small $q_i$ or $(1-q_i)$).
-For $\tau \in [2...N]$ let $I_\tau = \set{ i : \min\set{q_i, 1-q_i} < \frac{1}{\tau}}$.
-Define
-
-\eq{
-\label{eq:m-simple}
-m(\vec{q}) = \min \set{ \tau : |I_{\tau}| \leq \tau}\,.
-}
-
-\begin{algorithm}[H]
-\caption{Parallel Bandit Algorithm}\label{alg:simple}
-\begin{algorithmic}[1]
-\STATE {\bf Input:} Total rounds $T$ and $N$.
-\FOR{$t \in 1,\ldots,T / 2$}
-\STATE Perform empty intervention $do()$
-\STATE Observe $\vec{X}_t$ and $Y_t$
-\ENDFOR
-\FOR{$a = do(X_i = x) \in \actions$}
-\STATE Count times $X_i = x$ seen: $T_a = \sum_{t=1}^{T/2} \ind{X_{t,i} = x}$
-\STATE Estimate reward: $\hat{\mu}_a = \frac{1}{T_a} \sum_{t=1}^{T/2} \ind{X_{t,i} = x} Y_t$ \\[0.2cm]
-\STATE Estimate probabilities: $\hat{p}_a = \frac{2 T_a}{T}$,\,\, $\hat q_i = \hat p_{do(X_i = 1)}$
-\ENDFOR
-\STATE Compute $\hat{m} = m(\vec{\hat q})$ and $A = \set{a \in \actions \colon \hat{p}_a \leq \frac{1}{\hat m}}$.
-\STATE Let $T_A := \frac{T}{2 |A|}$ be times to sample each $a\in A$.
-\FOR{$a = do(X_i = x) \in A$}
-\FOR{$t \in 1,\ldots,T_A$}
-\STATE Intervene with $a$ and observe $Y_t$
-\ENDFOR
-\STATE Re-estimate $\hat{\mu}_a = \frac{1}{T_A} \sum_{t=1}^{T_A} Y_t$
-\ENDFOR
-\RETURN estimated optimal $\hat{a}^*_T \in \argmax_{a\in\actions} \hat{\mu}_a$
-\end{algorithmic}
-\end{algorithm}
-
-$I_\tau$ is the set of variables considered unbalanced and we tune $\tau$ to trade off identifying the low probability actions against not having too many of them, so as to minimise the worst-case simple regret.
-When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 2$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$.
-We do not assume that $\vec{q}$ is known, thus Algorithm \ref{alg:simple} also utilises the samples captured during the observational phase to estimate $m(\vec{q})$.
-Although very simple, the following two theorems show that this algorithm is effectively optimal.
-
-\begin{theorem}\label{thm:uq-simple}
-Algorithm \ref{alg:simple} satisfies
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\vec{q})}{T}\log\left(\frac{NT}{m(\vec{q})}\right)}}\,.
-}
-\end{theorem}
-
-\begin{theorem}\label{thm:lower}
-For all strategies and $T$, $\vec{q}$, there exist rewards such that
-$\displaystyle \simpleregret 
-\in \Omega\left(\sqrt{\frac{m(\vec{q})}{T}}\right)$.
-\end{theorem}
-
-The proofs of Theorems \ref{thm:uq-simple} and \ref{thm:lower} follow by carefully analysing the concentration
-of $\hat p_a$ and $\hat m$ about their true values and may be found in Sections \ref{sec:thm:uq-simple} and \ref{sec:thm:lower} respectively.
-
-%We prove a lower bound on the simple regret that matches up to logarithmic factors the upper bound given in Theorem \ref{thm:uq-simple}. 
-By utilising knowledge of the causal structure, Algorithm \ref{alg:simple} effectively only has to explore the $m(\vec{q})$ 'difficult' actions.
-Standard multi-armed bandit algorithms must explore all $2N$ actions and thus achieve regret  $\smash{\Omega(\sqrt{N/T})}$.
-Since $m$ is typically much smaller than $N$, the new algorithm can significantly outperform classical bandit algorithms in this setting.
-In practice, you would combine the data from both phases to estimate rewards for the low probability actions.
-We do not do so here as it slightly complicates the proofs and does not improve the worst case regret.
-
-\subsection{General graphs}
-\label{sec:simple-regret-general}
-We now consider the more general problem where the graph structure is known, but arbitrary.
-For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorisation formula \citep{Pearl2000}.
-\eq{
-\P{X_1...X_N|do(X_i=j)} = 
-\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
-} 
-where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the Dirac delta function.
-
-We could naively generalise our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorisation to 
-write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
-estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
-from rounds in which we act on a different variable. Consider the graph in Figure \ref{fig:causalchain} and suppose each variable deterministically 
-takes the value of its parent, $X_k = X_{k-1}$ for $k\in {2,\ldots,N}$ and $\P{X_1} = 0$. We can learn the reward for all the interventions $do(X_i = 1)$ 
-simultaneously by selecting $do(X_1 = 1)$, but not from $do()$. In addition, variance of the observational estimator for $a = do(X_i = j)$ can be 
-high even if $\P{X_i = j}$ is large. Given the causal graph in Figure \ref{fig:causalStructure_confounded}, $\P{Y|do(X_2= j)} = \sum_{X_1}\P{X_1}\P{Y|X_1, X_2 = j}$. 
-Suppose $X_2 = X_1$ deterministically, no matter how large $\P{X_2 = 1}$ is we will never observe $(X_2=1,X_1 = 0)$ and so cannot 
-get a good estimate for $\P{Y|do(X_2=1)}$.
-
-To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
-allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
-are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
-or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
-start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
-Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
-Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.
-
-\begin{algorithm}[H]
-\caption{General Algorithm}\label{alg:general}
-\begin{algorithmic}
-\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
-\FOR{$t \in \set{1,\ldots,T}$}
-\STATE Sample action $a_t$ from $\eta$
-\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
-\ENDFOR
-\FOR{$a \in \calA$}
-\STATE
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
-}
-\ENDFOR
-\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
-\end{algorithmic}
-\end{algorithm}
-
-Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator.
-Let $\parents{Y}(X)$ denote the realisation of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$
-
-\eq {
-\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}\,, 
-}
-
-where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently.
-The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.
-
-The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimised to minimise the worst-case simple regret. Abusing notation we define $m(\eta)$ by
-\eq{
-m(\eta) = \max_{a \in \calA} \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,,\text{ where } \EEa \text{ is the expectation with respect to } \Pn{a}.
-}
-
-We will show shortly that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.
-
-\begin{theorem}\label{thm:general}
-If Algorithm \ref{alg:general} is run with $B \in \R^{\calA}$ given by $B_a = \sqrt{\frac{m(\eta)T}{\log\left(2T|\calA|\right)}}\,.$
-
-\eq{
-\simpleregret \in \bigo{\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}}\,.
-}
-\end{theorem}
-
-The proof is in Section \ref{sec:thm:general}.
-
-Note the regret has the same form as that obtained for Algorithm \ref{alg:simple}, with $m(\eta)$ replacing $m(q)$.
-Algorithm \ref{alg:simple} assumes only the graph structure and not knowledge of the conditional distributions on $X$.
-Thus it has broader applicability to the parallel graph than the generic algorithm given here.
-We believe that Algorithm \ref{alg:general} with the optimal choice of $\eta$ is close to mini-max optimal, but leave lower bounds
-for future work.
-
-\paragraph{Choosing the Sampling Distribution} Algorithm \ref{alg:general} depends on a choice of sampling distribution $\operatorname{Q}$ that is determined by $\eta$.
-In light of Theorem \ref{thm:general}
-a natural choice of $\eta$ is the minimiser of $m(\eta)$.
-\eq{
-\eta^* 
-= \argmin_\eta m(\eta) = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{\sum_{b \in \calA} \eta_b \Pn{b}{\parents{Y}(X)}}\right]}_{m(\eta)}\,.
-}
-Since the mixture of convex functions is convex and the maximum of a set of convex functions is convex, we see that $m(\eta)$ is convex (in $\eta$).
-Therefore the minimisation problem may be tackled using standard techniques from convex optimisation.
-The quantity $m(\eta^*)$ may be interpreted as the minimum achievable worst-case variance of the importance weighted estimator.
-In the experimental section we present some special cases, but for now we give two simple results.
-The first shows that $|\calA|$ serves as an upper bound on $m(\eta^*)$.
-
-\begin{proposition}\label{pro:m-bound}
-$m(\eta^*) \leq |\calA|$. \textit{Proof.} 
-\textup{By definition, $m(\eta^*) \leq m(\eta)$ for all $\eta$. Let $\eta_a = 1/|\calA|\,\forall a$.}
-\eq{
-m(\eta) 
-= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\eta_a \Pn{a}{\parents{Y}(X)}}\right] 
-= \max_a \EEa\left[\frac{1}{\eta_a}\right] = |\calA| %\qedhere
-}
-\end{proposition}
-
-The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$.
-This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.
-The proof is in section \ref{sec:m-equivelence}.
-
-\begin{remark}\label{rem:truncate}
-The choice of $B_a$ given in Theorem \ref{thm:general} is not the only possibility. As we shall see in the experiments, it is 
-often possible to choose $B_a$ significantly
-larger when there is no heavy tail and this can drastically improve performance by eliminating the bias. This is especially true when the ratio $R_a$ is never too large
-and Bernstein's inequality could be used directly without the truncation. For another discussion see the article by \citet{Bottou2013} who also use importance weighted estimators
-to learn from observational data.
-\end{remark}
-
-\subsection{Experiments}
-\label{sec:causal-bandit-experiments}
-We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{Audibert2010}, Thompson Sampling and UCB under a variety of conditions.
-Thomson sampling and UCB are optimised to minimise cumulative regret.
-We apply them in the fixed horizon, best arm identification setting by running them up to horizon $T$ and then selecting the arm with the highest empirical mean.
-The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.
-
-Throughout we use a model in which $Y$ depends only on a single variable $X_1$ (this is unknown to the algorithms).
-$Y_t \sim \bernoulli(\frac{1}{2}+\epsilon)$ if $X_1=1$ and $Y_t \sim \bernoulli(\frac{1}{2}-\epsilon')$ otherwise, where $\epsilon' = q_1\epsilon/(1-q_1)$.
-This leads to an expected reward of $\frac{1}{2}+\epsilon$ for $do(X_1=1)$, $\frac{1}{2}-\epsilon'$ for $do(X_1=0)$ and $\frac{1}{2}$ for all other actions.
-We set $q_i = 0$ for $i \leq m$ and $\frac{1}{2}$ otherwise.
-Note that changing $m$ and thus $\boldsymbol{q}$ has no effect on the reward distribution.
-For each experiment, we show the average regret over 10,000 simulations with error bars displaying three standard errors.
-The code is available from \url{<https://github.com/finnhacks42/causal_bandits>}
-
-In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases.
-The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure.
-For the causal algorithms it increases approximately with $\sqrt{m}$.
-As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.
-
-\begin{figure}[h]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment1_20161020_1247.pdf}
-    		\caption{Simple regret vs $m(\boldsymbol{q})$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment2_20161020_1249.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and $\epsilon = \sqrt{\frac{N}{8T}}$}
-        \label{fig:simple_vs_T_vary_epsilon}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment3_20161020_1252.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$, $m=2$ and fixed $\epsilon = .3$}
-    		\label{fig:simple_vs_T}
-    \end{subfigure}
-    \caption{Experimental results}
-    \label{fig:experiments}
-\end{figure}
-
-Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
-This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
-In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
-The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.
-
-For the parallel bandit problem, the regression estimator used in the specific algorithm outperforms the truncated importance weighted estimator in the more general algorithm, despite the fact the specific algorithm must estimate $\boldsymbol{q}$ from the data.
-This is an interesting phenomenon that has been noted before in off-policy evaluation where the regression (and not the importance weighted) estimator is known to be mini-max optimal asymptotically \citep{li2014minimax}.
-
-\begin{figure}[H]
-	\centering    
-          \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.45cm,
-  thick,main node/.style={observed}, hidden/.style={empty},background rectangle/.style={fill=olive!45}]
-%every node/.style={scale=0.6}
- %nodes
-\node[main node](1){$X_{1}$};
-\node[main node, right=of 1](2){$X_{2}$};
-\node[hidden, right=of 2](3){$...$};
-\node[main node, right=of 3](4){$X_{N}$};
-\node[main node, below right=of 2](5){$Y$};
-\node[main node, above right=of 2](6){$Z$};
- \path[every node/.style={font=\tiny}]
-    (1) edge (5)
-    	(2) edge (5)
-    (4) edge (5)
-    (6) edge (1) edge (2) edge (4);
-\end{tikzpicture}
-        \caption{Confounded graph}
-        \label{fig:parallel_confounded} 
-\end{figure}
-
-We now compare the general algorithm with a range of standard bandit algorithms on the confounded graph in Figure \ref{fig:parallel_confounded}.
-All the variables are binary and the action space consists of the set of single variable interventions plus the do nothing action,
-
-\eq{
-\calA = \set{\set{do(X_i = j)} \cup \set{do(Z = j)} \cup \set{do()}: 1\leq i \leq N,\; j \in \set{0,1}}
-}
-
-We choose this setting because it generalises the parallel bandit, while simultaneously being sufficiently simple that we can compute the exact reward and interventional distributions for large $N$ (in general inference in graphical models is exponential in $N$).
-As before, we show the average regret over 10,000 simulations with error bars showing three standard errors.
-
-In Figure \ref{fig:simple_vs_m_general} we fix $N$ and $T$ and $P(Z=1) = .4$. For some $2 \leq N_1 \leq N$ we define 
-\eq{
-P(X_i = 1|Z = 0) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .4 & \text{ otherwise } \end{cases}\\
-P(X_i = 1|Z = 1) &= \begin{cases} 0 & \text{ if } i \in \set{1,...N_1} \\ .65 & \text{ otherwise } \end{cases}
-}
-As in the parallel bandit case, we let $Y$ depend only on $X_1$, $P(Y|do(X_1=1)) = \frac{1}{2} + \epsilon$ and $P(Y|do(X_1=0)) = \frac{1}{2}-\epsilon'$, where $\epsilon' = \epsilon P(X_1=1) / P(X_1=0)$.
-The value of $N_1$ determines $m$ and ranges between $2$ and $N$.
-The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
-
-In Figure \ref{fig:simple_vs_T_general} we fix the model and number of variables, $N$, and vary the horizon $T$. $P(Z)$ and $P(X|Z)$ are the same as for the previous experiment.  
-In Figure \ref{fig:simple_vs_T_misspecified} we additionally show the performance of Algorithm 1, but exclude actions on $Z$ from the set of allowable actions to demonstrate that Algorithm 1 can fail in the presence of a confounding variable, which occurs because it incorrectly assumes that $P(Y|do(X)) = P(Y|X)$.
-We let $P(Z) = .6$, $P(Y|\boldsymbol{X}) = X_7 \oplus X_N$ and $P(X|Z)$ be given by:
-
-\eq{
-P(X_i = 1|Z = 0) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.2 & \text{ if } i = 7 \\
-.7 & \text { otherwise} 
- \end{cases}\\
- P(X_i = 1|Z = 1) &= 
-\begin{cases} 
-.166 & \text{ if } i \in \set{1,..., 6} \\
-.8 & \text{ if } i = 7 \\
-.3 & \text { otherwise} 
- \end{cases}\\
-}
-
-In this setting $X_7$ tends to agree with $Z$ and $X_N$ tends to disagree.
-It is sub-optimal to act on either $X_7$ or $X_N$, while all other actions are optimal.
-The first group of $X$ variables with $i \leq 6$ will be identified by the parallel bandit as the most unbalanced ones and played explicitly.
-All remaining variables are likely to be identified as balanced and estimated from observational estimates.
-The CPD values have been chosen to demonstrate the worst case outcome, where the bias in the estimates leads Algorithm 1 to asymptotically select a sub-optimal action.
-
-\begin{figure}[H]
-    \begin{subfigure}[t]{0.3\textwidth}
-		\centering    
-    		\includegraphics[width=\textwidth]{experiment4_20161023_2120.pdf}
-    		\caption{Simple regret vs $m(\eta*)$ for fixed horizon $T=400$ and number of variables $N = 50$}
-        \label{fig:simple_vs_m_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-        \includegraphics[width=\textwidth]{experiment7_20161020_1257.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 50$ and $m(\eta*)=3.1$ }
-        \label{fig:simple_vs_T_general}
-    \end{subfigure}\hfill
-    \begin{subfigure}[t]{0.3\textwidth}
-    		\centering
-    		\includegraphics[width=\textwidth]{experiment5_20161023_2118.pdf}
-    		\caption{Simple regret vs horizon, $T$, with $N = 21$, $m(\eta*)=4.3$ with no actions setting $Z$}
-    		\label{fig:simple_vs_T_misspecified}
-    \end{subfigure}
-    \caption{Experimental results on the confounded graph}
-    \label{fig:experiments_confounded}
-\end{figure}
-
-\subsection{Discussion \& Future work}
-\label{sec:causal_bandit_discussion}
-Algorithm~\ref{alg:general} for general causal bandit problems 
-estimates the reward for all allowable interventions $a \in \calA$ over $T$ rounds by sampling and applying interventions from a distribution $\eta$.
-Theorem~\ref{thm:general} shows that this algorithm has (up to log factors) simple regret that is $\smash{\mathcal O(\sqrt{m(\eta)/T)}}$ where 
-the parameter $m(\eta)$ measures the difficulty of learning the causal model and is always less than $N$.
-The value of $m(\eta)$ is a uniform bound on the variance of the reward estimators $\hat{\mu}_a$ and, intuitively, problems where all variables' values in the causal model ``occur naturally'' when interventions are sampled from $\eta$ will have low values of $m(\eta)$.
-
-The main practical drawback of Algorithm~\ref{alg:general} is that both the estimator $\hat{\mu}_a$ and the optimal sampling distribution $\eta^*$ (\ie, the one that minimises $m(\eta)$) require knowledge of the conditional distributions $\Pn{a}{\parents{Y}}$ for all $a \in \calA$.
-In contrast, in the special case of parallel bandits, Algorithm~\ref{alg:simple} uses the $do()$ action to effectively estimate $m(\eta)$ and the rewards then re-samples the interventions with variances that are not bound by $\hat{m}(\eta)$.
-Despite these extra estimates, Theorem~\ref{thm:lower} shows that this approach is optimal (up to log factors).Finding an algorithm that only requires the causal graph and lower bounds for its simple regret in the general case is left as future work.
-
-\paragraph{Making Better Use of the Reward Signal}
-Existing algorithms for best arm identification are based on ``successive rejection'' (SR) of arms based on UCB-like bounds on their rewards~\citep{Even-Dar2002}.
-In contrast, our algorithms completely ignore the reward signal when developing their arm sampling policies and only use the rewards when estimating $\hat{\mu}_a$.
-Incorporating the reward signal into our sampling techniques or designing more adaptive reward estimators that focus on high reward interventions is an obvious next step.
-This would likely improve the poor performance of our causal algorithm relative to the successive rejects algorithm for large $m$, as seen in Figure~\ref{fig:simple_vs_m}.
-
-For the parallel bandit the required modifications should be quite straightforward.
-The idea would be to adapt the algorithm to essentially use successive elimination in the second phase so arms are eliminated as soon as they are provably no longer optimal with high probability.
-In the general case a similar modification is also possible by dividing the budget $T$ into phases and optimising the sampling distribution $\eta$, eliminating arms when their confidence intervals are no longer overlapping.
-This has now been done by \citet{Sen2017a}, leading to problem dependent regret bounds for causal bandit problems.
-Note that these modifications do not improve the mini-max regret, which at least for the parallel bandit is already optimal.
-For this reason we focused on emphasising the point that causal structure can and should be exploited when available.
-Another observation is that Algorithm \ref{alg:general} is actually using a fixed design, which in some cases may be preferred to a sequential design for logistical reasons.
-This is not possible for Algorithm \ref{alg:simple}, since the $\vec{q}$ vector is unknown.
-
-\paragraph{Cumulative Regret}
-Although we have focused on simple regret in our analysis, it would also be natural to consider the cumulative regret.
-In the case of the parallel bandit problem we can slightly modify the analysis from \citep{wu2015online} on bandits with side information
-to get near-optimal cumulative regret guarantees.
-They consider a finite-armed bandit model with side information where in reach round the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
-In this way the learner can gain information about actions it does not take with varying levels of accuracy.
-The reduction follows by substituting the importance weighted estimators in place of the Gaussian reward.
-In the case that $\vec{q}$ is known this would lead to a known variance and the only (insignificant) difference is the Bernoulli noise model.
-In the parallel bandit case we believe this would lead to near-optimal cumulative regret,
-at least asymptotically.
-
-%Their model assumes the rewards for all arms $a$ are Gaussian with mean $\mu_a$ and variance $\sigma^2_{ab}$ and that playing an arm $a$ will reveal a side observation $Y_{ab}$ of the reward for all arms $b$ distributed with mean $\mu_b$ and variance $\sigma^2_{ab}$.
-%We can build a similar dependence structure with variances for a Bernoulli reward variable that is derived from the $\vec{q}$ vector of probabilities.
-%\todom{Check this!}
-%Even though the original results are for Gaussian rewards we believe the analysis will go through largely unchanged.
-
-The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each time step the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
-The feedback graph is distinct from the causal graph.
-A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$.
-For this parallel bandit problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions.
-However, \citet{Alon2015,Kocak2014} give adversarial algorithms, which when applied to the parallel bandit problem obtain the standard bandit regret.
-A malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action.
-This is equivalent to a nominally stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$.
-
-\citet{Lelarge2012} consider a stochastic version of the graph feedback problem, but with a fixed graph available to the algorithm before it must select an action.
-In addition, their algorithm is not optimal for all graph structures and fails, in particular, to provide improvements for star like graphs as in our case.
-\citep{Buccapatnam2014} improve the dependence of the algorithm on the graph structure but still assume the graph is fixed and available to the algorithm before the action is selected.
-
-\paragraph{Causal Models with Non-Observable Variables}
-If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
-non-observable variables. Some of the interventional distributions may be non-identifiable meaning we can not obtain prior estimates for $\Pn{a}{\parents{Y}}$ from 
-even an infinite amount of observational data. Even if all variables are observable and the graph is known, if the conditional distributions are unknown, then Algorithm
-\ref{alg:general} cannot be used. Estimating these quantities while simultaneously minimising the simple regret is an interesting and challenging open problem.
-
-% For example, if we had access to a data set of \textit{experiments} in which the reward variable $Y$ was not available from which to build estimates of $P_a$.
-% In this case, some conditional distributions may be non-identifiable. 
-% The corresponding actions can be immediately added to the set $A$ prior to collecting any data. 
-% We can then use the same algorithm as in the case where there are no latent variables, except that we will have to use the more general do-calculus rather than simply adjusting for the parents to write the expression for each action in terms of observational data.
-% Combining our estimation techniques with insights from \citet{Bareinboim2015} for handling unobserved confounders would be worth investigation.
-
-% More generally, assuming causal structure creates more complex types of side information, such as that shown in equation \ref{eq:estimation_transfer}.
-In this case, selecting one action does not fully reveal an alternate action but provides some information towards an estimate.
-The quality of the estimate notably depends not only on the number of times that action was selected.
-For example, to get a good estimate for $X_1 = 1$ by intervening on $X_2$ requires us to sample both $X_2=0$ and $X_2=1$, in proportions dependent on $q_2$.
-This more complex side information does not fit within the graph feedback framework.
-
-\paragraph{Partially or Completely Unknown Causal Graph}
-A much more difficult generalisation would be to consider causal bandit problems where the causal graph is completely unknown or known to be a member of class of models.
-The latter case arises naturally if we assume free access to a large observational data set, from which the Markov equivalence class can be found via causal discovery techniques. 
-Work on the problem of selecting experiments to discover the correct causal graph from within a Markov equivalence class~\citep{eberhardt2005,eberhardt2010causal,hauser2014two,Hu2014} could potentially be incorporated into a causal bandit algorithm.
-In particular, \citet{Hu2014} show that only $\bigo{\log \log n}$ multi-variable interventions are required on average to recover a causal graph over $n$ variables once purely observational data is used to recover the ``essential graph''.
-Simultaneously learning a completely unknown causal model while estimating the rewards of interventions without a large observational data set would be much more challenging.
-
-\subsection{Proofs}
-\subsubsection{Proof of Theorem \ref{thm:uq-simple}}\label{sec:thm:uq-simple}
-
-Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$. The assumption is non-restrictive since all variables
-are independent and permutations of the variables can be pushed to the reward function.
-
-The proof of Theorem \ref{thm:uq-simple} requires some lemmas. \
-
-\begin{lemma}\label{lem:conc1}
-Let $i \in \set{1,\ldots, N}$ and $\delta > 0$. Then
-\eq{
-\P{\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2}{\delta}}} \leq \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-By definition, $\hat{q}_i = \frac{2}{T}\sum_{t=1}^{T/2}X_{t,i}$, where $X_{t,i} \sim Bernoulli(q_i)$. Therefore from the Chernoff bound (see equation 6 in \citet{Hagerup1990}),
-
-\eq{
-\P{\left|\hat q_i - q_i\right| \geq \epsilon} \leq 2e^{-\frac{T\epsilon^2}{6q_i}}
-}
-
-Letting $\delta = 2e^{-\frac{T\epsilon^2}{6q_i}}$ and solving for $\epsilon$ completes the proof.
-
-\end{proof}
-
-\begin{lemma}\label{lem:m_est}
-
-Let $\delta \in (0,1)$ and assume $T \geq 48m \log\frac{2N}{\delta}$. Then
-\eq{
-\P{2m(\vec{q}) / 3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})} \geq 1 - \delta\,.
-}
-\end{lemma}
-
-\begin{proof}
-Let $F$ be the event that there exists and $1 \leq i \leq N$ for which
-\eq{
-\left|\hat q_i - q_i\right| \geq \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}\,.
-}
-Then by the union bound and Lemma \ref{lem:conc1} we have $\P{F} \leq \delta$. The result will be completed by showing that
-when $F$ does not hold we have $2m(\vec{q})/3 \leq m(\vec{\hat q}) \leq 2m(\vec{q})$.
-From the definition of $m(\vec{q})$ and our assumption on $\vec{q}$ we have for $i > m(\vec{q})$ that $q_i \geq q_m \geq 1/m(\vec{q})$ and so by Lemma \ref{lem:conc1} we have
-\eq{
-\frac{3}{4} 
-&\geq \frac{1}{2} + \sqrt{\frac{3}{T} \log \frac{2N}{\delta}} 
-\geq q_i + \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}} 
-\geq \hat q_i \\
-&\geq q_i - \sqrt{\frac{6q_i}{T} \log \frac{2N}{\delta}}
-\geq q_i - \sqrt{\frac{q_i}{8m(\vec{q})}}
-\geq \frac{1}{2m(\vec{q})}\,.
-}
-Therefore by the pigeonhole principle we have $m(\vec{\hat q}) \leq 2m(\vec{q})$.
-For the other direction we proceed in a similar fashion. Since the failure event $F$ does not hold we have for $i \leq m(\vec{q})$ that
-\eq{
-\hat q_i 
-\leq q_i + \sqrt{\frac{6q_i}{T} \log\frac{2N}{\delta}} 
-\leq \frac{1}{m(\vec{q})} \left(1 + \sqrt{\frac{1}{8}}\right)
-\leq \frac{3}{2m(\vec{q})}\,.
-}
-Therefore $m(\vec{\hat q}) \geq 2m(\vec{q}) / 3$ as required. 
-\end{proof}
-
-\begin{proof}[Proof of Theorem \ref{thm:uq-simple}]
-Recall that $A = \set{a \in \actions : \hat p_a \leq 1/m(\vec{\hat q})}$. Then,
-for $a \in A$, the algorithm estimates $\mu_a$ from $T_A \doteq T/(2m(\vec{\hat q}))$ samples.
-From lemma \ref{lem:m_est}, $T_A  \geq T/(4m(\vec{q}))$ with probability $(1-\delta)$.
-Let $H$ be the event $T_A < T/(4m(\vec{q}))$ and $G$ be the event $\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}$
-
-\eq{
-\P{G} \leq \P{H} + \P{G|\neg H} \leq \delta + \P{G|\neg H}
-}
-
-Via Hoeffding's inequality and the union bound,
-
-\eq{
-\P{G|\neg H}&\doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}},\text{ given }T_A \geq T/(4m(\vec{q}))} \leq \delta\\
- \implies  \P{G} & \doteq \P{\exists a \in A : |\mu_a - \hat \mu_a| \geq \sqrt{\frac{2m(\vec{q})}{T} \log\frac{2N}{\delta}}} \leq 2\delta\,.
-}
-For arms not in $A$,
-
-\eq{
-\hat p_a = & \frac{2}{T} \sum_{t=1}^{T/2} \ind{X_i = j} \geq 1/m(\vec{\hat q}), \; \text {by definition of not being in $A$}\\
- \geq & \frac{1}{2m(\vec{q})},\; \text{ with probability } 1-\delta \\
-\implies  T_a \doteq & \sum_{t=1}^{T/2} \ind{X_i = j} \geq  \frac{T}{4m(\vec{q})},\; \text{ with probability } 1-\delta
-}
-
-Again applying Hoeffding's and the union bound
-
-\eq{
-\P{\exists a \notin A:  \left|\hat \mu_a - \mu_a\right| \geq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}}} \leq 2\delta
-}
-
-Therefore, combining this result with the bound for arms $a \in A$, we have with probability at least $1 - 4\delta$ that,
-\eq{
-(\forall a \in \actions) \qquad |\hat \mu_a - \mu_a| \leq \sqrt{\frac{2m(\vec{q})}{T} \log \frac{2N}{\delta}} \doteq \epsilon\,.
-}
-If this occurs, then 
-\eq{
-\mu_{\hat a^*_T} \geq \hat \mu_{\hat a^*_T} - \epsilon \geq \hat \mu_{a^*} - \epsilon \geq \mu_{a^*} - 2\epsilon\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_{\hat a^*_T}] 
-& \leq 4\delta + \epsilon\\
-& \leq \frac{8m(\vec{q})}{T} + \sqrt{\frac{2m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ letting } \delta = \frac{2m(\vec{q})}{T}\\
-& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jenson's Inequality}
-}
-which completes the result.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LOWER BOUND
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Proof of Theorem \ref{thm:lower}}\label{sec:thm:lower}
-
-We follow a relatively standard path by choosing multiple environments that have different optimal arms, but which cannot all be statistically
-separated in $T$ rounds.
-Assume without loss of generality that $q_1 \leq q_2 \leq \ldots \leq q_N \leq 1/2$.
-For each $i$ define reward function $r_i$ by
-\eq{
-r_0(\boldsymbol{X}) &= \frac{1}{2} &
-r_i(\boldsymbol{X}) &= \begin{cases}
-\frac{1}{2} + \epsilon & \text{if } X_i = 1 \\
-\frac{1}{2} & \text{otherwise}\,,
-\end{cases}
-}
-where $1/4 \geq \epsilon > 0$ is some constant to be chosen later.
-We abbreviate $R_{T,i}$ to be the expected simple regret incurred when interacting with the
-environment determined by $\boldsymbol{q}$ and $r_i$. Let $\operatorname{P}_i$ be the corresponding measure
-on all observations over all $T$ rounds and $\EE_i$ the expectation with respect to $\operatorname{P}_i$. By Lemma 2.6 by \citet{Tsy08} we have
-\eq{
-\Prz{\hat a^*_T = a^*} + \Pri{\hat a^*_T \neq a^*} \geq \exp\left(-\KL(\operatorname{P}_0, \operatorname{P}_i)\right)\,,
-}
-where $\KL(\Ps_0, \Ps_i)$ is the KL divergence between measures $\operatorname{P}_0$ and $\operatorname{P}_i$.
-Let $T_i(T) = \sum_{t=1}^T \ind{a_t = do(X_i = 1)}$ be the total number of times the learner intervenes on variable $i$ by setting it to $1$.
-Then for $i \leq m$ we have $q_i \leq 1/m$ and the KL divergence between $\Ps_0$ and $\Ps_i$ may be bounded using the telescoping property (chain rule) and
-by bounding the local KL divergence by the $\chi$-squared distance as by \citet{Auer1995}. This leads to 
-\eq{
-\KL(\Ps_0, \Ps_i) 
-&\leq 6\epsilon^2 \EE_0\left[\sum_{t=1}^T \ind{X_{t,i} = 1}\right] 
-\leq 6\epsilon^2 \left(\EE_0 T_i(T) + q_i T\right) 
-\leq 6\epsilon^2 \left(\EE_0 T_i(T) + \frac{T}{m}\right)\,.
-}
-Define set $A = \set{i \leq m : \EE_0 T_i(T) \leq 2T / m}$.
-Then for $i \in A$ and choosing $\epsilon = \min\set{1/4, \sqrt{m/(18T)}}$ we have
-\eq{
-\KL(\Ps_0, \Ps_i) \leq \frac{18T\epsilon^2}{m} = 1\,. 
-}
-Now $\sum_{i=1}^m \EE_0 T_i(T) \leq T$, which implies that $|A| \geq m/2$.
-Therefore
-\eq{
-\sum_{i \in A} \Pri{\hat a^*_T \neq a} 
-\geq \sum_{i \in A} \exp\left(-\KL(\Ps_0, \Ps_i)\right) - 1
-\geq \frac{|A|}{e} - 1 
-\geq \frac{m}{2e} - 1\,.
-}
-Therefore there exists an $i \in A$ such that
-$\Pri{\hat a^*_T \neq a^*} \geq \frac{\frac{m}{2e} - 1}{m}$. 
-Therefore if $\epsilon < 1/4$ we have
-\eq{
-R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{\frac{m}{2e} - 1}{2m} \sqrt{\frac{m}{18T}}\,.
-}
-Otherwise $m \geq 18T$ so $\sqrt{m/T} = \Omega(1)$ and
-\eq{
-R_{T,i} \geq \frac{1}{2} \Pn{i}{\hat a^*_T \neq a^*} \epsilon \geq \frac{1}{4} \frac{\frac{m}{2e} - 1}{2m} \in \Omega(1) 
-}
-as required.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% GENERAL-GRAPH UPPER BOUND
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Proof of Theorem \ref{thm:general}}\label{sec:thm:general}
-
-\begin{proof}
-First note that $X_t, Y_t$ are sampled from $\operatorname{Q}$.
-We define $Z_a(X_t) = Y_t R_a(X_t)\ind{R_a(X_t)\leq B_a}$ and abbreviate $Z_{at} = Z_a(X_t)$, $R_{at} = R_a(X_t)$ and $\Pn{a}{.} = \Pns{a}{.}$.
-By definition we have $|Z_{at}| \leq B_a$ and 
-\eq{
-\Var_Q[Z_{at}] 
-\leq \EE_Q[Z_{at}^2] 
-\leq \EE_Q[R_{at}^2]
-= \EEa[R_{at}]
-= \EEa\left[\frac{\Pns{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] 
-\leq m(\eta)\,.
-}
-Checking the expectation we have
-\eq{
-\EE_Q[Z_{at}] 
-= \EEa \left[Y \ind{R_{at} \leq B_a}\right] 
-= \EEa Y - \EEa \left[Y\ind{R_{at} > B_a}\right] 
-= \mu_a - \beta_a\,,
-}
-where 
-\eq{
-0 \leq \beta_a = \EEa[Y \ind{R_{at} > B_a}] \leq \Pns{a}{R_{at} > B_a}
-}
-is the negative bias. 
-The bias may be bounded in terms of $m(\eta)$ via an application of Markov's inequality.
-\eq{
-\beta_a \leq \Pns{a}{R_{at} > B_a} \leq \frac{\EEa[R_{at}]}{B_a} \leq \frac{m(\eta)}{B_a}\,.
-}
-Let $\epsilon_a > 0$ be given by
-\eq{
-\epsilon_a = \sqrt{\frac{2m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{3B_a}{T} \log\left(2T|\calA|\right)\,.
-}
-Then by the union bound and Bernstein's inequality 
-\eq{
-\P{\text{exists } a \in \calA : \left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} 
-\leq \sum_{a \in \calA} \P{\left|\hat \mu_a - \EE_Q[Z_{at}]\right| \geq \epsilon_a} \leq \frac{1}{T}\,.
-}
-
-Let $I = \hat{a}^*_T$ be the action selected by the algorithm, $a^* = \argmax_{a \in \calA} \mu_a$ be the true optimal action and recall that $\EE_Q[Z_{at}] = \mu_a - \beta_a$.
-Assuming the above event does not occur we have,
-
-\eq{
-\mu_I \geq \hat \mu_I - \epsilon_I  
-\geq \hat \mu_{a^*} - \epsilon_I  
-\geq \mu^* - \epsilon_{a^*} - \epsilon_I - \beta_{a^*}\,. 
-}
-By the definition of the truncation
-we have
-\eq{
-\epsilon_a \leq \left(\sqrt{2} + 3\right)\sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}
-}
-and
-\eq{
-\beta_a \leq \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}\,. 
-}
-Therefore for $C = \sqrt{2} + 4$ we have
-\eq{
-\P{\mu_I \geq \mu^* - C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)}} \leq \frac{1}{T}\,.
-}
-Therefore
-\eq{
-\mu^* - \EE[\mu_I] \leq C \sqrt{\frac{m(\eta)}{T} \log\left(2T|\calA|\right)} + \frac{1}{T}
-}
-as required.
-\end{proof}
-
-\subsubsection{Relationship between $m(\eta)$ and $m(\boldsymbol{q})$}\label{sec:m-equivelence}
-
-\begin{proposition} In the parallel bandit setting,
-$m(\eta^*) \leq 2m(\boldsymbol{q})$.
-\end{proposition}
-
-\begin{proof}
-
-Recall that in the parallel bandit setting,
-
-\eq{
-\mathcal{A} = \set{do()} \cup \set{ do(X_i = j) \colon 1 \leq i \leq N \text{ and } j \in \set{0,1}}
-}
-
-Let:
-
-\eq {
-\eta_a = \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}}\frac{1}{2m(\boldsymbol{q})} \text { for } a \in do(X_i = j)
-}
-
-Let $D =\sum_{a\in do(X_i=j)}\eta_a$. From the definition of $m(\boldsymbol{q})$, 
-\eq {
-\sum_{a\in do(X_i=j)} \ind{\P{X_i = j} < \frac{1}{m(\boldsymbol{q})}} \leq m(\boldsymbol{q}) \implies D \leq \frac{1}{2}
-}
-
-Let $\eta_a = \frac{1}{2} + (1-D)$ for $a = do()$ such that $\sum_{a \in \calA}\eta_a = 1$
-
-Recall that,
-
-\eq{
-m(\eta) &
-= \max_a \EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]
-}
-
-We now show that our choice of $\eta$ ensures $\EEa\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right] \leq 2m(\boldsymbol{q})$ for all actions $a$.
-
-For the actions $a: \eta_a > 0$, ie $do()$ and $do(X_i = j):\P{X_i=j}<\frac{1}{m(\boldsymbol{q})}$,
-\eq{
-\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq \EEa\left[\frac{\Pn{a}{X_1...X_N}}{\eta_a\Pn{a}{X_1...X_N}}\right] = \EEa\left[\frac{1}{\eta_a}\right] \leq 2m(\boldsymbol{q})
-}
-
-For the actions $a :\eta_a = 0$, ie $do(X_i=j):\P{X_i=j}\geq\frac{1}{m(\boldsymbol{q})}$,
-\eq{
-\EEa\left[\frac{\Pn{a}{X_1...X_N}}{\sum_{b}\eta_b\Pn{b}{X_1...X_N}}\right] \leq & \EEa\left[\frac{\ind{X_i=j}\prod_{k\neq i}\P{X_k}}{(1/2+D)\prod_k \P{X_k}}\right] \\=& \EEa\left[\frac{\ind{X_i=j}}{(1/2+D)\P{X_i = j}}\right]
-\leq  \EEa\left[\frac{\ind{X_i=j}}{(1/2)(1/m(\boldsymbol{q}))}\right] \leq 2m(\boldsymbol{q})
-}
-
-Therefore $m(\eta*) \leq m(\eta) \leq 2m(\boldsymbol{q})$ as required.
-
-\end{proof}
-
-\chapter{Conclusion}
-
-TODO - write conclusion \todo{}
-
-Here is the problem and why its important
-
-Here is how what I have done contributes to solving thes problems
-
-What remains to be done - directions for future work
-
-- Relaxing assumption that interventional distributions known
-- Algorithms for causal bandits in which both contextual and post-action feedback is available.
-- Off-policy evaluation for general causal bandits.
-
-Other approaches to identifying causal effects, such instrumental variables, the \quotes{front door} approach could also be applied to off-policy evaluation in some settings.
-
-Ideas developed within the off-policy selection liturature are also relevent classic causal problems within economics etc.
- - optimal action selection (rather than evaluation)
- - quantification of finite time properties of estimators (as opposed to asymptopic efficiency). eg \cite{li2014minimax}
-
-Warm start for (contextual) bandit problem based on potentially biased estimates. (We can detect the presense of a confounder with experimental data).
-
-Alternatives to cross-validation.
-
-\bibliographystyle{apalike}% Select the citation style e.g. ieeetr
-\bibliography{library}% write the directory to the .bib file
-\end{document}
-
-\chapter{Causation vs Prediction}
-
-The two cultures \citep{Breiman2001}
- - wrong because suggests we can just use the machine learning approach for everything.
-Does not address the problems that a) structural assumptions required for causal inference.
-b) we cannot use cross-validation to assess the validity of a causal model (really b is the whole point).
-
-To explain or to predict \citep{Shmueli2010}
-- wrong because they assume all the parts of explanatory modelling are required - so black box models cannot be used.
-
-Predict the relationship between X and Y as a way to stratify to reduce the variance in an experiment.
-
-They consider a different definition of explanatory models,
-
-Definition as backwards.
-Explanatory models focus on theory because some theory (or equivalently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
-
-Causal modelling frameworks allow us to separate assumptions about the structure of the relationships between variables from assumptions about the functional form.
-
-What can be data driven and what cannot? (for a given set of assumptions).
-
-Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the language of EoP but are built on a hybrid of both approaches.

### 2017-09-10 thesis/thesis_map.tex
fixes title pages

@@
 
-\author{Finnian Lattimore}
+\author{Finnian Rachel Lattimore}
 \title{Learning how to act: making good decisions with machine learning}
@@
 
-\begin{titlepage}
-    \begin{center}
-        \vspace*{1cm}
-
-        \Huge
-        {Learning how to act: making good decisions with machine learning}
-
-        \vspace{2cm}
-        \Large
-        {Finnian Lattimore}
-
-        \vfill
-        \Large
-        A thesis submitted for the degree of Doctor of Philosophy
-
-         The Australian National University
-
-        \vspace{1cm}
-
-        September 2017
-
-    \end{center}
-\end{titlepage}
+\input{titlepage}
+\input{original_statement}
 
@@
 
-\clearpage
-\vspace*{\fill}
-\begin{center}
-\begin{minipage}{.9\textwidth}
-Except where acknowledged in the customary manner, the material 
-presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any 
-university.
-
-\vspace{40mm}  % vertical space
-
-\hspace{80mm}\rule{40mm}{.15mm}\par   % horizontal space, line, start new line
-\hspace{80mm} Finnian Lattimore\par
-\hspace{80mm} September, 2017
-\end{minipage}
-\end{center}
-\vfill % equivalent to \vspace{\fill}
-\clearpage
-
-\begin{abstract}
-
-This thesis is about machine learning and statistical approaches to decision making.
-How can we learn from data to anticipate the consequence of, and optimally select, interventions or actions?
-Problems such as deciding which medication to prescribe to patients, who should be released on bail, and how much to charge for insurance are ubiquitous, and have far reaching impacts on our lives.
-There are two fundamental approaches to learning how to act: reinforcement learning, in which an agent directly intervenes in a system and learns from the outcome, and observational causal inference, whereby we seek to infer the outcome of an intervention from observing the system.
-
-The goal of this thesis to connect and unify these key approaches.
-I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis.
-I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be exploited by existing algorithms.
-I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.
-
-\end{abstract}
+\chapter*{Acknowledgements}
+\input{abstract}
 

### 2017-09-13 thesis/thesis_map.tex
fixes references

@@
 By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.
+Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}
 
@@
 However, this does not mean that the problem of identifying causes of effects is ill-posed in principle.
-It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see for example \citep{}.
-\todo{reference a review paper on causal discovery here}
+It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich literature, see \citet{Spirtes2016} for a recent review.
 
@@
 The state of the board following each action is deterministic and given by the rules of the game.
-The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem.
+The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win.\footnote{This is a challenging pattern recognition problem.
 There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win.
-It must identify higher level features of the board state that are associated with winning.}.
-Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of a player selecting action $a$, given board state $s$.
+It must identify higher level features of the board state that are associated with winning.} Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of a player selecting action $a$, given board state $s$.
 This allows the agent to estimate the likelihood of winning from a given starting state by simulating many times the remainder of the game, drawing actions from $\P{a|s}$ for both players.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 The equivalent causal Bayesian network additionally represents the post-interventional distributions for six possible single variable interventions and twelve possible two variable interventions.
-Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional thirty parameters \footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
+Encoding all this information without the assumptions implicit in the causal Bayesian network would require an additional thirty parameters.\footnote{After each single variable intervention we have a distribution over two variables, which can be represented by three parameters.
 After each two variable intervention, we have a distribution over one variables which requires one parameter.
-This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}.
+This takes us to a total of $6\times3+12\times1 = 30$ additional parameters.}
 
@@
 
-Clarify Cheng's Questions - is there a mechanical method to go from one framework to another, does it matter if variables are discrete or continuous. \todo{}
-
 Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \caption{Treatment results}
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \caption{Treatment results by gender}
@@
 However, if we know a student is receiving a scholarship, then learning that they are male  increases the probability that they are disadvantaged.
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 \citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem.
-If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$, but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.
-\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
+If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$, but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$.\footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift.
 Take the case where we have a binary intervention $x \in \set{0,1}$.
@@
 
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}.
+The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{Robins1994a,Kang2007}.
 Doubly robust methods are asymptotically unbiased so long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \center
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{subfigure}[t]{0.99\textwidth}
@@
 In its classic formulation \citep{Robbins1952,Lai1985} the (stochastic) k-armed bandit describes a sequential decision making problem, with $k$ possible actions or arms.
-Each arm $i$ is associated with a fixed but unknown reward distribution.
-\footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
-It sufficient (but not necessary) to assume they are sub-Gaussian.} For each time step up to some horizon $T$, the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
+Each arm $i$ is associated with a fixed but unknown reward distribution.\footnote{In order to quantify the performance of bandit algorithms, some assumptions are required on the distributions from which the rewards are generated.
+It sufficient (but not necessary) to assume they are sub-Gaussian.} For each timestep up to some horizon $T$, the learner selects an action and receives a reward, sampled i.i.d from the marginal distribution corresponding to that action.
 The goal of the learner is to maximise the total reward they receive.
@@
 \begin{enumerate}
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
+\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous timesteps and a (potentially stochastic) \emph{policy} $\pi$
 \item the world stochastically generates the rewards for each action, $[\cfb{Y}{1},...,\cfb{Y}{k}] \sim \P{\vec{y}}$
@@
 The total reward a bandit algorithm/policy can expect to achieve depends on the distributions from which the rewards for each action are sampled.
-To account for this, the performance of bandit algorithms is quantified by the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every time step.
-This difference is known as the (cumulative) regret \footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
+To account for this, the performance of bandit algorithms is quantified by the difference between the reward obtained by the algorithm and the reward that would have been obtained by an oracle that selects the arm with the highest expected reward at every timestep.
+This difference is known as the (cumulative) regret.\footnote{The term regret is somewhat overloaded in the reinforcement learning literature.
 There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.}.
+In addition, researches often refer to the expected regret as \quotes{the regret}.}
 \eqn{
@@
 \item the world (or adversary) generates, but does not reveal, a vector or rewards $\vec{Y_t}=[\cfb{Y}{1},...,\cfb{Y}{k}]$. 
-\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous time-steps and a (potentially stochastic) \emph{policy} $\pi$
+\item the learner selects an action $a_{t} \in \set{1,...,k}$, based on the actions and rewards from previous timesteps and a (potentially stochastic) \emph{policy} $\pi$
 \item the learner observes and receives (only) the reward for the selected action $\cfb{Y}{a_t}$ 
@@
 
-Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous time steps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
+Adversaries that generate rewards independently of the sequence of actions selected by the learner in previous timesteps are referred to as \emph{oblivious}, as opposed to \emph{non-oblivious} adversaries, which can generate rewards as a function of the history of the game.
 In the case of oblivious adversaries, we can also define the adversarial bandit problem by assuming the adversary generates the entire sequence of reward vectors before the game commences.
 
-For oblivious adversarial bandits, we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm \footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.}.
-We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because the sequence of rewards is fixed.
+For oblivious adversarial bandits, we can define regret analogously to stochastic bandits as the difference between the reward obtained by playing the single arm with the highest reward in every round and the expected reward obtained by the algorithm.\footnote{This is also referred to as the weak regret, since in the adversarial case, it can make more sense to compare against the best sequence of arms rather than the best single arm.} We do not have to take the expectation over the first term of equation \ref{eqn:pseudo_regret-adversarial} because the sequence of rewards is fixed.
 However the reward obtained by the algorithm is still a random variable as we are considering randomised algorithms.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \begin{subfigure}[t]{0.49\textwidth}
@@
 \item the world stochastically generates the vector of rewards for each action and the context, $(\vec{X}_t,[\cfb{Y}{1},...,\cfb{Y}{k}]) \sim \P{\vec{x},\vec{y}}$ and reveals $\vec{X}_t$ to the learner
-\item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous time-steps,
+\item the learner selects an action $A_{t} \in \set{1,...,k}$, based on the context as well as actions and rewards from previous timesteps,
 \item the learner observes and receives (only) the reward for the selected action $Y_t = \cfb{Y}{A_t}$ 
@@
 However, as this reduction treats the problem of learning the correct action for each context completely independently, it cannot leverage any structure in the relationships between different contexts and actions.
-As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional \footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
-For example, with $n$ binary variables, $N = 2^n$}.
-We expect some form of smoothness; that is, values of context that are similar should lead to comparable rewards for a given action.
+As in the supervised learning setting, the existence of some form of low-dimensional structure is key to learning in realistic problems, where the context is continuous or high-dimensional.\footnote{Even if the context is genuinely discrete, $N$ grows exponentially with the number of variables.
+For example, with $n$ binary variables, $N = 2^n$} We expect some form of smoothness; that is, values of context that are similar should lead to comparable rewards for a given action.
 We need algorithms that can leverage such assumptions.
@@
 An alternate reduction to the standard bandit problem, which allows us to constrain the hypothesis space to explore, is to treat each hypothesis $h$ as a bandit arm \citep{Langford2008}.
-At each time-step, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
+At each timestep, we select $h \in \mathcal{H}$ based on the rewards previously observed for each hypothesis, take action $h(\vec{x})$ and observe the associated reward.
 Although this approach removes the explicit dependence on the size of the context, the regret  grows linearly with the size of the hypothesis class considered, limiting our ability to learn any complex mappings from context to actions.
@@
 It selects actions uniformly at random during an exploration phase and leverages this data to estimate the value of each hypothesis, using inverse propensity weighted estimators to \quotes{fill in} the missing data.
-The hypothesis with the highest empirical reward is then used to select actions for the remaining time steps.
+The hypothesis with the highest empirical reward is then used to select actions for the remaining timesteps.
 The epsilon-greedy algorithm obtains worst case regret $\bigo{T^{\frac{2}{3}}(k \log{\card{\mathcal{H}}})^{\frac{1}{3}}}$, which has sub-optimal dependence on the horizon $T$.
@@
 The context $\vec{X}$ can only be useful in causal graphs of the form in figure \ref{fig:context-graph-anti-causal} if the components of $\vec{Y}$ are realised prior to the agent selecting an action, i.e when the components of $\vec{Y}$ are not truly counterfactual, it just that the agent is limited to only observing the outcome of the action they choose.
-To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each time-step $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
+To represent a realistic (contextual) bandit problem, where actions are not selected at random, we would need to \quotes{unroll} the graphs, such that there was a copy for each timestep $t$ and allow the action $A_t$ to depend on the context $\vec{X}_t$ and the previous observations $(\vec{X},A,Y)_1^{t-1}$.
 }
@@
 
-It is possible, as we demonstrated in example \ref{exm:anti-casual-context}, to construct bandit problems in which the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$ is actually generated at each time-step (and, at least in principle, is observable to someone although not the agent).
+It is possible, as we demonstrated in example \ref{exm:anti-casual-context}, to construct bandit problems in which the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$ is actually generated at each timestep (and, at least in principle, is observable to someone although not the agent).
 However, in most real applications for bandit problems, only the outcome associated with the selected action is ever realised and so the remaining components of the reward vector are counterfactual, justifying the use of the counterfactual notation to denote them.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \captionsetup[subfigure]{position=b}
@@
 The expectation of the regret as defined by equation \ref{eqn:pseudo_regret} also remains unchanged as both its definition and the learner's actions depend only on the marginal distributions.
-It seems therefor unfortunate that we should have to assume, for example, that the rewards for alternate actions are independent of one-another to be able to analyse the variance of the regret.
+It seems therefore unfortunate that we should have to assume, for example, that the rewards for alternate actions are independent of one-another to be able to analyse the variance of the regret.
 This is particularly so, as this assumption is likely to be violated in many realistic bandit problems.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \begin{subfigure}[t]{0.3\textwidth}
@@
 
-The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each time step the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
+The parallel bandit problem can also be viewed as an instance of a time varying graph feedback problem \citep{Alon2015,Kocak2014}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen.
 The feedback graph is distinct from the causal graph.
@@
 In particular, estimators that are geared towards optimal action selection (rather than evaluation of all actions) and approaches to quantify the finite time properties, as opposed to asymptotic efficiency, of estimators, for example \citep{li2014minimax}.
-
 An important line of research, that is relevant to both the observational and interventional approaches to causal inference, is developing methodologies for model evaluation and selection that provide something equivalent to what cross-validation does for supervised learning.
-
 My hope is that in the next years, combining the reinforcement learning approach to decision making, with causal graphical models and causal effect estimation techniques developed within statistics and economics, will revolutionise our ability to make good data driven decisions.

### 2017-09-16 thesis/thesis_map.tex
adds bit about bandits and RL

@@
 
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation).
+Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation) \citep{Pearl2000}.
 If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
@@
 
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{Robins1994a,Kang2007}.
+The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \c%fitep{Robins1994a,Kang2007}.
 Doubly robust methods are asymptotically unbiased so long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
@@
 Figure (a) shows the results with no prior (equivalently an improper prior) for both $w_yx$ and $\beta$.
-Figure (b) shows the results when we place a Gaussian prior centered around the $w_yz$ on the distribution for $\beta$.}\label{fig:careful-prior-example}
+Figure (b) shows the results when we place a Gaussian prior centered around the $w_yz$ on the distribution for $\beta$.
+All models were fit using Stan \citep{carpenter2016stan}}\label{fig:careful-prior-example}
 \end{figure}
@@
 This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.
+
+Reinforcement learning describes the problem of an agent interacting with an enviroment, learning by observing the outcome of its actions, with the goal of maximising some reward.
+These problems, which also incorporate planning, are extremely difficult because the value of an action is generally not immediately clear.
+The state of the enviroment may evolve over time and according to previously selected actions.
+Actions available at future timesteps can depend on those taken in the past, and rewards may be obtained only after a long sequence of actions.
+This makes it extremely difficult for the agent to acurately attribute value to each chosen action along the path to obtaining a given reward.
+
+We focus on a simpler class of problems within reinforcement learning known as multi-armed bandit problems.
+Multi-armed bandit problems also descibe an agent aiming to maximise some reward by interacting with an enviroment.
+However, the reward is observed immediately after the action is selected and the enviroment is stateless.
+Future rewards generated by the  enviroment in response to the agents actions do not depend on the previous actions of the agent.
+These assumptions are a reasonable approximation in settings where we face the same decision repeately with respect to many independent units or individuals and we have a good proxy for the outcome.
+In this chapter we describe the interventionalist approaches to learning to act for these types of problems, beginning with classical randomised experiments and then the, more adaptive, multi-armed bandit algorithms.
+We show how both observational causal inference and bandit algorithms can be viewed as generalisations of randomized experiments and highlight connections between the observational and interventional approaches to causal inference.
 

### 2017-09-20 thesis/offcuts_chapter6.tex
fixes references
addresses Tiberio's comments

@@
 \section*{Questions for Tor}
-- How do I fix the page numbering in my thesis?
 - How does one derive the regret decompostion
@@
 \begin{enumerate}
+Clarify Cheng's Questions - is there a mechanical method to go from one framework to another, does it matter if variables are discrete or continuous. \todo{}
+
 \item Write a talk - include slide on causal inference without do

### 2017-09-20 thesis/thesis_map.tex
addresses Tiberio's comments

@@
 Will asking people to pay upfront at the doctors reduce long term health expenditure?
-If we developed a drug to suppress particular genes, could we cure multiple sclerosis and would delaying teen-aged pregnancies improve the outcome for their children.
+If we developed a drug to suppress particular genes, could we cure multiple sclerosis?
+Would delaying teen-aged pregnancies improve the outcome for their children.
 
@@
 Correlation is not causation.
-Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, despite the fact that the variables involved are susceptible to human intuition and understanding.
+Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy - despite the fact that the variables involved are susceptible to human intuition and understanding.
 
@@
 This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
-However, the magnitude, direction and even existence of an association depends which variables are controlled for (or included in a regression).
+However, the magnitude, direction and even existence of an association depends on which variables are controlled for (or included in a regression).
 Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.
@@
 
-Causality is often liked to explanation; understanding how and why things happen.
-I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture.
+Causality is often linked to explanation; understanding how and why things happen.
+I view explanation in terms of compression and generalisation: the amount of information about the world a model can capture.
 This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction.
@@
 By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}
+Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}
 
@@
 There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models.
-However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models.
-In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.
+However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation of models.
+In the next section we emphasise the subtlety that can exist in determining if a problem is causal by examining some typical examples.
 
@@
 Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them.
-Machine learning techniques are being applied a huge range of problems, in both industry and academia.
+Machine learning techniques are being applied to a huge range of problems, in both industry and academia.
 The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to.
@@
 This was because doctors treated such patients very aggressively, thus actually lowering their risk.
-The issue is not with model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols.
+The issue is not with the model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols.
 However, using it to decide how to \emph{change} these protocols could kill.
-The actual question of interest in this case is what happens to patients with characteristics $X$ they are assigned treatment according to decision rule (or policy) $\pi(X)$.
+The actual question of interest in this case is what happens to patients with characteristics $X$ if they are assigned treatment according to decision rule (or policy) $\pi(X)$.
 
@@
 Predictive models can be applied to such problems where the most effective treatment is known for a given target group, and where deciding who to treat on the basis of the model predictions will not change the relationship between the features and outcome.
-
-Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc).
-Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.
 
@@
 The agent wishes to learn the probability of a win given an action they take.
-However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach.
-The actions the agent has to explore are the sames ones as human players explored to generate the training data, and both have the same objective - to win the game.
+However, there are some special characteristics of the go problem that make it amenable to a primarily supervised learning approach.
+The actions the agent has to explore are the same ones as human players explored to generate the training data, and both have the same objective - to win the game.
 In addition, the state of the board encapsulates all the information relevant to selecting a move.
@@
 As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning.
-Although the focus and terminology can differ substantially between fields these techniques, all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system.
+Although the focus and terminology can differ substantially between fields, these techniques all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system.
 Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.
@@
 A particular strand of research within reinforcement learning are multi-armed bandit problems.
-They describe settings in which there are a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action.
-They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.
+They describe settings in which there is a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action.
+They capture problems, such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of the outcome they observe.
 
@@
 The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system.
-This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another.
+This is impossible without some assumptions about how the original system and the system after intervention are related to one-another.
 The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.
@@
 Bandit algorithms deal with the sequential nature of the decision making process, and causal inference with problems where randomisation is not feasible, affordable or ethical.
-The similarities between the problems addressed by these techniques raises the question of whether there are problems best addressed by a combination of these approaches, and if so, how they can be combined.
+The similarities between the problems addressed by these techniques raise the question of whether there are problems best addressed by a combination of these approaches, and if so, how they can be combined.
 
@@
 I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
-I make a number of additional connective contributions that are not encompassed by my work on causal bandit problems.
+
+I also make a number of additional connective contributions that are not encompassed by my work on causal bandit problems.
 I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias.
 I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift.
-I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.
-
-\paragraph{Thesis overview} This thesis is divided into three key chapters, learning from observational data, learning from interventional data and unifying the approaches.
+Finally, I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.
+
+\paragraph{Thesis overview} This thesis is divided into three key chapters: learning from observational data, learning from interventional data and unifying the approaches.
 Chapter \ref{chap:causality} covers learning to act from observational data.
 In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems.
-Section \ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \ref{sec:defining-causal-effects} and \ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation.
-Section \ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.
-
-Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems.
+Section \S\ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \S\ref{sec:defining-causal-effects} and \S\ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation.
+Section \S\ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.
+
+Chapter \ref{chap:interventionalist_view} deals with the interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems.
 In \S\ref{sec:multi-armed-bandits}, I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}).
@@
 In \S\ref{sec:off-policy-evaluation}, I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of causal effect estimation from observational data.
-Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.
+Finally, in \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature of bandit regret.
 
@@
 Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model.
-I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.
+I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.
 
@@
 I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms.
-In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}.
+In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of the causal bandit problem, which I refer to as the \emph{parallel bandit problem}.
 I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning.
@@
 An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited.
-This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system from observing it.
+This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system by observing, rather than acting on it.
 This is the problem of observational causal inference.
@@
 
-Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it.
+Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening on) it.
 As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention.
-Mapping properties from one system to another requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate a system will respond to them.
+Mapping properties from one system to another requires some assumptions about how these two systems are related, or in other words, a way of describing actions and how we anticipate a system will respond to them.
 Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.
@@
 Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings.
-They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another.
+They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action versus what would happen if we took another.
 Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}.
@@
 An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}.
-However, the literature on causal inference techniques remains spit between the different frameworks.
+However, the literature on causal inference techniques remains split between the different frameworks.
 Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework.
@@
 By describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them.
-However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers.
+However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped.
 In order to demonstrate the notation and formalisms each framework provides, we will use them to describe the following simple examples.
@@
 The global Markov property and d-separation also hold in causal networks.
-D-separation, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
+D-separation, which characterises which conditional independencies must hold in any distribution that can be represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference.
 We include a brief review of D-separation in section \ref{sec:dsep}.
@@
 Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}.
-Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.
+Causal effects are differences in counterfactual variables: what the difference is between what would have happened if we did one thing versus what would have happened if we did something else.
 
@@
 \begin{aligned}
-&Y^{0} \equiv P(Y|do(X=0)) \\
-&Y^{1} \equiv P(Y|do(X=1))
+&\P{Y^{0}} \equiv P(Y|do(X=0)) \\
+&\P{Y^{1}} \equiv P(Y|do(X=1))
 \end{aligned}
@@
 
-\subsection{Careful with that prior}
+\subsection{Be careful with that prior}
 
@@
 Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$.
-If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and the problem reduces to a standard multi-armed bandit (over actions in in $\mathcal{A'}$).
+If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and Algorithm \ref{alg:general} cannot do better than uniform exploration over $\mathcal{A'}$.
 \begin{proof}
@@
 
-A better understanding of causality is also relevant for the discussions around transparency and ethics in machine learning, particularly with respect to the European Union's new General Data Protection Regulation, which requires that automated decision systems that significantly affect individuals provide provide \textit{"meaningful information about the logic involved."} \cite{Goodman2016}.
-The recognition that there is a fundamental trade-off between accuracy and transparency, unless we can build perfect causal models, when the interests of individuals and society diverge has implications for they way we design and regulate systems that have the potential to have major impacts on people's lives.
+A better understanding of causality is also relevant for the discussions around transparency and ethics in machine learning, particularly with respect to the European Union's new General Data Protection Regulation, which requires that automated decision systems that significantly affect individuals provide \textit{"meaningful information about the logic involved."} \cite{Goodman2016}.
+The recognition that there is a fundamental trade-off between accuracy and transparency, unless we can build perfect causal models, when the interests of individuals and society diverge has implications for the way we design and regulate systems that have the potential to have major impacts on people's lives.
 We must develop approaches to ensuring machine learning decisions are reasonable and ethical that allow effected individuals recourse to dispute or improve outcomes but do not undermine the ability of the system to function.
@@
 Observational causal inference provides, through the do-calculus, a formal means to map information from observational to interventional settings, as well as between different interventions.
-Bandit algorithms, capture the sequential nature of decision making processes and provide techniques to carefully balance exploration and exploitation.
-
-I have developed a framework that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this framework encode some key existing problems in the literature.
+Bandit algorithms capture the sequential nature of decision making processes and provide techniques to carefully balance exploration and exploitation.
+
+I have developed a framework that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this framework encodes some key existing problems in the literature.
 I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.
@@
 Relaxing this assumption entirely is a much more challenging problem.
-However, as is demonstrated by the specific example of the parallel causal bandit problem, it is possible to develop algorithms that require only the casual structure of the graph and yield substantially lower regret.
+However, as is demonstrated by the specific example of the parallel causal bandit problem, it is possible to develop algorithms that require only the causal structure of the graph and yield substantially lower regret.
 
@@
 As in the online case, knowledge of the causal structure between variables in the graph provides additional information about the reward for the actions that were not selected at each timestep.
-The problem differs from typical observational causal inference in several ways; the focus is on identifying an optimal policy, rather than unbiased estimation of all policies, the goal is to explore the value of interventions on a range of different variables, rather than the optimal setting of a single variable, and the data will, in general, be non-stationary in a rather special way due to the adaptive nature of policy that generated it.
+The problem differs from typical observational causal inference in several ways: the focus is on identifying an optimal policy, rather than unbiased estimation of all policies; the goal is to explore the value of interventions on a range of different variables, rather than the optimal setting of a single variable; and the data will, in general, be non-stationary in a rather special way due to the adaptive nature of policy that generated it.
 Existing work on off-policy evaluation focuses (at least implicitly) on estimating causal effects by adjusting for all variables that simultaneously effect both the action selection and outcome.
@@
 An important line of research, that is relevant to both the observational and interventional approaches to causal inference, is developing methodologies for model evaluation and selection that provide something equivalent to what cross-validation does for supervised learning.
-My hope is that in the next years, combining the reinforcement learning approach to decision making, with causal graphical models and causal effect estimation techniques developed within statistics and economics, will revolutionise our ability to make good data driven decisions.
+My hope is that, in the next years, combining the reinforcement learning approach to decision making with causal graphical models and causal effect estimation techniques developed within statistics, economics and epidemiology will lead to a revolution in our ability to make good data driven decisions.
 

### 2017-09-28 thesis/offcuts_chapter6.tex
adds section for bob, starting work on talk

@@
+Note that in the general case, where the interventional distributions over the parents of $Y$, the effect of a given intervention by be non-identifiable even with good overlap and thus it is not possible to infer ahead of time the extent to which estimation is biased via confounding and thus the relative value of observational versus interventional data.
+
 \section*{Questions for Tor}
@@
 \item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
-\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
-\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality.
-When is a problem causal, what bounds does predictive accuracy give on causal models?
-To what degree is predictive accuracy a metric on which causal models can be judged?
+\item  When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
 \item Benchmark estimation algorithms
@@
 \end{itemize}
+
+In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
 

### 2017-09-28 thesis/recipe_for_causal_inference.tex
adds section for bob, starting work on talk

--- 2016-04-22 review/recipe_for_causal_inference.tex
@@
 \usetikzlibrary{shapes,arrows}
+\usepackage{xcolor}
+
+\definecolor{talk}{HTML}{729FCF}
+\definecolor{talk2}{HTML}{E8A753}
 \begin{document}
@@
 % Define block styles
-\tikzstyle{block} = [rectangle, draw, fill=blue!20, 
+\tikzstyle{block} = [rectangle, draw, fill=talk!50, 
     text width=10em, text centered, rounded corners, minimum height=3em]
@@
      \node [block, below of=success](ident){Great! Apply the formula from the do-caclulus to estimate effect.};
-     \node [block, fill=red!10, dashed, right of=ident,node distance=4.5cm](stats){Deal with pesky statistical issues introduced by finite data as you see fit};
+     \node [block, fill=talk2!50, dashed, right of=ident,node distance=4.5cm](stats){Deal with pesky statistical issues introduced by finite data as you see fit};
 

### 2017-09-28 thesis/thesis_map.tex
adds section for bob, starting work on talk

@@
 \usepackage{centernot}
+
+\usepackage{geometry}
+ \geometry{
+ a4paper,
+ left=40mm,
+ right=20mm,
+ top=20mm,
+ bottom=30mm,
+ footskip=10mm
+ }
+
+\usepackage{setspace}
 
@@
 \chapter*{Acknowledgements}
+
+%There are many people without whose support this thesis would not have been possible. I would like to thank my supervisor Dr Cheng Song Ong,
+
+% Dr Mark Reid, Dr Tiberio Ceatano and Prof. Robert Williamson.
+
+%My brother Tor for many interesting discussions on bandits.
+
+%To my daughters, Freya and Anouk,
+
+% supervisors
+% family (Tor)
+% thesis boot camp
+% scholarship
+
+%I am grateful to have recieved funding from an Australian Postgraduate Award and NICTA/Data61 scholarship to support this research.
+
+\onehalfspacing
+
 \input{abstract}
-
-%\chapter*{Acknowledgements}
 
@@
 By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}
+Finally, a model that can distill how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}
 
@@
 \paragraph{Thesis overview} This thesis is divided into three key chapters: learning from observational data, learning from interventional data and unifying the approaches.
-Chapter \ref{chap:causality} covers learning to act from observational data.
+Chapter \ref{chap:causality} covers learning to act from observational data, where the goal is to learn the outcome of an external intervention in a system from data obtained by observing it without control over which actions are selected.
 In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems.
@@
 This is the problem of observational causal inference.
+The key feature that distinguishes observational from interventional data is that the learning agent does not control the action about which they are trying to learn.
 
@@
 
+\paragraph{Granger causality}
+A discussion of approaches to (observational) causal inference would not be complete without a mention of Granger causality, \citep{Granger1969}.
+The fundamental idea behind Granger causality is to leverage the assumption that the future does not cause the past to test the existence and direction of a causal link between two time series.
+The basic approach is to test, for a pair of time series variables $X$ and $Y$, if $Y_t \ci (X_{1},...,X_{t-1}) | (Y_{1},...,Y_{t-1})$ - that is if the history of $X$ helps to predict $Y$ given the history of $Y$.
+The original formulation considered only pairs of variables and linear causal relationships but recent work has generalised the key idea to multiple variables and non-linear relationships.
+Unlike the previous models we have discussed, Granger causality does not provide us with a means to specify our assumptions about the causal structure between variables.
+Rather it aims to infer the causal structure of a structural equation model from observational data - subject to some assumptions.
+I would categorise Granger causality as method for \emph{causal discovery} in time series data, see the discussion of causal discovery versus causal effect estimation in section \ref{sec:causal effect estimation}.
+
 \subsection{Comparing and unifying the models}
@@
 
-Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation) \citep{Pearl2000}.
+Remarkably for models developed relatively independently in fields with very different approaches and problems, causal Bayesian networks, counterfactuals and structural equation models can be nicely unified for interventional queries (those that can be expressed with the do-notation) \citep{Pearl2000}.
+These queries, and the assumptions required to answer them, can be mapped between the frameworks in a straightforward way, allowing techniques developed within one framework to be immediately applied within another.
 If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables.
@@
 These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.
+The graphical approach to representing causal knowledge can be extended to cover these types of questions via Single World Intervention Graphs \citep{Richardson2013}, which explicitly represent counterfactual variables in the graph.
 
@@
 The Neyman-Rubin framework builds on the understanding of randomised experiments and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment.
+Treatment variables are typically discrete (often binary).
 This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}.
@@
 Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}.
+The central approach to estimation is regression - which naturally handles continuous variables while discrete variables are typically encoded as indicator variables.
 In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.
+However, these differences are not fundamental to the frameworks.
+Functional assumptions can be specified on the conditional distributions of (causal) Bayesian networks, counterfactuals can readily represent continuous treatments (eg $\cf{Y}{x}$), and structural equation models can represent complex non-linear relationships between both continuous and discrete variables.
 
@@
 \label{fig:rule3_original}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,main node/.style={observed}]
@@
 \label{fig:rule3_augmented}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,main node/.style={observed}]
@@
 \label{fig:rule3_after}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,main node/.style={observed}]
@@
 \label{fig:id1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,lt/.style={latent},ob/.style={observed}]
@@
 \label{fig:id2}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,lt/.style={latent},ob/.style={observed}]
@@
 \label{fig:id3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,lt/.style={latent},ob/.style={observed}]
@@
 \label{fig:nid1}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,lt/.style={latent},ob/.style={observed}]
@@
 \label{fig:nid2_instrumental_variable}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,lt/.style={latent},ob/.style={observed}]
@@
 \label{fig:nid3}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
   thick,lt/.style={latent},ob/.style={observed}]
@@
 \centering
-\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
+\includegraphics[scale=.48]{covariate_shift_estimation.pdf}
 \caption{
@@
 
-Reinforcement learning describes the problem of an agent interacting with an enviroment, learning by observing the outcome of its actions, with the goal of maximising some reward.
+Reinforcement learning describes the problem of an agent interacting with an environment, learning by observing the outcome of its actions, with the goal of maximising some reward.
 These problems, which also incorporate planning, are extremely difficult because the value of an action is generally not immediately clear.
-The state of the enviroment may evolve over time and according to previously selected actions.
-Actions available at future timesteps can depend on those taken in the past, and rewards may be obtained only after a long sequence of actions.
-This makes it extremely difficult for the agent to acurately attribute value to each chosen action along the path to obtaining a given reward.
+The state of the environment may evolve over time and according to previously selected actions.
+Actions available at future time steps can depend on those taken in the past, and rewards may be obtained only after a long sequence of actions.
+This makes it extremely difficult for the agent to accurately attribute value to each chosen action along the path to obtaining a given reward.
 
 We focus on a simpler class of problems within reinforcement learning known as multi-armed bandit problems.
-Multi-armed bandit problems also descibe an agent aiming to maximise some reward by interacting with an enviroment.
-However, the reward is observed immediately after the action is selected and the enviroment is stateless.
-Future rewards generated by the  enviroment in response to the agents actions do not depend on the previous actions of the agent.
-These assumptions are a reasonable approximation in settings where we face the same decision repeately with respect to many independent units or individuals and we have a good proxy for the outcome.
+Multi-armed bandit problems also describe an agent aiming to maximise some reward by interacting with an environment.
+However, the reward is observed immediately after the action is selected and the environment is stateless.
+Future rewards generated by the  environment in response to the agents actions do not depend on the previous actions of the agent.
+These assumptions are a reasonable approximation in settings where we face the same decision repeatedly with respect to many independent units or individuals and we have a good proxy for the outcome.
 In this chapter we describe the interventionalist approaches to learning to act for these types of problems, beginning with classical randomised experiments and then the, more adaptive, multi-armed bandit algorithms.
-We show how both observational causal inference and bandit algorithms can be viewed as generalisations of randomized experiments and highlight connections between the observational and interventional approaches to causal inference.
+We show how both observational causal inference and bandit algorithms can be viewed as generalisations of randomised experiments and highlight connections between the observational and interventional approaches to causal inference.
 
@@
 where $N_i(T)$ is the number of times arm $i$ was played up to timestep $T$ and $\Delta_i$ is the degree to which arm $i$ is sub-optimal, $\Delta_i = \mu^* - \mu_i$.
-The expectation of this variant of regret is the same as for the standard version defined in equation \ref{eqn:raw_regret2}.
+The expectation of this variant of regret is the same as for the  version defined in equation \ref{eqn:raw_regret2}.
 It depends only on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions.
@@
 variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry.
-Where there is no intervenion, the temperature, soil, and moisture vary naturally from season to season due to weather conditions.
+Where there is no intervention, the temperature, soil, and moisture vary naturally from season to season due to weather conditions.
 These are all observed along with the final crop yield at the end of each season.
@@
 The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action, and on which variables we receive post-action feedback.
-If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant, and the problem reduces to a standard multi-armed bandit problem.
-
-%This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.
-
+If we can (at no cost) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant, and the problem can be treated as a standard multi-armed bandit problem over the set of actions that fully specify the values for the parents of $Y$.
 If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
@@
 However, the regret bounds are not optimal in our setting.
-They also focus on cumulative regret, which cannot be used to guarantee low simple regret~\citep{Bubeck2009a}.
+They also focus on cumulative regret.
 The partial monitoring approach taken by \citet{wu2015online} could be applied (up to modifications for the simple regret) to the parallel bandit, but the resulting strategy require knowledge about the likelihood of each factor in advance, while our strategy learns this online.
@@
 
-Two pieces of recent work also consider applying ideas from causal inference to bandit problems.\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been
+Two pieces of recent work also consider applying ideas from causal inference to bandit problems.
+\citet{Bareinboim2015} demonstrate that in the presence of confounding variables the value that a variable would have taken had it not been
 intervened on can provide important contextual information.
@@
 For example, the focus is on the cumulative regret, and the context is observed before the action is taken and cannot be controlled by the learning agent.
-
-%\todof{write one sentence to explain how their work is different to ours}
-
 \citet{Ortega2014thompson} present an analysis and extension of Thompson sampling, assuming actions are causal interventions.
@@
 Combining their handling of  causal induction with our analysis is left as future work.
-% Key to Elias' paper is: observing the action an agent would take if it were allowed to make its natural choice can provide some information about hidden confounders that influence both the reward and the choice of action.
-Therefore, incorporating an agents natural choice as context may outperform a standard bandit that does not use that context.
-(Note: even in the presence of hidden confounders, including the agents natural choice as context only may improve the results.
-It is easy to come up with a counter example in which it does not).
-
-The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal framework by \citet{Bottou2013}, 
+
+The truncated importance weighted estimators used in \S\ref{sec:simple-regret-general} have been studied before in a causal setting by \citet{Bottou2013}, 
 where the focus is on learning from observational data, but not controlling the sampling process. They also briefly discuss some of the issues 
@@
 
+\subsubsection{Combining interventional and observational data}
+
+In the previous sections we have demonstrated that knowledge of the causal structure between variables allows the rewards of multiple actions to be estimated simultaneously in an online, interventional setting, leading to more rapid identification of the optimal arm.
+In many cases we may also have access to observational data; in other words data sampled from $\P{\vec{X},Y|do()}$.
+Under the assumptions we made for algorithm \ref{alg:general}, we can incorporate such data in a straightforward way.
+Let $N$ be the number of observational data points previously collected and $T$ again be the number of rounds in which we can intervene.
+Combining these two periods, the values for the parents of $Y$ are sampled from;
+
+\eqn {
+Q'(\parents{Y}(X)) = \frac{N}{N+T}\Pn{do()}{\parents{Y}(X)} + \frac{T}{N+T}\sum_{b \in \calA}{\eta_b \Pn{b}{\parents{Y}(X)}}
+}
+
+We can then let $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{Q'(\parents{Y}(X))}$ and select weights to minimise the worst case variance of the importance weighted estimator with respect to $Q'$.
+
+\eqn{
+\eta^* 
+ = \argmin_\eta \underbrace{\max_{a \in \calA} \EEa \left[\frac{\Pn{a}{\parents{Y}(X)}}{Q'(\parents{Y}(X))}\right]}_{m'(\eta)}\,.
+}
+
+Algorithm 3 is modified to sum over both $N$ and $T$ \footnote{In practise, to minimise the problem dependent regret, we would also incorporate an elimination step based on the observational data prior to selecting interventions.
+See the discussion on making better use of the reward in section \ref{sec:causal_bandit_discussion}} and obtains regret bounds;
+
+\eqn{
+\simpleregret \in \bigo{\sqrt{\frac{m'(\eta)}{N+T} \log\left(2(N+T)|\calA|\right)}}\,.
+}
+
+In the worse case, the observational distribution over the parents of $Y$ has no overlap with the distributions for any of the actions in $\actionspace$.
+This results in $m'(\eta^*)$ that is a factor of $\frac{N+T}{T}$ larger than $m(\eta^*)$ (which assumes all actions were selected from the optimal sampling distribution) and results in regret that decays with $\sqrt{T^{-1}}$ as in the case where no observational data is provided.
+
+\subsubsection{The relative value of observational versus interventional data} Another interesting question is the relative value of observational data versus interventional data in a given setting.
+Obtaining interventional data often involves substantial fixed costs in setting up a system to control the allocation of interventions.
+Ideally, we would be able to estimate of the additional value interventional data would provide prior to setting up such a system.
+The quantity $m(\eta)$ also provides a means to this goal.
+The regret bound in theorem \ref{thm:general} holds for any $\eta$.
+Thus we can compare the relative value of purely observational data as opposed to optimally designed interventional data by considering the ratio:
+
+\eqn{
+v_{obs} \in [0,1] = \frac{m(\eta^*)}{m(\eta_b = \ind{b=do()})}
+}
+
+If $v_{obs} = 0$, then there exists an action for which the reward cannot be estimated from observational data.
+Thus we cannot guarantee that we will identify the optimal action regardless of the quantity of available observational data.
+This does not imply that observational data will not improve estimation in conjunction with interventional data as we discussed in the previous section - just that observational data \emph{alone} is insufficient for best arm identification.
+If $v_{obs} = 1$, then the worst case regret from purely observational data matches that for interventional data.
+A value of $v_{obs} = .5$ would imply we would need twice as many samples to obtain the same regret bound from observational data as compared to interventional data.
+
+The ratio $v_{obs}$ can be computed prior to collecting any data, observational or interventional, subject to the assumption that the distribution over the parents of $Y$ given each action are known - as we assumed for Algorithm \ref{alg:general}.
+The approach is also not limited to the comparison of observational data with optimised interventional data.
+It can equally be applied to evaluate the potential for improvement on any other distribution over actions, for example we might want to evaluate the benefit of replacing a system that uniformly explores all actions with Algorithm \ref{alg:general}.
+We should note however that theorem \ref{thm:general} bounds the worst case regret, which occurs when the rewards for each action are sufficiently close that we must obtain good estimates for the value of all of the actions.
+Where the rewards are better separated, the \emph{problem-dependent} regret can be reduced by using an adaptive algorithm that ceases to explore actions that are sub-optimal with high probability, as we discuss in \S\ref{sec:causal_bandit_discussion}.
+Although $v_{obs}$ accurately captures the relative value of interventional versus observational data for a fixed design algorithm like Algorithm \ref{alg:general}, in general, using interventional data with an adaptive algorithm will lead to lower regret than selecting the best action on the basis of observational data even when $v_{obs} = 1$.
+However, the gap between the problem dependent and worst case regret is not known in advance - as the reward distributions are unknown - so the additional benefit of adaptive control cannot be computed ahead of time.
+
 \subsection{Experiments}

### 2017-09-30 thesis/thesis_map.tex
adds epigraph

@@
 \setcounter{secnumdepth}{3}
+\setlength{\epigraphwidth}{.6\textwidth}
 
@@
 
-\epigraph{In vain the grave, with retrospective Eye,
-Would from the apparent what conclude the why,
-Infer the Motive from the Deed, and show
-That what we chanced, was what we meant, to do.}{Alexander Pope}
+\epigraph{In vain the Sage, with retrospective eye,\\
+Would from th' apparent What conclude the Why,\\
+Infer the Motive from the Deed, and show,\\
+That what we chanced, was what we meant to do.}{Alexander Pope}
 
@@
 
-The ratio $v_{obs}$ can be computed prior to collecting any data, observational or interventional, subject to the assumption that the distribution over the parents of $Y$ given each action are known - as we assumed for Algorithm \ref{alg:general}.
+The ratio $v_{obs}$ can be computed prior to collecting any data, observational or interventional, if the distribution over the parents of $Y$ given each action are known - as we assumed for Algorithm \ref{alg:general}.
 The approach is also not limited to the comparison of observational data with optimised interventional data.

### 2017-10-03 thesis/talk/do_calculus_examples.tex
backup

\documentclass{article}
\usepackage{tikz} 
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{standalone}

\usepackage{subcaption}
\usepackage{multicol,caption}
\usepackage{geometry}
 \geometry{
 a4paper,
 total={210mm,297mm},
 left=20mm,
 right=20mm,
 top=20mm,
 bottom=20mm,
 }

\usetikzlibrary{arrows,positioning,shapes} 
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
\thispagestyle{empty}
\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\definecolor{talk}{HTML}{729FCF}
\definecolor{talk2}{HTML}{E8A753}
\definecolor{grey}{HTML}{DBDCDD}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
   	circle,
   	rounded corners,
   	draw=black, thick,
   	minimum width=2.3em,
   	minimum height=2.3em,
   	font=\footnotesize,
   	text centered,
   	fill=talk!50
   },
   latent/.style={
   	circle,
   	rounded corners,
   	draw=black, thick, dashed,
   	minimum width=2.2em,
   	minimum height=2.2em,
   	font=\footnotesize,
   	text centered
   },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 },
observedrect/.style={
	rectangle,
	rounded corners,
	draw=black, thick,
	minimum width=3em,
	minimum height=1.5em,
	font=\footnotesize,
	text centered,
	fill=talk2!80
},  
 target/.style={
	circle,
	rounded corners,
	draw=black, thick,
	minimum width=2.3em,
	minimum height=2.3em,
	font=\footnotesize,
	text centered,
	fill=talk2!80
},
empty/.style={
	circle,
	rounded corners,
	minimum width=.5em,
	minimum height=.5em,
	font=\footnotesize,
	text centered,
},
}

\begin{document}
\pagestyle{empty}

\begin{figure}
	\centering
	\begin{subfigure}[t]{0.3\textwidth}
	\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=5em,thick,node/.style={observed}, tg/.style={target},act/.style={observed},lt/.style={empty}]
	\node[lt](3){};
	\node[act, below left of=3](1){X};
	\node[tg, below right of = 3](2){Y};
	\node[above=of 1,yshift=-4em](T){\includegraphics[width=1cm]{tick.png}};
	\path[]
	(1) edge (2);
	\end{tikzpicture}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=5em,thick,node/.style={observed}, tg/.style={target},act/.style={observed},lt/.style={latent}]
	\node[lt](3){U};
	\node[act, below left of=3](1){X};
	\node[tg, below right of = 3](2){Y};
	\node[above=of 1,yshift=-4em](T){\includegraphics[width=1cm]{cross.png}};
	\path[]
	(1) edge (2)
	(3) edge (1) edge (2);
	\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
	\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=5em,thick,node/.style={observed}, tg/.style={target},act/.style={observed},lt/.style={latent}]
	\node[lt](3){U};
	\node[act, below left of=3](1){X};
	\node[tg, below right of = 3](2){Y};
	\node[node,above left of=1](4){Z};
	\node[above=of 1,yshift=-4em](T){\includegraphics[width=1cm]{cross.png}};
	\path[]
	(1) edge (2)
	(3) edge (1) edge (2)
	(4) edge (1);
	\end{tikzpicture}
\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\vspace{3em}
	\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=5em,thick,node/.style={observed}, tg/.style={target},act/.style={observed},lt/.style={empty}]
\node[node](3){Z};
\node[act, below left of=3](1){X};
\node[tg, below right of = 3](2){Y};
\node[above=of 1,yshift=-4em](T){\includegraphics[width=1cm]{tick.png}};
\path[]
(1) edge (2)
(3) edge (1) edge (2);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
	\vspace{3em}
	\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=5em,thick,node/.style={observed}, tg/.style={target},act/.style={observed},lt/.style={latent}]
	\node[act](1){X};
	\node[node,right of = 1](4){Z};
	\node[lt, above of=4,yshift=-1.5em](3){U};
	\node[tg, right of = 4](2){Y};
	\node[above=of 1,yshift=-4em](T){\includegraphics[width=1cm]{tick.png}};
	\path[]
	(1) edge (4)
	(4) edge (2)
	(3) edge (1) edge (2);
	\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
	\vspace{3em}
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=5em,thick,node/.style={observed}, tg/.style={target},act/.style={observed},lt/.style={latent}]
\node[act](1){X};
\node[node,right of = 1](4){Z};
\node[lt, above of=4,yshift=-1.5em](3){U};
\node[tg, right of = 4](2){Y};
\node[lt, below right of = 4,xshift=-.9em,yshift=0.5em](5){U};
\node[above=of 1,yshift=-4em](T){\includegraphics[width=1cm]{cross.png}};
\path[]
(1) edge (4)
(4) edge (2)
(3) edge (1) edge (2)
(5) edge (4) edge (2);
\end{tikzpicture}
\end{subfigure}
\end{figure}

\end{document}

### 2017-10-03 thesis/talk/recipe_for_causal_inference.tex
backup

\documentclass[tikz]{standalone}

\usepackage[latin1]{inputenc}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{xcolor}

\definecolor{talk}{HTML}{729FCF}
\definecolor{talk2}{HTML}{E8A753}
\definecolor{grey}{HTML}{DBDCDD}
\begin{document}
\pagestyle{empty}

\tikzset{
	%Define standard arrow tip
	>=stealth',
	%Define style for boxes
	punkt/.style={
		circle,
		rounded corners,
		draw=black, 
		text width=1em,
		minimum height=1em,
		text centered},
	observed/.style={
		circle,
		rounded corners,
		draw=black, thick,
		minimum width=2.3em,
		minimum height=2.3em,
		font=\footnotesize,
		text centered,
		fill=talk
	},
	latent/.style={
		circle,
		rounded corners,
		draw=black, thick, dashed,
		minimum width=2.2em,
		minimum height=2.2em,
		font=\footnotesize,
		text centered
	},
	target/.style={
		circle,
		rounded corners,
		draw=black, thick,
		minimum width=2.2em,
		minimum height=2.2em,
		font=\footnotesize,
		text centered,
		fill=black!20!white,
	},
	observedrect/.style={
		rectangle,
		rounded corners,
		draw=black, thick,
		minimum width=6em,
		minimum height=2em,
		font=\footnotesize,
		text centered,
		fill=talk!50
	},
	latentrect/.style={
		rectangle,
		rounded corners,
		draw=black, thick, dashed,
		minimum width=2.2em,
		minimum height=2.2em,
		font=\footnotesize,
		text centered
	},
	targetrect/.style={
		rectangle,
		rounded corners,
		draw=black, thick,
		minimum width=6em,
		minimum height=2em,
		font=\footnotesize,
		text centered,
		fill=black!20!white,
	},
	empty/.style={
		circle,
		rounded corners,
		minimum width=.5em,
		minimum height=.5em,
		font=\footnotesize,
		text centered,
	},
	% Define arrow style
	pil/.style={
		o->,
		thick,
		shorten <=2pt,
		shorten >=2pt,},
	sh/.style={ shade, shading=axis, left color=red, right color=green,
		shading angle=45 },
	line/.style={
		draw,
		-latex
	},
    ln/.style={
    	draw
    },
	block/.style={
		rectangle, 
		draw, 
		fill=talk!50, 
		text width=8em, 
		text centered, 
		rounded corners, 
		minimum height=3em
	}   
}

\begin{tikzpicture}[node distance = 2cm, node/.style={punkt}]
% Place nodes
\node[node](or){or};
\node[block, right of = or](learngraph){Apply \emph{Causal Discovery} to learn the graph};
\node [block,left of=or] (drawgraph) {Use prior knowledge to draw the graph};
\node [block, below of=or] (docalc) {Apply the Do Calculus};
\node [block, below left of=docalc,node distance=3.5cm] (fail) {Non Identifiable};
\node [block, below right of=docalc, node distance=3.5cm] (success) {Identifiable};
\node [block, below of=fail](notident){Bugger! You will have to make additional assumptions or do experiments};
\node [block, below of=success](ident){Great! Apply the formula from the do-caclulus to estimate effect.};
\node [block, fill=talk2!50, dashed, right of=ident,node distance=4.5cm](stats){Deal with pesky statistical issues introduced by finite data as you see fit};

% Draw edges
\path[ln](or) -- (learngraph);
\path[ln](or) -- (drawgraph);
\path[line] (or) -- (docalc);
\path[line] (docalc) -- (fail);
\path[line] (docalc) -- (success);
\path[line] (fail) -- (notident);
\path[line] (success) -- (ident);
\path[line] (stats) -- (ident);

\end{tikzpicture}

\end{document}

### 2017-10-03 thesis/talk/smoking.tex
backup

--- 2015-09-29 smoking.tex
@@
 
+\definecolor{talk}{HTML}{729FCF}
+\definecolor{talk2}{HTML}{E8A753}
+\definecolor{grey}{HTML}{DBDCDD}
+
 \pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}
@@
            text centered,
-           fill=blue!20!white},
+           fill=talk!50},
      latent/.style={	(Switch) edge (X)
@@
 \node[main node, below right=of X](C){Cancer};
-\node[main node, fill=red!20!yellow, above left=of X](Switch){$\hat{X}$};
+\node[main node, fill=talk2!50, above left=of X](Switch){$\hat{X}$};
 \node[main node, below left=of X](A){Asthma};

### 2017-10-03 thesis/talk/multi-armed-bandit-figure.py

@@ -1,0 +1,42 @@
+# This scipt generates a bunch of plots of distributions representing
+# the rewards for a hypothetical bandit problem.
+
+import matplotlib.pyplot as plt
+import numpy as np
+from scipy.stats import beta
+
+params = [(2,2),(3,7),(1,1),(5,5),(4,2),(3,1)]
+
+# Two subplots, unpack the axes array immediately
+x = np.linspace(0,1,100)
+
+f, ax = plt.subplots(1, len(params),sharey=True)
+f.patch.set_facecolor('white')
+plt.setp(ax, xticks=[0.1, 0.5, 0.9], xticklabels=['a', 'b', 'c'],  yticks=[])
+ax[0].set_ylabel("$P(y)$")
+ax[0].set_xlabel("$y$")
+for indx,(a,b) in enumerate(params):
+    mean = a/(a+b)
+    ax[indx].plot(x,beta.pdf(x,a,b))
+    ax[indx].axvline(x = mean,c="red",linestyle="--")
+    ax[indx].set_xticks([mean])
+    ax[indx].set_xticklabels(['$\mu_{0}$'.format(indx)])
+
+    
+    
+
+#x = [0,1,2]
+#y = [90,40,65]
+#labels = ['high', 'low', 37337]
+#plt.plot(x,y, 'r')
+#plt.xticks(x, labels, rotation='vertical')
+
+
+plt.show()
+    
+
+#ax1.plot(x,beta.pdf(x,1,1),'r-', lw=5, alpha=0.6, label='beta pdf')
+#ax1.set_title('Sharing Y axis')
+
+#print(beta.pdf(x,1,1))
+#plt.show()

### 2017-10-12 thesis/thesis_map.tex
more talk stuff

--- 2017-09-28 thesis/thesis_map.tex
@@
 
-\epigraph{In vain the grave, with retrospective Eye,
+\epigraph{In vain the sage, with retrospective Eye,
 Would from the apparent what conclude the why,

### 2017-10-12 causalpy/causalpy/bandits/algorithms.py

@@ -111,8 +111,7 @@ class DiscreteContextualBandit(object):
         
         for context, bandit in enumerate(self.bandits):
             self.arms[self.contexts == context] = bandit.arms[0:bandit.t]
-        
-        
+            
 class ThompsonSampling(object):
     """ Sample actions via the Thomson sampling approach and return the empirically best arm 
         when the number of rounds is exhausted """
@@ -203,13 +202,6 @@ class UCB(object):
         self.best_action = argmax_rand(mu)
         return best_reward - model.expected_rewards[self.best_action]
     
-    
-        
-    
-        
-        
-        
-
 class AlphaUCB(UCB):
     """ 
     Implementation based on ... 

### 2017-10-12 causalpy/causalpy/bandits/enviroments.py

@@ -159,6 +159,12 @@ class BasicBernoulliModel(object):
         expected_rewards = np.full(K,.5)
         expected_rewards[-1] = .5 + epsilon # reward of one arm is just better than the others.
         return cls(expected_rewards)
+        
+    @classmethod
+    def create_random(cls,K):
+        """ creates a bandit where the expected rewards are sampled from a uniform distribution """
+        expected_rewards = np.random.uniform(size=K)
+        return cls(expected_rewards)
     
     @classmethod
     def create_worst_case(cls,K,T):

### 2017-10-14 thesis/talk/do_calculus_examples.tex

@@
 	text centered,
-	fill=talk2!80
+	fill=talk!50
 },  
@@
 	text centered,
-	fill=talk2!80
+	fill=talk!50
 },

### 2017-10-14 thesis/talk/latexit_segments.tex

At each timestep $t \in 1,...,T$:
\begin{enumerate}
\itemsep0em 
\item The agent sets the value of zero or one variables, $do()$ or $do(X_i = j)$.
\item Non-intervened variables sampled from underlying distributions.
\item The agent receives reward $Y$ sampled from $P(Y|x_1,...,x_N)$.
\item The values of all variables, $x_1,...,x_N$, are revealed to the agent.
\end{enumerate}
The variables are binary $\implies$ there are $2N+1$ arms.\\
If the agent opts to observe, $do()$, it will obtain an estimate for half the arms

\text{for } t \in & \set{1,...,T}:\\ 
&\text{Sample action } a_t \text{ from } \eta \\
\text{for } a \in & \actionspace:\\ 
&\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t \frac{\Pn{a}{\parents{Y}(X_t)}}{\sum_{a \in \calA} \eta_a \P{\parents{Y}(X_t)|a}}

### 2017-10-14 thesis/talk/unfaithful_chocolate_obesity.tex

--- 2015-09-29 smoking.tex
@@
 
-\begin{enumerate}
-\item D-separation still applies after intervention.
+%\multicolumn{'num_cols'}{'alignment'}{'contents'}
 
-${(Cancer \ci Asthma|Smoke)_{G_{\overline{X}}} \implies P(Cancer|do(Smoke),Asthma) = P(Cancer|do(Smoke)}$
-\item If there are no backdoor paths from $X$ to $Y$ then intervention$\equiv$observation.
-
-${(\hat{X} \ci Cancer|X,Poverty)_{G^{\dagger}} \implies P(Cancer|do(Smoke),Poverty) = P(Cancer|Smoke,Poverty)}$
-\item If there are only backdoor paths from $X$ to $Y$ then intervention doesn't change $P(Y)$.
-
-$(\hat{X} \ci Diet)_{G^{\dagger}} \implies P(Diet|do(Smoke)) = P(Diet)$
-\end{enumerate}
 \begin{figure}[h]
 \centering
-\begin{subfigure}[t]{0.45\textwidth}
+\begin{subfigure}[t]{0.6\textwidth}
 \centering
-\caption{$G^{\dagger}$}
+\caption{True causal graph generating $P$}
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,thick,main node/.style={observed}]
-\node[main node](X){X=Smoke};
-\node[main node, above right=of X](P){Poverty};
-\node[main node, below right=of P](D){Diet};
-\node[main node, below right=of X](C){Cancer};
-\node[main node, fill=red!20!yellow, above left=of X](Switch){$\hat{X}$};
-\node[main node, below left=of X](A){Asthma};
+\node[main node, align=center,minimum height=4.3em](C){\textbf{Chocolate} \\ \begin{tabular}{| c | c | c |}
+  \hline			
+  Income & H & L \\
+  \hline
+  P(C=H) & .8639 & .2\\
+  \hline
+\end{tabular}};
+\node[main node, above right=of C, align=center](I){\textbf{Income} \\ $P(I=High) = 0.5$};
+\node[main node, below right=of I, align=center,minimum height=4.3em](O){\textbf{Obesity} \\ \begin{tabular}{| c | c | c | c | c |}
+  \hline			
+  Income & \multicolumn{2}{| c |}{H} & \multicolumn{2}{| c |}{L} \\
+  \hline
+  Chocolate & H & L & H & L \\
+  \hline
+  $P(O=Y)$ & .4 & .2 & .7 & .5 \\
+  \hline
+\end{tabular}};
 \path[]
-	(Switch) edge (X)
-	(X) edge (C)
-	(D) edge (C)
-	(X) edge (A)
-	(P) edge (D) edge (X);
+	(I) edge (C) edge (O)
+	(C) edge (O);
 \end{tikzpicture}
 \end{subfigure}
-\begin{subfigure}[t]{0.45\textwidth}
+\begin{subfigure}[t]{0.3\textwidth}
 \centering
-\caption{$G_{\overline{X}}$}
+\caption{Perfect map for $P, \;(C \ci O)$}
 \begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=0.45cm,  thick,main node/.style={observed}, shaded/.style={latent}]
-\node[main node](X){X=Smoke};
-\node[main node, above right=of X](P){Poverty};
-\node[main node, below right=of P](D){Diet};
-\node[main node, below right=of X](C){Cancer};
-\node[main node, below left=of X](A){Asthma};
+\node[main node](C){Chocolate};
+\node[main node, below right=of C](I){Income};
+\node[main node, above right=of I](O){Obesity};
 \path[]
-	(X) edge (C) edge (A)
-	(D) edge (C)
-	(P) edge (D);
+	(C) edge (I)
+	(O) edge (I);
 \end{tikzpicture}

### 2017-10-20 thesis/abstract.tex
finalised talk, replaced most instances of the word framework, fixed spelling and repeated word errors, completes aknowledgements.

@@
 The goal of this thesis to connect and unify these key approaches.
-I introduce causal bandit problems: a framework that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis.
+I introduce causal bandit problems: a synthesis that combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcement learning problems that are simple enough to admit formal analysis.
 I show that knowledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be exploited by existing algorithms.
-I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.
+I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and better than algorithms that do not use the additional causal information.

### 2017-10-20 thesis/offcuts.tex
finalised talk, replaced most instances of the word framework, fixed spelling and repeated word errors, completes aknowledgements.

@@
 Yes, if the way in which features get their values changes, then causal models will be invariant to that in a way that non-causal ones are not.
+
+ \chapter{Causation vs Prediction}
+
+ Mixing continuous experimentation at an individual level with batch approaches.
+There may be fixed costs associated with \quotes{enabling} certain actions.
+For example, social programs may have some fixed overheads, no matter how many people are enrolled in them, etc.
+Reuse data obtained to optimise for one outcome to learn about optimising for another.
+We don't have a single outcome or loss-measure that fully incorporates everything we care about for the majority of problems.
+
+ The two cultures \citep{Breiman2001}
+ - wrong because suggests we can just use the machine learning approach for everything.
+Does not address the problems that a) structural assumptions required for causal inference.
+b) we cannot use cross-validation to assess the validity of a causal model (really b is the whole point).
+
+ To explain or to predict \citep{Shmueli2010}
+ - wrong because they assume all the parts of explanatory modelling are required - so black box models cannot be used.
+
+ Predict the relationship between X and Y as a way to stratify to reduce the variance in an experiment.
+
+ They consider a different definition of explanatory models,
+
+ Definition as backwards.
+Explanatory models focus on theory because some theory (or equivalently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
+
+ Causal modelling frameworks allow us to separate assumptions about the structure of the relationships between variables from assumptions about the functional form.
+
+ What can be data driven and what cannot? (for a given set of assumptions).
+
+ Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the language of EoP but are built on a hybrid of both approaches.

### 2017-10-20 thesis/thesis_map.tex
getting ready for talk
finalised talk, replaced most instances of the word framework, fixed spelling and repeated word errors, completes aknowledgements.

--- 2017-09-30 thesis/thesis_map.tex
@@
 
-%There are many people without whose support this thesis would not have been possible. I would like to thank my supervisor Dr Cheng Song Ong,
-
-% Dr Mark Reid, Dr Tiberio Ceatano and Prof. Robert Williamson.
-
-%My brother Tor for many interesting discussions on bandits.
-
-%To my daughters, Freya and Anouk,
-
-% supervisors
-% family (Tor)
-% thesis boot camp
-% scholarship
-
-%I am grateful to have recieved funding from an Australian Postgraduate Award and NICTA/Data61 scholarship to support this research.
+There are many people without whose support this thesis would not have been possible.
+I would like to thank my supervisors Dr Cheng Song Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof.
+Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters.
+I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I need to get past the blank page and get words out.
+Finally I want to thank my wonderful husband for his unwavering support and willingness to take on the household chaos, and my two beautiful daughters Freya and Anouk for their patience while mummy worked on her "tesis".
+
+I am also grateful for the financial support provided by an Australian Postgraduate Award and a NICTA/Data61 top up scholarship.
 
@@
 By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}.
-Finally, a model that can distill how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}
+Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}
 
@@
 \paragraph{Contributions}  The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints.
-My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
-This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
+My major contribution unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits.
+This synthesis allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem.
 I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.
@@
 
-In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems: a framework that unifies causal graphical models with multi-armed bandit problems.
+In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems that unify causal graphical models and multi-armed bandit problems.
 Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model.
-I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.
+I show how this framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.
 
@@
 I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning.
-In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}.
-Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems.
+In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known a priori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}.
+I also show how this measure can be used to quantify the value of optimised interventional data over purely observational data.
+Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of both the parallel and general causal bandit algorithms on causal bandit problems.
 Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.
@@
 A central component of estimating causal effects from observational data is determining if and how we can write expressions for the interventional distributions of interest in terms of observational quantities, which can be measured.
-We did this on an ad-hoc basis to resolve the examples discussed in section \ref{sec:causal_models}.
+We did this on an ad hoc basis to resolve the examples discussed in section \ref{sec:causal_models}.
 In this section we describe a principled approach to mapping observational quantities to interventional ones and then, in section \ref{sec:causal_effect_under_ignorability}, discuss the key issues involved in estimating such expressions from finite sample data.
@@
 \vspace*{.3cm}
-\begin{definition}[Non-parametric identifiablity]
+\begin{definition}[Non-parametric identifiability]
 Let $G$ be a causal graph containing observed variables $\boldsymbol{V}$ and latent variables $\boldsymbol{U}$ and let $\P{\cdot}$ be any positive distribution over $\boldsymbol{V}$.
@@
 \end{subfigure}
-\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
+\caption{Examples of identifiable and non-identifiable queries. In sub figures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In sub figures (d), (e) and (f) it is not.}
 \label{fig:identify}
@@
 The significance of this should not be underestimated.
-Cross-validation has allowed applied machine learning to succeed with a very atheoretical approach on the basis that we can identify when a model is successful.
+Cross-validation has allowed applied machine learning to succeed with a very a theoretical approach on the basis that we can identify when a model is successful.
 With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention.
@@
 
-Undertaking a Bayesian regression of $X$ and $Z$ against $Y$ and using intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ will lead to selection of something centered around $w_{yz}$.
+Undertaking a Bayesian regression of $X$ and $Z$ against $Y$ and using intuition based on the causal relationship between $Z$ and $Y$ to select a prior for $\beta$ will lead to selection of something centred around $w_{yz}$.
 However, this is incorrect and can introduce bias into our estimate of the coefficient for $w_{yx}$, as demonstrated in figure \ref{fig:careful-prior-example}.
@@
 \begin{subfigure}[t]{0.99\textwidth}
-     \caption{No prior on $\beta$, the posterior on $w_{yx}$ is centered around its true value}
+     \caption{No prior on $\beta$, the posterior on $w_{yx}$ is centred around its true value}
     \includegraphics[scale=1]{prior_no_prior}
@@
 \begin{subfigure}[t]{0.99\textwidth}
-     \caption{Prior on $\beta$ centered around $w_{yz}$, the causal effect of $Z$ on $Y$, $Prior(\beta)= N(w_{yz},\sigma = 0.5)$.
+     \caption{Prior on $\beta$ centred around $w_{yz}$, the causal effect of $Z$ on $Y$, $Prior(\beta)= N(w_{yz},\sigma = 0.5)$.
 The posterior on $w_{yx}$ is biased away from its true value (as is the posterior on $\beta$ but this is not the key quantity of interest).}
@@
 \end{subfigure}
-\caption{An example demonstrating how selecting a prior for a nuisance parameter centered around the causal effect of that parameter on the outcome, rather than its association with the outcome, can bias estimates for actual parameter of interest.
+\caption{An example demonstrating how selecting a prior for a nuisance parameter centred around the causal effect of that parameter on the outcome, rather than its association with the outcome, can bias estimates for actual parameter of interest.
 We sample $n=1000$ data points from the joint distribution defined by $P(U)P(Z|U)P(X|Z)P(Y|U,Z,X) \sim N(0,1)N(2U,0.09)N(0.5Z,1)N(3U -Z + 0.5X , 0.25)$ and fit the model $Y \sim N(w_yx X+\beta Z,\epsilon)$ in Stan and plot the posterior distributions over $w_yx$ and $\beta$.
 Figure (a) shows the results with no prior (equivalently an improper prior) for both $w_yx$ and $\beta$.
-Figure (b) shows the results when we place a Gaussian prior centered around the $w_yz$ on the distribution for $\beta$.
+Figure (b) shows the results when we place a Gaussian prior centred around the $w_yz$ on the distribution for $\beta$.
 All models were fit using Stan \citep{carpenter2016stan}}\label{fig:careful-prior-example}
@@
 More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, (figure \ref{fig:random_experiment_network_stratified}).
-This necessitates that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under under ignorability (\S\ref{sec:causal_effect_under_ignorability}).
+This necessitates that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under ignorability (\S\ref{sec:causal_effect_under_ignorability}).
 The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
@@
 \vspace{0.5cm}
-\begin{definition}[Stochasitic k-armed bandit problem]
+\begin{definition}[Stochastic k-armed bandit problem]
 Let $\actionspace = \set{1,...,k}$ be the set of available actions (or bandit arms) and $\P{\vec{y}} = \P{\cf{Y}{1},...,\cf{Y}{k}}$ be a joint distribution over the rewards for each action.
@@
 \caption{The actual performance of the UCB algorithm can be substantially better than suggested by the upper bound, particularly for small $T$.
-The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $bernoulli([.3,.7])$ over 1000 simulations.
+The solid curve shows the mean expected regret associated with the sequence of arms chosen by UCB-1 with $k=2$ arms and the rewards sampled from $Bernoulli([.3,.7])$ over 1000 simulations.
 The dashed curve shows the corresponding upper bound given by the minimum of $T\Delta_{max}$ and equation\ref{eqn:ucb_problem_dependent_regret}.}
@@
 
-\section{A framework to connect bandit problems with observational causal inference}
+\section{A synthesis that connects bandit problems and observational causal inference}
 \label{sec:intro_causal_bandit}
@@
 
-The classical $K$-armed stochastic bandit problem can be recovered in our framework by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
+The classical $K$-armed stochastic bandit problem can be recovered in the causal bandit setting by considering a simple causal model with one edge connecting a single variable $X$ that can take on $K$ values to a reward variable $Y \in \set{0,1}$ where $\P{Y = 1|X} = r(X)$ for some arbitrary but unknown, real-valued function $r$.
 The set of allowed actions in this case is $\mathcal{A} = \{ do(X = k) \colon k \in \{1, \ldots, K\}\}$.
@@
 
-Our framework bears a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
-However, a crucial difference is that in our model the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
+Causal bandit problems have a superficial similarity to contextual bandit problems, \S\ref{sec:contextual-bandits}, since the extra observations on non-intervened variables might be viewed as context for selecting an intervention.
+However, a crucial difference is that we have focused on analysing settings where we receive post-action feedback, in which the extra observations are only revealed \emph{after} selecting an intervention and hence cannot be used as context.
 
@@
 
-%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimize cumulative regret.
+%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimise cumulative regret.
 They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
-% Partial monitoring is a very general framework for for decoupling the feedback from the action and reward.
+ Partial monitoring is a very general framework for decoupling the feedback from the action and reward.
 It can be used to classify problems into one of four categories, trivial with no regret, easy with $R_T = \bigthetatilde{\sqrt{T}}$ , hard with $R_T = \bigtheta{T^{2/3}}$ and hopeless with $R_T = \bigomega{T}$ \citep{Bartok2014}.
-Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as $K$, which is the key focus of incorporating causal structure.
+Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as the number of actions, which is the key focus of incorporating causal structure.
 
@@
 
-I have developed a framework that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this framework encodes some key existing problems in the literature.
+I have developed an approach that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this conceptualisation encodes some key existing problems in the literature.
 I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.
+I introduce a metric that captures the difficulty of causal bandit problems and (under strong assumptions) allows the value of optimised interventional data over purely observational data to be computed prior to intervening in the system.
 
@@
 There is much exciting work remaining to be done.
-Although the causal bandit framework can capture contextual information as well as post-action feedback, I have formally analysed and developed algorithms only for the latter.
+Although the causal bandit setting can capture contextual information as well as post-action feedback, I have formally analysed and developed algorithms only for the latter.
 Additionally, to make the problem tractable, I made the (major) assumption that the interventional distribution over the parents of the outcome was known.
@@
 \end{document}
-
-\chapter{Causation vs Prediction}
-
-Mixing continuous experimentation at an individual level with batch approaches.
-There may be fixed costs associated with \quotes{enabling} certain actions.
-For example, social programs may have some fixed overheads, no matter how many people are enrolled in them, etc.
-Reuse data obtained to optimise for one outcome to learn about optimising for another.
-We don't have a single outcome or loss-measure that fully incorporates everything we care about for the majority of problems.
-
-The two cultures \citep{Breiman2001}
- - wrong because suggests we can just use the machine learning approach for everything.
-Does not address the problems that a) structural assumptions required for causal inference.
-b) we cannot use cross-validation to assess the validity of a causal model (really b is the whole point).
-
-To explain or to predict \citep{Shmueli2010}
-- wrong because they assume all the parts of explanatory modelling are required - so black box models cannot be used.
-
-Predict the relationship between X and Y as a way to stratify to reduce the variance in an experiment.
-
-They consider a different definition of explanatory models,
-
-Definition as backwards.
-Explanatory models focus on theory because some theory (or equivalently stated assumptions) are required in order to predict the outcome of an intervention from observational data.
-
-Causal modelling frameworks allow us to separate assumptions about the structure of the relationships between variables from assumptions about the functional form.
-
-What can be data driven and what cannot? (for a given set of assumptions).
-
-Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the language of EoP but are built on a hybrid of both approaches.

### 2017-10-22 thesis/thesis_map.tex
fixes typos

@@
 Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters.
-I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I need to get past the blank page and get words out.
+I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I needed to get past the blank page and get words out.
 Finally I want to thank my wonderful husband for his unwavering support and willingness to take on the household chaos, and my two beautiful daughters Freya and Anouk for their patience while mummy worked on her "tesis".
@@
 If we developed a drug to suppress particular genes, could we cure multiple sclerosis?
-Would delaying teen-aged pregnancies improve the outcome for their children.
+Would delaying teenage pregnancies improve the outcome for their children.
 
@@
 
-The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined.
+The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still viewed as poorly defined.
 This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}.
@@
 The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample.
-Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated.
+Given finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated.
 For example, based on data from the years 1999 to 2009, the age of Miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}.
@@
 Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model.
-I show how this framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.
+I show how the causal bandit framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate when causal bandit problems reduce to different existing bandit settings depending what information is observable and when.
 
 In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected.
-I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms.
+I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by existing bandit algorithms.
 In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of the causal bandit problem, which I refer to as the \emph{parallel bandit problem}.
@@
 Questions around designing policies or interventions play a central role in economics.
-Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised \citep{Pearl2000}) framework and associated set of methods for estimating causal effects.
+Thus they have transformed simultaneous equations into a powerful framework and associated set of methods for estimating causal effects.
+The is also a rich strand of work on using the assumptions that can be encoded in structural equation models, also known as functional causal models to discover the structure and direction of causal relationships - see for example \citet{mooij2016distinguishing}.
 Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks.
@@
 Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models.
-This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.
+This makes it worthwhile for researchers interested in causality to develop an understanding of all these viewpoints.
 
@@
 Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds.
-We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES).
-We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
+We have a large cross-sectional data set on a group of twenty year olds that records if they attended preschool, if they graduated high school and their parents socio-economic status (SES).
+We will let $X\in \set{0,1}$ indicate if an individual attended preschool, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
 \end{example}
@@
 Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$?
-The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient.
+The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease subtype of the patient.
 Since $X$ is assigned via deliberate randomisation, it cannot be affected by any latent variables.
@@
 Similarly $Y^{0}$ represents the potential outcome for the placebo.
-The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
+The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery if everyone received the placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$.
 We can estimate (from an experimental or observational study):
@@
 In general the treatment assignment will not be independent of the potential outcomes.
-In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e.
-$X \nci \cf{Y}{0}$.
-Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e.
-$X \nci \cf{Y}{1}$.
+In example \ref{exm:adjusting}, the children from wealthy families could be more likely to attend preschool but also more likely to do better in school regardless, i.e $X \nci \cf{Y}{0}$ and $X \nci \cf{Y}{1}$.
 A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.
@@
 
-Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}.
+Assuming that within each socio-economic status level, attendance at preschool is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal preschool program can be computed from equation \ref{eqn:counterfactual1}.
 Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.
@@
 This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual.
-As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school.
+As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend preschool.
 The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them.
@@
 The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}.
-Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
+Intervening to send all children to preschool replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.
 
@@
 \begin{theorem}
-\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
+\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-descendents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
 \end{theorem}
@@
 
-In some cases, the average causal effect for some sub-group of the population is of prime interest.
+In some cases, the average causal effect for some subgroup of the population is of prime interest.
 A particularly common example of this is the average treatment effect of the treatment of the treated (ATT).
@@
 The key approaches are based on matching on covariates, propensity score methods and regression.
-We now examine these approaches from a machine learning perceptive.
+We now examine these approaches from a machine learning perspective.
 
@@
 \subsubsection{Propensity scores}
-If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
+If the parametric model is missspecified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled.
 The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.
@@
 \caption{
-Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data.
+Parametric regression may yield poor estimates of causal effects if the model is missspecified, even if the model fits well over the domain of the training data.
 In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities.
@@
 
-The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \c%fitep{Robins1994a,Kang2007}.
+The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{Robins1994a,Kang2007}.
 Doubly robust methods are asymptotically unbiased so long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.
@@
 
-This formulation of the regression estimator highlights the missing data aspect of casual effect estimation.
+This formulation of the regression estimator highlights the missing data aspect of causal effect estimation.
 For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment.
@@
 With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.
-I have found no examples of such an approach.
 
@@
 Suppose data is generated by the linear Gaussian Bayesian network below.
-The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$.
+The goal is to estimate the causal effect of $X$ on $Y$, that is identify the value of the coefficient $w_{yx}$.
 The variable $U$ is latent.
@@
 
-where
-
-\eqn{
-\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and}\\
+where,
+
+\eqn{
+\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and,}\\
 \epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
@@
 Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning.
-For example, in a drug trial , assigning people a sub-optimal treatment has real world costs.
+For example, in a drug trial, assigning people a sub-optimal treatment has real world costs.
 This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain.
@@
 However, the reward is observed immediately after the action is selected and the environment is stateless.
-Future rewards generated by the  environment in response to the agents actions do not depend on the previous actions of the agent.
+Future rewards generated by the  environment in response to the agent's actions do not depend on the previous actions of the agent.
 These assumptions are a reasonable approximation in settings where we face the same decision repeatedly with respect to many independent units or individuals and we have a good proxy for the outcome.
@@
 More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, (figure \ref{fig:random_experiment_network_stratified}).
-This necessitates that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under ignorability (\S\ref{sec:causal_effect_under_ignorability}).
+This necessitates that one condition on these covariates in computing the average causal effect in the same way as for estimating causal effects under ignorability (\S\ref{sec:causal_effect_under_ignorability}).
 The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome).
@@
 
-It is not always possible or ethical to conduct a randomised controlled trial, as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights \ref{fig:parashute}.
+It is not always possible or ethical to conduct a randomised controlled trial, as is beautifully demonstrated by the paper of \citet{smith2003parachute} on randomised cross-over trials of parachute use for the reduction of the mortality and morbidity associated with falls from large heights (figure \ref{fig:parashute}).
 When experimentation is possible, it is frequently difficult or expensive.
@@
 In addition, they are often conducted on a convenient, but unrepresentative, sample of the broader population of interest (for example first year university students).
-This can result in estimates with high \emph{internal validity} \citep{Campbell1963}, in that they should replicate well in a similar population, but very low \emph{external validity} in that the results may not carry over to the general population of interest.
+This can result in estimates with high \emph{internal validity} \citep{Campbell1963}, in that they should replicate well in a similar population, but low \emph{external validity} in that the results may not carry over to the general population of interest.
 The question of whether an experiment conducted on one population can be mapped to another is referred to as the transportability problem \citep{Bareinboim2013} and relies on very similar assumptions and arguments to causal inference and the do-calculus.
@@
 There are alternative definitions that arise in the related problems of adversarial bandits and learning from expert advice.
-In addition, researches often refer to the expected regret as \quotes{the regret}.}
+In addition, researchers often refer to the expected regret as \quotes{the regret}.}
 \eqn{
@@
 
-The adaptive nature of multi-armed bandit algorithms complicates the design and analyse of estimators.
+The adaptive nature of multi-armed bandit algorithms complicates the design and analysis of estimators.
 The action selected by an algorithm at a given timestep can depend on the history of previous actions and rewards.
@@
 We now assume the rewards are bounded in $[0,1]$.
-The algorithm and regret bounds can be generalised to submission reward distributions, see \citet{Bubeck2012}.
+The algorithm and regret bounds can be generalised to sub-gaussian reward distributions, see \citet{Bubeck2012}.
 
@@
 
-Let $\Delta_i = \mu_i - \mu^*$ be degree to which each arm is sub-optimal.
+Let $\Delta_i = \mu_i - \mu^*$ be the degree to which each arm is sub-optimal.
 The problem-dependent pseudo-regret for UCB-1 is bounded by equation \ref{eqn:ucb_problem_dependent_regret} \citep{Bubeck2012},
@@
 This is because it becomes harder for the algorithm to identify the optimal arm.
-As the differences $\Delta_i \rightarrow 0$, the regret bound in \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
+As the differences $\Delta_i \rightarrow 0$, the regret bound in equation \ref{eqn:ucb_problem_dependent_regret} blows up, however the regret itself does not - since although we may not be able to distinguish arms with very small $\Delta_i$ from the optimal arm, we also do not lose much by selecting them.
 The worst case occurs if all arms have the same expected reward $\mu$ except for the optimal arm which has reward $\mu^*=\mu+\Delta$, where $\Delta$ is just too small for the algorithm to learn to identify which arm is optimal given the horizon $T$.
@@
 \caption{The regret bound in equation \ref{eqn:ucb_problem_dependent_regret} grows as the differences between the expected rewards for each arm shrink.
-The solid curve shows the mean (cumulative) regret, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
+The solid curve shows the mean (cumulative) regret for the UCB-1 algorithm, over a 1000 simulations for a 2-armed, Bernoulli bandit with fixed horizon, $T = 500$, as a function of the difference in the expected reward for the arms $\Delta$.
 The dashed curves show the corresponding upper bounds; $T\Delta$ and equation \ref{eqn:ucb_problem_dependent_regret}}
@@
 
-Subtle modifications to the UCB algorithm can eliminate the logarithmic term equation \ref{eqn:ucb_worst_case_regret}.
+Subtle modifications to the UCB algorithm can eliminate the logarithmic term in equation \ref{eqn:ucb_worst_case_regret}.
 This yields regret $ \bigo{\sqrt{TK}}$ and closes the gap with the worst case lower bound \citep{Audibert2009,Lattimore2015}, whilst retaining a good problem-dependent bound of the form achieved by UCB \citep{Lattimore2015}.
@@
 
-%\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}
-
-Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds. The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period.
+Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds.
+The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period.
+This means we should focus exploration on arms which are plausibly optimal, creating a form of exploration-exploitation trade-off, albeit subtlety different to that for the cumulative regret.
 
@@
 Adversarial bandits appear to be more applicable to real world problems because they do not assume that the rewards associated with each arm are constant over time or independent of the previous actions of the learner.
-However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, does not fully capture an algorithms performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
-In settings where the rewards change over time, the pseudo-regret can be negative (see figure \ref{fig:non-stationary-bandit}) so upper bounds on the pseudo-regret do not reflect how sub-optimal the algorithm may be.
+However, pseudo-regret, as defined in equation \ref{eqn:pseudo_regret-adversarial}, does not fully capture an algorithm's performance in such cases because it is defined with respect to playing the single arm with the best average return over the game.
+In settings where the rewards change over time, the pseudo-regret can be negative (see figure \ref{fig:non-stationary-bandit}) so upper bounds on the pseudo-regret do not fully reflect how sub-optimal the algorithm may be.
 An example of a setting that lies between stochastic and adversarial bandit problems is the non-stationary setting, in which the rewards are generated stochastically from a distribution that varies over time.
@@
 However, in most real life (sequential) decision making processes, the optimal action depends on some context.
-The best treatment to offer an individual patient could depend on their age, gender, disease sub-type or genetics and will not always align with the treatment that is best on average (or for the majority of people).
+The best treatment to offer an individual patient could depend on their age, gender, disease subtype or genetics and will not always align with the treatment that is best on average (or for the majority of people).
 Similarly, decisions on which ad or content to display on a webpage, or which product to recommend, can be \emph{personalised} based on the previous behaviour of the user.
@@
 
-Suppose that, at each timestep $t$, after selecting an action, the learner received the reward for chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
+Suppose that, at each timestep $t$, after selecting an action, the learner received the reward for the chosen action but observed the full vector of rewards $[\cfb{Y}{1},...,\cfb{Y}{k}]$.
 This is known as the full information setting.
@@
 
-The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ in both the stochastic and adversarial settings \citep{Auer2002a}.
+The Exp-4 algorithm, developed in the context of learning from expert advice (each $h \in \mathcal{H}$ can be viewed as an expert who recommends which action to take), achieves optimal worst case regret of $\bigo{\sqrt{kT log(\card{\mathcal{H}})}}$ for both stochastic and adversarial contextual bandit problems \citep{Auer2002a}.
 However, it involves maintaining a list of weights for each hypothesis $h$, resulting in time and memory requirements that grow linearly with the size of the hypothesis space and, unlike the epoch-greedy algorithm, it cannot be generalised to infinite dimensional hypothesis spaces in a straightforward way.
@@
 
-The definition \ref{def:contextual-bandit-problem} does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$ (see figure \ref{fig:contextual-bandit-causal-graph}).
-However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is the equivalent to adding irrelevant features to a supervised learning problem.
+The definition of contextual bandits (see definition \ref{def:contextual-bandit-problem}) does not make any assumptions about the \emph{causal} relationship between the context $\vec{X}$ and the reward $\vec{Y}$ (see figure \ref{fig:contextual-bandit-causal-graph}).
+However, the context should be relevant, such that $\P{\vec{y}|\vec{x}} \neq \P{\vec{y}}$, otherwise including it is equivalent to adding irrelevant features to a supervised learning problem.
 \citet{Bareinboim2015} demonstrate that, in some cases, policies that incorporate observations of the action an agent would have taken were their action not set by the bandit policy can achieve lower regret than those that ignore this information.
@@
 \label{sec:off-policy-evaluation}
-Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
+Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation \citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}.
 In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$.
-The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
+The goal of the learner is to evaluate the value of an alternative policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.
 
@@
 However, the problem does not reduce to supervised learning because of the bandit nature of the feedback; only the reward of the selected action is observed.
-In addition, if $\pi$ is allowed to depend $h$ then the samples are not i.i.d.
+In addition, if $\pi$ is allowed to depend on $h$ then the samples are not i.i.d.
 The majority of the literature considers the case where the original policy $\pi$ was stationary ($\pi(\vec{x}_t,h_t) = \pi(\vec{x})$).
-\citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original estimator $\pi$ did not depend on $\vec{X}$.
+\citet{Langford2008a} do allow the original policy to be adaptive and prove a high probability bound on the accuracy of their estimator for $\pi'$, albeit with the strong assumption that the original policy $\pi$ did not depend on $\vec{X}$.
 
@@
 However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces.
-This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \citep{filippi2010parametric}, dependent bandits \citep{Pandey2007}, X-armed bandits \citep{Bubeck2010} and Gaussian process bandits \citep{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback\citep{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \citep{Piccolboni2001,Bartok2014}.
-
-In the next chapter, I propose a very natural connection between causal graphs and bandit problems and show it induces a novel form of additional feedback and structure between arms that cannot be can exploited by any of these previous approaches.
+This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalised linear bandits \citep{filippi2010parametric}, dependent bandits \citep{Pandey2007}, X-armed bandits \citep{Bubeck2010} and Gaussian process bandits \citep{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \citep{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \citep{Piccolboni2001,Bartok2014}.
+
+In the next chapter, I propose a very natural connection between causal graphs and bandit problems and show it induces a novel form of additional feedback and structure between arms that cannot be can exploited by these previous approaches.
 
@@
 
-We conclude this chapter on bandits with a note on the counterfactual nature of regret, which has not been discussed with respect to bandit regret.
+We conclude this chapter on bandits with a note on the counterfactual nature of regret, which I have not seen discussed with respect to bandit regret.
 However, the issues raised are very closely related to the work by \citet{Dawid2000} on problems associated with the use of counterfactual variables for (observational) causal inference.
@@
 However, the covariance $\rho$ cannot, because we never simultaneously observe the rewards for both arms.
-As a result, even with full knowledge of the reward distributions for each arm, the distribution over regret cannot, in general, be computed without un-testable assumptions about the covariance between counterfactuals (see figure \ref{fig:metaphysical_distribution_difference}).
+As a result, even with full knowledge of the reward distributions for each arm, the distribution over regret cannot, in general, be computed without untestable assumptions about the covariance between counterfactuals (see figure \ref{fig:metaphysical_distribution_difference}).
 
@@
 For example, a given user may be more (or less) likely to buy something no matter which advertisement they are served, leading to a positive correlation between counterfactual rewards.
-Equally, an illness might have two (unobservable) sub-types, with each medication (bandit arm) effective only against one, resulting in a negative correlation between the counterfactual rewards.
+Equally, an illness might have two (unobservable) subtypes, with each medication (bandit arm) effective only against one, resulting in a negative correlation between the counterfactual rewards.
 
@@
 The expectation of this variant of regret is the same as for the  version defined in equation \ref{eqn:raw_regret2}.
-It depends only on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions.
+It depends on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions.
 Furthermore, this quantity has already been analysed in existing work on the concentration of bandit regret, \citep{Audibert2007,Audibert2009exploration} as a more tractable proxy to the standard regret.
-In conclusion, when selecting measures of bandit performance, it is worth noting whether they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.
+In conclusion, when selecting measures of bandit performance, it is worth noting whether they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem of interest.
 
@@
 In most key real world decision making processes, from algorithms recommending drug treatments to robots deciding how to open doors, we have access to both potentially huge observational data sets, as well as the ability to intervene in the system at some level.
-It is critical that we develop methods that allow us to incorporate the causal knowledge we have about the world, along with data sets collected under a range of different conditions, into our current decision making process.
+It is critical that we develop methods that allow us to incorporate the causal knowledge we have about the world, along with data sets collected under a range of different conditions, into our decision making processes.
 In this chapter, I introduce a very general framework that connects causal graphical models with multi-armed bandit problems and demonstrate how it can be leveraged to make better decisions.
@@
 In this thesis, I focus on the case where the causal graph is known.
-Extending this work to simultaneously learning the casual graph is discussed in \S\ref{sec:causal_bandit_discussion}.
+Extending this work to simultaneously learning the causal graph is discussed in \S\ref{sec:causal_bandit_discussion}.
 
@@
 \begin{definition}[Causal bandit problem]
-A learner for a casual bandit problem is given the casual model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
+A learner for a causal bandit problem is given the causal model’s graph $G$ over variables $\mathcal{X}$ and a set of allowed actions $\actionspace$.
 Each action $a \in \actionspace$, denoted $do(\vec{X}=\vec{x})$, assigns values $\vec{x}$ to a corresponding set of variables $\vec{X} \subset \mathcal{X}$ and incurs a known cost $C(a)$ on the learner.
@@
 The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action, and on which variables we receive post-action feedback.
-If we can (at no cost) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant, and the problem can be treated as a standard multi-armed bandit problem over the set of actions that fully specify the values for the parents of $Y$.
+If we can (at no cost) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant, and the problem can be treated as a standard multi-armed bandit problem over the set of actions that fully specify the values of the parents of $Y$.
 If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example.
-If we receive post-action feedback on additional nodes the problem can be more interesting.
+If we receive post-action feedback on additional variables the problem can be more interesting.
 In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected.
@@
 \end{tikzpicture}
-\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level .
+\caption{Example causal graph (based on \citet{Koller2009}) where the outcome of interest (reward) is cholesterol level.
 The do-calculus can be applied to eliminate some actions immediately without the need to do any experiments.
@@
 We only get this form of side information for actions that do not specify the value of every variable, for example those in the bottom half of the table in figure \ref{fig:unify_frameworks}.
-If additional variables are only observed before an intervention is selected, the causal bandit problem can be treated as a stochastic contextual bandit problem, which is already reasonably well understood \citep{Agarwal2014}.
 
@@
 It is also worth noting the connection between this setting and instrumental variables \cite{balke1997bounds}.
-By making some functional assumptions about the relationships between the variables, we can use $A$ as an instrumental variable to bound or estimate the (casual) effect of $X$ on $Y$.
+By making some functional assumptions about the relationships between the variables, we can use $A$ as an instrumental variable to bound or estimate the (causal) effect of $X$ on $Y$.
 The estimation will be complicated in the bandit setting because the action chosen at each timestep is dependent on the previous sequence of actions and rewards.}
@@
 To the best of our knowledge, our paper is the first to analyse simple regret in bandit problems with extra post-action feedback.
-
-%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimise cumulative regret.
-They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
 Partial monitoring is a very general framework for decoupling the feedback from the action and reward.
@@
 Partial monitoring algorithms yield results that are optimal with respect the horizon $T$ but not other parameters, such as the number of actions, which is the key focus of incorporating causal structure.
+
+%\citet{Yu2009} consider bandit problems where a learner chooses an arm to play as well as set of arms to observe rewards for in a stochastic setting where the reward distributions can change infrequently and the aim is to minimise cumulative regret.
+They use extra observations to detect changes whereas we assume fixed reward distributions and use extra observations to improve arm selection.
 
@@
 
-The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$ where 
+The value of the reward variable is distributed as $\P{Y = 1|\vec{X}} = r(\vec{X})$, where 
 $r : \{0,1\}^N \to [0,1]$ is an arbitrary, fixed, and unknown function. 
@@
 For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation).
-However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorisation formula \citep{Pearl2000}.
+However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated product formula \citep{Pearl2000}.
 \eq{
@@
 
-We could naively generalise our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product factorisation to 
-write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational 
+We could naively generalise our approach for parallel bandits by observing for $T/2$ rounds, applying the truncated product formula (or the do-calculus if there are latent variables) to write an expression for each $\P{Y|a}$ in terms of observational quantities and explicitly playing the actions for which the observational
 estimates were poor. However, it is no longer optimal to ignore the information we can learn about the reward for intervening on one variable 
@@
 To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
-allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
+allocate samples to actions. To address this difficult problem we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
 are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
@@
 
-\begin{remark} If the actionspace $\mathcal{A}$ is unconstrained, that is consists of all possible assignments of values to variables, (and all actions have equal cost) the optimal action will set the value of all the parents of $Y$ and Algorithm \ref{alg:general} cannot do better than uniform exploration over these arms.
-In this case, after we use the causal structure to eliminate irrelevant actions prior to taking any samples, the problem of selecting within the remaining actions can be treated as a standard multi-armed bandit problem.
+\begin{remark} If the action space $\mathcal{A}$ is unconstrained, that is consists of all possible assignments of values to variables, (and all actions have equal cost) the optimal action will set the value of all the parents of $Y$ and Algorithm \ref{alg:general} cannot do better than uniform exploration over these arms.
+In this case, after we use the causal structure to eliminate irrelevant actions prior to taking any samples, the problem of selecting within the remaining actions can be treated as a standard multi-armed bandit problem as we state more formally below.
 \end{remark}
@@
 Note that we now only have to obtain estimates for actions $a \in \mathcal{A'}$, since we know the others to be sub-optimal, so the $max$ is only over these actions.
-However we $b$ still sums over all possible actions in the denominator of the importance sampling estimator, to allow for the possibility that playing sub-optimal actions allows more efficient estimation of the optimal actions.
+However $b$ still sums over all possible actions in the denominator of the importance sampling estimator, to allow for the possibility that playing sub-optimal actions allows more efficient estimation of the optimal actions.
 We now prove that this is not the case and that, in this specific setting, Algorithm \ref{alg:general} cannot do better than uniform sampling over the actions $a \in \mathcal{A'}$.
-Now each action $a \in \mathcal{A'}$ consists of a given assignment $\vec{x}_a$ to $\parents{Y}$.
+Each action $a \in \mathcal{A'}$ consists of a given assignment $\vec{x}_a$ to $\parents{Y}$.
 
@@
 A value of $v_{obs} = .5$ would imply we would need twice as many samples to obtain the same regret bound from observational data as compared to interventional data.
+
+\begin{figure}
+	\centering
+	 \begin{subfigure}[t]{0.8\textwidth}
+	 \centering
+	 \includegraphics{parallel_bandit2-crop}
+	 \end{subfigure}
+     \begin{subfigure}[t]{0.8\textwidth} 
+     \centering  
+	\includegraphics{value_of_observation}
+	\end{subfigure}
+	\label{fig:value_of_observation}
+	\caption{An example of quantifying the value of purely observational data.
+The heat map shows the value of $v_{obs}$ for an instance of the parallel bandit problem where all the variables have equal probability, $q$, of taking the value $1$, as a function of $q$ and the number of variables $N$.
+If the \part{variables} are perfectly balanced, $q=0.5$, then purely observing is optimal and $v_{obs}=1$.
+If $q=0$ then we cannot learn the value of actions $do(X_i=1)$ without intervention, even with infinite observational data, and $v_{obs}=0$.
+For intermediate values of $q$, we see that the improvement in the worst case regret that optimised intervention yields over purely observational data drops as the number of variables, $N$, increases.}
+\end{figure}
 
@@
 We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{Audibert2010}, Thompson Sampling and UCB under a variety of conditions.
-Thomson sampling and UCB are optimised to minimise cumulative regret.
+Thompson sampling and UCB are optimised to minimise cumulative regret.
 We apply them in the fixed horizon, best arm identification setting by running them up to horizon $T$ and then selecting the arm with the highest empirical mean.
@@
 Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned.
-This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
+This gap is learnable by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$.
 In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially.
@@
 The value of $N_1$ determines $m$ and ranges between $2$ and $N$.
-The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta*$.
-This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
+The values for the CPD's have been chosen such that the reward distribution is independent of $m$ and so that we can analytically calculate $\eta^*$.
+This allows us to just show the dependence on $m$, removing the noise associated with different models selecting values for $\eta^*$ with the same $m$ (and also worst case performance), but different performance for a given reward distribution.
 
@@
 to get near-optimal cumulative regret guarantees.
-They consider a finite-armed bandit model with side information where in reach round the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
+They consider a finite-armed bandit model with side information where in each round the learner chooses an action and receives a Gaussian reward signal for all actions, but with a known variance that depends on the chosen action.
 In this way the learner can gain information about actions it does not take with varying levels of accuracy.
@@
 
-\paragraph{Causal Models with Non-Observable Variables}
+\paragraph{Causal Models with Unobservable Variables}
 If we assume knowledge of the conditional \textit{interventional} distributions $\Pn{a}{\parents{Y}}$ our analysis applies unchanged to the case of causal models with 
@@
 & \leq \frac{8m(\vec{q})}{T} + \sqrt{\frac{2m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ letting } \delta = \frac{2m(\vec{q})}{T}\\
-& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jenson's Inequality}
+& \leq \sqrt{\frac{20m(\vec{q})}{T} \log \frac{NT}{m(\vec{q})}},\; \text{ via Jensen's Inequality}
 }
@@
 The recognition that there is a fundamental trade-off between accuracy and transparency, unless we can build perfect causal models, when the interests of individuals and society diverge has implications for the way we design and regulate systems that have the potential to have major impacts on people's lives.
-We must develop approaches to ensuring machine learning decisions are reasonable and ethical that allow effected individuals recourse to dispute or improve outcomes but do not undermine the ability of the system to function.
+We must develop approaches to ensuring machine learning decisions are reasonable and ethical that allow affected individuals recourse to dispute or improve outcomes but do not undermine the ability of the system to function.
 
@@
 
-I have developed an approach that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this conceptualisation encodes some key existing problems in the literature.
-I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.
+I have developed an approach that formally connects causal graphical models with bandit problems in a natural way and demonstrated that this conceptualisation encodes some key existing problems in the literature.
+I showed that knowledge of the causal structure (but without knowledge of the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.
 I introduce a metric that captures the difficulty of causal bandit problems and (under strong assumptions) allows the value of optimised interventional data over purely observational data to be computed prior to intervening in the system.
 
-This work represents an important first step towards a unified approach to causal inference and optimal decision making.
+This work represents a first step towards a unified approach to causal inference and optimal decision making.
 There is much exciting work remaining to be done.
@@
 The problem differs from typical observational causal inference in several ways: the focus is on identifying an optimal policy, rather than unbiased estimation of all policies; the goal is to explore the value of interventions on a range of different variables, rather than the optimal setting of a single variable; and the data will, in general, be non-stationary in a rather special way due to the adaptive nature of policy that generated it.
-Existing work on off-policy evaluation focuses (at least implicitly) on estimating causal effects by adjusting for all variables that simultaneously effect both the action selection and outcome.
+Existing work on off-policy evaluation focuses (at least implicitly) on estimating causal effects by adjusting for all variables that simultaneously affect both the action selection and outcome.
 If additional information is available about the causal relationships between variables involved, other approaches to identifying causal effects such as instrumental variables or the \quotes{front door method} \citep{Pearl2000} could be also be applied to off-policy evaluation.

### 2017-11-04 thesis/thesis_map.tex
fixes latex build error
final version of submitted thesis

@@
 
+\definecolor{nodeblue}{HTML}{729FCF}
 \tikzset{
@@
            text centered,
+           %fill=nodeblue!50
            fill=black!10!white
@@
            text centered,
-           fill=black!20!white,
+           fill=black!10!white,
+           %fill=nodeblue!50
            },
@@
            text centered,
-           fill=black!10!white
+           fill=black!10!white,
+           %fill=nodeblue!50
            },
@@
 \input{titlepage}
-\input{original_statement}
+
+\clearpage
+\vspace*{\fill}
+\begin{center}
+\begin{minipage}{.9\textwidth}
+Except where acknowledged in the customary manner, the material 
+presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any 
+university.
+
+\vspace{40mm}  % vertical space
+\hspace{80mm}\includegraphics[scale=.1]{sig}
+\vspace{.1mm}
+\hspace{80mm}\rule{40mm}{.15mm}\par   % horizontal space, line, start new line
+\hspace{80mm} Finnian Lattimore\par
+\hspace{80mm} November, 2017
+\end{minipage}
+\end{center}
+\vfill % equivalent to \vspace{\fill}
+\clearpage
+%\input{original_statement}
 
@@
 There are many people without whose support this thesis would not have been possible.
-I would like to thank my supervisors Dr Cheng Song Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof.
+I would like to thank my supervisors Dr Cheng Soon Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof.
 Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters.
@@
 I am also grateful for the financial support provided by an Australian Postgraduate Award and a NICTA/Data61 top up scholarship.
+
+This research is supported by an Australian Research Training Program (RTP) Scholarship.
 
@@
 Thus they have transformed simultaneous equations into a powerful framework and associated set of methods for estimating causal effects.
-The is also a rich strand of work on using the assumptions that can be encoded in structural equation models, also known as functional causal models to discover the structure and direction of causal relationships - see for example \citet{mooij2016distinguishing}.
+The is also a rich strand of work on using the assumptions that can be encoded in structural equation models, also known as functional causal models to discover the structure and direction of causal relationships - see for example \citep{Mooij2016a,Peters2014}.
 Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks.
@@
 Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment.
-\citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
+\citet{Abadie2002} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.
 
@@
 \centering
-\begin{subfigure}[t]{0.49\textwidth}
-\centering
-\caption{Bandits with unobserved confounders:$\vec{X}^c=\set{I}$, $\vec{X}^o = \set{Y}$, $\actionspace = do(X=x)$}
+\begin{subfigure}[t]{0.45\textwidth}
+\centering
+\caption{Bandits with unobserved confounders: $\vec{X}^c=\set{I}$, $\vec{X}^o = \set{Y}$, $\actionspace = do(X=x)$}
 \label{fig:confounded-causal-bandit-graph}
@@
 \end{subfigure}
-\begin{subfigure}[t]{0.49\textwidth}
+\hspace{1cm}
+\begin{subfigure}[t]{0.45\textwidth}
 \centering
@@
 \end{subfigure}
-\caption{}
+\caption{Causal bandit representation of bandits with unobserved confounders and compliance aware bandits.}
 \label{fig:causal-bandit-version-confounded-compliance}
@@
 The heat map shows the value of $v_{obs}$ for an instance of the parallel bandit problem where all the variables have equal probability, $q$, of taking the value $1$, as a function of $q$ and the number of variables $N$.
-If the \part{variables} are perfectly balanced, $q=0.5$, then purely observing is optimal and $v_{obs}=1$.
+If the variables are perfectly balanced, $q=0.5$, then purely observing is optimal and $v_{obs}=1$.
 If $q=0$ then we cannot learn the value of actions $do(X_i=1)$ without intervention, even with infinite observational data, and $v_{obs}=0$.
